### 2011-08-25 doc/manual/parallax_model.tex
Moving the manual and adding the typing rules documentation

\documentclass{article}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{color}
\usepackage{latexsym}
\usepackage{listings}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{cite}
\usepackage[all]{xy}
\usepackage{fullpage}
\usepackage[left=1in,top=1in,right=1in,bottom=1in]{geometry}

\theoremstyle{definition}
\newtheorem{definition}{Definition}
% spellcheck using ``aspell -c filename'' in Unix
\title{The Parallax Programming Model and Language}
\author{Michael Bauer, Sean Treichler, Alex Aiken}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\lstset{ %
	language=C++,                % choose the language of the code
	basicstyle=\footnotesize,       % the size of the fonts that are used for the code
	numbers=left,                   % where to put the line-numbers
	numberstyle=\footnotesize,      % the size of the fonts that are used for the line-numbers
	stepnumber=1,                   % the step between two line-numbers. If it's 1 each line will be numbered
	numbersep=5pt,                  % how far the line-numbers are from the code
	backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
	showspaces=false,               % show spaces adding particular underscores
	showstringspaces=false,         % underline spaces within strings
	showtabs=false,                 % show tabs within strings adding particular underscores
	frame=single,	                % adds a frame around the code
	tabsize=2,	                % sets default tabsize to 2 spaces
	captionpos=b,                   % sets the caption-position to bottom
	breaklines=true,                % sets automatic line breaking
	breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
	title=\lstname,                 % show the filename of files included with \lstinputlisting; also try caption instead of title
	escapeinside={\%*}{*)},          % if you want to add a comment within your code
	morekeywords={tunable, }            % if you want to add more keywords to the set
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle
\pagebreak

\tableofcontents
\pagebreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Introduction

\section{Introduction \label{Intro}}
% Multicore revolution
% Memory-compute mismatch
% Hieararchical memory
% Locality-aware programming systems
% Regions for locality and independence

\subsection{Related Work \label{Related}}

\subsubsection{Locality-Aware Programming Models \label{LocalityAware}}
% Chapel
% X10/Hiearchical Place Trees
% Sequoia

\subsubsection{Region Languages \label{RegionLangs}}
% Re-heaps
% RC
% Cyclone

\subsection{Outline \label{Outline}}

\subsection{Document Scope and Completeness \label{Scope} }
The scope of this document is the design of the Parallax programming model
as well as some of the syntax of the Parallax language.  We describe in a 
separate document the implementation of a Parallax compiler and runtime system
capable of supporting the Parallax language.  Not all features and interactions
for the language have been fully specified.  Some outstanding issues are listed
in Section \ref{OutstandingQs}.

\pagebreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Motivation

\section{Motivating Applications \label{Applications}}
% The need for applications to drive research
% Applications other than arrays
% Ghost cells
% Recursive decomposition for hierarchical memory

% Mesh based applications
% Graph algorithms
% Tree data structures
\noindent
The motivation for creating the Parallax programming model and language came
directly from a class of applications that we found difficult to describe
using other parallel languages.  The characteristics of these applications were that
they all required distributed pointer data structures.  Furthermore these pointer data
structures could not easily be represented by arrays due to either irregularity
in the data structure or complex sharing patterns.  These characteristics
prevent us from using array-centric languages like Sequoia \cite{Fatahalian06} 
or Chapel \cite{Chamberlain07} to implement these applications and still
achieve high performance. \\

\noindent
To better understand this set of applications we have constructed a
classification of the various applications that we've encountered.  The
classification is determined primarily based on the data structures used in
the applications as opposed to the kind of parallelism that exists.  This
is a direct consequence of our goal to build locality-aware languages instead
of simply parallel languages.  To reason about the locality present in
application we must consider the data structures required for an application.\\

\noindent
In this section we will first describe the set of data structure classes
that we are interested in supporting in Parallax.  We then present two example
applications that will be used as running examples throughout this document
to illustrate important Parallax concepts.

\subsection{Data Structure Classification \label{DataStructures}}
\noindent
There are three primary classes of data structures that we are interested
in investigating in Parallax: meshes, graphs and trees.  These classes of data 
structures are not intended to be representative of all applications for
our targets machines.  However, we believe that they capture a 
much larger set of applications than could have been described in previous
languages.  We now define each of these data structure classes and provide
instances of applications that use these data structures.

\subsubsection{Meshes \label{Mesh}}
\noindent
The first set of applications that motivated our design of Parallax was 
applications which made use of mesh based data structures.  We define a mesh
to be any data structure which consists of a polygonal structures which we
call {\tt cells}.  These polygonal structures can exist in any number of dimensions.
The only requirement of cells is that they have a neighbor relationship which
operates as a manifold in the formal mathematical sense.  This allows us
to differentiate meshes from graphs described in Section \ref{Graphs} and indeed
the locality properties of these two classes of data structure are very 
different.  Meshes may also have shared sub-structures (i.e. two cells sharing
a face or an edge), but this isn't a necessary condition for a mesh. \\

\noindent
Given this description of a mesh, we believe that there are several different
sub-classes of applications that use mesh based data structures.  We break
these sub-classes down across two different dimensions:

\begin{enumerate}
\item Regular/Irregular - Formally we'll classify a mesh as regular if the basis
of the vector space corresponding to the neighbor relationship manifold is the same
for all cells in the mesh and irregular otherwise.  Intuitively, if there is 
a $O(1)$ function in time and space for computing the neighbor relationship for 
all cells in the mesh we'll classify the mesh as regular.
\item Gather/Scatter - This is a property of the computation being performed on
the mesh and determines whether or not the operation on a single cell requires
only reading neighbor cells (gather) or reading and writing neighbor cells (scatter).
\end{enumerate}

\noindent
Given these different dimensions for classifying mesh data structures and mesh
computations we can now give examples of computations that fit in each category.

\begin{center}
\begin{tabular}{|c||c|c|}
\hline
 & Regular & Irregular \\ \hline \hline
Gather & Reverse-Time Migration \cite{Micikevicius09} & Lattice-Boltzmann \cite{Heuveline07} \\ \hline
Scatter & Fluidanimate \cite{Bienia08} & Liszt DSL\cite{DeVito10} \\ \hline
\end{tabular}
\end{center}

\begin{itemize}
\item Reverse-Time Migration - This application relies on a mesh that consists of
a three dimensional array.  For each time step only reads of neighboring cells are
required to perform a computation.
\item Lattice-Boltzmann - Lattice-Boltzmann computations can involve either
regular or irregular meshes, but computations to update a cell for a given time
step involve gathering (reading) information from neighboring cells.
\item Fluidanimate - An application from the Parsec Benchmark Suite \cite{Bienia08}
that relies on a three dimensional array of cells, but scatters updates to
neighboring cells rather than reading from neighboring cells.
\item Liszt DSL - A domain specific language for mesh-based partial differential
equations.  Meshes can be either regular or irregular, but updates are scattered
to neighboring cells often with specific reduction phases.
\end{itemize}

\noindent
These applications are just a sampling of the many codes that could be 
classified under these sets of mesh computations.  Our goal in designing
Parallax is to be able to handle applications that fit under all four of
these classes of mesh based computations.

\subsubsection{Graphs \label{Graphs}}
\noindent
Another class of data structures that we want to be able to handle in Parallax
are graph data structures.  Graphs differ from meshes as the
notion of a neighbor relationship in graphs do not have to obey any nice 
mathematical requirements.  Instead nodes in a graph are neighbors
if they simply share an edge which makes it much more difficult to identify
locality in graph data structures.  Due to this property we have chosen
to separate applications which use graph data structures into a
separate class of applications. \\

\noindent
One current thinking on graph applications is that there are two classes of
graph applications \cite{Bader11}:

\begin{itemize}
\item Static Graph Algorithms - operate on graphs which once
loaded into an application do not change for the duration of the program.
\item Dynamic Graph Algorithms - operate on graphs which can change 
throughout the duration of a computation with new inputs, or modifications
to the graph by the algorithm itself.
\end{itemize}

\noindent
These two classes of graph algorithms will require different graph 
representations.  In the case of static graph algorithms, the graph
can be partitioned at load-time, while for dynamic graphs there may need
to be load balancing steps to move parts of the graph data structure
around at runtime.  Another goal of Parallax is to be capable of handling
both classes of graph algorithms and data structures.

\subsubsection{Trees \label{Trees}}
\noindent
The last class of data structures that we want to be able to support
with Parallax are tree data structures.  Trees are a special type of
graph data structure in which the locality is very clear.  Tree data
structures tend to be highly scalable in parallel environments due to 
their obvious division of work between independent sub-trees.  Tree
data structures also mirror the underlying memory hierarchy of parallel
machines which is often structured as a tree.\\

\noindent
While there are not many scientific applications which rely on tree
data structures, there are interesting parallel applications which use
tree data structures.  Other members of our research group are interested
in performing static analysis over large abstract syntax trees, often
describing multi-million line programs like the Linux kernel \cite{Dillig10}.
Many of these static analyses can be performed in parallel, but require
significant amounts of memory and benefit from being run over large machines
which can handle their working sets.  Handling tree data structures
for applications such as these is another Parallax goal.

\subsection{Simple Example Application: Linked List Fold \label{LinkedList}}
\noindent
In this section and the next we introduce two applications that will serve to
be running examples throughout the remainder of the document.  This first
example application is trivial and will only be used to illustrate basic
concepts.  The example in section \ref{GraphSim} will be used to demonstrate
the need for more advanced concepts and how the language actually works.\\

\noindent
Our simple example application has a single data structure consisting of a
linked list of integers.  Each node in the linked list contains a single integer
and a pointer to the next element in the linked list.  The computation that
we wish to perform over the linked list is a fold reduction to compute the 
sum of the linked list. \\

\noindent
Our goal is to be able to parallelize this computation for an arbitrarily
deep hierarchical memory machine.  To do so we want to be able to 
partition the linked list into independent sub-lists and then potentially
continue partitioning sub-lists for as many levels of the memory hierarchy
as necessary.  Ultimately a sequential summation will be performed on
each sub-list and results reduced by the tree of parent computations.

\subsection{Advanced Example Application: Graph Based Simulation \label{GraphSim}}
\noindent
As an example of a more advanced application inline with the different sets
of applications that we introduced in section \ref{DataStructures} we now
describe a graph based simulation.  This particular application is an instance
of a static graph application described in section \ref{Graphs}. \\

\noindent
Our application takes as input a large graph representing an electrical
circuit\footnote{This is what happens when two electrical engineers
design a programming model.}.  The nodes in the graph represent points of
constant voltage or connections between circuit elements.  The edges of
the graph represent circuit elements including voltage sources, current
sources, resistors, capacitors, and inductors.  The application then
simulates this circuit using Kirchhoff's current and voltage laws for
a large number of time steps.\\

\noindent
To perform this computation in parallel we will have to partition this graph
into sub-graphs.  For hierarchical memory we will want to recursively partition
the graph an arbitrary number of times.  Once we have partitioned the graph at the beginning
of the computation, we can then perform the simulation in a two phase approach.
In the first phase, each partition gets the most up to date information from
adjacent nodes that are not in the partition.  The second phase then uses
this information to simulate the local partition.  This can be done at every
level of partitioning.  Each phase is performed for every time step
of the simulation.

\pagebreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Regions

\section{A Region Programming Model \label{Regions}}
% Goals of region programming model

\noindent
In this section we propose a new region-based programming model for describing
computations to be performed on hierarchical memory machines.  
While our ultimate goal is to design a parallel language, we've explicitly
decoupled the region programming model from our model of parallelism.  Our region
programming model doesn't require parallelism to describe locality.  We even believe
it could be beneficial for exploiting locality in a sequential language as described
in section \ref{Sequential}. \\

\noindent
The primary difference between previous programming models and our region programming
model is its model of memory.  Most traditional programming models use the
abstractions of a globally visible heap (shared memory) and a stack for each thread
of control.  In distributed programming models there are independent heaps for each
process of control (MPI).  Instead of relying on these abstractions that are tied
to the machine, we raise the level of abstraction in how memory is managed.\\

\noindent
In our model all memory is allocated inside of a set of regions which are explicitly
under the control of the programmer.  We feel that regions and the operations on them
that we describe in this section are sufficient to provide the following information:

\begin{enumerate}
\item Locality - regions should be capable of describing data that
should be colocated for a computation to be performed.
\item Independence - regions should enable us to reason about which computations
use disjoint data and can be run independently.
\item Pointer Fidelity - to ensure independence and enable us to move regions
throughout the memory hierarchy we need to know exactly which regions pointers target.
\end{enumerate}

\noindent
More importantly this information can be conveyed by our region programming model in
a machine agnostic context.  This will allow the compiler, runtime system, and the
programmer to specify how to map region programs to different hierarchical memory
machines at a later period of time as described in section \ref{Mapping}. \\

\noindent
To introduce our region programming model, we first provide definitions and concepts
required for understanding our definition of regions.  We then describe the 
operations that the programmer is capable of performing on regions.  Next we
describe the need for region relationships.  Finally, we show how arrays
are just a special case of regions.

\subsection{Region Definition \label{RegionDef}}
% Basic definitions required for defining regions
\noindent
To describe our region programming model we first need to define some common terms
that will be used throughout our programming model description.  
We begin by defining an allocation and a pointer to an allocation.  
From these definitions we can then define a region.

\begin{definition}
An {\tt allocation} is a pair ($T$,$R$) where $T$ is an instance of a type $T$ from
the type system and $R$ is a region. 
\end{definition}

\begin{definition}
A {\tt pointer} of type $T$ into a region $R$ is a reference to a unique allocation
of type $T$ in region $R$.
\end{definition}

\noindent From these definitions we can now specify our definition of a region.

\begin{definition}
A {\tt region} $R$ is a potentially empty mapping of pointers to allocations of type 
$T$ in region $R$.  All allocations (and therefore all pointers) in a region must be of the
same type $T$.  This restriction is necessary for guaranteeing the type safety of 
pointers into the region\footnote{It also has an added implementation benefit described
in the technical report on the implementation the Parallax compiler and runtime system.}.
\end{definition}

\noindent
Unlike previous work on regions (see section \ref{RegionLangs}), our 
definition of a region is independent of the hardware we are targeting.  
We don't rely on the concept of memory pages anywhere in our definition of 
a region.  Our definition of a region is simply an abstract  mapping that 
enables us to use pointers to uniquely identify allocations.  We can
describe locality using regions by placing two allocations in the same region.
We can also describe data independence by performing two allocations in different
regions. \\

\noindent
From our definition of a region we can now define a sub-region relationship.

\begin{definition}
A region $R$ is said to be a {\tt sub-region} or {\tt child region} of a parent region $P$
if the mapping for $R$ is a subset of the mapping for $P$.
\end{definition}

\noindent
The sub-region relationship will be critical for providing the ability 
to recursively decompose data structures contained within a region.  Our approach
will allow us to mirror the way that the Sequoia language was able to recursively 
decompose arrays.  We describe in more detail how to create sub-regions in Section
\ref{RegOp}. \\

\subsection{Region Operations \label{RegOp}}
% Basic operations on regions
\noindent
There are three complementary pairs of region operations that are available to a
programmer.  The first two pairs of operations are basic operations on regions 
that allow the programmer to create/destroy regions and to allocate/free memory
inside of a region.  The third pair of operations partition/union allow for the recursive
decomposition of data structures into sub-regions.

\subsubsection{Creation and Destruction \label{CreateDestroy}}
\noindent
To be able to allocate memory the programmer must first have the ability to create
and destroy regions.  There are built-in calls to the language for creating a region
and destroying a region.  The call to create a region must take as a parameter the
type of region to create and then returns a unique region identifier.  The type
passed to a call to create a region can either be any type in the type system 
or a region relationship (described in more detail in section \ref{RegionRelationship}).
If the type passed is a type from the type system then the region identifier
is simply a handle, but if the type is a region relationship, then the region
identifier will have region handles for each of the regions in the region relationship. \\

\noindent
To delete a region the programmer calls a delete region function that takes
as an argument a region identifier.  Memory management of regions is explicit and
it is therefore possible to have memory leaks if a program does not delete
regions before their handles go out of scope.\\

\noindent
Regions are not first class objects and therefore region identifiers have a special
type that is not permitted to be written into memory or read out of memory.  This
restriction is enforced by the type system.  We may consider in the future making
regions first class objects, but to simplify our region analysis we currently
don't permit this.

\subsubsection{Allocation and Deletion \label{AllocDelete}}
\noindent
The only way to allocate memory in our programming model is to allocate it inside
of a region.  Calls to malloc and free in our language take a region in which to
perform an allocation.  Since the region is already typed with the type of element
that it contains, then a call to malloc on a given region will return a pointer to
an allocation of the given region's type.  The type of the returned pointer is
also qualified to point into region in which the allocation was performed. \\

\noindent
Allocation also has an interesting interaction with sub-region relationships.  If
an allocation is performed in a region that is a sub-region of a parent region
then the allocation is also added as a valid allocation of the parent region.  This
follows directly from our definition of sub-region given in section \ref{RegionDef}:
the mapping from pointers to allocations for a sub-region must be a subset of the
mapping of its parent.  This allocation operation is transitive which implies that
an allocation in any sub-region will automatically be valid in any of the ancestor
regions of the region in which the allocation was performed. \\

\noindent
Unlike allocation, deleting an element from a region does not have the same effect
as allocation of applying to any ancestor region.  Instead deletion operates on
all sub-regions of the region in which the deallocation is performed.  To understand
the reasoning behind this consider the following scenario: an allocation is performed in a
parent region and is added to two different sub-regions of the parent region (we
describe how this occurs in section \ref{Partition}).  If we delete the allocation
from one of the sub-regions, we cannot safely delete the allocation from the parent region
because the allocation is still valid in the other sub-region.  However, if we delete
the allocation from the parent region, then we must delete the allocation from all of
the child regions to maintain the definition of a sub-region relationship.

\subsubsection{Partition \label{Partition}}
\noindent
To this point we have only provided basic operations on regions that are a part of
any region language.  This section introduces the first of two operations that give us the
machinery necessary for describing how to recursively decompose data structures in
regions.  The first operation, partition, will take a region and break
it apart into a set of sub-regions.  Conversely the second operation, union, will
take two sub-regions with a common ancestor and allow us to combine them into a
new region and is described in section \ref{Union}.

\noindent
To describe how partitioning of a region is performed we first introduce the
concept of a coloring and several properties of a coloring.

\begin{definition}
A {\tt coloring} is a mapping from a (sub)set of pointers valid in a region
to $\mathbb{N}$.  The subset of $\mathbb{N}$ in the image of the mapping
is called the set of {\tt colors} used in the coloring.
\end{definition}

\begin{definition}
A coloring is said to be {\tt disjoint} if the color mapping is injective, otherwise
the coloring is said to be {\tt aliased}.
\end{definition}

\begin{definition}
A coloring is said to be {\tt total} if the color mapping is surjective, otherwise
the coloring is said to be {\tt subtotal}.
\end{definition}

\noindent
The process of creating a partition is a two step process.  In the first
step the programmer is responsible for creating the coloring.  In the second step
the programmer then invokes a built-in partition functioning that takes as
arguments the original region, the coloring, and the set of colors used to
create the coloring.  The built-in partitioning function then returns a
special data structure called a collection of regions.

\begin{definition}
A {\tt region collection} is a special data structure that holds the set
of sub-regions returned by a partitioning operation and can be indexed by
the set of colors used in the partitioning.  The type of a region collection
is also qualified by the properties of coloring that created it such
as disjoint/aliased and total/subtotal. 
\end{definition}

\noindent
It is legal for there to be more than one partition made of a region.  This allows
the programmer to use different partitions of a data structure during different
phases of a computation when different locality properties of the data structure
might be exploited.  The ability to have multiple views into a data structure by
creating separate partitions creates a parallax on the data structure.\footnote{This 
is the naming origin of our language.}\\

\noindent
To make the concept of a partitioning concrete we now describe how partitioning
works for both of the example programs introduced in section \ref{Applications}.
For the case of the linked list we create a coloring that maps allocations in
contiguous ranges of the linked list to different sub-regions as can be
seen in figure \ref{fig:LinkedListColor}.  We can then pass
this coloring to the partition function that will return a region collection consisting
of the green, blue, and red sub-regions of the initial region. \\

\begin{figure}[t]
\centering
\includegraphics[width=4in]{figs/LinkedListPartition.pdf}
%\vspace{-5mm}
\caption{Coloring for partitioning a linked list. \label{fig:LinkedListColor}}
\end{figure}

\noindent
To describe the sharing scheme for the circuit simulation we will need
multiple partitions.  Recall from section \ref{GraphSim} that we want to
partition up the graph in such a way that we can keep track of which nodes
have to be shared with other subsets of the graph so we can exchange information
as dictated by the first phase of the graph algorithm.  To do this we first want
to partition the graph into the nodes that are private to computations
and those that are shared between computations.\\

\noindent
Figure \ref{fig:GraphPartition}
shows the view of a set of nodes in the graph from the perspective of a
computation.  In this figure all the nodes inside the black circle
are nodes that the computation owns, while all the nodes outside the orange circle
are nodes that are shared between different computations.  This creates three
different sets of nodes that must be created from the set of all nodes: private
to a computation, owned by a computation but shared, and then read by a computation
but owned by a different computation. \\

\begin{figure}[t]
\centering
\includegraphics[width=4in]{figs/GraphPartition.pdf}
\caption{A subset of the circuit simulation graph from the perspective of
a computation.  Nodes inside of the black circle are owned
by the computation.  Nodes outside of the orange circle are shared between
computations.  There are then three different sets of nodes that must be
described: private to a computation (green), owned and shared (red), and
not owned but read (blue). \label{fig:GraphPartition}}
\end{figure}

\noindent
To create the sets described in figure \ref{fig:GraphPartition} we will
need several partitions of the set of nodes.  To show the different partitions
visually we have created a symbol to represent the partition operator which
can be seen in figure \ref{fig:PartitionOperator}.  The partition operator
indicates that a region should be partitioned into a set of sub-regions.  The
$N$ in figure \ref{fig:PartitionOperator} indicates the number of sub-regions
which will be created.  The partition operator can also specify whether the
partition is disjoint/aliased and total/subtotal.\\

\begin{figure}[t]
\centering
\includegraphics[width=3in]{figs/PartitionSymbol.pdf}
\caption{Symbol for a partition of a region into sub-regions.  The $N$
indicates the number of sub-regions to be created. \label{fig:PartitionOperator}}
\end{figure}

\noindent
The different partitions that are required to break up the set of all nodes
in the graph can now be seen in figure \ref{fig:GraphPartitionGraph}.  We
first partition the nodes into two sub-regions: those nodes which are private
to computations and those which are shared between computations.  This
corresponds to the orange circle in figure \ref{fig:GraphPartition} and
the orange partition in figure \ref{fig:GraphPartitionGraph}.  For the 
private sub-region we can again partition this region into the sets of private
nodes for each computation corresponding to the green nodes in figure 
\ref{fig:GraphPartition} and the green partition in figure
\ref{fig:GraphPartitionGraph}.\\

\begin{figure}[t]
\centering
\includegraphics[width=4in]{figs/GraphPartitionGraph.pdf}
\caption{Illustration of the different partitions required for
breaking up the set of nodes in the circuit application.
\label{fig:GraphPartitionGraph}}
\end{figure}

\noindent
For the shared sub-region, we now want to partition this group of nodes two
different ways, one to capture the set of a nodes that are shared by computations
but still owned, and one to capture the set of nodes that are not owned but
required for a computation.  We create a separate partition of all the shared
nodes to describe each of these sets.  The first partition to describe the owned
nodes is over all $N$ computations and is therefore total and disjoint because each shared
cell can only belong to one computation.  This partition corresponds to the
red nodes in figure \ref{fig:GraphPartition} and the red partition in figure
\ref{fig:GraphPartitionGraph}.  The second partition to describe the set of nodes
that have to be read but are not owned (ghost nodes), is over all $N$ computations
and corresponds to the blue nodes in figure \ref{fig:GraphPartition} and the blue 
partition operator in figure \ref{fig:GraphPartitionGraph}.  This partition is
total but aliased because all the shared nodes are ghost nodes for at least one
other computation, but can also be ghost nodes for multiple other computations.\\

\noindent
This example has demonstrate the power of the partitioning operation as it's shown
that we can capture complicated sharing patterns using multiple partitions.  The
ability partition regions isn't enough however if we want to be able to recursively
decompose data structures.  To achieve our goal of an arbitrary level of 
decomposition we need an additional operation as shown in the next section.

\subsubsection{Union \label{Union}}
\noindent
In this section we introduce the union operation as a complementary operation
to the partition operation.  To understand the need for a union operation consider
the graph-based circuit simulation from section \ref{GraphSim}.  In section \ref{Partition}
we described how to partition the graph of nodes ONE time, but our goal is to
be able to partition it as many times as necessary for hierarchical memory.  To
achieve this goal we need to be able to maintain a recursive invariant.\\

\noindent
Our initial computation wanted to partition a single set of nodes but after
the partitioning from section \ref{Partition} each computation now has three
sets of nodes: private, owned, and ghost.  Each of these computations may want
to again partition the set of nodes for which it is responsible.  In the
case of the circuit simulation this is the set of private nodes unioned with
the set of owned nodes.  To achieve this we provide a union operation to create
a new region which is the result of a union between two regions.  For the circuit
simulation we can then pass the result of this union operation to function
which performs the partitioning and recursively decompose our graph as many times
as necessary.  This recursive invariant can be seen in figure \ref{fig:UnionPartition}
where we apply a union for each pair of shared and private nodes to get a new
subset of the graph.  Note that because we maintained disjoint/aliased and
total/subtotal information about each partition we can infer that the region
collection resultant from unioning together {\tt Private[i]} and {\tt Owned[i]}
is both disjoint and total. \\

\begin{figure}[t]
\centering
\includegraphics[width=4in]{figs/UnionPartitionGraph.pdf}
\caption{Illustration of the union operator on regions to restore the recursive
invariant of a single set of nodes in the circuit graph simulation. \label{fig:UnionPartition}}
\end{figure}

\noindent
The only rule for region union is that the two regions being unioned must share
a common ancestor.  This is important for maintaining the invariants specified
by the sub-region relationship.  The created region is a parent region to
each of the unioned child regions and a child region of the least common ancestor
of the set of regions which were unioned together.  By adding the ability to perform region
unions we can no longer describe the set of all region relationships using a tree
like all previous region languages, but instead require a directed acyclic graph
to describe all the regions.  Note that this still maintains the sub-region
relationship we defined in section \ref {RegionDef}.

\subsection{Region Relationships \label{RegionRelationship}}
% Mutually recusive region relationships
\noindent
As stated at the beginning of this section one of the goals of our programming model
is to accurately model the set of regions a pointer can target.  To do this we
qualify every pointer type with a set of regions into which it can point.  By also
restricting regions to only contain allocations of a single type, we can parametrize
the type of the allocation on the region on which it is contained.  This can give us
very strong guarantees about where pointers within a region can point.\\

\noindent 
Consider the linked list example from section \ref{LinkedList}.  In this case we create
a region which will contain allocations of linked list nodes.  Our system however allows
us to guarantee that the pointers to the next element in every linked list
node also point into the same region in which the allocation occurs because we can
parameterize the type of allocations in the region on the region itself.  This allows
the compiler to reason about locality in this data structure by specifying that all
the nodes in the linked list are in a single region.\\

\noindent
The approach illustrated above works well when an entire data structure can be
represented by a single type in a single region.  However, many data structures require
multiple types to represent and often these types are self referential.  Consider
the case of the graph example from section \ref{GraphSim}.  In this example there
are two types to this data structure: nodes and edges.  Each node may need to point
to many edges and every edge will need to point to two nodes.  We now need to create
two regions whose types are mutually referential: the node type for the node region
needs to be parameterized on the edge region, and the edge type for the edge region
needs to be parameterized on the node region.  From this we conclude that both the
node region and the edge region must be created simultaneously.\\

\noindent
To be able to support mutually referential region creation we introduce the concept
of a {\tt region relationship}.  A region relationship is a special kind of struct that
can contain multiple region handles.  The types of all the regions in a region relationship
may be self-referential and the compiler will know how to type check this.  Region
relationships also make explicit how different regions can be used together to represent
a data structure.  This becomes beneficial when partitioning data structures as the
programmer must now partition an entire region relationship instead of simply a single
region, indicating that an entire data structure is being partitioned.  Region relationships
allow us to capture the locality in more complicated data structures which may require
multiple regions to represent.

\subsection{Arrays as a Special Kind of Region \label{Arrays}}
% Mapping from N^n to pointers
% Compactness? Convexity?
% Dense arrays?
\noindent
Having now given a description of our region programming model, we note that we
can describe arrays as simply a special kind of region.

\begin{definition}
A $d$-dimensional {\tt array} is a special kind of region in which there exists
a bijective mapping from $\mathbb{N}^d$ to the set of pointers used to index
the region.
\end{definition}

\noindent
By defining arrays in this way we can apply all of the operations on regions to
arrays as well including partitioning and unioning.  At the same time by defining
the mapping between pointers and points in the space of integer tuples used to
define an array, we also introduce the possibility of doing additional analysis
on arrays that is not possible on regions.  Our goal is that array based
programs from Sequoia will still perform just as well as in our programming model.\\

\pagebreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% A Sequential Region Language

\section{A Sequential Region Language \label{Sequential}}
% Implementation of region programming model
\noindent
In section \ref{Regions} we introduced a region programming model and described
how regions could be used for expressing complex locality and sharing patterns in a
machine agnostic way.  In this section we give an example of an imperative sequential language that
embodies our region programming model.  In section \ref{LangStructures} we first
describe the basic structures of the language.  In section \ref{TypeSystem} we
then introduce a region type system that will allow static type checking of pointers
that point into regions.

\subsection{Language Structures \label{LangStructures}}
% Functions and Statements
% Special region structures
% Built-in region operations
% Explicit region passing
\noindent
For our region language we begin with a restricted subset of the C language.  We then
incorporate regions into the language by adding a small sub-language for describing
regions operations.  The decision to keep the region part of the language separate
is a result of decision that regions are currently not first-class structures within
the language.  In our implementation this will simplify our analysis considerably.

\subsubsection{Base Imperative Language \label{Imperative}}
\noindent
We begin with a restricted subset of the C language and restrict several features
to match the semantics necessary for our region programming model.  One
operation that is permitted in C that we reject is the ability to perform pointer
arithmetic.  This restriction stems directly from our definition of a pointer not as
explicit pointer into memory, but instead as a reference into an allocation in a given
region.  Pointers in our language are therefore immutable values that can only be passed
around and never modified.\\

\noindent
Another restriction on C that we enforce is there is no ability to use the $\&$ operator 
to create a pointer.  The only way that pointers can be created is via allocations in
regions which allow us to reason about the pointer types.  We also do not permit arguments
to functions to be passed by value.  This restriction is a result of our desire to type
all statements with the set of regions that are accessed by that statement as described
in section \ref{TypeSystem}. \\

\noindent
The final restriction we enforce is that arrays are no longer declared the same way they
are in C.  Instead arrays will be created as a special kind of region and all of the
semantics of regions described in section \ref{Embedded} will be enforced.  However, we
will minimize the difference in syntax from classic array syntax.  For example, to retrieve
an array value the programmer can still make use of the $[]$ operator just like in C.  This
follows directly from our definition of arrays as regions in section \ref{Arrays} which
defines an array as a special region with a mapping from a tuple of integers to a pointer
in the region.  Our implementation will automatically convert the tuple of integers into
a pointer and use that to dereference the region\footnote{The actual implementation will
be much more efficient than this for dense arrays.}.

\subsubsection{Embedded Region Language \label{Embedded}}
\noindent
We now describe a small region language embedded into our imperative language for implementing
our region programming model.  This embedded region language first adds two new structures
to the language: a region identifier type that can be used to reference individual regions
or region relationships and a region collection type that can be used to store an unbounded
number of regions.  Since we don't want regions to be first class variables, we do not permit
objects of these types to be passed by value to functions normal types can be.  Instead we
introduce a new function syntax where region identifiers or region collections are explicitly
passed as a separate list of arguments to functions: {\tt fn\_name(region\_args):(by\_value\_args)}.
This syntax will be beneficial to the type system as it makes it explicit which regions can
be affected by a function as will be described in section \ref{TypeSystem}.  By placing the
region arguments first we can use the region arguments when defining the types passed in the
set of by-value arguments so that the pointers and the regions they point to can be aligned.\\

\noindent
In addition to adding the built-in types for regions and region collections, we also create
a set of built-in functions (similar to malloc/free) for performing the operations defined
in the region programming model.  Here are the definitions for each of the built-in functions:

\begin{itemize}
\item {\tt create\_region} - takes a type for a region or a region relationship and returns a
region identifier for the created region 
\item {\tt destroy\_region} - takes a region identifier and destroys the associated region or region relationship
\item {\tt region\_malloc} - takes a region identifier and returns a pointer to an allocation of the region's type
\item {\tt region\_free} - takes a pointer and frees an allocation in the region
\item {\tt region\_partition} - takes a region, a color set, a coloring map, and returns a region collection
of sub-regions corresponding to the partition
\item {\tt region\_union} - takes two region identifiers and performs a union of the two regions
\end{itemize}

These built-in functions obey all of the semantics defined in section \ref{Regions} for the
region programming model.  We have yet to fully specify the set of operations that can be performed
on region collections as well as the mechanisms used for differentiating arrays from standard
regions (reflected in questions in section \ref{OutstandingQs}.

\subsection{Region Type System \label{TypeSystem}}
% Basic region type structure
% Mapping pointers to regions
% Existential types
\noindent
To fully define our sequential region language, we also have to specify a type
system that will allow us to check the type safety of our language\footnote{Our
language cannot be fully type safe because we are doing explicit memory management
and the risk of dangling pointers and double-frees always exists.}.  Type checking
the base imperative language is done the same as in C.  Type checking the embedded
region language is also straightforward.  The difficult part of the type system comes 
in dealing with the interaction between these two domains. \\

\noindent
We stated in our goals for our region programming model the requirement that we
be able to reason about pointers and the regions into which they point.  The
type checker must therefore be able to handle this crossover between the base
imperative language and embedded region part of the language. 
There is a catch however to qualifying pointers with the set of regions they 
can target: regions are runtime values within the program and it's not obvious
how to refer to them as type qualifiers.  There are two solutions
to this problem that we considered.

\begin{enumerate}
\item Bounded Region Instances - Ensure that there are a bounded number of regions in
any program and that this set of regions is statically analyzable by the compiler.  We
could then treat regions as both values and types and prove at compile time the safety
of pointer dereferences.
\item Existential Types\cite{Pierce02} - Allow for an unbounded number of regions 
and treat pointer types as existential types whose {\tt witness types} are sets of regions.
The existential pointer types would then have implicit {\tt getter} and {\tt setter}
methods as part of the pointer type for whenever pointers were derefenced.  
\end{enumerate}

\noindent
From our experience programming in Sequoia we felt that the restriction for statically
bounding the set of available regions was too stringent.  We therefore opted for the
second approach to type check our region programming model.  This approach
allows us to dynamically create regions as needed at runtime while still ensuring
that our program is well typed. \\

\noindent
Existential types allow us to check the validity of pointer dereferences because
the act of actually dereferencing a pointer will require us to implicitly be able
to unpack the existential type.  To perform the unpack operation we must have
the witness type (in this case the set of regions the pointer can point to) available
in scope.  This set of regions is easily identifiable as the set of regions created
in the current block of code as well as the set of regions that were passed explicitly
to the function using the {\tt fn\_name(region\_args):(by\_value\_args)} syntax.  If
a pointer is passed to a function, but not the associated region then this pointer
cannot be dereferenced because the witness type for its existential type is not
available.  \\

\noindent
By using existential types to check the validity of pointer dereferences we also get
an additional benefit from the type system: we can specify for every statement in a
program the set of regions that are accessed and side-effected by the statement.  This
follows directly from the requirement that all pointer dereferences know the witness
regions for their existential type.  For all pointer dereference expressions in a
statement we can then figure out the regions they access and how they are accessed.
This information will be useful for cases where we may want to reorder statements for
locality purposes.  We know this operation will be safe because we know exactly the
regions that are accessed by every statement.

\pagebreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Parallax: A Parallel Region Language

\section{Parallax: A Parallel Region Language \label{Parallax}}
% Add parallelism to sequential region language
\noindent
In this section we now describe the Parallax programming language.
Parallax is a parallel version of the sequential region language introduced
in section \ref{Sequential}.  To incorporate parallelism into the language
there are several extensions to the language that must be made.  In section
\ref{Parallelism} we introduce extensions for creating parallel computation.  In
the presence of parallel computation we also need to have a concrete view
of the parallel memory semantics.  Section \ref{Coherence} describes our
mechanism for keeping the memory in a region coherent.  Section \ref{Consistency}
then describes our memory consistency model.  Finally, section \ref{Execution}
describes the Parallax execution model and how it enables programmers to
reason about performance.

\subsection{Parallelism Constructs \label{Parallelism}}
% Explicit but flexible parallelism
% C, sequential region, cilk, parallax diamond
\noindent
This section introduces the language features that make Parallax a parallel
language.  Our language extensions are designed to make the parallelism
in a program both explicit and flexible.  These design decisions stem
from our goal to minimize the amount of compiler and runtime analysis
that must be done to parallelize a program. \\

\noindent
In our experience explicit parallelism is preferable to implicit parallelism
because it allows the programmer to say exactly where parallelism exists
without needing the compiler or runtime to infer it.  Implicit parallelism in
imperative languages is often difficult to infer due to the problem of creating
scalable pointer-based dependency analyses.  With our region programming model
we can actually infer all these dependencies because we know which regions every
statement in the program will side-effect.  However, this introduces a new 
problem: too much parallelism.  Much like pure functional languages, we can actually
infer more parallelism than we might actually want to exploit on a given machine.
Rather than attempt to have the compiler and runtime decide how to marshall all of
this parallelism we want to the programmer to declare this information explicitly.\\

\noindent
While we want our parallelism to be declared explicitly, we also want there
to be some flexibility in the parallelism so that we can map our computations onto
different architectures.  The amount of parallelism declared by the programmer
therefore should be an upper bound on the amount of parallelism which the programmer
thinks could be exploited by a machine.  There are then mechanisms in the mapping
file (see section \ref{Mapping}), the compiler, and the runtime system for serializing
parallelism as necessary for a specific machine. \\

\noindent
To create explicit and flexible parallelism we introduce a modified version of the
{\tt spawn} statement from the Cilk programming language\cite{Blumofe95}.  The
spawn statement can be used in Parallax in an identical way as in Cilk.  The spawn
keyword can be used to prefix any function call in the language to indicate that
the function could potentially be run in parallel.  It is then up to the compiler and
the runtime system to determine whether or not function calls are actually executed
in parallel\footnote{In Parallax this can also be specified in the mapping file.}.
Unlike Cilk, we do not require explicit synchronization with spawned functions.  The
return type of a spawned function will be a continuation.  The first place that either
the continuation is used, or a side-effected region of the spawned function is used
we can automatically insert the necessary synchronization. \\

\noindent
Our spawn statement also permits two additional optional arguments of which either
or both can be used.  The first optional argument specifies whether or not the
spawn statement {\tt may} or {\tt must} create parallel work.  By default a spawn
statement indicates the work {\tt may} be done in parallel.  However, if the programmer
needs to create parallel workers that are capable of synchronizing with each other
then he has to use the {\tt must} option.  The must option mandates that parallelism be
created and creates a way for the programmer to reason about the execution model
described in section \ref{Execution}. \\

\noindent
The second optional argument to the spawn statement is an iteration space of functions
to be in parallel.  An iteration space is simply a set of tuples for which an 
independent version of function is run.  The arguments to each function in an 
iteration space may dependent on the position in the iteration space.
Used in a {\tt may} parallelism context this is analogous to the {\tt mappar} 
statement in Sequoia \cite{Fatahalian06}.  In a {\tt must} parallelism context
this will ensure that all points in the iteration space are run in parallel and
have the ability to synchronize with each other\footnote{This might be achieved
by providing different hardware contexts for each computation, or time-multiplexed
scheduling by the runtime or operating system depending on the machine.}.  We
do not provide any built-in synchronization mechanisms for {\tt must} parallel
environments.  Instead through the use of simultaneous access regions (see section
\ref{Coherence}) the programmer can create shared regions which are visible to
all functions in the {\tt must} parallel context.  In this shared region the programmer
can create synchronization primitives as necessary\footnote{We ultimately plan to
implement a library for supporting this.}.

\noindent
As a summary, figure \ref{fig:taxonomy} presents a taxonomy of how the sequential
and parallel versions of the parallax language relate to C and Cilk in terms
of memory management and parallelism.

\begin{figure}[t]
\centering
\includegraphics[width=3.5in]{figs/Taxonomy.pdf}
\caption{Relationship between tradition C, Cilk, and the sequential and
parallel versions of the Parallax language \label{fig:taxonomy}}.
\end{figure}

\subsection{Coherence Capabilities \label{Coherence}}
% Describe need for different ways of accessing regions
% Enables additional parallelism
% Tensor of coherence interactions
\noindent
In section \ref{Parallelism} we introduced mechanisms to create parallel computation
in Parallax.  We now need to present a memory model for understanding how
parallel operations can reason about memory accesses.  In this section we present
a memory coherency model for reasoning about the order of accesses to a single location
in memory.  In section \ref{Consistency} we describe a memory consistency model for
reasoning about the ordering of accesses to different locations. \\

\noindent
In most shared memory models coherence is maintained at the granularity of an address 
by hardware mechanisms.  This coherence model is not scalable and often times is 
overly restrictive.  In our approach coherence is maintained at the granularity of a 
region and can be controlled by the software.  Our approach to coherence allows us to
specify both how a function is going to access a region as well as who else is allowed
to access a region at the same time.  This will enable the compiler and runtime schedulers
to better exploit parallel work and only perform data movement when it is required. \\

\noindent
Our approach to memory coherence is governed by the idea of capabilities \cite{Stork09,Bierhoff07}.
Capabilities provide a type system solution to controlling access to a shared resource.
In our case the shared resource is access to a region and we want to manage how
computations are permitted to access it.  We extend our type system from section \ref{TypeSystem}
to now require every function declaration to be explicit about the capabilities required
for each region that will be accessed by the function.  When we go to type check the program
we enforce the following invariant: every statement in a function must require a subset of the
capabilities declared in the function declaration to be correct. \\

\noindent
The set of capabilities that can be acquired by a function is represented in a two dimensional
space.  The first dimension deals with how the function will access the region, while the
second dimension determines what other access may be permitted to the region while the function
is being executed.  This two dimensional space and the set of symbols that we use to represent
each can be seen in table \ref{tab:capabilities}.

\begin{table}
\centering
\begin{tabular}{|c||c|c|c|c|}
\hline
 & Exclusive & Atomic & Simultaneous & Relaxed \\ \hline \hline
Read-Only & ROE & ROA & ROS & ROR \\ \hline
Read-Write & RWE & RWA & RWS & RWR \\ \hline
Reduction(Write-Only)$<$op$>$ & WOE & WOA & WOS & WOR \\ \hline
\end{tabular}
\caption{Two dimensional space of coherence capabilities for regions and their symbols. \label{tab:capabilities}}
\end{table}

\noindent
Each row in table \ref{tab:capabilities} represents a different way that a function
might access a region.  If a function is only going to be reading values from a
region then it needs read-only capabilities.  If a function is going to be mutating
a region then it must have the region with read-write capabilities.  Finally if a
region is only going to be performing a reduction into a region then it needs
reduction capabilities. \\

\noindent
Each column in table \ref{tab:capabilities} corresponds to a different way in which
other functions are permitted to be accessing the region at the same time. 
\begin{itemize}
\item Exclusive - mandates that no other transactions may be using
the region in any form while the functions is accessing the region.
\item Atomic - other functions may be accessing the region at the same
time, but this function must appear to execute atomically (serializably) with respect
to other functions accessing the region.  This is equivalent to transactional memory semantics.
\item Simultaneous - other functions may be accessing the region at the
same time and all updates made by other functions MUST be made visible to the 
function accessing the region.  This is equivalent to traditional hardware
single-writer/multiple-reader coherence.
\item Relaxed - other functions may be accessing the region at the same time
and updates made by other functions MAY be visible to this function.  To force updates to be
visible there will be a set of memory operations (i.e. invalidate/flush) that
will be used to update values just like in software coherence schemes.
\end{itemize}

\noindent
By providing each of these different coherence capabilities we allow the programmer
to say explicitly what the coherence properties are of the function that is
being executed.  Ultimately this will allow us to extract as much parallelism
as possible from programs and to minimize the amount of memory movement that
might ultimately need to occur in machines with distributed hierarchical memory. \\

\noindent
We now have twelve possible memory coherence capabilities that can be requested
over two dimensions.  We now want to examine the concurrency rules for which
capabilities are able to be held by two functions executing at the same time.  Rather
than enumerate all 144 entries in this fourth-order tensor, we instead give rules
for which capabilities can held by different functions and these functions performed in parallel.

\begin{itemize}
\item exclusive capabilities guarantee that now other function is accessing the region
\item atomic capabilities can only be used in conjunction with other functions which permit
serializability with the statements in those functions (strong isolation in TM terminology)
\item reduction operations can only be performed in parallel with reduction functions of
the same operand, otherwise they must be serialized
\end{itemize}

\noindent
It is important to note that for a sequential stream of statements to a region, all
capabilities must be acquired and released in the sequence of statements in the
program.  This guarantees classic sequential code guarantees.  However, outside
of this restriction, the compiler is free to re-order statements as desired as long
as the expected memory coherence behavior is respected.  The details of re-orderings
that are permitted are discussed in more detail in section \ref{Consistency}.

\subsection{Memory Consistency Model \label{Consistency}}
% Weak consistency model based on location consistency, but on regions
% Add fences for guarantees 
\noindent
In section \ref{Coherence} we defined how capabilities could be used to
control the ordering of accesses to a single region.  In this section we deal
with the memory consistency model that describes what reordering of operations
are permitted across multiple regions.  Our view on memory consistency is
that Parallax is a low-level programming model that requires explicit input
from the programmer about the desired program output.  In conjunction with 
this view, we propose to use a very relaxed memory consistency model
called location consistency\cite{Frigo98}.   We will then provide additional
memory fence operations to the programmer to prevent unwanted reorderings.\\

\noindent
Traditionally memory consistency models have been defined over the set of
all addresses in a memory.  In our approach however, memory consistency
is now defined over the set of regions in a program.  While the number of
regions in this set is unbounded, in practice it is many orders of magnitude
smaller than total number of addresses available in a program.  This justifies
our choice to use a more relaxed memory model such as location consistency 
since the programmer will have many fewer reorderings to consider due to 
the limited number of regions in scope at any point in the program. \\

\noindent
Location consistency proposes that the only reorderings that are not permitted
are ones that would break the serial semantics of a serial execution of stream
of instructions.  We therefore must obey the serial set of operations to a region,
but reordering statements to different regions is acceptable.  This will come
in especially helpful in our implementation when it may be beneficial to overlap
computation with communication. \\

\noindent
In some cases the reordering of statements that side-effect different
regions is undesirable.  To avoid such cases the programmer can insert memory
fences.  Memory fences take an argument of either a region identifier or a region
collection to prevent any reordering of statements that affect those regions.  If
this argument list is left empty then no statements are permitted to be reordered
across the memory fence.

\pagebreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Mapping

\section{Mapping Parallax to Hierarchical Memory Machines \label{Mapping}}
% Tunables
% Adding tags to source code, separate mapping file for each architecture
% Complements dynamic runtime decision making
% Place constraints on functions and region instances
\noindent
Much like the original Sequoia language, Parallax will also use a mapping
file to specify how to map a Parallax program onto a given architecture.  Unlike
Sequoia's mapping file, the Parallax mapping file will provide more flexible
ways of performing this mapping and a different execution model for the
Parallax program.

\subsection{Tunable Parameters and Tagged Statements \label{Tunables}}
\noindent
Similar to Sequoia, Parallax allows for type declarations to be declared as
tunable variables.  These tunable variables will have their value fixed at 
a later time as specified by the Parallax execution model\footnote{This is different from
Sequoia where they had to be fixed at compile-time.}.  The tunable qualifier
can be applied to any plain-old-data (POD) type or to any function pointer.
The need for tunable function pointers is useful in the case where the algorithm
may either opt to recursively decompose a data structure, or perform the base computation.
In these scenarios a tunable function pointer can be used to represent both options. \\

\noindent
Unlike Sequoia tunable parameters can either be assigned explicit values or they
can be assigned a finite subset of values from which the compiler/runtime system
may choose.  These sets can either be specified in the original source code if they
are machine agnostic (i.e. some function pointers) or they can be specified in the
mapping file.  A mapping file must specify either a value or a set of values for every
tunable parameter in a source code or there will be a compile-time error.  This
ensures that the compiler/runtime will at least have a set of valid ranges over which
to perform a search if only a set is specified.\\

\noindent
In addition to specifying tunable variables, we also permit the programmer to tag
statements with labels as ways of providing a naming mechanism for use in the mapping
file.  By being able to tag statements, the programmer can add additional constraints
to the program that machine specific.  For example, a programmer can tag a {\tt spawn}
statement and in the mapping file indicate that for a specific machine the statement
must be executed in a {\tt must} context.  This would force the system to create
parallelism even if it potentially degraded performance.  The opposite would be to
tag a {\tt spawn} statement as a {\tt not} context in which case the compiler and 
runtime cannot create parallelism at this function call. \\

\noindent 
Tagged statements can also be used for specifying compiler and runtime plugins for
use with a specific statement as discussed in section \ref{Plugins}.

\subsection{Exact and Relative Mappings \label{ConstraintMap}}

\subsection{Compiler/Runtime Plugins \label{Plugins}}

\subsection{Constraint-Based Execution Model \label{Execution}}

\pagebreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Outstanding Questions

\section{Outstanding Issues \label{OutstandingQs}} 
\noindent
Below are some issues that we have yet to resolve.
\begin{itemize}
\item Region Slicing: How do we describe region slicing for regions whose 
types are complex data types?  Are the regions created by region slicing 
still sub-regions?  How does this relate to the sub-region relationship?
\item What operations are supported on region collections?  Can region collections
be passed everywhere that a region identifier is passed?
\item How do we differentiate arrays form base region types?  Are there special
operations for arrays that we want to support that cannot be supported on
general regions?  Are these built-in operations?
\end{itemize}

\pagebreak

\bibliographystyle{IEEEtran}
\bibliography{parallax_model}

\end{document}

### 2011-11-02 doc/pldi2012/code_ex.tex
initial text for code example and low-level runtime implementation

\lstset{
  language=C++,
  basicstyle=\scriptsize,
  numbers=left,
  numberstyle=\tiny,
  stepnumber=1,
  escapechar=\#,
  keepspaces=true,
  morekeywords={region_relation,region,coloring,partition},
  deletekeywords=float,
}
\begin{lstlisting}[float]
struct Node<rn>    { Node<rn>@rn next;    float charge, capacitance; }
struct Wire<rn,rn2,rw> { Wire<rn,rn2,rw>@rw next; Node<rn2>@rn in_node, out_node; float current, ... ; }
region_relation Circuit {
  region< Node<r_all_nodes> >                r_all_nodes;
  region< Wire<r_all_nodes,r_all_wires> >    r_all_wires;
  Node<r_all_nodes>@r_all_nodes              first_node;
  Wire<r_all_nodes,r_all_wires>@r_all_wires  first_wire;
}
region_relation CircuitPiece<rn, rw> {
  region< Node<rn_pvt+rn_shr> >                    rn_pvt (< rn), rn_shr (< rn);
  region< Node<rn> >                               rn_ghost (< rn);
  region< Wire<rn_pvt+rn_shr+rn_ghost,rn,rw_pvt> > rw_pvt (< rw);
  Node<rn_pvt+rn_shr>@(rn_pvt+rn_shr)              first_node;
  Wire<rn_pvt+rn_shr+rn_ghost,rn,rw_pvt>@rw_pvt    first_wire;
};
void simulate_circuit(Circuit c) : RWE(c.r_all_nodes,c.r_all_wires)
{
  coloring<c.r_all_wires> wire_owner_map = ... ; // wires colored by which piece they're in
  partition<c.r_all_wires> p_wires = partition(c.r_all_wires, wire_owner_map);
  coloring<c.r_all_nodes> node_owner_map = ... ; // nodes colored by which piece they're in
  coloring<c.r_all_wires> node_nghbr_map = ... ; // nodes colored by which pieces they neigbor
  coloring<c.r_all_wires> node_privacy_map = ... ; // nodes colored: 0 = no neighbors, 1 = some neighbors
  partition<c.r_all_nodes> p_nodes_pvs = partition(c.r_all_nodes, node_privacy_map);
  partition<p_nodes_pvs[0]> p_pvt_nodes = partition(p_nodes_pvs[0], node_owner_map);
  partition<p_nodes_pvs[1]> p_shr_nodes = partition(p_nodes_pvs[1], node_owner_map);
  partition<p_nodes_pvs[1]> p_ghost_nodes = partition(p_nodes_pvs[1], node_nghbr_map);

  CircuitPiece<c.r_all_nodes,c.r_all_wires> pieces[MAX_PIECES];
  for(i = 0; i < MAX_PIECES; i++) 
    pieces[i] <- { rn_pvt = p_pvt_nodes[i], rn_shr = p_shr_nodes[i],
                   rn_ghost = p_ghost_nodes[i], rw_pvt = p_wires[i] };

  while(!done) {
    for(i = 0; i < MAX_PIECES; i++) spawn(calc_new_currents(pieces[i]));
    for(i = 0; i < MAX_PIECES; i++) spawn(distribute_charge(pieces[i]));
    for(i = 0; i < MAX_PIECES; i++) spawn(update_voltages(pieces[i]));
  }
}

void calc_new_currents(CircuitPiece<rn,rw> piece): RWE(piece.rw_pvt), ROE(piece.rn_pvt,piece.rn_shr,piece.rn_ghost) {
  // read info from nodes connected to each wire, update state of wire
}

void distribute_charge(CircuitPiece<rn,rw> piece): ROE(piece.rw_pvt), RdA(piece.rn_pvt,piece.rn_shr,piece.rn_ghost) {
  // current moving through wires redistributes charge between nodes
}

void update_voltages(CircuitPiece<rn,rw> piece): RWE(piece.rn_pvt,piece.rn_shr)
{
  // total charge added to a node causes changes in voltage
}
\end{lstlisting}

### 2011-11-02 doc/pldi2012/lowlevel.tex
initial text for code example and low-level runtime implementation

\section{Low-Level Runtime}

The job of the low-level runtime is to provide a machine-independent layer
for the high-level runtime to sit on top of.  It needs to hide the
ideosynchracies of the underlying machine, operating system, and device
drivers.  At the same time, it needs to avoid getting too far away from the 
native capabilities of the hardware and software in order to obtain an 
efficient implementation.

The target machines for Legion include both clusters and GPU-equipped nodes,
and neither of these are able to efficiently provide a uniformly addressable
and accessible memory space across the system.  The magic of an
apparently-unified address space is left to the high-level runtime, and the
interface to the low-level runtime is done in terms of handles rather than 
pointers for runtime objects and relative offsets rather than absolute 
pointers for resource allocations.

The low-level runtime chooses to provide a ``deferred execution'' model rather
than the more common ``immediate'' model.  Requests made of the low-level 
runtime usually return to the caller immediately, with the requested action
performing asynchronously to the caller's thread of execution.  This is done
for two different, but equally important, reasons.  First, it matches the
standard GPU execution model.  GPUs tend to have very deep pipelines, and
maximum performance cannot be achieved if each command has to be run to
completion before the next command can be started.  Second, in clusters, the
latency of inter-node communication can be significant, and a caller that
waited for every request of a remote node to return before continuing on would
also be similarly underutilizing the computational resources of the system.

Although it does not show up explicitly in the interface, the low-level runtime
was also architected to permit a distributed and hierarchical implementation.
Critical resources can be replicated or migrated between nodes to reduce
both average latencies and the amount of traffic that use the scarce inter-node
communication bandwidth.

\subsection{Events}

The basic building block of the low-level runtime's deferred execution model
is the Event.  An Event is a placeholder for the completion of a deferred
operation.  When a call is made to the low-level runtime, an Event is usually
returned to the caller.  The caller may query whether the event has occurred,
and may also ask to wait until the event occurs, but the more common case is
to provide one or more Events to subsequent requests made of the runtime, 
causing those requests to be automatically deferred until the previous
operations corresponding to those Events have completed.  These requests will
also return Events, and the caller can set up a complicated, but coordinated,
set of operations to be performed by the low-level runtime with no babysitting
required of the original caller.

With nearly every single runtime operation creating a new Event and referring
to potentially any number of past Events, both the time and space overhead of
these Events is a serious concern.  The time overhead is addressed by a
mechanism that allows a node to create a new Event without needing to
communicate with any other node.  The Event's handle identifies which node
created it, allowing another node to lazily access it only when and if it 
is asked to perform an operation on it.  Checking to see if an Event has
occurred is done via a subscription mechanism.  Any number of threads or
deferred operations on one node can query or wait on an Event with a single
message sent to the originating node.  The requesting node tracks the multiple
local requestors and upon receiving the trigger message from the originating
node, distributes that message to all the local parties that are interested.

The potential space concern with Events is addressed by a generation scheme
that allows Events to be reclaimed and reused without tracking references.  The
naive implementation of an Event cannot reclaim the storage used by an Event
until the Event has triggered \emph{and} all waiters on the event have been
notified.  Some of these waiters may take quite a while to make their query,
and worse, we must somehow track how many queries to expect if we want to know
when they've all been performed.  However, if the Event handle is paired with a
``generation'' count that tracks how many times the Event has been triggered,
the Event can be reused as soon as the previous generation has been triggered.
Holders of an older (handle,generation) pair are able to query the newer
incarnation of the Event and determine from the generation mismatch that their
Event must have already triggered.  With this technique, there is no need to
count references to the Event at all.  An Event can only be reused until the
generation counter reaches its maximum value, the current implementation
supports $2^{32}-1$ generations, and should that not be sufficient, extending
it to $2^{64}-1$ is trivial.

\subsection{Deferred Locks}

Events are used to provide a deterministic ordering between operations, but in
some cases, enforcing only serializability is a sufficient for correctness and
can result in much better performance.  The traditional tool for this job is
the mutex (a.k.a. lock), and it is from the common reader/writer lock that
the low-level runtime's \emph{deferred lock} is derived.  The key visible 
difference between a mutex and a deferred lock is that the request does not
block the caller's execution.  Many libraries offer ``nonblocking locks'', but
they achieve this by returning without the lock if it is unavailable.  If
a deferred lock is currently unavailable, the request is guaranteed to be
granted at some time in the future (with no further action required of the 
caller).  That time in the future is described with an Event.  Like all other
Events, it can be given to other threads, or to the runtime, or even stored
in a data structure in memory to be extracted by an arbitrary thread at an 
arbitrary time.  This dissociation of a held lock from the thread that
requested it is another important difference between the deferred lock and
existing mutexes.

The deferred nature of a deferred lock also allows a distributed implementation
of locks.  Similar to Events, a deferred lock can be created on any node
without communication with the other nodes in the cluster.  That node becomes
the initial home for the lock, but the deferred lock (unlike events) can 
migrate to other nodes or even be shared between nodes, allowing repeated
lock requests from a single node to be kept local to that node after a single
message is sent to the current owner.  (Should the ownership of the lock change
while the message is in flight, the previous owner forwards the request to
the new owner.  The highly unlikely, but possible, case in which multiple 
forwardings are required is bounded at the number of nodes in the cluster.)
The ability to migrate locks between nodes allows the performance of a
heavily-contended cluster-level lock to approach that of a single-node lock.

The final interesting feature of the deferred lock is that it can be associated
with an object in memory and that object will automatically be kept coherent
across the cluster.  This is similar to MESI coherence in a distributed cache,
but is done per-object rather than on cache lines.  This is currently only 
used for object within the runtime itself, but we are exploring ways to expose
it at the Legion application level.

\subsection{Processors}

Hardware units capable of performing computation are captured as generic
processors in the low-level runtime.  Like other runtime objects, a processor
has a globally unique name and can be referred to anywhere in the system.
Processors come in different \emph{kinds} (e.g. X86 CPU core, CUDA-capable GPU),
but they all support the same
interface and execution model.  A request to spawn a task on a given processor
may come from anywhere in the system.  The request names which task should
be spawn and provides input arguments and zero or more events.  Once those
events have occurred, the task is placed into a queue of ready tasks, which
the processor works through in the order in which they were enqueued.  A
processor will generally run a task to completion before taking another, but
will try to run another task (resources permitting) if the current task needs
to wait for an event to occur.  Once a task is placed on a processor's queue,
it may not be moved or deleted.  Load-balancing of tasks is expected to be done
by the high-level runtime.  The primary purpose of the queue for the low-level
Processor is to be able to hide the latency of starting new tasks.

\subsection{Memories}

Application-visible storage in the low-level runtime is described in terms of
memory objects.  A memory object has a size and describes its affinity to 
processors (and other memories) in terms of relative bandwidth and latency.
The low-level runtime allows for the allocation and freeing of region instances
and allocators.  It also supports bulk transfers from one memory to another.
These bulk transfers can often be implemented using dedicated DMA (or RDMA)
hardware, freeing up the processors to perform more interesting
computation.

\subsection{Regions}

Application-level regions are broken into four inter-related pieces in the
low-level runtime:
\begin{description}

\item[MetaData] The metadata for a logical region captures attributes of a
logical region that are common across all instances.  These attributes include
the size of the logical region, the element type, and the allocation mask,
which tracks which portions of the logical region are actually in use.  Region
MetaDatas are created independently of any memory objects and are accessible
(via a globally unique name) to all nodes in the machine.

\item[Allocator] The dynamic memory allocation capability of a region instance
is provided via an allocator object.  An allocator is associated with a
region metadata and must be created in a particular memory.  Once created, it
may be used to allocate and free elements of the region by any task on any
processor that has non-zero affinity with the memory.

\item[Instance] Each physical instance of a region stores some version of all
the elements in a region.  As with allocators, a physical instance is
associated with a particular region metadata and spends its entire lifetime in
the memory objecy in which it was created.  When multiple physical instances
exist for the same region, the low-level runtime does nothing to keep them
coherent - it is the job of the high-level runtime to perform the appropriate
copies and set up the right inter-operation dependencies to provide the level
of coherence requested by the application.  Although a physical instance 
stores the contents of a region, it does not directly offer access to those
contents.

\item[Accessor] Access to the contents of a physical region instance are 
handled by accessor objects.  These are lightweight objects that allow a given
task to access a given physical region instance, regardless of the
implementation of that instance or the underlying memory.  In the most common
case, read and write access is done through a single array dereference, but the
accessor interface allows seamless use of other types of memory as well (e.g.
memory that's accessed via RMDA put and get calls).

\end{description}

\subsection{Initialization}

On startup, the low-level runtime allocates all the resources it needs from
the operating system and set up the communication channels between nodes.
Each processor in the system then automatically runs the high-level runtime's
initialization task, which will usually spawn more tasks.  These tasks are
started, and whenever any processor's task queue is empty , the high-level
runtime's scheduler task is automatically called to explore the possibility
of moving tasks around for load balancing.

\subsection{SMP Implementation}

The SMP implementation of the low-level runtime is a relatively simple one.
It runs on any SMP computer that supports POSIX threads.  A single memory
object is created for the shared memory, and a processor object is created for
each CPU core in the system.  Any GPUs in the system are ignore.  Events are 
mutex-protected data structures that maintain a list of callbacks to call when
the corresponding event is triggered.  Locks are similar, except that the list
is of pending lock requests to attempt when the previous lock is released.

\subsection{GPU Cluster Implementation}

The cluster implementation of the low-level runtime is implemented on a mix
of POSIX threads for intra-node threading and communication, GASnet for 
inter-node communication, and CUDA for access to resources of the GPUs in 
each node.  Locks and events have roughly the same implementation as the SMP
runtime for handling multiple local requestors, but use GASnet active messages
to coordinate lock ownership and event subscription and triggering between
nodes.  In addition to the processor objects for each CPU core, a single
processor object is created for each GPU in the system.  It would be preferable
to expose the GPU's computational resources at a finer granularity, but
current GPU hardware doesn't directly allow this flexibility and recent attempts
at implementing such control in a software layer on top of CUDA have suffered
from significant overhead.  The memory hierarchy is expanded to include a 
``global'' memory that is striped across all nodes and accessible via GASnet
RDMA calls, and two memories per GPU: one for the GPU's framebuffer that is
accessible only to that GPU and by bulk DMA operations, and one for the pinned
``zero copy'' system memory that is accessible to the GPU and all the CPU cores
in a single system.

### 2011-11-02 runtime/lowlevel.cc
checkpoint

@@ -7,6 +7,7 @@
 #endif
 
 GASNETT_THREADKEY_DEFINE(cur_thread);
+GASNETT_THREADKEY_DEFINE(gpu_thread);
 
 #if 0
 #include <assert.h>
@@ -933,11 +934,11 @@ namespace RegionRuntime {
 
       virtual void event_triggered(void)
       {
-	log_event(LEVEL_INFO, "recevied trigger merged event %x/%d",
-		  finish_event.id, finish_event.gen);
 	bool last_trigger = false;
 	{
 	  AutoHSLLock a(mutex);
+	  log_event(LEVEL_INFO, "recevied trigger merged event %x/%d (%d)",
+		    finish_event.id, finish_event.gen, count_needed);
 	  count_needed--;
 	  if(count_needed == 0) last_trigger = true;
 	}
@@ -2780,9 +2781,27 @@ namespace RegionRuntime {
 
       // figure out which thread we are - better be a local CPU thread!
       void *ptr = gasnett_threadkey_get(cur_thread);
-      assert(ptr != 0);
-      LocalProcessor::Thread *thr = (LocalProcessor::Thread *)ptr;
-      thr->sleep_on_event(*this);
+      if(ptr != 0) {
+	LocalProcessor::Thread *thr = (LocalProcessor::Thread *)ptr;
+	thr->sleep_on_event(*this);
+	return;
+      }
+      // maybe a GPU thread?
+      ptr = gasnett_threadkey_get(gpu_thread);
+      if(ptr != 0) {
+	printf("oh, good - we're a gpu thread - we'll spin for now\n");
+	printf("waiting for %x/%d\n", id, gen);
+	while(!e->has_triggered(gen)) usleep(1000);
+	printf("done\n");
+	return;
+      }
+      // we're probably screwed here - try waiting and polling gasnet while
+      //  we wait
+      printf("waiting on event, polling gasnet to hopefully not die\n");
+      while(!e->has_triggered(gen))
+	gasnet_AMPoll();
+      return;
+      //assert(ptr != 0);
     }
 
     // can't be static if it's used in a template...
@@ -3595,6 +3614,22 @@ namespace RegionRuntime {
 	    }
 	    break;
 
+	  case Memory::Impl::MKIND_GPUFB:
+	    {
+	      // all GPU operations are deferred, so we need an event if
+	      //  we don't already have one created
[…]

### 2011-11-02 runtime/lowlevel_gpu.cc

@@ -61,6 +61,7 @@ namespace RegionRuntime {
 
     class GPUProcessor::Internal {
     public:
+      GPUProcessor *gpu;
       int gpu_index;
       size_t zcmem_size, fbmem_size;
       size_t zcmem_reserve, fbmem_reserve;
@@ -69,6 +70,7 @@ namespace RegionRuntime {
       void *fbmem_gpu_base;
 
       bool initialized;
+      bool worker_enabled;
       bool shutdown_requested;
       pthread_t gpu_thread;
       gasnet_hsl_t mutex;
@@ -76,7 +78,7 @@ namespace RegionRuntime {
       std::list<GPUJob *> jobs;
 
       Internal(void)
-	: initialized(false), shutdown_requested(false)
+	: initialized(false), worker_enabled(false), shutdown_requested(false)
       {
 	gasnet_hsl_init(&mutex);
 	gasnett_cond_init(&parent_condvar);
@@ -85,6 +87,8 @@ namespace RegionRuntime {
 
       void thread_main(void)
       {
+	gasnett_threadkey_set(gpu_thread, this);
+
 	CHECK_CUDART( cudaSetDevice(gpu_index) );
 	CHECK_CUDART( cudaSetDeviceFlags(cudaDeviceMapHost |
 					 cudaDeviceScheduleBlockingSync) );
@@ -108,6 +112,12 @@ namespace RegionRuntime {
 	  AutoHSLLock a(mutex);
 	  initialized = true;
 	  gasnett_cond_signal(&parent_condvar);
+
+	  // wait until we've been told to proceed
+	  while(!worker_enabled) {
+	    log_gpu.info("waiting for enable signal");
+	    gasnett_cond_wait(&worker_condvar, &mutex.lock);
+	  }
 	}
 
 	while(!shutdown_requested) {
@@ -115,28 +125,42 @@ namespace RegionRuntime {
 	  GPUJob *job;
 	  {
 	    AutoHSLLock a(mutex);
-	    while(jobs.size() == 0) {
-	      printf("job queue empty - sleeping\n");
-	      gasnett_cond_wait(&worker_condvar, &mutex.lock);
-	      if(shutdown_requested) {
-		printf("awoke due to shutdown request...\n");
-		break;
+	    while((jobs.size() == 0) && !shutdown_requested) {
+	      // see if there's an idle task we should run
[…]

### 2011-11-02 doc/pldi2012/code_ex.tex

@@
 \lstset{
+  captionpos=b,
   language=C++,
@@
   numberstyle=\tiny,
+  columns=fullflexible,
   stepnumber=1,
@@
   keepspaces=true,
-  morekeywords={region_relation,region,coloring,partition},
+  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
+  morekeywords={region_relation,region,coloring,partition,spawn},
   deletekeywords=float,
 }
-\begin{lstlisting}[float]
+\begin{lstlisting}[float,caption={Circuit Simulation Code Example}]
 struct Node<rn>    { Node<rn>@rn next;    float charge, capacitance; }
-struct Wire<rn,rn2,rw> { Wire<rn,rn2,rw>@rw next; Node<rn2>@rn in_node, out_node; float current, ... ; }
+struct Wire<rn,rn2,rw> { Wire<rn,rn2,rw>@rw next;
+                                          Node<rn2>@rn in_node, out_node; float current, ... ; }
 region_relation Circuit {
@@
 region_relation CircuitPiece<rn, rw> {
-  region< Node<rn_pvt+rn_shr> >                    rn_pvt (< rn), rn_shr (< rn);
-  region< Node<rn> >                               rn_ghost (< rn);
-  region< Wire<rn_pvt+rn_shr+rn_ghost,rn,rw_pvt> > rw_pvt (< rw);
+  region< Node<rn_pvt+rn_shr> >                                 rn_pvt (#$\prec$# rn), rn_shr (#$\prec$# rn);
+  region< Node<rn> >                                                   rn_ghost (#$\prec$# rn);
+  region< Wire<rn_pvt+rn_shr+rn_ghost,rn,rw_pvt> >  rw_pvt (#$\prec$# rw);
   Node<rn_pvt+rn_shr>@(rn_pvt+rn_shr)                     first_node;
@@
 {
-  coloring<c.r_all_wires> wire_owner_map = ... ; // wires colored by which piece they're in
-  partition<c.r_all_wires> p_wires = partition(c.r_all_wires, wire_owner_map);
-  coloring<c.r_all_nodes> node_owner_map = ... ; // nodes colored by which piece they're in
-  coloring<c.r_all_wires> node_nghbr_map = ... ; // nodes colored by which pieces they neigbor
-  coloring<c.r_all_wires> node_privacy_map = ... ; // nodes colored: 0 = no neighbors, 1 = some neighbors
-  partition<c.r_all_nodes> p_nodes_pvs = partition(c.r_all_nodes, node_privacy_map);
-  partition<p_nodes_pvs[0]> p_pvt_nodes = partition(p_nodes_pvs[0], node_owner_map);
-  partition<p_nodes_pvs[1]> p_shr_nodes = partition(p_nodes_pvs[1], node_owner_map);
-  partition<p_nodes_pvs[1]> p_ghost_nodes = partition(p_nodes_pvs[1], node_nghbr_map);
+  partition<c.r_all_wires> p_wires(wire_owner_map); // colored by piece they're in
+  partition<c.r_all_nodes> p_nodes_pvs(node_sharing_map);
+                                                                    // true = some neighbors in other pieces
+  partition<p_nodes_pvs[false]> p_pvt_nodes(node_owner_map);
+                                                                    // colored by piece they're in
+  partition<p_nodes_pvs[true]> p_shr_nodes(node_owner_map);
+                                                                    // colored by piece they're in
+  partition<p_nodes_pvs[true]> p_ghost_nodes(node_nghbr_map);
+                                                                    // colored by pieces they neighbor
 
   CircuitPiece<c.r_all_nodes,c.r_all_wires> pieces[MAX_PIECES];
-  for(i = 0; i < MAX_PIECES; i++) 
-    pieces[i] <- { rn_pvt = p_pvt_nodes[i], rn_shr = p_shr_nodes[i],
+  for(i = 0; i #$<$# MAX_PIECES; i++) 
+    pieces[i] #$\gets$# { rn_pvt = p_pvt_nodes[i], rn_shr = p_shr_nodes[i],
                             rn_ghost = p_ghost_nodes[i], rw_pvt = p_wires[i] };
@@
   while(!done) {
-    for(i = 0; i < MAX_PIECES; i++) spawn(calc_new_currents(pieces[i]));
-    for(i = 0; i < MAX_PIECES; i++) spawn(distribute_charge(pieces[i]));
-    for(i = 0; i < MAX_PIECES; i++) spawn(update_voltages(pieces[i]));
+    for(i = 0; i #$<$# MAX_PIECES; i++) spawn calc_new_currents(pieces[i]);
+    for(i = 0; i #$<$# MAX_PIECES; i++) spawn distribute_charge(pieces[i]);
+    for(i = 0; i #$<$# MAX_PIECES; i++) spawn update_voltages(pieces[i]);
   }
@@
 
-void calc_new_currents(CircuitPiece<rn,rw> piece): RWE(piece.rw_pvt), ROE(piece.rn_pvt,piece.rn_shr,piece.rn_ghost) {
+void calc_new_currents(CircuitPiece<rn,rw> piece): RWE(piece.rw_pvt),
+                                                                         ROE(piece.rn_pvt,piece.rn_ghost) {
   // read info from nodes connected to each wire, update state of wire
@@
 
-void distribute_charge(CircuitPiece<rn,rw> piece): ROE(piece.rw_pvt), RdA(piece.rn_pvt,piece.rn_shr,piece.rn_ghost) {
+void distribute_charge(CircuitPiece<rn,rw> piece): ROE(piece.rw_pvt),
+                                                                         RdA(piece.rn_pvt,piece.rn_ghost) {
   // current moving through wires redistributes charge between nodes

### 2011-11-03 doc/pldi2012/highlevel.tex
Starting work on the high level runtime section

\section{High-Level Runtime} 
\label{sec:highlevel}
The high-level runtime is responsible for guaranteeing the semantics 
of programs written in the Legion programming model while at the same
time extracting high performance.  This goal is challenging because
the high-level runtime must be capable of operating on top of many 
different architectures abstracted by the low-level runtime.  This 
includes those with very large latencies for transferring data such
as distributed memory clusters.  To hide these latencies, the high
level runtime uses a deffered model of execution made possible by the
event model of the low-level runtime.  By deffering execution, the 
runtime can hide the latencies of data and task movement and acieve
good throughput and high performance.

However, a deffered execution model presents challenges to implementing
the semantics of the Legion programming model.  Deffered execution
does not necessarily imply that tasks will be executed in the order in
which they are scheduled, but only in the order in which low-level
event dependencies are created.  To orchestrate the execution of Legion
programs in this environment the high-level runtime is architected
in a manner similar to an out-of-order hardware processor.  There are
multiple stages to executing a task which allows the runtime to extract
as much task-level parallelism as possible from an application whithin
the constraints of the Legion programming model.  We now investigate
each of these stages in further detail.

\subsection{Dependence Analysis} 
\label{subsec:dep_analysis}
The first step in the execution of a task is dependence analysis.  When
a task is registered with the high level runtime the task call includes
information about both the logical regions that the task will require
when it is executed as well as the read-write access and coherence properties
for each of the regions.  The runtime uses this information to perform
conflict detection with each of the previous tasks that have been registered
in the same parent task.  Conflict detection only needs to be performed at
the scope of an enclosing task because the semantics of the Legion programming
model ensure that all tasks can only use regions which are subregions of
regions that the parent task uses.  This implies that if a child task were
to have a conflict with task at a higher level of the task tree, then the
parent task would also have had a conflict.

Lemma: If a task {\tt t} with ancestor task {\tt p} conflicts with task {\tt t'}
a sibling task of {\tt p}, then tasks {\tt p} and {\tt t'} must conflict.

This property of the Legion programming model enables the runtime to only 
have to perform dependence analysis between tasks which share the same 
parent task.  By restricting dependence analysis to tasks which share the same
parent task, the runtime can perform dependence analysis locally and not
have to be concerned with tasks being created in other parts of the machine.

\subsubsection{Detecting Dependences}
\label{subsec:dep_detect}
To detect dependences between the tasks, the runtime leverages its knowledge
about logical regions and their partitions.  To create a new logical region
or partition, the application must invoke the corresponding call in the
runtime.  As these calls occur, the runtime constructs a data strcture
called a {\em region tree}.  A region tree describes the relationship between
logical regions and partitions.  A region tree contains two types of nodes:
region nodes and partition nodes.  Every region tree is rooted with a region
node and alternates between region and partition nodes each corresponding level.
A region node represents a specific logical region and tracks the set of
partitions of that logical region.  Similarly a partition node represents 
a specific partitioning of a region and keeps track of the logical regions 
which are subregions of the partition.

For every task, the runtime keeps track of the subset of region trees that
the task uses.

\subsubsection{Managing Region Trees}

\subsection{Mapping}

\subsubsection{Placing Tasks}

\subsubsection{Tracking Instances}

\subsection{Execution}

\subsection{Task Completion}

### 2011-11-04 doc/pldi2012/highlevel.tex
More writing for the high level runtime

@@
 \subsection{Dependence Analysis} 
-\label{subsec:dep_analysis}
+\label{subsec:depanalysis}
 The first step in the execution of a task is dependence analysis.  When
@@
 
-\subsubsection{Detecting Dependences}
-\label{subsec:dep_detect}
+When performing dependence analysis the runtime relies on its knowledge about
+the relationships between logical regions and partitions.  We first describe
+the the components to dependence analysis in sections \ref{subsec:regiontree},
+\ref{subsec:cohconf}, and \ref{absinsts}.  We then describe the algorithm for performing 
+dependence analysis on this data structures in section \ref{subsec:depdetect}.
+
+\subsection{Region Trees}
+\label{subsec:regiontree}
 To detect dependences between the tasks, the runtime leverages its knowledge
@@
 region nodes and partition nodes.  Every region tree is rooted with a region
-node and alternates between region and partition nodes each corresponding level.
+node and alternates between region and partition nodes at each corresponding level.
 A region node represents a specific logical region and tracks the set of
 partitions of that logical region.  Similarly a partition node represents 
-a specific partitioning of a region and keeps track of the logical regions 
+a specific partitioning of a logical region and keeps track of the logical regions 
 which are subregions of the partition.
 
-For every task, the runtime keeps track of the subset of region trees that
-the task uses.
+\subsection{Coherence Conflicts}
+\label{subsec:cohconf}
+To detect conflicts between any two tasks using a logical region, we leverage
+the information contained in the access and cohrence properties provided by the programmer
+for how the region is to be used.  Conflict checks are performed pair-wise between regions
+to determine whether two tasks conflict with each other.  When checking for
+conflicts we first examine the access property.  If both tasks are reads, then 
+there is never a conflict.  If at least one of them is a write we use table
+\ref{tab:conflicts} to determine whether there is a conflict.
 
-\subsubsection{Managing Region Trees}
+\begin{center}
+\begin{table}
+\begin{tabular}{|c|c|c|c|c|} \hline
+             & Exclusive & Atomic   & Simultaneous & Relaxed \\ \hline 
+Exclusive    & Conflict  & Conflict & Conflict     & Conflict \\ \hline
+Atomic       & Conflict  & Same Instance & Contingent & Contingent \\ \hline
+Simultaneous & Conflict  & Contingent & Same Instance & None \\ \hline
+Relaxed      & Conflict  & Contingent & None & None \\ \hline
+\end{tabular}
+\caption{Conflict detection cases when at least one access is a write.\label{tab:conflicts}}
+\end{table}
+\end{center}
+In this table, {\em Conflict} indicates that there will always be a conflict between
+regions while {\em None} indicates that there will never be a conflict regardless
+of mapping decisions.  In the case of {\em Same Instance} there is a conflict contingent
+upon a mapping decision.  If both tasks map to the same physical instance (described in 
+more detail in section \ref{subsec:instmang}) then there is no conflict, which if
+they map to different instances there is a conflict as the system will have to generate
+a copy between the two tasks running.
+
+For the case of {\tt Contingent} there are two special cases that we handle.
+One special case is Write-After-Read (WAR) conflicts.  For the WAR case, if neither coherence property
+is {\em Exclusive} then we can ensure that different physical instances
+are used to avoid a conflict and increase parallelism.  If either property is {\em Exclusive} then the semantics
+of the programming model mandate that we detect a conflict and serialize the task.
+
+Another special case occurs if there is an {\tt Atomic} writer and either {\tt Simultaneous}
+or {\tt Relaxed} reader.  In this case it is acceptable for the tasks to run in
+parallel as it is valid for the reader to see updates form the writer, and the
+reader will not violate the atomic access semantics of the writer.
+
+\subsection{Abstract Instances}
+\label{subsec:absinsts}
+
+\subsection{Dependence Detection}
+\label{subsec:depdetect}
+When a task is registered, the runtime
+will perform conflict detection for every region that the task requests with
+prior tasks that have been registered in the same parent task context.  Referring
+back to our analogy of the high level runtime as an out of order processor, this
+corresponds to the wake-up logic that detects when an instruction is ready to be
+executed.  Continuing this analogy, the regions that a task requires are annalogous
+to the registers listed in an instruction.  However, this poses a difficulty for the
+high level runtime as two logical regions are not necessarily independent.  In
+order to do the conflict detection we rely on the region tree data structure to be
+able to detect conflicts between tasks which require different logical regions.
+
+To perform dependence detection, we first find the {\em root region} for the 
+requested region.  The root region is the ancestor region for which the parent task
+has privliges.  Note that the semantics of the Legion programming model state that
+we must always be able to find such a root region or else an error is thrown.
+
+Once we have found the root region for the requested region, we then proceed to walk
+down the region tree from the node corresponding to the root region to the node
+for the requested region.  At each node, we determine if any potential conflicts exist.
+
+For region nodes we first check to see if any tasks currently using this region
+conflict based on the coherence proprties the two task require.  We describe coherence
+conflicts in more detail in section \ref{subsec:cohconf}.  If a conflict is discovered
+we record that the task must wait on the previous task before executing.  We also check
+to see if any of the partitions of the region have been marked open.  If they have
+we must {\em close} the partitions to detect any possible conflicts
+
+\subsection{Correctness}
 
@@
 
-\subsubsection{Tracking Instances}
+\subsubsection{Instance Management}
+\label{subsec:instmang}
 

### 2011-11-04 doc/pldi2012/highlevel.tex

@@
 Exclusive    & Conflict  & Conflict & Conflict     & Conflict \\ \hline
-Atomic       & Conflict  & Same Instance & Contingent & Contingent \\ \hline
-Simultaneous & Conflict  & Contingent & Same Instance & None \\ \hline
-Relaxed      & Conflict  & Contingent & None & None \\ \hline
+Atomic       & Conflict  & Same Instance & Conflict & Conflict \\ \hline
+Simultaneous & Conflict  & Conflict & Same Instance & None \\ \hline
+Relaxed      & Conflict  & Conflict & None & None \\ \hline
 \end{tabular}
-\caption{Conflict detection cases when at least one access is a write.\label{tab:conflicts}}
+\caption{Conflict detection cases}
 \end{table}
 \end{center}
-In this table, {\em Conflict} indicates that there will always be a conflict between
-regions while {\em None} indicates that there will never be a conflict regardless
-of mapping decisions.  In the case of {\em Same Instance} there is a conflict contingent
-upon a mapping decision.  If both tasks map to the same physical instance (described in 
-more detail in section \ref{subsec:instmang}) then there is no conflict, which if
-they map to different instances there is a conflict as the system will have to generate
-a copy between the two tasks running.
-
-For the case of {\tt Contingent} there are two special cases that we handle.
-One special case is Write-After-Read (WAR) conflicts.  For the WAR case, if neither coherence property
-is {\em Exclusive} then we can ensure that different physical instances
-are used to avoid a conflict and increase parallelism.  If either property is {\em Exclusive} then the semantics
-of the programming model mandate that we detect a conflict and serialize the task.
-
-Another special case occurs if there is an {\tt Atomic} writer and either {\tt Simultaneous}
-or {\tt Relaxed} reader.  In this case it is acceptable for the tasks to run in
-parallel as it is valid for the reader to see updates form the writer, and the
-reader will not violate the atomic access semantics of the writer.
 
@@
 \subsubsection{Instance Management}
-\label{subsec:instmang}
 

### 2011-11-04 doc/pldi2012/highlevel.tex
Removing the word conflict from the high level runtime description

@@
 for each of the regions.  The runtime uses this information to perform
-conflict detection with each of the previous tasks that have been registered
-in the same parent task.  Conflict detection only needs to be performed at
+dependence detection with each of the previous tasks that have been registered
+in the same parent task.  Dependence detection only needs to be performed at
 the scope of an enclosing task because the semantics of the Legion programming
@@
 regions that the parent task uses.  This implies that if a child task were
-to have a conflict with task at a higher level of the task tree, then the
-parent task would also have had a conflict.
+to depend on a task at a higher level of the task tree, then the
+parent task would also have had a dependence.
 
-Lemma: If a task {\tt t} with ancestor task {\tt p} conflicts with task {\tt t'}
-a sibling task of {\tt p}, then tasks {\tt p} and {\tt t'} must conflict.
+Lemma: If a task {\tt t} with ancestor task {\tt p} depends on task {\tt t'}
+a sibling task of {\tt p}, then task {\tt p} must depend on task {\tt t'}.
 
@@
 the the components to dependence analysis in sections \ref{subsec:regiontree},
-\ref{subsec:cohconf}, and \ref{absinsts}.  We then describe the algorithm for performing 
+\ref{subsec:cohdep}, and \ref{absinsts}.  We then describe the algorithm for performing 
 dependence analysis on this data structures in section \ref{subsec:depdetect}.
 
-\subsection{Region Trees}
+\subsubsection{Region Trees}
 \label{subsec:regiontree}
@@
 
-\subsection{Coherence Conflicts}
-\label{subsec:cohconf}
-To detect conflicts between any two tasks using a logical region, we leverage
+\subsubsection{Dependencies From Coherence Properties}
+\label{subsec:cohdep}
+To detect dependencies between any two tasks using a logical region, we leverage
 the information contained in the access and cohrence properties provided by the programmer
-for how the region is to be used.  Conflict checks are performed pair-wise between regions
-to determine whether two tasks conflict with each other.  When checking for
-conflicts we first examine the access property.  If both tasks are reads, then 
-there is never a conflict.  If at least one of them is a write we use table
-\ref{tab:conflicts} to determine whether there is a conflict.
+for how the region is to be used.  Dependency checks are performed pair-wise between regions
+to determine whether one task depends on the other.  When checking for
+dependencies we first examine the access property.  If both tasks are reads, then 
+there is never a dependence.  If at least one of them is a write we use table
+\ref{tab:depsdetect} to determine whether there is a dependence.
 
@@
              & Exclusive & Atomic   & Simultaneous & Relaxed \\ \hline 
-Exclusive    & Conflict  & Conflict & Conflict     & Conflict \\ \hline
-Atomic       & Conflict  & Same Instance & Conflict & Conflict \\ \hline
-Simultaneous & Conflict  & Conflict & Same Instance & None \\ \hline
-Relaxed      & Conflict  & Conflict & None & None \\ \hline
+Exclusive    & Dependence & Dependence & Dependence & Dependence \\ \hline
+Atomic       & Dependence & Same Instance & Contingent & Contingent \\ \hline
+Simultaneous & Dependence & Contingent & Same Instance & None \\ \hline
+Relaxed      & Dependence & Contingent & None & None \\ \hline
 \end{tabular}
-\caption{Conflict detection cases}
+\caption{Dependence detection cases when at least one access is a write.\label{tab:depsdetect}}
 \end{table}
 \end{center}
+In this table, {\em Dependence} indicates that there will always be a dependence between
+regions while {\em None} indicates that there will never be a dependence regardless
+of mapping decisions.  In the case of {\em Same Instance} there is a dependence contingent
+upon a mapping decision.  If both tasks map to the same physical instance (described in 
+more detail in section \ref{subsec:instmang}) then there is no dependence, which if
+they map to different instances there is a dependence as the system will have to generate
+a copy between the two tasks running.
 
-\subsection{Abstract Instances}
+For the case of {\tt Contingent} there are two special cases that we handle.
+One special case is Write-After-Read (WAR) dependencies.  For the WAR case, if neither coherence property
+is {\em Exclusive} then we can ensure that different physical instances
+are used to avoid an anti-dependence and increase parallelism.  If either property is {\em Exclusive} then the semantics
+of the programming model mandate that we detect a dependence and serialize the task.
+
+Another special case occurs if there is an {\tt Atomic} writer and either {\tt Simultaneous}
+or {\tt Relaxed} reader.  In this case it is acceptable for the tasks to run in
+parallel as it is valid for the reader to see updates form the writer, and the
+reader will not violate the atomic access semantics of the writer.
+
+\subsubsection{Abstract Instances}
 \label{subsec:absinsts}
+For the cases where tasks do not depend, they are allowed to create different physical
+instances of the same logical region.  The decision to do this however does not
+happen until the mapping phase of a task execution.  However, we need to have a place
+holder that is capable of representing all the possible valid physical instances
+from which to choose even if those instances do not exist because the prior
+tasks have not been mapped.  We call these place holders {\tt abstract instances}.
 
-\subsection{Dependence Detection}
+Abstract instances represent the set of possible valid physical instances which a
+task will be capable of selecting from or adding to when it is mapped.  The value
+of abstract instances is that they allow us to remember which physical instances
+are valuable while continuing to perform dependence detection on newly registered
+tasks.  In the analogy to out-of-order processors abstract instances allow us
+to perform the same operation as register-renaming of instructions to avoid
+anti-dependences and to remember actual dependences even though our dependence detection
+analysis may run ahead of the actual execution of tasks.
+
+\subsubsection{Dependence Detection}
 \label{subsec:depdetect}
 When a task is registered, the runtime
-will perform conflict detection for every region that the task requests with
-prior tasks that have been registered in the same parent task context.  Referring
-back to our analogy of the high level runtime as an out of order processor, this
-corresponds to the wake-up logic that detects when an instruction is ready to be
-executed.  Continuing this analogy, the regions that a task requires are annalogous
-to the registers listed in an instruction.  However, this poses a difficulty for the
-high level runtime as two logical regions are not necessarily independent.  In
-order to do the conflict detection we rely on the region tree data structure to be
-able to detect conflicts between tasks which require different logical regions.
+will perform dependence detection for every region that the task requests with
+prior tasks that have been registered in the same parent task context.  Since
+tasks are registered in program order we will always perform dependence detection
+in program order which guarantees correctness even if tasks are executed out-of-order.  
+Performing dependence detection in order is also annalogous to how an out of order
+processor works.  Out-of-order processors load instructions in order and only after
+analyzing their instructions and register dependencies do they execute instructions
+out of order.  The regions that a task requires are annalogous
+to the registers listed in an instruction and the goal of the high level
+runtime is to detect the same dependences and use them to enforce scheudling constraints
+upon a task.  However, there is a catch to this analogy: in the Legion programming
+model two different logical regions are not necessarily independent from each other.  In
+order to do the dependence detection we rely on the region tree data structure to be
+able to detect dependences between tasks which require different logical regions.
 
@@
 down the region tree from the node corresponding to the root region to the node
-for the requested region.  At each node, we determine if any potential conflicts exist.
+for the requested region.  At each node, we determine if any potential dependences exist.
 
-For region nodes we first check to see if any tasks currently using this region
-conflict based on the coherence proprties the two task require.  We describe coherence
-conflicts in more detail in section \ref{subsec:cohconf}.  If a conflict is discovered
+For region nodes we first check to see if the required region has any dependences on
+tasks currently using the logical region by performing the check on coherence properties
+described in section \ref{subsec:cohdep}.  If a conflict is discovered
 we record that the task must wait on the previous task before executing.  We also check
@@
 we must {\em close} the partitions to detect any possible conflicts
-
-\subsection{Correctness}
 
@@
 \subsubsection{Instance Management}
+\label{subsec:instmang}
 

### 2011-11-04 doc/pldi2012/mapping_fig.tex
very rough draft of type system section - dummy figure in place for mapping description

\begin{figure}[t]
  \centering
  \subfigure[mapping of $dc_0$ task]{
    \label{sfig:mapping_fig:dc}
    \begin{tikzpicture}
      \partitiontree
    \end{tikzpicture}
  }
  \subfigure[mapping of $volt_0$ task]{
    \label{sfig:mapping_fig:volt}
    \begin{tikzpicture}
      \partitiontree
    \end{tikzpicture}
  }
  \subfigure[mapping completed]{
    \label{sfig:mapping_fig:done}
    \begin{tikzpicture}
      \partitiontree
    \end{tikzpicture}
  }
  \label{fig:mapping_fig}
  \caption{Mapping of tasks in region tree -- TODO: draw in little boxes and stuff}
\end{figure}

### 2011-11-04 doc/pldi2012/type_system.tex
very rough draft of type system section - dummy figure in place for mapping description

\section{Type System}

The key to both performance and correctness in the Legion system is the
accuracy of the region usage declarations.  The runtime will place data and
schedule tasks based on these declarations, and has no way to recover if a 
task attempts to access data that lies outside the mapped regions or attempts
to write to a region that the runtime expected would only be read.

The guarantee that these region usage declarations will be adhered to is
provided by Legion's type system which is able to statically verify that a 
task will stay within its declared boundaries.  The Legion type system is
flow-sensitive, and uses judgements of the form:

\begin{center}
$\Gamma, \Phi_i, C_i \vdash e : T, \Phi_o, C_o$
\end{center}

In addition to the mapping of variables to types (i.e. $\Gamma$), the
environment includes a set of \emph{privileges} $\Phi_i$ that are possessed
and a set of \emph{constraints} $C_i$ that are known to hold before the 
expression $e$ is evaluated.  The judgement says that $e$ has some type $T$,
but also tells us that the privileges possessed after the
evaluation are a potentially-different $\Phi_o$ and that the new set of 
constraints is $C_o$.

Table~\ref{tbl:priv_const} shows the form of the individual privileges and
constraints that make up $\Phi$ and $C$.  Privileges represent the ability to
perform some operation on a region (or, as we will see, on any of its
subregions).  Constraints are used to capture the subregion relationships that
result from partitioning ($a \prec b$ if $a$ was a result of a partitioning of
$b$), the more general anywhere-in-the-subregion-tree relationship ($a \le b$ if $a$ equals $b$ or is either a direct or indirect subregion), and equality (or
inequality) of integer expressions.  More complicated constraints can be formed
through conjunctions, but neither negation nor disjunctions are permitted.  For
an example of how these work, here is the typing rule for reading an element
from a region:

\begin{center}
{\small
\begin{math}
\infrule{
\begin{array}{lc}
  \multicolumn{2}{c}{\Gamma, \Phi_1, C_1 \vdash p : T@r_1, \Phi_2, C_2} \\
  C_2 \models r_1 \le r_2 & readable(r_2) \in \Phi_2
%  \envsub{1}{1} e_1 : \rtripsub{T_1}{2}{2} \\
%  \envsub{2}{2} e_2 : \rtripsub{T_2}{3}{3} 
\end{array}
}
{
  \Gamma, \Phi_1, C_1 \vdash read(r_2, p) : T, \Phi_2, C_2
%  \envsub{1}{1} e_1; e_2 : \rtripsub{T_2}{3}{3}
}
\end{math} 
}
\end{center}

This rule tells us that in order to read from a pointer, we must be able to
provide a region ($r_2$ in this case) for which we possess read privilege that 
is known to either be, or contain, the region to which the pointer is
constrained.  Similar rules exist for the other four region access operations.

The Legion type system also has to be able to express the self-referential
nature of region relationships.  It does this through existential quantification
over regions.  The quantifiers captures both types and constraints, but does
not capture privileges.  Region relations have the following form in the
type system:

\begin{center}
$RR = \exists r_1, \ldots, r_n.\left(T, \emptyset, C\right)$
\end{center}

Explicit pack and unpack operations are used in the type system
to add and remove the existential quantifiers, shown here:

\begin{center}
\begin{math}
\begin{array}{c}
\infrule{
\begin{array}{l}
RR = \exists r_1, \ldots, r_n.\left(T_1, \emptyset, C_1\right) \\
\Gamma, \Phi_2, C_2 \vdash e : \left(RR, \Phi_3, C_3\right) \\
r'_1, \ldots, r'_n \not\in \mathit{RegionsOf}\left(\Gamma, \Phi_3, C_3\right)
\end{array}
}{
\begin{array}{l@{}l}
\Gamma, \Phi_2, C_2 \vdash unpackrr{ }e : ( & [r'_1/r_1,\ldots,r'_n/r_n]T_1, \Phi_3, \\
& C_3 \wedge [r'_1/r_1,\ldots,r'_n/r_n]C_1)
\end{array}
}
\\
\\
\infrule{
\begin{array}{l}
RR = \exists r_1, \ldots, r_n.\left(T_1, \emptyset, C_1\right) \\
\Gamma, \Phi_2, C_2 \vdash e : \left([r'_1/r_1,\ldots,r'_n/r_n]T_1, \Phi_3, C_3\right) \\
C_3 \models [r'_1/r_1,\ldots,r'_n/r_n]C_1
\end{array}
}{
\begin{array}{l}
\Gamma, \Phi_2, C_2 \vdash packrr{ }RR\mbox{ }e : \left(RR, \Phi_3, C_3\right) \\
\end{array}
}
\end{array}
\end{math}
\end{center}

Any expression whose type is a region relationship can be unpacked.  Doing so
introduces a fresh region variable for each region that was bound in the 
quantification.  The fresh region variables are substitued into the region
relaionship's type, and a corresponding substituion is performed on the
constraints that were captured by the region relationship.  These substituted
constraints are added to the current function's (flow-sensitive) constraints.

Packing an expression into a region relationship works in reverse.  If the type
of an expression can be unified with the region relationship's type and the
corresponding constraints of the region relationship can be shown to hold, the
expression can be packed into the region relationship and used interchangably
with any other instance of that region relationship.

Although the type system uses explicit packing and unpacking operations, they 
are implicit in the Legion application code.  Unpacking operations are 
automatically inserted whenever a variable comes into scope.  Packing operations
are attemped whenever a variable is written to memory or passed as an argument
to a function.  Additionally, the simultaneous assignment operation essentially
performs a pack for the result of the assignments and will also perform an
unpack if one ore more fields of the structure are left unchanged.

\begin{table}
\centering
{\small
\begin{math}
\begin{array}{cc}
\begin{tabular}{ccc}
$\phi$ & ::= & $readable(r)$ \\
  &$\mid$&$writeable(r)$ \\
  &$\mid$&$reduceable(r,f)$ \\
  &$\mid$&$allocable(r)$ \\
  &$\mid$&$freeable(r)$ \\
\\
$\Phi$ & ::= & $\{ \phi_1, \ldots, \phi_n \}$
\end{tabular}
&
\begin{array}{ccc}
C & ::= & (r_1 \prec r_2) \\
  &\mid&(r_1 \le r_2) \\
  &\mid&(r_1 * r_2) \\
  &\mid&(i_1 = i_2) \\
  &\mid&(i_1 \ne i_2) \\
  &\mid&(C_1 \wedge C_2) \\
\end{array}
\end{array}
\end{math}
}
\label{tbl:priv_const}
\caption{Privileges and Constraints}
\end{table}

Region access privileges cannot be stored or packed.  Instead, they follow the
function call tree, and it is this property that allows the runtime to make
its scheduling decisions in a distributed way.  The original source of any
privilege for a given region is the \emph{newrr} operation that created that
region, but the way in which privileges can be provided to a subtask, or
returned to a parent task, is described in the function application rule:

\begin{center}
\begin{math}
\begin{array}{c}
\infrule
{
\begin{array}{lc}
  \Gamma, \Phi_1, C_1 \vdash e_1 : \left(\left(T_i, \Phi_i, C_i\right) \rightarrow \exists r_1, \ldots, r_n . \left(T_o, \Phi_o, C_o\right), \Phi_2, C_2\right) \\
  \Gamma, \Phi_2, C_2 \vdash e_2 : \left(T_i, \Phi_3, C_3\right) \\
  r'_1, \ldots, r'_n \not\in \mathit{RegionsOf}(\Gamma, \Phi_3, C_3) \hspace{1cm} 
  \Phi_i \subset \Phi_3 \hspace{1cm}
  C_3 \models C_i
\end{array}
}
{
\begin{array}{l@{}l}
%  \envsub{1}{1} e_1 e_2 : \rtriple{& \regionexpand T_o}{\left(\Phi_3 \cup \regionexpand \Phi_o\right)}{\\ & \left(C_3 \wedge \regionexpand C_o \wedge \bigwedge_{\substack{{\footnotesize r_f \in \{ r | \regionexpand C_o \models fresh(r) \} } \\ {\tiny r_o \in \mathit{RegionsOf}(\Gamma, \Phi_3, C_3)}}} r_f * r_o\right)}
  \Gamma, \Phi_1, C_1 \vdash e_1 e_2 : \Big( & [r'_1/r_1,\ldots,r'_n/r_n]T_o, \\
 & \left(\Phi_3 \cup [r'_1/r_1,\ldots,r'_n/r_n] \Phi_o\right) , \\
 & \left(C_3 \wedge [r'_1/r_1,\ldots,r'_n/r_n] C_o\right) \Big)
\end{array}
}
\end{array}
\end{math}
\end{center}

In this rule, expression $e_1$ has the type of a Legion function.  In addition
to the function's input and output types, it specifies the privileges and
constraints that must exist at the point where the function is called.  It also
describes the privileges and constraints that exist as post-conditions of the
function invocation.  The output type, privileges, and constraints can be
captured in an existential quantification, allowing a function to return one
or more regions that were previously unknown to the caller.

### 2011-11-04 doc/pldi2012/highlevel.tex

@@
 in program order which guarantees correctness even if tasks are executed out-of-order.  
-Performing dependence detection in order is also annalogous to how an out of order
+Performing dependence detection in program order is also annalogous to how an out of order
 processor works.  Out-of-order processors load instructions in order and only after
@@
 model two different logical regions are not necessarily independent from each other.  In
-order to do the dependence detection we rely on the region tree data structure to be
+order to do the dependence detection we rely on the forest of region tree data structures to be
 able to detect dependences between tasks which require different logical regions.
@@
 for the requested region.  At each node, we determine if any potential dependences exist.
+Determining whether dependencies exist involves checking information both about the current
+node being visited as well as whether there are any potential dependences farther down
+in the region tree.  We now describe the decision procedure performed at both region
+nodes and partition nodes in the region trees.
 
-For region nodes we first check to see if the required region has any dependences on
-tasks currently using the logical region by performing the check on coherence properties
-described in section \ref{subsec:cohdep}.  If a conflict is discovered
-we record that the task must wait on the previous task before executing.  We also check
-to see if any of the partitions of the region have been marked open.  If they have
-we must {\em close} the partitions to detect any possible conflicts
+For region nodes, we begin by checking if there are any tasks actively using the logical
+region which we depend.  We determine this using the coherence dependence analysis described 
+in section \ref{cohdep}.  For each dependence that is found, we record that the previous
+task must have completed prior to the start of the task we are registering.  After performing
+dependence analysis we then check if there is an open partition which may contain additional
+dependencies.  If the open partition is the partition we intend to traverse then we
+simply continue the traversal.  However, if the partition we want is not open then we must
+first {\em close} up any open partition and then open the partition we are going to
+traverse.  The act of closing a partition involves traversing the region tree and recording
+all active tasks as dependences.  The reason for this is that we have no disjointness
+information about tasks that use different partitions and therefore to be safe must register
+all task touching subregions as dependences.  After closing any open partitions, we
+can then open the partition on our path and continue the traversal.
+
+When we reach our target logical region we again perform the coherence dependence analysis.
+We also close up any open partitions as they might contain tasks which could contain
+a dependence.  If our task was dependent on all of the previous active tasks in a region
+we say that our task was a {\em dominator} of all the previous active tasks.  We can
+then clear the list of active tasks and replace it with the current task.  This is safe since
+any later tasks that depend on the current task will have a transitive dependence 
+through the dominator task.  If our task is not a dominator task, then we simply append
+it to the list of active tasks making it possible for future task registrations to detect
+it as a dependence.
+
+When traversing partition nodes we perform a similar, but less complicated analysis.  There
+are two cases to consider for partition nodes: disjoint partitions and aliased partitions.
+In the case of disjoint partitions, we simply open the subregion that we intend to traverse
+and then traverse it.  This is correct because we know by definition that any two logical
+subregions in a disjoint partition are disjoint and therefore a dependence can never exits
+between them.
+
+For aliased partitions we have to perform an additional dependence analysis.  At every
+aliased partition node we maintain a list of tasks active in all of the subregions.  We
+perform a coherence dependence analysis between every pair of tasks regardless of the subregion
+they are accessing to see if a dependence could exist.  We do this regardless of the subregion
+because we have no disjointness information in an aliased partition.  If there are any
+potential dependences, we close up all the subregions and then open the subregion required.  If
+no dependence was detected then we open the subregion we need, append the task to the list
+of active tasks, and continue the traversal.  Allowing multiple open subregions in an aliased
+partition is safe because we performed an analysis between all active tasks regardless of
+whether the particular subregions were aliased or not.
+
+\subsubsection{Putting It All Together: Circuit Example}
+\label{subsec:cirdependence}
+
+\subsection{Implicit Dataflow Construction}
+\label{subsec:dataflow}
+In addition to computing the dependences between tasks, we also have to compute where
+a task is permitted to acquire its data from when it goes to execute.
+It is insufficient to simply record the producer task as there
+may be multiple active tasks for a logical region, all of which may produce different physical
+instances.  In addition, it is likely that the producer tasks have not been mapped
+yet which makes it impossible to know the set of physical instances from which to pull data.
+
+To solve this problem we create the notion of an {\em abstract instance}.  An abstract instance
+is a place holder for the set of physical instances from which a task can choose to pull
+data when it creates its own physical instances before executing.  For each region in
+a task we record the source and destination abstract instances that correspond to the physical
+instances from which a task will read its input data and then place its output data.
+When the task is mapped as described in section \ref{subsec:mapping} we will update these
+abstract instances with the actual physical instances chosen.
+
+To determine the source and destination abstract instances of a region, we keep track of 
+the previous abstract instances that were observed on the way to our target region node
+when performing dependence analysis.  These abstract instances describe the set of physical
+instances from which valid data for our task can be chosen when it is mapped.  If a region
+subtree must be closed then all the abstract instances in that subtree are marked closed
+so no additional tasks can register them as sources.  We can also chose to create a new
+abstract instance to replace a currently valid abstract instance to avoid a WAR conflict.
+By creating an abstract instance we will force the mapper to choose a different physical
+instance from the existing ones when mapping is performed.
+
+Abstract instances correspond to a dynamic construction of the dataflow graph.  Each abstract
+instance is a node in the dataflow graph corresponding to a specific version of a logical 
+region and its instantiations as physical regions.  Tasks are then edges between abstract
+instances indicating operations on different versions of data.  By computing this information
+during conflict detection, we can record it before tasks begin executing out of order allowing
+us to guarantee program correctness in the presence of out of order task execution.  Continuing
+the analogy with out-of-order processors, this implicit construction of the dataflow graph
+is annalogous to register renaming which is done in order, but allows the processor to remove
+WAR dependences and execute instructions in parallel that may have had false dependences.
 
 \subsection{Mapping}
+\label{subsec:mapping}
 

### 2011-11-04 doc/pldi2012/highlevel.tex
More text for the high level runtime section

@@
 
-\subsubsection{Putting It All Together: Circuit Example}
-\label{subsec:cirdependence}
-
-\subsection{Implicit Dataflow Construction}
+\subsubsection{Implicit Dataflow Construction}
 \label{subsec:dataflow}
@@
 so no additional tasks can register them as sources.  We can also chose to create a new
-abstract instance to replace a currently valid abstract instance to avoid a WAR conflict.
+abstract instance to replace a currently valid abstract instance to avoid a WAR dependence.
 By creating an abstract instance we will force the mapper to choose a different physical
@@
 instances indicating operations on different versions of data.  By computing this information
-during conflict detection, we can record it before tasks begin executing out of order allowing
+during dependence detection, we can record it before tasks begin executing out of order allowing
 us to guarantee program correctness in the presence of out of order task execution.  Continuing
@@
 
+\subsubsection{Putting It All Together: Circuit Example}
+\label{subsec:cirdependence}
+To illustrate concretely how dependence analysis works, we describe the steps for
+performing dependence analysis for running circuit example.  When the {\tt simulate\_circuit}
+task begins executing there are two region trees rooted at the logical
+regions for {\tt all\_nodes} and {\tt all\_wires}.  For conciseness we will only
+show the region tree for {\tt all\_nodes}.  At the start of the task, there exists
+a single abstract instance at the root node representing the valid set of physical
+instances from which sub tasks will be allowed to pull data.  The state of the
+system at this point can be observed in figure \ref{fig:circdep1}.
+
+When each of the {\tt compute\_new\_current} tasks are issued they must perform dependence
+analysis for two node regions.  In the {\tt private\_node} case they will first open
+up the {\tt p\_nodes\_pvs} partition and then open up the {\tt p\_pvt\_nodes} sub partition
+for each of their subregions.  Since all of the private regions use a different subregion
+of a disjoint partition for the private node accesses, no dependence is detected.  Similarly,
+when all the dependence analysis is performed for the {\tt p\_ghost\_nodes} subregions,
+each of the necessary partitions are opened.  There is no dependence with any of the private
+nodes since all the ghost nodes traverse a different sub region of the disjoint partition
+{\tt p\_nodes\_pvs}.  There is also no dependence between any tasks even though the
+{\tt p\_ghost\_nodes} partition is aliased because all the ghost accesses are read-only.
+The state of the region tree after dependence analysis has been performed for the
+{\tt compute\_new\_current} nodes can be seen in figure \ref{fig:circdep2}.
+
+In the second phase of the computation, all of the {\tt distribute\_charge} tasks
+will be issued.  For these tasks they will first analyze traverse the region tree
+for their subregions containing their private nodes.  The path to these regions are
+already open, but when the dependence analysis arrives at the specific private node subregion it will
+discover a dependence between the task and the {\tt compute\_new\_currents} task that operated
+on that region previously.  This dependence will be recorded and the 
+{\tt distribute\_charge} task will be made the new active task as it dominated the
+previous task.  For the ghost node regions, the dependence analysis will discover that the
+there is a dependence between each of the {\tt distribute\_charge} tasks and all of the
+{\tt compute\_new\_current} tasks because the {\tt distribute\_charge} tasks are
+performing writes and the partition is aliased.  However, the runtime will detect this as
+anti-dependences which are resolved by forcing the {\tt distribute\_charge} tasks to
+create new valid abstract instances rather than use the same abstract instances as
+each of the {\tt compute\_new\_current} tasks.  There are no dependences detected
+between each of the {\tt distribute\_charge} tasks on the ghost regions because
+the {\tt Simultaneous} coherence property permits multiple tasks to be using the same
+partition at the same time even if the subregions they are accessing are aliased.  The
+state of the region tree after dependence analysis for {\tt distribute\_charge} can
+be seen in figure \ref{fig:circdep3}.
+
+For the final phase of the computation, all of the {\tt update\_voltage} tasks
+are issued requiring their private node region and their shared node region.  Again
+the private node paths are already open, but a WAR dependence is detected which
+requires a new abstract instance to be created to avoid a dependence.  The
+shared nodes present a more interesting case.  The first shared node to be registered
+will detect a dependence with every prior task since it must open the shared partition
+instead of the ghost partition.  A new abstract instance is created at the 
+{\tt all\_shared} node and copies are recorded for each valid instance in the ghost partition
+back to the new {\tt all\_shared} abstract instance.  After this is complete, the
+ghost partition can be opened.  Each corresponding {\tt update\_voltage} task will record
+a dependence on the copy operation to close ghost partition as it traverses the {\tt all\_shared}
+node ensuring that there is no requirement on the ordering of {\tt update\_voltage}
+tasks.  The state of the region tree after this phase can be seen in figure \ref{fig:circdep4}.
+
 \subsection{Mapping}
 \label{subsec:mapping}
+After tasks have undergone dependence analysis, the tasks are placed on either a ready
+queue to be mapped or waiting queue.  A task is placed on the ready queue if all the tasks with
+which it registered a dependence have mapped.  If any of the task's dependences have
+not mapped then the task is placed on the waiting queue.  The task also registers
+itself with the task it depends on so that when the dependent task is mapped it will notify
+the task on the waiting queue. Tasks are mapped based on the partial ordering created by 
+their data dependences.  By mapping tasks based on data dependences we guarantee the 
+semantics of the programming model, while still allowing tasks to execute out of order.
 
-\subsubsection{Placing Tasks}
+When the scheduler is invoked, it will pull a series of tasks off the ready queue, map them,
+and then issue them to be executed by the low level runtime.  The process of mapping a task
+consists of several different queries made by the runtime to an application level object
+called a {\em mapper}.  A mapper is an object which has control over how both tasks and
+data can be placed in the machine.  This idea is a dynamic instantiation of the idea
+initially presented in \cite{Fatahalian06} of creating mapping files for specifying how
+tasks and data were placed in a memory hierarchy.  Rather than rely on a static mapping
+we opted for a dynamic interface that could make decisions both based on application specific
+input data as well as the machine specifics both at start-up and throughout the execution
+of an application.  The interface for a mapper is described in section \ref{subsec:mapinter}.
 
-\subsubsection{Instance Management}
-\label{subsec:instmang}
+\subsubsection{Mapper Interface}
+\label{subsec:mapinter}
+The interface for a mapper is designed to give the programmer control over how data and tasks
+are placed in the machine.  Every mapper must support the following function calls:
+
+\begin{itemize}
+\item {\tt rank\_initial\_region\_locations} - When a new logical region is created, the runtime
+will ask the mapper to give a ranking of physical memories of where to place the inital instance.
+
+\item {\tt rank\_initial\_partition\_locations} - When a new partition is created, the mapper
+must provide a ranking of memory locations for each of the new sub partitions.
+
+\item {\tt select\_initial\_processor} - Given a task to be mapped, select a processor to
+send this task to for mapping.
+
+\item {\tt target\_task\_steal} - select a processor to attempt task stealing.  This is explained
+in more detail in section \ref{subsec:steal}.
+
+\item {\tt permit\_task\_steal} - also covered in section \ref{subsec:steal}
+
+\item {\tt map\_task\_region} - Given a specific task and the logical region to be mapped, select
+a source physical instance for the data, and provide a ranking of memories in which to create
+or reuse an existing physical instance.
+
+\item {\tt rank\_copy\_targets} - Given a task and a set of valid instances, generate a set of
+rankings for new memories in which to place new phsyical instances.  This operation is used to
+manage the copy operations performed when closing a subtree as they allow the programmer to
+eagerly migrate regions through the memory hierarchy as copy ups are occuring.
+
+\item {\tt select\_copy\_source} - Given a set of currently valid source instances and destination
+physical instance, select the instance from which a copy should be made.
+\end{itemize}
+
+These eight task calls give the programmer complete control over where all tasks are executed, and
+where all data is placed within the memory hierarchy.  While this interface does give the programmer
+complete control over data and task movement, it will never allow the programmer to affect the
+correctness of the program.  The highlevel runtime still manages all the task dependences and the
+ordering in which tasks are executed and copies are performed.  Therefore, regardless of choices
+made by any mapper, the resulting program will always be executed correctly.
+
+In order to make it easy for programmers to write programs, we provide an initial implementation 
+of the mapper interface that uses simple heuristics to place task and data.  This makes it easy
+for the programmer to get an initial program running, while allowing them to write their own
+mapper to tune for application or machine specific scenarios at a later time.
+
+Lastly, the runtime supports multiple mappers in use in the same application.  For each task 
+call the programmer can choose a mapper to be used for the runtime calls made inside that
+task.  This allows for the creation of library code which provides their own mappers indpendent
+from any application level mappers that may be created.
+
+\subsubsection{Task Stealing for Load Balance}
+\label{subsec:steal}
+In order to achieve load balancing in an application the runtime supports an interface
+for task stealing.  Our task stealing algorithm is loosely based on the stealing algorithm
+used by Cilk\ref{CILK95} with guidance from the mapper about how stealing should be
+performed.  Tasks are initially placed on the ready queue on the processor in which they
+were registered.  However, a task can be moved to a different processor either by a mapper
+explicitly choosing to relocate it, or by a steal request from another processor.
+
+When a processor notices that it has no work on its ready queue, it will inquire into each
+mapper asking for a target processor to attempt a steal via the {\tt target\_task\_steal}
+call in the mapper interface.  A mapper can either respond with
+a target processor, or indicate that no steal should be attempted.  The runtime will then
+send steal requests to each processor on behalf of the specific mapper.
+
+When a steal request is received at a processor it will determine the mapper(s) that generated
+the list of steal requests.  For each mapper that issued a steal request, the runtime will
+give to the local instance of that mapper a list of the tasks in the ready queue that that
+specific mapper is responsible for mapping.  The mapper must then determine the set which
+it is going to permit to be stolen if any.  Any tasks that the mapper marks as stolen are
+sent back to the stealing processor.  By only allowing mappers to make steal requests about
+the tasks they own, we are giving the mappers complete control over where tasks are mapped.
+In addition, we are also giving the mappers control over how steal requests are made for
+the case where the mapper has better knowledge about where additional work is than the 
+runtime.  Finally, note that this mapping interface is sufficiently powerful to allow a user to
+disable all stealing if desired, and to map tasks to explicit processors with no interference
+from the runtime.
 

### 2011-11-04 doc/pldi2012/highlevel.tex
More text for the high level runtime section

@@
 \subsection{Execution}
+\label{subsec:execution}
+Once task mapping is complete the high level runtime performs the necessary operations
+to be issue the task to the low level runtime.  It first creates any new physical instances
+for the regions which the task requires.  It then issues the necessary copies required
+for the task to execute contignent on the tasks producing the source physical instances
+have finished.  Then it issues the task itself to the low level processor contingent
+on all the copies for the task haveing finished as well as the the set of tasks
+on which the issuing task dependends having finished.  These dependences are all encoded
+using event dependencies in the low level runtime.  Once all the necessary events
+have triggered the task will be run.
+
+Once a task starts running, the runtime is asked to create the region accessors for
+the task.  The runtime investigates whether or not fast accessors can be created for
+each of the regions that the task requires.  Fast accessors are described in more
+detail in section \ref{sec:fastaccess}.  If fast accessors can be created for every
+region in a task, then the runtime chooses to execute a task instantiated with fast
+region accessor methods.  However, if any region can not be specialized into a fast
+accessor the runtime defaults to a task instantiation that relies on {\tt get} and
+{\tt put} operations to access regions.
+
+The runtime will also return to the executing task a unique identifier describing the
+context in which the task is executing.  All calls into the runtime require the task
+to pass this context identifier which is what permits the runtime to know the context
+for incoming calls.
 
 \subsection{Task Completion}
+\label{subsec:cleanup}

### 2011-11-04 runtime/highlevel.cc
Adding an assert to the high level runtime

@@ -3940,6 +3940,9 @@ namespace RegionRuntime {
         {
           // ask the mapper for where to place the task
           Processor target = mapper_objects[task->map_id]->select_initial_processor(task);
+#ifdef DEBUG_HIGH_LEVEL
+          assert(target.exists());
+#endif
           task->chosen = true;
           if (target == local_proc)
           {

### 2011-11-04 doc/pldi2012/highlevel.tex
Final bit of text for for the high level runtime tonight

@@
 \label{subsec:cleanup}
+We define a task to have {\em completed} once the actual code for the task has completed.
+We also define a task to have {\tt terminated} once all of the operations for a task
+have finished.  We can only consider a task to be terminated once all of the regions that
+the task accessed reflect all the changes made by a task and its child tasks.  However,
+even after a task has completed it is possible for child tasks of that task
+not to have been executed or even mapped.  We describe how we handle this scenario in
+section \ref{subsec:restore}.  We also describe how physical instances are garbage
+collected in seciont \ref{subsec:garbage}.
+
+The runtime needs to know the mappings for
+the child tasks as it must issue restoring copy operations so that the set of regions
+that the task operated on are up to date with any data generated by child tasks.
+
+\subsubsection{Restoring Copy Operations}
+\label{subsec:restore}
+An underlying assumption of the mapping process is that the resulting regions that
+a task access will represent the data produced by that task.  However, this may not
+be true if a task has child tasks which also modify subregions of the parent task.
+In order to handle this issue the runtime must compute the necessary copy operations
+from child task physical instances back to the parent task's physical instances.  To
+compute these copy operations, the runtime first will wait until all the child tasks
+have been mapped.  The runtime then issues a close operation to the each of the parent
+task's region trees.  These close operations record all of the necessary copies to
+restore the parent task's regions as well as the child tasks on which these copies
+are contingent.  We call these copy operations at the end of a task {\em restoring copies}.
+The runtime then issues these copies with the required dependences
+and marks the task terminated once the restoring copy operations are complete.
+
+\subsubsection{Garbage Collecting Physical Instances}
+\label{subsec:garbage}
+In addition to cleaning up task execution, the runtime must also deal with the problem
+of garbage collecting physical instances of regions.  Physical instances must be
+garbage collected since they are not visible to the application and would present
+a correctness problem if deleting instances was an option for the mapper.  
+Garbage collecting physical instances is a challenging problem
+due to both the distributed environment of the machine and the deffered execution
+model.  To solve this problem we employ a two-level reference counting scheme.  We
+reference count both abstract instances and as well as physical instances.
+
+An abstract instance maintains a count of the number of tasks that have referenced 
+that abstract instance, but have not yet been mapped.  An abstract instance also tracks
+whether or not it is open or closed.  An abstract instance is open if it is still
+a valid abstract instance in a region tree.  An abstract instance is marked closed
+once it is no longer a valid abstract instance and can therefore no longer be used
+by any new tasks that are registered.
+
+When a task is mapped it notifies an abstract instance that it has been used and can
+decrement its reference count.  The task mapping also notifies the abstract instance
+of the specific physical instance that was used and whether or not this use invalidates
+previous physical instances of the abstract instance.  The abstract instance maintains
+a reference count for each physical instance of it that has been made as well as
+the set that are currently valid physical instances to be used for mapping.  When a 
+task terminates, it notifies the abstract instances that it used to decrement the
+reference counts for the task's specific physical instances.
+
+There are two conditions under which a physical instance can be garbage collected.  The
+first condition is when a physical instance is not a valid physical instance of
+an abstract instance and its reference count is zero.  This condition occurs once no
+more tasks are using a physical instance and it can longer be mapped by any task.  The
+second condition is when an abstract instance is both closed and has a reference count
+of zero as well as the physical instance's reference count being zero.  This handles the
+case when no more tasks can be set to use an abstract instance, there are no more tasks
+to map to an abstract instance, and there are no current tasks using the physical
+instance.
+
+While reference counting normally doesn't allow for garbage collection of all data structures
+it will always work in this case because both our tasks and our regions are organized as 
+trees which ensures that it is impossible to create cyclic region references in our system.

### 2011-11-04 doc/pldi2012/exec.tex
exec

\section{Execution Model}
\label{sec:exec}

In this section we present the Legion execution model and argue that
it is correct in that it is guaranteed to preserve the sequential
execution semantics of programs.  The correctness argument is
instructive in that the properties that ensure correctness are also
exploited in the implementation of the Legion high-level runtime
system to achieve high performance and highly distributed scheduling
of parallel tasks.

For the purposes of describing and reasoning about the
execution model we consider an extremely restricted core language, so
simple in fact that it is not even Turing complete and is also extremely
inconvenient for writing programs.  The loss in expressive power, however,
is hopefully a gain in clarity and simplicity for illustrating the core ideas in
Legion.    We consider programs written in the following language:

\begin{eqnarray*}
\mbox{def}\ f(r_1,\ldots,r_n): P_1,\ldots,P_m & = & s_1; \ldots s_n; \\[.15in]
P & := & R(r) | W(r) \\[.15in]
s & := & f(r_1,\ldots,r_n) \\
&| & \deref r = x \\
& | & x = \deref r \\
&| & \mbox{\tt newregion}\ r \\
&| & (r_1, r_2)_{p} = \mbox{\tt partition}(r) \\ 
&| & \mbox{\tt close}\ p

\end{eqnarray*}

There are four kinds of names: functions $f$, regions $r$, local
variables $x$, and partitions $p$.  A function takes a number of
regions as arguments and the function definition $f$ specifies the
region permissions $f$ requires; the permission $R(r)$ means $f$ may
read $r$, and $W(r)$ means $f$ may write $r$.

A function body consists of an ordered list of statements.  The statement
$f(r_1,\ldots,r_n)$ calls a function, which is executed for its effects
on its region arguments.  The statement $x = \deref
r$ reads a value from region $r$ and stores it in local variable $x$,
and the statement $\deref r = x$) stores the value of $x$ in region
$r$.  A $\mbox{\tt newregion} r$ statement creates a new region with
name $r$.

The statement $(r_1, r_2)_{p} = \mbox{\tt partition}(r)$
creates two {\em subregions} $r_1$ and $r_$ partitioning the parent
region $r$.  All three region names ($r_1$,$r_2$,and $r$) can be used
after the partition.  Regions $r_1$ and $r_2$ are disjoint but neither
$r_1$ nor $r_2$ is disjoint from $r$.  The partition itself also has a name $p$,
which is used to {\em close} the partition in the $\mbox{\tt close}\ p$ statement 
Intuitively, closing a partition reconciles modifications to the subregions with the {\em parent} region: 
$r$ is updated to incorporate any changes to $r_1$ and $r_2$.  In our source language {\tt close} is not
available to the programmer, but is invoked by the language implementation to reconcile distributed copies
of regions whenever necessary, and it is impossible for the program to observe discrepencies between
a region and its subregions. This toy language, then, is closer to the level of ourintermediate representation,
which is most appropriate for discussing our runtime system.

A program is a list of functions, with the entry point the first function on the list.
We further require that every function except the entry poiint be called in exactly one place
and to have its formal parameters be identical to the actual parameters in its single call.
Region names in {\tt newregion} and partition names in {\tt partition} are chosen to be distinct from
all other region/partition names in the program.  These restrictions allow us to avoid dealing with
alpha renaming and also mean that each runtime call is uniquely named by its function name.

### 2011-11-04 doc/pldi2012/exec.tex
exec

@@
 \begin{eqnarray*}
-\mbox{def}\ f(r_1,\ldots,r_n): P_1,\ldots,P_m & = & s_1; \ldots s_n; \\[.15in]
+F & := & \mbox{def}\ f(r_1,\ldots,r_n): P_1,\ldots,P_m  =   s_1; \ldots s_n; \\[.15in]
 P & := & R(r) | W(r) \\[.15in]
@@
 &| & \mbox{\tt close}\ p 
-
 \end{eqnarray*}
@@
 The statement $(r_1, r_2)_{p} = \mbox{\tt partition}(r)$
-creates two {\em subregions} $r_1$ and $r_$ partitioning the parent
+creates two {\em subregions} $r_1$ and $r_2$ partitioning the parent
 region $r$.  All three region names ($r_1$,$r_2$,and $r$) can be used

### 2011-11-06 runtime/activemsg.h
detailed timing, other stuff

@@ -140,18 +140,21 @@ struct MessageRawArgs<MSGTYPE, MSGID, SHORT_HNDL_PTR, MED_HNDL_PTR, n> { \
 \
   void request_short(gasnet_node_t dest) \
   { \
+    RegionRuntime::LowLevel::DetailedTimer::ScopedPush sp(TIME_SYSTEM);	\
     MACROPROXY(gasnet_AMRequestShort ## n, dest, MSGID, HANDLERARG_VALS_ ## n); \
   } \
 \
   void request_medium(gasnet_node_t dest, \
 		      const void *data, size_t datalen) \
   { \
+    RegionRuntime::LowLevel::DetailedTimer::ScopedPush sp(TIME_SYSTEM); \
     MACROPROXY(gasnet_AMRequestMedium ## n, dest, MSGID, (void *)data, datalen, HANDLERARG_VALS_ ## n); \
   } \
 \
   void request_long(gasnet_node_t dest, \
 		    const void *data, size_t datalen, void *dstptr)	\
   { \
+    RegionRuntime::LowLevel::DetailedTimer::ScopedPush sp(TIME_SYSTEM); \
     MACROPROXY(gasnet_AMRequestLong ## n, dest, MSGID, (void *)data, datalen, dstptr, HANDLERARG_VALS_ ## n); \
   } \
 \
@@ -191,6 +194,7 @@ struct RequestRawArgs<REQTYPE, REQID, RPLTYPE, RPLID, FNPTR, RPL_N, n> { \
 \
   void request_short(gasnet_node_t dest) \
   { \
+    RegionRuntime::LowLevel::DetailedTimer::ScopedPush sp(TIME_SYSTEM); \
     MACROPROXY(gasnet_AMRequestShort ## n, dest, REQID, HANDLERARG_VALS_ ## n ); \
   } \
 \
@@ -221,6 +225,7 @@ template <class RPLTYPE, int RPLID> struct ReplyRawArgs<RPLTYPE, RPLID, n> { \
 \
   void reply_short(gasnet_token_t token) \
   { \
+    RegionRuntime::LowLevel::DetailedTimer::ScopedPush sp(TIME_SYSTEM); \
     MACROPROXY(gasnet_AMReplyShort ## n, token, RPLID, HANDLERARG_VALS_ ## n ); \
   } \
  \

### 2011-11-06 runtime/circuit_gpu.cu

@@ -13,6 +13,8 @@ using namespace RegionRuntime::HighLevel;
 
 #define TOP_LEVEL_TASK_ID   TASK_ID_REGION_MAIN 
 
+#define NO_SYNC_AFTER_KERNELS
+
 enum {
   TASKID_LOAD_CIRCUIT = TASK_ID_AVAILABLE,
   TASKID_CALC_NEW_CURRENTS,
@@ -261,6 +263,7 @@ public:
 					     MappingTagID tag,
 					     std::vector<Memory> &ranking)
   {
+    DetailedTimer::ScopedPush sp(TIME_MAPPER);
     //log_mapper("mapper: ranking initial region locations (%zd,%zd,%d)\n",
     //	       elmt_size, num_elmts, tag);
     //Mapper::rank_initial_region_locations(elmt_size, num_elmts, tag, ranking);
@@ -274,6 +277,7 @@ public:
 						MappingTagID tag,
 						std::vector<std::vector<Memory> > &rankings)
   {
+    DetailedTimer::ScopedPush sp(TIME_MAPPER);
     //    log_mapper("mapper: ranking initial partition locations (%zd,%d,%d)\n",
     //	       elmt_size, num_subregions, tag);
     //Mapper::rank_initial_partition_locations(elmt_size, num_subregions,
@@ -288,6 +292,7 @@ public:
   virtual bool compact_partition(const UntypedPartition &partition, 
 				 MappingTagID tag)
   {
+    DetailedTimer::ScopedPush sp(TIME_MAPPER);
     //    log_mapper("mapper: compact partition? (%d)\n",
     //	   tag);
     //return Mapper::compact_partition(partition, tag);
@@ -298,6 +303,7 @@ public:
 
   virtual Processor select_initial_processor(const Task *task)
   {
+    DetailedTimer::ScopedPush sp(TIME_MAPPER);
     //    log_mapper("mapper: select initial processor (%p)\n", task);
 
     switch(task->task_id) {
@@ -347,6 +353,7 @@ public:
 
   virtual Processor target_task_steal(void)
   {
+    DetailedTimer::ScopedPush sp(TIME_MAPPER);
     //log_mapper("mapper: select target of task steal\n");
     //return Mapper::target_task_steal();
 
@@ -358,6 +365,7 @@ public:
 				 const std::vector<const Task*> &tasks,
 				 std::set<const Task*> &to_steal)
   {
+    DetailedTimer::ScopedPush sp(TIME_MAPPER);
[…]

### 2011-11-06 runtime/fluid.cc

@@ -479,6 +479,7 @@ void main_task(const void *args, size_t arglen,
   struct timespec ts_start, ts_end;
   std::list<Future> futures;
   clock_gettime(CLOCK_MONOTONIC, &ts_start);
+  DetailedTimer::clear_timers();
 
   int cur_buffer = 0;  // buffer we're generating on this pass
   // Run the simulation
@@ -613,6 +614,7 @@ void main_task(const void *args, size_t arglen,
   double sim_time = ((1.0 * (ts_end.tv_sec - ts_start.tv_sec)) +
 		     (1e-9 * (ts_end.tv_nsec - ts_start.tv_nsec)));
   printf("ELAPSED TIME = %7.3f s\n", sim_time);
+  DetailedTimer::report_timers();
 
   log_app.info("all done!");
 

### 2011-11-06 runtime/highlevel.cc

@@ -3453,6 +3453,7 @@ namespace RegionRuntime {
     void HighLevelRuntime::enqueue_tasks(const void * args, size_t arglen, Processor p)
     //--------------------------------------------------------------------------------------------
     {
+      DetailedTimer::ScopedPush sp(TIME_HIGH_LEVEL);
       HighLevelRuntime::get_runtime(p)->process_tasks(args,arglen);
     }
 
@@ -3460,6 +3461,7 @@ namespace RegionRuntime {
     void HighLevelRuntime::steal_request(const void * args, size_t arglen, Processor p)
     //--------------------------------------------------------------------------------------------
     {
+      DetailedTimer::ScopedPush sp(TIME_HIGH_LEVEL);
       HighLevelRuntime::get_runtime(p)->process_steal(args,arglen);
     }
 
@@ -3467,6 +3469,7 @@ namespace RegionRuntime {
     void HighLevelRuntime::children_mapped(const void *result, size_t result_size, Processor p)
     //--------------------------------------------------------------------------------------------
     {
+      DetailedTimer::ScopedPush sp(TIME_HIGH_LEVEL);
       HighLevelRuntime::get_runtime(p)->process_mapped(result, result_size);
     }
 
@@ -3474,6 +3477,7 @@ namespace RegionRuntime {
     void HighLevelRuntime::finish_task(const void * args, size_t arglen, Processor p)
     //--------------------------------------------------------------------------------------------
     {
+      DetailedTimer::ScopedPush sp(TIME_HIGH_LEVEL);
       HighLevelRuntime::get_runtime(p)->process_finish(args, arglen);
     }
     
@@ -3481,6 +3485,7 @@ namespace RegionRuntime {
     void HighLevelRuntime::notify_start(const void * args, size_t arglen, Processor p)
     //--------------------------------------------------------------------------------------------
     {
+      DetailedTimer::ScopedPush sp(TIME_HIGH_LEVEL);
       HighLevelRuntime::get_runtime(p)->process_notify_start(args, arglen);
     }
 
@@ -3488,6 +3493,7 @@ namespace RegionRuntime {
     void HighLevelRuntime::notify_finish(const void * args, size_t arglen, Processor p)
     //--------------------------------------------------------------------------------------------
     {
+      DetailedTimer::ScopedPush sp(TIME_HIGH_LEVEL);
       HighLevelRuntime::get_runtime(p)->process_notify_finish(args, arglen);
     }
 
@@ -3495,6 +3501,7 @@ namespace RegionRuntime {
[…]

### 2011-11-06 runtime/highlevel.h

@@ -78,6 +78,7 @@ namespace RegionRuntime {
     typedef LowLevel::UserEvent UserEvent;
     typedef LowLevel::Lock Lock;
     typedef LowLevel::ElementMask Mask;
+    typedef LowLevel::DetailedTimer DetailedTimer;
     typedef unsigned int Color;
     typedef unsigned int MapperID;
     typedef unsigned int Context;
@@ -835,7 +836,11 @@ namespace RegionRuntime {
       arglen -= sizeof(Context);
       
       // Invoke the task with the given context
-      T return_value = (*TASK_PTR)((const void*)arg_ptr, arglen, regions, ctx, runtime);
+      T return_value;
+      {
+	DetailedTimer::ScopedPush sp(TIME_KERNEL);
+	return_value = (*TASK_PTR)((const void*)arg_ptr, arglen, regions, ctx, runtime);
+      }
 
       // Send the return value back
       runtime->end_task(ctx, (void*)(&return_value), sizeof(T));
@@ -859,7 +864,10 @@ namespace RegionRuntime {
       arglen -= sizeof(Context);
       
       // Invoke the task with the given context
-      (*TASK_PTR)((const void*)arg_ptr, arglen, regions, ctx, runtime);
+      {
+	DetailedTimer::ScopedPush sp(TIME_KERNEL);
+	(*TASK_PTR)((const void*)arg_ptr, arglen, regions, ctx, runtime);
+      }
 
       // Send an empty return value back
       runtime->end_task(ctx, NULL, 0); 
@@ -907,11 +915,17 @@ namespace RegionRuntime {
         {
           fast_regions.push_back(it->convert());
         }
-        return_value = (*FAST_TASK_PTR)((const void*)arg_ptr, arglen, fast_regions, ctx, runtime);
+	{
+	  DetailedTimer::ScopedPush sp(TIME_KERNEL);
+	  return_value = (*FAST_TASK_PTR)((const void*)arg_ptr, arglen, fast_regions, ctx, runtime);
+	}
       }
       else
       {
-        return_value = (*SLOW_TASK_PTR)((const void *)arg_ptr, arglen, regions, ctx, runtime);
+	{
+	  DetailedTimer::ScopedPush sp(TIME_KERNEL);
+	  return_value = (*SLOW_TASK_PTR)((const void *)arg_ptr, arglen, regions, ctx, runtime);
+	}
       }
 
[…]

### 2011-11-06 runtime/lowlevel.cc

@@ -327,7 +327,246 @@ namespace RegionRuntime {
       REMOTE_COPY_MSGID,
       VALID_MASK_REQ_MSGID,
       VALID_MASK_DATA_MSGID,
+      ROLL_UP_TIMER_MSGID,
+      ROLL_UP_DATA_MSGID,
     };
+
+    // detailed timer stuff
+
+    struct TimerStackEntry {
+      int timer_kind;
+      double start_time;
+      double accum_child_time;
+    };
+
+    struct PerThreadTimerData {
+    public:
+      PerThreadTimerData(void)
+      {
+	thread = pthread_self();
+	gasnet_hsl_init(&mutex);
+      }
+
+      pthread_t thread;
+      std::list<TimerStackEntry> timer_stack;
+      std::map<int, double> timer_accum;
+      gasnet_hsl_t mutex;
+    };
+
+    gasnet_hsl_t timer_data_mutex = GASNET_HSL_INITIALIZER;
+    std::vector<PerThreadTimerData *> timer_data;
+    __thread PerThreadTimerData *thread_timer_data;
+    
+#ifdef DETAILED_TIMING
+    /*static*/ void DetailedTimer::clear_timers(void)
+    {
+      // take global mutex because we need to walk the list
+      AutoHSLLock l1(timer_data_mutex);
+      for(std::vector<PerThreadTimerData *>::iterator it = timer_data.begin();
+	  it != timer_data.end();
+	  it++) {
+	// take each thread's data's lock too
+	AutoHSLLock l2((*it)->mutex);
+	(*it)->timer_accum.clear();
+      }
+    }
+
+    /*static*/ void DetailedTimer::push_timer(int timer_kind)
+    {
+      if(!thread_timer_data) {
+	//printf("creating timer data for thread %lx\n", pthread_self());
+	AutoHSLLock l1(timer_data_mutex);
+	thread_timer_data = new PerThreadTimerData;
+	timer_data.push_back(thread_timer_data);
+      }
+
+      // no lock needed here - only our thread touches the stack
+      TimerStackEntry entry;
+      entry.timer_kind = timer_kind;
+      struct timespec ts;
+      clock_gettime(CLOCK_MONOTONIC, &ts);
+      entry.start_time = (1.0 * ts.tv_sec + 1e-9 * ts.tv_nsec);
+      entry.accum_child_time = 0;
+      thread_timer_data->timer_stack.push_back(entry);
+    }
+	
[…]

### 2011-11-06 runtime/lowlevel.h

@@ -9,6 +9,19 @@
 
 #include "common.h"
 
+// outside of namespace because 50-letter-long enums are annoying
+enum {
+  TIME_NONE,
+  TIME_KERNEL,
+  TIME_COPY,
+  TIME_HIGH_LEVEL,
+  TIME_LOW_LEVEL,
+  TIME_MAPPER,
+  TIME_SYSTEM,
+};
+
+#define NO_DETAILED_TIMING
+
 namespace RegionRuntime {
   namespace LowLevel {
     // widget for generating debug/info messages
@@ -132,6 +145,30 @@ namespace RegionRuntime {
 	}
 	return index;
       }
+    };
+
+    class DetailedTimer {
+    public:
+#ifdef DETAILED_TIMING
+      static void clear_timers(void);
+      static void push_timer(int timer_kind);
+      static void pop_timer(void);
+      static void roll_up_timers(std::map<int, double>& timers, bool local_only);
+      static void report_timers(bool local_only = false);
+#else
+      static void clear_timers(void) {}
+      static void push_timer(int timer_kind) {}
+      static void pop_timer(void) {}
+      static void roll_up_timers(std::map<int, double>& timers, bool local_only) {}
+      static void report_timers(bool local_only = false) {}
+#endif
+
+      class ScopedPush {
+      public:
+	ScopedPush(int timer_kind) { push_timer(timer_kind); }
+	~ScopedPush(void) { pop_timer(); }
+      };
+
     };
 
     // forward class declarations because these things all refer to each other

### 2011-11-06 runtime/lowlevel_gpu.cc

@@ -270,6 +270,7 @@ namespace RegionRuntime {
 
       virtual void execute(void)
       {
+	DetailedTimer::ScopedPush sp(TIME_COPY);
 	log_gpu.info("gpu memcpy: dst=%p src=%p bytes=%zd kind=%d",
 		     dst, src, bytes, kind);
 	fflush(stdout);
@@ -295,6 +296,7 @@ namespace RegionRuntime {
 
       virtual void execute(void)
       {
+	DetailedTimer::ScopedPush sp(TIME_COPY);
 	log_gpu.info("gpu memcpy generic: gpuptr=%p mem=%x offset=%zd bytes=%zd kind=%d",
 		     gpu_ptr, memory->me.id, mem_offset, bytes, kind);
 	const size_t BUFFER_SIZE = 65536;

### 2011-11-06 runtime/runckt.pl

@@ -13,16 +13,20 @@ die "must specify output directory with -o!" unless($outdir);
 
 mkdir($outdir) || die "mkdir $outdir: $!\n";
 
-#for ([1, 1, 12], [1, 2, 12], [2, 1, 12], [3, 1, 12], [4, 1, 12]) {
-for (  [2, 1, 12], [3, 1, 12], [4, 1, 12]) {
+for ([1, 1, 12], [1, 2, 12], [2, 1, 12], [3, 1, 12], [4, 1, 12]) {
+#for (  [2, 1, 12], [3, 1, 12], [4, 1, 12]) {
     my($nn, $ng, $np) = @$_;
 
-    for my $run (1..5) {
+    for my $run (1..3) {
 	my $outfile = "${outdir}/ckt_${nn}_${ng}_${np}_run${run}.log";
-	my $cmd = "gasnetrun_ibv -n ${nn} ./cktgpu2 -ll:gpu ${ng} -level 1 -cat app -npp 25000 -wpp 100000 -ll:fsize 3072 -ll:zsize 1536 -ll:csize 8192 -ll:gsize 2000 -l 10 -p ${np} > $outfile";
+	my $cmd = "gasnetrun_ibv -n ${nn} ./cktgpu2 -ll:gpu ${ng} -level 1 -cat app -npp 25000 -wpp 100000 -ll:fsize 3072 -ll:zsize 1536 -ll:csize 8192 -ll:gsize 2000 -l 5 -p ${np} > $outfile";
 	print "$cmd\n";
-	my $res = system($cmd);
-	print "result = $res\n";
+	for my $tries (1..5) {
+	    my $res = system($cmd);
+	    print "result = $res\n";
+	    last if($res == 0);
+	    unlink($outfile);
+	}
     }
     #last;
 }

### 2011-11-06 doc/pldi2012/circuit.tex

\section{Example: Circuit Simulator}

%\include{code_ex}

Listing~\ref{lst:code_ex} shows an example of Legion code for an electrical
circuit simulation.  Although it does not use all the capabilities of Legion
(or all the aspects of the language), it covers the core set.  Legion code is
derived from C, but different in several important ways.

Lines 1-16 
define the data structures used in the rest of the code.  A {\tt Circuit}
is an irregular data structure that describes the wires in the circuit and
the 'nodes' (where the wires meet).  The basic nodes and wires are defined
first.  Nodes have a few fields for the simulation (e.g. charge, capacitance),
and are hooked up in a circular linked list to allow the simulation to be
able to iterate over all of them.  Pointers in data structures (e.g. the
``next'' fields in this linked list) must be constrained
to a particular region, so Legion uses a {\tt Type@region} syntax rather than
C's unconstrained {\tt Type~*}.  The example relies on the ability to
constrain different linked lists of nodes differently, so rather than using
the name of some static region the {\tt Node} definition is parameterized on
a region {\tt rn}.  The use of the same parameter on the type pointed to by
the next pointer means that every single node in the linked list is constrained
to reside in the specified region.

The definition of a {\tt Wire} is similar, except that a wire must refer to
the two nodes to which it is electrically connected.  The pointers to the nodes
must be constrained, and the type that the pointer points to requires a region
as well, but unlike in the linked list case, two different regions are used.
This is a common occurrance in Legion code, and arises when one needs to place
one constraint on the objects directly referenced, and a second (presumably
weaker) constraint on objects that are indirectly referencable through the
direct pointers.

With the definition of nodes and wires out of the way, lines 4-9 declare how
they are formed into a {\tt Circuit}, and is our first example of a
\emph{region relationship}.  At the language level, a region relationship is
similar to a structure, except that it is allowed to contain references to
dynamically allocated regions as fields, and the types of the fields of a region
relationship are allowed to refer to those region fields.  (This will be 
discussed in additional detail below.)  It is also allowed
to place constraints on those region fields, which we'll see in the next
declaration.  In this example, a circuit is defined
by two regions, one for its nodes and one for its wires.  The element type of
the node region (and of the {\tt first\_node} pointer) guarantees that every
node in the circuit reside in this region.  Simiarly, the type of the wire
region's element (and of the {\tt first\_wire} pointer) not only guarantees
that every wire in the circuit is in the wire region, but also guarantees that
every node referred to by any wire is in the circuit's node region.

The last definition is for the pieces that the circuit is partitioned into.
The original node and wire regions are specified as parameters, and the
four kinds of subregions that are created below are made fields of the
{\tt CircuitPiece}.  Constraints are placed on all four of these fields that
they must be subregions of the original node and wire regions that were
provided as parameters.  Since the simulation code will need to be able to
iterate over the nodes owned by a piece, the parameterization of the {\tt rn\_pvt},
{\tt rn\_shr}, and {\tt first\_node} fields constrains the linked list of
nodes to reside entirely in those two regions.  The linked list of wires is
similarly constrained, and the wire's node pointers are constrained to fall
into one of the three subregions defined in the {\tt CircuitPiece}.  Finally,
the use of {\tt rn} for {\tt rn\_ghost}'s parameter and the second parameter
in the wire type places no constraints on which nodes a ghost node might point
to.

Line 17 declares the main simulator function, which accepts a \emph{Circuit}
to be simulated.  In addition to specifying the return type and the types and
names of the formal parameters, it must also specify which regions of memory
it will access, how it will access them (e.g. read-only, read-write, or with
reductions), and what coherence guarantees are needed (e.g. exclusive access).
The regions listed in these access declarations may be anything in the static
scope of the function definition, the formal parameters, or fields of the formal
parameters.  In our example, the simulation of the circuit will need to read
and write all of the nodes and all of the wires in the circuit, and it must
be done with exclusive access to ensure a correct result.

%\include{part_fig}

Lines 19-27 are responsible for partitioning the circuit into \emph{MAX\_PIECES}
pieces that can be worked on in parallel.  In order to partition a region, we
need to provide a \emph{coloring}, which is a relation over the valid elements
in a region and a set of ``colors''.  Our current implementation uses small
integers or booleans for the names of colors, but any enumerable set would be
fine.  The coloring relation need not be total - it can leave out some elements entirely.  And although it is not required to be injective (it can map a single
element to multiple colors), the fact that injective mappings will always
result in disjoint subregions makes them strongly encouraged (when possible).

Our example uses four colorings to help create the partitions of the circuit.
A graph partitioning algorithm
(not shown) is run on to divide the nodes of the graph into \emph{MAX\_PIECES}
subsets.  (These subsets are hopefully reasonably compact, but the correctness
of the simulation is not dependent on that.)  The thick lines in \ref{sfig:part_fig:pvs}
show a how the nodes in a small graph might be partitioned into 3 subsets.
Once the subsets are known, the \emph{node\_owner\_map} is created by assigning
each node a color corresponding to which subset it was placed in.  The 
\emph{wire\_owner\_map} assigns each wire to the same subset as its ``in\_node''.
The \emph{node\_nghbr\_map} maps a node to color(s) of all wires that 
connect to it.  Finally, the \emph{node\_sharing\_map} is derived from the
\emph{node\_nghbr\_map}, with a node colored ``true'' if any colors other than 
its own were used, and ``false'' if the only wires that connect to a node are
in the same piece.
Figure~\ref{fig:part_fig} shows how the partitions are defined.

Line 19 uses the \emph{wire\_owner\_map} to partition the wires region of the
circuit into a subregion for each piece, but the partitioning of the nodes is
more complicated due to the sharing that is necessary between the pieces.
First, the \emph{node\_sharing\_map} is used to create two subregions: $p\_nodes\_pvs[false]$ contains all the nodes that are private to some piece (i.e. will
never be needed for the computations in any other piece), while $p\_nodes\_pvs[true]$ contains nodes that will be accessed by multiple pieces' computations.
Each of these subregions is then partitioned using the $node\_owner\_map$ to
create the subregions owned by each computation.  These are the $p_i$ and $s_i$
subregions, shown in figures \ref{sfig:part_fig:p_i} and \ref{sfig:part_fig:s_i}.  Finally, the $g_i$ subregions (shown in figure~\ref{sfig:part_fig:g_i}) are
created using the $node\_nghbr\_map$ to 
create subregions that include the ``ghost'' nodes needed to perform each
piece's calculations.

With the partitioning operations completed, the various subregions are
recorded in an array of $CircuitPiece$ structures.  The combination of the
choice of parameters on line 29 with the constraints from lines 11-13 allow 
this array to hold pieces of the circuit provided to the $simulate\_circuit$
function, but no others.

Lines 31-32 use a \emph{simultaneous assignment} operation to fill in multiple
fields of a $CircuitPiece$ at once.  Because a region relationship can have
fields with types that refer its own fields' values, it is often not possible
to assign fields one at a time without violating the type checking rules.
The simultaneous assignment operator asks the type checker only to make sure
that the fields in the structure would have the right (self-referential) types
after all the fields have been changed.

Lines 34-38 form the bulk of the actual simulation, performing three passes
over the circuit on each iteration.  For each pass, a for loop is used to 
spawn a task for each piece of the circuit.  There are no explicit requests for
parallel execution, nor is there explicit synchronization required between the
passes.  As we will see, bBoth the fact that the pieces can be run in parallel
for a single 
pass and the the required inter-task dependencies are determined automatically
by the runtime based on the region access annotations on the task declarations.

The declarations for the three subtasks are shown on lines 41-50.  The 
$calc\_new\_currents$ task declares that it will read and write the wires
for that piece, and will need to read both the nodes in its piece and any
node that is adjacent to the piece (i.e. any node that one of the wires in the
piece might connect to).  Since the wire subregions are known to be disjoint,
the write sets of invocations of $calc\_new\_currents$ do not overlap, and can
therefore be safely run in parallel.

The $distribute\_charge$ subtask turns things around, reading the piece's 
wires and updating all the nodes that those wires connect to.  However,
rather than requesting the ability to read and write the nodes (which would
require serialization of these tasks for correctness), the task declares that
it will use reorderable reduction operations and that the coherence requirement
can be reduced to atomicity rather than exclusive access.  As long as the
runtime can guarantee to apply the reductions from multiple subtasks safely, it
can run the subtasks themselves in parallel.  Each invocation of 
$distribute\_charge$ will be delayed until the corresponding invocation of 
$calc\_new\_currents$ has completed due to the read-after-write dependency on
the corresponding wire subregion.  However, despite the apparent 
write-after-read anti-dependency on the ghost node regions, $distribute\_charge$
tasks will generally not have to wait on the the completion of the other
$calc\_new\_current$ tasks.  If there is sufficient memory available to make
two copies of those nodes, the runtime can allow $distribute\_charge$ tasks to
start calculating a new version of the nodes while older $calc\_new\_currents$
tasks are still referring to the older version, all completely transparently to
the application code.

The final task is $update\_voltages$, in which each piece is able to update
voltage information on the nodes in that piece based on the reductions that
happened in the previous task.  Again, the disjointness of the $p_i$ and $s_i$
node subregions allows the runtime to safely run these tasks in parallel.  In
this case, the runtime does wait for the completion of all the tasks in the 
previous pass.  The read-after-write dependence on the $p_i$ is a guaranteed
conflict, but there is also potential overlap between the $s_i$ subregions
being reduced to in the previous pass and the $g_i$ subregions being accessed
in this pass.  Although not every pair of $s_i$ and $g_j$ conflict, the
runtime knows that they were created from two independent partitioning
operations and guarantees correctness by conservatively assuming they might
conflict.

%\include{code_ex}
%\include{part_fig}

### 2011-11-06 doc/pldi2012/code_ex.tex

@@
 }
-\begin{lstlisting}[float,caption={Circuit Simulation Code Example}]
+\begin{lstlisting}[float={t},label={lst:code_ex},caption={Circuit Simulation Code Example}]
 struct Node<rn>    { Node<rn>@rn next;    float charge, capacitance; }
@@
   region< Node<r_all_nodes> >                        r_all_nodes;
-  region< Wire<r_all_nodes,r_all_wires> >       r_all_wires;
+  region< Wire<r_all_nodes,r_all_nodes,r_all_wires> >       r_all_wires;
   Node<r_all_nodes>@r_all_nodes                   first_node;
@@
 
+// read info from nodes connected to each wire, update state of wire
 void calc_new_currents(CircuitPiece<rn,rw> piece): RWE(piece.rw_pvt),
-                                                                         ROE(piece.rn_pvt,piece.rn_ghost) {
-  // read info from nodes connected to each wire, update state of wire
-}
+                                                                         ROE(piece.rn_pvt,piece.rn_ghost)
 
+// current moving through wires redistributes charge between nodes
 void distribute_charge(CircuitPiece<rn,rw> piece): ROE(piece.rw_pvt),
-                                                                         RdA(piece.rn_pvt,piece.rn_ghost) {
-  // current moving through wires redistributes charge between nodes
-}
+                                                                         RdA(piece.rn_pvt,piece.rn_ghost)
 
+// total charge added to a node causes changes in voltage
 void update_voltages(CircuitPiece<rn,rw> piece): RWE(piece.rn_pvt,piece.rn_shr)
-{
-  // total charge added to a node causes changes in voltage
-}
 \end{lstlisting}

### 2011-11-06 doc/pldi2012/exec.tex

@@
 
-In this section we present the Legion execution model and argue that
+In this section we present the Legion execution model and show that
 it is correct in that it is guaranteed to preserve the sequential
@@
 execution model we consider an extremely restricted core language, so
-simple in fact that it is not even Turing complete and is also extremely
+simple in fact that it is not even Turing complete and is also very
 inconvenient for writing programs.  The loss in expressive power, however,
@@
 \begin{eqnarray*}
-F & := & \mbox{def}\ f(r_1,\ldots,r_n): P_1,\ldots,P_m  =   s_1; \ldots s_n; \\[.15in]
-P & := & R(r) | W(r) \\[.15in]
+P & := & F_1 \ldots F_n \\[.1in]
+F & := & \mbox{\tt def}\ f(r_1,\ldots,r_n) =   s_1; \ldots s_n; \\[.1in]
 s & := & f(r_1,\ldots,r_n) \\
@@
 & | & x = \deref r \\
-&| & \mbox{\tt newregion}\ r \\
 &| & (r_1, r_2)_{p} = \mbox{\tt partition}(r) \\ 
-&| & \mbox{\tt close}\ p 
+%&| & \mbox{\tt close}\ p 
 \end{eqnarray*}
@@
 There are four kinds of names: functions $f$, regions $r$, local
-variables $x$, and partitions $p$.  A function takes a number of
-regions as arguments and the function definition $f$ specifies the
-region permissions $f$ requires; the permission $R(r)$ means $f$ may
-read $r$, and $W(r)$ means $f$ may write $r$.
+variables $x$, and partitions $p$.  A function takes only regions as
+arguments and a function body consists of an ordered list of
+statements.  The core language has no permissions or coherence annotations
+on functions; for simplicity, in this section we consider only exclusive access in the case
+that a function writes a region.  The statement $f(r_1,\ldots,r_n)$ calls a function, which
+is executed for its effects on the region arguments.  The statement $x
+= \deref r$ reads a value from region $r$ and stores it in local
+variable $x$, and the statement $\deref r = x$ stores the value of $x$
+in region $r$.
 
-A function body consists of an ordered list of statements.  The statement
-$f(r_1,\ldots,r_n)$ calls a function, which is executed for its effects
-on its region arguments.  The statement $x = \deref
-r$ reads a value from region $r$ and stores it in local variable $x$,
-and the statement $\deref r = x$) stores the value of $x$ in region
-$r$.  A $\mbox{\tt newregion} r$ statement creates a new region with
-name $r$.
+The statement $(r_1, r_2)_{p} = \mbox{\tt partition}(r)$ creates two
+{\em subregions} $r_1$ and $r_2$ partitioning the parent region $r$.
+All three region names ($r_1$,$r_2$,and $r$) can be used after the
+partition.  Regions $r_1$ and $r_2$ are disjoint, meaning that writes
+to $r_1$ have no effect on the contents of $r_2$ and vice versa, but
+neither $r_1$ nor $r_2$ is disjoint from $r$.  The partition itself
+also has a name $p$.  We define a relation $\rleq$ which is the
+reflexive, transitive closure of
+\[ r_1 \rleq p \ \ \ r_2 \rleq p \ \ \ p \rleq r \]
+The $\rleq$ relation defines a {\em region forest} with regions at the
+roots, and alternating levels where partitions are always children of
+regions and regions are always children of partitions.  If we ignore
+partitions, $\rleq$ is just the subregion relationship: $r \rleq r'$
+means that $r$ is a subregion (either immmediately or transitively) of
+$r'$.  Including partitions also gives information about disjointness.
+Let $\lca{r}{r'}$ be the meet of $r$ and $r'$ (the least-common
+ancestor of $r$ and $r'$ in the region forest).  If $r \neq r'$ and
+$\lca{r}{r'} = p$ then $r$ and $r'$ are disjoint, as $r$ and $r'$ must
+be in distinct subregions of the partition $p$ and partitions
+guarantee disjoint subregions.  If $\lca{r}{r'} = r''$ then either $r
+\rleq r' = r''$ , $r \rleq r = r''$, or $r \leq p$, $r' \leq p'$, and
+$p$ and $p'$ are two distinct partitions of $r''$.  In any of these
+cases $r$ and $r'$ cannot be proven disjoint using $\rleq$.
 
-The statement $(r_1, r_2)_{p} = \mbox{\tt partition}(r)$
-creates two {\em subregions} $r_1$ and $r_2$ partitioning the parent
-region $r$.  All three region names ($r_1$,$r_2$,and $r$) can be used
-after the partition.  Regions $r_1$ and $r_2$ are disjoint but neither
-$r_1$ nor $r_2$ is disjoint from $r$.  The partition itself also has a name $p$,
-which is used to {\em close} the partition in the $\mbox{\tt close}\ p$ statement 
-Intuitively, closing a partition reconciles modifications to the subregions with the {\em parent} region: 
-$r$ is updated to incorporate any changes to $r_1$ and $r_2$.  In our source language {\tt close} is not
-available to the programmer, but is invoked by the language implementation to reconcile distributed copies
-of regions whenever necessary, and it is impossible for the program to observe discrepencies between
-a region and its subregions. This toy language, then, is closer to the level of ourintermediate representation,
-which is most appropriate for discussing our runtime system.
+%which is used to {\em close} the partition in the $\mbox{\tt close}\
+%p$ statement Intuitively, closing a partition reconciles modifications
+%to the subregions with the {\em parent} region: $r$ is updated to
+%incorporate any changes to $r_1$ and $r_2$.  In our source language
+%{\tt close} is not available to the programmer, but is invoked by the
+%language implementation to reconcile distributed copies of regions
+%whenever necessary, and it is impossible for the program to observe
+%discrepencies between a region and its subregions. This toy language,
+%then, is closer to the level of ourintermediate representation, which
+%is most appropriate for discussing our runtime system.
 
-A program is a list of functions, with the entry point the first function on the list.
-We further require that every function except the entry poiint be called in exactly one place
-and to have its formal parameters be identical to the actual parameters in its single call.
-Region names in {\tt newregion} and partition names in {\tt partition} are chosen to be distinct from
-all other region/partition names in the program.  These restrictions allow us to avoid dealing with
-alpha renaming and also mean that each runtime call is uniquely named by its function name.
+A program is a list of functions, with the entry point the first
+function on the list.  We further require that every function (except
+the entry point) is called in exactly one place and that the formal
+parameters are named identically to the actual parameters in the
+call.  Further, local variables, new region names and
+partition names in {\tt partition} are chosen to be distinct from all
+other names in the program.  These restrictions allow us to avoid
+dealing with variable renaming and also mean that each runtime call is
+uniquely named by its function name.
+
+The state of an execution step for a program is a map $M$ from the
+variable and region names of the program to their values.  We leave the values
+abstract, assuming only that each statement performs a deterministic
+update of the variables/regions it writes that depends only on the
+current state of the regions/variables it reads.
+The sets of regions read and written by statements are defined as follows:
+\[
+\begin{array}{rcl}
+\rread{x = \deref r} & = & \{ r \} \\
+\rwrite{\deref r = x} & = & \{ r \} \\
+\rwrite{(r_1, r_2)_{p} = \mbox{\tt partition}(r)} & = & \{ r_1,r_2,r \} \\[.15in]
+\rwrite{f(r_1,\ldots,r_n)} & = & \bigcup_{1 \leq i \leq n}{\rwrite{s_i}} \\
+\multicolumn{3}{l}{\ \ \ \ \mbox{where\ {\tt def}}\ f(r_1,\ldots,r_n): =   s_1; \ldots s_n;} \\[.15in]
+\rread{f(r_1,\ldots,r_n)} & = & \bigcup_{1 \leq i \leq n}{\rread{s_i}} \\
+\multicolumn{3}{l}{\ \ \ \ \mbox{where\ {\tt def}}\ f(r_1,\ldots,r_n): =   s_1; \ldots s_n;} \\[.15in]
+\end{array}
+\]
+In all other cases the read/write set is empty (e.g., $\rread{\deref r = x} = \emptyset$).
+
+The {\em sequential execution order} is the canonical order in which reads, writes, and partitions of regions are performed.
+Th sequential order is obtained by inlining all function calls---repeatedly
+replace function calls $g(r_1,\ldots,r_n)$ by their function bodies until the entry function contains no function calls.
+The order of statements in the entry function is then the sequential execution order. 
+Two programs $P_1$ and $P_2$ are equivalent, written $P_1 \equiv P_2$, if starting in the same initial states they halt in
+the same final states when evaluated in the sequential execution order.
+
+Consider the following program, where reads, writes, and partitions are numbered in the sequential execution order:
+{\small 
+\begin{tabbing}
+\ \ \ \ \ \= {\tt def} \= $\rm f(r_0) =$ \\
+1. \>\>   $\rm (r_1,r_2)_a = \mbox{\tt partition}(r_0);$   \\
+2. \>\>  $\rm (r_3,r_4)_b = \mbox{\tt partition}(r_1);$ \\
+\>\>     $\rm g(r_1); h(r_2); j(r_3); k(r_3,r_4);$ \\
+\> {\tt def} $\rm g(r_1) =$   \\
+3. \>\>  $\rm \ast r_1 = u;$ \\ \\
+\> {\tt def} $\rm h(r_2) =$ \\      
+4. \>\>   $\rm \ast r_2 = v;$\\ \\
+\> {\tt def} $\rm j(r_3) = $ \\
+5.\>\>   $\rm w = \ast r_3;$ \\ \\
+\> {\tt def} $\rm k(r_3,r_4) = $\\
+6.\>\>   $\rm x = \ast r_3;$ \\
+7.\>\>   $\rm (r_5,r_6)_c = \mbox{\tt partition}(r_4);$ \\ 
+\>\>     $\rm m(r_5);$ \\ \\
+\> {\tt def} $\rm m(r_5) = $ \\
+8. \>\>  $\rm \ast r_5 = y;$
+\end{tabbing}
+}
+
+Consider any two statements $s_1$ and $s_2$.  We say $s_1$ and $s_2$ are {\em independent} if
+\[
+\begin{array}{l}
+   (\forall r_1 \in \rread{s_1} \cup \rwrite{s_1}.\forall r_2 \in \rwrite{s_2}. \exists p. \lca{r_1}{r_2} = p) \wedge \\
+   (\forall r_2 \in \rread{s_2} \cup \rwrite{s_2}.\forall r_1 \in \rwrite{s_1}. \exists p. \lca{r_1}{r_2} = p)
+\end{array}
+\]
+The definition captures the usual idea that two statements are independent if whenever one of them is a write they do not access the same memory location, but the notion of ``location'' is generalized to regions and must take into account that two regions
+are disjoint only if they are in different components of some partition.
+
+\begin{lemma}
+\rm
+\label{lem:independence}
+If $s_1$ and $s_2$ are independent, then $s_1; s_2 \equiv s_2; s_1$.
+\end{lemma}
+\begin{proof}
+For a contradiction, assume that $s_1$ writes a region $r_1$ and $s_2$ may read from some $r_2$ where $r_1$ and $r_2$
+are not disjoint; then reversing the order of the statements clearly violates the claim.  Now $r_1 \in \rwrite{s_1}$
+and $r_2 \in \rread{s_2}$, and therefore $\lca{r_1}{r_2} = p$ for some partition $p$.  But then $r_1$ and $r_2$ are disjoint.
+The reasoning is symmetric if $s_1$ reads $r_1$ and $s_2$ writes $r_2$ or $s_1$ writes $r_1$ and $s_2$ writes $r_2$.
+Since the read-read case cannot violate independence, the result follows.
+\end{proof}
+We say $s_2$ {\em depends on} $s_1$ if $s_2$ follows $s_1$ in the sequential execution order and the statements are not independent.
+\begin{lemma}
+\rm
+\label{lem:dataflow}
+Let $S = s_1; \ldots s_n;$ be the sequential execution order of reads, writes, and partitions and let $S'$ be any permutation of $S$ such that the order of dependent statements is preserved.
+Then $S \equiv S'$.
+\end{lemma}
+\begin{proof}
+By induction on the number of swaps needed to transpose $S'$ into $S$.
+If $S = S'$ then we are done.  If $S$ and $S'$ are different then
+there must be two adjacent statements $s_1; s_2$ in $S'$ that are
+reversed with respect to the sequential execution order.  Since the order of
+dependent statements is preserved by the permutation, $s_1$ and $s_2$ are also independent.
+But then $s_1; s_2 \equiv s_2; s_1$ by Lemma~\ref{lem:independence} and the result follows.
+\end{proof}
+Lemma~\ref{lem:dataflow} captures the standard idea that any
+topological sort of the statements that preserves dependencies also
+preserves sequential execution semantics.  While this is the basis for
+dataflow parallelism, in the large distributed memory machines we
+target instruction-level dataflow parallelism is both too fine-grain
+and incurs too much communication, as we must compute a global
+dependence graph across the entire program.  In Legion we use a
+different notion of what can be executed in parallel that works at the
+granularity of functions and also requires only local reasoning.  We
+first need a few additional definitions.  Define $\context{s}$ to be
+the function call statement that invokes the function in which
+statement $s$ occurs (here we assume some way of distinguishing
+identical statements that occur in different functions).  Let
+$\interval{f}$ to be the set of statements executed by the one call to
+function $f$ and all the functions $f$ transitively calls; we define
+$\interval{s} = \{ s \}$ for any statement $s$ other than a function call.
+Finally we say statements $s_1$ and $s_2$ are {\em siblings} if they
+occur in the same function body.
+
+\begin{lemma}
+\rm
+\label{lem:locality}
+If $s_2$ and $s_1$ are not independent and $s_1$ and $s_2$ are not siblings, then $\context{s_2}$ and $s_1$ are not independent
+and $s_2$ and $\context{s_1}$ are not independent.
+\end{lemma}
+\begin{proof}
+Follows immediately from the fact that the read/writes sets of $\context{s}$ are supersetsn of the read/write sets of $s$.
+\end{proof}
+We use Lemma~\ref{lem:locality} in the following way.  Consider two
+statements $s_1$ and $s_2$ such that $s_2$ depends on $s_1$.  Let $f$ be the function that is the least
+common ancestor of $s_1$ and $s_2$ in the call tree of the
+program---the unique function $f$ with the smallest interval such that
+$s_1,s_2 \in \interval{f}$.  Then one of the following is true:
+\begin{itemize}
+\item $s_1$ and $s_2$ are siblings.
+
+\item $s_1$ occurs in $f$, there is a function call $f_2(\ldots)$ in $f$ such that $s_2 \in \interval{f_2}$, and $f_2(\ldots)$
+is dependent on $s_1$.
+
+\item $s_2$ occurs in $f$, there is a function call $f_1(\ldots)$ in $f$ such that $s_1 \in \interval{f_1}$, and
+$s_2$ is dependent on $f_1(\ldots)$.
+
+\item There are two distinct function call statements $f_1(\ldots)$ and $f_2(\ldots)$ in the body of $f$ such
+that $s_1 \in \interval{f_1}$ and $s_2 \in \interval{f_2}$ and $f_2(\ldots)$ is dependent on $f_1(\ldots)$.
+\end{itemize}
+In other words, any dependence between arbitrary statements
+in different functions can be identified at a coarser granularity as a
+dependence between two statements (one or both of which may be a function call) within the same function body.  For
+example, in the example program above, statements 3 and 8 are
+dependent because they write into regions that are in different
+partitions of region $r_0$.  This dependence means that any functions
+that include these statements in their interval cannot be independent,
+and in particular in the entry point function $f$ the function call
+$k(r_3,r_4)$ depends on sibling $g(r_1)$.
+
+The following lemma gives another class of statement orderings based on these observations.
+\begin{lemma}
+\rm
+\label{lem:scheduling}
+Let $S$ be the sequential execution order of reads, writes and partitions.  Let $S'$ be any permutation of $S$ such that
+for any sibling statements $s_1$ and $s_2$ such that $s_2$ depends on $s_1$, all statements in $\interval{s_1}$ precede
+all statements in $\interval{s_2}$.  Then $S \equiv S'$.
+\end{lemma}
+\begin{proof}
+It suffices to show that the order of any two dependent statements $s_1$ and $s_2$ is the same in $S$ and $S'$.
+From the discussion above, we know that if $s_2$ depends on $s_1$ then there are two sibling statements $s_1'$ and $s_2'$
+such that $s_1 \in \interval{s_1'}$, $s_2 \in \interval{s_2'}$, and $s_2'$ depends
+on $s_1'$.
+\end{proof}
+The advantage of Lemma~\ref{lem:scheduling} is that provided we have
+a way to efficiently obtain the read and write sets for function calls, scheduling
+decisions can be made completely locally for the statements in a
+single function body at a time, independently of any decisions for
+other function calls.  This observation eliminates any global
+computation to determine what statements can execute in parallel, and
+furtherore allows decisions to be made in parallel, as the statements
+in each function can be scheduled separately.  As a practical matter
+our Legion implementation only parallelizes function calls---the other
+statements within a function body are executed in sequential
+order.
+
+Returning to our example above, in function $f$ the function calls
+$g(r_1)$ and $h(r_2)$ are independent, as they work on disjoint
+regions $r_1$ and $r_2$.  Similarly $j(r_3)$ and $k(r_3,r_4)$ are
+also independent as they both only read from $r_3$.  However, $j$ and $k$
+are both dependent on $g$ and $h$, as the earlier functions modify a different
+partition of the same data.  Thus Legion will ensure functions $j$ and $k$
+begin only after $g$ and $h$ complete, while it will allow $g$ and $h$
+to run in parallel, and $j$ and $k$ to run in parallel.

### 2011-11-06 doc/pldi2012/related.tex
Adding a related work section to the paper

\section{Related Work}

The Chapel programming language \cite{Chamberlain:Chapel} is the closest related work to the Legion
programming model.  Chapel has several concepts built into the language to aid in
the expression of locality.  The first concept is that of domains.  Domains
are similar to logical regions in that they describe a collections of indexes
into items of the same type.  Domains can create sub-domains by slicing the index sets from
a parent domain.  There are two differences between domains and logical regions.
First, domains are a higher level concept than regions since they track index sets which
have concepts such as dimensionality and support iterators.  Logical regions in Legion 
are only allowed to be accessed with pointers.  Second, the act of creating
subdomains doesn't track disjointness information the way that partitioning of logical regions
does making it challenging for the compiler or the runtime to infer task independence.

In addition to domains, Chapel also supports the notion of domain maps and locales to enable 
the programmer to efficiently map domains onto hardware \cite{CHAPEL11}.  Locales are a flat array
of abstract locations.  Programmers can use locales by writing domain maps that specify
how domains are broken up and assigned to locales.  Domain maps provide the same functionality
as partitions and mappers in Legion, but require the user to correctly implement domain
maps in order for the program to be correct.  Legion explicitly isolates correctness
from performance by defining the Mapper interface.  In addition, Chapel's flat array of locales
makes it challenging to fully utilize deep memory hierarchies.

X10 is another parallel programming environment designed to operate on distributed
memory machines \cite{X1005}.  In X10 the concept of places enables programmers to talk about where
to place both data and tasks in the machine.  However, once data and tasks have
been placed they are fixed which mandates that data movement be explictly managed by
user level code or implicitly by the compiler or runtime \cite{X1008}.  Recently X10 has introduced
the concept of a region into the compiler's intermediate representation \cite{X1011}.
Unlike Legion, regions in X10 are not visible to the programmer but are inferred
from high level arrays through static analysis.

SPMD languages such as Titanium \cite{JV:Yel98} and UPC \cite{UPC99} have mechanisms for
describing array partitions in distributed memories.  However, the partition
operations supported only operate on two-level memory hierarchies consisting of local
and global memory and do not reason about pointer data structures.  Part of Legion's 
low-level runtime system is constructed using UPC's underlying GASNet 
runtime system \cite{GASNET07}.

In previous programming systems regions have primarily been used as a construct for
describing memory management schemes \cite{REAPS02}\cite{RC01}  
or for enforcing security policies \cite{CYCLONE01}.  In all cases, regions had
memory layout implications.  Logical regions in Legion enable the programmer to
describe locality and independence without any memory layout implications.

Marino introduces a generic flow-insensitive type-effect system capable of verifying 
accesses to statically tagged locations of memory \cite{PRIVLIGES09}.  Legion's programming
system allows for dynamic creation and partitioning of regions requiring a 
flow-sensitive analysis to track privleges.

### 2011-11-06 doc/pldi2012/exec.tex

@@
 statements.  The core language has no permissions or coherence annotations
-on functions; for simplicity, in this section we consider only exclusive access in the case
+on functions; for simplicity, in this section we simply use exclusive access in the case
 that a function writes a region.  The statement $f(r_1,\ldots,r_n)$ calls a function, which
@@
 A program is a list of functions, with the entry point the first
-function on the list.  We further require that every function (except
+function on the list.  We require that every function (except
 the entry point) is called in exactly one place and that the formal
@@
 The {\em sequential execution order} is the canonical order in which reads, writes, and partitions of regions are performed.
-Th sequential order is obtained by inlining all function calls---repeatedly
+The sequential order is obtained by inlining all function calls---repeatedly
 replace function calls $g(r_1,\ldots,r_n)$ by their function bodies until the entry function contains no function calls.
 The order of statements in the entry function is then the sequential execution order. 
-Two programs $P_1$ and $P_2$ are equivalent, written $P_1 \equiv P_2$, if starting in the same initial states they halt in
-the same final states when evaluated in the sequential execution order.
+Two programs $P_1$ and $P_2$ are equivalent, written $P_1 \equiv P_2$, if starting in the same initial state they halt in
+the same final state when evaluated in the sequential execution order.
 
@@
 different notion of what can be executed in parallel that works at the
-granularity of functions and also requires only local reasoning.  We
+granularity of functions and also requires analysis only within a function body.  We
 first need a few additional definitions.  Define $\context{s}$ to be
@@
 identical statements that occur in different functions).  Let
-$\interval{f}$ to be the set of statements executed by the one call to
+$\interval{f}$ be the set of statements executed by the one call to
 function $f$ and all the functions $f$ transitively calls; we define
@@
 \label{lem:locality}
-If $s_2$ and $s_1$ are not independent and $s_1$ and $s_2$ are not siblings, then $\context{s_2}$ and $s_1$ are not independent
+If $s_1$ and $s_2$ are not independent and not siblings, then $s_1$ and $\context{s_2}$ are not independent
 and $s_2$ and $\context{s_1}$ are not independent.
@@
 \begin{proof}
-Follows immediately from the fact that the read/writes sets of $\context{s}$ are supersetsn of the read/write sets of $s$.
+Follows immediately from the fact that the read/writes sets of $\context{s}$ are supersets of the read/write sets of $s$.
 \end{proof}
@@
 \begin{itemize}
-\item $s_1$ and $s_2$ are siblings.
+\item $s_1$ and $s_2$ are siblings in $f$.
 
@@
 in different functions can be identified at a coarser granularity as a
-dependence between two statements (one or both of which may be a function call) within the same function body.  For
-example, in the example program above, statements 3 and 8 are
+dependence between two statements (one or both of which may be a function call) within the same function body.  
+In the example program above, statements 3 and 8 are
 dependent because they write into regions that are in different
@@
 The advantage of Lemma~\ref{lem:scheduling} is that provided we have
-a way to efficiently obtain the read and write sets for function calls, scheduling
+a way to obtain the read and write sets for function calls, scheduling
 decisions can be made completely locally for the statements in a
 single function body at a time, independently of any decisions for
-other function calls.  This observation eliminates any global
+other functions.  This observation eliminates any global
 computation to determine what statements can execute in parallel, and
-furtherore allows decisions to be made in parallel, as the statements
-in each function can be scheduled separately.  As a practical matter
+furthermore allows scheduling decisions theselves to be parallelized, as the statements
+in different functions can be scheduled separately.  As a practical matter
 our Legion implementation only parallelizes function calls---the other

### 2011-11-06 doc/pldi2012/circuit.tex

@@
 \section{Example: Circuit Simulator}
+\label{sec:ex}
 
@@
 
-Listing~\ref{lst:code_ex} shows an example of Legion code for an electrical
-circuit simulation.  Although it does not use all the capabilities of Legion
-(or all the aspects of the language), it covers the core set.  Legion code is
-derived from C, but different in several important ways.
+In this section we give an informal overview of Legion through an example.
+Listing~\ref{lst:code_ex} shows Legion code for an electrical
+circuit simulation, which illustrates the core features of the programming model.
 
-Lines 1-16 
-define the data structures used in the rest of the code.  A {\tt Circuit}
-is an irregular data structure that describes the wires in the circuit and
-the 'nodes' (where the wires meet).  The basic nodes and wires are defined
-first.  Nodes have a few fields for the simulation (e.g. charge, capacitance),
-and are hooked up in a circular linked list to allow the simulation to be
-able to iterate over all of them.  Pointers in data structures (e.g. the
-``next'' fields in this linked list) must be constrained
-to a particular region, so Legion uses a {\tt Type@region} syntax rather than
-C's unconstrained {\tt Type~*}.  The example relies on the ability to
-constrain different linked lists of nodes differently, so rather than using
-the name of some static region the {\tt Node} definition is parameterized on
-a region {\tt rn}.  The use of the same parameter on the type pointed to by
-the next pointer means that every single node in the linked list is constrained
-to reside in the specified region.
+Lines 1-16 define data structures.  A circuit consists of wires and
+nodes where wires meet.  {\tt Node}s have fields for the simulation
+(e.g. charge, capacitance) and are linked in a circular list.
+Pointers in data structures (e.g. the {\tt next} fields in the list)
+point to a particular region given by the standard {\tt Type@region}
+syntax.  The program relies on different lists being in different
+regions, so the {\tt Node} definition is parameterized on a region
+{\tt rn}.  The {\tt next} field of a $\tt Node \langle rn \rangle$ has
+type $\tt Node \langle rn \rangle \mbox{\tt @} rn$, which implies that
+all nodes in the list are in region {\tt rn}.
 
-The definition of a {\tt Wire} is similar, except that a wire must refer to
-the two nodes to which it is electrically connected.  The pointers to the nodes
-must be constrained, and the type that the pointer points to requires a region
-as well, but unlike in the linked list case, two different regions are used.
-This is a common occurrance in Legion code, and arises when one needs to place
-one constraint on the objects directly referenced, and a second (presumably
-weaker) constraint on objects that are indirectly referencable through the
-direct pointers.
+{\tt Wire}s are similar, except that a wire refers to two nodes to
+which it is electrically connected.  The two {\tt Node}s are in region {\tt rn},
+but they may point to {\tt Node}s in a different region {\tt rn2}.  This is a common pattern in Legion code, and
+arises when one needs to place one constraint on the objects directly
+referenced, and a second (presumably weaker) constraint on objects
+that are indirectly reached through the direct pointers.
 
-With the definition of nodes and wires out of the way, lines 4-9 declare how
-they are formed into a {\tt Circuit}, and is our first example of a
-\emph{region relationship}.  At the language level, a region relationship is
-similar to a structure, except that it is allowed to contain references to
-dynamically allocated regions as fields, and the types of the fields of a region
-relationship are allowed to refer to those region fields.  (This will be 
-discussed in additional detail below.)  It is also allowed
-to place constraints on those region fields, which we'll see in the next
-declaration.  In this example, a circuit is defined
-by two regions, one for its nodes and one for its wires.  The element type of
-the node region (and of the {\tt first\_node} pointer) guarantees that every
-node in the circuit reside in this region.  Simiarly, the type of the wire
-region's element (and of the {\tt first\_wire} pointer) not only guarantees
-that every wire in the circuit is in the wire region, but also guarantees that
-every node referred to by any wire is in the circuit's node region.
+Lines 4-9 declare {\tt Circuit}s, which is our first example of a
+\emph{region relationship}.  A region relationship is just a structure
+with named fields, except that any region fields can be used in the
+type declarations.  Thus the region {\tt r\_all\_nodes} contains {\tt
+Node}s that point only into region {\tt r\_all\_nodes}, guaranteeing that
+all {\tt Node}s are in region {\tt r\_all\_nodes}.  (Note that regions are
+declared with the single data type they can contain.)  Simiarly, the
+type of the wire region's element not only guarantees that every wire
+in the circuit is in the wire region, but also guarantees that every
+node referred to by any wire is in the circuit's node region.
 
-The last definition is for the pieces that the circuit is partitioned into.
+The last definition is for the pieces into which the circuit is partitioned.
 The original node and wire regions are specified as parameters, and the
-four kinds of subregions that are created below are made fields of the
-{\tt CircuitPiece}.  Constraints are placed on all four of these fields that
-they must be subregions of the original node and wire regions that were
-provided as parameters.  Since the simulation code will need to be able to
-iterate over the nodes owned by a piece, the parameterization of the {\tt rn\_pvt},
-{\tt rn\_shr}, and {\tt first\_node} fields constrains the linked list of
+four kinds of subregions that are created are made fields of the
+{\tt CircuitPiece}.  This declaration also illustrates {\em subregion constraints}
+placed on the three region fields stating the region fields  must be subregions of the 
+node and wire regions provided as parameters.  Since the simulation code will need to be able to
+iterate over the nodes owned by a {\tt CircuitPiece}, the parameterization of the {\tt rn\_pvt} (private nodes),
+{\tt rn\_shr} (shared nodes), and {\tt first\_node} fields constrains the linked list of
 nodes to reside entirely in those two regions.  The linked list of wires is

### 2011-11-06 doc/pldi2012/circuit.tex
circuit

@@
 
-Line 17 declares the main simulator function, which accepts a \emph{Circuit}
-to be simulated.  In addition to specifying the return type and the types and
-names of the formal parameters, it must also specify which regions of memory
-it will access, how it will access them (e.g. read-only, read-write, or with
-reductions), and what coherence guarantees are needed (e.g. exclusive access).
-The regions listed in these access declarations may be anything in the static
-scope of the function definition, the formal parameters, or fields of the formal
-parameters.  In our example, the simulation of the circuit will need to read
-and write all of the nodes and all of the wires in the circuit, and it must
-be done with exclusive access to ensure a correct result.
+Line 17 declares the main simulator function, which takes a
+\emph{Circuit} to be simulated.  This function specifies the regions it
+will access as well as the {\em permissions} and {\em coherence}
+properties it requires of those regions.  In this case, the {\tt RWE}
+annotation specifies that the function accesses {\tt c.r\_all\_nodes}
+and {\tt c.r\_all\_wires} with read-write permissions and exclusive
+access (the strongest coherence mode).  The simulation of the circuit
+will read and write all of the nodes and all of the wires, and it must
+be done with exclusive access to ensure a correct result.  Functions that
+declare their accessed regions, permissions, and coherence properties are
+called {\em tasks} and will be considered by the Legion runtime for parallel
+execution.
 
@@
 need to provide a \emph{coloring}, which is a relation over the valid elements
-in a region and a set of ``colors''.  Our current implementation uses small
-integers or booleans for the names of colors, but any enumerable set would be
-fine.  The coloring relation need not be total - it can leave out some elements entirely.  And although it is not required to be injective (it can map a single
-element to multiple colors), the fact that injective mappings will always
-result in disjoint subregions makes them strongly encouraged (when possible).
-
-Our example uses four colorings to help create the partitions of the circuit.
+in a region and a set of colors.  
+%Our current implementation uses small
+%integers or booleans for the names of colors, but any enumerable set would be
+%fine.  The coloring relation need not be total - it can leave out some elements entirely.  And although it is not required to be injective (it can map a single
+%element to multiple colors), the fact that injective mappings will always
+%result in disjoint subregions makes them strongly encouraged (when possible).
+Our example uses four colorings to  create circuit partitions.
 A graph partitioning algorithm
-(not shown) is run on to divide the nodes of the graph into \emph{MAX\_PIECES}
-subsets.  (These subsets are hopefully reasonably compact, but the correctness
-of the simulation is not dependent on that.)  The thick lines in \ref{sfig:part_fig:pvs}
-show a how the nodes in a small graph might be partitioned into 3 subsets.
+(not shown) divides nodes of the graph into \emph{MAX\_PIECES}
+subsets.  
+%(These subsets are hopefully reasonably compact, but the correctness
+%of the simulation is not dependent on that.)  
+The thick lines in \ref{sfig:part_fig:pvs}
+show how the nodes in a small graph might be partitioned into 3 subsets.
 Once the subsets are known, the \emph{node\_owner\_map} is created by assigning
-each node a color corresponding to which subset it was placed in.  The 
+each node the color corresponding to its subset.  The 
 \emph{wire\_owner\_map} assigns each wire to the same subset as its ``in\_node''.
@@
 create the subregions owned by each computation.  These are the $p_i$ and $s_i$
-subregions, shown in figures \ref{sfig:part_fig:p_i} and \ref{sfig:part_fig:s_i}.  Finally, the $g_i$ subregions (shown in figure~\ref{sfig:part_fig:g_i}) are
+subregions, shown in Figures \ref{sfig:part_fig:p_i} and \ref{sfig:part_fig:s_i}.  Finally, the $g_i$ subregions (shown in Figure~\ref{sfig:part_fig:g_i}) are
 created using the $node\_nghbr\_map$ to 
@@
 parallel execution, nor is there explicit synchronization required between the
-passes.  As we will see, bBoth the fact that the pieces can be run in parallel
+passes.  As we will see, both the fact that the pieces can be run in parallel
 for a single 

### 2011-11-06 doc/pldi2012/related.tex

@@
 from performance by defining the Mapper interface.  In addition, Chapel's flat array of locales
-makes it challenging to fully utilize deep memory hierarchies.
+makes it challenging to fully utilize deep memory hierarchies.  Chapel currently supports
+clusters and GPUs in isolation \cite{CHAPELGPU}, but does not report results that 
+make use of both.
+
+Sequoia is an example of a locality-aware programming language capable of 
+expressing computations in deep memory hierarchies \cite{Fatahalian06}.  
+Sequoia is a special case
+of the Legion progamming model in which only arrays can be recursively partitioned.
+The tree of array partitions is also required to exactly match the shape of the task tree.  
+Sequoia requires the programmer to statically map arrays and task trees 
+using a machine specific mapping file.
+Legion generalizes the Sequoia model by allowing for partitioning of pointer data structures through 
+regions, decoupling the region tree from the task tree, and enabling dynamic 
+mappings through the mapper interface.  Sequoia is capable of running on
+combinations of many machines including MPI clusters of Cell processors, but
+is unable to use GPUs.
 
@@
 Unlike Legion, regions in X10 are not visible to the programmer but are inferred
-from high level arrays through static analysis.
+from high level arrays through static analysis.  X10 provides support for clusters of GPUs
+\cite{X10GPU}, but requires the programmer to write all code managing data movement
+through both the cluster and GPU memory hierarchies.
 
@@
 operations supported only operate on two-level memory hierarchies consisting of local
-and global memory and do not reason about pointer data structures.  Part of Legion's 
+and global memory and do not reason about pointer data structures.  Neither Titanium
+nor UPC currently support GPUs as part of their language standard \cite{TITANIUMSTANDARD}
+\cite{UPCSTANDARD}.  Part of Legion's 
 low-level runtime system is constructed using UPC's underlying GASNet 

### 2011-11-06 doc/pldi2012/results.tex
rough half-draft of experiments section

\section{Results}

To measure Legion's ability to distribute work across a distributed machine, 
three benchmarks were developed and run on a small cluster under a variety
of configurations.  The cluster consists of four systems connected by a 40~Gb/s
Infiniband fabric.  Each system is powered by 2 Intel Xeon X5680 CPUs, each
with 6 cores, and 2 NVIDIA Tesla C2070 GPUs.  The systems were running Linux
with the standard OFED software stack, with GASnet using its ``ibv'' conduit
to run directly on top.

\subsection{Circuit Simulator}

The circuit simulator benchmark was designed to be representative of the
general class of finite-element simulations on irregular grids.  It relies
on many iterations using small stencils (e.g. just the neighboring nodes) to
propagate changes through the system, and each iteration requires communication
between neighboring nodes.  Careful partitioning of the system can minimize
the amount of information that must be exchanged, but the irregularity of the
underlying graph mandates similar irregularity in the sharing pattern.

The structure of the circuit simulator was presented in section~\ref{sec:code_example}.  Our implementation uses CUDA to perform the computations on the GPU, and Legion
automatically handles the movement of data between the GPUs' memories and the
system memory, but only when necessary.  In fact, with the help of an
application-specific mapper, it is able to keep much of
the data permanently resident in a given GPU for the entirety of the simulation.
Figure~\ref{sfig:results1:perf} shows the relative speedup as the simulation is
spread over multiple nodes in the cluster.  The size of the circuit
being simulated is not changed as the node count increases.  This is therefore
a demonstration of ``strong scaling''.  Adding a second node to the simulation
speeds it up by 77\%, and with four nodes, the simulation is sped up by a total
of 216\% (i.e. a factor of 3.16x).

The bar graph in figure~\ref{sfig:results1:timers} breaks down the simulation's
run time by where it was spent.  The times are summed over all threads, so 
perfect linear scaling would keep the bars flat in this graph.  The total
overhead introduced by Legion is XX \% for the single-node case.
shows a breakdown of the
total amount of time spent

\subsection{Fluid}

To show that Legion remains efficient for regular partitioning cases as well,
we have implemented a particle simulation benchmark, based heavily on the
{\tt fluidanimate} benchmark from the PARSEC suite.  The simulation uses a
cyclic 2D grid of cells instead of PARSEC's 3D grid with boundaries, and 
The fluid benchmark used here differs slightly from the PARSEC version.  It
uses a 2D grid of cells instead of a 3D grid, but the computations being 
performed are preserved.
Although the 
simulation is moving and colliding individual particles, those particles
are placed in cells that make up a regular grid, and the time steps and grid
sizes are chosen carefully to ensure that a particle can only interact with
particles in the same cell or in one immediately adjacent.  The grid is
divided into blocks to be processed by individual threads, with only the
particles in the edge cells being of interest to other threads.  The PARSEC
implementation uses fine-grained locking and the assumption of cache
coherence to access neighboring particles, but neither of these are available
across a cluster.  Our fluid benchmark replaces these accesses with rings of
explicit edge, or \emph{ghost}, cells that are exchanged between adjacent
blocks several times in each iteration.  The capturing of these inter-block
interactions through edge cells and the partitioning of those cells into
subregions shoulds the runtime the ability to schedule data movement in
advance and do a more precise enforcement of ordering constraints (e.g.
requiring a task to wait only for the tasks in the previous pass that computed
one of the ghose cells this task needs).

Figure~\ref{sfig:results2:perf} shows the performance of the fluid benchmark
relative to the serial Legion version.  MORE TEXT HERE.

The bar chart in figure~\ref{sdif:results2:timers} shows MORE STUFF.

\subsection{Tree Search}

The circuit simulation was able to count on a good static, if irregular,
partition of its workload, but that is not possible in some cases.  One
such example arises in ray tracing, in which the components of a scene to be
rendered are partitioned in a binary tree to minimize the number of expensive
object intersection tests that must be performed.  By traversing its path 
through the tree, a ray is able to efficiently skip entire chunks of the 
scene that it is guaranteed not to intersect with.  The trees tend to be
highly imbalanced (often intentionally so), and the set of rays that traverse
the tree tend to have similar path.  Both of these factors contribute to
large fluctuations in the load on any particular part of the tree, complicating
attempts to statically partition the tree.  The {\tt Tree Search} benchmark
explores Legion's ability to load-balance tasks in a locality-aware manner.

The benchmark constructs an irregular binary tree, and then partitions it into
roughly subregions that have roughly equal numbers of nodes.  Partition
crossings are annotated on the tree and traversal tasks are re-spawned when
a ray crosses into another partition.  Load imbalances will occur, but when
they do, the default mapper's ability to provide a set of tasks with good
locality should minimize the amount of data that must be copied around.  To
measure this benefit, each run is also done with the tree partitioning 
disabled.  This denies the mapper the locality information it would need to 
make good choices, reducing it to making essentially random choices.

Figure~\ref{sfig:results3:perf} shows performance of the tree search test
relative to the single node with partitioning case. MORE TEXT HERE.

\begin{figure}
  \centering
  \subfigure[Relative Performance]{
    \begin{tikzpicture}
      \begin{axis}[xlabel=Processors per Node,ylabel=Normalized Speedup,height=6cm]
      \end{axis}
    \end{tikzpicture}
    \label{sfig:results1:perf}
  }
  \subfigure[Execution Time Breakdown]{
  \begin{tikzpicture}
    \begin{axis}[ybar stacked,bar width=5pt,ylabel=Execution Time (in s)]
    \addplot coordinates {
(1, 25.66)
(2, 32.59)
(3, 35.77)
(4, 36.65)
(5, 25.37)
(6, 31.58)
(7, 34.19)
(8, 36.09)
(9, 25.42)
(10, 31.70)
(11, 34.03)
(12, 35.56)
(13, 28.66)
(14, 32.38)
(15, 33.93)
(16, 37.31)
(17, 26.26)
(18, 34.34)
(19, 35.18)
(20, 36.33)
(21, 31.12)
(22, 36.91)
(23, 38.96)
(24, 41.52)
(25, 31.35)
(26, 45.08)
(27, 46.56)
(28, 47.01)
};
    \addplot coordinates {
(1, 0.01)
(2, 0.75)
(3, 0.89)
(4, 0.93)
(5, 0.01)
(6, 0.96)
(7, 1.26)
(8, 1.34)
(9, 0.01)
(10, 1.49)
(11, 1.53)
(12, 1.98)
(13, 0.02)
(14, 1.67)
(15, 1.60)
(16, 2.45)
(17, 0.02)
(18, 1.59)
(19, 1.86)
(20, 2.23)
(21, 0.03)
(22, 1.74)
(23, 2.06)
(24, 2.18)
(25, 0.05)
(26, 2.31)
(27, 2.12)
(28, 2.32)
};
    \addplot coordinates {
(1, 0.59)
(2, 2.09)
(3, 4.30)
(4, 6.02)
(5, 1.31)
(6, 3.83)
(7, 7.17)
(8, 10.96)
(9, 2.08)
(10, 5.24)
(11, 11.38)
(12, 14.28)
(13, 2.59)
(14, 7.40)
(15, 13.45)
(16, 19.04)
(17, 3.66)
(18, 13.03)
(19, 23.48)
(20, 28.21)
(21, 8.68)
(22, 16.42)
(23, 34.34)
(24, 46.12)
(25, 11.02)
(26, 32.69)
(27, 60.49)
(28, 87.13)
};
    \addplot coordinates {
(1, 0.58)
(2, 4.77)
(3, 4.53)
(4, 4.35)
(5, 1.17)
(6, 2.75)
(7, 2.02)
(8, 2.28)
(9, 0.99)
(10, 1.40)
(11, 1.49)
(12, 1.57)
(13, 2.48)
(14, 1.72)
(15, 1.56)
(16, 1.78)
(17, 1.05)
(18, 1.33)
(19, 1.74)
(20, 1.80)
(21, 1.61)
(22, 1.75)
(23, 1.80)
(24, 1.95)
(25, 1.76)
(26, 2.02)
(27, 2.12)
(28, 2.15)
};
    \addplot coordinates {
(1, 0.00)
(2, 0.00)
(3, 0.00)
(4, 0.00)
(5, 0.00)
(6, 0.00)
(7, 0.00)
(8, 0.00)
(9, 0.00)
(10, 0.00)
(11, 0.00)
(12, 0.00)
(13, 0.00)
(14, 0.00)
(15, 0.00)
(16, 0.00)
(17, 0.00)
(18, 0.00)
(19, 0.00)
(20, 0.00)
(21, 0.00)
(22, 0.00)
(23, 0.00)
(24, 0.00)
(25, 0.00)
(26, 0.00)
(27, 0.00)
(28, 0.00)
};
    \addplot coordinates {
(1, 0.00)
(2, 0.03)
(3, 0.04)
(4, 0.05)
(5, 0.00)
(6, 0.04)
(7, 0.06)
(8, 0.08)
(9, 0.00)
(10, 0.14)
(11, 0.11)
(12, 0.19)
(13, 0.00)
(14, 0.57)
(15, 0.27)
(16, 0.47)
(17, 0.00)
(18, 0.34)
(19, 0.61)
(20, 0.52)
(21, 0.00)
(22, 0.64)
(23, 0.63)
(24, 0.96)
(25, 0.00)
(26, 0.78)
(27, 1.06)
(28, 1.49)
};
    \legend{Application,Memory Copies,High-level,Low-level,Mapper,System}
    \end{axis}
  \end{tikzpicture}
    \label{sfig:results1:timers}
  }
  \caption{Circuit Simulator Results}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \begin{axis}[xlabel=Total Processors,ylabel=Normalized Speedup]
    \end{axis}
  \end{tikzpicture}
  \caption{This is a caption}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \begin{axis}[xlabel=Total Processors,ylabel=Normalized Speedup,xmin=0,xmax=24]
    \end{axis}
  \end{tikzpicture}
  \caption{This is a caption}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \begin{loglogaxis}[xlabel=Total Processors,ylabel=Normalized Speedup]
    \end{loglogaxis}
  \end{tikzpicture}
  \caption{This is a caption}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \begin{axis}[ybar stacked,bar width=5pt]
    \addplot coordinates {
(1, 25.66)
(2, 32.59)
(3, 35.77)
(4, 36.65)
(5, 25.37)
(6, 31.58)
(7, 34.19)
(8, 36.09)
(9, 25.42)
(10, 31.70)
(11, 34.03)
(12, 35.56)
(13, 28.66)
(14, 32.38)
(15, 33.93)
(16, 37.31)
(17, 26.26)
(18, 34.34)
(19, 35.18)
(20, 36.33)
(21, 31.12)
(22, 36.91)
(23, 38.96)
(24, 41.52)
(25, 31.35)
(26, 45.08)
(27, 46.56)
(28, 47.01)
};
    \addplot coordinates {
(1, 0.01)
(2, 0.75)
(3, 0.89)
(4, 0.93)
(5, 0.01)
(6, 0.96)
(7, 1.26)
(8, 1.34)
(9, 0.01)
(10, 1.49)
(11, 1.53)
(12, 1.98)
(13, 0.02)
(14, 1.67)
(15, 1.60)
(16, 2.45)
(17, 0.02)
(18, 1.59)
(19, 1.86)
(20, 2.23)
(21, 0.03)
(22, 1.74)
(23, 2.06)
(24, 2.18)
(25, 0.05)
(26, 2.31)
(27, 2.12)
(28, 2.32)
};
    \addplot coordinates {
(1, 0.59)
(2, 2.09)
(3, 4.30)
(4, 6.02)
(5, 1.31)
(6, 3.83)
(7, 7.17)
(8, 10.96)
(9, 2.08)
(10, 5.24)
(11, 11.38)
(12, 14.28)
(13, 2.59)
(14, 7.40)
(15, 13.45)
(16, 19.04)
(17, 3.66)
(18, 13.03)
(19, 23.48)
(20, 28.21)
(21, 8.68)
(22, 16.42)
(23, 34.34)
(24, 46.12)
(25, 11.02)
(26, 32.69)
(27, 60.49)
(28, 87.13)
};
    \addplot coordinates {
(1, 0.58)
(2, 4.77)
(3, 4.53)
(4, 4.35)
(5, 1.17)
(6, 2.75)
(7, 2.02)
(8, 2.28)
(9, 0.99)
(10, 1.40)
(11, 1.49)
(12, 1.57)
(13, 2.48)
(14, 1.72)
(15, 1.56)
(16, 1.78)
(17, 1.05)
(18, 1.33)
(19, 1.74)
(20, 1.80)
(21, 1.61)
(22, 1.75)
(23, 1.80)
(24, 1.95)
(25, 1.76)
(26, 2.02)
(27, 2.12)
(28, 2.15)
};
    \addplot coordinates {
(1, 0.00)
(2, 0.00)
(3, 0.00)
(4, 0.00)
(5, 0.00)
(6, 0.00)
(7, 0.00)
(8, 0.00)
(9, 0.00)
(10, 0.00)
(11, 0.00)
(12, 0.00)
(13, 0.00)
(14, 0.00)
(15, 0.00)
(16, 0.00)
(17, 0.00)
(18, 0.00)
(19, 0.00)
(20, 0.00)
(21, 0.00)
(22, 0.00)
(23, 0.00)
(24, 0.00)
(25, 0.00)
(26, 0.00)
(27, 0.00)
(28, 0.00)
};
    \addplot coordinates {
(1, 0.00)
(2, 0.03)
(3, 0.04)
(4, 0.05)
(5, 0.00)
(6, 0.04)
(7, 0.06)
(8, 0.08)
(9, 0.00)
(10, 0.14)
(11, 0.11)
(12, 0.19)
(13, 0.00)
(14, 0.57)
(15, 0.27)
(16, 0.47)
(17, 0.00)
(18, 0.34)
(19, 0.61)
(20, 0.52)
(21, 0.00)
(22, 0.64)
(23, 0.63)
(24, 0.96)
(25, 0.00)
(26, 0.78)
(27, 1.06)
(28, 1.49)
};
    \legend{Application,Memory Copies,High-level,Low-level,Mapper,System}
    \end{axis}
  \end{tikzpicture}
  \caption{This is a caption}
\end{figure}

### 2011-11-06 doc/pldi2012/circuit.tex
Updating the circuit example section

@@
 
-Lines 1-16 define data structures.  A circuit consists of wires and
-nodes where wires meet.  {\tt Node}s have fields for the simulation
-(e.g. charge, capacitance) and are linked in a circular list.
-Pointers in data structures (e.g. the {\tt next} fields in the list)
-point to a particular region given by the standard {\tt Type@region}
-syntax.  The program relies on different lists being in different
-regions, so the {\tt Node} definition is parameterized on a region
-{\tt rn}.  The {\tt next} field of a $\tt Node \langle rn \rangle$ has
+The circuit simulation is an application that takes a collection of
+circuit elements consisting of wires and nodes where wires meet.  
+The simulation is a three-phase algorithm that repeatedly updates
+currents, distributes charges, and updates voltages for as many
+time steps as the simulation demands.  We first
+describe the partitioning of regions for the simulation and how
+the program exececutes in section \ref{subsec:partitioning}.  
+We then describe the data type declarations 
+and the constraints they enforce on the simulation that enable
+the runtime to efficiently execute the application
+in section \ref{subsec:datatypes}.
+
+\subsection{Regions and Partitions}
+\label{subsec:partitioning}
+
+Line 18 declares the main simulator function, which takes a
+{\tt Circuit} to be simulated.  This function specifies the regions it
+will access as well as the {\em permissions} and {\em coherence}
+properties it requires of those regions.  In this case, the {\tt RWE}
+annotation specifies that the function accesses the regions {\tt c.r\_all\_nodes}
+and {\tt c.r\_all\_wires} with read-write permissions and exclusive
+access (the strongest coherence mode).  The simulation of the circuit
+will read and write all of the nodes and all of the wires, and it must
+be done with exclusive access to ensure a correct result.  Functions that
+declare their accessed regions, permissions, and coherence properties are
+called {\em tasks} and will be considered by the Legion runtime for parallel
+execution.
+
+Lines 19-27 are responsible for partitioning the circuit into {\tt MAX\_PIECES}
+pieces that can be worked on in parallel.  In order to partition a region, we
+need to provide a {\em coloring}, which is a relation over the valid elements
+in a region and a set of colors.  A {\em partition} is an object which given
+a coloring and a parent region, will contain a subregion of the parent region
+for every color with each subregion containing the elements corresponding 
+to the elements colored with the subregions color.
+
+The first step in partitioning the graph
+is to partition it into two sets of nodes consisting of 
+{\em private} and {\em shared} nodes (lines 20-21).  Private nodes
+will only be touched by a single task in a phase, while shared nodes may
+be referenced by multiple tasks in a phase.  The thick lines in figure \ref{sfig:part_fig:pvs}
+show how the nodes in a small graph might be partitioned into 3 subsets.  Gray
+nodes correspond to shared nodes while black nodes are private.  Note that
+this partition is disjoint since each node will only be colored once.
+%Our current implementation uses small
+%integers or booleans for the names of colors, but any enumerable set would be
+%fine.  The coloring relation need not be total - it can leave out some elements entirely.  And although it is not required to be injective (it can map a single
+%element to multiple colors), the fact that injective mappings will always
+%result in disjoint subregions makes them strongly encouraged (when possible).
+%Our example uses four colorings to create circuit partitions.
+
+After partitioning into all the nodes that will be private and shared, we now
+need to partition each of these subregions to generate {\tt MAX\_PIECES}-way
+parallelism.  We first parition the private nodes into each of the disjoint
+{\tt MAX\_PIECES} subregions (lines 23-24).  This partition can be seen in figure
+\ref{sfig:part_fig:p_i}.  We similarly partition the shared subregion to
+generate {\tt MAX\_PIECES} to correspond to the nodes that will be updated
+by a each parallel computation.  This partitioning is also disjoint since
+each node will only ever by updated by one parallel task and can be
+seen in figure \ref{sfig:part_fig:s_i}.  Finally, we partition
+the shared nodes again into the set of ghost nodes that we may want to make
+multiple copies of since they will be read by multiple tasks.  Since
+ghost nodes may be read by multiple parallel tasks, we may have to color
+them multiple times and therefore the ghost partition is aliased
+as can be seen in figure \ref{sfig:part_fig:g_i}.
+
+Figure \ref{sfig:part_fig:tree} shows the final shape of the regions
+and partitions.  The $*$ symbol indicates that a partition is 
+disjoint.
+
+%(These subsets are hopefully reasonably compact, but the correctness
+%of the simulation is not dependent on that.)  
+%Once the subsets are known, the \emph{node\_owner\_map} is created by assigning
+%each node the color corresponding to its subset.  The 
+%\emph{wire\_owner\_map} assigns each wire to the same subset as its ``in\_node''.
+%The \emph{node\_nghbr\_map} maps a node to color(s) of all wires that 
+%connect to it.  Finally, the \emph{node\_sharing\_map} is derived from the
+%\emph{node\_nghbr\_map}, with a node colored ``true'' if any colors other than 
+%its own were used, and ``false'' if the only wires that connect to a node are
+%in the same piece.
+%Figure~\ref{fig:part_fig} shows how the partitions are defined.
+
+%Line 19 uses the \emph{wire\_owner\_map} to partition the wires region of the
+%circuit into a subregion for each piece, but the partitioning of the nodes is
+%more complicated due to the sharing that is necessary between the pieces.
+%First, the \emph{node\_sharing\_map} is used to create two subregions: $p\_nodes\_pvs[false]$ contains all the nodes that are private to some piece (i.e. will
+%never be needed for the computations in any other piece), while $p\_nodes\_pvs[true]$ contains nodes that will be accessed by multiple pieces' computations.
+%Each of these subregions is then partitioned using the $node\_owner\_map$ to
+%create the subregions owned by each computation.  These are the $p_i$ and $s_i$
+%subregions, shown in Figures \ref{sfig:part_fig:p_i} and \ref{sfig:part_fig:s_i}.  Finally, the $g_i$ subregions (shown in Figure~\ref{sfig:part_fig:g_i}) are
+%created using the $node\_nghbr\_map$ to 
+%create subregions that include the ``ghost'' nodes needed to perform each
+%piece's calculations.
+
+With the partitioning operations completed, the various subregions are
+recorded in an array of {\tt CircuitPiece} structures.  Lines 32-33 fill in multiple
+fields of a {\tt CircuitPiece} at once.  The need for such a statement will be described in detail in 
+section \ref{subsec:datatypes}.
+
+%Because a region relationship can have
+%fields with types that refer its own fields' values, it is often not possible
+%to assign fields one at a time without violating the type checking rules.
+%The simultaneous assignment operator asks the type checker only to make sure
+%that the fields in the structure would have the right (self-referential) types
+%after all the fields have been changed.
+
+Lines 35-39 form the bulk of the actual simulation, performing three passes
+over the circuit on each iteration.  For each pass, a for loop is used to 
+spawn a task for each piece of the circuit.  There are no explicit requests for
+parallel execution, nor is there explicit synchronization required between the
+passes.  Both the fact that the pieces can be run in parallel for a single 
+pass and the the required inter-task dependencies are determined automatically
+by the runtime based on the region access annotations on the task declarations.
+We will describe how this process occurs in more detail in section 
+\ref{subsec:cirdependence}.
+
+The declarations for the three subtasks are shown on lines 42-51.  The 
+{\tt calc\_new\_currents} task declares that it will read and write the region containing wires
+for that piece, and will need to read both the region containing nodes in its piece and any
+node that is adjacent to the piece (i.e. the ghost node region).  
+The {\tt distribute\_charge} subtask turns things around, reading the piece's 
+region of wires and updating all the nodes that those wires connect to.  However,
+rather than requesting the ability to read and write the nodes (which would
+require serialization of these tasks for correctness), the task declares that
+it will use reorderable reduction operations and that the coherence requirement
+can be reduced to atomicity rather than exclusive access. The final task is 
+{\tt update\_voltages}, in which each piece is able to update
+voltage information on the nodes contained in regions that it owns (i.e. private and shared)
+in that piece based on the reductions that happened in the previous task.
+
+%Since the wire subregions are known to be disjoint,
+%the write sets of invocations of $calc\_new\_currents$ do not overlap, and can
+%therefore be safely run in parallel.
+
+%As long as the
+%runtime can guarantee to apply the reductions from multiple subtasks safely, it
+%can run the subtasks themselves in parallel.  Each invocation of 
+%$distribute\_charge$ will be delayed until the corresponding invocation of 
+%$calc\_new\_currents$ has completed due to the read-after-write dependency on
+%the corresponding wire subregion.  However, despite the apparent 
+%write-after-read anti-dependency on the ghost node regions, $distribute\_charge$
+%tasks will generally not have to wait on the the completion of the other
+%$calc\_new\_current$ tasks.  If there is sufficient memory available to make
+%two copies of those nodes, the runtime can allow $distribute\_charge$ tasks to
+%start calculating a new version of the nodes while older $calc\_new\_currents$
+%tasks are still referring to the older version, all completely transparently to
+%the application code.
+
+%Again, the disjointness of the $p_i$ and $s_i$
+%node subregions allows the runtime to safely run these tasks in parallel.  In
+%this case, the runtime does wait for the completion of all the tasks in the 
+%previous pass.  The read-after-write dependence on the $p_i$ is a guaranteed
+%conflict, but there is also potential overlap between the $s_i$ subregions
+%being reduced to in the previous pass and the $g_i$ subregions being accessed
+%in this pass.  Although not every pair of $s_i$ and $g_j$ conflict, the
+%runtime knows that they were created from two independent partitioning
+%operations and guarantees correctness by conservatively assuming they might
+%conflict.
+
+\subsection{Data Types and Constraints}
+\label{subsec:datatypes}
+There are two primitive data types for the simulation: nodes
+and wires.  {\tt Node}s are linked in a circular list.
+To reflect that pointers from one {\tt Node} to the next stay
+in the same region we use the {\tt Type@region} syntax (e.g.
+the {\tt next} field on line 1).
+The program relies on different lists being in different
+regions, and the {\tt Node} definition is parameterized on a region
+{\tt rn} to reflect this constraint.  
+The {\tt next} field of a $\tt Node \langle rn \rangle$ has
 type $\tt Node \langle rn \rangle \mbox{\tt @} rn$, which implies that
@@
 which it is electrically connected.  The two {\tt Node}s are in region {\tt rn},
-but they may point to {\tt Node}s in a different region {\tt rn2}.  This is a common pattern in Legion code, and
-arises when one needs to place one constraint on the objects directly
+but they may point to {\tt Node}s in a different region {\tt rn2}.  
+This is a common pattern in Legion code, and
+arises when a constraint needs to be placed on the objects directly
 referenced, and a second (presumably weaker) constraint on objects
@@
 
-Lines 4-9 declare {\tt Circuit}s, which is our first example of a
-\emph{region relationship}.  A region relationship is just a structure
-with named fields, except that any region fields can be used in the
-type declarations.  Thus the region {\tt r\_all\_nodes} contains {\tt
+In cases where we need to create compound data structures with types whose
+region parameters are self-referential we declare a {\em region relationship}.
+A region relationship is a structure with named fields, 
+except that any region fields can be used in the type declarations
+for other fields.  Lines 4-9 declare the region relationship {\tt Circuit}.
+In {\tt Circuit} the region {\tt r\_all\_nodes} contains {\tt
 Node}s that point only into region {\tt r\_all\_nodes}, guaranteeing that
@@
 
-The last definition is for the pieces into which the circuit is partitioned.
+Since a region relationship can have
+fields with types that refer its own fields' values, it is often not possible
+to assign fields one at a time without violating the type checking rules.
+The simultaneous assignment operator asks the type checker only to make sure
+that the fields in the structure would have the right (self-referential) types
+after all the fields have been changed.  Lines 32 and 33 illustrate a simultaneous
+assignment to update a {\tt CircuitPiece} region relationship.
+
+The last type definition is for the pieces into which the circuit is partitioned.
 The original node and wire regions are specified as parameters, and the
 four kinds of subregions that are created are made fields of the
-{\tt CircuitPiece}.  This declaration also illustrates {\em subregion constraints}
+{\tt CircuitPiece} (lines 10-17).  This declaration also illustrates {\em subregion constraints}
 placed on the three region fields stating the region fields must be subregions of the 
-node and wire regions provided as parameters.  Since the simulation code will need to be able to
+node and wire regions provided as parameters (line 16).  Since the simulation code will need to be able to
 iterate over the nodes owned by a {\tt CircuitPiece}, the parameterization of the {\tt rn\_pvt} (private nodes),
@@
 
-Line 17 declares the main simulator function, which takes a
-\emph{Circuit} to be simulated.  This function specifies the regions it
-will access as well as the {\em permissions} and {\em coherence}
-properties it requires of those regions.  In this case, the {\tt RWE}
-annotation specifies that the function accesses {\tt c.r\_all\_nodes}
-and {\tt c.r\_all\_wires} with read-write permissions and exclusive
-access (the strongest coherence mode).  The simulation of the circuit
-will read and write all of the nodes and all of the wires, and it must
-be done with exclusive access to ensure a correct result.  Functions that
-declare their accessed regions, permissions, and coherence properties are
-called {\em tasks} and will be considered by the Legion runtime for parallel
-execution.
-
 %\include{part_fig}
-
-Lines 19-27 are responsible for partitioning the circuit into \emph{MAX\_PIECES}
-pieces that can be worked on in parallel.  In order to partition a region, we
-need to provide a \emph{coloring}, which is a relation over the valid elements
-in a region and a set of colors.  
-%Our current implementation uses small
-%integers or booleans for the names of colors, but any enumerable set would be
-%fine.  The coloring relation need not be total - it can leave out some elements entirely.  And although it is not required to be injective (it can map a single
-%element to multiple colors), the fact that injective mappings will always
-%result in disjoint subregions makes them strongly encouraged (when possible).
-Our example uses four colorings to  create circuit partitions.
-A graph partitioning algorithm
-(not shown) divides nodes of the graph into \emph{MAX\_PIECES}
-subsets.  
-%(These subsets are hopefully reasonably compact, but the correctness
-%of the simulation is not dependent on that.)  
-The thick lines in \ref{sfig:part_fig:pvs}
-show how the nodes in a small graph might be partitioned into 3 subsets.
-Once the subsets are known, the \emph{node\_owner\_map} is created by assigning
-each node the color corresponding to its subset.  The 
-\emph{wire\_owner\_map} assigns each wire to the same subset as its ``in\_node''.
-The \emph{node\_nghbr\_map} maps a node to color(s) of all wires that 
-connect to it.  Finally, the \emph{node\_sharing\_map} is derived from the
-\emph{node\_nghbr\_map}, with a node colored ``true'' if any colors other than 
-its own were used, and ``false'' if the only wires that connect to a node are
-in the same piece.
-Figure~\ref{fig:part_fig} shows how the partitions are defined.
-
-Line 19 uses the \emph{wire\_owner\_map} to partition the wires region of the
-circuit into a subregion for each piece, but the partitioning of the nodes is
-more complicated due to the sharing that is necessary between the pieces.
-First, the \emph{node\_sharing\_map} is used to create two subregions: $p\_nodes\_pvs[false]$ contains all the nodes that are private to some piece (i.e. will
-never be needed for the computations in any other piece), while $p\_nodes\_pvs[true]$ contains nodes that will be accessed by multiple pieces' computations.
-Each of these subregions is then partitioned using the $node\_owner\_map$ to
-create the subregions owned by each computation.  These are the $p_i$ and $s_i$
-subregions, shown in Figures \ref{sfig:part_fig:p_i} and \ref{sfig:part_fig:s_i}.  Finally, the $g_i$ subregions (shown in Figure~\ref{sfig:part_fig:g_i}) are
-created using the $node\_nghbr\_map$ to 
-create subregions that include the ``ghost'' nodes needed to perform each
-piece's calculations.
-
-With the partitioning operations completed, the various subregions are
-recorded in an array of $CircuitPiece$ structures.  The combination of the
-choice of parameters on line 29 with the constraints from lines 11-13 allow 
-this array to hold pieces of the circuit provided to the $simulate\_circuit$
-function, but no others.
-
-Lines 31-32 use a \emph{simultaneous assignment} operation to fill in multiple
-fields of a $CircuitPiece$ at once.  Because a region relationship can have
-fields with types that refer its own fields' values, it is often not possible
-to assign fields one at a time without violating the type checking rules.
-The simultaneous assignment operator asks the type checker only to make sure
-that the fields in the structure would have the right (self-referential) types
-after all the fields have been changed.
-
-Lines 34-38 form the bulk of the actual simulation, performing three passes
-over the circuit on each iteration.  For each pass, a for loop is used to 
-spawn a task for each piece of the circuit.  There are no explicit requests for
-parallel execution, nor is there explicit synchronization required between the
-passes.  As we will see, both the fact that the pieces can be run in parallel
-for a single 
-pass and the the required inter-task dependencies are determined automatically
-by the runtime based on the region access annotations on the task declarations.
-
-The declarations for the three subtasks are shown on lines 41-50.  The 
-$calc\_new\_currents$ task declares that it will read and write the wires
-for that piece, and will need to read both the nodes in its piece and any
-node that is adjacent to the piece (i.e. any node that one of the wires in the
-piece might connect to).  Since the wire subregions are known to be disjoint,
-the write sets of invocations of $calc\_new\_currents$ do not overlap, and can
-therefore be safely run in parallel.
-
-The $distribute\_charge$ subtask turns things around, reading the piece's 
-wires and updating all the nodes that those wires connect to.  However,
-rather than requesting the ability to read and write the nodes (which would
-require serialization of these tasks for correctness), the task declares that
-it will use reorderable reduction operations and that the coherence requirement
-can be reduced to atomicity rather than exclusive access.  As long as the
-runtime can guarantee to apply the reductions from multiple subtasks safely, it
-can run the subtasks themselves in parallel.  Each invocation of 
-$distribute\_charge$ will be delayed until the corresponding invocation of 
-$calc\_new\_currents$ has completed due to the read-after-write dependency on
-the corresponding wire subregion.  However, despite the apparent 
-write-after-read anti-dependency on the ghost node regions, $distribute\_charge$
-tasks will generally not have to wait on the the completion of the other
-$calc\_new\_current$ tasks.  If there is sufficient memory available to make
-two copies of those nodes, the runtime can allow $distribute\_charge$ tasks to
-start calculating a new version of the nodes while older $calc\_new\_currents$
-tasks are still referring to the older version, all completely transparently to
-the application code.
-
-The final task is $update\_voltages$, in which each piece is able to update
-voltage information on the nodes in that piece based on the reductions that
-happened in the previous task.  Again, the disjointness of the $p_i$ and $s_i$
-node subregions allows the runtime to safely run these tasks in parallel.  In
-this case, the runtime does wait for the completion of all the tasks in the 
-previous pass.  The read-after-write dependence on the $p_i$ is a guaranteed
-conflict, but there is also potential overlap between the $s_i$ subregions
-being reduced to in the previous pass and the $g_i$ subregions being accessed
-in this pass.  Although not every pair of $s_i$ and $g_j$ conflict, the
-runtime knows that they were created from two independent partitioning
-operations and guarantees correctness by conservatively assuming they might
-conflict.
 

### 2011-11-06 doc/pldi2012/code_ex.tex
Updating the circuit example section

@@
 region_relation CircuitPiece<rn, rw> {
-  region< Node<rn_pvt+rn_shr> >                                 rn_pvt (#$\prec$# rn), rn_shr (#$\prec$# rn);
-  region< Node<rn> >                                                   rn_ghost (#$\prec$# rn);
-  region< Wire<rn_pvt+rn_shr+rn_ghost,rn,rw_pvt> >  rw_pvt (#$\prec$# rw);
+  region< Node<rn_pvt+rn_shr> >                                 rn_pvt, rn_shr;
+  region< Node<rn> >                                                   rn_ghost;
+  region< Wire<rn_pvt+rn_shr+rn_ghost,rn,rw_pvt> >  rw_pvt;
   Node<rn_pvt+rn_shr>@(rn_pvt+rn_shr)                     first_node;
   Wire<rn_pvt+rn_shr+rn_ghost,rn,rw_pvt>@rw_pvt    first_wire;
+  where rn_pvt $\le$ rn, rn_shr $\le$ rn, rn_ghost $\le$ rn, rw_pvt $\le$ rw;
 };

### 2011-11-06 doc/pldi2012/type_system.tex
intro
Making some minor changes to the type system section

@@
 \section{Type System}
+\label{sec:type}
 
@@
 through conjunctions, but neither negation nor disjunctions are permitted.  For
-an example of how these work, here is the typing rule for reading an element
+an example of how these work, below is the typing rule for reading an element
 from a region:
@@
 This rule tells us that in order to read from a pointer, we must be able to
-provide a region ($r_2$ in this case) for which we possess read privilege that 
+provide a region ($r_2$ in this case) for which we possess a read privilege that 
 is known to either be, or contain, the region to which the pointer is
@@
 constraints that were captured by the region relationship.  These substituted
-constraints are added to the current function's (flow-sensitive) constraints.
+constraints are added to the current task's (flow-sensitive) constraints.
 
@@
 are attemped whenever a variable is written to memory or passed as an argument
-to a function.  Additionally, the simultaneous assignment operation essentially
+to a task.  Additionally, the simultaneous assignment operation essentially
 performs a pack for the result of the assignments and will also perform an
@@
 Region access privileges cannot be stored or packed.  Instead, they follow the
-function call tree, and it is this property that allows the runtime to make
+task call tree, and it is this property that allows the runtime to make
 its scheduling decisions in a distributed way.  The original source of any
@@
 region, but the way in which privileges can be provided to a subtask, or
-returned to a parent task, is described in the function application rule:
+returned to a parent task, is described in the task application rule:
 
@@
 
-In this rule, expression $e_1$ has the type of a Legion function.  In addition
-to the function's input and output types, it specifies the privileges and
-constraints that must exist at the point where the function is called.  It also
+In this rule, expression $e_1$ has the type of a Legion task.  In addition
+to the task's input and output types, it specifies the privileges and
+constraints that must exist at the point where the task is called.  It also
 describes the privileges and constraints that exist as post-conditions of the
-function invocation.  The output type, privileges, and constraints can be
-captured in an existential quantification, allowing a function to return one
+task invocation.  The output type, privileges, and constraints can be
+captured in an existential quantification, allowing a task to return one
 or more regions that were previously unknown to the caller.

### 2011-11-06 doc/pldi2012/highlevel.tex
Some edits to the execution model and high level runtime

@@
 \label{sec:highlevel}
-The high-level runtime is responsible for guaranteeing the semantics 
-of programs written in the Legion programming model while at the same
-time extracting high performance.  This goal is challenging because
-the high-level runtime must be capable of operating on top of many 
-different architectures abstracted by the low-level runtime.  This 
-includes those with very large latencies for transferring data such
-as distributed memory clusters.  To hide these latencies, the high
-level runtime uses a deffered model of execution made possible by the
-event model of the low-level runtime.  By deffering execution, the 
-runtime can hide the latencies of data and task movement and acieve
-good throughput and high performance.
+In this section we describe a high level runtime system
+that is designed to implement the Legion programming
+and execution models while remaining machine agnostic.  
+The high level runtime system operates on top of the 
+low-level runtime which hides machine specific details
+and is described in detail in section \ref{sec:lowlevel}.
 
-However, a deffered execution model presents challenges to implementing
+The biggest challenge to a implementing the high level
+runtime is to avoid the latencies associated with distributed
+memory systems and deep memory hierarchies.  To aid
+in this battle, the high level runtime employs two useful 
+scheduling techniques.  The first useful technique was proved
+in section \ref{sec:exec}: all scheduling decisions for a
+task can be made locally in the context in which the
+task was created.  The second technique is a deferred execution
+model that allows the runtime to hide the latencies associated
+with moving data through deep memory hierarchies.
+
+%The high-level runtime is responsible for guaranteeing the semantics 
+%of programs written in the Legion programming model while at the same
+%time extracting high performance.  This goal is challenging because
+%the high-level runtime must be capable of operating on top of many 
+%different architectures abstracted by the low-level runtime.  This 
+%includes those with very large latencies for transferring data such
+%as distributed memory clusters.  To hide these latencies, the high
+%level runtime uses a deffered model of execution made possible by the
+%event model of the low-level runtime.  By deffering execution, the 
+%runtime can hide the latencies of data and task movement and acieve
+%good throughput and high performance.
+
+A deffered execution model however presents challenges to implementing
 the semantics of the Legion programming model.  Deffered execution
 does not necessarily imply that tasks will be executed in the order in
-which they are scheduled, but only in the order in which low-level
-event dependencies are created.  To orchestrate the execution of Legion
+which they are scheduled, but only in the order in which dependences
+are expressed.  To orchestrate the execution of Legion
 programs in this environment the high-level runtime is architected
-in a manner similar to an out-of-order hardware processor.  There are
-multiple stages to executing a task which allows the runtime to extract
+in a manner similar to an out-of-order hardware processor, but at
+a coarser granularity with task corresponding to instructions and
+regions corresponding to registers.  There are multiple stages to 
+executing a task which allows the runtime to extract
 as much task-level parallelism as possible from an application whithin
-the constraints of the Legion programming model.  We now investigate
-each of these stages in further detail.
+the constraints of the Legion programming model.  To illustrate
+the implementation of the high level runtime we now describe the
+details of how each stage of task execution is performed.
 
@@
 information about both the logical regions that the task will require
-when it is executed as well as the read-write access and coherence properties
+when it is executed as well as the read-write permissions and coherence properties
 for each of the regions.  The runtime uses this information to perform
 dependence detection with each of the previous tasks that have been registered
-in the same parent task.  Dependence detection only needs to be performed at
-the scope of an enclosing task because the semantics of the Legion programming
-model ensure that all tasks can only use regions which are subregions of
-regions that the parent task uses.  This implies that if a child task were
-to depend on a task at a higher level of the task tree, then the
-parent task would also have had a dependence.
+in the same context.  Dependence detection only needs to be performed at
+the scope of an enclosing context as was proved in section \ref{sec:exec}.
 
-Lemma: If a task {\tt t} with ancestor task {\tt p} depends on task {\tt t'}
-a sibling task of {\tt p}, then task {\tt p} must depend on task {\tt t'}.
+We initially describe how dependences are detected from logical regions
+in section \ref{subsec:cohdep}.  We then describe the algorithm that
+is used to perform dependence detection in section \ref{subsec:depdetect}.
+Section \ref{subsec:dataflow} describes how data is versioned correctly
+during deffered execution.  For concreteness we describe how dependence
+analysis is performed for the circuit simulation from section \ref{sec:ex}.
 
-This property of the Legion programming model enables the runtime to only 
-have to perform dependence analysis between tasks which share the same 
-parent task.  By restricting dependence analysis to tasks which share the same
-parent task, the runtime can perform dependence analysis locally and not
-have to be concerned with tasks being created in other parts of the machine.
+%When performing dependence analysis the runtime relies on its knowledge about
+%the relationships between logical regions and partitions.  We first describe
+%the the components to dependence analysis in sections \ref{subsec:regiontree},
+%\ref{subsec:cohdep}, and \ref{absinsts}.  We then describe the algorithm for performing 
+%dependence analysis on this data structures in section \ref{subsec:depdetect}.
 
-When performing dependence analysis the runtime relies on its knowledge about
-the relationships between logical regions and partitions.  We first describe
-the the components to dependence analysis in sections \ref{subsec:regiontree},
-\ref{subsec:cohdep}, and \ref{absinsts}.  We then describe the algorithm for performing 
-dependence analysis on this data structures in section \ref{subsec:depdetect}.
-
-\subsubsection{Region Trees}
-\label{subsec:regiontree}
-To detect dependences between the tasks, the runtime leverages its knowledge
-about logical regions and their partitions.  To create a new logical region
-or partition, the application must invoke the corresponding call in the
-runtime.  As these calls occur, the runtime constructs a data strcture
-called a {\em region tree}.  A region tree describes the relationship between
-logical regions and partitions.  A region tree contains two types of nodes:
-region nodes and partition nodes.  Every region tree is rooted with a region
-node and alternates between region and partition nodes at each corresponding level.
-A region node represents a specific logical region and tracks the set of
-partitions of that logical region.  Similarly a partition node represents 
-a specific partitioning of a logical region and keeps track of the logical regions 
-which are subregions of the partition.
+%\subsubsection{Region Trees}
+%\label{subsec:regiontree}
+%To detect dependences between the tasks, the runtime leverages its knowledge
+%about logical regions and their partitions.  To create a new logical region
+%or partition, the application must invoke the corresponding call in the
+%runtime.  As these calls occur, the runtime constructs a data strcture
+%called a {\em region tree}.  A region tree describes the relationship between
+%logical regions and partitions.  A region tree contains two types of nodes:
+%region nodes and partition nodes.  Every region tree is rooted with a region
+%node and alternates between region and partition nodes at each corresponding level.
+%A region node represents a specific logical region and tracks the set of
+%partitions of that logical region.  Similarly a partition node represents 
+%a specific partitioning of a logical region and keeps track of the logical regions 
+%which are subregions of the partition.
 
@@
 To detect dependencies between any two tasks using a logical region, we leverage
-the information contained in the access and cohrence properties provided by the programmer
-for how the region is to be used.  Dependency checks are performed pair-wise between regions
+the information contained in the permission and cohrence properties provided by the
+task call.  Dependency checks are performed pair-wise between regions
 to determine whether one task depends on the other.  When checking for
-dependencies we first examine the access property.  If both tasks are reads, then 
+dependencies we first examine the permission.  If both tasks are reads, then 
 there is never a dependence.  If at least one of them is a write we use table
 \ref{tab:depsdetect} to determine whether there is a dependence.
-
 \begin{center}
@@
              & Exclusive & Atomic   & Simultaneous & Relaxed \\ \hline 
-Exclusive    & Dependence & Dependence & Dependence & Dependence \\ \hline
-Atomic       & Dependence & Same Instance & Contingent & Contingent \\ \hline
-Simultaneous & Dependence & Contingent & Same Instance & None \\ \hline
-Relaxed      & Dependence & Contingent & None & None \\ \hline
+Exclusive    & Dep & Dep & Dep & Dep \\ \hline
+Atomic       & Dep & Same & Cont & Cont \\ \hline
+Simultaneous & Dep & Cont & Same & None \\ \hline
+Relaxed      & Dep & Cont & None & None \\ \hline
 \end{tabular}
-\caption{Dependence detection cases when at least one access is a write.\label{tab:depsdetect}}
+\caption{Dependence detection cases when at least one access is a write. {\tt Dep}
+indicates a dependence is detected.  {\tt Same} indicates a dependence is detected
+only if the task uses different instances.  {\tt Cont} means that the dependence
+is contingent upon the permissions.  {\tt None} means there will never
+be a conflict.\label{tab:depsdetect}}
 \end{table}
 \end{center}
-In this table, {\em Dependence} indicates that there will always be a dependence between
-regions while {\em None} indicates that there will never be a dependence regardless
-of mapping decisions.  In the case of {\em Same Instance} there is a dependence contingent
-upon a mapping decision.  If both tasks map to the same physical instance (described in 
-more detail in section \ref{subsec:instmang}) then there is no dependence, which if
-they map to different instances there is a dependence as the system will have to generate
-a copy between the two tasks running.
-
+%In this table, {\em Dependence} indicates that there will always be a dependence between
+%regions while {\em None} indicates that there will never be a dependence regardless
+%of mapping decisions.  In the case of {\em Same Instance} there is a dependence contingent
+%upon a mapping decision.  If both tasks map to the same physical instance (described in 
+%more detail in section \ref{subsec:instmang}) then there is no dependence, which if
+%they map to different instances there is a dependence as the system will have to generate
+%a copy between the two tasks running.
 For the case of {\tt Contingent} there are two special cases that we handle.
@@
 
-\subsubsection{Abstract Instances}
-\label{subsec:absinsts}
-For the cases where tasks do not depend, they are allowed to create different physical
-instances of the same logical region.  The decision to do this however does not
-happen until the mapping phase of a task execution.  However, we need to have a place
-holder that is capable of representing all the possible valid physical instances
-from which to choose even if those instances do not exist because the prior
-tasks have not been mapped.  We call these place holders {\tt abstract instances}.
+%\subsubsection{Abstract Instances}
+%\label{subsec:absinsts}
+%For the cases where tasks do not depend, they are allowed to create different physical
+%instances of the same logical region.  The decision to do this however does not
+%happen until the mapping phase of a task execution.  However, we need to have a place
+%holder that is capable of representing all the possible valid physical instances
+%from which to choose even if those instances do not exist because the prior
+%tasks have not been mapped.  We call these place holders {\tt abstract instances}.
 
-Abstract instances represent the set of possible valid physical instances which a
-task will be capable of selecting from or adding to when it is mapped.  The value
-of abstract instances is that they allow us to remember which physical instances
-are valuable while continuing to perform dependence detection on newly registered
-tasks.  In the analogy to out-of-order processors abstract instances allow us
-to perform the same operation as register-renaming of instructions to avoid
-anti-dependences and to remember actual dependences even though our dependence detection
-analysis may run ahead of the actual execution of tasks.
+%Abstract instances represent the set of possible valid physical instances which a
+%task will be capable of selecting from or adding to when it is mapped.  The value
+%of abstract instances is that they allow us to remember which physical instances
+%are valuable while continuing to perform dependence detection on newly registered
+%tasks.  In the analogy to out-of-order processors abstract instances allow us
+%to perform the same operation as register-renaming of instructions to avoid
+%anti-dependences and to remember actual dependences even though our dependence detection
+%analysis may run ahead of the actual execution of tasks.
 
@@
 will perform dependence detection for every region that the task requests with
-prior tasks that have been registered in the same parent task context.  Since
+prior tasks that have been registered in the same context.  Since
 tasks are registered in program order we will always perform dependence detection
 in program order which guarantees correctness even if tasks are executed out-of-order.  
-Performing dependence detection in program order is also annalogous to how an out of order
+Performing dependence detection in program order is annalogous to how an out of order
 processor works.  Out-of-order processors load instructions in order and only after
@@
 model two different logical regions are not necessarily independent from each other.  In
-order to do the dependence detection we rely on the forest of region tree data structures to be
+order to do the dependence detection we rely on the forest of regions to be
 able to detect dependences between tasks which require different logical regions.
@@
 requested region.  The root region is the ancestor region for which the parent task
-has privliges.  Note that the semantics of the Legion programming model state that
-we must always be able to find such a root region or else an error is thrown.
-
-Once we have found the root region for the requested region, we then proceed to walk
-down the region tree from the node corresponding to the root region to the node
+has privliges.  Once we have found the root region for the requested region, we then proceed to walk
+down the region tree from the root region to the node
 for the requested region.  At each node, we determine if any potential dependences exist.
@@
 \label{subsec:dataflow}
-In addition to computing the dependences between tasks, we also have to compute where
-a task is permitted to acquire its data from when it goes to execute.
+In addition to computing the dependences between tasks, we also have to compute the version
+of data that a task is permitted to use when it goes to execute.
 It is insufficient to simply record the producer task as there

### 2011-11-06 doc/pldi2012/experiments.tex
several figures done - references linked up

--- 2011-11-06 doc/pldi2012/results.tex
@@
-\section{Results}
+\section{Experiments}
+\label{sec:exp}
 
@@
 to run directly on top.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-The structure of the circuit simulator was presented in section~\ref{sec:code_example}.  Our implementation uses CUDA to perform the computations on the GPU, and Legion
+The structure of the circuit simulator was presented in section~\ref{sec:ex}.
+Our implementation uses CUDA to perform the computations on the GPU, and Legion
 automatically handles the movement of data between the GPUs' memories and the
@@
 total amount of time spent
+
+\pgfplotsset{every axis legend/.append style={
+  font=\tiny,
+  at={(0.02,0.98)},
+  anchor=north west}}
+
+\begin{figure}
+  \centering
+  \subfigure[Performance Scaling]{
+    \begin{tikzpicture}
+      \begin{axis}[xlabel=Number of Nodes,ylabel=Normalized Speedup,height=5cm,width=8.5cm,xtick={1,2,3,4}]
+        \addplot coordinates {
+% num gpus = 2
+  (1, 1.99)
+  (2, 3.46)
+  (3, 3.44)
+  (4, 5.90)
+        };
+        \addplot coordinates {
+% num gpus = 1
+  (1, 1.00)
+  (2, 1.86)
+  (3, 2.40)
+  (4, 3.41)
+        };
+      \legend{2 GPU/node,1 GPU/node}
+      \end{axis}
+    \end{tikzpicture}
+    \label{sfig:results1:perf}
+  }
+
+  \subtable[Execution Time Breakdown]{
+   \tiny
+    \begin{tabular}{c@{\hspace{2pt}}cD{.}{.}{3}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}}
+\toprule
+GPUs/ & \# \\
+node & nodes & \multicolumn{1}{c}{Wall Clock} & \multicolumn{1}{c}{Application} & \multicolumn{1}{c}{Copies} & \multicolumn{1}{c}{High-level} & \multicolumn{1}{c}{Low-level} & \multicolumn{1}{c}{Mapper} & \multicolumn{1}{c}{System} \\
+\midrule
+\multirow{4}{*}{$1$} & $1$ & 106.892 & 105.387 & 1.578 & 0.003 & 21.624 & 0 & 0\\
+ & $2$ & 57.388 & 105.407 & 10.348 & 0.003 & 15.177 & 0 & 0.001\\
+ & $3$ & 44.552 & 105.41 & 11.935 & 0.003 & 11.695 & 0 & 0.002\\
+ & $4$ & 31.368 & 105.42 & 12.875 & 0.004 & 8.419 & 0 & 0.003\\
+\midrule
+\multirow{4}{*}{$2$} & $1$ & 53.787 & 105.397 & 1.857 & 0.004 & 10.953 & 0 & 0\\
+ & $2$ & 30.923 & 105.415 & 10.716 & 0.004 & 7.937 & 0 & 0.002\\
+ & $3$ & 31.037 & 105.414 & 12.558 & 0.004 & 7.996 & 0 & 0.001\\
+ & $4$ & 18.103 & 105.42 & 13.747 & 0.004 & 4.638 & 0 & 0.003\\
+\bottomrule \\
+    \end{tabular}
+    \label{sfig:results1:timers}
+  }
+  \caption{Circuit Simulator Results}
+\end{figure}
+
+\pgfplotsset{every axis legend/.append style={
+  font=\tiny,
+  at={(0.02,0.98)},
+  anchor=north west}}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-The bar chart in figure~\ref{sdif:results2:timers} shows MORE STUFF.
+The bar chart in figure~\ref{sfig:results2:timers} shows MORE STUFF.
+
+\begin{figure}
+  \centering
+  \subfigure[Performance Scaling]{
+    \begin{tikzpicture}
+      \begin{axis}[xlabel=Number of Nodes,ylabel=Normalized Speedup,height=5cm,width=8.5cm,xtick={1,2,3,4}]
+        \addplot coordinates {
+% 12 CPUs
+  (1, 5.29)
+  (2, 7.34)
+  (3, 7.30)
+  (4, 10.14)
+        };
+        \addplot coordinates {
+% 6 CPUs
+  (1, 4.51)
+  (2, 6.00)
+  (3, 6.57)
+  (4, 8.83)
+        };
+        \addplot coordinates {
+% 4 CPUs
+  (1, 3.36)
+  (2, 4.05)
+  (3, 5.86)
+  (4, 6.68)
+        };
+        \addplot coordinates {
+% 3 CPUs
+  (1, 2.58)
+  (2, 3.55)
+  (3, 4.35)
+  (4, 5.61)
+        };
+        \addplot coordinates {
+% 2 CPUs
+  (1, 1.83)
+  (2, 2.43)
+  (3, 3.48)
+  (4, 4.15)
+        };
+        \addplot coordinates {
+% 1 CPUs
+  (1, 1.00)
+  (2, 1.61)
+  (3, 1.87)
+  (4, 2.33)
+        };
+      \legend{12 CPU/node,6 CPU/node,4 CPU/node,3 CPU/node,2 CPU/node,1 CPU/node}
+      \end{axis}
+    \end{tikzpicture}
+    \label{sfig:results2:perf}
+  }
+
+  \subtable[Execution Time Breakdown]{
+   \tiny
+    \begin{tabular}{c@{\hspace{2pt}}cD{.}{.}{3}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}}
+\toprule
+GPUs/ & \# \\
+node & nodes & \multicolumn{1}{c}{Wall Clock} & \multicolumn{1}{c}{Application} & \multicolumn{1}{c}{Copies} & \multicolumn{1}{c}{High-level} & \multicolumn{1}{c}{Low-level} & \multicolumn{1}{c}{Mapper} & \multicolumn{1}{c}{System} \\
+\midrule
+\multirow{4}{*}{$1$} & $1$ & 40.718 & 38.486 & 0.008 & 0.996 & 0.544 & 0 & 0\\
+ & $2$ & 25.365 & 43.602 & 0.455 & 2.027 & 0.682 & 0 & 0.043\\
+ & $3$ & 21.808 & 49.256 & 0.513 & 2.377 & 0.765 & 0 & 0.06\\
+ & $4$ & 17.476 & 53.181 & 0.501 & 2.912 & 0.743 & 0 & 0.072\\
+\midrule
+\multirow{4}{*}{$2$} & $1$ & 22.289 & 38.55 & 0.009 & 1.694 & 0.68 & 0 & 0\\
+ & $2$ & 16.787 & 51.081 & 0.573 & 2.373 & 0.756 & 0 & 0.056\\
+ & $3$ & 11.706 & 49.502 & 0.733 & 2.661 & 0.749 & 0 & 0.072\\
+ & $4$ & 9.817 & 50.798 & 0.86 & 2.755 & 0.735 & 0 & 0.1\\
+\midrule
+\multirow{4}{*}{$3$} & $1$ & 15.799 & 38.507 & 0.01 & 2.056 & 0.765 & 0 & 0\\
+ & $2$ & 11.462 & 48.698 & 0.951 & 2.682 & 0.819 & 0 & 0.109\\
+ & $3$ & 9.363 & 50.786 & 1.282 & 2.976 & 0.767 & 0 & 0.081\\
+ & $4$ & 7.26 & 51.778 & 1.285 & 3.172 & 0.818 & 0 & 0.117\\
+\midrule
+\multirow{4}{*}{$4$} & $1$ & 12.11 & 39.296 & 0.011 & 2.17 & 0.827 & 0 & 0\\
+ & $2$ & 10.05 & 49.799 & 1.145 & 3.023 & 0.82 & 0 & 0.24\\
+ & $3$ & 6.951 & 51.216 & 1.367 & 3.137 & 0.798 & 0 & 0.12\\
+ & $4$ & 6.093 & 52.183 & 1.216 & 3.317 & 0.782 & 0 & 0.135\\
+\midrule
+\multirow{4}{*}{$6$} & $1$ & 9.031 & 40.265 & 0.013 & 2.536 & 0.969 & 0 & 0\\
+ & $2$ & 6.79 & 48.992 & 0.946 & 3.182 & 0.937 & 0 & 0.145\\
+ & $3$ & 6.199 & 51.904 & 1.061 & 3.752 & 0.909 & 0 & 0.185\\
+ & $4$ & 4.612 & 52.603 & 1.238 & 3.712 & 0.872 & 0 & 0.222\\
+\midrule
+\multirow{4}{*}{$12$} & $1$ & 7.704 & 45.959 & 0.03 & 4.01 & 1.765 & 0 & 0\\
+ & $2$ & 5.547 & 65.432 & 1.019 & 4.762 & 1.328 & 0 & 0.232\\
+ & $3$ & 5.577 & 68.74 & 1.059 & 5.922 & 1.384 & 0 & 0.322\\
+ & $4$ & 4.017 & 67.662 & 1.19 & 4.882 & 1.22 & 0 & 1.485\\
+\bottomrule \\
+    \end{tabular}
+    \label{sfig:results2:timers}
+  }
+
+  \caption{Fluid Results}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
   \centering
-  \subfigure[Relative Performance]{
+  \subfigure[Performance Scaling]{
     \begin{tikzpicture}
-      \begin{axis}[xlabel=Processors per Node,ylabel=Normalized Speedup,height=6cm]
+      \begin{axis}[xlabel=Number of Nodes,ylabel=Normalized Speedup,height=5cm,width=8.5cm,xtick={1,2,3,4}]
+        \addplot coordinates {
+% num gpus = 2
+  (1, 1.99)
+  (2, 3.46)
+  (3, 3.44)
+  (4, 5.90)
+        };
+        \addplot coordinates {
+% num gpus = 1
+  (1, 1.00)
+  (2, 1.86)
+  (3, 2.40)
+  (4, 3.41)
+        };
+      \legend{2 GPU/node,1 GPU/node}
       \end{axis}
     \end{tikzpicture}
-    \label{sfig:results1:perf}
+    \label{sfig:results3:perf}
   }
-  \subfigure[Execution Time Breakdown]{
-  \begin{tikzpicture}
-    \begin{axis}[ybar stacked,bar width=5pt,ylabel=Execution Time (in s)]
-    \addplot coordinates {
-(1, 25.66)
-(2, 32.59)
-(3, 35.77)
-(4, 36.65)
-(5, 25.37)
-(6, 31.58)
-(7, 34.19)
-(8, 36.09)
-(9, 25.42)
-(10, 31.70)
-(11, 34.03)
-(12, 35.56)
-(13, 28.66)
-(14, 32.38)
-(15, 33.93)
-(16, 37.31)
-(17, 26.26)
-(18, 34.34)
-(19, 35.18)
-(20, 36.33)
-(21, 31.12)
-(22, 36.91)
-(23, 38.96)
-(24, 41.52)
-(25, 31.35)
-(26, 45.08)
-(27, 46.56)
-(28, 47.01)
-};
-    \addplot coordinates {
-(1, 0.01)
-(2, 0.75)
-(3, 0.89)
-(4, 0.93)
-(5, 0.01)
-(6, 0.96)
-(7, 1.26)
-(8, 1.34)
-(9, 0.01)
-(10, 1.49)
-(11, 1.53)
-(12, 1.98)
-(13, 0.02)
-(14, 1.67)
-(15, 1.60)
-(16, 2.45)
-(17, 0.02)
-(18, 1.59)
-(19, 1.86)
-(20, 2.23)
-(21, 0.03)
-(22, 1.74)
-(23, 2.06)
-(24, 2.18)
-(25, 0.05)
-(26, 2.31)
-(27, 2.12)
-(28, 2.32)
-};
-    \addplot coordinates {
-(1, 0.59)
-(2, 2.09)
-(3, 4.30)
-(4, 6.02)
-(5, 1.31)
-(6, 3.83)
-(7, 7.17)
-(8, 10.96)
-(9, 2.08)
-(10, 5.24)
-(11, 11.38)
-(12, 14.28)
-(13, 2.59)
-(14, 7.40)
-(15, 13.45)
-(16, 19.04)
-(17, 3.66)
-(18, 13.03)
-(19, 23.48)
-(20, 28.21)
-(21, 8.68)
-(22, 16.42)
-(23, 34.34)
-(24, 46.12)
-(25, 11.02)
-(26, 32.69)
-(27, 60.49)
-(28, 87.13)
-};
-    \addplot coordinates {
-(1, 0.58)
-(2, 4.77)
-(3, 4.53)
-(4, 4.35)
-(5, 1.17)
-(6, 2.75)
-(7, 2.02)
-(8, 2.28)
-(9, 0.99)
-(10, 1.40)
-(11, 1.49)
-(12, 1.57)
-(13, 2.48)
-(14, 1.72)
-(15, 1.56)
-(16, 1.78)
-(17, 1.05)
-(18, 1.33)
-(19, 1.74)
-(20, 1.80)
-(21, 1.61)
-(22, 1.75)
-(23, 1.80)
-(24, 1.95)
-(25, 1.76)
-(26, 2.02)
-(27, 2.12)
-(28, 2.15)
-};
-    \addplot coordinates {
-(1, 0.00)
-(2, 0.00)
-(3, 0.00)
-(4, 0.00)
-(5, 0.00)
-(6, 0.00)
-(7, 0.00)
-(8, 0.00)
-(9, 0.00)
-(10, 0.00)
-(11, 0.00)
-(12, 0.00)
-(13, 0.00)
-(14, 0.00)
-(15, 0.00)
-(16, 0.00)
-(17, 0.00)
-(18, 0.00)
-(19, 0.00)
-(20, 0.00)
-(21, 0.00)
-(22, 0.00)
-(23, 0.00)
-(24, 0.00)
-(25, 0.00)
-(26, 0.00)
-(27, 0.00)
-(28, 0.00)
-};
-    \addplot coordinates {
-(1, 0.00)
-(2, 0.03)
-(3, 0.04)
-(4, 0.05)
-(5, 0.00)
-(6, 0.04)
-(7, 0.06)
-(8, 0.08)
-(9, 0.00)
-(10, 0.14)
-(11, 0.11)
-(12, 0.19)
-(13, 0.00)
-(14, 0.57)
-(15, 0.27)
-(16, 0.47)
-(17, 0.00)
-(18, 0.34)
-(19, 0.61)
-(20, 0.52)
-(21, 0.00)
-(22, 0.64)
-(23, 0.63)
-(24, 0.96)
-(25, 0.00)
-(26, 0.78)
-(27, 1.06)
-(28, 1.49)
-};
-    \legend{Application,Memory Copies,High-level,Low-level,Mapper,System}
-    \end{axis}
-  \end{tikzpicture}
-    \label{sfig:results1:timers}
+
+  \subtable[Execution Time Breakdown]{
+   \tiny
+    \begin{tabular}{c@{\hspace{2pt}}cD{.}{.}{3}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}}
+\toprule
+GPUs/ & \# \\
+node & nodes & \multicolumn{1}{c}{Wall Clock} & \multicolumn{1}{c}{Application} & \multicolumn{1}{c}{Copies} & \multicolumn{1}{c}{High-level} & \multicolumn{1}{c}{Low-level} & \multicolumn{1}{c}{Mapper} & \multicolumn{1}{c}{System} \\
+\midrule
+\multirow{4}{*}{$1$} & $1$ & 106.892 & 105.387 & 1.578 & 0.003 & 21.624 & 0 & 0\\
+ & $2$ & 57.388 & 105.407 & 10.348 & 0.003 & 15.177 & 0 & 0.001\\
+ & $3$ & 44.552 & 105.41 & 11.935 & 0.003 & 11.695 & 0 & 0.002\\
+ & $4$ & 31.368 & 105.42 & 12.875 & 0.004 & 8.419 & 0 & 0.003\\
+\midrule
+\multirow{4}{*}{$2$} & $1$ & 53.787 & 105.397 & 1.857 & 0.004 & 10.953 & 0 & 0\\
+ & $2$ & 30.923 & 105.415 & 10.716 & 0.004 & 7.937 & 0 & 0.002\\
+ & $3$ & 31.037 & 105.414 & 12.558 & 0.004 & 7.996 & 0 & 0.001\\
+ & $4$ & 18.103 & 105.42 & 13.747 & 0.004 & 4.638 & 0 & 0.003\\
+\bottomrule \\
+    \end{tabular}
+    \label{sfig:results3:timers}
   }
-  \caption{Circuit Simulator Results}
+  \caption{Tree Search Results}
 \end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \begin{axis}[xlabel=Total Processors,ylabel=Normalized Speedup]
-    \end{axis}
-  \end{tikzpicture}
-  \caption{This is a caption}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \begin{axis}[xlabel=Total Processors,ylabel=Normalized Speedup,xmin=0,xmax=24]
-    \end{axis}
-  \end{tikzpicture}
-  \caption{This is a caption}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \begin{loglogaxis}[xlabel=Total Processors,ylabel=Normalized Speedup]
-    \end{loglogaxis}
-  \end{tikzpicture}
-  \caption{This is a caption}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \begin{axis}[ybar stacked,bar width=5pt]
-    \addplot coordinates {
-(1, 25.66)
-(2, 32.59)
-(3, 35.77)
-(4, 36.65)
-(5, 25.37)
-(6, 31.58)
-(7, 34.19)
-(8, 36.09)
-(9, 25.42)
-(10, 31.70)
-(11, 34.03)
-(12, 35.56)
-(13, 28.66)
-(14, 32.38)
-(15, 33.93)
-(16, 37.31)
-(17, 26.26)
-(18, 34.34)
-(19, 35.18)
-(20, 36.33)
-(21, 31.12)
-(22, 36.91)
-(23, 38.96)
-(24, 41.52)
-(25, 31.35)
-(26, 45.08)
-(27, 46.56)
-(28, 47.01)
-};
-    \addplot coordinates {
-(1, 0.01)
-(2, 0.75)
-(3, 0.89)
-(4, 0.93)
-(5, 0.01)
-(6, 0.96)
-(7, 1.26)
-(8, 1.34)
-(9, 0.01)
-(10, 1.49)
-(11, 1.53)
-(12, 1.98)
-(13, 0.02)
-(14, 1.67)
-(15, 1.60)
-(16, 2.45)
-(17, 0.02)
-(18, 1.59)
-(19, 1.86)
-(20, 2.23)
-(21, 0.03)
-(22, 1.74)
-(23, 2.06)
-(24, 2.18)
-(25, 0.05)
-(26, 2.31)
-(27, 2.12)
-(28, 2.32)
-};
-    \addplot coordinates {
-(1, 0.59)
-(2, 2.09)
-(3, 4.30)
-(4, 6.02)
-(5, 1.31)
-(6, 3.83)
-(7, 7.17)
-(8, 10.96)
-(9, 2.08)
-(10, 5.24)
-(11, 11.38)
-(12, 14.28)
-(13, 2.59)
-(14, 7.40)
-(15, 13.45)
-(16, 19.04)
-(17, 3.66)
-(18, 13.03)
-(19, 23.48)
-(20, 28.21)
-(21, 8.68)
-(22, 16.42)
-(23, 34.34)
-(24, 46.12)
-(25, 11.02)
-(26, 32.69)
-(27, 60.49)
-(28, 87.13)
-};
-    \addplot coordinates {
-(1, 0.58)
-(2, 4.77)
-(3, 4.53)
-(4, 4.35)
-(5, 1.17)
-(6, 2.75)
-(7, 2.02)
-(8, 2.28)
-(9, 0.99)
-(10, 1.40)
-(11, 1.49)
-(12, 1.57)
-(13, 2.48)
-(14, 1.72)
-(15, 1.56)
-(16, 1.78)
-(17, 1.05)
-(18, 1.33)
-(19, 1.74)
-(20, 1.80)
-(21, 1.61)
-(22, 1.75)
-(23, 1.80)
-(24, 1.95)
-(25, 1.76)
-(26, 2.02)
-(27, 2.12)
-(28, 2.15)
-};
-    \addplot coordinates {
-(1, 0.00)
-(2, 0.00)
-(3, 0.00)
-(4, 0.00)
-(5, 0.00)
-(6, 0.00)
-(7, 0.00)
-(8, 0.00)
-(9, 0.00)
-(10, 0.00)
-(11, 0.00)
-(12, 0.00)
-(13, 0.00)
-(14, 0.00)
-(15, 0.00)
-(16, 0.00)
-(17, 0.00)
-(18, 0.00)
-(19, 0.00)
-(20, 0.00)
-(21, 0.00)
-(22, 0.00)
-(23, 0.00)
-(24, 0.00)
-(25, 0.00)
-(26, 0.00)
-(27, 0.00)
-(28, 0.00)
-};
-    \addplot coordinates {
-(1, 0.00)
-(2, 0.03)
-(3, 0.04)
-(4, 0.05)
-(5, 0.00)
-(6, 0.04)
-(7, 0.06)
-(8, 0.08)
-(9, 0.00)
-(10, 0.14)
-(11, 0.11)
-(12, 0.19)
-(13, 0.00)
-(14, 0.57)
-(15, 0.27)
-(16, 0.47)
-(17, 0.00)
-(18, 0.34)
-(19, 0.61)
-(20, 0.52)
-(21, 0.00)
-(22, 0.64)
-(23, 0.63)
-(24, 0.96)
-(25, 0.00)
-(26, 0.78)
-(27, 1.06)
-(28, 1.49)
-};
-    \legend{Application,Memory Copies,High-level,Low-level,Mapper,System}
-    \end{axis}
-  \end{tikzpicture}
-  \caption{This is a caption}
-\end{figure}

### 2011-11-06 doc/pldi2012/highlevel.tex
several figures done - references linked up

--- 2011-11-04 doc/pldi2012/highlevel.tex
@@
 instances from which sub tasks will be allowed to pull data.  The state of the
-system at this point can be observed in figure \ref{fig:circdep1}.
+system at this point can be observed in figure~\ref{sfig:mapping_fig:cnc}.
 
@@
 The state of the region tree after dependence analysis has been performed for the
-{\tt compute\_new\_current} nodes can be seen in figure \ref{fig:circdep2}.
+{\tt compute\_new\_current} nodes can be seen in figure~\ref{sfig:mapping_fig:dc}.
 
@@
 state of the region tree after dependence analysis for {\tt distribute\_charge} can
-be seen in figure \ref{fig:circdep3}.
+be seen in figure~\ref{sfig:mapping_fig:volt}.
 

### 2011-11-06 doc/pldi2012/mapping_fig.tex
several figures done - references linked up

@@
+\usetikzlibrary{calc}
 \begin{figure}[t]
   \centering
+  \subfigure[mapping of $cnc_0$ task]{
+    \label{sfig:mapping_fig:cnc}
+    \begin{tikzpicture}
+      \partitiontree
+      \node(cnc_p0)[very thick,draw,fill=white] at (0.45,0.30) {\tiny $cnc_0$};
+      \node(cnc_g0)[very thick,draw,fill=white] at (5.45,0.30) {\tiny $cnc_0$};
+      \node(def_top)[very thin,draw,fill=white] at (3.3,4.35) {\tiny ~~---~~};
+      \draw[very thick] ($ (top.south)!.5!(ptp.center) $) circle (2pt);
+      \draw[very thick] ($ (pvsf.north)!.5!(ptf.center) $) circle (2pt);
+      \draw[very thick] ($ (pvsf.250)!.5!(ppp.center) $) circle (2pt);
+      \draw[very thick] ($ (p0.north)!.5!(pp0.center) $) circle (2pt);
+      \draw[very thick] ($ (pvst.north)!.5!(ptt.center) $) circle (2pt);
+      \draw[very thick] ($ (pvst.310)!.5!(pgp.center) $) circle (2pt);
+      \draw[very thick] ($ (g0.north)!.5!(pg0.center) $) circle (2pt);
+      \draw[very thick,->] (def_top) to [bend right=45] (cnc_p0);
+      \draw[very thick,->] (def_top) to [bend left=45] (cnc_g0);
+    \end{tikzpicture}
+  }
   \subfigure[mapping of $dc_0$ task]{
@@
       \partitiontree
+      %\node(cnc_p0)[very thick,draw,fill=white] at (0.45,0.30) {\tiny $\begin{array}{c}\cancel{cnc_0} \\ dc_0\end{array}$};
+      \node(dc_p0)[very thick,draw,fill=white] at (0.45,0.30) 
+{\tiny
+$\begin{array}{@{}c@{}}
+\cancel{cnc_0} \\
+dc_0
+\end{array}$};
+      \node(cnc_p1)[very thin,draw,fill=white] at (1.25,0.30) {\tiny $cnc_1$};
+      \node(cnc_pn)[very thin,draw,fill=white] at (2.25,0.30) {\tiny $cnc_{n-1}$};
+      \node(cnc_g0)[very thin,draw,fill=white] at (5.45,0.30) {\tiny $cnc_0$};
+      \node(dc_g0)[very thick,draw,fill=white] at (5.45,0.75) {\tiny $dc_0$};
+      \draw[very thick,->] (cnc_g0.west) to [bend left=70] (dc_g0.west);
+      \draw[very thick] ($ (cnc_g0.center) - (0.25,0.25) $) -- ++(0.5,0.5);
+      \draw[very thick] ($ (cnc_g0.center) - (0.25,-0.25) $) -- ++(0.5,-0.5);
+      \node(cnc_g1)[very thin,draw,fill=white] at (6.25,0.30) {\tiny $cnc_1$};
+      \node(cnc_gn)[very thin,draw,fill=white] at (7.25,0.30) {\tiny $cnc_{n-1}$};
+      \node(def_top)[very thin,draw,fill=white] at (3.3,4.35) {\tiny ~~---~~};
+      \draw[very thin] ($ (top.south)!.5!(ptp.center) $) circle (2pt);
+      \draw[very thin] ($ (pvsf.north)!.5!(ptf.center) $) circle (2pt);
+      \draw[very thin] ($ (pvsf.250)!.5!(ppp.center) $) circle (2pt);
+      \draw[very thin] ($ (p0.north)!.5!(pp0.center) $) circle (2pt);
+      \draw[very thin] ($ (p1.north)!.5!(pp1.center) $) circle (2pt);
+      \draw[very thin] ($ (pn.north)!.5!(ppn.center) $) circle (2pt);
+      \draw[very thin] ($ (pvst.north)!.5!(ptt.center) $) circle (2pt);
+      \draw[very thin] ($ (pvst.310)!.5!(pgp.center) $) circle (2pt);
+      \draw[very thin] ($ (g0.north)!.5!(pg0.center) $) circle (2pt);
+      \draw[very thin] ($ (g1.north)!.5!(pg1.center) $) circle (2pt);
+      \draw[very thin] ($ (gn.north)!.5!(pgn.center) $) circle (2pt);
     \end{tikzpicture}
@@
       \partitiontree
-    \end{tikzpicture}
-  }
-  \subfigure[mapping completed]{
-    \label{sfig:mapping_fig:done}
-    \begin{tikzpicture}
-      \partitiontree
+      \node(dc_p0)[very thin,draw,fill=white] at (0.45,0.30) {\tiny $dc_0$};
+      \node(dc_p1)[very thin,draw,fill=white] at (1.25,0.30) {\tiny $dc_1$};
+      \node(dc_pn)[very thin,draw,fill=white] at (2.25,0.30) {\tiny $dc_{n-1}$};
+      \node(dc_g0)[very thin,draw,fill=white] at (5.45,0.30) {\tiny $dc_0$};
+      \node(volt_p0)[very thick,draw,fill=white] at (0.45,0.75) {\tiny $volt_0$};
+      \draw[very thick,->] (dc_p0.west) to [bend left=70] (volt_p0.west);
+      \draw[very thick] ($ (dc_p0.center) - (0.25,0.25) $) -- ++(0.5,0.5);
+      \draw[very thick] ($ (dc_p0.center) - (0.25,-0.25) $) -- ++(0.5,-0.5);
+      \draw[very thick] ($ (dc_g0.center) - (0.25,0.25) $) -- ++(0.5,0.5);
+      \draw[very thick] ($ (dc_g0.center) - (0.25,-0.25) $) -- ++(0.5,-0.5);
+      \node(dc_g0)[very thin,draw,fill=white] at (5.45,0.30) {\tiny $dc_0$};
+      \node(dc_g1)[very thin,draw,fill=white] at (6.25,0.30) {\tiny $dc_1$};
+      \node(dc_gn)[very thin,draw,fill=white] at (7.25,0.30) {\tiny $dc_{n-1}$};
+      \draw[very thick] ($ (dc_g0.center) - (0.25,0.25) $) -- ++(0.5,0.5);
+      \draw[very thick] ($ (dc_g0.center) - (0.25,-0.25) $) -- ++(0.5,-0.5);
+      \draw[very thick] ($ (dc_g1.center) - (0.25,0.25) $) -- ++(0.5,0.5);
+      \draw[very thick] ($ (dc_g1.center) - (0.25,-0.25) $) -- ++(0.5,-0.5);
+      \draw[very thick] ($ (dc_gn.center) - (0.25,0.25) $) -- ++(0.5,0.5);
+      \draw[very thick] ($ (dc_gn.center) - (0.25,-0.25) $) -- ++(0.5,-0.5);
+      \node(def_top)[very thin,draw,fill=white] at (3.3,4.35) {\tiny ~~---~~};
+      \draw[very thin] ($ (top.south)!.5!(ptp.center) $) circle (2pt);
+      \draw[very thin] ($ (pvsf.north)!.5!(ptf.center) $) circle (2pt);
+      \draw[very thin] ($ (pvsf.250)!.5!(ppp.center) $) circle (2pt);
+      \draw[very thin] ($ (p0.north)!.5!(pp0.center) $) circle (2pt);
+      \draw[very thin] ($ (p1.north)!.5!(pp1.center) $) circle (2pt);
+      \draw[very thin] ($ (pn.north)!.5!(ppn.center) $) circle (2pt);
+      \draw[very thin] ($ (pvst.north)!.5!(ptt.center) $) circle (2pt);
+      \draw[very thin] ($ (pvst.310)!.5!(pgp.center) $) circle (2pt);
+      \draw[very thin] ($ (g0.north)!.5!(pg0.center) $) circle (2pt);
+      \draw[very thin] ($ (g1.north)!.5!(pg1.center) $) circle (2pt);
+      \draw[very thin] ($ (gn.north)!.5!(pgn.center) $) circle (2pt);
+      \draw[very thick] ($ (g0.north)!.5!(pg0.center) - (0.1,0.1) $) -- ++(0.2,0.2);
+      \draw[very thick] ($ (g0.north)!.5!(pg0.center) - (0.1,-0.1) $) -- ++(0.2,-0.2);
+      \draw[very thick] ($ (g1.north)!.5!(pg1.center) - (0.1,0.1) $) -- ++(0.2,0.2);
+      \draw[very thick] ($ (g1.north)!.5!(pg1.center) - (0.1,-0.1) $) -- ++(0.2,-0.2);
+      \draw[very thick] ($ (gn.north)!.5!(pgn.center) - (0.1,0.1) $) -- ++(0.2,0.2);
+      \draw[very thick] ($ (gn.north)!.5!(pgn.center) - (0.1,-0.1) $) -- ++(0.2,-0.2);
+      \draw[very thick] ($ (pvst.310)!.5!(pgp.center) - (0.1,0.1) $) -- ++(0.2,0.2);
+      \draw[very thick] ($ (pvst.310)!.5!(pgp.center) - (0.1,-0.1) $) -- ++(0.2,-0.2);
+      \draw[very thick] ($ (pvst.230)!.5!(psp.center) $) circle (2pt);
+      \draw[very thick] ($ (s0.north)!.5!(ps0.center) $) circle (2pt);
+      \node(def_pvst)[very thick,draw,fill=white] at (5.3,2.75) {\tiny ~~---~~};
+      \draw[very thick,->] (dc_g0.north) to [bend right=30] (def_pvst.270);
+      \draw[very thick,->] (dc_g1.north) to [bend right=30] (def_pvst.315);
+      \draw[very thick,->] (dc_gn.north) to [bend right=30] (def_pvst.0);
+      \node(volt_s0)[very thick,draw,fill=white] at (3.2,0.30) {\tiny $volt_0$};
+      \draw[very thick,->] (def_pvst.190) to [bend right=40] (volt_s0.130);
     \end{tikzpicture}
@@
   \label{fig:mapping_fig}
-  \caption{Mapping of tasks in region tree -- TODO: draw in little boxes and stuff}
+  \caption{Mapping of tasks in region tree}
 \end{figure}

### 2011-11-06 doc/pldi2012/exec.tex

@@
 
-In this section we present the Legion execution model and show that
-it is correct in that it is guaranteed to preserve the sequential
-execution semantics of programs.  The correctness argument is
+%In this section we present the Legion execution model and show that
+%it is correct in that it is guaranteed to preserve the sequential
+%execution semantics of programs.  The correctness argument is
+%instructive in that the properties that ensure correctness are also
+%exploited in the implementation of the Legion high-level runtime
+%system to achieve high performance and highly distributed scheduling
+%of parallel tasks.
+
+In section \ref{sec:type} we showed that the type system
+only enables privileges to be passed with task calls.  Based on
+this property of privileges we have claimed that our runtime
+will be able to make local scheduling decisions.  In this
+section we present the Legion execution model and show that
+this property of scheduling preserves the sequential execution 
+semantics of a program.  The correctness argument is
 instructive in that the properties that ensure correctness are also

### 2011-11-06 doc/pldi2012/highlevel.tex
Making some edits to the high level runtime section

@@
 
-The biggest challenge to a implementing the high level
+The biggest challenge to implementing the high level
 runtime is to avoid the latencies associated with distributed
 memory systems and deep memory hierarchies.  To aid
-in this battle, the high level runtime employs two useful 
-scheduling techniques.  The first useful technique was proved
+in solving this problem, the high level runtime employs two useful 
+scheduling techniques.  The first technique was proved
 in section \ref{sec:exec}: all scheduling decisions for a
 task can be made locally in the context in which the
-task was created.  The second technique is a deferred execution
-model that allows the runtime to hide the latencies associated
-with moving data through deep memory hierarchies.
+task was created.  The second technique is a {\em deferred execution model} 
+that allows the runtime to hide the latencies associated
+with moving data through deep memory hierarchies.  A
+deffered execution model decouples the issuing of operations
+from when they are actually performed.  This model is supported
+by features in the low-level runtime that are discussed in section \ref{sec:lowlevel}.
 
@@
 in a manner similar to an out-of-order hardware processor, but at
-a coarser granularity with task corresponding to instructions and
+a coarser granularity with tasks corresponding to instructions and
 regions corresponding to registers.  There are multiple stages to 
@@
 information about both the logical regions that the task will require
-when it is executed as well as the read-write permissions and coherence properties
+when it is executed as well as the read-write privileges and coherence properties
 for each of the regions.  The runtime uses this information to perform
@@
 analysis is performed for the circuit simulation from section \ref{sec:ex}.
-
 %When performing dependence analysis the runtime relies on its knowledge about
@@
 %dependence analysis on this data structures in section \ref{subsec:depdetect}.
-
 %\subsubsection{Region Trees}
@@
 %which are subregions of the partition.  
-
-\subsubsection{Dependencies From Coherence Properties}
+\subsubsection{Dependencies From Coherence and Privileges}
 \label{subsec:cohdep}
 To detect dependencies between any two tasks using a logical region, we leverage
-the information contained in the permission and cohrence properties provided by the
+the information contained in the privileges and cohrence properties provided by the
 task call.  Dependency checks are performed pair-wise between regions
 to determine whether one task depends on the other.  When checking for
-dependencies we first examine the permission.  If both tasks are reads, then 
-there is never a dependence.  If at least one of them is a write we use table
+dependencies we first examine the region privileges.  If both tasks have reads privileges, then 
+there is never a dependence.  If at least one task has write privileges we use table
 \ref{tab:depsdetect} to determine whether there is a dependence.
@@
 \end{tabular}
-\caption{Dependence detection cases when at least one access is a write. {\tt Dep}
+\caption{Dependence detection cases when at least one access has a write privilege. {\tt Dep}
 indicates a dependence is detected.  {\tt Same} indicates a dependence is detected
@@
 model two different logical regions are not necessarily independent from each other.  In
-order to do the dependence detection we rely on the forest of regions to be
+order to perform dependence detection we rely on knowledge of the forest of regions to be
 able to detect dependences between tasks which require different logical regions.
@@
 node being visited as well as whether there are any potential dependences farther down
-in the region tree.  We now describe the decision procedure performed at both region
-nodes and partition nodes in the region trees.
+in the region tree.
 
-For region nodes, we begin by checking if there are any tasks actively using the logical
-region which we depend.  We determine this using the coherence dependence analysis described 
-in section \ref{cohdep}.  For each dependence that is found, we record that the previous
-task must have completed prior to the start of the task we are registering.  After performing
+When traversing a region node, we begin by checking if there are any tasks actively using
+the node.  We determine this using the pair-wise dependence analysis described 
+in section \ref{cohdep} to detect dependences between the region we are looking for and the
+region represented by the current node.  For each dependence that is found, we record that the previous
+task must have completed prior to the start of the task we are registering.
+
+After performing
 dependence analysis we then check if there is an open partition which may contain additional
@@
 simply continue the traversal.  However, if the partition we want is not open then we must
-first {\em close} up any open partition and then open the partition we are going to
-traverse.  The act of closing a partition involves traversing the region tree and recording
-all active tasks as dependences.  The reason for this is that we have no disjointness
+first {\em close} up any open partition.   The act of closing a partition 
+involves traversing the region tree and recording
+all active tasks as dependences.  The reason for closing a partition is that we have no disjointness
 information about tasks that use different partitions and therefore to be safe must register
@@
 
-When we reach our target logical region we again perform the coherence dependence analysis.
+When we reach our target logical region we again perform the pair-wise dependence analysis.
 We also close up any open partitions as they might contain tasks which could contain
@@
 
-When traversing partition nodes we perform a similar, but less complicated analysis.  There
+When traversing partition nodes we perform a similar, but less expensive analysis.  There
 are two cases to consider for partition nodes: disjoint partitions and aliased partitions.
 In the case of disjoint partitions, we simply open the subregion that we intend to traverse
-and then traverse it.  This is correct because we know by definition that any two logical
+and then traverse it.  This is correct because we know by definition that any two 
 subregions in a disjoint partition are disjoint and therefore a dependence can never exits
@@
 aliased partition node we maintain a list of tasks active in all of the subregions.  We
-perform a coherence dependence analysis between every pair of tasks regardless of the subregion
+perform a pair-wise dependence analysis between every pair of tasks regardless of the subregion
 they are accessing to see if a dependence could exist.  We do this regardless of the subregion
@@
 
-\subsubsection{Implicit Dataflow Construction}
+\subsubsection{Managing Data Versions}
 \label{subsec:dataflow}
-In addition to computing the dependences between tasks, we also have to compute the version
-of data that a task is permitted to use when it goes to execute.
-It is insufficient to simply record the producer task as there
-may be multiple active tasks for a logical region, all of which may produce different physical
-instances.  In addition, it is likely that the producer tasks have not been mapped
-yet which makes it impossible to know the set of physical instances from which to pull data.
+One aspect of the deffered execution model is that it allows dependence analysis to run
+ahead of the actual execution of tasks.  As a result the runtime must maintain information
+about which {\em physical instances} of a logical region a task is allowed to use when executing.
+A physical instance is a copy of the data in a logical region.  There can be multiple
+physical instances of a logical region.  It is the responsibility of the high level runtime
+to ensure tasks use instances corresponding to the appropriate version of data.
 
-To solve this problem we create the notion of an {\em abstract instance}.  An abstract instance
-is a place holder for the set of physical instances from which a task can choose to pull
-data when it creates its own physical instances before executing.  For each region in
+%In addition to computing the dependences between tasks, we also have to compute the version
+%of data that a task is permitted to use when it goes to execute since the computation.
+%It is insufficient to simply record the producer task as there
+%may be multiple active tasks for a logical region, all of which may produce different physical
+%instances.  In addition, it is likely that the producer tasks have not been mapped
+%yet which makes it impossible to know the set of physical instances from which to pull data.
+
+In an out-of-order processor this problem is solved by register-renaming which maps logical
+registers to a larger set of physical registers.  Our solution employs a similar technique
+that maps logical regions to a larger set of {\em abstract instances}.  An abstract instance
+is a place holder for the set of physical instances which a task will reference.  We call these
+abstract instances because we may not know the actual physical instances until a task's predecessors
+are mapped as described in section \ref{subsec:mapping}.  For each region in
 a task we record the source and destination abstract instances that correspond to the physical
 instances from which a task will read its input data and then place its output data.
-When the task is mapped as described in section \ref{subsec:mapping} we will update these
-abstract instances with the actual physical instances chosen.
+%When the task is mapped as described in section \ref{subsec:mapping} we will update these
+%abstract instances with the actual physical instances chosen.
 
 To determine the source and destination abstract instances of a region, we keep track of 
-the previous abstract instances that were observed on the way to our target region node
-when performing dependence analysis.  These abstract instances describe the set of physical
-instances from which valid data for our task can be chosen when it is mapped.  If a region
-subtree must be closed then all the abstract instances in that subtree are marked closed
-so no additional tasks can register them as sources.  We can also chose to create a new
-abstract instance to replace a currently valid abstract instance to avoid a WAR dependence.
-By creating an abstract instance we will force the mapper to choose a different physical
-instance from the existing ones when mapping is performed.
+the previous valid abstract instances that were observed while traversing the region tree
+during dependence analysis.  Every region node in the region tree that has active tasks
+has a valid abstract instance.  Every time we close a region tree we close all the valid
+abstract instances.  When a node is reopened it must then create a new abstract instance
+to reflect that new physical instances must be created.
+%If a region
+%subtree must be closed then all the abstract instances in that subtree are marked closed
+%so no additional tasks can register them as sources.  We can also chose to create a new
+%abstract instance to replace a currently valid abstract instance to avoid a WAR dependence.
+%By creating an abstract instance we will force the mapper to choose a different physical
+%instance from the existing ones when mapping is performed.
 
@@
 during dependence detection, we can record it before tasks begin executing out of order allowing
-us to guarantee program correctness in the presence of out of order task execution.  Continuing
-the analogy with out-of-order processors, this implicit construction of the dataflow graph
-is annalogous to register renaming which is done in order, but allows the processor to remove
-WAR dependences and execute instructions in parallel that may have had false dependences.
+us to guarantee program correctness.  
+%Continuing
+%the analogy with out-of-order processors, this implicit construction of the dataflow graph
+%is annalogous to register renaming which is done in order, but allows the processor to remove
+%WAR dependences and execute instructions in parallel that may have had false dependences.
 
-\subsubsection{Putting It All Together: Circuit Example}
+\subsubsection{Circuit Execution Example}
 \label{subsec:cirdependence}

### 2011-11-06 doc/pldi2012/circuit.tex
2

@@
 In this section we give an informal overview of Legion through an example.
-Listing~\ref{lst:code_ex} shows Legion code for an electrical
+Listing~\ref{lst:code_ex} shows code for an electrical
 circuit simulation, which illustrates the core features of the programming model.
 
-The circuit simulation is an application that takes a collection of
-circuit elements consisting of wires and nodes where wires meet.  
-The simulation is a three-phase algorithm that repeatedly updates
+The circuit simulation takes a collection of
+of wires and nodes where wires meet.  
+The simulation repeatedly updates
 currents, distributes charges, and updates voltages for as many
 time steps as the simulation demands.  We first
-describe the partitioning of regions for the simulation and how
-the program exececutes in section \ref{subsec:partitioning}.  
-We then describe the data type declarations 
+describe the partitioning of regions in Section~\ref{subsec:partitioning}
+and then describe the data type declarations 
 and the constraints they enforce on the simulation that enable
 the runtime to efficiently execute the application
-in section \ref{subsec:datatypes}.
+in Section~\ref{subsec:datatypes}.
 
@@
 
-Line 18 declares the main simulator function, which takes a
-{\tt Circuit} to be simulated.  This function specifies the regions it
-will access as well as the {\em permissions} and {\em coherence}
-properties it requires of those regions.  In this case, the {\tt RWE}
-annotation specifies that the function accesses the regions {\tt c.r\_all\_nodes}
-and {\tt c.r\_all\_wires} with read-write permissions and exclusive
-access (the strongest coherence mode).  The simulation of the circuit
-will read and write all of the nodes and all of the wires, and it must
-be done with exclusive access to ensure a correct result.  Functions that
-declare their accessed regions, permissions, and coherence properties are
-called {\em tasks} and will be considered by the Legion runtime for parallel
-execution.
+Line 18 declares the main simulator function, which takes a {\tt
+Circuit} to be simulated.  This function specifies the regions it 
+accesses and the {\em privileges} and {\em coherence} it requires of those regions.
+In this case, the {\tt RWE} annotation specifies that the function
+accesses the regions {\tt c.r\_all\_nodes} and {\tt c.r\_all\_wires}
+with read-write privileges and {\em exclusive} coherence (i.e., no other
+function can access the regions concurrently).  The simulation 
+reads and writes all nodes and wires, and it must be done
+with exclusive access to ensure correct results.  Privileges specify what
+the function can do with the regions; coherence specifies what other
+functions can do with the regions concurrently.  Functions that
+declare their accessed regions, privileges, and coherence are called {\em tasks}
+and are the unit of parallel execution in Legion.
 
 Lines 19-27 are responsible for partitioning the circuit into {\tt MAX\_PIECES}
-pieces that can be worked on in parallel.  In order to partition a region, we
-need to provide a {\em coloring}, which is a relation over the valid elements
-in a region and a set of colors.  A {\em partition} is an object which given
-a coloring and a parent region, will contain a subregion of the parent region
-for every color with each subregion containing the elements corresponding 
-to the elements colored with the subregions color.
+pieces that can be worked on in parallel.  To partition a region, we
+provide a {\em coloring}, which is a relation on the elements
+of a region and a set of colors.  A {\em partition} is an object which given
+a coloring and a region $r$ contains for each color $c$ a {\em subregion} of $r$
+with all the elements of $r$ colored $c$.
 
-The first step in partitioning the graph
-is to partition it into two sets of nodes consisting of 
+The region of all nodes in the circuit is first partitioned into 
 {\em private} and {\em shared} nodes (lines 20-21).  Private nodes
-will only be touched by a single task in a phase, while shared nodes may
+are touched by a single task in a phase, while shared nodes may
 be referenced by multiple tasks in a phase.  The thick lines in figure \ref{sfig:part_fig:pvs}
-show how the nodes in a small graph might be partitioned into 3 subsets.  Gray
-nodes correspond to shared nodes while black nodes are private.  Note that
-this partition is disjoint since each node will only be colored once.
+show how the nodes in a small graph might be partitioned into three subsets.  Gray
+nodes are shared and black nodes are private.  Note that
+this partition is disjoint since each node has one color.
 %Our current implementation uses small

### 2011-11-06 doc/pldi2012/experiments.tex
finished first draft of experiments for first two experiments

@@
 with 6 cores, and 2 NVIDIA Tesla C2070 GPUs.  The systems were running Linux
-with the standard OFED software stack, with GASnet using its ``ibv'' conduit
-to run directly on top.
+with GASNet using the {\tt ibv} conduit to interface directly with the standard
+OFED software stack for low-latency access to the Infiniband hardware.
 
@@
 
-The structure of the circuit simulator was presented in section~\ref{sec:ex}.
+The structure of the circuit simulator was presented in Section~\ref{sec:ex}.
 Our implementation uses CUDA to perform the computations on the GPU, and Legion
@@
 a demonstration of ``strong scaling''.  Adding a second node to the simulation
-speeds it up by 77\%, and with four nodes, the simulation is sped up by a total
-of 216\% (i.e. a factor of 3.16x).
+speeds it up by 86\%, and with four nodes, the simulation is sped up by a total
+of 241\% (i.e. a factor of 3.41x).  When enabling the use of both GPUs in each
+node (a decision that requires no changes to the application), the we achieve
+a speedup of 5.90x when using all eight GPUs.
 
-The bar graph in figure~\ref{sfig:results1:timers} breaks down the simulation's
-run time by where it was spent.  The times are summed over all threads, so 
-perfect linear scaling would keep the bars flat in this graph.  The total
-overhead introduced by Legion is XX \% for the single-node case.
-shows a breakdown of the
-total amount of time spent
+The dip in the performance graph for the 6 GPU case is an artifact of the
+application code's decision to partition the circuit into 8 pieces.  The
+current runtime interface does not allow the application to make queries
+regarding the machine configuration, so it is not able to adapt to
+configurations that differ from the initial design target.  The Legion runtime
+will always guarantee correct execution, but performance can suffer if the
+tasks provided by the application do not divide evenly over the machine.
+
+Table\~ref{sfig:results1:timers} breaks down the execution time by where it was
+spent.  The times are summed over all threads, so the aggregate execution time
+will always exceed the elapsed wall clock time.  The time spent in the 
+application code is nearly constant in all configurations, but the time spent
+copying data jumps as soon as the simulation is spread across multiple nodes.
+This is not unexpected, and while the the overhead is responsible for the
+less-than-linear performance scaling, the slow growth in the absolute overhead
+(after the initial jump) suggests that larger clusters would enjoy additional
+speedup, even without increasing the size of the problem being tackled.
 
@@
 To show that Legion remains efficient for regular partitioning cases as well,
-we have implemented a particle simulation benchmark, based heavily on the
-{\tt fluidanimate} benchmark from the PARSEC suite.  The simulation uses a
-cyclic 2D grid of cells instead of PARSEC's 3D grid with boundaries, and 
-The fluid benchmark used here differs slightly from the PARSEC version.  It
-uses a 2D grid of cells instead of a 3D grid, but the computations being 
-performed are preserved.
+we have implemented a particle simulation benchmark based heavily on the
+{\tt fluidanimate} benchmark from the PARSEC \cite{bienia11benchmarking} suite.
+Our simulation uses a cyclic 2D grid of cells instead of PARSEC's 3D grid with
+boundaries, but the computations being performed are preserved.
 Although the 
 simulation is moving and colliding individual particles, those particles
-are placed in cells that make up a regular grid, and the time steps and grid
-sizes are chosen carefully to ensure that a particle can only interact with
-particles in the same cell or in one immediately adjacent.  The grid is
+are placed in cells that make up a regular grid.  The grid is
 divided into blocks to be processed by individual threads, with only the
@@
 implementation uses fine-grained locking and the assumption of cache
-coherence to access neighboring particles, but neither of these are available
-across a cluster.  Our fluid benchmark replaces these accesses with rings of
+coherence to access neighboring particles, but neither of these are conducive
+to performance scaling across a cluster.  Our fluid benchmark replaces these
+accesses with rings of
 explicit edge, or \emph{ghost}, cells that are exchanged between adjacent
-blocks several times in each iteration.  The capturing of these inter-block
+blocks several times per iteration.  The capturing of these inter-block
 interactions through edge cells and the partitioning of those cells into
-subregions shoulds the runtime the ability to schedule data movement in
-advance and do a more precise enforcement of ordering constraints (e.g.
-requiring a task to wait only for the tasks in the previous pass that computed
-one of the ghose cells this task needs).
+subregions provides the runtime the ability to schedule data movement in
+advance and perform a more precise enforcement of the ordering
+constraints necessary for correctness (e.g.
+requiring a task to wait only for the subset of tasks in the previous pass 
+that computed one of the ghost cells being read by this task).
 
-Figure~\ref{sfig:results2:perf} shows the performance of the fluid benchmark
-relative to the serial Legion version.  MORE TEXT HERE.
+Figure~\ref{sfig:results2:perf} shows the throughput of the fluid benchmark when
+run on various subsets of the CPUs and nodes in the cluster.  Performance is
+normalized with respect to the serial execution case.  As in the circuit
+simulation case, the same size problem is used for all runs, and we see 
+speedups of up to 10.14x when using the entire cluster (a total of 48 CPUs).
+The execution time breakdown in Table~\ref{sfig:results2:timers} shows some
+differences in scaling behavior compared to the circuit simulation.  The time
+spent performing copies is much smaller on absolute scale, but shows the same
+large step when leaving a single node, but then relative insensitivity to the
+number of nodes.  This indicates that the runtime is able to send data only to
+the task that needs it rather than having to perform conservative broadcasts of
+data to maintain consistency.
 
-The bar chart in figure~\ref{sfig:results2:timers} shows MORE STUFF.
+A more qualitative difference between the two simulations is that fluid shows
+significant variability in the time spent in the application code, and there
+are two separate issues at work.  First, the runtime is taking advantage of
+the CPU's ability to perform RDMA operations to place some of the shared data
+in the remotely-accessible memory windows provided by GASNet.  Although access
+to this memory is significantly slower than accesses to a node's system memory,
+being able to avoid copies entirely in some cases is a net win.
+
+The second significant jump in application time is when using all 12 CPUs in
+each node.  With application tasks running on all the available CPU cores,
+there are no ``extra'' cores for running the OS, device drivers, and the like.
+For example, GASNet's low-latency implementation counts on a CPU core being
+available for 
+a thread to spin on message arrival notifications.  Allowing that thread to
+sleep (and then wait for the OS rescheduling delay) degrades system performance
+much more than the penalty being observed in our fluid simulation.  Now that
+one can generally count on having many CPU cores in any new system, it might
+make sense to reserve one (or more) in the low-level runtime for background
+tasks like these in order to provide more predictable performance on the CPU
+cores that are exposed to the high-level runtime.
 

### 2011-11-06 doc/pldi2012/highlevel.tex
More modification to the high level runtime section

@@
 \label{sec:highlevel}
-In this section we describe a high level runtime system
+In this Section we describe a high level runtime system
 that is designed to implement the Legion programming
@@
 low-level runtime which hides machine specific details
-and is described in detail in section \ref{sec:lowlevel}.
+and is described in detail in Section \ref{sec:lowlevel}.
 
@@
 scheduling techniques.  The first technique was proved
-in section \ref{sec:exec}: all scheduling decisions for a
+in Section \ref{sec:exec}: all scheduling decisions for a
 task can be made locally in the context in which the
@@
 from when they are actually performed.  This model is supported
-by features in the low-level runtime that are discussed in section \ref{sec:lowlevel}.
+by features in the low-level runtime that are discussed in Section \ref{sec:lowlevel}.
 
@@
 in the same context.  Dependence detection only needs to be performed at
-the scope of an enclosing context as was proved in section \ref{sec:exec}.
+the scope of an enclosing context as was proved in Section \ref{sec:exec}.
 
 We initially describe how dependences are detected from logical regions
-in section \ref{subsec:cohdep}.  We then describe the algorithm that
-is used to perform dependence detection in section \ref{subsec:depdetect}.
+in Section \ref{subsec:cohdep}.  We then describe the algorithm that
+is used to perform dependence detection in Section \ref{subsec:depdetect}.
 Section \ref{subsec:dataflow} describes how data is versioned correctly
 during deffered execution.  For concreteness we describe how dependence
-analysis is performed for the circuit simulation from section \ref{sec:ex}.
+analysis is performed for the circuit simulation from Section \ref{sec:ex}.
 %When performing dependence analysis the runtime relies on its knowledge about
@@
 the node.  We determine this using the pair-wise dependence analysis described 
-in section \ref{cohdep} to detect dependences between the region we are looking for and the
+in Section \ref{cohdep} to detect dependences between the region we are looking for and the
 region represented by the current node.  For each dependence that is found, we record that the previous
@@
 abstract instances because we may not know the actual physical instances until a task's predecessors
-are mapped as described in section \ref{subsec:mapping}.  For each region in
+are mapped as described in Section \ref{subsec:mapping}.  For each region in
 a task we record the source and destination abstract instances that correspond to the physical
@@
 To illustrate concretely how dependence analysis works, we describe the steps for
-performing dependence analysis for running circuit example.  When the {\tt simulate\_circuit}
-task begins executing there are two region trees rooted at the logical
-regions for {\tt all\_nodes} and {\tt all\_wires}.  For conciseness we will only
-show the region tree for {\tt all\_nodes}.  At the start of the task, there exists
+performing dependence analysis for circuit example from Section \ref{sec:ex}.  
+For conciseness we will only show the node region tree.  At the start of the task, there exists
 a single abstract instance at the root node representing the valid set of physical
-instances from which sub tasks will be allowed to pull data.  The state of the
-system at this point can be observed in figure~\ref{sfig:mapping_fig:cnc}.
+instances from the parent task.
 
@@
 up the {\tt p\_nodes\_pvs} partition and then open up the {\tt p\_pvt\_nodes} sub partition
-for each of their subregions.  Since all of the private regions use a different subregion
-of a disjoint partition for the private node accesses, no dependence is detected.  Similarly,
-when all the dependence analysis is performed for the {\tt p\_ghost\_nodes} subregions,
+for each of their subregions.  We use a bold circle on an edge to indicate that it has been
+opened.  Since all of the private regions use a different subregion
+of a disjoint partition for the private node accesses, no dependence is detected.
+
+Similarly, when the dependence analysis is performed for the {\tt p\_ghost\_nodes} subregions,
 each of the necessary partitions are opened.  There is no dependence with any of the private
-nodes since all the ghost nodes traverse a different sub region of the disjoint partition
+nodes since all the ghost nodes traverse a different subregion of the disjoint partition
 {\tt p\_nodes\_pvs}.  There is also no dependence between any tasks even though the
 {\tt p\_ghost\_nodes} partition is aliased because all the ghost accesses are read-only.
-The state of the region tree after dependence analysis has been performed for the
-{\tt compute\_new\_current} nodes can be seen in figure~\ref{sfig:mapping_fig:dc}.
+The state of the region tree after dependence analysis has been performed for the first
+{\tt compute\_new\_current} task can be seen in figure~\ref{sfig:mapping_fig:cnc}.
 
 In the second phase of the computation, all of the {\tt distribute\_charge} tasks
-will be issued.  For these tasks they will first analyze traverse the region tree
-for their subregions containing their private nodes.  The path to these regions are
+will be issued.  For these tasks they will first traverse the region tree
+for the regions containing their private nodes.  The path to these regions are
 already open, but when the dependence analysis arrives at the specific private node subregion it will
@@
 {\tt distribute\_charge} task will be made the new active task as it dominated the
-previous task.  For the ghost node regions, the dependence analysis will discover that the
+previous task.
+
+For the ghost node regions, the dependence analysis will discover that the
 there is a dependence between each of the {\tt distribute\_charge} tasks and all of the
@@
 partition at the same time even if the subregions they are accessing are aliased.  The
-state of the region tree after dependence analysis for {\tt distribute\_charge} can
-be seen in figure~\ref{sfig:mapping_fig:volt}.
+state of the region tree after dependence analysis for first {\tt distribute\_charge} task can
+be seen in figure~\ref{sfig:mapping_fig:dc}.
 
@@
 node ensuring that there is no requirement on the ordering of {\tt update\_voltage}
-tasks.  The state of the region tree after this phase can be seen in figure \ref{fig:circdep4}.
+tasks.  The state of the region tree after the first update voltage task
+has been registered can be seen in figure \ref{sfig:mapping_fig:volt}.
 
@@
 not mapped then the task is placed on the waiting queue.  The task also registers
-itself with the task it depends on so that when the dependent task is mapped it will notify
+itself with the task on which it depends so that when that task is mapped it will notify
 the task on the waiting queue. Tasks are mapped based on the partial ordering created by 
@@
 initially presented in \cite{Fatahalian06} of creating mapping files for specifying how
-tasks and data were placed in a memory hierarchy.  Rather than rely on a static mapping
+tasks and data are placed in a memory hierarchy.  Rather than rely on a static mapping
 we opted for a dynamic interface that could make decisions both based on application specific
 input data as well as the machine specifics both at start-up and throughout the execution
-of an application.  The interface for a mapper is described in section \ref{subsec:mapinter}.
+of an application.  %The interface for a mapper is described in Section \ref{subsec:mapinter}.
 
 \subsubsection{Mapper Interface}
-\label{subsec:mapinter}
-The interface for a mapper is designed to give the programmer control over how data and tasks
-are placed in the machine.  Every mapper must support the following function calls:
+The mapper interface is a collection of eight function calls that allow the runtime to query
+the application for information about how to map tasks are regions.  Mapper calls are made
+from the runtime to ask where physical instances of regions should be place in the memory hierarchy,
+which physical instances should be the source for copy operations, on which processors
+tasks should be run, and how task stealing should be performed.
+%\label{subsec:mapinter}
+%The interface for a mapper is designed to give the programmer control over how data and tasks
+%are placed in the machine.  Every mapper must support the following function calls:
 
-\begin{itemize}
-\item {\tt rank\_initial\_region\_locations} - When a new logical region is created, the runtime
-will ask the mapper to give a ranking of physical memories of where to place the inital instance.
+%\begin{itemize}
+%\item {\tt rank\_initial\_region\_locations} - When a new logical region is created, the runtime
+%will ask the mapper to give a ranking of physical memories of where to place the inital instance.
 
-\item {\tt rank\_initial\_partition\_locations} - When a new partition is created, the mapper
-must provide a ranking of memory locations for each of the new sub partitions.
+%\item {\tt rank\_initial\_partition\_locations} - When a new partition is created, the mapper
+%must provide a ranking of memory locations for each of the new sub partitions.
 
-\item {\tt select\_initial\_processor} - Given a task to be mapped, select a processor to
-send this task to for mapping.
+%\item {\tt select\_initial\_processor} - Given a task to be mapped, select a processor to
+%send this task to for mapping.
 
-\item {\tt target\_task\_steal} - select a processor to attempt task stealing.  This is explained
-in more detail in section \ref{subsec:steal}.
+%\item {\tt target\_task\_steal} - select a processor to attempt task stealing.  This is explained
+%in more detail in Section \ref{subsec:steal}.
 
-\item {\tt permit\_task\_steal} - also covered in section \ref{subsec:steal}
+%\item {\tt permit\_task\_steal} - also covered in Section \ref{subsec:steal}
 
-\item {\tt map\_task\_region} - Given a specific task and the logical region to be mapped, select
-a source physical instance for the data, and provide a ranking of memories in which to create
-or reuse an existing physical instance.
+%\item {\tt map\_task\_region} - Given a specific task and the logical region to be mapped, select
+%a source physical instance for the data, and provide a ranking of memories in which to create
+%or reuse an existing physical instance.
 
-\item {\tt rank\_copy\_targets} - Given a task and a set of valid instances, generate a set of
-rankings for new memories in which to place new phsyical instances.  This operation is used to
-manage the copy operations performed when closing a subtree as they allow the programmer to
-eagerly migrate regions through the memory hierarchy as copy ups are occuring.
+%\item {\tt rank\_copy\_targets} - Given a task and a set of valid instances, generate a set of
+%rankings for new memories in which to place new phsyical instances.  This operation is used to
+%manage the copy operations performed when closing a subtree as they allow the programmer to
+%eagerly migrate regions through the memory hierarchy as copy ups are occuring.
 
-\item {\tt select\_copy\_source} - Given a set of currently valid source instances and destination
-physical instance, select the instance from which a copy should be made.
-\end{itemize}
+%\item {\tt select\_copy\_source} - Given a set of currently valid source instances and destination
+%physical instance, select the instance from which a copy should be made.
+%\end{itemize}
 
@@
 In order to make it easy for programmers to write programs, we provide an initial implementation 
-of the mapper interface that uses simple heuristics to place task and data.  This makes it easy
-for the programmer to get an initial program running, while allowing them to write their own
-mapper to tune for application or machine specific scenarios at a later time.
+of the mapper interface that uses simple heuristics to place tasks and data.  For example, when
+performing task stealing, the default mapper only encourages stealing tasks in groups that use
+the same logical regions, which improves locality by moving tasks that use the same regions
+together.  The default mapper implementation makes it easy for the programmer to get an 
+initial program running.  As the programmer tunes the application, he can write mappers that
+specialize for both application specific and/or machine specific parameters.
 
-Lastly, the runtime supports multiple mappers in use in the same application.  For each task 
+The high level runtime also supports the use of multiple mappers in the same application.  For each task 
 call the programmer can choose a mapper to be used for the runtime calls made inside that
-task.  This allows for the creation of library code which provides their own mappers indpendent
-from any application level mappers that may be created.
+task.  This allows for the creation of library code which provides its own mappers indpendent
+from any application level mappers.
 
@@
 for task stealing.  Our task stealing algorithm is loosely based on the stealing algorithm
-used by Cilk\ref{CILK95} with guidance from the mapper about how stealing should be
+used by Cilk \cite{CILK95} with guidance from the mapper about how stealing should be
 performed.  Tasks are initially placed on the ready queue on the processor in which they
 were registered.  However, a task can be moved to a different processor either by a mapper
-explicitly choosing to relocate it, or by a steal request from another processor.
+explicitly choosing to relocate it, or by a steal request from another processor's mappers.
 
 When a processor notices that it has no work on its ready queue, it will inquire into each
-mapper asking for a target processor to attempt a steal via the {\tt target\_task\_steal}
-call in the mapper interface.  A mapper can either respond with
+mapper asking for a target processor to attempt a steal.  A mapper can either respond with
 a target processor, or indicate that no steal should be attempted.  The runtime will then
-send steal requests to each processor on behalf of the specific mapper.
+send steal requests to each processor on behalf of each specific mapper.
 
@@
 the case where the mapper has better knowledge about where additional work is than the 
-runtime.  Finally, note that this mapping interface is sufficiently powerful to allow a user to
+runtime.  Note that this mapping interface is sufficiently powerful to allow a user to
 disable all stealing if desired, and to map tasks to explicit processors with no interference
@@
 Once task mapping is complete the high level runtime performs the necessary operations
-to be issue the task to the low level runtime.  It first creates any new physical instances
+to issue the task to the low level runtime.  It first creates any new physical instances
 for the regions which the task requires.  It then issues the necessary copies required
@@
 on which the issuing task dependends having finished.  These dependences are all encoded
-using event dependencies in the low level runtime.  Once all the necessary events
-have triggered the task will be run.
+using event dependencies in the low level runtime.  Events are described
+in detail in Section \ref{subsec:events}.  Continuing the analogy of an out-of-order
+processor, the execution phase is equivalent to issuing an instruction to an
+execution unit, and having the execution unit defer the instruction's execution until all its
+operands are ready.  The low-level runtime contains processors that act as execution
+units that will run a task once all of the events required for the task to run
+have occurred.
 
-Once a task starts running, the runtime is asked to create the region accessors for
-the task.  The runtime investigates whether or not fast accessors can be created for
-each of the regions that the task requires.  Fast accessors are described in more
-detail in section \ref{sec:fastaccess}.  If fast accessors can be created for every
-region in a task, then the runtime chooses to execute a task instantiated with fast
-region accessor methods.  However, if any region can not be specialized into a fast
-accessor the runtime defaults to a task instantiation that relies on {\tt get} and
-{\tt put} operations to access regions.
+%Once a task starts running, the runtime is asked to create the region accessors for
+%the task.  The runtime investigates whether or not fast accessors can be created for
+%each of the regions that the task requires.  Fast accessors are described in more
+%detail in Section \ref{sec:fastaccess}.  If fast accessors can be created for every
+%region in a task, then the runtime chooses to execute a task instantiated with fast
+%region accessor methods.  However, if any region can not be specialized into a fast
+%accessor the runtime defaults to a task instantiation that relies on {\tt get} and
+%{\tt put} operations to access regions.
 
-The runtime will also return to the executing task a unique identifier describing the
-context in which the task is executing.  All calls into the runtime require the task
+The runtime also assigns a unique identifier describing the context
+in which the task is executing.  All calls into the runtime require the task
 to pass this context identifier which is what permits the runtime to know the context
-for incoming calls.
+for incoming calls for creating/destroying regions and launching sub-tasks.
 
@@
 \label{subsec:cleanup}
-We define a task to have {\em completed} once the actual code for the task has completed.
-We also define a task to have {\tt terminated} once all of the operations for a task
-have finished.  We can only consider a task to be terminated once all of the regions that
-the task accessed reflect all the changes made by a task and its child tasks.  However,
-even after a task has completed it is possible for child tasks of that task
-not to have been executed or even mapped.  We describe how we handle this scenario in
-section \ref{subsec:restore}.  We also describe how physical instances are garbage
-collected in seciont \ref{subsec:garbage}.
+We define a task to have {\tt completed} once all of the task's regions reflect
+the updated state of the task and all its sub-tasks.  This reflects an underlying assumption 
+of the mapping process that the regions for which a task has priviliges will represent the data produced 
+by that task.  However, this is not inherently true that a physical instance of a task reflects
+this state after the task finishes executing.  If a task has child tasks these tasks may
+still be executing when the parent task finishes the execution phase.
 
-The runtime needs to know the mappings for
-the child tasks as it must issue restoring copy operations so that the set of regions
-that the task operated on are up to date with any data generated by child tasks.
-
-\subsubsection{Restoring Copy Operations}
-\label{subsec:restore}
-An underlying assumption of the mapping process is that the resulting regions that
-a task access will represent the data produced by that task.  However, this may not
-be true if a task has child tasks which also modify subregions of the parent task.
-In order to handle this issue the runtime must compute the necessary copy operations
+In order to handle this scenario the runtime must compute the necessary copy operations
 from child task physical instances back to the parent task's physical instances.  To
-compute these copy operations, the runtime first will wait until all the child tasks
+compute these copy operations, the runtime must first wait until all the child tasks
 have been mapped.  The runtime then issues a close operation to the each of the parent
 task's region trees.  These close operations record all of the necessary copies to
-restore the parent task's regions as well as the child tasks on which these copies
-are contingent.  We call these copy operations at the end of a task {\em restoring copies}.
+restore the parent task's regions.  We call these copy operations at the end of 
+a task {\em restoring copies}.
 The runtime then issues these copies with the required dependences
-and marks the task terminated once the restoring copy operations are complete.
+and marks the task completed once the restoring copy operations are complete.
+This is similar to the write-back phase of an out-of-order processor where store instructions cannot be
+considered complete until the store effects the state of memory.
 
-\subsubsection{Garbage Collecting Physical Instances}
-\label{subsec:garbage}
-In addition to cleaning up task execution, the runtime must also deal with the problem
-of garbage collecting physical instances of regions.  Physical instances must be
-garbage collected since they are not visible to the application and would present
-a correctness problem if deleting instances was an option for the mapper.  
-Garbage collecting physical instances is a challenging problem
-due to both the distributed environment of the machine and the deffered execution
-model.  To solve this problem we employ a two-level reference counting scheme.  We
-reference count both abstract instances and as well as physical instances.
+The last part of the completetion phase is garbage collecting physical instances.  Physical
+instances are not a visible part of the programming model and therefore cannot be managed
+directly by the programmer.  Having the mapper manage them would introduce correctness
+issues.  We use a two-level reference counting scheme on both abstract instances and physical
+instances to know when it is safe to reclaim physical instances.  Due to space constraints
+we omit the details of this algorithm.
 
-An abstract instance maintains a count of the number of tasks that have referenced 
-that abstract instance, but have not yet been mapped.  An abstract instance also tracks
-whether or not it is open or closed.  An abstract instance is open if it is still
-a valid abstract instance in a region tree.  An abstract instance is marked closed
-once it is no longer a valid abstract instance and can therefore no longer be used
-by any new tasks that are registered.
+%Even after a task has completed it is possible for child tasks of that task not to have been 
+%executed or even mapped.  This is similar
+%to the write-back phase in an out-of-order processor where store instructions can only
+%be considered terminated when they finish effecting the state of memory.
 
-When a task is mapped it notifies an abstract instance that it has been used and can
-decrement its reference count.  The task mapping also notifies the abstract instance
-of the specific physical instance that was used and whether or not this use invalidates
-previous physical instances of the abstract instance.  The abstract instance maintains
-a reference count for each physical instance of it that has been made as well as
-the set that are currently valid physical instances to be used for mapping.  When a 
-task terminates, it notifies the abstract instances that it used to decrement the
-reference counts for the task's specific physical instances.
+%The runtime needs to know the mappings for
+%the child tasks as it must issue restoring copy operations so that the set of regions
+%that the task operated on are up to date with any data generated by child tasks.
 
-There are two conditions under which a physical instance can be garbage collected.  The
-first condition is when a physical instance is not a valid physical instance of
-an abstract instance and its reference count is zero.  This condition occurs once no
-more tasks are using a physical instance and it can longer be mapped by any task.  The
-second condition is when an abstract instance is both closed and has a reference count
-of zero as well as the physical instance's reference count being zero.  This handles the
-case when no more tasks can be set to use an abstract instance, there are no more tasks
-to map to an abstract instance, and there are no current tasks using the physical
-instance.
+%\subsubsection{Restoring Copy Operations}
+%\label{subsec:restore}
+%\subsubsection{Garbage Collecting Physical Instances}
+%\label{subsec:garbage}
+%In addition to cleaning up task execution, the runtime must also deal with the problem
+%of garbage collecting physical instances of regions.  Physical instances must be
+%garbage collected since they are not visible to the application and would present
+%a correctness problem if deleting instances was an option for the mapper.  
+%Garbage collecting physical instances is a challenging problem
+%due to both the distributed environment of the machine and the deffered execution
+%model.  To solve this problem we employ a two-level reference counting scheme.  We
+%reference count both abstract instances and as well as physical instances.
+
+%An abstract instance maintains a count of the number of tasks that have referenced 
+%that abstract instance, but have not yet been mapped.  An abstract instance also tracks
+%whether or not it is open or closed.  An abstract instance is open if it is still
+%a valid abstract instance in a region tree.  An abstract instance is marked closed
+%once it is no longer a valid abstract instance and can therefore no longer be used
+%by any new tasks that are registered.
+
+%When a task is mapped it notifies an abstract instance that it has been used and can
+%decrement its reference count.  The task mapping also notifies the abstract instance
+%of the specific physical instance that was used and whether or not this use invalidates
+%previous physical instances of the abstract instance.  The abstract instance maintains
+%a reference count for each physical instance of it that has been made as well as
+%the set that are currently valid physical instances to be used for mapping.  When a 
+%task terminates, it notifies the abstract instances that it used to decrement the
+%reference counts for the task's specific physical instances.
+
+%There are two conditions under which a physical instance can be garbage collected.  The
+%first condition is when a physical instance is not a valid physical instance of
+%an abstract instance and its reference count is zero.  This condition occurs once no
+%more tasks are using a physical instance and it can longer be mapped by any task.  The
+%second condition is when an abstract instance is both closed and has a reference count
+%of zero as well as the physical instance's reference count being zero.  This handles the
+%case when no more tasks can be set to use an abstract instance, there are no more tasks
+%to map to an abstract instance, and there are no current tasks using the physical
+%instance.
 

### 2011-11-07 doc/pldi2012/circuit.tex
circuit

@@
 and then describe the data type declarations 
-and the constraints they enforce on the simulation that enable
-the runtime to efficiently execute the application
+and the constraints they enforce 
 in Section~\ref{subsec:datatypes}.
@@
 pieces that can be worked on in parallel.  To partition a region, we
-provide a {\em coloring}, which is a relation on the elements
+provide a {\em coloring}, which is a relation between the elements
 of a region and a set of colors.  A {\em partition} is an object which given
 a coloring and a region $r$ contains for each color $c$ a {\em subregion} of $r$
-with all the elements of $r$ colored $c$.
+containing all the elements of $r$ colored $c$.
 
-The region of all nodes in the circuit is first partitioned into 
-{\em private} and {\em shared} nodes (lines 20-21).  Private nodes
+Wires are simply partitioned in {\tt MAX\_PIECES} subregions (line 20).  
+The nodes are first partitioned into 
+{\em private} and {\em shared} nodes (line 21).  Private nodes
 are touched by a single task in a phase, while shared nodes may
@@
 
-After partitioning into all the nodes that will be private and shared, we now
-need to partition each of these subregions to generate {\tt MAX\_PIECES}-way
-parallelism.  We first parition the private nodes into each of the disjoint
-{\tt MAX\_PIECES} subregions (lines 23-24).  This partition can be seen in figure
-\ref{sfig:part_fig:p_i}.  We similarly partition the shared subregion to
-generate {\tt MAX\_PIECES} to correspond to the nodes that will be updated
-by a each parallel computation.  This partitioning is also disjoint since
-each node will only ever by updated by one parallel task and can be
-seen in figure \ref{sfig:part_fig:s_i}.  Finally, we partition
-the shared nodes again into the set of ghost nodes that we may want to make
-multiple copies of since they will be read by multiple tasks.  Since
-ghost nodes may be read by multiple parallel tasks, we may have to color
-them multiple times and therefore the ghost partition is aliased
-as can be seen in figure \ref{sfig:part_fig:g_i}.
-
-Figure \ref{sfig:part_fig:tree} shows the final shape of the regions
-and partitions.  The $*$ symbol indicates that a partition is 
+We next further partition the private and shared subregions into {\tt
+MAX\_PIECES} disjoint subregions each (lines 23-24); one subregion of the private and shared node partitions
+is shown in Figures~\ref{sfig:part_fig:p_i} and \ref{sfig:part_fig:s_i} respectively.  
+We partition the shared nodes again
+into the sets of ghost nodes for each task.  Since ghost nodes may be read by multiple tasks,
+nodes may be colored multiple times and the ghost regions are not disjoint; one of the ghost subregions is shown in Figure~\ref{sfig:part_fig:g_i}.
+Figure~\ref{sfig:part_fig:tree} shows the final hierarchy of node partitions.  The $*$ symbol indicates a partition is 
 disjoint.
@@
 
-With the partitioning operations completed, the various subregions are
-recorded in an array of {\tt CircuitPiece} structures.  Lines 32-33 fill in multiple
-fields of a {\tt CircuitPiece} at once.  The need for such a statement will be described in detail in 
-section \ref{subsec:datatypes}.
+%With the partitioning operations completed, the various subregions are
+%recorded in an array of {\tt CircuitPiece} structures.  Lines 32-33 fill in multiple
+%fields of a {\tt CircuitPiece} at once.  The need for such a statement will be described in detail in 
+%section \ref{subsec:datatypes}.
 
@@
 Lines 35-39 form the bulk of the actual simulation, performing three passes
-over the circuit on each iteration.  For each pass, a for loop is used to 
-spawn a task for each piece of the circuit.  There are no explicit requests for
-parallel execution, nor is there explicit synchronization required between the
-passes.  Both the fact that the pieces can be run in parallel for a single 
-pass and the the required inter-task dependencies are determined automatically
-by the runtime based on the region access annotations on the task declarations.
-We will describe how this process occurs in more detail in section 
-\ref{subsec:cirdependence}.
+over the circuit in each iteration.  Each pass loops over an array of pieces (constructed on lines 30-33 from the partitions),
+spawning a task for each piece.   There are no explicit requests for
+parallel execution, nor is there synchronization required between the
+passes.  Both the fact that the tasks can be run in parallel within a pass
+and the required inter-pass dependencies are determined automatically
+by the Legion runtime based on the region access annotations on the task declarations
+(see Section~\ref{subsec:cirdependence}).
 
-The declarations for the three subtasks are shown on lines 42-51.  The 
-{\tt calc\_new\_currents} task declares that it will read and write the region containing wires
-for that piece, and will need to read both the region containing nodes in its piece and any
-node that is adjacent to the piece (i.e. the ghost node region).  
+%The declarations for the three subtasks are shown on lines 42-51.  
+The  {\tt calc\_new\_currents} task reads and writes the wires subregion, 
+the nodes subregion, and the ghost node region for its piece.
 The {\tt distribute\_charge} subtask turns things around, reading the piece's 
-region of wires and updating all the nodes that those wires connect to.  However,
+wires subregion and updating all nodes those wires touch.  However,
 rather than requesting the ability to read and write the nodes (which would
 require serialization of these tasks for correctness), the task declares that
-it will use reorderable reduction operations and that the coherence requirement
-can be reduced to atomicity rather than exclusive access. The final task is 
-{\tt update\_voltages}, in which each piece is able to update
-voltage information on the nodes contained in regions that it owns (i.e. private and shared)
-in that piece based on the reductions that happened in the previous task.
+it will use reorderable {\em reduction} operations and that the coherence requirement
+is {\em atomic} rather than exclusive access. The final task is 
+{\tt update\_voltages}, which writes the shared and private nodes for a piece
+and reads the results of the previous task's reductions.
 
@@
 \label{subsec:datatypes}
-There are two primitive data types for the simulation: nodes
-and wires.  {\tt Node}s are linked in a circular list.
-To reflect that pointers from one {\tt Node} to the next stay
-in the same region we use the {\tt Type@region} syntax (e.g.
-the {\tt next} field on line 1).
-The program relies on different lists being in different
-regions, and the {\tt Node} definition is parameterized on a region
-{\tt rn} to reflect this constraint.  
-The {\tt next} field of a $\tt Node \langle rn \rangle$ has
-type $\tt Node \langle rn \rangle \mbox{\tt @} rn$, which implies that
-all nodes in the list are in region {\tt rn}.
+The two basic data types are {\tt Node}s and {\tt Wire}s.
+{\tt Node}s are linked in circular lists.
+The simulation relies on different lists being in different
+regions, so the {\tt Node} definition on line 1 is parameterized by a region
+{\tt rn}.  The {\tt next} field of a $\tt Node \langle rn \rangle$ has
+type $\tt Node \langle rn \rangle \mbox{\tt @} rn$, which means the field points
+to an object in region {\tt rn} of type $\tt Node \langle rn \rangle$, 
+implying that all nodes of the list are in region {\tt rn}.
 
@@
 all {\tt Node}s are in region {\tt r\_all\_nodes}.  (Note that regions are
-declared with the single data type they can contain.)  Simiarly, the
+declared with the single data type they can contain.)  Similarly, the
 type of the wire region's element not only guarantees that every wire
@@
 Since a region relationship can have
-fields with types that refer its own fields' values, it is often not possible
+fields with types that refer its own fields' regions, it is often not possible
 to assign fields one at a time without violating the type checking rules.
-The simultaneous assignment operator asks the type checker only to make sure
-that the fields in the structure would have the right (self-referential) types
-after all the fields have been changed.  Lines 32 and 33 illustrate a simultaneous
+The simultaneous assignment operator asks the type checker only to ensure
+that fields have the correct types after all the fields are updated.  
+Lines 32 and 33 illustrate a simultaneous
 assignment to update a {\tt CircuitPiece} region relationship.
@@
 placed on the three region fields stating the region fields must be subregions of the 
-node and wire regions provided as parameters (line 16).  Since the simulation code will need to be able to
-iterate over the nodes owned by a {\tt CircuitPiece}, the parameterization of the {\tt rn\_pvt} (private nodes),
+node and wire regions provided as parameters (line 16).  Since the simulation 
+iterates over the nodes owned by a {\tt CircuitPiece}, the parameterization of the {\tt rn\_pvt} (private nodes),
 {\tt rn\_shr} (shared nodes), and {\tt first\_node} fields constrains the linked list of
@@
 similarly constrained, and the wire's node pointers are constrained to fall
-into one of the three subregions defined in the {\tt CircuitPiece}.  Finally,
+into one of the three subregions.
+% defined in the {\tt CircuitPiece}.  
+Finally,
 the use of {\tt rn} for {\tt rn\_ghost}'s parameter and the second parameter

### 2011-11-07 doc/pldi2012/experiments.tex

@@
   at={(0.02,0.98)},
-  anchor=north west}}
+  anchor=north west},
+  cycle list={{black, mark=x}, {black, mark=+},
+              {black, mark=o}, {black, mark=diamond},
+              {black, mark=triangle}, {black, mark=star}}};
 
@@
 
+\usepgflibrary{plotmarks}
 \begin{figure}
@@
     \begin{tikzpicture}
-      \begin{axis}[xlabel=Number of Nodes,ylabel=Normalized Speedup,height=5cm,width=8.5cm,xtick={1,2,3,4}]
+      \begin{axis}[
+          xlabel=Number of Nodes,
+          ylabel=Normalized Speedup,
+          height=5cm,
+          width=8.5cm,
+          xtick={1,2,3,4},
+          legend columns=3,
+        ]
         \addplot coordinates {

### 2011-11-07 doc/pldi2012/highlevel.tex

@@
 \label{sec:highlevel}
-In this Section we describe a high level runtime system
+In this Section we describe a high-level runtime system
 that is designed to implement the Legion programming
 and execution models while remaining machine agnostic.  
-The high level runtime system operates on top of the 
+The high-level runtime system operates on top of the 
 low-level runtime which hides machine specific details
@@
 
-The biggest challenge to implementing the high level
+The biggest challenge to implementing the high-level
 runtime is to avoid the latencies associated with distributed
 memory systems and deep memory hierarchies.  To aid
-in solving this problem, the high level runtime employs two useful 
+in solving this problem, the high-level runtime employs two useful 
 scheduling techniques.  The first technique was proved
@@
 the constraints of the Legion programming model.  To illustrate
-the implementation of the high level runtime we now describe the
+the implementation of the high-level runtime we now describe the
 details of how each stage of task execution is performed.
@@
 The first step in the execution of a task is dependence analysis.  When
-a task is registered with the high level runtime the task call includes
+a task is registered with the high-level runtime the task call includes
 information about both the logical regions that the task will require
@@
 out of order.  The regions that a task requires are annalogous
-to the registers listed in an instruction and the goal of the high level
+to the registers listed in an instruction and the goal of the high-level
 runtime is to detect the same dependences and use them to enforce scheudling constraints
@@
 A physical instance is a copy of the data in a logical region.  There can be multiple
-physical instances of a logical region.  It is the responsibility of the high level runtime
+physical instances of a logical region.  It is the responsibility of the high-level runtime
 to ensure tasks use instances corresponding to the appropriate version of data.
@@
 complete control over data and task movement, it will never allow the programmer to affect the
-correctness of the program.  The highlevel runtime still manages all the task dependences and the
+correctness of the program.  The high-level runtime still manages all the task dependences and the
 ordering in which tasks are executed and copies are performed.  Therefore, regardless of choices
@@
 
-The high level runtime also supports the use of multiple mappers in the same application.  For each task 
+The high-level runtime also supports the use of multiple mappers in the same application.  For each task 
 call the programmer can choose a mapper to be used for the runtime calls made inside that
@@
 \label{subsec:execution}
-Once task mapping is complete the high level runtime performs the necessary operations
+Once task mapping is complete the high-level runtime performs the necessary operations
 to issue the task to the low level runtime.  It first creates any new physical instances

### 2011-11-07 doc/pldi2012/lowlevel.tex
intro
More modification to the high level runtime section

@@
 \section{Low-Level Runtime}
+\label{sec:lowlevel}
 
-The job of the low-level runtime is to provide a machine-independent layer
-for the high-level runtime to sit on top of.  It needs to hide the
-ideosynchracies of the underlying machine, operating system, and device
-drivers.  At the same time, it needs to avoid getting too far away from the 
-native capabilities of the hardware and software in order to obtain an 
-efficient implementation.
+The purpose of the low-level runtime is to provide a machine-independent layer
+for the high-level runtime.  This abstraction will provide portability to the Legion
+programming model and the high-level runtime.  To achieve this goal the low-level
+runtime needs to hide the ideosynchracies of the underlying machine.  At the same time, it needs to avoid 
+straying too far from the native capabilities of the underlying hardware 
+and software in order to obtain an efficient implementation.  The Legion
+low-level runtime does this by abstracting the machine into a base set
+of objects required to implement the Legion high-level.  These objects,
+include processors, memories, regions, events, and deffered locks.
 
-The target machines for Legion include both clusters and GPU-equipped nodes,
-and neither of these are able to efficiently provide a uniformly addressable
-and accessible memory space across the system.  The magic of an
-apparently-unified address space is left to the high-level runtime, and the
-interface to the low-level runtime is done in terms of handles rather than 
-pointers for runtime objects and relative offsets rather than absolute 
-pointers for resource allocations.
+%The target machines for Legion include both clusters and GPU-equipped nodes,
+%and neither of which are able to efficiently provide a uniformly addressable
+%and accessible memory space across the system.  The low-level runtime does
+%not hide this fact from the high-level runtime, but rather relies on the
+%high-level runtime to manage data movement.  Consequently all objects in
+%the low-level runtime are named by handles 
+%which can be copied by value and moved around the machine.
 
-The low-level runtime chooses to provide a ``deferred execution'' model rather
-than the more common ``immediate'' model.  Requests made of the low-level 
-runtime usually return to the caller immediately, with the requested action
-performing asynchronously to the caller's thread of execution.  This is done
-for two different, but equally important, reasons.  First, it matches the
-standard GPU execution model.  GPUs tend to have very deep pipelines, and
-maximum performance cannot be achieved if each command has to be run to
-completion before the next command can be started.  Second, in clusters, the
-latency of inter-node communication can be significant, and a caller that
-waited for every request of a remote node to return before continuing on would
-also be similarly underutilizing the computational resources of the system.
+%The low-level runtime provides a deferred execution model rather
+%than the more common immediate model where operations occur as soon
+%as requested.  Requests made of the low-level 
+%runtime usually return to the caller immediately, with the requested action
+%performing asynchronously to the caller's thread of execution.  This is done
+%for two different, but equally important, reasons.  First, it matches the
+%standard GPU execution model.  GPUs tend to have very deep pipelines, and
+%maximum performance cannot be achieved if each command has to be run to
+%completion before the next command can be started.  Second, in clusters, the
+%latency of inter-node communication can be significant, and a caller that
+%waited for every request of a remote node to return before continuing on would
+%also be similarly underutilizing the computational resources of the system.
 
-Although it does not show up explicitly in the interface, the low-level runtime
-was also architected to permit a distributed and hierarchical implementation.
-Critical resources can be replicated or migrated between nodes to reduce
-both average latencies and the amount of traffic that use the scarce inter-node
-communication bandwidth.
+%Although it does not show up explicitly in the interface, the low-level runtime
+%was also architected to permit a distributed and hierarchical implementation.
+%Critical resources can be replicated or migrated between nodes to reduce
+%both average latencies and the amount of traffic that use the scarce inter-node
+%communication bandwidth.
 
-\subsection{Events}
+\subsection{Low-Level Abstractions}
+\label{subsec:lowobjects}
+\subsubsection{Events}
+\label{subsec:events}
+The low-level runtime supports a deffered execution model to help hide
+the latencies associated with communicating in a distributed memory
+system.  The basic building block of the low-level runtime's deferred execution model
+is the {\em event}.  An event is a placeholder for the completion of a deferred
+operation.  When an asynchronous call is made to any of the low-level runtime objects, an event is 
+immediately returned to the caller.  The caller may query whether the event has {\em triggered}
+which happens when an even occurs.  The common use case for events is
+to provide them as prerequisites to subsequent requests made of the runtime. 
+These dependences cause the later requests to be automatically deferred until the previous
+operations have completed.  By using events as dependences, the caller can
+set up complicated, but coordinated chains of operations to be executed without
+having to wait for the low-level runtime to execute them.
+%These requests will
+%also return events, and the caller can set up a complicated, but coordinated,
+%set of operations to be performed by the low-level runtime with no babysitting
+%required of the original caller.
 
-The basic building block of the low-level runtime's deferred execution model
-is the Event.  An Event is a placeholder for the completion of a deferred
-operation.  When a call is made to the low-level runtime, an Event is usually
-returned to the caller.  The caller may query whether the event has occurred,
-and may also ask to wait until the event occurs, but the more common case is
-to provide one or more Events to subsequent requests made of the runtime, 
-causing those requests to be automatically deferred until the previous
-operations corresponding to those Events have completed.  These requests will
-also return Events, and the caller can set up a complicated, but coordinated,
-set of operations to be performed by the low-level runtime with no babysitting
-required of the original caller.
-
-With nearly every single runtime operation creating a new Event and referring
-to potentially any number of past Events, both the time and space overhead of
-these Events is a serious concern.  The time overhead is addressed by a
-mechanism that allows a node to create a new Event without needing to
-communicate with any other node.  The Event's handle identifies which node
+With nearly every single runtime operation creating a new event and referring
+to potentially any number of past events, both the time and space overhead of
+implementing events is a serious concern.  The time overhead is addressed by a
+mechanism that allows a node to create a new event without needing to
+communicate with any other node.  The event's handle identifies which node
 created it, allowing another node to lazily access it only when and if it 
-is asked to perform an operation on it.  Checking to see if an Event has
+is asked to perform an operation on it.  Checking to see if an event has
 occurred is done via a subscription mechanism.  Any number of threads or
-deferred operations on one node can query or wait on an Event with a single
+deferred operations on one node can query or wait on an event with a single
 message sent to the originating node.  The requesting node tracks the multiple
@@
 
-The potential space concern with Events is addressed by a generation scheme
-that allows Events to be reclaimed and reused without tracking references.  The
-naive implementation of an Event cannot reclaim the storage used by an Event
-until the Event has triggered \emph{and} all waiters on the event have been
-notified.  Some of these waiters may take quite a while to make their query,
-and worse, we must somehow track how many queries to expect if we want to know
-when they've all been performed.  However, if the Event handle is paired with a
-``generation'' count that tracks how many times the Event has been triggered,
-the Event can be reused as soon as the previous generation has been triggered.
+The potential space concern with events is addressed by a {\em generation scheme}
+that allows events to be reclaimed and reused without tracking references.  The
+naive implementation of an event cannot reclaim the storage used by an event
+until the event has triggered and all waiters on the event have been
+notified.  Some of these waiters may take quite a while to query the event.
+Additionaly, we must somehow track how many queries to expect to know
+when they've all been performed.  However, if the event handle is paired with a
+{\em generation count} that tracks how many times the event has been triggered,
+the event can be reused as soon as the previous generation has been triggered.
 Holders of an older (handle,generation) pair are able to query the newer
-incarnation of the Event and determine from the generation mismatch that their
-Event must have already triggered.  With this technique, there is no need to
-count references to the Event at all.  An Event can only be reused until the
-generation counter reaches its maximum value, the current implementation
-supports $2^{32}-1$ generations, and should that not be sufficient, extending
-it to $2^{64}-1$ is trivial.
+incarnation of the event and determine from the generation mismatch that their
+event must have already triggered.  With this technique, there is no need to
+count references to the event at all. 
+% An event can only be reused until the
+%generation counter reaches its maximum value, the current implementation
+%supports $2^{32}-1$ generations, and should that not be sufficient, extending
+%it to $2^{64}-1$ is trivial.
 
-\subsection{Deferred Locks}
-
+\subsubsection{Deferred Locks}
+\label{subsec:defferedlocks}
 Events are used to provide a deterministic ordering between operations, but in
-some cases, enforcing only serializability is a sufficient for correctness and
-can result in much better performance.  The traditional tool for this job is
+some cases is overly restrictive.  To implement the more relaxed coherence
+properties supported in Legion we need a less restrictive ordering semantics that
+can result in better performance.  The traditional tool for this job is
 the mutex (a.k.a. lock), and it is from the common reader/writer lock that
-the low-level runtime's \emph{deferred lock} is derived.  The key visible 
+the low-level runtime's {\em deferred lock} is derived.  The key visible 
 difference between a mutex and a deferred lock is that the request does not
-block the caller's execution.  Many libraries offer ``nonblocking locks'', but
+block the caller's execution.  Many libraries offer nonblocking locks, but
 they achieve this by returning without the lock if it is unavailable.  If
@@
 granted at some time in the future (with no further action required of the 
-caller).  That time in the future is described with an Event.  Like all other
-Events, it can be given to other threads, or to the runtime, or even stored
-in a data structure in memory to be extracted by an arbitrary thread at an 
-arbitrary time.  This dissociation of a held lock from the thread that
+caller).  That time in the future is described with an event.  Like all other
+events, it can be given to other threads, or even stored
+in a data structure in memory to be extracted by an arbitrary thread at a 
+later time.  This dissociation of a held lock from the thread that
 requested it is another important difference between the deferred lock and
@@
 The deferred nature of a deferred lock also allows a distributed implementation
-of locks.  Similar to Events, a deferred lock can be created on any node
+of locks.  Similar to events, a deferred lock can be created on any node
 without communication with the other nodes in the cluster.  That node becomes
@@
 
-\subsection{Processors}
-
+\subsubsection{Processors}
+\label{subsec:processors}
 Hardware units capable of performing computation are captured as generic
@@
 may come from anywhere in the system.  The request names which task should
-be spawn and provides input arguments and zero or more events.  Once those
+be spawned and provides input arguments and zero or more events.  Once those
 events have occurred, the task is placed into a queue of ready tasks, which
@@
 
-\subsection{Memories}
-
+\subsubsection{Memories}
+\label{subsec:memories}
 Application-visible storage in the low-level runtime is described in terms of
@@
 
-\subsection{Regions}
-
+\subsubsection{Region Meta-Data Objects}
+\label{subsec:regionmeta}
 Application-level regions are broken into four inter-related pieces in the
@@
 the size of the logical region, the element type, and the allocation mask,
-which tracks which portions of the logical region are actually in use.  Region
-MetaDatas are created independently of any memory objects and are accessible
-(via a globally unique name) to all nodes in the machine.
+which tracks which portions of the logical region are actually in use.  
+%Region
+%MetaDatas are created independently of any memory objects and are accessible
+%(via a globally unique name) to all nodes in the machine.
 
@@
 associated with a particular region metadata and spends its entire lifetime in
-the memory objecy in which it was created.  When multiple physical instances
-exist for the same region, the low-level runtime does nothing to keep them
-coherent - it is the job of the high-level runtime to perform the appropriate
-copies and set up the right inter-operation dependencies to provide the level
-of coherence requested by the application.  Although a physical instance 
-stores the contents of a region, it does not directly offer access to those
-contents.
+the memory objecy in which it was created.  
+%When multiple physical instances
+%exist for the same region, the low-level runtime does nothing to keep them
+%copies and set up the right inter-operation dependencies to provide the level
+%of coherence requested by the application.  Although a physical instance 
+%stores the contents of a region, it does not directly offer access to those
+%contents.
 
@@
 
-\subsection{Initialization}
+\subsection{Low-Level Implementations}
+\label{subsec:lowimpl}
+The primary goal of the low-level runtime is to provide an
+abstraction that enables efficient execution of Legion programs
+on a wide array of machines.  To support this claim we describe
+two implementation of the low-level runtime: one based on POSIX
+threads for a shared memory SMP and one based on GASNet and CUDA
+for distributed memory clusters with multiple GPUs.
+%The magic of an
+%apparently-unified address space is left to the high-level runtime, and the
+%interface to the low-level runtime is done in terms of handles rather than 
+%pointers for runtime objects and relative offsets rather than absolute 
+%pointers for resource allocations.
 
-On startup, the low-level runtime allocates all the resources it needs from
-the operating system and set up the communication channels between nodes.
-Each processor in the system then automatically runs the high-level runtime's
-initialization task, which will usually spawn more tasks.  These tasks are
-started, and whenever any processor's task queue is empty , the high-level
-runtime's scheduler task is automatically called to explore the possibility
-of moving tasks around for load balancing.
+%\subsection{Initialization}
 
-\subsection{SMP Implementation}
+%On startup, the low-level runtime allocates all the resources it needs from
+%the operating system and set up the communication channels between nodes.
+%Each processor in the system then automatically runs the high-level runtime's
+%initialization task, which will usually spawn more tasks.  These tasks are
+%started, and whenever any processor's task queue is empty , the high-level
+%runtime's scheduler task is automatically called to explore the possibility
+%of moving tasks around for load balancing.
 
-The SMP implementation of the low-level runtime is a relatively simple one.
-It runs on any SMP computer that supports POSIX threads.  A single memory
-object is created for the shared memory, and a processor object is created for
-each CPU core in the system.  Any GPUs in the system are ignore.  Events are 
+\subsubsection{SMP Implementation}
+\label{subsec:smpimpl}
+The SMP implementation of the low-level runs on any SMP computer that supports 
+POSIX threads.  A single memory object is created for the shared memory, and 
+a processor object is created for
+each CPU core in the system.  Events are 
 mutex-protected data structures that maintain a list of callbacks to call when
@@
 
-\subsection{GPU Cluster Implementation}
-
+\subsubsection{GPU Cluster Implementation}
+\label{subsec:clusterimpl}
 The cluster implementation of the low-level runtime is implemented on a mix
@@
 from significant overhead.  The memory hierarchy is expanded to include a 
-``global'' memory that is striped across all nodes and accessible via GASnet
-RDMA calls, and two memories per GPU: one for the GPU's framebuffer that is
+{\em global} memory that is striped across all nodes and accessible via GASnet
+RDMA calls, one process local memory per cluster node, and two memories per GPU: one for the GPU's framebuffer that is
 accessible only to that GPU and by bulk DMA operations, and one for the pinned
-``zero copy'' system memory that is accessible to the GPU and all the CPU cores
+{\em zero copy} system memory that is accessible to the GPU and all the CPU cores
 in a single system.

### 2011-11-07 doc/pldi2012/experiments.tex
Some modifications to the experimental section

--- 2011-11-06 doc/pldi2012/experiments.tex
@@
 three benchmarks were developed and run on a small cluster under a variety
-of configurations.  The cluster consists of four systems connected by a 40~Gb/s
-Infiniband fabric.  Each system is powered by 2 Intel Xeon X5680 CPUs, each
-with 6 cores, and 2 NVIDIA Tesla C2070 GPUs.  The systems were running Linux
+of configurations.  The cluster consists of four nodes connected by a 40~Gb/s
+Infiniband fabric.  Each node is powered by 2 Intel Xeon X5680 CPUs, each
+with 6 cores, and 2 NVIDIA Tesla C2070 GPUs.  The cluster is running Linux
 with GASNet using the {\tt ibv} conduit to interface directly with the standard
@@
 on many iterations using small stencils (e.g. just the neighboring nodes) to
-propagate changes through the system, and each iteration requires communication
+propagate changes through the system.  Each iteration requires communication
 between neighboring nodes.  Careful partitioning of the system can minimize
@@
 automatically handles the movement of data between the GPUs' memories and the
-system memory, but only when necessary.  In fact, with the help of an
-application-specific mapper, it is able to keep much of
+system memory when communication is necessary.  By utilizing an  
+application-specific mapper, our implementation is able to keep much of
 the data permanently resident in a given GPU for the entirety of the simulation.
-Figure~\ref{sfig:results1:perf} shows the relative speedup as the simulation is
-spread over multiple nodes in the cluster.  The size of the circuit
-being simulated is not changed as the node count increases.  This is therefore
-a demonstration of ``strong scaling''.  Adding a second node to the simulation
+Figure ~\ref{sfig:results1:perf} shows the relative speedup as the simulation is
+spread over multiple nodes in the cluster.  We demonstrate strong scaling by
+keeping the size of the circuit fixed as node count increases.
+Adding a second node to the simulation
 speeds it up by 86\%, and with four nodes, the simulation is sped up by a total
 of 241\% (i.e. a factor of 3.41x).  When enabling the use of both GPUs in each
-node (a decision that requires no changes to the application), the we achieve
+node (a decision that requires no changes to the application), we achieve
 a speedup of 5.90x when using all eight GPUs.
@@
 
-Table\~ref{sfig:results1:timers} breaks down the execution time by where it was
+Table~\ref{sfig:results1:timers} breaks down the execution time by where it was
 spent.  The times are summed over all threads, so the aggregate execution time
@@
 To show that Legion remains efficient for regular partitioning cases as well,
-we have implemented a particle simulation benchmark based heavily on the
+we have implemented a particle simulation benchmark based on the
 {\tt fluidanimate} benchmark from the PARSEC \cite{bienia11benchmarking} suite.
@@
 each node.  With application tasks running on all the available CPU cores,
-there are no ``extra'' cores for running the OS, device drivers, and the like.
+there are no extra cores for running the OS or device drivers.
 For example, GASNet's low-latency implementation counts on a CPU core being
 available for 
-a thread to spin on message arrival notifications.  Allowing that thread to
+a thread to spin on active message arrival notifications.  Allowing that thread to
 sleep (and then wait for the OS rescheduling delay) degrades system performance
-much more than the penalty being observed in our fluid simulation.  Now that
-one can generally count on having many CPU cores in any new system, it might
-make sense to reserve one (or more) in the low-level runtime for background
-tasks like these in order to provide more predictable performance on the CPU
-cores that are exposed to the high-level runtime.
+much more than the penalty being observed in our fluid simulation.  In future
+work we plan to reserve low-level runtime processors for these background
+tasks to provide more predictable performance when using all CPU cores.
+
+%Now that
+%one can generally count on having many CPU cores in any new system, it might
+%make sense to reserve one (or more) in the low-level runtime for background
+%tasks like these in order to provide more predictable performance on the CPU
+%cores that are exposed to the high-level runtime.
 
@@
 
-The circuit simulation was able to count on a good static, if irregular,
+The circuit simulation was able to count on a static, if irregular,
 partition of its workload, but that is not possible in some cases.  One

### 2011-11-07 doc/pldi2012/highlevel.tex
Fixing a missing reference

@@
 that is designed to implement the Legion programming
-and execution models while remaining machine agnostic.  
+and execution models while remaining machine agnostic for portability.  
 The high-level runtime system operates on top of the 
@@
 the node.  We determine this using the pair-wise dependence analysis described 
-in Section \ref{cohdep} to detect dependences between the region we are looking for and the
+in Section \ref{subsec:cohdep} to detect dependences between the region we are looking for and the
 region represented by the current node.  For each dependence that is found, we record that the previous

### 2011-11-07 doc/pldi2012/type_system.tex
types

@@
 
-The key to both performance and correctness in the Legion system is the
-accuracy of the region usage declarations.  The runtime will place data and
-schedule tasks based on these declarations, and has no way to recover if a 
-task attempts to access data that lies outside the mapped regions or attempts
-to write to a region that the runtime expected would only be read.
-
-The guarantee that these region usage declarations will be adhered to is
-provided by Legion's type system which is able to statically verify that a 
-task will stay within its declared boundaries.  The Legion type system is
-flow-sensitive, and uses judgements of the form:
-
+The key to both performance and correctness in Legion is the
+accuracy of the region usage declarations.  The runtime places data and
+schedules tasks based on these declarations and has no way to recover if a 
+task attempts to access data outside the declared regions or
+writes a region the runtime expected would only be read.
+The purpose of the Legion type system is to statically check that region usage declarations are adhered to.
+In this section we summarize the main features of the type system; due to space limitations we omit more than
+we include.
+Type judgments have the form:
 \begin{center}
@@
 \end{center}
-
-In addition to the mapping of variables to types (i.e. $\Gamma$), the
-environment includes a set of \emph{privileges} $\Phi_i$ that are possessed
-and a set of \emph{constraints} $C_i$ that are known to hold before the 
-expression $e$ is evaluated.  The judgement says that $e$ has some type $T$,
-but also tells us that the privileges possessed after the
-evaluation are a potentially-different $\Phi_o$ and that the new set of 
-constraints is $C_o$.
+In addition to tracking the types of variables $\Gamma$, the
+type environment includes the current privileges $\Phi_i$ 
+and region constraints $C_i$.  A judgment concludes $e$ has some type $T$,
+but also tells us how the evaluation of $e$ changes the privileges $\Phi_o$ and constraints $C_o$.
 
@@
 constraints that make up $\Phi$ and $C$.  Privileges represent the ability to
-perform some operation on a region (or, as we will see, on any of its
-subregions).  Constraints are used to capture the subregion relationships that
-result from partitioning ($a \prec b$ if $a$ was a result of a partitioning of
-$b$), the more general anywhere-in-the-subregion-tree relationship ($a \le b$ if $a$ equals $b$ or is either a direct or indirect subregion), and equality (or
-inequality) of integer expressions.  More complicated constraints can be formed
-through conjunctions, but neither negation nor disjunctions are permitted.  For
-an example of how these work, below is the typing rule for reading an element
-from a region:
+perform some operation on a region and its subregions.  Constraints capture relationships between regions
+resulting from partitioning: $a \rleq b$ if $a$ is a subregion $b$ (directly or transitively) and $a \ast b$ if
+$a$ and $b$ are disjoint (e.g., because they are distinct subregions of a disjoint partition).
+Our first type rule checks region reads:
 
@@
 
-This rule tells us that in order to read from a pointer, we must be able to
+This rule says that to read from a pointer, we must 
 provide a region ($r_2$ in this case) for which we possess a read privilege that 
 is known to either be, or contain, the region to which the pointer is
-constrained.  Similar rules exist for the other four region access operations.
+constrained.  The rule for region writes is similar.
 
-The Legion type system also has to be able to express the self-referential
-nature of region relationships.  It does this through existential quantification
-over regions.  The quantifiers captures both types and constraints, but does
-not capture privileges.  Region relations have the following form in the
-type system:
-
+Region relationships are first-class values and so can be stored in data structures
+and passed around the program.  However, the Legion type
+system cannot afford to lose information about regions that are stored
+in the heap.  To track the regions
+in region relationships accurately we use existential quantification
+over regions.  The quantifiers capture both types and constraints, but 
+not privileges.  The type of a region relation has the form:
 \begin{center}
@@
 \end{center}
-
-Explicit pack and unpack operations are used in the type system
-to add and remove the existential quantifiers, shown here:
-
+Pack and unpack operations are used in the type system
+to add and remove the existential quantifiers:
 \begin{center}
+{\small
 \begin{math}
@@
 RR = \exists r_1, \ldots, r_n.\left(T_1, \emptyset, C_1\right) \\
-\Gamma, \Phi_2, C_2 \vdash e : \left(RR, \Phi_3, C_3\right) \\
+\Gamma, \Phi_2, C_2 \vdash e : RR, \Phi_3, C_3  \\
 r'_1, \ldots, r'_n \not\in \mathit{RegionsOf}\left(\Gamma, \Phi_3, C_3\right)
@@
 \begin{array}{l@{}l}
-\Gamma, \Phi_2, C_2 \vdash unpackrr{ }e : ( & [r'_1/r_1,\ldots,r'_n/r_n]T_1, \Phi_3, \\
-& C_3 \wedge [r'_1/r_1,\ldots,r'_n/r_n]C_1)
+\Gamma, \Phi_2, C_2 \vdash unpackrr{ }e :  & [r'_1/r_1,\ldots,r'_n/r_n]T_1, \Phi_3, \\
+& C_3 \wedge [r'_1/r_1,\ldots,r'_n/r_n]C_1
 \end{array}
@@
 RR = \exists r_1, \ldots, r_n.\left(T_1, \emptyset, C_1\right) \\
-\Gamma, \Phi_2, C_2 \vdash e : \left([r'_1/r_1,\ldots,r'_n/r_n]T_1, \Phi_3, C_3\right) \\
+\Gamma, \Phi_2, C_2 \vdash e : [r'_1/r_1,\ldots,r'_n/r_n]T_1, \Phi_3, C_3 \\
 C_3 \models [r'_1/r_1,\ldots,r'_n/r_n]C_1
@@
 \begin{array}{l}
-\Gamma, \Phi_2, C_2 \vdash packrr{ }RR\mbox{ }e : \left(RR, \Phi_3, C_3\right) \\
+\Gamma, \Phi_2, C_2 \vdash packrr{ }RR\mbox{ }e : RR, \Phi_3, C_3 \\
 \end{array}
@@
 \end{math}
+}
 \end{center}
+Any expression whose type is a region relationship can be unpacked.
+Doing so introduces a fresh region variable into the current type
+environment for each region that was bound in the quantification.  The
+fresh region variables are substituted into the region relationship's
+type and constraints.  Packing an expression into a region
+relationship does the reverse.  If the type of an expression can be
+unified with the region relationship's type and the corresponding
+constraints of the region relationship can be shown to hold, the
+expression can be packed into the region relationship and used
+interchangeably with any other instance of that region relationship.
 
-Any expression whose type is a region relationship can be unpacked.  Doing so
-introduces a fresh region variable for each region that was bound in the 
-quantification.  The fresh region variables are substitued into the region
-relaionship's type, and a corresponding substituion is performed on the
-constraints that were captured by the region relationship.  These substituted
-constraints are added to the current task's (flow-sensitive) constraints.
+Although the type system uses explicit packing and unpacking operations, 
+note that these are implicit in Legion application code, such as the
+circuit simulation in Listing~\ref{lst:code_ex}.  Unpacking is
+inserted whenever a variable comes into scope, and packing is used
+whenever a variable is written to the heap or passed as a task argument.
 
-Packing an expression into a region relationship works in reverse.  If the type
-of an expression can be unified with the region relationship's type and the
-corresponding constraints of the region relationship can be shown to hold, the
-expression can be packed into the region relationship and used interchangably
-with any other instance of that region relationship.
-
-Although the type system uses explicit packing and unpacking operations, they 
-are implicit in the Legion application code.  Unpacking operations are 
-automatically inserted whenever a variable comes into scope.  Packing operations
-are attemped whenever a variable is written to memory or passed as an argument
-to a task.  Additionally, the simultaneous assignment operation essentially
-performs a pack for the result of the assignments and will also perform an
-unpack if one ore more fields of the structure are left unchanged.
+%Additionally, the simultaneous assignment operation essentially
+%performs a pack for the result of the assignments and will also perform an
+%unpack if one ore more fields of the structure are left unchanged.
 
@@
   &$\mid$&$reduceable(r,f)$ \\
-  &$\mid$&$allocable(r)$ \\
-  &$\mid$&$freeable(r)$ \\
+%  &$\mid$&$allocable(r)$ \\
+%  &$\mid$&$freeable(r)$ \\
 \\
@@
 \begin{array}{ccc}
-C & ::= & (r_1 \prec r_2) \\
-  &\mid&(r_1 \le r_2) \\
+C & ::= & (r_1 \rleq r_2) \\
+%  &\mid&(r_1 \le r_2) \\
   &\mid&(r_1 * r_2) \\
-  &\mid&(i_1 = i_2) \\
-  &\mid&(i_1 \ne i_2) \\
+%  &\mid&(i_1 = i_2) \\
+%  &\mid&(i_1 \ne i_2) \\
   &\mid&(C_1 \wedge C_2) \\
@@
 
-Region access privileges cannot be stored or packed.  Instead, they follow the
-task call tree, and it is this property that allows the runtime to make
-its scheduling decisions in a distributed way.  The original source of any
-privilege for a given region is the \emph{newrr} operation that created that
-region, but the way in which privileges can be provided to a subtask, or
-returned to a parent task, is described in the task application rule:
+As mentioned above, region privileges are not stored in packed types;
+instead, privileges are passed through the task call tree.  This is a key property of the type
+system, and is what allows the runtime to 
+track which regions are in use by which tasks and to make distributed scheduling decisions.
+Each subtask gets the privileges it requires from
+the task that calls it.  For example, in Listing~\ref{lst:code_ex},
+the task {\tt calc\_new\_currents} takes regions {\tt piece.rw\_ovt} with {\tt RW} privileges
+and {\tt piece.rn\_pvt} and {\tt piece.rn\_ghost} with {\tt RO} (read-only) privileges
+as arguments.  Because the calling task has {\tt RW} privileges on the parent regions of these
+region arguments, the type checker confirms that the subtask can execute with the needed privileges.
+The coherence property for a task argument is always identical to the one of the parent task.
 
-\begin{center}
-\begin{math}
-\begin{array}{c}
-\infrule
-{
-\begin{array}{lc}
-  \Gamma, \Phi_1, C_1 \vdash e_1 : \left(\left(T_i, \Phi_i, C_i\right) \rightarrow \exists r_1, \ldots, r_n . \left(T_o, \Phi_o, C_o\right), \Phi_2, C_2\right) \\
-  \Gamma, \Phi_2, C_2 \vdash e_2 : \left(T_i, \Phi_3, C_3\right) \\
-  r'_1, \ldots, r'_n \not\in \mathit{RegionsOf}(\Gamma, \Phi_3, C_3) \hspace{1cm} 
-  \Phi_i \subset \Phi_3 \hspace{1cm}
-  C_3 \models C_i
-\end{array}
-}
-{
-\begin{array}{l@{}l}
+Because privileges are forgotten when a region relationship is packed, one may wonder how
+a task can ever actually use a region $r$ in the region relationship when it is unpacked.
+The answer is that if $r \rleq r'$ and the task already holds privileges for $r'$, then it is able
+to use $r'$ with those same privileges.
+
+%bprivilege for a given region is the \emph{newrr} operation that created that
+%region, but the way in which privileges can be provided to a subtask, or
+%returned to a parent task, is described in the task application rule:
+%
+%\begin{center}
+%\begin{math}
+%\begin{array}{c}
+%\infrule
+%{
+%\begin{array}{lc}
+%  \Gamma, \Phi_1, C_1 \vdash e_1 : \left(\left(T_i, \Phi_i, C_i\right) \rightarrow \exists r_1, \ldots, r_n . \left(T_o, \Phi_o, C_o\right), \Phi_2, C_2\right) \\
+%  \Gamma, \Phi_2, C_2 \vdash e_2 : \left(T_i, \Phi_3, C_3\right) \\
+%  r'_1, \ldots, r'_n \not\in \mathit{RegionsOf}(\Gamma, \Phi_3, C_3) \hspace{1cm} 
+%  \Phi_i \subset \Phi_3 \hspace{1cm}
+%  C_3 \models C_i
+%\end{array}
+%}
+%{
+%\begin{array}{l@{}l}
 %  \envsub{1}{1} e_1 e_2 : \rtriple{& \regionexpand T_o}{\left(\Phi_3 \cup \regionexpand \Phi_o\right)}{\\ & \left(C_3 \wedge \regionexpand C_o \wedge \bigwedge_{\substack{{\footnotesize r_f \in \{ r | \regionexpand C_o \models fresh(r) \} } \\ {\tiny r_o \in \mathit{RegionsOf}(\Gamma, \Phi_3, C_3)}}} r_f * r_o\right)}
-  \Gamma, \Phi_1, C_1 \vdash e_1 e_2 : \Big( & [r'_1/r_1,\ldots,r'_n/r_n]T_o, \\
- & \left(\Phi_3 \cup [r'_1/r_1,\ldots,r'_n/r_n] \Phi_o\right) , \\
- & \left(C_3 \wedge [r'_1/r_1,\ldots,r'_n/r_n] C_o\right) \Big)
-\end{array}
-}
-\end{array}
-\end{math}
-\end{center}
-
-In this rule, expression $e_1$ has the type of a Legion task.  In addition
-to the task's input and output types, it specifies the privileges and
-constraints that must exist at the point where the task is called.  It also
-describes the privileges and constraints that exist as post-conditions of the
-task invocation.  The output type, privileges, and constraints can be
-captured in an existential quantification, allowing a task to return one
-or more regions that were previously unknown to the caller.
+%  \Gamma, \Phi_1, C_1 \vdash e_1 e_2 : \Big( & [r'_1/r_1,\ldots,r'_n/r_n]T_o, \\
+% & \left(\Phi_3 \cup [r'_1/r_1,\ldots,r'_n/r_n] \Phi_o\right) , \\
+% & \left(C_3 \wedge [r'_1/r_1,\ldots,r'_n/r_n] C_o\right) \Big)
+%\end{array}
+%}
+%\end{array}
+%\end{math}
+%\end{center}
+%
+%In this rule, expression $e_1$ has the type of a Legion task.  In addition
+%to the task's input and output types, it specifies the privileges and
+%constraints that must exist at the point where the task is called.  It also
+%describes the privileges and constraints that exist as post-conditions of the
+%task invocation.  The output type, privileges, and constraints can be
+%captured in an existential quantification, allowing a task to return one
+%or more regions that were previously unknown to the caller.

### 2011-11-07 doc/pldi2012/highlevel.tex

@@
 \label{sec:highlevel}
-In this Section we describe a high-level runtime system
+In this Section we describe the high-level runtime system
 that is designed to implement the Legion programming
@@
 low-level runtime which hides machine specific details
-and is described in detail in Section \ref{sec:lowlevel}.
+and is described in Section \ref{sec:lowlevel}.
 
@@
 from when they are actually performed.  This model is supported
-by features in the low-level runtime that are discussed in Section \ref{sec:lowlevel}.
+by features in the low-level runtime.
 
@@
 programs in this environment the high-level runtime is architected
-in a manner similar to an out-of-order hardware processor, but at
+in a manner similar to a pipelined out-of-order hardware processor, but at
 a coarser granularity with tasks corresponding to instructions and
-regions corresponding to registers.  There are multiple stages to 
+regions corresponding to registers.  There is a pipeline of stages to 
 executing a task which allows the runtime to extract
@@
 information about both the logical regions that the task will require
-when it is executed as well as the read-write privileges and coherence properties
+as well as the read-write privileges and coherence properties
 for each of the regions.  The runtime uses this information to perform
@@
 \ref{tab:depsdetect} to determine whether there is a dependence.
-\begin{center}
-\begin{table}
-\begin{tabular}{|c|c|c|c|c|} \hline
-             & Exclusive & Atomic   & Simultaneous & Relaxed \\ \hline 
-Exclusive    & Dep & Dep & Dep & Dep \\ \hline
-Atomic       & Dep & Same & Cont & Cont \\ \hline
-Simultaneous & Dep & Cont & Same & None \\ \hline
-Relaxed      & Dep & Cont & None & None \\ \hline
-\end{tabular}
-\caption{Dependence detection cases when at least one access has a write privilege. {\tt Dep}
-indicates a dependence is detected.  {\tt Same} indicates a dependence is detected
-only if the task uses different instances.  {\tt Cont} means that the dependence
-is contingent upon the permissions.  {\tt None} means there will never
-be a conflict.\label{tab:depsdetect}}
-\end{table}
-\end{center}
 %In this table, {\em Dependence} indicates that there will always be a dependence between
@@
 reader will not violate the atomic access semantics of the writer.
+\begin{center}
+\begin{table}
+\small
+\begin{tabular}{|c|c|c|c|c|} \hline
+             & Exclusive & Atomic   & Simultaneous & Relaxed \\ \hline 
+Exclusive    & Dep & Dep & Dep & Dep \\ \hline
+Atomic       & Dep & Same & Cont & Cont \\ \hline
+Simultaneous & Dep & Cont & Same & None \\ \hline
+Relaxed      & Dep & Cont & None & None \\ \hline
+\end{tabular}
+\caption{Dependence detection cases when at least one access has a write privilege. {\tt Dep}
+indicates a dependence is detected.  {\tt Same} indicates a dependence is detected
+only if the task uses different instances.  {\tt Cont} means that the dependence
+is contingent upon the permissions.  {\tt None} means there will never
+be a conflict.\label{tab:depsdetect}}
+\end{table}
+\end{center}
 
@@
 
-The runtime also assigns a unique identifier describing the context
-in which the task is executing.  All calls into the runtime require the task
-to pass this context identifier which is what permits the runtime to know the context
-for incoming calls for creating/destroying regions and launching sub-tasks.
+%The runtime also assigns a unique identifier describing the context
+%in which the task is executing.  All calls into the runtime require the task
+%to pass this context identifier which is what permits the runtime to know the context
+%for incoming calls for creating/destroying regions and launching sub-tasks.
 
@@
 
-While reference counting normally doesn't allow for garbage collection of all data structures
-it will always work in this case because both our tasks and our regions are organized as 
-trees which ensures that it is impossible to create cyclic region references in our system.
+%While reference counting normally doesn't allow for garbage collection of all data structures
+%it will always work in this case because both our tasks and our regions are organized as 
+%trees which ensures that it is impossible to create cyclic region references in our system.

### 2011-11-07 doc/pldi2012/experiments.tex
removal of tree search benchmark, scaling of some figures/tables

--- 2011-11-07 doc/pldi2012/experiments.tex
@@
 To measure Legion's ability to distribute work across a distributed machine, 
-three benchmarks were developed and run on a small cluster under a variety
-of configurations.  The cluster consists of four nodes connected by a 40~Gb/s
-Infiniband fabric.  Each node is powered by 2 Intel Xeon X5680 CPUs, each
-with 6 cores, and 2 NVIDIA Tesla C2070 GPUs.  The cluster is running Linux
+two real-world benchmarks were developed and run on a small cluster under a
+variety of configurations.  The cluster consists of four nodes connected by a
+40~Gb/s Infiniband fabric.  Each node is powered by 2 Intel Xeon X5680 CPUs,
+each with 6 cores, and 2 NVIDIA Tesla C2070 GPUs.  The cluster is running Linux
 with GASNet using the {\tt ibv} conduit to interface directly with the standard
@@
   at={(0.02,0.98)},
-  anchor=north west}}
+  anchor=north west},
+  cycle list={{black, mark=x}, {black, mark=+},
+              {black, mark=o}, {black, mark=diamond},
+              {black, mark=triangle}, {black, mark=star}}}
 
@@
   \subfigure[Performance Scaling]{
-    \begin{tikzpicture}
-      \begin{axis}[xlabel=Number of Nodes,ylabel=Normalized Speedup,height=5cm,width=8.5cm,xtick={1,2,3,4}]
+    \begin{tikzpicture}[scale=0.8]
+      \begin{axis}[xlabel=Number of Nodes,ylabel=Normalized Speedup,height=5cm,width=8.5cm,xtick={1,2,3,4},font=\small]
         \addplot coordinates {
@@
   \subtable[Execution Time Breakdown]{
+    \scalebox{0.8}{
    \tiny
@@
     \end{tabular}
+}
     \label{sfig:results1:timers}
@@
   \subfigure[Performance Scaling]{
-    \begin{tikzpicture}
-      \begin{axis}[xlabel=Number of Nodes,ylabel=Normalized Speedup,height=5cm,width=8.5cm,xtick={1,2,3,4}]
+    \begin{tikzpicture}[scale=0.8]
+      \begin{axis}[
+          xlabel=Number of Nodes,
+          ylabel=Normalized Speedup,
+          height=5cm,
+          width=8.5cm,
+          xtick={1,2,3,4},
+          legend columns=3,
+          font=\small
+        ]
         \addplot coordinates {
@@
     \end{tabular}
+}
     \label{sfig:results2:timers}
@@
 \end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Tree Search}
-
-The circuit simulation was able to count on a static, if irregular,
-partition of its workload, but that is not possible in some cases.  One
-such example arises in ray tracing, in which the components of a scene to be
-rendered are partitioned in a binary tree to minimize the number of expensive
-object intersection tests that must be performed.  By traversing its path 
-through the tree, a ray is able to efficiently skip entire chunks of the 
-scene that it is guaranteed not to intersect with.  The trees tend to be
-highly imbalanced (often intentionally so), and the set of rays that traverse
-the tree tend to have similar path.  Both of these factors contribute to
-large fluctuations in the load on any particular part of the tree, complicating
-attempts to statically partition the tree.  The {\tt Tree Search} benchmark
-explores Legion's ability to load-balance tasks in a locality-aware manner.
-
-The benchmark constructs an irregular binary tree, and then partitions it into
-roughly subregions that have roughly equal numbers of nodes.  Partition
-crossings are annotated on the tree and traversal tasks are re-spawned when
-a ray crosses into another partition.  Load imbalances will occur, but when
-they do, the default mapper's ability to provide a set of tasks with good
-locality should minimize the amount of data that must be copied around.  To
-measure this benefit, each run is also done with the tree partitioning 
-disabled.  This denies the mapper the locality information it would need to 
-make good choices, reducing it to making essentially random choices.
-
-Figure~\ref{sfig:results3:perf} shows performance of the tree search test
-relative to the single node with partitioning case. MORE TEXT HERE.
-
-\begin{figure}
-  \centering
-  \subfigure[Performance Scaling]{
-    \begin{tikzpicture}
-      \begin{axis}[xlabel=Number of Nodes,ylabel=Normalized Speedup,height=5cm,width=8.5cm,xtick={1,2,3,4}]
-        \addplot coordinates {
-% num gpus = 2
-  (1, 1.99)
-  (2, 3.46)
-  (3, 3.44)
-  (4, 5.90)
-        };
-        \addplot coordinates {
-% num gpus = 1
-  (1, 1.00)
-  (2, 1.86)
-  (3, 2.40)
-  (4, 3.41)
-        };
-      \legend{2 GPU/node,1 GPU/node}
-      \end{axis}
-    \end{tikzpicture}
-    \label{sfig:results3:perf}
-  }
-
-  \subtable[Execution Time Breakdown]{
-   \tiny
-    \begin{tabular}{c@{\hspace{2pt}}cD{.}{.}{3}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}@{\hspace{2pt}}D{.}{.}{3}}
-\toprule
-GPUs/ & \# \\
-node & nodes & \multicolumn{1}{c}{Wall Clock} & \multicolumn{1}{c}{Application} & \multicolumn{1}{c}{Copies} & \multicolumn{1}{c}{High-level} & \multicolumn{1}{c}{Low-level} & \multicolumn{1}{c}{Mapper} & \multicolumn{1}{c}{System} \\
-\midrule
-\multirow{4}{*}{$1$} & $1$ & 106.892 & 105.387 & 1.578 & 0.003 & 21.624 & 0 & 0\\
- & $2$ & 57.388 & 105.407 & 10.348 & 0.003 & 15.177 & 0 & 0.001\\
- & $3$ & 44.552 & 105.41 & 11.935 & 0.003 & 11.695 & 0 & 0.002\\
- & $4$ & 31.368 & 105.42 & 12.875 & 0.004 & 8.419 & 0 & 0.003\\
-\midrule
-\multirow{4}{*}{$2$} & $1$ & 53.787 & 105.397 & 1.857 & 0.004 & 10.953 & 0 & 0\\
- & $2$ & 30.923 & 105.415 & 10.716 & 0.004 & 7.937 & 0 & 0.002\\
- & $3$ & 31.037 & 105.414 & 12.558 & 0.004 & 7.996 & 0 & 0.001\\
- & $4$ & 18.103 & 105.42 & 13.747 & 0.004 & 4.638 & 0 & 0.003\\
-\bottomrule \\
-    \end{tabular}
-    \label{sfig:results3:timers}
-  }
-  \caption{Tree Search Results}
-\end{figure}

### 2011-11-07 doc/pldi2012/exec.tex
text cleanup, compaction

@@
 
-In section \ref{sec:type} we showed that the type system
-only enables privileges to be passed with task calls.  Based on
-this property of privileges we have claimed that our runtime
-will be able to make local scheduling decisions.  In this
+In section \ref{sec:type} we showed that the Legion type system
+only enables privileges to be passed with task calls. 
+ In this
 section we present the Legion execution model and show that
-this property of scheduling preserves the sequential execution 
-semantics of a program.  The correctness argument is
+the sequential execution 
+semantics of a program can be preserved while making only local scheduling
+decisions.
+  The correctness argument is
 instructive in that the properties that ensure correctness are also
@@
 statements.  The core language has no permissions or coherence annotations
-on functions; for simplicity, in this section we simply use exclusive access in the case
-that a function writes a region.  The statement $f(r_1,\ldots,r_n)$ calls a function, which
+on functions; in this section we simply use exclusive access in all cases.
+ The statement $f(r_1,\ldots,r_n)$ calls a function, which
 is executed for its effects on the region arguments.  The statement $x
@@
 Let $\lca{r}{r'}$ be the meet of $r$ and $r'$ (the least-common
-ancestor of $r$ and $r'$ in the region forest).  If $r \neq r'$ and
+ancestor of $r$ and $r'$ in the region forest).  If 
 $\lca{r}{r'} = p$ then $r$ and $r'$ are disjoint, as $r$ and $r'$ must
 be in distinct subregions of the partition $p$ and partitions
-guarantee disjoint subregions.  If $\lca{r}{r'} = r''$ then either $r
-\rleq r' = r''$ , $r \rleq r = r''$, or $r \leq p$, $r' \leq p'$, and
-$p$ and $p'$ are two distinct partitions of $r''$.  In any of these
-cases $r$ and $r'$ cannot be proven disjoint using $\rleq$.
+guarantee disjoint subregions.  If $r$ and $r'$ meet at a region (i.e. $\lca{r}{r'} = r''$) then either one of $r$ or $r'$ is a subregion of the other, or they
+lie in two distinct partitions of $r''$ and disjointness cannot be guaranteed.
 

### 2011-11-07 doc/pldi2012/type_system.tex
text cleanup, compaction

@@
 The key to both performance and correctness in Legion is the
-accuracy of the region usage declarations.  The runtime places data and
-schedules tasks based on these declarations and has no way to recover if a 
-task attempts to access data outside the declared regions or
-writes a region the runtime expected would only be read.
-The purpose of the Legion type system is to statically check that region usage declarations are adhered to.
+accuracy of the region usage declarations.  The Legion type system has been
+designed to statically check that region usage declarations are adhered to.
 In this section we summarize the main features of the type system; due to space limitations we omit more than
@@
 
-This rule says that to read from a pointer, we must 
-provide a region ($r_2$ in this case) for which we possess a read privilege that 
-is known to either be, or contain, the region to which the pointer is
-constrained.  The rule for region writes is similar.
+In order to safely read from a pointer into $r_1$, we must 
+provide a region $r_2$ which is known to contain $r_1$ and for which we possess
+a read privilege.  Rules for region writes and reductions are similar.
 
-Region relationships are first-class values and so can be stored in data structures
+Region relationships are first-class values and can be stored in data structures
 and passed around the program.  However, the Legion type
@@
 \begin{array}{l@{}l}
-\Gamma, \Phi_2, C_2 \vdash unpackrr{ }e :  & [r'_1/r_1,\ldots,r'_n/r_n]T_1, \Phi_3, \\
+\Gamma, \Phi_2, C_2 \vdash unpackrr{~}e :~  & [r'_1/r_1,\ldots,r'_n/r_n]T_1, \Phi_3, \\
 & C_3 \wedge [r'_1/r_1,\ldots,r'_n/r_n]C_1
@@
 Although the type system uses explicit packing and unpacking operations, 
-note that these are implicit in Legion application code, such as the
+these are implicit in Legion application code, such as the
 circuit simulation in Listing~\ref{lst:code_ex}.  Unpacking is
@@
 As mentioned above, region privileges are not stored in packed types;
-instead, privileges are passed through the task call tree.  This is a key property of the type
-system, and is what allows the runtime to 
-track which regions are in use by which tasks and to make distributed scheduling decisions.
-Each subtask gets the privileges it requires from
+instead, privileges are passed through the task call tree.  This is a key
+property of the type system, and is what allows the runtime to 
+track which regions are in use by which tasks and to make distributed
+scheduling decisions.
+Each subtask must obtain its privileges from
 the task that calls it.  For example, in Listing~\ref{lst:code_ex},
-the task {\tt calc\_new\_currents} takes regions {\tt piece.rw\_ovt} with {\tt RW} privileges
-and {\tt piece.rn\_pvt} and {\tt piece.rn\_ghost} with {\tt RO} (read-only) privileges
-as arguments.  Because the calling task has {\tt RW} privileges on the parent regions of these
-region arguments, the type checker confirms that the subtask can execute with the needed privileges.
-The coherence property for a task argument is always identical to the one of the parent task.
+the task {\tt calc\_new\_currents} requires regions {\tt piece.rw\_pvt} with {\tt RW} privileges
+and {\tt piece.rn\_pvt} and {\tt piece.rn\_ghost} with {\tt RO} (read-only) privileges.  Because the calling task has {\tt RW} privileges on the parent regions of these
+region arguments, the type checker is able to confirm that the subtask can execute with the needed privileges.
 
-Because privileges are forgotten when a region relationship is packed, one may wonder how
-a task can ever actually use a region $r$ in the region relationship when it is unpacked.
-The answer is that if $r \rleq r'$ and the task already holds privileges for $r'$, then it is able
-to use $r'$ with those same privileges.
+%The coherence property for a task argument is always identical to the one of the parent task.
+
+Due the the inability for a packed region relationship to hold privileges, any
+useful region relationship will leave at least one region free (i.e.
+unquantified) and include constraints that relate the quantified region(s) to the free one(s).
+As long as a task holds a privilege for the region that is kept free, the
+region relationship can be unpacked and used.
 

### 2011-11-07 doc/pldi2012/circuit.tex

@@
 the nodes subregion, and the ghost node region for its piece.
-The {\tt distribute\_charge} subtask turns things around, reading the piece's 
-wires subregion and updating all nodes those wires touch.  However,
+The {\tt distribute\_charge} subtask is a different phase that reads the piece's 
+wires subregion and updates all nodes those wires touch.  However,
 rather than requesting the ability to read and write the nodes (which would
@@
 Since a region relationship can have
-fields with types that refer its own fields' regions, it is often not possible
+fields with types that refer to its own fields' regions, it is often not possible
 to assign fields one at a time without violating the type checking rules.

### 2011-11-07 doc/pldi2012/exec.tex
exec

@@
 
-In section \ref{sec:type} we showed that the Legion type system
-only enables privileges to be passed with task calls. 
- In this
-section we present the Legion execution model and show that
-the sequential execution 
-semantics of a program can be preserved while making only local scheduling
-decisions.
-  The correctness argument is
-instructive in that the properties that ensure correctness are also
-exploited in the implementation of the Legion high-level runtime
-system to achieve high performance and highly distributed scheduling
-of parallel tasks.
-
-For the purposes of describing and reasoning about the
-execution model we consider an extremely restricted core language, so
-simple in fact that it is not even Turing complete and is also very
-inconvenient for writing programs.  The loss in expressive power, however,
-is hopefully a gain in clarity and simplicity for illustrating the core ideas in
-Legion.    We consider programs written in the following language:
-
-\begin{eqnarray*}
-P & := & F_1 \ldots F_n \\[.1in]
-F & := & \mbox{\tt def}\ f(r_1,\ldots,r_n) =   s_1; \ldots s_n; \\[.1in]
+In this section we present the Legion parallel execution model and
+show that the sequential execution semantics of a program can be
+preserved while making only local scheduling decisions, provided we
+know the set of regions accessed by a task at scheduling time---the
+property that the type system described in Section~\ref{sec:type}
+guarantees.
+To make our arguments precise we consider an extremely restricted core language
+that nevertheless illustrates the main idea.
+Programs are sets of functions, one of which is a distinguished entry point:
+{ \small
+\[
+\begin{array}{rcl}
+F & := & \mbox{\tt def}\ f(r_1,\ldots,r_n) =   s_1; \ldots s_n; \\
 s & := & f(r_1,\ldots,r_n) \\
@@
 & | & x = \deref r \\
-&| & (r_1, r_2)_{p} = \mbox{\tt partition}(r) \\ 
+&| & (r_1, r_2)_{p} = \mbox{\tt partition}(r)
 %&| & \mbox{\tt close}\ p 
-\end{eqnarray*}
-
-There are four kinds of names: functions $f$, regions $r$, local
-variables $x$, and partitions $p$.  A function takes only regions as
-arguments and a function body consists of an ordered list of
-statements.  The core language has no permissions or coherence annotations
-on functions; in this section we simply use exclusive access in all cases.
+\end{array}
+\]
+}
+There are four kinds of names: functions $f$, regions $r$, 
+variables $x$, and partitions $p$.  A function body is a list of statements.
+The core language has no permissions or coherence annotations;
+in this section we simply use exclusive access.
 The statement $f(r_1,\ldots,r_n)$ calls a function, which
 is executed for its effects on the region arguments.  The statement $x
-= \deref r$ reads a value from region $r$ and stores it in local
-variable $x$, and the statement $\deref r = x$ stores the value of $x$
+= \deref r$ reads a value from region $r$ into
+$x$, and the statement $\deref r = x$ stores the value of $x$
 in region $r$.
 
-The statement $(r_1, r_2)_{p} = \mbox{\tt partition}(r)$ creates two
-{\em subregions} $r_1$ and $r_2$ partitioning the parent region $r$.
-All three region names ($r_1$,$r_2$,and $r$) can be used after the
-partition.  Regions $r_1$ and $r_2$ are disjoint, meaning that writes
-to $r_1$ have no effect on the contents of $r_2$ and vice versa, but
-neither $r_1$ nor $r_2$ is disjoint from $r$.  The partition itself
-also has a name $p$.  We define a relation $\rleq$ which is the
-reflexive, transitive closure of
-\[ r_1 \rleq p \ \ \ r_2 \rleq p \ \ \ p \rleq r \]
-The $\rleq$ relation defines a {\em region forest} with regions at the
-roots, and alternating levels where partitions are always children of
-regions and regions are always children of partitions.  If we ignore
-partitions, $\rleq$ is just the subregion relationship: $r \rleq r'$
-means that $r$ is a subregion (either immmediately or transitively) of
-$r'$.  Including partitions also gives information about disjointness.
-Let $\lca{r}{r'}$ be the meet of $r$ and $r'$ (the least-common
-ancestor of $r$ and $r'$ in the region forest).  If 
+The statement $(r_1, r_2)_{p} = \mbox{\tt partition}(r)$ creates
+regions $r_1$ and $r_2$ partitioning the parent region $r$.  For
+simplicity we assume partitions result in disjoint subregions.  The
+partition has a name $p$.  We extend the subregion
+relation $\rleq$ to include partitions, so $r_1 \rleq p$, $r_2 \rleq
+p$, and $p \rleq r$.  The $\rleq$ relation defines a {\em region
+forest} with regions at the roots, and alternating levels where
+partitions are always children of regions and regions are always
+children of partitions.  If we ignore partitions, $\rleq$ is just the
+subregion relation.  Including partitions gives information about
+disjointness.  Let $\lca{r}{r'}$ be the
+least-common ancestor of $r$ and $r'$ in the region forest.  If
 $\lca{r}{r'} = p$ then $r$ and $r'$ are disjoint, as $r$ and $r'$ must
-be in distinct subregions of the partition $p$ and partitions
-guarantee disjoint subregions.  If $r$ and $r'$ meet at a region (i.e. $\lca{r}{r'} = r''$) then either one of $r$ or $r'$ is a subregion of the other, or they
-lie in two distinct partitions of $r''$ and disjointness cannot be guaranteed.
+be in distinct subregions of the partition $p$.  If $\lca{r}{r'} = r''$ then either one of $r$ or $r'$
+is a subregion of the other, or they lie in two different partitions of
+$r''$ and disjointness cannot be guaranteed.  (In the full
+language, the test for disjointness has one more case for
+partitions with non-disjoint subregions.)
 
@@
 
-A program is a list of functions, with the entry point the first
-function on the list.  We require that every function (except
-the entry point) is called in exactly one place and that the formal
+To make the following development as simple as possible, we assume every
+function except the entry point is called in exactly one place and that the formal
 parameters are named identically to the actual parameters in the
-call.  Further, local variables, new region names and
-partition names in {\tt partition} are chosen to be distinct from all
+call.  Variable and partition names are chosen to be distinct from all
 other names in the program.  These restrictions allow us to avoid
 dealing with variable renaming and also mean that each runtime call is
-uniquely named by its function name.
+uniquely named by its function.
 
-The state of an execution step for a program is a map $M$ from the
-variable and region names of the program to their values.  We leave the values
-abstract, assuming only that each statement performs a deterministic
-update of the variables/regions it writes that depends only on the
-current state of the regions/variables it reads.
-The sets of regions read and written by statements are defined as follows:
+The sets of regions read and written by statements are:
+{\small
 \[
@@
 \rwrite{\deref r = x} & = & \{ r \} \\
-\rwrite{(r_1, r_2)_{p} = \mbox{\tt partition}(r)} & = & \{ r_1,r_2,r \} \\[.15in]
+%\rwrite{(r_1, r_2)_{p} = \mbox{\tt partition}(r)} & = & \{ r_1,r_2,r \} \\[.15in]
 \rwrite{f(r_1,\ldots,r_n)} & = & \bigcup_{1 \leq i \leq n}{\rwrite{s_i}} \\
-\multicolumn{3}{l}{\ \ \ \ \mbox{where\ {\tt def}}\ f(r_1,\ldots,r_n): =   s_1; \ldots s_n;} \\[.15in]
+\multicolumn{3}{l}{\ \ \ \ \mbox{where\ {\tt def}}\ f(r_1,\ldots,r_n): =   s_1; \ldots s_n;} \\
 \rread{f(r_1,\ldots,r_n)} & = & \bigcup_{1 \leq i \leq n}{\rread{s_i}} \\
-\multicolumn{3}{l}{\ \ \ \ \mbox{where\ {\tt def}}\ f(r_1,\ldots,r_n): =   s_1; \ldots s_n;} \\[.15in]
+\multicolumn{3}{l}{\ \ \ \ \mbox{where\ {\tt def}}\ f(r_1,\ldots,r_n): =   s_1; \ldots s_n;} \\
 \end{array}
 \]
-In all other cases the read/write set is empty (e.g., $\rread{\deref r = x} = \emptyset$).
-
-The {\em sequential execution order} is the canonical order in which reads, writes, and partitions of regions are performed.
-The sequential order is obtained by inlining all function calls---repeatedly
-replace function calls $g(r_1,\ldots,r_n)$ by their function bodies until the entry function contains no function calls.
-The order of statements in the entry function is then the sequential execution order. 
-Two programs $P_1$ and $P_2$ are equivalent, written $P_1 \equiv P_2$, if starting in the same initial state they halt in
-the same final state when evaluated in the sequential execution order.
-
-Consider the following program, where reads, writes, and partitions are numbered in the sequential execution order:
-{\small 
-\begin{tabbing}
-\ \ \ \ \ \= {\tt def} \= $\rm f(r_0) =$ \\
-1. \>\>   $\rm (r_1,r_2)_a = \mbox{\tt partition}(r_0);$   \\
-2. \>\>  $\rm (r_3,r_4)_b = \mbox{\tt partition}(r_1);$ \\
-\>\>     $\rm g(r_1); h(r_2); j(r_3); k(r_3,r_4);$ \\
-\> {\tt def} $\rm g(r_1) =$   \\
-3. \>\>  $\rm \ast r_1 = u;$ \\ \\
-\> {\tt def} $\rm h(r_2) =$ \\      
-4. \>\>   $\rm \ast r_2 = v;$\\ \\
-\> {\tt def} $\rm j(r_3) = $ \\
-5.\>\>   $\rm w = \ast r_3;$ \\ \\
-\> {\tt def} $\rm k(r_3,r_4) = $\\
-6.\>\>   $\rm x = \ast r_3;$ \\
-7.\>\>   $\rm (r_5,r_6)_c = \mbox{\tt partition}(r_4);$ \\ 
-\>\>     $\rm m(r_5);$ \\ \\
-\> {\tt def} $\rm m(r_5) = $ \\
-8. \>\>  $\rm \ast r_5 = y;$
-\end{tabbing}
 }
-
-Consider any two statements $s_1$ and $s_2$.  We say $s_1$ and $s_2$ are {\em independent} if
+We say two statements $s_1$ and $s_2$ are {\em independent} if
 \[
@@
 \]
-The definition captures the usual idea that two statements are independent if whenever one of them is a write they do not access the same memory location, but the notion of ``location'' is generalized to regions and must take into account that two regions
-are disjoint only if they are in different components of some partition.
+The definition captures the usual idea that statements are independent if whenever one of them is a write they do not access the same memory location, where the notion of ``location'' is generalized to our setting with regions and partitions.
+If $s_1$ and $s_2$ are independent, then $s_1; s_2 \equiv s_2; s_1$, meaning the two sequences are semantically
+equivalent (have the same net effect on regions when executed).  We omit the simple proof of this claim.
 
-\begin{lemma}
-\rm
-\label{lem:independence}
-If $s_1$ and $s_2$ are independent, then $s_1; s_2 \equiv s_2; s_1$.
-\end{lemma}
-\begin{proof}
-For a contradiction, assume that $s_1$ writes a region $r_1$ and $s_2$ may read from some $r_2$ where $r_1$ and $r_2$
-are not disjoint; then reversing the order of the statements clearly violates the claim.  Now $r_1 \in \rwrite{s_1}$
-and $r_2 \in \rread{s_2}$, and therefore $\lca{r_1}{r_2} = p$ for some partition $p$.  But then $r_1$ and $r_2$ are disjoint.
-The reasoning is symmetric if $s_1$ reads $r_1$ and $s_2$ writes $r_2$ or $s_1$ writes $r_1$ and $s_2$ writes $r_2$.
-Since the read-read case cannot violate independence, the result follows.
-\end{proof}
-We say $s_2$ {\em depends on} $s_1$ if $s_2$ follows $s_1$ in the sequential execution order and the statements are not independent.
-\begin{lemma}
-\rm
-\label{lem:dataflow}
-Let $S = s_1; \ldots s_n;$ be the sequential execution order of reads, writes, and partitions and let $S'$ be any permutation of $S$ such that the order of dependent statements is preserved.
-Then $S \equiv S'$.
-\end{lemma}
-\begin{proof}
-By induction on the number of swaps needed to transpose $S'$ into $S$.
-If $S = S'$ then we are done.  If $S$ and $S'$ are different then
-there must be two adjacent statements $s_1; s_2$ in $S'$ that are
-reversed with respect to the sequential execution order.  Since the order of
-dependent statements is preserved by the permutation, $s_1$ and $s_2$ are also independent.
-But then $s_1; s_2 \equiv s_2; s_1$ by Lemma~\ref{lem:independence} and the result follows.
-\end{proof}
-Lemma~\ref{lem:dataflow} captures the standard idea that any
-topological sort of the statements that preserves dependencies also
-preserves sequential execution semantics.  While this is the basis for
-dataflow parallelism, in the large distributed memory machines we
-target instruction-level dataflow parallelism is both too fine-grain
-and incurs too much communication, as we must compute a global
-dependence graph across the entire program.  In Legion we use a
-different notion of what can be executed in parallel that works at the
-granularity of functions and also requires analysis only within a function body.  We
-first need a few additional definitions.  Define $\context{s}$ to be
-the function call statement that invokes the function in which
-statement $s$ occurs (here we assume some way of distinguishing
-identical statements that occur in different functions).  Let
-$\interval{f}$ be the set of statements executed by the one call to
-function $f$ and all the functions $f$ transitively calls; we define
-$\interval{s} = \{ s \}$ for any statement $s$ other than a function call.
-Finally we say statements $s_1$ and $s_2$ are {\em siblings} if they
-occur in the same function body.
+The {\em sequential execution order} $S$ is the usual call-by-value
+sequence of region read and write statements performed by the program.  Let
+$S'$ be a permutation of $S$ where the order of dependent (i.e., not
+independent) statements is preserved.  It is easy to prove that
+$S \equiv S'$ by a sequence of swaps of adjacent independent
+statements.  
+Preserving dependencies between statement reads and writes captures
+dataflow parallelism, but in the large
+distributed memory machines we target instruction-level dataflow
+is both too fine-grain and incurs too much communication,
+as we must compute a global dependence graph across the entire
+program.
 
-\begin{lemma}
-\rm
-\label{lem:locality}
-If $s_1$ and $s_2$ are not independent and not siblings, then $s_1$ and $\context{s_2}$ are not independent
-and $s_2$ and $\context{s_1}$ are not independent.
-\end{lemma}
-\begin{proof}
-Follows immediately from the fact that the read/writes sets of $\context{s}$ are supersets of the read/write sets of $s$.
-\end{proof}
-We use Lemma~\ref{lem:locality} in the following way.  Consider two
-statements $s_1$ and $s_2$ such that $s_2$ depends on $s_1$.  Let $f$ be the function that is the least
-common ancestor of $s_1$ and $s_2$ in the call tree of the
+In Legion we use a different notion of what can be executed
+in parallel that works at the granularity of functions and also
+requires only local analysis within a function body.  We first need a few
+additional definitions.  Define $\context{s}$ to be the function call
+statement that invokes the function in which statement $s$ occurs
+(here we assume some way of distinguishing identical statements that
+occur in different functions).  Let $\interval{f}$ be the set of
+read and write statements executed by $f$ and all the
+functions $f$ transitively calls; we define $\interval{s} = \{ s \}$
+for any statement $s$ other than a function call.  Finally we say
+statements $s_1$ and $s_2$ are {\em siblings} if they occur in the
+same function body.
+
+Consider two statements $s_1$ and $s_2$ where $s_2$ {\em depends on}
+$s_1$ (i.e., $s_1$ and $s_2$ are dependent and $s_2$ follows $s_1$ in
+the sequential execution order).  Let $f$ be the function that is the
+least common ancestor of $s_1$ and $s_2$ in the call tree of the
 program---the unique function $f$ with the smallest interval such that
-$s_1,s_2 \in \interval{f}$.  Then one of the following is true:
-\begin{itemize}
-\item $s_1$ and $s_2$ are siblings in $f$.
+$s_1,s_2 \in \interval{f}$.  Then one of the following is true: (1) $s_1$ and $s_2$ are siblings in $f$;
+(2) $s_1$ occurs in $f$, there is a function call $f_2(\ldots)$ in $f$ such that $s_2 \in \interval{f_2}$, and $f_2(\ldots)$
+is dependent on $s_1$;
+(3) $s_2$ occurs in $f$, there is a function call $f_1(\ldots)$ in $f$ such that $s_1 \in \interval{f_1}$, and
+$s_2$ is dependent on $f_1(\ldots)$; or
+(4) there are two distinct function call statements $f_1(\ldots)$ and $f_2(\ldots)$ in the body of $f$ such
+that $s_1 \in \interval{f_1}$ and $s_2 \in \interval{f_2}$ and $f_2(\ldots)$ is dependent on $f_1(\ldots)$.
+The proof of this statement follows from the observation that if $s \in \interval{f}$, then either $s$ is in the function body of $f$ or $f$ contains a function call to $g$ and $s \in \interval{g}$ and $\rread{s} \subseteq \rread{g}$
+and $\rwrite{s} \subseteq \rwrite{g}$.
 
-\item $s_1$ occurs in $f$, there is a function call $f_2(\ldots)$ in $f$ such that $s_2 \in \interval{f_2}$, and $f_2(\ldots)$
-is dependent on $s_1$.
-
-\item $s_2$ occurs in $f$, there is a function call $f_1(\ldots)$ in $f$ such that $s_1 \in \interval{f_1}$, and
-$s_2$ is dependent on $f_1(\ldots)$.
-
-\item There are two distinct function call statements $f_1(\ldots)$ and $f_2(\ldots)$ in the body of $f$ such
-that $s_1 \in \interval{f_1}$ and $s_2 \in \interval{f_2}$ and $f_2(\ldots)$ is dependent on $f_1(\ldots)$.
-\end{itemize}
 In other words, any dependence between arbitrary statements
@@
 dependence between two statements (one or both of which may be a function call) within the same function body.  
-In the example program above, statements 3 and 8 are
-dependent because they write into regions that are in different
-partitions of region $r_0$.  This dependence means that any functions
-that include these statements in their interval cannot be independent,
-and in particular in the entry point function $f$ the function call
-$k(r_3,r_4)$ depends on sibling $g(r_1)$.
-
-The following lemma gives another class of statement orderings based on these observations.
+The following lemma gives a class of statement orderings based on these observations.
 \begin{lemma}
@@
 \label{lem:scheduling}
-Let $S$ be the sequential execution order of reads, writes and partitions.  Let $S'$ be any permutation of $S$ such that
+Let $S$ be the sequential execution order of region reads and writes.  Let $S'$ be any permutation of $S$ such that
 for any sibling statements $s_1$ and $s_2$ such that $s_2$ depends on $s_1$, all statements in $\interval{s_1}$ precede
@@
 \end{proof}
-The advantage of Lemma~\ref{lem:scheduling} is that provided we have
-a way to obtain the read and write sets for function calls, scheduling
-decisions can be made completely locally for the statements in a
-single function body at a time, independently of any decisions for
-other functions.  This observation eliminates any global
-computation to determine what statements can execute in parallel, and
-furthermore allows scheduling decisions theselves to be parallelized, as the statements
-in different functions can be scheduled separately.  As a practical matter
-our Legion implementation only parallelizes function calls---the other
-statements within a function body are executed in sequential
-order.
-
-Returning to our example above, in function $f$ the function calls
-$g(r_1)$ and $h(r_2)$ are independent, as they work on disjoint
-regions $r_1$ and $r_2$.  Similarly $j(r_3)$ and $k(r_3,r_4)$ are
-also independent as they both only read from $r_3$.  However, $j$ and $k$
-are both dependent on $g$ and $h$, as the earlier functions modify a different
-partition of the same data.  Thus Legion will ensure functions $j$ and $k$
-begin only after $g$ and $h$ complete, while it will allow $g$ and $h$
-to run in parallel, and $j$ and $k$ to run in parallel.
+The advantage of Lemma~\ref{lem:scheduling} is that given
+the region read and write sets for function calls, the decision about
+whether to execute two tasks in parallel can be made completely locally for the statements in a
+single function body at a time, independent of any decisions for
+any other functions.  
+%This observation eliminates any global
+%computation to determine what statements can execute in parallel, and
+%furthermore allows scheduling decisions theselves to be parallelized, as the statements
+%in different functions can be scheduled separately.  As a practical matter
+%our Legion implementation only parallelizes function calls---the other
+%statements within a function body are executed in sequential
+%order.
+Consider just the first two loops with subtasks in the example in Listing~\ref{lst:code_ex}.
+The calls to {\tt calc\_new\_currents} are independent because the region arguments that are written
+are all disjoint, and the Legion scheduler will run them in parallel, resources permitting.  Each of the following calls to
+{\tt distribute\_charge} will be dependent on some of the preceding calls to {\tt calc\_new\_currents} and will be delayed until
+all the tasks they depend on have completed.
+If {\tt calc\_new\_current} or {\tt distribute\_charge} have tasks in their implementation, the scheduling for those tasks will be handled only considering the local dependences within their function bodies, and
+correctness is still guaranteed by Lemma~\ref{lem:scheduling}.

### 2011-11-07 doc/pldi2012/highlevel.tex

@@
 
-The biggest challenge to implementing the high-level
+The biggest challenge in implementing the high-level
 runtime is to avoid the latencies associated with distributed
@@
 in solving this problem, the high-level runtime employs two useful 
-scheduling techniques.  The first technique was proved
-in Section \ref{sec:exec}: all scheduling decisions for a
-task can be made locally in the context in which the
+scheduling techniques.  The first technique, validated in 
+Section~\ref{sec:exec}, is to make all scheduling decisions for a
+task locally, i.e. in the context in which the
 task was created.  The second technique is a {\em deferred execution model} 
 that allows the runtime to hide the latencies associated
-with moving data through deep memory hierarchies.  A
-deffered execution model decouples the issuing of operations
+with moving data through deep memory hierarchies by decoupling the issuing
+of operations
 from when they are actually performed.  This model is supported
@@
 
-A deffered execution model however presents challenges to implementing
-the semantics of the Legion programming model.  Deffered execution
-does not necessarily imply that tasks will be executed in the order in
-which they are scheduled, but only in the order in which dependences
-are expressed.  To orchestrate the execution of Legion
-programs in this environment the high-level runtime is architected
-in a manner similar to a pipelined out-of-order hardware processor, but at
-a coarser granularity with tasks corresponding to instructions and
-regions corresponding to registers.  There is a pipeline of stages to 
-executing a task which allows the runtime to extract
-as much task-level parallelism as possible from an application whithin
-the constraints of the Legion programming model.  To illustrate
+A deferred execution model however presents challenges to implementing
+the semantics of the Legion programming model.  Deferred operations
+may not be executed in the order they were scheduled if no dependency
+is expressed between them.  To orchestrate the execution of Legion
+programs in this environment the high-level runtime uses an architecture
+similar to that of a pipelined out-of-order hardware processor, but at
+a coarser granularity.  Tasks can be though of as corresponding to instructions
+and regions to registers.  The runtime uses a pipeline of task execution stages
+to extract
+as much task-level parallelism as possible from an application, while
+preserving the guarantees of the Legion programming model.  To illustrate
 the implementation of the high-level runtime we now describe the
@@
 \label{subsec:depanalysis}
-The first step in the execution of a task is dependence analysis.  When
-a task is registered with the high-level runtime the task call includes
-information about both the logical regions that the task will require
-as well as the read-write privileges and coherence properties
-for each of the regions.  The runtime uses this information to perform
+The first step in the execution of a task is dependence analysis.  Tasks are
+registered with the high-level runtime in sequential program order.  The
+registration lists the logical
+regions to be accessed, specifying the needed privileges and the coherence.
+The runtime uses this information to perform
 dependence detection with each of the previous tasks that have been registered
-in the same context.  Dependence detection only needs to be performed at
-the scope of an enclosing context as was proved in Section \ref{sec:exec}.
+in the same context (i.e. called by the same parent task).  These checks will
+be sufficient to ensure correct program order, as was proved in Section~\ref{sec:exec}.
 
-We initially describe how dependences are detected from logical regions
-in Section \ref{subsec:cohdep}.  We then describe the algorithm that
-is used to perform dependence detection in Section \ref{subsec:depdetect}.
-Section \ref{subsec:dataflow} describes how data is versioned correctly
-during deffered execution.  For concreteness we describe how dependence
-analysis is performed for the circuit simulation from Section \ref{sec:ex}.
+%We initially describe how dependences are detected from logical regions
+%in Section \ref{subsec:cohdep}.  We then describe the algorithm that
+%is used to perform dependence detection in Section \ref{subsec:depdetect}.
+%Section \ref{subsec:dataflow} describes how data is versioned correctly
+%during deferred execution.  For concreteness we describe how dependence
+%analysis is performed for the circuit simulation from Section \ref{sec:ex}.
+
 %When performing dependence analysis the runtime relies on its knowledge about
@@
 \label{subsec:cohdep}
-To detect dependencies between any two tasks using a logical region, we leverage
-the information contained in the privileges and cohrence properties provided by the
-task call.  Dependency checks are performed pair-wise between regions
-to determine whether one task depends on the other.  When checking for
-dependencies we first examine the region privileges.  If both tasks have reads privileges, then 
-there is never a dependence.  If at least one task has write privileges we use table
-\ref{tab:depsdetect} to determine whether there is a dependence.
+Conceptually, dependence detection with a previous task requires comparing the two sets of
+logical regions being accessed.  For each logical reason from the previous task
+that may overlap with a logical region needed by the new task, the privileges
+and coherence modes are compared to determine whether a dependence exists.  If
+only read privileges are in use, there is never a conflict, but if either or
+both task needs write or reduction privileges, the coherence modes are compared
+according to this table:
+
+\vspace{2mm}
+{\small
+\begin{tabular}{c|cccc}
+             & Exclusive & Atomic   & Simultaneous & Relaxed \\
+\midrule
+Exclusive    & Dep & Dep & Dep & Dep \\ 
+Atomic       & Dep & Same & Cont & Cont \\
+Simultaneous & Dep & Cont & Same & None \\
+Relaxed      & Dep & Cont & None & None \\
+\end{tabular}
+}
+\vspace{2mm}
+
+An entry of {\tt Dep} means there is always a dependence, whereas {\tt None}
+means that there is no dependency created by this region pair.  An entry of
+{\tt Same} indicates a dependency that can be avoided if the two tasks are
+able to use the same physical region.  Finally, an entry labeled {\tt Cont}
+indicates a dependency that can be avoided if a single writer is using the
+atomic coherence mode.  When a dependence is detected, the new task will
+generally have to wait for the older task to complete before it can begin
+execution.
+
+%For each previously registered task, 
+%To detect dependencies between any two tasks using a logical region, we leverage
+%the information contained in the privileges and cohrence properties provided by the
+%task call.  Dependency checks are performed pair-wise between regions
+%to determine whether one task depends on the other.  When checking for
+%dependencies we first examine the region privileges.  If both tasks have reads privileges, then 
+%there is never a dependence.  If at least one task has write privileges we use table
+%\ref{tab:depsdetect} to determine whether there is a dependence.
+
 %In this table, {\em Dependence} indicates that there will always be a dependence between
@@
 %a copy between the two tasks running.
-For the case of {\tt Contingent} there are two special cases that we handle.
-One special case is Write-After-Read (WAR) dependencies.  For the WAR case, if neither coherence property
-is {\em Exclusive} then we can ensure that different physical instances
-are used to avoid an anti-dependence and increase parallelism.  If either property is {\em Exclusive} then the semantics
-of the programming model mandate that we detect a dependence and serialize the task.
+%For the case of {\tt Contingent} there are two special cases that we handle.
+%One special case is Write-After-Read (WAR) dependencies.  For the WAR case, if neither coherence property
+%is {\em Exclusive} then we can ensure that different physical instances
+%are used to avoid an anti-dependence and increase parallelism.  If either property is {\em Exclusive} then the semantics
+%of the programming model mandate that we detect a dependence and serialize the task.
 
-Another special case occurs if there is an {\tt Atomic} writer and either {\tt Simultaneous}
-or {\tt Relaxed} reader.  In this case it is acceptable for the tasks to run in
-parallel as it is valid for the reader to see updates form the writer, and the
-reader will not violate the atomic access semantics of the writer.
-\begin{center}
-\begin{table}
-\small
-\begin{tabular}{|c|c|c|c|c|} \hline
-             & Exclusive & Atomic   & Simultaneous & Relaxed \\ \hline 
-Exclusive    & Dep & Dep & Dep & Dep \\ \hline
-Atomic       & Dep & Same & Cont & Cont \\ \hline
-Simultaneous & Dep & Cont & Same & None \\ \hline
-Relaxed      & Dep & Cont & None & None \\ \hline
-\end{tabular}
-\caption{Dependence detection cases when at least one access has a write privilege. {\tt Dep}
-indicates a dependence is detected.  {\tt Same} indicates a dependence is detected
-only if the task uses different instances.  {\tt Cont} means that the dependence
-is contingent upon the permissions.  {\tt None} means there will never
-be a conflict.\label{tab:depsdetect}}
-\end{table}
-\end{center}
+%Another special case occurs if there is an {\tt Atomic} writer and either {\tt Simultaneous}
+%or {\tt Relaxed} reader.  In this case it is acceptable for the tasks to run in
+%parallel as it is valid for the reader to see updates form the writer, and the
+%reader will not violate the atomic access semantics of the writer.
+%XS\begin{center}
+%\begin{table}
+%\small
+%\begin{tabular}{|c|c|c|c|c|} \hline
+%             & Exclusive & Atomic   & Simultaneous & Relaxed \\ \hline 
+%Exclusive    & Dep & Dep & Dep & Dep \\ \hline
+%Atomic       & Dep & Same & Cont & Cont \\ \hline
+%Simultaneous & Dep & Cont & Same & None \\ \hline
+%Relaxed      & Dep & Cont & None & None \\ \hline
+%\end{tabular}
+%\caption{Dependence detection cases when at least one access has a write privilege. {\tt Dep}
+%indicates a dependence is detected.  {\tt Same} indicates a dependence is detected
+%only if the task uses different instances.  {\tt Cont} means that the dependence
+%is contingent upon the permissions.  {\tt None} means there will never
+%be a conflict.\label{tab:depsdetect}}
+%\end{table}
+%\end{center}
 
@@
 \label{subsec:depdetect}
-When a task is registered, the runtime
-will perform dependence detection for every region that the task requests with
-prior tasks that have been registered in the same context.  Since
-tasks are registered in program order we will always perform dependence detection
-in program order which guarantees correctness even if tasks are executed out-of-order.  
-Performing dependence detection in program order is annalogous to how an out of order
-processor works.  Out-of-order processors load instructions in order and only after
-analyzing their instructions and register dependencies do they execute instructions
-out of order.  The regions that a task requires are annalogous
-to the registers listed in an instruction and the goal of the high-level
-runtime is to detect the same dependences and use them to enforce scheudling constraints
-upon a task.  However, there is a catch to this analogy: in the Legion programming
-model two different logical regions are not necessarily independent from each other.  In
-order to perform dependence detection we rely on knowledge of the forest of regions to be
-able to detect dependences between tasks which require different logical regions.
 
-To perform dependence detection, we first find the {\em root region} for the 
+Although the above algorithm for dependence detection is correct, the
+high-level runtime is able to perform the detection much more efficiently with
+the help of a custom data structure.
+
+To perform dependence detection, the runtime will first find the {\em root region} for the 
 requested region.  The root region is the ancestor region for which the parent task
-has privliges.  Once we have found the root region for the requested region, we then proceed to walk
-down the region tree from the root region to the node
-for the requested region.  At each node, we determine if any potential dependences exist.
-Determining whether dependencies exist involves checking information both about the current
-node being visited as well as whether there are any potential dependences farther down
-in the region tree.
+has priviliges.  Once the root region for the requested region is found, 
+the runtime can walk down the region tree from the root region to the node
+for the requested region.  Each node contains a list of tasks that are
+actively using the region corresponding to the node, potential dependencies are
+discovered by comparing privileges and coherence modes with any tasks on that
+list.
 
-When traversing a region node, we begin by checking if there are any tasks actively using
-the node.  We determine this using the pair-wise dependence analysis described 
-in Section \ref{subsec:cohdep} to detect dependences between the region we are looking for and the
-region represented by the current node.  For each dependence that is found, we record that the previous
-task must have completed prior to the start of the task we are registering.
+After check the list of tasks actively using the current node, the runtime
+checks to see if one of the partitions of current node is {\em open}.  An
+open partition indicates that there are some active tasks using subregions
+of the current region.  No more than one partition of a node can be open
+at any given time.  If the open partition is is along the path to the desired
+region's node, traversal continues.  If the open partition lies along another 
+path, or if the current node is the intended destination, the runtime must
+{\em close} the currently active partition because it cannot guarantee
+disjointness with the region the new task wishes to access.  To close a
+partition in the region tree, the runtime must find all active tasks in that
+subtree and consider them all to be dependencies for the task currently being
+registered.  With the previously active partition closed, the runtime can 
+then open up the desired partition and traverse further down the tree as
+needed.
 
-After performing
-dependence analysis we then check if there is an open partition which may contain additional
-dependencies.  If the open partition is the partition we intend to traverse then we
-simply continue the traversal.  However, if the partition we want is not open then we must
-first {\em close} up any open partition.   The act of closing a partition 
-involves traversing the region tree and recording
-all active tasks as dependences.  The reason for closing a partition is that we have no disjointness
-information about tasks that use different partitions and therefore to be safe must register
-all task touching subregions as dependences.  After closing any open partitions, we
-can then open the partition on our path and continue the traversal.
+When the runtime traverses an open partition, its behavior depends on the kind
+of partition being traversed.  If the partition is disjoint, no other subregion
+can overlap with the target region, so the state of the other subregions can
+be ignored in the traversal.  However, if the partition is aliases, any one
+of those other subregions could overlap with the current partition, so
+the runtime must close all of the open subregions (recording new dependencies
+as it does so) before opening the subregion it intends to.
 
-When we reach our target logical region we again perform the pair-wise dependence analysis.
-We also close up any open partitions as they might contain tasks which could contain
-a dependence.  If our task was dependent on all of the previous active tasks in a region
-we say that our task was a {\em dominator} of all the previous active tasks.  We can
-then clear the list of active tasks and replace it with the current task.  This is safe since
-any later tasks that depend on the current task will have a transitive dependence 
-through the dominator task.  If our task is not a dominator task, then we simply append
-it to the list of active tasks making it possible for future task registrations to detect
-it as a dependence.
-
-When traversing partition nodes we perform a similar, but less expensive analysis.  There
-are two cases to consider for partition nodes: disjoint partitions and aliased partitions.
-In the case of disjoint partitions, we simply open the subregion that we intend to traverse
-and then traverse it.  This is correct because we know by definition that any two 
-subregions in a disjoint partition are disjoint and therefore a dependence can never exits
-between them.
-
-For aliased partitions we have to perform an additional dependence analysis.  At every
-aliased partition node we maintain a list of tasks active in all of the subregions.  We
-perform a pair-wise dependence analysis between every pair of tasks regardless of the subregion
-they are accessing to see if a dependence could exist.  We do this regardless of the subregion
-because we have no disjointness information in an aliased partition.  If there are any
-potential dependences, we close up all the subregions and then open the subregion required.  If
-no dependence was detected then we open the subregion we need, append the task to the list
-of active tasks, and continue the traversal.  Allowing multiple open subregions in an aliased
-partition is safe because we performed an analysis between all active tasks regardless of
-whether the particular subregions were aliased or not.
+After the runtime reaches the target logical region, it performs dependence
+checks against any active tasks listed for that region.  If a dependence is 
+discovered, it is added to the current task's dependence list.  It can then
+be removed from the region tree's active task list, as the current task (which
+will be pushed onto the active task list) {\em dominates} the previous task.
+In a case where the new task has dependencies with all existing active tasks
+for the target node, that whole list is replaced by a single entry for the
+new task.
 

### 2011-11-07 doc/pldi2012/lowlevel.tex

@@
 
-The purpose of the low-level runtime is to provide a machine-independent layer
-for the high-level runtime.  This abstraction will provide portability to the Legion
-programming model and the high-level runtime.  To achieve this goal the low-level
-runtime needs to hide the ideosynchracies of the underlying machine.  At the same time, it needs to avoid 
-straying too far from the native capabilities of the underlying hardware 
-and software in order to obtain an efficient implementation.  The Legion
-low-level runtime does this by abstracting the machine into a base set
-of objects required to implement the Legion high-level.  These objects,
+The low-level runtime is a machine-independent layer that provides portability
+to the high-level runtime and consequently the Legion programming model.
+The low-level runtime hides the ideosynchracies of the underlying machine while
+at the same time not straying 
+too far from the native capabilities of the underlying hardware.  
+The machine abstraction that the Legion low-level runtime presents is a base set
+of objects.  These objects,
 include processors, memories, regions, events, and deffered locks.
@@
 \label{subsec:events}
-The low-level runtime supports a deffered execution model to help hide
+The low-level runtime supports a deffered execution model to hide
 the latencies associated with communicating in a distributed memory
@@
 immediately returned to the caller.  The caller may query whether the event has {\em triggered}
-which happens when an even occurs.  The common use case for events is
+which happens when the operation corresponding to the event completes.  The common use case for events is
 to provide them as prerequisites to subsequent requests made of the runtime. 
 These dependences cause the later requests to be automatically deferred until the previous
-operations have completed.  By using events as dependences, the caller can
+events have triggered.  By using events as dependences, the caller can
 set up complicated, but coordinated chains of operations to be executed without
@@
 Events are used to provide a deterministic ordering between operations, but in
-some cases is overly restrictive.  To implement the more relaxed coherence
+some cases are overly restrictive.  To implement the more relaxed coherence
 properties supported in Legion we need a less restrictive ordering semantics that
@@
 
-\subsubsection{Region Meta-Data Objects}
+\subsubsection{Regions}
 \label{subsec:regionmeta}
@@
 \label{subsec:smpimpl}
-The SMP implementation of the low-level runs on any SMP computer that supports 
+The SMP implementation of the low-level runs on any SMP machine that supports 
 POSIX threads.  A single memory object is created for the shared memory, and 

### 2011-11-07 doc/pldi2012/highlevel.tex
high level

@@
 \label{sec:highlevel}
-In this Section we describe the high-level runtime system
-that is designed to implement the Legion programming
-and execution models while remaining machine agnostic for portability.  
-The high-level runtime system operates on top of the 
-low-level runtime which hides machine specific details
-and is described in Section \ref{sec:lowlevel}.
+%In this Section we describe the high-level runtime system
+%that is designed to implement the Legion programming
+%and execution models while remaining machine agnostic for portability.  
+%The high-level runtime system operates on top of the 
+%low-level runtime which hides machine specific details
+%and is described in Section \ref{sec:lowlevel}.
 
-The biggest challenge in implementing the high-level
+The major challenge in implementing the high-level
 runtime is to avoid the latencies associated with distributed
@@
 in solving this problem, the high-level runtime employs two useful 
-scheduling techniques.  The first technique, validated in 
+scheduling techniques.  The first, validated in 
 Section~\ref{sec:exec}, is to make all scheduling decisions for a
 task locally, i.e. in the context in which the
-task was created.  The second technique is a {\em deferred execution model} 
+task was created.  The second is a {\em deferred execution model} 
 that allows the runtime to hide the latencies associated
@@
 of operations
-from when they are actually performed.  This model is supported
-by features in the low-level runtime.
+from when they are performed.  This model is supported
+by features in the low-level runtime (see Section~\ref{sec:lowlevel}).
 
@@
 
-A deferred execution model however presents challenges to implementing
-the semantics of the Legion programming model.  Deferred operations
-may not be executed in the order they were scheduled if no dependency
-is expressed between them.  To orchestrate the execution of Legion
-programs in this environment the high-level runtime uses an architecture
-similar to that of a pipelined out-of-order hardware processor, but at
-a coarser granularity.  Tasks can be though of as corresponding to instructions
-and regions to registers.  The runtime uses a pipeline of task execution stages
-to extract
+A problem with deferred execution is that deferred operations must
+still be executed in the correct order if they are dependent.
+To ensure correctness the high-level runtime uses an architecture
+similar to a pipelined out-of-order hardware processor, but at
+the coarser granularity of tasks and regions rather than instructions and registers.
+The stages of the task execution pipeline seek to extract
 as much task-level parallelism as possible from an application, while
-preserving the guarantees of the Legion programming model.  To illustrate
-the implementation of the high-level runtime we now describe the
-details of how each stage of task execution is performed.
+preserving dependencies.  We describe some
+details of the four pipeline stages: dependence analysis, mapping, execution, and
+task completion.
 
@@
 \label{subsec:depanalysis}
-The first step in the execution of a task is dependence analysis.  Tasks are
-registered with the high-level runtime in sequential program order.  The
-registration lists the logical
-regions to be accessed, specifying the needed privileges and the coherence.
+The first step in the execution of a task is dependence analysis.  Within a task call, subtasks are
+registered with the high-level runtime in sequential execution order.  The
+registration lists the logical regions to be accessed with needed privileges and coherence.
 The runtime uses this information to perform
-dependence detection with each of the previous tasks that have been registered
-in the same context (i.e. called by the same parent task).  These checks will
-be sufficient to ensure correct program order, as was proved in Section~\ref{sec:exec}.
+dependence detection with each of the subtasks already registered
+in the same context (i.e. called by the same parent task).  
+%These checks will
+%be sufficient to ensure correct program order, as was proved in Section~\ref{sec:exec}.
 
@@
 \label{subsec:cohdep}
-Conceptually, dependence detection with a previous task requires comparing the two sets of
-logical regions being accessed.  For each logical reason from the previous task
-that may overlap with a logical region needed by the new task, the privileges
-and coherence modes are compared to determine whether a dependence exists.  If
-only read privileges are in use, there is never a conflict, but if either or
-both task needs write or reduction privileges, the coherence modes are compared
-according to this table:
+Conceptually, dependence detection between a new task $t$ and a previous task $t'$ requires comparing the two sets of
+logical regions accessed.  For each logical region from $t'$
+that may overlap with a logical region needed by $t$, the privileges
+and coherence modes are compared to determine whether a dependence exists.  Read
+privileges never cause conflicts, but if either 
+task needs write or reduction privileges, the coherence modes are compared:
 
@@
 
-An entry of {\tt Dep} means there is always a dependence, whereas {\tt None}
-means that there is no dependency created by this region pair.  An entry of
-{\tt Same} indicates a dependency that can be avoided if the two tasks are
-able to use the same physical region.  Finally, an entry labeled {\tt Cont}
-indicates a dependency that can be avoided if a single writer is using the
-atomic coherence mode.  When a dependence is detected, the new task will
-generally have to wait for the older task to complete before it can begin
-execution.
+An {\tt Dep} entry indicates a dependence, whereas {\tt None}
+means no dependence.  A
+{\tt Same} entry indicates a dependence unless the two tasks 
+use the same {\em physical region}.  Finally, a {\tt Cont} entry
+indicates a dependence unless a single writer is using the
+atomic coherence mode.  When a dependence is detected, the new task 
+waits for the older task to complete before it begins execution.
+
+The table lists coherence modes we have not yet discussed.  Atomic
+allows tasks to be reordered as long as the result is
+serializable.  Both simultaneous and relaxed allow other tasks to
+execute at the same time.  The difference comes in the restriction of
+what data must be observed in the different modes.  In Simultaneous
+mode, a task must see all updates to the logical region made by other
+tasks operating on the same region at the same time.  In Relaxed mode,
+a task is ambivalent as to whether it observes any updates to the
+logical region by other tasks executing at the same time.
 
@@
 
-Although the above algorithm for dependence detection is correct, the
-high-level runtime is able to perform the detection much more efficiently with
-the help of a custom data structure.
+To detect dependencies efficiently, the high-level runtime maintains a
+local region forest (recall Section~\ref{sec:ex}) for each task.  The
+root regions are the task's region arguments, and the forest includes
+all partitions and subregions created within the task.  To perform
+dependence detection on a region $r$, the runtime first finds a root
+region $r'$ such that $r \rleq r'$.  The runtime then walks the region
+forest from $r'$ to $r$.  Each region $r''$ on the path contains a
+list $L$ of registered tasks using $r''$, and potential dependencies
+are discovered by comparing privileges and coherence modes with the
+tasks in $L$.
 
-To perform dependence detection, the runtime will first find the {\em root region} for the 
-requested region.  The root region is the ancestor region for which the parent task
-has priviliges.  Once the root region for the requested region is found, 
-the runtime can walk down the region tree from the root region to the node
-for the requested region.  Each node contains a list of tasks that are
-actively using the region corresponding to the node, potential dependencies are
-discovered by comparing privileges and coherence modes with any tasks on that
-list.
-
-After check the list of tasks actively using the current node, the runtime
-checks to see if one of the partitions of current node is {\em open}.  An
-open partition indicates that there are some active tasks using subregions
-of the current region.  No more than one partition of a node can be open
-at any given time.  If the open partition is is along the path to the desired
-region's node, traversal continues.  If the open partition lies along another 
-path, or if the current node is the intended destination, the runtime must
-{\em close} the currently active partition because it cannot guarantee
+The runtime next checks if a partition of $r''$ is {\em open}.  An
+open partition indicates that there are active tasks using subregions
+of $r''$.  No more than one partition of a node can be open
+at any time.  If the open partition is is along the path to $r$,
+traversal continues.  If the open partition lies along another 
+path the runtime must {\em close} the currently active partition because it cannot guarantee
 disjointness with the region the new task wishes to access.  To close a
-partition in the region tree, the runtime must find all active tasks in that
-subtree and consider them all to be dependencies for the task currently being
-registered.  With the previously active partition closed, the runtime can 
-then open up the desired partition and traverse further down the tree as
+partition, the runtime finds all active tasks in that
+subtree and considers them all dependencies for the new task.  
+With the previously active partition closed, the runtime can 
+then opens the desired partition and traverses further down the tree as
 needed.
@@
 of partition being traversed.  If the partition is disjoint, no other subregion
-can overlap with the target region, so the state of the other subregions can
-be ignored in the traversal.  However, if the partition is aliases, any one
-of those other subregions could overlap with the current partition, so
-the runtime must close all of the open subregions (recording new dependencies
-as it does so) before opening the subregion it intends to.
+can overlap with the target region, so the state of other subregions is
+ignored.  However, if the partition is not disjoint, any 
+of subregion could overlap with the desired subregion, so
+the runtime must close all open subregions (recording new dependencies
+as it does so) before continuing the traversal.
 
@@
 \label{subsec:dataflow}
-One aspect of the deffered execution model is that it allows dependence analysis to run
-ahead of the actual execution of tasks.  As a result the runtime must maintain information
-about which {\em physical instances} of a logical region a task is allowed to use when executing.
-A physical instance is a copy of the data in a logical region.  There can be multiple
-physical instances of a logical region.  It is the responsibility of the high-level runtime
-to ensure tasks use instances corresponding to the appropriate version of data.
 
+A {\em physical instance} is a copy of the data in a logical region.  Multiple
+physical instances of a logical region may exist simultaneously, and it is the responsibility of the high-level runtime
+to ensure tasks use the correct version.
+In an out-of-order processor this problem is solved by register-renaming which maps logical
+registers to a larger set of physical registers.  Our solution employs a similar technique
+that maps logical regions to a larger set of {\em abstract instances}.  An abstract instance
+is a place holder for the set of physical instances a task will reference when it actually runs.  
+Abstract instances are needed because we may not know the actual physical instances a task will
+use until its predecessors in the dependence graph are {\em mapped} 
+(see Section \ref{subsec:mapping}).  
+To determine the abstract instances of a region, we keep track of 
+the previous valid abstract instances that were observed while traversing the region forest
+during dependence analysis.  Every region in the region forest that has active tasks
+has a valid abstract instance.  Every time we close a region we close all the valid
+abstract instances.  When a region is reopened it must then create a new abstract instance
+to reflect that new physical instances must be created (i.e., a fresh version of the data must be copied from
+the parent region).  In addition, whenever a task writes a region, a new abstract instance is usually created
+to represent the new version of the data.
+
+%For each region in
+%a task we record the source and destination abstract instances that correspond to the physical
+%instances from which a task will read its input data and then place its output data.
+%When the task is mapped as described in section \ref{subsec:mapping} we will update these
+%abstract instances with the actual physical instances chosen.
 %In addition to computing the dependences between tasks, we also have to compute the version
@@
 
-In an out-of-order processor this problem is solved by register-renaming which maps logical
-registers to a larger set of physical registers.  Our solution employs a similar technique
-that maps logical regions to a larger set of {\em abstract instances}.  An abstract instance
-is a place holder for the set of physical instances which a task will reference.  We call these
-abstract instances because we may not know the actual physical instances until a task's predecessors
-are mapped as described in Section \ref{subsec:mapping}.  For each region in
-a task we record the source and destination abstract instances that correspond to the physical
-instances from which a task will read its input data and then place its output data.
-%When the task is mapped as described in section \ref{subsec:mapping} we will update these
-%abstract instances with the actual physical instances chosen.
-
-To determine the source and destination abstract instances of a region, we keep track of 
-the previous valid abstract instances that were observed while traversing the region tree
-during dependence analysis.  Every region node in the region tree that has active tasks
-has a valid abstract instance.  Every time we close a region tree we close all the valid
-abstract instances.  When a node is reopened it must then create a new abstract instance
-to reflect that new physical instances must be created.
 %If a region
@@
 
-Abstract instances correspond to a dynamic construction of the dataflow graph.  Each abstract
-instance is a node in the dataflow graph corresponding to a specific version of a logical 
-region and its instantiations as physical regions.  Tasks are then edges between abstract
-instances indicating operations on different versions of data.  By computing this information
-during dependence detection, we can record it before tasks begin executing out of order allowing
-us to guarantee program correctness.  
+%Abstract instances correspond to a dynamic construction of the dataflow graph.  Each abstract
+%instance is a node in the dataflow graph corresponding to a specific version of a logical 
+%region and its instantiations as physical regions.  Tasks are then edges between abstract
+%instances indicating operations on different versions of data.  By computing this information
+%during dependence detection, we can record it before tasks begin executing out of order allowing
+%us to guarantee program correctness.  
 %Continuing
@@
 To illustrate concretely how dependence analysis works, we describe the steps for
-performing dependence analysis for circuit example from Section \ref{sec:ex}.  
-For conciseness we will only show the node region tree.  At the start of the task, there exists
-a single abstract instance at the root node representing the valid set of physical
+performing dependence analysis for the circuit simulation in Listing~\ref{lst:code_ex}.
+For conciseness we will only show the node region forest.  At the start of the task, there exists
+a single abstract instance at the root representing the valid set of physical
 instances from the parent task. 
-
 When each of the {\tt compute\_new\_current} tasks are issued they must perform dependence
 analysis for two node regions.  In the {\tt private\_node} case they will first open
-up the {\tt p\_nodes\_pvs} partition and then open up the {\tt p\_pvt\_nodes} sub partition
+up the {\tt p\_nodes\_pvs} partition and then open up the {\tt p\_pvt\_nodes} sub-partition
 for each of their subregions.  We use a bold circle on an edge to indicate that it has been
@@
 {\tt p\_ghost\_nodes} partition is aliased because all the ghost accesses are read-only.
-The state of the region tree after dependence analysis has been performed for the first
+The state of the region forest after dependence analysis has been performed for the first
 {\tt compute\_new\_current} task can be seen in figure~\ref{sfig:mapping_fig:cnc}.
 
-In the second phase of the computation, all of the {\tt distribute\_charge} tasks
-will be issued.  For these tasks they will first traverse the region tree
+In the second phase of the computation, the {\tt distribute\_charge} tasks
+are issued.  For these tasks they will first traverse the region forest
 for the regions containing their private nodes.  The path to these regions are
-already open, but when the dependence analysis arrives at the specific private node subregion it will
-discover a dependence between the task and the {\tt compute\_new\_currents} task that operated
-on that region previously.  This dependence will be recorded and the 
-{\tt distribute\_charge} task will be made the new active task as it dominated the
-previous task.
+already open, but when the dependence analysis arrives at the specific private node subregion it 
+discovers a dependence between the task and the {\tt compute\_new\_currents} task that operated
+on that region previously.  This dependence is recorded and the 
+{\tt distribute\_charge} task becomes the new active task.
 
-For the ghost node regions, the dependence analysis will discover that the
-there is a dependence between each of the {\tt distribute\_charge} tasks and all of the
+For the ghost node regions, dependence analysis discovers 
+a dependence between each of the {\tt distribute\_charge} tasks and all of the
 {\tt compute\_new\_current} tasks because the {\tt distribute\_charge} tasks are
@@
 partition at the same time even if the subregions they are accessing are aliased.  The
-state of the region tree after dependence analysis for first {\tt distribute\_charge} task can
+state of the region forest after dependence analysis for the first {\tt distribute\_charge} task can
 be seen in figure~\ref{sfig:mapping_fig:dc}.
@@
 are issued requiring their private node region and their shared node region.  Again
-the private node paths are already open, but a WAR dependence is detected which
+the private node paths are already open, but a write-after-read dependence is detected which
 requires a new abstract instance to be created to avoid a dependence.  The
 shared nodes present a more interesting case.  The first shared node to be registered
-will detect a dependence with every prior task since it must open the shared partition
+detects a dependence with every prior task since it must open the shared partition
 instead of the ghost partition.  A new abstract instance is created at the 
@@
 node ensuring that there is no requirement on the ordering of {\tt update\_voltage}
-tasks.  The state of the region tree after the first update voltage task
-has been registered can be seen in figure \ref{sfig:mapping_fig:volt}.
+tasks.  The state of the region forest after the first update voltage task
+has been registered can be seen in Figure~\ref{sfig:mapping_fig:volt}.
 
@@
 \label{subsec:mapping}
-After tasks have undergone dependence analysis, the tasks are placed on either a ready
-queue to be mapped or waiting queue.  A task is placed on the ready queue if all the tasks with
-which it registered a dependence have mapped.  If any of the task's dependences have
-not mapped then the task is placed on the waiting queue.  The task also registers
-itself with the task on which it depends so that when that task is mapped it will notify
-the task on the waiting queue. Tasks are mapped based on the partial ordering created by 
-their data dependences.  By mapping tasks based on data dependences we guarantee the 
-semantics of the programming model, while still allowing tasks to execute out of order.
+After dependence analysis a task is placed on either a {\em ready queue} to be mapped or a 
+The high-level runtime maintains the invariant that a task {\em maps} only after all tasks on which it depends have mapped. 
+While some predecessor tasks have not mapped a task remains in the {\em waiting queue}; otherwise it is placed on the {\em ready queue}.
 
-When the scheduler is invoked, it will pull a series of tasks off the ready queue, map them,
+%Tasks are mapped observing the partial ordering created by 
+%their data dependences.  By mapping tasks based on data dependences we guarantee the 
+%semantics of the programming model, while still allowing tasks to execute out of order.
+
+When the scheduler is invoked, it pulls tasks off the ready queue, maps them,
 and then issue them to be executed by the low level runtime.  The process of mapping a task
-consists of several different queries made by the runtime to an application level object
-called a {\em mapper}.  A mapper is an object which has control over how both tasks and
-data can be placed in the machine.  This idea is a dynamic instantiation of the idea
-initially presented in \cite{Fatahalian06} of creating mapping files for specifying how
-tasks and data are placed in a memory hierarchy.  Rather than rely on a static mapping
-we opted for a dynamic interface that could make decisions both based on application specific
-input data as well as the machine specifics both at start-up and throughout the execution
-of an application.  %The interface for a mapper is described in Section \ref{subsec:mapinter}.
+consists of several different queries made by the runtime to an application level {\em mapper} object.
+A mapper controls how tasks and data can be placed in the machine.  Mappers are a dynamic version of 
+the static mapping files of \cite{Fatahalian06}.  
+%Rather than rely on a static mapping
+%we opted for a dynamic interface that could make decisions both based on application specific
+%input data as well as the machine specifics both at start-up and throughout the execution
+%of an application.  %The interface for a mapper is described in Section \ref{subsec:mapinter}.
 
-\subsubsection{Mapper Interface}
+%\subsubsection{Mapper Interface}
 The mapper interface is a collection of eight function calls that allow the runtime to query
-the application for information about how to map tasks are regions.  Mapper calls are made
-from the runtime to ask where physical instances of regions should be place in the memory hierarchy,
+the application for information about how to map tasks and regions.  Mapper calls are made
+from the runtime to ask where physical instances of regions should be placed in the memory hierarchy,
 which physical instances should be the source for copy operations, on which processors
@@
 
-These eight task calls give the programmer complete control over where all tasks are executed, and
-where all data is placed within the memory hierarchy.  While this interface does give the programmer
-complete control over data and task movement, it will never allow the programmer to affect the
-correctness of the program.  The high-level runtime still manages all the task dependences and the
+These eight mapper calls give the programmer complete control over where all tasks are executed, and
+where all data is placed within the memory hierarchy.  While this interface allows the programmer
+to tune performance, it cannot affect correctness.  The high-level runtime still manages all the task dependences and the
 ordering in which tasks are executed and copies are performed.  Therefore, regardless of choices
 made by any mapper, the resulting program will always be executed correctly.
+To make it easy for programmers to write programs, we also provide a default
+ mapper that uses simple heuristics to place tasks and data.  
+%For example, when
+%performing task stealing, the default mapper only encourages stealing tasks in groups that use
+%the same logical regions, which improves locality by moving tasks that use the same regions
+%together.  The default mapper implementation makes it easy for the programmer to get an 
+%initial program running.  As the programmer tunes the application, he can write mappers that
+%specialize for both application specific and/or machine specific parameters.
 
-In order to make it easy for programmers to write programs, we provide an initial implementation 
-of the mapper interface that uses simple heuristics to place tasks and data.  For example, when
-performing task stealing, the default mapper only encourages stealing tasks in groups that use
-the same logical regions, which improves locality by moving tasks that use the same regions
-together.  The default mapper implementation makes it easy for the programmer to get an 
-initial program running.  As the programmer tunes the application, he can write mappers that
-specialize for both application specific and/or machine specific parameters.
-
-The high-level runtime also supports the use of multiple mappers in the same application.  For each task 
-call the programmer can choose a mapper to be used for the runtime calls made inside that
-task.  This allows for the creation of library code which provides its own mappers indpendent
+The high-level runtime also supports the use of multiple mappers in the same application.  Each task 
+call may (but does not need to) have an associated mapper,
+%runtime  can choose a mapper to be used for the runtime calls made inside that
+%task.  
+which allows for the creation of library code that provides its own mappers indpendent
 from any application level mappers.
@@
 \label{subsec:steal}
-In order to achieve load balancing in an application the runtime supports an interface
-for task stealing.  Our task stealing algorithm is loosely based on the stealing algorithm
-used by Cilk \cite{CILK95} with guidance from the mapper about how stealing should be
+To achieve load balancing in an application the runtime uses
+task stealing.  Our task stealing algorithm is loosely based on the Cilk algorithm
+\cite{CILK95} with guidance from the mapper about how stealing should be
 performed.  Tasks are initially placed on the ready queue on the processor in which they
@@
 the list of steal requests.  For each mapper that issued a steal request, the runtime will
-give to the local instance of that mapper a list of the tasks in the ready queue that that
+give the local instance of that mapper a list of the tasks in the ready queue that that
 specific mapper is responsible for mapping.  The mapper must then determine the set which
-it is going to permit to be stolen if any.  Any tasks that the mapper marks as stolen are
+it is going to permit to be stolen, if any.  Any tasks that the mapper marks as stolen are
 sent back to the stealing processor.  By only allowing mappers to make steal requests about
-the tasks they own, we are giving the mappers complete control over where tasks are mapped.
-In addition, we are also giving the mappers control over how steal requests are made for
+the tasks they own, we give mappers complete control over where tasks are mapped.
+In addition, we also give the mappers control over how steal requests are made for
 the case where the mapper has better knowledge about where additional work is than the 
 runtime.  Note that this mapping interface is sufficiently powerful to allow a user to
-disable all stealing if desired, and to map tasks to explicit processors with no interference
+disable all stealing if desired, and to map tasks to specific processors with no interference
 from the runtime.
@@
 to issue the task to the low level runtime.  It first creates any new physical instances
-for the regions which the task requires.  It then issues the necessary copies required
-for the task to execute contignent on the tasks producing the source physical instances
-have finished.  Then it issues the task itself to the low level processor contingent
+for the regions the task requires.  It then issues copy operations required
+for the task to execute contignent on predecessor tasks producing the source physical instances
+having finished.  Then it issues the task itself to the low level processor contingent
 on all the copies for the task haveing finished as well as the the set of tasks
-on which the issuing task dependends having finished.  These dependences are all encoded
-using event dependencies in the low level runtime.  Events are described
-in detail in Section \ref{subsec:events}.  Continuing the analogy of an out-of-order
+on which the issuing task depends having finished.  These dependences are all encoded
+using {\em events} (see Section \ref{subsec:events}).  Continuing the analogy of an out-of-order
 processor, the execution phase is equivalent to issuing an instruction to an
@@
 We define a task to have {\tt completed} once all of the task's regions reflect
-the updated state of the task and all its sub-tasks.  This reflects an underlying assumption 
-of the mapping process that the regions for which a task has priviliges will represent the data produced 
-by that task.  However, this is not inherently true that a physical instance of a task reflects
-this state after the task finishes executing.  If a task has child tasks these tasks may
-still be executing when the parent task finishes the execution phase.
-
-In order to handle this scenario the runtime must compute the necessary copy operations
+the updated state of the task and all its sub-tasks.  
+%This reflects an underlying assumption 
+%of the mapping process that the regions for which a task has priviliges will represent the data produced 
+%by that task.  However, this is not inherently true that a physical instance of a task reflects
+%this state after the task finishes executing.  
+It is possible for a task to terminate but to have subtasks that are still in the execution phase.
+%If a task has child tasks these tasks may
+%still be executing when the parent task finishes the execution phase.  
+To handle this scenario the runtime must compute the necessary copy operations
 from child task physical instances back to the parent task's physical instances.  To
@@
 have been mapped.  The runtime then issues a close operation to the each of the parent
-task's region trees.  These close operations record all of the necessary copies to
-restore the parent task's regions.  We call these copy operations at the end of 
-a task {\em restoring copies}.
-The runtime then issues these copies with the required dependences
-and marks the task completed once the restoring copy operations are complete.
+task's region forests.  These close operations record all of the necessary copies to
+restore the parent task's regions.  
+%We call these copy operations at the end of  a task {\em restoring copies}.
+The runtime issues these copies with the required dependences
+and marks the task completed once the copy operations are complete.
 This is similar to the write-back phase of an out-of-order processor where store instructions cannot be

### 2011-11-07 doc/pldi2012/code_ex.tex

@@
   for(i = 0; i #$<$# MAX_PIECES; i++) 
-    pieces[i] #$\gets$# { rn_pvt = p_pvt_nodes[i], rn_shr = p_shr_nodes[i],
+    pieces[i] #$\gets$# region_cast<CircuitPiece<c.r_all_nodes,c.r_all_wires> >
+                          { rn_pvt = p_pvt_nodes[i], rn_shr = p_shr_nodes[i],
                              rn_ghost = p_ghost_nodes[i], rw_pvt = p_wires[i] };

### 2011-11-07 doc/pldi2012/type_system.tex
Fixing a few small typos

@@
 and {\tt piece.rn\_pvt} and {\tt piece.rn\_ghost} with {\tt RO} (read-only) privileges.  Because the calling task has {\tt RW} privileges on the parent regions of these
-region arguments, the type checker is able to confirm that the subtask can execute with the needed privileges.
+region arguments, the type checker can confirm that the subtask can execute with the needed privileges.
 
@@
 
-Due the the inability for a packed region relationship to hold privileges, any
+Due to the inability for a packed region relationship to hold privileges, any
 useful region relationship will leave at least one region free (i.e.

### 2011-11-07 doc/pldi2012/highlevel.tex
spell check

@@
 %task.  
-which allows for the creation of library code that provides its own mappers indpendent
+which allows for the creation of library code that provides its own mappers ind pendent
 from any application level mappers.
@@
 for the regions the task requires.  It then issues copy operations required
-for the task to execute contignent on predecessor tasks producing the source physical instances
+for the task to execute contingent on predecessor tasks producing the source physical instances
 having finished.  Then it issues the task itself to the low level processor contingent
-on all the copies for the task haveing finished as well as the the set of tasks
+on all the copies for the task having finished as well as the the set of tasks
 on which the issuing task depends having finished.  These dependences are all encoded
@@
 
-The last part of the completetion phase is garbage collecting physical instances.  Physical
+The last part of the completion phase is garbage collecting physical instances.  Physical
 instances are not a visible part of the programming model and therefore cannot be managed

### 2011-11-07 doc/pldi2012/experiments.tex
low-level

@@
 To measure Legion's ability to distribute work across a distributed machine, 
-two real-world benchmarks were developed and run on a small cluster under a
+two real-world benchmarks were run on a cluster under a
 variety of configurations.  The cluster consists of four nodes connected by a
@@
 
-  \subtable[Execution Time Breakdown]{
+  \subtable[Execution Time Breakdown (all times in s)]{
     \scalebox{0.8}{

### 2011-11-07 doc/pldi2012/lowlevel.tex
low-level

@@
 The low-level runtime is a machine-independent layer that provides portability
-to the high-level runtime and consequently the Legion programming model.
-The low-level runtime hides the ideosynchracies of the underlying machine while
-at the same time not straying 
-too far from the native capabilities of the underlying hardware.  
-The machine abstraction that the Legion low-level runtime presents is a base set
-of objects.  These objects,
-include processors, memories, regions, events, and deffered locks.
+for our Legion implementation.
+%The low-level runtime hides idiosyncracies of the underlying machine while
+%at the same time not straying 
+%too far from the native capabilities of the underlying hardware.  
+The low-level runtime abstracts the machine as a base set of objects:
+events, deferred locks, processors, memories, and regions.
 
@@
 \label{subsec:events}
-The low-level runtime supports a deffered execution model to hide
-the latencies associated with communicating in a distributed memory
-system.  The basic building block of the low-level runtime's deferred execution model
+%The low-level runtime supports a deffered execution model to hide
+%the latencies associated with communicating in a distributed memory
+%system.  
+The basic building block of deferred execution 
 is the {\em event}.  An event is a placeholder for the completion of a deferred
 operation.  When an asynchronous call is made to any of the low-level runtime objects, an event is 
-immediately returned to the caller.  The caller may query whether the event has {\em triggered}
+immediately returned to the caller.  The caller may query whether the event has {\em triggered},
 which happens when the operation corresponding to the event completes.  The common use case for events is
@@
 until the event has triggered and all waiters on the event have been
-notified.  Some of these waiters may take quite a while to query the event.
-Additionaly, we must somehow track how many queries to expect to know
-when they've all been performed.  However, if the event handle is paired with a
+notified.  
+%Some of these waiters may take quite a while to query the event.
+%Additionaly, we must somehow track how many queries to expect to know
+%when they've all been performed.  
+Our events include a
 {\em generation count} that tracks how many times the event has been triggered,
-the event can be reused as soon as the previous generation has been triggered.
+which allows an event to ve reused with a new generation as soon as the previous generation has triggered.
 Holders of an older (handle,generation) pair are able to query the newer
@@
 \label{subsec:defferedlocks}
-Events are used to provide a deterministic ordering between operations, but in
-some cases are overly restrictive.  To implement the more relaxed coherence
-properties supported in Legion we need a less restrictive ordering semantics that
-can result in better performance.  The traditional tool for this job is
+Events provide a deterministic ordering between operations, 
+but the more relaxed coherence do not require strict ordering.  The traditional tool for this job is
 the mutex (a.k.a. lock), and it is from the common reader/writer lock that
@@
 requested it is another important difference between the deferred lock and
-existing mutexes.
+mutexes.
 
@@
 of locks.  Similar to events, a deferred lock can be created on any node
-without communication with the other nodes in the cluster.  That node becomes
+without communication with the other nodes.  That node becomes
 the initial home for the lock, but the deferred lock (unlike events) can 
@@
 the new owner.  The highly unlikely, but possible, case in which multiple 
-forwardings are required is bounded at the number of nodes in the cluster.)
+forwardings are required is bounded at the number of nodes in the machine.)
 The ability to migrate locks between nodes allows the performance of a
@@
 associated with a particular region metadata and spends its entire lifetime in
-the memory objecy in which it was created.  
+the memory object in which it was created.  
 %When multiple physical instances

### 2011-11-07 doc/pldi2012/lowlevel.tex
lowlevel

@@
 {\em generation count} that tracks how many times the event has been triggered,
-which allows an event to ve reused with a new generation as soon as the previous generation has triggered.
+which allows an event to be reused with a new generation as soon as the previous generation has triggered.
 Holders of an older (handle,generation) pair are able to query the newer
@@
 lock requests from a single node to be kept local to that node after a single
-message is sent to the current owner.  (Should the ownership of the lock change
-while the message is in flight, the previous owner forwards the request to
-the new owner.  The highly unlikely, but possible, case in which multiple 
-forwardings are required is bounded at the number of nodes in the machine.)
+message is sent to the current owner.  
+%(Should the ownership of the lock change
+%while the message is in flight, the previous owner forwards the request to
+%the new owner.  The highly unlikely, but possible, case in which multiple 
+%forwardings are required is bounded at the number of nodes in the machine.)
 The ability to migrate locks between nodes allows the performance of a

### 2011-11-07 doc/pldi2012/highlevel.tex
Some minor fixes to the high level section

@@
 that may overlap with a logical region needed by $t$, the privileges
-and coherence modes are compared to determine whether a dependence exists.  Read
-privileges never cause conflicts, but if either 
+and coherence modes are compared to determine whether a dependence exists.  If
+both regions need read
+privileges there is never a dependence, but if either 
 task needs write or reduction privileges, the coherence modes are compared:
@@
 To detect dependencies efficiently, the high-level runtime maintains a
-local region forest (recall Section~\ref{sec:ex}) for each task.  The
+local region forest (recall Section~\ref{sec:exec}) for each task.  The
 root regions are the task's region arguments, and the forest includes
@@
 With the previously active partition closed, the runtime can 
-then opens the desired partition and traverses further down the tree as
+then open the desired partition and traverse further down the tree as
 needed.
@@
 to reflect that new physical instances must be created (i.e., a fresh version of the data must be copied from
-the parent region).  In addition, whenever a task writes a region, a new abstract instance is usually created
-to represent the new version of the data.
+the parent region).  
+%In addition, whenever a task writes a region, a new abstract instance is usually created
+%to represent the new version of the data.
 
@@
 {\tt compute\_new\_current} tasks because the {\tt distribute\_charge} tasks are
-performing writes and the partition is aliased.  However, the runtime will detect this as
+performing writes and the partition is aliased.  However, the runtime will detect these as
 anti-dependences which are resolved by forcing the {\tt distribute\_charge} tasks to
@@
 \label{subsec:mapping}
-After dependence analysis a task is placed on either a {\em ready queue} to be mapped or a 
+%After dependence analysis a task is placed on either a {\em ready queue} to be mapped or a 
 The high-level runtime maintains the invariant that a task {\em maps} only after all tasks on which it depends have mapped. 
@@
 made by any mapper, the resulting program will always be executed correctly.
-To make it easy for programmers to write programs, we also provide a default
+To make it easy for programmers to write programs, we provide a default
  mapper that uses simple heuristics to place tasks and data.  
@@
 %task.  
-which allows for the creation of library code that provides its own mappers ind pendent
+which allows for the creation of library code that provides its own mappers independent
 from any application level mappers.
@@
 a target processor, or indicate that no steal should be attempted.  The runtime will then
-send steal requests to each processor on behalf of each specific mapper.
+send steal requests to each processor on behalf of each mapper.
 
@@
 give the local instance of that mapper a list of the tasks in the ready queue that that
-specific mapper is responsible for mapping.  The mapper must then determine the set which
+specific mapper is responsible for mapping.  The mapper must then determine the set of tasks which
 it is going to permit to be stolen, if any.  Any tasks that the mapper marks as stolen are
@@
 for the task to execute contingent on predecessor tasks producing the source physical instances
-having finished.  Then it issues the task itself to the low level processor contingent
+having finished.  Then it issues the task to the low level processor contingent
 on all the copies for the task having finished as well as the the set of tasks
@@
 \label{subsec:cleanup}
-We define a task to have {\tt completed} once all of the task's regions reflect
+We define a task to have {\em completed} once all of the task's regions reflect
 the updated state of the task and all its sub-tasks.  

### 2011-11-07 doc/pldi2012/experiments.tex

@@
 differences in scaling behavior compared to the circuit simulation.  The time
-spent performing copies is much smaller on absolute scale, but shows the same
+spent performing copies is much smaller on an absolute scale, but shows the same
 large step when leaving a single node, but then relative insensitivity to the
@@
 in the remotely-accessible memory windows provided by GASNet.  Although access
-to this memory is significantly slower than accesses to a node's system memory,
-being able to avoid copies entirely in some cases is a net win.
+to this memory is slower than accesses to a node's system memory,
+avoiding copies entirely in some cases is a net win.
 

### 2011-11-07 doc/pldi2012/lowlevel.tex
removing fluff

@@
 %forwardings are required is bounded at the number of nodes in the machine.)
-The ability to migrate locks between nodes allows the performance of a
+%The ability to migrate locks between nodes 
+This allows the performance of a
 heavily-contended cluster-level lock to approach that of a single-node lock.
@@
 but is done per-object rather than on cache lines.  This is currently only 
-used for object within the runtime itself, but we are exploring ways to expose
+used for objects within the runtime itself, but we are exploring ways to expose
 it at the Legion application level.
 
-\subsubsection{Processors}
-\label{subsec:processors}
-Hardware units capable of performing computation are captured as generic
-processors in the low-level runtime.  Like other runtime objects, a processor
-has a globally unique name and can be referred to anywhere in the system.
-Processors come in different \emph{kinds} (e.g. X86 CPU core, CUDA-capable GPU),
-but they all support the same
-interface and execution model.  A request to spawn a task on a given processor
-may come from anywhere in the system.  The request names which task should
-be spawned and provides input arguments and zero or more events.  Once those
-events have occurred, the task is placed into a queue of ready tasks, which
-the processor works through in the order in which they were enqueued.  A
-processor will generally run a task to completion before taking another, but
-will try to run another task (resources permitting) if the current task needs
-to wait for an event to occur.  Once a task is placed on a processor's queue,
-it may not be moved or deleted.  Load-balancing of tasks is expected to be done
-by the high-level runtime.  The primary purpose of the queue for the low-level
-Processor is to be able to hide the latency of starting new tasks.
+\subsubsection{Regions}
+\label{subsec:regionmeta}
+Application-level regions are broken into three inter-related pieces in the
+low-level runtime.  The metadata object tracks global information such as
+region size and valid mask.  Allocators are used to dynamically allocate and
+free entries in a region.  Finally, instances hold versions of the data for 
+a logical region.
 
@@
 hardware, freeing up the processors to perform more interesting
-computation.
+computation.  Finally, the runtime is able to provide memory-specific accessors,
+lightweight objects that allow a given task to access a given physical region
+instance.  In the most common case, reads and write access is done through a 
+single array dereference, but the interface allows seamless use of other
+access modes (e.g. use of RDMA puts and gets for access to remote memory).
 
-\subsubsection{Regions}
-\label{subsec:regionmeta}
-Application-level regions are broken into four inter-related pieces in the
-low-level runtime:
-\begin{description}
-
-\item[MetaData] The metadata for a logical region captures attributes of a
-logical region that are common across all instances.  These attributes include
-the size of the logical region, the element type, and the allocation mask,
-which tracks which portions of the logical region are actually in use.  
-%Region
-%MetaDatas are created independently of any memory objects and are accessible
-%(via a globally unique name) to all nodes in the machine.
-
-\item[Allocator] The dynamic memory allocation capability of a region instance
-is provided via an allocator object.  An allocator is associated with a
-region metadata and must be created in a particular memory.  Once created, it
-may be used to allocate and free elements of the region by any task on any
-processor that has non-zero affinity with the memory.
-
-\item[Instance] Each physical instance of a region stores some version of all
-the elements in a region.  As with allocators, a physical instance is
-associated with a particular region metadata and spends its entire lifetime in
-the memory object in which it was created.  
-%When multiple physical instances
-%exist for the same region, the low-level runtime does nothing to keep them
-%copies and set up the right inter-operation dependencies to provide the level
-%of coherence requested by the application.  Although a physical instance 
-%stores the contents of a region, it does not directly offer access to those
-%contents.
-
-\item[Accessor] Access to the contents of a physical region instance are 
-handled by accessor objects.  These are lightweight objects that allow a given
-task to access a given physical region instance, regardless of the
-implementation of that instance or the underlying memory.  In the most common
-case, read and write access is done through a single array dereference, but the
-accessor interface allows seamless use of other types of memory as well (e.g.
-memory that's accessed via RMDA put and get calls).
-
-\end{description}
+\subsubsection{Processors}
+\label{subsec:processors}
+Hardware units capable of performing computation are captured as generic
+processors in the low-level runtime.  Like other runtime objects, a processor
+has a globally unique name and can be referred to anywhere in the system.
+Processors come in different \emph{kinds} (e.g. X86 CPU core, CUDA-capable GPU),
+but they all support the same
+interface and execution model.  A request to spawn a task on a given processor
+may come from anywhere in the system, and names a task along with zero or
+more events that must happen before the task may start.  Once those
+events have occurred, the task is placed into a queue of ready tasks, which
+the processor works through in the order in which they were enqueued.  A
+processor will generally run a task to completion before taking another, but
+will try to run another task (resources permitting) if the current task needs
+to wait for an event to occur.  Once a task is placed on a processor's queue,
+it may not be moved or deleted.  Load-balancing of tasks is expected to be done
+by the high-level runtime.  The primary purpose of the queue for the low-level
+Processor is to be able to hide the latency of starting new tasks.
 
@@
 two implementation of the low-level runtime: one based on POSIX
-threads for a shared memory SMP and one based on GASNet and CUDA
-for distributed memory clusters with multiple GPUs.
+threads for a shared memory SMP and one based on GASNet \cite{GASNET07} 
+and CUDA for distributed memory clusters with multiple GPUs.
 %The magic of an
@@
 The cluster implementation of the low-level runtime is implemented on a mix
-of POSIX threads for intra-node threading and communication, GASnet for 
+of POSIX threads for intra-node threading and communication, GASNet for 
 inter-node communication, and CUDA for access to resources of the GPUs in 
 each node.  Locks and events have roughly the same implementation as the SMP
-runtime for handling multiple local requestors, but use GASnet active messages
+runtime for handling multiple local requestors, but use GASNet active messages
 to coordinate lock ownership and event subscription and triggering between
@@
 from significant overhead.  The memory hierarchy is expanded to include a 
-{\em global} memory that is striped across all nodes and accessible via GASnet
+{\em global} memory that is striped across all nodes and accessible via GASNet
 RDMA calls, one process local memory per cluster node, and two memories per GPU: one for the GPU's framebuffer that is

### 2011-11-07 doc/pldi2012/related.tex
Updating the bibliography to save some space

--- 2011-11-06 doc/pldi2012/related.tex
@@
 
-The Chapel programming language \cite{Chamberlain:Chapel} is the closest related work to the Legion
-programming model.  Chapel has several concepts built into the language to aid in
-the expression of locality.  The first concept is that of domains.  Domains
-are similar to logical regions in that they describe a collections of indexes
-into items of the same type.  Domains can create sub-domains by slicing the index sets from
-a parent domain.  There are two differences between domains and logical regions.
-First, domains are a higher level concept than regions since they track index sets which
-have concepts such as dimensionality and support iterators.  Logical regions in Legion 
-are only allowed to be accessed with pointers.  Second, the act of creating
-subdomains doesn't track disjointness information the way that partitioning of logical regions
-does making it challenging for the compiler or the runtime to infer task independence.
+Chapel \cite{Chamberlain:Chapel} is the closest related work to Legion.  
+Chapel has several concepts to support the expression of locality.  
+Domains are similar to logical regions in that they describe maps from indexes
+to same-typed objects.  Domains can create sub-domains by slicing the index sets from
+a parent domain.  Domains are a higher level concept than regions; the domain index sets 
+support dimensionality and iterators, whereas  logical regions can only be accessed by pointers.  
+Second, the act of creating subdomains in Chapel does not track disjointness information,
+making it more challenging for the Chapel compiler or runtime to infer task independence.
 
@@
 of abstract locations.  Programmers can use locales by writing domain maps that specify
-how domains are broken up and assigned to locales.  Domain maps provide the same functionality
+how domains are subdivided and assigned to locales.  Domain maps provide the same functionality
 as partitions and mappers in Legion, but require the user to correctly implement domain
-maps in order for the program to be correct.  Legion explicitly isolates correctness
+maps for the program to be correct.  Legion explicitly isolates correctness
 from performance by defining the Mapper interface.  In addition, Chapel's flat array of locales
-makes it challenging to fully utilize deep memory hierarchies.
+makes it challenging to fully utilize deep memory hierarchies.  Chapel currently supports
+clusters and GPUs in isolation \cite{CHAPELGPU}, but we are not aware of any results that 
+make use of both.
+
+Sequoia is an example of a locality-aware programming language capable
+of expressing computations in deep memory hierarchies
+\cite{Fatahalian06}.  Sequoia is a special case of the Legion
+programming model in which only arrays can be recursively partitioned,
+all access is exclusive and the task tree and tree of array
+decompositions are identical.  Sequoia requires the programmer to
+statically map arrays and task trees using a machine specific mapping
+file.  Legion generalizes the Sequoia model by allowing for
+partitioning of pointer data structures through regions, decoupling
+the region tree from the task tree, enabling dynamic mappings through
+the mapper interface, and allowing different coherence
+properties.
 
 X10 is another parallel programming environment designed to operate on distributed
-memory machines \cite{X1005}.  In X10 the concept of places enables programmers to talk about where
+memory machines \cite{X1005}.  X10's {\em places} enable programmers to talk about where
 to place both data and tasks in the machine.  However, once data and tasks have
-been placed they are fixed which mandates that data movement be explictly managed by
+been placed they are fixed, which mandates that data movement be explicitly managed by
 user level code or implicitly by the compiler or runtime \cite{X1008}.  Recently X10 has introduced
@@
 Unlike Legion, regions in X10 are not visible to the programmer but are inferred
-from high level arrays through static analysis.
+from high level arrays through static analysis.  X10 provides support for clusters of GPUs
+\cite{X10GPU}, but requires the programmer to write all code managing data movement
+through both the cluster and GPU memory hierarchies.
 
-SPMD languages such as Titanium \cite{JV:Yel98} and UPC \cite{UPC99} have mechanisms for
-describing array partitions in distributed memories.  However, the partition
-operations supported only operate on two-level memory hierarchies consisting of local
-and global memory and do not reason about pointer data structures.  Part of Legion's 
+SPMD languages such as Titanium \cite{TIT98} and UPC \cite{UPC99} have
+mechanisms for describing array partitions in distributed memories.
+However, the partition operations supported only operate on two-level
+memory hierarchies consisting of local and global memory. 
+% Neither
+%Titanium nor UPC currently support GPUs as part of their language
+%standard \cite{TITANIUMSTANDARD} \cite{UPCSTANDARD}.  
+Part of Legion's
 low-level runtime system is constructed using UPC's underlying GASNet
@@
 describing memory management schemes \cite{REAPS02}\cite{RC01}  
-or for enforcing security policies \cite{CYCLONE01}.  In all cases, regions had
+or for enforcing safety policies \cite{CYCLONE01}.  In all cases, regions had
 memory layout implications.  Logical regions in Legion enable the programmer to
@@
 
-Marino introduces a generic flow-insensitive type-effect system capable of verifying 
-accesses to statically tagged locations of memory \cite{PRIVLIGES09}.  Legion's programming
-system allows for dynamic creation and partitioning of regions requiring a 
-flow-sensitive analysis to track privleges.
+%Marino introduces a generic flow-insensitive type-effect system capable of verifying 
+%accesses to statically tagged locations of memory \cite{PRIVLIGES09}.  Legion's programming
+%system allows for dynamic creation and partitioning of regions requiring a 
+%flow-sensitive analysis to track privileges.

### 2011-11-07 doc/pldi2012/highlevel.tex
figure 2 moved to same page as text that references it - had to inline mapping_fig.tex into highlevel.tex to make this happen

@@
 new task.
+
+\usetikzlibrary{calc}
+\begin{figure*}[t]
+  \centering
+  \subfigure[mapping of $cnc_0$ task]{
+    \label{sfig:mapping_fig:cnc}
+    \tiny
+    \begin{tikzpicture}[scale=0.7]
+      \partitiontree
+      \node(cnc_p0)[very thick,draw,fill=white] at (0.45,0.30) {\tiny $cnc_0$};
+      \node(cnc_g0)[very thick,draw,fill=white] at (5.45,0.30) {\tiny $cnc_0$};
+      \node(def_top)[very thin,draw,fill=white] at (3.3,4.35) {\tiny ~~---~~};
+      \draw[very thick] ($ (top.south)!.5!(ptp.center) $) circle (2pt);
+      \draw[very thick] ($ (pvsf.north)!.5!(ptf.center) $) circle (2pt);
+      \draw[very thick] ($ (pvsf.250)!.5!(ppp.center) $) circle (2pt);
+      \draw[very thick] ($ (p0.north)!.5!(pp0.center) $) circle (2pt);
+      \draw[very thick] ($ (pvst.north)!.5!(ptt.center) $) circle (2pt);
+      \draw[very thick] ($ (pvst.310)!.5!(pgp.center) $) circle (2pt);
+      \draw[very thick] ($ (g0.north)!.5!(pg0.center) $) circle (2pt);
+      \draw[very thick,->] (def_top) to [bend right=45] (cnc_p0);
+      \draw[very thick,->] (def_top) to [bend left=45] (cnc_g0);
+    \end{tikzpicture}
+  }
+  \subfigure[mapping of $dc_0$ task]{
+    \label{sfig:mapping_fig:dc}
+    \tiny
+    \begin{tikzpicture}[scale=0.7]
+      \partitiontree
+      %\node(cnc_p0)[very thick,draw,fill=white] at (0.45,0.30) {\tiny $\begin{array}{c}\cancel{cnc_0} \\ dc_0\end{array}$};
+      \node(cnc_pn)[very thin,draw,fill=white] at (2.25,0.30) {\tiny $cnc_{n-1}$};
+      \node(cnc_p1)[very thin,draw,fill=white] at (1.25,0.30) {\tiny $cnc_1$};
+      \node(dc_p0)[very thick,draw,fill=white] at (0.45,0.30) 
+{\tiny
+$\begin{array}{@{}c@{}}
+\cancel{cnc_0} \\
+dc_0
+\end{array}$};
+      \node(cnc_gn)[very thin,draw,fill=white] at (7.25,0.30) {\tiny $cnc_{n-1}$};
+      \node(cnc_g1)[very thin,draw,fill=white] at (6.25,0.30) {\tiny $cnc_1$};
+      \node(cnc_g0)[very thin,draw,fill=white] at (5.45,0.30) {\tiny $cnc_0$};
+      \node(dc_g0)[very thick,draw,fill=white] at (5.45,0.75) {\tiny $dc_0$};
+      \draw[very thick,->] (cnc_g0.west) to [bend left=70] (dc_g0.west);
+      \draw[very thick] ($ (cnc_g0.center) - (0.25,0.25) $) -- ++(0.5,0.5);
+      \draw[very thick] ($ (cnc_g0.center) - (0.25,-0.25) $) -- ++(0.5,-0.5);
+      \node(def_top)[very thin,draw,fill=white] at (3.3,4.35) {\tiny ~~---~~};
+      \draw[very thin] ($ (top.south)!.5!(ptp.center) $) circle (2pt);
+      \draw[very thin] ($ (pvsf.north)!.5!(ptf.center) $) circle (2pt);
+      \draw[very thin] ($ (pvsf.250)!.5!(ppp.center) $) circle (2pt);
+      \draw[very thin] ($ (p0.north)!.5!(pp0.center) $) circle (2pt);
+      \draw[very thin] ($ (p1.north)!.5!(pp1.center) $) circle (2pt);
+      \draw[very thin] ($ (pn.north)!.5!(ppn.center) $) circle (2pt);
+      \draw[very thin] ($ (pvst.north)!.5!(ptt.center) $) circle (2pt);
+      \draw[very thin] ($ (pvst.310)!.5!(pgp.center) $) circle (2pt);
+      \draw[very thin] ($ (g0.north)!.5!(pg0.center) $) circle (2pt);
+      \draw[very thin] ($ (g1.north)!.5!(pg1.center) $) circle (2pt);
+      \draw[very thin] ($ (gn.north)!.5!(pgn.center) $) circle (2pt);
+    \end{tikzpicture}
+  }
+  \subfigure[mapping of $volt_0$ task]{
+    \label{sfig:mapping_fig:volt}
+    \tiny
+    \begin{tikzpicture}[scale=0.7]
+      \partitiontree
+      \node(dc_p0)[very thin,draw,fill=white] at (0.45,0.30) {\tiny $dc_0$};
+      \node(dc_p1)[very thin,draw,fill=white] at (1.25,0.30) {\tiny $dc_1$};
+      \node(dc_pn)[very thin,draw,fill=white] at (2.25,0.30) {\tiny $dc_{n-1}$};
+      \node(dc_g0)[very thin,draw,fill=white] at (5.45,0.30) {\tiny $dc_0$};
+      \node(volt_p0)[very thick,draw,fill=white] at (0.45,0.75) {\tiny $volt_0$};
+      \draw[very thick,->] (dc_p0.west) to [bend left=70] (volt_p0.west);
+      \draw[very thick] ($ (dc_p0.center) - (0.25,0.25) $) -- ++(0.5,0.5);
+      \draw[very thick] ($ (dc_p0.center) - (0.25,-0.25) $) -- ++(0.5,-0.5);
+      \draw[very thick] ($ (dc_g0.center) - (0.25,0.25) $) -- ++(0.5,0.5);
+      \draw[very thick] ($ (dc_g0.center) - (0.25,-0.25) $) -- ++(0.5,-0.5);
+      \node(dc_g0)[very thin,draw,fill=white] at (5.45,0.30) {\tiny $dc_0$};
+      \node(dc_g1)[very thin,draw,fill=white] at (6.25,0.30) {\tiny $dc_1$};
+      \node(dc_gn)[very thin,draw,fill=white] at (7.25,0.30) {\tiny $dc_{n-1}$};
+      \draw[very thick] ($ (dc_g0.center) - (0.25,0.25) $) -- ++(0.5,0.5);
+      \draw[very thick] ($ (dc_g0.center) - (0.25,-0.25) $) -- ++(0.5,-0.5);
+      \draw[very thick] ($ (dc_g1.center) - (0.25,0.25) $) -- ++(0.5,0.5);
+      \draw[very thick] ($ (dc_g1.center) - (0.25,-0.25) $) -- ++(0.5,-0.5);
+      \draw[very thick] ($ (dc_gn.center) - (0.25,0.25) $) -- ++(0.5,0.5);
+      \draw[very thick] ($ (dc_gn.center) - (0.25,-0.25) $) -- ++(0.5,-0.5);
+      \node(def_top)[very thin,draw,fill=white] at (3.3,4.35) {\tiny ~~---~~};
+      \draw[very thin] ($ (top.south)!.5!(ptp.center) $) circle (2pt);
+      \draw[very thin] ($ (pvsf.north)!.5!(ptf.center) $) circle (2pt);
+      \draw[very thin] ($ (pvsf.250)!.5!(ppp.center) $) circle (2pt);
+      \draw[very thin] ($ (p0.north)!.5!(pp0.center) $) circle (2pt);
+      \draw[very thin] ($ (p1.north)!.5!(pp1.center) $) circle (2pt);
+      \draw[very thin] ($ (pn.north)!.5!(ppn.center) $) circle (2pt);
+      \draw[very thin] ($ (pvst.north)!.5!(ptt.center) $) circle (2pt);
+      \draw[very thin] ($ (pvst.310)!.5!(pgp.center) $) circle (2pt);
+      \draw[very thin] ($ (g0.north)!.5!(pg0.center) $) circle (2pt);
+      \draw[very thin] ($ (g1.north)!.5!(pg1.center) $) circle (2pt);
+      \draw[very thin] ($ (gn.north)!.5!(pgn.center) $) circle (2pt);
+      \draw[very thick] ($ (g0.north)!.5!(pg0.center) - (0.1,0.1) $) -- ++(0.2,0.2);
+      \draw[very thick] ($ (g0.north)!.5!(pg0.center) - (0.1,-0.1) $) -- ++(0.2,-0.2);
+      \draw[very thick] ($ (g1.north)!.5!(pg1.center) - (0.1,0.1) $) -- ++(0.2,0.2);
+      \draw[very thick] ($ (g1.north)!.5!(pg1.center) - (0.1,-0.1) $) -- ++(0.2,-0.2);
+      \draw[very thick] ($ (gn.north)!.5!(pgn.center) - (0.1,0.1) $) -- ++(0.2,0.2);
+      \draw[very thick] ($ (gn.north)!.5!(pgn.center) - (0.1,-0.1) $) -- ++(0.2,-0.2);
+      \draw[very thick] ($ (pvst.310)!.5!(pgp.center) - (0.1,0.1) $) -- ++(0.2,0.2);
+      \draw[very thick] ($ (pvst.310)!.5!(pgp.center) - (0.1,-0.1) $) -- ++(0.2,-0.2);
+      \draw[very thick] ($ (pvst.230)!.5!(psp.center) $) circle (2pt);
+      \draw[very thick] ($ (s0.north)!.5!(ps0.center) $) circle (2pt);
+      \node(def_pvst)[very thick,draw,fill=white] at (5.3,2.75) {\tiny ~~---~~};
+      \draw[very thick,->] (dc_g0.north) to [bend right=30] (def_pvst.270);
+      \draw[very thick,->] (dc_g1.north) to [bend right=30] (def_pvst.315);
+      \draw[very thick,->] (dc_gn.north) to [bend right=30] (def_pvst.0);
+      \node(volt_s0)[very thick,draw,fill=white] at (3.2,0.30) {\tiny $volt_0$};
+      \draw[very thick,->] (def_pvst.190) to [bend right=40] (volt_s0.130);
+    \end{tikzpicture}
+  }
+  \label{fig:mapping_fig}
+  \caption{Mapping of tasks in region tree}
+\end{figure*}
 

### 2011-11-07 doc/pldi2012/lowlevel.tex
described valid mask
More edits to the low level

@@
 The low-level runtime abstracts the machine as a base set of objects:
-events, deferred locks, processors, memories, and regions.
+events, deferred locks, regions, memories, and processors.
 
@@
 event must have already triggered.  With this technique, there is no need to
-count references to the event at all. 
+count references to the event. 
 % An event can only be reused until the
@@
 Events provide a deterministic ordering between operations, 
-but the more relaxed coherence do not require strict ordering.  The traditional tool for this job is
+but the more relaxed coherence properties do not require strict ordering.  
+The traditional tool for this job is
 the mutex (a.k.a. lock), and it is from the common reader/writer lock that
@@
 region size and valid mask.  Allocators are used to dynamically allocate and
-free entries in a region.  Finally, instances hold versions of the data for 
+free entries in a region.  Finally, physical instances hold versions of the data for 
 a logical region.

### 2011-11-09 doc/pldi2012/code_ex.tex
fixed typo in code example - math stuff needs to be escaped from listing

@@
   Wire<rn_pvt+rn_shr+rn_ghost,rn,rw_pvt>@rw_pvt    first_wire;
-  where rn_pvt $\le$ rn, rn_shr $\le$ rn, rn_ghost $\le$ rn, rw_pvt $\le$ rw;
+  where rn_pvt #$\le$# rn, rn_shr #$\le$# rn, rn_ghost #$\le$# rn, rw_pvt #$\le$# rw;
 };

### 2011-11-10 doc/pldi2012/code_ex.tex
fixed type of wires->nodes in code example

--- 2011-11-07 doc/pldi2012/code_ex.tex
@@
   region< Node<rn> >                                                   rn_ghost;
-  region< Wire<rn_pvt+rn_shr+rn_ghost,rn,rw_pvt> >  rw_pvt;
+  region< Wire<rn_pvt+rn_ghost,rn,rw_pvt> >               rw_pvt;
   Node<rn_pvt+rn_shr>@(rn_pvt+rn_shr)                     first_node;
-  Wire<rn_pvt+rn_shr+rn_ghost,rn,rw_pvt>@rw_pvt    first_wire;
-  where rn_pvt $\le$ rn, rn_shr $\le$ rn, rn_ghost $\le$ rn, rw_pvt $\le$ rw;
+  Wire<rn_pvt+rn_ghost,rn,rw_pvt>@rw_pvt                first_wire;
+  where rn_pvt #$\le$# rn, rn_shr #$\le$# rn, rn_ghost #$\le$# rn, rw_pvt #$\le$# rw;
 };

### 2012-01-02 doc/pldi2012/circuit.tex

@@
 
-Lines 35-39 form the bulk of the actual simulation, performing three passes
-over the circuit in each iteration.  Each pass loops over an array of pieces (constructed on lines 30-33 from the partitions),
+Lines 36-40 form the bulk of the actual simulation, performing three passes
+over the circuit in each iteration.  Each pass loops over an array of pieces (constructed on lines 30-34 from the partitions),
 spawning a task for each piece.   There are no explicit requests for
@@
 The simultaneous assignment operator asks the type checker only to ensure
-that fields have the correct types after all the fields are updated.  
-Lines 32 and 33 illustrate a simultaneous
+that fields have the correct types after all the fields are updated.  In
+many cases, the target type also tries to describe tighter constraints on
+the region parameters used for field types.  The {\tt region\_cast} operator
+allows the programmer to assert the correctness of this transformation, which
+can be verified with a linear scan over the elements of a region.
+Lines 32-34 illustrate a simultaneous
 assignment to update a {\tt CircuitPiece} region relationship.

### 2012-01-02 doc/pldi2012/lowlevel.tex

@@
 
-The final interesting feature of the deferred lock is that it can be associated
-with an object in memory and that object will automatically be kept coherent
-across the cluster.  This is similar to MESI coherence in a distributed cache,
-but is done per-object rather than on cache lines.  This is currently only 
-used for objects within the runtime itself, but we are exploring ways to expose
-it at the Legion application level.
+%The final interesting feature of the deferred lock is that it can be associated
+%with an object in memory and that object will automatically be kept coherent
+%across the cluster.  This is similar to MESI coherence in a distributed cache,
+%but is done per-object rather than on cache lines.  This is currently only 
+%used for objects within the runtime itself, but we are exploring ways to expose
+%it at the Legion application level.
 
@@
 low-level runtime.  The metadata object tracks global information such as
-region size and valid mask.  Allocators are used to dynamically allocate and
+region size and the set of valid elements, which is used in partitioning
+operations.  Allocators are used to dynamically allocate and
 free entries in a region.  Finally, physical instances hold versions of the data for 

### 2012-03-02 doc/runtimes/hlr.tex
Beginning to try to write up a description for the high level runtime

\section{Context Initialization}
When a task is first registered with the runtime we create a new {\em context} for the task.  
A context stores all information about that task.  The context for a task will always 
live on the processor in which the task was created even if copies of the task's context 
are made wherever the task is moved.  We call the context on the home processor of the 
task the {\em original} context.  The context for a task also knows the context of the parent 
task that created it.  This parent-child context relationship is fundamental to the propagation 
of information through the runtime.

After we've set up a context and stored all the information from the task launch, we make 
our first mapper call.  We ask the mapper to tell us whether or not this task should be spawned.  
If a task is spawned we will later ask the mapper if it would like to move the task to a remote 
processor.  Spawned tasks are also eligible to be stolen.  If a task is not spawned then it 
is required to be run on the processor on which it was created (cannot be sent anywhere and 
cannot be stolen).

\section{Dependence Analysis}
After the task's context has been initialized, we perform a dependence analysis.  This first 
dependence analysis does NOT determine any actual dependences between tasks.  This first dependence 
analysis determines {\em mapping dependences} between tasks.  A mapping dependence only implies that two 
tasks must be mapped in a specific order.  This may or may not result in them having an actual 
dependence later.  One task has a mapping dependence on a prior task if the later task uses a logical 
region which may alias with a logical region being used by the prior task and at least one of the two 
tasks is using their logical region in a write mode.  Note at this point we can almost completely 
ignore coherence properties.  If two tasks both want a region in simultaneous mode with at least 
one of them requesting a write, they will have a mapping dependence even if they ultimately don't 
have a dependence because they choose to use the same physical instance.  The one exception to this 
rule is with relaxed coherence.  Relaxed coherence will allow two tasks using aliased logical regions 
with at least one write to be mapped in any order.

To find a task's mapping dependences, we must determine which other tasks are using logical regions 
that alias with logical regions for the task we are currently registering.  This is where the type 
system makes it possible for us to only have to perform this analysis on other tasks that share 
the same parent context.  To find the tasks on which we have a mapping dependence, we will use the 
{\em logical state} of the region trees in the parent's context.  For each of the region trees for which 
the parent task has a privilege (any privilege), the parent task maintains two pieces of logical 
state information for each node in the region tree: which tasks in the parent's context are using 
the node and whether which children of the node are open.   A child node is {\em open} if it has users 
in the logical state of the region sub-tree.  Regions are only permitted to have one partition open 
at a time.  Aliased partitions are only permitted to have one child region open at a time.  Disjoint 
partitions are allowed to have any number of child regions open at a time.  Note also that both 
regions and partitions can have users.  Partitions can only be used by index space tasks.

To find mapping dependences, for each logical region that a task needs we first find the 
corresponding region for which the parent has privileges.  This can be either a region that the 
parent was passed as an argument or a region which the parent created.  We call these regions the 
{\em top-most regions} in the parent's context.  For each region that a task requires, we walk from the 
parent's top-most region down to the logical region required by the child task.  At each node we 
examine the users in the logical state of the node.  If there are any dependences between the 
tasks, we register them as mapping dependences.  As we traverse down the tree we open up nodes 
as necessary to arrive at our destination.  Once we arrive at our destination node we mark that 
we are using the current node.

If it at any point we find a path open that we don't intend to traverse, we must first {\em close}
the open subtree, since it's possible that we might use aliased logical regions from that sub-tree.  
This only occurs for regions and aliased partitions.  To close a logical sub-tree, we record all 
users of the logical subtree and move them all to be users of node in our current location.  
This is a valid operation since all these users were really using sub-pieces of the current logical 
region.  We must keep the users of the closed subtree because later tasks will also need to record 
mapping dependences on them.  We record any dependences on the tasks in the closed logical subtree 
and then continue our traversal.  Once we arrive at our destination, we also must close up all 
logical subtrees below as well to record mapping dependences on any tasks using sub-logical regions 
of our intended region.

One optimization that we perform is to replace all the users of a logical region with a single user 
as long as the single user is said to {\em dominate} all the prior users.  A task dominates all the prior 
users of a logical region if it has a mapping dependence on all prior users and all users in any 
open sub-tree.  We only check for a task being a dominator when it reaches its target node since 
it is unlikely that a task dominates all tasks in any open subtree (assuming that most partitions 
in the tree are disjoint).

After we have have found all the mapping dependences for a task, we are ready to place it on one of 
the runtime's scheduling queues.  If a task has mapping dependences we place it on the runtime's 
waiting queue of tasks that are not ready to be mapped yet.  If the task has no mapping dependences 
we place it on the runtime's ready queue of tasks that are ready to map.  As an optimization, for 
tasks that have been spawned and have no mapping dependences, we eagerly query the mapper 
for the target processor of the task.  This allows us to move tasks as soon as possible to their 
target processor.  If the target processor is not the current processor, the task along with a copy 
of its context are relocated to the target processor and then enqueued on the ready queue.

\section{Scheduling and Mapping}
When the application is started the high level runtime registers an {\em idle task} with the low-level 
runtime as a callback for whenever the low-level runtime has little or no-work to perform.  The 
high-level runtime's idle task will choose to perform one of two operations whenever it is invoked.  
If it has any tasks on its ready queue, it will map some of those tasks and launch them on the current 
processor.  If the runtime has little no ready tasks it will ask each of its mappers whether or 
not they want to attempt a task steal.  Task stealing will be discussed in more detail in a 
later section.

Tasks are currently scheduled in a very simple manner.  The oldest task on the ready queue 
will always be the next one mapped.  Changing the scheduling algorithm is easy, but may require 
coordination between mappers over how to weight tasks which is not something we've thought about 
deeply.  Scheduling at the high-level has less of an impact because the order in which tasks are 
mapped isn't necessarily the order in which they are run.  The high-level runtime simply passes 
tasks down to the low-level runtime with their corresponding event dependences.  The low-level 
runtime is the one that ultimately decides the order in which tasks run based primarily on when 
the event dependences are satisfied.

Once a task has been selected to be mapped, we first check to see if it has been spawned and if 
so whether the mapper has already selected its target processor.  If the mapper hasn't had an 
opportunity to a direct a spawned task yet, we ask it where to send the task and move it to the 
target processor.  I'll describe how remote mapping happens later in this section.

\subsection{Mapping}
To map a task, we first ask the mapper to select a location for a physical instance of each of 
the logical regions that the task needs to execute.  The mapper has the option of specifying 
a list of memories in which to try getting a physical instance or specify no memories.  If the 
mapper provides a list memories, the runtime will attempt to create or find a physical instance 
in one memory and only move onto the next if it fails to create or find a physical instance in 
the current one.  If the runtime can't create a physical instance in any of the memories it 
will fail.  If the mapper doesn't specify any memories, the runtime uses this as a signal that 
the task does not want a physical instance for the task.

To aid in the mapper's decision, the runtime provides a list of memories were physical instances with 
{\em valid} data currently reside in the memory hierarchy. A physical instance is said to be valid if
is a the physical instance or a copy the physical instance of the last task to write to the logical
region which the physical instance is representing.

\subsection{Physical Region Tree Traversal}
To be capable of providing this information and finding actual data dependences,
each context also maintains a {\em physical} state of the its topmost regions and their 
corresponding region trees.  The physical state of a node in the region tree is different for 
partitions and regions.  For regions, the physical state includes the state and set of open partitions, 
the set of valid physical instances of the particular logical region, and whether or not the physical 
instances for the region are dirty.  Regions can have multiple partitions open in a read-only state 
and only a single partition open in a write state.  Regions physical instances are marked dirty as 
soon as they have a task that performs a write.  Having dirty physical instance indicates that a 
copy will need to be made back to a parent physical instance at a later point in time.  The physical 
state of a partition node is somewhat simpler.  Partitions only keep track of their open child 
regions.  Aliased partitions can have multiple children in read-only mode or a single child open in 
write mode.  Disjoint partitions can have an arbitrary number of regions open in either read-only 
or write mode.

Similar to the algorithm for dependence analysis on the logical state, for each region that the 
mapper has requested a physical instance we walk the physical region tree from the parent context's 
top-most region down to our target region.  The steps that we perform along the way however are 
different.  Unlike before we don't need to check for any dependences explicitly on regions.  Instead 
we will discover these dependences implicitly as we build up a chain of events that will be required 
to occur before our task can be executed.  As we traverse the tree, we keep a monad corresponding to 
the last event in the chain of dependences that must occur before our task can be launched.  We call
this monad the precondition since it will ultimately define a precondition for launching the task.

As we traverse the tree, we must perform open and close operations analogous to the operations 
performed on the logical region tree.  Open operations just denote opening the physical state 
of the node.  Close operations are slightly more complicated.  To perform a close operation we must 
copy all dirty data back from the physical instances for each of the open subtrees into a physical 
instance of the current region node.  To do this we invoke the help of the mapper.  We first ask the mapper 
to provide a list of memories in which to try to create a physical instance of the current region.  We 
then go through this list of memories until we find or make a physical instance of the current region which
becomes the {\em target} physical instance.  It is an error to fail to make a target instance.

Once we have made a target instance we traverse down the subtree to be closed with the precondition monad
initialized to a null event.  At each node we issue a copy from one of the valid physical instances back
to the target instance conditioned on the precondition monad as well as the termination events of all prior
users of the physical instance.  To select the source physical instance for the copy operation we invoke
the mapper.  The event corresponding to the finish of the copy operation becomes the new state of the
precondition monad and we continue our traversal down the tree.  Once we have traversed the entire tree,
we merge the preconditions for all different subtrees back into our precondition monad and make the corresponding
event the new precondition as we continue traversing the tree.

When we arrive at our target node, we first check to see if the physical instance is a newly created.  If
it has been newly created we must copy valid data from one of the currently valid physical instances.  If the
state of our region is dirty, then we can only use valid instances from our node.  If the state of our instance
is clean, then we can traverse back up the region tree to find physical instances of parent regions that can
also be used as valid sources for our physical instance.  We must also check to see if there are any open sub-trees.  
If there are, then we must issue close operations to them as well.  If there is dirty data in the subtrees this
will result in copies back to the instance we are creating as well which will update the precondition monad.

The last step is to find the set of other tasks which are using the same physical instance as us on which we have
a dependence.  This is necessary because we didn't record any actual dependences in the original mapping dependence
analysis.  The mapping dependence analysis is an over-approximation of the actual dependences we might have since its
possible for tasks on which we have a mapping analysis to complete before we are even mapped.

To track the set of users of a physical instance, the runtime wraps a physical instance in an object called an
{\em InstanceInfo} (need a better name).  InstanceInfos track all the users of a physical instance including both
task users and copys users.  In the case of tasks, InstanceInfos track both the mode and coherence in which the
task is using a physical instance.  In the case of copies, InstanceInfos track whether the copy is a copy that
is reading or writing from the instance.  For all of these users, the InstanceInfo tracks the event corresponding
to when the user will be finished using the InstanceInfo.  Therefore whenever a new user is registered with an
InstanceInfo, the InstanceInfo returns an event that will be bound to the precondition monad indicating the event
that will trigger when all other users on which the new user has a dependence will be finished.

One complication to this is that a physical instance of a logical region can also be used as a physical instance
of any logical subregion.  To handle this case, we can have multiple InstanceInfos for the same physical
instance in the physical state of different regions in the region tree.  Each of these InstanceInfos keeps track
of the users of a physical instance for a specific logical region.  When checking for a dependences on a previous
users of a physical instance it is necessary to find dependences on users using the instance for a logical
region farther up in the region tree.  To handle this case, InstanceInfos also know about all the InstanceInfos
for the same physical instance in logical regions about them in the region tree.  When dependences are checked
on a given InstanceInfo they are also checked in all InstanceInfos farther up the tree.  This is how we are
able to avoid checking dependences when traversing down the physical state of the region tree.

Once the physical region tree state traversal is complete, the resulting event in the precondition monad
becomes a precondition for the task to be executed.  This is repeated for all the regions that the mapper
requested that the task map and the resulting events are merged to create one event that is passed as the
precondition event to the low-level runtime as the task is launched on the low-level processor.

Once the task has been mapped and launched, it notifies all other tasks which had mapping dependences on it.
If these tasks have had all their mapping dependences satisfied then they are moved to the ready queue to
be mapped.  It is important to note that a task can only perform this step if the mapper requested a physical
instance for all of the logical regions of the task.  If the mapper didn't ask for a physical instance for any
of the logical regions, then the task will not be considered mapped until all of its child tasks have been mapped.

\subsection{Remote Mapping}
Tasks can also be mapped remotely from their original context.  In order for this to be possible the region trees
as well as the physical state in the parent task's context must be moved to the region trees.  Before they can be
moved however we first must perform a {\em sanitizing} operation.  If a task is moved remotely it can be mapped
in parallel with other tasks in the same context that don't have any mapping dependence.  However, it is possible
that two or more tasks may rely on the same close operation in traversing to their needed region.  In this case
we don't want both tasks issuing the same close operation on different processors as this can lead to an
inconsistent state of the physical region tree.  Therefore before a task can be sent remotely it traverses the
physical region trees to each of its needed regions and performs any necessary close operations.  In the process
if it opens any regions it will also create valid physical instances at its needed regions using preexisting physical
instances from parent regions.  This way when the task is sent remotely, we only need to send the region trees
and physical state corresponding to the child's top-most regions instead of the parent's top-most regions.

Mapping a task remotely follows the same procedure already described, but the physical state of the region tree
only needs to be traversed from the child's top-most regions since we've already sanitized the region tree
prior to moving it.  After the mapping is complete, the child packages back up the state of all region trees
in which a physical instance was requested by the mapper.  The state is then sent back to the child's original
context which propagates the information back to the parent context on the original processor.  Note that performing
mappings remotely is a safe operation because the dependence analysis on logical regions ensures that no other
tasks in the same context could possibly map on the original processor at the same time if they touched any
of the same logical regions in the set of the child's top-most regions.

\section{Execution}

\section{Children Mapped}

\section{Restoring Copies}

\section{Termination}

\section{Garbage Collection}

\section{Region Tree Updates}

\section{Task Stealing}

\section{Utility Processors and Thread Safety}

\section{TODO}

### 2012-03-14 tools/legion_spy.py
added #!'s at the top so these scripts can be executed directly

@@ -1,3 +1,4 @@
+#!/usr/bin/python
 
 import subprocess
 import pygame, sys, os, shutil

### 2012-03-14 tools/parser.py

@@ -1,3 +1,4 @@
+#!/usr/bin/python
 
 from state import *
 import sys, re

### 2012-03-14 tools/state.py

@@ -1,3 +1,4 @@
+#!/usr/bin/python
 
 import subprocess
 

### 2012-03-30 doc/sc2012/legion.tex
Adding files for the sc12 paper

\documentclass[conference]{IEEEtran}

\bibliographystyle{plainnat}
\usepackage{amsmath,amsthm}
\usepackage{listings}
\usepackage{tikz}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{cancel}

\newcommand{\infrule}[2]{\displaystyle\frac{\displaystyle\strut{#1}}{\displaystyle\strut {#2}}}
\newcommand{\deref}{\ast}
\newcommand{\rread}[1]{\mbox{\em Read}(#1)}
\newcommand{\rwrite}[1]{\mbox{\em Write}(#1)}
\newcommand{\lca}[2]{#1 \sqcup #2}
\newcommand{\rleq}{\leq}
\newcommand{\interval}[1]{\mbox{\em interval}(#1)}
\newcommand{\context}[1]{\mbox{\em context}(#1)}
\newtheorem{theorem}{Theorem} 
\newtheorem{lemma}[theorem]{Lemma}

\begin{document}

\title{Legion: Expressing Locality and Independence with Logical Regions}
\author{Author names omitted}

\maketitle

\begin{abstract}
%As transistor counts have continued to scale with Moore's Law, computer
%architectures are becoming increasingly parallel to avoid fundamental
%physical limitations.  
Modern parallel architectures have both
heterogeneous processors and deep, complex memory hierarchies.  
%New
%programming abstractions are required to grant programmers the power
%to leverage these architectures.  With this problem in mind 
We present Legion, 
a programming model, type system, and runtime for programming these machines.
Legion is organized around {\em logical regions}, which express both locality and independence of program data.  
Legion also enables explicit, programmer controlled movement of data through
the memory hierarchy and placement of tasks based on locality information
via a novel mapping interface.  Running on a 4 node cluster with 8 total GPUs 
and 4 levels of memory hierarchy, our implementation of Legion 
achieves up to 5.9X speedup over a single CPU-GPU node on real-world applications.
\end{abstract}

\section{Introduction}
\label{sect:intro}
Put intro text here \cite{Fatahalian06}

%\input{circuit}
%\input{code_ex}
%\input{part_fig}

%\input{type_system}
%\input{exec}

%\input{highlevel}
%\input{mapping_fig}

%\input{lowlevel}
%\input{experiments}
%\input{related}
%\input{conclusion}

\bibstyle{IEEEtran}
{
\tiny
\bibliography{bibliography}
}

\end{document}

### 2012-04-03 doc/sc2012/code_ex.tex
Working on the code example for the paper

\lstset{
  captionpos=b,
  language=C++,
  basicstyle=\scriptsize,
  numbers=left,
  numberstyle=\tiny,
  columns=fullflexible,
  stepnumber=1,
  escapechar=\#,
  keepspaces=true,
  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
  morekeywords={region,coloring,partition,spawn,disjoint,aliased},
  deletekeywords=float,
}
\begin{lstlisting}[float={t},label={lst:code_ex},caption={Circuit Simulation Code Example}]
struct Node { float charge, capacitance; };
struct Wire { Node in_node, out_node; float current, ... ; };
struct Circuit {
  region   r_all_nodes;
  region   r_all_wires; };

struct CircuitPiece {
  region  rn_pvt, rn_shr, rn_ghost; // private, shared, ghost node regions
  region  rw_pvt;                   // private wires region };

void simulate_circuit(Circuit c) : RWE(c.r_all_nodes,c.r_all_wires)
{
  // Partition of wires into pieces
  partition<disjoint> p_wires = c.r_all_wires.partition(wire_owner_map); 
  // Partition nodes into two parts for all-private vs. all-shared
  partition<disjoint> p_nodes_pvs = c.r_all_nodes.partition(node_sharing map);

  // Partition all-private into separate circuit pieces
  partition<disjoint> p_pvt_nodes = p_nodes_pvs[0].partition(node_owner_map);
  // Partition all-shared into separate circuit pieces
  partition<disjoint> p_shr_nodes = p_nodes_pvs[1].partition(node_owner_map);
  // Partition all-shared into ghost regions which may be aliased
  partition<aliased> p_ghost_nodes = p_nodes_pvs[1].partition(node_neighbor_map);

  CircuitPiece pieces[MAX_PIECES];
  for(i = 0; i #$<$# MAX_PIECES; i++) 
    pieces[i] = { rn_pvt = p_pvt_nodes[i], rn_shr = p_shr_nodes[i],
                  rn_ghost = p_ghost_nodes[i], rw_pvt = p_wires[i] };
  while(!done) {
    spawn (i = 0; i #$<$# MAX_PIECES; i++) calc_new_currents(pieces[i]);
    spawn (i = 0; i #$<$# MAX_PIECES; i++) distribute_charge(pieces[i]);
    spawn (i = 0; i #$<$# MAX_PIECES; i++) update_voltages(pieces[i]);
  }
}

// read info from nodes connected to each wire, update state of wire
void calc_new_currents(CircuitPiece piece): RWE(piece.rw_pvt),
                                            ROE(piece.rn_pvt,piece.rn_shr,piece.rn_ghost)
// current moving through wires redistributes charge between nodes
void distribute_charge(CircuitPiece piece): ROE(piece.rw_pvt),
                                            RWE(piece.rn_pvt),
                                            RdA(piece.rn_shr,piece.rn_ghost)
// total charge added to a node causes changes in voltage
void update_voltages(CircuitPiece piece): RWE(piece.rn_pvt,piece.rn_shr)
\end{lstlisting}

### 2012-04-04 doc/sc2012/code_ex.tex
most of the copy code cleaned up - reductions not quite right

@@
   for(i = 0; i #$<$# MAX_PIECES; i++) 
-    pieces[i] = { rn_pvt = p_pvt_nodes[i], rn_shr = p_shr_nodes[i],
-                  rn_ghost = p_ghost_nodes[i], rw_pvt = p_wires[i] };
+    pieces[i] = { rn_pvt: p_pvt_nodes[i], rn_shr: p_shr_nodes[i],
+                  rn_ghost: p_ghost_nodes[i], rw_pvt: p_wires[i] };
   while(!done) {

### 2012-04-21 doc/sc2012/code_ex.tex
added code for the child tasks

@@
 \begin{lstlisting}[float={t},label={lst:code_ex},caption={Circuit Simulation Code Example}]
-struct Node { float charge, capacitance; };
-struct Wire { Node in_node, out_node; float current, ... ; };
-struct Circuit {
-  region   r_all_nodes;
-  region   r_all_wires; };
+struct Node { float voltage, new_charge, capacitance; };
+struct Wire<rn> { Node@rn in_node, out_node; float current, ... ; };
+struct Circuit { region   r_all_nodes; /* contains all nodes for the circuit */
+                 region   r_all_wires; /* contains all circuit wires */ };
+struct CircuitPiece {
+  region  rn_pvt, rn_shr, rn_ghost; /* private, shared, ghost node regions */
+  region  rw_pvt;                   /* private wires region */ };
 
-struct CircuitPiece {
-  region  rn_pvt, rn_shr, rn_ghost; // private, shared, ghost node regions
-  region  rw_pvt;                   // private wires region };
-
-void simulate_circuit(Circuit c) : RWE(c.r_all_nodes,c.r_all_wires)
+void simulate_circuit(Circuit c, float dt) : RWE(c.r_all_nodes,c.r_all_wires)
 {
@@
     spawn (i = 0; i #$<$# MAX_PIECES; i++) calc_new_currents(pieces[i]);
-    spawn (i = 0; i #$<$# MAX_PIECES; i++) distribute_charge(pieces[i]);
+    spawn (i = 0; i #$<$# MAX_PIECES; i++) distribute_charge(pieces[i], dt);
     spawn (i = 0; i #$<$# MAX_PIECES; i++) update_voltages(pieces[i]);
@@
 
-// read info from nodes connected to each wire, update state of wire
-void calc_new_currents(CircuitPiece piece): RWE(piece.rw_pvt),
-                                            ROE(piece.rn_pvt,piece.rn_shr,piece.rn_ghost)
-// current moving through wires redistributes charge between nodes
-void distribute_charge(CircuitPiece piece): ROE(piece.rw_pvt),
-                                            RWE(piece.rn_pvt),
-                                            RdA(piece.rn_shr,piece.rn_ghost)
-// total charge added to a node causes changes in voltage
-void update_voltages(CircuitPiece piece): RWE(piece.rn_pvt,piece.rn_shr)
+void calc_new_currents(CircuitPiece piece):
+        RWE(piece.rw_pvt), ROE(piece.rn_pvt,piece.rn_shr,piece.rn_ghost) {
+  foreach(w : piece.rw_pvt)
+    w->current = (w->in_node->voltage - w->out_node->voltage) / w->resistance;
+}
+
+void distribute_charge(CircuitPiece piece, float dt):
+        ROE(piece.rw_pvt), RdA(piece.rn_pvt, piece.rn_shr,piece.rn_ghost) {
+  foreach(w : piece.rw_pvt) {
+    w->in_node->new_charge += -dt * w->current;
+    w->out_node->new_charge +=  dt * w->current;
+  }
+}
+
+void update_voltages(CircuitPiece piece): RWE(piece.rn_pvt,piece.rn_shr) {
+  foreach(n : piece.rn_pvt, piece.rn_shr) {
+    n->voltage += n->new_charge / n->capacitance;
+    n->new_charge = 0;
+  }
+}
 \end{lstlisting}

### 2012-04-26 doc/sc2012/legion.tex
Working on the code example for the paper

@@
 
-\bibliographystyle{plainnat}
+\bibliographystyle{IEEEtran}
 \usepackage{amsmath,amsthm}
@@
 
+\pdfpagewidth=8.5in
+\pdfpageheight=11in
+
 \title{Legion: Expressing Locality and Independence with Logical Regions}
-\author{Author names omitted}
+\author{
+\IEEEauthorblockN{Michael Bauer\IEEEauthorrefmark{1},
+%\authorinfo(Stanford University, email: \mbox{mebauer@cs.stanford.edu}
+Sean Treichler\IEEEauthorrefmark{1}, 
+%\authorinfo(Stanford University, email: \mbox{sjt@cs.stanford.edu}
+Elliott Slaughter\IEEEauthorrefmark{1}, 
+% \authorinfo(Stanford University, email: \mbox{slaughter@cs.stanford.edu}
+Alex Aiken\IEEEauthorrefmark{1}  
+%\authorinfo(Stanford University, email: \mbox{aiken@cs.stanford.edu}
+%\email{aiken@cs.stanford.edu}
+}
+\IEEEauthorblockA{
+\IEEEauthorrefmark{1}{Computer Science Department, Stanford University}
+}
+}
 
@@
 \begin{abstract}
-%As transistor counts have continued to scale with Moore's Law, computer
-%architectures are becoming increasingly parallel to avoid fundamental
-%physical limitations.  
-Modern parallel architectures have both
-heterogeneous processors and deep, complex memory hierarchies.  
-%New
-%programming abstractions are required to grant programmers the power
-%to leverage these architectures.  With this problem in mind 
-We present Legion, 
-a programming model, type system, and runtime for programming these machines.
-Legion is organized around {\em logical regions}, which express both locality and independence of program data.  
-Legion also enables explicit, programmer controlled movement of data through
-the memory hierarchy and placement of tasks based on locality information
-via a novel mapping interface.  Running on a 4 node cluster with 8 total GPUs 
-and 4 levels of memory hierarchy, our implementation of Legion 
-achieves up to 5.9X speedup over a single CPU-GPU node on real-world applications.
+Modern parallel architectures have both heterogeneous processors and
+deep, complex memory hierarchies.  We present Legion, a programming
+model and runtime system for achieving high performance on these
+machines.  Legion is organized around {\em logical regions}, which
+express both locality and independence of program data and {\em
+  tasks}, functions that perform computations on regions.  We describe
+a runtime system that dynamically extracts parallelism from
+Legion programs, using a distributed, parallel scheduling algorithm
+that identifies both independent tasks and nested parallelism.  Legion
+also enables explicit, programmer controlled movement of data through
+the memory hierarchy and placement of tasks based on locality
+information via a novel mapping interface.  We evaluate our Legion
+implementation using four applications: fluid-flow on a regular grid,
+a three-level AMR code solving a simple heat transfer equation, a circuit simulation,
+and LINPACK.
+%Running on a 4 node cluster with 8 total GPUs 
+%and 4 levels of memory hierarchy, our implementation of Legion 
+%achieves up to 5.9X speedup over a single CPU-GPU node on real-world applications.
 \end{abstract}
@@
 \label{sect:intro}
-Put intro text here \cite{Fatahalian06}
+%Put intro text here \cite{Fatahalian06}
 
+Modern parallel machines are increasingly complex, with deep,
+distributed memory hierarchies and heterogeneous processing units.  On
+these architectures, it is crucial for performance that the programmer
+and the compiler be able to reason about {\em locality} (data resident
+close to computation that uses it) and {\em independence} (computations
+operating on disjoint data).  Most contemporary
+languages have no facilities for the programmer to express locality
+and independence.  The few languages that do focus primarily on
+array-based locality \cite{Fatahalian06,CHAPEL04,UPC99} and 
+avoid irregular pointer data structures.
+%encode such structures in arrays, which 
+%data structures encoded in arrays which are impossible for compilers
+%and runtime systems to reason about with regard to locality and independence.
+
+In this paper we describe Legion, a parallel programming system based
+on using {\em logical regions} to describe the organization of data.
+A logical region names a set of objects.  Logical regions are first
+class values in Legion, and in particular may be passed as arguments
+to functions that access the data in those regions, providing locality
+information.  Logical regions may be {\em partitioned} into (usually)
+disjoint {\em subregions}, providing information for determining independence of computations.  Furthermore,
+computations access logical regions with particular {\em privileges} (e.g., {\em
+  read-only} and {\em read-write}, among others) 
+and {\em coherence} (e.g., {\em exclusive access} and {\em atomic access}, among others).
+
+We begin with an extended example of a Legion program in Section~\ref{sec:ex}.
+Each subsequent section describes a level in the Legion programming
+system from the application down to the machine:
+
+\begin{itemize}
+
+\item We summarize Legion's static type system (Section~\ref{sec:type}), which serves two primary
+  purposes.  First, pointers to regions are checked
+  statically, eliminating the need for expensive runtime checks.
+  Second, the invariant that a function $f$ accesses only
+  regions passed as arguments to $f$, or subregions of those regions, is enforced.
+  This property guarantees the Legion implementation can rely on
+  preservation of independence---if two functions are passed disjoint
+  region arguments, their computations access disjoint
+data.
+
+\item We describe the execution semantics for Legion and show that
+analyzing region dependencies within a single function is sufficient to 
+guarantee that a Legion execution preserves sequential program
+behavior if regions are accessed with the strongest (exclusive)
+coherence privilege (Section~\ref{sec:exec}).  This local scheduling property is central to 
+the scalability of Legion on distributed memory machines.
+
+\item We describe the design and implementation of the Legion {\em high-level runtime}
+system (Section~\ref{sec:highlevel}), which performs scheduling of {\em tasks} (functions to be executed in parallel)
+and implements each logical region as one or more {\em physical regions}.
+The high-level runtime's scheduling algorithm is analogous to an out-of-order processor,
+but works at the granularity of the tasks and regions in a function body rather than the instructions and registers
+in a code block.  We describe the stages of the task execution pipeline, and how the high-level runtime coordinates parallel
+scheduling decisions across the machine.  We also present the {\em mapping} interface, an API for incorporating
+application-specific knowledge into the high-level runtime's decisions.
+
+\item We describe the design and implementation of the Legion {\em low-level runtime} (Section~\ref{sec:lowlevel}),
+a portability layer designed to abstract a wide variety of hardware, including multicore chips, clusters, and accelerators
+such as GPUs.
+
+\item We present the results of experiments on two applications running on a cluster of multicore processors with
+attached GPUs (Section~\ref{sec:exp}).  The applications are both irregular and illustrate the capability of Legion to 
+exploit locality and independence on such platforms.
+Running on a 4 node cluster with 8 total GPUs 
+and 4 levels of memory hierarchy, our implementation of Legion 
+achieves up to 5.9X speedup over a single CPU-GPU node on these applications.
+
+\end{itemize}
 %\input{circuit}
-%\input{code_ex}
+\input{code_ex}
 %\input{part_fig}

### 2012-04-26 doc/sc2012/legion.tex
intro first pass

@@
 class values in Legion, and in particular may be passed as arguments
-to functions that access the data in those regions, providing locality
+to distinguished functions called {\em tasks} that access the data in those regions, providing locality
 information.  Logical regions may be {\em partitioned} into (usually)
@@
 and {\em coherence} (e.g., {\em exclusive access} and {\em atomic access}, among others).
+Priveleges express how a task may use its region arguments, providing data dependence information
+that is used to guide the extraction of parallelism (e.g., if two tasks access the same region with
+{\em read-only} priveleges the two tasks can potentially be run in parallel), and coherence annotations
+express the required semantics of concurrent accesses to the region.  For example, if tasks $x$ and $y$ access region $r$
+with {\em exclusive} access, then the effects of $x$ and $y$ on $r$ are guaranteed to be the same as if they had executed
+in the sequential order expressed by the program; on the other hand, if the tasks access $r$ with {\em atomic} coherence, then 
+Legion guarantees only  atomicity of the tasks with respect to $r$: either $x$ appears to run entirely before $y$ or vice versa.
 
-We begin with an extended example of a Legion program in Section~\ref{sec:ex}.
-Each subsequent section describes a level in the Legion programming
-system from the application down to the machine:
+To introduce the programming model, we begin with an extended example of a circuit simulation programmed in Legion
+in Section~\ref{sec:ex}. This example illustrates regions, tasks, permissions/coherence properties, the interactions
+between them, and how these building blocks are assembled into a realistic Legion program.   Subsequent sections each
+describe a contribution in the implementation and evaluation of Legion:
 
@@
 
-\item We summarize Legion's static type system (Section~\ref{sec:type}), which serves two primary
-  purposes.  First, pointers to regions are checked
-  statically, eliminating the need for expensive runtime checks.
-  Second, the invariant that a function $f$ accesses only
-  regions passed as arguments to $f$, or subregions of those regions, is enforced.
-  This property guarantees the Legion implementation can rely on
-  preservation of independence---if two functions are passed disjoint
-  region arguments, their computations access disjoint
-data.
+\item We define a {\em software out-of order processor}, or SOOP, for scheduling tasks with region arguments in a manner
+analogous to how out-of-order hardware schedulers process instructions with register arguments.  In addition to pipelining
+the scheduling of tasks over several stages, our SOOP is also hierarchical and can extract nested parallelism (because tasks may recursively spawn subtasks) as well as concurrent and distributed across the machine (Section~\ref{sec:soop}).
 
-\item We describe the execution semantics for Legion and show that
-analyzing region dependencies within a single function is sufficient to 
-guarantee that a Legion execution preserves sequential program
-behavior if regions are accessed with the strongest (exclusive)
-coherence privilege (Section~\ref{sec:exec}).  This local scheduling property is central to 
-the scalability of Legion on distributed memory machines.
+\item A crucial part of orchestrating a program's execution is how
+tasks are assigned (or {\em mapped}) to processors and how regions are
+mapped to specific memory units within the machine.  Often there is
+application-specific information that, if exploited, results in better
+mappings than a generic, one-size-fits-all mapping strategy.  We describe
+a {\em mapping interface} that allows programmers to give the SOOP a specification
+of how to map tasks and regions.  This mapping API is carefully defined so that
+any use-defined mapping strategy can only affect the performance of applications,
+not their correctness.  Thus, user mappings only affect the time and resources required
+to run the program and cannot change the program's output (Section~\ref{sec:mapping}).
 
-\item We describe the design and implementation of the Legion {\em high-level runtime}
-system (Section~\ref{sec:highlevel}), which performs scheduling of {\em tasks} (functions to be executed in parallel)
-and implements each logical region as one or more {\em physical regions}.
-The high-level runtime's scheduling algorithm is analogous to an out-of-order processor,
-but works at the granularity of the tasks and regions in a function body rather than the instructions and registers
-in a code block.  We describe the stages of the task execution pipeline, and how the high-level runtime coordinates parallel
-scheduling decisions across the machine.  We also present the {\em mapping} interface, an API for incorporating
-application-specific knowledge into the high-level runtime's decisions.
-
-\item We describe the design and implementation of the Legion {\em low-level runtime} (Section~\ref{sec:lowlevel}),
-a portability layer designed to abstract a wide variety of hardware, including multicore chips, clusters, and accelerators
-such as GPUs.
-
-\item We present the results of experiments on two applications running on a cluster of multicore processors with
-attached GPUs (Section~\ref{sec:exp}).  The applications are both irregular and illustrate the capability of Legion to 
-exploit locality and independence on such platforms.
-Running on a 4 node cluster with 8 total GPUs 
-and 4 levels of memory hierarchy, our implementation of Legion 
-achieves up to 5.9X speedup over a single CPU-GPU node on these applications.
+\item We present the results of experiments on four applications: fluid-flow on a regular grid,
+a three-level AMR code solving a simple heat transfer equation, a circuit simulation,
+and LINPACK.  We run each of these four codes and best available reference versions on three different
+clusters of multicore processors with attached GPUs: a small cluster of our own, a medium-sized cluster designed for visualization applications, and up to half of the Keeneland supercomputer (Section~\ref{sec:results}).
 

### 2012-04-27 doc/sc2012/soop.tex
new files

\section{The Software Out-of-Order Processor}
\label{sec:soop}

### 2012-04-27 doc/sc2012/circuit.tex
fix up dependenencies and references

--- 2012-01-02 doc/pldi2012/circuit.tex
@@
 are touched by a single task in a phase, while shared nodes may
-be referenced by multiple tasks in a phase.  The thick lines in figure \ref{sfig:part_fig:pvs}
+be referenced by multiple tasks in a phase.  The thick lines in Figure~\ref{sfig:part_fig:pvs}
 show how the nodes in a small graph might be partitioned into three subsets.  Gray
@@
 by the Legion runtime based on the region access annotations on the task declarations
-(see Section~\ref{subsec:cirdependence}).
+(see Section~\ref{sec:soop}).
 

### 2012-04-27 doc/sc2012/code_ex.tex
fix up dependenencies and references

@@
 }
-\begin{lstlisting}[float={t},label={lst:code_ex},caption={Circuit Simulation Code Example}]
+\begin{lstlisting}[float={t},label={lst:code_ex},caption={Circuit simulation.}]
 struct Node { float voltage, new_charge, capacitance; };

### 2012-04-27 doc/sc2012/legion.tex
fix up dependenencies and references

@@
 and LINPACK.  We run each of these four codes and best available reference versions on three different
-clusters of multicore processors with attached GPUs: a small cluster of our own, a medium-sized cluster designed for visualization applications, and up to half of the Keeneland supercomputer (Section~\ref{sec:results}).
+clusters of multicore processors with attached GPUs: a small cluster of our own, a medium-sized cluster designed for visualization applications, and up to half of the Keeneland supercomputer (Section~\ref{sec:exp}).
 
 \end{itemize}
-%\input{circuit}
 \input{code_ex}
-%\input{part_fig}
+\input{circuit}
+\input{part_fig}
+\input{soop}
+\input{mapping}cir
+\input{experiments}
+\input{related}
+\input{conclusion}
 
-%\input{type_system}
-%\input{exec}
-
-%\input{highlevel}
 %\input{mapping_fig}
 
-%\input{lowlevel}
-%\input{experiments}
-%\input{related}
-%\input{conclusion}
-
 \bibstyle{IEEEtran}
-{
-\tiny
 \bibliography{bibliography}
-}
 

### 2012-04-27 doc/sc2012/legion.tex
secs 1 and 2

@@
 machines.  Legion is organized around {\em logical regions}, which
-express both locality and independence of program data and {\em
+express both locality and independence of program data, and {\em
   tasks}, functions that perform computations on regions.  We describe
@@
 implementation using four applications: fluid-flow on a regular grid,
-a three-level AMR code solving a simple heat transfer equation, a circuit simulation,
+a three-level AMR code solving a standard heat diffusion equation, a circuit simulation,
 and LINPACK.
@@
 class values in Legion, and in particular may be passed as arguments
-to distinguished functions called {\em tasks} that access the data in those regions, providing locality
-information.  Logical regions may be {\em partitioned} into (usually)
-disjoint {\em subregions}, providing information for determining independence of computations.  Furthermore,
-computations access logical regions with particular {\em privileges} (e.g., {\em
-  read-only} and {\em read-write}, among others) 
-and {\em coherence} (e.g., {\em exclusive access} and {\em atomic access}, among others).
-Priveleges express how a task may use its region arguments, providing data dependence information
-that is used to guide the extraction of parallelism (e.g., if two tasks access the same region with
-{\em read-only} priveleges the two tasks can potentially be run in parallel), and coherence annotations
-express the required semantics of concurrent accesses to the region.  For example, if tasks $x$ and $y$ access region $r$
-with {\em exclusive} access, then the effects of $x$ and $y$ on $r$ are guaranteed to be the same as if they had executed
-in the sequential order expressed by the program; on the other hand, if the tasks access $r$ with {\em atomic} coherence, then 
-Legion guarantees only  atomicity of the tasks with respect to $r$: either $x$ appears to run entirely before $y$ or vice versa.
+to distinguished functions called {\em tasks} that access the data in
+those regions, providing locality information.  Logical regions may be
+{\em partitioned} into (usually) disjoint {\em subregions}, providing
+information for determining independence of computations.
+Furthermore, computations access logical regions with particular {\em
+  privileges} (e.g., {\em read-only} and {\em read-write}, among
+others) and {\em coherence} (e.g., {\em exclusive access} and {\em
+  atomic access}, among others).  Priveleges express how a task may
+use its region arguments, providing data dependence information that
+is used to guide the extraction of parallelism (e.g., if two tasks
+access the region $r$with {\em read-only} priveleges the two tasks
+can potentially be run in parallel, at least with respect to $r$),
+and coherence properties express the required semantics of concurrent
+accesses to the region.  For example, if tasks $x$ and $y$ access
+region $r$ with {\em exclusive} access, then the effects of $x$ and
+$y$ on $r$ are guaranteed to be the same as if they had executed in
+the sequential order expressed by the program; on the other hand, if
+the tasks access $r$ with {\em atomic} coherence, then Legion
+guarantees only atomicity of the tasks with respect to $r$: either $x$
+appears to run entirely before $y$ or vice versa.
 
-To introduce the programming model, we begin with an extended example of a circuit simulation programmed in Legion
-in Section~\ref{sec:ex}. This example illustrates regions, tasks, permissions/coherence properties, the interactions
-between them, and how these building blocks are assembled into a realistic Legion program.   Subsequent sections each
-describe a contribution in the implementation and evaluation of Legion:
+To introduce the programming model, we begin with an extended example
+of a circuit simulation programmed in Legion in
+Section~\ref{sec:ex}. This example illustrates regions, tasks,
+permissions and coherence properties, the interactions between them,
+and how these building blocks are assembled into a Legion
+program.  Subsequent sections each describe a contribution in the
+implementation and evaluation of Legion:
 
@@
 
-\item We define a {\em software out-of order processor}, or SOOP, for scheduling tasks with region arguments in a manner
-analogous to how out-of-order hardware schedulers process instructions with register arguments.  In addition to pipelining
-the scheduling of tasks over several stages, our SOOP is also hierarchical and can extract nested parallelism (because tasks may recursively spawn subtasks) as well as concurrent and distributed across the machine (Section~\ref{sec:soop}).
+\item We define a {\em software out-of-order processor}, or SOOP, for
+  scheduling tasks with region arguments in a manner analogous to how
+  out-of-order hardware schedulers process instructions with register
+  arguments.  In addition to pipelining the execution of tasks over
+  several stages, our SOOP is hierarchical and naturally extracts nested
+  parallelism (because tasks may recursively spawn subtasks) as well
+  as concurrent and distributed across the machine
+  (Section~\ref{sec:soop}).
 
@@
 a {\em mapping interface} that allows programmers to give the SOOP a specification
-of how to map tasks and regions.  This mapping API is carefully defined so that
-any use-defined mapping strategy can only affect the performance of applications,
+of how to map tasks and regions for a specific application, or even part of an
+application.  This mapping API is carefully defined so that
+any user-supplied mapping strategy can only affect the performance of applications,
 not their correctness.  Thus, user mappings only affect the time and resources required
@@
 
-\item We present the results of experiments on four applications: fluid-flow on a regular grid,
-a three-level AMR code solving a simple heat transfer equation, a circuit simulation,
-and LINPACK.  We run each of these four codes and best available reference versions on three different
-clusters of multicore processors with attached GPUs: a small cluster of our own, a medium-sized cluster designed for visualization applications, and up to half of the Keeneland supercomputer (Section~\ref{sec:exp}).
+\item We present the results of experiments on four applications:
+  fluid-flow on a regular grid, a three-level AMR code solving a
+  standard heat diffusion equation, a circuit simulation, and LINPACK.
+  We run each of these four codes and best available reference
+  versions on three different clusters of multicore processors with
+  attached GPUs: a small cluster of our own (4 nodes with 2 6-core CPUs and 2 GPUs per node), a medium-sized cluster
+  designed for visualization applications (10 nodes with 2 6-core CPUs and 5 GPUs per node), and up to a third of the
+  Keeneland supercomputer (32 nodes, 2 6-core CPUs and 3 GPUs per node, see Section~\ref{sec:exp}).
 
@@
 \input{soop}
-\input{mapping}cir
+\input{mapping}
 \input{experiments}

### 2012-04-27 doc/sc2012/code_ex.tex

@@
 {
-  // Partition of wires into pieces
+  // The construction of the colorings is not shown.  The colorings wire_owner_map,
+  // node_owner_map, and node_neighbor_map have MAX_PIECES colors 
+  // 0..MAX_PIECES #$-$# 1. The coloring node_sharing map has two colors 0 and 1.
+  //
+  // Partition of wires into MAX_PIECES pieces
   partition<disjoint> p_wires = c.r_all_wires.partition(wire_owner_map); 
@@
 
-  // Partition all-private into separate circuit pieces
+  // Partition all-private into MAX_PIECES disjoint circuit pieces
   partition<disjoint> p_pvt_nodes = p_nodes_pvs[0].partition(node_owner_map);
-  // Partition all-shared into separate circuit pieces
+  // Partition all-shared into MAX_PIECES disjoint circuit pieces
   partition<disjoint> p_shr_nodes = p_nodes_pvs[1].partition(node_owner_map);
-  // Partition all-shared into ghost regions which may be aliased
+  // Partition all-shared into MAX_PIECES ghost regions, which may be aliased
   partition<aliased> p_ghost_nodes = p_nodes_pvs[1].partition(node_neighbor_map);
@@
   foreach(w : piece.rw_pvt)
-    w->current = (w->in_node->voltage - w->out_node->voltage) / w->resistance;
+    w#$\rightarrow$#current = (w#$\rightarrow$#in_node#$\rightarrow$#voltage - w#$\rightarrow$#out_node#$\rightarrow$#voltage) / w#$\rightarrow$#resistance;
 }
@@
   foreach(w : piece.rw_pvt) {
-    w->in_node->new_charge += -dt * w->current;
-    w->out_node->new_charge +=  dt * w->current;
+    w#$\rightarrow$#in_node#$\rightarrow$#new_charge += -dt * w#$\rightarrow$#current;
+    w#$\rightarrow$#out_node#$\rightarrow$#new_charge +=  dt * w#$\rightarrow$#current;
   }
@@
   foreach(n : piece.rn_pvt, piece.rn_shr) {
-    n->voltage += n->new_charge / n->capacitance;
-    n->new_charge = 0;
+    n#$\rightarrow$#voltage += n#$\rightarrow$#new_charge / n#$\rightarrow$#capacitance;
+    n#$\rightarrow$#new_charge = 0;
   }

### 2012-04-27 doc/sc2012/circuit.tex

@@
 
-%\include{code_ex}
-
-In this section we give an informal overview of Legion through an example.
+In this section we give an overview of Legion using an example.
 Listing~\ref{lst:code_ex} shows code for an electrical
@@
 describe the partitioning of regions in Section~\ref{subsec:partitioning}
-and then describe the data type declarations 
-and the constraints they enforce 
-in Section~\ref{subsec:datatypes}.
+and then describe the data type declarations in Section~\ref{subsec:datatypes}.
 
@@
 
-Line 18 declares the main simulator function, which takes a {\tt
+Line 9 declares the main simulator function, which takes a {\tt
 Circuit} to be simulated.  This function specifies the regions it 
@@
 with read-write privileges and {\em exclusive} coherence (i.e., no other
-function can access the regions concurrently).  The simulation 
+function can access these two regions concurrently).  The simulation 
 reads and writes all nodes and wires, and it must be done
@@
 
-Lines 19-27 are responsible for partitioning the circuit into {\tt MAX\_PIECES}
+Lines 15-21 are responsible for partitioning the circuit into {\tt MAX\_PIECES}
 pieces that can be worked on in parallel.  To partition a region, we
@@
 
-Wires are simply partitioned in {\tt MAX\_PIECES} subregions (line 20).  
+Wires are simply partitioned in {\tt MAX\_PIECES} subregions (line 16).  
 The nodes are first partitioned into 
@@
 We next further partition the private and shared subregions into {\tt
-MAX\_PIECES} disjoint subregions each (lines 23-24); one subregion of the private and shared node partitions
-is shown in Figures~\ref{sfig:part_fig:p_i} and \ref{sfig:part_fig:s_i} respectively.  
-We partition the shared nodes again
-into the sets of ghost nodes for each task.  Since ghost nodes may be read by multiple tasks,
-nodes may be colored multiple times and the ghost regions are not disjoint; one of the ghost subregions is shown in Figure~\ref{sfig:part_fig:g_i}.
-Figure~\ref{sfig:part_fig:tree} shows the final hierarchy of node partitions.  The $*$ symbol indicates a partition is 
-disjoint.
+  MAX\_PIECES} disjoint subregions each (lines 20-23); one subregion
+of the private and shared node partitions is shown in
+Figures~\ref{sfig:part_fig:p_i} and \ref{sfig:part_fig:s_i}
+respectively.
+
+We partition the shared nodes a second time using a
+different coloring into the sets of ghost nodes for each task.  This 
+partition is noteworthy because the subregions are {\em not} disjoint:
+ghost nodes may be read by multiple tasks, and thus a node $n$  may belong 
+to more than one ghost region, which corresponds to $n$ being assigned
+more than one color by the coloring {\tt node\_neighbor\_map}.
+One of the ghost subregions is shown in Figure~\ref{sfig:part_fig:g_i}.
+
+Figure~\ref{sfig:part_fig:tree} shows the final hierarchy of node
+partitions;  the $*$ symbol indicates a partition is disjoint. This
+{\em region tree} data structure plays an important role in scheduling
+tasks for out-of-order execution (see Section~\ref{sec:soop}).
 
@@
 
-Lines 36-40 form the bulk of the actual simulation, performing three passes
-over the circuit in each iteration.  Each pass loops over an array of pieces (constructed on lines 30-34 from the partitions),
+Lines 31-57 form the bulk of the actual simulation, performing three passes
+over the circuit in each iteration.  Each pass loops over an array of pieces (constructed on lines 27-30 from the partitions),
 spawning a task for each piece.   There are no explicit requests for
-parallel execution, nor is there synchronization required between the
+parallel execution (the default semantics of {\tt spawn} is sequential 
+execution of the spawned tasks) nor is there synchronization required between the
 passes.  Both the fact that the tasks can be run in parallel within a pass
@@
 
-\subsection{Data Types and Constraints}
+\subsection{Data Types}
 \label{subsec:datatypes}
-The two basic data types are {\tt Node}s and {\tt Wire}s.
-{\tt Node}s are linked in circular lists.
-The simulation relies on different lists being in different
-regions, so the {\tt Node} definition on line 1 is parameterized by a region
-{\tt rn}.  The {\tt next} field of a $\tt Node \langle rn \rangle$ has
-type $\tt Node \langle rn \rangle \mbox{\tt @} rn$, which means the field points
-to an object in region {\tt rn} of type $\tt Node \langle rn \rangle$, 
-implying that all nodes of the list are in region {\tt rn}.
+The two basic data types in the circuit simulation are {\tt Node}s and
+{\tt Wire}s declared on lines 1 and 2 of Listing~\ref{lst:code_ex}.
+As is standard in region-based systems, allocating a {\tt Node}, {\tt
+  Wire}, or any value in the heap requires naming the logical region
+in which the value is allocated.  Furthermore, pointers to heap values
+declare the region to which they point.  For example, the definition
+of {\tt Wire} is parameterized on the region {\tt rn} to which the
+{\tt Node} pointers in fields {\tt in\_nodes} and {\tt out\_nodes}
+point, which guarantees that the two endpoints of a wire are always in
+the same region.
 
-{\tt Wire}s are similar, except that a wire refers to two nodes to
-which it is electrically connected.  The two {\tt Node}s are in region {\tt rn},
-but they may point to {\tt Node}s in a different region {\tt rn2}.  
-This is a common pattern in Legion code, and
-arises when a constraint needs to be placed on the objects directly
-referenced, and a second (presumably weaker) constraint on objects
-that are indirectly reached through the direct pointers.
-
-In cases where we need to create compound data structures with types whose
-region parameters are self-referential we declare a {\em region relationship}.
-A region relationship is a structure with named fields, 
-except that any region fields can be used in the type declarations
-for other fields.  Lines 4-9 declare the region relationship {\tt Circuit}.
-In {\tt Circuit} the region {\tt r\_all\_nodes} contains {\tt
-Node}s that point only into region {\tt r\_all\_nodes}, guaranteeing that
-all {\tt Node}s are in region {\tt r\_all\_nodes}.  (Note that regions are
-declared with the single data type they can contain.)  Similarly, the
-type of the wire region's element not only guarantees that every wire
-in the circuit is in the wire region, but also guarantees that every
-node referred to by any wire is in the circuit's node region.
-
-Since a region relationship can have
-fields with types that refer to its own fields' regions, it is often not possible
-to assign fields one at a time without violating the type checking rules.
-The simultaneous assignment operator asks the type checker only to ensure
-that fields have the correct types after all the fields are updated.  In
-many cases, the target type also tries to describe tighter constraints on
-the region parameters used for field types.  The {\tt region\_cast} operator
-allows the programmer to assert the correctness of this transformation, which
-can be verified with a linear scan over the elements of a region.
-Lines 32-34 illustrate a simultaneous
-assignment to update a {\tt CircuitPiece} region relationship.
-
-The last type definition is for the pieces into which the circuit is partitioned.
-The original node and wire regions are specified as parameters, and the
-four kinds of subregions that are created are made fields of the
-{\tt CircuitPiece} (lines 10-17).  This declaration also illustrates {\em subregion constraints}
-placed on the three region fields stating the region fields must be subregions of the 
-node and wire regions provided as parameters (line 16).  Since the simulation 
-iterates over the nodes owned by a {\tt CircuitPiece}, the parameterization of the {\tt rn\_pvt} (private nodes),
-{\tt rn\_shr} (shared nodes), and {\tt first\_node} fields constrains the linked list of
-nodes to reside entirely in those two regions.  The linked list of wires is
-similarly constrained, and the wire's node pointers are constrained to fall
-into one of the three subregions.
-% defined in the {\tt CircuitPiece}.  
-Finally,
-the use of {\tt rn} for {\tt rn\_ghost}'s parameter and the second parameter
-in the wire type places no constraints on which nodes a ghost node might point
-to.
+Regions are first-class values in Legion and can be stored in heap
+data structures.  For example, a {\tt Circuit} (defined on line 3)
+holds the regions hold all nodes and wires of the circuit and a {\tt
+  CircuitPiece} (defined on line 4) hold pointers to the private,
+shared, and ghost nodes as well as the private wires of its piece of
+the circuit.
 

### 2012-04-27 doc/sc2012/circuit.tex
edits

@@
 
-In this section we give an overview of Legion using an example.
 Listing~\ref{lst:code_ex} shows code for an electrical
-circuit simulation, which illustrates the core features of the programming model.
-
+circuit simulation, which illustrates the core features of the Legion programming model.
 The circuit simulation takes a collection of
@@
 different coloring into the sets of ghost nodes for each task.  This 
-partition is noteworthy because the subregions are {\em not} disjoint:
+partition is noteworthy because the subregions are not disjoint:
 ghost nodes may be read by multiple tasks, and thus a node $n$  may belong 
@@
   Wire}, or any value in the heap requires naming the logical region
-in which the value is allocated.  Furthermore, pointers to heap values
-declare the region to which they point.  For example, the definition
-of {\tt Wire} is parameterized on the region {\tt rn} to which the
-{\tt Node} pointers in fields {\tt in\_nodes} and {\tt out\_nodes}
-point, which guarantees that the two endpoints of a wire are always in
-the same region.
+in which the value is allocated.  For example, {\tt new Node@r} returns
+a reference to a new value of {\tt Node} type in region {\tt r}.  
+Listing~\ref{lst:code_ex} illustrates one way of constructing partitioned
+data structures in Legion: populate a region with data (using {\tt new}) and then partition it
+into subregions.  Legion also supports partitioning an empty region into subregions
+and then allocating data in the subregions; we use both approaches in our
+application discussed in Section~\ref{sec:ex}.
+
+Pointers to region values declare the region to which they point.  For
+example, the definition of {\tt Wire} is parameterized on the region
+{\tt rn} to which the {\tt Node} pointers in fields {\tt in\_nodes}
+and {\tt out\_nodes} point, which guarantees that the two endpoints of
+a wire are always in the same region.
 

### 2012-04-27 doc/sc2012/legion.tex
edits

@@
 Modern parallel machines are increasingly complex, with deep,
-distributed memory hierarchies and heterogeneous processing units.  On
-these architectures, it is crucial for performance that the programmer
+distributed memory hierarchies and heterogeneous processing units.  Because the
+costs of communication within these architectures vary by several orders of magnitude, 
+the penalty for mistakes in the placement of data or computation is very poor performance.  Thus,
+to achieve high performance it is crucial that the programmer
 and the compiler be able to reason about {\em locality} (data resident
 close to computation that uses it) and {\em independence} (computations
-operating on disjoint data).  Most contemporary
-languages have no facilities for the programmer to express locality
+operating on disjoint data, and therefore not requiring communication and able to be placed in possibly distant parts of the machine).  Most contemporary
+programming systems have no facilities for the programmer to express locality
 and independence.  The few languages that do focus primarily on
@@
 others) and {\em coherence} (e.g., {\em exclusive access} and {\em
-  atomic access}, among others).  Priveleges express how a task may
-use its region arguments, providing data dependence information that
-is used to guide the extraction of parallelism (e.g., if two tasks
-access the region $r$with {\em read-only} priveleges the two tasks
-can potentially be run in parallel, at least with respect to $r$),
-and coherence properties express the required semantics of concurrent
+  atomic access}, among others).  Priveleges express how a task may use
+its region arguments, providing data dependence information that
+is used to guide the extraction of parallelism.  For example, if two tasks
+access the same region with {\em read-only} privileges the two tasks
+can potentially be run in parallel.
+Coherence properties express the required semantics of concurrent
 accesses to the region.  For example, if tasks $x$ and $y$ access
-region $r$ with {\em exclusive} access, then the effects of $x$ and
+region $r$ with {\em exclusive} access, then Legion guarantees that the effects of $x$ and
 $y$ on $r$ are guaranteed to be the same as if they had executed in
-the sequential order expressed by the program; on the other hand, if
+the sequential order expressed by the program. On the other hand, if
 the tasks access $r$ with {\em atomic} coherence, then Legion
@@
 
+Legion programs are written using logical regions, each of which
+has one or more {\em physical instances} when the program executes.  Often
+logical regions correspond to a single physical instance, but there are common situations
+in which it is desirable to have multiple physical copies of a logical region's
+contents in different parts of the machine (e.g., to duplicate and distribute read-only
+data) or even to have physical instances of a logical region that temporarily diverge
+in their contents (e.g., to keep an old copy of a region to break a write-after-read dependency, or to 
+allow independent, local reductions that are later combined).
+
 To introduce the programming model, we begin with an extended example
-of a circuit simulation programmed in Legion in
+of a circuit simulation in
 Section~\ref{sec:ex}. This example illustrates regions, tasks,
@@
 implementation and evaluation of Legion:
+
+\input{code_ex}
 
@@
 \item A crucial part of orchestrating a program's execution is how
-tasks are assigned (or {\em mapped}) to processors and how regions are
+tasks are assigned (or {\em mapped}) to processors and how physical instances of logical regions are
 mapped to specific memory units within the machine.  Often there is
@@
 \end{itemize}
-\input{code_ex}
+
 \input{circuit}

### 2012-04-28 doc/sc2012/legion.tex
intro

@@
 costs of communication within these architectures vary by several orders of magnitude, 
-the penalty for mistakes in the placement of data or computation is very poor performance.  Thus,
+the penalty for mistakes in the placement of data or computation is usually very poor performance.  Thus,
 to achieve high performance it is crucial that the programmer
@@
 is used to guide the extraction of parallelism.  For example, if two tasks
-access the same region with {\em read-only} privileges the two tasks
+access the same region with read-only privileges the two tasks
 can potentially be run in parallel.
@@
 region $r$ with {\em exclusive} access, then Legion guarantees that the effects of $x$ and
-$y$ on $r$ are guaranteed to be the same as if they had executed in
+$y$ on $r$ are the same as if they had executed in
 the sequential order expressed by the program. On the other hand, if
 the tasks access $r$ with {\em atomic} coherence, then Legion
-guarantees only atomicity of the tasks with respect to $r$: either $x$
+guarantees only atomicity of the tasks with respect to $r$: either task $x$
 appears to run entirely before $y$ or vice versa.
@@
 has one or more {\em physical instances} when the program executes.  Often
-logical regions correspond to a single physical instance, but there are common situations
-in which it is desirable to have multiple physical copies of a logical region's
-contents in different parts of the machine (e.g., to duplicate and distribute read-only
-data) or even to have physical instances of a logical region that temporarily diverge
-in their contents (e.g., to keep an old copy of a region to break a write-after-read dependency, or to 
-allow independent, local reductions that are later combined).
+logical regions correspond to a single physical instance, but it is also common and useful
+to have multiple physical copies of a logical region's contents in different parts of the machine 
+to duplicate and distribute read-only data, to keep an old copy of a region to break a 
+write-after-read dependency, or to allow independent, local reductions that are later combined.
 
@@
   parallelism (because tasks may recursively spawn subtasks) as well
-  as concurrent and distributed across the machine
-  (Section~\ref{sec:soop}).
+  as concurrent and distributed across the machine (Section~\ref{sec:soop}).
+
+\item We describe a {\em deferred execution} model that extracts parallelism
+even from highly irregular, unbalanced computations.  Instead of blocking,
+an operation $o$ that cannot immediately execute because it is dependent
+on some other operation $o'$ waits on an event that occurs when $o'$ completes.
+Events can be created for the completion of any operation in Legion, and 
+the scheduler can build long chains of deferred operations.  In Section~\ref{sec:exp}
+we show that dataflow-like deferred execution gives us a significant performance improvement
+over builk synchronous global synchronization primitives such as barriers.
 

### 2012-04-29 doc/sc2012/circuit.tex
more mods

@@
 Circuit} to be simulated.  This function specifies the regions it 
-accesses and the {\em privileges} and {\em coherence} it requires of those regions.
+accesses and the privileges and coherence it requires of those regions.
 In this case, the {\tt RWE} annotation specifies that the function
@@
 be referenced by multiple tasks in a phase.  The thick lines in Figure~\ref{sfig:part_fig:pvs}
-show how the nodes in a small graph might be partitioned into three subsets.  Gray
-nodes are shared and black nodes are private.  Note that
+show how the nodes in a small graph might be partitioned into three subsets.  Orange
+nodes are shared and blue nodes are private.  Note that
 this partition is disjoint since each node has one color.
@@
 We next further partition the private and shared subregions into {\tt
-  MAX\_PIECES} disjoint subregions each (lines 20-23); one subregion
-of the private and shared node partitions is shown in
+  MAX\_PIECES} disjoint subregions each (lines 20-23); the private and shared node subregions
+for the upper-left patch of the graph is shown in
 Figures~\ref{sfig:part_fig:p_i} and \ref{sfig:part_fig:s_i}
@@
 more than one color by the coloring {\tt node\_neighbor\_map}.
-One of the ghost subregions is shown in Figure~\ref{sfig:part_fig:g_i}.
+The ghost subregion for the upper left patch of the graph
+ is shown in Figure~\ref{sfig:part_fig:g_i}.
 
@@
 As is standard in region-based systems, allocating a {\tt Node}, {\tt
-  Wire}, or any value in the heap requires naming the logical region
+  Wire}, or any value in the heap requires naming the region
 in which the value is allocated.  For example, {\tt new Node@r} returns
@@
 Listing~\ref{lst:code_ex} illustrates one way of constructing partitioned
-data structures in Legion: populate a region with data (using {\tt new}) and then partition it
-into subregions.  Legion also supports partitioning an empty region into subregions
-and then allocating data in the subregions; we use both approaches in our
-application discussed in Section~\ref{sec:ex}.
+data structures in Legion: populate a region with data (using {\tt new}, note the allocation
+of the data is omitted from the example) and then partition it
+into subregions.  One can also first partition an empty region into subregions
+and then allocate data in the subregions. We use both approaches in our
+applications discussed in Section~\ref{sec:exp}.
 
@@
 data structures.  For example, a {\tt Circuit} (defined on line 3)
-holds the regions hold all nodes and wires of the circuit and a {\tt
+holds the regions for all nodes and wires of the circuit and a {\tt
   CircuitPiece} (defined on line 4) hold pointers to the private,
-shared, and ghost nodes as well as the private wires of its piece of
+shared, and ghost node regions as well as the private wires regions of its piece of
 the circuit.

### 2012-04-29 doc/sc2012/legion.tex
more mods

@@
 \usepackage{cancel}
+\usepackage{graphics}
 
@@
 the penalty for mistakes in the placement of data or computation is usually very poor performance.  Thus,
-to achieve high performance it is crucial that the programmer
+to achieve good performance it is crucial that the programmer
 and the compiler be able to reason about {\em locality} (data resident
 close to computation that uses it) and {\em independence} (computations
-operating on disjoint data, and therefore not requiring communication and able to be placed in possibly distant parts of the machine).  Most contemporary
+operating on disjoint data, and therefore not requiring communication and able to be placed in possibly distant parts of the machine).  
+Most contemporary
 programming systems have no facilities for the programmer to express locality
@@
 others) and {\em coherence} (e.g., {\em exclusive access} and {\em
-  atomic access}, among others).  Priveleges express how a task may use
+  atomic access}, among others).  Privileges express how a task may use
 its region arguments, providing data dependence information that
@@
 Coherence properties express the required semantics of concurrent
-accesses to the region.  For example, if tasks $x$ and $y$ access
-region $r$ with {\em exclusive} access, then Legion guarantees that the effects of $x$ and
-$y$ on $r$ are the same as if they had executed in
-the sequential order expressed by the program. On the other hand, if
+accesses to the region.  For example, if the program executes $f_1(r); f_2(r)$ and tasks $f_1$
+and $f_2$ both require {\em exclusive} access to region $r$, then Legion guarantees that 
+$f_1(r)$ completes before $f_2(r)$ begins.   On the other hand, if
 the tasks access $r$ with {\em atomic} coherence, then Legion
-guarantees only atomicity of the tasks with respect to $r$: either task $x$
-appears to run entirely before $y$ or vice versa.
+guarantees only atomicity of the tasks with respect to $r$: either task $f_1(r)$
+appears to run entirely before $f_2(r)$ or vice versa.
 
@@
   out-of-order hardware schedulers process instructions with register
-  arguments.  In addition to pipelining the execution of tasks over
-  several stages, our SOOP is hierarchical and naturally extracts nested
-  parallelism (because tasks may recursively spawn subtasks) as well
-  as concurrent and distributed across the machine (Section~\ref{sec:soop}).
+  arguments (Section~\ref{sec:soop}).  In addition to pipelining the execution of tasks over
+  several stages, our SOOP is concurrent and distributed across the machine and is
+  also hierarchical and naturally extracts nested
+  parallelism (because tasks may recursively spawn subtasks).
 
-\item We describe a {\em deferred execution} model that extracts parallelism
+\item We use a {\em deferred execution} model that extracts parallelism
 even from highly irregular, unbalanced computations.  Instead of blocking,
@@
 we show that dataflow-like deferred execution gives us a significant performance improvement
-over builk synchronous global synchronization primitives such as barriers.
+over bulk synchronous global synchronization primitives such as barriers.
 
-\item A crucial part of orchestrating a program's execution is how
+\item A central part of orchestrating a program's execution is how
 tasks are assigned (or {\em mapped}) to processors and how physical instances of logical regions are
-mapped to specific memory units within the machine.  Often there is
+mapped to specific memory units within the machine  (Section~\ref{sec:mapping}).  Often there is
 application-specific information that, if exploited, results in better
@@
 any user-supplied mapping strategy can only affect the performance of applications,
-not their correctness.  Thus, user mappings only affect the time and resources required
-to run the program and cannot change the program's output (Section~\ref{sec:mapping}).
+not their correctness.  
+%Thus, user mappings only affect the time and resources required
+%to run the program and cannot change the program's output.
 
@@
   fluid-flow on a regular grid, a three-level AMR code solving a
-  standard heat diffusion equation, a circuit simulation, and LINPACK.
+  standard heat diffusion equation, a circuit simulation, and LINPACK (Section~\ref{sec:exp}).
   We run each of these four codes and best available reference
@@
   designed for visualization applications (10 nodes with 2 6-core CPUs and 5 GPUs per node), and up to a third of the
-  Keeneland supercomputer (32 nodes, 2 6-core CPUs and 3 GPUs per node, see Section~\ref{sec:exp}).
+  Keeneland supercomputer (32 nodes, 2 6-core CPUs and 3 GPUs per node)
 

### 2012-04-29 doc/sc2012/related.tex
new files
more mods

--- 2011-11-07 doc/pldi2012/related.tex
@@
 \section{Related Work}
+\label{sec:related}
 
-Chapel \cite{Chamberlain:Chapel} is the closest related work to Legion.  
-Chapel has several concepts to support the expression of locality.  
+Legion is an outgrowth of previous work on Sequoia, a locality-aware programming language capable
+of expressing computations in deep memory hierarchies
+\cite{Fatahalian06}.  Sequoia is a special case of the Legion
+programming model in which only arrays can be recursively partitioned,
+all access is exclusiv, there is a static mapping of tasks and data on to the machine (though
+extensions to Sequoia make this mapping more dynamic \cite{BCSA11}) and, most fundamentally, the decomposition of tasks and the decomposition
+of data is one-to-one.  Legion generalizes the Sequoia model by allowing for
+partitioning of pointer data structures through regions, enabling dynamic mappings through
+the mapper interface, and allowing different coherence
+properties.  Legion's decoupling of the task tree from the region tree in Legion leads directly to the scheduling problem solved by our software out-of-order processor for tasks with region arguments.
+
+Chapel has several concepts to support the expression of locality \cite{Chamberlain:Chapel}.
 Domains are similar to logical regions in that they describe maps from indexes
@@
 
-Sequoia is an example of a locality-aware programming language capable
-of expressing computations in deep memory hierarchies
-\cite{Fatahalian06}.  Sequoia is a special case of the Legion
-programming model in which only arrays can be recursively partitioned,
-all access is exclusive and the task tree and tree of array
-decompositions are identical.  Sequoia requires the programmer to
-statically map arrays and task trees using a machine specific mapping
-file.  Legion generalizes the Sequoia model by allowing for
-partitioning of pointer data structures through regions, decoupling
-the region tree from the task tree, enabling dynamic mappings through
-the mapper interface, and allowing different coherence
-properties.
-
 X10 is another parallel programming environment designed to operate on distributed
@@
 through both the cluster and GPU memory hierarchies.
+
+Deterministic Parallel Java (DPJ) is a parallel extension of Java
+that, like Legion, uses regions to express locality and does
+dependence analysis at the granularity of conflicts on region
+arguments to functions \cite{Bocchino09}.  As the name suggests, one
+of the primary goals of DPJ is to provide a programming model that
+guarantees determinism while also providing supporting parallelism.
+As a result the programming model is more restrictive than Legion, as
+there is no notion of weaker coherence properties of regions than
+exclusive access (which is essential to determinism).  DPJ can express
+the populate-then-partition style of using regions, but not the
+partition-then-populate; overall, the language is more static, and
+thus provides more guarantees and less flexibility than Legion.  Most
+significantly, however, DPJ efforts have so far focused on JVM
+implementations on on shared memory machines; there is nothing
+corresponding to our distributed software out-of-order processor or
+the mapping interface.
 

### 2012-04-29 runtime/legion.cc
Doing some common subexpression elimination for the high level runtime and replacing the low level locks with pthread locks

@@ -70,17 +70,31 @@ namespace RegionRuntime {
     // the object goes out of scope
     class AutoLock {
     private:
-      Lock lock;
+      const bool is_low;
+      Lock low_lock;
+      ImmovableLock immov_lock;
     public:
       AutoLock(Lock l, unsigned mode = 0, bool exclusive = true, Event wait_on = Event::NO_EVENT)
-        : lock(l)
+        : is_low(true), low_lock(l)
       {
         Event lock_event = l.lock(mode,exclusive,wait_on);
         lock_event.wait(true/*block*/);
       }
-      ~AutoLock()
+      AutoLock(ImmovableLock l)
+        : is_low(false), immov_lock(l)
       {
-        lock.unlock();
+        l.lock();
+      }
+      ~AutoLock(void)
+      {
+        if (is_low)
+        {
+          low_lock.unlock();
+        }
+        else
+        {
+          immov_lock.unlock();
+        }
       }
     };
 
@@ -1447,8 +1461,12 @@ namespace RegionRuntime {
       {
         if (need_lock)
         {
+#ifdef LOW_LEVEL_LOCKS
           Event lock_event = context_lock.lock(0,true/*exclusive*/);
           lock_event.wait(true/*block*/);
+#else
+          context_lock.lock();
+#endif
         }
         // Mark that the region has been unmapped
         unmapped_event.trigger();
@@ -1946,7 +1964,11 @@ namespace RegionRuntime {
       handle = LogicalRegion::NO_REGION;
       parent_ctx = NULL;
       active = false;
+#ifdef LOW_LEVEL_LOCKS
       current_lock = Lock::NO_LOCK;
+#else
+      current_lock.clear();
+#endif
       current_gen++;
       runtime->free_deletion(this);
     }
@@ -1957,8 +1979,12 @@ namespace RegionRuntime {
     {
       if (acquire_lock)
       {
+#ifdef LOW_LEVEL_LOCKS
         Event lock_event = current_lock.lock(0,true/*exclusive*/);
         lock_event.wait(true/*block*/);
+#else
+        current_lock.lock();
+#endif
       }
 #ifdef DEBUG_HIGH_LEVEL
       assert(active);
@@ -2139,7 +2165,11 @@ namespace RegionRuntime {
     HighLevelRuntime::HighLevelRuntime(LowLevel::Machine *m, Processor local)
       : local_proc(local), proc_kind (m->get_processor_kind(local)), machine(m),
[…]

### 2012-04-29 runtime/legion.h

@@ -940,18 +940,31 @@ namespace RegionRuntime {
       const Processor::Kind proc_kind;
       Machine *const machine;
       std::vector<Mapper*> mapper_objects;
+#ifdef LOW_LEVEL_LOCKS
       std::vector<Lock> mapper_locks;
-      Lock mapping_lock; // Protect mapping data structures
+      Lock mapping_lock;
+#else
+      std::vector<ImmovableLock> mapper_locks;
+      ImmovableLock mapping_lock; // Protect mapping data structures
+#endif
       // Colorize Functions
       std::vector<ColorizeFnptr> colorize_functions;
       // Task Contexts
       bool idle_task_enabled; // Keep track if the idle task enabled or not
       std::vector<TaskContext*> mapping_queue; // tasks that need to be mapped (IN ORDER!)
       std::list<TaskContext*> ready_queue; // Tasks ready to be mapped/stolen
-      Lock queue_lock; // Protect ready and waiting queues and idle_task_enabled
+#ifdef LOW_LEVEL_LOCKS
+      Lock queue_lock;
+#else
+      ImmovableLock queue_lock; // Protect ready and waiting queues and idle_task_enabled
+#endif
       unsigned total_contexts;
       std::list<TaskContext*> available_contexts; // open task descriptions
-      Lock available_lock; // Protect available contexts
+#ifdef LOW_LEVEL_LOCKS
+      Lock available_lock;
+#else
+      ImmovableLock available_lock; // Protect available contexts
+#endif
       // Region Mappings 
       std::vector<RegionMappingImpl*> ready_maps;
       std::list<RegionMappingImpl*> available_maps;
@@ -959,7 +972,11 @@ namespace RegionRuntime {
       std::vector<DeletionOp*> ready_deletions;
       std::list<DeletionOp*> available_deletions;
       // Keep track of how to do partition numbering
-      Lock unique_lock; // Make sure all unique values are actually unique
+#ifdef LOW_LEVEL_LOCKS
+      Lock unique_lock;
+#else
+      ImmovableLock unique_lock; // Make sure all unique values are actually unique
+#endif
       PartitionID next_partition_id; // The next partition id for this instance (unique)
       UniqueID next_task_id; // Give all tasks a unique id for debugging purposes
       InstanceID next_instance_id;
@@ -967,9 +984,14 @@ namespace RegionRuntime {
       // Information for stealing
       const unsigned int max_outstanding_steals;
[…]

### 2012-04-29 doc/sc2012/related.tex

@@
 
-Legion is an outgrowth of previous work on Sequoia, a locality-aware programming language capable
+Legion began as an outgrowth of Sequoia, a locality-aware programming language capable
 of expressing computations in deep memory hierarchies
@@
 programming model in which only arrays can be recursively partitioned,
-all access is exclusiv, there is a static mapping of tasks and data on to the machine (though
+all access is exclusive, there is a static mapping of tasks and data on to the machine (though
 extensions to Sequoia make this mapping more dynamic \cite{BCSA11}) and, most fundamentally, the decomposition of tasks and the decomposition
@@
 of the primary goals of DPJ is to provide a programming model that
-guarantees determinism while also providing supporting parallelism.
+guarantees determinism while also supporting parallelism.
 As a result the programming model is more restrictive than Legion, as
@@
 partition-then-populate; overall, the language is more static, and
-thus provides more guarantees and less flexibility than Legion.  Most
-significantly, however, DPJ efforts have so far focused on JVM
-implementations on on shared memory machines; there is nothing
-corresponding to our distributed software out-of-order processor or
-the mapping interface.
+thus provides more guarantees and less flexibility than Legion.  
+DPJ efforts have so far focused on JVM implementations on shared memory machines.
+There have also been several other recent efforts that propose task schedulers
+based on dynamic detection of memory conflicts, but these also have assumed
+underlying shared memory hardware \cite{VTN11,Tzenakis12,EYJD12}.
 
@@
 In previous programming systems regions have primarily been used as a construct for
-describing memory management schemes \cite{REAPS02}\cite{RC01}  
-or for enforcing safety policies \cite{CYCLONE01}.  In all cases, regions had
+describing memory management schemes \cite{REAPS02,RC01}  
+or for enforcing safety policies \cite{CYCLONE01}.  We follow \cite{RC01} in Legion's
+decision to make regions first class.  In these works, however, regions have
 memory layout implications.  Logical regions in Legion enable the programmer to
-describe locality and independence without any memory layout implications.
+describe locality independent of memory layout.
 

### 2012-04-29 runtime/legion.cc
Fixing some typos

@@ -2334,7 +2334,9 @@ namespace RegionRuntime {
           delete mapper_objects[i];
           mapper_objects[i] = NULL;
 #ifdef DEBUG_HIGH_LEVEL
+#ifdef LOW_LEVEL_LOCKS
           assert(mapper_locks[i].exists());
+#endif
 #endif
 #ifdef LOW_LEVEL_LOCKS
           mapper_locks[i].destroy_lock();
@@ -4497,7 +4499,9 @@ namespace RegionRuntime {
       sanitized = false;
 #ifdef DEBUG_HIGH_LEVEL
       assert(_mapper != NULL);
+#ifdef LOW_LEVEL_LOCKS
       assert(_map_lock.exists());
+#endif
 #endif
       mapper = _mapper;
       mapper_lock = _map_lock;

### 2012-04-29 runtime/legion.h

@@ -987,7 +987,7 @@ namespace RegionRuntime {
       std::multimap<MapperID,Processor> failed_thiefs;
 #ifdef LOW_LEVEL_LOCKS
       Lock stealing_lock;
-      Lock theiving_lock;
+      Lock thieving_lock;
 #else
       ImmovableLock stealing_lock;
       ImmovableLock thieving_lock;

### 2012-04-29 doc/sc2012/soop.tex
more mods
soop

@@
 \label{sec:soop}
+
+There are two major challenges in implementing an efficient task
+scheduler for Legion:
+\begin{itemize}
+\item  For correctness, Legion must guarantee that every pair of {\em dependent} tasks is executed 
+in sequential order.
+
+\item Legion must deal with and, to
+  the extent possible, hide the extremely long latencies associated
+  with machines that have both distributed memory and many levels of
+  memory hierarchy.
+\end{itemize}
+We use three techniques to deal with these problems:
+\begin{itemize}
+\item Nested tasks (subtasks) enjoy an isolation property that allows Legion
+to avoid comparing all pairs of tasks for dependences.
+Specifically, the only tasks that need to be checked for dependences 
+are the immediate subtasks of a common
+parent task (see Section~\ref{sec:exec}).
+
+\item Legion uses a {\em deferred execution model} that decouples the issuing
+of operations from when operations are performed.  Issued operations wait for other operations on
+which they are dependent to complete before executing.  For example, Legion may issue a copy
+operation to move the results of a task $a$ to the place where a task $b$ will take the copied data as
+an argument.  Tasks $a$ and $b$ and the copy can all be issued, but the copy will not start until
+task $a$ completes and task $b$ will not start until the copy completes.
+
+\item Legion uses a {\em  software out-of-order processor}, or SOOP, to schedule tasks.  The SOOP 
+is distributed and concurrent, and also extracts nested parallelism from subtasks.
+\end{itemize}
+Legion's SOOP has a five stage pipeline that  decouples the management of resources from actual
+task execution.  The five stages are: dependence analysis (Section~\ref{sec:dep}),
+distribution (Section~\ref{sec:dist}),
+mapping (Section~\ref{sec:map}),
+execution (Section~\ref{sec:exec}),
+and clean-up (Section~\ref{sec:clean}).
+We discuss each in turn.
+
+\input{mapping_fig}
+
+\subsection{Depedence Analysis}
+\label{sec:dep}
+{\small
+\begin{tabular}{c|cccc}
+             & Exclusive & Atomic   & Simultaneous & Relaxed \\
+\midrule
+Exclusive    & Dep & Dep & Dep & Dep \\ 
+Atomic       & Dep & Same & Cont & Cont \\
+Simultaneous & Dep & Cont & Same & None \\
+Relaxed      & Dep & Cont & None & None \\
+\end{tabular}
+}
+
+\subsection{Distribution}
+\label{sec:dist}
+
+\subsection{Mapping}
+\label{sec:map}
+
+\subsection{Execution}
+\label{sec:exec}
+
+\begin{figure}
+\includegraphics[scale=0.48]{figs/CircuitMem.pdf}
+\caption{Tasks and data for the circuit simulation on a cluster of GPUs.}
+\end{figure}
+
+\subsection{Clean-Up}
+\label{sec:clean}

### 2012-04-30 doc/sc2012/circuit.tex

@@
 and the required inter-pass dependencies are determined automatically
-by the Legion runtime based on the region access annotations on the task declarations
-(see Section~\ref{sec:soop}).
+by the Legion runtime based on the region access annotations on the task declarations.
+The  tasks spawned on lines 32-34 are {\em subtasks} of the main
+{\tt simulate\_circuit} task.
+The region arguments to a subtask  must be either regions of, or subregions of, the arguments to its parent task with a subset of the permissions and the same coherence properties.
+This restriction plays a central role in our task scheduling algorithm (see Section~\ref{sec:soop}).
 

### 2012-04-30 apps/linpack/linpack.cc
changes

@@ -6,6 +6,7 @@
 #include <cstdarg>
 #include <mkl.h>
 #include <mkl_cblas.h>
+#include <sched.h>
 
 #include "legion.h"
 
@@ -48,6 +49,8 @@ enum {
   LMAP_PICK_GPU_BASE = 0x200000,
   LMAP_PICK_GPU_END  = 0x2FFFFF,
 
+  LMAP_PICK_SCHED    = 0x500000,
+
   LMAP_IDXCOL_MODE   = 0x010000,
   LMAP_IDXROW_MODE   = 0x020000,
 
@@ -56,6 +59,7 @@ enum {
   LMAP_PLACE_ZCMEM,
   LMAP_PLACE_FBMEM,
   LMAP_PLACE_NONE,
+  LMAP_INLINE        = 0x600000,
 
   LMAP_PLACE_ATSYS_BASE = 0x400000,
   LMAP_PLACE_ATSYS_END  = 0x4FFFFF,
@@ -82,11 +86,18 @@ public:
 
   std::map<Processor, CPUProcInfo> cpu_procs;
   std::map<Processor, GPUProcInfo> gpu_procs;
-  std::vector<Processor> cpu_list, gpu_list;
+  std::vector<Processor> cpu_list, gpu_list, sched_list;
+  int sched_local_id;
+
+  bool is_sched_processor(Processor p)
+  {
+    return((((int)p.id) & 0xffff) == sched_local_id);
+  }
 
   LinpackMapper(Machine *m, HighLevelRuntime *r, Processor p)
     : Mapper(m, r, p)
   {
+    sched_local_id = -1;
     const std::set<Processor> &all_procs = m->get_all_processors(); 
     for (std::set<Processor>::const_iterator it = all_procs.begin();
           it != all_procs.end(); it++)
@@ -94,7 +105,16 @@ public:
       Processor::Kind k = m->get_processor_kind(*it);
       if (k == Processor::LOC_PROC)
       {
-	cpu_list.push_back(*it);
+	// terrible hack - take the first processor we see from each node
+	//  and use it as a scheduling processor
+	if(sched_local_id == -1)
+	  sched_local_id = it->id & 0xffff;
+	if(is_sched_processor(*it)) {
+	  log_mapper.info("processor %x hijacked for scheduling", it->id);
+	  sched_list.push_back(*it);
+	} else {
+	  cpu_list.push_back(*it);
+	}
[…]

### 2012-04-30 runtime/lowlevel.cc

@@ -3359,10 +3359,10 @@ namespace RegionRuntime {
 	virtual void sleep_on_event(Event wait_for, bool block = false)
 	{
 	  // create an entry to go on our event stack (on our stack)
-	  EventStackEntry entry(this, wait_for, event_stack);
-	  event_stack = &entry;
+	  EventStackEntry *entry = new EventStackEntry(this, wait_for, event_stack);
+	  event_stack = entry;
 
-	  entry.add_waiter();
+	  entry->add_waiter();
 
 	  // now take the processor lock that controls everything and see
 	  //  what we can do while we're waiting
@@ -3375,7 +3375,7 @@ namespace RegionRuntime {
 	    assert(state == STATE_RUN);
 
 	    // loop until our event has triggered
-	    while(!entry.triggered) {
+	    while(!entry->triggered) {
 	      // first step - if some other thread is resumable, give up
 	      //   our spot and let him run
 	      if(proc->resumable_threads.size() > 0) {
@@ -3461,8 +3461,9 @@ namespace RegionRuntime {
 			  this, proc->me.id, wait_for.id, wait_for.gen);
 	  }
 
-	  assert(event_stack == &entry);
-	  event_stack = entry.next;
+	  assert(event_stack == entry);
+	  event_stack = entry->next;
+	  delete entry;
 	}
 
 	virtual void thread_main(void)
@@ -3999,7 +4000,7 @@ namespace RegionRuntime {
 	while(!wait_for.has_triggered()) {
 	  log_util.info("utility thread polling on event %x/%d",
 			wait_for.id, wait_for.gen);
-	  usleep(1000);
+	  //usleep(1000);
 	}
       }
 
@@ -5091,12 +5092,33 @@ namespace RegionRuntime {
       log_copy.debug("sending remote write request: mem=%x, offset=%zd, size=%zd, event=%x/%d",
 		     mem.id, offset, datalen,
 		     event.id, event.gen);
-      RemoteWriteArgs args;
-      args.mem = mem;
-      args.offset = offset;
-      args.event = event;
-      RemoteWriteMessage::request(ID(mem).node(), args,
-				  data, datalen, PAYLOAD_KEEP);
+      const size_t MAX_SEND_SIZE = 4 << 20; // should be <= LMB_SIZE
+      if(datalen > MAX_SEND_SIZE) {
[…]

### 2012-04-30 runtime/lowlevel.h

@@ -430,7 +430,7 @@ namespace RegionRuntime {
 	}
 
       template <class T>
-      void write(ptr_t<T> ptr, T newval) const
+      void write(ptr_t<T> ptr, const T& newval) const
 	{
 	  assert(!is_reduction_only());
 	  put_untyped(ptr.value*sizeof(T), &newval, sizeof(T));
@@ -492,7 +492,7 @@ namespace RegionRuntime {
       T read(ptr_t<T> ptr) const { return ((T*)array_base)[ptr.value]; }
 
       template <class T>
-      void write(ptr_t<T> ptr, T newval) const { ((T*)array_base)[ptr.value] = newval; }
+      void write(ptr_t<T> ptr, const T& newval) const { ((T*)array_base)[ptr.value] = newval; }
 
       template <class REDOP, class T, class RHS>
       void reduce(ptr_t<T> ptr, RHS newval) const { REDOP::apply<false>(((T*)array_base)[ptr.value], newval); }
@@ -558,7 +558,7 @@ namespace RegionRuntime {
 
       template <class T>
       __device__ __forceinline__
-      void write(ptr_t<T> ptr, T newval) const { 
+      void write(ptr_t<T> ptr, const T& newval) const { 
 #ifdef DEBUG_LOW_LEVEL
         bounds_check(ptr);
 #endif
@@ -654,7 +654,7 @@ namespace RegionRuntime {
       RegionInstanceAccessorUntyped<AT> ria;
 
       ET read(ptr_t<ET> ptr) const { return ria.read(ptr); }
-      void write(ptr_t<ET> ptr, ET newval) const { ria.write(ptr, newval); }
+      void write(ptr_t<ET> ptr, const ET& newval) const { ria.write(ptr, newval); }
       ET& ref(ptr_t<ET> ptr) const { return ria.ref(ptr); }
 
       void read_partial(ptr_t<ET> ptr, off_t offset, void *dst, size_t size) const
@@ -683,7 +683,7 @@ namespace RegionRuntime {
       RegionInstanceAccessorUntyped<AccessorGPU> ria;
 
       __device__ ET read(ptr_t<ET> ptr) const { return ria.read(ptr); }
-      __device__ void write(ptr_t<ET> ptr, ET newval) const { ria.write(ptr, newval); }
+      __device__ void write(ptr_t<ET> ptr, const ET& newval) const { ria.write(ptr, newval); }
 
       //template <class REDOP, class RHS>
[…]

### 2012-04-30 doc/sc2012/circuit.tex

@@
 
-Lines 31-57 form the bulk of the actual simulation, performing three passes
-over the circuit in each iteration.  Each pass loops over an array of pieces (constructed on lines 27-30 from the partitions),
+Lines 31-57 form the bulk of the actual simulation, performing three
+passes over the circuit in each iteration.  Each pass loops over an
+array of pieces (constructed on lines 27-30 from the partitions),
 spawning a task for each piece.  There are no explicit requests for
 parallel execution (the default semantics of {\tt spawn} is sequential
-execution of the spawned tasks) nor is there synchronization required between the
-passes.  Both the fact that the tasks can be run in parallel within a pass
-and the required inter-pass dependencies are determined automatically
-by the Legion runtime based on the region access annotations on the task declarations.
-The  tasks spawned on lines 32-34 are {\em subtasks} of the main
-{\tt simulate\_circuit} task.
-The region arguments to a subtask  must be either regions of, or subregions of, the arguments to its parent task with a subset of the permissions and the same coherence properties.
-This restriction plays a central role in our task scheduling algorithm (see Section~\ref{sec:soop}).
+execution of the spawned tasks) nor is there synchronization required
+between the passes.  Both the fact that the tasks can be run in
+parallel within a pass and the required inter-pass dependencies are
+determined automatically by the Legion runtime based on the region
+access annotations on the task declarations.  The tasks spawned on
+lines 32-34 are {\em subtasks} of the main {\tt simulate\_circuit}
+task. A subtask can only take region arguments that are either also region arguments of or subregions
+of region arguments of the parent task.  This requirement
+plays an important role in our task scheduling algorithm (see
+Section~\ref{sec:soop}).
 

### 2012-04-30 runtime/legion.h
Fixing a bug in premapping and adding some operators for the high level interface

@@ -354,6 +354,8 @@ namespace RegionRuntime {
       TaskArgument(void) : args(NULL), arglen(0) { }
       TaskArgument(const void *arg, size_t argsize)
         : args(const_cast<void*>(arg)), arglen(argsize) { }
+      TaskArgument(const TaskArgument &rhs)
+        : args(rhs.args), arglen(rhs.arglen) { }
     public:
       inline size_t get_size(void) const { return arglen; }
       inline void*  get_ptr(void) const { return args; }
@@ -362,6 +364,12 @@ namespace RegionRuntime {
         { return args == arg.args; }
       bool operator<(const TaskArgument &arg) const
         { return args < arg.args; }
+      TaskArgument& operator=(const TaskArgument &rhs)
+      {
+        args = rhs.args;
+        arglen = rhs.arglen;
+        return *this;
+      }
     private:
       void *args;
       size_t arglen;

### 2012-04-30 doc/sc2012/soop.tex

@@
 \begin{itemize}
-\item  For correctness, Legion must guarantee that every pair of {\em dependent} tasks is executed 
-in sequential order.
+\item  For correctness, Legion must guarantee that every pair of {\em dependent} tasks is serialized in the correct order.
 
-\item Legion must deal with and, to
-  the extent possible, hide the extremely long latencies associated
+\item For performance, Legion must hide the extremely long latencies associated
   with machines that have both distributed memory and many levels of
@@
 \end{itemize}
-We use three techniques to deal with these problems:
+Our scheduler uses three techniques to achieve both correctness and high performance:
 \begin{itemize}
-\item Nested tasks (subtasks) enjoy an isolation property that allows Legion
+\item Subtasks enjoy an isolation property that allows Legion
 to avoid comparing all pairs of tasks for dependences.
@@
 \end{itemize}
-Legion's SOOP has a five stage pipeline that  decouples the management of resources from actual
+
+During execution of a Legion program, each processor runs its own instance of the Legion SOOP.
+A processor's SOOP is the Legion runtime system for that processor and handles all requests
+for Legion services from tasks executing on the processor.
+Legion's SOOP has a five stage pipeline that  separates the management of system resources from actual
 task execution.  The five stages are: dependence analysis (Section~\ref{sec:dep}),
@@
 \label{sec:dep}
+
+When a Legion task runs, it may spawn new subtasks to be executed.
+Each subtask is {\em registered} with the local SOOP in sequential execution
+order, which records the subtask's logical region arguments and associated privileges and coherence properties.  If two subtasks work on indepedent data Legion can schedule
+them to out of order, in parallel.  Thus, the first stage in the
+execution of a task is dependence analysis between tasks.
+
+Conceptually, detecting dependences between a newly registered task $t$ and a previous task 
+$t'$ requires comparing the two sets of logical regions accessed.  For each logical region used by
+$t'$ that may overlap with a logical region used by $t$, the privileges
+and coherence modes are compared to determine whether a dependence exists.  If
+both regions need only read privileges there is never a dependence, but if either
+task needs write or reduction privileges, the coherence modes are compared:
 {\small
@@
 }
+
+An {\tt Dep} entry indicates dependence while {\tt None}
+indicates independence.  A {\tt Same} entry is a dependence unless the two tasks
+use the same {\em physical region}.  Finally, a {\tt Cont} entry
+indicates a dependence unless a single writer is using the
+atomic coherence mode.  When a dependence is detected, the new task
+waits for the older task to complete before it begins execution.
+
+The table lists {\em simultaneous} and {\em relaxed} coherence modes
+that we have not yet discussed.  Both simultaneous and relaxed
+allow other tasks using the region to execute at the same time and differ
+only in what data must be observed.  With simultaneous coherence, a task must 
+see all updates to the logical region made by other tasks operating on the same region 
+simultaneously (i.e., shared memory semantics).  With relaxed coherence, 
+a task may or may not observe updates to the logical region by other tasks executing at
+the same time.
+
+% *****  isolation, efficient data structure ******
+
+To detect dependencies efficiently, the high-level runtime maintains a
+local region forest (recall Section~\ref{sec:exec}) for each task.  The
+root regions are the task's region arguments, and the forest includes
+all partitions and subregions created within the task.  To perform
+dependence detection on a region $r$, the runtime first finds a root
+region $r'$ such that $r \rleq r'$.  The runtime then walks the region
+forest from $r'$ to $r$.  Each region $r''$ on the path contains a
+list $L$ of registered tasks using $r''$, and potential dependencies
+are discovered by comparing privileges and coherence modes with the
+tasks in $L$.
+
+The runtime next checks if a partition of $r''$ is {\em open}.  An
+open partition indicates that there are active tasks using subregions
+of $r''$.  No more than one partition of a node can be open
+at any time.  If the open partition is is along the path to $r$,
+traversal continues.  If the open partition lies along another
+path the runtime must {\em close} the currently active partition because it cannot guarantee
+disjointness with the region the new task wishes to access.  To close a
+partition, the runtime finds all active tasks in that
+subtree and considers them all dependencies for the new task.
+With the previously active partition closed, the runtime can
+then open the desired partition and traverse further down the tree as
+needed.
+
+When the runtime traverses an open partition, its behavior depends on the kind
+of partition being traversed.  If the partition is disjoint, no other subregion
+can overlap with the target region, so the state of other subregions is
+ignored.  However, if the partition is not disjoint, any
+of subregion could overlap with the desired subregion, so
+the runtime must close all open subregions (recording new dependencies
+as it does so) before continuing the traversal.
+
+After the runtime reaches the target logical region, it performs dependence
+checks against any active tasks listed for that region.  If a dependence is
+discovered, it is added to the current task's dependence list.  It can then
+be removed from the region tree's active task list, as the current task (which
+will be pushed onto the active task list) {\em dominates} the previous task.
+In a case where the new task has dependencies with all existing active tasks
+for the target node, that whole list is replaced by a single entry for the
+new task.
 

### 2012-05-01 apps/circuit/plot_circuit.py
plotting script for the circuit experiments

@@ -1,0 +1,258 @@
+#!/usr/bin/python
+
+import subprocess
+import sys, os, shutil
+import string, re
+import numpy as np
+import matplotlib.pyplot as plt
+
+home_dir="/home/mebauer/region/apps/circuit/"
+out_dir="/home/mebauer/region/apps/circuit/figs/"
+baseline="baseline"
+expr_name="ckt_sim"
+problems=[48,96]
+#speedup_axis=[1,97,1,97]
+speedup_axis=[1,128,1,128]
+linear=[1,2,4,8,16,32,48,96]
+
+markers = ['o','+','x','s','D','*','v','^','p','1','2','3','4','<','>','d']
+
+class Machine(object):
+    def __init__(self,directory,name,gpu_set):
+        self.name = name
+        self.directory = directory
+        self.baselines = dict()
+        self.experiments = set()
+        self.gpu_set = gpu_set
+
+    def add_experiment(self,e):
+        assert e not in self.experiments
+        self.experiments.add(e)
+
+    def add_baseline(self,b):
+        assert b.pieces not in self.baselines
+        self.baselines[b.pieces] = b
+
+    def read_experiments(self):
+        for f in os.listdir(home_dir+self.directory):
+            m = expr_pat.match(f)
+            if m <> None:
+                e = Experiment(self.directory+'/'+f,
+                               int(m.group('pieces')),
+                               int(m.group('nodes')),
+                               int(m.group('cpus')),
+                               int(m.group('gpus')))
+                if e.success:
+                    self.add_experiment(e)
+                continue
+            m = base_pat.match(f)
+            if m <> None:
+                e = Experiment(self.directory+'/'+f,
+                               int(m.group('pieces')),
+                               int(m.group('nodes')),
+                               int(m.group('cpus')),
+                               int(m.group('gpus')))
+                self.add_baseline(e)
+                continue
+
+    def print_summary(self):
+        for e in sorted(self.experiments,key=lambda x: x.pieces+x.nodes*x.gpus):
+            assert e.pieces in self.baselines
+            speedup = self.baselines[e.pieces].time / e.time
+            print "Pieces: "+str(e.pieces)+" Nodes: "+str(e.nodes)+" GPUs: "+str(e.gpus)+ \
[…]

### 2012-05-01 doc/sc2012/soop.tex

@@
 \begin{itemize}
-\item  For correctness, Legion must guarantee that every pair of {\em dependent} tasks is serialized in the correct order.
+\item  For correctness, Legion must guarantee that every pair of {\em dependent} tasks is serialized in program order.
 
@@
 \begin{itemize}
-\item Subtasks enjoy an isolation property that allows Legion
-to avoid comparing all pairs of tasks for dependences.
-Specifically, the only tasks that need to be checked for dependences 
-are the immediate subtasks of a common
-parent task (see Section~\ref{sec:exec}).
+\item If it were necessary to compare all pairs of tasks for dependences that alone
+would be a serial bottleneck in all but the simplest Legion programs.  Fortunately,
+subtasks enjoy an isolation property that allows Legion to check only task siblings for dependences
+(see Section~\ref{sec:dep}).
+
+\item Legion uses a {\em  software out-of-order processor}, or SOOP, to schedule tasks.  The SOOP 
+is pipelined and distributed, and also extracts nested parallelism from subtasks.
 
 \item Legion uses a {\em deferred execution model} that decouples the issuing
-of operations from when operations are performed.  Issued operations wait for other operations on
-which they are dependent to complete before executing.  For example, Legion may issue a copy
-operation to move the results of a task $a$ to the place where a task $b$ will take the copied data as
-an argument.  Tasks $a$ and $b$ and the copy can all be issued, but the copy will not start until
-task $a$ completes and task $b$ will not start until the copy completes.
+of operations from when operations are performed.  An issued operation waits for other operations on
+which it is dependent to complete before executing, but a waiting operation does not cause the SOOP
+to block, allowing Legion to perform other useful work while the operation is pending.
 
-\item Legion uses a {\em  software out-of-order processor}, or SOOP, to schedule tasks.  The SOOP 
-is distributed and concurrent, and also extracts nested parallelism from subtasks.
+%  For example, Legion may issue a copy
+%operation to move the results of a task $a$ to the place where a task $b$ will take the copied data as
+%an argument.  Tasks $a$ and $b$ and the copy can all be issued, but the copy will not start until
+%task $a$ completes and task $b$ will not start until the copy completes.
+
 \end{itemize}
 
-During execution of a Legion program, each processor runs its own instance of the Legion SOOP.
-A processor's SOOP is the Legion runtime system for that processor and handles all requests
-for Legion services from tasks executing on the processor.
-Legion's SOOP has a five stage pipeline that  separates the management of system resources from actual
-task execution.  The five stages are: dependence analysis (Section~\ref{sec:dep}),
+Each processor in the machine runs an instance of the SOOP, which handles all requests
+for Legion services from tasks executing on the processor.  The SOOP is analagous to the out-of-order
+instruction scheduler in superscalar processors, but working at the granularity of tasks and their region
+arguments instead of instructions and their register arguments.
+Legion's SOOP has a five stage pipeline: dependence analysis (Section~\ref{sec:dep}),
 distribution (Section~\ref{sec:dist}),
@@
 
-When a Legion task runs, it may spawn new subtasks to be executed.
-Each subtask is {\em registered} with the local SOOP in sequential execution
-order, which records the subtask's logical region arguments and associated privileges and coherence properties.  If two subtasks work on indepedent data Legion can schedule
-them to out of order, in parallel.  Thus, the first stage in the
-execution of a task is dependence analysis between tasks.
+When a Legion task runs, it may spawn new subtasks to be executed.  We refer to the
+subtasks as {\em children} of the {\em parent} task.  The children of a parent task
+are {\em siblings} of each other.
 
+When a parent task spawns a child task, the child is {\em registered} with the SOOP 
+on which the parent is executing; registration records the subtask's logical region
+arguments and associated privileges and coherence properties.
+Children are registered in the sequential execution order in which the parent task
+spawns them.
+
+If two children have a dependence between their region arguments, the
+SOOP will schedule them to run serially to preserve the program's
+sequential semantics.  If two child tasks have no dependence then
+Legion can schedule them to execute out of order, in parallel.  Thus,
+the first stage in the execution of a task is dependence analysis
+between tasks.
+
+\subsubsection{Dependences}
 Conceptually, detecting dependences between a newly registered task $t$ and a previous task 
@@
 
-An {\tt Dep} entry indicates dependence while {\tt None}
+A {\tt Dep} entry indicates dependence while {\tt None}
 indicates independence.  A {\tt Same} entry is a dependence unless the two tasks
-use the same {\em physical region}.  Finally, a {\tt Cont} entry
+use the same physical instance of the logical region.\footnote{For example, if two tasks wish to acess the region with atomic coherence, then if they are working on the same physical copy of the data atomicity can be guaranteed using, e.g., locking.
+If they are working on different copies of the data then the only safe execution strategy for Legion is to serialize the two tasks.}
+A {\tt Cont} entry
 indicates a dependence unless a single writer is using the
@@
 
-% *****  isolation, efficient data structure ******
+\subsubsection{Isolation}
+The reader may wonder why we only compare sibling tasks for
+dependences: can't arbitrary pairs of tasks have dependences?  The
+answer is ``yes'', but the following lemma shows that
+any non-sibling tasks are always detected first as a dependence between
+siblings that are registered earlier in execution!
+\begin{lemma}
+\label{lem:isolation}
+\rm
+Let $t_1$ and $t_2$ be two distinct, non-sibling tasks other than the root task with parents $p_1$ and $p_2$, respectively.
+If $t_1$ and $t_2$ have a dependence, then $p_1$ and $p_2$ have a dependence.
+\end{lemma}
+\begin{proof}
+Recall from Section~\ref{sec:ex} that the parents' region arguments include the children's region arguments and have
+only more permissions.
+\end{proof}
+Lemma~\ref{lem:isolation} justifies checking dependences only between sibling tasks, because any dependences between 
+arbitrary tasks are detected first between children of the tasks' least common ancestor.  Legion
+guarantees that all dependence analysis is performed for all children of a task $t$ on the instance of the SOOP where $t$
+runs, which by Lemma~\ref{lem:isolation} is sufficient to detect all required dependences across all tasks.
+
+\subsubsection{Efficient Dependence Analysis}
 

### 2012-05-01 runtime/lowlevel.cc
faster (and hopefully still correct) mask iterator

@@ -4695,8 +4695,8 @@ namespace RegionRuntime {
 	//printf("ENABLE %p %d %d %d %x\n", raw_data, offset, start, count, impl->bits[0]);
 	int pos = start - first_element;
 	for(int i = 0; i < count; i++) {
-	  unsigned *ptr = &(impl->bits[pos >> 5]);
-	  *ptr |= (1U << (pos & 0x1f));
+	  uint64_t *ptr = &(impl->bits[pos >> 6]);
+	  *ptr |= (1ULL << (pos & 0x3f));
 	  pos++;
 	}
 	//printf("ENABLED %p %d %d %d %x\n", raw_data, offset, start, count, impl->bits[0]);
@@ -4706,11 +4706,11 @@ namespace RegionRuntime {
 
 	int pos = start - first_element;
 	for(int i = 0; i < count; i++) {
-	  off_t ofs = offset + ((pos >> 5) << 2);
-	  unsigned val;
+	  off_t ofs = offset + ((pos >> 6) << 3);
+	  uint64_t val;
 	  m_impl->get_bytes(ofs, &val, sizeof(val));
 	  //printf("ENABLED(2) %d,  %x\n", ofs, val);
-	  val |= (1U << (pos & 0x1f));
+	  val |= (1ULL << (pos & 0x3f));
 	  m_impl->put_bytes(ofs, &val, sizeof(val));
 	  pos++;
 	}
@@ -4729,8 +4729,8 @@ namespace RegionRuntime {
 	ElementMaskImpl *impl = (ElementMaskImpl *)raw_data;
 	int pos = start - first_element;
 	for(int i = 0; i < count; i++) {
-	  unsigned *ptr = &(impl->bits[pos >> 5]);
-	  *ptr &= ~(1U << (pos & 0x1f));
+	  uint64_t *ptr = &(impl->bits[pos >> 6]);
+	  *ptr &= ~(1ULL << (pos & 0x3f));
 	  pos++;
 	}
       } else {
@@ -4739,11 +4739,11 @@ namespace RegionRuntime {
 
 	int pos = start - first_element;
 	for(int i = 0; i < count; i++) {
-	  off_t ofs = offset + ((pos >> 5) << 2);
-	  unsigned val;
+	  off_t ofs = offset + ((pos >> 6) << 3);
+	  uint64_t val;
 	  m_impl->get_bytes(ofs, &val, sizeof(val));
[…]

### 2012-05-01 runtime/lowlevel_impl.h

@@ -219,12 +219,13 @@ namespace RegionRuntime {
 
     struct ElementMaskImpl {
       //int count, offset;
-      int dummy;
-      unsigned bits[0];
+      typedef unsigned long long uint64;
+      uint64_t dummy;
+      uint64_t bits[0];
 
       static size_t bytes_needed(off_t offset, off_t count)
       {
-	size_t need = sizeof(ElementMaskImpl) + (((count + 31) >> 5) << 2);
+	size_t need = sizeof(ElementMaskImpl) + (((count + 63) >> 6) << 3);
 	return need;
       }
 	

### 2012-05-01 doc/sc2012/circuit.tex

@@
 lines 32-34 are {\em subtasks} of the main {\tt simulate\_circuit}
-task. A subtask can only take region arguments that are either also region arguments of or subregions
-of region arguments of the parent task.  This requirement
+task. A subtask can only take a region argument that is a region argument of the
+parent task or a subregion of a region argument of the parent task;
+furthermore, the subtask 
+can only have permissions on the region that are compatible with the parent's 
+permissions.   This requirement
 plays an important role in our task scheduling algorithm (see
@@
 
-%Since the wire subregions are known to be disjoint,
-%the write sets of invocations of $calc\_new\_currents$ do not overlap, and can
+%Since the wire subregions are known to be disjoint, the write sets of invocations of $calc\_new\_currents$ do not overlap, and can
 %therefore be safely run in parallel.

### 2012-05-01 doc/sc2012/soop.tex
soop

@@
 \subsubsection{Dependences}
-Conceptually, detecting dependences between a newly registered task $t$ and a previous task 
+Detecting dependences between a newly registered task $t$ and a previous task 
 $t'$ requires comparing the two sets of logical regions accessed.  For each logical region used by
@@
 both regions need only read privileges there is never a dependence, but if either
-task needs write or reduction privileges, the coherence modes are compared:
+task needs write or reduction privileges, the coherence modes are compared using
+the table in Figure~\ref{fig:dependence}.
+
+\begin{figure}
 {\small
@@
 }
-
+\caption{Dependence table.}
+\label{fig:dependence}
+\end{figure}
 A {\tt Dep} entry indicates dependence while {\tt None}
@@
 \subsubsection{Isolation}
+
+\begin{figure}
+\centering
+\begin{tikzpicture}
+\node(top) at (3.5,3.5) { $p$ };
+
+\node(s1) at (1.9,2.5) { $s1$ };
+\node(s2) at (5.1,2.5) { $s2$ };
+
+\node(m1) at (1.9,1.5) { $\ldots$ };
+\node(m2) at (5.1,1.5) { $\ldots$ };
+\node(t1) at (1.9,0.5) { $t_1$ };
+\node(t2) at (5.1,0.5) { $t_2$ };
+
+\draw
+  (top.south) edge (s1.north)
+  (s2.north) edge (top.south)
+  (m1.north) edge (s1.south)
+  (m2.north) edge (s2.south)
+  (t1.north) edge (m1.south)
+  (t2.north) edge (m2.south)
+  ;
+
+\draw[dashed]
+  (t1.east) edge (t2.west)
+  (s1.east) edge (s2.west)
+  ;
+
+\end{tikzpicture}
+\caption{Dependence between tasks implies a dependence between siblings of the least common ancestor.}
+\label{fig:independence}
+\end{figure}
+
 The reader may wonder why we only compare sibling tasks for
@@
 any non-sibling tasks are always detected first as a dependence between
-siblings that are registered earlier in execution!
+siblings that are registered earlier in execution.
 \begin{lemma}
@@
 \rm
-Let $t_1$ and $t_2$ be two distinct, non-sibling tasks other than the root task with parents $p_1$ and $p_2$, respectively.
-If $t_1$ and $t_2$ have a dependence, then $p_1$ and $p_2$ have a dependence.
+Let $t_1$ and $t_2$ be two distinct, non-sibling tasks other than the root task, let $p$ be the least
+common ancestor task of $t_1$ and $t_2$, and let $s_1$ and $s_2$ be the sibling children of $p$ 
+that are ancestors of $t_1$ and $t_2$, respectively.
+If $t_1$ and $t_2$ have a dependence, then $s_1$ and $s_2$ have a dependence.
 \end{lemma}
 \begin{proof}
-Recall from Section~\ref{sec:ex} that the parents' region arguments include the children's region arguments and have
-only more permissions.
+Recall from Section~\ref{sec:ex} that a task's parent, takes region arguments that include the children's region arguments and have only more permissions.  The same is true of any ancestor of a task.
 \end{proof}
-Lemma~\ref{lem:isolation} justifies checking dependences only between sibling tasks, because any dependences between 
-arbitrary tasks are detected first between children of the tasks' least common ancestor.  Legion
-guarantees that all dependence analysis is performed for all children of a task $t$ on the instance of the SOOP where $t$
-runs, which by Lemma~\ref{lem:isolation} is sufficient to detect all required dependences across all tasks.
+Lemma~\ref{lem:isolation} is illustrated in Figure~\ref{fig:isolation}:
+if arbitrary tasks $t_1$ and $t_2$ have a dependence, then, because tasks always work on a subset
+of the data passed to their parents, two siblings $s_1$ and $s_2$ of the task's
+least common ancestor also have a dependence.  Since $s_1$ and $s_2$ are issued earlier in the program
+execution than $t_1$ and $t_2$, $s_1$ and $s_2$ are serialized by the SOOP and the dependence between
+$t_1$ and $t_2$ is observed.
+Note that Lemma~\ref{lem:isolation} also guarantees a dual independence property: if two sibling tasks $s_1$ and $s_2$ have no dependence, then there can never be a dependence between the descendants of $s_1$ and the descendants of $s_2$.
 

### 2012-05-02 doc/sc2012/soop.tex

--- 2012-05-01 doc/sc2012/soop.tex
@@
 \begin{itemize}
-\item If it were necessary to compare all pairs of tasks for dependences that alone
-would be a serial bottleneck in all but the simplest Legion programs.  Fortunately,
+\item If it were necessary to compare all pairs of tasks for dependences that 
+would be a bottleneck in most Legion programs.  Fortunately,
 subtasks enjoy an isolation property that allows Legion to check only task siblings for dependences
@@
 and clean-up (Section~\ref{sec:clean}).
-We discuss each in turn.
+We discuss each stage in turn.
 
@@
 subtasks as {\em children} of the {\em parent} task.  The children of a parent task
-are {\em siblings} of each other.
+are {\em siblings} of each other.  A task always executes on a single processor, but
+its subtasks may execute elsewhere.
 
-When a parent task spawns a child task, the child is {\em registered} with the SOOP 
-on which the parent is executing; registration records the subtask's logical region
-arguments and associated privileges and coherence properties.
-Children are registered in the sequential execution order in which the parent task
-spawns them.
+When a parent task spawns a child task, the child is {\em registered}
+with the SOOP on the processor where the parent is executing;
+registration records the subtask's logical region arguments and
+associated privileges and coherence properties.  Children are
+registered in the sequential execution order in which the parent task
+spawns them and enqueued for dependence analysis.  For example, in the
+circuit simulation in Listing~\ref{lst:code_ex}, the three spawn
+statements on lines 32-34 register will register all of the three
+kinds of subtasks (in program order) on the processor where {\tt
+simulate\_circuit} executes.
 
@@
 Legion can schedule them to execute out of order, in parallel.  Thus,
-the first stage in the execution of a task is dependence analysis
-between tasks.
+the first stage in the execution of a task is dependence analysis on
+the logical regins used by tasks.
 
 \subsubsection{Dependences}
-Conceptually, detecting dependences between a newly registered task $t$ and a previous task 
+Detecting dependences between a newly registered task $t$ and a previously registered task 
 $t'$ requires comparing the two sets of logical regions accessed.  For each logical region used by
@@
 \subsubsection{Isolation}
+
+%
+%\begin{figure}
+%
+%\centering
+%\begin{tikzpicture}
+%\node(top) at (3.5,3.5) { $p$ };
+%
+%\node(s1) at (1.9,2.5) { $s1$ };
+%\node(s2) at (5.1,2.5) { $s2$ };
+%
+%\node(m1) at (1.9,1.5) { $\ldots$ };
+%\node(m2) at (5.1,1.5) { $\ldots$ };
+%\node(t1) at (1.9,0.5) { $t_1$ };
+%\node(t2) at (5.1,0.5) { $t_2$ };
+%
+%\draw
+%  (top.south) edge (s1.north)
+%  (s2.north) edge (top.south)
+%  (m1.north) edge (s1.south)
+%  (m2.north) edge (s2.south)
+%  (t1.north) edge (m1.south)
+%  (t2.north) edge (m2.south)
+%  ;
+%
+%\draw[dashed]
+%  (t1.east) edge (t2.west)
+%  (s1.east) edge (s2.west)
+%  ;
+%
+%\end{tikzpicture}
+%\caption{Dependence between tasks implies a dependence between siblings of the least common ancestor.}
+%\label{fig:independence}
+%\end{figure}
+
 The reader may wonder why we only compare sibling tasks for
 dependences: can't arbitrary pairs of tasks have dependences?  The
-answer is ``yes'', but the following lemma shows that
-any non-sibling tasks are always detected first as a dependence between
-siblings that are registered earlier in execution!
-\begin{lemma}
-\label{lem:isolation}
+answer is ``yes'', but the following observation shows why it is sufficient
+to check only sibling tasks for dependences.
+\begin{observation}
+\label{obs:isolation}
 \rm
-Let $t_1$ and $t_2$ be two distinct, non-sibling tasks other than the root task with parents $p_1$ and $p_2$, respectively.
-If $t_1$ and $t_2$ have a dependence, then $p_1$ and $p_2$ have a dependence.
-\end{lemma}
-\begin{proof}
-Recall from Section~\ref{sec:ex} that the parents' region arguments include the children's region arguments and have
-only more permissions.
-\end{proof}
-Lemma~\ref{lem:isolation} justifies checking dependences only between sibling tasks, because any dependences between 
-arbitrary tasks are detected first between children of the tasks' least common ancestor.  Legion
-guarantees that all dependence analysis is performed for all children of a task $t$ on the instance of the SOOP where $t$
-runs, which by Lemma~\ref{lem:isolation} is sufficient to detect all required dependences across all tasks.
+Let $t_1$ and $t_2$ be two sibling tasks with no dependence.  Then no subtask of $t_1$ has a dependence with any subtask
+of $t_2$.
+\end{observation}
+Recall from Section~\ref{sec:ex} that a subtasks can only access subregions of their parent task's region arguments (and with fewer
+privileges).  Thus if the regions used by $t_1$ and $t_2$ do not overlap, the regions used by any subtasks of $t_1$ cannot
+overlap with the regions used by any subtask of $t_2$.  In the case where the regions used by $t_1$ and $t_2$ do overlap
+but there is no dependence because the regions are used with simultaneous or relaxed coherence, then by definition there is no dependence between the subtasks.
 
 \subsubsection{Efficient Dependence Analysis}
+
+Consider the first two subtasks spawned on line 32 of the circuit
+simulation in Listing~\ref{lst:code_ex}, {\tt
+calc\_new\_currents(pieces[0])} and {\tt
+calc\_new\_currents(pieces[1])}.  The first of these tasks reads and
+writes the private wires subregion {\tt pieces[0].rw\_pvt} in
+exclusive mode, and reads in exclusive mode the node subregions {\tt
+pieces[0].rn\_pvt}, {\tt pieces[0].rn\_shr}, and {\tt
+pieces[0].rn\_ghost} (see line 39).  The second subtask must be
+checked against the first for any dependences; the second subtask uses
+{\tt pieces[1].rw\_pvt} (read/write exclusive) and {\tt
+pieces[1].rn\_pvt}, {\tt pieces[1].rn\_shr}, and {\tt pieces[1].rn\_ghost} (read-only exclusive).
+Recall the region tree in Figure~\ref{sfig:part_fig:tree}.  We reason about the dependences as follows:
+\begin{enumerate}
+\item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rw\_pvt} do not overlap as they are subregions of a disjoint partition.
+\item {\tt pieces[0].rn\_pvt} and {\tt pieces[1].rn\_pvt} do not overlap for the same reason.
+\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_shr} do not overlap, also for the same reason.
+\item {\tt pieces[0].rn\_ghost} may overlap with {\tt pieces[1].rn\_shr}, because they are in two different
+  partitions of the same region.
+\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_ghost} may overlap because they are in two different
+  partitions of the same region.
+\item {\tt pieces[0].rn\_ghost} may overlap with {\tt pieces[1].rn\_ghost} because they are in different subregions of
+a non-disjoint, or {\em aliased}, partition.
+\item All the remaining pairs of regions used by the two tasks (e.g., {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rn\_pvt})
+do not overlap because they are in different region trees and have no common region ancestor.
+\end{enumerate}
+Cases 4-6 indicate a possible dependence, however in each case the regions are accessed with only read privileges and
+so no dependence exists.  Thus, these two subtasks are independent.
+
+A closer look at this example shows that whether there is overlap
+between two regions $r_1$ and $r_2$ can be determined by examining
+their least common ancestor $\lca{r_1}{r_2}$ in the region tree.  Note
+that the region tree alternates between regions and partitions at each
+level.  That is, regions are always children of partitions and
+partitions are always children of regions in the region tree.  There
+are four cases.  If $\lca{r_1}{r_2}$ either does not exist (the regions
+are in different region trees) or is a disjoint partition, then $r_1$
+and $r_2$ are guaranteed to be disjoint.  If $\lca{r_1}{r_2}$ is an
+aliased partition or a region (because $r_1$ and $r_2$ are in
+different partitions of the region, as in cases 4 and 5
+above), then $r_1$ and $r_2$ may not be disjoint.
 

### 2012-05-02 doc/sc2012/soop.tex
soop

@@
 We discuss each stage in turn.
-
-\input{mapping_fig}
 
@@
 The reader may wonder why we only compare sibling tasks for
-dependences: can't arbitrary pairs of tasks have dependences?  The
-answer is ``yes'', but the following observation shows why it is sufficient
-to check only sibling tasks for dependences.
+dependences when presumably arbitrary pairs of tasks may be dependent. 
+The following observation shows why dependence analysis can be limited to
+sibling tasks.
 \begin{observation}
@@
 \end{observation}
-Recall from Section~\ref{sec:ex} that a subtasks can only access subregions of their parent task's region arguments (and with fewer
+Recall from Section~\ref{sec:ex} that subtasks only access subregions of their parent task's region arguments (and with fewer
 privileges).  Thus if the regions used by $t_1$ and $t_2$ do not overlap, the regions used by any subtasks of $t_1$ cannot
@@
 pieces[1].rn\_pvt}, {\tt pieces[1].rn\_shr}, and {\tt pieces[1].rn\_ghost} (read-only exclusive).
-Recall the region tree in Figure~\ref{sfig:part_fig:tree}.  We reason about the dependences as follows:
+It may be helpful to refer to the region tree for nodes in Figure~\ref{sfig:part_fig:tree} (recall that the wires region tree consists of a single disjoint partition).
+We reason about the dependences as follows:
 \begin{enumerate}
@@
 \end{enumerate}
-Cases 4-6 indicate a possible dependence, however in each case the regions are accessed with only read privileges and
-so no dependence exists.  Thus, these two subtasks are independent.
+Cases 4-6 indicate a possible dependence, however in each case the regions are accessed with only read privileges.  Thus, these two subtasks are independent.
 
@@
 their least common ancestor $\lca{r_1}{r_2}$ in the region tree.  Note
-that the region tree alternates between regions and partitions at each
-level.  That is, regions are always children of partitions and
+that the region tree alternates between regions and partitions:
+regions are always children of partitions and
 partitions are always children of regions in the region tree.  There
@@
 
-To detect dependencies efficiently, the high-level runtime maintains a
-local region forest (recall Section~\ref{sec:exec}) for each task.  The
-root regions are the task's region arguments, and the forest includes
-all partitions and subregions created within the task.  To perform
-dependence detection on a region $r$, the runtime first finds a root
-region $r'$ such that $r \rleq r'$.  The runtime then walks the region
-forest from $r'$ to $r$.  Each region $r''$ on the path contains a
-list $L$ of registered tasks using $r''$, and potential dependencies
-are discovered by comparing privileges and coherence modes with the
-tasks in $L$.
+Consider the subtasks {\tt distribute\_charge(pieces[0],dt)} and
+{\tt update\_voltages[1]} in Listing~\ref{lst:code_ex}.  The former task
+uses region {\tt pieces[0].rn\_ghost} and the latter uses region {\tt pieces[1].rn\_shr}.
+The least common ancestor is the region of all shared nodes {\tt p\_nodes\_pvs[1]},
+so these two regions are not disjoint.  Since both tasks write their respective subregions,
+there is a dependence and the {\tt update\_voltages} task can only run after the
+{\tt distribute\_charges} task has completed.  There is an important additional piece 
+of information, as well.  Because the {\tt update\_voltages} task possibly reads the values
+written by the {\tt distribute\_charges} task, the updates done by {\tt distribute\_charges} must
+be reconciled on least common ancestor region {\tt p\_nodes\_pvs[1]}; our depedence analysis
+tracks this information as well.
 
-The runtime next checks if a partition of $r''$ is {\em open}.  An
-open partition indicates that there are active tasks using subregions
-of $r''$.  No more than one partition of a node can be open
-at any time.  If the open partition is is along the path to $r$,
-traversal continues.  If the open partition lies along another
-path the runtime must {\em close} the currently active partition because it cannot guarantee
-disjointness with the region the new task wishes to access.  To close a
-partition, the runtime finds all active tasks in that
-subtree and considers them all dependencies for the new task.
-With the previously active partition closed, the runtime can
-then open the desired partition and traverse further down the tree as
-needed.
+We can now describe the algorithm for the dependence analysis phase of
+the SOOP.  The SOOP maintains a local region forest for each task $t$,
+the roots of which are $t$'s region arguments and including all
+partitions and subregions used within the task.  Dependence analysis
+decorates the region tree with the subtasks of $t$, maintaining the
+following invariant. Let $r'$ be a region argument to subtask $t'$,
+and let $R$ be the set of regions $r''$ such that $r''$ is an argument
+to a task $t''$ registered after $t'$ and $t''$ depends on $t'$ because
+of overlap between $r''$ and $r'$.  Then $t'$ is listed on the region that
+is the least common ancestor of the set $R \cup \{r'\}$.
 
-When the runtime traverses an open partition, its behavior depends on the kind
-of partition being traversed.  If the partition is disjoint, no other subregion
-can overlap with the target region, so the state of other subregions is
-ignored.  However, if the partition is not disjoint, any
-of subregion could overlap with the desired subregion, so
-the runtime must close all open subregions (recording new dependencies
-as it does so) before continuing the traversal.
+We can identify all dependences between sibling tasks and place tasks
+in the correct place in the region tree in amortized time ${\cal O}(nd)$,
+where $n$ is the total number of region arguments to tasks and $d$ is the depth
+of the region tree.  For region $r'$ an argument of task $t'$, the analysis
+walks the unique path in the region forest from a root to $r'$.  For each node $m$
+on the walk from the root to $r'$ the following actions are taken:
+\begin{itemize}
+\item Assume $m$ is a region with multiple child partitions, and let $m'$ be the child on the
+path to $r'$.  Any task in a subtree of $m$ other than $m'$ may conflict with $t'$; these tasks
+are all moved to $m$.
 
-After the runtime reaches the target logical region, it performs dependence
-checks against any active tasks listed for that region.  If a dependence is
-discovered, it is added to the current task's dependence list.  It can then
-be removed from the region tree's active task list, as the current task (which
-will be pushed onto the active task list) {\em dominates} the previous task.
-In a case where the new task has dependencies with all existing active tasks
-for the target node, that whole list is replaced by a single entry for the
-new task.
+\item Assume $m$ is a non-disjoint partition, and let $m'$ be the child on the
+path to $r'$.  Any task in a subtree of $m$ other than $m'$ may conflict with $t'$; these tasks
+are all moved to $m$.
+
+\item Assume $m$ is a disjoint partition, and let $m'$ be the child on the
+path to $r'$.  No tasks in a subtree of $m$ other than $m'$ can conflict wtih $t'$; we simply move
+on to $m'$.
+
+\item Assume $m = r'$.  In this case any dependent tasks in $r'$'s subtree must be moved to $r'$.
+\end{itemize}
+
+\begin{figure*}
+\centering
+\subfigure[Two independent {\tt calc\_new\_currents} tasks.]{
+\label{fig:depexamp:a}
+}
+\subfigure[A {\tt distribute\_charge} task depends on the previous two.]{
+\label{fig:depexamp:b}
+}
+\subfigure[An {\tt update\_voltages} task depends on the previous one.]{
+\label{fig:depexamp:c}
+}
+\label{fig:depexamp}
+\caption{Dependence analysis examples from {\tt circuit\_simulation}.}
+\end{figure*}
+
+Thus, there are two parts to depedence analysis of a region argument
+$r'$ to a task: the walk from a region root to $r'$ and the off-path work
+to move dependent tasks to the least common ancestor on the path.
+Clearly the walk to $r'$ is bounded by the depth of the region tree.
+The off-path work can be made efficient by maintaining at each node a
+record of which subtrees have tasks in them and with what priveleges,
+enabling the off-path search to traverse directly and only to
+dependent tasks.  The off-path work is proportional to how far a task
+is moved back up the region tree.  Thus, for each region argument a
+task is initially placed at some depth in the tree and subsequently
+only moved up towards the root, so the overall work is (amortized)
+${\cal O}(nd)$.  In practice, we have found this level of efficiency,
+where dependence analysis is linear in the number of region arguments (and not
+the square) and proportional to the depth of the region tree (and not its size)
+to be very important in keeping the overhead of the SOOP low.
+
+We now illustrate dependence analysis using the four tasks discussed
+above.  Many more tasks are executed by the circuit simulation in Listing~\ref{lst:code_ex},
+but this subset illustrates the important
+points. Figure~\ref{fig:depexamp} shows the node region tree in three
+stages.  In Figure~\ref{fig:depexamp:a}, the two tasks
+{\tt calc\_new\_currents(pieces[0])} ($c_0$) and {\tt
+    calc\_new\_currents(pieces[1])} ($c_1$) have been placed on their region arguments,
+  which determines these two tasks are independent.
+Next, in Figure~\ref{fig:depexamp:b}, the task {\tt distribute\_charge(pieces[0],dt)} ($d_0$)
+has dependences with both $c_0$ and $c_1$ caused by possible overlap between the pieces of the
+shared node partition and the ghost node partition, so both $c_0$ and $c_1$ are moved to
+the {\tt p\_nodes\_pvs[1]} region.  Finally, the task {\tt update\_voltages[1]} ($u_1$) causes
+$d_0$ to also be moved to {\tt p\_nodes\_pvs[1]} for essentially the same reason; the final
+state of the node region tree is shown in Figure~\ref{fig:depexamp:c}.
 
@@
 \label{sec:map}
+
+\input{mapping_fig}
 

### 2012-05-02 doc/sc2012/circuit.tex
pass up to soop

@@
 {\tt rn} to which the {\tt Node} pointers in fields {\tt in\_nodes}
-and {\tt out\_nodes} point, which guarantees that the two endpoints of
-a wire are always in the same region.
+and {\tt out\_nodes} point.
+% which guarantees that the two endpoints of
+%a wire are always in the same region.
 

### 2012-05-02 doc/sc2012/legion.tex
soop
pass up to soop

@@
 \newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{observation}[theorem]{Observation}
 
@@
 
-Legion programs are written using logical regions, each of which
-has one or more {\em physical instances} when the program executes.  Often
+Legion programs are written using logical regions, which do not commit
+to any particular placement of the data in the machine.  At runtime,
+each logical reagion has one or more {\em physical instances}.  Often
 logical regions correspond to a single physical instance, but it is also common and useful
@@
 to duplicate and distribute read-only data, to keep an old copy of a region to break a 
-write-after-read dependency, or to allow independent, local reductions that are later combined.
+write-after-read dependence, or to allow independent, local reductions that are later combined.
 
@@
 
-\item We use a {\em deferred execution} model that extracts parallelism
-even from highly irregular, unbalanced computations.  Instead of blocking,
-an operation $o$ that cannot immediately execute because it is dependent
-on some other operation $o'$ waits on an event that occurs when $o'$ completes.
-Events can be created for the completion of any operation in Legion, and 
-the scheduler can build long chains of deferred operations.  In Section~\ref{sec:exp}
-we show that dataflow-like deferred execution gives us a significant performance improvement
-over bulk synchronous global synchronization primitives such as barriers.
+%\item We use a {\em deferred execution} model that extracts parallelism
+%even from highly irregular, unbalanced computations.  Instead of blocking,
+%an operation $o$ that cannot immediately execute because it is dependent
+%on some other operation $o'$ waits on an event that occurs when $o'$ completes.
+%Events can be created for the completion of any operation in Legion, and 
+%the scheduler can build long chains of deferred operations.  In Section~\ref{sec:exp}
+%we show that dataflow-like deferred execution gives us a significant performance improvement
+%over bulk synchronous global synchronization primitives such as barriers.
 
@@
 
-%\input{mapping_fig}
-
 \bibstyle{IEEEtran}

### 2012-05-02 doc/sc2012/soop.tex
pass up to soop

@@
 the first stage in the execution of a task is dependence analysis on
-the logical regins used by tasks.
+the logical regions used by tasks.
 
@@
 \end{enumerate}
-Cases 4-6 indicate a possible dependence, however in each case the regions are accessed with only read privileges.  Thus, these two subtasks are independent.
+Cases 4-6 indicate a possible dependence, however in each case the regions are accessed with only read privileges.  Thus, the two subtasks are independent.
 
@@
 regions are always children of partitions and
-partitions are always children of regions in the region tree.  There
+partitions are always children of regions.  There
 are four cases.  If $\lca{r_1}{r_2}$ either does not exist (the regions
@@
 
-Consider the subtasks {\tt distribute\_charge(pieces[0],dt)} and
+Consider subtasks {\tt distribute\_charge(pieces[0],dt)} and
 {\tt update\_voltages[1]} in Listing~\ref{lst:code_ex}.  The former task
@@
 there is a dependence and the {\tt update\_voltages} task can only run after the
-{\tt distribute\_charges} task has completed.  There is an important additional piece 
-of information, as well.  Because the {\tt update\_voltages} task possibly reads the values
-written by the {\tt distribute\_charges} task, the updates done by {\tt distribute\_charges} must
-be reconciled on least common ancestor region {\tt p\_nodes\_pvs[1]}; our depedence analysis
-tracks this information as well.
+{\tt distribute\_charges} task has completed.
 
@@
 of overlap between $r''$ and $r'$.  Then $t'$ is listed on the region that
-is the least common ancestor of the set $R \cup \{r'\}$.
+is the least common ancestor of the set $R \cup \{r'\}$.  Moving $t'$ to an
+ancestor of $r'$ has the effect of coarsening the dependences involving $t'$
+for tasks registered in the future, and this can result in false dependences.
+However, the loss of information can be shown to be small and, as we show next, the
+benefit is that it dramatically speeds up dependence analysis overall.
 
@@
 \item Assume $m$ is a region with multiple child partitions, and let $m'$ be the child on the
-path to $r'$.  Any task in a subtree of $m$ other than $m'$ may conflict with $t'$; these tasks
-are all moved to $m$.
+path to $r'$.  Any task in a subtree of $m$ other than $m'$ on which $t'$ depends 
+is moved to $m$.
 
 \item Assume $m$ is a non-disjoint partition, and let $m'$ be the child on the
-path to $r'$.  Any task in a subtree of $m$ other than $m'$ may conflict with $t'$; these tasks
-are all moved to $m$.
+path to $r'$.  Any task in a subtree of $m$ other than $m'$ that $t'$ depends on is
+moved to $m$.
 
@@
 
-\item Assume $m = r'$.  In this case any dependent tasks in $r'$'s subtree must be moved to $r'$.
+\item Assume $m = r'$.  In this case any dependent tasks in $r'$'s subtree are moved to $r'$.
 \end{itemize}
@@
 the square) and proportional to the depth of the region tree (and not its size)
-to be very important in keeping the overhead of the SOOP low.
+to be very important in minimizing the overhead of the SOOP.
 

### 2012-05-02 doc/sc2012/soop.tex
distribution subsection

@@
 
+Once task $t$'s dependences have been analyzed, the distribution phase
+determines on which processor $t$ will run.  Subsequently the
+mapping phase chooses the physical instances of its logical
+regions that $t$ will use.  However, in choosing the processor and
+physical instances for $t$, it is helpful to know the processors and
+physical instances used by the tasks $t$ depends on---and not just for
+performance, but for correctness.  Thus, the first step in
+distribution is that $t$ waits for all tasks on which $t$ depends to
+map; nothing happens to $t$ during this time.  Once $t$'s mapping
+dependences are satisfied, $t$ is ready to be mapped and places
+itself in the mapping queue.
+
+Once in the mapping queue, SOOPs on other processors in the system may
+ask to steal $t$ from its home SOOP.  The home SOOP may decline the
+request, but if the request is granted, task $t$ along with a copy of
+$t$'s region forest is sent to the remote SOOP.  If $t$ has not been
+stolen when it reaches the head of the queue, the SOOP decides whether
+to execute $t$ on the local processor or send it to another processor.
+
+Thus, task distribution to processors in Legion is both ``pull'' (stealing) and ``push''.
+The various policies (whether steal requests are granted, when to send
+a task to another processor and which processor to send it to, and when to
+try to steal tasks from other SOOPs) are part of the {\em mapping interface}
+and can be defined by the user (see Section~\ref{sec:mapping}).
+
 \subsection{Mapping}

### 2012-05-03 doc/sc2012/soop.tex
full draft of SOOP

@@
 
-\input{mapping_fig}
+After a processor $p$ is chosen for a task, $t$ executes entirely on $p$.
+Before execution, however, we must {\em map} $t$, i.e., choose the
+physical instances for $t$'s logical regions.  A physical instance is assigned
+to a particular memory in the machine and holds a snapshot of the region's data
+from some point in time; a {\em valid} physical instance has data that reflects
+all the updates of all previously mapped tasks.
+
+The region tree used to calculate logical dependences for $t$ is also
+used to hold mapping information.  Each logical region $r$ maintains a
+list of its physical instances and which sibling tasks are using which
+instances and with what permissions.  The important correctness
+property established by the dependence analysis phase is that all
+tasks $t$ depends on map before $t$ maps and no task that depends on
+$t$ maps before $t$ finishes mapping.  Thus, the region tree at the
+moment $t$ maps contains a consistent view of what the state of memory
+will be after $t$'s dependence predecessors have executed.
+
+There are two steps to mapping a particular region argument $r$ of
+task $t$.  First, the mapper must ensure there is at least one {\em valid}
+physical instance of $r$.  For example, assume there is one physical instance $s$ of
+logical region $r$ and that a task $t'$ that mapped prior to $t$ is recorded as writing a
+physical instance $s_i$ of subregion $r_i$ of $r$.  At this point there
+is no valid instance of $r$, because $s$ does not include
+the updates to $s_i$.  To restore $s$ to a valid instance of $r$ the
+system must shedule an operation after $t'$ to copy
+$s_i$ back into $s$.
+
+In the second step, the mapping interface is used to select one of the
+valid physical instances of $r$ or to create a new one.  Typical
+policies are to to choose the existing physical instance that is
+closest to $t$'s processor, or to create a new instance (by issuing a
+copy operation) in the memory closest to $t$'s processor.
+We focus here on the first step, establishing a valid physical instance.
+The mapping interface is discussed further in Section~\ref{sec:mapping}.
+
+Detecting or creating a valid instance is, at its core, a dependence analysis on
+physical instances and thus similar to the mapping dependence analysis on logical
+regions already described in Section~\ref{sec:dep}.  Consider any task $t'$ using
+physical instance $s'$ of logical region $r'$.  There are four cases.
+\begin{enumerate}
+\item If $t'$ has only read permission
+for $r'$ it cannot invalidate any instance of region $r$ and nothing needs to be done.
+
+\item If $t'$ has reduce permission for $r'$ and $t$ has reduce
+  permission for $r$ and both use the same reduction operator, then
+  nothing needs to be done.  This is the only case in which a writer
+  does not require a valid instance---because reductions can be
+  reordered, the recombining of the instances of $r$ and $r'$ (if they
+  overlap) can be deferred to a later consumer of the data (see case
+  4).
+
+\item If $t'$ has write permission for $r'$ and $r$ and $r'$ may
+  overlap (i.e., $\lca{r}{r'}$ is either a region or an aliased
+  partition), then $s'$ must be copied to an instance $s''$ of
+  $\lca{r}{r'}$ and then a fresh instance of $r'$ created from $s''$
+  (if either $r = \lca{r}{r'}$ or $r' = \lca{r}{r'}$ one of the two
+  copy operations is omitted).
+ IS s' REMOVED FROM THE TREE?
+
+\item If $t'$ has reduce permission for $r'$, $t$ has read or write
+  permission for $r$, and $r$ and $r'$ may overlap, then $s'$ must be
+  reduced (using $t'$'s reduction operator) to an instance $s''$ of
+  $\lca{r}{r'}$ and then a fresh instance of $r'$ created from $s''$.
+  (If $r' = \lca{r}{r'}$ the reduction is unneeded, and if $r =
+  \lca{r}{r'}$ the copy is unneeded.)
+ IS s' REMOVED FROM THE TREE?
+\end{enumerate}
+
+Not all of these cases can occur simultaneously for a particular
+region argument $r$ of task $t$.  There may always be multiple
+previous readers (case 1), but the mapping dependence analysis
+guarantees that there would only be one previous writer (case 3) or
+one or more reducers (cases 2 and 4).  To map $r$, we walk from $r$'s
+root ancestor in the region forest to $r$, along the way exploring
+off-path subtrees to find region instances satisfying cases 3 and
+4; the details are similar to the implementation of dependence
+analysis in Section~\ref{sec:dep}.
+
+As part of the walk any required copy and reduction operations are
+issued to construct a valid instance of $r$.  These operations are
+deferred, waiting to execute on the tasks that produce the data they
+copy or reduce. Similarly, the start of $t$'s execution is made
+dependent on completion of the copies/reductions that construct the
+instance of $r$ it will use.
+
+%
+%possibly make this consistent with the way we actually map circuit?
+%
+Consider again the four example tasks from Listing~\ref{lst:code_ex}.
+We focus only on the mapping of the node regions of the region tree in
+Figure~\ref{sfig:part_fig:tree}.  For this example, we assume a simple 
+machine with $n$ processors each with a local memory, and we map the $i$th
+piece of the circuit to the $i$th processor's memory.  Thus, all the
+region arguments of 
+{\tt cal\_new\_currents(pieces[0])}
+and
+{\tt cal\_new\_currents(pieces[1])}
+are mapped to instances in the memory's of processor's 0 and 1, respectively.  The tasks themselves
+are also mapped to processors 0 and 1, respectively.
+
+When {\tt distribute\_charge(pieces[0])} executes, it uses the same
+instances on processor 0 and the task also runs on processor 0.
+Enlarging the example to include more tasks, note that the region
+arguments of {\tt distribute\_charge(pieces[i])} overlap with {\tt
+  pieces[0]} both because the shared and ghost partitions are of the
+same data and because te ghost partitions are not disjoint.  However,
+{\tt distribute\_charge} only performs commutative reductions on all
+of these regions and so the subregion instances can be used as is on
+each of the processors for all {\tt distribute\_charge} tasks.
+Finally, {\tt update\_voltages(pieces[0])} reads and writes {\tt
+  pieces[0].rn\_shr}.  To guarantee the task sees the correct data,
+all of the shared and ghost partitions must be reduced back to the
+region of all shared nodes {\tt p\_nodes\_pvs[1]} (see
+Figure~\ref{sfig:part_fig:tree}, after which a fresh instance of {\tt
+  pieces[0].rn\_shr} can be copied from {\tt p\_nodes\_pvs[1]} to node
+0.
+
+%\input{mapping_fig}
+
+Figure~\ref{fig:gpumapping} shows a more involved mapping of the circuit simulation
+on a cluster of GPUs.  The $i$th private node and wire regions are mapped to the
+$i$th GPU's DRAM and simply left there for all three phases of the simulation.  The $i$th
+shared and ghost regions are mapped to zero-copy memory most of the time, but have to be
+reduced out to the global (GASNet) memory and a new instances of the  shared  regions
+created again with the updated information before the {\tt update\_voltage} tasks execute.
+A similar situation arises before {\tt calc\_new\_currents} tasks run again: the updates
+to the shared nodes by {\tt update\_voltage} have to be propagated to the ghost nodes before
+continuing.
+
+\begin{figure}
+\includegraphics[scale=0.48]{figs/CircuitMem.pdf}
+\caption{Tasks and data for the circuit simulation on a cluster of GPUs.}
+\label{fig:gpumapping}
+\end{figure}
 
@@
 
-\begin{figure}
-\includegraphics[scale=0.48]{figs/CircuitMem.pdf}
-\caption{Tasks and data for the circuit simulation on a cluster of GPUs.}
-\end{figure}
+Once a task $t$ has been mapped it enters the execution stage of the SOOP.  When all of
+the operations (other tasks and copies) on which $t$ depends have completed, $t$ is
+launched on the processor.  When $t$ launches subtasks, they are recursively scheduled as siblings by
+the SOOP using $t$ as the parent task and $t$'s region arguments as the roots of the region forest.
 
@@
 \label{sec:clean}
+
+Once a task is done executing the various pieces of state associated with it can be reclaimed.
+Dependence and mapping information can be removed from the region tree; in point of fact, 
+as an optimization we are often able to remove tasks from the region tree well before the task terminates.
+The most involved aspect of clean-up is colleting physical instances that are no longer in use.
+We use a distributed reference counting scheme to track the number of tasks using physical instances;
+when this number goes to 0 the physical intance can be reclaimed.

### 2012-05-04 doc/sc2012/mapping.tex
Initial draft of the mapping section of the paper

\section{Mapping}
\label{sec:mapping}

While the Legion programming model describes locality and independence information abstractly,
to run a Legion application on a machine we must make concrete decisions about
where tasks will be run and where region instances will be placed.  Instead of placing
the burden of decision making on the programming system, we introduce a programmable mapping
interface that gives a programmer total control over how these decisions are made.  Armed
with this interface, the programmer can then specify application- or machine-specific
mapping decisions that would be difficult for a general-purpose programming system to infer.
In this section we describe the general mapping interface (Section \ref{sec:mapinterface}),
our base implmentation of this interface (Section \ref{sec:defmapper}), and the benefits
of creating custom mappers (Section \ref{sec:custommap}).

\subsection{Mapping Interface}
\label{sec:mapinterface}
A mapper is any object that implements the mapping interface.  The mapping interface
consists of ten function calls that the runtime will make to request direction from
the programmer about how a program should be executed.  To aid in performing the 
decisions made in these calls the mapper has access to a simple 
interface for inspecting the properties of the machine on which it is running.  This includes 
information about all the processors and their kinds (i.e. CPU,GPU), as well as the memories visible
to each processor and their latencies and bandwidths.  For brevity we only
discuss three of the mapping interface calls.

\begin{itemize}
\item {\tt select\_initial\_processor} - for each task the runtime system will
ask the mapper to which processor the task should be sent.  The mapper 
can choose to keep the task on the local processor or send it to any other processor
in the system.

\item {\tt permit\_task\_steal} - when handling a steal request from another processor
the runtime will ask the mapper to decide which tasks are allowed to be stolen.  
By always returning an empty set, the mapper can disable stealing.

\item {\tt map\_task\_region} - for each logical region requested for a task, the
runtime will query the mapper for a prioritized list of memories where a physical instance for this
task should be placed.  To aid the mapping decision the runtime provides a list of the current valid physical intances
of the logical region.  The mapper returns a list of memories in which the runtime
should attempt to either reuse or manifest a physical instance of the logical region.  The
runtime traverses this list and searches for a currently valid instance.  If none can be
found, it attempts to allocate a physical instance and then issue the necessary copies
to retrive the valid data.  If both options fail, it moves on to the next memory in the list.
\end{itemize}

There are two very elegant properties of this mapping interface.  The first property is
that all decisions made by a mapper are orthogonal to the correctness of the program and can only 
impact performance.  Regardless of where a mapper chooses to place a task or map a region, 
the runtime will always schedule tasks and copies in accordance with the privileges and 
coherence properties specified in the Legion program.  Therefore when writing an application
in Legion, a programmer can begin by using the default mapper and later optimize his program
by creating a custom mapper and gradually refining its mapping decisions.

The second useful property of the mapping interface is that it isolates machine-specific decisions
to a specific module of code.  As a result, Legion programs are highly
portable.  To port a Legion program to a new architecture, a programmer only has to
implement a new mapper with decisions specific to the new architecture.

\subsection{Default Mapper}
\label{sec:defmapper}
To make writing Legion applications easier, we provide a default mapper implementation that
can be used to get an application working quickly at a moderate performance level.  In addition,
the default mapper serves as the base class for all custom mappers which makes implementing
custom mappers easier as they only have to override the functions that need to be specialized.

The default mapper employs a simple scheme for mapping tasks and regions.  When
{\tt select\_initial\_processor} is invoked for the default mapper, it first checks for which kind of processors
the task has variants (i.e. GPU).  If the fastest variant is for the kind of processor the mapper is managing
it will choose to keep the task local, otherwise it will send the task to the closest processor of the
fastest variant kind.

The default mapper employs a Cilk-like algorithm for task stealing\cite{CILK95}.  Tasks are kept local
to their target processor whenever possible and only moved when stolen.  Unlike Cilk, the default mapper
has the information necessary for locality-aware stealing.  When {\tt permit\_task\_steal} is called for a task, 
the default mapper inspects the logical regions for the task being stolen, and will mark that other tasks using 
the same logical regions should be stolen as well.

For calls to {\tt map\_task\_region}, the default mapper constructs a stack of memories ordered from best-to-worst
by bandwidth from the local processor.  This stack is then returned as the location of memories to be used for
mapping each region.  Note that this is a greedy algorithm that works well for the common case, but can cause some 
regions to be pulled unnecessarily close to the processor and consuming precious fast memory space.

\subsection{Custom Mappers}
\label{sec:custommap}
To optimize a Legion program or library, programmers can create one or more custom mappers.  
For each Legion task call the programmer specifies which mapper should be invoked by the runtime for
mapping that particular task.  This allows for the composition of Legion applications and libraries
each with their own custom mappers.

Each custom mapper extends the default mapper.  A programmer only has to override the functions necessary 
for creating a custom mapping.  Custom mappers can be used to create totally static mappings, 
mappings that memoize their results, or even totally dynamic mappings.  We describe examples of 
custom mappers in Section \ref{sec:exp}.

### 2012-05-04 doc/sc2012/circuit.tex
revision

@@
 Listing~\ref{lst:code_ex} shows code for an electrical
-circuit simulation, which illustrates the core features of the Legion programming model.
-The circuit simulation takes a collection of
+circuit simulation, which takes a collection of
 of wires and nodes where wires meet.  
@@
 currents, distributes charges, and updates voltages for as many
-time steps as the simulation demands.  We first
-describe the partitioning of regions in Section~\ref{subsec:partitioning}
-and then describe the data type declarations in Section~\ref{subsec:datatypes}.
+time steps as the simulation demands.
 
-\subsection{Regions and Partitions}
-\label{subsec:partitioning}
+The key decisions in a Legion program are how data is grouped into
+regions and how those regions are {\em partitioned} into {\em
+subregions}.  The goal is to pick an organization that makes explicit
+in the algorithms which computations are independent.  A {\tt Circuit}
+has two regions: undifferentiated collections of nodes and wires (line
+3 of Listing~\ref{lst:code_ex}). An efficient parallel
+implementation breaks this unstructured graph into pieces that can be
+processed (mostly) independently. There is always some computation
+across the wires between pieces, but an appropriate
+region organization makes explicit which nodes and wires are
+involved in inter-piece computation and which are involved in only
+intra-piece computation.
 
+% FIXME   We add here a description of how the colors print when printed greyscale.
+%
+Figure~\ref{sfig:part_fig:pvs} shows how the nodes
+in a small graph might be split into three pieces.  Blue nodes are
+attached by wires only to nodes in the same piece and are 
+{\em private} to the piece.  Orange nodes on the boundary of a
+piece are {\em shared} with (connected to) other pieces.
+In the simulation, computations on the
+private nodes of different pieces will be independent, while
+computations on the shared nodes will require communication.  To make
+this explicit in the program, we partition the nodes region into
+private and shared subregions (line 18).  To partition a region, we
+provide a {\em coloring}, which is a relation between the elements of
+a region and a set of colors.  A partition is an object which given a
+coloring and a region $r$ contains for each color $c$ a subregion of
+$r$ containing all the elements of $r$ colored $c$. Note that the
+partition into shared and private nodes is disjoint because each node has one color.
+
+The private nodes and the shared nodes are partitioned again into the private and shared nodes for each circuit piece (lines 21
+and 23);  both partitions are disjoint.  There is another useful partition of the shared nodes:
+for a piece $i$, we will need the shared nodes that border $i$ in other pieces of the circuit.
+This {\em ghost node} partition (line 25) has two interesting properties.  First, it is a second partition of the shared nodes:
+we have two views on to the same collection of data.  Second, the ghost node partition is {\em aliased}, meaning the subregions
+are not disjoint: a node may border several different circuit pieces and belong to more than one ghost node subregion (thus, {\tt node\_neighbor\_map} on line 25 assigns more than one color to some nodes).
+The private, shared, and ghost node subregions
+for the upper-left patch of the example graph are shown in
+Figures~\ref{sfig:part_fig:p_i}, \ref{sfig:part_fig:s_i}, and
+\ref{sfig:part_fig:g_i} respectively.
+
+Figure~\ref{sfig:part_fig:tree} shows the final hierarchy of node
+partitions and subregions. The $*$ symbol indicates a partition is disjoint. This
+{\em region tree} data structure plays an important role in scheduling
+tasks for out-of-order execution (see Section~\ref{sec:soop}).
+The organization of the wires is much simpler: a single disjoint partition
+that assigns each wire to one piece (line 16).
+
+%
+% FIXME
+% Sean wants to use the term task through this paragraph; however we haven't defined it yet.
+% I think it is OK as is since the term is introduced at the end of the paragraph.
+%
 Line 9 declares the main simulator function, which takes a {\tt
@@
 
-Lines 15-21 are responsible for partitioning the circuit into {\tt MAX\_PIECES}
-pieces that can be worked on in parallel.  To partition a region, we
-provide a {\em coloring}, which is a relation between the elements
-of a region and a set of colors.  A {\em partition} is an object which given
-a coloring and a region $r$ contains for each color $c$ a {\em subregion} of $r$
-containing all the elements of $r$ colored $c$.
-
-Wires are simply partitioned in {\tt MAX\_PIECES} subregions (line 16).  
-The nodes are first partitioned into 
-{\em private} and {\em shared} nodes (line 21).  Private nodes
-are touched by a single task in a phase, while shared nodes may
-be referenced by multiple tasks in a phase.  The thick lines in Figure~\ref{sfig:part_fig:pvs}
-show how the nodes in a small graph might be partitioned into three subsets.  Orange
-nodes are shared and blue nodes are private.  Note that
-this partition is disjoint since each node has one color.
-%Our current implementation uses small
-%integers or booleans for the names of colors, but any enumerable set would be
-%fine.  The coloring relation need not be total - it can leave out some elements entirely.  And although it is not required to be injective (it can map a single
-%element to multiple colors), the fact that injective mappings will always
-%result in disjoint subregions makes them strongly encouraged (when possible).
-%Our example uses four colorings to create circuit partitions.
-
-We next further partition the private and shared subregions into {\tt
-  MAX\_PIECES} disjoint subregions each (lines 20-23); the private and shared node subregions
-for the upper-left patch of the graph is shown in
-Figures~\ref{sfig:part_fig:p_i} and \ref{sfig:part_fig:s_i}
-respectively.
-
-We partition the shared nodes a second time using a
-different coloring into the sets of ghost nodes for each task.  This 
-partition is noteworthy because the subregions are not disjoint:
-ghost nodes may be read by multiple tasks, and thus a node $n$  may belong 
-to more than one ghost region, which corresponds to $n$ being assigned
-more than one color by the coloring {\tt node\_neighbor\_map}.
-The ghost subregion for the upper left patch of the graph
- is shown in Figure~\ref{sfig:part_fig:g_i}.
-
-Figure~\ref{sfig:part_fig:tree} shows the final hierarchy of node
-partitions;  the $*$ symbol indicates a partition is disjoint. This
-{\em region tree} data structure plays an important role in scheduling
-tasks for out-of-order execution (see Section~\ref{sec:soop}).
-
-%(These subsets are hopefully reasonably compact, but the correctness
-%of the simulation is not dependent on that.)  
-%Once the subsets are known, the \emph{node\_owner\_map} is created by assigning
-%each node the color corresponding to its subset.  The 
-%\emph{wire\_owner\_map} assigns each wire to the same subset as its ``in\_node''.
-%The \emph{node\_nghbr\_map} maps a node to color(s) of all wires that 
-%connect to it.  Finally, the \emph{node\_sharing\_map} is derived from the
-%\emph{node\_nghbr\_map}, with a node colored ``true'' if any colors other than 
-%its own were used, and ``false'' if the only wires that connect to a node are
-%in the same piece.
-%Figure~\ref{fig:part_fig} shows how the partitions are defined.
-
-%Line 19 uses the \emph{wire\_owner\_map} to partition the wires region of the
-%circuit into a subregion for each piece, but the partitioning of the nodes is
-%more complicated due to the sharing that is necessary between the pieces.
-%First, the \emph{node\_sharing\_map} is used to create two subregions: $p\_nodes\_pvs[false]$ contains all the nodes that are private to some piece (i.e. will
-%never be needed for the computations in any other piece), while $p\_nodes\_pvs[true]$ contains nodes that will be accessed by multiple pieces' computations.
-%Each of these subregions is then partitioned using the $node\_owner\_map$ to
-%create the subregions owned by each computation.  These are the $p_i$ and $s_i$
-%subregions, shown in Figures \ref{sfig:part_fig:p_i} and \ref{sfig:part_fig:s_i}.  Finally, the $g_i$ subregions (shown in Figure~\ref{sfig:part_fig:g_i}) are
-%created using the $node\_nghbr\_map$ to 
-%create subregions that include the ``ghost'' nodes needed to perform each
-%piece's calculations.
-
-%With the partitioning operations completed, the various subregions are
-%recorded in an array of {\tt CircuitPiece} structures.  Lines 32-33 fill in multiple
-%fields of a {\tt CircuitPiece} at once.  The need for such a statement will be described in detail in 
-%section \ref{subsec:datatypes}.
-
-%Because a region relationship can have
-%fields with types that refer its own fields' values, it is often not possible
-%to assign fields one at a time without violating the type checking rules.
-%The simultaneous assignment operator asks the type checker only to make sure
-%that the fields in the structure would have the right (self-referential) types
-%after all the fields have been changed.
-
 Lines 31-57 form the bulk of the actual simulation, performing three
@@
 spawning a task for each piece.  There are no explicit requests for
-parallel execution (the default semantics of {\tt spawn} is sequential
-execution of the spawned tasks) nor is there synchronization required
+parallel execution ({\tt spawn} simply indicates a task call and does
+not require parallel execution) nor is there synchronization required
 between the passes.  Both the fact that the tasks can be run in
@@
 lines 32-34 are {\em subtasks} of the main {\tt simulate\_circuit}
-task. A subtask can only take a region argument that is a region argument of the
-parent task or a subregion of a region argument of the parent task;
-furthermore, the subtask 
-can only have permissions on the region that are compatible with the parent's 
-permissions.   This requirement
-plays an important role in our task scheduling algorithm (see
-Section~\ref{sec:soop}).
+task. A subtask can only take a region argument that is a region
+argument of the parent task or a subregion of a region argument of the
+parent task; furthermore, the subtask can only have permissions on the
+region that are compatible with the parent's permissions.  
+%This
+%requirement plays an important role in our task scheduling algorithm
+%(see Section~\ref{sec:soop}).
 
-%The declarations for the three subtasks are shown on lines 42-51.  
-The  {\tt calc\_new\_currents} task reads and writes the wires subregion, 
-the nodes subregion, and the ghost node region for its piece.
+The  {\tt calc\_new\_currents} task reads and writes the wires subregion 
+and reads the private, shared, and ghost node regions for its piece.
 The {\tt distribute\_charge} subtask is a different phase that reads the piece's 
@@
 
-\subsection{Data Types}
-\label{subsec:datatypes}
+%\subsection{Data Types}
+%\label{subsec:datatypes}
+
 The two basic data types in the circuit simulation are {\tt Node}s and
@@
 Listing~\ref{lst:code_ex} illustrates one way of constructing partitioned
-data structures in Legion: populate a region with data (using {\tt new}, note the allocation
-of the data is omitted from the example) and then partition it
-into subregions.  One can also first partition an empty region into subregions
+data structures in Legion: populate a region with data (using {\tt new}, note the example
+makes use of whatever data has been allocated by the caller of {\tt simulate\_circuit})
+and then partition it into subregions.  One can also first partition an empty region into subregions
 and then allocate data in the subregions. We use both approaches in our
-applications discussed in Section~\ref{sec:exp}.
+applications (see Section~\ref{sec:exp}).
 
@@
 %a wire are always in the same region.
-
 Regions are first-class values in Legion and can be stored in heap
@@
 the circuit.
-
-%\include{part_fig}
-
-%\include{code_ex}
-%\include{part_fig}

### 2012-05-04 doc/sc2012/legion.tex
revision

@@
 information via a novel mapping interface.  We evaluate our Legion
-implementation using four applications: fluid-flow on a regular grid,
-a three-level AMR code solving a standard heat diffusion equation, a circuit simulation,
-and LINPACK.
-%Running on a 4 node cluster with 8 total GPUs 
-%and 4 levels of memory hierarchy, our implementation of Legion 
-%achieves up to 5.9X speedup over a single CPU-GPU node on real-world applications.
+implementation on three applications: fluid-flow on a regular grid,
+a three-level AMR code solving a heat diffusion equation, and 
+a circuit simulation.
+%
+% FIXME: put some performance numbers here
+%
 \end{abstract}
@@
 avoid irregular pointer data structures.  
+%
+% FIXME
+% Sean would like a bit more on why other systems don't support pointer data structures (e.g., the difficulty of
+% alias analysis.  I thought about it and feel it is too much of a digression: mentioning that in the first paragraph
+% makes it sound like it will be an important part of the paper.
+%
 %encode such structures in arrays, which 
@@
 In this paper we describe Legion, a parallel programming system based
-on using {\em logical regions} to describe the organization of data.
-A logical region names a set of objects.  Logical regions are first
-class values in Legion, and in particular may be passed as arguments
+on using {\em logical regions} to describe the organization of data
+and to make explicit relationships useful for reasoning about locality
+and independence.
+A logical region names a set of objects.  Logical regions are first-class 
+values in Legion and may be dynamically allocated, deleted and stored
+in data structures. Regions can also be passed as arguments
 to distinguished functions called {\em tasks} that access the data in
@@
 Furthermore, computations access logical regions with particular {\em
-  privileges} (e.g., {\em read-only} and {\em read-write}, among
-others) and {\em coherence} (e.g., {\em exclusive access} and {\em
+  privileges} ({\em read-only}, {\em read-write}, and {\em reduce}) and {\em coherence} (e.g., {\em exclusive access} and {\em
   atomic access}, among others).  Privileges express how a task may use
@@
 can potentially be run in parallel.
-Coherence properties express the required semantics of concurrent
-accesses to the region.  For example, if the program executes $f_1(r); f_2(r)$ and tasks $f_1$
-and $f_2$ both require {\em exclusive} access to region $r$, then Legion guarantees that 
-$f_1(r)$ completes before $f_2(r)$ begins.   On the other hand, if
+Coherence properties express the required semantics of concurrent region accesses.
+For example, if the program executes $f_1(r); f_2(r)$ and tasks $f_1$
+and $f_2$ both require {\em exclusive} access to region $r$, then Legion guarantees the result will
+be as if $f_1(r)$ completes before $f_2(r)$ begins.   On the other hand, if
 the tasks access $r$ with {\em atomic} coherence, then Legion
@@
 
-To introduce the programming model, we begin with an extended example
-of a circuit simulation in
-Section~\ref{sec:ex}. This example illustrates regions, tasks,
+To introduce the programming model, we present
+a circuit simulation in
+Section~\ref{sec:ex}, illustrating regions, tasks,
 permissions and coherence properties, the interactions between them,
@@
   arguments (Section~\ref{sec:soop}).  In addition to pipelining the execution of tasks over
-  several stages, our SOOP is concurrent and distributed across the machine and is
+  several stages, our SOOP is distributed across the machine and is
   also hierarchical and naturally extracts nested
@@
 
-\item A central part of orchestrating a program's execution is how
+\item Of central importance is how
 tasks are assigned (or {\em mapped}) to processors and how physical instances of logical regions are
-mapped to specific memory units within the machine  (Section~\ref{sec:mapping}).  Often there is
-application-specific information that, if exploited, results in better
-mappings than a generic, one-size-fits-all mapping strategy.  We describe
+mapped to specific memory units (Section~\ref{sec:mapping}).  Often using
+application-specific information results in better
+mappings than a generic mapping strategy.  We describe
 a {\em mapping interface} that allows programmers to give the SOOP a specification
 of how to map tasks and regions for a specific application, or even part of an
-application.  This mapping API is carefully defined so that
+application.  This mapping API is designed so that
 any user-supplied mapping strategy can only affect the performance of applications,
@@
 
-\item We present the results of experiments on four applications:
+\item We present the results of experiments on three applications:
   fluid-flow on a regular grid, a three-level AMR code solving a
-  standard heat diffusion equation, a circuit simulation, and LINPACK (Section~\ref{sec:exp}).
+  standard heat diffusion equation and a circuit simulation.
   We run each of these four codes and best available reference
   versions on three different clusters of multicore processors with
-  attached GPUs: a small cluster of our own (4 nodes with 2 6-core CPUs and 2 GPUs per node), a medium-sized cluster
-  designed for visualization applications (10 nodes with 2 6-core CPUs and 5 GPUs per node), and up to a third of the
-  Keeneland supercomputer (32 nodes, 2 6-core CPUs and 3 GPUs per node)
+  attached GPUs: a small cluster of our own (4 nodes with 12 CPUs and 2 GPUs per node), a medium-sized cluster
+  designed for visualization applications (10 nodes with 12 CPUs and 5 GPUs per node), and up to a third of the
+  Keeneland supercomputer (32 nodes, 12 CPUs and 3 GPUs per node).
+%
+% FIXME Give the same numbers as in the abstract.
+%
 

### 2012-05-04 doc/sc2012/soop.tex
revision

@@
 
-There are two major challenges in implementing an efficient task
-scheduler for Legion:
+Legion uses a {\em  software out-of-order processor}, or SOOP, to schedule tasks.  The SOOP 
+is pipelined, distributed and extracts nested parallelism from subtasks.
+There are two major challenges in implementing an efficient task scheduler:
 \begin{itemize}
-\item  For correctness, Legion must guarantee that every pair of {\em dependent} tasks is serialized in program order.
+\item  For correctness, Legion must preserve data dependences between tasks.
 
@@
 \end{itemize}
-Our scheduler uses three techniques to achieve both correctness and high performance:
-\begin{itemize}
-\item If it were necessary to compare all pairs of tasks for dependences that 
-would be a bottleneck in most Legion programs.  Fortunately,
-subtasks enjoy an isolation property that allows Legion to check only task siblings for dependences
-(see Section~\ref{sec:dep}).
 
-\item Legion uses a {\em  software out-of-order processor}, or SOOP, to schedule tasks.  The SOOP 
-is pipelined and distributed, and also extracts nested parallelism from subtasks.
-
-\item Legion uses a {\em deferred execution model} that decouples the issuing
+To solve the second problem Legion uses a {\em deferred execution model} that decouples the issuing
 of operations from when operations are performed.  An issued operation waits for other operations on
 which it is dependent to complete before executing, but a waiting operation does not cause the SOOP
-to block, allowing Legion to perform other useful work while the operation is pending.
+to block.  With deferred execution, it is worthwhile for the SOOP to run (potentially far) ahead of actual execution,
+allocating resources to and issuing tasks that must wait because they have data dependences on earlier tasks that have not 
+finished executing.   Pipelining and distributing the SOOP further improves throughput and hides latency.
 
-%  For example, Legion may issue a copy
-%operation to move the results of a task $a$ to the place where a task $b$ will take the copied data as
-%an argument.  Tasks $a$ and $b$ and the copy can all be issued, but the copy will not start until
-%task $a$ completes and task $b$ will not start until the copy completes.
+To solve the first problem, detecting and enforcing data dependences
+efficiently at run time, the SOOP analyzes, for each task $t$, the
+permissions on the regions $t$ uses and ensures that $t$ waits on any
+earlier task whose region uses may conflict with $t$'s.  There is a
+difficulty, however.  Recall that the programmer's regions are
+logical: a region names a set of objects, but a region does not say
+where those objects are physically allocated.
+Thus, to {\em map} $t$ (assign physical instances for each logical region used by $t$) the SOOP must know
+both what tasks $t$ depends on how those other tasks are mapped.
 
-\end{itemize}
+Our SOOP design solves this problem by dividing the analysis of
+dependences between two pipeline stages.  The first SOOP stage
+computes task dependences at the granularity of logical regions (Section~\ref{sec:dep}).  This
+does not give enough information to map $t$, but it does identify the
+other tasks that must map before $t$ can map. 
+The third SOOP stage
+maps $t$ by carrying out a more refined analysis of the (already
+mapped) tasks on which $t$ depends (Section~\ref{sec:map}).
+Between these stages, the second stage distributes tasks to other processors
+to improve locality or load balance the machine (Section~\ref{sec:dist}).
+Once the task has been assigned a processor and physical regions it is executed
+using deferred execution (Section~\ref{sec:exec}).  The fifth and final SOOP stage
+reclaims resources from completed tasks (Section~\ref{sec:clean}).
 
-Each processor in the machine runs an instance of the SOOP, which handles all requests
-for Legion services from tasks executing on the processor.  The SOOP is analagous to the out-of-order
-instruction scheduler in superscalar processors, but working at the granularity of tasks and their region
-arguments instead of instructions and their register arguments.
-Legion's SOOP has a five stage pipeline: dependence analysis (Section~\ref{sec:dep}),
-distribution (Section~\ref{sec:dist}),
-mapping (Section~\ref{sec:map}),
-execution (Section~\ref{sec:exec}),
-and clean-up (Section~\ref{sec:clean}).
-We discuss each stage in turn.
-
-\subsection{Depedence Analysis}
+\subsection{Stage 1: Mapping Depedences}
 \label{sec:dep}
 
-When a Legion task runs, it may spawn new subtasks to be executed.  We refer to the
-subtasks as {\em children} of the {\em parent} task.  The children of a parent task
-are {\em siblings} of each other.  A task always executes on a single processor, but
+When a Legion task runs, it may spawn new subtasks to be executed.  
+These subtasks are {\em children} of the {\em parent} task and {\em siblings} of each other.  
+A task always executes on a single processor, but
 its subtasks may execute elsewhere.
 
+Each processor in the system runs an instance of the SOOP.
 When a parent task spawns a child task, the child is {\em registered}
-with the SOOP on the processor where the parent is executing;
-registration records the subtask's logical region arguments and
-associated privileges and coherence properties.  Children are
-registered in the sequential execution order in which the parent task
-spawns them and enqueued for dependence analysis.  For example, in the
-circuit simulation in Listing~\ref{lst:code_ex}, the three spawn
-statements on lines 32-34 register will register all of the three
+with the SOOP on the parent's processor;
+registration records the subtask's logical regions,
+privileges and coherence properties.  Children are
+registered in the sequential order the parent
+spawns them and enqueued for mapping dependence analysis.  In the
+circuit simulation in Listing~\ref{lst:code_ex}, the spawn
+statements on lines 32-34 register all three
 kinds of subtasks (in program order) on the processor where {\tt
@@
 
-If two children have a dependence between their region arguments, the
-SOOP will schedule them to run serially to preserve the program's
-sequential semantics.  If two child tasks have no dependence then
-Legion can schedule them to execute out of order, in parallel.  Thus,
-the first stage in the execution of a task is dependence analysis on
-the logical regions used by tasks.
-
-\subsubsection{Dependences}
-Detecting dependences between a newly registered task $t$ and a previously registered task 
-$t'$ requires comparing the two sets of logical regions accessed.  For each logical region used by
-$t'$ that may overlap with a logical region used by $t$, the privileges
-and coherence modes are compared to determine whether a dependence exists.  If
-both regions need only read privileges there is never a dependence, but if either
-task needs write or reduction privileges, the coherence modes are compared using
-the table in Figure~\ref{fig:dependence}.
+Detecting mapping dependences between a newly registered task $t$ and
+a previously registered task $t'$ requires comparing the two sets of
+logical regions accessed.  For each logical region used by $t'$ that
+may overlap with a logical region used by $t$, the privileges and
+coherence modes are compared to determine whether a dependence exists.
+If both regions need only read privileges there is never a dependence,
+but if either task needs write or reduction privileges, the coherence
+modes are compared using the table in Figure~\ref{fig:dependence}.
+{\tt Dep} indicates dependence while {\tt None}
+indicates independence.  {\tt Same} is a dependence unless the two tasks
+use the same physical instance of the logical region.\footnote{For example, if two tasks wish to acess the region with atomic coherence, then if they are working on the same physical copy of the data atomicity can be guaranteed using, e.g., locking.
+If they are working on different copies of the data then the only safe execution strategy for Legion is to serialize the two tasks.}
+{\tt Cont} 
+indicates a dependence unless a single writer has
+atomic coherence.
 
@@
 \end{figure}
-A {\tt Dep} entry indicates dependence while {\tt None}
-indicates independence.  A {\tt Same} entry is a dependence unless the two tasks
-use the same physical instance of the logical region.\footnote{For example, if two tasks wish to acess the region with atomic coherence, then if they are working on the same physical copy of the data atomicity can be guaranteed using, e.g., locking.
-If they are working on different copies of the data then the only safe execution strategy for Legion is to serialize the two tasks.}
-A {\tt Cont} entry
-indicates a dependence unless a single writer is using the
-atomic coherence mode.  When a dependence is detected, the new task
-waits for the older task to complete before it begins execution.
 
 The table lists {\em simultaneous} and {\em relaxed} coherence modes
-that we have not yet discussed.  Both simultaneous and relaxed
+that we have not yet discussed.  Both modes
 allow other tasks using the region to execute at the same time and differ
@@
 simultaneously (i.e., shared memory semantics).  With relaxed coherence, 
-a task may or may not observe updates to the logical region by other tasks executing at
-the same time.
-
-\subsubsection{Isolation}
+a task may or may not observe concurrent updates.
 
@@
 
-The reader may wonder why we only compare sibling tasks for
-dependences when presumably arbitrary pairs of tasks may be dependent. 
-The following observation shows why dependence analysis can be limited to
-sibling tasks.
+A key property of Legion, that dependence analysis is only needed
+between sibling tasks:
 \begin{observation}
@@
 privileges).  Thus if the regions used by $t_1$ and $t_2$ do not overlap, the regions used by any subtasks of $t_1$ cannot
-overlap with the regions used by any subtask of $t_2$.  In the case where the regions used by $t_1$ and $t_2$ do overlap
+overlap with the regions used by any subtask of $t_2$.  If the regions used by $t_1$ and $t_2$ do overlap
 but there is no dependence because the regions are used with simultaneous or relaxed coherence, then by definition there is no dependence between the subtasks.
-
-\subsubsection{Efficient Dependence Analysis}
 
@@
 \item {\tt pieces[0].rn\_ghost} may overlap with {\tt pieces[1].rn\_ghost} because they are in different subregions of
-a non-disjoint, or {\em aliased}, partition.
-\item All the remaining pairs of regions used by the two tasks (e.g., {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rn\_pvt})
+an aliased partition.
+\item All remaining pairs of regions (e.g., {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rn\_pvt})
 do not overlap because they are in different region trees and have no common region ancestor.
@@
 between two regions $r_1$ and $r_2$ can be determined by examining
-their least common ancestor $\lca{r_1}{r_2}$ in the region tree.  Note
-that the region tree alternates between regions and partitions:
-regions are always children of partitions and
-partitions are always children of regions.  There
-are four cases.  If $\lca{r_1}{r_2}$ either does not exist (the regions
+their least common ancestor $\lca{r_1}{r_2}$ in the region tree.  
+There are four cases.  If $\lca{r_1}{r_2}$ either does not exist (the regions
 are in different region trees) or is a disjoint partition, then $r_1$
@@
 so these two regions are not disjoint.  Since both tasks write their respective subregions,
-there is a dependence and the {\tt update\_voltages} task can only run after the
-{\tt distribute\_charges} task has completed.
+there is a dependence and the {\tt update\_voltages} task can only map after the
+{\tt distribute\_charges} task has mapped.
 
@@
 and let $R$ be the set of regions $r''$ such that $r''$ is an argument
-to a task $t''$ registered after $t'$ and $t''$ depends on $t'$ because
-of overlap between $r''$ and $r'$.  Then $t'$ is listed on the region that
-is the least common ancestor of the set $R \cup \{r'\}$.  Moving $t'$ to an
-ancestor of $r'$ has the effect of coarsening the dependences involving $t'$
-for tasks registered in the future, and this can result in false dependences.
-However, the loss of information can be shown to be small and, as we show next, the
-benefit is that it dramatically speeds up dependence analysis overall.
+to a task $t''$ registered after $t'$ and $t''$ depends on $t'$
+because of overlap between $r''$ and $r'$.  Then $t'$ is listed on the
+region that is the least common ancestor of the set $R \cup \{r'\}$.
+Moving $t'$ to an ancestor of $r'$ has the effect of coarsening the
+dependences involving $t'$ for tasks registered in the future, and
+this can result in false dependences.  However, the loss of
+information can be shown to be small and only affects when $t$ can
+map; any loss of information here will be recovered in the mapping
+phase.  Furthermore, the benefit is that it dramatically speeds up
+dependence analysis overall.
 
-We can identify all dependences between sibling tasks and place tasks
-in the correct place in the region tree in amortized time ${\cal O}(nd)$,
-where $n$ is the total number of region arguments to tasks and $d$ is the depth
-of the region tree.  For region $r'$ an argument of task $t'$, the analysis
+We can identify all mappoing dependences between sibling tasks and place tasks
+in the correct place in the region tree in amortized time ${\cal O}(d)$ per region,
+whre $d$ is the depth of the region tree.  For region $r'$ an argument of task $t'$, the analysis
 walks the unique path in the region forest from a root to $r'$.  For each node $m$
@@
 \begin{itemize}
-\item Assume $m$ is a region with multiple child partitions, and let $m'$ be the child on the
-path to $r'$.  Any task in a subtree of $m$ other than $m'$ on which $t'$ depends 
+\item If $m$ is a region with multiple child partitions and $m'$ is the child on the
+path to $r'$, then any task in a subtree of $m$ other than $m'$ on which $t'$ depends 
 is moved to $m$.
 
-\item Assume $m$ is a non-disjoint partition, and let $m'$ be the child on the
-path to $r'$.  Any task in a subtree of $m$ other than $m'$ that $t'$ depends on is
+\item If $m$ is a non-disjoint partition and $m'$ is the child on the
+path to $r'$, then any task in a subtree of $m$ other than $m'$ that $t'$ depends on is
 moved to $m$.
 
-\item Assume $m$ is a disjoint partition, and let $m'$ be the child on the
-path to $r'$.  No tasks in a subtree of $m$ other than $m'$ can conflict wtih $t'$; we simply move
+\item If $m$ is a disjoint partition and $m'$ is the child on the
+path to $r'$, then no tasks in a subtree of $m$ other than $m'$ can conflict wtih $t'$; we simply move
 on to $m'$.
 
-\item Assume $m = r'$.  In this case any dependent tasks in $r'$'s subtree are moved to $r'$.
+\item If $m = r'$ then any dependent tasks in $r'$'s subtree are moved to $r'$.
 \end{itemize}
@@
 
-Thus, there are two parts to depedence analysis of a region argument
-$r'$ to a task: the walk from a region root to $r'$ and the off-path work
+There are two parts to this analysis:
+the walk from a region root to $r'$ and the off-path work
 to move dependent tasks to the least common ancestor on the path.
@@
 The off-path work can be made efficient by maintaining at each node a
-record of which subtrees have tasks in them and with what priveleges,
+record of which subtrees have tasks in them and with what privileges,
 enabling the off-path search to traverse directly and only to
@@
 task is initially placed at some depth in the tree and subsequently
-only moved up towards the root, so the overall work is (amortized)
-${\cal O}(nd)$.  In practice, we have found this level of efficiency,
-where dependence analysis is linear in the number of region arguments (and not
-the square) and proportional to the depth of the region tree (and not its size)
+only moved up towards the root, so the overall work per region is (amortized)
+${\cal O}(d)$.  In practice, we have found this level of efficiency,
+where analyzing mapping dependences is linear in the number regions used (and not the square) and
+proportional to the depth of the region tree (and not its size)
 to be very important in minimizing the overhead of the SOOP.

### 2012-05-04 doc/sc2012/soop.tex
soop

@@
 To solve the first problem, detecting and enforcing data dependences
-efficiently at run time, the SOOP analyzes, for each task $t$, the
-permissions on the regions $t$ uses and ensures that $t$ waits on any
+efficiently at run time, the SOOP considers, for each task $t$, the
+the regions $t$ uses and ensures that $t$ waits on any
 earlier task whose region uses may conflict with $t$'s.  There is a
-difficulty, however.  Recall that the programmer's regions are
-logical: a region names a set of objects, but a region does not say
-where those objects are physically allocated.
-Thus, to {\em map} $t$ (assign physical instances for each logical region used by $t$) the SOOP must know
-both what tasks $t$ depends on how those other tasks are mapped.
+difficulty, however.  The programmer's regions are logical; a region
+names a set of objects, but does not say where those objects
+are located.  To ensure dependences are satisfied the SOOP must also {\em map} $t$: assign an
+appropriate physical instance to each of $t$'s logical region.  Thus, dependence
+analysis requires knowing both which tasks $t$ depends on and how those
+other tasks are mapped.
 
@@
 dependences between two pipeline stages.  The first SOOP stage
-computes task dependences at the granularity of logical regions (Section~\ref{sec:dep}).  This
-does not give enough information to map $t$, but it does identify the
-other tasks that must map before $t$ can map. 
-The third SOOP stage
-maps $t$ by carrying out a more refined analysis of the (already
-mapped) tasks on which $t$ depends (Section~\ref{sec:map}).
-Between these stages, the second stage distributes tasks to other processors
-to improve locality or load balance the machine (Section~\ref{sec:dist}).
-Once the task has been assigned a processor and physical regions it is executed
-using deferred execution (Section~\ref{sec:exec}).  The fifth and final SOOP stage
-reclaims resources from completed tasks (Section~\ref{sec:clean}).
+computes task dependences at the granularity of logical regions
+(Section~\ref{sec:dep}), which does not give enough information to map
+$t$, but does identify other tasks that must map before $t$ can map.
+The third SOOP stage maps $t$ by carrying out a more refined analysis
+of the (already mapped) tasks on which $t$ depends
+(Section~\ref{sec:map}).  Between these stages, the second stage
+distributes tasks to other processors (Section~\ref{sec:dist}).  Once
+a task has been assigned a processor and physical instancess it is
+issued for deferred execution (Section~\ref{sec:exec}).  The final
+SOOP stage reclaims resources from completed tasks
+(Section~\ref{sec:clean}).
 
-\subsection{Stage 1: Mapping Depedences}
+\subsection{Stage 1: Mapping Dependences}
 \label{sec:dep}
@@
 the roots of which are $t$'s region arguments and including all
-partitions and subregions used within the task.  Dependence analysis
-decorates the region tree with the subtasks of $t$, maintaining the
-following invariant. Let $r'$ be a region argument to subtask $t'$,
-and let $R$ be the set of regions $r''$ such that $r''$ is an argument
-to a task $t''$ registered after $t'$ and $t''$ depends on $t'$
+partitions and subregions used by $t$.  Dependence analysis
+places the child tasks of $t$ on the regions they use, maintaining the
+following invariant. Let $r'$ be a region used by child task $t'$,
+and let $R$ be the set of regions $r''$ such that $r''$ is used
+by a task $t''$ registered after $t'$ and $t''$ depends on $t'$
 because of overlap between $r''$ and $r'$.  Then $t'$ is listed on the
 region that is the least common ancestor of the set $R \cup \{r'\}$.
-Moving $t'$ to an ancestor of $r'$ has the effect of coarsening the
-dependences involving $t'$ for tasks registered in the future, and
+Placing $t'$ on an ancestor of $r'$ has the effect of coarsening the
+dependences involving $t'$, which
 this can result in false dependences.  However, the loss of
 information can be shown to be small and only affects when $t$ can
-map; any loss of information here will be recovered in the mapping
+map; any information lost is recovered in the subsequent mapping
 phase.  Furthermore, the benefit is that it dramatically speeds up
@@
 in the correct place in the region tree in amortized time ${\cal O}(d)$ per region,
-whre $d$ is the depth of the region tree.  For region $r'$ an argument of task $t'$, the analysis
+where $d$ is the depth of the region tree.  For region $r'$ an argument of task $t'$, the analysis
 walks the unique path in the region forest from a root to $r'$.  For each node $m$
@@
 only moved up towards the root, so the overall work per region is (amortized)
-${\cal O}(d)$.  In practice, we have found this level of efficiency,
-where analyzing mapping dependences is linear in the number regions used (and not the square) and
-proportional to the depth of the region tree (and not its size)
-to be very important in minimizing the overhead of the SOOP.
+${\cal O}(d)$.  
+%In practice, we have found this level of efficiency,
+%where analyzing mapping dependences is linear in the number regions used (and not the square) and
+%proportional to the depth of the region tree (and not its size)
+%to be very important in minimizing the overhead of the SOOP.
 
-We now illustrate dependence analysis using the four tasks discussed
-above.  Many more tasks are executed by the circuit simulation in Listing~\ref{lst:code_ex},
+We illustrate dependence analysis using the four tasks discussed
+above.  Many more tasks are executed by the circuit simulation,
 but this subset illustrates the important
@@
 
-\subsection{Distribution}
+\subsection{Stage 2: Distribution}
 \label{sec:dist}
 
-Once task $t$'s dependences have been analyzed, the distribution phase
-determines on which processor $t$ will run.  Subsequently the
-mapping phase chooses the physical instances of its logical
-regions that $t$ will use.  However, in choosing the processor and
-physical instances for $t$, it is helpful to know the processors and
-physical instances used by the tasks $t$ depends on---and not just for
-performance, but for correctness.  Thus, the first step in
-distribution is that $t$ waits for all tasks on which $t$ depends to
+The first step in distribution is that $t$ waits for all tasks on which $t$ depends to
 map; nothing happens to $t$ during this time.  Once $t$'s mapping
-dependences are satisfied, $t$ is ready to be mapped and places
-itself in the mapping queue.
+dependences are satisfied, $t$ is ready to be mapped and is placed
+in the mapping queue.
 
@@
 The various policies (whether steal requests are granted, when to send
-a task to another processor and which processor to send it to, and when to
-try to steal tasks from other SOOPs) are part of the {\em mapping interface}
+a task to another processor, etc.) are part of the {\em mapping interface}
 and can be defined by the user (see Section~\ref{sec:mapping}).
 
-\subsection{Mapping}
+\subsection{Stage 3: Mapping}
 \label{sec:map}
 
-After a processor $p$ is chosen for a task, $t$ executes entirely on $p$.
-Before execution, however, we must {\em map} $t$, i.e., choose the
-physical instances for $t$'s logical regions.  A physical instance is assigned
-to a particular memory in the machine and holds a snapshot of the region's data
-from some point in time; a {\em valid} physical instance has data that reflects
-all the updates of all previously mapped tasks.
+%After a processor $p$ is chosen for a task, $t$ executes entirely on $p$.
+%nBefore execution, however, we must {\em map} $t$, i.e., choose the
+%physical instances for $t$'s logical regions.  A physical instance is assigned
+%to a particular memory in the machine and holds a snapshot of the region's data
+%from some point in time; a {\em valid} physical instance has data that reflects
+%all the updates of all previously mapped tasks.
 
-The region tree used to calculate logical dependences for $t$ is also
-used to hold mapping information.  Each logical region $r$ maintains a
+The region tree is also used to hold mapping information.  Each logical region $r$ maintains a
 list of its physical instances and which sibling tasks are using which
 instances and with what permissions.  The important correctness
-property established by the dependence analysis phase is that all
+property established by stage 1 is that all
 tasks $t$ depends on map before $t$ maps and no task that depends on
@@
 moment $t$ maps contains a consistent view of what the state of memory
-will be after $t$'s dependence predecessors have executed.
+will be after $t$'s dependence predecessors execute.
 
-There are two steps to mapping a particular region argument $r$ of
+There are two steps to mapping a region $r$ used by
 task $t$.  First, the mapper must ensure there is at least one {\em valid}
-physical instance of $r$.  For example, assume there is one physical instance $s$ of
-logical region $r$ and that a task $t'$ that mapped prior to $t$ is recorded as writing a
-physical instance $s_i$ of subregion $r_i$ of $r$.  At this point there
-is no valid instance of $r$, because $s$ does not include
-the updates to $s_i$.  To restore $s$ to a valid instance of $r$ the
-system must shedule an operation after $t'$ to copy
-$s_i$ back into $s$.
+physical instance of $r$.  If an instance of a subregion of $r$ has been written by a previous task,
+for example, it is necessary to copy that subregion back to an instance of the parent $r$ so that $t$
+sees the correct data. In the second step, the mapping interface selects one of the
+valid physical instances of $r$ or creates a new one.  
+We focus here on the first step; the mapping interface is discussed further in Section~\ref{sec:mapping}.
 
-In the second step, the mapping interface is used to select one of the
-valid physical instances of $r$ or to create a new one.  Typical
-policies are to to choose the existing physical instance that is
-closest to $t$'s processor, or to create a new instance (by issuing a
-copy operation) in the memory closest to $t$'s processor.
-We focus here on the first step, establishing a valid physical instance.
-The mapping interface is discussed further in Section~\ref{sec:mapping}.
+%For example, assume there is one physical instance $s$ of
+%logical region $r$ and that a task $t'$ that mapped prior to $t$ is recorded as writing a
+%physical instance $s_i$ of subregion $r_i$ of $r$.  At this point there
+%is no valid instance of $r$, because $s$ does not include
+%the updates to $s_i$.  To restore $s$ to a valid instance of $r$ the
+%system must shedule an operation after $t'$ to copy
+%$s_i$ back into $s$.
 
-Detecting or creating a valid instance is, at its core, a dependence analysis on
+%Typical
+%policies are to to choose the existing physical instance that is
+%closest to $t$'s processor, or to create a new instance (by issuing a
+%copy operation) in the memory closest to $t$'s processor.
+
+Detecting or creating a valid instance of $r$ is, at its core, a dependence analysis on
 physical instances and thus similar to the mapping dependence analysis on logical
-regions already described in Section~\ref{sec:dep}.  Consider any task $t'$ using
-physical instance $s'$ of logical region $r'$.  There are four cases.
+regions described in Section~\ref{sec:dep}.  Consider any task $t'$ using
+physical instance $s'$ of logical region $r'$ in the region tree.  There are four cases.
 \begin{enumerate}
-\item If $t'$ has only read permission
-for $r'$ it cannot invalidate any instance of region $r$ and nothing needs to be done.
+\item If $t'$ has only read privileges
+for $r'$ it cannot invalidate any instance of $r$.
 
-\item If $t'$ has reduce permission for $r'$ and $t$ has reduce
-  permission for $r$ and both use the same reduction operator, then
-  nothing needs to be done.  This is the only case in which a writer
+\item If $t'$ has read/write privileges for $r'$, and $r, r'$ may
+  overlap (i.e., $\lca{r}{r'}$ is a region or aliased
+  partition), then $s'$ must be copied to an instance $s''$ of
+  $\lca{r}{r'}$ and a fresh instance of $r'$ created from $s''$.
+%  (if either $r = \lca{r}{r'}$ or $r' = \lca{r}{r'}$ one of the two
+%  copies is omitted)
+Instance $s'$ is removed from the region tree
+  but not deallocated (see Section~\ref{sec:clean}).
+
+\item If $t'$ has reduce privileges for $r'$ and $t$ has reduce
+  privileges for $r$ and both use the same reduction operator, then
+  nothing is done.  This is the only case in which a writer
   does not require a valid instance---because reductions can be
-  reordered, the recombining of the instances of $r$ and $r'$ (if they
+  reordered, combining the instances of $r$ and $r'$ (if they
   overlap) can be deferred to a later consumer of the data (see case
@@
 
-\item If $t'$ has write permission for $r'$ and $r$ and $r'$ may
-  overlap (i.e., $\lca{r}{r'}$ is either a region or an aliased
-  partition), then $s'$ must be copied to an instance $s''$ of
-  $\lca{r}{r'}$ and then a fresh instance of $r'$ created from $s''$
-  (if either $r = \lca{r}{r'}$ or $r' = \lca{r}{r'}$ one of the two
-  copy operations is omitted).
- IS s' REMOVED FROM THE TREE?
-
-\item If $t'$ has reduce permission for $r'$, $t$ has read or write
-  permission for $r$, and $r$ and $r'$ may overlap, then $s'$ must be
+\item If $t'$ has reduce privilege for $r'$ and $t$ has read, write, 
+  or reduce privileges with a different operator than $t'$
+  for $r$, and $r$ and $r'$ may overlap, then $s'$ must be
   reduced (using $t'$'s reduction operator) to an instance $s''$ of
   $\lca{r}{r'}$ and then a fresh instance of $r'$ created from $s''$.
-  (If $r' = \lca{r}{r'}$ the reduction is unneeded, and if $r =
-  \lca{r}{r'}$ the copy is unneeded.)
- IS s' REMOVED FROM THE TREE?
+%  (If $r' = \lca{r}{r'}$ the reduction is unneeded, and if $r =
+%  \lca{r}{r'}$ the copy is unneeded.) 
+Instance $s'$ is removed from the region
+  tree but not deallocated.
 \end{enumerate}
 
-Not all of these cases can occur simultaneously for a particular
-region argument $r$ of task $t$.  There may always be multiple
-previous readers (case 1), but the mapping dependence analysis
-guarantees that there would only be one previous writer (case 3) or
-one or more reducers (cases 2 and 4).  To map $r$, we walk from $r$'s
+%Not all of these cases can occur simultaneously for a particular
+%region argument $r$ of task $t$.  There may always be multiple
+%previous readers (case 1), but the mapping dependence analysis
+%guarantees that there would only be one previous writer (case 3) or
+%one or more reducers (cases 2 and 4).  
+To map $r$, we walk from $r$'s
 root ancestor in the region forest to $r$, along the way exploring
 off-path subtrees to find region instances satisfying cases 3 and
-4; the details are similar to the implementation of dependence
-analysis in Section~\ref{sec:dep}.
+4. The details and analysis are similar to the implementation of dependence
+analysis in Section~\ref{sec:dep}; the amortized work per region mapped
+is proportional to the depth of the region tree.
 
@@
 
-%
-%possibly make this consistent with the way we actually map circuit?
-%
-Consider again the four example tasks from Listing~\ref{lst:code_ex}.
-We focus only on the mapping of the node regions of the region tree in
-Figure~\ref{sfig:part_fig:tree}.  For this example, we assume a simple 
-machine with $n$ processors each with a local memory, and we map the $i$th
-piece of the circuit to the $i$th processor's memory.  Thus, all the
-region arguments of 
-{\tt cal\_new\_currents(pieces[0])}
-and
-{\tt cal\_new\_currents(pieces[1])}
-are mapped to instances in the memory's of processor's 0 and 1, respectively.  The tasks themselves
-are also mapped to processors 0 and 1, respectively.
+Figure~\ref{fig:depexamp:c} shows part of the mapping of task {\tt update\_voltages[0]}; we focus
+only on the shared nodes.
+Because the immediately preceding {\tt distribute\_charges} tasks all performed the same reduction
+on the shared and ghost nodes they were allowed to run in parallel.  But {\tt update\_voltages[0]}
+needs read/write privilege for {\tt pieces[0].rn\_shr}, which forces all of the reductions to be merged
+back to an instance of the {\tt all-shared} nodes region, from which a new instance of {\tt pieces[0].rn\_shr} is copied and used by {\tt update\_voltages[0]}.
 
-When {\tt distribute\_charge(pieces[0])} executes, it uses the same
-instances on processor 0 and the task also runs on processor 0.
-Enlarging the example to include more tasks, note that the region
-arguments of {\tt distribute\_charge(pieces[i])} overlap with {\tt
-  pieces[0]} both because the shared and ghost partitions are of the
-same data and because te ghost partitions are not disjoint.  However,
-{\tt distribute\_charge} only performs commutative reductions on all
-of these regions and so the subregion instances can be used as is on
-each of the processors for all {\tt distribute\_charge} tasks.
-Finally, {\tt update\_voltages(pieces[0])} reads and writes {\tt
-  pieces[0].rn\_shr}.  To guarantee the task sees the correct data,
-all of the shared and ghost partitions must be reduced back to the
-region of all shared nodes {\tt p\_nodes\_pvs[1]} (see
-Figure~\ref{sfig:part_fig:tree}, after which a fresh instance of {\tt
-  pieces[0].rn\_shr} can be copied from {\tt p\_nodes\_pvs[1]} to node
-0.
-
-%\input{mapping_fig}
-
-Figure~\ref{fig:gpumapping} shows a more involved mapping of the circuit simulation
-on a cluster of GPUs.  The $i$th private node and wire regions are mapped to the
-$i$th GPU's DRAM and simply left there for all three phases of the simulation.  The $i$th
-shared and ghost regions are mapped to zero-copy memory most of the time, but have to be
-reduced out to the global (GASNet) memory and a new instances of the  shared  regions
-created again with the updated information before the {\tt update\_voltage} tasks execute.
-A similar situation arises before {\tt calc\_new\_currents} tasks run again: the updates
-to the shared nodes by {\tt update\_voltage} have to be propagated to the ghost nodes before
-continuing.
+Figure~\ref{fig:gpumapping} shows a timeline of our mapping of the
+circuit simulation on a cluster of GPUs.  The $i$th private node and
+wire regions are mapped to the $i$th GPU's DRAM and simply left there
+for all three phases of the simulation.  The $i$th shared and ghost
+regions are mapped to zero-copy memory most of the time, but, as just
+illustrated, must be reduced out to the all-shared node region in the
+global (GASNet) memory and then a new instance of the shared
+subregions created with the updated information before the {\tt
+  update\_voltages} tasks execute.  A similar situation arises before
+{\tt calc\_new\_currents} tasks run again: the updates to the shared
+nodes by {\tt update\_voltage} must be propagated to the ghost nodes
+before continuing.
 
@@
 
-\subsection{Execution}
+\subsection{Stage 4: Execution}
 \label{sec:exec}
@@
 the operations (other tasks and copies) on which $t$ depends have completed, $t$ is
-launched on the processor.  When $t$ launches subtasks, they are recursively scheduled as siblings by
+launched on the processor.  When $t$ spawns subtasks, they are recursively scheduled as siblings by
 the SOOP using $t$ as the parent task and $t$'s region arguments as the roots of the region forest.
 
-\subsection{Clean-Up}
+\subsection{Stage 5: Clean-Up}
 \label{sec:clean}
 
-Once a task is done executing the various pieces of state associated with it can be reclaimed.
-Dependence and mapping information can be removed from the region tree; in point of fact, 
-as an optimization we are often able to remove tasks from the region tree well before the task terminates.
-The most involved aspect of clean-up is colleting physical instances that are no longer in use.
-We use a distributed reference counting scheme to track the number of tasks using physical instances;
-when this number goes to 0 the physical intance can be reclaimed.
+Once task $t$ is done executing its state can be reclaimed Dependence
+and mapping information is removed from the region tree (it can
+usually be removed even earlier).  The most involved aspect of
+clean-up is collecting physical instances that are no longer in use,
+for which we use a distributed reference counting scheme.

### 2012-05-04 doc/sc2012/experiments.tex

\section{Experiments}
\label{sec:exp}

The efficiency and scalability of Legion was examined by measuring the performance
of three applications on three clusters.  Table \ref{table:systems} summarizes 
the three clusters that were used.

\begin{tabular}{lccccc}
 & Nodes & CPUs/Node & GPUs/Node & DRAM/Node & Infiniband \\
sapling & 4 & 2x Xeon 5680 (HT enabled) & 2x Tesla C2070 & 48 GB & 2x QDR \\
viz & 10 & 2x Xeon 5680 (HT disabled) & 5x Quadro Q5000 & 24 GB & QDR \\
keeneland & 32* & 2x Xeon 5660 (HT disabled) & 3x Tesla M2090 & 24 GB & QDR 
\end{tabular}

Nodes with multiple instances of multiple resources are a challenge for 
existing programming models that have a flat model of the system (e.g. MPI).
They are faced with a choice of lumping all 
the resources together and ignoring the internal irregularities (e.g. NUMA
in multi-socket x86 systems) or of separating a single physical node into
multiple smaller nodes, ignoring the better affinity between sockets/GPUs/etc
on the same node and either wasting or over-subscribing some resources when
the quantities of different resources don't share a common divisor.

In contrast, the machine model used by the Legion runtime is based on an
affinity graph of CPUs and GPUs and is able to accurately capture complex
machine hierarchies.

For each application, multiple problem sizes were used, and each size problem was
run on subsets of each machine ranging from the smallest (a single CPU core or GPU)
to the largest or near-largest.  (Keeneland was an exception here - although it has
115 nodes, we limited our runs to at most 32 nodes to get sufficient cluster time.)
By looking at the performance of the same size problem over progressively larger
machines, we are able to measure the ``strong scaling'' that Legion is able to
achive.  By increasing the problem size as well, we also measure ``weak scaling''.

\subsection{Circuit Simulation}
\label{subsec:exp_ckt}

The first experiment run was the distributed circuit simulation, described in 
section \ref{sec:ex}, with all application code running on the CUDA-capable GPUs in
each node.  The application kernels are written in CUDA, while the Legion runtime
handles all the resource allocation, scheduling, and data movement.  In particular,
Legion's ability to efficiently manage, access, and move the irregularly partitioned
shared data around the system while keeping the private nodes and wires resident in
each GPU's framebuffer memory is critical in keeping the overhead low, a neccessity
for good scalability.

Circuits of two different sizes were simulated.  The first had 480K wires, connecting
120K nodes.  The second is twice as large, with nearly a million wires connecting a
quarter of a million nodes.  In addition to running these tests on varying number of
nodes, the number of GPUs used by the runtime was also varied.  In no case did the 
changes to nodes or number of GPUs per node require any changes to the application code.

The circuit simulation used a simple, but application-specific, mapper.  At initialization
time, the mapper queries the list of GPUs in the machine and identifies each GPU's
framebuffer memory and ``zero-copy'' memory (i.e. the pinned memory that both the GPU and
CPUs in the system can access directly).  Once the circuit is partitioned, the partitions
are assigned a ``home'' GPU in round-robin fashion.  Every task related to that partition is
then sent to that GPU, with no task stealing allowed.  (If the circuit is partitioned well,
there should be minimal load imbalance between the pieces, and the rebalancing benefit of 
moving a partition is unlikely to be worth the cost of moving all the 
private data for a piece from one GPU to another.)  The regions for a task are mapped as 
shown in figure \ref{fig:gpumapping}.

Figure \ref{fig:ckt_speed} shows the performance of the Legion circuit simulation relative
to a single-GPU implmementation hand-coded in CUDA.  Each line shows the scaling of
a particular problem size as the number of nodes is varied.  Our results demonstrate
excellent strong scaling behavior, with speedups of *??*x (compared to the single-GPU CUDA
version) for the small problem on 48 GPUs and *??*x for the larger problem size on 96 GPUs.

Figure \ref{fig:ckt_overhead} shows the fraction of the overall simulation time (summed over
all nodes) spent in the application kernels compared to the various pieces of the Legion
runtime.  As the node count increases, the non-communication overhead stays relatively constant.
Only the communication overhead grows, and it grows nearly linearly with the node count.
A linear growth in communication cost as a simulation is split over multiple nodes is to
be expected.

\begin{figure}
\subfigure[Circuit Simulation Speed Relative to Single-GPU Implementation]
{
\includegraphics[scale=0.4]{figs/circuit_speedups.pdf}
\label{fig:ckt_speed}
}

\subfigure[Overhead of Circuit Simulation on Keeneland with 3 GPUs/Node]
{
\includegraphics[scale=0.4]{figs/circuit_overhead.pdf}
\label{fig:ckt_overhead}
}
\caption{Circuit Simulation Results}
\end{figure}

\subsection{Particle Simulation}
\label{subsec:exp_fluid}

Our second experiment is a port of the \emph{fluidanimate} benchmark from the PARSEC suite\cite{bienia11benchmarking},
which does a particle-based simulation of a fluid.  Each particle interacts only with other nearby
particles, so the benchmark divides the space in which the fluid
can move into a three-dimensional array of cells such that the range of interaction is limited to just the
cells adjacent (including diagonals) to the one a particle resides in.  Originally designed to be a
chip multiprocessor benchmark, the application divides the array into ``grids'' and assigns each one to 
a thread.  Per particle locking is used to safely access particles in cells that lie
on the edge of a grid.  This fine-grained locking along with the assumption of a shared address space lead to
good scaling in a multi-core processor, but rule out any attempt to scale beyond a single node.  To extend the
scaling further, our port divides uses region partitioning to divide the array into grids in a similar
way, but handles the interaction between grids in a way that doesn't rely on access to uniform shared memory.
Instead, the Legion version of the benchmark creates explicit ghost copies of cells on the boundaries of grids
and uses those ghost cells to exchange information between the grids.

The mapper for the particle simulation is very simple - it just maps one grid's tasks onto each processor and
maps all regions for that grid (both the internal the ghost cells) into the system memory used by that processor.
The exchange of the ghost cell data between processors is handled by the Legion runtime as a ghost cell region
gets alternately mapped in two different memories.

Figure \ref{fig:fluid_single} compares the performance of the Legion implementation against the PARSEC version,
using a relatively small problem (300K particles on a 15x21x15 array of cells).  Between 1 and 4 threads, the
results for sapling are nearly indistinguishable, indicating 
neither the Legion runtime nor the the restructuring of the implementation to allow multi-node scaling impose
any significant overhead.  In fact, it's possible that the use of explicit ghost cells rather than 
fine-grained sharing of cache lines might be a net win.  At 8 threads and above, the performance begins to vary.
Both the Legion and PARSEC version on viz flatten out as they over-subscribe the 12 physical cores.  On sapling,
which has HyperThreading enabled, deviations from linear begin sooner as the operating system's scheduler's
thread placement choices begin to matter.

To measure scaling beyond a single node, three different problem sizes were run for each of the three systems,
 with the results plotted in figure
\ref{fig:fluid_multi}.  For the smallest problem (300K particles), we observe a 20\% speedup when going to 1
node to 2 (a total of 16 threads), but begin slowing down beyond that due to communication overhead - at 4 nodes
there are twice as many ghost cells as interior grid cells.  The larger problem sizes (2.4M and 19M
particles) do much better, with scaling of up to 5.4x when going from 1 node up to 16.

%Although the particle simulation being performed is on a regular array of cells, it turns out that the
%distribution of particles amongst the cells is very irregular.  The simulation models gravity, which points in
%the -Y direction, so the particles are clustered mostly in the lower half of the cell array.  The PARSEC
%implementation works around this imbalance by only slicing the cell array through the X and Z axes, yielding
%grids that are uniformly populated, even if the number of boundary cells (for which locks must be used) is
%increased above the minimum. MORE TEXT NEEDED HERE

\begin{figure}
\subfigure[Single-Node Particle Simulation Speed]
{
\includegraphics[scale=0.4]{figs/fluid_singlenode.pdf}
\label{fig:fluid_single}
}

\subfigure[Multi-Node Scaling]
{
\includegraphics[scale=0.4]{figs/fluid_multinode.pdf}
\label{fig:fluid_multi}
}

%\subfigure[Effect of Load-Balanced Partitioning]
%{
%\includegraphics[scale=0.4]{figs/fluid_balance.pdf}
%\label{fig:fluid_balance}
%}
\caption{Fluid Simulation Results}
\end{figure}

\subsection{Adaptive Mesh Refinement}
\label{subsec:exp_amr}

\begin{figure}
\subfigure[Sapling Results]
{
\includegraphics[scale=0.4]{figs/Sapling_amr.pdf}
\label{fig:amr_sapling}
}

\subfigure[Viz Results]
{
\includegraphics[scale=0.4]{figs/Viz_amr.pdf}
\label{fig:amr_viz}
}

\subfigure[Keeneland Results]
{
\includegraphics[scale=0.4]{figs/Keeneland_amr.pdf}
\label{fig:amr_keeneland}
}
\caption{Adaptive Mesh Refinement Results}
\end{figure}

### 2012-05-04 doc/sc2012/circuit.tex
Minor edits for the paper

@@
 circuit simulation, which takes a collection of
-of wires and nodes where wires meet.  
-The simulation repeatedly updates
+wires and nodes where wires meet.  
+The simulation repeatedly calculates 
 currents, distributes charges, and updates voltages for as many
@@
 with read-write privileges and {\em exclusive} coherence (i.e., no other
-function can access these two regions concurrently).  The simulation 
+task can access these two regions concurrently or be reordered around this
+task if they use either of the regions).  The simulation 
 reads and writes all nodes and wires, and it must be done
@@
 not require parallel execution) nor is there synchronization required
-between the passes.  Both the fact that the tasks can be run in
+between the passes.  Which tasks can be run in
 parallel within a pass and the required inter-pass dependencies are

### 2012-05-04 doc/sc2012/soop.tex
figs

@@
 \centering
-\subfigure[Two independent {\tt calc\_new\_currents} tasks.]{
+
+\subfigure[Two {\tt calc\_new\_currents} tasks.]{
+\includegraphics[scale=0.45]{figs/CNC_State.pdf}
 \label{fig:depexamp:a}
 }
-\subfigure[A {\tt distribute\_charge} task depends on the previous two.]{
+\subfigure[{\tt distribute\_charge} depends on {\tt calc\_new\_currents}.]{ 
+\includegraphics[scale=0.45]{figs/DC_state.pdf}
 \label{fig:depexamp:b}
 }
-\subfigure[An {\tt update\_voltages} task depends on the previous one.]{
+\subfigure[{\tt update\_voltages} depends on {\tt distribute\_charge}.]{
+\includegraphics[scale=0.45]{figs/UV_State.pdf}
 \label{fig:depexamp:c}
+}
+\subfigure[An {\tt update\_voltages} mapping.]{
+\includegraphics[scale=0.45]{figs/UV_Update.pdf}
+\label{fig:depexamp:d}
 }

### 2012-05-04 doc/sc2012/experiments.tex

@@
 a particular problem size as the number of nodes is varied.  Our results demonstrate
-excellent strong scaling behavior, with speedups of *??*x (compared to the single-GPU CUDA
-version) for the small problem on 48 GPUs and *??*x for the larger problem size on 96 GPUs.
+excellent strong scaling behavior, with speedups of 39.0X (compared to the single-GPU CUDA
+version) for the small problem on 48 GPUs and 62.5X for the larger problem size on 96 GPUs.
 
@@
 good scaling in a multi-core processor, but rule out any attempt to scale beyond a single node.  To extend the
-scaling further, our port divides uses region partitioning to divide the array into grids in a similar
+scaling further, our port uses region partitioning to divide the array into grids in a similar
 way, but handles the interaction between grids in a way that doesn't rely on access to uniform shared memory.

### 2012-05-04 doc/sc2012/legion.tex

@@
 to achieve good performance it is crucial that the programmer
-and the compiler be able to reason about {\em locality} (data resident
+and the programming system be able to reason about {\em locality} (data resident
 close to computation that uses it) and {\em independence} (computations
@@
 those regions, providing locality information.  Logical regions may be
-{\em partitioned} into (usually) disjoint {\em subregions}, providing
+{\em partitioned} into disjoint or aliased {\em subregions}, providing
 information for determining independence of computations.
@@
 Legion programs are written using logical regions, which do not commit
-to any particular placement of the data in the machine.  At runtime,
+to any particular layout of the data or placement of the data in the machine.  At runtime,
 each logical reagion has one or more {\em physical instances}.  Often
@@
   designed for visualization applications (10 nodes with 12 CPUs and 5 GPUs per node), and up to a third of the
-  Keeneland supercomputer (32 nodes, 12 CPUs and 3 GPUs per node).
+  Keeneland supercomputer (32 nodes, 12 CPUs and 3 GPUs per node) \cite{Keeneland}.
 %

### 2012-05-04 doc/sc2012/mapping.tex

@@
 To make writing Legion applications easier, we provide a default mapper implementation that
-can be used to get an application working quickly at a moderate performance level.  In addition,
-the default mapper serves as the base class for all custom mappers which makes implementing
-custom mappers easier as they only have to override the functions that need to be specialized.
-
-The default mapper employs a simple scheme for mapping tasks and regions.  When
+can be used to get an application working quickly at a moderate performance level.  The 
+default mapper employs a simple scheme for mapping tasks and regions.  When
 {\tt select\_initial\_processor} is invoked for the default mapper, it first checks for which kind of processors
@@
 mapping each region.  Note that this is a greedy algorithm that works well for the common case, but can cause some 
-regions to be pulled unnecessarily close to the processor and consuming precious fast memory space.
+regions to be pulled unnecessarily close to the processor and consume precious fast memory space.
 

### 2012-05-04 doc/sc2012/soop.tex
soop

@@
 indicates independence.  {\tt Same} is a dependence unless the two tasks
-use the same physical instance of the logical region.\footnote{For example, if two tasks wish to acess the region with atomic coherence, then if they are working on the same physical copy of the data atomicity can be guaranteed using, e.g., locking.
+use the same physical instance of the logical region.
+%\footnote{For example, if two tasks wish to acess the region with atomic coherence, then if they are working on the same physical copy of the data atomicity can be guaranteed using, e.g., locking.
 If they are working on different copies of the data then the only safe execution strategy for Legion is to serialize the two tasks.}
@@
 
-A key property of Legion, that dependence analysis is only needed
+A key property of Legion is that dependence analysis is only needed
 between sibling tasks:
@@
 exclusive mode, and reads in exclusive mode the node subregions {\tt
-pieces[0].rn\_pvt}, {\tt pieces[0].rn\_shr}, and {\tt
-pieces[0].rn\_ghost} (see line 39).  The second subtask must be
+pieces[0].rn\_pvt}, {\tt .rn\_shr}, and {\tt
+.rn\_ghost} (see line 39).  The second subtask must be
 checked against the first for any dependences; the second subtask uses
 {\tt pieces[1].rw\_pvt} (read/write exclusive) and {\tt
-pieces[1].rn\_pvt}, {\tt pieces[1].rn\_shr}, and {\tt pieces[1].rn\_ghost} (read-only exclusive).
+pieces[1].rn\_pvt}, {\tt .rn\_shr}, and {\tt .rn\_ghost} (read-only exclusive).
 It may be helpful to refer to the region tree for nodes in Figure~\ref{sfig:part_fig:tree} (recall that the wires region tree consists of a single disjoint partition).
-We reason about the dependences as follows:
+We give a few representative examples
+of reasoning about pairwise dependences (not all pairs are covered):
 \begin{enumerate}
 \item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rw\_pvt} do not overlap as they are subregions of a disjoint partition.
-\item {\tt pieces[0].rn\_pvt} and {\tt pieces[1].rn\_pvt} do not overlap for the same reason.
-\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_shr} do not overlap, also for the same reason.
+%\item {\tt pieces[0].rn\_pvt} and {\tt pieces[1].rn\_pvt} do not overlap for the same reason.
+%\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_shr} do not overlap, also for the same reason.
 \item {\tt pieces[0].rn\_ghost} may overlap with {\tt pieces[1].rn\_shr}, because they are in two different
   partitions of the same region.
-\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_ghost} may overlap because they are in two different
+%\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_ghost} may overlap because they are in two different
   partitions of the same region.
@@
 an aliased partition.
-\item All remaining pairs of regions (e.g., {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rn\_pvt})
+\item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rn\_pvt})
 do not overlap because they are in different region trees and have no common region ancestor.
 \end{enumerate}
-Cases 4-6 indicate a possible dependence, however in each case the regions are accessed with only read privileges.  Thus, the two subtasks are independent.
+Cases 2 and 3 indicate a possible dependence, however the regions are
+accessed with only read privileges. All other cases for these two
+subtasks are similar to one of the examples above; thus, the two
+subtasks are independent.
 
@@
 are in different region trees) or is a disjoint partition, then $r_1$
-and $r_2$ are guaranteed to be disjoint.  If $\lca{r_1}{r_2}$ is an
-aliased partition or a region (because $r_1$ and $r_2$ are in
-different partitions of the region, as in cases 4 and 5
-above), then $r_1$ and $r_2$ may not be disjoint.
+and $r_2$ are disjoint.  If $\lca{r_1}{r_2}$ is an
+aliased partition or a region, 
+%(because $r_1$ and $r_2$ are in
+%different partitions of the region, as in cases 4 and 5
+%above), 
+then $r_1$ and $r_2$ may not be disjoint.
 
-Consider subtasks {\tt distribute\_charge(pieces[0],dt)} and
+Consider tasks {\tt distribute\_charge(pieces[0],dt)} and
 {\tt update\_voltages[1]} in Listing~\ref{lst:code_ex}.  The former task
@@
 dependences involving $t'$, which
-this can result in false dependences.  However, the loss of
+can result in false dependences.  However, the loss of
 information can be shown to be small and only affects when $t$ can
-map; any information lost is recovered in the subsequent mapping
-phase.  Furthermore, the benefit is that it dramatically speeds up
+map; any information lost is recovered during mapping.
+Furthermore, the benefit is that it dramatically speeds up
 dependence analysis overall.

### 2012-05-04 doc/sc2012/experiments.tex
Adding the AMR section for the paper

@@
 \label{subsec:exp_amr}
+Our final application is based on the third heat equation example from the Berkeley Labs 
+BoxLib project \cite{BoxLib}.  This application is a three-level adaptive-mesh-refinement (AMR)
+code that computes a first order stencil on a 2D mesh of cells.  The simulation iterates for many time
+steps.  Updating the simulation for one time step consists of three phases.  In the first phase,
+the boundary cells around a box at a refined level linearly interpolate their values from the nearby cells at the
+next coarsest level.  The second phase computes the stencil computation on each cell in all levels.
+In the third phase, cells at a coarser level that have been refined are restricted to the average
+of the cells that they physically contain at the next finest level of refinement.
+
+Achieving high-performance on this application is particularly challenging for several reasons.  
+First, the application has a very high communication-to-computation ratio which for a fixed problem
+size begins as being memory bound, and with increasing node count becomes network bound as the
+perimeter-to-area ratio of cell grids increases.  Second, when chosing how to partition cells into grids, the
+programmer must consider the locality both between cells in the same level, as well as for cells
+across levels.  In the case of cross-level cell dependences, mapping decisions must be made at runtime as the
+location of refinements are determined based on application input.  Finally, this application has
+parallelism both between tasks running at the same level and tasks running across levels, leading
+to complicated input-dependent data dependences.
+
+BoxLib's implementation of this benchmark partitions cells within the same level into a number of grids
+based on the number of nodes in the machine and distributes one grid from each level to each node.  This
+optimizes for memory bandwidth and load balance, but does not take into account cross-level locality 
+that exists between grids from different levels of refinement.  Furthermore, BoxLib does not block grids
+into sub-grids to take advantage of intra-grid locality.
+
+Our Legion implementation performs two optimizations that allow us to achieve better performance than BoxLib.
+First, for each level of refinement we recursively partion the logical region of cells based on 
+the number of nodes in the machine and the size of the L3 cache and L2 caches.  Legion allows us to describe 
+locality at many levels of the memory hierarchy instead of just at the node-level.
+
+Our second optimization takes advantage of the cross-level locality present in the application.  We implemented
+an application specific mapper that inspects the arguments of tasks to discover the relationships between
+grids at different levels of refinement.  The mapper dynamically performs intersection tests between logical regions
+containing grids of different refinement levels to determine if they overlap or not.  
+If grids from different levels overlap, the mapper
+ensures that they are always placed on the same node in the machine to leverage cross-level locality.  The mapper
+memoizes the results of these intersection tests to ammortize their cost.  The mapper also dynamically
+load balances by distributing unconstrained grids from the coarsest level onto under-loaded nodes.
+
+We compared our Legion implementation against BoxLib on three different problem sizes with a fixed number
+of cells per level of refinement, but with randomly chosen refinements locations.  BoxLib also supports OpenMP
+and we took their best number from using 1, 2, 4, or 8 OpenMP threads per node.  Our Legion implementation
+always uses one thread per node to illustrate that in this application locality is significantly more important 
+than parallelism.  The results from our experiments on all three clusters can be seen in Figure \ref{fig:amr_total}.
+
+On just one node, blocking for caches using Legion achieves up to 2.6X speedup over BoxLib.  As the node count 
+increases, the ability of our mapper to leverage cross-level locality allows us to further increase
+our performance advantage to 5.4X over BoxLib by reducing the total communication costs.
+
+As the node count increased the AMR code becomes highly dependent on interconnect performance.  BoxLib performs much better
+on Keeneland than it does on our viz cluster due to the better interconnect.  This can also be seen as we scale
+to a larger number of nodes and BoxLib begins to catch up as can be seen in Figure \ref{fig:amr_keeneland}.  
+This occurs because our intra-level ghost-cell exchange algorithm relies on GASNet memory to exchange ghost cells which 
+involves a linear increase in network traffic with the number of nodes.  BoxLib instead uses direct
+node-to-node exchanges of ghost cells similar to our fluid application.  A future implementation of our AMR code 
+will employ a similar ghost cell exchange algorithm to improve scalability.
 
@@
 }
-\caption{Adaptive Mesh Refinement Results}
+\caption{Throuphput of Adaptive Mesh Refinement in Millions of Cells/s}
+\label{fig:amr_total}
 \end{figure}

### 2012-05-04 doc/sc2012/code_ex.tex
Updating the code example to fix a bug

@@
 void distribute_charge(CircuitPiece piece, float dt):
-        ROE(piece.rw_pvt), RdA(piece.rn_pvt, piece.rn_shr,piece.rn_ghost) {
+        ROE(piece.rw_pvt), RWE(piece.rn_pvt), RdA(piece.rn_shr,piece.rn_ghost) {
   foreach(w : piece.rw_pvt) {

### 2012-05-04 doc/sc2012/code_ex.tex
Reverting changes made earlier for dc state

--- 2012-04-27 doc/sc2012/code_ex.tex
@@
 void distribute_charge(CircuitPiece piece, float dt):
-        ROE(piece.rw_pvt), RdA(piece.rn_pvt, piece.rn_shr,piece.rn_ghost) {
+        ROE(piece.rw_pvt), RdA(piece.rn_pvt,piece.rn_shr,piece.rn_ghost) {
   foreach(w : piece.rw_pvt) {

### 2012-05-04 doc/sc2012/legion.tex
Some minor changes to the paper

@@
 to any particular layout of the data or placement of the data in the machine.  At runtime,
-each logical reagion has one or more {\em physical instances}.  Often
+each logical region has one or more {\em physical instances}.  Often
 logical regions correspond to a single physical instance, but it is also common and useful
 to have multiple physical copies of a logical region's contents in different parts of the machine 
-to duplicate and distribute read-only data, to keep an old copy of a region to break a 
+to duplicate read-only data, to keep an old copy of a region to break a 
 write-after-read dependence, or to allow independent, local reductions that are later combined.

### 2012-05-04 doc/sc2012/mapping.tex
Some minor changes to the paper

@@
 consists of ten function calls that the runtime will make to request direction from
-the programmer about how a program should be executed.  To aid in performing the 
+the mapper about how a program should be executed.  To aid in performing the 
 decisions made in these calls the mapper has access to a simple 

### 2012-05-04 doc/sc2012/soop.tex
Some minor changes to the paper

@@
 
-We can identify all mappoing dependences between sibling tasks and place tasks
+We can identify all mapping dependences between sibling tasks and place tasks
 in the correct place in the region tree in amortized time ${\cal O}(d)$ per region,
@@
 property established by stage 1 is that all
-tasks $t$ depends on map before $t$ maps and no task that depends on
+tasks which $t$ may depend on map before $t$ maps and no task that may depend on
 $t$ maps before $t$ finishes mapping.  Thus, the region tree at the
@@
 task $t$.  First, the mapper must ensure there is at least one {\em valid}
-physical instance of $r$.  If an instance of a subregion of $r$ has been written by a previous task,
+physical instance of $r$ for $t$ to use.  If an instance of a subregion of $r$ has been written by a previous task,
 for example, it is necessary to copy that subregion back to an instance of the parent $r$ so that $t$
@@
 
-Figure~\ref{fig:depexamp:c} shows part of the mapping of task {\tt update\_voltages[0]}; we focus
+Figure~\ref{fig:depexamp:d} shows part of the mapping of task {\tt update\_voltages[0]}; we focus
 only on the shared nodes.

### 2012-05-04 doc/sc2012/experiments.tex
Some edits to the experiments section

@@
 
-Nodes with multiple instances of multiple resources are a challenge for 
+Nodes with multiple instances of different resources are a challenge for 
 existing programming models that have a flat model of the system (e.g. MPI).
@@
 run on subsets of each machine ranging from the smallest (a single CPU core or GPU)
-to the largest or near-largest.  (Keeneland was an exception here - although it has
-115 nodes, we limited our runs to at most 32 nodes to get sufficient cluster time.)
+to the largest or near-largest (except Keeneland which has 115 nodes, but we limited 
+our runs to at most 32 nodes to get sufficient cluster time.)
 By looking at the performance of the same size problem over progressively larger
-machines, we are able to measure the ``strong scaling'' that Legion is able to
-achive.  By increasing the problem size as well, we also measure ``weak scaling''.
+machines, we are able to measure the strong scaling that Legion is able to
+achive.  By increasing the problem size as well, we also measure weak scaling.
 
@@
 
-The first experiment run was the distributed circuit simulation, described in 
-section \ref{sec:ex}, with all application code running on the CUDA-capable GPUs in
-each node.  The application kernels are written in CUDA, while the Legion runtime
-handles all the resource allocation, scheduling, and data movement.  In particular,
-Legion's ability to efficiently manage, access, and move the irregularly partitioned
+The first experiment we investigate is the distributed circuit simulation described in 
+Section \ref{sec:ex}.  The Legion runtime handles all of the resource allocation, 
+scheduling, and data movement across the cluster of GPUs.  In particular,  
+Legion's ability to efficiently move the irregularly partitioned
 shared data around the system while keeping the private nodes and wires resident in
-each GPU's framebuffer memory is critical in keeping the overhead low, a neccessity
-for good scalability.
+each GPU's framebuffer memory is critical in keeping the overhead low and achieving 
+good scalability.
 
@@
 time, the mapper queries the list of GPUs in the machine and identifies each GPU's
-framebuffer memory and ``zero-copy'' memory (i.e. the pinned memory that both the GPU and
-CPUs in the system can access directly).  Once the circuit is partitioned, the partitions
-are assigned a ``home'' GPU in round-robin fashion.  Every task related to that partition is
-then sent to that GPU, with no task stealing allowed.  (If the circuit is partitioned well,
-there should be minimal load imbalance between the pieces, and the rebalancing benefit of 
+framebuffer memory and {\em zero-copy} memory, the pinned memory that both the GPU and
+CPUs in the system can access directly.  Once the circuit is partitioned, the partitions
+are assigned a home GPU in round-robin fashion.  Every task related to that partition is
+then sent to the home GPU, with no task stealing allowed.  (If the circuit is partitioned well,
+there is minimal load imbalance between the pieces, and the rebalancing benefit of 
 moving a partition is unlikely to be worth the cost of moving all the 
@@
 Figure \ref{fig:ckt_speed} shows the performance of the Legion circuit simulation relative
-to a single-GPU implmementation hand-coded in CUDA.  Each line shows the scaling of
+to a hand-coded single-GPU implmementation written in CUDA.  Each line shows the scaling of
 a particular problem size as the number of nodes is varied.  Our results demonstrate
-excellent strong scaling behavior, with speedups of 39.0X (compared to the single-GPU CUDA
-version) for the small problem on 48 GPUs and 62.5X for the larger problem size on 96 GPUs.
+excellent strong scaling, with speedups of 39.0X for the small problem on 48 GPUs and 
+62.5X for the larger problem size on 96 GPUs.  On a single GPU, our Legion implementation
+was within 5\% of the performance of the hand-coded example.
 
@@
 runtime.  As the node count increases, the non-communication overhead stays relatively constant.
-Only the communication overhead grows, and it grows nearly linearly with the node count.
-A linear growth in communication cost as a simulation is split over multiple nodes is to
-be expected.
+As expected the communication overhead grow linearly with number of nodes.
 
@@
 Our second experiment is a port of the \emph{fluidanimate} benchmark from the PARSEC suite\cite{bienia11benchmarking},
-which does a particle-based simulation of a fluid.  Each particle interacts only with other nearby
-particles, so the benchmark divides the space in which the fluid
-can move into a three-dimensional array of cells such that the range of interaction is limited to just the
+which does a particle-based simulation of an incompressible fluid.  Each particle interacts only with other nearby
+particles. The benchmark divides the space in which the fluid can move into a three-dimensional array of cells 
+such that the range of interaction is limited to just the
 cells adjacent (including diagonals) to the one a particle resides in.  Originally designed to be a
-chip multiprocessor benchmark, the application divides the array into ``grids'' and assigns each one to 
-a thread.  Per particle locking is used to safely access particles in cells that lie
-on the edge of a grid.  This fine-grained locking along with the assumption of a shared address space lead to
-good scaling in a multi-core processor, but rule out any attempt to scale beyond a single node.  To extend the
+chip multiprocessor benchmark, the application divides the array into {\em grids} and assigns each one to 
+a thread.  Per cell locking is used to safely access particles in cells that lie
+on the edge of a grid.  This fine-grained locking scheme along with the assumption of a shared address space lead to
+good scaling in a multi-core processor, but rule out any attempt run beyond a single node.  To extend the
 scaling further, our port uses region partitioning to divide the array into grids in a similar
@@
 
-The mapper for the particle simulation is very simple - it just maps one grid's tasks onto each processor and
-maps all regions for that grid (both the internal the ghost cells) into the system memory used by that processor.
+The mapper for the particle simulation is very simple - it maps one grid's tasks onto each processor and
+maps all regions for that grid (both the internal and the ghost regions) into the system memory used by that processor.
 The exchange of the ghost cell data between processors is handled by the Legion runtime as a ghost cell region
@@
 neither the Legion runtime nor the the restructuring of the implementation to allow multi-node scaling impose
-any significant overhead.  In fact, it's possible that the use of explicit ghost cells rather than 
+any significant overhead.  It's possible that the use of explicit ghost cells rather than 
 fine-grained sharing of cache lines might be a net win.  At 8 threads and above, the performance begins to vary.
@@
 there are twice as many ghost cells as interior grid cells.  The larger problem sizes (2.4M and 19M
-particles) do much better, with scaling of up to 5.4x when going from 1 node up to 16.
+particles) do much better, with scaling of up to 5.4x when going from 1 node up to 16 because the
+communication-to-computation ratio is lower.
 
@@
 Achieving high-performance on this application is particularly challenging for several reasons.  
-First, the application has a very high communication-to-computation ratio which for a fixed problem
-size begins as being memory bound, and with increasing node count becomes network bound as the
+First, the application has a very high communication-to-computation ratio which, for a fixed problem
+size begins as being memory bound and with increasing node count becomes network bound as the
 perimeter-to-area ratio of cell grids increases.  Second, when chosing how to partition cells into grids, the
@@
 the number of nodes in the machine and the size of the L3 cache and L2 caches.  Legion allows us to describe 
-locality at many levels of the memory hierarchy instead of just at the node-level.
+locality for many levels of the memory hierarchy instead of just at the node-level.
 
@@
 grids at different levels of refinement.  The mapper dynamically performs intersection tests between logical regions
-containing grids of different refinement levels to determine if they overlap or not.  
+containing grids of different refinement levels to determine whether they overlap.  
 If grids from different levels overlap, the mapper

### 2012-05-04 doc/sc2012/legion.tex
first two sections

--- 2012-05-04 doc/sc2012/legion.tex
@@
 the penalty for mistakes in the placement of data or computation is usually very poor performance.  Thus,
-to achieve good performance it is crucial that the programmer
-and the compiler be able to reason about {\em locality} (data resident
+to achieve good performance the programmer
+and the programming system must reason about {\em locality} (data resident
 close to computation that uses it) and {\em independence} (computations
@@
 those regions, providing locality information.  Logical regions may be
-{\em partitioned} into (usually) disjoint {\em subregions}, providing
+{\em partitioned} into disjoint or aliased (overlapping) {\em subregions}, providing
 information for determining independence of computations.
@@
 
-Legion programs are written using logical regions, which do not commit
-to any particular placement of the data in the machine.  At runtime,
-each logical reagion has one or more {\em physical instances}.  Often
-logical regions correspond to a single physical instance, but it is also common and useful
-to have multiple physical copies of a logical region's contents in different parts of the machine 
-to duplicate and distribute read-only data, to keep an old copy of a region to break a 
-write-after-read dependence, or to allow independent, local reductions that are later combined.
+Logical regions do not commit to any particular layout of the data or
+placement in the machine.  At runtime, each logical region has one or
+more {\em physical instances} assigned to specific memories .  It is often useful to have multiple
+physical instances of a logical region (e.g., to duplicate read-only data, or to allow
+independent reductions that are later combined).
 
@@
 
-\item We present the results of experiments on three applications:
+\item We present results of experiments on three applications:
   fluid-flow on a regular grid, a three-level AMR code solving a
-  standard heat diffusion equation and a circuit simulation.
-  We run each of these four codes and best available reference
+  heat diffusion equation and a circuit simulation.
+  We compare each application with best available reference
   versions on three different clusters of multicore processors with
-  attached GPUs: a small cluster of our own (4 nodes with 12 CPUs and 2 GPUs per node), a medium-sized cluster
-  designed for visualization applications (10 nodes with 12 CPUs and 5 GPUs per node), and up to a third of the
-  Keeneland supercomputer (32 nodes, 12 CPUs and 3 GPUs per node).
+  GPUs, including the Keeneland supercomputer.
 %

### 2012-05-04 doc/sc2012/experiments.tex
spelling

@@
 machines, we are able to measure the strong scaling that Legion is able to
-achive.  By increasing the problem size as well, we also measure weak scaling.
+archive.  By increasing the problem size as well, we also measure weak scaling.
 
@@
 Figure \ref{fig:ckt_speed} shows the performance of the Legion circuit simulation relative
-to a hand-coded single-GPU implmementation written in CUDA.  Each line shows the scaling of
+to a hand-coded single-GPU implementation written in CUDA.  Each line shows the scaling of
 a particular problem size as the number of nodes is varied.  Our results demonstrate
@@
 size begins as being memory bound and with increasing node count becomes network bound as the
-perimeter-to-area ratio of cell grids increases.  Second, when chosing how to partition cells into grids, the
+perimeter-to-area ratio of cell grids increases.  Second, when choosing how to partition cells into grids, the
 programmer must consider the locality both between cells in the same level, as well as for cells
@@
 Our Legion implementation performs two optimizations that allow us to achieve better performance than BoxLib.
-First, for each level of refinement we recursively partion the logical region of cells based on 
+First, for each level of refinement we recursively partition the logical region of cells based on 
 the number of nodes in the machine and the size of the L3 cache and L2 caches.  Legion allows us to describe 
@@
 ensures that they are always placed on the same node in the machine to leverage cross-level locality.  The mapper
-memoizes the results of these intersection tests to ammortize their cost.  The mapper also dynamically
+memoizes the results of these intersection tests to amortize their cost.  The mapper also dynamically
 load balances by distributing unconstrained grids from the coarsest level onto under-loaded nodes.

### 2012-05-04 doc/sc2012/soop.tex
spelling

@@
 distributes tasks to other processors (Section~\ref{sec:dist}).  Once
-a task has been assigned a processor and physical instancess it is
+a task has been assigned a processor and physical instances it is
 issued for deferred execution (Section~\ref{sec:exec}).  The final
@@
 \item If $m$ is a disjoint partition and $m'$ is the child on the
-path to $r'$, then no tasks in a subtree of $m$ other than $m'$ can conflict wtih $t'$; we simply move
+path to $r'$, then no tasks in a subtree of $m$ other than $m'$ can conflict with $t'$; we simply move
 on to $m'$.

### 2012-05-04 doc/sc2012/experiments.tex
fixed systems table

--- 2012-05-04 doc/sc2012/experiments.tex
@@
 The efficiency and scalability of Legion was examined by measuring the performance
-of three applications on three clusters.  Table \ref{table:systems} summarizes 
+of three applications on three clusters.  The table in Figure~\ref{fig:systems} summarizes 
 the three clusters that were used.
 
-\begin{tabular}{lccccc}
- & Nodes & CPUs/Node & GPUs/Node & DRAM/Node & Infiniband \\
-sapling & 4 & 2x Xeon 5680 (HT enabled) & 2x Tesla C2070 & 48 GB & 2x QDR \\
-viz & 10 & 2x Xeon 5680 (HT disabled) & 5x Quadro Q5000 & 24 GB & QDR \\
-keeneland & 32* & 2x Xeon 5660 (HT disabled) & 3x Tesla M2090 & 24 GB & QDR 
+\begin{figure}
+{\footnotesize
+\begin{tabular}{l|ccc}
+Cluster & Sapling & Viz & Keeneland \\
+\midrule
+Nodes   &   4     &  10 &  32 (120) \\
+CPUs/Node & 2x Xeon 5680 & 2x Xeon 5680 & 2x Xeon 5660 \\
+HyperThreading & on & off & off \\
+GPUs/Node & 2x Tesla C2070 & 5x Quadro Q5000 & 3x Tesla M2090 \\
+DRAM/Node & 48 GB & 24 GB & 24 GB \\
+Infiniband & 2x QDR & QDR & QDR \\
 \end{tabular}
+}
+\label{fig:systems}
+\caption{Configurations of Systems Used for Experiments}
+\end{figure}
 

### 2012-05-04 doc/sc2012/circuit.tex
figs

@@
 wires and nodes where wires meet.  
-The simulation repeatedly calculates 
-currents, distributes charges, and updates voltages for as many
-time steps as the simulation demands.
+At each time step the simulation calculates 
+currents, distributes charges, and updates voltages.
 
 The key decisions in a Legion program are how data is grouped into
-regions and how those regions are {\em partitioned} into {\em
-subregions}.  The goal is to pick an organization that makes explicit
-in the algorithms which computations are independent.  A {\tt Circuit}
-has two regions: undifferentiated collections of nodes and wires (line
-3 of Listing~\ref{lst:code_ex}). An efficient parallel
+regions and how  regions are {\em partitioned} into {\em
+subregions}.  
+%
+%FIXME
+% the following sentence oversimplifies; it's one of the goals, and regions tell us
+% useful things about dependent computations as well
+%
+The goal is to pick an organization that makes explicit
+which computations are independent.  
+A {\tt Circuit}
+has two regions: collections of nodes and wires (line
+3 of Listing~\ref{lst:code_ex}).\footnote{Note that pointers to region values declare the region to which they point.  For
+example, the definition of {\tt Wire} (line 2) is parametrized on the region
+{\tt rn} to which the {\tt Node} pointers in fields {\tt in\_nodes}
+and {\tt out\_nodes} point.}
+An efficient parallel
 implementation breaks this unstructured graph into pieces that can be
-processed (mostly) independently. There is always some computation
-across the wires between pieces, but an appropriate
+processed (mostly) independently. An appropriate
 region organization makes explicit which nodes and wires are
-involved in inter-piece computation and which are involved in only
-intra-piece computation.
+involved in intra-piece computation and, where wires connect different pieces,
+which are involved in inter-piece computation.
 
@@
 Figure~\ref{sfig:part_fig:pvs} shows how the nodes
-in a small graph might be split into three pieces.  Blue nodes are
-attached by wires only to nodes in the same piece and are 
-{\em private} to the piece.  Orange nodes on the boundary of a
-piece are {\em shared} with (connected to) other pieces.
+in a small graph might be split into three pieces.  Blue nodes,
+attached by wires only to nodes in the same piece, are 
+{\em private} to the piece.  Orange nodes, on the boundary of a
+piece, are {\em shared} with (connected to) other pieces.
 In the simulation, computations on the
-private nodes of different pieces will be independent, while
-computations on the shared nodes will require communication.  To make
+private nodes of different pieces are independent, while
+computations on the shared nodes require communication.  To make
 this explicit in the program, we partition the nodes region into
@@
 
-The private nodes and the shared nodes are partitioned again into the private and shared nodes for each circuit piece (lines 21
+The private and shared nodes are partitioned again into private and shared nodes for each circuit piece (lines 21
 and 23);  both partitions are disjoint.  There is another useful partition of the shared nodes:
@@
 The private, shared, and ghost node subregions
-for the upper-left patch of the example graph are shown in
+for the upper-left piece of the example graph are shown in
 Figures~\ref{sfig:part_fig:p_i}, \ref{sfig:part_fig:s_i}, and
@@
 %
-Line 9 declares the main simulator function, which takes a {\tt
-Circuit} to be simulated.  This function specifies the regions it 
+Line 9 declares the main simulator function, which specifies the regions it 
 accesses and the privileges and coherence it requires of those regions.
-In this case, the {\tt RWE} annotation specifies that the function
-accesses the regions {\tt c.r\_all\_nodes} and {\tt c.r\_all\_wires}
-with read-write privileges and {\em exclusive} coherence (i.e., no other
+The {\tt RWE} annotation specifies that
+the regions {\tt c.r\_all\_nodes} and {\tt c.r\_all\_wires}
+are accessed with read-write privileges and {\em exclusive} coherence (i.e., no other
 task can access these two regions concurrently or be reordered around this
-task if they use either of the regions).  The simulation 
+task if they use either region).  The simulation 
 reads and writes all nodes and wires, and it must be done
@@
 parallel execution ({\tt spawn} simply indicates a task call and does
-not require parallel execution) nor is there synchronization required
+not require parallel execution) nor is there synchronization 
 between the passes.  Which tasks can be run in
@@
 lines 32-34 are {\em subtasks} of the main {\tt simulate\_circuit}
-task. A subtask can only take a region argument that is a region
-argument of the parent task or a subregion of a region argument of the
-parent task; furthermore, the subtask can only have permissions on the
-region that are compatible with the parent's permissions.  
+task. A subtask can only access regions (or subregions) that its parent task
+could access; furthermore, the subtask can only have permissions on a
+region compatible with the parent's permissions.  
 %This
@@
 The  {\tt calc\_new\_currents} task reads and writes the wires subregion 
-and reads the private, shared, and ghost node regions for its piece.
-The {\tt distribute\_charge} subtask is a different phase that reads the piece's 
+and reads the private, shared, and ghost node subregions for its piece.
+The {\tt distribute\_charge} task reads the piece's 
 wires subregion and updates all nodes those wires touch.  However,
-rather than requesting the ability to read and write the nodes (which would
-require serialization of these tasks for correctness), the task declares that
-it will use reorderable {\em reduction} operations and that the coherence requirement
-is {\em atomic} rather than exclusive access. The final task is 
-{\tt update\_voltages}, which writes the shared and private nodes for a piece
+rather than using read/write permission for the nodes (which would
+serialize these tasks for correctness), the task
+uses reorderable reduction operations and 
+atomic rather than exclusive access. The final task 
+{\tt update\_voltages} writes the shared and private nodes for a piece
 and reads the results of the previous task's reductions.
@@
 
-The two basic data types in the circuit simulation are {\tt Node}s and
-{\tt Wire}s declared on lines 1 and 2 of Listing~\ref{lst:code_ex}.
-As is standard in region-based systems, allocating a {\tt Node}, {\tt
-  Wire}, or any value in the heap requires naming the region
-in which the value is allocated.  For example, {\tt new Node@r} returns
-a reference to a new value of {\tt Node} type in region {\tt r}.  
+%The two basic data types in the circuit simulation are {\tt Node}s and
+%{\tt Wire}s declared on lines 1 and 2 of Listing~\ref{lst:code_ex}.
+%As is standard in region-based systems, allocating a {\tt Node}, {\tt
+%  Wire}, or any value in the heap requires naming the region
+%in which the value is allocated.  For example, {\tt new Node@r} returns
+%a reference to a new value of {\tt Node} type in region {\tt r}.  
 Listing~\ref{lst:code_ex} illustrates one way of constructing partitioned
-data structures in Legion: populate a region with data (using {\tt new}, note the example
+data structures in Legion: populate a region with data (the example
 makes use of whatever data has been allocated by the caller of {\tt simulate\_circuit})
@@
 
-Pointers to region values declare the region to which they point.  For
-example, the definition of {\tt Wire} is parameterized on the region
-{\tt rn} to which the {\tt Node} pointers in fields {\tt in\_nodes}
-and {\tt out\_nodes} point.
 % which guarantees that the two endpoints of
 %a wire are always in the same region.
-Regions are first-class values in Legion and can be stored in heap
-data structures.  For example, a {\tt Circuit} (defined on line 3)
-holds the regions for all nodes and wires of the circuit and a {\tt
-  CircuitPiece} (defined on line 4) hold pointers to the private,
-shared, and ghost node regions as well as the private wires regions of its piece of
-the circuit.
+%Regions are first-class values in Legion and can be stored in heap
+%data structures.  For example, a {\tt Circuit} (defined on line 3)
+%holds the regions for all nodes and wires of the circuit and a {\tt
+%  CircuitPiece} (defined on line 4) hold pointers to the private,
+%shared, and ghost node regions as well as the private wires regions of its piece of
+%the circuit.

### 2012-05-04 doc/sc2012/mapping.tex

@@
 In this section we describe the general mapping interface (Section \ref{sec:mapinterface}),
-our base implmentation of this interface (Section \ref{sec:defmapper}), and the benefits
+our base implementation of this interface (Section \ref{sec:defmapper}), and the benefits
 of creating custom mappers (Section \ref{sec:custommap}).
@@
 runtime will query the mapper for a prioritized list of memories where a physical instance for this
-task should be placed.  To aid the mapping decision the runtime provides a list of the current valid physical intances
+task should be placed.  To aid the mapping decision the runtime provides a list of the current valid physical instances
 of the logical region.  The mapper returns a list of memories in which the runtime
@@
 found, it attempts to allocate a physical instance and then issue the necessary copies
-to retrive the valid data.  If both options fail, it moves on to the next memory in the list.
+to retrieve the valid data.  If both options fail, it moves on to the next memory in the list.
 \end{itemize}

### 2012-05-04 doc/sc2012/soop.tex
soop

@@
 
-Legion uses a {\em  software out-of-order processor}, or SOOP, to schedule tasks.  The SOOP 
-is pipelined, distributed and extracts nested parallelism from subtasks.
+Legion uses a {\em  software out-of-order processor}, or SOOP, to schedule tasks.  A SOOP 
+dynamically schedules a stream of tasks in a manner akin to an out-of-order processor scheduling a stream of instructions:
+just as an instruction scheduler is constrained by register dependences,
+the SOOP is constrained by region dependences.
+The SOOP  is pipelined, distributed and extracts nested parallelism from subtasks.
 There are two major challenges in implementing an efficient task scheduler:
 \begin{itemize}
-\item  For correctness, Legion must preserve data dependences between tasks.
+\item  For correctness, Legion must preserve data dependences between tasks, which is non-trivial
+because the same data may be in multiple different regions or subregions.
 
@@
 To solve the second problem Legion uses a {\em deferred execution model} that decouples the issuing
-of operations from when operations are performed.  An issued operation waits for other operations on
+of operations from when operations are performed.  Using a low-level runtime event system (which we do
+not describe further in this paper), an issued operation waits for other operations on
 which it is dependent to complete before executing, but a waiting operation does not cause the SOOP
@@
 
-To solve the first problem, detecting and enforcing data dependences
-efficiently at run time, the SOOP considers, for each task $t$, the
-the regions $t$ uses and ensures that $t$ waits on any
-earlier task whose region uses may conflict with $t$'s.  There is a
+To dynamically detect and enforce data dependences,
+the SOOP considers, for each task $t$, the
+the regions $t$ uses and ensures $t$ waits on any
+earlier task whose region uses conflict with $t$'s.  There is a
 difficulty, however.  The programmer's regions are logical; a region
 names a set of objects, but does not say where those objects
-are located.  To ensure dependences are satisfied the SOOP must also {\em map} $t$: assign an
-appropriate physical instance to each of $t$'s logical region.  Thus, dependence
+are in memory.  To ensure dependences are satisfied the SOOP must also {\em map} $t$: assign an
+appropriate physical instance to each of $t$'s logical regions.  Thus, dependence
 analysis requires knowing both which tasks $t$ depends on and how those
@@
 (Section~\ref{sec:dep}), which does not give enough information to map
-$t$, but does identify other tasks that must map before $t$ can map.
+$t$, but does identify all tasks that must map before $t$ can map.
 The third SOOP stage maps $t$ by carrying out a more refined analysis
@@
 
-When a Legion task runs, it may spawn new subtasks to be executed.  
-These subtasks are {\em children} of the {\em parent} task and {\em siblings} of each other.  
-A task always executes on a single processor, but
-its subtasks may execute elsewhere.
+%When a Legion task runs, it may spawn new subtasks to be executed.  
+%These subtasks are {\em children} of the {\em parent} task and {\em siblings} of each other.  
+%A task always executes on a single processor, but
+%its subtasks may execute elsewhere.
 
 Each processor in the system runs an instance of the SOOP.
-When a parent task spawns a child task, the child is {\em registered}
+When a {\em parent} task spawns a {\em child} task, the child is {\em registered}
 with the SOOP on the parent's processor;
@@
 logical regions accessed.  For each logical region used by $t'$ that
-may overlap with a logical region used by $t$, the privileges and
+may {\em alias} (may share data with) a logical region used by $t$, the privileges and
 coherence modes are compared to determine whether a dependence exists.
@@
 allow other tasks using the region to execute at the same time and differ
-only in what data must be observed.  With simultaneous coherence, a task must 
+only in what updates must be observed.  With simultaneous coherence, a task must 
 see all updates to the logical region made by other tasks operating on the same region 
@@
 Recall from Section~\ref{sec:ex} that subtasks only access subregions of their parent task's region arguments (and with fewer
-privileges).  Thus if the regions used by $t_1$ and $t_2$ do not overlap, the regions used by any subtasks of $t_1$ cannot
-overlap with the regions used by any subtask of $t_2$.  If the regions used by $t_1$ and $t_2$ do overlap
+privileges).  Thus if the regions used by $t_1$ and $t_2$ do not alias, the regions used by any subtasks of $t_1$ cannot
+alias the regions used by any subtask of $t_2$.  If the regions used by $t_1$ and $t_2$ alias
 but there is no dependence because the regions are used with simultaneous or relaxed coherence, then by definition there is no dependence between the subtasks.
@@
 \begin{enumerate}
-\item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rw\_pvt} do not overlap as they are subregions of a disjoint partition.
-%\item {\tt pieces[0].rn\_pvt} and {\tt pieces[1].rn\_pvt} do not overlap for the same reason.
-%\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_shr} do not overlap, also for the same reason.
-\item {\tt pieces[0].rn\_ghost} may overlap with {\tt pieces[1].rn\_shr}, because they are in two different
+\item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rw\_pvt} do not alias as they are subregions of a disjoint partition.
+%\item {\tt pieces[0].rn\_pvt} and {\tt pieces[1].rn\_pvt} do not alias for the same reason.
+%\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_shr} do not alias, also for the same reason.
+\item {\tt pieces[0].rn\_ghost} aliases {\tt pieces[1].rn\_shr}, because they are in two different
   partitions of the same region.
-%\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_ghost} may overlap because they are in two different
+%\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_ghost} alias because they are in two different
   partitions of the same region.
-\item {\tt pieces[0].rn\_ghost} may overlap with {\tt pieces[1].rn\_ghost} because they are in different subregions of
+\item {\tt pieces[0].rn\_ghost} aliases {\tt pieces[1].rn\_ghost} because they are in different subregions of
 an aliased partition.
 \item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rn\_pvt})
-do not overlap because they are in different region trees and have no common region ancestor.
+do not alias because they are in different region trees and have no common region ancestor.
 \end{enumerate}
@@
 
-A closer look at this example shows that whether there is overlap
+A closer look at this example shows that whether there is aliasing
 between two regions $r_1$ and $r_2$ can be determined by examining
@@
 by a task $t''$ registered after $t'$ and $t''$ depends on $t'$
-because of overlap between $r''$ and $r'$.  Then $t'$ is listed on the
+because of aliasing between $r''$ and $r'$.  Then $t'$ is listed on the
 region that is the least common ancestor of the set $R \cup \{r'\}$.
@@
 Next, in Figure~\ref{fig:depexamp:b}, the task {\tt distribute\_charge(pieces[0],dt)} ($d_0$)
-has dependences with both $c_0$ and $c_1$ caused by possible overlap between the pieces of the
+has dependences with both $c_0$ and $c_1$ caused by aliasing between the pieces of the
 shared node partition and the ghost node partition, so both $c_0$ and $c_1$ are moved to
@@
 
-\item If $t'$ has read/write privileges for $r'$, and $r, r'$ may
-  overlap (i.e., $\lca{r}{r'}$ is a region or aliased
-  partition), then $s'$ must be copied to an instance $s''$ of
+\item If $t'$ has read/write privileges for $r'$, and $r$ aliases $r'$,
+   then $s'$ must be copied to an instance $s''$ of
   $\lca{r}{r'}$ and a fresh instance of $r'$ created from $s''$.
@@
   reordered, combining the instances of $r$ and $r'$ (if they
-  overlap) can be deferred to a later consumer of the data (see case
+  alias) can be deferred to a later consumer of the data (see case
   4).
@@
   or reduce privileges with a different operator than $t'$
-  for $r$, and $r$ and $r'$ may overlap, then $s'$ must be
+  for $r$, and $r$ aliases $r'$, then $s'$ must be
   reduced (using $t'$'s reduction operator) to an instance $s''$ of

### 2012-05-04 doc/sc2012/soop.tex
soop.tex

@@
 
-A key property of Legion is that dependence analysis is only needed
-between sibling tasks:
+A key property is that dependence analysis is not needed between arbitrary pairs of tasks.
+In fact, it suffices to check only  {\em siblings} (children with the same parent) for dependences.
 \begin{observation}
@@
 \end{observation}
-Recall from Section~\ref{sec:ex} that subtasks only access subregions of their parent task's region arguments (and with fewer
-privileges).  Thus if the regions used by $t_1$ and $t_2$ do not alias, the regions used by any subtasks of $t_1$ cannot
-alias the regions used by any subtask of $t_2$.  If the regions used by $t_1$ and $t_2$ alias
-but there is no dependence because the regions are used with simultaneous or relaxed coherence, then by definition there is no dependence between the subtasks.
+Recall from Section~\ref{sec:ex} that subtasks only access regions (or subregions) that their parent accesses.  
+Thus if the regions used by $t_1$ and $t_2$ do not alias, the regions used by any subtasks of $t_1$ cannot
+alias the regions used by any subtask of $t_2$.  If regions of $t_1$ and $t_2$ alias
+but there is no dependence because the regions have simultaneous or relaxed coherence, then by definition there is no dependence between the subtasks.
 
-Consider the first two subtasks spawned on line 32 of the circuit
-simulation in Listing~\ref{lst:code_ex}, {\tt
+Consider the first two subtasks spawned on line 32 
+in Listing~\ref{lst:code_ex}, {\tt
 calc\_new\_currents(pieces[0])} and {\tt
@@
 We give a few representative examples
-of reasoning about pairwise dependences (not all pairs are covered):
+of reasoning about pairwise aliasing (not all pairs are covered):
 \begin{enumerate}
@@
 %\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_shr} do not alias, also for the same reason.
-\item {\tt pieces[0].rn\_ghost} aliases {\tt pieces[1].rn\_shr}, because they are in two different
+\item {\tt pieces[0].rn\_ghost} and {\tt pieces[1].rn\_shr} alias as they are in different
   partitions of the same region.
 %\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_ghost} alias because they are in two different
-  partitions of the same region.
-\item {\tt pieces[0].rn\_ghost} aliases {\tt pieces[1].rn\_ghost} because they are in different subregions of
+%  partitions of the same region.
+\item {\tt pieces[0].rn\_ghost} and {\tt pieces[1].rn\_ghost} alias because they are in different subregions of
 an aliased partition.
-\item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rn\_pvt})
-do not alias because they are in different region trees and have no common region ancestor.
+\item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rn\_pvt}
+do not alias because they are in different region trees.
 \end{enumerate}
@@
 
-A closer look at this example shows that whether there is aliasing
-between two regions $r_1$ and $r_2$ can be determined by examining
+A closer look at this example shows that whether 
+$r_1$ aliases $r_2$ can be determined by examining
 their least common ancestor $\lca{r_1}{r_2}$ in the region tree.  
-There are four cases.  If $\lca{r_1}{r_2}$ either does not exist (the regions
+If $\lca{r_1}{r_2}$ either does not exist (the regions
 are in different region trees) or is a disjoint partition, then $r_1$
@@
 %above), 
-then $r_1$ and $r_2$ may not be disjoint.
+then $r_1$ and $r_2$ alias.
 
@@
 The least common ancestor is the region of all shared nodes {\tt p\_nodes\_pvs[1]},
-so these two regions are not disjoint.  Since both tasks write their respective subregions,
+so these two regions alias.  Since both tasks write their respective subregions,
 there is a dependence and the {\tt update\_voltages} task can only map after the
@@
 
-We can now describe the algorithm for the dependence analysis phase of
-the SOOP.  The SOOP maintains a local region forest for each task $t$,
+We can now describe the algorithm for mapping dependence analysis.
+The SOOP maintains a local region forest for each task $t$,
 the roots of which are $t$'s region arguments and including all
@@
 region that is the least common ancestor of the set $R \cup \{r'\}$.
-Placing $t'$ on an ancestor of $r'$ has the effect of coarsening the
+Placing $t'$ on an ancestor of $r'$ coarsens the
 dependences involving $t'$, which
@@
 in the correct place in the region tree in amortized time ${\cal O}(d)$ per region,
-where $d$ is the depth of the region tree.  For region $r'$ an argument of task $t'$, the analysis
-walks the unique path in the region forest from a root to $r'$.  For each node $m$
+where $d$ is the depth of the region forest.  For region $r'$ used by task $t'$, the analysis
+walks the unique path in the region forest from a root to $r'$; task $t'$ depends on all tasks
+on this path when the walk completes.  For each node $m$
 on the walk from the root to $r'$ the following actions are taken:
@@
 \subfigure[Two {\tt calc\_new\_currents} tasks.]{
-\includegraphics[scale=0.45]{figs/CNC_State.pdf}
+\includegraphics[scale=0.46]{figs/CNC_State.pdf}
 \label{fig:depexamp:a}
@@
 \subfigure[{\tt distribute\_charge} depends on {\tt calc\_new\_currents}.]{ 
-\includegraphics[scale=0.45]{figs/DC_state.pdf}
+\includegraphics[scale=0.46]{figs/DC_state.pdf}
 \label{fig:depexamp:b}
@@
 \subfigure[{\tt update\_voltages} depends on {\tt distribute\_charge}.]{
-\includegraphics[scale=0.45]{figs/UV_State.pdf}
+\includegraphics[scale=0.46]{figs/UV_State.pdf}
 \label{fig:depexamp:c}
@@
 \subfigure[An {\tt update\_voltages} mapping.]{
-\includegraphics[scale=0.45]{figs/UV_Update.pdf}
+\includegraphics[scale=0.46]{figs/UV_Update.pdf}
 \label{fig:depexamp:d}
@@
 to move dependent tasks to the least common ancestor on the path.
-Clearly the walk to $r'$ is bounded by the depth of the region tree.
+The walk to $r'$ is bounded by the depth of the region tree.
 The off-path work can be made efficient by maintaining at each node a
-record of which subtrees have tasks in them and with what privileges,
+record of the {\em mode} of each subtree (whether the subtree has any tasks and with what privileges),
 enabling the off-path search to traverse directly and only to
@@
 is moved back up the region tree.  Thus, for each region argument a
-task is initially placed at some depth in the tree and subsequently
-only moved up towards the root, so the overall work per region is (amortized)
+task is initially placed at some depth in the forest and subsequently
+only moved up towards a root, so the overall work per region is (amortized)
 ${\cal O}(d)$.  
@@
 shared node partition and the ghost node partition, so both $c_0$ and $c_1$ are moved to
-the {\tt p\_nodes\_pvs[1]} region.  Finally, the task {\tt update\_voltages[1]} ($u_1$) causes
+the {\tt p\_nodes\_pvs[1]} region.  Finally, the task {\tt update\_voltages[1]} causes
 $d_0$ to also be moved to {\tt p\_nodes\_pvs[1]} for essentially the same reason; the final

### 2012-05-04 doc/sc2012/soop.tex
soop

@@
 dependences are satisfied, $t$ is ready to be mapped and is placed
-in the mapping queue.
-
-Once in the mapping queue, SOOPs on other processors in the system may
+in the SOOP's mapping queue. At this point SOOPs on other processors may
 ask to steal $t$ from its home SOOP.  The home SOOP may decline the
@@
 to execute $t$ on the local processor or send it to another processor.
-
 Thus, task distribution to processors in Legion is both ``pull'' (stealing) and ``push''.
@@
 
-The region tree is also used to hold mapping information.  Each logical region $r$ maintains a
-list of its physical instances and which sibling tasks are using which
+There are two steps to mapping a region $r$ used by
+task $t$.  First, the mapper must ensure there is at least one {\em valid}
+physical instance of $r$ for $t$ to use.  Not all intances have up-to-date
+data at all times.  For example, if an instance of a subregion of $r$ has been written by a previous task,
+it is necessary to copy that subregion back to an instance of $r$ so that $t$
+sees the correct data. In the second step, the mapping interface selects one of the
+valid physical instances of $r$ for $t$ to use or creates a new one.  
+We focus on the first step; the mapping interface is discussed further in Section~\ref{sec:mapping}.
+
+Each logical region in the region tree maintains a
+list of its physical instances and which sibling tasks are using those
 instances and with what permissions.  The important correctness
 property established by stage 1 is that all
-tasks which $t$ may depend on map before $t$ maps and no task that may depend on
-$t$ maps before $t$ finishes mapping.  Thus, the region tree at the
-moment $t$ maps contains a consistent view of what the state of memory
+tasks $t$ depends on map before $t$ maps and no task that depends on
+$t$ maps before $t$ finishes mapping.  Thus, the region tree when
+$t$ maps contains a consistent view of what the state of physical instances
 will be after $t$'s dependence predecessors execute.
-
-There are two steps to mapping a region $r$ used by
-task $t$.  First, the mapper must ensure there is at least one {\em valid}
-physical instance of $r$ for $t$ to use.  If an instance of a subregion of $r$ has been written by a previous task,
-for example, it is necessary to copy that subregion back to an instance of the parent $r$ so that $t$
-sees the correct data. In the second step, the mapping interface selects one of the
-valid physical instances of $r$ or creates a new one.  
-We focus here on the first step; the mapping interface is discussed further in Section~\ref{sec:mapping}.
 
@@
 
-Detecting or creating a valid instance of $r$ is, at its core, a dependence analysis on
+Detecting or creating a valid instance of $r$ is, at its core, dependence analysis on
 physical instances and thus similar to the mapping dependence analysis on logical
-regions described in Section~\ref{sec:dep}.  Consider any task $t'$ using
+regions in Section~\ref{sec:dep}.  Consider any task $t'$ using
 physical instance $s'$ of logical region $r'$ in the region tree.  There are four cases.
@@
    then $s'$ must be copied to an instance $s''$ of
-  $\lca{r}{r'}$ and a fresh instance of $r'$ created from $s''$.
+  $\lca{r}{r'}$ and a fresh instance of $r$ created from $s''$.
 %  (if either $r = \lca{r}{r'}$ or $r' = \lca{r}{r'}$ one of the two
 %  copies is omitted)
-Instance $s'$ is removed from the region tree
-  but not deallocated (see Section~\ref{sec:clean}).
+Instance $s'$ is removed from the region tree.
 
@@
   reduced (using $t'$'s reduction operator) to an instance $s''$ of
-  $\lca{r}{r'}$ and then a fresh instance of $r'$ created from $s''$.
+  $\lca{r}{r'}$ and then a fresh instance of $r$ created from $s''$.
 %  (If $r' = \lca{r}{r'}$ the reduction is unneeded, and if $r =
@@
 Instance $s'$ is removed from the region
-  tree but not deallocated.
+  tree.
 \end{enumerate}
@@
 root ancestor in the region forest to $r$, along the way exploring
-off-path subtrees to find region instances satisfying cases 3 and
+off-path subtrees to find region instances satisfying cases 2 and
 4. The details and analysis are similar to the implementation of dependence
 analysis in Section~\ref{sec:dep}; the amortized work per region mapped
-is proportional to the depth of the region tree.
+is proportional to the depth of the region forest.
 
@@
 illustrated, must be reduced out to the all-shared node region in the
-global (GASNet) memory and then a new instance of the shared
+global memory (which is managed by GASNet) and then a new instance of the shared
 subregions created with the updated information before the {\tt
@@
 
-Once task $t$ is done executing its state can be reclaimed Dependence
+Once task $t$ is done executing its state is reclaimed. Dependence
 and mapping information is removed from the region tree (it can

### 2012-05-04 doc/sc2012/legion.tex
spelling
legion
Adding the citation to Keeneland back into the introduction

@@
   heat diffusion equation and a circuit simulation.
-  We compare each application with best available reference
+  We compare each application with the best reference
   versions on three different clusters of multicore processors with
-  GPUs, including the Keeneland supercomputer.
+  GPUs, including the Keeneland supercomputer \cite{Keeneland}.
 %
@@
 \input{related}
-\input{conclusion}
+%\input{conclusion}
 

### 2012-05-04 doc/sc2012/experiments.tex
Small fixes to the experiments and soop sections

@@
 }
-\label{fig:systems}
-\caption{Configurations of Systems Used for Experiments}
+
+\caption{Configurations of Systems Used for Experiments \label{fig:systems}}
 \end{figure}
@@
 The first experiment we investigate is the distributed circuit simulation described in 
-Section \ref{sec:ex}.  The Legion runtime handles all of the resource allocation, 
+Section~\ref{sec:ex}.  The Legion runtime handles all of the resource allocation, 
 scheduling, and data movement across the cluster of GPUs.  In particular,  
@@
 private data for a piece from one GPU to another.)  The regions for a task are mapped as 
-shown in figure \ref{fig:gpumapping}.
+shown in Figure~\ref{fig:gpumapping}.
 
-Figure \ref{fig:ckt_speed} shows the performance of the Legion circuit simulation relative
+Figure~\ref{fig:ckt_speed} shows the performance of the Legion circuit simulation relative
 to a hand-coded single-GPU implementation written in CUDA.  Each line shows the scaling of
@@
 
-Figure \ref{fig:ckt_overhead} shows the fraction of the overall simulation time (summed over
+Figure~\ref{fig:ckt_overhead} shows the fraction of the overall simulation time (summed over
 all nodes) spent in the application kernels compared to the various pieces of the Legion
@@
 \includegraphics[scale=0.4]{figs/circuit_speedups.pdf}
+\makebox[0pt][r]{
+\raisebox{0.25 in}{
+\includegraphics[scale=0.18]{figs/circuit_speedups_zoom.pdf}
+}
+}
 \label{fig:ckt_speed}
@@
 
-Figure \ref{fig:fluid_single} compares the performance of the Legion implementation against the PARSEC version,
+Figure~\ref{fig:fluid_single} compares the performance of the Legion implementation against the PARSEC version,
 using a relatively small problem (300K particles on a 15x21x15 array of cells).  Between 1 and 4 threads, the
@@
 To measure scaling beyond a single node, three different problem sizes were run for each of the three systems,
- with the results plotted in figure
-\ref{fig:fluid_multi}.  For the smallest problem (300K particles), we observe a 20\% speedup when going to 1
+ with the results plotted in Figure~\ref{fig:fluid_multi}.  
+ For the smallest problem (300K particles), we observe a 20\% speedup when going to 1
 node to 2 (a total of 16 threads), but begin slowing down beyond that due to communication overhead - at 4 nodes
@@
 always uses one thread per node to illustrate that in this application locality is significantly more important 
-than parallelism.  The results from our experiments on all three clusters can be seen in Figure \ref{fig:amr_total}.
+than parallelism.  The results from our experiments on all three clusters can be seen in Figure~\ref{fig:amr_total}.
 
@@
 on Keeneland than it does on our viz cluster due to the better interconnect.  This can also be seen as we scale
-to a larger number of nodes and BoxLib begins to catch up as can be seen in Figure \ref{fig:amr_keeneland}.  
+to a larger number of nodes and BoxLib begins to catch up as can be seen in Figure~\ref{fig:amr_keeneland}.  
 This occurs because our intra-level ghost-cell exchange algorithm relies on GASNet memory to exchange ghost cells which 
@@
 }
-\caption{Throuphput of Adaptive Mesh Refinement in Millions of Cells/s}
-\label{fig:amr_total}
+\caption{Throuphput of Adaptive Mesh Refinement in Millions of Cells/s \label{fig:amr_total}}
 \end{figure}

### 2012-05-04 doc/sc2012/soop.tex
Small fixes to the experiments and soop sections

@@
 property established by stage 1 is that all
-tasks $t$ depends on map before $t$ maps and no task that depends on
+tasks on which $t$ depends map before $t$ maps and no task that depends on
 $t$ maps before $t$ finishes mapping.  Thus, the region tree when

### 2012-05-04 doc/sc2012/mapping.tex
mapping

--- 2012-05-04 doc/sc2012/mapping.tex
@@
-\section{Mapping}
+\section{Mapping Interface}
 \label{sec:mapping}
 
-While the Legion programming model describes locality and independence information abstractly,
-to run a Legion application on a machine we must make concrete decisions about
-where tasks will be run and where region instances will be placed.  Instead of placing
-the burden of decision making on the programming system, we introduce a programmable mapping
-interface that gives a programmer total control over how these decisions are made.  Armed
-with this interface, the programmer can then specify application- or machine-specific
-mapping decisions that would be difficult for a general-purpose programming system to infer.
-In this section we describe the general mapping interface (Section \ref{sec:mapinterface}),
-our base implmentation of this interface (Section \ref{sec:defmapper}), and the benefits
+As mentioned previously, a novel aspect of Legion is the {\em mapping interface},
+which gives programmers control over where tasks run and where region instances are placed,
+making possible application- or machine-specific mapping decisions that would be difficult for a 
+general-purpose programming system to infer.
+We describe the interface (Section \ref{sec:mapinterface}),
+our base implementation (Section \ref{sec:defmapper}), and the benefits
 of creating custom mappers (Section \ref{sec:custommap}).
 
-\subsection{Mapping Interface}
+\subsection{The Interface}
 \label{sec:mapinterface}
-A mapper is any object that implements the mapping interface.  The mapping interface
-consists of ten function calls that the runtime will make to request direction from
-the mapper about how a program should be executed.  To aid in performing the 
-decisions made in these calls the mapper has access to a simple 
-interface for inspecting the properties of the machine on which it is running.  This includes 
-information about all the processors and their kinds (i.e. CPU,GPU), as well as the memories visible
-to each processor and their latencies and bandwidths.  For brevity we only
-discuss three of the mapping interface calls.
+
+The mapping interface consists of ten methods that SOOPs call for
+mapping decisions.  A {\em mapper} implementing these methods has access to a simple interface
+for inspecting properties of the machine, including a list of
+processors and their type (i.e. CPU, GPU) and a list of
+memories visible to each processor and their latencies and bandwidths.
+For brevity we only discuss three of the mapping interface calls:
 
 \begin{itemize}
-\item {\tt select\_initial\_processor} - for each task the runtime system will
-ask the mapper to which processor the task should be sent.  The mapper 
-can choose to keep the task on the local processor or send it to any other processor
-in the system.
+\item {\tt select\_initial\_processor} - For each task $t$ in its mapping queue a SOOP will
+ask for a processor for $t$.  The mapper can keep the task on the local processor or send it to 
+any other processor in the system.
 
-\item {\tt permit\_task\_steal} - when handling a steal request from another processor
-the runtime will ask the mapper to decide which tasks are allowed to be stolen.  
-By always returning an empty set, the mapper can disable stealing.
+\item {\tt permit\_task\_steal} - When handling a steal request a SOOP
+asks which tasks may be stolen.  Stealing can be disabled by always returning the empty set.
 
-\item {\tt map\_task\_region} - for each logical region requested for a task, the
-runtime will query the mapper for a prioritized list of memories where a physical instance for this
-task should be placed.  To aid the mapping decision the runtime provides a list of the current valid physical intances
-of the logical region.  The mapper returns a list of memories in which the runtime
-should attempt to either reuse or manifest a physical instance of the logical region.  The
-runtime traverses this list and searches for a currently valid instance.  If none can be
-found, it attempts to allocate a physical instance and then issue the necessary copies
-to retrive the valid data.  If both options fail, it moves on to the next memory in the list.
+\item {\tt map\_task\_region} - For each logical region $r$ used by a task, a
+SOOP asks for a prioritized list of memories where a physical instance of $r$ should be placed.  
+The SOOP provides a list of $r$'s current valid physical instances;
+the mapper returns a priority list of memories in which the SOOP
+should attempt to either reuse or create a physical instance of $r$.  Beginning with
+the first memory, the SOOP uses a current valid instance if one is present.  Otherwise,
+the SOOP attempts to allocate a physical instance and issue copies
+to retrieve the valid data.  If that also fails, the SOOP moves on to the next memory in the list.
 \end{itemize}
 
-There are two very elegant properties of this mapping interface.  The first property is
-that all decisions made by a mapper are orthogonal to the correctness of the program and can only 
-impact performance.  Regardless of where a mapper chooses to place a task or map a region, 
-the runtime will always schedule tasks and copies in accordance with the privileges and 
-coherence properties specified in the Legion program.  Therefore when writing an application
-in Legion, a programmer can begin by using the default mapper and later optimize his program
-by creating a custom mapper and gradually refining its mapping decisions.
-
-The second useful property of the mapping interface is that it isolates machine-specific decisions
-to a specific module of code.  As a result, Legion programs are highly
-portable.  To port a Legion program to a new architecture, a programmer only has to
+The mapping interface has two desirable properties.  First, all mapper
+decisions are orthogonal to program correctness and can only impact
+performance.  Regardless of where a mapper places a task or region,
+the SOOPs schedule tasks and copies in accordance with the privileges
+and coherence properties specified in the program.  Therefore when
+writing a Legion application, a programmer can begin by using the
+default mapper and later improve performance by creating and refining a custom mapper.
+Second, the mapping interface isolates machine-specific decisions
+to the mapper.  As a result, Legion programs are highly
+portable.  To port a Legion program to a new architecture, a programmer need only
 implement a new mapper with decisions specific to the new architecture.
@@
 \label{sec:defmapper}
-To make writing Legion applications easier, we provide a default mapper implementation that
-can be used to get an application working quickly at a moderate performance level.  The 
-default mapper employs a simple scheme for mapping tasks and regions.  When
-{\tt select\_initial\_processor} is invoked for the default mapper, it first checks for which kind of processors
-the task has variants (i.e. GPU).  If the fastest variant is for the kind of processor the mapper is managing
-it will choose to keep the task local, otherwise it will send the task to the closest processor of the
-fastest variant kind.
+To make writing Legion applications easier, we provide a default
+mapper that can quickly get an application working
+with moderate performance.  The default mapper employs a
+simple scheme for mapping tasks.  When {\tt
+select\_initial\_processor} is invoked, the mapper
+checks the type of processors for which task $t$ has
+implementations (e.g., GPU).  If the fastest implementation is for the
+local processor the mapper keeps $t$ local, otherwise
+it sends $t$ to the closest processor of the fastest kind that can run
+$t$.
 
-The default mapper employs a Cilk-like algorithm for task stealing\cite{CILK95}.  Tasks are kept local
-to their target processor whenever possible and only moved when stolen.  Unlike Cilk, the default mapper
-has the information necessary for locality-aware stealing.  When {\tt permit\_task\_steal} is called for a task, 
-the default mapper inspects the logical regions for the task being stolen, and will mark that other tasks using 
-the same logical regions should be stolen as well.
+The default mapper employs a Cilk-like algorithm for task
+stealing \cite{CILK95}.  Tasks are kept local 
+whenever possible and only moved when stolen.  Unlike Cilk, the
+default mapper has the information necessary for locality-aware
+stealing.  When {\tt permit\_task\_steal} is called for a task, the
+default mapper inspects the logical regions for the task being stolen
+and marks that other tasks using the same logical regions should
+be stolen as well.
 
@@
 by bandwidth from the local processor.  This stack is then returned as the location of memories to be used for
-mapping each region.  Note that this is a greedy algorithm that works well for the common case, but can cause some 
-regions to be pulled unnecessarily close to the processor and consume precious fast memory space.
+mapping each region.  This greedy algorithm works well in common cases, but can cause some 
+regions to be pulled unnecessarily close to the processor, consuming precious fast memory.
 
@@
 
-Each custom mapper extends the default mapper.  A programmer only has to override the functions necessary 
+Each custom mapper extends the default mapper.  A programmer need only override the functions necessary 
 for creating a custom mapping.  Custom mappers can be used to create totally static mappings, 

### 2012-05-04 doc/sc2012/soop.tex
mapping

--- 2012-05-04 doc/sc2012/soop.tex
@@
 
-Figure~\ref{fig:gpumapping} shows a timeline of our mapping of the
-circuit simulation on a cluster of GPUs.  The $i$th private node and
-wire regions are mapped to the $i$th GPU's DRAM and simply left there
-for all three phases of the simulation.  The $i$th shared and ghost
-regions are mapped to zero-copy memory most of the time, but, as just
-illustrated, must be reduced out to the all-shared node region in the
-global memory (which is managed by GASNet) and then a new instance of the shared
-subregions created with the updated information before the {\tt
-  update\_voltages} tasks execute.  A similar situation arises before
-{\tt calc\_new\_currents} tasks run again: the updates to the shared
-nodes by {\tt update\_voltage} must be propagated to the ghost nodes
-before continuing.
-
-\begin{figure}
-\includegraphics[scale=0.48]{figs/CircuitMem.pdf}
-\caption{Tasks and data for the circuit simulation on a cluster of GPUs.}
-\label{fig:gpumapping}
-\end{figure}
+%
+% FIXME   CUT
+%
+%Figure~\ref{fig:gpumapping} shows a timeline of our mapping of the
+%circuit simulation on a cluster of GPUs.  The $i$th private node and
+%wire regions are mapped to the $i$th GPU's DRAM and simply left there
+%for all three phases of the simulation.  The $i$th shared and ghost
+%regions are mapped to zero-copy memory most of the time, but, as just
+%illustrated, must be reduced out to the all-shared node region in the
+%global memory (which is managed by GASNet) and then a new instance of the shared
+%subregions created with the updated information before the {\tt
+%  update\_voltages} tasks execute.  A similar situation arises before
+%{\tt calc\_new\_currents} tasks run again: the updates to the shared
+%nodes by {\tt update\_voltage} must be propagated to the ghost nodes
+%before continuing.
+%
+%\begin{figure}
+%\includegraphics[scale=0.48]{figs/CircuitMem.pdf}
+%\caption{Tasks and data for the circuit simulation on a cluster of GPUs.}
+%\label{fig:gpumapping}
+%\end{figure}
 

### 2012-05-04 doc/sc2012/experiments.tex
Modifications to the experiments section

@@
 
-In contrast, the machine model used by the Legion runtime is based on an
+In contrast, the machine model used by the Legion SOOP runtime is based on an
 affinity graph of CPUs and GPUs and is able to accurately capture complex
@@
 run on subsets of each machine ranging from the smallest (a single CPU core or GPU)
-to the largest or near-largest (except Keeneland which has 115 nodes, but we limited 
-our runs to at most 32 nodes to get sufficient cluster time.)
+to the largest or near-largest (except Keeneland which has 120 nodes, but we limited 
+to at most 32 nodes to get sufficient cluster time.)
 By looking at the performance of the same size problem over progressively larger
@@
 The first experiment we investigate is the distributed circuit simulation described in 
-Section~\ref{sec:ex}.  The Legion runtime handles all of the resource allocation, 
+Section~\ref{sec:ex}.  The Legion SOOP runtime handles all of the resource allocation, 
 scheduling, and data movement across the cluster of GPUs.  In particular,  
@@
 shared data around the system while keeping the private nodes and wires resident in
-each GPU's framebuffer memory is critical in keeping the overhead low and achieving 
-good scalability.
+each GPU's framebuffer memory is critical to achieving good scalability.
 
@@
 all nodes) spent in the application kernels compared to the various pieces of the Legion
-runtime.  As the node count increases, the non-communication overhead stays relatively constant.
+SOOP runtime.  As the node count increases, the non-communication overhead stays relatively constant.
 As expected the communication overhead grow linearly with number of nodes.
@@
 such that the range of interaction is limited to just the
-cells adjacent (including diagonals) to the one a particle resides in.  Originally designed to be a
-chip multiprocessor benchmark, the application divides the array into {\em grids} and assigns each one to 
+cells adjacent (including diagonals) to the one a particle resides in.  The application divides the array 
+into {\em grids} and assigns each one to 
 a thread.  Per cell locking is used to safely access particles in cells that lie
-on the edge of a grid.  This fine-grained locking scheme along with the assumption of a shared address space lead to
-good scaling in a multi-core processor, but rule out any attempt run beyond a single node.  To extend the
-scaling further, our port uses region partitioning to divide the array into grids in a similar
-way, but handles the interaction between grids in a way that doesn't rely on access to uniform shared memory.
+on the edge of a grid.  This fine-grained locking scheme along with the assumption of a shared address space leads to
+good scaling in a multi-core processor, but rules out any attempt run beyond a single node.
+
+To extend the scaling further, our port uses region partitioning to divide the array into grids in a similar
+way, but handles the interaction between grids in a way that doesn't rely on access to shared memory.
 Instead, the Legion version of the benchmark creates explicit ghost copies of cells on the boundaries of grids
@@
 using a relatively small problem (300K particles on a 15x21x15 array of cells).  Between 1 and 4 threads, the
-results for sapling are nearly indistinguishable, indicating 
-neither the Legion runtime nor the the restructuring of the implementation to allow multi-node scaling impose
-any significant overhead.  It's possible that the use of explicit ghost cells rather than 
-fine-grained sharing of cache lines might be a net win.  At 8 threads and above, the performance begins to vary.
-Both the Legion and PARSEC version on viz flatten out as they over-subscribe the 12 physical cores.  On sapling,
+results for Sapling are nearly indistinguishable, indicating 
+neither the Legion runtime nor the restructuring of the implementation to allow multi-node scaling impose
+any significant overhead.  
+%It's possible that the use of explicit ghost cells rather than fine-grained sharing of cache lines might be a net win.  
+At 8 threads and above, the performance begins to vary.
+Both the Legion and PARSEC version on Viz flatten out as they over-subscribe the 12 physical cores.  On Sapling,
 which has HyperThreading enabled, deviations from linear begin sooner as the operating system's scheduler's
@@
 there are twice as many ghost cells as interior grid cells.  The larger problem sizes (2.4M and 19M
-particles) do much better, with scaling of up to 5.4x when going from 1 node up to 16 because the
-communication-to-computation ratio is lower.
+particles) do much better, with scaling of up to 5.4x when going from 1 node up to 16 because of a lower 
+communication-to-computation ratio.
 
@@
 the boundary cells around a box at a refined level linearly interpolate their values from the nearby cells at the
-next coarsest level.  The second phase computes the stencil computation on each cell in all levels.
+next coarsest level.  The second phase performs the stencil computation on each cell in every level.
 In the third phase, cells at a coarser level that have been refined are restricted to the average
@@
 Our second optimization takes advantage of the cross-level locality present in the application.  We implemented
-an application specific mapper that inspects the arguments of tasks to discover the relationships between
+an application specific mapper that dynamically discovers the relationships between
 grids at different levels of refinement.  The mapper dynamically performs intersection tests between logical regions
@@
 of cells per level of refinement, but with randomly chosen refinements locations.  BoxLib also supports OpenMP
-and we took their best number from using 1, 2, 4, or 8 OpenMP threads per node.  Our Legion implementation
+and we took their best performance from using 1, 2, 4, or 8 threads per node.  Our Legion implementation
 always uses one thread per node to illustrate that in this application locality is significantly more important 
@@
 As the node count increased the AMR code becomes highly dependent on interconnect performance.  BoxLib performs much better
-on Keeneland than it does on our viz cluster due to the better interconnect.  This can also be seen as we scale
-to a larger number of nodes and BoxLib begins to catch up as can be seen in Figure~\ref{fig:amr_keeneland}.  
+on Keeneland than it does on our Viz cluster due to the better interconnect.  As we scale
+to a larger number of nodes BoxLib begins to catch up as can be seen in Figure~\ref{fig:amr_keeneland}.  
 This occurs because our intra-level ghost-cell exchange algorithm relies on GASNet memory to exchange ghost cells which 

### 2012-05-04 doc/sc2012/experiments.tex
experiments

--- 2012-05-04 doc/sc2012/experiments.tex
@@
 \label{sec:exp}
-
-The efficiency and scalability of Legion was examined by measuring the performance
-of three applications on three clusters.  The table in Figure~\ref{fig:systems} summarizes 
-the three clusters that were used.
 
@@
 }
-
-\caption{Configurations of Systems Used for Experiments \label{fig:systems}}
+\caption{System configurations used for the experiments. \label{fig:systems}}
 \end{figure}
 
-Nodes with multiple instances of different resources are a challenge for 
-existing programming models that have a flat model of the system (e.g. MPI).
-They are faced with a choice of lumping all 
-the resources together and ignoring the internal irregularities (e.g. NUMA
-in multi-socket x86 systems) or of separating a single physical node into
-multiple smaller nodes, ignoring the better affinity between sockets/GPUs/etc
-on the same node and either wasting or over-subscribing some resources when
-the quantities of different resources don't share a common divisor.
+% FIXME
+% This is a nice paragraph, but it doesn't belong here; perhaps in the mapping section
+% if we have space.
+%
+%Nodes with multiple instances of different resources are challenging
+%for existing programming models with flat system views (e.g., MPI).
+%They must either lump resources together and ignore internal irregularities (e.g. NUMA
+%in multi-socket x86 systems), or divide a single physical node into
+%multiple smaller nodes, ignoring the better affinity between sockets/GPUs/etc.
+%on the same node. In contrast, the machine model used by Legion is based on an
+%affinity graph of CPUs and GPUs and accurately captures complex
+%machine hierarchies.
+%
+% FIXME
+%   I cut this from the above, commented-out paragraph because it was a digression. 
+%  
+% and either wasting or over-subscribing some resources when
+%the quantities of different resources don't share a common divisor.
 
-In contrast, the machine model used by the Legion runtime is based on an
-affinity graph of CPUs and GPUs and is able to accurately capture complex
-machine hierarchies.
-
+We evaluate the efficiency and scalability of Legion using
+three applications on three clusters (see Figure~\ref{fig:systems}).
 For each application, multiple problem sizes were used, and each size problem was
 run on subsets of each machine ranging from the smallest (a single CPU core or GPU)
-to the largest or near-largest (except Keeneland which has 115 nodes, but we limited 
-our runs to at most 32 nodes to get sufficient cluster time.)
-By looking at the performance of the same size problem over progressively larger
-machines, we are able to measure the strong scaling that Legion is able to
-archive.  By increasing the problem size as well, we also measure weak scaling.
+to the largest or near-largest (except Keeneland, where we limited
+runs to 32 nodes to get sufficient cluster time).
+By examining performance of the same size problem over progressively larger
+machines, we measure Legion's strong scaling.
+By increasing the problem size as well, we also measure weak scaling.
 
@@
 shared data around the system while keeping the private nodes and wires resident in
-each GPU's framebuffer memory is critical in keeping the overhead low and achieving 
+each GPU's framebuffer memory is critical to keeping the overhead low and achieving 
 good scalability.
@@
 Circuits of two different sizes were simulated.  The first had 480K wires, connecting
-120K nodes.  The second is twice as large, with nearly a million wires connecting a
-quarter of a million nodes.  In addition to running these tests on varying number of
+120K nodes.  The second is twice as large, with nearly 1M wires connecting 
+250K nodes.  In addition to running these tests on varying numbers of
 nodes, the number of GPUs used by the runtime was also varied.  In no case did the 
-changes to nodes or number of GPUs per node require any changes to the application code.
+changes to nodes or number of GPUs per node require changes to the application code.
 
-The circuit simulation used a simple, but application-specific, mapper.  At initialization
+The circuit simulation has a simple application-specific mapper.  At initialization
 time, the mapper queries the list of GPUs in the machine and identifies each GPU's
-framebuffer memory and {\em zero-copy} memory, the pinned memory that both the GPU and
-CPUs in the system can access directly.  Once the circuit is partitioned, the partitions
+framebuffer memory and {\em zero-copy} memory (pinned memory that both the GPU and
+CPUs in the system can access directly).  Once the circuit is partitioned, the partitions
 are assigned a home GPU in round-robin fashion.  Every task related to that partition is
-then sent to the home GPU, with no task stealing allowed.  (If the circuit is partitioned well,
-there is minimal load imbalance between the pieces, and the rebalancing benefit of 
-moving a partition is unlikely to be worth the cost of moving all the 
-private data for a piece from one GPU to another.)  The regions for a task are mapped as 
-shown in Figure~\ref{fig:gpumapping}.
+then sent to the home GPU, with no task stealing allowed.  
+%
+% FIXME CUT
+% We could put this nice sentence back in if we have space.
+%
+%(In a well-partitioned circuit,
+%load imbalance is low enough that the cost of moving the private data for a piece from one
+%GPU to another outweighs any benefits.)
+%
+% FIXME CUT  
+% If we find enough space this sentence and the figure can be reinserted here.
+%  The regions for a task are mapped as 
+%shown in Figure~\ref{fig:gpumapping}.
 
@@
 62.5X for the larger problem size on 96 GPUs.  On a single GPU, our Legion implementation
-was within 5\% of the performance of the hand-coded example.
+is within 5\% of the performance of the hand-coded simulation.
 
@@
 runtime.  As the node count increases, the non-communication overhead stays relatively constant.
-As expected the communication overhead grow linearly with number of nodes.
+As expected the communication overhead grows linearly with number of nodes.
 
 \begin{figure}
-\subfigure[Circuit Simulation Speed Relative to Single-GPU Implementation]
+\subfigure[Circuit simulation speed relative to single-GPU implementation.]
 {
@@
 
-\subfigure[Overhead of Circuit Simulation on Keeneland with 3 GPUs/Node]
+\subfigure[Overhead of circuit simulation on Keeneland with 3 GPUs/node.]
 {
@@
 }
-\caption{Circuit Simulation Results}
+\caption{Circuit simulation results.}
 \end{figure}
@@
 \begin{figure}
-\subfigure[Single-Node Particle Simulation Speed]
+\subfigure[Single-node particle simulation speed.]
 {
@@
 
-\subfigure[Multi-Node Scaling]
+\subfigure[Multi-node scaling.]
 {
@@
 
-%\subfigure[Effect of Load-Balanced Partitioning]
+%\subfigure[Effect of load-balanced partitioning.]
 %{
@@
 %}
-\caption{Fluid Simulation Results}
+\caption{Fluid simulation results.}
 \end{figure}
@@
 
-As the node count increased the AMR code becomes highly dependent on interconnect performance.  BoxLib performs much better
+As the node count increases the AMR code becomes highly dependent on interconnect performance.  BoxLib performs much better
 on Keeneland than it does on our viz cluster due to the better interconnect.  This can also be seen as we scale
@@
 \begin{figure}
-\subfigure[Sapling Results]
+\subfigure[Sapling results.]
 {
@@
 
-\subfigure[Viz Results]
+\subfigure[Viz results.]
 {
@@
 
-\subfigure[Keeneland Results]
+\subfigure[Keeneland results.]
 {
@@
 }
-\caption{Throuphput of Adaptive Mesh Refinement in Millions of Cells/s \label{fig:amr_total}}
+\caption{Throuphput of adaptive mesh refinement in millions of cells/s \label{fig:amr_total}}
 \end{figure}

### 2012-05-04 doc/sc2012/code_ex.tex

@@
                   rn_ghost: p_ghost_nodes[i], rw_pvt: p_wires[i] };
-  while(!done) {
+  for (t = 0; t #$<$# TIME_STEPS; t++) {
     spawn (i = 0; i #$<$# MAX_PIECES; i++) calc_new_currents(pieces[i]);

### 2012-05-04 doc/sc2012/legion.tex
Minor edits

@@
 more {\em physical instances} assigned to specific memories .  It is often useful to have multiple
-physical instances of a logical region (e.g., to duplicate read-only data, or to allow
+physical instances of a logical region (e.g., to replicate read-only data, or to allow
 independent reductions that are later combined).

### 2012-05-04 doc/sc2012/soop.tex
Fixing a figure bug in the soop section
Minor edits

@@
 to block.  With deferred execution, it is worthwhile for the SOOP to run (potentially far) ahead of actual execution,
-allocating resources to and issuing tasks that must wait because they have data dependences on earlier tasks that have not 
+allocating resources to and issuing tasks that may wait because they have data dependences on earlier tasks that have not 
 finished executing.   Pipelining and distributing the SOOP further improves throughput and hides latency.
@@
 checked against the first for any dependences; the second subtask uses
-{\tt pieces[1].rw\_pvt} (read/write exclusive) and {\tt
+{\tt pieces[1].rw\_pvt} (read~/~write exclusive) and {\tt
 pieces[1].rn\_pvt}, {\tt .rn\_shr}, and {\tt .rn\_ghost} (read-only exclusive).
@@
 The least common ancestor is the region of all shared nodes {\tt p\_nodes\_pvs[1]},
-so these two regions alias.  Since both tasks write their respective subregions,
+so these two regions alias.  Since both tasks modify their respective subregions,
 there is a dependence and the {\tt update\_voltages} task can only map after the
@@
 }
-\label{fig:depexamp}
-\caption{Dependence analysis examples from {\tt circuit\_simulation}.}
+\caption{Dependence analysis examples from {\tt circuit\_simulation}. \label{fig:depexamp}}
 \end{figure*}
@@
 tasks on which $t$ depends map before $t$ maps and no task that depends on
-$t$ maps before $t$ finishes mapping.  Thus, the region tree when
-$t$ maps contains a consistent view of what the state of physical instances
-will be after $t$'s dependence predecessors execute.
+$t$ maps before $t$ finishes mapping.  Thus, when $t$ maps, the region tree
+will be in a state consistent with an execution in which all tasks
+mapped in program order.
+%Thus, the region tree when
+%$t$ maps contains a consistent view of what the state of physical instances
+%will be after $t$'s dependence predecessors execute.
 
@@
 physical instances and thus similar to the mapping dependence analysis on logical
-regions in Section~\ref{sec:dep}.  Consider any task $t'$ using
+regions in Section~\ref{sec:dep}.  Consider mapping task $t$ with task $t'$ using
 physical instance $s'$ of logical region $r'$ in the region tree.  There are four cases.
@@
 analysis in Section~\ref{sec:dep}; the amortized work per region mapped
-is proportional to the depth of the region forest.
+is proportional to the depth of the region forest and independent of
+the number of tasks or physical instances.
 
@@
 Once task $t$ is done executing its state is reclaimed. Dependence
-and mapping information is removed from the region tree (it can
-usually be removed even earlier).  The most involved aspect of
+and mapping information is removed from the region tree.  The most involved aspect of
 clean-up is collecting physical instances that are no longer in use,

### 2012-05-04 doc/sc2012/experiments.tex

@@
 shared data around the system while keeping the private nodes and wires resident in
-each GPU's framebuffer memory is critical to keeping the overhead low and achieving 
-good scalability.
+each GPU's framebuffer memory is critical to achieving good scalability.
 
@@
 all nodes) spent in the application kernels compared to the various pieces of the Legion
-runtime.  As the node count increases, the non-communication overhead stays relatively constant.
-As expected the communication overhead grows linearly with number of nodes.
+SOOP runtime.  As the node count increases, the non-communication overhead stays relatively constant.
+As expected the communication overhead grow linearly with number of nodes.
 

### 2012-05-05 doc/sc2012/experiments.tex
updates

@@
 
-Our second experiment is a port of the \emph{fluidanimate} benchmark from the PARSEC suite\cite{bienia11benchmarking},
-which does a particle-based simulation of an incompressible fluid.  Each particle interacts only with other nearby
-particles. The benchmark divides the space in which the fluid can move into a three-dimensional array of cells 
-such that the range of interaction is limited to just the
-cells adjacent (including diagonals) to the one a particle resides in.  The application divides the array 
-into {\em grids} and assigns each one to 
-a thread.  Per cell locking is used to safely access particles in cells that lie
-on the edge of a grid.  This fine-grained locking scheme along with the assumption of a shared address space leads to
-good scaling in a multi-core processor, but rules out any attempt run beyond a single node.
+Our second experiment is a port of the \emph{fluidanimate} benchmark
+from the PARSEC suite\cite{bienia11benchmarking}, which does a
+particle-based simulation of an incompressible fluid.  Each particle
+interacts only with other nearby particles. The benchmark divides the
+space in which the fluid can move into a three-dimensional array of
+cells such that the range of interaction is limited to just the cells
+adjacent (including diagonals) to the one a particle resides in.  The
+application divides the array into {\em grids} and assigns each grid to
+a thread.  Per-cell locking safely accesses particles in
+cells that lie on the edge of a grid.  This fine-grained locking
+scheme along with the assumption of a shared address space gives
+good scaling in a multi-core processor, but rules out any attempt to run
+beyond a single node.
 
-To extend the scaling further, our port uses region partitioning to divide the array into grids in a similar
-way, but handles the interaction between grids in a way that doesn't rely on access to shared memory.
-Instead, the Legion version of the benchmark creates explicit ghost copies of cells on the boundaries of grids
-and uses those ghost cells to exchange information between the grids.
+To extend the scaling further, our port uses region partitioning to
+divide the array into grids in a similar way, but avoids relying
+on shared memory to handle interactions between grids.
+Instead, the Legion version creates
+explicit ghost copies of cells on grid boundaries and uses
+those ghost cells to exchange information between the grids.
 
-The mapper for the particle simulation is very simple - it maps one grid's tasks onto each processor and
-maps all regions for that grid (both the internal and the ghost regions) into the system memory used by that processor.
-The exchange of the ghost cell data between processors is handled by the Legion runtime as a ghost cell region
-gets alternately mapped in two different memories.
+The particle simulation's mapper is very simple: it maps one grid's
+tasks onto each processor and maps all that grid's regions (both
+internal and ghost regions) into that processor's system memory.  The
+exchange of ghost cell data between processors is handled by the
+Legion runtime as a ghost cell region is alternately mapped to two
+different memories.
 
-Figure~\ref{fig:fluid_single} compares the performance of the Legion implementation against the PARSEC version,
-using a relatively small problem (300K particles on a 15x21x15 array of cells).  Between 1 and 4 threads, the
-results for Sapling are nearly indistinguishable, indicating 
-neither the Legion runtime nor the restructuring of the implementation to allow multi-node scaling impose
-any significant overhead.  
+Figure~\ref{fig:fluid_single} compares the performance of the Legion
+implementation against the PARSEC version, using a relatively small
+problem (300K particles on a 15x21x15 array of cells).  Between 1 and
+4 threads, the results for Sapling are nearly indistinguishable,
+indicating neither the Legion runtime nor the restructuring of the
+implementation to allow multi-node scaling impose any significant
+overhead.
 %It's possible that the use of explicit ghost cells rather than fine-grained sharing of cache lines might be a net win.  
-At 8 threads and above, the performance begins to vary.
-Both the Legion and PARSEC version on Viz flatten out as they over-subscribe the 12 physical cores.  On Sapling,
-which has HyperThreading enabled, deviations from linear begin sooner as the operating system's scheduler's
+At 8 threads and above, performance begins to vary.  Both the Legion
+and PARSEC versions on Viz flatten out as they over-subscribe the 12
+physical cores.  On Sapling, which has HyperThreading enabled,
+deviations from linear begin sooner as the operating system's
 thread placement choices begin to matter.
 
-To measure scaling beyond a single node, three different problem sizes were run for each of the three systems,
- with the results plotted in Figure~\ref{fig:fluid_multi}.  
- For the smallest problem (300K particles), we observe a 20\% speedup when going to 1
-node to 2 (a total of 16 threads), but begin slowing down beyond that due to communication overhead - at 4 nodes
-there are twice as many ghost cells as interior grid cells.  The larger problem sizes (2.4M and 19M
-particles) do much better, with scaling of up to 5.4x when going from 1 node up to 16 because of a lower 
+To measure scaling beyond a single node, three different problem sizes
+were run for each of the three systems (Figure~\ref{fig:fluid_multi}).
+For the smallest problem (300K particles), we observe a 20\% speedup
+from 1 to 2 nodes (16 threads total), but slow
+down beyond that due to communication overhead---at 4 nodes there are
+twice as many ghost cells as interior grid cells.  The larger problem
+sizes (2.4M and 19M particles) do much better, with scaling of up to
+5.4x when going from 1 node up to 16 because of a lower
 communication-to-computation ratio.
@@
 \label{subsec:exp_amr}
-Our final application is based on the third heat equation example from the Berkeley Labs 
-BoxLib project \cite{BoxLib}.  This application is a three-level adaptive-mesh-refinement (AMR)
-code that computes a first order stencil on a 2D mesh of cells.  The simulation iterates for many time
-steps.  Updating the simulation for one time step consists of three phases.  In the first phase,
-the boundary cells around a box at a refined level linearly interpolate their values from the nearby cells at the
-next coarsest level.  The second phase performs the stencil computation on each cell in every level.
-In the third phase, cells at a coarser level that have been refined are restricted to the average
-of the cells that they physically contain at the next finest level of refinement.
+Our final application is based on the third heat equation example from
+the Berkeley Labs BoxLib project \cite{BoxLib}.  This application is a
+three-level adaptive-mesh-refinement (AMR) code that computes a first
+order stencil on a 2D mesh of cells.  
+%The simulation iterates for many time steps.  
+Updating the simulation for one time step consists of
+three phases.  In the first phase, the boundary cells around a box at
+a refined level linearly interpolate their values from the nearby
+cells at the next coarsest level.  The second phase performs the
+stencil computation on each cell in every level.  In the third phase,
+cells at a coarser level that have been refined are restricted to the
+average of the cells that they physically contain at the next finest
+level of refinement.
 
-Achieving high-performance on this application is particularly challenging for several reasons.  
-First, the application has a very high communication-to-computation ratio which, for a fixed problem
-size begins as being memory bound and with increasing node count becomes network bound as the
-perimeter-to-area ratio of cell grids increases.  Second, when choosing how to partition cells into grids, the
-programmer must consider the locality both between cells in the same level, as well as for cells
-across levels.  In the case of cross-level cell dependences, mapping decisions must be made at runtime as the
-location of refinements are determined based on application input.  Finally, this application has
-parallelism both between tasks running at the same level and tasks running across levels, leading
-to complicated input-dependent data dependences.
+Achieving high-performance on this application is particularly
+challenging for several reasons.  First, the application has a very
+high communication-to-computation ratio which, for a fixed problem
+size, begins as being memory bound and with increasing node count
+becomes network bound as the perimeter-to-area ratio of cell grids
+increases.  Second, when choosing how to partition cells into grids,
+the programmer must consider the locality between cells within a
+level as well as across levels.  For
+cross-level cell dependences, mapping decisions must be made at
+runtime as the location of refinements are dynamically determined.
+Finally, this application has parallelism both
+between tasks running at the same level and tasks running across
+levels, leading to complicated input-dependent data dependences.
 
-BoxLib's implementation of this benchmark partitions cells within the same level into a number of grids
-based on the number of nodes in the machine and distributes one grid from each level to each node.  This
-optimizes for memory bandwidth and load balance, but does not take into account cross-level locality 
-that exists between grids from different levels of refinement.  Furthermore, BoxLib does not block grids
-into sub-grids to take advantage of intra-grid locality.
+BoxLib's implementation partitions cells within a level 
+into a number of grids based on the number of nodes in the
+machine and distributes one grid from each level to each node.  This
+optimizes for memory bandwidth and load balance, but does not 
+exploit cross-level locality between grids from
+different levels of refinement.  Furthermore, BoxLib does not block
+grids into sub-grids to take advantage of intra-grid locality.
 
-Our Legion implementation performs two optimizations that allow us to achieve better performance than BoxLib.
-First, for each level of refinement we recursively partition the logical region of cells based on 
-the number of nodes in the machine and the size of the L3 cache and L2 caches.  Legion allows us to describe 
-locality for many levels of the memory hierarchy instead of just at the node-level.
+Our Legion implementation performs two optimizations that allow us to
+outperform BoxLib.  First, for each level of
+refinement we recursively partition the logical region of cells based
+on the number of nodes in the machine and the sizes of the L2 and L3 caches.
+%Legion allows us to describe locality for many levels of
+%the memory hierarchy instead of just at the node-level.
+Our second optimization takes advantage of the cross-level locality.
+We wrote an application-specific mapper that dynamically discovers
+relationships between grids at different levels of refinement.  The
+mapper dynamically performs intersection tests between logical regions
+containing grids of different refinement levels.  If the mapper
+discovers overlaps between grids from different levels, the mapper
+places them on the same node in the machine.  The
+mapper memoizes the intersection tests to amortize their cost.  The
+mapper also dynamically load balances by distributing unconstrained
+grids from the coarsest level onto under-loaded nodes.
 
-Our second optimization takes advantage of the cross-level locality present in the application.  We implemented
-an application specific mapper that dynamically discovers the relationships between
-grids at different levels of refinement.  The mapper dynamically performs intersection tests between logical regions
-containing grids of different refinement levels to determine whether they overlap.  
-If grids from different levels overlap, the mapper
-ensures that they are always placed on the same node in the machine to leverage cross-level locality.  The mapper
-memoizes the results of these intersection tests to amortize their cost.  The mapper also dynamically
-load balances by distributing unconstrained grids from the coarsest level onto under-loaded nodes.
+We compared our Legion implementation against BoxLib on three
+different problem sizes with a fixed number of cells per level of
+refinement, but with randomly chosen refinement locations.  BoxLib
+also supports OpenMP and we took their best performance from using 1,
+2, 4, or 8 threads per node.  Our Legion implementation always uses
+one thread per node to illustrate that in this application locality is
+significantly more important than parallelism.
 
-We compared our Legion implementation against BoxLib on three different problem sizes with a fixed number
-of cells per level of refinement, but with randomly chosen refinements locations.  BoxLib also supports OpenMP
-and we took their best performance from using 1, 2, 4, or 8 threads per node.  Our Legion implementation
-always uses one thread per node to illustrate that in this application locality is significantly more important 
-than parallelism.  The results from our experiments on all three clusters can be seen in Figure~\ref{fig:amr_total}.
+Figure~\ref{fig:amr_total} gives the results.
+On just one node, blocking for caches using Legion achieves up to 2.6X
+speedup over BoxLib.  As node count increases, the mapper's
+ability to exploit cross-level locality further increases
+the performance advantage to 5.4X by reducing the total
+communication costs.
 
-On just one node, blocking for caches using Legion achieves up to 2.6X speedup over BoxLib.  As the node count 
-increases, the ability of our mapper to leverage cross-level locality allows us to further increase
-our performance advantage to 5.4X over BoxLib by reducing the total communication costs.
-
-As the node count increases the AMR code becomes highly dependent on interconnect performance.  BoxLib performs much better
-on Keeneland than it does on our viz cluster due to the better interconnect.  This can also be seen as we scale
-to a larger number of nodes and BoxLib begins to catch up as can be seen in Figure~\ref{fig:amr_keeneland}.  
-This occurs because our intra-level ghost-cell exchange algorithm relies on GASNet memory to exchange ghost cells which 
-involves a linear increase in network traffic with the number of nodes.  BoxLib instead uses direct
-node-to-node exchanges of ghost cells similar to our fluid application.  A future implementation of our AMR code 
-will employ a similar ghost cell exchange algorithm to improve scalability.
+As the node count increases the AMR code becomes highly dependent on
+interconnect performance.  BoxLib performs much better on Keeneland
+than on Viz due to the better interconnect.  
+At higher node counts BoxLib
+begins to catch up (see Figure~\ref{fig:amr_keeneland})
+because our intra-level ghost-cell exchange 
+uses GASNet memory to exchange ghost cells, requiring a
+linear increase in network traffic with the number of nodes.  BoxLib
+uses direct node-to-node exchanges of ghost cells, similar to
+our fluid application.  A future implementation of our AMR code will
+employ a similar ghost cell exchange algorithm to improve scalability.
 

### 2012-05-05 doc/sc2012/circuit.tex

@@
 A {\tt Circuit}
-has two regions: collections of nodes and wires (line
-3 of Listing~\ref{lst:code_ex}).\footnote{Note that pointers to region values declare the region to which they point.  For
+has two regions: a collection of nodes and a collection of wires (line
+3 of Listing~\ref{lst:code_ex}).\footnote{Note that all pointers declare the region to which they point.  For
 example, the definition of {\tt Wire} (line 2) is parametrized on the region
@@
 provide a {\em coloring}, which is a relation between the elements of
-a region and a set of colors.  A partition is an object which given a
-coloring and a region $r$ contains for each color $c$ a subregion of
-$r$ containing all the elements of $r$ colored $c$. Note that the
+a region and a set of colors.  
+%A partition is an object which given a
+%coloring and a region $r$ contains for each color $c$ a subregion of
+%$r$ containing all the elements of $r$ colored $c$. 
+For each color $c$ in the coloring, the partition contains a subregion $r$
+of the region being partitioned, with $r$ consisting of the elements colored $c$.  Note that the
 partition into shared and private nodes is disjoint because each node has one color.
@@
 task can access these two regions concurrently or be reordered around this
-task if they use either region).  The simulation 
-reads and writes all nodes and wires, and it must be done
-with exclusive access to ensure correct results.  Privileges specify what
+task if they use either region).  
+%The simulation 
+%reads and writes all nodes and wires, and it must be done
+%with exclusive access to ensure correct results.  
+Privileges specify what
 the function can do with the regions; coherence specifies what other
@@
 parallel execution ({\tt spawn} simply indicates a task call and does
-not require parallel execution) nor is there synchronization 
+not mandate parallel execution) nor is there synchronization 
 between the passes.  Which tasks can be run in
@@
 and then partition it into subregions.  One can also first partition an empty region into subregions
-and then allocate data in the subregions. We use both approaches in our
-applications (see Section~\ref{sec:exp}).
+and then allocate data in the subregions. 
+%We use both approaches in our
+%applications (see Section~\ref{sec:exp}).
 

### 2012-05-05 doc/sc2012/soop.tex

@@
 
-Once a task $t$ has been mapped it enters the execution stage of the SOOP.  When all of
+After a task $t$ has been mapped it enters the execution stage of the SOOP.  When all of
 the operations (other tasks and copies) on which $t$ depends have completed, $t$ is
-launched on the processor.  When $t$ spawns subtasks, they are recursively scheduled as siblings by
-the SOOP using $t$ as the parent task and $t$'s region arguments as the roots of the region forest.
+launched on the processor it was mapped to.  When $t$ spawns subtasks, 
+they are registered by the SOOP on which $t$ is executing, using $t$ as the parent task 
+and $t$'s region arguments as the roots of the region forest.  Each child of $t$ traverses
+the SOOP pipeline on the same processor as $t$, possibly being mapped to a 
+SOOP instance on a different processor to execute.
 

### 2012-05-05 doc/sc2012/related.tex

@@
 programming model in which only arrays can be recursively partitioned,
-all access is exclusive, there is a static mapping of tasks and data on to the machine (though
+all access is exclusive, there is a static mapping of tasks and data (though
 extensions to Sequoia make this mapping more dynamic \cite{BCSA11}) and, most fundamentally, the decomposition of tasks and the decomposition
@@
 the mapper interface, and allowing different coherence
-properties.  Legion's decoupling of the task tree from the region tree in Legion leads directly to the scheduling problem solved by our software out-of-order processor for tasks with region arguments.
+properties.  Legion's decoupling of the task tree from the region tree leads directly to the scheduling problem solved by our software out-of-order processor for tasks with region arguments.
 
@@
 
-X10 is another parallel programming environment designed to operate on distributed
-memory machines \cite{X1005}.  X10's {\em places} enable programmers to talk about where
-to place both data and tasks in the machine.  However, once data and tasks have
-been placed they are fixed, which mandates that data movement be explicitly managed by
-user level code or implicitly by the compiler or runtime \cite{X1008}.  Recently X10 has introduced
-the concept of a region into the compiler's intermediate representation \cite{X1011}.
-Unlike Legion, regions in X10 are not visible to the programmer but are inferred
-from high level arrays through static analysis.  X10 provides support for clusters of GPUs
-\cite{X10GPU}, but requires the programmer to write all code managing data movement
-through both the cluster and GPU memory hierarchies.
+X10 is another parallel programming language designed to operate on
+distributed memory machines \cite{X1005}.  X10's {\em places} enable
+programmers to talk about where to place both data and tasks.
+However, once data and tasks have been placed they are fixed, which
+mandates that data movement be explicitly managed by user level code
+or implicitly by the compiler or runtime \cite{X1008}.  Recently X10
+has introduced regions into the compiler's
+intermediate representation \cite{X1011}.  Unlike Legion, regions in
+X10 are not visible to the programmer but are inferred from high level
+arrays through static analysis.  X10 provides support for clusters of
+GPUs \cite{X10GPU}, but requires the programmer to write all code
+managing data movement through both the cluster and GPU memory
+hierarchies.
 
@@
 Part of Legion's
-low-level runtime system is constructed using UPC's underlying GASNet
+low-level runtime system is constructed using UPC's GASNet
 runtime system \cite{GASNET07}.

### 2012-05-05 doc/sc2012/experiments.tex
Updating spacing on figures in the experiments section
inset for circuit plot
lots of plot changes

@@
 }
+\vspace{-2mm}
 \caption{System configurations used for the experiments. \label{fig:systems}}
+\vspace{-6mm}
 \end{figure}
@@
 We evaluate the efficiency and scalability of Legion using
-three applications on three clusters (see Figure~\ref{fig:systems}).
+three applications on three clusters (see Figure~\ref{fig:systems}).  All three
+clusters were Linux-based, and the Legion runtime was built using pthreads for
+managing CPU threads, CUDA\cite{cuda} for GPUs, and GASNet\cite{GASNET07} for
+inter-node communication.  The RDMA features of GASNet were used to create a 
+globally addressable, but relatively slow, ``GASNet memory'' that is accessible
+by all nodes.
 For each application, multiple problem sizes were used, and each size problem was
@@
 %
-%(In a well-partitioned circuit,
-%load imbalance is low enough that the cost of moving the private data for a piece from one
-%GPU to another outweighs any benefits.)
+(In a well-partitioned circuit,
+load imbalance is low enough that the cost of moving the private data for a piece from one
+GPU to another outweighs any benefits.)
 %
@@
 % If we find enough space this sentence and the figure can be reinserted here.
-%  The regions for a task are mapped as 
-%shown in Figure~\ref{fig:gpumapping}.
+  The regions for a task are mapped as 
+shown in Figure~\ref{fig:gpumapping}.
+
+\begin{figure}
+\includegraphics[scale=0.48]{figs/CircuitMem.pdf}
+\caption{Tasks and data for the circuit simulation on a cluster of GPUs.}
+\label{fig:gpumapping}
+\end{figure}
 
@@
 \raisebox{0.25 in}{
-\includegraphics[scale=0.18]{figs/circuit_speedups_zoom.pdf}
+\includegraphics[scale=0.4]{figs/circuit_speedups_zoom.pdf}
 }
@@
 }
+\vspace{-2mm}
 \caption{Circuit simulation results.}
+\vspace{-6mm}
 \end{figure}
@@
 %}
+\vspace{-2mm}
 \caption{Fluid simulation results.}
+\vspace{-6mm}
 \end{figure}
@@
 }
-\caption{Throuphput of adaptive mesh refinement in millions of cells/s \label{fig:amr_total}}
+\vspace{-2mm}
+\caption{Throughput of adaptive mesh refinement code. \label{fig:amr_total}}
+\vspace{-6mm}
 \end{figure}

### 2012-05-05 apps/fluid3d/plot_graph_multinode.py

@@ -13,27 +13,45 @@ def plot_init(title, xlabel, ylabel, pp):
 
 markers = ['o','s','D','*','v','^','p','<','>','d']
 
-def plot(cpus, speedups, label, mark_index):
-    plt.plot(cpus, speedups, label = label, linestyle = 'dashed', markersize = 7, marker = markers[mark_index], linewidth=0.5)
+def plot(cpus, speedups, label, marker, color, size, fill):
+    plt.plot(cpus, speedups, label = label, linestyle = 'dashed', 
+             markersize = size, 
+             marker = marker, 
+             color = color, 
+             markerfacecolor = (color if fill else "None"),
+             linewidth=0.5)
 
 total_particles = {'300K': 305809, '2400K': 2446472, '19M': 19554392}
 total_cells = {'300K': 135424, '2400K': 1115721, '19M': 9056971}
 total_steps = 10
 
 large_plot_nodes = [1, 2, 4, 8, 16]
-large_plot = [('Keeneland',
-               [('19M', np.array([46.377, 21.444, 13.483, 10.891, 10.68])),
-                ('2400K', np.array([5.169, 3.206, 2.673, 2.764, 3.801])),
-                ('300K', np.array([0.912, 0.847, 0.836, 1.158, 2.415]))]),
-              ('Viz',
-               [('19M', np.array([43.058, 20.849, 13.434, 10.459])),
-                ('2400K', np.array([5.449, 3.331, 2.691, 2.519])),
-                ('300K', np.array([0.734, 0.684, 0.873, 1.122]))]),
-              ('Sapling',
-               [('19M', np.array([39.89, 21.859, 13.323])),
-                ('2400K', np.array([4.957, 2.896, 2.413])),
-                ('300K', np.array([0.788, 0.653, 0.761]))]),
-]
+large_plot = [('Keeneland', '19M', np.array([46.377, 21.444, 13.483, 10.891, 10.68])),
+              ('Viz', '19M', np.array([43.058, 20.849, 13.434, 10.459])),
+              ('Sapling', '19M', np.array([39.89, 21.859, 13.323])),
+              ('Keeneland', '2400K', np.array([5.169, 3.206, 2.673, 2.764, 3.801])),
+              ('Viz', '2400K', np.array([5.449, 3.331, 2.691, 2.519])),
+              ('Sapling', '2400K', np.array([4.957, 2.896, 2.413])),
+              ('Keeneland', '300K', np.array([0.912, 0.847, 0.836, 1.158, 2.415])),
[…]

### 2012-05-05 doc/sc2012/experiments.tex

@@
 DRAM/Node & 48 GB & 24 GB & 24 GB \\
-Infiniband & 2x QDR & QDR & QDR \\
+Infiniband & 2x QDR & QDR & 2x QDR \\
 \end{tabular}

### 2012-05-05 doc/sc2012/experiments.tex
Some updates to the experiment section

--- 2012-05-05 doc/sc2012/experiments.tex
@@
 DRAM/Node & 48 GB & 24 GB & 24 GB \\
-Infiniband & 2x QDR & QDR & QDR \\
+Infiniband & 2x QDR & QDR & 2x QDR \\
 \end{tabular}
 }
+\vspace{-2mm}
 \caption{System configurations used for the experiments. \label{fig:systems}}
+\vspace{-6mm}
 \end{figure}
@@
 }
+\vspace{-2mm}
 \caption{Throuphput of adaptive mesh refinement in millions of cells/s \label{fig:amr_total}}
+\vspace{-6mm}
 \end{figure}

### 2012-05-05 doc/sc2012/experiments.tex
Fixing the machine table
more changes

--- 2012-05-05 doc/sc2012/experiments.tex
@@
 clusters were Linux-based, and the Legion runtime was built using pthreads for
-managing CPU threads, CUDA\cite{cuda} for GPUs, and GASNet\cite{GASNET07} for
+managing CPU threads, CUDA\cite{CUDA} for GPUs, and GASNet\cite{GASNET07} for
 inter-node communication.  The RDMA features of GASNet were used to create a 
-globally addressable, but relatively slow, ``GASNet memory'' that is accessible
+globally addressable, but relatively slow, {\em GASNet memory} that is accessible
 by all nodes.
@@
 Figure~\ref{fig:ckt_speed} shows the performance of the Legion circuit simulation relative
-to a hand-coded single-GPU implementation written in CUDA.  Each line shows the scaling of
+to a hand-coded single-GPU implementation written in CUDA.  The hand-coded implementation is
+able to keep the entire simulation state in fast framebuffer memory.  Each line shows the scaling of
 a particular problem size as the number of nodes is varied.  Our results demonstrate
 excellent strong scaling, with speedups of 39.0X for the small problem on 48 GPUs and 
-62.5X for the larger problem size on 96 GPUs.  On a single GPU, our Legion implementation
+62.5X for the larger problem size on 96 GPUs.  The inset in the graph shows the relative
+performance for small numbers of GPUs.  On a single GPU, our Legion implementation
 is within 5\% of the performance of the hand-coded simulation.
@@
 implementation against the PARSEC version, using a relatively small
-problem (300K particles on a 15x21x15 array of cells).  Between 1 and
-4 threads, the results for Sapling are nearly indistinguishable,
+problem (300K particles on a 15x21x15 array of cells).  Speedups for
+both the Legion and threaded PARSEC implementations are measured 
+relative to PARSEC's serial version, which eliminates all locking
+operations.  Between 1 and
+4 threads, the PARSEC and Legion results are nearly indistinguishable,
 indicating neither the Legion runtime nor the restructuring of the

### 2012-05-05 apps/AMR/plot_amr.py
update to amr plot script

@@ -120,6 +120,8 @@ class Machine(object):
         legion_label = 'Legion Cells='+str(prob_size)
         plt.plot(nodes,boxlib_updates,'--',label=boxlib_label,linestyle='dashed',markersize=7,marker=markers[(mark_index)%len(markers)],linewidth=0.5)
         plt.plot(nodes,legion_updates,'--',label=legion_label,linestyle='dashed',markersize=7,marker=markers[(mark_index+1)%len(markers)],linewidth=0.5)
+        for i in range(len(nodes)):
+            print "Speedup over BoxLib on "+str(nodes[i])+" on "+self.name+" = "+str(legion_updates[i]/boxlib_updates[i])
 
         return mark_index+2
 
@@ -227,8 +229,8 @@ def make_plots(show = True, save = True):
         plt.xlabel('Node Count')
         plt.ylabel('Millions of Cell Updates/s')
         plt.grid(True)
-        if save:
-            fig.savefig(out_dir+mach.name+"_amr.pdf", format="pdf", bbox_inches="tight")
+        #if save:
+        #    fig.savefig(out_dir+mach.name+"_amr.pdf", format="pdf", bbox_inches="tight")
         #pp.savefig()
         #pp.close()
     

### 2012-05-05 doc/sc2012/experiments.tex
more changes

@@
 {\footnotesize
-\begin{tabular}{l|ccc}
+\begin{tabular*}{3.5in}{l|ccc}
 Cluster & Sapling & Viz & Keeneland \\
@@
 Infiniband & 2x QDR & QDR & 2x QDR \\
-\end{tabular}
+\end{tabular*}
 }
@@
 then sent to the home GPU, with no task stealing allowed.  
-%
-% FIXME CUT
-% We could put this nice sentence back in if we have space.
-%
 (In a well-partitioned circuit,
@@
 GPU to another outweighs any benefits.)
-%
-% FIXME CUT  
-% If we find enough space this sentence and the figure can be reinserted here.
-  The regions for a task are mapped as 
-shown in Figure~\ref{fig:gpumapping}.
+
+  The regions for the tasks are mapped as 
+shown in Figure~\ref{fig:gpumapping}.  Wires and private node data are kept in each GPU's
+framebuffer at all times.  An instance of the ``all-shared nodes'' region is placed in
+GASNet memory and instances for just the shared and ghost nodes needed by each GPU
+are placed into that GPU's zero-copy memory.
 

### 2012-05-05 apps/fluid3d/plot_graph_multinode.py

@@ -67,6 +67,7 @@ if __name__ == '__main__':
              psize_size[psize],
              psize_fill[psize],
              )
-    plt.legend(loc=4)
+    plt.legend(bbox_to_anchor=(0.40, 0.3, 0.6, 0.6))
+    #plt.legend(loc=4)
     fig.savefig("figs/fluid_multinode.pdf", format="pdf", bbox_inches="tight");
     #pp.close()

### 2012-05-05 doc/sc2012/soop.tex
Saving some space on figures

@@
 indicates independence.  {\tt Same} is a dependence unless the two tasks
-use the same physical instance of the logical region.
+use the same physical instance of the logical region.  Since tasks
+haven't mapped physical instances at this point in the pipeline {\tt Same} is always
+a mapping dependence.
 %\footnote{For example, if two tasks wish to acess the region with atomic coherence, then if they are working on the same physical copy of the data atomicity can be guaranteed using, e.g., locking.
@@
 {\tt Cont} 
-indicates a dependence unless a single writer has
-atomic coherence.
+indicates a dependence contingent on whether it is a reader or
+writer that is Atomic (e.g. Atomic Read-Write and Simultaneous
+Read-Only will not cause a dependence).
 
@@
 }
-\caption{Dependence table.}
-\label{fig:dependence}
+\vspace{-2mm}
+\caption{Dependence table. \label{fig:dependence}}
+\vspace{-6mm}
 \end{figure}
@@
 }
+\vspace{-2mm}
 \caption{Dependence analysis examples from {\tt circuit\_simulation}. \label{fig:depexamp}}
+\vspace{-6mm}
 \end{figure*}

### 2012-05-05 apps/AMR/plot_amr.py
unclobber plotting script

@@ -1,0 +1,252 @@
+#!/usr/bin/python
+
+import subprocess
+import sys, os, shutil
+import string, re
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.backends.backend_pdf import PdfPages
+
+home_dir="./"
+out_dir = home_dir + "figs/"
+expr_name="heat"
+legion_name="legion"
+
+problem_sizes=[4096,8192,16384]
+
+boxlib_pat = re.compile(expr_name+"_(?P<size>[0-9]+)_(?P<div>[0-9]+)_(?P<nodes>[0-9]+)\.stdio")
+boxlib_pat2= re.compile(expr_name+"_(?P<size>[0-9]+)_(?P<div>[0-9]+)_(?P<nodes>[0-9]+)_(?P<threads>[0-9]+)\.stdio")
+legion_pat = re.compile(legion_name+"_(?P<size>[0-9]+)_(?P<nodes>[0-9]+)\.stdio")
+success_pat = re.compile("\[[0-9]+ - [0-9a-f]+\] \{\w+\}\{\w+\}: SUCCESS!")
+time_pat    = re.compile("ELAPSED TIME\s+=\s+(?P<time>[0-9\.]+) s")
+updates_pat = re.compile("UPDATES PER SEC\s+=\s+(?P<updates>[0-9\.]+)")
+kernel_pat  = re.compile("\s+KERNEL\s+-\s+(?P<kernel>[0-9\.]+) s")
+copy_pat    = re.compile("\s+COPY\s+-\s+(?P<copy>[0-9\.]+) s")
+high_pat    = re.compile("\s+HIGH-LEVEL\s+-\s+(?P<high>[0-9\.]+) s")
+low_pat     = re.compile("\s+LOW-LEVEL\s+-\s+(?P<low>[0-9\.]+) s")
+mapper_pat  = re.compile("\s+MAPPER\s+-\s+(?P<mapper>[0-9\.]+) s")
+system_pat  = re.compile("\s+SYSTEM\s+-\s+(?P<system>[0-9\.]+) s")
+other_pat   = re.compile("\s+(?P<id>[0-9]+)\s+-\s+(?P<other>[0-9\.]+) s")
+runtime_pat = re.compile("\s+Run time \(s\) =\s+(?P<time>[0-9\.]+)")
+update_pat  = re.compile("\s+Cells Updates/s =\s+(?P<updates>[0-9\.E+]+)")
+
[…]

### 2012-05-05 doc/sc2012/experiments.tex
minor change to circuit-region-usage stuff
Minor edit

--- 2012-05-05 doc/sc2012/experiments.tex
@@
 Our second experiment is a port of the \emph{fluidanimate} benchmark
-from the PARSEC suite\cite{bienia11benchmarking}, which does a
+from the PARSEC benchmark suite\cite{bienia11benchmarking}, which does a
 particle-based simulation of an incompressible fluid.  Each particle

### 2012-05-05 doc/sc2012/experiments.tex

--- 2012-05-05 doc/sc2012/experiments.tex
@@
 GASNet memory and instances for just the shared and ghost nodes needed by each GPU
-are placed into that GPU's zero-copy memory.
+are placed into that GPU's zero-copy memory.  This enables the application kernels
+to access the data as well as the necessary inter-node exchanges of data via the GASNet
+memory.
 

### 2012-05-05 doc/sc2012/circuit.tex
Lots of little changes

@@
 Figure~\ref{sfig:part_fig:pvs} shows how the nodes
-in a small graph might be split into three pieces.  Blue nodes,
+in a small graph might be split into three pieces.  Blue (lighter) nodes,
 attached by wires only to nodes in the same piece, are 
-{\em private} to the piece.  Orange nodes, on the boundary of a
+{\em private} to the piece.  Orange (darker) nodes, on the boundary of a
 piece, are {\em shared} with (connected to) other pieces.

### 2012-05-05 doc/sc2012/experiments.tex
Lots of little changes

@@
 time, the mapper queries the list of GPUs in the machine and identifies each GPU's
-framebuffer memory and {\em zero-copy} memory (pinned memory that both the GPU and
-CPUs in the system can access directly).  Once the circuit is partitioned, the partitions
+framebuffer memory and {\em zero-copy} memory (pinned memory that both the GPUs and
+CPUs on a node can access directly).  Once the circuit is partitioned, the partitions
 are assigned a home GPU in round-robin fashion.  Every task related to that partition is
@@
 particle-based simulation of an incompressible fluid.  Each particle
-interacts only with other nearby particles. The benchmark divides the
+interacts only with nearby particles. The benchmark divides the
 space in which the fluid can move into a three-dimensional array of
@@
 application divides the array into {\em grids} and assigns each grid to
-a thread.  Per-cell locking safely accesses particles in
+a thread.  Per-particle locking safely accesses particles in
 cells that lie on the edge of a grid.  This fine-grained locking
 scheme along with the assumption of a shared address space gives
-good scaling in a multi-core processor, but rules out any attempt to run
+good scaling in a multi-core processor, but prohibits running
 beyond a single node.
@@
 a refined level linearly interpolate their values from the nearby
-cells at the next coarsest level.  The second phase performs the
+cells at the next coarser level.  The second phase performs the
 stencil computation on each cell in every level.  In the third phase,
@@
 level as well as across levels.  For
-cross-level cell dependences, mapping decisions must be made at
+cross-level cell dependences, optimal mapping decisions can only be made at
 runtime as the location of refinements are dynamically determined.
@@
 one thread per node to illustrate that in this application locality is
-significantly more important than parallelism.
+significantly more important than fine-grained data-parallelism.
 
@@
 begins to catch up (see Figure~\ref{fig:amr_keeneland})
-because our intra-level ghost-cell exchange 
-uses GASNet memory to exchange ghost cells, requiring a
+because our application's intra-level ghost-cell exchange algorithm
+uses GASNet memory to communicate ghost cells, requiring a
 linear increase in network traffic with the number of nodes.  BoxLib

### 2012-05-05 doc/sc2012/related.tex
Lots of little changes

@@
 of data is one-to-one.  Legion generalizes the Sequoia model by allowing for
-partitioning of pointer data structures through regions, enabling dynamic mappings through
+dynamic partitioning of pointer data structures through regions, enabling dynamic mappings through
 the mapper interface, and allowing different coherence
@@
 Domains are similar to logical regions in that they describe maps from indexes
-to same-typed objects.  Domains can create sub-domains by slicing the index sets from
+to objects.  Domains can create sub-domains by slicing the index sets from
 a parent domain.  Domains are a higher level concept than regions; the domain index sets 
 support dimensionality and iterators, whereas logical regions can only be accessed by pointers.  
-Second, the act of creating subdomains in Chapel does not track disjointness information,
+Also, the act of creating subdomains in Chapel does not track disjointness information,
 making it more challenging for the Chapel compiler or runtime to infer task independence.

### 2012-05-05 doc/sc2012/experiments.tex
More small edits

@@
 shown in Figure~\ref{fig:gpumapping}.  Wires and private node data are kept in each GPU's
-framebuffer at all times.  An instance of the ``all-shared nodes'' region is placed in
+framebuffer at all times.  An instance of the all-shared-nodes region is placed in
 GASNet memory and instances for just the shared and ghost nodes needed by each GPU

### 2012-05-05 doc/sc2012/soop.tex
punctuation

@@
 the SOOP considers, for each task $t$, the
-the regions $t$ uses and ensures $t$ waits on any
+regions $t$ uses and ensures $t$ waits on any
 earlier task whose region uses conflict with $t$'s.  There is a
@@
 use the same physical instance of the logical region.  Since tasks
-haven't mapped physical instances at this point in the pipeline {\tt Same} is always
-a mapping dependence.
+haven't mapped physical instances at this point in the pipeline,
+{\tt Same} is always a mapping dependence.
 %\footnote{For example, if two tasks wish to acess the region with atomic coherence, then if they are working on the same physical copy of the data atomicity can be guaranteed using, e.g., locking.

### 2012-05-05 doc/sc2012/soop.tex
Minor edits

--- 2012-05-05 doc/sc2012/soop.tex
@@
 {\tt Cont} 
-indicates a dependence contingent on whether it is a reader or
-writer that is Atomic (e.g. Atomic Read-Write and Simultaneous
-Read-Only will not cause a dependence).
+indicates a dependence contingent on the privileges of the two tasks
+(e.g. an Atomic Read-Write task and Simultaneous Read-Only task will not have a dependence).
 

### 2012-05-14 doc/popl2013/legion-types.tex
templates for popl2013 and asplos2013

--- 2012-05-04 doc/sc2012/legion.tex
@@
-\documentclass[conference]{IEEEtran}
+\documentclass[9pt,nocopyrightspace,preprint]{sigplanconf}
 
-\bibliographystyle{IEEEtran}
+\bibliographystyle{plainnat}
 \usepackage{amsmath,amsthm}
@@
 \usepackage{cancel}
-\usepackage{graphics}
 
@@
 \newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{observation}[theorem]{Observation}
 
@@
 
-\pdfpagewidth=8.5in
-\pdfpageheight=11in
-
 \title{Legion: Expressing Locality and Independence with Logical Regions}
-\author{
-\IEEEauthorblockN{Michael Bauer\IEEEauthorrefmark{1},
-%\authorinfo(Stanford University, email: \mbox{mebauer@cs.stanford.edu}
-Sean Treichler\IEEEauthorrefmark{1}, 
-%\authorinfo(Stanford University, email: \mbox{sjt@cs.stanford.edu}
-Elliott Slaughter\IEEEauthorrefmark{1}, 
-% \authorinfo(Stanford University, email: \mbox{slaughter@cs.stanford.edu}
-Alex Aiken\IEEEauthorrefmark{1}  
-%\authorinfo(Stanford University, email: \mbox{aiken@cs.stanford.edu}
-%\email{aiken@cs.stanford.edu}
-}
-\IEEEauthorblockA{
-\IEEEauthorrefmark{1}{Computer Science Department, Stanford University}
-}
-}
-
+\authorinfo{}{}{}
 \maketitle
@@
 \begin{abstract}
-Modern parallel architectures have both heterogeneous processors and
-deep, complex memory hierarchies.  We present Legion, a programming
-model and runtime system for achieving high performance on these
-machines.  Legion is organized around {\em logical regions}, which
-express both locality and independence of program data, and {\em
-  tasks}, functions that perform computations on regions.  We describe
-a runtime system that dynamically extracts parallelism from
-Legion programs, using a distributed, parallel scheduling algorithm
-that identifies both independent tasks and nested parallelism.  Legion
-also enables explicit, programmer controlled movement of data through
-the memory hierarchy and placement of tasks based on locality
-information via a novel mapping interface.  We evaluate our Legion
-implementation on three applications: fluid-flow on a regular grid,
-a three-level AMR code solving a heat diffusion equation, and 
-a circuit simulation.
-%
-% FIXME: put some performance numbers here
-%
+Limit is 12 pages, including bibliography and appendices.
+Deadlines: \\
+
+\begin{tabular}{ll}
+Paper registration & July 7, 2012 3:59am PDT \\
+Paper submission & July 11, 2012 3:59am PDT \\
+Author response & September 10 - September 13, 2012 \\
+Author notification & October 1, 2012
+\end{tabular}
 \end{abstract}
@@
 \label{sect:intro}
-%Put intro text here \cite{Fatahalian06}
+Introduction goes here.
 
-Modern parallel machines are increasingly complex, with deep,
-distributed memory hierarchies and heterogeneous processing units.  Because the
-costs of communication within these architectures vary by several orders of magnitude, 
-the penalty for mistakes in the placement of data or computation is usually very poor performance.  Thus,
-to achieve good performance the programmer
-and the programming system must reason about {\em locality} (data resident
-close to computation that uses it) and {\em independence} (computations
-operating on disjoint data, and therefore not requiring communication and able to be placed in possibly distant parts of the machine).  
-Most contemporary
-programming systems have no facilities for the programmer to express locality
-and independence.  The few languages that do focus primarily on
-array-based locality \cite{Fatahalian06,CHAPEL04,UPC99} and 
-avoid irregular pointer data structures.  
-%
-% FIXME
-% Sean would like a bit more on why other systems don't support pointer data structures (e.g., the difficulty of
-% alias analysis.  I thought about it and feel it is too much of a digression: mentioning that in the first paragraph
-% makes it sound like it will be an important part of the paper.
-%
-%encode such structures in arrays, which 
-%data structures encoded in arrays which are impossible for compilers
-%and runtime systems to reason about with regard to locality and independence.
-
-In this paper we describe Legion, a parallel programming system based
-on using {\em logical regions} to describe the organization of data
-and to make explicit relationships useful for reasoning about locality
-and independence.
-A logical region names a set of objects.  Logical regions are first-class 
-values in Legion and may be dynamically allocated, deleted and stored
-in data structures. Regions can also be passed as arguments
-to distinguished functions called {\em tasks} that access the data in
-those regions, providing locality information.  Logical regions may be
-{\em partitioned} into disjoint or aliased (overlapping) {\em subregions}, providing
-information for determining independence of computations.
-Furthermore, computations access logical regions with particular {\em
-  privileges} ({\em read-only}, {\em read-write}, and {\em reduce}) and {\em coherence} (e.g., {\em exclusive access} and {\em
-  atomic access}, among others).  Privileges express how a task may use
-its region arguments, providing data dependence information that
-is used to guide the extraction of parallelism.  For example, if two tasks
-access the same region with read-only privileges the two tasks
-can potentially be run in parallel.
-Coherence properties express the required semantics of concurrent region accesses.
-For example, if the program executes $f_1(r); f_2(r)$ and tasks $f_1$
-and $f_2$ both require {\em exclusive} access to region $r$, then Legion guarantees the result will
-be as if $f_1(r)$ completes before $f_2(r)$ begins.   On the other hand, if
-the tasks access $r$ with {\em atomic} coherence, then Legion
-guarantees only atomicity of the tasks with respect to $r$: either task $f_1(r)$
-appears to run entirely before $f_2(r)$ or vice versa.
-
-Logical regions do not commit to any particular layout of the data or
-placement in the machine.  At runtime, each logical region has one or
-more {\em physical instances} assigned to specific memories .  It is often useful to have multiple
-physical instances of a logical region (e.g., to duplicate read-only data, or to allow
-independent reductions that are later combined).
-
-To introduce the programming model, we present
-a circuit simulation in
-Section~\ref{sec:ex}, illustrating regions, tasks,
-permissions and coherence properties, the interactions between them,
-and how these building blocks are assembled into a Legion
-program.  Subsequent sections each describe a contribution in the
-implementation and evaluation of Legion:
-
-\input{code_ex}
-
-\begin{itemize}
-
-\item We define a {\em software out-of-order processor}, or SOOP, for
-  scheduling tasks with region arguments in a manner analogous to how
-  out-of-order hardware schedulers process instructions with register
-  arguments (Section~\ref{sec:soop}).  In addition to pipelining the execution of tasks over
-  several stages, our SOOP is distributed across the machine and is
-  also hierarchical and naturally extracts nested
-  parallelism (because tasks may recursively spawn subtasks).
-
-%\item We use a {\em deferred execution} model that extracts parallelism
-%even from highly irregular, unbalanced computations.  Instead of blocking,
-%an operation $o$ that cannot immediately execute because it is dependent
-%on some other operation $o'$ waits on an event that occurs when $o'$ completes.
-%Events can be created for the completion of any operation in Legion, and 
-%the scheduler can build long chains of deferred operations.  In Section~\ref{sec:exp}
-%we show that dataflow-like deferred execution gives us a significant performance improvement
-%over bulk synchronous global synchronization primitives such as barriers.
-
-\item Of central importance is how
-tasks are assigned (or {\em mapped}) to processors and how physical instances of logical regions are
-mapped to specific memory units (Section~\ref{sec:mapping}).  Often using
-application-specific information results in better
-mappings than a generic mapping strategy.  We describe
-a {\em mapping interface} that allows programmers to give the SOOP a specification
-of how to map tasks and regions for a specific application, or even part of an
-application.  This mapping API is designed so that
-any user-supplied mapping strategy can only affect the performance of applications,
-not their correctness.  
-%Thus, user mappings only affect the time and resources required
-%to run the program and cannot change the program's output.
-
-\item We present results of experiments on three applications:
-  fluid-flow on a regular grid, a three-level AMR code solving a
-  heat diffusion equation and a circuit simulation.
-  We compare each application with best available reference
-  versions on three different clusters of multicore processors with
-  GPUs, including the Keeneland supercomputer.
-%
-% FIXME Give the same numbers as in the abstract.
-%
-
-\end{itemize}
-
-\input{circuit}
-\input{part_fig}
-\input{soop}
-\input{mapping}
-\input{experiments}
-\input{related}
 \input{conclusion}
 
-\bibstyle{IEEEtran}
+{
+\small
 \bibliography{bibliography}
+}
 

### 2012-05-25 doc/popl2013/outline.tex
initial pass at outline of proofs

\section{Proof Outline}

\begin{itemize}

\item A logical region $L$ is a set of abstract memory locations $a_i, a_j, ... \in A$.

\item A coloring function $c : A \rightarrow C$ is a function from abstract memory locations to ``colors''.

\item A partitioning $P_{L,c} : C \rightarrow 2^A$ of a region $L$ with a coloring function $c$ is a function from a color to a subregion of L, satisfying:

\begin{itemize}

\item $\forall L,c,i,a . a \in L \wedge c(a) = i \leftrightarrow a \in P_{L,c}(i)$

\item $\forall L,c,i_1,i_2 . i_1 \neq i_2 \rightarrow P_{L,c}(i_1) * P_{L,c}(i_2)$

\end{itemize}

\item A static effect $E$ is a set of tuples $\langle L, op \rangle$ where $op \in \{ read, write \} \cup \{ reduce_f : f \text{is a reduction function} \}$.

\item Physical memory locations are represented by the set $m_1, m_2, ... \in M$, along with a mapping function $\alpha : M \rightarrow A$ that describes which abstract location a physical memory location corresponds to.

\item A dynamic trace $D = \langle \hat E, \hat O \rangle$ is a directed acyclic graph whose nodes $\hat E$ are actual memory operations $\langle m, op \rangle$ and edges $\hat O$ describe a partial ordering $\hat e_1 \prec \hat e_2$ of those memoory operations.  (Hmm...  Need notation that makes it clear that the same memory operation can be performed on the same memory address multiple times.)

\item Soundness of effects: If $\vdash t : ^ET$ then the mapping function $\alpha$ and dynamic trace $D = \langle \hat E, \hat O \rangle$ that results from evaluating $t$ have the following properties:

\begin{itemize}

\item $\forall m . \langle m, read \rangle \in \hat E \rightarrow \exists L . \alpha(m) \in L \wedge \langle L, read \rangle \in E$

\item $\forall m . \langle m, write \rangle \in \hat E \rightarrow \exists L . \alpha(m) \in L \wedge \langle L, write \rangle \in E$

\item $\forall m, f . \langle m, reduce_f \rangle \in \hat E \rightarrow ( \exists L . \alpha(m) \in L \wedge \langle L, reduce_f \rangle \in E ) \vee ( \exists L_1, L_2 . \alpha(m) \in L_1 \wedge \langle L_1, read \rangle \in E \wedge \alpha(m) \in L_2 \wedge \langle L_2, write \rangle \in E )$

\end{itemize}

\item Two dynamic subtraces $D_1 = \langle \hat E_1, \hat O_1 \rangle$, and $D_2 = \langle \hat E_2, \hat O_2 \rangle$ are ``memory ordered'' (written $D_1 \prec_D D_2$) within a larger trace $D = \langle \hat E_1 \cup \hat E_2 \cup \hat E', \hat O_1 \cup \hat O_2 \cup \hat O' \rangle$ if $D_2$ sees all the results of $D_1$'s memory operations and $D_1$ sees none of the results of $D_2$'s memory operations:

\begin{tabular}{l@{}l@{}l}
$D_1 \prec_D D_2 \leftrightarrow \forall$ & $\hat e_1 = \langle m_1, op_1 \rangle \in \hat E_1,$ \\
& $\hat e_2 = \langle m_2, op_2 \rangle \in \hat E_2 . \big($ & $m_1 \neq m_2 \vee$  \\
&& $( op_1 = read \wedge op_2 = read ) \vee$ \\
&& $( op_1 = reduce_f \wedge op_2 = reduce_f ) \vee$ \\
&& $( \hat e_1 \prec \hat e_2 \in \hat O' ) \big)$
\end{tabular}

\item Note that if $D_1$ and $D_2$ have no memory addresses in common, then you have $D_1 \prec_D D_2$ and $D_2 \prec_D D_1$ for all D.  Maybe $\prec$ is the wrong symbol to use?

\item Tasks are annotated with a coherence requirements $H_{excl}, H_{atom} \subseteq A$.  The default annotation is $H_{excl} = \bigcup_{\langle L, op \rangle \in E} L, H_{atom} = \emptyset$.

\item The runtime enforces an execution order $\prec_E$ between two tasks $S_1$ and $S_2$ as follows:

\begin{itemize}

\item Strict ordering: when the two tasks have exclusive coherence requirements on two regions that can't be proven disjoint (i.e. $\not\vdash H_{excl_1} * H_{excl_2}$), we enforce $S_1 \prec_E S_2$.

\item Serializability: when the two tasks have atomic coherence requirements on two regions that can't be proven disjoint, we enforce $S_1 \prec_E S_2 \vee S_2 \prec_E S_1$.

\end{itemize}

\item Execution order is stronger than memory order: $(S_1 \prec_E S_2) \rightarrow ( \forall \hat e_1 \in \hat E_1, \hat e_2 \in \hat E_2 . \hat e_1 \prec \hat e_2 ) \rightarrow (\forall D. D_1 \prec_D D_2)$.

\item Coherence of sibling tasks:  If sibling tasks $S_1$ and $S_2$ are program ordered (i.e. $S_1 \prec_P S_2$) within their parent task:

\begin{itemize}

\item Overlap in exclusivity requirements guarantees memory ordering: $H_{excl_1} \cap H_{excl_2} \neq \emptyset \rightarrow D_1 \prec_D D_2$.  (If $\vdash (E_1 \cap H_{excl_1}) * (E_2 \cap H_{excl_2})$, soundness of effects guarantees disjointness of memory addresses and therefore memory ordering.  If not, the runtime enforces execution order and therefore memory ordering.)

\item Overlap in atomic requirements guarantees serializability: $H_{atom_1} \cap H_{atom_2} \neq \emptyset \rightarrow D_1 \prec_D D_2 \vee D_2 \prec_D D_1$.  (Parallels proof above.)

\end{itemize}

\end{itemize}

### 2012-05-26 doc/sc2012/legion.tex
more changes
Minor edits
thanks

@@
 
-\title{Legion: Expressing Locality and Independence with Logical Regions}
+\title{Legion: Expressing Locality and Independence with Logical Regions\thanks{This work was supported by the Army High Performance Research Center, DARPA contract HR0011-10-9-0008, and NVIDIA.
+The views and conclusions contained in this document are those of the authors and should not be interpreted as representing the official policies, either expressed or implied, of the U.S.
+Government.}}
 \author{

### 2012-05-26 doc/asplos2013/legion-llr.tex

--- 2012-05-14 doc/popl2013/legion-types.tex
@@
-\documentclass[9pt,nocopyrightspace,preprint]{sigplanconf}
+\documentclass{sig-alternate}
 
 \bibliographystyle{plainnat}
-\usepackage{amsmath,amsthm}
+\usepackage{amsmath}
+%\usepackage{amsthm}
 \usepackage{listings}
@@
 \title{Legion: Expressing Locality and Independence with Logical Regions}
-\authorinfo{}{}{}
+\numberofauthors{3}
+\author{}
 \maketitle
@@
 \begin{abstract}
-Limit is 12 pages, including bibliography and appendices.
-Deadlines: \\
-
+Last year's limit was 12 pages, including bibliography and appendices.
+Deadlines:
 \begin{tabular}{ll}
-Paper registration & July 7, 2012 3:59am PDT \\
-Paper submission & July 11, 2012 3:59am PDT \\
-Author response & September 10 - September 13, 2012 \\
-Author notification & October 1, 2012
+Abstract deadline & July 16, 2012 3pm PDT \\
+Paper deadline & July 23, 2012 3pm PDT \\
+Author response & around October 15, 2012 \\
+Author notification & November 9, 2012
 \end{tabular}

### 2012-05-26 doc/sc2012/circuit.tex

@@
 wires subregion and updates all nodes those wires touch.  However,
-rather than using read/write permission for the nodes (which would
+rather than using read/write privilege for the nodes (which would
 serialize these tasks for correctness), the task

### 2012-05-26 doc/sc2012/code_ex.tex

@@
 
-void simulate_circuit(Circuit c, float dt) : RWE(c.r_all_nodes,c.r_all_wires)
+void simulate_circuit(Circuit c, float dt) : RWE(c.r_all_nodes, c.r_all_wires)
 {
@@
 void calc_new_currents(CircuitPiece piece):
-        RWE(piece.rw_pvt), ROE(piece.rn_pvt,piece.rn_shr,piece.rn_ghost) {
+        RWE(piece.rw_pvt), ROE(piece.rn_pvt, piece.rn_shr, piece.rn_ghost) {
   foreach(w : piece.rw_pvt)
@@
 void distribute_charge(CircuitPiece piece, float dt):
-        ROE(piece.rw_pvt), RdA(piece.rn_pvt,piece.rn_shr,piece.rn_ghost) {
+        ROE(piece.rw_pvt), RdA(piece.rn_pvt, piece.rn_shr, piece.rn_ghost) {
   foreach(w : piece.rw_pvt) {
@@
 
-void update_voltages(CircuitPiece piece): RWE(piece.rn_pvt,piece.rn_shr) {
+void update_voltages(CircuitPiece piece): RWE(piece.rn_pvt, piece.rn_shr) {
   foreach(n : piece.rn_pvt, piece.rn_shr) {

### 2012-05-26 doc/sc2012/experiments.tex

@@
 scheme along with the assumption of a shared address space gives
-good scaling in a multi-core processor, but prohibits running
+good scaling in a multicore processor, but prohibits running
 beyond a single node.

### 2012-05-26 doc/sc2012/mapping.tex

@@
 for inspecting properties of the machine, including a list of
-processors and their type (i.e. CPU, GPU) and a list of
-memories visible to each processor and their latencies and bandwidths.
-For brevity we only discuss three of the mapping interface calls:
+processors and their type (i.e. CPU, GPU), a list of
+memories visible to each processor, and their latencies and bandwidths.
+For brevity we only discuss the three most important interface calls:
 
@@
 
-The mapping interface has two desirable properties.  First, all mapper
-decisions are orthogonal to program correctness and can only impact
+The mapping interface has two desirable properties.  
+First, program correctness is unaffected by mapper decisions, which can only impact 
+%First, all mapper
+%decisions are orthogonal to program correctness and can only impact
 performance.  Regardless of where a mapper places a task or region,
 the SOOPs schedule tasks and copies in accordance with the privileges
-and coherence properties specified in the program.  Therefore when
+and coherence properties specified in the program.  Therefore, when
 writing a Legion application, a programmer can begin by using the
@@
 
-Each custom mapper extends the default mapper.  A programmer need only override the functions necessary 
-for creating a custom mapping.  Custom mappers can be used to create totally static mappings, 
+Each custom mapper extends the default mapper.  A programmer need only override the functions 
+%necessary for creating a custom mapping.  
+he wishes to customize.  Custom mappers can be used to create totally static mappings, 
 mappings that memoize their results, or even totally dynamic mappings.  We describe examples of 

### 2012-06-25 shll/shll_ast.py
updates

@@ -6,6 +6,53 @@ class DuplicateTypeDefinition(Exception):
 class DuplicateTaskDefinition(Exception):
     pass
 
+class TypeError(Exception):
+    pass
+
+class UnimplementedTypeCheck(TypeError):
+    def __init__(self, obj):
+        self.obj = obj
+
+    def __str__(self):
+        return "type checking not implemented for class " + self.obj.__class__.__name__
+
+class IsNullArgTypeError(TypeError):
+    def __init__(self, expr, argtype):
+        self.expr = expr
+        self.argtype = argtype
+
+    def __str__(self):
+        return "isnull requires a pointer argument - got: %s" % (self.argtype,)
+
+class NoReadPriviledgeError(TypeError):
+    def __init__(self, expr, ptrtype, privs):
+        self.expr = expr
+        self.ptrtype = ptrtype
+        self.privs = privs
+
+    def __str__(self):
+        return "read of %s attempted with the following priviledges:\n%s" % (self.ptrtype, self.privs)
+
+class LetInitTypeError(TypeError):
+    def __init__(self, expr, exptype, acttype):
+        self.expr = expr
+        self.exptype = exptype
+        self.acttype = acttype
+
+    def __str__(self):
+        return ("expected type %s for let initialization, got %s" %
+                (self.exptype,
+                 self.acttype))
+
+class WrongTypeClassError(TypeError):
+    def __init__(self, expr, expclass, acttype):
+        self.expr = expr
+        self.expclass = expclass
+        self.acttype = acttype
+
+    def __str__(self):
+        return ("expected a %s type, got: %s" % (self.expclass, self.acttype))
+
 class Program:
     def __init__(self):
         self.types = dict()
@@ -24,6 +71,13 @@ class Program:
         print "adding task", str(t)
         self.tasks[t.name] = t
         return self
+
+    def type_check(self):
+        for t in self.tasks.itervalues():
+            try:
+                t.type_check(self)
+            except TypeError as e:
+                print "type checking of %s failed: %s" % (t.name, e)
 
 class Typedef:
     def __init__(self, name, params, innertype):
@@ -50,6 +104,13 @@ class Taskdef:
         s += str(self.effects)
         return s
 
+    def type_check(self, pgrm):
+        env = dict(self.formals.byname)
+        privs = self.effects
+        consts = set()  # no constraints initially
+        bodytype = self.body.get_type(pgrm, env, privs, consts)
+        print "type of task %s body = %s" % (self.name, bodytype)
+
 class Type:
[…]

### 2012-06-25 shll/shll_lex.py

@@ -18,7 +18,7 @@ class Lexer:
         'INTVAL',
         ] + list(x.upper() for x in keywords)
 
-    literals = '={}()<>@*,:.+&|'
+    literals = '={}()<>@*,:.+&|[]'
 
     whitespace = " \t\r"
 

### 2012-06-25 shll/shll_parse.py

@@ -127,8 +127,8 @@ class Parser:
 
     def p_letexpr(self, p):
         'expr : LET ID ":" type "=" expr IN expr'
-        p[0] = LetExpr(valname = p[1], valtype = p[3], valexpr = p[5],
-                       body = p[7])
+        p[0] = LetExpr(valname = p[2], valtype = p[4], valexpr = p[6],
+                       body = p[8])
         
     def p_identexpr(self, p):
         'expr : ID'
@@ -192,13 +192,13 @@ class Parser:
                              body = p[11])
 
     def p_packexpr(self, p):
-        'expr : PACK expr AS type "(" region_list ")"'
+        'expr : PACK expr AS type "[" region_list "]"'
         p[0] = PackExpr(body = p[2],
                         rrtype = p[4],
                         rrparams = p[6])
 
     def p_unpackexpr(self, p):
-        'expr : UNPACK expr AS ID ":" type "(" region_list ")" IN expr'
+        'expr : UNPACK expr AS ID ":" type "[" region_list "]" IN expr'
         p[0] = UnpackExpr(argexpr = p[2],
                           name = p[4],
                           rrtype = p[6],
@@ -243,9 +243,10 @@ class Parser:
 
     def parse(self, src, **kwargs):
         self.lexer.input(src)
-        self.parser.parse(lexer = self.lexer)
+        return self.parser.parse(lexer = self.lexer)
 
 if __name__ == '__main__':
     import sys
     p = Parser()
-    p.parse(sys.stdin)
+    pgrm = p.parse(sys.stdin)
+    pgrm.type_check()

### 2012-06-25 doc/sc2012/legion.tex
thanks

@@
 \usepackage{graphics}
+\usepackage{url}
 
@@
 \newtheorem{observation}[theorem]{Observation}
+\IEEEoverridecommandlockouts
 
@@
 
-\title{Legion: Expressing Locality and Independence with Logical Regions\thanks{This work was supported by the Army High Performance Research Center, DARPA contract HR0011-10-9-0008, and NVIDIA.
-The views and conclusions contained in this document are those of the authors and should not be interpreted as representing the official policies, either expressed or implied, of the U.S.
-Government.}}
-\author{
-\IEEEauthorblockN{Michael Bauer\IEEEauthorrefmark{1},
+\title{Legion: Expressing Locality and Independence with Logical
+  Regions\thanks{This work was supported by the Army High Performance
+    Research Center (DARPA contract HR0011-10-9-0008), a subcontract
+    from Los Alamos National Laboratory for the DOE Office of Science,
+    Advanced Scientific Computing Research, and NVIDIA.  The views and
+    conclusions contained in this document are those of the authors
+    and should not be interpreted as representing the official
+    policies, either expressed or implied, of the U.S. Government.}}
+\author{ \IEEEauthorblockN{Michael Bauer\IEEEauthorrefmark{1},
 %\authorinfo(Stanford University, email: \mbox{mebauer@cs.stanford.edu}
@@
 placement in the machine.  At runtime, each logical region has one or
-more {\em physical instances} assigned to specific memories .  It is often useful to have multiple
+more {\em physical instances} assigned to specific memories.  It is often useful to have multiple
 physical instances of a logical region (e.g., to replicate read-only data, or to allow

### 2012-06-26 shll/shll_ast.py
fix for null pointers

@@ -216,7 +216,7 @@ class PtrType(Type):
                        #bindings.get(self.region))
 
     def isnullptrtype(self):
-        return ((self.elemtype is None) and (self.region is None))
+        return (self.elemtype is None)
 
     def equals(self, other, **kwargs):
         if not isinstance(other, PtrType):
@@ -895,7 +895,7 @@ class NullConstExpr(Expr):
 
     def get_type(self, pgrm, env, privs, consts):
         # TODO: is there a better way to handle this?
-        return PtrType(elemtype = None, region = None)
+        return PtrType(elemtype = None, regions = [ None ])
 
 class IntConstExpr(Expr):
     def __init__(self, value):

### 2012-06-26 doc/sc2012/legion.tex

@@
     from Los Alamos National Laboratory for the DOE Office of Science,
-    Advanced Scientific Computing Research, and NVIDIA.  The views and
+    Advanced Scientific Computing Research (DE-AC52-06NA25396), and NVIDIA.  The views and
     conclusions contained in this document are those of the authors

### 2012-06-27 doc/popl2013/langdef.tex
mostly-complete draft of language definition and typing rules

\begin{table*}
\centering
{\small
\begin{tabular}{cclr}

$T$ & ::= &  & types \\
  &$\mid$& bool $\;\;\;\mid\;\;\;$ int & base types \\
  &$\mid$& $\langle T_1, T_2 \rangle$ & tuple \\
  &$\mid$& $T@(r_1, \ldots, r_n)$ & pointer \\
  &$\mid$& $rr[r_1, \ldots, r_n] T where C$ & region relationship \\
\\
$C$ & ::= &  & constraints \\
  &$\mid$& $r_1 \leq r_2$ & subregion \\
  &$\mid$& $r_1 * r_2$ & disjointness \\
  &$\mid$& $C_1 \wedge C_2$ & conjunction \\
\\
$v$ & ::= & & values \\
  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & base values \\
  &$\mid$& $\langle v_1, v_2 \rangle$ & tuple \\
  &$\mid$& null $\;\;\;\mid\;\;\;$ $a$ & address \\
  &$\mid$& $\langle \langle r_1, \ldots, r_n, v\rangle \rangle$ & region relationship \\
\\
$bv$ & ::= & false $\;\;\;\mid\;\;\;$ true \\
\\
$iv$ & ::= & 0 $\;\;\;\mid\;\;\;$ 1 $\ldots$ \\
\\
$e$ & ::= & & expressions \\
  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ $\;\;\;\mid\;\;\;$ null & constants \\
  &$\mid$& $\langle e_1, e_2 \rangle$ $\;\;\;\mid\;\;\;$ $e$.1 $\;\;\;\mid\;\;\;$ $e$.2 & tuple \\
  &$\mid$& $id$ &  \\
  &$\mid$& $\text{new}\ T@r$ & allocation \\
  &$\mid$& $\text{isnull}(e)$ & \\
  &$\mid$& $\text{upregion}(e, r_1,\ldots,r_n)$ $\;\;\;\mid\;\;\;$ $\text{downregion}(e, r_1,\ldots,r_n)$ & \\
  &$\mid$& $\text{read}(e_1)$ $\;\;\;\mid\;\;\;$ $\text{write}(e_1, e_2)$ $\;\;\;\mid\;\;\;$ $\text{reduce}(id, e_1, e_2)$ & memory access \\
  &$\mid$& $e_1 + e_2$ & integer operations \\
  &$\mid$& $e_1 < e_2$ $\;\;\;\mid\;\;\;$ $e_1 > e_2$ & comparison operations \\
  &$\mid$& $\text{let}\ id : T = e_1 \text{in}\ e_2$ &  \\
  &$\mid$& $\text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3$ &  \\
  &$\mid$& $id(e_1, \ldots, e_n)$ $\;\;\;\mid\;\;\;$ $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & task calls \\
  &$\mid$& $\text{partition}\ r_p\ \text{using}\ id(e_1,\ldots,e_n)\ \text{as}\ r_1,\ldots,r_n\ \text{in}\ e_{n+1}$ &  \\
  &$\mid$& $\text{pack}\ e_1\ \text{as}\ T[r_1,\ldots,r_n]$ &  \\
  &$\mid$& $\text{unpack}\ e_1\ \text{as}\ id : T[r_1,\ldots,r_n]\ \text{in}\ e_2$ &  \\

\end{tabular}
}
\caption{Types}
\end{table*}

%\newcommand{\infrule}[2]{\displaystyle\frac{\displaystyle\strut{#1}}{\displaystyle\strut {#2}}}
\newcommand{\cinfrule}[3]{\parbox{14cm}{\hfil$\infrule{#1}{#2}$\hfil}\parbox{4cm}{$\,#3$\hfil}}
\newcommand{\finfrule}[2]{\vspace{10pt}\framebox{$\infrule{#1}{#2}$}\vspace{10pt}}

\newcommand{\infx}[2]{\infrule{\begin{array}{l}{#1}\end{array}}{#2}}

\newcommand{\regionexpand}[0]{[r'_1/r_1, \ldots, r'_n/r_n]}
\newcommand{\rtriple}[3]{\left({#1},{#2},{#3}\right)}
\newcommand{\rsingle}[1]{\rtriple{#1}{\emptyset}{\top}}
\newcommand{\rtripsub}[3]{\rtriple{#1}{\Phi_{#2}}{C_{#3}}}
\newcommand{\envsub}[2]{\Gamma, \Phi_{#1}, C_{#2} \vdash}
\newcommand{\typeenv}[0]{\Gamma, \Phi, C \vdash}
\newcommand{\oton}[1]{{#1}_1,\ldots,{#1}_n}
\newcommand{\otok}[2]{{#2}_1,\ldots,{#2}_{#1}}

\pagebreak

Type rules:

\framebox{
$
\begin{array}{c}
\typeenv bv : bool \\
\typeenv iv : int \\
\typeenv \text{null}\ : ??? \\
\typeenv \text{new}\ T@r : T@r \\

\end{array}
$
}

\finfrule{\typeenv e : \langle T_1,T_2 \rangle}{\typeenv e\text{.1}\ : T_1} \hspace{1cm}
\finfrule{\typeenv e : \langle T_1,T_2 \rangle}{\typeenv e\text{.2}\ : T_2} \\

\finfrule{\Gamma(id) = T}{\typeenv id : T} \\

\finfrule{\typeenv e : T@(\oton{r})}{\typeenv \text{isnull}(e) : bool} \\

\finfrule{\begin{array}{l}\typeenv e : T@(\otok{k}{r'}) \\ C \models \bigwedge_{i \in [1,k]}\bigvee_{j \in [1,n]} r'_i <= r_j\end{array}}{\typeenv upregion(e,\oton{r}) : T@(\oton{r})} \\

\finfrule{\begin{array}{l}\typeenv e : T@(\otok{k}{r'}) \\ C \models \bigwedge_{j \in [1,n]}\bigvee_{i \in [1,k]} r_j <= r'_i\end{array}}{\typeenv downregion(e,\oton{r}) : T@(\oton{r})} \\

\finfrule{\begin{array}{l}\typeenv e_1 : T@(\oton{r}) \\ \forall i, reads(r_i) \in \Phi\end{array}}{\typeenv \text{read}(e_1) : T} \\

\finfrule{\begin{array}{l}\typeenv e_1 : T@(\oton{r}) \\ \typeenv e_2 : T \\ \forall i, writes(r_i) \in \Phi\end{array}}{\typeenv \text{write}(e_1, e_2) : bool} \\

\finfrule{\begin{array}{l}\typeenv e_1 : T_1@(\oton{r}) \\ \typeenv e_2 : T_2 \\ \typeenv id : (T_1,T_2) \rightarrow T_1 !!! \\ \forall i, (reduces(id, r_i) \in \Phi) \vee ((reads(r_i) \in \Phi) \wedge (writes(r_i) \in \Phi))\end{array}}{\typeenv \text{reduce}(id, e_1, e_2) : bool} \\

\finfrule{\begin{array}{l}\typeenv e_1 : int \\ \typeenv e_2 : int\end{array}}{\typeenv e_1 + e_2 : int} \\

\finfrule{\begin{array}{l}\typeenv e_1 : int \\ \typeenv e_2 : int\end{array}}{\typeenv e_1 < e_2 : bool} \hspace{1cm}
\finfrule{\begin{array}{l}\typeenv e_1 : int \\ \typeenv e_2 : int\end{array}}{\typeenv e_1 > e_2 : bool} \\

\finfrule{\begin{array}{l}\typeenv e_1 : T_1 \\ \Gamma' = \Gamma[id/T_1] \\ \Gamma', \Phi, C \vdash e_2 : T_2 \end{array}}{\typeenv : \text{let}\ id : T_1 \text{in}\ e_2 : T_2} \\

\finfrule{\begin{array}{l}\typeenv e_1 : bool \\ \typeenv e_2 : T \\ \typeenv e_3 : T\end{array}}{\typeenv \text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3 : T} \\

\finfrule{
\begin{array}{l}
id = \text{task}(f_1 : T_1, \ldots f_n : T_n) : T_r, E \\
\typeenv e_i : T_i \\
E \subset \Phi !!!
\end{array}}
{\typeenv id(\oton{e}) : T_r} \\

\finfrule{
\begin{array}{l}
id = \text{task}[\otok{k}{r'}](f_1 : T_1, \ldots f_n : T_n) : T_r, E \\
\typeenv e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\
E[r_1/r'_1,\ldots,r_k/r'_k] \subset \Phi !!!
\end{array}}
{\typeenv id[\otok{k}{r}](\oton{e}) : T_r[r_1/r'_1,\ldots,r_k/r'_k]} \\

\finfrule{
\begin{array}{l}
id = \text{task}(f_1 : T_1, \ldots f_n : T_n) : int, E \\
\typeenv e_i : T_i \\
T_n = ??? \\\
E \subset \Phi !!! \\
C' = C \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{i \not= j} r_i * r_j \\
\Gamma, \Phi, C' \vdash e_{n+1} : T_{n+1}
\end{array}}
{\typeenv \text{partition}\ r_p\ \text{using}\ id(\otok{n-1}{e})\ \text{as}\ \otok{k}{r}\ \text{in}\ e_{n+1} : T_{n+1}} \\

\finfrule{
\begin{array}{l}
T_1 = rr[\otok{k}{r'}] T_2 where C_1 \\
C \models C_1[r_1/r'_1,\ldots,r_k/r'_k] \\
\typeenv e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]
\end{array}}
{\typeenv \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] : T_1} \\

\finfrule{
\begin{array}{l}
T_1 = rr[\otok{k}{r'}] T_2 where C_1 \\
\typeenv e_1 : T_1 \\
\Gamma' = \Gamma[T_2[r_1/r'_1,\ldots,r_k/r'_k] / id] \\
C' = C \wedge C_1[r_1/r'_1,\ldots,r_k/r'_k] \\
\Gamma', \Phi, C' \vdash e_2 : T_3
\end{array}}
{\typeenv \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 : T_3} \\

### 2012-06-27 doc/popl2013/langdef.tex
updates to type rules - on to operational semantics next

@@
+\newcommand{\oton}[1]{{#1}_1,\ldots,{#1}_n}
+\newcommand{\otok}[2]{{#2}_1,\ldots,{#2}_{#1}}
+
 \begin{table*}
@@
   &$\mid$& $\langle T_1, T_2 \rangle$ & tuple \\
-  &$\mid$& $T@(r_1, \ldots, r_n)$ & pointer \\
-  &$\mid$& $rr[r_1, \ldots, r_n] T where C$ & region relationship \\
+  &$\mid$& $T@(\oton{r})$ & pointer \\
+  &$\mid$& $\text{coloring}(r)$ & region coloring \\
+  &$\mid$& $\exists \oton{r}. T\text{ where }R$ & region relationship \\
+  &$\mid$& $\forall \oton{r}. (\oton{T}), \Phi \rightarrow T_r$ & functions \\
 \\
-$C$ & ::= &  & constraints \\
+$R$ & ::= & true & region constraints \\
   &$\mid$& $r_1 \leq r_2$ & subregion \\
   &$\mid$& $r_1 * r_2$ & disjointness \\
-  &$\mid$& $C_1 \wedge C_2$ & conjunction \\
+  &$\mid$& $R_1 \wedge R_2$ & conjunction \\
+\\
+$\Phi$ & ::= & $\{ \oton{\phi} \}$ & privileges \\
+$\phi$ & ::= & reads$(r)$ $\;\;\;\mid\;\;\;$ writes$(r)$ $\;\;\;\mid\;\;\;$ reduces$(id,r)$ & \\
 \\
@@
   &$\mid$& null $\;\;\;\mid\;\;\;$ $a$ & address \\
-  &$\mid$& $\langle \langle r_1, \ldots, r_n, v\rangle \rangle$ & region relationship \\
+  &$\mid$& $\{ (a_i, iv), \ldots \}$ & coloring \\
+  &$\mid$& $\langle \langle \oton{\rho}, v\rangle \rangle$ & region relationship instance \\
 \\
@@
 $e$ & ::= & & expressions \\
-  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ $\;\;\;\mid\;\;\;$ null & constants \\
+  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & constants \\
   &$\mid$& $\langle e_1, e_2 \rangle$ $\;\;\;\mid\;\;\;$ $e$.1 $\;\;\;\mid\;\;\;$ $e$.2 & tuple \\
   &$\mid$& $id$ &  \\
-  &$\mid$& $\text{new}\ T@r$ & allocation \\
-  &$\mid$& $\text{isnull}(e)$ & \\
+  &$\mid$& $\text{new}\ T@r$ $\;\;\;\mid\;\;\;$ $\text{null }T@r$ $\;\;\;\mid\;\;\;$ $\text{isnull}(e)$ & \\
   &$\mid$& $\text{upregion}(e, r_1,\ldots,r_n)$ $\;\;\;\mid\;\;\;$ $\text{downregion}(e, r_1,\ldots,r_n)$ & \\
   &$\mid$& $\text{read}(e_1)$ $\;\;\;\mid\;\;\;$ $\text{write}(e_1, e_2)$ $\;\;\;\mid\;\;\;$ $\text{reduce}(id, e_1, e_2)$ & memory access \\
+  &$\mid$& $\text{newcolor}\ r$ $\;\;\;\mid\;\;\;$ $\text{color}(e_1, e_2, e_3)$ & coloring operations \\
   &$\mid$& $e_1 + e_2$ & integer operations \\
-  &$\mid$& $e_1 < e_2$ $\;\;\;\mid\;\;\;$ $e_1 > e_2$ & comparison operations \\
+  &$\mid$& $e_1 < e_2$ & comparison operations \\
   &$\mid$& $\text{let}\ id : T = e_1 \text{in}\ e_2$ &  \\
   &$\mid$& $\text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3$ &  \\
-  &$\mid$& $id(e_1, \ldots, e_n)$ $\;\;\;\mid\;\;\;$ $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & task calls \\
-  &$\mid$& $\text{partition}\ r_p\ \text{using}\ id(e_1,\ldots,e_n)\ \text{as}\ r_1,\ldots,r_n\ \text{in}\ e_{n+1}$ &  \\
+  &$\mid$& $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & task calls \\
+  &$\mid$& $\text{partition}\ r_p\text{ using }e_1\text{ as }\oton{r}\text{ in }\ e_2$ &  \\
   &$\mid$& $\text{pack}\ e_1\ \text{as}\ T[r_1,\ldots,r_n]$ &  \\
@@
 \newcommand{\envsub}[2]{\Gamma, \Phi_{#1}, C_{#2} \vdash}
-\newcommand{\typeenv}[0]{\Gamma, \Phi, C \vdash}
-\newcommand{\oton}[1]{{#1}_1,\ldots,{#1}_n}
-\newcommand{\otok}[2]{{#2}_1,\ldots,{#2}_{#1}}
+\newcommand{\typeenv}[0]{\Gamma, \Phi, R \vdash}
 
@@
 \typeenv iv : int \\
-\typeenv \text{null}\ : ??? \\
-\typeenv \text{new}\ T@r : T@r \\
 
@@
 
+\framebox{$\typeenv \text{null }T@r : T@r$} \hspace{1cm}
+\framebox{$\typeenv \text{new }T@r : T@r$} \\
+
 \finfrule{\typeenv e : T@(\oton{r})}{\typeenv \text{isnull}(e) : bool} \\
 
-\finfrule{\begin{array}{l}\typeenv e : T@(\otok{k}{r'}) \\ C \models \bigwedge_{i \in [1,k]}\bigvee_{j \in [1,n]} r'_i <= r_j\end{array}}{\typeenv upregion(e,\oton{r}) : T@(\oton{r})} \\
+\finfrule{\begin{array}{l}\typeenv e : T@(\otok{k}{r'}) \\ R \models \bigwedge_{i \in [1,k]}\bigvee_{j \in [1,n]} r'_i \leq r_j\end{array}}{\typeenv upregion(e,\oton{r}) : T@(\oton{r})} \\
 
-\finfrule{\begin{array}{l}\typeenv e : T@(\otok{k}{r'}) \\ C \models \bigwedge_{j \in [1,n]}\bigvee_{i \in [1,k]} r_j <= r'_i\end{array}}{\typeenv downregion(e,\oton{r}) : T@(\oton{r})} \\
+\finfrule{\begin{array}{l}\typeenv e : T@(\otok{k}{r'}) \\ R \models \bigwedge_{j \in [1,n]}\bigvee_{i \in [1,k]} r_j \leq r'_i\end{array}}{\typeenv downregion(e,\oton{r}) : T@(\oton{r})} \\
 
@@
 
-\finfrule{\begin{array}{l}\typeenv e_1 : T_1@(\oton{r}) \\ \typeenv e_2 : T_2 \\ \typeenv id : (T_1,T_2) \rightarrow T_1 !!!
-\\ \forall i, (reduces(id, r_i) \in \Phi) \vee ((reads(r_i) \in \Phi) \wedge (writes(r_i) \in \Phi))\end{array}}{\typeenv \text{reduce}(id, e_1, e_2) : bool} \\
+\finfrule{\begin{array}{l}
+\Gamma(id) = (T_1, T_2), \emptyset \rightarrow T_1 \\
+\typeenv e_1 : T_1@(\oton{r}) \\
+\typeenv e_2 : T_2 \\
+\forall i, (reduces(id, r_i) \in \Phi) \vee ((reads(r_i) \in \Phi) \wedge (writes(r_i) \in \Phi))\end{array}}{\typeenv \text{reduce}(id, e_1, e_2) : bool} \\
 
-\finfrule{\begin{array}{l}\typeenv e_1 : int \\ \typeenv e_2 : int\end{array}}{\typeenv e_1 + e_2 : int} \\
+\framebox{$\typeenv \text{newcolor }r : \text{coloring}(r)$} \\
 
-\finfrule{\begin{array}{l}\typeenv e_1 : int \\ \typeenv e_2 : int\end{array}}{\typeenv e_1 < e_2 : bool} \hspace{1cm}
-\finfrule{\begin{array}{l}\typeenv e_1 : int \\ \typeenv e_2 : int\end{array}}{\typeenv e_1 > e_2 : bool} \\
+\finfrule{\begin{array}{l}
+\typeenv e_1 : \text{coloring}(r) \\
+\typeenv e_2 : T@r \\
+\typeenv e_3 : int
+\end{array}}
+{\typeenv \text{color}(e_1, e_2, e_3) : \text{coloring}(r)} \\
 
-\finfrule{\begin{array}{l}\typeenv e_1 : T_1 \\ \Gamma' = \Gamma[id/T_1] \\ \Gamma', \Phi, C \vdash e_2 : T_2 \end{array}}{\typeenv : \text{let}\ id : T_1 \text{in}\ e_2 : T_2} \\
+\finfrule{\begin{array}{l}\typeenv e_1 : int \\ \typeenv e_2 : int\end{array}}{\typeenv e_1 + e_2 : int} \hspace{1 cm}
+\finfrule{\begin{array}{l}\typeenv e_1 : int \\ \typeenv e_2 : int\end{array}}{\typeenv e_1 < e_2 : bool} \\
+
+\finfrule{\begin{array}{l}\typeenv e_1 : T_1 \\ \Gamma' = \Gamma[id/T_1] \\ \Gamma', \Phi, R \vdash e_2 : T_2 \end{array}}{\typeenv : \text{let}\ id : T_1 \text{in}\ e_2 : T_2} \\
 
@@
 \begin{array}{l}
-id = \text{task}(f_1 : T_1, \ldots f_n : T_n) : T_r, E \\
-\typeenv e_i : T_i \\
-E \subset \Phi !!!
-\end{array}}
-{\typeenv id(\oton{e}) : T_r} \\
-
-\finfrule{
-\begin{array}{l}
-id = \text{task}[\otok{k}{r'}](f_1 : T_1, \ldots f_n : T_n) : T_r, E \\
+\Gamma(id) = \forall \otok{k}{r'}.(\oton{T}),\Phi_i \rightarrow T_r \\
 \typeenv e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\
-E[r_1/r'_1,\ldots,r_k/r'_k] \subset \Phi !!!
+\Phi_i[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi
 \end{array}}
@@
 \begin{array}{l}
-id = \text{task}(f_1 : T_1, \ldots f_n : T_n) : int, E \\
-\typeenv e_i : T_i \\
-T_n = ??? \\\
-E \subset \Phi !!! \\
-C' = C \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{i \not= j} r_i * r_j \\
-\Gamma, \Phi, C' \vdash e_{n+1} : T_{n+1}
+\typeenv e_1 : \text{coloring}(r_p) \\
+R' = R \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{i \not= j} r_i * r_j \\
+\Gamma, \Phi, R' \vdash e_2 : T
 \end{array}}
-{\typeenv \text{partition}\ r_p\ \text{using}\ id(\otok{n-1}{e})\ \text{as}\ \otok{k}{r}\ \text{in}\ e_{n+1} : T_{n+1}} \\
+{\typeenv \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 : T} \\
 
@@
 \begin{array}{l}
-T_1 = rr[\otok{k}{r'}] T_2 where C_1 \\
-C \models C_1[r_1/r'_1,\ldots,r_k/r'_k] \\
+T_1 = \exists \otok{k}{r'}. T_2\text{ where }R_1 \\
+R \models R_1[r_1/r'_1,\ldots,r_k/r'_k] \\
 \typeenv e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]
@@
 \begin{array}{l}
-T_1 = rr[\otok{k}{r'}] T_2 where C_1 \\
+T_1 = \exists \otok{k}{r'}. T_2\text{ where }R_1 \\
 \typeenv e_1 : T_1 \\
 \Gamma' = \Gamma[T_2[r_1/r'_1,\ldots,r_k/r'_k] / id] \\
-C' = C \wedge C_1[r_1/r'_1,\ldots,r_k/r'_k] \\
-\Gamma', \Phi, C' \vdash e_2 : T_3
+R' = R \wedge R_1[r_1/r'_1,\ldots,r_k/r'_k] \\
+\Gamma', \Phi, R' \vdash e_2 : T_3
 \end{array}}

### 2012-06-29 doc/popl2013/langdef.tex
operational semantics

@@
   &$\mid$& $\text{coloring}(r)$ & region coloring \\
-  &$\mid$& $\exists \oton{r}. T\text{ where }R$ & region relationship \\
+  &$\mid$& $\exists \oton{r}. T\text{ where }\Omega$ & region relationship \\
   &$\mid$& $\forall \oton{r}. (\oton{T}), \Phi \rightarrow T_r$ & functions \\
 \\
-$R$ & ::= & true & region constraints \\
-  &$\mid$& $r_1 \leq r_2$ & subregion \\
+$\Omega$ & ::= & $\{ \oton{\omega} \}$ & region constraints \\
+$\omega$ & ::= & $r_1 \leq r_2$ & subregion \\
   &$\mid$& $r_1 * r_2$ & disjointness \\
-  &$\mid$& $R_1 \wedge R_2$ & conjunction \\
 \\
@@
 \newcommand{\envsub}[2]{\Gamma, \Phi_{#1}, C_{#2} \vdash}
-\newcommand{\typeenv}[0]{\Gamma, \Phi, R \vdash}
+\newcommand{\typeenv}[0]{\Gamma, \Phi, \Omega \vdash}
 
@@
 
-\framebox{
-$
-\begin{array}{c}
-\typeenv bv : bool \\
-\typeenv iv : int \\
-
-\end{array}
-$
-}
+\framebox{$\typeenv bv : bool$} \hspace{1cm}
+\framebox{$\typeenv iv : int$} \\
 
@@
 
-\finfrule{\begin{array}{l}\typeenv e : T@(\otok{k}{r'}) \\ R \models \bigwedge_{i \in [1,k]}\bigvee_{j \in [1,n]} r'_i \leq r_j\end{array}}{\typeenv upregion(e,\oton{r}) : T@(\oton{r})} \\
+\finfrule
+{\begin{array}{l}
+\typeenv e : T@(\otok{k}{r'}) \\
+\forall i. \exists j, r'_i \leq r_j \in \Omega^* \\
+\end{array}}
+{\typeenv upregion(e,\oton{r}) : T@(\oton{r})} \\
 
-\finfrule{\begin{array}{l}\typeenv e : T@(\otok{k}{r'}) \\ R \models \bigwedge_{j \in [1,n]}\bigvee_{i \in [1,k]} r_j \leq r'_i\end{array}}{\typeenv downregion(e,\oton{r}) : T@(\oton{r})} \\
+\finfrule
+{\begin{array}{l}
+\typeenv e : T@(\otok{k}{r'}) \\
+\forall j. \exists i, r_j \leq r'_i \in \Omega^* \\
+\end{array}}
+{\typeenv downregion(e,\oton{r}) : T@(\oton{r})} \\
 
-\finfrule{\begin{array}{l}\typeenv e_1 : T@(\oton{r}) \\ \forall i, reads(r_i) \in \Phi\end{array}}{\typeenv \text{read}(e_1) : T} \\
+\finfrule
+{\begin{array}{l}
+\typeenv e_1 : T@(\oton{r}) \\
+\forall i, reads(r_i) \in \Phi^*\end{array}}
+{\typeenv \text{read}(e_1) : T} \\
 
-\finfrule{\begin{array}{l}\typeenv e_1 : T@(\oton{r}) \\ \typeenv e_2 : T \\ \forall i, writes(r_i) \in \Phi\end{array}}{\typeenv \text{write}(e_1, e_2) : bool} \\
+\finfrule
+{\begin{array}{l}
+\typeenv e_1 : T@(\oton{r}) \\
+\typeenv e_2 : T \\
+\forall i, writes(r_i) \in \Phi^*
+\end{array}}
+{\typeenv \text{write}(e_1, e_2) : bool} \\
 
-\finfrule{\begin{array}{l}
+\finfrule
+{\begin{array}{l}
 \Gamma(id) = (T_1, T_2), \emptyset \rightarrow T_1 \\
@@
 \typeenv e_2 : T_2 \\
-\forall i, (reduces(id, r_i) \in \Phi) \vee ((reads(r_i) \in \Phi) \wedge (writes(r_i) \in \Phi))\end{array}}{\typeenv \text{reduce}(id, e_1, e_2) : bool} \\
+\forall i, reduces(id, r_i) \in \Phi^*
+\end{array}}
+{\typeenv \text{reduce}(id, e_1, e_2) : bool} \\
 
@@
 
-\finfrule{\begin{array}{l}\typeenv e_1 : T_1 \\ \Gamma' = \Gamma[id/T_1] \\ \Gamma', \Phi, R \vdash e_2 : T_2 \end{array}}{\typeenv : \text{let}\ id : T_1 \text{in}\ e_2 : T_2} \\
+\finfrule{\begin{array}{l}\typeenv e_1 : T_1 \\ \Gamma' = \Gamma[id/T_1] \\ \Gamma', \Phi, \Omega \vdash e_2 : T_2 \end{array}}{\typeenv : \text{let}\ id : T_1 \text{in}\ e_2 : T_2} \\
 
@@
 \typeenv e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\
-\Phi_i[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi
+\Phi_i[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*
 \end{array}}
@@
 \typeenv e_1 : \text{coloring}(r_p) \\
-R' = R \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{i \not= j} r_i * r_j \\
-\Gamma, \Phi, R' \vdash e_2 : T
+\Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{i \not= j} r_i * r_j \\
+\Gamma, \Phi, \Omega' \vdash e_2 : T
 \end{array}}
@@
 \begin{array}{l}
-T_1 = \exists \otok{k}{r'}. T_2\text{ where }R_1 \\
-R \models R_1[r_1/r'_1,\ldots,r_k/r'_k] \\
+T_1 = \exists \otok{k}{r'}. T_2\text{ where }\Omega_1 \\
+\Omega \models \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \\
 \typeenv e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]
@@
 \begin{array}{l}
-T_1 = \exists \otok{k}{r'}. T_2\text{ where }R_1 \\
+T_1 = \exists \otok{k}{r'}. T_2\text{ where }\Omega_1 \\
 \typeenv e_1 : T_1 \\
 \Gamma' = \Gamma[T_2[r_1/r'_1,\ldots,r_k/r'_k] / id] \\
-R' = R \wedge R_1[r_1/r'_1,\ldots,r_k/r'_k] \\
-\Gamma', \Phi, R' \vdash e_2 : T_3
+\Omega' = \Omega \wedge \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \\
+\Gamma', \Phi, \Omega' \vdash e_2 : T_3
 \end{array}}
 {\typeenv \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 : T_3} \\
+
+\subsection{Closure of region constraints and priviledges}
+
+\framebox{
+$\begin{array}{l}
+\Omega \subseteq \Omega^* \\
+r_i \leq r_j \in \Omega^*  \Rightarrow r_i \leq r_i \in \Omega^*\wedge r_j \leq r_j \in \Omega^* \\
+r_i \leq r_j \in \Omega^* \wedge r_j \leq r_k \in \Omega^* \Rightarrow r_i \leq r_k \in \Omega^* \\
+r_i \leq r_j \in \Omega^* \wedge r_j * r_k \in \Omega^* \Rightarrow r_i * r_k \in \Omega^* \\
+r_i * r_j \in \Omega^* \Rightarrow r_j * r_i \in \Omega^* \\
+\end{array}$
+} \\
+\framebox{
+$\begin{array}{l}
+\Phi \subseteq \Phi^* \\
+r_i \leq r_j \in \Omega^* \wedge reads(r_j) \in \Phi^* \Rightarrow reads(r_i) \in \Phi^* \\
+r_i \leq r_j \in \Omega^* \wedge writes(r_j) \in \Phi^* \Rightarrow writes(r_i) \in \Phi^* \\
+r_i \leq r_j \in \Omega^* \wedge reduces(id, r_j) \in \Phi^* \Rightarrow reduces(id, r_i) \in \Phi^* \\
+reads(r) \in \Phi^* \wedge writes(r) \in \Phi^* \wedge \\
+\hspace{1cm} \Gamma(id) = \forall \otok{k}{r}. (\oton{T}), E \rightarrow T_r \Rightarrow reduces(id, r) \in \Phi^*
+\end{array}$
+}
+
+\subsection{Operational Semantics}
+
+Operational semantics rules have the following form: \\
+
+\newcommand{\opsenv}{M, L, H, S, C \vdash}
+\newcommand{\opsenvm}[1]{{#1}, L, H, S, C \vdash}
+\newcommand{\opsenvs}[1]{M, L, H, {#1}, C \vdash}
+\newcommand{\opsenvls}[2]{M, {#1}, H, {#2}, C \vdash}
+\newcommand{\opsenvms}[2]{{#1}, L, H, {#2}, C \vdash}
+\newcommand{\opsenvmls}[3]{{#1}, {#2}, H, {#3}, C \vdash}
+
+\framebox{$\opsenv e \mapsto v, S', E$}
+
+where: 
+\begin{itemize}
+\item $M$ is the mapping from logical regions $r_i$ in types to physical regions $\rho_i$, which are sets of memory locations
+\item $L$ is the mapping of local variable names to their values
+\item $H$ is the heap typing, mapping from memory locations to types
+\item $S$ is the store, mapping from memory locations $l_i$ to values
+\item $C$ is the ``clobber set'', a set of memory locations that may be accessed by concurrently executing operations (??? does this need to be split into read/write/reduce?)
+\item $e$ is the expression being evaluated
+\item $v$ is the result of the evaluation
+\item $E$ is a memory trace (i.e. the effects) of the execution, a list of read/write/reduce operations in the order they were performed
+\end{itemize}
+
+\framebox{$\opsenv bv \mapsto bv, []$} \hspace{1cm}
+\framebox{$\opsenv iv \mapsto iv, []$} \\
+
+\finfrule
+{\opsenv e \mapsto \langle v_1, v_2 \rangle, E}
+{\opsenv e\text{.1} \mapsto v_1, E} \hspace{1cm}
+\finfrule
+{\opsenv e \mapsto \langle v_1, v_2 \rangle, E}
+{\opsenv e\text{.2} \mapsto v_2, E} \\
+
+\finfrule
+{L(id) = v}
+{\opsenv id \mapsto v, []} \\
+
+\framebox{$\opsenv \text{null }T@r \mapsto null, []$} \\
+
+\finfrule
+{\begin{array}{l}
+l \in M(r) \\
+H(l) = T[M]
+\end{array}}
+{\opsenv \text{new }T@r \mapsto l, []} \\
+
+\finfrule
+{\opsenv e \mapsto v, E}
+{\opsenv \text{isnull}(e) \mapsto false, E} \\
+
+\finfrule
+{\opsenv e \mapsto null, E}
+{\opsenv \text{isnull}(e) \mapsto true, E} \\
+
+\finfrule
+{\opsenv e \mapsto v, E}
+{\opsenv \text{upregion}(e, \oton{r}) \mapsto v, E}
+
+\finfrule
+{\begin{array}{l}
+\opsenv e \mapsto v, E \\
+v' = \begin{cases}
+v, & \text{if $v \not= null$ and $\exists i, v \in M(r_i)$}. \\
+null, & \text{otherwise}.
+\end{cases}
+\end{array}}
+{\opsenv \text{downregion}(e, \oton{r}) \mapsto v', E} \\
+
+\finfrule
+{\begin{array}{l}
+\opsenv e_1 \mapsto l, E \\
+S' = \text{apply}(S, E, C) \\
+v = S'[l]
+\end{array}}
+{\opsenv \text{read}(e_1) \mapsto v, E ++ [ read(l, v) ]} \\
+
+\finfrule
+{\begin{array}{l}
+\opsenv e_1 \mapsto l, E_1 \\
+S' = \text{apply}(S, E_1, C) \\
+\opsenvs{S'} e_2 \mapsto v, E_2 \\
+\end{array}}
+{\opsenv \text{write}(e_1, e_2) \mapsto true, E_1 ++ E_2 ++ [ write(l, v) ]} \\
+
+\finfrule
+{\begin{array}{l}
+\opsenv e_1 \mapsto l, E_1 \\
+S' = \text{apply}(S, E_1, C) \\
+\opsenvs{S'} e_2 \mapsto v, E_2 \\
+\end{array}}
+{\opsenv \text{reduce}(id, e_1, e_2) \mapsto true, E_1 ++ E_2 ++ [ reduce_{id}(l, v) ]} \\
+
+\framebox{$\opsenv \text{newcolor }r \mapsto \emptyset, []$} \\
+
+\finfrule
+{\begin{array}{l}
+\opsenv e_1 \mapsto K, E_1 \\
+S' = \text{apply}(S, E_1, C) \\
+\opsenvs{S'} e_2 \mapsto l, E_2 \\
+S'' = \text{apply}(S', E_2, C) \\
+\opsenvs{S''} e_3 \mapsto v, E_3 \\
+K' = \{ (l,v) \} \cup \{ (l_i,v_i) : (l_i,v_i) \in K \wedge l \not= l_i \}
+\end{array}}
+{\opsenv \text{color}(e_1, e_2, e_3) \mapsto K', E_1 ++ E_2 ++ E_3} \\
+
+\finfrule
+{\begin{array}{l}
+\opsenv e_1 \mapsto v_1, E_1 \\
+S' = \text{apply}(S, E_1, C) \\
+\opsenvs{S'} e_2 \mapsto v_2, E_2 \\
+v' = v_1 + v_2
+\end{array}}
+{\opsenv e_1 + e_2 \mapsto v', E_1 ++ E_2} \hspace{1cm}
+
+\finfrule
+{\begin{array}{l}
+\opsenv e_1 \mapsto v_1, E_1 \\
+S' = \text{apply}(S, E_1, C) \\
+\opsenvs{S'} e_2 \mapsto v_2, E_2 \\
+v' = \begin{cases}
+true, & \text{if $v_1 < v_2$}. \\
+false, & \text{otherwise}.
+\end{cases}
+\end{array}}
+{\opsenv e_1 < e_2 \mapsto v', E_1 ++ E_2} \\
+
+\finfrule
+{\begin{array}{l}
+\opsenv e_1 \mapsto v_1, E_1 \\
+L' = L[v_1/id] \\
+S' = \text{apply}(S, E_1, C) \\
+\opsenvls{L'}{S'} e_2 \mapsto v_2, E_2
+\end{array}}
+{\opsenv \text{let }id : T = e_1\text{ in }e_2 \mapsto v_2, E_1 ++ E_2} \\
+
+\finfrule
+{\begin{array}{l}
+\opsenv e_1 \mapsto true, E_1 \\
+S' = \text{apply}(S, E_1, C) \\
+\opsenvls{L'}{S'} e_2 \mapsto v_2, E_2
+\end{array}}
+{\opsenv \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_2, E_1 ++ E_2} \\
+
+\finfrule
+{\begin{array}{l}
+\opsenv e_1 \mapsto false, E_1 \\
+S' = \text{apply}(S, E_1, C) \\
+\opsenvls{L'}{S'} e_3 \mapsto v_3, E_3
+\end{array}}
+{\opsenv \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_3, E_1 ++ E_3} \\
+
+\finfrule
+{\begin{array}{l}
+\opsenv e_1 \mapsto v_1, E_1 \\
+S_1 = \text{apply}(S, E_1, C) \\
+\opsenv e_2 \mapsto v_2, E_2 \\
+\ldots \\
+S_n = \text{apply}(S_{n-1}, E_n, C) \\
+task id[\otok{k}{r'}](f_1 : T_1, \ldots, f_n : T_n) : T_r, E = e_{n+1} \\
+M' = \{ (r'_1, M(r_1)), \ldots (r'_k, M(r_k)) \} \\
+L' = \{ (f_1, v_1), \ldots, (f_n, v_n) \} \\
+\opsenvmls{M'}{L'}{S_n} e_{n+1} \mapsto v_{n+1}, E_{n+1}
+\end{array}}
+{\opsenv id[\otok{k}{r}](\oton{e}) \mapsto v_{n+1}, E_1 ++ \ldots ++ E_{n+1}} \\
+
+\finfrule{
+\begin{array}{l}
+\opsenv e_1 \mapsto K, E_1 \\
+\rho_i = \{ l : (l, i) \in K \} \\
+M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \\
+S' = \text{apply}(S, E_1, C) \\
+\opsenvms{M'}{S'} e_2 \mapsto v, E_2
+\end{array}}
+{\opsenv \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 \mapsto v, E_1 ++ E_2} \\
+
+\finfrule{
+\begin{array}{l}
+\opsenv e_1 \mapsto v, E \\
+\rho_i = M[r_i] \\
+v' = \langle \langle \otok{k}{\rho}, v \rangle \rangle
+\end{array}}
+{\opsenv \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] \mapsto v', E} \\
+
+\finfrule{
+\begin{array}{l}
+\opsenv e_1 \mapsto \langle \langle \otok{k}{\rho} , v_1 \rangle \rangle, E_1 \\
+M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \\
+L' = L[v_1/id] \\
+S' = \text{apply}(S, E_1, C) \\
+\opsenvmls{M'}{L'}{S'} e_2 \mapsto v_2, E_2
+\end{array}}
+{\opsenv \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 \mapsto v_2, E_1 ++ E_2 } \\
+
+\subsection{Soundness of Effects}
+
+If
+\begin{enumerate}
+\item $\typeenv e : T$
+\item $\opsenv e \mapsto v, E$
+\end{enumerate}
+then for all effects $\epsilon$ in $E$,
+\begin{enumerate}
+\item if $\epsilon = read(l, v)$, then for some $r$, $reads(r) \in \Phi \wedge l \in M(r)$
+\item if $\epsilon = write(l, v)$, then for some $r$, $writes(r) \in \Phi \wedge l \in M(r)$
+\item if $\epsilon = reduce(id, l, v)$, then for some $r$, $reduces(id,r) \in \Phi \wedge l \in M(r)$
+\end{enumerate}

### 2012-06-29 doc/popl2013/langdef.tex
updates

@@
 \Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{i \not= j} r_i * r_j \\
-\Gamma, \Phi, \Omega' \vdash e_2 : T
+\Gamma, \Phi, \Omega' \vdash e_2 : T \\
+r_i \not\in \textit{regions\_of}(\Gamma, T)
 \end{array}}
@@
 \Omega' = \Omega \wedge \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \\
-\Gamma', \Phi, \Omega' \vdash e_2 : T_3
+\Gamma', \Phi, \Omega' \vdash e_2 : T_3 \\
+r_i \not\in \textit{regions\_of}(\Gamma, T_1, T_3)
 \end{array}}
@@
 \end{enumerate}
+
+\subsection{Strategy}
+
+\newcommand{\llbracket}{[\![}
+\newcommand{\rrbracket}{]\!]}
+
+The mapping $M$ is a function from logical region names $r$ to physical regions $\rho$.  It will
+be convenient to apply $M$ to types and even functions that operate on types by defining
+$\hat x = M \llbracket x \rrbracket$ to be $x$ with all logical region names $r_i$ replaced by their
+corresponding physical region $\rho_i = M(r_i)$.  Note that $M \llbracket x \rrbracket$ is not
+well-formed if it contains a logical region name $r_j \not\in dom(M)$. \\
+
+Soundness is proven using induction over the structure of an expression.  A number of properties
+are assumed and/or proven at each step: 
+\begin{enumerate}
+\item $M \sim \Omega \Leftrightarrow (\forall r_i, r_j.
+r_i \leq r_j \in \Omega \rightarrow M(r_i) \subseteq M(r_j)) \wedge (\forall r_i, r_j.
+r_i * r_j \in \Omega \rightarrow M(r_i) \cap M(r_j) = \emptyset)$, that M is an $\Omega$-satisfying mapping of logical regions to physical regions
+\item $\hat T = M \llbracket T \rrbracket$, a {\em physical type}, constructed by replacing all of the free region names $r$ in T with the physical regions $\rho = M(r)$
+\item $v \sim_H \hat T$, value consistency with respect to a heap typing.  Value consistency is
+itself defined inductively over the structure of a type:
+\begin{enumerate}
+\item $bv \sim_H bool$
+\item $iv \sim_H int$
+\item $null \ sim_H T@\rho$
+\item $l \sim_H T@\rho \Leftrightarrow (l \in \rho) \wedge (H(l) \sim_H T)$
+\item $\langle v_1, v_2 \rangle \sim_H \langle \hat T_1, \hat T_2 \rangle \Leftrightarrow (v_1 \sim_H \hat T_1) \wedge (v_2 \sim_H \hat T_2)$
+\item $\langle \langle \oton{\rho}, v \rangle \rangle \sim_H \text{rr}[\oton{r}] \hat T\text{ where }\Omega \Leftrightarrow (v \sim_H \hat T[\rho_1/r_1, \ldots \rho_n/r_n]) \wedge (\{ (r_i, \rho_i) \} \sim \Omega)$
+\end{enumerate}
+\item $L \sim_H M \llbracket \Gamma \rrbracket \Leftrightarrow \forall (id,v) \in L. v \sim_H M \llbracket \Gamma \rrbracket (id) $, local stack consistency with respect to a heap typing
+\item $S \sim H \Leftrightarrow \forall (l,v) \in S. v \sim_H H(l)$, store consistency with respect to a given heap typing
+\item $E \sim H \Leftrightarrow (\forall l, v.
+write(l,v) \in E \rightarrow v \sim_H H(l)) \wedge (\forall id, l, v.
+reduce_{id}(l, v) \in E \rightarrow (\hat \Gamma (id) = (\hat T_1, \hat T_2), \emptyset \rightarrow \hat T_1) \wedge H(l) = \hat T_1 \wedge v \sim_H \hat T_2)$, effect consistnecy with respect to a given heap typing
+\end{enumerate}
+
+Armed with all of these, we look at the following property:
+\begin{enumerate}
+\item $E \sim_H \Phi$, containment of effects within privileges
+\end{enumerate}
+
+and prove
+
+\finfrule
+{\begin{array}{l}
+\typeenv e : T \\
+\opsenv e \mapsto v, E \\
+M \sim \Omega \hspace{1cm}
+L \sim_H \Gamma \hspace{1cm}
+S \sim H
+\end{array}}
+{
+v \sim_H M \llbracket T \rrbracket \hspace{1cm}
+E \sim_H \Phi
+}
+
+A few lemmas will be useful:
+
+\finfrule
+{\begin{array}{l}
+S \sim H \hspace{1cm}
+E \sim H
+\end{array}}
+{ \text{apply}(S, E, C) \sim H }
+
+\finfrule
+{\begin{array}{l}
+E_1 \sim_H \Phi \hspace{1cm}
+E_2 \sim_H \Phi
+\end{array}}
+{ E_1 ++ E_2 \sim_H \Phi }

### 2012-06-30 doc/popl2013/langdef.tex
proof table to fill in

@@
 \newcommand{\otok}[2]{{#2}_1,\ldots,{#2}_{#1}}
+\newcommand{\dplus}{\text{++}}
 
@@
 \end{array}}
-{\opsenv \text{read}(e_1) \mapsto v, E ++ [ read(l, v) ]} \\
+{\opsenv \text{read}(e_1) \mapsto v, E \dplus [ read(l, v) ]} \\
 
@@
 \end{array}}
-{\opsenv \text{write}(e_1, e_2) \mapsto true, E_1 ++ E_2 ++ [ write(l, v) ]} \\
+{\opsenv \text{write}(e_1, e_2) \mapsto true, E_1 \dplus E_2 \dplus [ write(l, v) ]} \\
 
@@
 \end{array}}
-{\opsenv \text{reduce}(id, e_1, e_2) \mapsto true, E_1 ++ E_2 ++ [ reduce_{id}(l, v) ]} \\
+{\opsenv \text{reduce}(id, e_1, e_2) \mapsto true, E_1 \dplus E_2 \dplus [ reduce_{id}(l, v) ]} \\
 
@@
 \end{array}}
-{\opsenv \text{color}(e_1, e_2, e_3) \mapsto K', E_1 ++ E_2 ++ E_3} \\
+{\opsenv \text{color}(e_1, e_2, e_3) \mapsto K', E_1 \dplus E_2 \dplus E_3} \\
 
@@
 \end{array}}
-{\opsenv e_1 + e_2 \mapsto v', E_1 ++ E_2} \hspace{1cm}
+{\opsenv e_1 + e_2 \mapsto v', E_1 \dplus E_2} \hspace{1cm}
 
@@
 \end{array}}
-{\opsenv e_1 < e_2 \mapsto v', E_1 ++ E_2} \\
+{\opsenv e_1 < e_2 \mapsto v', E_1 \dplus E_2} \\
 
@@
 \end{array}}
-{\opsenv \text{let }id : T = e_1\text{ in }e_2 \mapsto v_2, E_1 ++ E_2} \\
+{\opsenv \text{let }id : T = e_1\text{ in }e_2 \mapsto v_2, E_1 \dplus E_2} \\
 
@@
 \end{array}}
-{\opsenv \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_2, E_1 ++ E_2} \\
+{\opsenv \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_2, E_1 \dplus E_2} \\
 
@@
 \end{array}}
-{\opsenv \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_3, E_1 ++ E_3} \\
+{\opsenv \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_3, E_1 \dplus E_3} \\
 
@@
 \end{array}}
-{\opsenv id[\otok{k}{r}](\oton{e}) \mapsto v_{n+1}, E_1 ++ \ldots ++ E_{n+1}} \\
+{\opsenv id[\otok{k}{r}](\oton{e}) \mapsto v_{n+1}, E_1 \dplus \ldots \dplus E_{n+1}} \\
 
@@
 \end{array}}
-{\opsenv \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 \mapsto v, E_1 ++ E_2} \\
+{\opsenv \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 \mapsto v, E_1 \dplus E_2} \\
 
@@
 \end{array}}
-{\opsenv \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 \mapsto v_2, E_1 ++ E_2 } \\
+{\opsenv \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 \mapsto v_2, E_1 \dplus E_2 } \\
 
@@
 \end{enumerate}
-then for all effects $\epsilon$ in $E$,
+we would like to prove that all of the dynamic effects described by $E$ are
+included in the static privileges $\Phi$.  A judgement $E : \Phi$ means that
+for all effects $\epsilon$ in $E$,
 \begin{enumerate}
@@
 v \sim_H M \llbracket T \rrbracket \hspace{1cm}
-E \sim_H \Phi
+E \sim H \hspace{1cm}
+E : \Phi
 }
@@
 {\begin{array}{l}
-E_1 \sim_H \Phi \hspace{1cm}
-E_2 \sim_H \Phi
+E_1 : \Phi \hspace{1cm}
+E_2 : \Phi
 \end{array}}
-{ E_1 ++ E_2 \sim_H \Phi }
+{ E_1 \dplus E_2 : \Phi } \\
+
+\framebox{$M \sim \Omega \Leftrightarrow M \sim \Omega^*$} \\
+
+\framebox{$E : \Phi \Leftrightarrow E : \Phi^*$}
+
+\begin{table*}
+\centering
+{\small
+\begin{tabular}{l|p{5cm}|p{5cm}|p{5cm}}
+{\bf Rule} & 
+\multicolumn{1}{|c|}{$\mathbf{v \sim_H M \llbracket T \rrbracket}$} &
+\multicolumn{1}{|c|}{$\mathbf{E \sim H}$} &
+\multicolumn{1}{|c|}{$\mathbf{E : \Phi}$} \\
+\hline
+BoolConst & & & \\
+IntConst & & & \\
+TupleBuild & & & \\
+TupleField & & & \\
+LocalVar & & & \\
+NullPtr & & & \\
+NewPtr & & & \\
+IsNull & & & \\
+UpRegion & & & \\
+DownRegion & & & \\
+Read & & & \\
+Write & & & \\
+Reduce & & & \\
+Newcolor & & & \\
+Color & & & \\
+IntMath & & & \\
+IntCmp & & & \\
+Let & & & \\
+IfElse & & & \\
+TaskCall & & & \\
+Partition & & & \\
+Pack & & & \\
+Unpack & & &
+\end{tabular}}
+\caption{Proof pieces}
+\end{table*}

### 2012-06-30 doc/popl2013/langdef.tex
more stuff (not all of it good)

@@
 {\begin{array}{l}
+E_1 \sim H \hspace{1cm}
+E_2 \sim H
+\end{array}}
+{ E_1 \dplus E_2 \sim H } \\
+
+\finfrule
+{\begin{array}{l}
 E_1 : \Phi \hspace{1cm}
@@
 \end{table*}
+
+\subsection{Noninterference}
+
+For a given expression $e$, we can consider $v^o$ to be a ``sequential'' result and $E^o$ to be the corresponding ``sequential'' effects if: \\
+
+\framebox{$M, L, H, S, \emptyset \vdash e \mapsto v^o, E^o$} \\
+
+(i.e. if there are no manipulations of the store other than what's done by
+the evaluation of the expression $e$.)  We now wish to consider what parallel
+evaluations of this expression with other expression evaluations will still
+yield our ``sequential'' result. \\
+
+Two effects are considered to be noninterfering, written $E_1 \# E_2$ if for
+every $\epsilon_1$ in $E_1$ and every $\epsilon_2$ in $E_2$,
+\begin{enumerate}
+\item $\epsilon_1$ and $\epsilon_2$ are to different memory locations, or
+\item $\epsilon_1 = read(l, v_1)$ and $\epsilon_2 = read(l, v_2)$, or
+\item $\epsilon_1 = reduce_{id}(l, v_1)$ and $\epsilon_2 = reduce_{id}(l, v_2)$
+\end{enumerate}
+
+We define nonintereference (under a given mapping $M$) of static privileges
+$\Phi_1$ and $\Phi_2$ similarly.  We use the following table for determining
+noninterference of two individual priviledges $\phi_1$ and $\phi_2$:
+
+\begin{tabular}{ccccc}
+\multicolumn{2}{l}{$\phi_1 \#_M \phi_2 ?$} & \multicolumn{3}{c}{$\phi_2$} \\
+\multicolumn{2}{l}{} & $reads(r_2)$ & $writes(r_2)$ & $reduces(id_2, r_2)$ \\
+\multirow{3}{*}{$\phi_1$} & $reads(r_1)$ & true & $M(r_1) \cap M(r_2) = \emptyset$ & $M(r_1) \cap M(r_2) = \emptyset$\\
+& $writes(r_1)$ & $M(r_1) \cap M(r_2) = \emptyset$ & $M(r_1) \cap M(r_2) = \emptyset$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
+& $reduces(id_1, r_1)$ & $M(r_1) \cap M(r_2) = \emptyset$ & $M(r_1) \cap M(r_2) = \emptyset$ & $\begin{array}{l}id_1 = id_2 \vee \\ M(r_1) \cap M(r_2) = \emptyset\end{array}$
+\end{tabular}
+
+We can then define: \\
+
+\framebox{$\Phi_1 \#_M \Phi_2 \Leftrightarrow \displaystyle\bigwedge_{\begin{array}{c}\phi_1 \in \Phi_1 \\ \phi_2 \in \Phi_2 \end{array}} \phi_1 \#_M \phi_2$} \\
+
+From the soundness of effects, result, we can show that for two expressions
+$e_1$ and $e_2$, if their privileges are noninterfering, any potential memory
+traces from their evaluation must also be noninterfering.  Formally:
+
+\finfrule
+{\begin{array}{l}
+\Gamma, \Phi_1, \Omega \vdash e_1 : T_1 \\
+\Gamma, \Phi_2, \Omega \vdash e_2 : T_2 \\
+\opsenv e_1 \mapsto v_1, E_1 \\
+\opsenv e_2 \mapsto v_2, E_2 \\
+\Phi_1 \#_M \Phi_2
+\end{array}}
+{E_1 \# E_2}
+
+(TODO: Generalize this further and allow $M_1 \not= M_2$?) \\
+
+Note that this result holds even if $C \not= \emptyset$.  This allows 
+two ``locally independent'' subtasks to run in parallel even if one or both
+of them is interacting (in a programmer-permitted way) with a third subtask.
+
+Noninterference of effects is something that can only be determined after
+evaluation of an expression is completed, and only at great expense (see STM).
+In contrast, the noninterference of privileges can be efficiently determined at
+runtime.  Even though the privileges themselves are static, the
+mapping $M$ can be dynamically determined, allowing noninterference to be 
+judged in many more cases than a purely static analysis can achieve.
+
+(TODO: Talk about coherence modes and how the runtime orders subtasks.  Maybe
+describe it in terms of a ``canonical'' memory trace and then allow
+reorderings?)
+
+\subsection{Hierarchical Scheduling}
+
+Let $e_1$ and $e_2$ be expressions that require privileges $\Phi_1$ and
+$\Phi_2$.  Now let $e_3$ be some subexpression of $e_1$ and $e_4$ be some 
+subexpression of $e_2$, with corresponding privileges $\Phi_3$ and $\Phi_4$.
+
+Proper scheduling of the evaluation of $e_3$ and $e_4$ depends on whether
+they interfere.  If they do intefere (and must therefore be serialized), we
+will have $E_3 \cancel\# E_4$, and by the above result, $\Phi_3 \cancel\# \Phi_4$.
+
+We now show that this will also imply that $\Phi_1 \cancel\# \Phi_2$, and
+unless sequential behavior has been relaxed by the programmer, the
+evaluations of $e_1$ and $e_2$ will be serialized.  Since the evaluation of
+$e_3$ is a subsequence of the evaluation of $e_1$ and the evaluation of $e_4$
+is a subsequence of the evaluation of $e_2$, the evaluations of $e_3$ and $e_4$
+are properly serialized as well.
+
+Conversely, if $e_1$ and $e_2$ are judged to be noninterfering, the schedule
+does not need to check for interference between $e_3$ and $e_4$, as their
+noninterference is guaranteed.
+
+Proof is straightforward once armed with the following lemmas:
+
+\framebox{$\Phi_1 \#_M \Phi_2 \Leftrightarrow \Phi_1^* \#_M \Phi_2^*$}
+
+\framebox{$\Phi_1 \#_M \Phi_2 \wedge \Phi_3 \subseteq \Phi_2 \Rightarrow \Phi_1 \#_M \Phi_3$}

### 2012-07-01 runtime/lowlevel.cc
initial swag at valid-pointer-check for GPU

@@ -4620,7 +4620,7 @@ namespace RegionRuntime {
       instance.impl()->memory.impl()->destroy_instance(instance, true);
     }
 
-    const ElementMask &RegionMetaDataUntyped::get_valid_mask(void)
+    const ElementMask &RegionMetaDataUntyped::get_valid_mask(void) const
     {
       DetailedTimer::ScopedPush sp(TIME_LOW_LEVEL);
       RegionMetaDataUntyped::Impl *r_impl = impl();

### 2012-07-01 runtime/lowlevel.h

@@ -364,7 +364,7 @@ namespace RegionRuntime {
       void destroy_allocator_untyped(RegionAllocatorUntyped allocator) const;
       void destroy_instance_untyped(RegionInstanceUntyped instance) const;
 
-      const ElementMask &get_valid_mask(void);
+      const ElementMask &get_valid_mask(void) const;
     };
 
     class RegionAllocatorUntyped {
@@ -534,6 +534,7 @@ namespace RegionRuntime {
 #ifdef DEBUG_LOW_LEVEL
       size_t first_elmt;
       size_t last_elmt;
+      unsigned *valid_mask_base;
 #endif
 #ifdef __CUDACC__
       // Need copy constructors so we can move things around
@@ -602,6 +603,9 @@ namespace RegionRuntime {
       void bounds_check(ptr_t<T> ptr) const
       {
         assert((first_elmt <= ptr.value) && (ptr.value <= last_elmt));
+	off_t rel_ptr = ptr.value - first_elmt;
+	unsigned bits = valid_mask_base[rel_ptr >> 5];
+	assert(bits & (1U << (rel_ptr & 0x1f)));
       }
 #endif
     };

### 2012-07-01 runtime/lowlevel_gpu.cc

@@ -666,6 +666,18 @@ namespace RegionRuntime {
 #ifdef DEBUG_LOW_LEVEL
         ria.first_elmt = i_data->first_elmt;
         ria.last_elmt  = i_data->last_elmt;
+
+	// also need to grab a copy of the element mask for pointer checks
+	const ElementMask &mask = i_data->region.get_valid_mask();
+	unsigned *valid_mask_base;
+	CHECK_CUDART( cudaMalloc((void **)&valid_mask_base, mask.raw_size()) );
+	log_gpu.info("copy of valid mask (%zd bytes) created at %p",
+		     mask.raw_size(), valid_mask_base);
+	CHECK_CUDART( cudaMemcpy(valid_mask_base,
+				 mask.get_raw(),
+				 mask.raw_size(),
+				 cudaMemcpyHostToDevice) );
+	ria.valid_mask_base = valid_mask_base;
 #endif
 	return ria;
       }

### 2012-07-01 doc/popl2013/langdef.tex

@@
 \newcommand{\dplus}{\text{++}}
+
+% fun latex tricks to make typeenv and opsenv more friendly
+\makeatletter
+\define@key{typeenv}{G}{\def\typeenv@G{#1}}
+\define@key{typeenv}{P}{\def\typeenv@P{#1}}
+\define@key{typeenv}{O}{\def\typeenv@O{#1}}
+\newcommand{\typeenvx}[1][]{
+{
+% default values
+\def\typeenv@G{\Gamma}
+\def\typeenv@P{\Phi}
+\def\typeenv@O{\Omega}
+\setkeys{typeenv}{#1}
+\typeenv@G, \typeenv@P, \typeenv@O \vdash \,
+}}
+\newcommand{\typeenv}[3][]{\typeenvx[#1] {#2} : {#3}}
+\define@key{opsenv}{M}{\def\opsenv@M{#1}}
+\define@key{opsenv}{L}{\def\opsenv@L{#1}}
+\define@key{opsenv}{H}{\def\opsenv@H{#1}}
+\define@key{opsenv}{S}{\def\opsenv@S{#1}}
+\define@key{opsenv}{C}{\def\opsenv@C{#1}}
+\newcommand{\opsenvx}[1][]{
+{
+% default values
+\def\opsenv@M{M}
+\def\opsenv@L{L}
+\def\opsenv@H{H}
+\def\opsenv@S{S}
+\def\opsenv@C{C}
+\setkeys{opsenv}{#1}
+\opsenv@M, \opsenv@L, \opsenv@H, \opsenv@S, \opsenv@C \vdash \,
+}}
+\newcommand{\opsenv}[4][]{\opsenvx[#1] {#2} \mapsto {#3}, {#4}}
+\makeatother
 
@@
 
-\newcommand{\regionexpand}[0]{[r'_1/r_1, \ldots, r'_n/r_n]}
-\newcommand{\rtriple}[3]{\left({#1},{#2},{#3}\right)}
-\newcommand{\rsingle}[1]{\rtriple{#1}{\emptyset}{\top}}
-\newcommand{\rtripsub}[3]{\rtriple{#1}{\Phi_{#2}}{C_{#3}}}
-\newcommand{\envsub}[2]{\Gamma, \Phi_{#1}, C_{#2} \vdash}
-\newcommand{\typeenv}[0]{\Gamma, \Phi, \Omega \vdash}
-
 \pagebreak
@@
 
-\framebox{$\typeenv bv : bool$} \hspace{1cm}
-\framebox{$\typeenv iv : int$} \\
-
-\finfrule{\typeenv e : \langle T_1,T_2 \rangle}{\typeenv e\text{.1}\ : T_1} \hspace{1cm}
-\finfrule{\typeenv e : \langle T_1,T_2 \rangle}{\typeenv e\text{.2}\ : T_2} \\
-
-\finfrule{\Gamma(id) = T}{\typeenv id : T} \\
-
-\framebox{$\typeenv \text{null }T@r : T@r$} \hspace{1cm}
-\framebox{$\typeenv \text{new }T@r : T@r$} \\
-
-\finfrule{\typeenv e : T@(\oton{r})}{\typeenv \text{isnull}(e) : bool} \\
+\framebox{$\typeenv{bv}{bool}$} \hspace{1cm}
+\framebox{$\typeenv{iv}{int}$} \\
 
@@
 {\begin{array}{l}
-\typeenv e : T@(\otok{k}{r'}) \\
-\forall i. \exists j, r'_i \leq r_j \in \Omega^* \\
+\typeenvx e_1 : T_1 \\
+\typeenvx e_2 : T_2
 \end{array}}
-{\typeenv upregion(e,\oton{r}) : T@(\oton{r})} \\
+{\typeenvx \langle e_1, e_2 \rangle : \langle T_1, T_2 \rangle} \\
+
+\finfrule{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.1}\ : T_1} \hspace{1cm}
+\finfrule{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.2}\ : T_2} \\
+
+\finfrule{\Gamma(id) = T}{\typeenvx id : T} \\
+
+\framebox{$\typeenvx \text{null }T@r : T@r$} \hspace{1cm}
+\framebox{$\typeenvx \text{new }T@r : T@r$} \\
+
+\finfrule{\typeenvx e : T@(\oton{r})}{\typeenvx \text{isnull}(e) : bool} \\
 
@@
 {\begin{array}{l}
-\typeenv e : T@(\otok{k}{r'}) \\
-\forall j. \exists i, r_j \leq r'_i \in \Omega^* \\
+\typeenvx e : T@(\otok{k}{r'}) \\
+\forall i. \exists j, r'_i \leq r_j \in \Omega^* \\
 \end{array}}
-{\typeenv downregion(e,\oton{r}) : T@(\oton{r})} \\
+{\typeenvx upregion(e,\oton{r}) : T@(\oton{r})} \\
 
@@
 {\begin{array}{l}
-\typeenv e_1 : T@(\oton{r}) \\
-\forall i, reads(r_i) \in \Phi^*\end{array}}
-{\typeenv \text{read}(e_1) : T} \\
+\typeenvx e : T@(\otok{k}{r'}) \\
+\forall j. \exists i, r_j \leq r'_i \in \Omega^* \\
+\end{array}}
+{\typeenvx downregion(e,\oton{r}) : T@(\oton{r})} \\
 
@@
 {\begin{array}{l}
-\typeenv e_1 : T@(\oton{r}) \\
-\typeenv e_2 : T \\
+\typeenvx e_1 : T@(\oton{r}) \\
+\forall i, reads(r_i) \in \Phi^*\end{array}}
+{\typeenvx \text{read}(e_1) : T} \\
+
+\finfrule
+{\begin{array}{l}
+\typeenvx e_1 : T@(\oton{r}) \\
+\typeenvx e_2 : T \\
 \forall i, writes(r_i) \in \Phi^*
 \end{array}}
-{\typeenv \text{write}(e_1, e_2) : bool} \\
+{\typeenvx \text{write}(e_1, e_2) : bool} \\
 
@@
 \Gamma(id) = (T_1, T_2), \emptyset \rightarrow T_1 \\
-\typeenv e_1 : T_1@(\oton{r}) \\
-\typeenv e_2 : T_2 \\
+\typeenvx e_1 : T_1@(\oton{r}) \\
+\typeenvx e_2 : T_2 \\
 \forall i, reduces(id, r_i) \in \Phi^*
 \end{array}}
-{\typeenv \text{reduce}(id, e_1, e_2) : bool} \\
+{\typeenvx \text{reduce}(id, e_1, e_2) : bool} \\
 
-\framebox{$\typeenv \text{newcolor }r : \text{coloring}(r)$} \\
+\framebox{$\typeenvx \text{newcolor }r : \text{coloring}(r)$} \\
 
 \finfrule{\begin{array}{l}
-\typeenv e_1 : \text{coloring}(r) \\
-\typeenv e_2 : T@r \\
-\typeenv e_3 : int
+\typeenvx e_1 : \text{coloring}(r) \\
+\typeenvx e_2 : T@r \\
+\typeenvx e_3 : int
 \end{array}}
-{\typeenv \text{color}(e_1, e_2, e_3) : \text{coloring}(r)} \\
+{\typeenvx \text{color}(e_1, e_2, e_3) : \text{coloring}(r)} \\
 
-\finfrule{\begin{array}{l}\typeenv e_1 : int \\ \typeenv e_2 : int\end{array}}{\typeenv e_1 + e_2 : int} \hspace{1 cm}
-\finfrule{\begin{array}{l}\typeenv e_1 : int \\ \typeenv e_2 : int\end{array}}{\typeenv e_1 < e_2 : bool} \\
+\finfrule{\begin{array}{l}\typeenvx e_1 : int \\ \typeenvx e_2 : int\end{array}}{\typeenvx e_1 + e_2 : int} \hspace{1 cm}
+\finfrule{\begin{array}{l}\typeenvx e_1 : int \\ \typeenvx e_2 : int\end{array}}{\typeenvx e_1 < e_2 : bool} \\
 
-\finfrule{\begin{array}{l}\typeenv e_1 : T_1 \\ \Gamma' = \Gamma[id/T_1] \\ \Gamma', \Phi, \Omega \vdash e_2 : T_2 \end{array}}{\typeenv : \text{let}\ id : T_1 \text{in}\ e_2 : T_2} \\
+\finfrule{\begin{array}{l}
+\typeenvx e_1 : T_1 \\
+\Gamma' = \Gamma[id/T_1] \\
+\typeenvx[G=\Gamma'] e_2 : T_2
+\end{array}}
+{\typeenvx : \text{let}\ id : T_1 \text{in}\ e_2 : T_2} \\
 
-\finfrule{\begin{array}{l}\typeenv e_1 : bool \\ \typeenv e_2 : T \\ \typeenv e_3 : T\end{array}}{\typeenv \text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3 : T} \\
+\finfrule{\begin{array}{l}\typeenvx e_1 : bool \\ \typeenvx e_2 : T \\ \typeenvx e_3 : T\end{array}}{\typeenvx \text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3 : T} \\
 
@@
 \Gamma(id) = \forall \otok{k}{r'}.(\oton{T}),\Phi_i \rightarrow T_r \\
-\typeenv e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\
+\typeenvx e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\
 \Phi_i[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*
 \end{array}}
-{\typeenv id[\otok{k}{r}](\oton{e}) : T_r[r_1/r'_1,\ldots,r_k/r'_k]} \\
+{\typeenvx id[\otok{k}{r}](\oton{e}) : T_r[r_1/r'_1,\ldots,r_k/r'_k]} \\
 
@@
 \begin{array}{l}
-\typeenv e_1 : \text{coloring}(r_p) \\
+\typeenvx e_1 : \text{coloring}(r_p) \\
 \Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{i \not= j} r_i * r_j \\
-\Gamma, \Phi, \Omega' \vdash e_2 : T \\
+\typeenvx[O=\Omega'] e_2 : T \\
 r_i \not\in \textit{regions\_of}(\Gamma, T)
 \end{array}}
-{\typeenv \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 : T} \\
+{\typeenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 : T} \\
 
@@
 \Omega \models \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \\
-\typeenv e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]
+\typeenvx e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]
 \end{array}}
-{\typeenv \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] : T_1} \\
+{\typeenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] : T_1} \\
 
@@
 T_1 = \exists \otok{k}{r'}. T_2\text{ where }\Omega_1 \\
-\typeenv e_1 : T_1 \\
+\typeenvx e_1 : T_1 \\
 \Gamma' = \Gamma[T_2[r_1/r'_1,\ldots,r_k/r'_k] / id] \\
 \Omega' = \Omega \wedge \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \\
-\Gamma', \Phi, \Omega' \vdash e_2 : T_3 \\
+\typeenvx[G=\Gamma',O=\Omega'] e_2 : T_3 \\
 r_i \not\in \textit{regions\_of}(\Gamma, T_1, T_3)
 \end{array}}
-{\typeenv \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 : T_3} \\
+{\typeenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 : T_3} \\
 
@@
 
-\newcommand{\opsenv}{M, L, H, S, C \vdash}
-\newcommand{\opsenvm}[1]{{#1}, L, H, S, C \vdash}
-\newcommand{\opsenvs}[1]{M, L, H, {#1}, C \vdash}
-\newcommand{\opsenvls}[2]{M, {#1}, H, {#2}, C \vdash}
-\newcommand{\opsenvms}[2]{{#1}, L, H, {#2}, C \vdash}
-\newcommand{\opsenvmls}[3]{{#1}, {#2}, H, {#3}, C \vdash}
-
-\framebox{$\opsenv e \mapsto v, S', E$}
+\framebox{$\opsenv{e}{v}{E}$}
 
@@
 
-\framebox{$\opsenv bv \mapsto bv, []$} \hspace{1cm}
-\framebox{$\opsenv iv \mapsto iv, []$} \\
+\framebox{$\opsenvx bv \mapsto bv, []$} \hspace{1cm}
+\framebox{$\opsenvx iv \mapsto iv, []$} \\
 
 \finfrule
-{\opsenv e \mapsto \langle v_1, v_2 \rangle, E}
-{\opsenv e\text{.1} \mapsto v_1, E} \hspace{1cm}
+{\begin{array}{l}
+\opsenvx e_1 \mapsto v_1, E_1 \\
+S' = apply(S, E_1, C) \\
+\opsenvx[S=S'] e_2 \mapsto v_2, E_2
+\end{array}}
+{\opsenvx \langle e_1, e_2 \rangle \mapsto \langle v_1, v_2 \rangle, E_1 \dplus E_2} \\
+
 \finfrule
-{\opsenv e \mapsto \langle v_1, v_2 \rangle, E}
-{\opsenv e\text{.2} \mapsto v_2, E} \\
+{\opsenvx e \mapsto \langle v_1, v_2 \rangle, E}
+{\opsenvx e\text{.1} \mapsto v_1, E} \hspace{1cm}
+\finfrule
+{\opsenvx e \mapsto \langle v_1, v_2 \rangle, E}
+{\opsenvx e\text{.2} \mapsto v_2, E} \\
 
@@
 {L(id) = v}
-{\opsenv id \mapsto v, []} \\
+{\opsenvx id \mapsto v, []} \\
 
-\framebox{$\opsenv \text{null }T@r \mapsto null, []$} \\
+\framebox{$\opsenvx \text{null }T@r \mapsto null, []$} \\
 
@@
 \end{array}}
-{\opsenv \text{new }T@r \mapsto l, []} \\
+{\opsenvx \text{new }T@r \mapsto l, []} \\
 
 \finfrule
-{\opsenv e \mapsto v, E}
-{\opsenv \text{isnull}(e) \mapsto false, E} \\
+{\opsenvx e \mapsto v, E}
+{\opsenvx \text{isnull}(e) \mapsto false, E} \\
 
 \finfrule
-{\opsenv e \mapsto null, E}
-{\opsenv \text{isnull}(e) \mapsto true, E} \\
+{\opsenvx e \mapsto null, E}
+{\opsenvx \text{isnull}(e) \mapsto true, E} \\
 
 \finfrule
-{\opsenv e \mapsto v, E}
-{\opsenv \text{upregion}(e, \oton{r}) \mapsto v, E}
+{\opsenvx e \mapsto v, E}
+{\opsenvx \text{upregion}(e, \oton{r}) \mapsto v, E}
 
@@
 {\begin{array}{l}
-\opsenv e \mapsto v, E \\
+\opsenvx e \mapsto v, E \\
 v' = \begin{cases}
@@
 \end{array}}
-{\opsenv \text{downregion}(e, \oton{r}) \mapsto v', E} \\
+{\opsenvx \text{downregion}(e, \oton{r}) \mapsto v', E} \\
 
@@
 {\begin{array}{l}
-\opsenv e_1 \mapsto l, E \\
+\opsenvx e_1 \mapsto l, E \\
 S' = \text{apply}(S, E, C) \\
@@
 \end{array}}
-{\opsenv \text{read}(e_1) \mapsto v, E \dplus [ read(l, v) ]} \\
+{\opsenvx \text{read}(e_1) \mapsto v, E \dplus [ read(l, v) ]} \\
 
@@
 {\begin{array}{l}
-\opsenv e_1 \mapsto l, E_1 \\
+\opsenvx e_1 \mapsto l, E_1 \\
 S' = \text{apply}(S, E_1, C) \\
-\opsenvs{S'} e_2 \mapsto v, E_2 \\
+\opsenvx[S=S'] e_2 \mapsto v, E_2 \\
 \end{array}}
-{\opsenv \text{write}(e_1, e_2) \mapsto true, E_1 \dplus E_2 \dplus [ write(l, v) ]} \\
+{\opsenvx \text{write}(e_1, e_2) \mapsto true, E_1 \dplus E_2 \dplus [ write(l, v) ]} \\
 
@@
 {\begin{array}{l}
-\opsenv e_1 \mapsto l, E_1 \\
+\opsenvx e_1 \mapsto l, E_1 \\
 S' = \text{apply}(S, E_1, C) \\
-\opsenvs{S'} e_2 \mapsto v, E_2 \\
+\opsenvx[S=S'] e_2 \mapsto v, E_2 \\
 \end{array}}
-{\opsenv \text{reduce}(id, e_1, e_2) \mapsto true, E_1 \dplus E_2 \dplus [ reduce_{id}(l, v) ]} \\
+{\opsenvx \text{reduce}(id, e_1, e_2) \mapsto true, E_1 \dplus E_2 \dplus [ reduce_{id}(l, v) ]} \\
 
-\framebox{$\opsenv \text{newcolor }r \mapsto \emptyset, []$} \\
+\framebox{$\opsenvx \text{newcolor }r \mapsto \emptyset, []$} \\
 
@@
 {\begin{array}{l}
-\opsenv e_1 \mapsto K, E_1 \\
+\opsenvx e_1 \mapsto K, E_1 \\
 S' = \text{apply}(S, E_1, C) \\
-\opsenvs{S'} e_2 \mapsto l, E_2 \\
+\opsenvx[S=S'] e_2 \mapsto l, E_2 \\
 S'' = \text{apply}(S', E_2, C) \\
-\opsenvs{S''} e_3 \mapsto v, E_3 \\
+\opsenvx[S=S''] e_3 \mapsto v, E_3 \\
 K' = \{ (l,v) \} \cup \{ (l_i,v_i) : (l_i,v_i) \in K \wedge l \not= l_i \}
 \end{array}}
-{\opsenv \text{color}(e_1, e_2, e_3) \mapsto K', E_1 \dplus E_2 \dplus E_3} \\
+{\opsenvx \text{color}(e_1, e_2, e_3) \mapsto K', E_1 \dplus E_2 \dplus E_3} \\
 
@@
 {\begin{array}{l}
-\opsenv e_1 \mapsto v_1, E_1 \\
+\opsenvx e_1 \mapsto v_1, E_1 \\
 S' = \text{apply}(S, E_1, C) \\
-\opsenvs{S'} e_2 \mapsto v_2, E_2 \\
+\opsenvx[S=S'] e_2 \mapsto v_2, E_2 \\
 v' = v_1 + v_2
 \end{array}}
-{\opsenv e_1 + e_2 \mapsto v', E_1 \dplus E_2} \hspace{1cm}
+{\opsenvx e_1 + e_2 \mapsto v', E_1 \dplus E_2} \hspace{1cm}
 
@@
 {\begin{array}{l}
-\opsenv e_1 \mapsto v_1, E_1 \\
+\opsenvx e_1 \mapsto v_1, E_1 \\
 S' = \text{apply}(S, E_1, C) \\
-\opsenvs{S'} e_2 \mapsto v_2, E_2 \\
+\opsenvx[S=S'] e_2 \mapsto v_2, E_2 \\
 v' = \begin{cases}
@@
 \end{array}}
-{\opsenv e_1 < e_2 \mapsto v', E_1 \dplus E_2} \\
+{\opsenvx e_1 < e_2 \mapsto v', E_1 \dplus E_2} \\
 
@@
 {\begin{array}{l}
-\opsenv e_1 \mapsto v_1, E_1 \\
+\opsenvx e_1 \mapsto v_1, E_1 \\
 L' = L[v_1/id] \\
 S' = \text{apply}(S, E_1, C) \\
-\opsenvls{L'}{S'} e_2 \mapsto v_2, E_2
+\opsenvx[L=L',S=S'] e_2 \mapsto v_2, E_2
 \end{array}}
-{\opsenv \text{let }id : T = e_1\text{ in }e_2 \mapsto v_2, E_1 \dplus E_2} \\
+{\opsenvx \text{let }id : T = e_1\text{ in }e_2 \mapsto v_2, E_1 \dplus E_2} \\
 
@@
 {\begin{array}{l}
-\opsenv e_1 \mapsto true, E_1 \\
+\opsenvx e_1 \mapsto true, E_1 \\
 S' = \text{apply}(S, E_1, C) \\
-\opsenvls{L'}{S'} e_2 \mapsto v_2, E_2
+\opsenvx[L=L',S=S'] e_2 \mapsto v_2, E_2
 \end{array}}
-{\opsenv \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_2, E_1 \dplus E_2} \\
+{\opsenvx \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_2, E_1 \dplus E_2} \\
 
@@
 {\begin{array}{l}
-\opsenv e_1 \mapsto false, E_1 \\
+\opsenvx e_1 \mapsto false, E_1 \\
 S' = \text{apply}(S, E_1, C) \\
-\opsenvls{L'}{S'} e_3 \mapsto v_3, E_3
+\opsenvx[L=L',S=S'] e_3 \mapsto v_3, E_3
 \end{array}}
-{\opsenv \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_3, E_1 \dplus E_3} \\
+{\opsenvx \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_3, E_1 \dplus E_3} \\
 
@@
 {\begin{array}{l}
-\opsenv e_1 \mapsto v_1, E_1 \\
+\opsenvx e_1 \mapsto v_1, E_1 \\
 S_1 = \text{apply}(S, E_1, C) \\
-\opsenv e_2 \mapsto v_2, E_2 \\
+\opsenvx e_2 \mapsto v_2, E_2 \\
 \ldots \\
@@
 L' = \{ (f_1, v_1), \ldots, (f_n, v_n) \} \\
-\opsenvmls{M'}{L'}{S_n} e_{n+1} \mapsto v_{n+1}, E_{n+1}
+\opsenvx[M=M',L=L',S=S_n] e_{n+1} \mapsto v_{n+1}, E_{n+1}
 \end{array}}
-{\opsenv id[\otok{k}{r}](\oton{e}) \mapsto v_{n+1}, E_1 \dplus \ldots \dplus E_{n+1}} \\
+{\opsenvx id[\otok{k}{r}](\oton{e}) \mapsto v_{n+1}, E_1 \dplus \ldots \dplus E_{n+1}} \\
 
@@
 \begin{array}{l}
-\opsenv e_1 \mapsto K, E_1 \\
+\opsenvx e_1 \mapsto K, E_1 \\
 \rho_i = \{ l : (l, i) \in K \} \\
@@
 S' = \text{apply}(S, E_1, C) \\
-\opsenvms{M'}{S'} e_2 \mapsto v, E_2
+\opsenvx[M=M',S=S'] e_2 \mapsto v, E_2
 \end{array}}
-{\opsenv \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 \mapsto v, E_1 \dplus E_2} \\
+{\opsenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 \mapsto v, E_1 \dplus E_2} \\
 
@@
 \begin{array}{l}
-\opsenv e_1 \mapsto v, E \\
+\opsenvx e_1 \mapsto v, E \\
 \rho_i = M[r_i] \\
@@
 \end{array}}
-{\opsenv \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] \mapsto v', E} \\
+{\opsenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] \mapsto v', E} \\
 
@@
 \begin{array}{l}
-\opsenv e_1 \mapsto \langle \langle \otok{k}{\rho} , v_1 \rangle \rangle, E_1 \\
+\opsenvx e_1 \mapsto \langle \langle \otok{k}{\rho} , v_1 \rangle \rangle, E_1 \\
 M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \\
@@
 S' = \text{apply}(S, E_1, C) \\
-\opsenvmls{M'}{L'}{S'} e_2 \mapsto v_2, E_2
+\opsenvx[M=M',L=L',S=S'] e_2 \mapsto v_2, E_2
 \end{array}}
-{\opsenv \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 \mapsto v_2, E_1 \dplus E_2 } \\
+{\opsenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 \mapsto v_2, E_1 \dplus E_2 } \\
 
@@
 \begin{enumerate}
-\item $\typeenv e : T$
-\item $\opsenv e \mapsto v, E$
+\item $\typeenvx e : T$
+\item $\opsenvx e \mapsto v, E$
 \end{enumerate}
@@
 \item $iv \sim_H int$
-\item $null \ sim_H T@\rho$
+\item $null \sim_H T@\rho$
 \item $l \sim_H T@\rho \Leftrightarrow (l \in \rho) \wedge (H(l) \sim_H T)$
@@
 \item $\langle \langle \oton{\rho}, v \rangle \rangle \sim_H \text{rr}[\oton{r}] \hat T\text{ where }\Omega \Leftrightarrow (v \sim_H \hat T[\rho_1/r_1, \ldots \rho_n/r_n]) \wedge (\{ (r_i, \rho_i) \} \sim \Omega)$
+\item $K \sim_H coloring(\rho) \Leftrightarrow \forall l_1, v_1. (l_1, v_1) \in K \rightarrow (l_1 \in \rho \wedge \forall l_2, v_2. (l_2, v_2) \in K \rightarrow (l_1 \not= l_2) \vee (v_1 = v_2))$
 \end{enumerate}
@@
 {\begin{array}{l}
-\typeenv e : T \\
-\opsenv e \mapsto v, E \\
+\typeenvx e : T \\
+\opsenvx e \mapsto v, E \\
 M \sim \Omega \hspace{1cm}
@@
 {\begin{array}{l}
-\Gamma, \Phi_1, \Omega \vdash e_1 : T_1 \\
-\Gamma, \Phi_2, \Omega \vdash e_2 : T_2 \\
-\opsenv e_1 \mapsto v_1, E_1 \\
-\opsenv e_2 \mapsto v_2, E_2 \\
+\typeenvx[P=\Phi_1] e_1 : T_1 \\
+\typeenvx[P=\Phi_2] e_2 : T_2 \\
+\opsenvx e_1 \mapsto v_1, E_1 \\
+\opsenvx e_2 \mapsto v_2, E_2 \\
 \Phi_1 \#_M \Phi_2

### 2012-07-02 doc/popl2013/langdef.tex
checkpoint

@@
 \newcommand{\dplus}{\text{++}}
+\newcommand{\llbracket}{[\![}
+\newcommand{\rrbracket}{]\!]}
 
@@
   &$\mid$& $\exists \oton{r}. T\text{ where }\Omega$ & region relationship \\
-  &$\mid$& $\forall \oton{r}. (\oton{T}), \Phi \rightarrow T_r$ & functions \\
+  &$\mid$& $\forall \oton{r}. (\oton{T}), \Phi, Q \rightarrow T_r$ & functions \\
 \\
@@
 $\phi$ & ::= & reads$(r)$ $\;\;\;\mid\;\;\;$ writes$(r)$ $\;\;\;\mid\;\;\;$ reduces$(id,r)$ & \\
+\\
+$Q$ & ::= & $\{ \oton{q} \}$ & coherence modes \\
+$q$ & ::= & atomic$(r)$ $\;\;\;\mid\;\;\;$ simult$(r)$ & \\
 \\
@@
 \begin{array}{l}
-\Gamma(id) = \forall \otok{k}{r'}.(\oton{T}),\Phi_i \rightarrow T_r \\
+\Gamma(id) = \forall \otok{k}{r'}.(\oton{T}),\Phi_i, Q \rightarrow T_r \\
 \typeenvx e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\
@@
 \begin{itemize}
-\item $M$ is the mapping from logical regions $r_i$ in types to physical regions $\rho_i$, which are sets of memory locations
+\item $M$ is the mapping from logical regions $r_i$ in types to physical regions $\rho_i$, which are sets of memory locations.  It will be useful to apply this logical-to-physical mapping to other
+things that defined in terms of logical regions to yield a corresponding definition in terms of
+physical regions.  The notation $\hat x = M \llbracket x \rrbracket$ is used to describe the
+result of replacing all free logical region names $r_i$ in $x$ with the corresponding physical
+region $\rho_i$.  (Note that $M \llbracket x \rrbracket$ is not
+well-formed if it contains a free logical region name $r_j \not\in dom(M)$.)
 \item $L$ is the mapping of local variable names to their values
@@
 \item $v$ is the result of the evaluation
-\item $E$ is a memory trace (i.e. the effects) of the execution, a list of read/write/reduce operations in the order they were performed
+\item $E$ is a memory trace (i.e. the effects) of the execution, a list of read/write/reduce operations in the order they were performed.  An operation $\epsilon$ is one of the following:
+\begin{itemize}
+\item $read(l, c, v)$, a read of location $l$ with coherence mode $c$, returning value $v$
+\item $write(l, c, v)$, a write to location $l$ with coherence mode $c$, storing value $v$
+\item $reduce_{id}(l, c, v)$, a reduction into location $l$ with coherence mode $c$ using a
+commutative function $id$ with value $v$
+\end{itemize}
+where coherence modes are one of ({\em excl}, {\em atomic}, {\em simult}).
+
+The following helper functions are used in some of the rules that deal with memory effects:
+\begin{itemize}
+\item $apply(S, E, C)$ is used to determine the effect on the store of a sequence of memory operations
+\item $mark\_coherence(E, \hat Q)$ marks the memory operations in a subtask's trace to
+constrain how the operations in that trace can be interleaved with other sibling tasks' operations
+\item $valid\_interleave(S, C, E', \oton{E})$ is used to determine if $E'$ is a valid way to
+interleave the memory traces of two or more subexpressions
+\end{itemize}
+Figure~\ref{fig:opsemfns} gives the definitions of {\em apply} and {\em mark\_coherence}.  For
+the {\em valid\_interleave} predicate, we state some of the necessary conditions here, but leave 
+the construction of actual valid interleavings until a later section.  Informally, a interleaving
+of two or more memory traces must include all the operations from each trace (and no additional
+operations), with each trace's operations remaining in order.  A valid interleaving must further
+satisfy the requirement that the value read from any location that's not in the clobber set $C$
+yield the most recently stored/reduced value.
+
+\begin{figure*}
+\centering{
+$
+\begin{array}{lll}
+apply(S, [], C) & = & S \\
+apply(S, E \dplus [ read(l, c, v) ]) & = & apply(S, E, C) \\
+apply(S, E \dplus [ write(l, c, v) ]) & = & apply(S, E, C)[v/l] \\
+apply(S, E \dplus [ reduce_{id}(l, c, v) ]) & = & apply(S, E, C)[id(S[l], v)/l] \\
+mark\_coherence([], \hat Q) & = & [] \\
+mark\_coherence(E \dplus [ op(l, c, v) ], \hat Q) & = & mark\_coherence(E, \hat Q) \dplus \\
+& & op(l, c', v), \text{ where } \\
+& & c' =
+\begin{cases}
+simult, & \text{if }\exists \rho. l \in \rho \wedge simult(\rho) \in \hat Q \\
+atomic, & \text{if }\exists \rho. l \in \rho \wedge atomic(\rho) \in \hat Q \\
+excl, & \text{otherwise}
+\end{cases} \\
+any\_interleave(E', [], \ldots, []) & \rightarrow & true \\
+any\_interleave([ op(l, c, v) ] \dplus E', E_1, \ldots, [ op(l, c, v) ] \dplus E_i, \ldots, E_n) & \rightarrow & any\_interleave(E', E_1, \ldots, E_i, \ldots, E_n) \\
+valid\_interleave(S, C, E', \oton{E}) & \rightarrow & any\_interleave(E', \oton{E}) \\
+valid\_interleave(S, C, E' \dplus [ read(l, c, v) ] \dplus E'', \oton{E}) & \rightarrow & l \in C \vee S'(l) = v, \text{ where } S' = apply(S, E', C)
+\end{array}
+$}
+\caption{Operational Semantics Helper Functions}
+\label{fig:opsemfns}
+\end{figure*}
+
+A few helper functions are used in the operational semantics rules, and defined in 
+A number of rules will also use the property $valid\_interleave(S, C, E', E_1, E_2)$ to 
+constrain how the memory operations of two subexpressions can be interleaved.  We will 
+consider actual constructions of valid interleavings later.  For now, we simply rely on
+the following necessary conditions for any valid interleaving:
+\begin{enumerate}
+\item A valid interleaving $E'$ of $E_1$ and $E_2$ contains exactly all operations from both
+$E_1$ and $E_2$, in some order.
+\item The operations in each of $E_1$ and $E_2$ remain in order in $E'$.
+\item For every $read(l, c, v)$ in $E'$ for which $l$ is not in the clobber set $C$, the value 
+returned by the read is consistent with the original store $S$ and any preceding writes or
+reductions.
+\end{enumerate}
 \end{itemize}
@@
 S' = apply(S, E_1, C) \\
-\opsenvx[S=S'] e_2 \mapsto v_2, E_2
+\opsenvx[S=S'] e_2 \mapsto v_2, E_2 \\
+valid\_interleave(S, C, E', E_1, E_2)
 \end{array}}
-{\opsenvx \langle e_1, e_2 \rangle \mapsto \langle v_1, v_2 \rangle, E_1 \dplus E_2} \\
+{\opsenvx \langle e_1, e_2 \rangle \mapsto \langle v_1, v_2 \rangle, E'} \\
 
@@
 \end{array}}
-{\opsenvx \text{read}(e_1) \mapsto v, E \dplus [ read(l, v) ]} \\
+{\opsenvx \text{read}(e_1) \mapsto v, E \dplus [ read(l, excl, v) ]} \\
 
@@
 \end{array}}
-{\opsenvx \text{write}(e_1, e_2) \mapsto true, E_1 \dplus E_2 \dplus [ write(l, v) ]} \\
+{\opsenvx \text{write}(e_1, e_2) \mapsto true, E_1 \dplus E_2 \dplus [ write(l, excl, v) ]} \\
 
@@
 \end{array}}
-{\opsenvx \text{reduce}(id, e_1, e_2) \mapsto true, E_1 \dplus E_2 \dplus [ reduce_{id}(l, v) ]} \\
+{\opsenvx \text{reduce}(id, e_1, e_2) \mapsto true, E_1 \dplus E_2 \dplus [ reduce_{id}(l, excl, v) ]} \\
 
@@
 \opsenvx[S=S''] e_3 \mapsto v, E_3 \\
-K' = \{ (l,v) \} \cup \{ (l_i,v_i) : (l_i,v_i) \in K \wedge l \not= l_i \}
+K' = \{ (l,v) \} \cup \{ (l_i,v_i) : (l_i,v_i) \in K \wedge l \not= l_i \} \\
+valid\_interleave(S, C, E', E_1, E_2, E_3)
 \end{array}}
-{\opsenvx \text{color}(e_1, e_2, e_3) \mapsto K', E_1 \dplus E_2 \dplus E_3} \\
+{\opsenvx \text{color}(e_1, e_2, e_3) \mapsto K', E'} \\
 
@@
 \opsenvx[S=S'] e_2 \mapsto v_2, E_2 \\
-v' = v_1 + v_2
+v' = v_1 + v_2 \\
+valid\_interleave(S, C, E', E_1, E_2)
 \end{array}}
-{\opsenvx e_1 + e_2 \mapsto v', E_1 \dplus E_2} \hspace{1cm}
+{\opsenvx e_1 + e_2 \mapsto v', E'} \hspace{1cm}
 
@@
 false, & \text{otherwise}.
-\end{cases}
+\end{cases} \\
+valid\_interleave(S, C, E', E_1, E_2)
 \end{array}}
-{\opsenvx e_1 < e_2 \mapsto v', E_1 \dplus E_2} \\
+{\opsenvx e_1 < e_2 \mapsto v', E'} \\
 
@@
 S' = \text{apply}(S, E_1, C) \\
-\opsenvx[L=L',S=S'] e_2 \mapsto v_2, E_2
+\opsenvx[L=L',S=S'] e_2 \mapsto v_2, E_2 \\
+valid\_interleave(S, C, E', E_1, E_2)
 \end{array}}
-{\opsenvx \text{let }id : T = e_1\text{ in }e_2 \mapsto v_2, E_1 \dplus E_2} \\
+{\opsenvx \text{let }id : T = e_1\text{ in }e_2 \mapsto v_2, E'} \\
 
@@
 S' = \text{apply}(S, E_1, C) \\
-\opsenvx[L=L',S=S'] e_2 \mapsto v_2, E_2
+\opsenvx[L=L',S=S'] e_2 \mapsto v_2, E_2 \\
+valid\_interleave(S, C, E', E_1, E_2)
 \end{array}}
-{\opsenvx \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_2, E_1 \dplus E_2} \\
+{\opsenvx \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_2, E'} \\
 
@@
 S' = \text{apply}(S, E_1, C) \\
-\opsenvx[L=L',S=S'] e_3 \mapsto v_3, E_3
+\opsenvx[L=L',S=S'] e_3 \mapsto v_3, E_3 \\
+valid\_interleave(S, C, E', E_1, E_3)
 \end{array}}
-{\opsenvx \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_3, E_1 \dplus E_3} \\
+{\opsenvx \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_3, E'} \\
 
@@
 S_n = \text{apply}(S_{n-1}, E_n, C) \\
-task id[\otok{k}{r'}](f_1 : T_1, \ldots, f_n : T_n) : T_r, E = e_{n+1} \\
+valid\_interleave(S, C, E', \oton{E}) \\
+S' = \text{apply}(S, E', C) \\
+task id[\otok{k}{r'}](f_1 : T_1, \ldots, f_n : T_n), \Phi_i, Q : T_r = e_{n+1} \\
 M' = \{ (r'_1, M(r_1)), \ldots (r'_k, M(r_k)) \} \\
 L' = \{ (f_1, v_1), \ldots, (f_n, v_n) \} \\
-\opsenvx[M=M',L=L',S=S_n] e_{n+1} \mapsto v_{n+1}, E_{n+1}
+\opsenvx[M=M',L=L',S=S'] e_{n+1} \mapsto v_{n+1}, E_{n+1} \\
+E'_{n+1} = mark\_coherence(E_{n+1}, M \llbracket Q \rrbracket) \\
+valid\_interleave(S, C, E'', E', E_{n+1})
 \end{array}}
-{\opsenvx id[\otok{k}{r}](\oton{e}) \mapsto v_{n+1}, E_1 \dplus \ldots \dplus E_{n+1}} \\
+{\opsenvx id[\otok{k}{r}](\oton{e}) \mapsto v_{n+1}, E''} \\
 
@@
 S' = \text{apply}(S, E_1, C) \\
-\opsenvx[M=M',S=S'] e_2 \mapsto v, E_2
+\opsenvx[M=M',S=S'] e_2 \mapsto v, E_2 \\
+valid\_interleave(S, C, E', E_1, E_2)
 \end{array}}
-{\opsenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 \mapsto v, E_1 \dplus E_2} \\
+{\opsenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 \mapsto v, E'} \\
 
@@
 S' = \text{apply}(S, E_1, C) \\
-\opsenvx[M=M',L=L',S=S'] e_2 \mapsto v_2, E_2
+\opsenvx[M=M',L=L',S=S'] e_2 \mapsto v_2, E_2 \\
+valid\_interleave(S, C, E', E_1, E_2)
 \end{array}}
-{\opsenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 \mapsto v_2, E_1 \dplus E_2 } \\
+{\opsenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 \mapsto v_2, E' } \\
 
@@
 
+An important property of the legion type system is that any expression that is well-typed is
+guaranteed to access the heap only in ways permitted by the privileges under which it was typed.
+A judgement $M \vdash E : \Phi$
 If
@@
 
-\newcommand{\llbracket}{[\![}
-\newcommand{\rrbracket}{]\!]}
-
 The mapping $M$ is a function from logical region names $r$ to physical regions $\rho$.  It will
@@
 $\hat x = M \llbracket x \rrbracket$ to be $x$ with all logical region names $r_i$ replaced by their
-corresponding physical region $\rho_i = M(r_i)$.  Note that $M \llbracket x \rrbracket$ is not
-well-formed if it contains a logical region name $r_j \not\in dom(M)$. \\
+corresponding physical region $\rho_i = M(r_i)$.   \\
 
@@
 {\begin{array}{l}
+E_1 \sim H \hspace{1cm}
+E_2 \sim H \\
+valid\_interleave(S, C, E', E_1, E_2)
+\end{array}}
+{ E' \sim H } \\
+
+\finfrule
+{\begin{array}{l}
 E_1 : \Phi \hspace{1cm}
@@
 \subsection{Noninterference}
+
+The soundness proof above depended, in part, on the properties of a valid interleaving of two
+memory traces.  We now consider algorithms that create interleavings that satisfy those
+conditions.
+
+The simplest way to achieve a valid interleaving is to concatenate the traces in order.  This
+will always result in a valid interleaving, but implies a serialization of the evaluation of
+subexpressions.
+
+\framebox{$\forall S, C, E_1, E_2. valid\_interleave(S, C, E_1 \dplus E_2, E_1, E_2)$}
+
+Next step: For noninterfering traces $E_1 \# E_2$, show any interleaving is valid.
+
+Third step: For interactions only on atomic-coherence locations, serializability is sufficient.
 

### 2012-07-03 doc/popl2013/langdef.tex
checkpoint

@@
 
-\begin{table*}
+\begin{figure*}
 \centering
@@
 }
-\caption{Types}
-\end{table*}
+\caption{Legion Core Language}
+\label{fig:langdef}
+\end{figure*}
 
@@
 \newcommand{\cinfrule}[3]{\parbox{14cm}{\hfil$\infrule{#1}{#2}$\hfil}\parbox{4cm}{$\,#3$\hfil}}
-\newcommand{\finfrule}[2]{\vspace{10pt}\framebox{$\infrule{#1}{#2}$}\vspace{10pt}}
+\newcommand{\finfrule}[2]{\framebox{$\infrule{#1}{#2}$}}
+\newcommand{\oldfinfrule}[2]{\vspace{10pt}\framebox{$\infrule{#1}{#2}$}\vspace{10pt}}
 
@@
 
-Type rules:
+\section{Legion Core Language}
 
-\framebox{$\typeenv{bv}{bool}$} \hspace{1cm}
-\framebox{$\typeenv{iv}{int}$} \\
+\subsection{Language Definition}
 
+Figure~\ref{fig:langdef} defines the core subset of Legion application language.
+Legion includes basic boolean, integer and tuple types.  Pointers are included as well, but in
+addition to specifying the type of value it points to, a pointer's type also constrains where in
+memory it can point (i.e. only into regions listed in the type).  A special type exists for
+{\em colorings}, which are used to specify how a region should be partitioned into subregions.
+Functions in Legion are named, accept one or more arguments and return a value of some type.
+A function also declares what region access privileges and coherence it requires.  Function types
+are universally quantified over all logical region bindings, with all argument and result types,
+as well as the privileges and coherence requirements being expressed relative to the quantified
+region names.
+The final additional value type that Legion uses is a region relation.  Although logical regions
+cannot be stored directly in variables, they can be captured as part of a region relation 
+instance, which stores a value and one or more logical regions that satisfy a particular set of
+constraints.  An instance of a region relation can be read out of the heap and unpacked, giving
+new local names to the logical regions contained in the instance and tracking the constraints
+between those new logical regions.
+
+\subsection{Legion Core Type System}
+
+The Legion Core language is explicitly typed, using judgments of the following form:
+
+\begin{center}
+$\typeenv{e}{T}$
+\end{center}
+
+where:
+\begin{itemize}
+\item $\Gamma$ is the type environment, mapping the names of local variables and global functions
+to their types,
+\item $\Phi$ is the region access privileges under which the expression will be evaluated, and
+\item $\Omega$ is the set of constraints that must hold between regions
+\end{itemize}
+
+As regions are hierarchical, privileges and/or constraints on a region will often imply similar
+privileges or constraints on subregions.  Figure~\ref{fig:closure} defines transitive closure
+operations for $\Phi^*$ and $\Omega^*$ that will be used in some of the type checking rules.
+
+The full set of type checking rules are in Figure~\ref{fig:types}.  Many of the rules are 
+straightforward, but several deserve more explanation.
+
+The Legion core language has no subtyping, even for pointers into regions that are part of a 
+known region hierarchy.  The region ``scope'' of a pointer can only be changed via explicit
+{\em upregion} and {\em downregion} operations, which check that the ``from'' and ``to'' regions
+are in the right relationship.  It would have been possible to safely infer {\em upregion} calls,
+but {\em downregion} calls must be accompanied by a run-time check that the pointer really does
+point into the subregion(s), so {\em upregion} was made explicit as well for simplicity.
+
+Accesses to heap memory are performed via the {\em read}, {\em write}, and {\em reduce}
+operations, which are only legal if the region(s) in a pointers type are ones for which the
+enclosing function has privileges to access.
+
+Similarly, a call to another function is only permitted if the privileges required by the 
+callee are a subset of those held by the caller.  There is no way to create privileges within
+a Legion function - it is limited to those given to it by its caller.
+
+The {\em partition} operation uses a {\em coloring} to split a region into one or more
+disjoint subregions.  The subregions are given names within the body of the {\em partition},
+as well as constraints that assert that each subregion is contained by the parent region and
+the pairwise disjointness of the subregions.  These constraints can be used locally, or packed
+into region relation instances that can be unpacked and used later.
+
+The {\em pack} expression packs together a value and the corresponding set of logical regions,
+as long as the constraints in the region relationship type can be shown to hold.  The
+corresponding {\em unpack} expression gives (new) names to those regions.  Neither the 
+{\em partition} nor the {\em unpack} expression allows aliasing of region names, simplifying the
+handling of existing variables and region names.
+
+\begin{figure*}
+\centering{
+\framebox{$\typeenv{bv}{bool}$}
+\framebox{$\typeenv{iv}{int}$}
 \finfrule
@@
 \end{array}}
-{\typeenvx \langle e_1, e_2 \rangle : \langle T_1, T_2 \rangle} \\
-
-\finfrule{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.1}\ : T_1} \hspace{1cm}
-\finfrule{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.2}\ : T_2} \\
-
-\finfrule{\Gamma(id) = T}{\typeenvx id : T} \\
-
-\framebox{$\typeenvx \text{null }T@r : T@r$} \hspace{1cm}
-\framebox{$\typeenvx \text{new }T@r : T@r$} \\
-
-\finfrule{\typeenvx e : T@(\oton{r})}{\typeenvx \text{isnull}(e) : bool} \\
-
+{\typeenvx \langle e_1, e_2 \rangle : \langle T_1, T_2 \rangle}
+\finfrule{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.1}\ : T_1}
+\finfrule{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.2}\ : T_2}
+\finfrule{\Gamma(id) = T}{\typeenvx id : T}
+\framebox{$\typeenvx \text{null }T@r : T@r$}
+\framebox{$\typeenvx \text{new }T@r : T@r$}
+\finfrule{\typeenvx e : T@(\oton{r})}{\typeenvx \text{isnull}(e) : bool}
 \finfrule
@@
 \end{array}}
-{\typeenvx upregion(e,\oton{r}) : T@(\oton{r})} \\
-
+{\typeenvx upregion(e,\oton{r}) : T@(\oton{r})}
 \finfrule
@@
 \end{array}}
-{\typeenvx downregion(e,\oton{r}) : T@(\oton{r})} \\
-
+{\typeenvx downregion(e,\oton{r}) : T@(\oton{r})}
 \finfrule
@@
 \forall i, reads(r_i) \in \Phi^*\end{array}}
-{\typeenvx \text{read}(e_1) : T} \\
-
+{\typeenvx \text{read}(e_1) : T}
 \finfrule
@@
 \end{array}}
-{\typeenvx \text{write}(e_1, e_2) : bool} \\
-
+{\typeenvx \text{write}(e_1, e_2) : bool}
 \finfrule
 {\begin{array}{l}
-\Gamma(id) = (T_1, T_2), \emptyset \rightarrow T_1 \\
+\Gamma(id) = (T_1, T_2), \emptyset, \emptyset \rightarrow T_1 \\
 \typeenvx e_1 : T_1@(\oton{r}) \\
@@
 \end{array}}
-{\typeenvx \text{reduce}(id, e_1, e_2) : bool} \\
-
-\framebox{$\typeenvx \text{newcolor }r : \text{coloring}(r)$} \\
-
+{\typeenvx \text{reduce}(id, e_1, e_2) : bool}
+\framebox{$\typeenvx \text{newcolor }r : \text{coloring}(r)$}
 \finfrule{\begin{array}{l}
@@
 \end{array}}
-{\typeenvx \text{color}(e_1, e_2, e_3) : \text{coloring}(r)} \\
-
-\finfrule{\begin{array}{l}\typeenvx e_1 : int \\ \typeenvx e_2 : int\end{array}}{\typeenvx e_1 + e_2 : int} \hspace{1 cm}
-\finfrule{\begin{array}{l}\typeenvx e_1 : int \\ \typeenvx e_2 : int\end{array}}{\typeenvx e_1 < e_2 : bool} \\
-
+{\typeenvx \text{color}(e_1, e_2, e_3) : \text{coloring}(r)}
+\finfrule{\begin{array}{l}\typeenvx e_1 : int \\ \typeenvx e_2 : int\end{array}}{\typeenvx e_1 + e_2 : int}
+\finfrule{\begin{array}{l}\typeenvx e_1 : int \\ \typeenvx e_2 : int\end{array}}{\typeenvx e_1 < e_2 : bool}
 \finfrule{\begin{array}{l}
@@
 \end{array}}
-{\typeenvx : \text{let}\ id : T_1 \text{in}\ e_2 : T_2} \\
-
-\finfrule{\begin{array}{l}\typeenvx e_1 : bool \\ \typeenvx e_2 : T \\ \typeenvx e_3 : T\end{array}}{\typeenvx \text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3 : T} \\
-
+{\typeenvx : \text{let}\ id : T_1 \text{in}\ e_2 : T_2}
+\finfrule{\begin{array}{l}\typeenvx e_1 : bool \\ \typeenvx e_2 : T \\ \typeenvx e_3 : T\end{array}}{\typeenvx \text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3 : T}
 \finfrule{
@@
 \end{array}}
-{\typeenvx id[\otok{k}{r}](\oton{e}) : T_r[r_1/r'_1,\ldots,r_k/r'_k]} \\
-
+{\typeenvx id[\otok{k}{r}](\oton{e}) : T_r[r_1/r'_1,\ldots,r_k/r'_k]}
 \finfrule{
@@
 \end{array}}
-{\typeenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 : T} \\
-
+{\typeenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 : T}
 \finfrule{
@@
 \end{array}}
-{\typeenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] : T_1} \\
-
+{\typeenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] : T_1}
 \finfrule{
@@
 \end{array}}
-{\typeenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 : T_3} \\
+{\typeenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 : T_3}
+}
+\caption{Legion Core Type System}
+\label{fig:types}
+\end{figure*}
 
-\subsection{Closure of region constraints and priviledges}
-
-\framebox{
+\begin{figure}
+\centering{
 $\begin{array}{l}
@@
 r_i * r_j \in \Omega^* \Rightarrow r_j * r_i \in \Omega^* \\
-\end{array}$
-} \\
-\framebox{
-$\begin{array}{l}
+\\
 \Phi \subseteq \Phi^* \\
@@
 }
+\caption{Privilege and Constraint Closure}
+\label{fig:closure}
+\end{figure}
 
@@
 
-Operational semantics rules have the following form: \\
+Operational semantics rules have the following form:
 
-\framebox{$\opsenv{e}{v}{E}$}
+\begin{center}
+$\opsenv{e}{v}{E}$
+\end{center}
 
@@
 \begin{itemize}
-\item $M$ is the mapping from logical regions $r_i$ in types to physical regions $\rho_i$, which are sets of memory locations.  It will be useful to apply this logical-to-physical mapping to other
+\item $M$ is the mapping from logical regions $r_i$ in types to physical regions $\rho_i$, which are sets of memory locations.
+
+It will be useful to apply this logical-to-physical mapping to other
 things that defined in terms of logical regions to yield a corresponding definition in terms of
@@
 well-formed if it contains a free logical region name $r_j \not\in dom(M)$.)
-\item $L$ is the mapping of local variable names to their values
-\item $H$ is the heap typing, mapping from memory locations to types
-\item $S$ is the store, mapping from memory locations $l_i$ to values
+\item $L$ is the mapping of local variable names to their values on the stack.  These values are
+immutable, and private to the function - it is not possible to get a pointer to a local value.
+\item $H$ is the heap typing, mapping from memory locations to types.  The heap typing does not
+change during the execution of a legion program.  The Legion type system is strong enough to
+support the use of garbage collection
+to allow the reuse of a memory location (potentially with a different type stored in that
+location) once all existing references to it are dead, but we
+do not consider that further in this paper.
+\item $S$ is the store, mapping from memory locations $l_i$ to the current value stored.
 \item $C$ is the ``clobber set'', a set of memory locations that may be accessed by concurrently executing operations (??? does this need to be split into read/write/reduce?)
@@
 \item $v$ is the result of the evaluation
-\item $E$ is a memory trace (i.e. the effects) of the execution, a list of read/write/reduce operations in the order they were performed.  An operation $\epsilon$ is one of the following:
+\item $E$ is a memory trace (i.e. the effects) of the execution, a list of memory operations
+in the order they were performed.  An operation $\epsilon$ is one of the following:
 \begin{itemize}
@@
 operations), with each trace's operations remaining in order.  A valid interleaving must further
-satisfy the requirement that the value read from any location that's not in the clobber set $C$
-yield the most recently stored/reduced value.
+satisfy the requirement that the value read from the store must yield the most recently 
+stored/reduced value unless that location is part of the clobber set $C$.
+\end{itemize}
+
+A number of the rules are worth discussing in detail.
+
+Because we would like to keep the heap typing $H$ constant, the semantics for the expression
+new $T@r$ allow the choice of any memory location that falls within the right physical region
+and has the appropriate heap type.  The soundness of the type system does not depend on that
+location being one that isn't already in use, although most implementations will want to 
+add that additional guarantee.
+
+A similar choice appears in the semantics of the read operation.  If the location being read
+from is within the clobber set $C$, the semantics appear to allow any value to be returned, as
+long as it has the right type.  In reality, the choices are more constrained.  If it's not the
+most recently written/reduced value from the current expression evaluation, it must at least be
+a value written by some concurrent evaluation.  The constraint is applied further up in the 
+expression tree when the result of the function is interleaved with other memory traces.
+
+Most of the heavy lifting in the Legion core language occurs in the function call expression.
+The expressions for all the function's arguments are evaluated in the caller's context, but a
+new context is assembled for the evaluation of the function's body.  The region mapping used
+in the callee only uses the caller's mapping indirectly - the only logical region names available
+are those over which the called function is quantified.  Similarly, the set of local values is
+replaced with a mapping of formal parameter names to the values of the arguments.  The store
+is updated to account for side-effects of those argument evaluations.  Finally, if the function
+being called operates with relaxed coherence (i.e. either {\em atomic} or {\em simult} instead of
+{\em excl}), the locations with relaxed coherence are added to the clobber set, allowing the
+impact of concurrent modifications of the store to be handled by the caller.  This is done by
+annotating the called function's memory trace with the relaxed coherence on a per-location basis
+and using the {\em valid\_interleave} predicate to control what parallel execution is allowed.
 
@@
 
-A few helper functions are used in the operational semantics rules, and defined in 
-A number of rules will also use the property $valid\_interleave(S, C, E', E_1, E_2)$ to 
-constrain how the memory operations of two subexpressions can be interleaved.  We will 
-consider actual constructions of valid interleavings later.  For now, we simply rely on
-the following necessary conditions for any valid interleaving:
-\begin{enumerate}
-\item A valid interleaving $E'$ of $E_1$ and $E_2$ contains exactly all operations from both
-$E_1$ and $E_2$, in some order.
-\item The operations in each of $E_1$ and $E_2$ remain in order in $E'$.
-\item For every $read(l, c, v)$ in $E'$ for which $l$ is not in the clobber set $C$, the value 
-returned by the read is consistent with the original store $S$ and any preceding writes or
-reductions.
-\end{enumerate}
-\end{itemize}
-
-\framebox{$\opsenvx bv \mapsto bv, []$} \hspace{1cm}
-\framebox{$\opsenvx iv \mapsto iv, []$} \\
-
+\begin{figure*}
+\centering{\small
+\framebox{$\opsenvx bv \mapsto bv, []$}
+\framebox{$\opsenvx iv \mapsto iv, []$}
 \finfrule
@@
 \end{array}}
-{\opsenvx \langle e_1, e_2 \rangle \mapsto \langle v_1, v_2 \rangle, E'} \\
-
+{\opsenvx \langle e_1, e_2 \rangle \mapsto \langle v_1, v_2 \rangle, E'}
 \finfrule
@@
 {\opsenvx e \mapsto \langle v_1, v_2 \rangle, E}
-{\opsenvx e\text{.2} \mapsto v_2, E} \\
-
+{\opsenvx e\text{.2} \mapsto v_2, E}
 \finfrule
 {L(id) = v}
-{\opsenvx id \mapsto v, []} \\
-
-\framebox{$\opsenvx \text{null }T@r \mapsto null, []$} \\
-
+{\opsenvx id \mapsto v, []}
+\framebox{$\opsenvx \text{null }T@r \mapsto null, []$}
 \finfrule
@@
 l \in M(r) \\
-H(l) = T[M]
+H(l) = M \llbracket T \rrbracket
 \end{array}}
-{\opsenvx \text{new }T@r \mapsto l, []} \\
-
+{\opsenvx \text{new }T@r \mapsto l, []}
 \finfrule
-{\opsenvx e \mapsto v, E}
-{\opsenvx \text{isnull}(e) \mapsto false, E} \\
-
+{\opsenvx e \mapsto l, E}
+{\opsenvx \text{isnull}(e) \mapsto false, E}
 \finfrule
 {\opsenvx e \mapsto null, E}
-{\opsenvx \text{isnull}(e) \mapsto true, E} \\
-
+{\opsenvx \text{isnull}(e) \mapsto true, E}
 \finfrule
@@
 {\opsenvx \text{upregion}(e, \oton{r}) \mapsto v, E}
-
 \finfrule
 {\begin{array}{l}
-\opsenvx e \mapsto v, E \\
-v' = \begin{cases}
-v, & \text{if $v \not= null$ and $\exists i, v \in M(r_i)$}. \\
+\opsenvx e \mapsto l, E \\
+v = \begin{cases}
+l, & \text{if $\exists i, l \in M(r_i)$}. \\
 null, & \text{otherwise}.
@@
 \end{array}}
-{\opsenvx \text{downregion}(e, \oton{r}) \mapsto v', E} \\
-
+{\opsenvx \text{downregion}(e, \oton{r}) \mapsto v, E}
 \finfrule
@@
 S' = \text{apply}(S, E, C) \\
-v = S'[l]
+v = \begin{cases}
+S'(l), & \text{if } l \not\in C \\
+v' : H(l), & \text{otherwise}
+\end{cases}
 \end{array}}
-{\opsenvx \text{read}(e_1) \mapsto v, E \dplus [ read(l, excl, v) ]} \\
-
+{\opsenvx \text{read}(e_1) \mapsto v, E \dplus [ read(l, excl, v) ]}
 \finfrule
@@
 \opsenvx[S=S'] e_2 \mapsto v, E_2 \\
+valid\_interleave(S, C, E', E_1, E_2)
 \end{array}}
-{\opsenvx \text{write}(e_1, e_2) \mapsto true, E_1 \dplus E_2 \dplus [ write(l, excl, v) ]} \\
-
+{\opsenvx \text{write}(e_1, e_2) \mapsto true, E' \dplus [ write(l, excl, v) ]}
 \finfrule
@@
 \opsenvx[S=S'] e_2 \mapsto v, E_2 \\
+valid\_interleave(S, C, E', E_1, E_2)
 \end{array}}
-{\opsenvx \text{reduce}(id, e_1, e_2) \mapsto true, E_1 \dplus E_2 \dplus [ reduce_{id}(l, excl, v) ]} \\
-
-\framebox{$\opsenvx \text{newcolor }r \mapsto \emptyset, []$} \\
-
+{\opsenvx \text{reduce}(id, e_1, e_2) \mapsto true, E' \dplus [ reduce_{id}(l, excl, v) ]}
+\framebox{$\opsenvx \text{newcolor }r \mapsto \emptyset, []$}
 \finfrule
@@
 \end{array}}
-{\opsenvx \text{color}(e_1, e_2, e_3) \mapsto K', E'} \\
-
+{\opsenvx \text{color}(e_1, e_2, e_3) \mapsto K', E'}
 \finfrule
@@
 \end{array}}
-{\opsenvx e_1 + e_2 \mapsto v', E'} \hspace{1cm}
-
+{\opsenvx e_1 + e_2 \mapsto v', E'}
 \finfrule
@@
 \end{array}}
-{\opsenvx e_1 < e_2 \mapsto v', E'} \\
-
+{\opsenvx e_1 < e_2 \mapsto v', E'}
 \finfrule
@@
 \end{array}}
-{\opsenvx \text{let }id : T = e_1\text{ in }e_2 \mapsto v_2, E'} \\
-
+{\opsenvx \text{let }id : T = e_1\text{ in }e_2 \mapsto v_2, E'}
 \finfrule
@@
 S' = \text{apply}(S, E_1, C) \\
-\opsenvx[L=L',S=S'] e_2 \mapsto v_2, E_2 \\
-valid\_interleave(S, C, E', E_1, E_2)
+\opsenvx[L=L',S=S'] e_2 \mapsto v_2, E_2
 \end{array}}
-{\opsenvx \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_2, E'} \\
-
+{\opsenvx \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_2, E_1 \dplus E_2}
 \finfrule
@@
 S' = \text{apply}(S, E_1, C) \\
-\opsenvx[L=L',S=S'] e_3 \mapsto v_3, E_3 \\
-valid\_interleave(S, C, E', E_1, E_3)
+\opsenvx[L=L',S=S'] e_3 \mapsto v_3, E_3
 \end{array}}
-{\opsenvx \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_3, E'} \\
-
+{\opsenvx \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_3, E_1 \dplus E_3}
 \finfrule
@@
 S_n = \text{apply}(S_{n-1}, E_n, C) \\
-valid\_interleave(S, C, E', \oton{E}) \\
-S' = \text{apply}(S, E', C) \\
-task id[\otok{k}{r'}](f_1 : T_1, \ldots, f_n : T_n), \Phi_i, Q : T_r = e_{n+1} \\
+valid\_interleave(S, C, E', \oton{E})
+\vspace{1.5mm} \\
+\text{task }id[\otok{k}{r'}](f_1 : T_1, \ldots, f_n : T_n), \Phi_i, Q : T_r = e_{n+1} \\
 M' = \{ (r'_1, M(r_1)), \ldots (r'_k, M(r_k)) \} \\
 L' = \{ (f_1, v_1), \ldots, (f_n, v_n) \} \\
-\opsenvx[M=M',L=L',S=S'] e_{n+1} \mapsto v_{n+1}, E_{n+1} \\
-E'_{n+1} = mark\_coherence(E_{n+1}, M \llbracket Q \rrbracket) \\
+S' = \text{apply}(S, E', C) \\
+C' = C \cup \{ l : \exists \rho. atomic(\rho) \in M' \llbracket Q \rrbracket \vee simult(\rho) \in M' \llbracket Q \rrbracket \} \\
+\opsenvx[M=M',L=L',S=S'] e_{n+1} \mapsto v_{n+1}, E_{n+1}
+\vspace{1.5mm} \\
+E'_{n+1} = mark\_coherence(E_{n+1}, M' \llbracket Q \rrbracket) \\
 valid\_interleave(S, C, E'', E', E_{n+1})
 \end{array}}
-{\opsenvx id[\otok{k}{r}](\oton{e}) \mapsto v_{n+1}, E''} \\
-
+{\opsenvx id[\otok{k}{r}](\oton{e}) \mapsto v_{n+1}, E''}
 \finfrule{
@@
 \end{array}}
-{\opsenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 \mapsto v, E'} \\
-
+{\opsenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 \mapsto v, E'}
 \finfrule{
@@
 \end{array}}
-{\opsenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] \mapsto v', E} \\
-
+{\opsenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] \mapsto v', E}
 \finfrule{
@@
 \end{array}}
-{\opsenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 \mapsto v_2, E' } \\
+{\opsenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 \mapsto v_2, E' }
+}
+\caption{Legion Core Operational Semantics}
+\label{fig:semantics}
+\end{figure*}
 

### 2012-07-03 doc/popl2013/langdef.tex
checkpoint - done with first draft text through soundness proof

@@
+\newtheorem{thm}{Theorem}
+\newtheorem{lem}{Lemma}
+
 \newcommand{\oton}[1]{{#1}_1,\ldots,{#1}_n}
@@
 T_1 = \exists \otok{k}{r'}. T_2\text{ where }\Omega_1 \\
-\Omega \models \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \\
+\Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^* \\
 \typeenvx e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]
@@
 S' = \text{apply}(S, E_1, C) \\
-\opsenvx[L=L',S=S'] e_2 \mapsto v_2, E_2
+\opsenvx[S=S'] e_2 \mapsto v_2, E_2
 \end{array}}
@@
 S' = \text{apply}(S, E_1, C) \\
-\opsenvx[L=L',S=S'] e_3 \mapsto v_3, E_3
+\opsenvx[S=S'] e_3 \mapsto v_3, E_3
 \end{array}}
@@
 guaranteed to access the heap only in ways permitted by the privileges under which it was typed.
-A judgement $M \vdash E : \Phi$
-If
-\begin{enumerate}
-\item $\typeenvx e : T$
-\item $\opsenvx e \mapsto v, E$
-\end{enumerate}
-we would like to prove that all of the dynamic effects described by $E$ are
-included in the static privileges $\Phi$.  A judgement $E : \Phi$ means that
-for all effects $\epsilon$ in $E$,
-\begin{enumerate}
-\item if $\epsilon = read(l, v)$, then for some $r$, $reads(r) \in \Phi \wedge l \in M(r)$
-\item if $\epsilon = write(l, v)$, then for some $r$, $writes(r) \in \Phi \wedge l \in M(r)$
-\item if $\epsilon = reduce(id, l, v)$, then for some $r$, $reduces(id,r) \in \Phi \wedge l \in M(r)$
-\end{enumerate}
+A judgement $M \vdash E : \Phi$ holds if all the memory operations in $E$ are of the types and
+to locations covered by the privileges in $\Phi$:
 
-\subsection{Strategy}
+\begin{center}
+\begin{tabular}{l@{ }l@{ }l}
+$M \vdash E : \Phi$ & $\Leftrightarrow$ & $\forall \epsilon\text { in } E.$ \\
+& & $(\epsilon = read(l, c, v) \rightarrow \exists r. l \in M(r) \wedge reads(r) \in \Phi)\ \wedge$ \\
+& & $(\epsilon = write(l, c, v) \rightarrow \exists r. l \in M(r) \wedge writes(r) \in \Phi)\ \wedge$ \\
+& & $(\epsilon = reduce_{id}(l, c, v) \rightarrow \exists r. l \in M(r) \wedge reduces(id, r) \in \Phi)$
+\end{tabular}
+\end{center}
 
-The mapping $M$ is a function from logical region names $r$ to physical regions $\rho$.  It will
-be convenient to apply $M$ to types and even functions that operate on types by defining
-$\hat x = M \llbracket x \rrbracket$ to be $x$ with all logical region names $r_i$ replaced by their
-corresponding physical region $\rho_i = M(r_i)$.   \\
+The proof that $M \vdash E : \Phi$ holds for all well-typed Legion expressions is achieved by
+induction over the structure of the expression.  Soundness is contingent on the following
+consistent properties, which must be shown to hold for each subexpression:
+\begin{itemize}
+\item $M \sim \Omega$, a region mapping must satisfy the region constraints under which the expression was typed
+\item $L \sim_H \Gamma$, the local values must have types consistent with the environment under which the expression was typed
+\item $S \sim H$, the initial state of the store must contain values whose types are consisten with the heap typing
+\end{itemize}
 
-Soundness is proven using induction over the structure of an expression.  A number of properties
-are assumed and/or proven at each step: 
-\begin{enumerate}
-\item $M \sim \Omega \Leftrightarrow (\forall r_i, r_j.
-r_i \leq r_j \in \Omega \rightarrow M(r_i) \subseteq M(r_j)) \wedge (\forall r_i, r_j.
-r_i * r_j \in \Omega \rightarrow M(r_i) \cap M(r_j) = \emptyset)$, that M is an $\Omega$-satisfying mapping of logical regions to physical regions
-\item $\hat T = M \llbracket T \rrbracket$, a {\em physical type}, constructed by replacing all of the free region names $r$ in T with the physical regions $\rho = M(r)$
-\item $v \sim_H \hat T$, value consistency with respect to a heap typing.  Value consistency is
-itself defined inductively over the structure of a type:
-\begin{enumerate}
-\item $bv \sim_H bool$
-\item $iv \sim_H int$
-\item $null \sim_H T@\rho$
-\item $l \sim_H T@\rho \Leftrightarrow (l \in \rho) \wedge (H(l) \sim_H T)$
-\item $\langle v_1, v_2 \rangle \sim_H \langle \hat T_1, \hat T_2 \rangle \Leftrightarrow (v_1 \sim_H \hat T_1) \wedge (v_2 \sim_H \hat T_2)$
-\item $\langle \langle \oton{\rho}, v \rangle \rangle \sim_H \text{rr}[\oton{r}] \hat T\text{ where }\Omega \Leftrightarrow (v \sim_H \hat T[\rho_1/r_1, \ldots \rho_n/r_n]) \wedge (\{ (r_i, \rho_i) \} \sim \Omega)$
-\item $K \sim_H coloring(\rho) \Leftrightarrow \forall l_1, v_1. (l_1, v_1) \in K \rightarrow (l_1 \in \rho \wedge \forall l_2, v_2. (l_2, v_2) \in K \rightarrow (l_1 \not= l_2) \vee (v_1 = v_2))$
-\end{enumerate}
-\item $L \sim_H M \llbracket \Gamma \rrbracket \Leftrightarrow \forall (id,v) \in L. v \sim_H M \llbracket \Gamma \rrbracket (id) $, local stack consistency with respect to a heap typing
-\item $S \sim H \Leftrightarrow \forall (l,v) \in S. v \sim_H H(l)$, store consistency with respect to a given heap typing
-\item $E \sim H \Leftrightarrow (\forall l, v.
-write(l,v) \in E \rightarrow v \sim_H H(l)) \wedge (\forall id, l, v.
-reduce_{id}(l, v) \in E \rightarrow (\hat \Gamma (id) = (\hat T_1, \hat T_2), \emptyset \rightarrow \hat T_1) \wedge H(l) = \hat T_1 \wedge v \sim_H \hat T_2)$, effect consistnecy with respect to a given heap typing
-\end{enumerate}
+The proof requires two additional properties be proved for each subexpression:
+\begin{itemize}
+\item $v \sim_H M \llbracket T \rrbracket$, a subexpresion must evaluate to a value of the right type
+\item $E \sim H$, the values written to the store must match the heap types of the location to which they're stored
+\end{itemize}
+Figure~\ref{fig:constprop} gives the precise definition of these properties.
 
-Armed with all of these, we look at the following property:
-\begin{enumerate}
-\item $E \sim_H \Phi$, containment of effects within privileges
-\end{enumerate}
+\begin{figure*}
+\centering{\footnotesize
+$
+\begin{array}{l}
+\begin{array}{ll}
+\begin{array}{lll}
+M \sim \Omega & \Leftrightarrow & (\forall r_i, r_j. r_i \leq r_j \in \Omega \rightarrow M(r_i) \subseteq M(r_j))\ \wedge \\
+              &                 & (\forall r_i, r_j. r_i * r_j \in \Omega \rightarrow M(r_i) \cap M(r_j) = \emptyset) \\
+L \sim_H \Gamma & \Leftrightarrow & \forall (id,v) \in L. v \sim_H M \llbracket \Gamma \rrbracket (id) \\
+S \sim H & \Leftrightarrow & \forall (l,v) \in S. v \sim_H H(l) \\
+\end{array} & 
+\begin{array}{lll}
+E \sim H & \Leftrightarrow & (\forall l, c, v. write(l,c, v) \in E \rightarrow v \sim_H H(l))\ \wedge \\
+& & (\forall id, l, v. reduce_{id}(l, v) \in E \rightarrow \\
+& & (M \llbracket \Gamma \rrbracket (id) = (\hat T_1, \hat T_2), \emptyset, \emptyset \rightarrow \hat T_1) \wedge H(l) = \hat T_1 \wedge v \sim_H \hat T_2) \\
+\end{array}
+\end{array} \vspace{3mm} \\
+\begin{array}{l@{\hspace{1cm}}l}
+\begin{array}{l}
+bv \sim_H bool \\
+iv \sim_H int \\
+null \sim_H \hat T@\rho \\
+\end{array} &
+\begin{array}{lll}
+l \sim_H \hat T@\rho & \Leftrightarrow & l \in \rho \wedge H(l) \sim_H \hat T \\
+\langle v_1, v_2 \rangle \sim_H \langle \hat T_1, \hat T_2 \rangle & \Leftrightarrow & (v_1 \sim_H \hat T_1) \wedge (v_2 \sim_H \hat T_2) \\
+\langle \langle \oton{\rho}, v \rangle \rangle \sim_H \text{rr}[\oton{r}] \hat T\text{ where }\Omega & \Leftrightarrow & (v \sim_H \hat T[\rho_1/r_1, \ldots \rho_n/r_n]) \wedge (\{ (r_i, \rho_i) \} \sim \Omega) \\
+K \sim_H coloring(\rho) & \Leftrightarrow & \forall l_1, v_1. (l_1, v_1) \in K \rightarrow (l_1 \in \rho\ \wedge \\
+& & \forall l_2, v_2. (l_2, v_2) \in K \rightarrow (l_1 \not= l_2) \vee (v_1 = v_2)) \\
+\end{array}
+\end{array}
+\end{array}
+$
+}
+\caption{Consistency Properties}
+\label{fig:constprop}
+\end{figure*}
 
-and prove
+\begin{thm}
+For any expression $e$, if:
+\begin{center}
+$
+\begin{array}{l} \\
+\typeenv{e}{T} \\
+\opsenv{e}{v}{E} \\
+M \sim \Omega \\
+L \sim_H \Gamma \\
+S \sim H
+\end{array}
+$
+\end{center}
+then:
+\begin{center}
+$
+\begin{array}{l}
+v \sim_H M \llbracket T \rrbracket \\
+E \sim H \\
+M \vdash E : \Phi
+\end{array}
+$
+\end{center}
+\end{thm}
 
-\finfrule
-{\begin{array}{l}
-\typeenvx e : T \\
-\opsenvx e \mapsto v, E \\
-M \sim \Omega \hspace{1cm}
-L \sim_H \Gamma \hspace{1cm}
-S \sim H
-\end{array}}
-{
-v \sim_H M \llbracket T \rrbracket \hspace{1cm}
-E \sim H \hspace{1cm}
-E : \Phi
-}
+Although there are 23 different cases to be proven, many of them are similar, and benefit from
+the use of the following lemmas:
 
-A few lemmas will be useful:
+\begin{lem}
+\label{lemma:heapconst:apply}
+If $S \sim H$ and $E \sim H$, then $\text{apply}(S, E, C) \sim H$.
+\end{lem}
 
-\finfrule
-{\begin{array}{l}
-S \sim H \hspace{1cm}
-E \sim H
-\end{array}}
-{ \text{apply}(S, E, C) \sim H }
+\begin{lem}
+\label{lemma:heapconst:effects1}
+If $E_1 \sim H$ and $E_2 \sim H$, then $E_1 \dplus E_2 \sim H$.
+\end{lem}
 
-\finfrule
-{\begin{array}{l}
-E_1 \sim H \hspace{1cm}
-E_2 \sim H
-\end{array}}
-{ E_1 \dplus E_2 \sim H } \\
+\begin{lem}
+\label{lemma:heapconst:effects2}
+If $E_1 \sim H$ and $E_2 \sim H$ and $valid\_interleave(S, C, E', E_1, E_2)$, then $E' \sim H$.
+\end{lem}
 
-\finfrule
-{\begin{array}{l}
-E_1 \sim H \hspace{1cm}
-E_2 \sim H \\
-valid\_interleave(S, C, E', E_1, E_2)
-\end{array}}
-{ E' \sim H } \\
+\begin{lem}
+\label{lemma:effsound:effects1}
+If $M \vdash E_1 : \Phi$ and $M \vdash E_2 : \Phi$, then $M \vdash E_1 \dplus E_2 : \Phi$.
+\end{lem}
 
-\finfrule
-{\begin{array}{l}
-E_1 : \Phi \hspace{1cm}
-E_2 : \Phi
-\end{array}}
-{ E_1 \dplus E_2 : \Phi } \\
+\begin{lem}
+\label{lemma:effsound:effects2}
+If $M \vdash E_1 : \Phi$ and $M \vdash E_2 : \Phi$ and $valid\_interleave(S, C, E', E_1, E_2)$, then $M \vdash E' : \Phi$.
+\end{lem}
 
-\framebox{$M \sim \Omega \Leftrightarrow M \sim \Omega^*$} \\
+\begin{lem}
+\label{lemma:omegaclosure}
+$M \sim \Omega^*$ if and only if $M \sim \Omega$.
+\end{lem}
 
-\framebox{$E : \Phi \Leftrightarrow E : \Phi^*$}
+\begin{lem}
+\label{lemma:phiclosure}
+$M \vdash E : \Phi^*$ if ant only if $M \vdash E : \Phi$.
+\end{lem}
 
+The complete soundness proof can be found here\cite{xyz} - for space reasons we just discuss
+the general strategy here.
+
+\begin{itemize}
+
+\item Only three expressions have subexpressions that modify $M$ or $\Omega$.  For the
+{\em partition} expression, the assumed consistency of the coloring can be shown to preserve
+the region mapping consistency with respect to the constraints.  Similarly, the consistency of 
+a region relation instance guarantees the maintenance of consistency in an {\em unpack}
+expression.  Finally, the body of a called function executes with a different region mapping,
+but with an initially-empty set of constraints, so they are trivially satisfied.
+
+\item Only three expression have subexpressions that modify $L$, $\Gamma$, $M$.  The
+{\em partition} expression only modifies $M$, and the requirement that it not reuse existing
+names ensures that $M \llbracket \Gamma \rrbracket$ does not change.  The value and type of the
+binding created in a {\em let} expression is obviously consistent, while the binding created in
+an {\em unpack} expression is less obviously so, requiring an induction over the type of the 
+unpacked value to show equivalence under the new mapping.  The last case is again the body of a
+called function, which requires the same style of proof as for the {\em unpack} expression to
+be performed for each formal parameter.
+
+\item The heap typing consistency of all stores used in subexpressions follows directly from
+Lemma~\ref{lemma:heapconst:apply}.
+
+\item With the necessary consistency conditions all satified, the inductive hypothesis asserts
+the type consistency of all subexpression result values.  The consistency of the result of 
+the {\em upregion} is guaranteed by the type checking requirement of appropriate subregion
+constraints and the mapping's consistency with those constraints, whereas {\em downregion}
+expression's result is consistent due to the run time check that is performed.  (The requirement
+for appropriate constraints in the {\em downregion} typing rule is primarily for performance
+(i.e. knowing which part of the region tree to examine) and to catch downcasts that will never
+work at runtime.)
+
+The consistency of a {\em read} expression's result is trivial when an address falls in the
+clobber set, and counts on the consistency of the store otherwise.  The consistency of a 
+{\em color} expression's result depends on the pointer subexpression's consistency and the 
+removal of any previous coloring of that location from the coloring set.  The remaining
+interesting cases arise from changes to the mapping $M$ rather than transformations on the
+value $v$.  In the case of {\em partition} and {\em unpack} expressions, the type system's
+requirement that the body subexpression's result not use the regions that were added to the
+mapping allow the changes to the mapping to be ignored.  The last case is, yet again, the body
+of a called function, and the same strategy that was used for the type consistency of the
+formal parameters works in reverse for the function's result.
+
+\item The type consistency of the value in an expression's effect follows from
+Lemmas~\ref{lemma:heapconst:effects1} and~\ref{lemma:heapconst:effects2}.  New effects are
+added by the {\em write} and {\em reduce} expressions, but their values' consistency follows
+directly from the inductive hypothesis on their subexpressions.  Finally, the consistency of the
+values in a called function's memory trace is addressed in the same way as the return value.
+
+\item The proof of the crucial property of containment of effects within the available privileges
+is very similar to the previous step, with Lemmas~\ref{lemma:effsound:effects1}
+and~\ref{lemma:effsound:effects2} covering most cases, straightforward proofs for the {\em read},
+{\em write}, and {\em reduce} expressions, and one final special case for function calls.  The 
+inductive hypothesis asserts that the function body's effects are contained within its own
+privileges.  Those are a subset of the caller's $\Phi^*$, and then Lemma~\ref{lemma:phiclosure}
+is used to complete the proof.
+\end{itemize}
+
+modify $M$
 \begin{table*}

### 2012-07-03 doc/popl2013/langdef.tex
some edits suggested by Alex

@@
 {\begin{array}{l}
-\typeenvx e : T@(\otok{k}{r'}) \\
+\typeenvx e : T@(r'_1, \ldots r'_k) \\
 \forall i. \exists j, r'_i \leq r_j \in \Omega^* \\
@@
 {\begin{array}{l}
-\typeenvx e : T@(\otok{k}{r'}) \\
+\typeenvx e : T@(r'_1, \ldots r'_k) \\
 \forall j. \exists i, r_j \leq r'_i \in \Omega^* \\
@@
 \typeenvx e_1 : T_1 \\
-\Gamma' = \Gamma[id/T_1] \\
-\typeenvx[G=\Gamma'] e_2 : T_2
+\typeenvx[G={\Gamma[id/T_1]}] e_2 : T_2
 \end{array}}
@@
 \begin{array}{l}
-\Gamma(id) = \forall \otok{k}{r'}.(\oton{T}),\Phi_i, Q \rightarrow T_r \\
+\Gamma(id) = \forall r'_1, \ldots r'_k.(\oton{T}),\Phi', Q' \rightarrow T_r \\
 \typeenvx e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\
-\Phi_i[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*
+\Phi'[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*
 \end{array}}
@@
 \typeenvx e_1 : \text{coloring}(r_p) \\
-\Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{i \not= j} r_i * r_j \\
+\Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{1 \leq i < j \leq k} r_i * r_j \\
 \typeenvx[O=\Omega'] e_2 : T \\
-r_i \not\in \textit{regions\_of}(\Gamma, T)
+\{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T) = \emptyset
 \end{array}}
@@
 \begin{array}{l}
-T_1 = \exists \otok{k}{r'}. T_2\text{ where }\Omega_1 \\
+T_1 = \exists r'_1, \ldots r'_n.\ T_2\text{ where }\Omega_1 \\
 \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^* \\
@@
 \begin{array}{l}
-T_1 = \exists \otok{k}{r'}. T_2\text{ where }\Omega_1 \\
+T_1 = \exists r'_1, \ldots, r'_n.\ T_2\text{ where }\Omega_1 \\
 \typeenvx e_1 : T_1 \\
 \Gamma' = \Gamma[T_2[r_1/r'_1,\ldots,r_k/r'_k] / id] \\
-\Omega' = \Omega \wedge \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \\
+\Omega' = \Omega \cup \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \\
 \typeenvx[G=\Gamma',O=\Omega'] e_2 : T_3 \\
-r_i \not\in \textit{regions\_of}(\Gamma, T_1, T_3)
+\{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T_1, T_3) = \emptyset
 \end{array}}
@@
 reads(r) \in \Phi^* \wedge writes(r) \in \Phi^* \wedge \\
-\hspace{1cm} \Gamma(id) = \forall \otok{k}{r}. (\oton{T}), E \rightarrow T_r \Rightarrow reduces(id, r) \in \Phi^*
+\hspace{1cm} \Gamma(id) = \forall \otok{k}{r}. (\oton{T}), \Phi', Q' \rightarrow T_r \Rightarrow reduces(id, r) \in \Phi^*
 \end{array}$
@@
 \begin{itemize}
-\item $apply(S, E, C)$ is used to determine the effect on the store of a sequence of memory operations
+\item $apply(S, E)$ is used to determine the effect on the store of a sequence of memory operations
 \item $mark\_coherence(E, \hat Q)$ marks the memory operations in a subtask's trace to
@@
 \begin{array}{lll}
-apply(S, [], C) & = & S \\
-apply(S, E \dplus [ read(l, c, v) ]) & = & apply(S, E, C) \\
-apply(S, E \dplus [ write(l, c, v) ]) & = & apply(S, E, C)[v/l] \\
-apply(S, E \dplus [ reduce_{id}(l, c, v) ]) & = & apply(S, E, C)[id(S[l], v)/l] \\
+apply(S, []) & = & S \\
+apply(S, E \dplus [ read(l, c, v) ]) & = & apply(S, E) \\
+apply(S, E \dplus [ write(l, c, v) ]) & = & apply(S, E)[v/l] \\
+apply(S, E \dplus [ reduce_{id}(l, c, v) ]) & = & S'[id(S'(l), v)/l], \text{ where } S' = apply(S, E) \\
 mark\_coherence([], \hat Q) & = & [] \\
@@
 valid\_interleave(S, C, E', \oton{E}) & \rightarrow & any\_interleave(E', \oton{E}) \\
-valid\_interleave(S, C, E' \dplus [ read(l, c, v) ] \dplus E'', \oton{E}) & \rightarrow & l \in C \vee S'(l) = v, \text{ where } S' = apply(S, E', C)
+valid\_interleave(S, C, E' \dplus [ read(l, c, v) ] \dplus E'', \oton{E}) & \rightarrow & l \in C \vee S'(l) = v, \text{ where } S' = apply(S, E')
 \end{array}
@@
 \opsenvx e_1 \mapsto v_1, E_1 \\
-S' = apply(S, E_1, C) \\
+S' = apply(S, E_1) \\
 \opsenvx[S=S'] e_2 \mapsto v_2, E_2 \\
@@
 {\opsenvx e \mapsto l, E}
-{\opsenvx \text{isnull}(e) \mapsto false, E}
+{\opsenvx \text{isnull}(e) \mapsto \textit{false}, E}
 \finfrule
@@
 \opsenvx e_1 \mapsto l, E \\
-S' = \text{apply}(S, E, C) \\
+S' = \text{apply}(S, E) \\
 v = \begin{cases}
@@
 \opsenvx e_1 \mapsto l, E_1 \\
-S' = \text{apply}(S, E_1, C) \\
+S' = \text{apply}(S, E_1) \\
 \opsenvx[S=S'] e_2 \mapsto v, E_2 \\
@@
 \opsenvx e_1 \mapsto K, E_1 \\
-S' = \text{apply}(S, E_1, C) \\
+S' = \text{apply}(S, E_1) \\
 \opsenvx[S=S'] e_2 \mapsto l, E_2 \\
-S'' = \text{apply}(S', E_2, C) \\
+S'' = \text{apply}(S', E_2) \\
 \opsenvx[S=S''] e_3 \mapsto v, E_3 \\
@@
 \opsenvx e_1 \mapsto v_1, E_1 \\
-S' = \text{apply}(S, E_1, C) \\
+S' = \text{apply}(S, E_1) \\
 \opsenvx[S=S'] e_2 \mapsto v_2, E_2 \\
@@
 true, & \text{if $v_1 < v_2$}. \\
-false, & \text{otherwise}.
+\textit{false}, & \text{otherwise}.
 \end{cases} \\
@@
 L' = L[v_1/id] \\
-S' = \text{apply}(S, E_1, C) \\
+S' = \text{apply}(S, E_1) \\
 \opsenvx[L=L',S=S'] e_2 \mapsto v_2, E_2 \\
@@
 \opsenvx e_1 \mapsto true, E_1 \\
-S' = \text{apply}(S, E_1, C) \\
+S' = \text{apply}(S, E_1) \\
 \opsenvx[S=S'] e_2 \mapsto v_2, E_2
@@
 {\begin{array}{l}
-\opsenvx e_1 \mapsto false, E_1 \\
-S' = \text{apply}(S, E_1, C) \\
+\opsenvx e_1 \mapsto \textit{false}, E_1 \\
+S' = \text{apply}(S, E_1) \\
 \opsenvx[S=S'] e_3 \mapsto v_3, E_3
@@
 \opsenvx e_1 \mapsto v_1, E_1 \\
-S_1 = \text{apply}(S, E_1, C) \\
+S_1 = \text{apply}(S, E_1) \\
 \opsenvx e_2 \mapsto v_2, E_2 \\
 \ldots \\
-S_n = \text{apply}(S_{n-1}, E_n, C) \\
+S_n = \text{apply}(S_{n-1}, E_n) \\
 valid\_interleave(S, C, E', \oton{E})
 \vspace{1.5mm} \\
-\text{task }id[\otok{k}{r'}](f_1 : T_1, \ldots, f_n : T_n), \Phi_i, Q : T_r = e_{n+1} \\
+\text{task }id[\otok{k}{r'}](a_1 : T_1, \ldots, a_n : T_n), \Phi', Q' : T_r = e_{n+1} \\
 M' = \{ (r'_1, M(r_1)), \ldots (r'_k, M(r_k)) \} \\
-L' = \{ (f_1, v_1), \ldots, (f_n, v_n) \} \\
-S' = \text{apply}(S, E', C) \\
-C' = C \cup \{ l : \exists \rho. atomic(\rho) \in M' \llbracket Q \rrbracket \vee simult(\rho) \in M' \llbracket Q \rrbracket \} \\
+L' = \{ (a_1, v_1), \ldots, (a_n, v_n) \} \\
+S' = \text{apply}(S, E') \\
+C' = C \cup \{ l : \exists \rho. atomic(\rho) \in M' \llbracket Q' \rrbracket \vee simult(\rho) \in M' \llbracket Q' \rrbracket \} \\
 \opsenvx[M=M',L=L',S=S'] e_{n+1} \mapsto v_{n+1}, E_{n+1}
 \vspace{1.5mm} \\
-E'_{n+1} = mark\_coherence(E_{n+1}, M' \llbracket Q \rrbracket) \\
+E'_{n+1} = mark\_coherence(E_{n+1}, M' \llbracket Q' \rrbracket) \\
 valid\_interleave(S, C, E'', E', E_{n+1})
@@
 \opsenvx e_1 \mapsto K, E_1 \\
-\rho_i = \{ l : (l, i) \in K \} \\
+\rho_i = \{ l : (l, i) \in K \}, \text{ for } 1 \leq i \leq k \\
 M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \\
-S' = \text{apply}(S, E_1, C) \\
+S' = \text{apply}(S, E_1) \\
 \opsenvx[M=M',S=S'] e_2 \mapsto v, E_2 \\
@@
 \opsenvx e_1 \mapsto v, E \\
-\rho_i = M[r_i] \\
+\rho_i = M[r_i], \text{ for } 1 \leq i \leq k \\
 v' = \langle \langle \otok{k}{\rho}, v \rangle \rangle
@@
 L' = L[v_1/id] \\
-S' = \text{apply}(S, E_1, C) \\
+S' = \text{apply}(S, E_1) \\
 \opsenvx[M=M',L=L',S=S'] e_2 \mapsto v_2, E_2 \\
@@
 \label{lemma:heapconst:apply}
-If $S \sim H$ and $E \sim H$, then $\text{apply}(S, E, C) \sim H$.
+If $S \sim H$ and $E \sim H$, then $\text{apply}(S, E) \sim H$.
 \end{lem}

### 2012-07-03 doc/popl2013/legion-types.tex
mostly-complete draft of language definition and typing rules
more stuff (not all of it good)

@@
 \usepackage{cancel}
+\usepackage{multirow}
 
@@
 
-\title{Legion: Expressing Locality and Independence with Logical Regions}
+\title{A Type System for Safe and Scalable Parallelism using Logical Regions}
+%\title{Scalable and Safe Parallelism using Logical Regions}
 \authorinfo{}{}{}
@@
 \begin{abstract}
-Limit is 12 pages, including bibliography and appendices.
-Deadlines: \\
+Legion is a programming model that introduces 
+{\em logical regions} as a mechanism for targeting large heterogeneous machines 
+with complex memory hierarchies.  Logical regions are a first-class 
+feature of Legion that allow programmers to specify independence 
+and locality properties of program data dynamically.  We present
+a type and effect system for reasoning about logical regions in Legion.  
+We prove a soundness result for our type system and use it to show
+that all memory accesses in Legion are safe.  Futhermore, we prove that
+the soundess of our type system enables a scalable hierarchical
+scheduling algorithm for Legion programs.  We implement a type checker
+for Legion programs and improve performance up to 85\% by eliding
+unncessary memory checks.  We demonstrate an implementation of the Legion 
+runtime based on the hierarchical scheduling algorithm scales for
+three real-world applications.
 
-\begin{tabular}{ll}
-Paper registration & July 7, 2012 3:59am PDT \\
-Paper submission & July 11, 2012 3:59am PDT \\
-Author response & September 10 - September 13, 2012 \\
-Author notification & October 1, 2012
-\end{tabular}
+%Limit is 12 pages, including bibliography and appendices.
+%Deadlines: \\
+
+%\begin{tabular}{ll}
+%Paper registration & July 7, 2012 3:59am PDT \\
+%Paper submission & July 11, 2012 3:59am PDT \\
+%Author response & September 10 - September 13, 2012 \\
+%Author notification & October 1, 2012
+%\end{tabular}
 \end{abstract}
@@
 Introduction goes here.
+
+%\input{outline}
+
+\input{langdef}
 

### 2012-07-03 doc/popl2013/langdef.tex
\rightarrow \rightarrow \Rightarrow

@@
 \end{cases} \\
-any\_interleave(E', [], \ldots, []) & \rightarrow & true \\
-any\_interleave([ op(l, c, v) ] \dplus E', E_1, \ldots, [ op(l, c, v) ] \dplus E_i, \ldots, E_n) & \rightarrow & any\_interleave(E', E_1, \ldots, E_i, \ldots, E_n) \\
-valid\_interleave(S, C, E', \oton{E}) & \rightarrow & any\_interleave(E', \oton{E}) \\
-valid\_interleave(S, C, E' \dplus [ read(l, c, v) ] \dplus E'', \oton{E}) & \rightarrow & l \in C \vee S'(l) = v, \text{ where } S' = apply(S, E')
+any\_interleave(E', [], \ldots, []) & = & true \\
+any\_interleave([ op(l, c, v) ] \dplus E', E_1, \ldots, [ op(l, c, v) ] \dplus E_i, \ldots, E_n) & = & any\_interleave(E', E_1, \ldots, E_i, \ldots, E_n) \\
+valid\_interleave(S, C, E', \oton{E}) & \Rightarrow & any\_interleave(E', \oton{E}) \\
+valid\_interleave(S, C, E' \dplus [ read(l, c, v) ] \dplus E'', \oton{E}) & \Rightarrow & l \in C \vee S'(l) = v, \text{ where } S' = apply(S, E')
 \end{array}
@@
 $M \vdash E : \Phi$ & $\Leftrightarrow$ & $\forall \epsilon\text { in } E.$ \\
-& & $(\epsilon = read(l, c, v) \rightarrow \exists r. l \in M(r) \wedge reads(r) \in \Phi)\ \wedge$ \\
-& & $(\epsilon = write(l, c, v) \rightarrow \exists r. l \in M(r) \wedge writes(r) \in \Phi)\ \wedge$ \\
-& & $(\epsilon = reduce_{id}(l, c, v) \rightarrow \exists r. l \in M(r) \wedge reduces(id, r) \in \Phi)$
+& & $(\epsilon = read(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reads(r) \in \Phi)\ \wedge$ \\
+& & $(\epsilon = write(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge writes(r) \in \Phi)\ \wedge$ \\
+& & $(\epsilon = reduce_{id}(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reduces(id, r) \in \Phi)$
 \end{tabular}
@@
 \begin{array}{lll}
-M \sim \Omega & \Leftrightarrow & (\forall r_i, r_j. r_i \leq r_j \in \Omega \rightarrow M(r_i) \subseteq M(r_j))\ \wedge \\
-              &                 & (\forall r_i, r_j. r_i * r_j \in \Omega \rightarrow M(r_i) \cap M(r_j) = \emptyset) \\
+M \sim \Omega & \Leftrightarrow & (\forall r_i, r_j. r_i \leq r_j \in \Omega \Rightarrow M(r_i) \subseteq M(r_j))\ \wedge \\
+              &                 & (\forall r_i, r_j. r_i * r_j \in \Omega \Rightarrow M(r_i) \cap M(r_j) = \emptyset) \\
 L \sim_H \Gamma & \Leftrightarrow & \forall (id,v) \in L. v \sim_H M \llbracket \Gamma \rrbracket (id) \\
@@
 \begin{array}{lll}
-E \sim H & \Leftrightarrow & (\forall l, c, v. write(l,c, v) \in E \rightarrow v \sim_H H(l))\ \wedge \\
-& & (\forall id, l, v. reduce_{id}(l, v) \in E \rightarrow \\
-& & (M \llbracket \Gamma \rrbracket (id) = (\hat T_1, \hat T_2), \emptyset, \emptyset \rightarrow \hat T_1) \wedge H(l) = \hat T_1 \wedge v \sim_H \hat T_2) \\
+E \sim H & \Leftrightarrow & (\forall l, c, v. write(l,c, v) \in E \Rightarrow v \sim_H H(l))\ \wedge \\
+& & (\forall id, l, v. reduce_{id}(l, v) \in E \Rightarrow \\
+& & (M \llbracket \Gamma \rrbracket (id) = (\hat T_1, \hat T_2), \emptyset, \emptyset \Rightarrow \hat T_1) \wedge H(l) = \hat T_1 \wedge v \sim_H \hat T_2) \\
 \end{array}
@@
 \langle \langle \oton{\rho}, v \rangle \rangle \sim_H \text{rr}[\oton{r}] \hat T\text{ where }\Omega & \Leftrightarrow & (v \sim_H \hat T[\rho_1/r_1, \ldots \rho_n/r_n]) \wedge (\{ (r_i, \rho_i) \} \sim \Omega) \\
-K \sim_H coloring(\rho) & \Leftrightarrow & \forall l_1, v_1. (l_1, v_1) \in K \rightarrow (l_1 \in \rho\ \wedge \\
-& & \forall l_2, v_2. (l_2, v_2) \in K \rightarrow (l_1 \not= l_2) \vee (v_1 = v_2)) \\
+K \sim_H coloring(\rho) & \Leftrightarrow & \forall l_1, v_1. (l_1, v_1) \in K \Rightarrow (l_1 \in \rho\ \wedge \\
+& & \forall l_2, v_2. (l_2, v_2) \in K \Rightarrow (l_1 \not= l_2) \vee (v_1 = v_2)) \\
 \end{array}

### 2012-07-03 doc/popl2013/intro.tex
Creating a separate tex file for the introduction

\section{Introduction}
\label{sec:intro}
Introduction goes here.

### 2012-07-03 doc/popl2013/langdef.tex
checkpoint

@@
 \newcommand{\otok}[2]{{#2}_1,\ldots,{#2}_{#1}}
-\newcommand{\dplus}{\text{++}}
+\newcommand{\dplus}{\text{+\!+}}
 \newcommand{\llbracket}{[\![}
@@
 \end{array}}
-{\typeenvx \text{write}(e_1, e_2) : bool}
+{\typeenvx \text{write}(e_1, e_2) : T@(\oton{r})}
 \finfrule
@@
 \end{array}}
-{\typeenvx \text{reduce}(id, e_1, e_2) : bool}
+{\typeenvx \text{reduce}(id, e_1, e_2) : T_1@(\oton{r})}
 \framebox{$\typeenvx \text{newcolor }r : \text{coloring}(r)$}
@@
 {\typeenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 : T_3}
+\finfrule{
+\begin{array}{l}
+\text{for }1 \leq i \leq p, \\
+\Gamma(id_i) = \forall r_{i1}, \ldots r_{ik_i}. (T_{i1}, \ldots, T_{in_i}), \Phi_i, Q_i \rightarrow T_{ir} \\
+\Gamma_i = \Gamma[a_{i1}/T_{i1}, \ldots, a_{in_i}/T_{in_i}] \\
+\typeenv[G={\Gamma_i},P={\Phi_i},O=\emptyset]{e_i}{T_{ir}}
+\end{array}}
+{ 
+\begin{array}{l@{ }l}
+\vdash \{ & \text{task }id_1[r_{11}, \ldots, r_{1k_1}]( a_{11} : T_{11}, \ldots a_{1n_1} : T_{1n_1} ), \Phi_1, Q_1 : T_{1r} : e_1, \\
+          & \ldots \\
+          & \text{task }id_p[r_{p1}, \ldots, r_{pk_1}]( a_{p1} : T_{p1}, \ldots a_{pn_p} : T_{pn_p} ), \Phi_p, Q_p : T_{pr} : e_p \} : \bullet
+\end{array} }
 }
@@
 \end{array}}
-{\opsenvx \text{write}(e_1, e_2) \mapsto true, E' \dplus [ write(l, excl, v) ]}
+{\opsenvx \text{write}(e_1, e_2) \mapsto l, E' \dplus [ write(l, excl, v) ]}
 \finfrule
@@
 \end{array}}
-{\opsenvx \text{reduce}(id, e_1, e_2) \mapsto true, E' \dplus [ reduce_{id}(l, excl, v) ]}
+{\opsenvx \text{reduce}(id, e_1, e_2) \mapsto l, E' \dplus [ reduce_{id}(l, excl, v) ]}
 \framebox{$\opsenvx \text{newcolor }r \mapsto \emptyset, []$}
@@
 
+\begin{figure*}
+$
+\begin{array}{l@{}l}
+valid\_interleave(S, C, E', E_1, E_2) = & any\_interleave(E', E_1, E_2) \wedge \\
+& coherent\_interleave(S, apply(S, E_1), C, excl, E', E_1, E_2) \wedge \\
+& (coherent\_interleave(S, apply(S, E_1), C, atomic, E', E_1, E_2) \vee \\
+& coherent\_interleave(S, apply(S, E_2), C, atomic, E', E_2, E_1)) \\
+
+coherent\_interleave(S_1, S_2, C, c, [], [], []) = & true \\
+coherent\_interleave(S_1, S_2, C, c, [ op(l, c', v) ] \dplus E', [ op(l, c, v) ] \dplus E_1, E_2) = & (op = read \wedge l \not\in C \wedge c = c' \Rightarrow S_1(l) = v) \wedge \\
+& coherent\_interleave(apply(S_1, op(l, c', v)), S_2, C, c, E', E_1, E_2) \\
+coherent\_interleave(S_1, S_2, C, c, [ op(l, c', v) ] \dplus E', E_1, [ op(l, c, v) ] \dplus E_2) = & (op = read \wedge l \not\in C \wedge c = c' \Rightarrow S_2(l) = v) \wedge \\
+& coherent\_interleave(S_1, apply(S_2, op(l, c', v)), C, c, E', E_1, E_2)
+\end{array}
+$
+\caption{work in progress}
+\end{figure*}
+
 \framebox{$\forall S, C, E_1, E_2. valid\_interleave(S, C, E_1 \dplus E_2, E_1, E_2)$}

### 2012-07-03 doc/popl2013/intro.tex

@@
 Introduction goes here.
+
+% This is a description of how the listings should be formatted.
+% It can go anywhere before the listings.
+\lstset{
+  captionpos=b,
+  language=Haskell,
+  basicstyle=\scriptsize,
+  numbers=left,
+  numberstyle=\tiny,
+  columns=fullflexible,
+  stepnumber=1,
+  escapechar=\#,
+  keepspaces=true,
+  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
+  morekeywords={task,rr,int,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,newcolor,atomic,simultaneous},
+  deletekeywords={float,head,min,max}
+}
+
+% The linked list example
+% Add float={t} to the list of options if you want the listing at the top
+\begin{lstlisting}[label={lst:linkedlist},caption={Linked List Example}]
+type list<r> = < int, list<r>@r >
+
+task sum_list[r] ( l:list<r>@r ), reads(r) : int =
+    if isnull(l) then 0 else
+    let v : list<r> = read(l) in
+        v.1 + sum_list[r](v.2)
+\end{lstlisting}
+
+% The tree example with partitioning and tree reversal
+\begin{lstlisting}[float={t},label={lst:treereverse},caption={Tree Partitioning and Reversal Example}]
+type tree<rl,rls,rt> = rr[rll, rlr, rtl, rtr] 
+                       < list<rl>@rls, < tree<rl,rll,rtl>@rtl, tree<rl,rlr,rtr>@rtr > >
+     		       where rll #$\le$# rls and rlr #$\le$# rls and rll #$*$# rlr and 
+                                  rtl #$\le$# rt and rtr #$\le$# rt  and rtl #$*$# rtr
+
+task color_list[r,rs] ( clr: coloring(rs), head: list<r>@r, 
+                        splitval: int ), reads(r) : coloring(rs) =
+    if isnull(head) then clr else
+    let v : list<r> = read(head) in
+    let ps : list<r>@rs = downregion(head, rs) in
+        color_list[r,rs](if isnull(ps) then clr 
+                                else color(clr, ps, if v.1 #$<$# splitval then 1 else 2), v.2, splitval)
+
+task make_tree[rl,rls,rt] ( l: list<rl>@rl, min: int, 
+                            max: int), reads(rl,rt), writes(rt) : tree<rl,rls,rt>@rt =
+    if isnull(l) then null tree<rl,rls,rt>@rt else 
+    let v : list<rl> = read(l) in
+    if (v.1 #$<$# min) | (v.1 #$>$# max) then
+    make_tree[rl,rls,rt](v.2, min, max) else
+    let root : tree<rl,rls,rt>@rt = new tree<rl,rls,rt>@rt in
+    partition rt using newcolor rt as rtl,rtr in
+    partition rls using color_list[rl,rls](newcolor rls, v.2, v.1) as rll,rlr in
+    let ptr : list<rl>@rls = downregion(l, rls) in
+    let lhs : tree<rl,rll,rtl>@rtl = make_tree[rl,rll,rtl](v.2, min, v.1) in
+    let rhs : tree<rl,rlr,rtr>@rtr = make_tree[rl,rlr,rtr](v.2, v.1, max) in
+    let v2 : tree<rl,rls,rt> = pack < ptr, < lhs, rhs > > as tree<rl,rls,rt>[rll, rlr, rtl, rtr] in
+        write(root, v2)
+
+task reverse_tree[rl,rls,rt] ( root: tree<rl,rls,rt>@rt ), reads(rt), writes(rt) : tree<rl,rls,rt>@rt =
+     if isnull(root) then null tree<rl,rls,rt>@rt else
+     let v : tree<rl,rls,rt> = read(root) in
+     unpack v as u : tree<rl,rls,t>[rll, rlr, rtl, rtr] in
+     let lhs_r : tree<rl,rll,rtl>@rtl = reverse_tree[rl,rll,rtl](u.2.1) in
+     let rhs_r : tree<rl,rlr,rtr>@rtr = reverse_tree[rl,rlr,rtr](u.2.2) in
+     let v_r : tree<rl,rls,rt> = pack < u.1, < rhs_r, lhs_r > > as tree<rl,rls,rt>[rlr, rll, rtr, rtl] in
+        write(root, v_r)
+\end{lstlisting}
+
+\begin{lstlisting}[label={lst:histogram},caption={Histogram with Reductions and Atomic Coherence}]
+type hist = < int, int >
+
+task hist_tree1[rl,rls,rt,rh] ( root: tree<rl,rls,rt>@rt, hptr: hist@rh, 
+                                t : int ), reads(rt,rls,rh), writes(rh), atomic(rh) : int =
+    if isnull(root) then 0 else
+    let v : tree<rl,rls,rt> = read(root) in
+    unpack v as u : tree<rl,rls,t>[rll, rlr, rtl, rtr] in
+    let lval : list<rl> = read(u.1) in
+    let hval : hist = read(hptr) in
+    let hval2 : hist = if lval.1 #$<$# t then < hval.1 + 1, hval.2 >
+                                     else < hval.1, hval.2 + 1 > in
+    let _ : hist@rh = write(hptr, hval2) in
+        1 + hist_tree1[rl,rll,rtl,rh](u.2.1, hptr, t) + hist_tree1[rl,rlr,rtr,rh](u.2.2, hptr, t)
+
+task inc_bucket ( hval: hist, b: int ) : hist =
+    if b #$<$# 2 then < hval.1 + 1, hval.2 > else < hval.1, hval.2 + 1 >
+
+task hist_tree2[rl,rls,rt,rh] ( root: tree<rl,rls,rt>@rt, hptr: hist@rh, 
+                                t : int ), reads(rt,rls), reduces(inc_bucket, rh) : int =
+    if isnull(root) then 0 else
+    let v : tree<rl,rls,rt> = read(root) in
+    unpack v as u : tree<rl,rls,t>[rll, rlr, rtl, rtr] in
+    let lval : list<rl> = read(u.1) in
+    let _ : hist@rh = reduce(inc_bucket, hptr, if lval.1 #$<$# t then 1 else 2) in
+        1 + hist_tree1[rl,rll,rtl,rh](u.2.1, hptr, t) + hist_tree1[rl,rlr,rtr,rh](u.2.2, hptr, t)
+\end{lstlisting}

### 2012-07-04 doc/popl2013/langdef.tex
reworded soundness theorem statement, changes to nondeterminism in operational semantics

@@
 
-A number of the rules are worth discussing in detail.
+There are a number of rules that appear to be nondeterministic, allowing multiple choices for
+valid evaluation.  In reality, only the memory allocation case (i.e. new $T@r$) has a truly
+nondeterministic choice (i.e. which currently-unused memory location of the right type to 
+use).  A particular memory allocation algorithm is likely to be deterministic, but the
+design of such algorithms is unrelated to the subject at hand, and by abstracting the location 
+choice, we show that our results hold for any correct memory allocation algorithm.  (In fact,
+our criteria are even weaker than that.  Our soundness result even holds if the location returned
+by ``new $T@r$'' is one that is already in use, as long as the heap typing of that location is
+correct.)
 
-Because we would like to keep the heap typing $H$ constant, the semantics for the expression
-new $T@r$ allow the choice of any memory location that falls within the right physical region
-and has the appropriate heap type.  The soundness of the type system does not depend on that
-location being one that isn't already in use, although most implementations will want to 
-add that additional guarantee.
+The remaining cases of apparent nondeterminism are the result of how we handle the semantics of
+concurrent execution.  An operational semantics that explicitly tracks the state of all 
+concurrently executing operations would allow precise specification of the result of each 
+evaluation step, but would be extremely cumbersome.  Instead, we identify the cases in which
+nonlocal effects (i.e. those that result from the concurrent evaluation of other expressions)
+can impact the local evaluation, and make a ``guess'' of what value would result from an 
+operation.  The check that the guess was consistent with the state of the system is deferred until
+the previously nonlocal effects become local (i.e. contained within the evaluation of an enclosing
+expression tree).  If the guess is not consistent, then the attempted proof of the subexpression's
+behavior is flawed and must be reattempted.
 
-A similar choice appears in the semantics of the read operation.  If the location being read
-from is within the clobber set $C$, the semantics appear to allow any value to be returned, as
-long as it has the right type.  In reality, the choices are more constrained.  If it's not the
-most recently written/reduced value from the current expression evaluation, it must at least be
-a value written by some concurrent evaluation.  The constraint is applied further up in the 
-expression tree when the result of the function is interleaved with other memory traces.
+In the case of ``read'' operations whose locations are in the clobber set $C$, we allow the local
+evaluation to guess any value as long as it is of the right type.  The consistency of the value
+is eventually checked by the $valid\_interleave$ predicate of some enclosing expression, but it
+may not be the closest enclosing one.  The $valid\_interleave$ test itself allows some (apparent)
+freedom of choice, deferring some of the consistency checks to the corresponding check further up
+the expression tree.  The final case of apparent nondeterminism is in the choice of what 
+additional locations to include in each subregion that results from a {\em partition} operation.
+The expansion of the coloring set is necessary for the success of subsequent memory allocation
+within a subregion, but that allocation might not occur within the {\em partition} expression's
+body.  Although the correctness of these partitioning guesses cannot be fully verified until the
+entire execution is known, when an inconsistency is encountered (i.e. there isn't a location of
+the right type in the right region for a ``new $T@r$'' expression) it is possible to ``patch'' the 
+proof (by adding a previously unnamed location to the appropriate expanded colorings and giving
+it the right type in $H$) rather than having to start the proof over from the beginning.
 
@@
 \begin{array}{l}
-\opsenvx e_1 \mapsto K, E_1 \\
-\rho_i = \{ l : (l, i) \in K \}, \text{ for } 1 \leq i \leq k \\
+\opsenv{e_1}{K}{E_1} \\
+K' = K \cup \{ (l_1, v_1), \ldots, (l_p, v_p) \}, \text{ where } \forall i \in [1,p]. (\forall v'. (l_i, v') \not\in K) \wedge (\forall j \in [1,p]. l_i \not= l_j) \wedge v_i \in [1,k] \\
+\rho_i = \{ l : (l, i) \in K' \}, \text{ for } 1 \leq i \leq k \\
 M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \\
@@
 \begin{thm}
-For any expression $e$, if:
-\begin{center}
-$
-\begin{array}{l} \\
-\typeenv{e}{T} \\
-\opsenv{e}{v}{E} \\
+Let $e$ be a well-typed Legion expression.  Let $M$ be a region mapping, $L$ be local value mapping,
+$H$ be a heap typing, and $S$ be an initial state of the store satisfying the consistency properties:
+$$
+\begin{array}{l}
 M \sim \Omega \\
@@
 \end{array}
-$
-\end{center}
-then:
-\begin{center}
-$
+$$
+If $v$ and $E$ are the value and memory trace resulting from a permissable terminating evaluation of
+$e$, then:
+$$
 \begin{array}{l}
@@
 \end{array}
-$
-\end{center}
+$$
 \end{thm}
 
-Although there are 23 different cases to be proven, many of them are similar, and benefit from
+Due to space constraints, we limit the discussion to the general strategy of the proof.  The full
+proof can be found here\cite{xyz}.  If $v$ and $E$ are the result of a permissible evaluation of
+$e$, then a proof exists of:
+$$\opsenv{e}{v}{E}$$
+The soundness of the result is shown by induction over the structure of this proof.  For each of 
+the 23 different forms a Legion expression can have, we assume the soundness of the subexpressions'
+evaluations (provided we can show the consistency of the mappings and store used by that
+subexpression) and prove that the result of the expression's evaluation is sound.  Although there
+are a multitude of cases to be proven, many of them are similar, and benefit from
 the use of the following lemmas:
@@
 
-The complete soundness proof can be found here\cite{xyz} - for space reasons we just discuss
-the general strategy here.
+The proofs of these lemmas can also be found here\cite{xyz}.  The proof of the theorem is organized
+into six pieces, one for each of the three consistency properties needed for subexpressions and one
+for each of the three soundness properties we wish to show.
 
@@
 
-\item Only three expressions have subexpressions that modify $M$ or $\Omega$.  For the
+\item Only three expressions have subexpressions that modify $M$ or $\Omega$ - those that don't
+trivially satisfy the region mapping consistency requirement.  For the
 {\em partition} expression, the assumed consistency of the coloring can be shown to preserve

### 2012-07-04 doc/popl2013/related.tex

\section{Related Work}
\label{sec:related}
Regions have been used for many purposes in the type system literature.  We
cover the ones most closely related to the Legion type system.

Deterministic Parallel Java (DPJ) is the only other region system that we are
aware of that uses regions for extracting parallelism from a 
program to improve performance \cite{Bocchino09}.  There are several
differences between DPJ and Legion.  First, regions in Legion are first-class 
which allows them to be created, deleted, partitioned, packed, and unpacked 
dynamically.  This allows programmers to dynamically determine partitioning
schemes and allocation strategies based on program data at runtime unlike DPJ
where partition schemes must be decided at compile-time.  Despite the dynamic nature
of Legion where regions can be packed and unpacked, we still maintain disjointness
information through constraints on region relationships.  Legion also allows 
programmers to create multiple partitions of the same region to give different 
views onto the same data which is not possible in DPJ.  DPJ only supports the 
equivalent to Legion's exclusive and atomic coherence modes \cite{Bocchino11} 
while Legion provides safe execution even in simultaneous environments.  The 
implementation of DPJ only runs on shared memory machines, while Legion's runtime
uses its knowledge of dynamic regions to run on distributed heterogenous machines.

% Unclear to me what subtleties need to be pointed out here and explained
% to properly differentiate our work
There have been many region systems that use regions for memory management either
to prove safety properties or for performance.  Tofte and Talpin introduced the
first region type system for memory management\cite{Tofte94}.  Aiken et. al. 
improved this system to decouple region allocation from lexical scoping\cite{Aiken95}.
Henglein et. al. introduced a type system for reasoning about memory management
in imperative object-oriented languages \cite{Henglein01}.  Walker et. al. add
capablities to regions to give provable safety guarantees about how regions
can be manipulated \cite{Walker00}.

In addition to region languages with static type systems, there have been several
dynamic region languages.  Cyclone used both a static type system and dynamic
region checks to enforce memory safety properties of C programs\cite{Grossman02}.
Gay and Aiken introduced RC which reference counts regions dynamically and uses
a static type system to reason about effeciently garbage collecting regions\cite{Gay01}.

% I'm also not sure how much detail to go into here.  DPJ spends a lot of time
% on these papers, but I'm not sure I understand all the details of these papers.
% DPJ also cites Lu06 here, but I'm not sure if we have to
Type and effect systems have also been used to reason about regions.  FX presented
the original type and effect system on regions \cite{Lucassen88}, but was restricted 
to using a finite number of regions and was incapable of describing nested data
structures.  There have been many type and effect systems for ownership types
\cite{Boyapati03} including ones that leverage nested regions for describing
relationships \cite{Clarke02,Cameron07}.  However, ownership type and effect systems
are primarily used for reasoning about determinism in object oriented languages and
don't capture the range of disjointness properties that can be specified in Legion.
% Do we need to enumerate what these disjointness properties are (DPJ does)
Type and effect systems have also been used in the context of parallelism to discover
deadlocks and race conditions \cite{Boyapati02,Abadi06,Jacobs08}, but do not present 
any mechanism for discovering parallelism.

Separation Logic is another mechanism that can be used for reasoning about disjointness
of program data \cite{Reynolds02}. %Still not sure what we want to put here.
% DPJ sites additional separation logic papers but they didn't seem very similar.
% Let me know if you think we should include them as well.

Chapel\cite{Chamberlain:Chapel} and X10 \cite{X1005} are similar parallel programming
models to Legion.  Chapel's domains and X10's places provide the programmer with a 
mechanism for expressing locality similar to regions in Legion.  However, domains
and places are not used in any way for independence analysis for discovering parallelism.
Jade is a parallel programming mode that relies on programmer annotations to describe
data disjointness\cite{Rinard98}.  Jade then leverages this disjointness information
to implicitly discover parallelism similar to Legion.

### 2012-07-04 doc/popl2013/related.tex
Adding some more related work

@@
 Separation Logic is another mechanism that can be used for reasoning about disjointness
-of program data \cite{Reynolds02}. %Still not sure what we want to put here.
+of program data \cite{Reynolds02}.  Concurrent separation logic\cite{Brookes04} has been 
+used both to parallelize sequential programs\cite{Raza09,Gotsman07} and to provide 
+a mechanism for reasoning about independence\cite{Hayman06}.
+While we adopted some notation from separation logic, we ultimately chose to use a 
+type and effect system as our primary formalism because separation logic cannot easily 
+describe the interleaving of effects to non-disjoint regions of memory.
+%Still not sure what we want to put here.
 % DPJ sites additional separation logic papers but they didn't seem very similar.
@@
 
-Chapel\cite{Chamberlain:Chapel} and X10 \cite{X1005} are similar parallel programming
+Chapel \cite{Chamberlain:Chapel} and X10 \cite{X1005} are similar parallel programming
 models to Legion.  Chapel's domains and X10's places provide the programmer with a 

### 2012-07-05 doc/popl2013/langdef.tex
checkpoint

@@
 Figure~\ref{fig:opsemfns} gives the definitions of {\em apply} and {\em mark\_coherence}.  For
-the {\em valid\_interleave} predicate, we state some of the necessary conditions here, but leave 
-the construction of actual valid interleavings until a later section.  Informally, a interleaving
-of two or more memory traces must include all the operations from each trace (and no additional
-operations), with each trace's operations remaining in order.  A valid interleaving must further
-satisfy the requirement that the value read from the store must yield the most recently 
-stored/reduced value unless that location is part of the clobber set $C$.
+the {\em valid\_interleave} predicate, we leave the precise definition to a later section.  For this
+section, we need only one of the necessary conditions of a valid interleaving, that a valid
+interleaving is at least an interleaving - it contains all the memory operations from each of the
+constituent traces and no more, and that each trace's operations remain in order in the interleaved
+trace.
 \end{itemize}
@@
 freedom of choice, deferring some of the consistency checks to the corresponding check further up
-the expression tree.  The final case of apparent nondeterminism is in the choice of what 
-additional locations to include in each subregion that results from a {\em partition} operation.
-The expansion of the coloring set is necessary for the success of subsequent memory allocation
-within a subregion, but that allocation might not occur within the {\em partition} expression's
+the expression tree.  The final case of apparent nondeterminism is in the choice of what initial
+locations to include in the result of a ``newcolor $r$'' expression.  The inclusion of locations
+beyond those which are explicitly requested by the application is necessary for the success of
+subsequent memory allocation within a subregion created from a partitioning using the coloring, but
+that allocation might not occur within the {\em partition} expression's
 body.  Although the correctness of these partitioning guesses cannot be fully verified until the
@@
 \end{cases} \\
-any\_interleave(E', [], \ldots, []) & = & true \\
-any\_interleave([ op(l, c, v) ] \dplus E', E_1, \ldots, [ op(l, c, v) ] \dplus E_i, \ldots, E_n) & = & any\_interleave(E', E_1, \ldots, E_i, \ldots, E_n) \\
+any\_interleave([], [], \ldots, []) & = & true \\
+any\_interleave([ \epsilon ] \dplus E', E_1, \ldots, [ \epsilon ] \dplus E_i, \ldots, E_n) & = & any\_interleave(E', E_1, \ldots, E_i, \ldots, E_n) \\
 valid\_interleave(S, C, E', \oton{E}) & \Rightarrow & any\_interleave(E', \oton{E}) \\
-valid\_interleave(S, C, E' \dplus [ read(l, c, v) ] \dplus E'', \oton{E}) & \Rightarrow & l \in C \vee S'(l) = v, \text{ where } S' = apply(S, E')
 \end{array}
@@
 {\opsenvx \text{reduce}(id, e_1, e_2) \mapsto l, E' \dplus [ reduce_{id}(l, excl, v) ]}
-\framebox{$\opsenvx \text{newcolor }r \mapsto \emptyset, []$}
+\finfrule{
+\begin{array}{l@{}l}
+K = \{ (l_1, iv_1), & \ldots, (l_p, v_p) \}, \text{ where } \\
+& (\forall i \in [1,p]. l_i \in M(r)) \wedge \\
+& (\forall i,j \in [1,p]. l_i \not= l_j)
+\end{array}
+}
+{\opsenvx \text{newcolor }r \mapsto K, []}
 \finfrule
@@
 \opsenv{e_1}{K}{E_1} \\
-K' = K \cup \{ (l_1, v_1), \ldots, (l_p, v_p) \}, \text{ where } \forall i \in [1,p]. (\forall v'. (l_i, v') \not\in K) \wedge (\forall j \in [1,p]. l_i \not= l_j) \wedge v_i \in [1,k] \\
-\rho_i = \{ l : (l, i) \in K' \}, \text{ for } 1 \leq i \leq k \\
+\rho_i = \{ l : (l, i) \in K \}, \text{ for } 1 \leq i \leq k \\
 M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \\
@@
 
-\subsection{Noninterference}
+\subsection{Coherence}
+
+The soundness proof above guarantees that the evaluation of an expression produces a value of the
+right type, and do so within the regions of memory for which privileges are held, but for the
+Legion programming model to be useful, we need to also guarantee that the evaluation of an expression
+produces the (or at least, a) correct value.  The key to doing this is to place additional 
+constraints on what it means to be a valid interleaving of two or more memory traces.  In
+addition to a valid interleaving being an interleaving of the individual traces, we will 
+require that it be:
+\begin{itemize}
+\item {\em coherent} - each memory read operation should see the effect of all writes and reductions 
+that occur earlier in the interleaved trace to the same location, and
+\item {\em sequentially equivalent} - the value read by each memory read operation and the result of applying the interleaved trace to the original store should be the same as if each constituent trace
+was applied to the store in a valid {\em program order}.
+\end{itemize}
+These requirements are relaxed in some cases due to either a non-empty clobber set (representing
+operations that may occur concurrently as part of evaluating other functions) or due to coherence
+modes other than {\em excl} on operations with the traces (representing operations that may 
+occur concurrently as part of evaluating other parts of the same function).
+
+Figure~\ref{fig:validinterleave} gives the precise definition of the {\em valid\_interleave}
+predicate, but there are a lot of moving parts, so description in ``plain'' English may help.
+Consider a set of constituent traces $\oton{E}$.  Let the {\em exclusive location set} $L_{excl}$ be
+the set of all locations that appear in at least one operation with coherence mode $excl$ in at least
+one constituent trace, excepting any locations in the clobber set $C$.  Let the {\em atomic location
+set} $L_{atomic}$ similarly be the set of locations that appear in at least one operation with 
+coherence mode $atomic$, but no operations with coherence mode $excl$, again excepting any location
+in the clobber set.  A memory trace $E'$ is considered a valid interleaving of $\oton{E}$ if all of
+the following are true:
+\begin{enumerate}
+\item $E'$ is an interleaving of $\oton{E}$.
+\item For every read in $E'$ with a location in $L_{excl}$, the value read is what would be in the
+ store at that location if all the previous operations of $E'$ were applied to the original
+store $S$.
+\item For every read in $E_i$ with a location in $L_{excl}$, the value read is what would be in the
+store at that location if all preceding constituent
+traces (i.e. $E_1, \ldots, E_{i-1}$) were applied to the store $S$, followed by the operations from
+$E_i$ that precede the read.
+\item For every location $l$ in $L_{excl}$, the value
+at that location after the trace $E'$ is applied to $S$ must match the value that would be in
+the store if each of $E_i$ were successively applied to $S$.
+\item For every read in $E'$ with a location in $L_{atomic}$ for which a write operation to the same
+location appears earlier in $E'$, the value read is what would be in the store if all the previous
+operations of $E'$ were applied to the initial store $S$.
+\item There is some permutation $\oton{\pi}$ of the integers $1,\ldots,n$ such that, for every read
+in $E_{\pi_i}$ with a location in $L_{atomic}$, the value read is what would be in the store at that
+location if the traces $E_{\pi_1}, \ldots E_{\pi_{i-1}}$ were applied to $S$, following by the operations 
+from $E_{\pi_i}$ that precede the read.
+\item Using the same permutation $\oton{\pi}$, for every location in $L_{atomic}$, the value at that
+location after the trace $E'$ is applied to $S$ must match the value that would be in the store
+if each of $E_{\pi_i}$ were successively applied to $S$.
+\end{enumerate}
+
+Flexibility in interleaving operations is the key to achieving parallel execution, but
+hopefully the reader will agree that the exact criteria for creating a valid interleaving is
+much too expensive to implement at runtime, so we now consider whether there are simpler ways to
+generate valid interleavings, and under what conditions those approaches may be used.
+
+The first option available to us is to serialize the constituent traces into the original
+program order.  Although each constituent trace can be assumed to be coherent, that coherence is
+in the context of a clobber set that may differ from the one in which the interleaving must occur.
+As discussed earlier, this difference can result in cases where no valid interleaving exists.
+However, we can show that as long as any valid interleavings exist, the serialization of the
+constituent traces will be one of them.
+
+To prove this result, we will use the following lemma:
+
+\begin{lem}
+\label{lemma:applychain}
+For any initial store $S$ and memory traces $E_1$ and $E_2$,
+$$apply(S, E_1 \dplus E_2) = apply(apply(S, E_1), E_2).$$
+\end{lem}
+
+The proof of this follows directly from the definition of $apply$.
+
+\begin{thm}
+Let $S_0$ be an initial store, $C$ be a clobber set and $\oton{E}$ be memory traces to be
+interleaved.  Let $E' = E_1 \dplus \ldots \dplus E_n$.  If there exists any $E''$ that is a
+valid interleaving of $\oton{E}$, then $E'$ is also valid interleaving of $\oton{E}$.
+\end{thm}
+
+\begin{proof}
+To prove $valid\_interleave(S_0, C, E', \oton{E})$, we need to show the three conditions hold:
+\begin{enumerate}
+\item $any\_interleave(E', \oton{E})$ is intuitively obvious, and can be proven by induction over 
+the length of $E_1$, followed by $E_2$, and so on.
+\item $coherent(S_0, C, E')$ is shown by considering each constituent trace's subsequence in $E'$
+in order, using Lemma~\ref{lemma:applychain} along with the coherence of each $E_i$.
+\item $seq\_equiv(S_0, C, E', \oton{E})$ is trivial, as each leading subsequence of $E'$ corresponds
+to the appropriate pieces of $\otok{i}{E}$.
+\item $coherent(S_0, C, E')$ can be shown from the sequential equivalence of $E''$.  The value 
+returned by each read in $E''$ has been shown to be equal to the value in the store that results
+from the application of all previous constituent traces and the operations before the read in its
+own trace.  This is exactly the sequence of preceding operations of the corresponding read in $E'$,
+and Lemma~\ref{lemma:applychain} allows us to trainsform the chain of $apply$ operations into a 
+single $apply$ operation on the whole sequence, satisfying the coherence condition for the read
+in $E'$.  This property applies for all reads in $E''$, and since both $E'$ and $E''$ are
+interleavings of the same traces, they have exactly the same reads, so we have shown all the reads
+in $E'$ to be coherent.
+\end{enumerate}
+\end{proof}
+
+Now that we have a way to construct a valid interleaving, we can look at how to modify that
+interleaving to get interleavings that are still valid, but allow parallel execution.  To do this,
+we consider the criteria under which we may swap the order of two adjacent memory operations in an
+existing valid interleaving.
+\begin{itemize}
+\item First, the two operations must have come from different constituent
+traces, or the swap would violate the $any\_interleave$ criterion.
+\item Second, if one is a read of
+a location (that is not in the clobber set), the other may not be a write or reduce to the same
+location, as changing the order would almost certainly change the value that is read.
+\item Third, if one is a write of a location, the other may not be a write or reduce to the same
+location, as the store that would result from applying those operations in the opposite order would
+show a difference in that location.
+\item Finally, if both operations are reductions to the same location, they may be swapped if they
+use the same reduction function, as reduction functions are assumed to be commutative.  (We 
+follow the example of other work (i.e. DPJ) and leave proofs of commutativity to the programmer,
+possibly with the aid of other tools.)  However, if they use different reduction functions (by
+name - we make no attempt to compare the operation of functions), commutative cannot be 
+guaranteed and swapping the two operations might result in a different resulting store.
+\end{itemize}
+
+To summarize this, we define a {\em noninterference} operator on two effects that have been
+annotated with which constituent trace they came from as follows:
+$$
+\begin{array}{l@{ }l}
+op_1^{E_1}(l_1, c_1,& v_1)\ \#\ op_2^{E_2}(l_2, c_2, v_2) \Leftrightarrow E_1 \not= E_2\ \wedge \\
+& \big( l_1 \not= l_2 \vee (op_1 = read \wedge op_2 = read)\ \vee \\
+& (op_1 = reduce_{id_1} \wedge op_2 = reduce_{id_2} \wedge id_1 = id_2) \big)
+\end{array}
+$$
+
+\begin{lem}
+Let $S$ be an initial store, $C$ be a clobber set, $\oton{E}, E'_a, E'_b$ be memory traces, and
+$\epsilon_1, \epsilon_2$ be two annotated memory operations such that $\epsilon_1 \# \epsilon_2$.
+Then,
+$$valid\_interleave(S, C, E'_a \dplus \epsilon_1 \dplus \epsilon_2 \dplus E'_b, \oton{E})$$
+if and only if
+$$valid\_interleave(S, C, E'_a \dplus \epsilon_2 \dplus \epsilon_1 \dplus E'_b, \oton{E}).$$
+\end{lem}
+
+\begin{proof}
+As before, we consider the three conditions needed to be a valid interleaving.  We describe the
+implication in the ``forward'' direction, but the arguments are all symmetric.
+\begin{enumerate}
+\item If $E'_a \dplus \epsilon_1 \dplus \epsilon_2 \dplus E'_b$ is an interleaving of $\oton{E}$,
+the proof must first ``pop off'' all of $E'_a$, leaving $\epsilon_1 \dplus \epsilon_2 \dplus E'_b$ for
+the interleaved subtrace.  The next two steps must pop $\epsilon_1$ and $\epsilon_2$, leaving $E'_b$.
+They cannot be in the same subtrace because $\epsilon_1 \# \epsilon_2$, so we could also pop
+$\epsilon_2$ and then $\epsilon_1$, resulting in a proof that $E'_a \dplus \epsilon_2 \dplus \epsilon_1 \dplus E'_b$ is also an interleaving of the constituent traces.
+\item For coherence, we must show the value read (if any) by $\epsilon_1$ does not change, the value
+read (if any) by $\epsilon_2$ does not change, and that any read in $E'_b$ sees no change.  With the
+help of Lemma~\ref{lemma:applychain}, we can eliminate the common $E'_a$ from all sequences by
+using $S' = apply(S, E'_a)$.
+Similarly to show that $apply(S', \epsilon_1 \dplus \epsilon_2 \dplus E) = apply(S', \epsilon_2 \dplus \epsilon_1 \dplus E)$ for any $E$, we need only to show that it holds
+for $E = []$.  The proof is now reduced to showing the following:
+\begin{enumerate}
+\item $\epsilon_1 = read(l, c, v) \Rightarrow S'(l) = apply(S',\epsilon_2)(l)$ - since $\epsilon_1$
+is a read, either $\epsilon_2$ must be a read (in which case $apply(S, \epsilon_2) = S'$) or must be
+to a different location, so the result of applying just it to $S'$ cannot change the value in 
+location $l$.
+\item $\epsilon_2 = read(l, c, v) \Rightarrow apply(S', \epsilon_1)(l) = S'(l)$ - this is the same
+as above, with $\epsilon_1$ and $\epsilon_2$ switched.
+\item $apply(S', \epsilon_1 \dplus \epsilon_2) = apply(S', \epsilon_2 \dplus \epsilon_1)$ - we must
+show that the two store operations $apply(\bullet, \epsilon_1)$ and $apply(\bullet, \epsilon_2)$
+commute.  Modifications to two different locations commute, as do two reads (which make no change
+to the store).  Finally, two reductions using the same function to the same location also commute.
+\end{enumerate}
+\item The proof that sequential equivalence is maintained comes down to showing the final store
+is identical after the swap of the two operations, and this is merely a special case of the third
+piece of the coherence proof.
+\end{enumerate}
+\end{proof}
+
+First, we would like a fallback approach that will always result in a valid interleaving, even if
+we lack information about the constituent traces (or perhaps lack the time to analyze it).  The
 
@@
 
-coherent\_interleave(S_1, S_2, C, c, [], [], []) = & true \\
-coherent\_interleave(S_1, S_2, C, c, [ op(l, c', v) ] \dplus E', [ op(l, c, v) ] \dplus E_1, E_2) = & (op = read \wedge l \not\in C \wedge c = c' \Rightarrow S_1(l) = v) \wedge \\
-& coherent\_interleave(apply(S_1, op(l, c', v)), S_2, C, c, E', E_1, E_2) \\
+coherent\_interleave(S, C, c, E', [], E_1, [], E_2, []) = & \forall l. l \in C \vee apply(S, E')(l) = apply(S, E_1 \dplus E_2)(l) \\
+coherent\_interleave(S, C, c, E'_a, \epsilon \dplus E'_b, E_{1a}, \epsilon \dplus E_{1b}, E_{2a}, E_{2b}) = & (\epsilon = read(l, c, v) \wedge l \not\in C \Rightarrow apply(S,E'_a)(l) = v \wedge apply(S,E_{1a})(l) = v) \wedge \\
+& coherent\_interleave(S, C, c,  E'_a \dplus \epsilon, E'_b, E_{1a} \dplus \epsilon, E_{1b}, E_{2a}, E_{2b}) \\
 coherent\_interleave(S_1, S_2, C, c, [ op(l, c', v) ] \dplus E', E_1, [ op(l, c, v) ] \dplus E_2) = & (op = read \wedge l \not\in C \wedge c = c' \Rightarrow S_2(l) = v) \wedge \\

### 2012-07-05 doc/popl2013/intro.tex

@@
   literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
-  morekeywords={task,rr,int,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,newcolor,atomic,simultaneous},
+  morekeywords={task,rr,int,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,color,newcolor,atomic,simultaneous},
   deletekeywords={float,head,min,max}

### 2012-07-05 doc/popl2013/evaluation.tex
Working on the evaluation section

\section{Evaluation}
\label{sec:evaluation}

To determine the benefits of the Legion type system we evaluate
the ability of the type system on three criteria:
\begin{itemize}
\item Expressivity: can real applications be expressed in the type system (Section \ref{subsec:expressivity})
\item Overhead: can the type system reduce the checking overhead for ensuring safety (Section \ref{subsec:overhead})
\item Scalability: does our system leverage hierarchical scheduling to scale (Section \ref{subsec:scalability})
\end{itemize}
Our evaluation uses a prototype implementation of the Legion langauge and programming model.
The prototype consists of two components: an implementation of a type checker
for the langauge introduced in Section \ref{subsec:langdef} and a C++ runtime library
for executing programs written in the Legion programming model\cite{Legion12}.  All experiments
are conducted on the Keeneland supercomputer\cite{Keeneland}.  Each node of the Keeneland
machine consists of two Xeon 5660 processors, three Tesla M2090 GPUs, and 24 GB of DRAM.  Nodes
are connected by a QDR Infiniband interconnect.

\subsection{Expressivity}
\label{subsec:expressivity}
We evaluate Legion on three real-world applications.  To qualitatively gauge the 
expressivity of the Legion type system, we introduce each of these applications in
the context of which features of the Legion type system are used in their implementations.

\subsubsection{Circuit Simulation}
\label{subsec:circuit}
The circuit application simulates the wires of a large integrated circuit using an RLC
model.  The computation consists of three phases for each time step: compute the current in each wire using
an iterative model, updated the charge in each node, and compute the voltage of each node.
The primary data structure in the circuit graph is a large irregular graph.  To execute
this computation in parallel the graph is dynamically partitioned into pieces to be
performed in parallel.  Our implementation creates separate regions for the wires and
nodes in the graph.  The wire and node regions are recusrively partitioned up into
subregions for pieces of the graph.  The node region is also partitioned an additional way to 
describe the sets of ghost nodes required for each piece providing multiple views onto the
same data.  The information about each piece of the graph is stored
in a region relationship that remembers the disjointness information for each piece
from other pieces.  Launching the tasks for the three phases of the computation make 
use of read, write, and reduce privileges as well as exclusive and simultaneous coherence.

\subsubsection{Fluid Simulation}
\label{subsec:fluid}
Our fluid application is based on the fluidanimate benchmark for the PARSEC benchmark
suite\cite{bienia11benchmarking}.  Fluid simulates the flow of an incompressible fluid
using particles that move between a regular grid of cells.  To perform operations in 
parallel the cells are partitioned into grids that can be executed in parallel.  Unlike
the Circuit application, the Fluid first creates regions and partitions them before
allocating cells in Regions.  Another difference between Fluid and Circuit is that Fluid
maintains separate regions for ghost cells rather using multiple views of
the regions containing shared data.  Region relationships are again used to remember
the disjointness of information for the regions of each grid.

\subsubsection{Adaptive Mesh Refinement}
\label{subsec:amr}
The third application is an adaptive mesh refinement (AMR) benchmark based on the third heat flow
example from the Berkeley Labs BoxLib project\cite{BoxLib}.  The algorithm solves the two
dimensional heat equation on a grid of cells using three levels of refinement with subrefinements
randomly placed on the surface.  For each time step in the application cells at the boundary of
a refinement are interpolated from cells at a coarser level of refinement, energy fluxes between
cells are computed, energy is transferred, and cells at a coarser level of refinement are restricted
to the values of refined cells.  The AMR application uses separate regions for every level of
refinement.  The regions at each level are partitioned several ways to provide multiple views of
the cells.  One partitioning separates cells into pieces that can be updated in
parallel.  Additional partitions are created for viewing data from coarser and finer levels of
the simulation.  Two types of region relationships are created: one for describing the pieces of
each level of refinement, and another for describing the relationship between pieces at different
levels of refinement.  These region relationships capture both intra- and inter-level disjointness
information.  The dynamic nature of AMR requires that both regions can be created, partitioned,
and destroyed at runtime.  The many ways in which cells are used also mandates that we multiple
partitions of regions can be created.

These applications illustrate the expressivity of the Legion type system.  Legion is capable of 
expressing applications with both regular (Fluid,AMR) and irregular (Circuit) pointer data structres.
Being able to dynamically create, partition, and destroy regions at runtime is crucial to Legion's
ability to handle applications that make runtime decisions about how to partition data (AMR).  Legion
is able to capture both allocate-then-partition (Circuit) and partition-then-allocate (Fluid) ways 
of loading data.  Having multiple partitions of regions is necessary for describing the many ways 
that data can be accessed (Circuit,AMR).  All types of privileges and coherence are necessary for 
expressing the various applications presented in this paper (Circuit,Fluid,AMR,Histogram).

\subsection{Checking Overhead}
\label{subsec:overhead}

\subsection{Scalability}
\label{subsec:scalability}

### 2012-07-05 doc/popl2013/langdef.tex
Adding some labels to the langdef
checkpoint

@@
 
-modify $M$
-\begin{table*}
-\centering
-{\small
-\begin{tabular}{l|p{5cm}|p{5cm}|p{5cm}}
-{\bf Rule} & 
-\multicolumn{1}{|c|}{$\mathbf{v \sim_H M \llbracket T \rrbracket}$} &
-\multicolumn{1}{|c|}{$\mathbf{E \sim H}$} &
-\multicolumn{1}{|c|}{$\mathbf{E : \Phi}$} \\
-\hline
-BoolConst & & & \\
-IntConst & & & \\
-TupleBuild & & & \\
-TupleField & & & \\
-LocalVar & & & \\
-NullPtr & & & \\
-NewPtr & & & \\
-IsNull & & & \\
-UpRegion & & & \\
-DownRegion & & & \\
-Read & & & \\
-Write & & & \\
-Reduce & & & \\
-Newcolor & & & \\
-Color & & & \\
-IntMath & & & \\
-IntCmp & & & \\
-Let & & & \\
-IfElse & & & \\
-TaskCall & & & \\
-Partition & & & \\
-Pack & & & \\
-Unpack & & &
-\end{tabular}}
-\caption{Proof pieces}
-\end{table*}
+%% \begin{table*}
+%% \centering
+%% {\small
+%% \begin{tabular}{l|p{5cm}|p{5cm}|p{5cm}}
+%% {\bf Rule} & 
+%% \multicolumn{1}{|c|}{$\mathbf{v \sim_H M \llbracket T \rrbracket}$} &
+%% \multicolumn{1}{|c|}{$\mathbf{E \sim H}$} &
+%% \multicolumn{1}{|c|}{$\mathbf{E : \Phi}$} \\
+%% \hline
+%% BoolConst & & & \\
+%% IntConst & & & \\
+%% TupleBuild & & & \\
+%% TupleField & & & \\
+%% LocalVar & & & \\
+%% NullPtr & & & \\
+%% NewPtr & & & \\
+%% IsNull & & & \\
+%% UpRegion & & & \\
+%% DownRegion & & & \\
+%% Read & & & \\
+%% Write & & & \\
+%% Reduce & & & \\
+%% Newcolor & & & \\
+%% Color & & & \\
+%% IntMath & & & \\
+%% IntCmp & & & \\
+%% Let & & & \\
+%% IfElse & & & \\
+%% TaskCall & & & \\
+%% Partition & & & \\
+%% Pack & & & \\
+%% Unpack & & &
+%% \end{tabular}}
+%% \caption{Proof pieces}
+%% \end{table*}
 
@@
 \begin{thm}
+\label{thm:sequential}
 Let $S_0$ be an initial store, $C$ be a clobber set and $\oton{E}$ be memory traces to be
@@
 \begin{lem}
+\label{lem:nonintswap}
 Let $S$ be an initial store, $C$ be a clobber set, $\oton{E}, E'_a, E'_b$ be memory traces, and
@@
 
-First, we would like a fallback approach that will always result in a valid interleaving, even if
-we lack information about the constituent traces (or perhaps lack the time to analyze it).  The
-
-The soundness proof above depended, in part, on the properties of a valid interleaving of two
-memory traces.  We now consider algorithms that create interleavings that satisfy those
-conditions.
-
-The simplest way to achieve a valid interleaving is to concatenate the traces in order.  This
-will always result in a valid interleaving, but implies a serialization of the evaluation of
-subexpressions.
+We now extend the noninterference operator to entire memory traces, providing a way to say that
+no operation in a trace interferes with any operation in other:
+$$E_1 \# E_2 \Leftrightarrow \displaystyle\bigwedge_{\epsilon_1 \text{ in } E_1, \epsilon_2 \text{ in } E_2} \epsilon_1 \# \epsilon_2$$
+and observe that this property permits arbitrary ordering (i.e. parallel execution) of the 
+corresponding subexpressions.
+
+\begin{lem}
+\label{lem:noninteffects}
+Let $S$ be an initial store, $C$ be a clobber set, $\oton{E}$ be memory traces such that $E_i \# E_j$
+for every $1 \leq i < j \leq n$.  Then any trace $E'$ satisfying:
+$$any\_interleave(E', \oton{E})$$
+is a valid interleaving.
+\end{lem}
+
+\begin{proof}
+We will use Theorem~\ref{thm:sequential} and Lemma~\ref{lem:nonintswap} to make repeated swaps
+of adjacent operations to transform $E'$ into $E_1 \dplus \ldots \dplus E_n$.  The symmetry of
+Lemma~\ref{lem:nonintswap} is then used to show that the validity of $E_1 \dplus \ldots \dplus E_n$
+under Theorem~\ref{thm:sequential} implies the validity of $E'$.
+
+The swapping algorithm as follows: Associate with each operation in $E'$ the index of that operation
+in $E_1 \dplus \ldots \dplus E_n$.  Any consectutive pair of operations $\epsilon_1$ and $\epsilon_2$
+for which $index(\epsilon_1) > index(\epsilon_2)$ is a {\em misordered pair}.  For any such pair,
+$\epsilon_1$ and $\epsilon_2$ will be from different constituent traces (if they were from the same
+trace, their misordering would mean that $E'$ was not an interleaving of $\oton{E}$).  Every pair of
+traces is noninterfering, so $\epsilon_1 \# \epsilon_2$ and they can be swapped while preserving
+validity of the trace.  At most $n^2/2$ swaps are needed to eliminate all misordered pairs.
+\end{proof}
+
+Our next step is to extend noninterference to the privileges used in the type system.  We will make
+use of the region mapping $M$ to translate the logical regions named in privileges into sets of
+locations (i.e. physical regions), and define noninterference between individual privileges as
+follows:
+\begin{center}
+\begin{tabular}{llc}
+\multicolumn{1}{c}{$\phi_1$} & \multicolumn{1}{c}{$\phi_2$} & $\phi_1 \#_M \phi_2$? \\
+\multirow{3}{*}{$reads(r_1)$} & $reads(r_2)$ & true \\
+& $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
+& $reduces(id_2,r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
+\cline{1-1}
+\multirow{3}{*}{$writes(r_1)$} & $reads(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
+& $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
+& $reduces(id_2,r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
+\cline{1-1}
+\multirow{3}{*}{$reduces(id_1r_1)$} & $reads(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
+& $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
+& $reduces(id_2,r_2)$ & $\begin{array}{c}id_1 = id_2 \vee \\ M(r_1) \cap M(r_2) = \emptyset\end{array}$ \\
+\end{tabular}
+\end{center}
+and noninterference between sets of privileges in the same way as memory traces:
+$$\Phi_1 \#_M \Phi_2 \Leftrightarrow \displaystyle\bigwedge_{\phi_1 \in \Phi_1, \phi_2 \in \Phi_2} \phi_1 \#_M \phi_2$$
+
+\begin{lem}
+\label{lem:nonintpriv}
+Let $M$ be a region mapping and $E_1$ and $E_2$ two memory traces such that $M \vdash E_1 : \Phi_1$ and $M \vdash E_2 : \Phi_2$.
+If $\Phi_1$ and $\Phi_2$ are noninterfering under $M$, then $E_1$ and $E_2$ must be noninterfering.
+\end{lem}
+
+\begin{proof}
+The proof is by contradiction.
+Assume $E_1 \cancel\# E_2$.
+Then there must be some $\epsilon_1$ in $E_1$ and some $\epsilon_2$ in $E_2$ such that $\epsilon_1 \cancel\# \epsilon_2$.
+This requires that
+$\epsilon_1$ and $\epsilon_2$ be operations to the same location $l$ and not both be reads or both
+reductions using the same function name.
+Assume $\epsilon_1 = read(l, c_1, v_1)$ and $\epsilon_2 = write(l, c_2, v_2)$.
+The definition of $M \vdash E_1 : \Phi_1$ tells us that there must be some $r_1$ satisfying $l \in M(r_1) \wedge reads(r_1) \in \Phi_1$.
+Similarly, there must be some $r_2$ satisfying $l \in M(r_2) \wedge writes(r_2) \in \Phi_2$.
+Letting $\phi_1 = reads(r_1)$ and $\phi_2 = writes(r_2)$, we have $M(r_1) \cap M(r_2) \not= \emptyset$ (it contains at least $l$), so $\phi_1 \cancel\#_M \phi_2$ and $\Phi_1 \cancel\#_M \Phi_2$.
+There are seven other cases to consider (using $id_1 \not= id_2$ for the reduce-reduce case), but all yield the same result.
+\end{proof}
+
+We are now able to connect the dots and get to the key result that allows the Legion runtime to 
+safely run subtasks in parallel.
+
+\begin{thm}
+Let $\oton{e}$ be well-typed Legion expressions, each with its own privileges $\Phi_i$.
+Let $M$ be a region mapping, $L$ be local value mapping,
+$H$ be a heap typing, and $S$ be an initial state of the store satisfying the consistency properties:
+$$
+\begin{array}{l}
+M \sim \Omega \\
+L \sim_H \Gamma \\
+S \sim H
+\end{array}
+$$
+If $\Phi_i \#_M \Phi_j$ for $1 \leq i < j \leq n$, then any parallel execution of expressions
+$\oton{e}$ will result in a valid interleaving of memory operations.
+\end{thm}
+
+The proof follows directly from Lemma~\ref{lem:nonintpriv} and Lemma~\ref{lem:noninteffects}. \\
+
+Noninterference of effects is something that can only be determined after
+evaluation of an expression is completed, and only at great expense (see STM).
+In contrast, the noninterference of privileges can be efficiently determined at
+runtime.  Even though the privileges themselves are static, the
+mapping $M$ can be dynamically determined, allowing noninterference to be 
+judged in many more cases than a purely static analysis can achieve.
+
+Note that this result holds even if the clobber set $C$ is non-empty.  This allows 
+two ``locally independent'' subtasks to run in parallel even if one or both
+of them is interacting (in a programmer-permitted way) with a third subtask.
+
+TODO: talk about atomic case
 
@@
 
-\framebox{$\forall S, C, E_1, E_2. valid\_interleave(S, C, E_1 \dplus E_2, E_1, E_2)$}
-
-Next step: For noninterfering traces $E_1 \# E_2$, show any interleaving is valid.
-
-Third step: For interactions only on atomic-coherence locations, serializability is sufficient.
-
-For a given expression $e$, we can consider $v^o$ to be a ``sequential'' result and $E^o$ to be the corresponding ``sequential'' effects if: \\
-
-\framebox{$M, L, H, S, \emptyset \vdash e \mapsto v^o, E^o$} \\
-
-(i.e. if there are no manipulations of the store other than what's done by
-the evaluation of the expression $e$.)  We now wish to consider what parallel
-evaluations of this expression with other expression evaluations will still
-yield our ``sequential'' result. \\
-
-Two effects are considered to be noninterfering, written $E_1 \# E_2$ if for
-every $\epsilon_1$ in $E_1$ and every $\epsilon_2$ in $E_2$,
-\begin{enumerate}
-\item $\epsilon_1$ and $\epsilon_2$ are to different memory locations, or
-\item $\epsilon_1 = read(l, v_1)$ and $\epsilon_2 = read(l, v_2)$, or
-\item $\epsilon_1 = reduce_{id}(l, v_1)$ and $\epsilon_2 = reduce_{id}(l, v_2)$
-\end{enumerate}
-
-We define nonintereference (under a given mapping $M$) of static privileges
-$\Phi_1$ and $\Phi_2$ similarly.  We use the following table for determining
-noninterference of two individual priviledges $\phi_1$ and $\phi_2$:
-
-\begin{tabular}{ccccc}
-\multicolumn{2}{l}{$\phi_1 \#_M \phi_2 ?$} & \multicolumn{3}{c}{$\phi_2$} \\
-\multicolumn{2}{l}{} & $reads(r_2)$ & $writes(r_2)$ & $reduces(id_2, r_2)$ \\
-\multirow{3}{*}{$\phi_1$} & $reads(r_1)$ & true & $M(r_1) \cap M(r_2) = \emptyset$ & $M(r_1) \cap M(r_2) = \emptyset$\\
-& $writes(r_1)$ & $M(r_1) \cap M(r_2) = \emptyset$ & $M(r_1) \cap M(r_2) = \emptyset$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
-& $reduces(id_1, r_1)$ & $M(r_1) \cap M(r_2) = \emptyset$ & $M(r_1) \cap M(r_2) = \emptyset$ & $\begin{array}{l}id_1 = id_2 \vee \\ M(r_1) \cap M(r_2) = \emptyset\end{array}$
-\end{tabular}
-
-We can then define: \\
-
-\framebox{$\Phi_1 \#_M \Phi_2 \Leftrightarrow \displaystyle\bigwedge_{\begin{array}{c}\phi_1 \in \Phi_1 \\ \phi_2 \in \Phi_2 \end{array}} \phi_1 \#_M \phi_2$} \\
-
-From the soundness of effects, result, we can show that for two expressions
-$e_1$ and $e_2$, if their privileges are noninterfering, any potential memory
-traces from their evaluation must also be noninterfering.  Formally:
-
-\finfrule
-{\begin{array}{l}
-\typeenvx[P=\Phi_1] e_1 : T_1 \\
-\typeenvx[P=\Phi_2] e_2 : T_2 \\
-\opsenvx e_1 \mapsto v_1, E_1 \\
-\opsenvx e_2 \mapsto v_2, E_2 \\
-\Phi_1 \#_M \Phi_2
-\end{array}}
-{E_1 \# E_2}
-
-(TODO: Generalize this further and allow $M_1 \not= M_2$?) \\
-
-Note that this result holds even if $C \not= \emptyset$.  This allows 
-two ``locally independent'' subtasks to run in parallel even if one or both
-of them is interacting (in a programmer-permitted way) with a third subtask.
-
-Noninterference of effects is something that can only be determined after
-evaluation of an expression is completed, and only at great expense (see STM).
-In contrast, the noninterference of privileges can be efficiently determined at
-runtime.  Even though the privileges themselves are static, the
-mapping $M$ can be dynamically determined, allowing noninterference to be 
-judged in many more cases than a purely static analysis can achieve.
-
-(TODO: Talk about coherence modes and how the runtime orders subtasks.  Maybe
-describe it in terms of a ``canonical'' memory trace and then allow
-reorderings?)
-
 \subsection{Hierarchical Scheduling}

### 2012-07-05 doc/popl2013/evaluation.tex
Updating the evaluation section for popl

@@
 To determine the benefits of the Legion type system we evaluate
-the ability of the type system on three criteria:
+it on three criteria:
 \begin{itemize}
-\item Expressivity: can real applications be expressed in the type system (Section \ref{subsec:expressivity})
-\item Overhead: can the type system reduce the checking overhead for ensuring safety (Section \ref{subsec:overhead})
-\item Scalability: does our system leverage hierarchical scheduling to scale (Section \ref{subsec:scalability})
+\item Expressivity: can real applications be expressed (Section \ref{subsec:expressivity})
+\item Overhead: can the it reduce checking overhead (Section \ref{subsec:overhead})
+\item Scalability: can it leverage hierarchical scheduling to scale (Section \ref{subsec:scalability})
 \end{itemize}
@@
 levels of refinement.  These region relationships capture both intra- and inter-level disjointness
-information.  The dynamic nature of AMR requires that both regions can be created, partitioned,
+information.  The dynamic nature of AMR requires that regions can be created, partitioned,
 and destroyed at runtime.  The many ways in which cells are used also mandates that we multiple
@@
 that data can be accessed (Circuit,AMR).  All types of privileges and coherence are necessary for 
-expressing the various applications presented in this paper (Circuit,Fluid,AMR,Histogram).
+expressing the various applications presented in this paper (Circuit,Fluid,AMR,Histogram).  In
+all applications region relationships allow for disjointness information to be conveyed despite
+regions being a first-class feature.
 
@@
 \label{subsec:overhead}
+Our initial implementation of Legion consisted only of a C++ library for Legion 
+programs\cite{Legion12} which contained no checking of region memory accesses.  In the process
+of implementing applications in Legion we frequently encountered memory corruption due to
+illegal region accesses because of bugs and the lack of checking.  Many times these corruptions 
+occurred on remote nodes or on GPUs.  Finding these bugs required either multiple {\tt gdb} windows
+for observing execution on multiple nodes simultaneously or using {\tt printf} for debugging on
+GPUs.  To alleviate this problem we added a feature to our runtime system that dynamically checked region 
+accesses for both CPUs and GPUs to guarantee safety.  However, in many cases this 
+added considerable overhead.
+
+To avoid the overhead of dynamic checks for all memory accesses we implemented a type checker of the
+language in Section \ref{subsec:langdef}.  Using this checker we were able to implement programs
+in this language and type check them.  The soundness proof for the type system allowed us to safely
+elide all dynamic checks of region accesses.
+
+%\begin{figure}
+%\subfigure[48 Piece Data Set]
+%{
+%\includegraphics[scale=0.4]{figs/circuit_48_popl.pdf}
+%\label{fig:ckt48}
+%}
+%\subfigure[96 Piece Data Set]
+%{
+%\includegraphics[scale=0.4]{figs/circuit_96_popl.pdf}
+%\label{fig:ckt96}
+%}
+%\caption{Checking overhead of the Circuit simulation.\label{fig:ckt_overhead}}
+%\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.40]{figs/circuit_96_popl.pdf}
+\end{center}
+\caption{Overhead for the Circuit simulation with 96 total pieces.\label{fig:ckt_overhead}}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.40]{figs/fluid_19200_popl.pdf}
+\end{center}
+\caption{Overhead for the Fluid simulation with 19200 cells.\label{fig:fluid_overhead}}
+\end{figure}
+
+Figures \ref{fig:ckt_overhead}, \ref{fig:fluid_overhead}, and \ref{fig:amr_overhead} show 
+the total time spent by all CPUs and GPUs in each phase of the application.  In 
+each figure the problem size stays the same as the number of processors are increased
+(strong scaling).  Figure \ref{fig:amr_overhead} includes multiple problem sizes to show
+how overhead is affected by changing problem size (weak scaling).  For cases where there
+is an existing implementation to compare against we have included a dotted line indicating
+baseline performance.  In a few cases (e.g. Figure \ref{fig:amr4096}) checking overhead is the difference
+between better and worse performance than the baseline.  For an explanation of overall
+performance relative to the baseline implementations please refer to \cite{Legion12}.
+
+\begin{figure}
+\begin{center}
+\subfigure[4096 cells per dimension]
+{
+\includegraphics[scale=0.40]{figs/amr_4096_popl.pdf}
+\label{fig:amr4096}
+}
+\subfigure[8192 cells per dimension]
+{
+\includegraphics[scale=0.40]{figs/amr_8192_popl.pdf}
+\label{fig:amr8192}
+}
+\subfigure[16384 cells per dimension]
+{
+\includegraphics[scale=0.40]{figs/amr_16384_popl.pdf}
+\label{fig:amr16384}
+}
+\end{center}
+\caption{Overhead for the AMR application across problem sizes.\label{fig:amr_overhead}}
+\end{figure}
+
+In addition to total processor overhead, we also measured performance gain in terms of wall-clock
+time.  Note that since most region accesses occur in leaf tasks they often parallelize
+very well.  Wall-clock performance gains from eliding memory checks ranged from 1-10\%, 1-15\%, 
+and 2-71\% for Circuit, Fluid, and AMR respectively.  Performance gains for AMR were larger than
+the other applications because AMR was already memory bound and the additional checks intensified
+memory pressure.  We also note that for GPU kernels, checking required up to 8 additional registers.  
+While the GPU kernels in Circuit where not bound by 
+available on-chip memory, kernels that are would be susceptible to extreme performance 
+degradation due to the extra registers required for checking.  Finally, we measured the overhead
+of the dynamic checks associated with unpack operations but found them to be negligible relative
+to the overhead of region access checks.
 

### 2012-07-05 apps/fluid3d/popl_plots.py
Updating the script for popl plots for fluid

@@ -116,6 +116,7 @@ def read_experiments(ps, directory):
 def print_experiments(exprs):
     for e in exprs:
         print "Nodes "+str(e.nodes)+" Time "+str(e.time)
+    print ""
 
 def make_plot(fig, orig_exprs, check_exprs, prob_size):
     assert len(orig_exprs) == len(check_exprs)
@@ -178,6 +179,11 @@ def make_plot(fig, orig_exprs, check_exprs, prob_size):
     plt.xticks(offset+ind+width/2.0,cpus16)
     plt.ylabel('Processor Time (seconds)')
     plt.legend(loc=2)
+    # Print the overheads
+    print "Overheads for "+str(prob_size)+" cells"
+    for o,c in zip_list:
+        print "Processors "+str(o.nodes*o.cpus)+" "+str(c.time/o.time)
+    print ""
     
 
 def make_plots(show = True, save = True, out_dir="figs/"):
@@ -186,9 +192,8 @@ def make_plots(show = True, save = True, out_dir="figs/"):
         check_exprs = read_experiments(ps,'check_results/')
         print "Results for problem size "+str(ps)+" for original experiments"
         print_experiments(orig_exprs)
-        print "\nResults for problem size "+str(ps)+" for checking experiments"
+        print "Results for problem size "+str(ps)+" for checking experiments"
         print_experiments(check_exprs)
-        print "\n"
         # new figure
         fig = plt.figure()
         make_plot(fig,orig_exprs,check_exprs,ps)

### 2012-07-05 doc/popl2013/legion-types.tex
Creating a separate tex file for the introduction
Fixing a number in the abstract and adding an evaluation section
Fixing some typos in the abstract

@@
 %\title{Scalable and Safe Parallelism using Logical Regions}
-\authorinfo{}{}{}
+\authorinfo{Sean Treichler \and Michael Bauer \and Alex Aiken}{Stanford University}{\{sjt,mebauer,aiken\}@cs.stanford.edu}
+%\authorinfo{Sean Treichler}{Stanford Unviersity}{sjt@cs.stanford.edu}
+%\authorinfo{Michael Bauer}{Stanford University}{mebauer@cs.stanford.edu}
+%\authorinfo{Alex Aiken}{Stanford University}{aiken@cs.stanford.edu}
 \maketitle
@@
 We prove a soundness result for our type system and use it to show
-that all memory accesses in Legion are safe.  Futhermore, we prove that
-the soundess of our type system enables a scalable hierarchical
+that all memory accesses in Legion are safe.  Furthermore, we prove that
+the soundness of our type system enables a scalable hierarchical
 scheduling algorithm for Legion programs.  We implement a type checker
-for Legion programs and improve performance up to 85\% by eliding
-unncessary memory checks.  We demonstrate an implementation of the Legion 
+for Legion programs and improve performance up to 71\% by eliding
+unnecessary memory checks.  We demonstrate an implementation of the Legion 
 runtime based on the hierarchical scheduling algorithm scales for
@@
 
-\section{Introduction}
-\label{sect:intro}
-Introduction goes here.
+\input{intro}
 
@@
 \input{langdef}
+
+\input{evaluation}
+
+\input{related}
 

### 2012-07-05 doc/popl2013/evaluation.tex
Messing with figure sizes in the evaluation
Working some more on the evaluation

@@
 \item Expressivity: can real applications be expressed (Section \ref{subsec:expressivity})
-\item Overhead: can the it reduce checking overhead (Section \ref{subsec:overhead})
+\item Overhead: can it reduce checking overhead (Section \ref{subsec:overhead})
 \item Scalability: can it leverage hierarchical scheduling to scale (Section \ref{subsec:scalability})
@@
 of implementing applications in Legion we frequently encountered memory corruption due to
-illegal region accesses because of bugs and the lack of checking.  Many times these corruptions 
-occurred on remote nodes or on GPUs.  Finding these bugs required either multiple {\tt gdb} windows
+illegal region accesses caused by bugs and a lack of checking.  Many times these corruptions 
+occurred on remote nodes or on GPUs.  Finding these bugs required either multiple {\tt gdb} sessions 
 for observing execution on multiple nodes simultaneously or using {\tt printf} for debugging on
-GPUs.  To alleviate this problem we added a feature to our runtime system that dynamically checked region 
-accesses for both CPUs and GPUs to guarantee safety.  However, in many cases this 
-added considerable overhead.
+GPUs.  To alleviate this problem we added a feature to our runtime system that dynamically checked all region 
+accesses for both CPUs and GPUs to guarantee safety.  However, this feature
+added considerable runtime overhead.
 
 To avoid the overhead of dynamic checks for all memory accesses we implemented a type checker of the
-language in Section \ref{subsec:langdef}.  Using this checker we were able to implement programs
-in this language and type check them.  The soundness proof for the type system allowed us to safely
-elide all dynamic checks of region accesses.
+language in Section \ref{subsec:langdef}.  Using the type checker we were able to statically verify
+the correctness of all region accesses and elide all the dynamic checks.
 
@@
 \begin{center}
-\includegraphics[scale=0.40]{figs/circuit_96_popl.pdf}
+\includegraphics[scale=0.35]{figs/circuit_96_popl.pdf}
 \end{center}
+\vspace{-2mm}
 \caption{Overhead for the Circuit simulation with 96 total pieces.\label{fig:ckt_overhead}}
+\vspace{-6mm}
 \end{figure}
@@
 \begin{center}
-\includegraphics[scale=0.40]{figs/fluid_19200_popl.pdf}
+\includegraphics[scale=0.35]{figs/fluid_19200_popl.pdf}
 \end{center}
+\vspace{-2mm}
 \caption{Overhead for the Fluid simulation with 19200 cells.\label{fig:fluid_overhead}}
+\vspace{-6mm}
 \end{figure}
@@
 Figures \ref{fig:ckt_overhead}, \ref{fig:fluid_overhead}, and \ref{fig:amr_overhead} show 
-the total time spent by all CPUs and GPUs in each phase of the application.  In 
+the total time spent by all CPUs and GPUs in each phase of the application.  The topmost
+component of each bar shows the overhead of the dynamic checks.  In 
 each figure the problem size stays the same as the number of processors are increased
@@
 {
-\includegraphics[scale=0.40]{figs/amr_4096_popl.pdf}
+\includegraphics[scale=0.35]{figs/amr_4096_popl.pdf}
 \label{fig:amr4096}
@@
 {
-\includegraphics[scale=0.40]{figs/amr_8192_popl.pdf}
+\includegraphics[scale=0.35]{figs/amr_8192_popl.pdf}
 \label{fig:amr8192}
@@
 {
-\includegraphics[scale=0.40]{figs/amr_16384_popl.pdf}
+\includegraphics[scale=0.35]{figs/amr_16384_popl.pdf}
 \label{fig:amr16384}
@@
 \end{center}
-\caption{Overhead for the AMR application across problem sizes.\label{fig:amr_overhead}}
+\vspace{-2mm}
+\caption{Overhead for the AMR application for three different problem sizes.\label{fig:amr_overhead}}
+\vspace{-6mm}
 \end{figure}

### 2012-07-05 doc/popl2013/intro.tex
intro

@@
 \label{sec:intro}
-Introduction goes here.
+
+In the past decade machine architecture, particularly at the high
+performance end of the spectrum, has undergone a revolution.  The
+latest supercomputers are now complex hierarchies of many different
+kinds of computing technologies: networks of nodes at the top level,
+multiple chips per node, mutiple cores within a chip, and, most
+recently, multiple accelerators (usually GPUs) per node, which are
+themselves can be further decomposed into several levels of machine
+hierarchy.  In this paper, we present the operational and static
+semantics of Legion, a programming language targeted at providing an
+appropriate level of abstraction for programming such machines, one
+that is both sufficiently high-level to be portable while still
+exposing aspects that are crucial to performance. A very brief summary
+of Legion is that data is organized in a hierarchy of {\em regions}
+and subregions and computation is organized in a hierarchy of {\em
+tasks} and subtasks operating on regions.  Regions and tasks interact
+through a static system of {\em region permissions} specifying the
+operations a task may perform on a region argument (e.g., read or
+write, among others) and novel {\em region coherence} annotations that
+expresses whether other tasks may perform operations concurrently on
+the region (e.g., {\em exclusive} access for task $t$ guarantees no
+other task's updates to a region are visible to $t$).  We prove the
+soundness of the permissions and coherence system and use these two
+results to prove a third result: if two tasks $t_1$ and $t_2$ are {\em
+dependent} (roughly, must be executed in program order to guarantee
+the correct answer), then the unique ancestors of $t_1$ and $t_2$ in
+the task hierarchy that are {\em siblings} (have the same parent task)
+are also dependent.  This theoremis the basis for distributed
+scheduling in a Legion implementation, which is crucial for
+performance; any centralized scheduler would be a severe bottleneck
+because of the huge latencies involved in moving data and computation
+in the target class of machines.
+
+Legion programs organize data into {\em logical regions}.  Logical
+regions (which we usually refer to simply as {\em regions}) express
+locality, which is just data that will be used together and therefore
+should be colocated in the machine.  Listing~\ref{lst:linkedlist}
+gives an example that defines a region containing a linked list of integers and a task
+that sums the elements of the list.  As is standard in region systems, we use
+$T @ r$ for the type of a pointer to a value of type $T$ in region $r$.
+As is also
+
+Two of the basic properties programmers need to express in writing
+programs for highly parallel, distributed machines are the {\em
+locality} and {\em independence} of data. Locality identifies data
+that will be used together and so should be placed in the same part of
+the machine; independence identifies data sets that are disjoint and
+so can be operated on in parallel.  Legion programs use {\em logical
+regions} to express both.  A logical region (which we will usually
+refer to as simply a {\em region} for brevity) is a
+programmer-specified collection of data; we use the notation of
+separating conjunction $r_1 \ast r_2$ for disjoint regions $r_1$ and
+$r_2$.
+
+%Programming
+%these machines is currently done by using a heterogeneous collection
+%of software technologies targeted at the individual level; a typical
+%melange is MPI for the network, OpenMP for the node, and CUDA for the GPUs.
 

### 2012-07-05 doc/popl2013/legion-types.tex
text complete for langdef.tex
intro

@@
 \begin{abstract}
-Legion is a programming model that introduces 
-{\em logical regions} as a mechanism for targeting large heterogeneous machines 
-with complex memory hierarchies.  Logical regions are a first-class 
-feature of Legion that allow programmers to specify independence 
-and locality properties of program data dynamically.  We present
-a type and effect system for reasoning about logical regions in Legion.  
-We prove a soundness result for our type system and use it to show
-that all memory accesses in Legion are safe.  Furthermore, we prove that
-the soundness of our type system enables a scalable hierarchical
-scheduling algorithm for Legion programs.  We implement a type checker
-for Legion programs and improve performance up to 71\% by eliding
-unnecessary memory checks.  We demonstrate an implementation of the Legion 
-runtime based on the hierarchical scheduling algorithm scales for
-three real-world applications.
+Legion is a programming model that introduces logical regions as a mechanism for targeting large heterogeneous machines with complex memory hierarchies.
+Logical regions are first-class values in Legion, allowing programmers to specify independence and locality properties of program data dynamically.
+We present a type system with permission and coherence properties for reasoning about logical regions.
+We prove the soundness of the  type system and use it to show that all memory accesses in Legion are safe.
+Furthermore, we show that type soundness directly enables a scalable, hierarchical scheduling algorithm for Legion programs.
+We implement a type checker for Legion programs and improve performance up to 71% by eliding memory checks the type system proves safe.
+We also demonstrate that an implementation of the Legion runtime based on the hierarchical scheduling algorithm scales well for three real-world applications.
 

### 2012-07-05 doc/popl2013/evaluation.tex

@@
 
-To determine the benefits of the Legion type system we evaluate
-it on three criteria:
+We evaluate the Legion type system on three criteria:
 \begin{itemize}
-\item Expressivity: can real applications be expressed (Section \ref{subsec:expressivity})
-\item Overhead: can it reduce checking overhead (Section \ref{subsec:overhead})
-\item Scalability: can it leverage hierarchical scheduling to scale (Section \ref{subsec:scalability})
+\item Expressivity: can real applications be encoded (Section \ref{subsec:expressivity})
+\item Overhead: can it reduce checking costs (Section \ref{subsec:overhead})
+\item Scalability: can it leverage hierarchical scheduling (Section \ref{subsec:scalability})
 \end{itemize}
-Our evaluation uses a prototype implementation of the Legion langauge and programming model.
+Our evaluation uses a prototype implementation of the Legion language and programming model.
 The prototype consists of two components: an implementation of a type checker
-for the langauge introduced in Section \ref{subsec:langdef} and a C++ runtime library
+for the language introduced in Section \ref{subsec:langdef} and a C++ runtime library
 for executing programs written in the Legion programming model\cite{Legion12}.  All experiments
@@
 We evaluate Legion on three real-world applications.  To qualitatively gauge the 
-expressivity of the Legion type system, we introduce each of these applications in
-the context of which features of the Legion type system are used in their implementations.
+expressivity of the Legion type system, we introduce each of these applications
+by describing which features of the Legion type system are used in their implementations.
 
@@
 \label{subsec:circuit}
-The circuit application simulates the wires of a large integrated circuit using an RLC
+The Circuit application simulates the wires of a large integrated circuit using an RLC
 model.  The computation consists of three phases for each time step: compute the current in each wire using
 an iterative model, updated the charge in each node, and compute the voltage of each node.
-The primary data structure in the circuit graph is a large irregular graph.  To execute
-this computation in parallel the graph is dynamically partitioned into pieces to be
-performed in parallel.  Our implementation creates separate regions for the wires and
-nodes in the graph.  The wire and node regions are recusrively partitioned up into
-subregions for pieces of the graph.  The node region is also partitioned an additional way to 
-describe the sets of ghost nodes required for each piece providing multiple views onto the
-same data.  The information about each piece of the graph is stored
+The primary data structure in the Circuit application is a large irregular graph.  To perform 
+this computation in parallel the graph is dynamically partitioned into pieces.  
+Our implementation creates separate regions for the wires and
+nodes in the graph.  The wire and node regions are recursively partitioned into
+subregions for pieces of the graph.  The node region is partitioned an additional way to 
+describe the sets of ghost nodes required for each piece.  
+The information about each piece of the graph is stored
 in a region relationship that remembers the disjointness information for each piece
-from other pieces.  Launching the tasks for the three phases of the computation make 
-use of read, write, and reduce privileges as well as exclusive and simultaneous coherence.
+from other pieces.  The tasks for the three phases of the computation use different combinations of
+read, write, and reduce privileges as well as exclusive and simultaneous coherence.
 
@@
 \label{subsec:fluid}
-Our fluid application is based on the fluidanimate benchmark for the PARSEC benchmark
+Our Fluid application is based on the fluidanimate benchmark from the PARSEC benchmark
 suite\cite{bienia11benchmarking}.  Fluid simulates the flow of an incompressible fluid
 using particles that move between a regular grid of cells.  To perform operations in 
-parallel the cells are partitioned into grids that can be executed in parallel.  Unlike
-the Circuit application, the Fluid first creates regions and partitions them before
-allocating cells in Regions.  Another difference between Fluid and Circuit is that Fluid
-maintains separate regions for ghost cells rather using multiple views of
+parallel the cells are partitioned into grids.  Unlike
+the Circuit application, the Fluid application first creates regions and partitions them before
+allocating cells in regions.  Another difference between Fluid and Circuit is that Fluid
+maintains separate regions for ghost cells rather using multiple partitions of
 the regions containing shared data.  Region relationships are again used to remember
@@
 example from the Berkeley Labs BoxLib project\cite{BoxLib}.  The algorithm solves the two
-dimensional heat equation on a grid of cells using three levels of refinement with subrefinements
-randomly placed on the surface.  For each time step in the application cells at the boundary of
+dimensional heat diffusion equation on a grid of cells using three levels of refinement with subrefinements
+randomly placed on the surface.  For each time step in the application, cells at the boundary of
 a refinement are interpolated from cells at a coarser level of refinement, energy fluxes between
@@
 parallel.  Additional partitions are created for viewing data from coarser and finer levels of
-the simulation.  Two types of region relationships are created: one for describing the pieces of
+the simulation.  Two types of region relationships are created: one for describing the pieces at 
 each level of refinement, and another for describing the relationship between pieces at different
@@
 information.  The dynamic nature of AMR requires that regions can be created, partitioned,
-and destroyed at runtime.  The many ways in which cells are used also mandates that we multiple
+and destroyed at runtime.  The many ways in which cells are used also mandates that multiple
 partitions of regions can be created.
@@
 These applications illustrate the expressivity of the Legion type system.  Legion is capable of 
-expressing applications with both regular (Fluid,AMR) and irregular (Circuit) pointer data structres.
+expressing applications with both regular (Fluid,AMR) and irregular (Circuit) pointer data structures.
 Being able to dynamically create, partition, and destroy regions at runtime is crucial to Legion's
-ability to handle applications that make runtime decisions about how to partition data (AMR).  Legion
+ability to handle applications that make runtime decisions about data organization(AMR).  Legion
 is able to capture both allocate-then-partition (Circuit) and partition-then-allocate (Fluid) ways 
@@
 expressing the various applications presented in this paper (Circuit,Fluid,AMR,Histogram).  In
-all applications region relationships allow for disjointness information to be conveyed despite
-regions being a first-class feature.
+all applications, region relationships convey disjointness information despite
+regions being a first-class feature that cannot be fully tracked statically.
 
@@
 of implementing applications in Legion we frequently encountered memory corruption due to
-illegal region accesses caused by bugs and a lack of checking.  Many times these corruptions 
+illegal region accesses caused by a confluence of bugs and lack of checking.  Many times these corruptions 
 occurred on remote nodes or on GPUs.  Finding these bugs required either multiple {\tt gdb} sessions 
-for observing execution on multiple nodes simultaneously or using {\tt printf} for debugging on
-GPUs.  To alleviate this problem we added a feature to our runtime system that dynamically checked all region 
-accesses for both CPUs and GPUs to guarantee safety.  However, this feature
+for observing execution on different nodes simultaneously or using {\tt printf} for debugging on
+GPUs.  To mitigate this problem and guarantee safety we added a feature to our runtime system that 
+dynamically checked all region accesses for both CPUs and GPUs.  However, this feature
 added considerable runtime overhead.
 
-To avoid the overhead of dynamic checks for all memory accesses we implemented a type checker of the
-language in Section \ref{subsec:langdef}.  Using the type checker we were able to statically verify
-the correctness of all region accesses and elide all the dynamic checks.
+To avoid the overhead of dynamic checks for all memory accesses we implemented a type checker for the
+language in Section \ref{subsec:langdef}.  We wrote all of our applications in this language and
+used our type checker to verify them.  After proving that all the applications type checked we were
+able to safely elide all the dynamic region access checks.
 
@@
 the total time spent by all CPUs and GPUs in each phase of the application.  The topmost
-component of each bar shows the overhead of the dynamic checks.  In 
+component of each bar shows the overhead added by the dynamic checks.  In 
 each figure the problem size stays the same as the number of processors are increased
@@
 is an existing implementation to compare against we have included a dotted line indicating
-baseline performance.  In a few cases (e.g. Figure \ref{fig:amr4096}) checking overhead is the difference
+baseline performance.  In a few cases (Figures \ref{fig:fluid_overhead} and 
+\ref{fig:amr4096}) checking overhead is the difference
 between better and worse performance than the baseline.  For an explanation of overall
@@
 
-In addition to total processor overhead, we also measured performance gain in terms of wall-clock
-time.  Note that since most region accesses occur in leaf tasks they often parallelize
-very well.  Wall-clock performance gains from eliding memory checks ranged from 1-10\%, 1-15\%, 
+In addition to total processor overhead, we also measured performance gain from eliding checks in 
+terms of wall-clock time.  Since most region accesses occur in leaf tasks the checks parallelize 
+well.  Wall-clock performance gains from eliding memory checks ranged from 1-10\%, 1-15\%, 
 and 2-71\% for Circuit, Fluid, and AMR respectively.  Performance gains for AMR were larger than
 the other applications because AMR was already memory bound and the additional checks intensified
-memory pressure.  We also note that for GPU kernels, checking required up to 8 additional registers.  
-While the GPU kernels in Circuit where not bound by 
+memory pressure.  For the GPU kernels in the Circuit application checking required up to 8 additional 
+registers per thread.  While the GPU kernels in Circuit where not bound by 
 available on-chip memory, kernels that are would be susceptible to extreme performance 
-degradation due to the extra registers required for checking.  Finally, we measured the overhead
+degradation due to the extra registers required for checking.  We also measured the overhead
 of the dynamic checks associated with unpack operations but found them to be negligible relative

### 2012-07-05 doc/popl2013/langdef.tex

@@
 \newcommand{\rrbracket}{]\!]}
+\makeatletter
+\newcommand{\nonint}[1][]{\#\@ifmtarg{#1}{}{_{\!\!#1}}}
+\makeatother
 
@@
 $\Phi$ & ::= & $\{ \oton{\phi} \}$ & privileges \\
-$\phi$ & ::= & reads$(r)$ $\;\;\;\mid\;\;\;$ writes$(r)$ $\;\;\;\mid\;\;\;$ reduces$(id,r)$ & \\
+$\phi$ & ::= & reads$(r)$ $\;\;\;\mid\;\;\;$ writes$(r)$ $\;\;\;\mid\;\;\;$ $\text{reduces}_{id}(r)$ & \\
 \\
@@
 \section{Legion Core Language}
+\label{sec:legioncore}
 
 \subsection{Language Definition}
+\label{subsec:langdef}
 
@@
 \typeenvx e_2 : T_2 \\
-\forall i, reduces(id, r_i) \in \Phi^*
+\forall i, reduces_{id}(r_i) \in \Phi^*
 \end{array}}
@@
 r_i \leq r_j \in \Omega^* \wedge writes(r_j) \in \Phi^* \Rightarrow writes(r_i) \in \Phi^* \\
-r_i \leq r_j \in \Omega^* \wedge reduces(id, r_j) \in \Phi^* \Rightarrow reduces(id, r_i) \in \Phi^* \\
+r_i \leq r_j \in \Omega^* \wedge reduces_{id}(r_j) \in \Phi^* \Rightarrow reduces_{id}(r_i) \in \Phi^* \\
 reads(r) \in \Phi^* \wedge writes(r) \in \Phi^* \wedge \\
-\hspace{1cm} \Gamma(id) = \forall \otok{k}{r}. (\oton{T}), \Phi', Q' \rightarrow T_r \Rightarrow reduces(id, r) \in \Phi^*
+\hspace{1cm} \Gamma(id) = \forall \otok{k}{r}. (\oton{T}), \Phi', Q' \rightarrow T_r \Rightarrow reduces_{id}(r) \in \Phi^*
 \end{array}$
@@
 & & $(\epsilon = write(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge writes(r) \in \Phi)\ \wedge$ \\
-& & $(\epsilon = reduce_{id}(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reduces(id, r) \in \Phi)$
+& & $(\epsilon = reduce_{id}(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reduces_{id}(r) \in \Phi)$
 \end{tabular}
@@
 \label{lemma:phiclosure}
-$M \vdash E : \Phi^*$ if ant only if $M \vdash E : \Phi$.
+$M \vdash E : \Phi^*$ if and only if $M \vdash E : \Phi$.
 \end{lem}
@@
 
+\begin{figure*}
+$$
+\begin{array}{lll}
+valid\_interleave(S, C, E', \oton{E}) & = & any\_interleave(E', \oton{E}) \wedge \\
+& & coherent(S, L_{excl}(E', C), L_{excl}(E', C), E')\ \wedge \\
+& & seq\_equiv(S, L_{excl}(E', C), L_{excl}(E', C), E', \oton{E})\ \wedge \\
+& & coherent(S, L_{atomic}(E', C), \emptyset, E')\ \wedge \\
+& & \exists \oton{\pi}, (\oton{\pi}) \text{ is a permutation of } (1,\ldots,n)\ \wedge \\
+& & \ \ \ seq\_equiv(S, L_{atomic}(E', C), \emptyset, E', E_{\pi_1}, \ldots, E_{\pi_n}) \\
+coherent(S, L_1, L_2, []) & = & \text{true} \\
+coherent(S, L_1, L_2, \epsilon \dplus E) & = &\
+\begin{cases}
+(l \in L_2 \Rightarrow S(l) = v)\ \wedge \\
+\ \ \ coherent(S, L_1, L_2, E), & \text{if $\epsilon = read(l, c, v)$} \\
+coherent(apply(S, \epsilon), L_1, L_2 \cup \{ l \}, E), & \text{if $\epsilon = write(l, c, v)$ and $l \in L_1$} \\
+coherent(apply(S, \epsilon), L_1, L_2, E), & \text{otherwise}
+\end{cases} \\
+seq\_equiv(S, L_1, L_2, E', \oton{E}) & = & coherent(S, L_1, L_2, E_1 \dplus \ldots \dplus E_2)\ \wedge \\
+& & \forall l \in L_1. apply(S, E')(l) = apply(S, E_1 \dplus \ldots \dplus E_n)(l) \\
+L_{excl}(E, C) & = & \{ l : op(l, excl, v) \text{ in } E  \} \setminus C \\
+L_{atomic}(E, C) & = & \{ l : op(l, atomic, v) \text{ in } E \} \setminus (C \cup L_{excl}(E, C)) \\
+\end{array}
+$$
+\caption{Valid Interleaving Test}
+\label{fig:validinterleave}
+\end{figure*}
+
 Figure~\ref{fig:validinterleave} gives the precise definition of the {\em valid\_interleave}
@@
 coherence mode $atomic$, but no operations with coherence mode $excl$, again excepting any location
-in the clobber set.  A memory trace $E'$ is considered a valid interleaving of $\oton{E}$ if all of
-the following are true:
-\begin{enumerate}
-\item $E'$ is an interleaving of $\oton{E}$.
-\item For every read in $E'$ with a location in $L_{excl}$, the value read is what would be in the
- store at that location if all the previous operations of $E'$ were applied to the original
-store $S$.
-\item For every read in $E_i$ with a location in $L_{excl}$, the value read is what would be in the
-store at that location if all preceding constituent
-traces (i.e. $E_1, \ldots, E_{i-1}$) were applied to the store $S$, followed by the operations from
-$E_i$ that precede the read.
-\item For every location $l$ in $L_{excl}$, the value
-at that location after the trace $E'$ is applied to $S$ must match the value that would be in
-the store if each of $E_i$ were successively applied to $S$.
-\item For every read in $E'$ with a location in $L_{atomic}$ for which a write operation to the same
-location appears earlier in $E'$, the value read is what would be in the store if all the previous
-operations of $E'$ were applied to the initial store $S$.
-\item There is some permutation $\oton{\pi}$ of the integers $1,\ldots,n$ such that, for every read
-in $E_{\pi_i}$ with a location in $L_{atomic}$, the value read is what would be in the store at that
-location if the traces $E_{\pi_1}, \ldots E_{\pi_{i-1}}$ were applied to $S$, following by the operations 
-from $E_{\pi_i}$ that precede the read.
-\item Using the same permutation $\oton{\pi}$, for every location in $L_{atomic}$, the value at that
-location after the trace $E'$ is applied to $S$ must match the value that would be in the store
-if each of $E_{\pi_i}$ were successively applied to $S$.
-\end{enumerate}
+in the clobber set.  The $coherent(S,L_1,L_2,E)$ test checks that all reads of locations in $L_2$ and any read of a location in $L_1$ that comes after a write to the same location return values
+that match what was most recently written into the store.  (The difference betwen $L_1$ and $L_2$
+defers the checking of atomic reads that don't have a preceeding write until any other traces that
+may be ordered before $E$ are included.)  The $seq\_equiv$ test verifies that every read in $E'$ with a location in $L_1$ would also have be coherent if the operations were performed in the original program order and that the store that results from applying $E'$ matches the store that would
+have resulted from applying the constituent traces in program order for all locations in $L_1$.
+For the atomic location set, we allow the constituent traces to be permuted, but all locations
+must be checked with the same permutation.
 
@@
 \begin{array}{l@{ }l}
-op_1^{E_1}(l_1, c_1,& v_1)\ \#\ op_2^{E_2}(l_2, c_2, v_2) \Leftrightarrow E_1 \not= E_2\ \wedge \\
+op_1^{E_1}(l_1, c_1,& v_1)\ \nonint\ op_2^{E_2}(l_2, c_2, v_2) \Leftrightarrow E_1 \not= E_2\ \wedge \\
 & \big( l_1 \not= l_2 \vee (op_1 = read \wedge op_2 = read)\ \vee \\
@@
 Let $S$ be an initial store, $C$ be a clobber set, $\oton{E}, E'_a, E'_b$ be memory traces, and
-$\epsilon_1, \epsilon_2$ be two annotated memory operations such that $\epsilon_1 \# \epsilon_2$.
+$\epsilon_1, \epsilon_2$ be two annotated memory operations such that $\epsilon_1 \nonint \epsilon_2$.
 Then,
@@
 the interleaved subtrace.  The next two steps must pop $\epsilon_1$ and $\epsilon_2$, leaving $E'_b$.
-They cannot be in the same subtrace because $\epsilon_1 \# \epsilon_2$, so we could also pop
+They cannot be in the same subtrace because $\epsilon_1 \nonint \epsilon_2$, so we could also pop
 $\epsilon_2$ and then $\epsilon_1$, resulting in a proof that $E'_a \dplus \epsilon_2 \dplus \epsilon_1 \dplus E'_b$ is also an interleaving of the constituent traces.
@@
 no operation in a trace interferes with any operation in other:
-$$E_1 \# E_2 \Leftrightarrow \displaystyle\bigwedge_{\epsilon_1 \text{ in } E_1, \epsilon_2 \text{ in } E_2} \epsilon_1 \# \epsilon_2$$
+$$E_1 \nonint E_2 \Leftrightarrow \displaystyle\bigwedge_{\epsilon_1 \text{ in } E_1, \epsilon_2 \text{ in } E_2} \epsilon_1 \nonint \epsilon_2$$
 and observe that this property permits arbitrary ordering (i.e. parallel execution) of the 
@@
 \label{lem:noninteffects}
-Let $S$ be an initial store, $C$ be a clobber set, $\oton{E}$ be memory traces such that $E_i \# E_j$
+Let $S$ be an initial store, $C$ be a clobber set, $\oton{E}$ be memory traces such that $E_i \nonint E_j$
 for every $1 \leq i < j \leq n$.  Then any trace $E'$ satisfying:
@@
 trace, their misordering would mean that $E'$ was not an interleaving of $\oton{E}$).  Every pair of
-traces is noninterfering, so $\epsilon_1 \# \epsilon_2$ and they can be swapped while preserving
+traces is noninterfering, so $\epsilon_1 \nonint \epsilon_2$ and they can be swapped while preserving
 validity of the trace.  At most $n^2/2$ swaps are needed to eliminate all misordered pairs.
@@
 follows:
-\begin{center}
-\begin{tabular}{llc}
-\multicolumn{1}{c}{$\phi_1$} & \multicolumn{1}{c}{$\phi_2$} & $\phi_1 \#_M \phi_2$? \\
-\multirow{3}{*}{$reads(r_1)$} & $reads(r_2)$ & true \\
-& $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
-& $reduces(id_2,r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
-\cline{1-1}
-\multirow{3}{*}{$writes(r_1)$} & $reads(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
-& $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
-& $reduces(id_2,r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
-\cline{1-1}
-\multirow{3}{*}{$reduces(id_1r_1)$} & $reads(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
-& $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
-& $reduces(id_2,r_2)$ & $\begin{array}{c}id_1 = id_2 \vee \\ M(r_1) \cap M(r_2) = \emptyset\end{array}$ \\
-\end{tabular}
-\end{center}
+$$
+\begin{array}{l@{ }l}
+priv_1(r_1) \nonint[M] & priv_2(r_2) \Leftrightarrow (M(r_1) \cap M(r_2) = \emptyset)\ \vee \\
+& (priv_1 = reads \wedge priv_2 = reads)\ \vee \\
+& (priv_1 = reduces_{id_1} \wedge priv_2 = reduces_{id_2} \wedge id_1 = id_2)
+\end{array}
+$$
+%% \begin{center}
+%% \begin{tabular}{llc}
+%% \multicolumn{1}{c}{$\phi_1$} & \multicolumn{1}{c}{$\phi_2$} & $\phi_1 \nonint[M] \phi_2$? \\
+%% \multirow{3}{*}{$reads(r_1)$} & $reads(r_2)$ & true \\
+%% & $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
+%% & $reduces_{id_2}(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
+%% \cline{1-1}
+%% \multirow{3}{*}{$writes(r_1)$} & $reads(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
+%% & $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
+%% & $reduces_{id_2}(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
+%% \cline{1-1}
+%% \multirow{3}{*}{$reduces_{id_1}(r_1)$} & $reads(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
+%% & $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
+%% & $reduces_{id_2}(r_2)$ & $\begin{array}{c}id_1 = id_2 \vee \\ M(r_1) \cap M(r_2) = \emptyset\end{array}$ \\
+%% \end{tabular}
+%% \end{center}
 and noninterference between sets of privileges in the same way as memory traces:
-$$\Phi_1 \#_M \Phi_2 \Leftrightarrow \displaystyle\bigwedge_{\phi_1 \in \Phi_1, \phi_2 \in \Phi_2} \phi_1 \#_M \phi_2$$
+$$\Phi_1 \nonint[M] \Phi_2 \Leftrightarrow \displaystyle\bigwedge_{\phi_1 \in \Phi_1, \phi_2 \in \Phi_2} \phi_1 \nonint[M] \phi_2$$
 
@@
 The proof is by contradiction.
-Assume $E_1 \cancel\# E_2$.
-Then there must be some $\epsilon_1$ in $E_1$ and some $\epsilon_2$ in $E_2$ such that $\epsilon_1 \cancel\# \epsilon_2$.
+Assume $E_1 \cancel\nonint E_2$.
+Then there must be some $\epsilon_1$ in $E_1$ and some $\epsilon_2$ in $E_2$ such that $\epsilon_1 \cancel\nonint \epsilon_2$.
 This requires that
@@
 Similarly, there must be some $r_2$ satisfying $l \in M(r_2) \wedge writes(r_2) \in \Phi_2$.
-Letting $\phi_1 = reads(r_1)$ and $\phi_2 = writes(r_2)$, we have $M(r_1) \cap M(r_2) \not= \emptyset$ (it contains at least $l$), so $\phi_1 \cancel\#_M \phi_2$ and $\Phi_1 \cancel\#_M \Phi_2$.
+Letting $\phi_1 = reads(r_1)$ and $\phi_2 = writes(r_2)$, we have $M(r_1) \cap M(r_2) \not= \emptyset$ (it contains at least $l$), so $\phi_1 \cancel{\nonint[M]} \phi_2$ and $\Phi_1 \cancel{\nonint[M]} \Phi_2$.
 There are seven other cases to consider (using $id_1 \not= id_2$ for the reduce-reduce case), but all yield the same result.
@@
 \begin{thm}
+\label{thm:parallelexec}
 Let $\oton{e}$ be well-typed Legion expressions, each with its own privileges $\Phi_i$.
@@
 $$
-If $\Phi_i \#_M \Phi_j$ for $1 \leq i < j \leq n$, then any parallel execution of expressions
+If $\Phi_i \nonint[M] \Phi_j$ for $1 \leq i < j \leq n$, then any parallel execution of expressions
 $\oton{e}$ will result in a valid interleaving of memory operations.
@@
 
-TODO: talk about atomic case
-
-\begin{figure*}
-$
-\begin{array}{l@{}l}
-valid\_interleave(S, C, E', E_1, E_2) = & any\_interleave(E', E_1, E_2) \wedge \\
-& coherent\_interleave(S, apply(S, E_1), C, excl, E', E_1, E_2) \wedge \\
-& (coherent\_interleave(S, apply(S, E_1), C, atomic, E', E_1, E_2) \vee \\
-& coherent\_interleave(S, apply(S, E_2), C, atomic, E', E_2, E_1)) \\
-
-coherent\_interleave(S, C, c, E', [], E_1, [], E_2, []) = & \forall l. l \in C \vee apply(S, E')(l) = apply(S, E_1 \dplus E_2)(l) \\
-coherent\_interleave(S, C, c, E'_a, \epsilon \dplus E'_b, E_{1a}, \epsilon \dplus E_{1b}, E_{2a}, E_{2b}) = & (\epsilon = read(l, c, v) \wedge l \not\in C \Rightarrow apply(S,E'_a)(l) = v \wedge apply(S,E_{1a})(l) = v) \wedge \\
-& coherent\_interleave(S, C, c,  E'_a \dplus \epsilon, E'_b, E_{1a} \dplus \epsilon, E_{1b}, E_{2a}, E_{2b}) \\
-coherent\_interleave(S_1, S_2, C, c, [ op(l, c', v) ] \dplus E', E_1, [ op(l, c, v) ] \dplus E_2) = & (op = read \wedge l \not\in C \wedge c = c' \Rightarrow S_2(l) = v) \wedge \\
-& coherent\_interleave(S_1, apply(S_2, op(l, c', v)), C, c, E', E_1, E_2)
-\end{array}
-$
-\caption{work in progress}
-\end{figure*}
+\subsection{Relaxed Coherence}
+
+The analysis above did not consider the coherence modes on each operation in the memory traces,
+making the worst-case assumption that all were using the $excl$ mode.  However, if two or more
+subexpressions contain calls to functions that use $atomic$ and/or $simult$ coherence modes
+on some of their regions, options other than serial execution may be guaranteed to 
+result in valid interleavings even if constituent traces have some locations in common.  There
+are two cases of particular interest that we will cover.  In each case, the entirety of the 
+reasoning from the previous section is repeated with minor modifications, so we will limit our
+discussion to a general description of the differences rather than the complete proofs.
+
+The first important case occurs when all apparently-interfering operations have coherence modes
+of either $atomic$ or $simult$.  In this case, we can see that the conditions required for
+coherence and sequential equivalence are satisfied for all locations in $L_{excl}$.  The
+requirements for locations in $L_{atomic}$ allow for any permutation of the constituent traces,
+so the runtime is free to serialize the execution of the subexpressions in any order, not just
+the original program order.
+
+In order to convert the required analysis to one that can be performed on regions, we modify
+the noninterference operator for effects to consider only locations in $L_{excl}$ and show that
+traces that a noninterfering under this modified check satisfy the requirement for atomic
+execution.  We then modify the noninterference operator on privileges to allow 
+$M(r_1) \cap M(r_2) \not= \emptyset$ if either $atomic(r_1)$ or $simult(r_1)$ is in $Q_1$ and
+either $atomic(r_2)$ or $simult(r_2)$ is in $Q_2$, and can show that subexpressions with 
+privileges that can be shown to be ``atomically noninterfering'' (under a dynamic region mapping
+$M$) will always have effects that are compatible with atomically-reordered execution.  Finally,
+in the special case where each subexpression has $atomic(r)$ or $simult(r)$ in $Q_i$ for every $r$
+in $\Phi_i$, we can eliminate the dynamic check entirely.
+
+The second case is a subset of the first, in which all apparently-interfering operations have
+the $simult$ coherence mode.  In this case, all the coherence and sequential equivalence 
+checks are trivially satisfied, and any interleaving of the constituent traces will be valid.
+We again modify the noninterference criteria for effects, traces, and privileges, this time
+only allowing $simult$ coherence annotations, and can show that subexpression with privileges
+that can be shown to be ``simulataneously noninterfering'' (under mapping $M$) can be executed
+in parallel.  This will include all the cases covered by Theorem~\ref{thm:parallelexec} as well
+as those where the programmer has used the $simult$ coherence mode to explicitly allow parallel
+execution.  Again, a special (but important) case exists where all privileges for each
+subexpression are paired with $simult$ coherence mode declarations, allowing the safety of
+parallel execution to be determined at compile time.
 
@@
 
-Let $e_1$ and $e_2$ be expressions that require privileges $\Phi_1$ and
+The enforcement of a (locally) valid interleaving at every step in the operational semantics
+captures the hierarchical nature of the scheduler used by the Legion runtime.  As the cost of
+a scheduling decision is quadratic in the number of tasks (more precisely, the number of 
+privileges for those tasks), it should be clear that a hierarchical scheduling algorithm will
+be vastly more efficient than a global algorithm that must consider the privileges of every
+concurrently executing expression.
+However, a scheduling algorithm that makes decisions hierarchically must do so with limited
+information, and must respect all required scheduling dependencies while ideally imposing as
+few extra dependencies as possible.
+
+Let $e_1$ and $e_2$ be well-typed Legion expressions that using privileges $\Phi_1$ and
 $\Phi_2$.  Now let $e_3$ be some subexpression of $e_1$ and $e_4$ be some 
@@
 
-Proper scheduling of the evaluation of $e_3$ and $e_4$ depends on whether
-they interfere.  If they do intefere (and must therefore be serialized), we
-will have $E_3 \cancel\# E_4$, and by the above result, $\Phi_3 \cancel\# \Phi_4$.
-
-We now show that this will also imply that $\Phi_1 \cancel\# \Phi_2$, and
-unless sequential behavior has been relaxed by the programmer, the
-evaluations of $e_1$ and $e_2$ will be serialized.  Since the evaluation of
-$e_3$ is a subsequence of the evaluation of $e_1$ and the evaluation of $e_4$
-is a subsequence of the evaluation of $e_2$, the evaluations of $e_3$ and $e_4$
-are properly serialized as well.
-
-Conversely, if $e_1$ and $e_2$ are judged to be noninterfering, the schedule
-does not need to check for interference between $e_3$ and $e_4$, as their
-noninterference is guaranteed.
-
-Proof is straightforward once armed with the following lemmas:
-
-\framebox{$\Phi_1 \#_M \Phi_2 \Leftrightarrow \Phi_1^* \#_M \Phi_2^*$}
-
-\framebox{$\Phi_1 \#_M \Phi_2 \wedge \Phi_3 \subseteq \Phi_2 \Rightarrow \Phi_1 \#_M \Phi_3$}
+By Theorem~\ref{thm:parallelexec}, if the runtime can verify $\Phi_1 \nonint[M] \Phi_2$, then 
+expressions $e_1$ and $e_2$ may be evaluated in parallel.  The evaluation of the subexpressions
+$e_3$ and $e_4$ will therefore be allowed to run in parallel, and we must show that this is
+permissible.  We do so by proving that any memory traces that can result from the evaluation
+of the subexpression must be noninterfering (allowing parallel execution by
+Lemma~\ref{lem:noninteffects}).
+
+\begin{thm}
+\label{thm:hiersched}
+Let $e_1$ and $e_2$ be well-typed Legion expressions that using privileges $\Phi_1$ and
+$\Phi_2$.  Now let $e_3$ be some subexpression of $e_1$ and $e_4$ be some 
+subexpression of $e_2$.  For any memory traces $E_3$ and $E_4$ that result from evaluation of
+$e_1$ and $e_2$ (with the usual consistent $M$, $L$, $H$, and $S$), $E_3$ and $E_4$ must be
+noninterfering.
+\end{thm}
+
+\begin{proof}
+Let $E_1$ be the result of an evaluation of $e_1$, with $E_3$ being the memory trace from the
+evaluation of $e_3$ within the expression tree.  $E_1$ will be formed from a tree of valid
+interleavings, each of which must include all the memory operations of each consitituent trace.
+By induction over the number of intermediate subexpressions between $e_1$ and $e_3$, we
+can show that any memory operation $\epsilon_3$ that is in $E_3$ must also be in $E_1$.
+Similarly, any memory operation $\epsilon_4$ that is in $E_4$ must also be in $E_2$.  We have
+$E_1 \nonint E_2$ from Lemma~\ref{lem:nonintpriv}, so and therefore $\epsilon_3 \nonint \epsilon_4$.  Since
+this holds for all $\epsilon_3$ in $E_3$ and all $\epsilon_4$ in $E_4$, we have $E_3 \nonint E_4$.
+\end{proof}
+
+Although a decision that $e_1$ and $e_2$ are noninterfering guarantees that $e_3$ and $e_4$ are
+noninterfering, the converse is not true.  It is possible for $e_1$ and $e_2$ to be judged to be
+potentially interfering and be serialized in their entirety even when the subexpressions could
+have been proven to be noninterfering.  This loss of precision can impact performance, and did
+for a few times during the implementation of our experiments.  When the performance impact is
+significant, two options exist to improve the precision.  In cases where the $e_1$ and $e_2$ are
+in fact noninterfering, the partitioning of regions can be further refined to make the
+disjointness visible in the privileges.  Alternately, and this is the only option when the 
+interference is dynamically determined, the code can be refactored to life the subexpressions
+(or some expression containing them) up to the same level, allowing the dynamic test based on
+the region mapping to infer noninterference in more cases.  Although this can increase the
+overhead of scheduling decisions, judicious use of this technique in practice has not shown
+any measurable performance degradation.

### 2012-07-05 doc/popl2013/langdef.tex

--- 2012-07-05 doc/popl2013/langdef.tex
@@
 
-TODO: talk about atomic case
+\subsection{Relaxed Coherence}
+
+The analysis above did not consider the coherence modes on each operation in the memory traces,
+making the worst-case assumption that all were using the $excl$ mode.  However, if two or more
+subexpressions contain calls to functions that use $atomic$ and/or $simult$ coherence modes
+on some of their regions, options other than serial execution may be guaranteed to 
+result in valid interleavings even if constituent traces have some locations in common.  There
+are two cases of particular interest that we will cover.  In each case, the entirety of the 
+reasoning from the previous section is repeated with minor modifications, so we will limit our
+discussion to a general description of the differences rather than the complete proofs.
+
+The first important case occurs when all apparently-interfering operations have coherence modes
+of either $atomic$ or $simult$.  In this case, we can see that the conditions required for
+coherence and sequential equivalence are satisfied for all locations in $L_{excl}$.  The
+requirements for locations in $L_{atomic}$ allow for any permutation of the constituent traces,
+so the runtime is free to serialize the execution of the subexpressions in any order, not just
+the original program order.
+
+In order to convert the required analysis to one that can be performed on regions, we modify
+the noninterference operator for effects to consider only locations in $L_{excl}$ and show that
+traces that a noninterfering under this modified check satisfy the requirement for atomic
+execution.  We then modify the noninterference operator on privileges to allow 
+$M(r_1) \cap M(r_2) \not= \emptyset$ if either $atomic(r_1)$ or $simult(r_1)$ is in $Q_1$ and
+either $atomic(r_2)$ or $simult(r_2)$ is in $Q_2$, and can show that subexpressions with 
+privileges that can be shown to be ``atomically noninterfering'' (under a dynamic region mapping
+$M$) will always have effects that are compatible with atomically-reordered execution.  Finally,
+in the special case where each subexpression has $atomic(r)$ or $simult(r)$ in $Q_i$ for every $r$
+in $\Phi_i$, we can eliminate the dynamic check entirely.
+
+The second case is a subset of the first, in which all apparently-interfering operations have
+the $simult$ coherence mode.  In this case, all the coherence and sequential equivalence 
+checks are trivially satisfied, and any interleaving of the constituent traces will be valid.
+We again modify the noninterference criteria for effects, traces, and privileges, this time
+only allowing $simult$ coherence annotations, and can show that subexpression with privileges
+that can be shown to be ``simulataneously noninterfering'' (under mapping $M$) can be executed
+in parallel.  This will include all the cases covered by Theorem~\ref{thm:parallelexec} as well
+as those where the programmer has used the $simult$ coherence mode to explicitly allow parallel
+execution.  Again, a special (but important) case exists where all privileges for each
+subexpression are paired with $simult$ coherence mode declarations, allowing the safety of
+parallel execution to be determined at compile time.
 
@@
 
-Proper scheduling of the evaluation of $e_3$ and $e_4$ depends on whether
-they interfere.  If they do intefere (and must therefore be serialized), we
-will have $E_3 \cancel\# E_4$, and by the above result, $\Phi_3 \cancel\# \Phi_4$.
-
-We now show that this will also imply that $\Phi_1 \cancel\# \Phi_2$, and
-unless sequential behavior has been relaxed by the programmer, the
-evaluations of $e_1$ and $e_2$ will be serialized.  Since the evaluation of
-$e_3$ is a subsequence of the evaluation of $e_1$ and the evaluation of $e_4$
-is a subsequence of the evaluation of $e_2$, the evaluations of $e_3$ and $e_4$
-are properly serialized as well.
-
-Conversely, if $e_1$ and $e_2$ are judged to be noninterfering, the schedule
-does not need to check for interference between $e_3$ and $e_4$, as their
-noninterference is guaranteed.
-
-Proof is straightforward once armed with the following lemmas:
-
-\framebox{$\Phi_1 \#_M \Phi_2 \Leftrightarrow \Phi_1^* \#_M \Phi_2^*$}
-
-\framebox{$\Phi_1 \#_M \Phi_2 \wedge \Phi_3 \subseteq \Phi_2 \Rightarrow \Phi_1 \#_M \Phi_3$}
+By Theorem~\ref{thm:parallelexec}, if the runtime can verify $\Phi_1 \#_M \Phi_2$, then 
+expressions $e_1$ and $e_2$ may be evaluated in parallel.  The evaluation of the subexpressions
+$e_3$ and $e_4$ will therefore be allowed to run in parallel, and we must show that this is
+permissible.  We do so by proving that any memory traces that can result from the evaluation
+of the subexpression must be noninterfering (allowing parallel execution by
+Lemma~\ref{lem:noninteffects}).
+
+\begin{thm}
+Let $e_1$ and $e_2$ be well-typed Legion expressions that using privileges $\Phi_1$ and
+$\Phi_2$.  Now let $e_3$ be some subexpression of $e_1$ and $e_4$ be some 
+subexpression of $e_2$.  For any memory traces $E_3$ and $E_4$ that result from evaluation of
+$e_1$ and $e_2$ (with the usual consistent $M$, $L$, $H$, and $S$), $E_3$ and $E_4$ must be
+noninterfering.
+\end{thm}
+
+\begin{proof}
+Let $E_1$ be the result of an evaluation of $e_1$, with $E_3$ being the memory trace from the
+evaluation of $e_3$ within the expression tree.  $E_1$ will be formed from a tree of valid
+interleavings, each of which must include all the memory operations of each consitituent trace.
+By induction over the number of intermediate subexpressions between $e_1$ and $e_3$, we
+can show that any memory operation $\epsilon_3$ that is in $E_3$ must also be in $E_1$.
+Similarly, any memory operation $\epsilon_4$ that is in $E_4$ must also be in $E_2$.  We have
+$E_1 \# E_2$ from Lemma~\ref{lem:nonintpriv}, so and therefore $\epsilon_3 \# \epsilon_4$.  Since
+this holds for all $\epsilon_3$ in $E_3$ and all $\epsilon_4$ in $E_4$, we have $E_3 \# E_4$.
+\end{proof}
+
+Although a decision that $e_1$ and $e_2$ are noninterfering guarantees that $e_3$ and $e_4$ are
+noninterfering, the converse is not true.  It is possible for $e_1$ and $e_2$ to be judged to be
+potentially interfering and be serialized in their entirety even when the subexpressions could
+have been proven to be noninterfering.  This loss of precision can impact performance, and did
+for a few times during the implementation of our experiments.  When the performance impact is
+significant, two options exist to improve the precision.  In cases where the $e_1$ and $e_2$ are
+in fact noninterfering, the partitioning of regions can be further refined to make the
+disjointness visible in the privileges.  Alternately, and this is the only option when the 
+interference is dynamically determined, the code can be refactored to life the subexpressions
+(or some expression containing them) up to the same level, allowing the dynamic test based on
+the region mapping to infer noninterference in more cases.  Although this can increase the
+overhead of scheduling decisions, judicious use of this technique in practice has not shown
+any measurable performance degradation.

### 2012-07-06 doc/popl2013/evaluation.tex

@@
 
-Figures \ref{fig:ckt_overhead}, \ref{fig:fluid_overhead}, and \ref{fig:amr_overhead} show 
+Figures~\ref{fig:ckt_overhead}, \ref{fig:fluid_overhead}, and \ref{fig:amr_overhead} show 
 the total time spent by all CPUs and GPUs in each phase of the application.  The topmost
@@
 each figure the problem size stays the same as the number of processors are increased
-(strong scaling).  Figure \ref{fig:amr_overhead} includes multiple problem sizes to show
+(strong scaling).  Figure~\ref{fig:amr_overhead} includes multiple problem sizes to show
 how overhead is affected by changing problem size (weak scaling).  For cases where there
 is an existing implementation to compare against we have included a dotted line indicating
-baseline performance.  In a few cases (Figures \ref{fig:fluid_overhead} and 
+baseline performance.  In a few cases (Figures~\ref{fig:fluid_overhead} and 
 \ref{fig:amr4096}) checking overhead is the difference
@@
 \vspace{-2mm}
-\caption{Overhead for the AMR application for three different problem sizes.\label{fig:amr_overhead}}
+\caption{Overhead for the AMR application.\label{fig:amr_overhead}}
 \vspace{-6mm}
@@
 \label{subsec:scalability}
+The scalability of Legion derives directly from Theorem~\ref{thm:hiersched}.  This 
+theorem makes it safe for the runtime to only perform dependence checks between two
+tasks that share the same immediate parent task instead of having to consider all
+possible interactions with other tasks anywhere in the system.  Since all tasks that
+share the same immediate parent are all created on the node where the parent is executing,
+all dependence checks and scheduling decisions can be performed locally with no communication.  
+Without this property the runtime would be forced to inform all other processors in the system 
+of every task that it launches using expensive broadcasts.
+
+Figures~\ref{fig:ckt_overhead}, \ref{fig:fluid_overhead}, and \ref{fig:amr_overhead} show that
+the overhead of the Legion runtime is always less than 7\% of the total execution time
+of the applications.  In some applications communication doesn't scale well, but in these cases communication is
+the result of the algorithm being performed and is not caused by runtime communication.
+Even in the case of the Circuit application which exhibits quadratic communication, the Legion
+runtime is able to achieve a 62.5X speedup on 96 GPUs over a hand-coded single GPU implementation\cite{Legion12}.

### 2012-07-06 doc/popl2013/intro.tex

@@
 regions (which we usually refer to simply as {\em regions}) express
-locality, which is just data that will be used together and therefore
-should be colocated in the machine.  Listing~\ref{lst:linkedlist}
-gives an example that defines a region containing a linked list of integers and a task
-that sums the elements of the list.  As is standard in region systems, we use
-$T @ r$ for the type of a pointer to a value of type $T$ in region $r$.
-As is also
+{\em locality}, which is just data that will be used together and
+therefore should be colocated in the machine.
+Listing~\ref{lst:linkedlist} gives an example that defines a region
+containing a linked list of integers and a task that sums the elements
+of the list.  As is standard in region systems, we use $T @ r$ for the
+type of a pointer to a value of type $T$ in region $r$.  The task also
+illustrates a simple use of the permissions system; in this case the
+task {\tt sum\_list} must have permission {\tt reads(r)}, the ability
+to read the values in region {\tt r}.
+
+Listing~\ref{lst:treereverse} gives a more elaborate example.  While
+artificial (it is designed to use the greatest number of Legion
+features in the smallest space), the program illustrates typical
+Legion programming idioms.  Given a list of integers as defined in Listing~\ref{lst:linkedlist}, 
+the task {\tt make\_tree} builds a binary tree of pointers to list elements.
+Each internal node of the tree stores a pointer to a list element with value $i$
+and the left and right subtrees hold pointers to list elements with values less than $i$ and greater than $i$,
+respectively.  The task {\tt reverse\_tree} then does an in-place, recursive reversal of the children of each interior
+node of the tree.  We focus on a description of the structure of types and permissions in this example.
+
+Line 1 defines a type that is a {\em region relationship} {\tt rr[ rll, rlr, rtl, rtr]}.  Region relationships are
+constrained existential types.  The {\tt tree} region relationship defines the type of a node of the tree: it has a pointer
+to a list node and two pointers to the left and right subtrees. The crucial part is the {\em subregion} and {\em disjointness}
+constraints that describe the relationships between the various regions.  Here {\tt rl} is the region holding all list elements,
+{\tt rls} is the region holding
 

### 2012-07-06 doc/popl2013/legion-types.tex
Fixing broken references

@@
 \usepackage{multirow}
+\usepackage{ifmtarg}
+\usepackage{url}
 
@@
 \begin{abstract}
-Legion is a programming model that introduces logical regions as a mechanism for targeting large heterogeneous machines with complex memory hierarchies.
-Logical regions are first-class values in Legion, allowing programmers to specify independence and locality properties of program data dynamically.
-We present a type system with permission and coherence properties for reasoning about logical regions.
-We prove the soundness of the  type system and use it to show that all memory accesses in Legion are safe.
-Furthermore, we show that type soundness directly enables a scalable, hierarchical scheduling algorithm for Legion programs.
-We implement a type checker for Legion programs and improve performance up to 71% by eliding memory checks the type system proves safe.
-We also demonstrate that an implementation of the Legion runtime based on the hierarchical scheduling algorithm scales well for three real-world applications.
+
+Legion is a programming model that introduces 
+{\em logical regions} as a mechanism for targeting large heterogeneous machines 
+with complex memory hierarchies.  Logical regions are first-class 
+values in Legion, allowing programmers to specify independence 
+and locality properties of program data dynamically.  We present
+a type system with permissions and coherence properties for reasoning about logical regions.  
+We prove the soundness of the type system and use it to show
+that all memory accesses in Legion are safe.  Furthermore, we prove that
+type soundness directly enables a scalable, hierarchical
+scheduling algorithm for Legion programs.  We implement a type checker
+for Legion programs and improve performance up to 71\% by eliding
+memory checks the type system proves safe.  We also demonstrate that an implementation of the Legion 
+runtime based on the hierarchical scheduling algorithm scales well for
+three real-world applications.
 

### 2012-07-06 doc/popl2013/intro.tex

@@
      		       where rll #$\le$# rls and rlr #$\le$# rls and rll #$*$# rlr and 
-                                  rtl #$\le$# rt and rtr #$\le$# rt  and rtl #$*$# rtr
+                                  rtl #$\le$# rt and rtr #$\le$# rt and rtl #$*$# rtr and rls #$\le$# rl
 

### 2012-07-06 doc/popl2013/legion-types.tex
Updating the abstract to match the one alex submitted

@@
 \begin{abstract}
-
 Legion is a programming model that introduces 

### 2012-07-06 doc/popl2013/intro.tex
Adding a constraint to the tree region relation example in the paper

--- 2012-07-05 doc/popl2013/intro.tex
@@
 \label{sec:intro}
-
-In the past decade machine architecture, particularly at the high
-performance end of the spectrum, has undergone a revolution.  The
-latest supercomputers are now complex hierarchies of many different
-kinds of computing technologies: networks of nodes at the top level,
-multiple chips per node, mutiple cores within a chip, and, most
-recently, multiple accelerators (usually GPUs) per node, which are
-themselves can be further decomposed into several levels of machine
-hierarchy.  In this paper, we present the operational and static
-semantics of Legion, a programming language targeted at providing an
-appropriate level of abstraction for programming such machines, one
-that is both sufficiently high-level to be portable while still
-exposing aspects that are crucial to performance. A very brief summary
-of Legion is that data is organized in a hierarchy of {\em regions}
-and subregions and computation is organized in a hierarchy of {\em
-tasks} and subtasks operating on regions.  Regions and tasks interact
-through a static system of {\em region permissions} specifying the
-operations a task may perform on a region argument (e.g., read or
-write, among others) and novel {\em region coherence} annotations that
-expresses whether other tasks may perform operations concurrently on
-the region (e.g., {\em exclusive} access for task $t$ guarantees no
-other task's updates to a region are visible to $t$).  We prove the
-soundness of the permissions and coherence system and use these two
-results to prove a third result: if two tasks $t_1$ and $t_2$ are {\em
-dependent} (roughly, must be executed in program order to guarantee
-the correct answer), then the unique ancestors of $t_1$ and $t_2$ in
-the task hierarchy that are {\em siblings} (have the same parent task)
-are also dependent.  This theoremis the basis for distributed
-scheduling in a Legion implementation, which is crucial for
-performance; any centralized scheduler would be a severe bottleneck
-because of the huge latencies involved in moving data and computation
-in the target class of machines.
-
-Legion programs organize data into {\em logical regions}.  Logical
-regions (which we usually refer to simply as {\em regions}) express
-locality, which is just data that will be used together and therefore
-should be colocated in the machine.  Listing~\ref{lst:linkedlist}
-gives an example that defines a region containing a linked list of integers and a task
-that sums the elements of the list.  As is standard in region systems, we use
-$T @ r$ for the type of a pointer to a value of type $T$ in region $r$.
-As is also
-
-Two of the basic properties programmers need to express in writing
-programs for highly parallel, distributed machines are the {\em
-locality} and {\em independence} of data. Locality identifies data
-that will be used together and so should be placed in the same part of
-the machine; independence identifies data sets that are disjoint and
-so can be operated on in parallel.  Legion programs use {\em logical
-regions} to express both.  A logical region (which we will usually
-refer to as simply a {\em region} for brevity) is a
-programmer-specified collection of data; we use the notation of
-separating conjunction $r_1 \ast r_2$ for disjoint regions $r_1$ and
-$r_2$.
-
-%Programming
-%these machines is currently done by using a heterogeneous collection
-%of software technologies targeted at the individual level; a typical
-%melange is MPI for the network, OpenMP for the node, and CUDA for the GPUs.
+Introduction goes here.
 

### 2012-07-06 doc/popl2013/intro.tex
intro

--- 2012-07-06 doc/popl2013/intro.tex
@@
 
-Line 1 defines a type that is a {\em region relationship} {\tt rr[ rll, rlr, rtl, rtr]}.  Region relationships are
-constrained existential types.  The {\tt tree} region relationship defines the type of a node of the tree: it has a pointer
-to a list node and two pointers to the left and right subtrees. The crucial part is the {\em subregion} and {\em disjointness}
-constraints that describe the relationships between the various regions.  Here {\tt rl} is the region holding all list elements,
-{\tt rls} is the region holding
+Line 1 defines a type that is a {\em region relationship} {\tt rr[
+rll, rlr, rtl, rtr]}.  Region relationships are constrained
+existential types.  The {\tt tree} region relationship defines the
+type of a node of the tree. The naming convention followed throughout
+the example is that regions that begin with {\tt rl} name regions
+holding list elements and regions that begin with {\tt rt} name
+regions holding tree elements.  The most important part is the {\em
+subregion} and {\em disjointness} constraints that describe the
+relationships between the various regions.  Here {\tt rl} is the
+region holding all list elements and {\tt rls} and {\tt rt} are the
+regions holding just the list elements pointed to by and the tree
+nodes included in the tree.  Note that {\tt rls} is required to be a
+subregion of {\tt rl}, which means that all the values in region {\tt
+rls} are also in region {\tt rl}.  The body of the {\tt tree} region
+relationship declares the shape of a node of the tree, which includes
+a pointer to a list element in {\tt rls} and pointers to two
+subtrees. The region relation says that there exist four regions {\tt
+rll, rlr, rtl}, and {\tt rtr} such that the left subtree has list and
+tree elements that are in subregions {\tt rll} and {\tt rtl} of the
+corresponding list and tree regions {\tt rls} and {\tt rt} for the
+entire tree; the right subtree's list and tree elements in subregions
+{\tt rlr} and {\tt rtr} are also subregions of {\tt rls} and {\tt rt}
+respectively.  Furthermore, the constraints $\tt rll \ast rlr$ and
+$\tt rtl \ast rtr$ guarantee that the left and right subtrees are
+disjoint.
+
+The {\tt make\_tree} task, which builds a {\tt tree} indexing a list
+of integers, illustrates region {\em partitions}, another important
+building block of Legion programs.  The region arguments to {\tt
+make\_tree} are {\tt rl}, the region holding all list elements, {\tt
+rls}, the subregion of {\tt rls} holding those list elements to be
+indexed by this call to {\tt make\_tree}, and {\tt rt}, the region to
+hold the tree itself.  Line 21 divides the list elements in {\tt rls}
+into those that are less than and those that are greater than a split
+value (which is arbitrarily chosen to be the value of the first
+element of the list).  Partitioning is done by constructing a {\em
+coloring} using the task {\tt color\_list}, which traverses the list
+and for each list element $e$ in the subregion (called {\tt rs} in
+{\tt color\_region}) either colors the element 1 or 2 (colors are
+integers) depending on its relationship to the split value.  A
+coloring is just a distinghuished function, built using the Legion
+primitive {\tt color}, that maps elements of a region to integers.
+When used in a {\tt partition} expression the coloring defines how to
+populate new subregtions of the partitioned region.  On line 22, the
+list elements in {\tt rls} are partitioned into two subregions {\tt
+rll} and {\tt rlr} of list elements less than, and greater than or
+equal to, the split value, respectively.  Another use of {\tt
+partition} is seen on line 21 where the {\tt rt} region is partitioned
+into subregions to hold the left {\tt rlt} and right {\tt rtr}
+subtrees, respectively.  Because region {\tt rt} is empty at this
+point (has no elements), the empty coloring function {\tt newcolor rt}
+for the region {\tt rt} is given to the partitioning operation; the
+two new subregions are also empty.  These examples illustrate that
+partitioning can be used for two distinct, and equally important,
+styles of computation: {\em allocate-then-partition}, in which a
+global collection of data is created which is then divided up for
+(usually parallel) processing, and {\em partition-then-allocate}, in
+which a collection of empty but disjoint subregions is created and
+then populated with data on which (usually parallel) processing will
+take place.
+
+Lines 26-27 illustrates {\em packing} a region relationship into the heap.
+Regions are first-class values in Legion and can be stored in data structures.
+Here a tree node is written to the {\tt root} pointer after being {\tt packed}.
+Opertionally, the four existentially quantified regions {\tt rll, rlr, rtl} and {\tt rtr} are
+stored with the tree value.  The type system guarantees statically that the packed value
+satisfies the subregion and disjointness constraints declared on lines 3-4 at the point
+the packed value is stored.  The advantage is that when the region relationship value is read 
+from the heap and {\tt unpack} is applied (line 32) the newly introduced regions are already known to 
+satisfy these constraints.  A key design decision in Legion, one that turns out to be
+central to all the soundness results as well as the ability to perform distributed, parallel
+scheduling of tasks, is that permissions and coherence annotations (which we have not yet discussed)
+are {\em not} captured by region relationships.  When a region is stored in the heap it has no
+permissions or coherence associated with it; these belong to the task.  When the region relationship
+is read out of the heap and unpacked, the task must already hold permissions for an unpacked region 
+to use it. Thus, on line 32, the task {\tt reverse\_tree} unpacks the region {\tt rtl} (among others).
+The reads and writes of {\tt rtl} typecheck, however, because {\tt reverse\_tree} has read and
+write permission for region {\tt rt} and $\tt rtl \leq rt}$ is a constraint of the region relationship (note
+that the region {\tt rt} is not one of the existentially quantified regions!).
 

### 2012-07-06 doc/popl2013/intro.tex
intro

@@
+% This is a description of how the listings should be formatted.
+% It can go anywhere before the listings.
+\lstset{
+  captionpos=b,
+  language=Haskell,
+  basicstyle=\scriptsize,
+  numbers=left,
+  numberstyle=\tiny,
+  columns=fullflexible,
+  stepnumber=1,
+  escapechar=\#,
+  keepspaces=true,
+  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
+  morekeywords={task,rr,int,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,color,newcolor,atomic,simultaneous},
+  deletekeywords={float,head,min,max}
+}
+
 \section{Introduction}
@@
 multiple chips per node, mutiple cores within a chip, and, most
-recently, multiple accelerators (usually GPUs) per node, which are
-themselves can be further decomposed into several levels of machine
-hierarchy.  In this paper, we present the operational and static
+recently, multiple accelerators (usually GPUs) per node, which 
+can be further decomposed into several levels.   In this paper, we present the operational and static
 semantics of Legion, a programming language targeted at providing an
@@
 the task hierarchy that are {\em siblings} (have the same parent task)
-are also dependent.  This theoremis the basis for distributed
-scheduling in a Legion implementation, which is crucial for
+are also dependent.  This theorem is the basis for correct distributed
+task scheduling in a Legion implementation, which is crucial for
 performance; any centralized scheduler would be a severe bottleneck
@@
 in the target class of machines.
-
-Legion programs organize data into {\em logical regions}.  Logical
-regions (which we usually refer to simply as {\em regions}) express
-{\em locality}, which is just data that will be used together and
-therefore should be colocated in the machine.
-Listing~\ref{lst:linkedlist} gives an example that defines a region
-containing a linked list of integers and a task that sums the elements
-of the list.  As is standard in region systems, we use $T @ r$ for the
-type of a pointer to a value of type $T$ in region $r$.  The task also
-illustrates a simple use of the permissions system; in this case the
-task {\tt sum\_list} must have permission {\tt reads(r)}, the ability
-to read the values in region {\tt r}.
-
-Listing~\ref{lst:treereverse} gives a more elaborate example.  While
-artificial (it is designed to use the greatest number of Legion
-features in the smallest space), the program illustrates typical
-Legion programming idioms.  Given a list of integers as defined in Listing~\ref{lst:linkedlist}, 
-the task {\tt make\_tree} builds a binary tree of pointers to list elements.
-Each internal node of the tree stores a pointer to a list element with value $i$
-and the left and right subtrees hold pointers to list elements with values less than $i$ and greater than $i$,
-respectively.  The task {\tt reverse\_tree} then does an in-place, recursive reversal of the children of each interior
-node of the tree.  We focus on a description of the structure of types and permissions in this example.
-
-Line 1 defines a type that is a {\em region relationship} {\tt rr[
-rll, rlr, rtl, rtr]}.  Region relationships are constrained
-existential types.  The {\tt tree} region relationship defines the
-type of a node of the tree. The naming convention followed throughout
-the example is that regions that begin with {\tt rl} name regions
-holding list elements and regions that begin with {\tt rt} name
-regions holding tree elements.  The most important part is the {\em
-subregion} and {\em disjointness} constraints that describe the
-relationships between the various regions.  Here {\tt rl} is the
-region holding all list elements and {\tt rls} and {\tt rt} are the
-regions holding just the list elements pointed to by and the tree
-nodes included in the tree.  Note that {\tt rls} is required to be a
-subregion of {\tt rl}, which means that all the values in region {\tt
-rls} are also in region {\tt rl}.  The body of the {\tt tree} region
-relationship declares the shape of a node of the tree, which includes
-a pointer to a list element in {\tt rls} and pointers to two
-subtrees. The region relation says that there exist four regions {\tt
-rll, rlr, rtl}, and {\tt rtr} such that the left subtree has list and
-tree elements that are in subregions {\tt rll} and {\tt rtl} of the
-corresponding list and tree regions {\tt rls} and {\tt rt} for the
-entire tree; the right subtree's list and tree elements in subregions
-{\tt rlr} and {\tt rtr} are also subregions of {\tt rls} and {\tt rt}
-respectively.  Furthermore, the constraints $\tt rll \ast rlr$ and
-$\tt rtl \ast rtr$ guarantee that the left and right subtrees are
-disjoint.
-
-The {\tt make\_tree} task, which builds a {\tt tree} indexing a list
-of integers, illustrates region {\em partitions}, another important
-building block of Legion programs.  The region arguments to {\tt
-make\_tree} are {\tt rl}, the region holding all list elements, {\tt
-rls}, the subregion of {\tt rls} holding those list elements to be
-indexed by this call to {\tt make\_tree}, and {\tt rt}, the region to
-hold the tree itself.  Line 21 divides the list elements in {\tt rls}
-into those that are less than and those that are greater than a split
-value (which is arbitrarily chosen to be the value of the first
-element of the list).  Partitioning is done by constructing a {\em
-coloring} using the task {\tt color\_list}, which traverses the list
-and for each list element $e$ in the subregion (called {\tt rs} in
-{\tt color\_region}) either colors the element 1 or 2 (colors are
-integers) depending on its relationship to the split value.  A
-coloring is just a distinghuished function, built using the Legion
-primitive {\tt color}, that maps elements of a region to integers.
-When used in a {\tt partition} expression the coloring defines how to
-populate new subregtions of the partitioned region.  On line 22, the
-list elements in {\tt rls} are partitioned into two subregions {\tt
-rll} and {\tt rlr} of list elements less than, and greater than or
-equal to, the split value, respectively.  Another use of {\tt
-partition} is seen on line 21 where the {\tt rt} region is partitioned
-into subregions to hold the left {\tt rlt} and right {\tt rtr}
-subtrees, respectively.  Because region {\tt rt} is empty at this
-point (has no elements), the empty coloring function {\tt newcolor rt}
-for the region {\tt rt} is given to the partitioning operation; the
-two new subregions are also empty.  These examples illustrate that
-partitioning can be used for two distinct, and equally important,
-styles of computation: {\em allocate-then-partition}, in which a
-global collection of data is created which is then divided up for
-(usually parallel) processing, and {\em partition-then-allocate}, in
-which a collection of empty but disjoint subregions is created and
-then populated with data on which (usually parallel) processing will
-take place.
-
-Lines 26-27 illustrates {\em packing} a region relationship into the heap.
-Regions are first-class values in Legion and can be stored in data structures.
-Here a tree node is written to the {\tt root} pointer after being {\tt packed}.
-Opertionally, the four existentially quantified regions {\tt rll, rlr, rtl} and {\tt rtr} are
-stored with the tree value.  The type system guarantees statically that the packed value
-satisfies the subregion and disjointness constraints declared on lines 3-4 at the point
-the packed value is stored.  The advantage is that when the region relationship value is read 
-from the heap and {\tt unpack} is applied (line 32) the newly introduced regions are already known to 
-satisfy these constraints.  A key design decision in Legion, one that turns out to be
-central to all the soundness results as well as the ability to perform distributed, parallel
-scheduling of tasks, is that permissions and coherence annotations (which we have not yet discussed)
-are {\em not} captured by region relationships.  When a region is stored in the heap it has no
-permissions or coherence associated with it; these belong to the task.  When the region relationship
-is read out of the heap and unpacked, the task must already hold permissions for an unpacked region 
-to use it. Thus, on line 32, the task {\tt reverse\_tree} unpacks the region {\tt rtl} (among others).
-The reads and writes of {\tt rtl} typecheck, however, because {\tt reverse\_tree} has read and
-write permission for region {\tt rt} and $\tt rtl \leq rt}$ is a constraint of the region relationship (note
-that the region {\tt rt} is not one of the existentially quantified regions!).
-
-Two of the basic properties programmers need to express in writing
-programs for highly parallel, distributed machines are the {\em
-locality} and {\em independence} of data. Locality identifies data
-that will be used together and so should be placed in the same part of
-the machine; independence identifies data sets that are disjoint and
-so can be operated on in parallel.  Legion programs use {\em logical
-regions} to express both.  A logical region (which we will usually
-refer to as simply a {\em region} for brevity) is a
-programmer-specified collection of data; we use the notation of
-separating conjunction $r_1 \ast r_2$ for disjoint regions $r_1$ and
-$r_2$.
-
-%Programming
-%these machines is currently done by using a heterogeneous collection
-%of software technologies targeted at the individual level; a typical
-%melange is MPI for the network, OpenMP for the node, and CUDA for the GPUs.
-
-% This is a description of how the listings should be formatted.
-% It can go anywhere before the listings.
-\lstset{
-  captionpos=b,
-  language=Haskell,
-  basicstyle=\scriptsize,
-  numbers=left,
-  numberstyle=\tiny,
-  columns=fullflexible,
-  stepnumber=1,
-  escapechar=\#,
-  keepspaces=true,
-  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
-  morekeywords={task,rr,int,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,color,newcolor,atomic,simultaneous},
-  deletekeywords={float,head,min,max}
-}
 
@@
 \end{lstlisting}
+
+Legion programs organize data into {\em logical regions}.  Logical
+regions (which we usually refer to simply as {\em regions}) express
+{\em locality}, which is simply data that will be used together.
+Listing~\ref{lst:linkedlist} gives an example that defines a region
+containing a linked list of integers and a task that sums the elements
+of the list.  As is standard in region systems, we use $T @ r$ for the
+type of a pointer to a value of type $T$ in region $r$.  The task also
+illustrates a simple use of the permissions system:
+task {\tt sum\_list} must have permission {\tt reads(r)}, the ability
+to read the values in region {\tt r}.
 
@@
 
+Listing~\ref{lst:treereverse} gives a more elaborate example.  While
+artificial (it is designed to use the greatest number of Legion
+features in the smallest space), the program illustrates typical
+Legion programming idioms.  Given a list of integers as defined in Listing~\ref{lst:linkedlist}, 
+the task {\tt make\_tree} builds a binary tree of pointers to list elements.
+Each internal node of the tree stores a pointer to a list element with value $i$
+and the left and right subtrees hold pointers to list elements with values less than $i$ and greater than $i$,
+respectively.  The task {\tt reverse\_tree} then does an in-place, recursive reversal of the children of each interior
+node of the tree.  We focus on a description of the structure of types, regions and permissions in this example.
+
+Line 1 defines a {\em region relationship} {\tt rr[rll, rlr, rtl, rtr]}.  
+Region relationships are bounded existential types.  The
+{\tt tree} region relationship defines the type of a binary tree
+node. The naming convention followed throughout the example is that
+regions beginning with {\tt rl} name regions 
+holding list elements and regions beginning with {\tt rt} name
+regions holding tree elements.  The most important part is the {\em
+  subregion} and {\em disjointness} constraints that describe the
+relationships between the various regions.  Here {\tt rl} is the
+region holding all list elements and {\tt rls} and {\tt rt} are the
+regions holding, respetively, the subset of list elements and the tree
+nodes for this tree.  Note that {\tt rls} is required to be a
+subregion of {\tt rl}, which means that all the values in region {\tt
+  rls} are also in region {\tt rl}.  The body of the {\tt tree} region
+relationship declares the shape of a node of the tree, which includes
+a pointer to a list element in {\tt rls} and pointers to two
+subtrees. The region relation says that there exist four regions {\tt
+  rll, rlr, rtl}, and {\tt rtr} such that the left subtree has list
+and tree elements that are in subregions {\tt rll} and {\tt rtl} of
+the corresponding list and tree regions {\tt rls} and {\tt rt} for the
+entire tree; the right subtree's list and tree elements in subregions
+{\tt rlr} and {\tt rtr} are also subregions of {\tt rls} and {\tt rt}
+respectively.  Furthermore, the constraints $\tt rll \ast rlr$ and
+$\tt rtl \ast rtr$ guarantee that the left and right subtrees are
+disjoint.
+
+The {\tt make\_tree} task, which builds a {\tt tree} indexing a list
+of integers, illustrates region {\em partitions}, another important
+building block of Legion programs.  The region arguments to {\tt
+make\_tree} are {\tt rl}, the region holding all list elements, {\tt
+rls}, the subregion of {\tt rl} holding those list elements to be
+indexed by this call to {\tt make\_tree}, and {\tt rt}, the region to
+hold the tree itself.  Line 21 divides the list elements in {\tt rls}
+into those that are less than and those that are greater than a split
+value (which is arbitrarily chosen to be the value of the first
+element of the list).  Partitioning is done by constructing a {\em
+coloring} using the task {\tt color\_list} (line 6), which traverses the list
+and for each list element $e$ in the subregion (called {\tt rs} in
+{\tt color\_region}) either colors the element 1 or 2 (colors are
+integers) depending on its relationship to the split value.  A
+coloring is just a distinghuished function, built using the Legion
+primitive {\tt color}, that maps elements of a region to integers.
+When used in a {\tt partition} expression the coloring defines how to
+populate new subregtions of the partitioned region.  On line 22, the
+list elements in {\tt rls} are partitioned into two subregions {\tt
+rll} and {\tt rlr} of list elements less than, and greater than or
+equal to, the split value, respectively.  Another use of {\tt
+partition} is seen on line 21 where the {\tt rt} region is partitioned
+into subregions to hold the left {\tt rlt} and right {\tt rtr}
+subtrees, respectively.  Because region {\tt rt} is empty at this
+point (has no elements), the empty coloring function {\tt newcolor rt}
+for the region {\tt rt} is given to the partitioning operation; the
+two new subregions are also empty.  These examples illustrate that
+partitioning can be used for two distinct, and equally important,
+styles of computation: {\em allocate-then-partition}, in which a
+global collection of data is created an then divided for
+(usually parallel) processing, and {\em partition-then-allocate}, in
+which a collection of empty but disjoint subregions is created and
+then populated with data on which (usually parallel) processing will
+take place.
+
+Lines 26-27 illustrates {\em packing} a region relationship into the heap.
+Regions are first-class values in Legion and can be stored in data structures.
+Here a tree node is written to the {\tt root} pointer after being {\tt packed}.
+Opertionally, the four existentially quantified regions {\tt rll, rlr, rtl} and {\tt rtr} are
+stored with the tree value.  The type system guarantees statically that the packed value
+satisfies the subregion and disjointness constraints declared on lines 3-4 at the point
+the packed value is stored.  When the region relationship value is read 
+from the heap and {\tt unpack} is applied (line 32) the newly introduced regions are known to 
+satisfy these constraints.  A key design decision in Legion, one that turns out to be
+central to all the soundness results as well as the ability to perform distributed, parallel
+scheduling of tasks, is that permissions and coherence annotations (which we have not yet discussed)
+are not captured by region relationships.  When a region is stored in the heap it has no
+permissions or coherence associated with it; these belong to the task.  When the region relationship
+is read out of the heap and unpacked, the task must already hold permissions for an unpacked region 
+to use it. Thus, on line 32, the task {\tt reverse\_tree} unpacks the region {\tt rtl} (among others).
+The reads and writes of {\tt rtl} typecheck, however, because {\tt reverse\_tree} has read and
+write permission for region {\tt rt} and $\tt rtl \leq rt$ is a constraint of the region relationship; holding
+permission for a region $r$ implies the same permission for any subregion of $r$ (note
+that the region {\tt rt} is not one of the existentially quantified regions).
+
+The {\tt reverse\_tree} task swaps the two subtrees of a tree in place (lines 35-36) after calling two
+subtasks to recursively swap all pointers in the subtrees (lines 33-34).  Beside the unpack operations, there
+are two other interesting features of this task.  First, the Legion scheduler will detect at runtime that the two
+subtasks are independent and may choose to run them in parallel.  In this example the test can even be done
+statically (because {\tt rtl} and {\tt rtr} are statically known to be disjoint) but, in general, testing the
+runtime {\em physical regions} is more precise than the compile-time logical regions.  Furthermore, it is a theorem
+that it is sufficient for the scheduler to check only {\em sibling} tasks (subtasks with the same parent task,
+such as those on lines 33-34) for dependences.  Thus, further subtasks within the recursive calls can be scheduled
+without consideration of their subtask cousins, which serves as the basis of a distributed scheduler for Legion.
+The second point is that
+even after destructively updating the subtrees the type system can prove using all of the region constraints in the
+{\tt tree} region relationship that the result is a well-formed {\tt tree}.
+
 \begin{lstlisting}[label={lst:histogram},caption={Histogram with Reductions and Atomic Coherence}]
@@
 \end{lstlisting}
+
+The last example, shown in Listing~\ref{lst:histogram}, is a histogram
+computation that recursively walks a {\tt tree}.  The first version 
+uses a {\em coherence annotation} to relax the
+required ordering of tasks.  When there is no coherence annotation on
+a region $r$ used by a task $t$, the task implicitly requires $r$
+with {\em exclusive} coherence, meaning that any other task $t'$ that might
+update $r$ in a way visible to $t$ must appear to execute in normal sequential
+order with respect to $t$.  The task {\tt hist\_tree1} only requires the region {\tt rh}
+with {\tt atomic} coherence, meaning that other tasks that also use {\tt rh} with {\tt atomic} coherence
+may be reordered to appear to execute atomically before or after the task.  The task {\tt hist\_tree2} achieves
+a similar affect in a finer-grain way by accessing {\tt rh} only with {\tt reduce} privileges, which allows reductions
+with the same associative-commutative operator to be reordered.
+
+The rest of the paper is organized as follows.  Section~\ref{sec:legioncore} defines the core legion language, type system,
+opertional semantics, and presents our main results, proving the soundness of permissions and coherence annotations, and proving
+that it is sufficient for the scheduler to test dependences only between sibling tasks.
+Section~\ref{sec:evaluation} presents an experimental evaluation, measuring both the improvement achieved by eliding checks the static type system can prove safe as well as demonstrating the performance benefits of the distributed scheduling.
+Section~\ref{sec:related} discusses the most closely related work and Section~\ref{sect:conclusion} concludes.

### 2012-07-06 doc/popl2013/legion-types.tex
intro

--- 2012-07-06 doc/popl2013/legion-types.tex
@@
 values in Legion, allowing programmers to specify independence 
-and locality properties of program data dynamically.  We present
+and locality properties of program data.  We present
 a type system with permissions and coherence properties for reasoning about logical regions.  
@@
 that all memory accesses in Legion are safe.  Furthermore, we prove that
-type soundness directly enables a scalable, hierarchical
-scheduling algorithm for Legion programs.  We implement a type checker
-for Legion programs and improve performance up to 71\% by eliding
+type soundness enables a scalable, hierarchical
+scheduling algorithm for Legion programs.  We show that an implementation of the type checker
+for Legion programs improves performance up to 71\% by eliding
 memory checks the type system proves safe.  We also demonstrate that an implementation of the Legion 

### 2012-07-06 doc/popl2013/intro.tex
intro

@@
 can be further decomposed into several levels.   In this paper, we present the operational and static
-semantics of Legion, a programming language targeted at providing an
+semantics of Legion, a programming model targeted at providing an
 appropriate level of abstraction for programming such machines, one
 that is both sufficiently high-level to be portable while still
-exposing aspects that are crucial to performance. A very brief summary
-of Legion is that data is organized in a hierarchy of {\em regions}
+exposing aspects that are crucial to performance.
+
+Briefly, in Legion data is organized in a hierarchy of {\em regions}
 and subregions and computation is organized in a hierarchy of {\em
@@
 through a static system of {\em region permissions} specifying the
-operations a task may perform on a region argument (e.g., read or
-write, among others) and novel {\em region coherence} annotations that
-expresses whether other tasks may perform operations concurrently on
-the region (e.g., {\em exclusive} access for task $t$ guarantees no
-other task's updates to a region are visible to $t$).  We prove the
+operations a task may perform on a region argument (read,
+write, or reduce) and novel {\em region coherence} annotations that
+express whether other tasks may perform operations concurrently on
+the region (exclusive, atomic, or simultaneous).  We prove the
 soundness of the permissions and coherence system and use these two
@@
 task scheduling in a Legion implementation, which is crucial for
-performance; any centralized scheduler would be a severe bottleneck
+performance; a centralized scheduler would be a bottleneck
 because of the huge latencies involved in moving data and computation
@@
 % Add float={t} to the list of options if you want the listing at the top
-\begin{lstlisting}[label={lst:linkedlist},caption={Linked List Example}]
+\begin{lstlisting}[float={t},label={lst:linkedlist},caption={Linked List Example}]
 type list<r> = < int, list<r>@r >
@@
 coloring} using the task {\tt color\_list} (line 6), which traverses the list
-and for each list element $e$ in the subregion (called {\tt rs} in
+and for each list element in the subregion (region {\tt rs} in
 {\tt color\_region}) either colors the element 1 or 2 (colors are
@@
 When used in a {\tt partition} expression the coloring defines how to
-populate new subregtions of the partitioned region.  On line 22, the
+populate new subregions of the partitioned region.  On line 22, the
 list elements in {\tt rls} are partitioned into two subregions {\tt
@@
 styles of computation: {\em allocate-then-partition}, in which a
-global collection of data is created an then divided for
+global collection of data is created and then subdivided for
 (usually parallel) processing, and {\em partition-then-allocate}, in
@@
 satisfy these constraints.  A key design decision in Legion, one that turns out to be
-central to all the soundness results as well as the ability to perform distributed, parallel
-scheduling of tasks, is that permissions and coherence annotations (which we have not yet discussed)
+central to the soundness results as well as the ability to perform distributed, parallel
+scheduling of tasks, is that permissions and coherence annotations 
 are not captured by region relationships.  When a region is stored in the heap it has no
@@
 may be reordered to appear to execute atomically before or after the task.  The task {\tt hist\_tree2} achieves
-a similar affect in a finer-grain way by accessing {\tt rh} only with {\tt reduce} privileges, which allows reductions
+a similar effect in a finer-grain way by accessing {\tt rh} only with {\tt reduce} privileges, which allows reductions
 with the same associative-commutative operator to be reordered.

### 2012-07-06 doc/popl2013/langdef.tex
intro
cleanup

--- 2012-07-05 doc/popl2013/langdef.tex
@@
 
-\pagebreak
-
-\section{Legion Core Language}
+\section{Legion}
 \label{sec:legioncore}
 
-\subsection{Language Definition}
-\label{subsec:langdef}
-
-Figure~\ref{fig:langdef} defines the core subset of Legion application language.
-Legion includes basic boolean, integer and tuple types.  Pointers are included as well, but in
-addition to specifying the type of value it points to, a pointer's type also constrains where in
-memory it can point (i.e. only into regions listed in the type).  A special type exists for
-{\em colorings}, which are used to specify how a region should be partitioned into subregions.
-Functions in Legion are named, accept one or more arguments and return a value of some type.
-A function also declares what region access privileges and coherence it requires.  Function types
-are universally quantified over all logical region bindings, with all argument and result types,
-as well as the privileges and coherence requirements being expressed relative to the quantified
-region names.
-The final additional value type that Legion uses is a region relation.  Although logical regions
-cannot be stored directly in variables, they can be captured as part of a region relation 
-instance, which stores a value and one or more logical regions that satisfy a particular set of
-constraints.  An instance of a region relation can be read out of the heap and unpacked, giving
-new local names to the logical regions contained in the instance and tracking the constraints
-between those new logical regions.
-
-\subsection{Legion Core Type System}
-
-The Legion Core language is explicitly typed, using judgments of the following form:
-
+Figure~\ref{fig:langdef} defines the core subset of Legion.  Types
+include booleans, integers, tuples, and pointers annotated with a list
+of regions into which they may point.  There is a special type for
+{\em colorings}, which are used to specify partitions of regions into
+subregions.  Functions/tasks in Legion are named, accept one or more
+arguments and return a value of some type.  A function also declares
+what region access privileges and coherence it requires.  Function
+types are universally quantified over all logical region bindings,
+with all argument and result types, as well as the privileges and
+coherence requirements being expressed relative to the quantified
+region names.  The final additional value type that Legion uses is a
+region relation, which stores a value and one or more logical regions
+satisfying associated region constraints.  An instance of a region
+relation can be read out of the heap and unpacked, giving new local
+names to the logical regions contained in the instance.
+% and tracking
+%the constraints between those new logical regions.
+
+\subsection{Core Type System}
+
+Core Legion is explicitly typed using judgments of the form
 \begin{center}
@@
 \end{center}
-
-where:
+where
 \begin{itemize}
@@
 \end{itemize}
-
-As regions are hierarchical, privileges and/or constraints on a region will often imply similar
-privileges or constraints on subregions.  Figure~\ref{fig:closure} defines transitive closure
-operations for $\Phi^*$ and $\Omega^*$ that will be used in some of the type checking rules.
-
-The full set of type checking rules are in Figure~\ref{fig:types}.  Many of the rules are 
+As regions are hierarchical, privileges and/or constraints on a region often imply similar
+privileges or constraints on subregions.  Figure~\ref{fig:closure} defines closure
+operations for $\Phi^*$ and $\Omega^*$ that add implied constraints.
+The type checking rules are in Figure~\ref{fig:types}.  Many of the rules are 
 straightforward, but several deserve more explanation.
 
-The Legion core language has no subtyping, even for pointers into regions that are part of a 
+Core Legion has no subtyping, even for pointers into regions that are part of a 
 known region hierarchy.  The region ``scope'' of a pointer can only be changed via explicit
@@
 Due to space constraints, we limit the discussion to the general strategy of the proof.  The full
-proof can be found here\cite{xyz}.  If $v$ and $E$ are the result of a permissible evaluation of
+proof can be found here\cite{LegionTypes12}.  If $v$ and $E$ are the result of a permissible evaluation of
 $e$, then a proof exists of:
@@
 
-The proofs of these lemmas can also be found here\cite{xyz}.  The proof of the theorem is organized
+The proofs of these lemmas can also be found here\cite{LegionTypes12}.  The proof of the theorem is organized
 into six pieces, one for each of the three consistency properties needed for subexpressions and one

### 2012-07-06 doc/popl2013/intro.tex
Moving one of the constraints in listing 2

@@
                        < list<rl>@rls, < tree<rl,rll,rtl>@rtl, tree<rl,rlr,rtr>@rtr > >
-     		       where rll #$\le$# rls and rlr #$\le$# rls and rll #$*$# rlr and 
-                                  rtl #$\le$# rt and rtr #$\le$# rt and rtl #$*$# rtr and rls #$\le$# rl
+     		       where rll #$\le$# rls and rlr #$\le$# rls and rll #$*$# rlr and rls #$\le$# rl and
+                                  rtl #$\le$# rt and rtr #$\le$# rt and rtl #$*$# rtr
 

### 2012-07-06 doc/popl2013/intro.tex
updates

@@
 recently, multiple accelerators (usually GPUs) per node, which 
-can be further decomposed into several levels.   In this paper, we present the operational and static
+can themselves be decomposed into several levels.   We present the operational and static
 semantics of Legion, a programming model targeted at providing an
@@
 tasks} and subtasks operating on regions.  Regions and tasks interact
-through a static system of {\em region permissions} specifying the
+through a static system of {\em region permissions} specifying 
 operations a task may perform on a region argument (read,
 write, or reduce) and novel {\em region coherence} annotations that
-express whether other tasks may perform operations concurrently on
+express what other tasks may do concurrently with
 the region (exclusive, atomic, or simultaneous).  We prove the
@@
 
-Line 1 defines a {\em region relationship} {\tt rr[rll, rlr, rtl, rtr]}.  
-Region relationships are bounded existential types.  The
+Line 1 defines a {\em region relationship}, which is a bounded existential type.  The
 {\tt tree} region relationship defines the type of a binary tree

### 2012-07-06 doc/popl2013/langdef.tex
updates

@@
 Figure~\ref{fig:langdef} defines the core subset of Legion.  Types
-include booleans, integers, tuples, and pointers annotated with a list
+include booleans, integers, tuples, and pointers, which are annotated with a list
 of regions into which they may point.  There is a special type for
@@
 arguments and return a value of some type.  A function also declares
-what region access privileges and coherence it requires.  Function
+the region access privileges and coherence it requires.  Function
 types are universally quantified over all logical region bindings,
@@
 to their types,
-\item $\Phi$ is the region access privileges under which the expression will be evaluated, and
+\item $\Phi$ is the region access privileges under which the expression is evaluated, and
 \item $\Omega$ is the set of constraints that must hold between regions
@@
 
-Core Legion has no subtyping, even for pointers into regions that are part of a 
-known region hierarchy.  The region ``scope'' of a pointer can only be changed via explicit
-{\em upregion} and {\em downregion} operations, which check that the ``from'' and ``to'' regions
-are in the right relationship.  It would have been possible to safely infer {\em upregion} calls,
-but {\em downregion} calls must be accompanied by a run-time check that the pointer really does
-point into the subregion(s), so {\em upregion} was made explicit as well for simplicity.
+Core Legion has no subtyping, but a pointer's region can be changed via explicit
+{\em upregion} and {\em downregion} operations.  {\em Upregion} calls can be inferred,
+but {\em downregion} must be accompanied by a run-time check that the pointer really does
+point into the subregion(s), so {\em upregion} is made explicit as well for simplicity.
 
 Accesses to heap memory are performed via the {\em read}, {\em write}, and {\em reduce}
-operations, which are only legal if the region(s) in a pointers type are ones for which the
-enclosing function has privileges to access.
-
-Similarly, a call to another function is only permitted if the privileges required by the 
+operations, which are only legal if the regions in a pointer's type are ones for which the
+enclosing function has privileges.
+Similarly, a call to another function is permitted only if the privileges required by the 
 callee are a subset of those held by the caller.  There is no way to create privileges within
-a Legion function - it is limited to those given to it by its caller.
-
-The {\em partition} operation uses a {\em coloring} to split a region into one or more
-disjoint subregions.  The subregions are given names within the body of the {\em partition},
-as well as constraints that assert that each subregion is contained by the parent region and
-the pairwise disjointness of the subregions.  These constraints can be used locally, or packed
-into region relation instances that can be unpacked and used later.
-
-The {\em pack} expression packs together a value and the corresponding set of logical regions,
-as long as the constraints in the region relationship type can be shown to hold.  The
-corresponding {\em unpack} expression gives (new) names to those regions.  Neither the 
-{\em partition} nor the {\em unpack} expression allows aliasing of region names, simplifying the
-handling of existing variables and region names.
+a Legion function - it is limited to privileges given to it by its caller.
+
+The {\em partition} operation uses a {\em coloring} to split a region into 
+disjoint subregions.  The subregions are given names within the body of the partition,
+as well as constraints asserting subregions are pairwise disjoint and contained within the parent region.
+%These constraints can be used locally, or packed into region relation instances that can be unpacked and used later.
+
+The {\em pack} expression packs together a value and the corresponding logical regions,
+provided the region relation's constraints hold.  The
+corresponding {\em unpack} expression gives (new) names to those regions.  For simplicitly, neither the 
+{\em partition} nor the {\em unpack} expression allows shadowing of region names.
 
@@
 r_i \leq r_j \in \Omega^* \wedge reduces_{id}(r_j) \in \Phi^* \Rightarrow reduces_{id}(r_i) \in \Phi^* \\
-reads(r) \in \Phi^* \wedge writes(r) \in \Phi^* \wedge \\
-\hspace{1cm} \Gamma(id) = \forall \otok{k}{r}. (\oton{T}), \Phi', Q' \rightarrow T_r \Rightarrow reduces_{id}(r) \in \Phi^*
+reads(r) \in \Phi^* \wedge writes(r) \in \Phi^* \wedge id \in \mbox{\em functions}(\Gamma) \Rightarrow reduces_{id}(r) \in \Phi^*
 \end{array}$
@@
 
-Operational semantics rules have the following form:
+Core Legion's operational semantics rules have the following form:
 

### 2012-07-06 doc/popl2013/langdef.tex
figure

@@
 {\small
-\begin{tabular}{cclr}
-
-$T$ & ::= &  & types \\
-  &$\mid$& bool $\;\;\;\mid\;\;\;$ int & base types \\
-  &$\mid$& $\langle T_1, T_2 \rangle$ & tuple \\
-  &$\mid$& $T@(\oton{r})$ & pointer \\
-  &$\mid$& $\text{coloring}(r)$ & region coloring \\
-  &$\mid$& $\exists \oton{r}. T\text{ where }\Omega$ & region relationship \\
-  &$\mid$& $\forall \oton{r}. (\oton{T}), \Phi, Q \rightarrow T_r$ & functions \\
-\\
-$\Omega$ & ::= & $\{ \oton{\omega} \}$ & region constraints \\
-$\omega$ & ::= & $r_1 \leq r_2$ & subregion \\
-  &$\mid$& $r_1 * r_2$ & disjointness \\
-\\
-$\Phi$ & ::= & $\{ \oton{\phi} \}$ & privileges \\
-$\phi$ & ::= & reads$(r)$ $\;\;\;\mid\;\;\;$ writes$(r)$ $\;\;\;\mid\;\;\;$ $\text{reduces}_{id}(r)$ & \\
-\\
-$Q$ & ::= & $\{ \oton{q} \}$ & coherence modes \\
-$q$ & ::= & atomic$(r)$ $\;\;\;\mid\;\;\;$ simult$(r)$ & \\
-\\
-$v$ & ::= & & values \\
-  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & base values \\
-  &$\mid$& $\langle v_1, v_2 \rangle$ & tuple \\
-  &$\mid$& null $\;\;\;\mid\;\;\;$ $a$ & address \\
-  &$\mid$& $\{ (a_i, iv), \ldots \}$ & coloring \\
-  &$\mid$& $\langle \langle \oton{\rho}, v\rangle \rangle$ & region relationship instance \\
-\\
-$bv$ & ::= & false $\;\;\;\mid\;\;\;$ true \\
-\\
-$iv$ & ::= & 0 $\;\;\;\mid\;\;\;$ 1 $\ldots$ \\
-\\
-$e$ & ::= & & expressions \\
-  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & constants \\
-  &$\mid$& $\langle e_1, e_2 \rangle$ $\;\;\;\mid\;\;\;$ $e$.1 $\;\;\;\mid\;\;\;$ $e$.2 & tuple \\
-  &$\mid$& $id$ &  \\
-  &$\mid$& $\text{new}\ T@r$ $\;\;\;\mid\;\;\;$ $\text{null }T@r$ $\;\;\;\mid\;\;\;$ $\text{isnull}(e)$ & \\
-  &$\mid$& $\text{upregion}(e, r_1,\ldots,r_n)$ $\;\;\;\mid\;\;\;$ $\text{downregion}(e, r_1,\ldots,r_n)$ & \\
-  &$\mid$& $\text{read}(e_1)$ $\;\;\;\mid\;\;\;$ $\text{write}(e_1, e_2)$ $\;\;\;\mid\;\;\;$ $\text{reduce}(id, e_1, e_2)$ & memory access \\
-  &$\mid$& $\text{newcolor}\ r$ $\;\;\;\mid\;\;\;$ $\text{color}(e_1, e_2, e_3)$ & coloring operations \\
-  &$\mid$& $e_1 + e_2$ & integer operations \\
-  &$\mid$& $e_1 < e_2$ & comparison operations \\
+\begin{tabular}{cclr|cclr}
+
+$T$ & ::= &  & types & $bv$ & ::= & false $\;\;\;\mid\;\;\;$ true & \\
+  &$\mid$& bool $\;\;\;\mid\;\;\;$ int & base types & & & & \\
+  &$\mid$& $\langle T_1, T_2 \rangle$ & tuple & $iv$ & ::= & 0 $\;\;\;\mid\;\;\;$ 1 $\ldots$ & \\
+  &$\mid$& $T@(\oton{r})$ & pointer & & & & \\
+  &$\mid$& $\text{coloring}(r)$ & region coloring & $e$ & ::= & & expressions \\
+  &$\mid$& $\exists \oton{r}. T\text{ where }\Omega$ & region relationship &   &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & constants \\
+  &$\mid$& $\forall \oton{r}. (\oton{T}), \Phi, Q \rightarrow T_r$ & functions &   &$\mid$& $\langle e_1, e_2 \rangle$ $\;\;\;\mid\;\;\;$ $e$.1 $\;\;\;\mid\;\;\;$ $e$.2 & tuple \\
+& & & &   &$\mid$& $id$ &  \\
+$\Omega$ & ::= & $\{ \oton{\omega} \}$ & region constraints &   &$\mid$& $\text{new}\ T@r$ $\;\;\;\mid\;\;\;$ $\text{null }T@r$ $\;\;\;\mid\;\;\;$ $\text{isnull}(e)$ & \\
+$\omega$ & ::= & $r_1 \leq r_2$ & subregion   &   
+  &$\mid$& $\text{upregion}(e, r_1,\ldots,r_n)$ & \\
+  &$\mid$& $r_1 * r_2$ & disjointness &   
+  &$\mid$ & $\text{downregion}(e, r_1,\ldots,r_n)$ & \\  
+& & &  & 
+  &$\mid$& $\text{read}(e_1)$ $\;\mid\;$ $\text{write}(e_1, e_2)$ & memory access \\
+$\Phi$ & ::= & $\{ \oton{\phi} \}$ & privileges & 
+  & $\mid$ & $\text{reduce}(id, e_1, e_2)$ & \\
+$\phi$ & ::= & reads$(r)$ $\;\mid\;$ writes$(r)$ $\;\mid\;$ $\text{reduces}_{id}(r)$ &  &  
+  &$\mid$& $\text{newcolor}\ r$ $\;\mid\;$ $\text{color}(e_1, e_2, e_3)$ & coloring \\
+& & & &  
+  &   $\mid$& $e_1 + e_2$ & integer ops \\
+$Q$ & ::= & $\{ \oton{q} \}$ & coherence modes  &   
+       & $\mid$& $e_1 < e_2$ & comparisons \\
+$q$ & ::= & atomic$(r)$ $\;\;\;\mid\;\;\;$ simult$(r)$ & &   
   &$\mid$& $\text{let}\ id : T = e_1 \text{in}\ e_2$ &  \\
-  &$\mid$& $\text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3$ &  \\
-  &$\mid$& $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & task calls \\
-  &$\mid$& $\text{partition}\ r_p\text{ using }e_1\text{ as }\oton{r}\text{ in }\ e_2$ &  \\
-  &$\mid$& $\text{pack}\ e_1\ \text{as}\ T[r_1,\ldots,r_n]$ &  \\
-  &$\mid$& $\text{unpack}\ e_1\ \text{as}\ id : T[r_1,\ldots,r_n]\ \text{in}\ e_2$ &  \\
-
+& & & &   
+       & $\mid$& $\text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3$ &  \\
+$v$ & ::= & & values &   
+       & $\mid$& $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & task calls \\
+  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & base values &   
+       & $\mid$ & $\text{partition}\ r_p\text{ using }e_1\text{ as }\oton{r}\text{ in }\ e_2$ &  \\
+  &$\mid$& $\langle v_1, v_2 \rangle$ & tuple & 
+       & $\mid$& $\text{pack}\ e_1\ \text{as}\ T[r_1,\ldots,r_n]$ &  \\
+  &$\mid$& null $\;\;\;\mid\;\;\;$ $a$ & address & 
+       & $\mid$& $\text{unpack}\ e_1\ \text{as}\ id : T[r_1,\ldots,r_n]\ \text{in}\ e_2$ &  \\
+  &$\mid$& $\{ (a_i, iv), \ldots \}$ & coloring & 
+    & & & \\
+  &$\mid$& $\langle \langle \oton{\rho}, v\rangle \rangle$ & reg. relation instance & 
+   & & & \\
+%$bv$ & ::= & false $\;\;\;\mid\;\;\;$ true \\
+%\\
+%$iv$ & ::= & 0 $\;\;\;\mid\;\;\;$ 1 $\ldots$ \\
+%\\
+%$e$ & ::= & & expressions \\
+%  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & constants \\
+%  &$\mid$& $\langle e_1, e_2 \rangle$ $\;\;\;\mid\;\;\;$ $e$.1 $\;\;\;\mid\;\;\;$ $e$.2 & tuple \\
+%  &$\mid$& $id$ &  \\
+%  &$\mid$& $\text{new}\ T@r$ $\;\;\;\mid\;\;\;$ $\text{null }T@r$ $\;\;\;\mid\;\;\;$ $\text{isnull}(e)$ & \\
+%  &$\mid$& $\text{upregion}(e, r_1,\ldots,r_n)$ $\;\;\;\mid\;\;\;$ $\text{downregion}(e, r_1,\ldots,r_n)$ & \\
+%  &$\mid$& $\text{read}(e_1)$ $\;\;\;\mid\;\;\;$ $\text{write}(e_1, e_2)$ $\;\;\;\mid\;\;\;$ $\text{reduce}(id, e_1, e_2)$ & memory access \\
+%  &$\mid$& $\text{newcolor}\ r$ $\;\;\;\mid\;\;\;$ $\text{color}(e_1, e_2, e_3)$ & coloring operations \\
+%  &$\mid$& $e_1 + e_2$ & integer operations \\
+%  &$\mid$& $e_1 < e_2$ & comparison operations \\
+%  &$\mid$& $\text{let}\ id : T = e_1 \text{in}\ e_2$ &  \\
+%  &$\mid$& $\text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3$ &  \\
+%  &$\mid$& $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & task calls \\
+%  &$\mid$& $\text{partition}\ r_p\text{ using }e_1\text{ as }\oton{r}\text{ in }\ e_2$ &  \\
+%  &$\mid$& $\text{pack}\ e_1\ \text{as}\ T[r_1,\ldots,r_n]$ &  \\
+%  &$\mid$& $\text{unpack}\ e_1\ \text{as}\ id : T[r_1,\ldots,r_n]\ \text{in}\ e_2$ &  \\
 \end{tabular}

### 2012-07-07 doc/popl2013/langdef.tex
more updates

@@
 Figure~\ref{fig:langdef} defines the core subset of Legion.  Types
-include booleans, integers, tuples, and pointers, which are annotated with a list
-of regions into which they may point.  There is a special type for
-{\em colorings}, which are used to specify partitions of regions into
-subregions.  Functions/tasks in Legion are named, accept one or more
-arguments and return a value of some type.  A function also declares
-the region access privileges and coherence it requires.  Function
-types are universally quantified over all logical region bindings,
-with all argument and result types, as well as the privileges and
-coherence requirements being expressed relative to the quantified
-region names.  The final additional value type that Legion uses is a
-region relation, which stores a value and one or more logical regions
-satisfying associated region constraints.  An instance of a region
-relation can be read out of the heap and unpacked, giving new local
-names to the logical regions contained in the instance.
+include booleans, integers, tuples, and pointers.  Note that pointers
+are annotated with a list of regions into which they may point; in a
+well-typed program, the pointer is guaranteed to point to at least one
+of those regions. There is a special type for {\em colorings}, which
+are used to specify partitions of regions into subregions.
+Functions/tasks in Legion are named, accept one or more arguments and
+return a value of some type.  A function also declares the region
+access privileges and coherence it requires.  Function types are
+universally quantified over all region names appearing in the type.
+The final value is a region relation, which stores a value and one or
+more logical regions satisfying associated region constraints.  An
+instance of a region relation can be read out of the heap and
+unpacked, giving new local names to the logical regions contained in
+the instance.
 % and tracking
@@
 
-Core Legion has no subtyping, but a pointer's region can be changed via explicit
+Core Legion has no subtyping. A pointer's region is changed only via explicit
 {\em upregion} and {\em downregion} operations.  {\em Upregion} calls can be inferred,
@@
 r_i \leq r_j \in \Omega^* \wedge reduces_{id}(r_j) \in \Phi^* \Rightarrow reduces_{id}(r_i) \in \Phi^* \\
-reads(r) \in \Phi^* \wedge writes(r) \in \Phi^* \wedge id \in \mbox{\em functions}(\Gamma) \Rightarrow reduces_{id}(r) \in \Phi^*
+reads(r) \in \Phi^* \wedge writes(r) \in \Phi^* \Rightarrow reduces_{id}(r) \in \Phi^* \\
+\ \ \ \mbox{for every function identifier $id$}
 \end{array}$
@@
 \begin{itemize}
-\item $M$ is the mapping from logical regions $r_i$ in types to physical regions $\rho_i$, which are sets of memory locations.
-
-It will be useful to apply this logical-to-physical mapping to other
-things that defined in terms of logical regions to yield a corresponding definition in terms of
-physical regions.  The notation $\hat x = M \llbracket x \rrbracket$ is used to describe the
-result of replacing all free logical region names $r_i$ in $x$ with the corresponding physical
-region $\rho_i$.  (Note that $M \llbracket x \rrbracket$ is not
-well-formed if it contains a free logical region name $r_j \not\in dom(M)$.)
-\item $L$ is the mapping of local variable names to their values on the stack.  These values are
-immutable, and private to the function - it is not possible to get a pointer to a local value.
-\item $H$ is the heap typing, mapping from memory locations to types.  The heap typing does not
-change during the execution of a legion program.  The Legion type system is strong enough to
-support the use of garbage collection
-to allow the reuse of a memory location (potentially with a different type stored in that
-location) once all existing references to it are dead, but we
-do not consider that further in this paper.
-\item $S$ is the store, mapping from memory locations $l_i$ to the current value stored.
-\item $C$ is the ``clobber set'', a set of memory locations that may be accessed by concurrently executing operations (??? does this need to be split into read/write/reduce?)
-\item $e$ is the expression being evaluated
-\item $v$ is the result of the evaluation
-\item $E$ is a memory trace (i.e. the effects) of the execution, a list of memory operations
-in the order they were performed.  An operation $\epsilon$ is one of the following:
+\item $M$ is a mapping from logical regions $r_i$ to {\em physical regions} $\rho_i$, which are sets of concrete memory locations.
+We write $\hat x = M \llbracket x \rrbracket$ for the application of $M$ to other objects (e.g., types), 
+replacing all free logical region names $r_i$ in $x$ with corresponding physical regions $\rho_i$.  
+(Note $M \llbracket x \rrbracket$ is ill-formed if it contains a free logical region name $r_j \not\in dom(M)$.)
+
+\item $L$ is a mapping of local variables to (immutable) values.
+
+\item $H$ is a {\em heap typing}, assigning an invariant type to every memory location.
+
+\item $S$ is a store, mapping memory locations $l_i$ to values.
+
+\item $C$ is a {\em clobber set}, memory locations that may be accessed by concurrently executing operations.
+
+\item $e$ is an expression.
+
+\item $v$ is the result of evaluatinig $e$.
+
+\item $E$ is a {\em memory trace} of the execution, a list of {\em memory operations}
+in the order performed.  A memory operation $\epsilon$ is either
+$read(l, c, v)$ (a read of location $l$ with coherence mode $c$, returning value $v$),
+$write(l, c, v)$ (a write to location $l$ with coherence mode $c$, storing value $v$),
+or $reduce_{id}(l, c, v)$ (a reduction into location $l$ with coherence mode $c$ using an
+associative function $id$ with value $v$).  Coherence modes are one of ({\em excl}, {\em atomic}, {\em simult}).
+\end{itemize}
+
+Helper functions are used in some rules:
 \begin{itemize}
-\item $read(l, c, v)$, a read of location $l$ with coherence mode $c$, returning value $v$
-\item $write(l, c, v)$, a write to location $l$ with coherence mode $c$, storing value $v$
-\item $reduce_{id}(l, c, v)$, a reduction into location $l$ with coherence mode $c$ using a
-commutative function $id$ with value $v$
+\item $apply(S, E) = S'$ computes the store $S'$ resulting from applying memory trace $E$ to initial store $S$.
+
+\item $mark\_coherence(E, \hat Q)$ records the coherence of operations in trace $E$ according to coherence annotations $Q$.
+
+\item $valid\_interleave(S, C, E', \oton{E})$ is true if $E'$ is a valid interleaving
+of the memory traces $\oton{E}$.
 \end{itemize}
-where coherence modes are one of ({\em excl}, {\em atomic}, {\em simult}).
-
-The following helper functions are used in some of the rules that deal with memory effects:
-\begin{itemize}
-\item $apply(S, E)$ is used to determine the effect on the store of a sequence of memory operations
-\item $mark\_coherence(E, \hat Q)$ marks the memory operations in a subtask's trace to
-constrain how the operations in that trace can be interleaved with other sibling tasks' operations
-\item $valid\_interleave(S, C, E', \oton{E})$ is used to determine if $E'$ is a valid way to
-interleave the memory traces of two or more subexpressions
-\end{itemize}
-Figure~\ref{fig:opsemfns} gives the definitions of {\em apply} and {\em mark\_coherence}.  For
-the {\em valid\_interleave} predicate, we leave the precise definition to a later section.  For this
-section, we need only one of the necessary conditions of a valid interleaving, that a valid
-interleaving is at least an interleaving - it contains all the memory operations from each of the
-constituent traces and no more, and that each trace's operations remain in order in the interleaved
-trace.
-\end{itemize}
-
-There are a number of rules that appear to be nondeterministic, allowing multiple choices for
-valid evaluation.  In reality, only the memory allocation case (i.e. new $T@r$) has a truly
-nondeterministic choice (i.e. which currently-unused memory location of the right type to 
-use).  A particular memory allocation algorithm is likely to be deterministic, but the
-design of such algorithms is unrelated to the subject at hand, and by abstracting the location 
-choice, we show that our results hold for any correct memory allocation algorithm.  (In fact,
-our criteria are even weaker than that.  Our soundness result even holds if the location returned
-by ``new $T@r$'' is one that is already in use, as long as the heap typing of that location is
-correct.)
-
-The remaining cases of apparent nondeterminism are the result of how we handle the semantics of
-concurrent execution.  An operational semantics that explicitly tracks the state of all 
-concurrently executing operations would allow precise specification of the result of each 
-evaluation step, but would be extremely cumbersome.  Instead, we identify the cases in which
-nonlocal effects (i.e. those that result from the concurrent evaluation of other expressions)
-can impact the local evaluation, and make a ``guess'' of what value would result from an 
-operation.  The check that the guess was consistent with the state of the system is deferred until
-the previously nonlocal effects become local (i.e. contained within the evaluation of an enclosing
-expression tree).  If the guess is not consistent, then the attempted proof of the subexpression's
-behavior is flawed and must be reattempted.
-
-In the case of ``read'' operations whose locations are in the clobber set $C$, we allow the local
-evaluation to guess any value as long as it is of the right type.  The consistency of the value
-is eventually checked by the $valid\_interleave$ predicate of some enclosing expression, but it
-may not be the closest enclosing one.  The $valid\_interleave$ test itself allows some (apparent)
-freedom of choice, deferring some of the consistency checks to the corresponding check further up
-the expression tree.  The final case of apparent nondeterminism is in the choice of what initial
-locations to include in the result of a ``newcolor $r$'' expression.  The inclusion of locations
-beyond those which are explicitly requested by the application is necessary for the success of
-subsequent memory allocation within a subregion created from a partitioning using the coloring, but
-that allocation might not occur within the {\em partition} expression's
-body.  Although the correctness of these partitioning guesses cannot be fully verified until the
-entire execution is known, when an inconsistency is encountered (i.e. there isn't a location of
-the right type in the right region for a ``new $T@r$'' expression) it is possible to ``patch'' the 
-proof (by adding a previously unnamed location to the appropriate expanded colorings and giving
-it the right type in $H$) rather than having to start the proof over from the beginning.
-
-Most of the heavy lifting in the Legion core language occurs in the function call expression.
-The expressions for all the function's arguments are evaluated in the caller's context, but a
-new context is assembled for the evaluation of the function's body.  The region mapping used
-in the callee only uses the caller's mapping indirectly - the only logical region names available
-are those over which the called function is quantified.  Similarly, the set of local values is
-replaced with a mapping of formal parameter names to the values of the arguments.  The store
-is updated to account for side-effects of those argument evaluations.  Finally, if the function
-being called operates with relaxed coherence (i.e. either {\em atomic} or {\em simult} instead of
-{\em excl}), the locations with relaxed coherence are added to the clobber set, allowing the
-impact of concurrent modifications of the store to be handled by the caller.  This is done by
-annotating the called function's memory trace with the relaxed coherence on a per-location basis
-and using the {\em valid\_interleave} predicate to control what parallel execution is allowed.
+Figure~\ref{fig:opsemfns} defines {\em apply} and {\em mark\_coherence}.  The precise definition of
+{\em valid\_interleave} is given in Section~\ref{sec:coherence}.  For now, we need only that a valid interleaving
+is at least an interleaving, containing each constituent trace as a subequence and no other operations.
+
+Some rules modeling concurrent execution are locally
+non-deterministic, guessing values that result from
+concurrent tasks.  For example, the rule for {\tt read(e)} returns an
+arbitrary value $v'$ of the correct type if the location being read is
+in the clobber set $C$, the set of locations concurrent tasks may
+clobber (i.e., update).\footnote{Note that this is also the reason for
+including a global heap typing.}  The guess is eventually checked for
+consistency by the $valid\_interleave$ predicate of some enclosing
+expression, at a point where the task that wrote $v'$ is also local to
+the expression being evaluated.
+
+Another prescient operation is {\tt newcolor r}, which guesses initial
+memory locations and their colors for a partition.  These locations
+support partition-then-allocate, ensuring the memory mappping $M$ has
+locations for the new subregions that can be allocated by {\tt new}.
+As is usual, memory allocation is also non-deterministic, choosing any
+location of the correct type in the region.  For simplicity we do not
+require locations to be ``fresh''; our soundness results hold even if
+{\tt new T@r} returns locations already in use, provided the heap
+typing of the location is correct.
+
+The most involved rule in Core Legion's semantics is for task calls.
+Note that the only regions accessible to the callee (through $M'$) are the region arguments.
+Also, if any regions are used with relaxed coherence (either {\em atomic} or {\em simult} instead of
+{\em excl}), those locations are added to the clobber set, allowing the
+impact of concurrent modifications of the store to be handled by the caller.  
+%This is done by
+%annotating the called function's memory trace with the relaxed coherence on a per-location basis
+%and using the {\em valid\_interleave} predicate to control what parallel execution is allowed.
 
@@
 {\begin{array}{l}
-\opsenvx e_1 \mapsto l, E \\
+\opsenvx e \mapsto l, E \\
 S' = \text{apply}(S, E) \\
@@
 \end{array}}
-{\opsenvx \text{read}(e_1) \mapsto v, E \dplus [ read(l, excl, v) ]}
+{\opsenvx \text{read}(e) \mapsto v, E \dplus [ read(l, excl, v) ]}
 \finfrule
@@
 
-An important property of the legion type system is that any expression that is well-typed is
+A key property of the Legion type system is that any expression that is well-typed is
 guaranteed to access the heap only in ways permitted by the privileges under which it was typed.
-A judgement $M \vdash E : \Phi$ holds if all the memory operations in $E$ are of the types and
+A judgement $M \vdash E : \Phi$ holds if all the memory operations in memory trace $E$ are of the types and
 to locations covered by the privileges in $\Phi$:
-
-\begin{center}
-\begin{tabular}{l@{ }l@{ }l}
-$M \vdash E : \Phi$ & $\Leftrightarrow$ & $\forall \epsilon\text { in } E.$ \\
-& & $(\epsilon = read(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reads(r) \in \Phi)\ \wedge$ \\
-& & $(\epsilon = write(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge writes(r) \in \Phi)\ \wedge$ \\
-& & $(\epsilon = reduce_{id}(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reduces_{id}(r) \in \Phi)$
-\end{tabular}
-\end{center}
-
-The proof that $M \vdash E : \Phi$ holds for all well-typed Legion expressions is achieved by
-induction over the structure of the expression.  Soundness is contingent on the following
-consistent properties, which must be shown to hold for each subexpression:
+\[ 
+\begin{array}{rl}
+M \vdash & E : \Phi \Leftrightarrow \forall \epsilon \in E. \\
+& (\epsilon = read(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reads(r) \in \Phi)\ \wedge \\
+& (\epsilon = write(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge writes(r) \in \Phi)\ \wedge \\
+& (\epsilon = reduce_{id}(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reduces_{id}(r) \in \Phi)
+\end{array}
+\]
+
+Our goal is to show $M \vdash E : \Phi$ holds for the execution of an expression $e$ with static permissions $\Phi$.
+As usual, even stating such a claim requires that corresponding parts of the initial static and dynamic environments be consistent.
+For our results, three consistency properties are needed:
 \begin{itemize}
-\item $M \sim \Omega$, a region mapping must satisfy the region constraints under which the expression was typed
-\item $L \sim_H \Gamma$, the local values must have types consistent with the environment under which the expression was typed
-\item $S \sim H$, the initial state of the store must contain values whose types are consisten with the heap typing
+\item $M \sim \Omega$, a region mapping satisfies the region constraints under which the expression was typed
+\item $L \sim_H \Gamma$, local values have types consistent with the environment under which the expression was typed
+\item $S \sim H$, the initial state of the store contains values whose types are consisten with the heap typing
 \end{itemize}
 
-The proof requires two additional properties be proved for each subexpression:
+Two additional properties are proven for each subexpression:
 \begin{itemize}
-\item $v \sim_H M \llbracket T \rrbracket$, a subexpresion must evaluate to a value of the right type
-\item $E \sim H$, the values written to the store must match the heap types of the location to which they're stored
+\item $v \sim_H M \llbracket T \rrbracket$, the expression yields a value of the right type
+\item $E \sim H$, values written to the store match the heap types of their locations
 \end{itemize}
@@
 \begin{thm}
-Let $e$ be a well-typed Legion expression.  Let $M$ be a region mapping, $L$ be local value mapping,
-$H$ be a heap typing, and $S$ be an initial state of the store satisfying the consistency properties:
-$$
-\begin{array}{l}
-M \sim \Omega \\
-L \sim_H \Gamma \\
-S \sim H
-\end{array}
-$$
-If $v$ and $E$ are the value and memory trace resulting from a permissable terminating evaluation of
-$e$, then:
-$$
-\begin{array}{l}
-v \sim_H M \llbracket T \rrbracket \\
-E \sim H \\
-M \vdash E : \Phi
-\end{array}
-$$
+\label{thm:effects}
+\rm
+If $\typeenv{e}{T}$ and $\opsenv{e}{v}{E}$  and $M \sim \Omega$, $L \sim_H \Gamma$ and $S \sim H$,
+then
+$v \sim_H M \llbracket T \rrbracket$, $E \sim H$ and $M \vdash E : \Phi$.
 \end{thm}
 
-Due to space constraints, we limit the discussion to the general strategy of the proof.  The full
-proof can be found here\cite{LegionTypes12}.  If $v$ and $E$ are the result of a permissible evaluation of
-$e$, then a proof exists of:
-$$\opsenv{e}{v}{E}$$
-The soundness of the result is shown by induction over the structure of this proof.  For each of 
+We spare the reader the lengthy proof, which can be found in
+\cite{LegionTypes12}, and merely outline the general strategy.  The
+proof is a standard induction on the structure of the derivation in
+the operational semantics.  The proof is organized into six pieces,
+one for each of the three consistency properties that must be shown
+for the execution environment of subexpressions and one for each of
+the three properties of the execution's results we wish to show.
+
+\texcomment{
+For each of 
 the 23 different forms a Legion expression can have, we assume the soundness of the subexpressions'
@@
 
-The proofs of these lemmas can also be found here\cite{LegionTypes12}.  The proof of the theorem is organized
-into six pieces, one for each of the three consistency properties needed for subexpressions and one
-for each of the three soundness properties we wish to show.
+The proofs of these lemmas can also be found here\cite{LegionTypes12}.  
+}
 
@@
 
-\item Only three expressions have subexpressions that modify $M$ or $\Omega$ - those that don't
-trivially satisfy the region mapping consistency requirement.  For the
-{\em partition} expression, the assumed consistency of the coloring can be shown to preserve
-the region mapping consistency with respect to the constraints.  Similarly, the consistency of 
-a region relation instance guarantees the maintenance of consistency in an {\em unpack}
-expression.  Finally, the body of a called function executes with a different region mapping,
-but with an initially-empty set of constraints, so they are trivially satisfied.
-
-\item Only three expression have subexpressions that modify $L$, $\Gamma$, $M$.  The
-{\em partition} expression only modifies $M$, and the requirement that it not reuse existing
-names ensures that $M \llbracket \Gamma \rrbracket$ does not change.  The value and type of the
-binding created in a {\em let} expression is obviously consistent, while the binding created in
-an {\em unpack} expression is less obviously so, requiring an induction over the type of the 
-unpacked value to show equivalence under the new mapping.  The last case is again the body of a
-called function, which requires the same style of proof as for the {\em unpack} expression to
-be performed for each formal parameter.
-
-\item The heap typing consistency of all stores used in subexpressions follows directly from
-Lemma~\ref{lemma:heapconst:apply}.
-
-\item With the necessary consistency conditions all satified, the inductive hypothesis asserts
-the type consistency of all subexpression result values.  The consistency of the result of 
-the {\em upregion} is guaranteed by the type checking requirement of appropriate subregion
-constraints and the mapping's consistency with those constraints, whereas {\em downregion}
-expression's result is consistent due to the run time check that is performed.  (The requirement
-for appropriate constraints in the {\em downregion} typing rule is primarily for performance
-(i.e. knowing which part of the region tree to examine) and to catch downcasts that will never
-work at runtime.)
-
-The consistency of a {\em read} expression's result is trivial when an address falls in the
-clobber set, and counts on the consistency of the store otherwise.  The consistency of a 
-{\em color} expression's result depends on the pointer subexpression's consistency and the 
-removal of any previous coloring of that location from the coloring set.  The remaining
-interesting cases arise from changes to the mapping $M$ rather than transformations on the
-value $v$.  In the case of {\em partition} and {\em unpack} expressions, the type system's
-requirement that the body subexpression's result not use the regions that were added to the
-mapping allow the changes to the mapping to be ignored.  The last case is, yet again, the body
+\item[$M \sim \Omega$] Three expressions have subexpressions that modify $M$ or $\Omega$ and therefore do not
+trivially satisfy region mapping consistency.  For {\tt partition},
+the consistency of the coloring preserves region mapping consistency
+with respect to the constraints.  For {\tt unpack}, the consistency of
+a region relation instance guarantees consistency of region mapping.
+Finally, the body of a called function with an initially-empty set of
+constraints, which are trivially satisfied.
+
+\item[$L \sim_H \Gamma$] Four expressions have subexpressions that modify $L$, $\Gamma$, or $M$.  For
+{\tt partition}, which only modifies $M$, the requirement that it not reuse existing
+names ensures that $M \llbracket \Gamma \rrbracket$ does not change.  For {\tt let},
+the value and type of the binding is obviously consistent, while the binding created in
+an {\tt unpack} is less obviously so, requiring an induction over the type of the 
+unpacked value to show equivalence under the new mapping.  The last case is the body of a
+called function, which requires the same style of proof as for {\tt unpack} 
+for each formal parameter.
+
+\item[$S \sim H$] The heap typing consistency of all stores used in subexpressions follows directly from the following
+lemma: If $S \sim H$ and $E \sim H$, then $\text{apply}(S, E) \sim H$.
+
+\end{itemize}
+
+With the necessary consistency conditions proven for a subexpression, the conclusion of the theorem follows for
+the subexpression by the inductive hypothesis.  The remaining obligation is to prove the the three parts of the
+conclusion for an entire expression:
+\begin{itemize}
+
+\item[$v \sim_H M \llbracket T \rrbracket$]    The consistency of 
+{\tt upregion} is guaranteed by the type checking requirement of appropriate subregion
+constraints and the mapping's consistency with those constraints, and {\tt downregion}'s
+result is consistent because of the run time check.  
+%(The requirement
+%for appropriate constraints in the {\em downregion} typing rule is primarily for performance
+%(i.e. knowing which part of the region tree to examine) and to catch downcasts that will never
+%work at runtime.) The consistency of a {\tt read}'s result is trivial for an address in the
+clobber set and uses the consistency of the store otherwise.  The consistency of a 
+{\tt color}'s result depends on the pointer subexpression's consistency and the 
+removal of any previous coloring of that location from the coloring set.
+
+The remaining interesting cases arise from changes to the mapping $M$ rather than transformations on the
+value $v$.  In the case of {\tt partition} and {\tt unpack}, the type system
+guarantees that the subexpression's result cannot use the regions that were added to the
+mapping, allowing the changes to the mapping to be ignored.  The last case is, yet again, the body
 of a called function, and the same strategy that was used for the type consistency of the
@@
 
-\item The type consistency of the value in an expression's effect follows from
-Lemmas~\ref{lemma:heapconst:effects1} and~\ref{lemma:heapconst:effects2}.  New effects are
-added by the {\em write} and {\em reduce} expressions, but their values' consistency follows
-directly from the inductive hypothesis on their subexpressions.  Finally, the consistency of the
+\item[$E \sim H$] The type consistency of the value in an expression's effect follows from two lemmas applied to the effects
+of subexpressions:
+if $E_1 \sim H$ and $E_2 \sim H$, then $E_1 \dplus E_2 \sim H$; and 
+if $E_1 \sim H$ and $E_2 \sim H$ and $valid\_interleave(S, C, E', E_1, E_2)$, then $E' \sim H$.
+New effects are added by {\tt write} and {\tt reduce} expressions, but consistency follows
+directly from the induction hypothesis.  Finally, the consistency of the
 values in a called function's memory trace is addressed in the same way as the return value.
 
-\item The proof of the crucial property of containment of effects within the available privileges
-is very similar to the previous step, with Lemmas~\ref{lemma:effsound:effects1}
-and~\ref{lemma:effsound:effects2} covering most cases, straightforward proofs for the {\em read},
-{\em write}, and {\em reduce} expressions, and one final special case for function calls.  The 
-inductive hypothesis asserts that the function body's effects are contained within its own
-privileges.  Those are a subset of the caller's $\Phi^*$, and then Lemma~\ref{lemma:phiclosure}
-is used to complete the proof.
+\item[$M \vdash E : \Phi$] The proof of the crucial property of containment of effects within the available privileges
+is similar in outline to the previous step.  The easy cases are covered by lemmas combining the corresponding results
+for subexpressions: if $M \vdash E_1 : \Phi$ and $M \vdash E_2 : \Phi$, then $M \vdash E_1 \dplus E_2 : \Phi$, and
+if $M \vdash E_1 : \Phi$ and $M \vdash E_2 : \Phi$ and $valid\_interleave(S, C, E', E_1, E_2)$, then $M \vdash E' : \Phi$.
+Straightforward proofs cover {\tt read}, {\tt write}, {\tt reduce}, and one final special case for function calls. 
 \end{itemize}
@@
 \subsection{Coherence}
-
-The soundness proof above guarantees that the evaluation of an expression produces a value of the
-right type, and do so within the regions of memory for which privileges are held, but for the
+\label{sec:coherence}
+
+Theorem~\ref{thm:effects} guarantees expression evaluation produces a value of the
+correct type and does so within the regions of memory for which privileges are held.
+
+ but for the
 Legion programming model to be useful, we need to also guarantee that the evaluation of an expression
@@
 only allowing $simult$ coherence annotations, and can show that subexpression with privileges
-that can be shown to be ``simulataneously noninterfering'' (under mapping $M$) can be executed
+that can be shown to be ``simultaneously noninterfering'' (under mapping $M$) can be executed
 in parallel.  This will include all the cases covered by Theorem~\ref{thm:parallelexec} as well

### 2012-07-07 doc/popl2013/langdef.tex
new sections

@@
 
-\subsection{Soundness of Effects}
+\section{Soundness of Effects}
 
@@
 
-\subsection{Coherence}
+\section{Coherence}
 \label{sec:coherence}
 
-Theorem~\ref{thm:effects} guarantees expression evaluation produces a value of the
+Theorem~\ref{thm:effects} guarantees that a task produces a value of the
 correct type and does so within the regions of memory for which privileges are held.
-
- but for the
-Legion programming model to be useful, we need to also guarantee that the evaluation of an expression
-produces the (or at least, a) correct value.  The key to doing this is to place additional 
-constraints on what it means to be a valid interleaving of two or more memory traces.  In
-addition to a valid interleaving being an interleaving of the individual traces, we will 
-require that it be:
+It does not, however, address conherence, what concurrent operations on the regions are permitted.
+An innovation in Legion is that programmers can control coherence on a per-region basis.  Intuitively,
+the different coherence modes permit different interleavings of memory traces.  We begin by formalizing
+exclusive access, the strictest coherence mode.  Exclusive access requires valid interleavings be:
 \begin{itemize}
-\item {\em coherent} - each memory read operation should see the effect of all writes and reductions 
-that occur earlier in the interleaved trace to the same location, and
-\item {\em sequentially equivalent} - the value read by each memory read operation and the result of applying the interleaved trace to the original store should be the same as if each constituent trace
-was applied to the store in a valid {\em program order}.
+\item {\em coherent} - a read of location $l$ should see the effect of all writes and reductions to $l$
+occuring earlier in the interleaved trace, and
+\item {\em sequentially equivalent} - the value read by each read 
+operation and the result of applying the interleaved trace to the original store should be the same as if each constituent trace
+was applied to the store in a valid sequential execution of the program.
 \end{itemize}
 These requirements are relaxed in some cases due to either a non-empty clobber set (representing
-operations that may occur concurrently as part of evaluating other functions) or due to coherence
-modes other than {\em excl} on operations with the traces (representing operations that may 
-occur concurrently as part of evaluating other parts of the same function).
+operations that may occur concurrently as part of evaluating other functions) or for coherence
+modes other than {\em excl}.
 
@@
 
-\subsection{Hierarchical Scheduling}
+\section{Hierarchical Scheduling}
+\label{sec:scheduling}
 

### 2012-07-07 doc/popl2013/langdef.tex
updates

@@
 correct type and does so within the regions of memory for which privileges are held.
-It does not, however, address conherence, what concurrent operations on the regions are permitted.
+It does not, however, address {\em coherence}, what concurrent operations on the regions are permitted.
 An innovation in Legion is that programmers can control coherence on a per-region basis.  Intuitively,
-the different coherence modes permit different interleavings of memory traces.  We begin by formalizing
-exclusive access, the strictest coherence mode.  Exclusive access requires valid interleavings be:
+the different coherence modes permit different interleavings of memory traces.  For example,
+in addition to be an interleaving, exclusive access requires interleaved memory traces to be:
 \begin{itemize}
@@
 \end{itemize}
-These requirements are relaxed in some cases due to either a non-empty clobber set (representing
+These requirements are relaxed for either a non-empty clobber set (representing
 operations that may occur concurrently as part of evaluating other functions) or for coherence
@@
 
-Figure~\ref{fig:validinterleave} gives the precise definition of the {\em valid\_interleave}
-predicate, but there are a lot of moving parts, so description in ``plain'' English may help.
-Consider a set of constituent traces $\oton{E}$.  Let the {\em exclusive location set} $L_{excl}$ be
-the set of all locations that appear in at least one operation with coherence mode $excl$ in at least
-one constituent trace, excepting any locations in the clobber set $C$.  Let the {\em atomic location
-set} $L_{atomic}$ similarly be the set of locations that appear in at least one operation with 
-coherence mode $atomic$, but no operations with coherence mode $excl$, again excepting any location
-in the clobber set.  The $coherent(S,L_1,L_2,E)$ test checks that all reads of locations in $L_2$ and any read of a location in $L_1$ that comes after a write to the same location return values
-that match what was most recently written into the store.  (The difference betwen $L_1$ and $L_2$
-defers the checking of atomic reads that don't have a preceeding write until any other traces that
-may be ordered before $E$ are included.)  The $seq\_equiv$ test verifies that every read in $E'$ with a location in $L_1$ would also have be coherent if the operations were performed in the original program order and that the store that results from applying $E'$ matches the store that would
-have resulted from applying the constituent traces in program order for all locations in $L_1$.
-For the atomic location set, we allow the constituent traces to be permuted, but all locations
-must be checked with the same permutation.
+Figure~\ref{fig:validinterleave} formally defines {\em
+valid\_interleave}, which we briefly explain.  Consider
+traces $\oton{E}$.  The {\em exclusive location set} $L_{excl}$ is the
+set of locations used in an operation with coherence mode $excl$ in
+some trace $E_i$, excepting any locations in the clobber set $C$.
+Similarly, the {\em atomic location set} $L_{atomic}$ is the set of
+locations in an operation with coherence mode $atomic$, but no
+operations with coherence mode $excl$, again excepting any location in
+the clobber set.  The $coherent(S,L_1,L_2,E)$ test checks that all
+reads of locations in $L_2$, and any read of a location $l \in L_1$
+that comes after a write to $l$, return the value most recently written
+to the store.  (If $L_2 = \emptyset$ initially, atomic reads without a preceding write are not checked, effectively defering
+these checks to point where traces ordered before $E$ are included.)  The $seq\_equiv$
+test verifies every read in $E'$ with a location in $L_1$ 
+is coherent if operations are performed in 
+program order and that the store that results from applying $E'$
+matches the store resulting from applying the
+constituent traces in program order for all locations in $L_1$.  For
+the atomic location set, we allow the constituent traces to be
+permuted, but all locations must be checked with the same permutation.
 
 Flexibility in interleaving operations is the key to achieving parallel execution, but
-hopefully the reader will agree that the exact criteria for creating a valid interleaving is
-much too expensive to implement at runtime, so we now consider whether there are simpler ways to
-generate valid interleavings, and under what conditions those approaches may be used.
-
-The first option available to us is to serialize the constituent traces into the original
-program order.  Although each constituent trace can be assumed to be coherent, that coherence is
-in the context of a clobber set that may differ from the one in which the interleaving must occur.
-As discussed earlier, this difference can result in cases where no valid interleaving exists.
-However, we can show that as long as any valid interleavings exist, the serialization of the
-constituent traces will be one of them.
-
-To prove this result, we will use the following lemma:
+implementing the precise criteria for checking a valid interleaving is
+much too expensive.  We now consider practical mechanisms that generate valid interleavings and 
+under what conditions these approaches may be used.
+
+%
+% FIX ME   I don't understand this paragraph.  I didn't see this discussed
+% earlier; I took that comment out.  I don't know what the pronoun ``the one in which''
+% refers to.  The clobber set?  The interleaving?
+%
+The first option is to serialize the constituent traces in program
+order.  Although each constituent trace can be assumed to be coherent,
+that coherence is in the context of a clobber set that may differ from
+the one in which the interleaving must occur. 
+This difference can result in cases where no valid interleaving
+exists.  
+%
+% FIX ME  Why?
+%
+However, we can show that as long as any valid interleaving
+exists, the serialization of the constituent traces is valid.
+To prove this result, we use the following lemma, the proof
+of which follows directly from the definition of $apply$.
 
@@
 \label{lemma:applychain}
+\rm
 For any initial store $S$ and memory traces $E_1$ and $E_2$,
@@
 
-The proof of this follows directly from the definition of $apply$.
-
 \begin{thm}
 \label{thm:sequential}
-Let $S_0$ be an initial store, $C$ be a clobber set and $\oton{E}$ be memory traces to be
-interleaved.  Let $E' = E_1 \dplus \ldots \dplus E_n$.  If there exists any $E''$ that is a
+\rm
+Let $S_0$ be a store, $C$ a clobber set and $\oton{E}$ memory traces to be
+interleaved.  Let $E' = E_1 \dplus \ldots \dplus E_n$, in program order.  If there exists any $E''$ that is a
 valid interleaving of $\oton{E}$, then $E'$ is also valid interleaving of $\oton{E}$.
@@
 \begin{proof}
-To prove $valid\_interleave(S_0, C, E', \oton{E})$, we need to show the three conditions hold:
+To prove $valid\_interleave(S_0, C, E', \oton{E})$, we need to show the three conditions:
 \begin{enumerate}
-\item $any\_interleave(E', \oton{E})$ is intuitively obvious, and can be proven by induction over 
+\item $any\_interleave(E', \oton{E})$ is proven by induction on
 the length of $E_1$, followed by $E_2$, and so on.
 \item $coherent(S_0, C, E')$ is shown by considering each constituent trace's subsequence in $E'$
-in order, using Lemma~\ref{lemma:applychain} along with the coherence of each $E_i$.
-\item $seq\_equiv(S_0, C, E', \oton{E})$ is trivial, as each leading subsequence of $E'$ corresponds
-to the appropriate pieces of $\otok{i}{E}$.
-\item $coherent(S_0, C, E')$ can be shown from the sequential equivalence of $E''$.  The value 
-returned by each read in $E''$ has been shown to be equal to the value in the store that results
+in order, using Lemma~\ref{lemma:applychain} with the coherence of each $E_i$.
+\item $seq\_equiv(S_0, C, E', \oton{E})$ follows trivially by matching the structure of $E'$
+to the corresponding  pieces $\oton{E}$.
+\item $coherent(S_0, C, E')$ is shown using the sequential equivalence of $E''$.  The value 
+returned by each read in $E''$ is equal to the value in the store that results
 from the application of all previous constituent traces and the operations before the read in its
 own trace.  This is exactly the sequence of preceding operations of the corresponding read in $E'$,
-and Lemma~\ref{lemma:applychain} allows us to trainsform the chain of $apply$ operations into a 
+and Lemma~\ref{lemma:applychain} allows us to transform the chain of $apply$ operations into a 
 single $apply$ operation on the whole sequence, satisfying the coherence condition for the read
-in $E'$.  This property applies for all reads in $E''$, and since both $E'$ and $E''$ are
-interleavings of the same traces, they have exactly the same reads, so we have shown all the reads
-in $E'$ to be coherent.
+in $E'$.  This property applies for all reads in $E''$, and since $E'$ and $E''$ are
+interleavings of the same traces, they have the same reads, so all reads
+in $E'$ are coherent.
 \end{enumerate}
@@
 
-Now that we have a way to construct a valid interleaving, we can look at how to modify that
-interleaving to get interleavings that are still valid, but allow parallel execution.  To do this,
-we consider the criteria under which we may swap the order of two adjacent memory operations in an
-existing valid interleaving.
+Next we examine how to modify the serialized interleaving to produce valid interleavings allowing parallel execution. 
+We consider when we may swap the order of two adjacent memory operations in a valid interleaving.
 \begin{itemize}
-\item First, the two operations must have come from different constituent
-traces, or the swap would violate the $any\_interleave$ criterion.
-\item Second, if one is a read of
-a location (that is not in the clobber set), the other may not be a write or reduce to the same
-location, as changing the order would almost certainly change the value that is read.
-\item Third, if one is a write of a location, the other may not be a write or reduce to the same
-location, as the store that would result from applying those operations in the opposite order would
-show a difference in that location.
+\item The two operations must have come from different constituent
+traces, or the swap would violate $any\_interleave$.
+\item If one is a read of
+a location $l$ (that is not in the clobber set), the other may not be a write or reduce to $l$, 
+as changing the order would almost certainly change the value read.
+\item If one is a write of a location $l$, the other may not be a write or reduce to $l$,
+as the store that would result from applying those operations in the opposite order may
+differ in $l$.
 \item Finally, if both operations are reductions to the same location, they may be swapped if they
-use the same reduction function, as reduction functions are assumed to be commutative.  (We 
-follow the example of other work (i.e. DPJ) and leave proofs of commutativity to the programmer,
-possibly with the aid of other tools.)  However, if they use different reduction functions (by
-name - we make no attempt to compare the operation of functions), commutative cannot be 
-guaranteed and swapping the two operations might result in a different resulting store.
+use the same reduction function, as reduction functions are assumed to commute.  
+%(We 
+%follow the example of other work (i.e. DPJ) and leave proofs of commutativity to the programmer,
+%possibly with the aid of other tools.)  
+However, if the reduction functions have different names, swapping the two operations might result in a different store.
 \end{itemize}
@@
 \label{lem:nonintswap}
+\rm
 Let $S$ be an initial store, $C$ be a clobber set, $\oton{E}, E'_a, E'_b$ be memory traces, and
@@
 \end{lem}
-
+\texcomment{
 \begin{proof}
@@
 \end{proof}
-
+}
+The proof is included in \cite{LegionTypes12}.
 We now extend the noninterference operator to entire memory traces, providing a way to say that
@@
 \label{lem:noninteffects}
+\rm
 Let $S$ be an initial store, $C$ be a clobber set, $\oton{E}$ be memory traces such that $E_i \nonint E_j$
-for every $1 \leq i < j \leq n$.  Then any trace $E'$ satisfying:
-$$any\_interleave(E', \oton{E})$$
+for every $1 \leq i < j \leq n$.  Then any trace $E'$ satisfying $any\_interleave(E', \oton{E})$
 is a valid interleaving.
 \end{lem}
-
+\texcomment{
 \begin{proof}
@@
 \end{proof}
-
-Our next step is to extend noninterference to the privileges used in the type system.  We will make
+}
+The next step is to extend noninterference to privileges.  We make
 use of the region mapping $M$ to translate the logical regions named in privileges into sets of
-locations (i.e. physical regions), and define noninterference between individual privileges as
-follows:
+locations (i.e. physical regions), and define noninterference between individual privileges and
+sets of privileges:
 $$
 \begin{array}{l@{ }l}
-priv_1(r_1) \nonint[M] & priv_2(r_2) \Leftrightarrow (M(r_1) \cap M(r_2) = \emptyset)\ \vee \\
+priv_1(r_1) & \nonint[M] priv_2(r_2) \Leftrightarrow (M(r_1) \cap M(r_2) = \emptyset)\ \vee \\
 & (priv_1 = reads \wedge priv_2 = reads)\ \vee \\
@@
 %% \end{center}
-and noninterference between sets of privileges in the same way as memory traces:
+
 $$\Phi_1 \nonint[M] \Phi_2 \Leftrightarrow \displaystyle\bigwedge_{\phi_1 \in \Phi_1, \phi_2 \in \Phi_2} \phi_1 \nonint[M] \phi_2$$
@@
 \label{lem:nonintpriv}
+\rm
 Let $M$ be a region mapping and $E_1$ and $E_2$ two memory traces such that $M \vdash E_1 : \Phi_1$ and $M \vdash E_2 : \Phi_2$.
@@
 \end{proof}
-
-We are now able to connect the dots and get to the key result that allows the Legion runtime to 
-safely run subtasks in parallel.
+}
+We are now able to state the key result that allows the Legion runtime to safely run subtasks in parallel.
 
@@
 \label{thm:parallelexec}
+\rm
 Let $\oton{e}$ be well-typed Legion expressions, each with its own privileges $\Phi_i$.
-Let $M$ be a region mapping, $L$ be local value mapping,
-$H$ be a heap typing, and $S$ be an initial state of the store satisfying the consistency properties:
-$$
-\begin{array}{l}
-M \sim \Omega \\
-L \sim_H \Gamma \\
-S \sim H
-\end{array}
-$$
+Let $M$ be a region mapping, $L$ a local value mapping,
+$H$ a heap typing, and $S$ be an initial store satisfying
+$M \sim \Omega$, $L \sim_H \Gamma$, and $S \sim H$.
 If $\Phi_i \nonint[M] \Phi_j$ for $1 \leq i < j \leq n$, then any parallel execution of expressions
-$\oton{e}$ will result in a valid interleaving of memory operations.
+$\oton{e}$ results in a valid interleaving of memory operations.
 \end{thm}
 
-The proof follows directly from Lemma~\ref{lem:nonintpriv} and Lemma~\ref{lem:noninteffects}. \\
-
-Noninterference of effects is something that can only be determined after
-evaluation of an expression is completed, and only at great expense (see STM).
-In contrast, the noninterference of privileges can be efficiently determined at
-runtime.  Even though the privileges themselves are static, the
-mapping $M$ can be dynamically determined, allowing noninterference to be 
-judged in many more cases than a purely static analysis can achieve.
-
-Note that this result holds even if the clobber set $C$ is non-empty.  This allows 
-two ``locally independent'' subtasks to run in parallel even if one or both
-of them is interacting (in a programmer-permitted way) with a third subtask.
+The proof follows directly from Lemmas~\ref{lem:noninteffects}
+and \ref{lem:nonintpriv}.  This result holds even if the
+clobber set $C$ is non-empty, allowing two ``locally independent''
+subtasks to run in parallel even if one or both of them is interacting
+(in a programmer-permitted way) with a third subtask.
+
+We are now able to explain an important aspect of the Legion
+implementation that is different from other systems and relies on the
+soundness of privileges.  Noninterference of memory effects is
+something that can only be determined after evaluation of an
+expression is completed, and only at great expense, as illustrated at
+one extreme by work on transactional
+memory \cite{transactionalmemory}.  At the other extreme are systems
+that check privileges statically, such as Jade \cite{} and
+DPJ \cite{}.  In contrast, Legion checks noninterference of privileges
+at runtime, which is much simpler and more efficient than checking
+noninterference of dynamic effects.  Even though the privileges
+themselves are static, the region mapping $M$ is dynamic.  Testing
+noninterference on the privileges with physical regions allows
+parallel execution in many more cases than a purely static analysis
+can achieve.
 
@@
 \label{thm:hiersched}
+\rm
 Let $e_1$ and $e_2$ be well-typed Legion expressions that using privileges $\Phi_1$ and

### 2012-07-07 doc/popl2013/langdef.tex
section 4

@@
 \rm
-Let $S$ be an initial store, $C$ be a clobber set, $\oton{E}, E'_a, E'_b$ be memory traces, and
-$\epsilon_1, \epsilon_2$ be two annotated memory operations such that $\epsilon_1 \nonint \epsilon_2$.
+Let $S$ be a store, $C$ a clobber set, $\oton{E}, E'_a, E'_b$ memory traces, and
+$\epsilon_1, \epsilon_2$ two annotated memory operations such that $\epsilon_1 \nonint \epsilon_2$.
 Then,
-$$valid\_interleave(S, C, E'_a \dplus \epsilon_1 \dplus \epsilon_2 \dplus E'_b, \oton{E})$$
+$valid\_interleave(S, C, E'_a \dplus \epsilon_1 \dplus \epsilon_2 \dplus E'_b, \oton{E})$
 if and only if
-$$valid\_interleave(S, C, E'_a \dplus \epsilon_2 \dplus \epsilon_1 \dplus E'_b, \oton{E}).$$
+$valid\_interleave(S, C, E'_a \dplus \epsilon_2 \dplus \epsilon_1 \dplus E'_b, \oton{E}).$
 \end{lem}
@@
 }
-The next step is to extend noninterference to privileges.  We make
-use of the region mapping $M$ to translate the logical regions named in privileges into sets of
+We next extend noninterference to privileges.  We 
+use the region mapping $M$ to translate the logical regions named in privileges into sets of
 locations (i.e. physical regions), and define noninterference between individual privileges and
@@
 memory \cite{transactionalmemory}.  At the other extreme are systems
-that check privileges statically, such as Jade \cite{} and
-DPJ \cite{}.  In contrast, Legion checks noninterference of privileges
+that check privileges statically, such as Jade \cite{Rinard98} and
+DPJ \cite{Bocchino11}.  In contrast, Legion checks noninterference of privileges
 at runtime, which is much simpler and more efficient than checking
@@
 
-The analysis above did not consider the coherence modes on each operation in the memory traces,
-making the worst-case assumption that all were using the $excl$ mode.  However, if two or more
-subexpressions contain calls to functions that use $atomic$ and/or $simult$ coherence modes
-on some of their regions, options other than serial execution may be guaranteed to 
-result in valid interleavings even if constituent traces have some locations in common.  There
-are two cases of particular interest that we will cover.  In each case, the entirety of the 
-reasoning from the previous section is repeated with minor modifications, so we will limit our
-discussion to a general description of the differences rather than the complete proofs.
+The analysis above does not consider the coherence modes of operations,
+making the worst-case assumption that all operations use $excl$ mode.  However, if two or more
+subexpressions contain calls to tasks using $atomic$ or $simult$ coherence modes
+on some regions, executions besides serial program order may 
+result in valid interleavings even if constituent traces have locations in common.  We discuss
+two cases of interest.  In both  the previous section's
+is repeated with minor modifications, so for brevity we limit
+discussion to a description of the differences rather than complete proofs.
 
 The first important case occurs when all apparently-interfering operations have coherence modes
-of either $atomic$ or $simult$.  In this case, we can see that the conditions required for
-coherence and sequential equivalence are satisfied for all locations in $L_{excl}$.  The
-requirements for locations in $L_{atomic}$ allow for any permutation of the constituent traces,
+of either $atomic$ or $simult$.  In this case, the conditions required for
+coherence and sequential equivalence are trivially satisfied for locations in $L_{excl}$.  The
+requirements for locations in $L_{atomic}$ allow any permutation of the constituent traces,
 so the runtime is free to serialize the execution of the subexpressions in any order, not just
-the original program order.
-
-In order to convert the required analysis to one that can be performed on regions, we modify
+program order.
+
+To convert the required analysis to one that can be performed on regions, we modify
 the noninterference operator for effects to consider only locations in $L_{excl}$ and show that
-traces that a noninterfering under this modified check satisfy the requirement for atomic
+traces that are noninterfering under this modified check satisfy the requirement for atomic
 execution.  We then modify the noninterference operator on privileges to allow 
 $M(r_1) \cap M(r_2) \not= \emptyset$ if either $atomic(r_1)$ or $simult(r_1)$ is in $Q_1$ and
-either $atomic(r_2)$ or $simult(r_2)$ is in $Q_2$, and can show that subexpressions with 
-privileges that can be shown to be ``atomically noninterfering'' (under a dynamic region mapping
-$M$) will always have effects that are compatible with atomically-reordered execution.  Finally,
+either $atomic(r_2)$ or $simult(r_2)$ is in $Q_2$. We then show that subexpressions with 
+privileges that are {\em atomically noninterfering} (under a dynamic region mapping
+$M$) always have effects compatible with atomically-reordered execution.  Finally,
 in the special case where each subexpression has $atomic(r)$ or $simult(r)$ in $Q_i$ for every $r$
@@
 the $simult$ coherence mode.  In this case, all the coherence and sequential equivalence 
-checks are trivially satisfied, and any interleaving of the constituent traces will be valid.
+checks are trivially satisfied, and any interleaving of the constituent traces is valid.
 We again modify the noninterference criteria for effects, traces, and privileges, this time
-only allowing $simult$ coherence annotations, and can show that subexpression with privileges
-that can be shown to be ``simultaneously noninterfering'' (under mapping $M$) can be executed
-in parallel.  This will include all the cases covered by Theorem~\ref{thm:parallelexec} as well
+only allowing $simult$ coherence annotations. We show that subexpression with privileges
+that are {\em simultaneously noninterfering} (under mapping $M$) can be executed
+in parallel, which includes all the cases covered by Theorem~\ref{thm:parallelexec} as well
 as those where the programmer has used the $simult$ coherence mode to explicitly allow parallel

### 2012-07-07 doc/popl2013/langdef.tex
section 5

@@
 information, and must respect all required scheduling dependencies while ideally imposing as
-few extra dependencies as possible.
-
-Let $e_1$ and $e_2$ be well-typed Legion expressions that using privileges $\Phi_1$ and
+few extra dependencies as possible.  The following theorem forms the basis of the 
+Legion scheduler.
+
+\texcomment{
+Let $e_1$ and $e_2$ be well-typed Legion expressions using privileges $\Phi_1$ and
 $\Phi_2$.  Now let $e_3$ be some subexpression of $e_1$ and $e_4$ be some 
 subexpression of $e_2$, with corresponding privileges $\Phi_3$ and $\Phi_4$.
-
 By Theorem~\ref{thm:parallelexec}, if the runtime can verify $\Phi_1 \nonint[M] \Phi_2$, then 
@@
 Lemma~\ref{lem:noninteffects}).
-
+}
 \begin{thm}
@@
 \rm
-Let $e_1$ and $e_2$ be well-typed Legion expressions that using privileges $\Phi_1$ and
-$\Phi_2$.  Now let $e_3$ be some subexpression of $e_1$ and $e_4$ be some 
-subexpression of $e_2$.  For any memory traces $E_3$ and $E_4$ that result from evaluation of
-$e_1$ and $e_2$ (with the usual consistent $M$, $L$, $H$, and $S$), $E_3$ and $E_4$ must be
+Let $e_1$ and $e_2$ be well-typed Legion expressions using privileges $\Phi_1$ and
+$\Phi_2$ respectively .  Let $e_3$ be a subexpression of $e_1$ and $e_4$ be a 
+subexpression of $e_2$.  Any memory traces $E_3$ of $e_3$ and $E_4$ of $e_4$ resulting from evaluation of
+$e_1$ and $e_2$ (with the usual consistent $M$, $L$, $H$, and $S$), must be
 noninterfering.
 \end{thm}
-
+\texcomment{
 \begin{proof}
@@
 \end{proof}
-
-Although a decision that $e_1$ and $e_2$ are noninterfering guarantees that $e_3$ and $e_4$ are
-noninterfering, the converse is not true.  It is possible for $e_1$ and $e_2$ to be judged to be
-potentially interfering and be serialized in their entirety even when the subexpressions could
-have been proven to be noninterfering.  This loss of precision can impact performance, and did
-for a few times during the implementation of our experiments.  When the performance impact is
-significant, two options exist to improve the precision.  In cases where the $e_1$ and $e_2$ are
-in fact noninterfering, the partitioning of regions can be further refined to make the
-disjointness visible in the privileges.  Alternately, and this is the only option when the 
-interference is dynamically determined, the code can be refactored to life the subexpressions
-(or some expression containing them) up to the same level, allowing the dynamic test based on
-the region mapping to infer noninterference in more cases.  Although this can increase the
-overhead of scheduling decisions, judicious use of this technique in practice has not shown
-any measurable performance degradation.
+} The Legion task scheduler uses Theorem~\ref{thm:hiersched} in the
+following way.  Sibling task calls (tasks invoked within the same
+function body) are checked for non-interference of their (dynamic)
+privileges.  If they interfere they are serialized; otherwise they are
+considered for parallel execution.  Although determining that $e_1$
+and $e_2$ are noninterfering guarantees $e_3$ and $e_4$ are
+noninterfering, the converse is not true.  It is possible for $e_1$
+and $e_2$ to be judged to be potentially interfering and be serialized
+even when the subexpressions are noninterfering.  This loss of
+precision can impact performance, and did in a few instances in our
+experiments.  
+%When the performance impact is significant, two options
+%exist to improve matters.  
+In cases where $e_1$ and $e_2$ are in fact
+noninterfering, region partitions can be further refined to make the
+disjointness visible in the privileges.  Alternately, the code can be
+refactored to lift the subexpressions up to the same level, allowing
+the dynamic test based on the region mapping to infer noninterference
+in more cases.  
+%Although this can increase the overhead of scheduling
+%decisions, judicious use of this technique in practice has not shown
+%any measurable performance degradation.

### 2012-07-07 doc/popl2013/evaluation.tex
evaluation

@@
 \begin{itemize}
-\item Expressivity: can real applications be encoded (Section \ref{subsec:expressivity})
-\item Overhead: can it reduce checking costs (Section \ref{subsec:overhead})
-\item Scalability: can it leverage hierarchical scheduling (Section \ref{subsec:scalability})
+\item Expressivity: Can real applications be written (Section \ref{subsec:expressivity})?
+\item Overhead: Does it reduce checking costs (Section \ref{subsec:overhead})?
+\item Scalability: Can we leverage hierarchical scheduling (Section \ref{subsec:scalability})?
 \end{itemize}
-Our evaluation uses a prototype implementation of the Legion language and programming model.
-The prototype consists of two components: an implementation of a type checker
+%Our evaluation uses a prototype implementation of the Legion language and programming model.
+Our prototype implementation has two components: a type checker
 for the language introduced in Section \ref{subsec:langdef} and a C++ runtime library
@@
 expressivity of the Legion type system, we introduce each of these applications
-by describing which features of the Legion type system are used in their implementations.
+by describing features of the Legion type system used in their implementations.
 
@@
 \label{subsec:circuit}
-The Circuit application simulates the wires of a large integrated circuit using an RLC
-model.  The computation consists of three phases for each time step: compute the current in each wire using
-an iterative model, updated the charge in each node, and compute the voltage of each node.
-The primary data structure in the Circuit application is a large irregular graph.  To perform 
-this computation in parallel the graph is dynamically partitioned into pieces.  
+Circuit simulates the wires of a large integrated circuit.
+% using an RLC model.  
+%The computation consists of three phases for each time step: compute the current in each wire using
+%an iterative model, updated the charge in each node, and compute the voltage of each node.
+The primary data structure is a large irregular graph, which is
+dynamically partitioned into pieces to enable parallel simulation.
 Our implementation creates separate regions for the wires and
-nodes in the graph.  The wire and node regions are recursively partitioned into
+nodes,  which are recursively partitioned into
 subregions for pieces of the graph.  The node region is partitioned an additional way to 
@@
 The information about each piece of the graph is stored
-in a region relationship that remembers the disjointness information for each piece
+in a region relationship that remembers the disjointness information of each piece
 from other pieces.  The tasks for the three phases of the computation use different combinations of
@@
 \label{subsec:fluid}
-Our Fluid application is based on the fluidanimate benchmark from the PARSEC benchmark
+Fluid is based on the {\tt fluidanimate} benchmark from the PARSEC 
 suite\cite{bienia11benchmarking}.  Fluid simulates the flow of an incompressible fluid
-using particles that move between a regular grid of cells.  To perform operations in 
-parallel the cells are partitioned into grids.  Unlike
-the Circuit application, the Fluid application first creates regions and partitions them before
-allocating cells in regions.  Another difference between Fluid and Circuit is that Fluid
+using particles that move within a regular grid of cells.  To perform operations in 
+parallel the cells are partitioned.  Unlike
+Circuit, Fluid first creates regions and partitions them before
+allocating cells in regions.  Another difference is that Fluid
 maintains separate regions for ghost cells rather using multiple partitions of
@@
 \label{subsec:amr}
-The third application is an adaptive mesh refinement (AMR) benchmark based on the third heat flow
-example from the Berkeley Labs BoxLib project\cite{BoxLib}.  The algorithm solves the two
+The third application is an adaptive mesh refinement (AMR) benchmark 
+from BoxLib \cite{BoxLib}.  The algorithm solves the two
 dimensional heat diffusion equation on a grid of cells using three levels of refinement with subrefinements
-randomly placed on the surface.  For each time step in the application, cells at the boundary of
-a refinement are interpolated from cells at a coarser level of refinement, energy fluxes between
-cells are computed, energy is transferred, and cells at a coarser level of refinement are restricted
-to the values of refined cells.  The AMR application uses separate regions for every level of
+randomly placed on the surface.  
+%For each time step in the application, cells at the boundary of
+%a refinement are interpolated from cells at a coarser level of refinement, energy fluxes between
+%cells are computed, energy is transferred, and cells at a coarser level of refinement are restricted
+%to the values of refined cells.  
+The AMR application uses separate regions for every level of
 refinement.  The regions at each level are partitioned several ways to provide multiple views of
@@
 levels of refinement.  These region relationships capture both intra- and inter-level disjointness
-information.  The dynamic nature of AMR requires that regions can be created, partitioned,
-and destroyed at runtime.  The many ways in which cells are used also mandates that multiple
-partitions of regions can be created.
+information.  The dynamic nature of AMR requires that regions can be created and partitioned
+at runtime.  
+%The many ways in which cells are used also requires multiple simultaneous
+%partitions of regions.
 
-These applications illustrate the expressivity of the Legion type system.  Legion is capable of 
+These applications show Legion is capable of 
 expressing applications with both regular (Fluid,AMR) and irregular (Circuit) pointer data structures.
-Being able to dynamically create, partition, and destroy regions at runtime is crucial to Legion's
+Being able to dynamically create and partition regions at runtime is crucial to Legion's
 ability to handle applications that make runtime decisions about data organization(AMR).  Legion
-is able to capture both allocate-then-partition (Circuit) and partition-then-allocate (Fluid) ways 
-of loading data.  Having multiple partitions of regions is necessary for describing the many ways 
-that data can be accessed (Circuit,AMR).  All types of privileges and coherence are necessary for 
-expressing the various applications presented in this paper (Circuit,Fluid,AMR,Histogram).  In
-all applications, region relationships convey disjointness information despite
-regions being a first-class feature that cannot be fully tracked statically.
+is able to capture both allocate-then-partition (Circuit) and partition-then-allocate (Fluid) algorithms. 
+Having multiple partitions of regions is necessary for describing the many ways 
+that data can be accessed (Circuit,AMR).  All the types of privileges and coherence we have
+described are needed in different combinations by the applications.  In
+all applications, region relationships convey disjointness information.
+% despite
+%regions being a first-class feature that cannot be fully tracked statically.
 
@@
 \label{subsec:overhead}
-Our initial implementation of Legion consisted only of a C++ library for Legion 
-programs\cite{Legion12} which contained no checking of region memory accesses.  In the process
-of implementing applications in Legion we frequently encountered memory corruption due to
-illegal region accesses caused by a confluence of bugs and lack of checking.  Many times these corruptions 
-occurred on remote nodes or on GPUs.  Finding these bugs required either multiple {\tt gdb} sessions 
-for observing execution on different nodes simultaneously or using {\tt printf} for debugging on
-GPUs.  To mitigate this problem and guarantee safety we added a feature to our runtime system that 
-dynamically checked all region accesses for both CPUs and GPUs.  However, this feature
+Our first Legion implementation consisted of a C++ library of Legion primitives
+\cite{Legion12} with no checking of region memory accesses.  
+When testing applications we frequently encountered memory corruption due to
+illegal region accesses caused by a confluence of application bugs and the lack of checking.  Many times these corruptions 
+occurred on remote nodes or on GPUs, requiring heroic debugging efforts.
+%  Finding these bugs required either multiple {\tt gdb} sessions 
+%for observing execution on different nodes simultaneously or using {\tt printf} for debugging on
+%GPUs.  
+To mitigate this problem and guarantee safety we added 
+dynamic checks on  all region accesses for both CPUs and GPUs.  However, these checks
 added considerable runtime overhead.  
-
-To avoid the overhead of dynamic checks for all memory accesses we implemented a type checker for the
-language in Section \ref{subsec:langdef}.  We wrote all of our applications in this language and
-used our type checker to verify them.  After proving that all the applications type checked we were
+To avoid the overhead of dynamic checks we implemented the type, privilege, and coherence checker we have described. 
+We rewrote all of our applications in this language and used our type checker to verify them.  Once the applications type checked we were
 able to safely elide all the dynamic region access checks.
@@
 \begin{center}
-\includegraphics[scale=0.35]{figs/circuit_96_popl.pdf}
+\includegraphics[scale=0.33]{figs/circuit_96_popl.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[scale=0.35]{figs/fluid_19200_popl.pdf}
+\includegraphics[scale=0.33]{figs/fluid_19200_popl.pdf}
 \end{center}
@@
 component of each bar shows the overhead added by the dynamic checks.  In 
-each figure the problem size stays the same as the number of processors are increased
+each figure the problem size stays the same as the number of processors increases
 (strong scaling).  Figure~\ref{fig:amr_overhead} includes multiple problem sizes to show
@@
 between better and worse performance than the baseline.  For an explanation of overall
-performance relative to the baseline implementations please refer to \cite{Legion12}.
+performance relative to the baseline implementations see \cite{Legion12}.
 
@@
 {
-\includegraphics[scale=0.35]{figs/amr_4096_popl.pdf}
+\includegraphics[scale=0.33]{figs/amr_4096_popl.pdf}
 \label{fig:amr4096}
@@
 {
-\includegraphics[scale=0.35]{figs/amr_8192_popl.pdf}
+\includegraphics[scale=0.33]{figs/amr_8192_popl.pdf}
 \label{fig:amr8192}
@@
 {
-\includegraphics[scale=0.35]{figs/amr_16384_popl.pdf}
+\includegraphics[scale=0.33]{figs/amr_16384_popl.pdf}
 \label{fig:amr16384}
@@
 degradation due to the extra registers required for checking.  We also measured the overhead
-of the dynamic checks associated with unpack operations but found them to be negligible relative
-to the overhead of region access checks.
+of the dynamic checks associated with unpack operations but found them to be negligible.
+% relative to the overhead of region access checks.
 
@@
 the overhead of the Legion runtime is always less than 7\% of the total execution time
-of the applications.  In some applications communication doesn't scale well, but in these cases communication is
-the result of the algorithm being performed and is not caused by runtime communication.
-Even in the case of the Circuit application which exhibits quadratic communication, the Legion
+of the applications.  In some applications communication doesn't scale well, but this is
+a result of the algorithm chosen by the programmer, not the Legion runtime.
+Even in the case of the Circuit application, which exhibits quadratic communication scaling, the Legion
 runtime is able to achieve a 62.5X speedup on 96 GPUs over a hand-coded single GPU implementation\cite{Legion12}.

### 2012-07-07 doc/popl2013/intro.tex
Minor edits

@@
 
-Briefly, in Legion data is organized in a hierarchy of {\em regions}
-and subregions and computation is organized in a hierarchy of {\em
+In Legion data is organized in a hierarchy of {\em regions}
+and subregions while computation is organized in a hierarchy of {\em
 tasks} and subtasks operating on regions.  Regions and tasks interact
@@
 soundness of the permissions and coherence system and use these two
-results to prove a third result: if two tasks $t_1$ and $t_2$ are {\em
-dependent} (roughly, must be executed in program order to guarantee
-the correct answer), then the unique ancestors of $t_1$ and $t_2$ in
-the task hierarchy that are {\em siblings} (have the same parent task)
-are also dependent.  This theorem is the basis for correct distributed
+results to prove a third result: if two {\em siblings} (tasks that
+have the same immediate parent task) $t_1$ and $t_2$ are 
+{\em independent} (have no ordering requirements for correctness), 
+then the unique ancestors of $t_1$ and $t_2$ in the task hierarchy 
+are also independent.  This theorem is the basis for correct distributed
 task scheduling in a Legion implementation, which is crucial for
 performance; a centralized scheduler would be a bottleneck
-because of the huge latencies involved in moving data and computation
-in the target class of machines.
+because of the huge latencies involved in communicating in the target
+class of machines.
 

### 2012-07-07 doc/popl2013/langdef.tex
Minor edits

@@
 %(i.e. knowing which part of the region tree to examine) and to catch downcasts that will never
-%work at runtime.) The consistency of a {\tt read}'s result is trivial for an address in the
+%work at runtime.) 
+The consistency of a {\tt read}'s result is trivial for an address in the
 clobber set and uses the consistency of the store otherwise.  The consistency of a 
@@
 guarantees that the subexpression's result cannot use the regions that were added to the
-mapping, allowing the changes to the mapping to be ignored.  The last case is, yet again, the body
+mapping, allowing the changes to the mapping to be ignored.  The last case is again the body
 of a called function, and the same strategy that was used for the type consistency of the

### 2012-07-07 doc/popl2013/related.tex
related

--- 2012-07-04 doc/popl2013/related.tex
@@
 \label{sec:related}
-Regions have been used for many purposes in the type system literature.  We
-cover the ones most closely related to the Legion type system.
+Legion is most directly related to Sequoia \cite{Fatahalian06}.  Sequoia is array-based, with
+a limited repetoire of ways to partition arrays.  Sequoia is a very static language with a single unified hierarchy
+of tasks and data; Legion is more dynamic with separate task and region hierarchies.
 
-Deterministic Parallel Java (DPJ) is the only other region system that we are
-aware of that uses regions for extracting parallelism from a 
-program to improve performance \cite{Bocchino09}.  There are several
-differences between DPJ and Legion.  First, regions in Legion are first-class 
-which allows them to be created, deleted, partitioned, packed, and unpacked 
-dynamically.  This allows programmers to dynamically determine partitioning
-schemes and allocation strategies based on program data at runtime unlike DPJ
-where partition schemes must be decided at compile-time.  Despite the dynamic nature
-of Legion where regions can be packed and unpacked, we still maintain disjointness
-information through constraints on region relationships.  Legion also allows 
+Deterministic Parallel Java (DPJ) is the only other region-based parallel system we are
+aware of \cite{Bocchino09}.  While there are similarities in the permission system and we have
+reused some DPJ notation, there are differences stemming from DPJ's more static approach.
+Regions in Legion are first-class and can be created, partitioned, packed, and unpacked 
+dynamically, allowing programmers to compute data organization at runtime; like Sequoia, DPJ
+partition schemes must be decided at compile-time.  Legion also allows 
 programmers to create multiple partitions of the same region to give different 
-views onto the same data which is not possible in DPJ.  DPJ only supports the 
-equivalent to Legion's exclusive and atomic coherence modes \cite{Bocchino11} 
-while Legion provides safe execution even in simultaneous environments.  The 
-implementation of DPJ only runs on shared memory machines, while Legion's runtime
-uses its knowledge of dynamic regions to run on distributed heterogenous machines.
+views onto the same data, which is not possible in DPJ.  
+%DPJ only supports the 
+%equivalent to Legion's exclusive and atomic coherence modes \cite{Bocchino11} 
+%while Legion provides safe execution even in simultaneous environments.  
+There is also a difference in emphasis: DPJ targets shared memory machines, while the Legion system
+is designed for distributed heterogenous machines.
+
+Chapel \cite{Chamberlain:Chapel} and X10 \cite{X1005} also provide some Legion-like facilities.
+Chapel's domains and X10's places provide the programmer with a 
+mechanism for expressing locality, similar to regions in Legion.  However, domains
+and places are not used for independence analysis to discover parallelism.
+In contrast, Jade uses annotations to describe
+data disjointness,  and like Legion leverages the disjointness information
+to discover parallelism, but lacks a region system to name and organize unbounded collections of objects \cite{Rinard98}.
 
@@
 % to properly differentiate our work
-There have been many region systems that use regions for memory management either
-to prove safety properties or for performance.  Tofte and Talpin introduced the
-first region type system for memory management\cite{Tofte94}.  Aiken et. al. 
-improved this system to decouple region allocation from lexical scoping\cite{Aiken95}.
-Henglein et. al. introduced a type system for reasoning about memory management
-in imperative object-oriented languages \cite{Henglein01}.  Walker et. al. add
-capablities to regions to give provable safety guarantees about how regions
-can be manipulated \cite{Walker00}.
+Many efforta use static region systems for  memory management (e.g., \cite{Tofte94, Grossman02}).
+Our system is more closely affiliated with dynamic region systems used for expressing locality for performance \cite{Gay01}.
 
-In addition to region languages with static type systems, there have been several
-dynamic region languages.  Cyclone used both a static type system and dynamic
-region checks to enforce memory safety properties of C programs\cite{Grossman02}.
-Gay and Aiken introduced RC which reference counts regions dynamically and uses
-a static type system to reason about effeciently garbage collecting regions\cite{Gay01}.
+%In addition to region languages with static type systems, there have been several
+%dynamic region languages.  Cyclone uses both a static type system and dynamic
+%region checks to enforce memory safety properties of C programs\cite{Grossman02}.
+%Gay and Aiken introduced RC which reference counts regions dynamically and uses
+%a static type system to reason about effeciently garbage collecting regions\cite{Gay01}.
+
+There have been many type and effect systems for ownership types
+\cite{Boyapati03} including ones that leverage nested regions for describing
+relationships \cite{Clarke02,Cameron07}.  However, ownership type and effect systems
+are primarily used for reasoning about determinism in object oriented languages and
+don't capture the range of disjointness properties that can be specified in Legion.
+
+Reasoning about disjoint heap data is the strong suit of separation logic \cite{Reynolds02}.  
+Concurrent separation logic\cite{Brookes04} has been 
+used both to parallelize sequential programs\cite{Raza09,Gotsman07} and to provide 
+a mechanism for reasoning about independence\cite{Hayman06}.
+While we have borrowed some separation logic notation, we ultimately chose to use a 
+permissions system as our primary formalism because separation logic does not easily
+support the interleaving of effects to non-disjoint regions of memory.
 
@@
 % DPJ also cites Lu06 here, but I'm not sure if we have to
-Type and effect systems have also been used to reason about regions.  FX presented
-the original type and effect system on regions \cite{Lucassen88}, but was restricted 
-to using a finite number of regions and was incapable of describing nested data
-structures.  There have been many type and effect systems for ownership types
-\cite{Boyapati03} including ones that leverage nested regions for describing
-relationships \cite{Clarke02,Cameron07}.  However, ownership type and effect systems
-are primarily used for reasoning about determinism in object oriented languages and
-don't capture the range of disjointness properties that can be specified in Legion.
+%Type and effect systems have also been used to reason about regions.  FX presented
+%the original type and effect system on regions \cite{Lucassen88}, but was restricted 
+%to using a finite number of regions and was incapable of describing nested data
+%structures.
+
 % Do we need to enumerate what these disjointness properties are (DPJ does)
-Type and effect systems have also been used in the context of parallelism to discover
-deadlocks and race conditions \cite{Boyapati02,Abadi06,Jacobs08}, but do not present 
-any mechanism for discovering parallelism.
+%Type and effect systems have also been used in the context of parallelism to discover
+%deadlocks and race conditions \cite{Boyapati02,Abadi06,Jacobs08}, but do not present 
+%any mechanism for discovering parallelism.
 
-Separation Logic is another mechanism that can be used for reasoning about disjointness
-of program data \cite{Reynolds02}. %Still not sure what we want to put here.
+%Still not sure what we want to put here.
 % DPJ sites additional separation logic papers but they didn't seem very similar.
 % Let me know if you think we should include them as well.
-
-Chapel\cite{Chamberlain:Chapel} and X10 \cite{X1005} are similar parallel programming
-models to Legion.  Chapel's domains and X10's places provide the programmer with a 
-mechanism for expressing locality similar to regions in Legion.  However, domains
-and places are not used in any way for independence analysis for discovering parallelism.
-Jade is a parallel programming mode that relies on programmer annotations to describe
-data disjointness\cite{Rinard98}.  Jade then leverages this disjointness information
-to implicitly discover parallelism similar to Legion.

### 2012-07-07 doc/popl2013/evaluation.tex
Minor edits and fixing of broken things

@@
 Our prototype implementation has two components: a type checker
-for the language introduced in Section \ref{subsec:langdef} and a C++ runtime library
+for the language introduced in Section \ref{sec:legioncore} and a C++ runtime library
 for executing programs written in the Legion programming model\cite{Legion12}.  All experiments

### 2012-07-07 doc/popl2013/langdef.tex
Minor edits and fixing of broken things

@@
 \subsection{Core Type System}
+\label{subsec:coretypes}
 
@@
 \subsection{Operational Semantics}
+\label{subsec:opsemantics}
 
@@
 \section{Soundness of Effects}
+\label{sec:soundness}
 
@@
 \begin{proof}
-To prove $valid\_interleave(S_0, C, E', \oton{E})$, we need to show the three conditions:
+To prove $valid\_interleave(S_0, C, E', \oton{E})$, we need to show the four conditions:
 \begin{enumerate}
@@
 We now extend the noninterference operator to entire memory traces, providing a way to say that
-no operation in a trace interferes with any operation in other:
+no operation in a trace interferes with any operation in another:
 $$E_1 \nonint E_2 \Leftrightarrow \displaystyle\bigwedge_{\epsilon_1 \text{ in } E_1, \epsilon_2 \text{ in } E_2} \epsilon_1 \nonint \epsilon_2$$
@@
 one extreme by work on transactional
-memory \cite{transactionalmemory}.  At the other extreme are systems
+memory \cite{Harris05}.  At the other extreme are systems
 that check privileges statically, such as Jade \cite{Rinard98} and
@@
 \subsection{Relaxed Coherence}
+\label{subsec:relaxed}
 
@@
 result in valid interleavings even if constituent traces have locations in common.  We discuss
-two cases of interest.  In both  the previous section's
+two cases of interest.  In both the previous sections noninterference % Is noninterference the right subject here?
 is repeated with minor modifications, so for brevity we limit

### 2012-07-07 doc/popl2013/related.tex
Minor edits and fixing of broken things

@@
 dynamically, allowing programmers to compute data organization at runtime; like Sequoia, DPJ
-partition schemes must be decided at compile-time.  Legion also allows 
+partition schemes must be decided at compile-time.  Legion allows 
 programmers to create multiple partitions of the same region to give different 
@@
 % to properly differentiate our work
-Many efforta use static region systems for  memory management (e.g., \cite{Tofte94, Grossman02}).
+Many efforts use static region systems for  memory management (e.g., \cite{Tofte94, Grossman02}).
 Our system is more closely affiliated with dynamic region systems used for expressing locality for performance \cite{Gay01}.

### 2012-07-08 doc/popl2013/langdef.tex
Getting rid of the ugly boxes for the type system

@@
 %\newcommand{\infrule}[2]{\displaystyle\frac{\displaystyle\strut{#1}}{\displaystyle\strut {#2}}}
-\newcommand{\cinfrule}[3]{\parbox{14cm}{\hfil$\infrule{#1}{#2}$\hfil}\parbox{4cm}{$\,#3$\hfil}}
-\newcommand{\finfrule}[2]{\framebox{$\infrule{#1}{#2}$}}
-\newcommand{\oldfinfrule}[2]{\vspace{10pt}\framebox{$\infrule{#1}{#2}$}\vspace{10pt}}
-
-\newcommand{\infx}[2]{\infrule{\begin{array}{l}{#1}\end{array}}{#2}}
+%\newcommand{\cinfrule}[3]{\parbox{14cm}{\hfil$\infrule{#1}{#2}$\hfil}\parbox{4cm}{$\,#3$\hfil}}
+%\newcommand{\finfrule}[2]{\framebox{$\infrule{#1}{#2}$}}
+%\newcommand{\oldfinfrule}[2]{\vspace{10pt}\framebox{$\infrule{#1}{#2}$}\vspace{10pt}}
+
+%\newcommand{\infx}[2]{\infrule{\begin{array}{l}{#1}\end{array}}{#2}}
 
@@
 
+\begin{comment}
 \begin{figure*}
@@
 }
+\caption{Legion Core Type System}
+\label{fig:types}
+\end{figure*}
+\end{comment}
+
+\begin{figure*}
+\begin{adjustwidth}{-4em}{-4em}
+\begin{center}
+% Row 1: bool, int, tuple1, tuple2
+\begin{minipage}{\linewidth}
+%\centering
+\begin{tabular}{m{4cm}m{3.5cm}m{5cm}m{5cm}}
+\infax[BOOL]{\typeenv{bv}{bool}} & 
+\infax[INT]{\typeenv{iv}{int}} & 
+\infrule[TUPLE1]{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.1}\ : T_1} & 
+\infrule[TUPLE2]{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.2}\ : T_2} 
+\end{tabular}
+\end{minipage}
+% Row 2:  var, null, isnull, maketuple
+\begin{minipage}{\linewidth}
+\vspace{-0.4cm}
+%\centering
+\begin{tabular}{m{3cm}m{3cm}m{5cm}m{6cm}}
+\infrule[VAR]{\Gamma(id) = T}{\typeenvx id : T} &
+\infax[NULL]{\typeenvx \text{null }T@r : T@r} &
+\infrule[ISNULL]{\typeenvx e : T@(\oton{r})}{\typeenvx \text{isnull}(e) : bool} &
+\infrule[MAKETUPLE]{\typeenvx e_1 : T_1 \andalso \typeenvx e_2 : T_2}{\typeenvx \langle e_1, e_2 \rangle : \langle T_1, T_2 \rangle}
+\end{tabular}
+\end{minipage}
+% Row 3: new, newcolor, coloring
+\begin{minipage}{\linewidth}
+\vspace{-0.6cm}
+%\centering
+\begin{tabular}{m{4cm}m{5cm}m{9cm}}
+\infax[NEW]{\typeenvx \text{new }T@r : T@r} &
+\infax[NEWCOLOR]{\typeenvx \text{newcolor }r : \text{coloring}(r)} &
+\infrule[COLORING]{\typeenvx e_1 : \text{coloring}(r) \andalso \typeenvx e_2 : T@r \andalso \typeenvx e_3 : int}{\typeenvx \text{color}(e_1, e_2, e_3) : \text{coloring}(r)}
+\end{tabular}
+\end{minipage}
+% Row 4: let, add, compare
+\begin{minipage}{\linewidth}
+\vspace{-0.7cm}
+%\centering
+\begin{tabular}{m{6cm}m{5cm}m{6cm}}
+\infrule[LET]{\typeenvx e_1 : T_1 \\ \typeenvx[G={\Gamma[id/T_1]}] e_2 : T_2}{\typeenvx : \text{let}\ id : T_1 \text{in}\ e_2 : T_2} &
+\infrule[ADD]{\typeenvx e_1 : int \\ \typeenvx e_2 : int}{\typeenvx e_1 + e_2 : int} &
+\infrule[COMPARE]{\typeenvx e_1 : int \\ \typeenvx e_2 : int}{\typeenvx e_1 < e_2 : bool}
+\end{tabular}
+\end{minipage}
+% Row 5: upregion, downregion
+\begin{minipage}{\linewidth}
+\vspace{-0.4cm}
+\begin{tabular}{m{9cm}m{10cm}}
+\infrule[UPREG]{\typeenvx e : T@(r'_1, \ldots r'_k) \andalso \forall i. \exists j, r'_i \leq r_j \in \Omega^*}{\typeenvx upregion(e,\oton{r}) : T@(\oton{r})} &
+\infrule[DOWNREG]{\typeenvx e : T@(r'_1, \ldots r'_k) \andalso \forall j. \exists i, r_j \leq r'_i \in \Omega^*}{\typeenvx downregion(e,\oton{r}) : T@(\oton{r})}
+\end{tabular}
+\end{minipage}
+% Row 6: read, write, reduce
+\begin{minipage}{\linewidth}
+\vspace{-0.3cm}
+\begin{tabular}{m{5cm}m{5cm}m{8cm}}
+\infrule[READ]{\typeenvx e_1 : T@(\oton{r}) \\ \forall i, reads(r_i) \in \Phi^*}{\typeenvx \text{read}(e_1) : T} &
+\infrule[WRITE]{\typeenvx e_1 : T@(\oton{r}) \\ \typeenvx e_2 : T \\ \forall i, writes(r_i) \in \Phi^*}{\typeenvx \text{write}(e_1, e_2) : T@(\oton{r})} &
+\infrule[REDUCE]{\Gamma(id) = (T_1, T_2), \emptyset, \emptyset \rightarrow T_1 \andalso \typeenvx e_2 : T_2 \\
+ \typeenvx e_1 : T_1@(\oton{r}) \andalso \forall i, reduces_{id}(r_i) \in \Phi^*}{\typeenvx \text{reduce}(id, e_1, e_2) : T_1@(\oton{r})}
+\end{tabular}
+\end{minipage}
+% Row 7: pack, if-else, task call
+\begin{minipage}{\linewidth}
+\vspace{-0.6cm}
+\begin{tabular}{m{5cm}m{5cm}m{8cm}}
+\infrule[PACK]{T_1 = \exists r'_1, \ldots r'_n.\ T_2\text{ where }\Omega_1 \\ \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^* \\ \typeenvx e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]}{\typeenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] : T_1} &
+\infrule[IF-ELSE]{\typeenvx e_1 : bool \\ \typeenvx e_2 : T \\ \typeenvx e_3 : T}{\typeenvx \text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3 : T} &
+\infrule[TASK CALL]{\Gamma(id) = \forall r'_1, \ldots r'_k.(\oton{T}),\Phi', Q' \rightarrow T_r \\ \typeenvx e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\ \Phi'[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*}{\typeenvx id[\otok{k}{r}](\oton{e}) : T_r[r_1/r'_1,\ldots,r_k/r'_k]}
+\end{tabular}
+\end{minipage}
+% Row 8: unpack, partition
+\begin{minipage}{\linewidth}
+\vspace{-0.7cm}
+\begin{tabular}{m{10cm}m{8cm}}
+\infrule[UNPACK]{T_1 = \exists r'_1, \ldots, r'_n.\ T_2\text{ where }\Omega_1 \andalso \typeenvx e_1 : T_1 \\
+  \Gamma' = \Gamma[T_2[r_1/r'_1,\ldots,r_k/r'_k] / id] \andalso \Omega' = \Omega \cup \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \\
+  \typeenvx[G=\Gamma',O=\Omega'] e_2 : T_3 \andalso \{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T_1, T_3) = \emptyset}
+  {\typeenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 : T_3} &
+\infrule[PARTITION]{\typeenvx e_1 : \text{coloring}(r_p) \\ \Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{1 \leq i < j \leq k} r_i * r_j \\ \typeenvx[O=\Omega'] e_2 : T \\ \{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T) = \emptyset}
+  {\typeenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 : T}
+\end{tabular}
+\end{minipage}
+% Row 9: program
+\begin{minipage}{\linewidth}
+\vspace{-0.5cm}
+\infrule[PROGRAM]{\text{for }1 \leq i \leq p, \\
+  \Gamma(id_i) = \forall r_{i1}, \ldots r_{ik_i}. (T_{i1}, \ldots, T_{in_i}), \Phi_i, Q_i \rightarrow T_{ir} \andalso
+  \Gamma_i = \Gamma[a_{i1}/T_{i1}, \ldots, a_{in_i}/T_{in_i}] \andalso
+  \typeenv[G={\Gamma_i},P={\Phi_i},O=\emptyset]{e_i}{T_{ir}}}
+  {\vdash \{  \text{task }id_1[r_{11}, \ldots, r_{1k_1}]( a_{11} : T_{11}, \ldots a_{1n_1} : T_{1n_1} ), \Phi_1, Q_1 : T_{1r} : e_1, 
+           \ldots
+           \text{task }id_p[r_{p1}, \ldots, r_{pk_1}]( a_{p1} : T_{p1}, \ldots a_{pn_p} : T_{pn_p} ), \Phi_p, Q_p : T_{pr} : e_p \} : \bullet}
+\end{minipage}
+\end{center}
+\end{adjustwidth}
 \caption{Legion Core Type System}
@@
 }
+\caption{Legion Core Operational Semantics}
+\label{fig:semantics}
+\end{figure*}
+\end{comment}
+
+\begin{figure*}
+\begin{adjustwidth}{-4em}{-4em}
+\begin{center}
+% Row 1: bool, int, val
+\begin{minipage}{\linewidth}
+\begin{tabular}{m{6cm}m{6cm}m{6cm}}
+\infax[OP-BOOL]{\opsenvx bv \mapsto bv, []} &
+\infax[OP-INT]{\opsenvx iv \mapsto iv, []} &
+\infax[OP-NULL]{\opsenvx \text{null }T@r \mapsto null, []} 
+\end{tabular}
+\end{minipage}
+% Row 2: tuple1, tuple2, make tuple
+\begin{minipage}{\linewidth}
+\vspace{-0.6cm}
+\begin{tabular}{m{9cm}m{4.5cm}m{4.5cm}}
+\infrule[OP-MAKETUPLE]{\opsenvx e_1 \mapsto v_1, E_1 \andalso S' = apply(S, E_1) \\
+  \opsenvx[S=S'] e_2 \mapsto v_2, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)}
+  {\opsenvx \langle e_1, e_2 \rangle \mapsto \langle v_1, v_2 \rangle, E'} &
+\infrule[OP-TUPLE1]{\opsenvx e \mapsto \langle v_1, v_2 \rangle, E}{\opsenvx e\text{.1} \mapsto v_1, E} &
+\infrule[OP-TUPLE2]{\opsenvx e \mapsto \langle v_1, v_2 \rangle, E}{\opsenvx e\text{.2} \mapsto v_2, E}
+\end{tabular}
+\end{minipage}
+% Row 3: new, null, isnull, isnull
+\begin{minipage}{\linewidth}
+\vspace{-0.6cm}
+\begin{tabular}{m{4cm}m{4.5cm}m{5cm}m{5cm}}
+\infrule[OP-NEW]{l \in M(r) \\ H(l) = M \llbracket T \rrbracket}{\opsenvx \text{new }T@r \mapsto l, []} &
+\infrule[OP-VAL]{L(id) = v}{\opsenvx id \mapsto v, []} &
+\infrule[OP-ISNULL-FALSE]{\opsenvx e \mapsto l, E}{\opsenvx \text{isnull}(e) \mapsto \textit{false}, E} &
+\infrule[OP-ISNULL-TRUE]{\opsenvx e \mapsto null, E}{\opsenvx \text{isnull}(e) \mapsto true, E}
+\end{tabular}
+\end{minipage}
+% Row 4: let, add, compare
+\begin{minipage}{\linewidth}
+\vspace{-0.7cm}
+\begin{tabular}{m{6cm}m{6cm}m{6cm}}
+\infrule[OP-LET]{\opsenvx e_1 \mapsto v_1, E_1 \\ L' = L[v_1/id] \\ S' = \text{apply}(S, E_1) \\ \opsenvx[L=L',S=S'] e_2 \mapsto v_2, E_2 \\ valid\_interleave(S, C, E', E_1, E_2)}{\opsenvx \text{let }id : T = e_1\text{ in }e_2 \mapsto v_2, E'} &
+\infrule[OP-ADD]{\opsenvx e_1 \mapsto v_1, E_1 \\ S' = \text{apply}(S, E_1) \\ \opsenvx[S=S'] e_2 \mapsto v_2, E_2 \\ v' = v_1 + v_2 \\ valid\_interleave(S, C, E', E_1, E_2)}{\opsenvx e_1 + e_2 \mapsto v', E'} &
+\infrule[OP-COMPARE]{\opsenvx e_1 \mapsto v_1, E_1 \\ S' = \text{apply}(S, E_1) \\ \opsenvx[S=S'] e_2 \mapsto v_2, E_2 \\ v' = \begin{cases} true, & \text{if $v_1 < v_2$}.
+\\ \textit{false}, & \text{otherwise}.
+\end{cases} \\ valid\_interleave(S, C, E', E_1, E_2)}{\opsenvx e_1 < e_2 \mapsto v', E'}
+\end{tabular}
+\end{minipage}
+% Row 5: if-else, if-else, new color
+\begin{minipage}{\linewidth}
+\vspace{-0.7cm}
+\begin{tabular}{m{6.5cm}m{6.5cm}m{6cm}}
+\infrule[OP-IF-ELSE-TRUE]{\opsenvx e_1 \mapsto true, E_1 \\ S' = \text{apply}(S, E_1) \\ \opsenvx[S=S'] e_2 \mapsto v_2, E_2}{\opsenvx \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_2, E_1 \dplus E_2} &
+\infrule[OP-IF-ELSE-FALSE]{\opsenvx e_1 \mapsto \textit{false}, E_1 \\ S' = \text{apply}(S, E_1) \\ \opsenvx[S=S'] e_3 \mapsto v_3, E_3}{\opsenvx \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_3, E_1 \dplus E_3} &
+\infrule[OP-NEW-COLOR]{K = \{ (l_1, iv_1), \ldots, (l_p, v_p) \}, \text{ where } \\ (\forall i \in [1,p].
+l_i \in M(r)) \wedge (\forall i,j \in [1,p].
+l_i \not= l_j)}{\opsenvx \text{newcolor }r \mapsto K, []}
+\end{tabular}
+\end{minipage}
+% Row 6: partition, color
+\begin{minipage}{\linewidth}
+\vspace{-0.6cm}
+\begin{tabular}{m{8.5cm}m{10.5cm}}
+\infrule[OP-PARTITION]{\opsenv{e_1}{K}{E_1} \andalso \rho_i = \{ l : (l, i) \in K \}, \text{ for } 1 \leq i \leq k \\
+  M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \andalso S' = \text{apply}(S, E_1) \\
+  \opsenvx[M=M',S=S'] e_2 \mapsto v, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)}
+  {\opsenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 \mapsto v, E'} &
+\infrule[OP-COLOR]{\opsenvx e_1 \mapsto K, E_1 \andalso S' = \text{apply}(S, E_1) \\
+  \opsenvx[S=S'] e_2 \mapsto l, E_2 \andalso S'' = \text{apply}(S', E_2) \\
+  \opsenvx[S=S''] e_3 \mapsto v, E_3 \\
+  K' = \{ (l,v) \} \cup \{ (l_i,v_i) : (l_i,v_i) \in K \wedge l \not= l_i \} \\
+  valid\_interleave(S, C, E', E_1, E_2, E_3)}
+  {\opsenvx \text{color}(e_1, e_2, e_3) \mapsto K', E'} 
+\end{tabular}
+\end{minipage}
+% Row 7: upregion, downregion, read
+\begin{minipage}{\linewidth}
+\vspace{-0.2cm}
+\begin{tabular}{m{6cm}m{6.5cm}m{6.5cm}}
+\infrule[OP-UPREG]{\opsenvx e \mapsto v, E}{\opsenvx \text{upregion}(e, \oton{r}) \mapsto v, E} &
+\infrule[OP-DOWNREG]{\opsenvx e \mapsto l, E \\ v = \begin{cases} l, & \text{if $\exists i, l \in M(r_i)$}.
+\\ null, & \text{otherwise}.
+\end{cases}}{\opsenvx \text{downregion}(e, \oton{r}) \mapsto v, E} &
+\infrule[OP-READ]{\opsenvx e \mapsto l, E \andalso S' = \text{apply}(S, E) \\ v = \begin{cases} S'(l), & \text{if } l \not\in C \\ v' : H(l), & \text{otherwise} \end{cases}}{\opsenvx \text{read}(e) \mapsto v, E \dplus [ read(l, excl, v) ]}
+\end{tabular}
+\end{minipage}
+% Row 8: write, reduce
+\begin{minipage}{\linewidth}
+\vspace{-0.5cm}
+\begin{tabular}{m{9cm}m{9cm}}
+\infrule[OP-WRITE]{\opsenvx e_1 \mapsto l, E_1 \andalso S' = \text{apply}(S, E_1) \\
+  \opsenvx[S=S'] e_2 \mapsto v, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)}
+  {\opsenvx \text{write}(e_1, e_2) \mapsto l, E' \dplus [ write(l, excl, v) ]} &
+\infrule[OP-REDUCE]{\opsenvx e_1 \mapsto l, E_1 \andalso S' = \text{apply}(S, E_1) \\
+  \opsenvx[S=S'] e_2 \mapsto v, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)}
+  {\opsenvx \text{reduce}(id, e_1, e_2) \mapsto l, E' \dplus [ reduce_{id}(l, excl, v) ]}
+\end{tabular}
+\end{minipage}
+% Row 9: pack, unpack
+\begin{minipage}{\linewidth}
+\vspace{-0.5cm}
+\begin{tabular}{m{8cm}m{10cm}}
+\infrule[OP-PACK]{\opsenvx e_1 \mapsto v, E \andalso \rho_i = M[r_i], \text{ for } 1 \leq i \leq k \\
+  v' = \langle \langle \otok{k}{\rho}, v \rangle \rangle}
+  {\opsenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] \mapsto v', E} &
+\infrule[OP-UNPACK]{\opsenvx e_1 \mapsto \langle \langle \otok{k}{\rho} , v_1 \rangle \rangle, E_1 \andalso M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \\
+  L' = L[v_1/id] \andalso S' = \text{apply}(S, E_1) \\
+  \opsenvx[M=M',L=L',S=S'] e_2 \mapsto v_2, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)}
+  {\opsenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 \mapsto v_2, E'}
+\end{tabular}
+\end{minipage}
+% Row 10: task call
+\begin{minipage}{\linewidth}
+%\vspace{-0.5cm}
+\infrule[OP-TASK-CALL]{\opsenvx e_1 \mapsto v_1, E_1 \andalso S_1 = \text{apply}(S, E_1) \andalso \opsenvx e_2 \mapsto v_2, E_2 \andalso \ldots \andalso S_n = \text{apply}(S_{n-1}, E_n) \\
+  valid\_interleave(S, C, E', \oton{E}) \andalso \text{task }id[\otok{k}{r'}](a_1 : T_1, \ldots, a_n : T_n), \Phi', Q' : T_r = e_{n+1} \\
+  M' = \{ (r'_1, M(r_1)), \ldots (r'_k, M(r_k)) \} \andalso L' = \{ (a_1, v_1), \ldots, (a_n, v_n) \} \andalso S' = \text{apply}(S, E') \\
+  C' = C \cup \{ l : \exists \rho. atomic(\rho) \in M' \llbracket Q' \rrbracket \vee simult(\rho) \in M' \llbracket Q' \rrbracket \} \\
+  \opsenvx[M=M',L=L',S=S'] e_{n+1} \mapsto v_{n+1}, E_{n+1} \andalso E'_{n+1} = mark\_coherence(E_{n+1}, M' \llbracket Q' \rrbracket) \andalso valid\_interleave(S, C, E'', E', E_{n+1})}
+  {\opsenvx id[\otok{k}{r}](\oton{e}) \mapsto v_{n+1}, E''}
+\end{minipage}
+\end{center}
+\end{adjustwidth}
 \caption{Legion Core Operational Semantics}

### 2012-07-08 doc/popl2013/evaluation.tex
Editting the evaluation section to be more precise about disjointness information

@@
 maintains separate regions for ghost cells rather using multiple partitions of
-the regions containing shared data.  Region relationships are again used to remember
-the disjointness of information for the regions of each grid.
+the regions containing shared data.  Region relationships are used for remembering
+which regions are required for each grid.
 
@@
 each level of refinement, and another for describing the relationship between pieces at different
-levels of refinement.  These region relationships capture both intra- and inter-level disjointness
-information.  The dynamic nature of AMR requires that regions can be created and partitioned
-at runtime.  
+levels of refinement.  
+%These region relationships capture both intra- and inter-level disjointness information.  
+The dynamic nature of AMR requires that regions can be created and partitioned at runtime.  
 %The many ways in which cells are used also requires multiple simultaneous
@@
 that data can be accessed (Circuit,AMR).  All the types of privileges and coherence we have
-described are needed in different combinations by the applications.  In
-all applications, region relationships convey disjointness information.
+described are needed in different combinations by the applications.  Region relationships
+are used in all applications. %, region relationships convey disjointness information.
 % despite

### 2012-07-09 doc/popl2013/intro.tex
intro

@@
 operations a task may perform on a region argument (read,
-write, or reduce) and novel {\em region coherence} annotations that
+write, or reduce) and  {\em region coherence} annotations that
 express what other tasks may do concurrently with
@@
 
-% The linked list example
-% Add float={t} to the list of options if you want the listing at the top
-\begin{lstlisting}[float={t},label={lst:linkedlist},caption={Linked List Example}]
-type list<r> = < int, list<r>@r >
+\lstset{
+  captionpos=b,
+  language=C++,
+  basicstyle=\scriptsize,
+  numbers=left,
+  numberstyle=\tiny,
+  columns=fullflexible,
+  stepnumber=1,
+  escapechar=\#,
+  keepspaces=true,
+  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
+  morekeywords={region,coloring,partition,spawn,disjoint,aliased},
+  deletekeywords=float,
+}
+\begin{lstlisting}[float={t},label={lst:code_ex},caption={Circuit simulation.}]
+struct Node { float voltage, new_charge, capacitance; };
+struct Wire<rn> { Node@rn in_node, out_node; float current, ... ; };
+struct Circuit { region   r_all_nodes; /* contains all nodes for the circuit */
+                 region   r_all_wires; /* contains all circuit wires */ };
+struct CircuitPiece {
+  region  rn_pvt, rn_shr, rn_ghost; /* private, shared, ghost node regions */
+  region  rw_pvt;                   /* private wires region */ };
 
-task sum_list[r] ( l:list<r>@r ), reads(r) : int =
-    if isnull(l) then 0 else
-    let v : list<r> = read(l) in
-        v.1 + sum_list[r](v.2)
+void simulate_circuit(Circuit c, float dt) : RWE(c.r_all_nodes, c.r_all_wires)
+{
+  // The construction of the colorings is not shown.  The colorings wire_owner_map,
+  // node_owner_map, and node_neighbor_map have MAX_PIECES colors 
+  // 0..MAX_PIECES #$-$# 1. The coloring node_sharing map has two colors 0 and 1.
+  //
+  // Partition of wires into MAX_PIECES pieces
+  partition<disjoint> p_wires = c.r_all_wires.partition(wire_owner_map); 
+  // Partition nodes into two parts for all-private vs. all-shared
+  partition<disjoint> p_nodes_pvs = c.r_all_nodes.partition(node_sharing map);
+
+  // Partition all-private into MAX_PIECES disjoint circuit pieces
+  partition<disjoint> p_pvt_nodes = p_nodes_pvs[0].partition(node_owner_map);
+  // Partition all-shared into MAX_PIECES disjoint circuit pieces
+  partition<disjoint> p_shr_nodes = p_nodes_pvs[1].partition(node_owner_map);
+  // Partition all-shared into MAX_PIECES ghost regions, which may be aliased
+  partition<aliased> p_ghost_nodes = p_nodes_pvs[1].partition(node_neighbor_map);
+
+  CircuitPiece pieces[MAX_PIECES];
+  for(i = 0; i #$<$# MAX_PIECES; i++) 
+    pieces[i] = { rn_pvt: p_pvt_nodes[i], rn_shr: p_shr_nodes[i],
+                  rn_ghost: p_ghost_nodes[i], rw_pvt: p_wires[i] };
+  for (t = 0; t #$<$# TIME_STEPS; t++) {
+    for (i = 0; i #$<$# MAX_PIECES; i++) calc_new_currents(pieces[i]);
+    for (i = 0; i #$<$# MAX_PIECES; i++) distribute_charge(pieces[i], dt);
+    for (i = 0; i #$<$# MAX_PIECES; i++) update_voltages(pieces[i]);
+  }
+}
+
+void calc_new_currents(CircuitPiece piece):
+        RWE(piece.rw_pvt), ROE(piece.rn_pvt, piece.rn_shr, piece.rn_ghost) {
+  foreach(w : piece.rw_pvt)
+    w#$\rightarrow$#current = (w#$\rightarrow$#in_node#$\rightarrow$#voltage - w#$\rightarrow$#out_node#$\rightarrow$#voltage) / w#$\rightarrow$#resistance;
+}
+
+void distribute_charge(CircuitPiece piece, float dt):
+        ROE(piece.rw_pvt), RdA(piece.rn_pvt, piece.rn_shr, piece.rn_ghost) {
+  foreach(w : piece.rw_pvt) {
+    w#$\rightarrow$#in_node#$\rightarrow$#new_charge += -dt * w#$\rightarrow$#current;
+    w#$\rightarrow$#out_node#$\rightarrow$#new_charge +=  dt * w#$\rightarrow$#current;
+  }
+}
+
+void update_voltages(CircuitPiece piece): RWE(piece.rn_pvt, piece.rn_shr) {
+  foreach(n : piece.rn_pvt, piece.rn_shr) {
+    n#$\rightarrow$#voltage += n#$\rightarrow$#new_charge / n#$\rightarrow$#capacitance;
+    n#$\rightarrow$#new_charge = 0;
+  }
+}
 \end{lstlisting}
 
-Legion programs organize data into {\em logical regions}.  Logical
-regions (which we usually refer to simply as {\em regions}) express
-{\em locality}, which is simply data that will be used together.
-Listing~\ref{lst:linkedlist} gives an example that defines a region
+Legion programs organize data into a forest of {\em logical regions}.  Logical
+regions express {\em locality} (data that will be used together, and therefore may be colocated in the same
+physical memory) and {\em independence} (disjoint data that may be operated on in parallel).
+Beginning with Sequoia \cite{Fatahalion06} and Deterministic Parallel Java \cite{Bocchino09}, researchers
+have explored hierarchical decomposition of data to express locality and independence.  Two
+
 containing a linked list of integers and a task that sums the elements

### 2012-07-09 doc/popl2013/intro.tex
intro

@@
 Legion programs organize data into a forest of {\em logical regions}.  Logical
-regions express {\em locality} (data that will be used together, and therefore may be colocated in the same
+regions express {\em locality} (data that will be used together, and therefore colocated if possible in the same
 physical memory) and {\em independence} (disjoint data that may be operated on in parallel).
-Beginning with Sequoia \cite{Fatahalion06} and Deterministic Parallel Java \cite{Bocchino09}, researchers
-have explored hierarchical decomposition of data to express locality and independence.  Two
+Researchers have previously explored language designs with hierarchical decomposition of data to express
+locality and independence for high performance programming. Two recent examples, Sequoia \cite{Fatahalion06} and 
+Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a mechanism to partition the heap recursively
+into a tree of collections of data.  The two designs are different in many aspects, but agree that there is
+a single tree-shaped partitioning of data that can be checked fully statically (see Section~\ref{sec:related} for more
+discussion of related work).
 
-containing a linked list of integers and a task that sums the elements
-of the list.  As is standard in region systems, we use $T @ r$ for the
-type of a pointer to a value of type $T$ in region $r$.  The task also
-illustrates a simple use of the permissions system:
-task {\tt sum\_list} must have permission {\tt reads(r)}, the ability
-to read the values in region {\tt r}.
+Our own experience with Sequoia and many high-performance applications
+written in the current industry standard mix of MPI, shared-memory
+threads, and CUDA, has taught us that for many practical situations a
+single, fairly static partitioning is insufficient.  In practice, the
+best way to partition data is often a function of the data
+itself---i.e., the partitions need to be dynamically computed.
+Furthermore, it is useful to allow mutliple partitions simultaneously of
+the same data, providing different views on to that data. Thus, in Legion the 
+handling of regions and partitioning of regions is much more dynamic than in 
+previous designs.  In particular,
+\begin{itemize}
+\item  logical regions are first-class values in Legion
+and may be dynamically allocated and stored in data structures;
 
-% The tree example with partitioning and tree reversal
-\begin{lstlisting}[float={t},label={lst:treereverse},caption={Tree Partitioning and Reversal Example}]
-type tree<rl,rls,rt> = rr[rll, rlr, rtl, rtr] 
-                       < list<rl>@rls, < tree<rl,rll,rtl>@rtl, tree<rl,rlr,rtr>@rtr > >
-     		       where rll #$\le$# rls and rlr #$\le$# rls and rll #$*$# rlr and rls #$\le$# rl and
-                                  rtl #$\le$# rt and rtr #$\le$# rt and rtl #$*$# rtr
+\item logical regions can be {\em partitioned} into {\em subregions}; the programmer can express arbitrary partitions of
+regions;
 
-task color_list[r,rs] ( clr: coloring(rs), head: list<r>@r, 
-                        splitval: int ), reads(r) : coloring(rs) =
-    if isnull(head) then clr else
-    let v : list<r> = read(head) in
-    let ps : list<r>@rs = downregion(head, rs) in
-        color_list[r,rs](if isnull(ps) then clr 
-                                else color(clr, ps, if v.1 #$<$# splitval then 1 else 2), v.2, splitval)
+\item  a logical region may be dynamically partitioned in multiple different ways.
+\end{itemize}
+A consequence of these decisions is that, through partitioning, a
+given datum may belong to multiple different regions.  For example, if
+a region $R$ participates in three separate disjoint partition
+operations and each partition allocates every element of $R$ to a
+subregion (partitions need not be total in Legion), then every element
+of $R$ is included in three different subregions.  This is why we use
+the term {\em logical} regions: language-level regions serve to name
+sets of data but do not imply anything about physical layout.  There
+is a separate system of {\em physical regions} at runtime that hold
+copies of the data in the language-level logical regions.  In fact,
+the Legion implementation routinely maintains multiple distinct
+physical regions, each with a copy of the data in the corresponding
+logical region, for performance reasons.
 
-task make_tree[rl,rls,rt] ( l: list<rl>@rl, min: int, 
-                            max: int), reads(rl,rt), writes(rt) : tree<rl,rls,rt>@rt =
-    if isnull(l) then null tree<rl,rls,rt>@rt else 
-    let v : list<rl> = read(l) in
-    if (v.1 #$<$# min) | (v.1 #$>$# max) then
-    make_tree[rl,rls,rt](v.2, min, max) else
-    let root : tree<rl,rls,rt>@rt = new tree<rl,rls,rt>@rt in
-    partition rt using newcolor rt as rtl,rtr in
-    partition rls using color_list[rl,rls](newcolor rls, v.2, v.1) as rll,rlr in
-    let ptr : list<rl>@rls = downregion(l, rls) in
-    let lhs : tree<rl,rll,rtl>@rtl = make_tree[rl,rll,rtl](v.2, min, v.1) in
-    let rhs : tree<rl,rlr,rtr>@rtr = make_tree[rl,rlr,rtr](v.2, v.1, max) in
-    let v2 : tree<rl,rls,rt> = pack < ptr, < lhs, rhs > > as tree<rl,rls,rt>[rll, rlr, rtl, rtr] in
-        write(root, v2)
+A further consequence of placing the same data in multiple logical
+regions as a result of runtime computation is that Legion has a very
+hard static alias analysis problem on regions, which poses a problem
+for parallel computation because knowing regions do not alias is the
+basis for safe parallel execution.  This is why languages such as
+Sequoia and DPJ allow only a single static partition of each
+region---it simplifies static analysis of regions to the point that it
+is practical.
 
-task reverse_tree[rl,rls,rt] ( root: tree<rl,rls,rt>@rt ), reads(rt), writes(rt) : tree<rl,rls,rt>@rt =
-     if isnull(root) then null tree<rl,rls,rt>@rt else
-     let v : tree<rl,rls,rt> = read(root) in
-     unpack v as u : tree<rl,rls,t>[rll, rlr, rtl, rtr] in
-     let lhs_r : tree<rl,rll,rtl>@rtl = reverse_tree[rl,rll,rtl](u.2.1) in
-     let rhs_r : tree<rl,rlr,rtr>@rtr = reverse_tree[rl,rlr,rtr](u.2.2) in
-     let v_r : tree<rl,rls,rt> = pack < u.1, < rhs_r, lhs_r > > as tree<rl,rls,rt>[rlr, rll, rtr, rtl] in
-        write(root, v_r)
-\end{lstlisting}
+Legion's execution model is that by default, a program's semantics is
+its meaning as a sequential program, which we refer to as the {\em
+program order} execution; this is similar to DPJ.  While the
+implementation is not our focus here, if two functions work on
+disjoint data (because all of the regions they use are disjoint), then
+Legion may execute them simultaneously.  Also like DPJ,
+Legion has a system of {\em permissions} ({\em read}, {\em write}, and
+{\em reduce}) on regions and an orthogonal system of region {\em
+coherence} modes ({\em excl}, {\em atomic}, and {\em simultaneous})
+that increase the number of situations in which functions can be
+executed in parallel.
 
-Listing~\ref{lst:treereverse} gives a more elaborate example.  While
-artificial (it is designed to use the greatest number of Legion
-features in the smallest space), the program illustrates typical
-Legion programming idioms.  Given a list of integers as defined in Listing~\ref{lst:linkedlist}, 
-the task {\tt make\_tree} builds a binary tree of pointers to list elements.
-Each internal node of the tree stores a pointer to a list element with value $i$
-and the left and right subtrees hold pointers to list elements with values less than $i$ and greater than $i$,
-respectively.  The task {\tt reverse\_tree} then does an in-place, recursive reversal of the children of each interior
-node of the tree.  We focus on a description of the structure of types, regions and permissions in this example.
+The important difference bewteen Legion and the more static approaches
+is that the hardest part of the tests to justify parallel execution is
+done dynamically instead of statically.  The key observation is that
+alias analysis is hard statically but very easy and relatively cheap
+dynamically.  Thus, Legion falls between fully static systems, such as
+Sequoia and DPJ, and fully dynamic approaches such as transactional
+memory.  Legion still has a significant static component; in
+particular, the safety of permissions is checked statically and region
+pointer dereferences are checked statically.  The dynamic alias tests
+are done at the granularity of regions and one check on a region
+itself often serves to prove the safety of many subsequent region
+accesses.  In contrast, because transactional memory has no mechanism
+for grouping heap data into larger collections, it much test the
+overlap between two sets of data on each individual element, which is
+much more expensive.
 
-Line 1 defines a {\em region relationship}, which is a bounded existential type.  The
-{\tt tree} region relationship defines the type of a binary tree
-node. The naming convention followed throughout the example is that
-regions beginning with {\tt rl} name regions 
-holding list elements and regions beginning with {\tt rt} name
-regions holding tree elements.  The most important part is the {\em
-  subregion} and {\em disjointness} constraints that describe the
-relationships between the various regions.  Here {\tt rl} is the
-region holding all list elements and {\tt rls} and {\tt rt} are the
-regions holding, respetively, the subset of list elements and the tree
-nodes for this tree.  Note that {\tt rls} is required to be a
-subregion of {\tt rl}, which means that all the values in region {\tt
-  rls} are also in region {\tt rl}.  The body of the {\tt tree} region
-relationship declares the shape of a node of the tree, which includes
-a pointer to a list element in {\tt rls} and pointers to two
-subtrees. The region relation says that there exist four regions {\tt
-  rll, rlr, rtl}, and {\tt rtr} such that the left subtree has list
-and tree elements that are in subregions {\tt rll} and {\tt rtl} of
-the corresponding list and tree regions {\tt rls} and {\tt rt} for the
-entire tree; the right subtree's list and tree elements in subregions
-{\tt rlr} and {\tt rtr} are also subregions of {\tt rls} and {\tt rt}
-respectively.  Furthermore, the constraints $\tt rll \ast rlr$ and
-$\tt rtl \ast rtr$ guarantee that the left and right subtrees are
-disjoint.
+We begin in Section~\ref{sec:example} with an example program that illustrates
+a typical style of programming in Legion as well as motivating the need for multiple, dynamically computed  partitions of
+a region.  We define Core Legion, a small language suitable for our formal development, in Section~\ref{sec:legioncore}.
+The next four sections each state and prove one of our main results and contributions:
+\begin{itemize}
+\item We prove the soundness of Legion's permissions system (Section~\ref{sec:soundness}).
 
-The {\tt make\_tree} task, which builds a {\tt tree} indexing a list
-of integers, illustrates region {\em partitions}, another important
-building block of Legion programs.  The region arguments to {\tt
-make\_tree} are {\tt rl}, the region holding all list elements, {\tt
-rls}, the subregion of {\tt rl} holding those list elements to be
-indexed by this call to {\tt make\_tree}, and {\tt rt}, the region to
-hold the tree itself.  Line 21 divides the list elements in {\tt rls}
-into those that are less than and those that are greater than a split
-value (which is arbitrarily chosen to be the value of the first
-element of the list).  Partitioning is done by constructing a {\em
-coloring} using the task {\tt color\_list} (line 6), which traverses the list
-and for each list element in the subregion (region {\tt rs} in
-{\tt color\_region}) either colors the element 1 or 2 (colors are
-integers) depending on its relationship to the split value.  A
-coloring is just a distinghuished function, built using the Legion
-primitive {\tt color}, that maps elements of a region to integers.
-When used in a {\tt partition} expression the coloring defines how to
-populate new subregions of the partitioned region.  On line 22, the
-list elements in {\tt rls} are partitioned into two subregions {\tt
-rll} and {\tt rlr} of list elements less than, and greater than or
-equal to, the split value, respectively.  Another use of {\tt
-partition} is seen on line 21 where the {\tt rt} region is partitioned
-into subregions to hold the left {\tt rlt} and right {\tt rtr}
-subtrees, respectively.  Because region {\tt rt} is empty at this
-point (has no elements), the empty coloring function {\tt newcolor rt}
-for the region {\tt rt} is given to the partitioning operation; the
-two new subregions are also empty.  These examples illustrate that
-partitioning can be used for two distinct, and equally important,
-styles of computation: {\em allocate-then-partition}, in which a
-global collection of data is created and then subdivided for
-(usually parallel) processing, and {\em partition-then-allocate}, in
-which a collection of empty but disjoint subregions is created and
-then populated with data on which (usually parallel) processing will
-take place.
+\item We use the soundness of permissions to prove the soundness of Legion's region coherence modes (Section~\ref{sec:coherence}).
 
-Lines 26-27 illustrates {\em packing} a region relationship into the heap.
-Regions are first-class values in Legion and can be stored in data structures.
-Here a tree node is written to the {\tt root} pointer after being {\tt packed}.
-Opertionally, the four existentially quantified regions {\tt rll, rlr, rtl} and {\tt rtr} are
-stored with the tree value.  The type system guarantees statically that the packed value
-satisfies the subregion and disjointness constraints declared on lines 3-4 at the point
-the packed value is stored.  When the region relationship value is read 
-from the heap and {\tt unpack} is applied (line 32) the newly introduced regions are known to 
-satisfy these constraints.  A key design decision in Legion, one that turns out to be
-central to the soundness results as well as the ability to perform distributed, parallel
-scheduling of tasks, is that permissions and coherence annotations 
-are not captured by region relationships.  When a region is stored in the heap it has no
-permissions or coherence associated with it; these belong to the task.  When the region relationship
-is read out of the heap and unpacked, the task must already hold permissions for an unpacked region 
-to use it. Thus, on line 32, the task {\tt reverse\_tree} unpacks the region {\tt rtl} (among others).
-The reads and writes of {\tt rtl} typecheck, however, because {\tt reverse\_tree} has read and
-write permission for region {\tt rt} and $\tt rtl \leq rt$ is a constraint of the region relationship; holding
-permission for a region $r$ implies the same permission for any subregion of $r$ (note
-that the region {\tt rt} is not one of the existentially quantified regions).
+\item We show that if expressions $e_1$ and $e_2$ are {\em non-interfering} (can be executed in parallel), then subexpressions
+$e_1'$ of $e_1$ and $e_2'$ of $e_2$ are also non-interfering (Section~\ref{sec:scheduling}).  This result is the basis
+for a hierarchical, distributed scheduler in the Legion implementation; on the target class of machines, any centralized
+scheduler would be a serious bottleneck.
 
-The {\tt reverse\_tree} task swaps the two subtrees of a tree in place (lines 35-36) after calling two
-subtasks to recursively swap all pointers in the subtrees (lines 33-34).  Beside the unpack operations, there
-are two other interesting features of this task.  First, the Legion scheduler will detect at runtime that the two
-subtasks are independent and may choose to run them in parallel.  In this example the test can even be done
-statically (because {\tt rtl} and {\tt rtr} are statically known to be disjoint) but, in general, testing the
-runtime {\em physical regions} is more precise than the compile-time logical regions.  Furthermore, it is a theorem
-that it is sufficient for the scheduler to check only {\em sibling} tasks (subtasks with the same parent task,
-such as those on lines 33-34) for dependences.  Thus, further subtasks within the recursive calls can be scheduled
-without consideration of their subtask cousins, which serves as the basis of a distributed scheduler for Legion.
-The second point is that
-even after destructively updating the subtrees the type system can prove using all of the region constraints in the
-{\tt tree} region relationship that the result is a well-formed {\tt tree}.
+\item We give experimental evidence for the Legion design.  On three real-world applications, we show that 
+dynamic region pointer checks would be expensive, justifying checking this aspect of the type system statically.
+We also show that the cost of checking aliasing on region permissions is low, showing that a much more expressive
+and dynamic langauge is not incompatible with high performance.
 
-\begin{lstlisting}[label={lst:histogram},caption={Histogram with Reductions and Atomic Coherence}]
-type hist = < int, int >
-
-task hist_tree1[rl,rls,rt,rh] ( root: tree<rl,rls,rt>@rt, hptr: hist@rh, 
-                                t : int ), reads(rt,rls,rh), writes(rh), atomic(rh) : int =
-    if isnull(root) then 0 else
-    let v : tree<rl,rls,rt> = read(root) in
-    unpack v as u : tree<rl,rls,t>[rll, rlr, rtl, rtr] in
-    let lval : list<rl> = read(u.1) in
-    let hval : hist = read(hptr) in
-    let hval2 : hist = if lval.1 #$<$# t then < hval.1 + 1, hval.2 >
-                                     else < hval.1, hval.2 + 1 > in
-    let _ : hist@rh = write(hptr, hval2) in
-        1 + hist_tree1[rl,rll,rtl,rh](u.2.1, hptr, t) + hist_tree1[rl,rlr,rtr,rh](u.2.2, hptr, t)
-
-task inc_bucket ( hval: hist, b: int ) : hist =
-    if b #$<$# 2 then < hval.1 + 1, hval.2 > else < hval.1, hval.2 + 1 >
-
-task hist_tree2[rl,rls,rt,rh] ( root: tree<rl,rls,rt>@rt, hptr: hist@rh, 
-                                t : int ), reads(rt,rls), reduces(inc_bucket, rh) : int =
-    if isnull(root) then 0 else
-    let v : tree<rl,rls,rt> = read(root) in
-    unpack v as u : tree<rl,rls,t>[rll, rlr, rtl, rtr] in
-    let lval : list<rl> = read(u.1) in
-    let _ : hist@rh = reduce(inc_bucket, hptr, if lval.1 #$<$# t then 1 else 2) in
-        1 + hist_tree1[rl,rll,rtl,rh](u.2.1, hptr, t) + hist_tree1[rl,rlr,rtr,rh](u.2.2, hptr, t)
-\end{lstlisting}
-
-The last example, shown in Listing~\ref{lst:histogram}, is a histogram
-computation that recursively walks a {\tt tree}.  The first version 
-uses a {\em coherence annotation} to relax the
-required ordering of tasks.  When there is no coherence annotation on
-a region $r$ used by a task $t$, the task implicitly requires $r$
-with {\em exclusive} coherence, meaning that any other task $t'$ that might
-update $r$ in a way visible to $t$ must appear to execute in normal sequential
-order with respect to $t$.  The task {\tt hist\_tree1} only requires the region {\tt rh}
-with {\tt atomic} coherence, meaning that other tasks that also use {\tt rh} with {\tt atomic} coherence
-may be reordered to appear to execute atomically before or after the task.  The task {\tt hist\_tree2} achieves
-a similar effect in a finer-grain way by accessing {\tt rh} only with {\tt reduce} privileges, which allows reductions
-with the same associative-commutative operator to be reordered.
-
-The rest of the paper is organized as follows.  Section~\ref{sec:legioncore} defines the core legion language, type system,
-opertional semantics, and presents our main results, proving the soundness of permissions and coherence annotations, and proving
-that it is sufficient for the scheduler to test dependences only between sibling tasks.
-Section~\ref{sec:evaluation} presents an experimental evaluation, measuring both the improvement achieved by eliding checks the static type system can prove safe as well as demonstrating the performance benefits of the distributed scheduling.
-Section~\ref{sec:related} discusses the most closely related work and Section~\ref{sect:conclusion} concludes.
+\end{itemize}
+Finally, Section~\ref{sec:related} discusses the most closely related work and
+Section~\ref{sect:conclusion} concludes.

### 2012-07-09 doc/popl2013/intro.tex
minor

@@
 its meaning as a sequential program, which we refer to as the {\em
-program order} execution; this is similar to DPJ.  While the
-implementation is not our focus here, if two functions work on
+program order} execution.  While the
+implementation is not our focus here, if two functions use
 disjoint data (because all of the regions they use are disjoint), then

### 2012-07-09 doc/popl2013/langdef.tex
Small edit
minor
updates

@@
 universally quantified over all region names appearing in the type.
-The final value is a region relation, which stores a value and one or
-more logical regions satisfying associated region constraints.  An
+The final type is a region relationship, instance of which capture a value and
+one or more regions satisfying associated constraints.  An
 instance of a region relation can be read out of the heap and
-unpacked, giving new local names to the logical regions contained in
+unpacked, giving new local names to the regions contained in
 the instance.
-% and tracking
-%the constraints between those new logical regions.
 
@@
 callee are a subset of those held by the caller.  There is no way to create privileges within
-a Legion function - it is limited to privileges given to it by its caller.
+a Legion function - it is limited to privileges given to it by its caller.  (The
+top-level task of a Legion program begins with a single region from with other
+regions can be partitioned and full read and
+write privileges.)
 
@@
 \infrule[UPREG]{\typeenvx e : T@(r'_1, \ldots r'_k) \andalso \forall i. \exists j, r'_i \leq r_j \in \Omega^*}{\typeenvx upregion(e,\oton{r}) : T@(\oton{r})} &
-\infrule[DOWNREG]{\typeenvx e : T@(r'_1, \ldots r'_k) \andalso \forall j. \exists i, r_j \leq r'_i \in \Omega^*}{\typeenvx downregion(e,\oton{r}) : T@(\oton{r})}
+\infrule[DOWNREG]{\typeenvx e : T@(r'_1, \ldots r'_k)
+% \andalso \forall j. \exists i, r_j \leq r'_i \in \Omega^*
+}{\typeenvx downregion(e,\oton{r}) : T@(\oton{r})}
 \end{tabular}
@@
 \infrule[IF-ELSE]{\typeenvx e_1 : bool \\ \typeenvx e_2 : T \\ \typeenvx e_3 : T}{\typeenvx \text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3 : T} &
-\infrule[TASK CALL]{\Gamma(id) = \forall r'_1, \ldots r'_k.(\oton{T}),\Phi', Q' \rightarrow T_r \\ \typeenvx e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\ \Phi'[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*}{\typeenvx id[\otok{k}{r}](\oton{e}) : T_r[r_1/r'_1,\ldots,r_k/r'_k]}
+\infrule[TASK-CALL]{\Gamma(id) = \forall r'_1, \ldots r'_k.(\oton{T}),\Phi', Q' \rightarrow T_r \\ \typeenvx e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\ \Phi'[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*}{\typeenvx id[\otok{k}{r}](\oton{e}) : T_r[r_1/r'_1,\ldots,r_k/r'_k]}
 \end{tabular}
@@
 replacing all free logical region names $r_i$ in $x$ with corresponding physical regions $\rho_i$.  
-(Note $M \llbracket x \rrbracket$ is ill-formed if it contains a free logical region name $r_j \not\in dom(M)$.)
+(Note $M \llbracket x \rrbracket$ is ill-formed if $x$ contains a free logical region name $r_j \not\in dom(M)$.)
 
@@
 
-\item $v$ is the result of evaluatinig $e$.
+\item $v$ is the result of evaluating $e$.
 
@@
 \begin{itemize}
-\item $apply(S, E) = S'$ computes the store $S'$ resulting from applying memory trace $E$ to initial store $S$.
+\item $apply(S, E)$ computes the store that results from applying memory trace $E$ to initial store $S$.
 
@@
 \infrule[OP-IF-ELSE-FALSE]{\opsenvx e_1 \mapsto \textit{false}, E_1 \\ S' = \text{apply}(S, E_1) \\ \opsenvx[S=S'] e_3 \mapsto v_3, E_3}{\opsenvx \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_3, E_1 \dplus E_3} &
-\infrule[OP-NEW-COLOR]{K = \{ (l_1, iv_1), \ldots, (l_p, v_p) \}, \text{ where } \\ (\forall i \in [1,p].
+\infrule[OP-NEW-COLOR]{K = \{ (l_1, iv_1), \ldots, (l_p, iv_p) \}, \text{ where } \\ (\forall i \in [1,p].
 l_i \in M(r)) \wedge (\forall i,j \in [1,p].
@@
 %\vspace{-0.5cm}
-\infrule[OP-TASK-CALL]{\opsenvx e_1 \mapsto v_1, E_1 \andalso S_1 = \text{apply}(S, E_1) \andalso \opsenvx e_2 \mapsto v_2, E_2 \andalso \ldots \andalso S_n = \text{apply}(S_{n-1}, E_n) \\
+\infrule[OP-Task-CALL]{\opsenvx e_1 \mapsto v_1, E_1 \andalso S_1 = \text{apply}(S, E_1) \andalso \opsenvx e_2 \mapsto v_2, E_2 \andalso \ldots \andalso S_n = \text{apply}(S_{n-1}, E_n) \\
   valid\_interleave(S, C, E', \oton{E}) \andalso \text{task }id[\otok{k}{r'}](a_1 : T_1, \ldots, a_n : T_n), \Phi', Q' : T_r = e_{n+1} \\
@@
   C' = C \cup \{ l : \exists \rho. atomic(\rho) \in M' \llbracket Q' \rrbracket \vee simult(\rho) \in M' \llbracket Q' \rrbracket \} \\
-  \opsenvx[M=M',L=L',S=S'] e_{n+1} \mapsto v_{n+1}, E_{n+1} \andalso E'_{n+1} = mark\_coherence(E_{n+1}, M' \llbracket Q' \rrbracket) \andalso valid\_interleave(S, C, E'', E', E_{n+1})}
+  \opsenvx[M=M',L=L',S=S',C=C'] e_{n+1} \mapsto v_{n+1}, E_{n+1} \andalso E'_{n+1} = mark\_coherence(E_{n+1}, M' \llbracket Q' \rrbracket) \andalso valid\_interleave(S, C, E'', E', E_{n+1})}
   {\opsenvx id[\otok{k}{r}](\oton{e}) \mapsto v_{n+1}, E''}
@@
 
-\item[$E \sim H$] The type consistency of the value in an expression's effect follows from two lemmas applied to the effects
-of subexpressions:
-if $E_1 \sim H$ and $E_2 \sim H$, then $E_1 \dplus E_2 \sim H$; and 
-if $E_1 \sim H$ and $E_2 \sim H$ and $valid\_interleave(S, C, E', E_1, E_2)$, then $E' \sim H$.
+\item[$E \sim H$] The type consistency of the values in an expression's memory
+trace follows from the following lemma applied to the traces of subexpressions:
+if $E_1 \sim H$ and $E_2 \sim H$ and $any\_interleave(E', E_1, E_2)$, then $E' \sim H$.
 New effects are added by {\tt write} and {\tt reduce} expressions, but consistency follows
@@
 the different coherence modes permit different interleavings of memory traces.  For example,
-in addition to be an interleaving, exclusive access requires interleaved memory traces to be:
+in addition to being an interleaving of the constituent traces, exclusive access requires interleaved memory traces to be:
 \begin{itemize}
@@
 coherent(S, L_1, L_2, []) & = & \text{true} \\
-coherent(S, L_1, L_2, \epsilon \dplus E) & = &\
+coherent(S, L_1, L_2, [ \epsilon ] \dplus E) & = &\
 \begin{cases}
@@
 to the store.  (If $L_2 = \emptyset$ initially, atomic reads without a preceding write are not checked, effectively defering
-these checks to point where traces ordered before $E$ are included.)  The $seq\_equiv$
+these checks to point where traces that might be ordered before $E$ are included.)  The $seq\_equiv$
 test verifies every read in $E'$ with a location in $L_1$ 
@@
 
-%
-% FIX ME   I don't understand this paragraph.  I didn't see this discussed
-% earlier; I took that comment out.  I don't know what the pronoun ``the one in which''
-% refers to.  The clobber set?  The interleaving?
-%
 The first option is to serialize the constituent traces in program
-order.  Although each constituent trace can be assumed to be coherent,
-that coherence is in the context of a clobber set that may differ from
-the one in which the interleaving must occur. 
-This difference can result in cases where no valid interleaving
-exists.  
-%
-% FIX ME  Why?
-%
+order.  Although each constituent trace has been checked for coherence, the
+checking of some operations may have been deferred due to non-empty
+clobber sets.  It is possible that when an interleaving of these traces
+is attemped in the parent expression's context, there may no valid interleaving
+if the wrong guesses were made in the derivation of a subexpression.
 However, we can show that as long as any valid interleaving
@@
 
+% FIX ME FIX ME FIX ME
 \begin{proof}
@@
 Then,
-$valid\_interleave(S, C, E'_a \dplus \epsilon_1 \dplus \epsilon_2 \dplus E'_b, \oton{E})$
+$$valid\_interleave(S, C, E'_a \dplus [ \epsilon_1, \epsilon_2 ] \dplus E'_b, \oton{E})$$
 if and only if
-$valid\_interleave(S, C, E'_a \dplus \epsilon_2 \dplus \epsilon_1 \dplus E'_b, \oton{E}).$
+$$valid\_interleave(S, C, E'_a \dplus [ \epsilon_2, \epsilon_1 ] \dplus E'_b, \oton{E}).$$
 \end{lem}
@@
 \begin{enumerate}
-\item If $E'_a \dplus \epsilon_1 \dplus \epsilon_2 \dplus E'_b$ is an interleaving of $\oton{E}$,
-the proof must first ``pop off'' all of $E'_a$, leaving $\epsilon_1 \dplus \epsilon_2 \dplus E'_b$ for
+\item If $E'_a \dplus [ \epsilon_1 ] \dplus [ \epsilon_2 ] \dplus E'_b$ is an interleaving of $\oton{E}$,
+the proof must first ``pop off'' all of $E'_a$, leaving $[ \epsilon_1 ] \dplus [ \epsilon_2 ] \dplus E'_b$ for
 the interleaved subtrace.  The next two steps must pop $\epsilon_1$ and $\epsilon_2$, leaving $E'_b$.
 They cannot be in the same subtrace because $\epsilon_1 \nonint \epsilon_2$, so we could also pop
-$\epsilon_2$ and then $\epsilon_1$, resulting in a proof that $E'_a \dplus \epsilon_2 \dplus \epsilon_1 \dplus E'_b$ is also an interleaving of the constituent traces.
+$\epsilon_2$ and then $\epsilon_1$, resulting in a proof that $E'_a \dplus [ \epsilon_2 ] \dplus [ \epsilon_1 ] \dplus E'_b$ is also an interleaving of the constituent traces.
 \item For coherence, we must show the value read (if any) by $\epsilon_1$ does not change, the value
@@
 using $S' = apply(S, E'_a)$.
-Similarly to show that $apply(S', \epsilon_1 \dplus \epsilon_2 \dplus E) = apply(S', \epsilon_2 \dplus \epsilon_1 \dplus E)$ for any $E$, we need only to show that it holds
+Similarly to show that $apply(S', [ \epsilon_1  ] \dplus [ \epsilon_2 ] \dplus E) = apply(S', [ \epsilon_2 ] \dplus [ \epsilon_1 ] \dplus E)$ for any $E$, we need only to show that it holds
 for $E = []$.  The proof is now reduced to showing the following:

### 2012-07-10 doc/popl2013/intro.tex

@@
-% This is a description of how the listings should be formatted.
-% It can go anywhere before the listings.
-\lstset{
-  captionpos=b,
-  language=Haskell,
-  basicstyle=\scriptsize,
-  numbers=left,
-  numberstyle=\tiny,
-  columns=fullflexible,
-  stepnumber=1,
-  escapechar=\#,
-  keepspaces=true,
-  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
-  morekeywords={task,rr,int,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,color,newcolor,atomic,simultaneous},
-  deletekeywords={float,head,min,max}
-}
-
 \section{Introduction}
@@
 
+% This is a description of how the listings should be formatted.
+% It can go anywhere before the listings.
 \lstset{
   captionpos=b,
-  language=C++,
+  language=Haskell,
   basicstyle=\scriptsize,
@@
   literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
-  morekeywords={region,coloring,partition,spawn,disjoint,aliased},
-  deletekeywords=float,
-}
-\begin{lstlisting}[float={t},label={lst:code_ex},caption={Circuit simulation.}]
-struct Node { float voltage, new_charge, capacitance; };
-struct Wire<rn> { Node@rn in_node, out_node; float current, ... ; };
-struct Circuit { region   r_all_nodes; /* contains all nodes for the circuit */
-                 region   r_all_wires; /* contains all circuit wires */ };
-struct CircuitPiece {
-  region  rn_pvt, rn_shr, rn_ghost; /* private, shared, ghost node regions */
-  region  rw_pvt;                   /* private wires region */ };
-
-void simulate_circuit(Circuit c, float dt) : RWE(c.r_all_nodes, c.r_all_wires)
-{
-  // The construction of the colorings is not shown.  The colorings wire_owner_map,
-  // node_owner_map, and node_neighbor_map have MAX_PIECES colors 
-  // 0..MAX_PIECES #$-$# 1. The coloring node_sharing map has two colors 0 and 1.
-  //
-  // Partition of wires into MAX_PIECES pieces
-  partition<disjoint> p_wires = c.r_all_wires.partition(wire_owner_map); 
-  // Partition nodes into two parts for all-private vs. all-shared
-  partition<disjoint> p_nodes_pvs = c.r_all_nodes.partition(node_sharing map);
-
-  // Partition all-private into MAX_PIECES disjoint circuit pieces
-  partition<disjoint> p_pvt_nodes = p_nodes_pvs[0].partition(node_owner_map);
-  // Partition all-shared into MAX_PIECES disjoint circuit pieces
-  partition<disjoint> p_shr_nodes = p_nodes_pvs[1].partition(node_owner_map);
-  // Partition all-shared into MAX_PIECES ghost regions, which may be aliased
-  partition<aliased> p_ghost_nodes = p_nodes_pvs[1].partition(node_neighbor_map);
-
-  CircuitPiece pieces[MAX_PIECES];
-  for(i = 0; i #$<$# MAX_PIECES; i++) 
-    pieces[i] = { rn_pvt: p_pvt_nodes[i], rn_shr: p_shr_nodes[i],
-                  rn_ghost: p_ghost_nodes[i], rw_pvt: p_wires[i] };
-  for (t = 0; t #$<$# TIME_STEPS; t++) {
-    for (i = 0; i #$<$# MAX_PIECES; i++) calc_new_currents(pieces[i]);
-    for (i = 0; i #$<$# MAX_PIECES; i++) distribute_charge(pieces[i], dt);
-    for (i = 0; i #$<$# MAX_PIECES; i++) update_voltages(pieces[i]);
-  }
+  morekeywords={task,rr,int,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,color,newcolor,atomic,simultaneous},
+  deletekeywords={float,head,min,max}
 }
 
-void calc_new_currents(CircuitPiece piece):
-        RWE(piece.rw_pvt), ROE(piece.rn_pvt, piece.rn_shr, piece.rn_ghost) {
-  foreach(w : piece.rw_pvt)
-    w#$\rightarrow$#current = (w#$\rightarrow$#in_node#$\rightarrow$#voltage - w#$\rightarrow$#out_node#$\rightarrow$#voltage) / w#$\rightarrow$#resistance;
-}
+\begin{lstlisting}[float={t},label={lst:code_ex},caption={Circuit simulation.}]
+type CircuitNode        = < float, --voltage
+                            int >
+type CircuitWire<rn,rg>  = < CircuitNode@rn,
+                             CircuitNode@(rn,rg),
+                             float> -- current
 
-void distribute_charge(CircuitPiece piece, float dt):
-        ROE(piece.rw_pvt), RdA(piece.rn_pvt, piece.rn_shr, piece.rn_ghost) {
-  foreach(w : piece.rw_pvt) {
-    w#$\rightarrow$#in_node#$\rightarrow$#new_charge += -dt * w#$\rightarrow$#current;
-    w#$\rightarrow$#out_node#$\rightarrow$#new_charge +=  dt * w#$\rightarrow$#current;
-  }
-}
+type node_list<rl,rn>       = < CircuitNode@rn, node_list<rl,rn>@rl >
+type wire_list<rl,rw,rn,rg> = < CircuitWire<rn,rg>@rw, wire_list<rl,rw,rn,rg>@rl >
 
-void update_voltages(CircuitPiece piece): RWE(piece.rn_pvt, piece.rn_shr) {
-  foreach(n : piece.rn_pvt, piece.rn_shr) {
-    n#$\rightarrow$#voltage += n#$\rightarrow$#new_charge / n#$\rightarrow$#capacitance;
-    n#$\rightarrow$#new_charge = 0;
-  }
-}
+type CircuitPiece<rl,rw,rn> = rr[rpw,rpn,rg]
+        < wire_list<rl,rpw,rpn,rg>@rl,
+          node_list<rl,rpn>@rl >         
+          where rpn <= rn and rg <= rn and rpw <= rw and
+                rpn * rg and rn * rw and rl * rn and rl * rw
+
+type piece_list<rl,rp,rw,rn> = < CircuitPiece<rl,rw,rn>@rp, piece_list<rl,rp,rw,rn>@rl >
+
+function color_circuit[rn,rw,rl] ( all_nodes : node_list<rl,rn>@rl,
+                               all_wires : wire_list<rl,rw,rn>@rl ), reads(rn,rw,rl) : 
+                               <coloring(rn), coloring(rn), coloring(rw)> =  
+  -- Invoke programmer chosen coloring algorithm (e.g. METIS)
+
+function build_lists[rl,rw,rn,rpw,rpn,rg] ( all_nodes : node_list<rl,rn>@rl, all_wires : wire_list<rl,rw,rn>@rl,
+                                 owned_coloring : coloring(rn), ghost_coloring : coloring(rn),
+                                 wires_coloring : coloring(rw), c : int), reads(rn,rw,rl), writes(rl) :
+                                 < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl > = 
+  -- Construct node and wire lists for each piece based on coloring
+
+function build_piece_list[rl,rp,rw,rn] ( one: CircuitPiece<rl,rw,rn>, two : CircuitPiece<rl,rw,rn>,
+                                 three: CircuitPiece<rl,rw,rn>, four : CircuitPiece<rl,rw,rn> ),
+                                 reads(rl,rp,rw,rn), writes(rl,rp) : piece_list<rl,rp,rw,rn>@rl =
+  -- Build a list of the pieces
+
+function simulate_circuit[rl,rw,rn] ( all_nodes : node_list<rl,rn>@rl,
+                                  all_wires : wire_list<rl,rw,rn,rn>@rl, steps : int ),
+                          reads(rn,rw,rl), writes(rn,rw,rl) : bool = 
+  let pc : <coloring(rn),<coloring(rn),coloring(rw)> >
+                      = color_circuit[rn,rw,rl](all_nodes,all_wires) in
+  partition rn using pc.1 as rn0,rn1,rn2,rn3 in
+  partition rn using pc.2 as rg0,rg1,rg2,rg3 in
+  partition rw using pc.3 as rw0,rw1,rw2,rw3 in
+  let piece0_lists : < wire_list<rl,rw0,rn0,rg0>@rl, node_list<rl,rn0>@rl > = 
+           build_lists[rl,rw,rn,rw0,rn0,rg0](all_nodes,all_wires,pc.1,pc.2.1,pc.2.2,0) in
+  let piece0 : CircuitPiece<rl,rw,rn> = pack piece0_lists as CircuitPiece<rl,rw,rn>[rw0,rn0,rg0] in
+  let piece1_lists : < wire_list<rl,rw1,rn1,rg1>@rl, node_list<rl,rn1>@rl > = 
+           build_lists[rl,rw,rn,rw1,rn1,rg1](all_nodes,all_wires,pc.1,pc.2.1,pc.2.2,1) in
+  let piece1 : CircuitPiece<rl,rw,rn> = pack piece1_lists as CircuitPiece<rl,rw,rn>[rw1,rn1,rg1] in
+  let piece2_lists : < wire_list<rl,rw2,rn2,rg2>@rl, node_list<rl,rn2>@rl > = 
+           build_lists[rl,rw,rn,rw2,rn2,rg2](all_nodes,all_wires,pc.1,pc.2.1,pc.2.2,2) in
+  let piece2 : CircuitPiece<rl,rw,rn> = pack piece2_lists as CircuitPiece<rl,rw,rn>[rw2,rn2,rg2] in
+  let piece3_lists : < wire_list<rl,rw3,rn3,rg3>@rl, node_list<rl,rn3>@rl > = 
+           build_lists[rl,rw,rn,rw3,rn3,rg3](all_nodes,all_wires,pc.1,pc.2.1,pc.2.2,3) in
+  let piece3 : CircuitPiece<rl,rw,rn> = pack piece3_lists as CircuitPiece<rl,rw,rn>[rw3,rn3,rg3] in
+  let all_pieces : pieces_list<rl,rl,rw,rn> = build_piece_list[rl,rl,rw,rn](piece0,piece1,piece2,piece3) in
+      execute_time_step[rl,rl,rw,rn](all_pieces,steps)
+
+function execute_time_step[rl,rp,rw,rn] ( all_pieces : pieces_list<rl,rp,rw,rn>@rl ),
+                                          reads(rl,rp,rw,rn), writes(rw,rn) : bool =
+  if steps < 1 then true else
+  let all_pieces' : pieces_list<rl,rp,rw,rn>@rl = map calc_new_currents[rl,rw,rn] all_pieces in
+  let all_pieces'' : pieces_list<rl,rp,rw,rn>@rl = map distribute_charge[rl,rw,rn] all_pieces' in
+  let all_pieces''' : pieces_list<rl,rp,rw,rn>@rl = map update_voltage[rl,rw,rn] all_pieces'' in
+      execute_time_step[rl,rp,rw,rn](all_pieces''',steps-1)
+
+function calc_new_currents[rl,rw,rn] ( p : CircuitPiece<rl,rw,rn> ), 
+                                      reads(rl,rw,rn), writes(rw) : CircuitPiece<rl,rw,rn> =
+  -- Calculate new currents for each wire
+
+function distribute_charge[rl,rw,rn] ( p : CircuitPiece<rl,rw,rn> ), 
+                      reads(rl,rw,rn), reduces(reduce_charge,rn), atomic(rn) : CircuitPiece<rl,rw,rn> =
+  -- Distribute charge from each wire to all nodes
+
+function update_voltage[rl,rw,rn] ( piece : CircuitPiece<rl,rw,rn> ), reads(rl,rw,rn), writes(rn) : bool = 
+  -- Update voltage for all our owned nodes
+
 \end{lstlisting}

### 2012-07-10 doc/popl2013/langdef.tex

@@
 \newcommand{\rrbracket}{]\!]}
+\newcommand{\tuple}[2]{\langle #1, #2 \rangle}
 \makeatletter
+% macros for consistency flavors
+\newcommand{\foo}[1]{\cancel{#1}}
+\newcommand{\simH}{\sim_{\!H}}
+% usually: \mapconsist{\Omega}
+\newcommand{\mapconsist}[2][M]{#1 \foo\sim #2}
+% usually: \localconsist{L}{\Gamma}
+\newcommand{\localconsist}[3][M]{#2 \foo\simH \@ifmtarg{#1}{#3}{#1 \llbracket #3 \rrbracket}}
+% usually: \storeconsist{S}{H}
+\newcommand{\storeconsist}[2]{#1 \foo\sim #2}
+% usually: \valueconsist{v}{T}
+\newcommand{\valueconsist}[3][M]{#2 \foo\simH \@ifmtarg{#1}{#3}{#1 \llbracket #3 \rrbracket}}
+% usually: \traceconsist{E}
+\newcommand{\traceconsist}[2][H]{#2 \foo\sim #1}
+% usually: \privconsist{E}{\Phi}
+\newcommand{\privconsist}[3][M]{#2 \@ifmtarg{#1}{:}{:_{\!#1}} #3}
+
 \newcommand{\nonint}[1][]{\#\@ifmtarg{#1}{}{_{\!\!#1}}}
@@
 }
-\caption{Legion Core Language}
+\caption{Core Legion}
 \label{fig:langdef}
@@
 
-Figure~\ref{fig:langdef} defines the core subset of Legion.  Types
+Figure~\ref{fig:langdef} defines Core Legion, a subset of the full Legion language
+that still illustrates all of the important issues.  Types
 include booleans, integers, tuples, and pointers.  Note that pointers
@@
 guaranteed to access the heap only in ways permitted by the privileges under which it was typed.
-A judgement $M \vdash E : \Phi$ holds if all the memory operations in memory trace $E$ are of the types and
+A judgement $\privconsist{E}{\Phi}$ holds if all the memory operations in memory trace $E$ are of the types and
 to locations covered by the privileges in $\Phi$:
 \[ 
-\begin{array}{rl}
-M \vdash & E : \Phi \Leftrightarrow \forall \epsilon \in E. \\
-& (\epsilon = read(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reads(r) \in \Phi)\ \wedge \\
-& (\epsilon = write(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge writes(r) \in \Phi)\ \wedge \\
-& (\epsilon = reduce_{id}(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reduces_{id}(r) \in \Phi)
+\begin{array}{l}
+\privconsist{E}{\Phi} \Leftrightarrow \forall \epsilon \in E. \\
+\quad\quad (\epsilon = read(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reads(r) \in \Phi)\ \wedge \\
+\quad\quad (\epsilon = write(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge writes(r) \in \Phi)\ \wedge \\
+\quad\quad (\epsilon = reduce_{id}(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reduces_{id}(r) \in \Phi)
 \end{array}
@@
 
-Our goal is to show $M \vdash E : \Phi$ holds for the execution of an expression $e$ with static permissions $\Phi$.
+Our goal is to show $\privconsist{E}{\Phi}$ holds for any execution of an expression $e$ with static permissions $\Phi$.
 As usual, even stating such a claim requires that corresponding parts of the initial static and dynamic environments be consistent.
@@
 \begin{itemize}
-\item $M \sim \Omega$, a region mapping satisfies the region constraints under which the expression was typed
-\item $L \sim_H \Gamma$, local values have types consistent with the environment under which the expression was typed
-\item $S \sim H$, the initial state of the store contains values whose types are consisten with the heap typing
+\item $\mapconsist{\Omega}$, a region mapping satisfies the region constraints under which the expression was typed
+\item $\localconsist{L}{\Gamma}$, local values have types consistent with the environment under which the expression was typed
+\item $\storeconsist{S}{H}$, the initial state of the store contains values whose types are consisten with the heap typing
 \end{itemize}
@@
 \begin{itemize}
-\item $v \sim_H M \llbracket T \rrbracket$, the expression yields a value of the right type
-\item $E \sim H$, values written to the store match the heap types of their locations
+\item $\valueconsist{v}{T}$, the expression yields a value of the right type
+\item $\traceconsist{E}$, values written to the store match the heap types of their locations
 \end{itemize}
@@
 \begin{array}{lll}
-M \sim \Omega & \Leftrightarrow & (\forall r_i, r_j. r_i \leq r_j \in \Omega \Rightarrow M(r_i) \subseteq M(r_j))\ \wedge \\
+\mapconsist{\Omega} & \Leftrightarrow & (\forall r_i, r_j. r_i \leq r_j \in \Omega \Rightarrow M(r_i) \subseteq M(r_j))\ \wedge \\
               &                 & (\forall r_i, r_j. r_i * r_j \in \Omega \Rightarrow M(r_i) \cap M(r_j) = \emptyset) \\
-L \sim_H \Gamma & \Leftrightarrow & \forall (id,v) \in L. v \sim_H M \llbracket \Gamma \rrbracket (id) \\
-S \sim H & \Leftrightarrow & \forall (l,v) \in S. v \sim_H H(l) \\
+\localconsist{L}{\Gamma} & \Leftrightarrow & \forall (id,v) \in L. \valueconsist[]{v}{M \llbracket \Gamma \rrbracket (id)} \\
+\storeconsist{S}{H} & \Leftrightarrow & \forall (l,v) \in S. \valueconsist[]{v}{H(l)} \\
 \end{array} & 
 \begin{array}{lll}
-E \sim H & \Leftrightarrow & (\forall l, c, v. write(l,c, v) \in E \Rightarrow v \sim_H H(l))\ \wedge \\
+\traceconsist{E} & \Leftrightarrow & (\forall l, c, v. write(l,c, v) \in E \Rightarrow \valueconsist[]{v}{H(l)})\ \wedge \\
 & & (\forall id, l, v. reduce_{id}(l, v) \in E \Rightarrow \\
-& & (M \llbracket \Gamma \rrbracket (id) = (\hat T_1, \hat T_2), \emptyset, \emptyset \Rightarrow \hat T_1) \wedge H(l) = \hat T_1 \wedge v \sim_H \hat T_2) \\
+& & (M \llbracket \Gamma \rrbracket (id) = (\hat T_1, \hat T_2), \emptyset, \emptyset \Rightarrow \hat T_1) \wedge H(l) = \hat T_1 \wedge \valueconsist[]{v}{\hat T_2}) \\
 \end{array}
@@
 \begin{array}{l}
-bv \sim_H bool \\
-iv \sim_H int \\
-null \sim_H \hat T@\rho \\
+\valueconsist[]{bv}{bool} \\
+\valueconsist[]{iv}{int} \\
+\valueconsist[]{null}{\hat T@\rho} \\
 \end{array} &
 \begin{array}{lll}
-l \sim_H \hat T@\rho & \Leftrightarrow & l \in \rho \wedge H(l) \sim_H \hat T \\
-\langle v_1, v_2 \rangle \sim_H \langle \hat T_1, \hat T_2 \rangle & \Leftrightarrow & (v_1 \sim_H \hat T_1) \wedge (v_2 \sim_H \hat T_2) \\
-\langle \langle \oton{\rho}, v \rangle \rangle \sim_H \text{rr}[\oton{r}] \hat T\text{ where }\Omega & \Leftrightarrow & (v \sim_H \hat T[\rho_1/r_1, \ldots \rho_n/r_n]) \wedge (\{ (r_i, \rho_i) \} \sim \Omega) \\
-K \sim_H coloring(\rho) & \Leftrightarrow & \forall l_1, v_1. (l_1, v_1) \in K \Rightarrow (l_1 \in \rho\ \wedge \\
+\valueconsist[]{l}{\hat T@\rho} & \Leftrightarrow & l \in \rho \wedge H(l) = \hat T \\
+\valueconsist[]{\tuple{v_1}{v_2}}{\tuple{\hat T_1}{\hat T_2}} & \Leftrightarrow & (\valueconsist[]{v_1}{\hat T_1}) \wedge (\valueconsist[]{v_2}{\hat T_2}) \\
+\valueconsist[]{\langle \langle \oton{\rho}, v \rangle \rangle}{\text{rr}[\oton{r}]\ \hat T\text{ where }\Omega} & \Leftrightarrow & (\valueconsist[]{v}{T[\rho_1/r_1, \ldots \rho_n/r_n]}) \wedge (\mapconsist[\{ (r_i, \rho_i) \}]{\Omega}) \\
+\valueconsist[]{K}{coloring(\rho)} & \Leftrightarrow & \forall l_1, v_1. (l_1, v_1) \in K \Rightarrow (l_1 \in \rho\ \wedge \\
 & & \forall l_2, v_2. (l_2, v_2) \in K \Rightarrow (l_1 \not= l_2) \vee (v_1 = v_2)) \\
@@
 \rm
-If $\typeenv{e}{T}$ and $\opsenv{e}{v}{E}$  and $M \sim \Omega$, $L \sim_H \Gamma$ and $S \sim H$,
+If $\typeenv{e}{T}$ and $\opsenv{e}{v}{E}$  and $\mapconsist{\Omega}$, $\localconsist{L}{\Gamma}$ and $\storeconsist{S}{H}$,
 then
-$v \sim_H M \llbracket T \rrbracket$, $E \sim H$ and $M \vdash E : \Phi$.
+$\valueconsist{v}{T}$, $\traceconsist{E}$ and $\privconsist{E}{\Phi}$.
 \end{thm}
@@
 \label{lemma:heapconst:apply}
-If $S \sim H$ and $E \sim H$, then $\text{apply}(S, E) \sim H$.
+If $\storeconsist{S}{H}$ and $\traceconsist{E}$, then $\storeconsist{\text{apply}(S, E)}{H}$.
 \end{lem}
@@
 \label{lemma:heapconst:effects1}
-If $E_1 \sim H$ and $E_2 \sim H$, then $E_1 \dplus E_2 \sim H$.
+If $\traceconsist{E_1}$ and $\traceconsist{E_2}$, then $\traceconsist{E_1 \dplus E_2}$.
 \end{lem}
@@
 \label{lemma:heapconst:effects2}
-If $E_1 \sim H$ and $E_2 \sim H$ and $valid\_interleave(S, C, E', E_1, E_2)$, then $E' \sim H$.
+If $\traceconsist{E_1}$ and $\traceconsist{E_2}$ and $valid\_interleave(S, C, E', E_1, E_2)$, then $\traceconsist{E'}$.
 \end{lem}
@@
 \label{lemma:effsound:effects1}
-If $M \vdash E_1 : \Phi$ and $M \vdash E_2 : \Phi$, then $M \vdash E_1 \dplus E_2 : \Phi$.
+If $\privconsist{E_1}{\Phi}$ and $\privconsist{E_2}{\Phi}$, then $\privconsist{E_1 \dplus E_2}{\Phi}$.
 \end{lem}
@@
 \label{lemma:effsound:effects2}
-If $M \vdash E_1 : \Phi$ and $M \vdash E_2 : \Phi$ and $valid\_interleave(S, C, E', E_1, E_2)$, then $M \vdash E' : \Phi$.
+If $\privconsist{E_1}{\Phi}$ and $\privconsist{E_2}{\Phi}$ and $valid\_interleave(S, C, E', E_1, E_2)$, then $\privconsist{E'}{\Phi}$.
 \end{lem}
@@
 \label{lemma:omegaclosure}
-$M \sim \Omega^*$ if and only if $M \sim \Omega$.
+$\mapconsist{\Omega^*}$ if and only if $\mapconsist{\Omega}$.
 \end{lem}
@@
 \label{lemma:phiclosure}
-$M \vdash E : \Phi^*$ if and only if $M \vdash E : \Phi$.
+$\privconsist{E}{\Phi^*}$ if and only if $\privconsist{E}{\Phi}$.
 \end{lem}
@@
 
-\item[$M \sim \Omega$] Three expressions have subexpressions that modify $M$ or $\Omega$ and therefore do not
+\item[$\mapconsist{\Omega}$] Three expressions have subexpressions that modify $M$ or $\Omega$ and therefore do not
 trivially satisfy region mapping consistency.  For {\tt partition},
@@
 
-\item[$L \sim_H \Gamma$] Four expressions have subexpressions that modify $L$, $\Gamma$, or $M$.  For
+\item[$\localconsist{L}{\Gamma}$] Four expressions have subexpressions that modify $L$, $\Gamma$, or $M$.  For
 {\tt partition}, which only modifies $M$, the requirement that it not reuse existing
@@
 
-\item[$S \sim H$] The heap typing consistency of all stores used in subexpressions follows directly from the following
-lemma: If $S \sim H$ and $E \sim H$, then $\text{apply}(S, E) \sim H$.
+\item[$\storeconsist{S}{H}$] The heap typing consistency of all stores used in subexpressions follows directly from the following
+lemma: If $\storeconsist{S}{H}$ and $\traceconsist{E}$, then $\storeconsist{\text{apply}(S, E)}{H}$.
 
@@
 
-\item[$v \sim_H M \llbracket T \rrbracket$]    The consistency of 
+\item[$\valueconsist{v}{T}$]    The consistency of 
 {\tt upregion} is guaranteed by the type checking requirement of appropriate subregion
@@
 
-\item[$E \sim H$] The type consistency of the values in an expression's memory
+\item[$\traceconsist{E}$] The type consistency of the values in an expression's memory
 trace follows from the following lemma applied to the traces of subexpressions:
-if $E_1 \sim H$ and $E_2 \sim H$ and $any\_interleave(E', E_1, E_2)$, then $E' \sim H$.
+if $\traceconsist{E_1}$ and $\traceconsist{E_2}$ and $any\_interleave(E', E_1, E_2)$, then $\traceconsist{E'}$.
 New effects are added by {\tt write} and {\tt reduce} expressions, but consistency follows
@@
 
-\item[$M \vdash E : \Phi$] The proof of the crucial property of containment of effects within the available privileges
-is similar in outline to the previous step.  The easy cases are covered by lemmas combining the corresponding results
-for subexpressions: if $M \vdash E_1 : \Phi$ and $M \vdash E_2 : \Phi$, then $M \vdash E_1 \dplus E_2 : \Phi$, and
-if $M \vdash E_1 : \Phi$ and $M \vdash E_2 : \Phi$ and $valid\_interleave(S, C, E', E_1, E_2)$, then $M \vdash E' : \Phi$.
+\item[$\privconsist{E}{\Phi}$] The proof of the crucial property of containment of effects within the available privileges
+is similar in outline to the previous step.  The easy cases are covered by a lemma combining the corresponding results
+for subexpressions: if $\privconsist{E_1}{\Phi}$ and $\privconsist{E_2}{\Phi}$ and $any\_interleave(E', E_1, E_2)$, then $\privconsist{E'}{\Phi}$.
 Straightforward proofs cover {\tt read}, {\tt write}, {\tt reduce}, and one final special case for function calls. 
@@
 \rm
-Let $M$ be a region mapping and $E_1$ and $E_2$ two memory traces such that $M \vdash E_1 : \Phi_1$ and $M \vdash E_2 : \Phi_2$.
+Let $M$ be a region mapping and $E_1$ and $E_2$ two memory traces such that $\privconsist{E_1}{\Phi_1}$ and $\privconsist{E_2}{\Phi_2}$.
 If $\Phi_1$ and $\Phi_2$ are noninterfering under $M$, then $E_1$ and $E_2$ must be noninterfering.
@@
 Assume $\epsilon_1 = read(l, c_1, v_1)$ and $\epsilon_2 = write(l, c_2, v_2)$.
-The definition of $M \vdash E_1 : \Phi_1$ tells us that there must be some $r_1$ satisfying $l \in M(r_1) \wedge reads(r_1) \in \Phi_1$.
+The definition of $\privconsist{E_1}{\Phi_1}$ tells us that there must be some $r_1$ satisfying $l \in M(r_1) \wedge reads(r_1) \in \Phi_1$.
 Similarly, there must be some $r_2$ satisfying $l \in M(r_2) \wedge writes(r_2) \in \Phi_2$.
@@
 $H$ a heap typing, and $S$ be an initial store satisfying
-$M \sim \Omega$, $L \sim_H \Gamma$, and $S \sim H$.
+$\mapconsist{\Omega}$, $\localconsist{L}{\Gamma}$, and $\storeconsist{S}{H}$.
 If $\Phi_i \nonint[M] \Phi_j$ for $1 \leq i < j \leq n$, then any parallel execution of expressions

### 2012-07-10 doc/popl2013/intro.tex
Major revision to the circuit example code

--- 2012-07-09 doc/popl2013/intro.tex
@@
   literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
-  morekeywords={region,coloring,partition,spawn,disjoint,aliased},
-  deletekeywords=float,
-}
-\begin{lstlisting}[float={t},label={lst:code_ex},caption={Circuit simulation.}]
-struct Node { float voltage, new_charge, capacitance; };
-struct Wire<rn> { Node@rn in_node, out_node; float current, ... ; };
-struct Circuit { region   r_all_nodes; /* contains all nodes for the circuit */
-                 region   r_all_wires; /* contains all circuit wires */ };
-struct CircuitPiece {
-  region  rn_pvt, rn_shr, rn_ghost; /* private, shared, ghost node regions */
-  region  rw_pvt;                   /* private wires region */ };
-
-void simulate_circuit(Circuit c, float dt) : RWE(c.r_all_nodes, c.r_all_wires)
-{
-  // The construction of the colorings is not shown.  The colorings wire_owner_map,
-  // node_owner_map, and node_neighbor_map have MAX_PIECES colors 
-  // 0..MAX_PIECES #$-$# 1. The coloring node_sharing map has two colors 0 and 1.
-  //
-  // Partition of wires into MAX_PIECES pieces
-  partition<disjoint> p_wires = c.r_all_wires.partition(wire_owner_map); 
-  // Partition nodes into two parts for all-private vs. all-shared
-  partition<disjoint> p_nodes_pvs = c.r_all_nodes.partition(node_sharing map);
-
-  // Partition all-private into MAX_PIECES disjoint circuit pieces
-  partition<disjoint> p_pvt_nodes = p_nodes_pvs[0].partition(node_owner_map);
-  // Partition all-shared into MAX_PIECES disjoint circuit pieces
-  partition<disjoint> p_shr_nodes = p_nodes_pvs[1].partition(node_owner_map);
-  // Partition all-shared into MAX_PIECES ghost regions, which may be aliased
-  partition<aliased> p_ghost_nodes = p_nodes_pvs[1].partition(node_neighbor_map);
-
-  CircuitPiece pieces[MAX_PIECES];
-  for(i = 0; i #$<$# MAX_PIECES; i++) 
-    pieces[i] = { rn_pvt: p_pvt_nodes[i], rn_shr: p_shr_nodes[i],
-                  rn_ghost: p_ghost_nodes[i], rw_pvt: p_wires[i] };
-  for (t = 0; t #$<$# TIME_STEPS; t++) {
-    for (i = 0; i #$<$# MAX_PIECES; i++) calc_new_currents(pieces[i]);
-    for (i = 0; i #$<$# MAX_PIECES; i++) distribute_charge(pieces[i], dt);
-    for (i = 0; i #$<$# MAX_PIECES; i++) update_voltages(pieces[i]);
-  }
+  morekeywords={function,rr,int,float,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,multicoloring,color,newcolor,atomic,simultaneous},
+  deletekeywords={float,head,min,max}
 }
 
-void calc_new_currents(CircuitPiece piece):
-        RWE(piece.rw_pvt), ROE(piece.rn_pvt, piece.rn_shr, piece.rn_ghost) {
-  foreach(w : piece.rw_pvt)
-    w#$\rightarrow$#current = (w#$\rightarrow$#in_node#$\rightarrow$#voltage - w#$\rightarrow$#out_node#$\rightarrow$#voltage) / w#$\rightarrow$#resistance;
-}
+\begin{lstlisting}[float={t},label={lst:circuit_ex},caption={Circuilt Simulation}]
+type CircuitNode        = <float/*voltage*/,float/*current*/,
+                           float/*charge*/,float/*capacitance*/>
+type CircuitWire<rn,rg>  = <CircuitNode@rn,      -- Owned node
+                            CircuitNode@(rn,rg), -- Owned or Ghost node
+                            float/*resistance*/, float/*current*/>
 
-void distribute_charge(CircuitPiece piece, float dt):
-        ROE(piece.rw_pvt), RdA(piece.rn_pvt, piece.rn_shr, piece.rn_ghost) {
-  foreach(w : piece.rw_pvt) {
-    w#$\rightarrow$#in_node#$\rightarrow$#new_charge += -dt * w#$\rightarrow$#current;
-    w#$\rightarrow$#out_node#$\rightarrow$#new_charge +=  dt * w#$\rightarrow$#current;
-  }
-}
+type node_list<rl,rn>       = < CircuitNode@rn, node_list<rl,rn>@rl >
+type wire_list<rl,rw,rn,rg>= < CircuitWire<rn,rg>@rw, wire_list<rl,rw,rn,rg>@rl >
 
-void update_voltages(CircuitPiece piece): RWE(piece.rn_pvt, piece.rn_shr) {
-  foreach(n : piece.rn_pvt, piece.rn_shr) {
-    n#$\rightarrow$#voltage += n#$\rightarrow$#new_charge / n#$\rightarrow$#capacitance;
-    n#$\rightarrow$#new_charge = 0;
-  }
-}
+type CircuitPiece<rl,rw,rn> = rr[rpw,rpn,rg]
+                            < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl >         
+                            where rpn <= rn and rg <= rn and rpw <= rw and
+                                  rpn * rg and rn * rw and rl * rn and rl * rw
+
+-- Simulation initialization and invocation
+function simulate_circuit[rl,rw,rn] ( all_nodes : node_list<rl,rn>@rl, 
+      all_wires : wire_list<rl,rw,rn,rn>@rl, steps : int ), 
+      reads(rn,rw,rl), writes(rn,rw,rl) : bool = 
+  let pc : <coloring(rn),coloring(rn),coloring(rw)> 
+            = color_circuit[rn,rw,rl](all_nodes,all_wires) in
+  -- Disjoint partition for the owned nodes of each piece
+  partition rn using pc.1 as rn0,rn1 in
+  -- Aliased partition for ghost nodes of each piece
+  partition rn using pc.2 as rg0,rg1 in
+  -- Disjoint partition for the owned wires of each piece
+  partition rw using pc.3 as rw0,rw1 in
+  let piece0 : CircuitPiece<rl,rw,rn> = 
+                build_piece[rl,rw,rn,rw0,rn0,rg0](all_nodes,all_wires,pc.1,pc.2,pc.3,0) in
+  let piece1 : CircuitPiece<rl,rw,rn> = 
+                build_piece[rl,rw,rn,rw1,rn1,rg1](all_nodes,all_wires,pc.1,pc.2,pc.3,1) in
+      execute_time_steps[rl,rw,rn](piece0,piece1,steps)
+
+-- Time Step Loop
+function execute_time_steps[rl,rw,rn] ( p0 : CircuitPiece<rl,rw,rn>, 
+      p1 : CircuitPiece<rl,rw,rn>, steps : int ) , reads(rn,rw,rl), writes(rn,rw) : bool = 
+  if steps #$<$# 1 then true else
+  unpack p0 as piece0 : CircuitPiece<rl,rw,rn>[rw0,rn0,rg0] in 
+  unpack p1 as piece1 : CircuitPiece<rl,rw,rn>[rw1,rn1,rg1] in
+  let _ : bool = calc_new_currents[rl,rw0,rn0,rg0](piece0.1) in
+  let _ : bool = calc_new_currents[rl,rw1,rn1,rg1](piece1.1) in
+  let _ : bool = distribute_charge[rl,rw0,rn0,rg0](piece0.1) in
+  let _ : bool = distribute_charge[rl,rw1,rn1,rg1](piece1.1) in
+  let _ : bool = update_voltage[rl,rn0](piece0.2) in
+  let _ : bool = update_voltage[rl,rn1](piece1.2) in
+      execute_time_steps[rl,rw,rn](p0,p1,steps-1)
+
+-- Helper methods
+function color_circuit[rn,rw,rl] ( all_nodes : node_list<rl,rn>@rl, 
+                               all_wires : wire_list<rl,rw,rn>@rl ), 
+        reads(rn,rw,rl) : <coloring(rn), multicoloring(rn), coloring(rw)> =  
+  -- Invoke programmer chosen coloring algorithm (e.g. METIS)
+  -- return owned, ghost, wire colorings
+
+function build_lists[rl,rw,rn,rpw,rpn,rg] ( nodes : node_list<rl,rn>@rl, 
+       wires : wire_list<rl,rw,rn>@rl, oc : coloring(rn), gc : multicoloring(rn), 
+       wc : coloring(rw), c : int), reads(rn,rw,rl), writes(rl) 
+       : < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl > = 
+  -- Construct lists of node and wire pointers for the given colorings
+
+function build_piece[rl,rw,rn,rpw,rpn,rg] ( nodes : node_list<rl,rn>@rl, 
+      wires : wire_list<rl,rw,rn>@rl, oc : multicoloring(rn), gc : coloring(rn), 
+      wc : coloring(rw), c : int), reads(rn,rw,rl), writes(rl) : CircuitPiece<rl,rw,rn> =
+  let piece_lists : < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl > = 
+    build_lists[rl,rw,rn,rpw,rpn,rg](nodes,wires,oc,gc,wc,c) in
+      pack piece_lists as CircuitPiece<rl,rw,rn>[rpw,rpn,rg] 
 \end{lstlisting}
@@
 discussion of related work).
+
+\begin{lstlisting}[float={t},label={lst:circuit_leaf},caption={Circuilt Leaf Tasks}]
+function calc_new_currents[rl,rw,rn,rg] ( ptr_list : wire_list<rl,rw,rn,rg>@rl ), 
+      reads(rl,rw,rn,rg), writes(rw) : bool =
+  if isnull(ptr_list) then true else
+  let wire_node : wire_list<rl,rw,rn,rg> = read(ptr_list) in
+  let wire : CircuitWire<rn,rg> = read(wire_node.1) in
+  let in_node : CircuitNode = read(wire.1) in
+  let out_node : CircuitNode = read(wire.2) in
+  let current : float = (in_node.1 - out_node.1) /  wire.3 in 
+  let new_wire : CircuitWire<rn,rg> = <wire.1,wire.2,wire.3,current> in
+  let _ : CircuitWire<rn,rg>@rw = write(wire_node.1, new_wire) in
+      calc_new_currents[rl,rw,rn,rg](wire_node.2)
+
+function distribute_charge[rl,rw,rn,rg] ( ptr_list : wire_list<rl,rw,rn,rg>@rl ), 
+      reads(rl,rw,rn), reduces(reduce_charge,rn,rg), atomic(rn,rg) : bool =
+  if isnull(ptr_list) then true else
+  let wire_node : wire_list<rl,rw,rn,rg> = read(ptr_list) in
+  let wire : CircuitWire<rn,rg> = read(wire_node.1) in
+  let _ : CircuitNode@rn = reduce(reduce_charge, wire.1, wire.4) in
+  let _ : CircuitNode@(rn,rg) = reduce(reduce_charge, wire.2, wire.4) in
+      distribute_charge[rl,rw,rn,rg](wire_node.2)
+
+function update_voltage[rl,rn] ( ptr_list : node_list<rl,rn>@rl ), 
+      reads(rl,rn), writes(rn) : bool = 
+  if isnull(ptr_list) then true else
+  let node_node : node_list<rl,rn> = read(ptr_list) in
+  let node : CircuitNode = read(node_node.1) in
+  let voltage : float = (node.3/node.4) in
+  let new_node : CircuitNode = <voltage,node.2,node.3,node.4> in
+  let _ : CircuitNode@rn = write(node_node.1, new_node) in
+      update_voltage[rl,rn](node_node.2)
+
+-- Reduction function for distribute charge
+function reduce_charge ( node : CircuitNode, current : float ) : CircuitNode =
+    let new_charge : float = node.3 + current in
+        < node.1,new_charge,node.3,node.4>
+\end{lstlisting}
 

### 2012-07-10 doc/popl2013/intro.tex
More edits to the circuit example

@@
 \begin{lstlisting}[float={t},label={lst:circuit_ex},caption={Circuilt Simulation}]
-type CircuitNode        = <float/*voltage*/,float/*current*/,
-                           float/*charge*/,float/*capacitance*/>
-type CircuitWire<rn,rg>  = <CircuitNode@rn,      -- Owned node
-                            CircuitNode@(rn,rg), -- Owned or Ghost node
-                            float/*resistance*/, float/*current*/>
+-- <voltage,current,charge,capacitance>
+type CircuitNode        = <float,float,float,float>
+-- < owned node, owned or ghost node, resistance, current>
+type CircuitWire<rn,rg>  = <CircuitNode@rn, CircuitNode@(rn,rg),float,float>
 

### 2012-07-10 doc/popl2013/intro.tex
More circuit example edits

@@
       reads(rn,rw,rl), writes(rn,rw,rl) : bool = 
-  let pc : <coloring(rn),coloring(rn),coloring(rw)> 
+  let pc : <coloring(rn),multicoloring(rn),coloring(rw)> 
             = color_circuit[rn,rw,rl](all_nodes,all_wires) in

### 2012-07-10 doc/popl2013/intro.tex
Updating circuit example in the intro to fix some bugs

--- 2012-07-10 doc/popl2013/intro.tex
@@
   partition rw using pc.3 as rw0,rw1 in
+  let lists0 : <wire_list<rl,rw0,rn0,rg0>@rl,node_list<rl,rn0>@rl> = 
+        build_lists[rl,rw,rn,rw0,rn0,rg0](all_nodes,all_wires,pc.1,pc.2,pc.3,0) in
   let piece0 : CircuitPiece<rl,rw,rn> = 
-                build_piece[rl,rw,rn,rw0,rn0,rg0](all_nodes,all_wires,pc.1,pc.2,pc.3,0) in
+        pack lists0 as CircuitPiece<rl,rw,rn>[rw0,rn0,rg0] in
+  let lists1 : <wire_list<rl,rw1,rn1,rg1>@rl,node_list<rl,rn1>@rl> =
+        build_lists[rl,rw,rn,rw1,rn1,rg1](all_nodes,all_wires,pc.1,pc.2,pc.3,1) in
   let piece1 : CircuitPiece<rl,rw,rn> = 
-                build_piece[rl,rw,rn,rw1,rn1,rg1](all_nodes,all_wires,pc.1,pc.2,pc.3,1) in
+        pack lists1 as CircuitPiece<rl,rw,rn>[rw1,rn1,rg1] in
       execute_time_steps[rl,rw,rn](piece0,piece1,steps)
@@
   -- Construct lists of node and wire pointers for the given colorings
-
-function build_piece[rl,rw,rn,rpw,rpn,rg] ( nodes : node_list<rl,rn>@rl, 
-      wires : wire_list<rl,rw,rn>@rl, oc : multicoloring(rn), gc : coloring(rn), 
-      wc : coloring(rw), c : int), reads(rn,rw,rl), writes(rl) : CircuitPiece<rl,rw,rn> =
-  let piece_lists : < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl > = 
-    build_lists[rl,rw,rn,rpw,rpn,rg](nodes,wires,oc,gc,wc,c) in
-      pack piece_lists as CircuitPiece<rl,rw,rn>[rpw,rpn,rg] 
 \end{lstlisting}

### 2012-07-10 doc/popl2013/example.tex
Adding a separate example section

--- 2012-07-10 doc/popl2013/intro.tex
@@
-\section{Introduction}
-\label{sec:intro}
-
-In the past decade machine architecture, particularly at the high
-performance end of the spectrum, has undergone a revolution.  The
-latest supercomputers are now complex hierarchies of many different
-kinds of computing technologies: networks of nodes at the top level,
-multiple chips per node, mutiple cores within a chip, and, most
-recently, multiple accelerators (usually GPUs) per node, which 
-can themselves be decomposed into several levels.   We present the operational and static
-semantics of Legion, a programming model targeted at providing an
-appropriate level of abstraction for programming such machines, one
-that is both sufficiently high-level to be portable while still
-exposing aspects that are crucial to performance.
-
-In Legion data is organized in a hierarchy of {\em regions}
-and subregions while computation is organized in a hierarchy of {\em
-tasks} and subtasks operating on regions.  Regions and tasks interact
-through a static system of {\em region permissions} specifying 
-operations a task may perform on a region argument (read,
-write, or reduce) and  {\em region coherence} annotations that
-express what other tasks may do concurrently with
-the region (exclusive, atomic, or simultaneous).  We prove the
-soundness of the permissions and coherence system and use these two
-results to prove a third result: if two {\em siblings} (tasks that
-have the same immediate parent task) $t_1$ and $t_2$ are 
-{\em independent} (have no ordering requirements for correctness), 
-then the unique ancestors of $t_1$ and $t_2$ in the task hierarchy 
-are also independent.  This theorem is the basis for correct distributed
-task scheduling in a Legion implementation, which is crucial for
-performance; a centralized scheduler would be a bottleneck
-because of the huge latencies involved in communicating in the target
-class of machines.
+\section{Circuit Example}
+\label{sec:example}
 
@@
 
--- Helper methods
 function color_circuit[rn,rw,rl] ( all_nodes : node_list<rl,rn>@rl, 
@@
 
+-- Helper method
 function build_lists[rl,rw,rn,rpw,rpn,rg] ( nodes : node_list<rl,rn>@rl, 
@@
 \end{lstlisting}
-
-Legion programs organize data into a forest of {\em logical regions}.  Logical
-regions express {\em locality} (data that will be used together, and therefore colocated if possible in the same
-physical memory) and {\em independence} (disjoint data that may be operated on in parallel).
-Researchers have previously explored language designs with hierarchical decomposition of data to express
-locality and independence for high performance programming. Two recent examples, Sequoia \cite{Fatahalion06} and 
-Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a mechanism to partition the heap recursively
-into a tree of collections of data.  The two designs are different in many aspects, but agree that there is
-a single tree-shaped partitioning of data that can be checked fully statically (see Section~\ref{sec:related} for more
-discussion of related work).
 
@@
 \end{lstlisting}
-
-Our own experience with Sequoia and many high-performance applications
-written in the current industry standard mix of MPI, shared-memory
-threads, and CUDA, has taught us that for many practical situations a
-single, fairly static partitioning is insufficient.  In practice, the
-best way to partition data is often a function of the data
-itself---i.e., the partitions need to be dynamically computed.
-Furthermore, it is useful to allow mutliple partitions simultaneously of
-the same data, providing different views on to that data. Thus, in Legion the 
-handling of regions and partitioning of regions is much more dynamic than in 
-previous designs.  In particular,
-\begin{itemize}
-\item  logical regions are first-class values in Legion
-and may be dynamically allocated and stored in data structures;
-
-\item logical regions can be {\em partitioned} into {\em subregions}; the programmer can express arbitrary partitions of
-regions;
-
-\item  a logical region may be dynamically partitioned in multiple different ways.
-\end{itemize}
-A consequence of these decisions is that, through partitioning, a
-given datum may belong to multiple different regions.  For example, if
-a region $R$ participates in three separate disjoint partition
-operations and each partition allocates every element of $R$ to a
-subregion (partitions need not be total in Legion), then every element
-of $R$ is included in three different subregions.  This is why we use
-the term {\em logical} regions: language-level regions serve to name
-sets of data but do not imply anything about physical layout.  There
-is a separate system of {\em physical regions} at runtime that hold
-copies of the data in the language-level logical regions.  In fact,
-the Legion implementation routinely maintains multiple distinct
-physical regions, each with a copy of the data in the corresponding
-logical region, for performance reasons.
-
-A further consequence of placing the same data in multiple logical
-regions as a result of runtime computation is that Legion has a very
-hard static alias analysis problem on regions, which poses a problem
-for parallel computation because knowing regions do not alias is the
-basis for safe parallel execution.  This is why languages such as
-Sequoia and DPJ allow only a single static partition of each
-region---it simplifies static analysis of regions to the point that it
-is practical.
-
-Legion's execution model is that by default, a program's semantics is
-its meaning as a sequential program, which we refer to as the {\em
-program order} execution.  While the
-implementation is not our focus here, if two functions use
-disjoint data (because all of the regions they use are disjoint), then
-Legion may execute them simultaneously.  Also like DPJ,
-Legion has a system of {\em permissions} ({\em read}, {\em write}, and
-{\em reduce}) on regions and an orthogonal system of region {\em
-coherence} modes ({\em excl}, {\em atomic}, and {\em simultaneous})
-that increase the number of situations in which functions can be
-executed in parallel.
-
-The important difference bewteen Legion and the more static approaches
-is that the hardest part of the tests to justify parallel execution is
-done dynamically instead of statically.  The key observation is that
-alias analysis is hard statically but very easy and relatively cheap
-dynamically.  Thus, Legion falls between fully static systems, such as
-Sequoia and DPJ, and fully dynamic approaches such as transactional
-memory.  Legion still has a significant static component; in
-particular, the safety of permissions is checked statically and region
-pointer dereferences are checked statically.  The dynamic alias tests
-are done at the granularity of regions and one check on a region
-itself often serves to prove the safety of many subsequent region
-accesses.  In contrast, because transactional memory has no mechanism
-for grouping heap data into larger collections, it much test the
-overlap between two sets of data on each individual element, which is
-much more expensive.
-
-We begin in Section~\ref{sec:example} with an example program that illustrates
-a typical style of programming in Legion as well as motivating the need for multiple, dynamically computed  partitions of
-a region.  We define Core Legion, a small language suitable for our formal development, in Section~\ref{sec:legioncore}.
-The next four sections each state and prove one of our main results and contributions:
-\begin{itemize}
-\item We prove the soundness of Legion's permissions system (Section~\ref{sec:soundness}).
-
-\item We use the soundness of permissions to prove the soundness of Legion's region coherence modes (Section~\ref{sec:coherence}).
-
-\item We show that if expressions $e_1$ and $e_2$ are {\em non-interfering} (can be executed in parallel), then subexpressions
-$e_1'$ of $e_1$ and $e_2'$ of $e_2$ are also non-interfering (Section~\ref{sec:scheduling}).  This result is the basis
-for a hierarchical, distributed scheduler in the Legion implementation; on the target class of machines, any centralized
-scheduler would be a serious bottleneck.
-
-\item We give experimental evidence for the Legion design.  On three real-world applications, we show that 
-dynamic region pointer checks would be expensive, justifying checking this aspect of the type system statically.
-We also show that the cost of checking aliasing on region permissions is low, showing that a much more expressive
-and dynamic langauge is not incompatible with high performance.
-
-\end{itemize}
-Finally, Section~\ref{sec:related} discusses the most closely related work and
-Section~\ref{sect:conclusion} concludes.

### 2012-07-10 doc/popl2013/intro.tex
Adding a separate example section

@@
 
-% This is a description of how the listings should be formatted.
-% It can go anywhere before the listings.
-\lstset{
-  captionpos=b,
-  language=Haskell,
-  basicstyle=\scriptsize,
-  numbers=left,
-  numberstyle=\tiny,
-  columns=fullflexible,
-  stepnumber=1,
-  escapechar=\#,
-  keepspaces=true,
-  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
-  morekeywords={function,rr,int,float,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,multicoloring,color,newcolor,atomic,simultaneous},
-  deletekeywords={float,head,min,max}
-}
-
-\begin{lstlisting}[float={t},label={lst:circuit_ex},caption={Circuilt Simulation}]
--- <voltage,current,charge,capacitance>
-type CircuitNode        = <float,float,float,float>
--- < owned node, owned or ghost node, resistance, current>
-type CircuitWire<rn,rg>  = <CircuitNode@rn, CircuitNode@(rn,rg),float,float>
-
-type node_list<rl,rn>       = < CircuitNode@rn, node_list<rl,rn>@rl >
-type wire_list<rl,rw,rn,rg>= < CircuitWire<rn,rg>@rw, wire_list<rl,rw,rn,rg>@rl >
-
-type CircuitPiece<rl,rw,rn> = rr[rpw,rpn,rg]
-                            < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl >         
-                            where rpn <= rn and rg <= rn and rpw <= rw and
-                                  rpn * rg and rn * rw and rl * rn and rl * rw
-
--- Simulation initialization and invocation
-function simulate_circuit[rl,rw,rn] ( all_nodes : node_list<rl,rn>@rl, 
-                            all_wires : wire_list<rl,rw,rn,rn>@rl, steps : int ), 
-      reads(rn,rw,rl), writes(rn,rw,rl) : bool = 
-  let pc : <coloring(rn),multicoloring(rn),coloring(rw)> 
-            = color_circuit[rn,rw,rl](all_nodes,all_wires) in
-  -- Disjoint partition for the owned nodes of each piece
-  partition rn using pc.1 as rn0,rn1 in
-  -- Aliased partition for ghost nodes of each piece
-  partition rn using pc.2 as rg0,rg1 in
-  -- Disjoint partition for the owned wires of each piece
-  partition rw using pc.3 as rw0,rw1 in
-  let lists0 : <wire_list<rl,rw0,rn0,rg0>@rl,node_list<rl,rn0>@rl> = 
-        build_lists[rl,rw,rn,rw0,rn0,rg0](all_nodes,all_wires,pc.1,pc.2,pc.3,0) in
-  let piece0 : CircuitPiece<rl,rw,rn> = 
-        pack lists0 as CircuitPiece<rl,rw,rn>[rw0,rn0,rg0] in
-  let lists1 : <wire_list<rl,rw1,rn1,rg1>@rl,node_list<rl,rn1>@rl> =
-        build_lists[rl,rw,rn,rw1,rn1,rg1](all_nodes,all_wires,pc.1,pc.2,pc.3,1) in
-  let piece1 : CircuitPiece<rl,rw,rn> = 
-        pack lists1 as CircuitPiece<rl,rw,rn>[rw1,rn1,rg1] in
-      execute_time_steps[rl,rw,rn](piece0,piece1,steps)
-
--- Time Step Loop
-function execute_time_steps[rl,rw,rn] ( p0 : CircuitPiece<rl,rw,rn>, 
-      p1 : CircuitPiece<rl,rw,rn>, steps : int ) , reads(rn,rw,rl), writes(rn,rw) : bool = 
-  if steps #$<$# 1 then true else
-  unpack p0 as piece0 : CircuitPiece<rl,rw,rn>[rw0,rn0,rg0] in 
-  unpack p1 as piece1 : CircuitPiece<rl,rw,rn>[rw1,rn1,rg1] in
-  let _ : bool = calc_new_currents[rl,rw0,rn0,rg0](piece0.1) in
-  let _ : bool = calc_new_currents[rl,rw1,rn1,rg1](piece1.1) in
-  let _ : bool = distribute_charge[rl,rw0,rn0,rg0](piece0.1) in
-  let _ : bool = distribute_charge[rl,rw1,rn1,rg1](piece1.1) in
-  let _ : bool = update_voltage[rl,rn0](piece0.2) in
-  let _ : bool = update_voltage[rl,rn1](piece1.2) in
-      execute_time_steps[rl,rw,rn](p0,p1,steps-1)
-
--- Helper methods
-function color_circuit[rn,rw,rl] ( all_nodes : node_list<rl,rn>@rl, 
-                               all_wires : wire_list<rl,rw,rn>@rl ), 
-        reads(rn,rw,rl) : <coloring(rn), multicoloring(rn), coloring(rw)> =  
-  -- Invoke programmer chosen coloring algorithm (e.g. METIS)
-  -- return owned, ghost, wire colorings
-
-function build_lists[rl,rw,rn,rpw,rpn,rg] ( nodes : node_list<rl,rn>@rl, 
-       wires : wire_list<rl,rw,rn>@rl, oc : coloring(rn), gc : multicoloring(rn), 
-       wc : coloring(rw), c : int), reads(rn,rw,rl), writes(rl) 
-       : < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl > = 
-  -- Construct lists of node and wire pointers for the given colorings
-\end{lstlisting}
-
 Legion programs organize data into a forest of {\em logical regions}.  Logical
@@
 Researchers have previously explored language designs with hierarchical decomposition of data to express
-locality and independence for high performance programming. Two recent examples, Sequoia \cite{Fatahalion06} and 
+locality and independence for high performance programming. Two recent examples, Sequoia \cite{Fatahalian06} and 
 Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a mechanism to partition the heap recursively
@@
 discussion of related work).
-
-\begin{lstlisting}[float={t},label={lst:circuit_leaf},caption={Circuilt Leaf Tasks}]
-function calc_new_currents[rl,rw,rn,rg] ( ptr_list : wire_list<rl,rw,rn,rg>@rl ), 
-      reads(rl,rw,rn,rg), writes(rw) : bool =
-  if isnull(ptr_list) then true else
-  let wire_node : wire_list<rl,rw,rn,rg> = read(ptr_list) in
-  let wire : CircuitWire<rn,rg> = read(wire_node.1) in
-  let in_node : CircuitNode = read(wire.1) in
-  let out_node : CircuitNode = read(wire.2) in
-  let current : float = (in_node.1 - out_node.1) /  wire.3 in 
-  let new_wire : CircuitWire<rn,rg> = <wire.1,wire.2,wire.3,current> in
-  let _ : CircuitWire<rn,rg>@rw = write(wire_node.1, new_wire) in
-      calc_new_currents[rl,rw,rn,rg](wire_node.2)
-
-function distribute_charge[rl,rw,rn,rg] ( ptr_list : wire_list<rl,rw,rn,rg>@rl ), 
-      reads(rl,rw,rn), reduces(reduce_charge,rn,rg), atomic(rn,rg) : bool =
-  if isnull(ptr_list) then true else
-  let wire_node : wire_list<rl,rw,rn,rg> = read(ptr_list) in
-  let wire : CircuitWire<rn,rg> = read(wire_node.1) in
-  let _ : CircuitNode@rn = reduce(reduce_charge, wire.1, wire.4) in
-  let _ : CircuitNode@(rn,rg) = reduce(reduce_charge, wire.2, wire.4) in
-      distribute_charge[rl,rw,rn,rg](wire_node.2)
-
-function update_voltage[rl,rn] ( ptr_list : node_list<rl,rn>@rl ), 
-      reads(rl,rn), writes(rn) : bool = 
-  if isnull(ptr_list) then true else
-  let node_node : node_list<rl,rn> = read(ptr_list) in
-  let node : CircuitNode = read(node_node.1) in
-  let voltage : float = (node.3/node.4) in
-  let new_node : CircuitNode = <voltage,node.2,node.3,node.4> in
-  let _ : CircuitNode@rn = write(node_node.1, new_node) in
-      update_voltage[rl,rn](node_node.2)
-
--- Reduction function for distribute charge
-function reduce_charge ( node : CircuitNode, current : float ) : CircuitNode =
-    let new_charge : float = node.3 + current in
-        < node.1,new_charge,node.3,node.4>
-\end{lstlisting}
 
@@
 hard static alias analysis problem on regions, which poses a problem
-for parallel computation because knowing regions do not alias is the
+for parallel computation because knowing which regions do not alias is the
 basis for safe parallel execution.  This is why languages such as

### 2012-07-10 doc/popl2013/langdef.tex
additional changes

@@
 \makeatother
+
+% override BCP's typesetrule function to left-justify multiple lines of hypotheses
+\renewcommand{\typesetrule}[2]{%
+   \setrulebody{%
+      \frac{\begin{array}{@{}l@{}}#1\end{array}}%
+           {\begin{array}{@{}l@{}}#2\end{array}}}}
 
@@
 $v$ & ::= & & values &   
-       & $\mid$& $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & task calls \\
+       & $\mid$& $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & function calls \\
   &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & base values &   
@@
 %  &$\mid$& $\text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3$ &  \\
-%  &$\mid$& $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & task calls \\
+%  &$\mid$& $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & function calls \\
 %  &$\mid$& $\text{partition}\ r_p\text{ using }e_1\text{ as }\oton{r}\text{ in }\ e_2$ &  \\
@@
 are used to specify partitions of regions into subregions.
-Functions/tasks in Legion are named, accept one or more arguments and
+Functions in Legion are named, accept one or more arguments and
 return a value of some type.  A function also declares the region
-access privileges and coherence it requires.  Function types are
+access privileges and coherence it requires.  A function may be executed
+within the current task or run as a separate subtask - this decision is made
+by the Legion task scheduler.  Function types are
 universally quantified over all region names appearing in the type.
-The final type is a region relationship, instance of which capture a value and
-one or more regions satisfying associated constraints.  An
-instance of a region relation can be read out of the heap and
+The final type is a region relationship, an instance of which captures a value and
+one or more regions satisfying associated constraints.  A region relation instance
+can be written into the heap and later read and
 unpacked, giving new local names to the regions contained in
@@
 \begin{array}{l@{ }l}
-\vdash \{ & \text{task }id_1[r_{11}, \ldots, r_{1k_1}]( a_{11} : T_{11}, \ldots a_{1n_1} : T_{1n_1} ), \Phi_1, Q_1 : T_{1r} : e_1, \\
+\vdash \{ & \text{function }id_1[r_{11}, \ldots, r_{1k_1}]( a_{11} : T_{11}, \ldots a_{1n_1} : T_{1n_1} ), \Phi_1, Q_1 : T_{1r} : e_1, \\
           & \ldots \\
-          & \text{task }id_p[r_{p1}, \ldots, r_{pk_1}]( a_{p1} : T_{p1}, \ldots a_{pn_p} : T_{pn_p} ), \Phi_p, Q_p : T_{pr} : e_p \} : \bullet
+          & \text{function }id_p[r_{p1}, \ldots, r_{pk_1}]( a_{p1} : T_{p1}, \ldots a_{pn_p} : T_{pn_p} ), \Phi_p, Q_p : T_{pr} : e_p \} : \bullet
 \end{array} }
@@
 \begin{tabular}{m{4cm}m{3.5cm}m{5cm}m{5cm}}
-\infax[BOOL]{\typeenv{bv}{bool}} & 
-\infax[INT]{\typeenv{iv}{int}} & 
-\infrule[TUPLE1]{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.1}\ : T_1} & 
-\infrule[TUPLE2]{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.2}\ : T_2} 
+\infax[T-Bool]{\typeenv{bv}{bool}} & 
+\infax[T-Int]{\typeenv{iv}{int}} & 
+\infrule[T-Tuple1]{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.1}\ : T_1} & 
+\infrule[T-Tuple2]{\typeenvx e : \langle T_1,T_2 \rangle}{\typeenvx e\text{.2}\ : T_2} 
 \end{tabular}
@@
 \begin{tabular}{m{3cm}m{3cm}m{5cm}m{6cm}}
-\infrule[VAR]{\Gamma(id) = T}{\typeenvx id : T} &
-\infax[NULL]{\typeenvx \text{null }T@r : T@r} &
-\infrule[ISNULL]{\typeenvx e : T@(\oton{r})}{\typeenvx \text{isnull}(e) : bool} &
-\infrule[MAKETUPLE]{\typeenvx e_1 : T_1 \andalso \typeenvx e_2 : T_2}{\typeenvx \langle e_1, e_2 \rangle : \langle T_1, T_2 \rangle}
+\infrule[T-Var]{\Gamma(id) = T}{\typeenvx id : T} &
+\infax[T-Null]{\typeenvx \text{null }T@r : T@r} &
+\infrule[T-IsNull]{\typeenvx e : T@(\oton{r})}{\typeenvx \text{isnull}(e) : bool} &
+\infrule[T-MakeTuple]{\typeenvx e_1 : T_1 \andalso \typeenvx e_2 : T_2}{\typeenvx \langle e_1, e_2 \rangle : \langle T_1, T_2 \rangle}
 \end{tabular}
@@
 \begin{tabular}{m{4cm}m{5cm}m{9cm}}
-\infax[NEW]{\typeenvx \text{new }T@r : T@r} &
-\infax[NEWCOLOR]{\typeenvx \text{newcolor }r : \text{coloring}(r)} &
-\infrule[COLORING]{\typeenvx e_1 : \text{coloring}(r) \andalso \typeenvx e_2 : T@r \andalso \typeenvx e_3 : int}{\typeenvx \text{color}(e_1, e_2, e_3) : \text{coloring}(r)}
+\infax[T-New]{\typeenvx \text{new }T@r : T@r} &
+\infax[T-Newcolor]{\typeenvx \text{newcolor }r : \text{coloring}(r)} &
+\infrule[T-Color]{\typeenvx e_1 : \text{coloring}(r) \andalso \typeenvx e_2 : T@r \andalso \typeenvx e_3 : int}{\typeenvx \text{color}(e_1, e_2, e_3) : \text{coloring}(r)}
 \end{tabular}
@@
 \begin{tabular}{m{6cm}m{5cm}m{6cm}}
-\infrule[LET]{\typeenvx e_1 : T_1 \\ \typeenvx[G={\Gamma[id/T_1]}] e_2 : T_2}{\typeenvx : \text{let}\ id : T_1 \text{in}\ e_2 : T_2} &
-\infrule[ADD]{\typeenvx e_1 : int \\ \typeenvx e_2 : int}{\typeenvx e_1 + e_2 : int} &
-\infrule[COMPARE]{\typeenvx e_1 : int \\ \typeenvx e_2 : int}{\typeenvx e_1 < e_2 : bool}
+\infrule[T-Let]{\typeenvx e_1 : T_1 \\ \typeenvx[G={\Gamma[id/T_1]}] e_2 : T_2}{\typeenvx : \text{let}\ id : T_1 \text{in}\ e_2 : T_2} &
+\infrule[T-Add]{\typeenvx e_1 : int \\ \typeenvx e_2 : int}{\typeenvx e_1 + e_2 : int} &
+\infrule[T-Compare]{\typeenvx e_1 : int \\ \typeenvx e_2 : int}{\typeenvx e_1 < e_2 : bool}
 \end{tabular}
@@
 \begin{tabular}{m{9cm}m{10cm}}
-\infrule[UPREG]{\typeenvx e : T@(r'_1, \ldots r'_k) \andalso \forall i. \exists j, r'_i \leq r_j \in \Omega^*}{\typeenvx upregion(e,\oton{r}) : T@(\oton{r})} &
-\infrule[DOWNREG]{\typeenvx e : T@(r'_1, \ldots r'_k)
+\infrule[T-UpRgn]{\typeenvx e : T@(r'_1, \ldots r'_k) \andalso \forall i. \exists j, r'_i \leq r_j \in \Omega^*}{\typeenvx upregion(e,\oton{r}) : T@(\oton{r})} &
+\infrule[T-DownRgn]{\typeenvx e : T@(r'_1, \ldots r'_k)
 % \andalso \forall j. \exists i, r_j \leq r'_i \in \Omega^*
@@
 \begin{tabular}{m{5cm}m{5cm}m{8cm}}
-\infrule[READ]{\typeenvx e_1 : T@(\oton{r}) \\ \forall i, reads(r_i) \in \Phi^*}{\typeenvx \text{read}(e_1) : T} &
-\infrule[WRITE]{\typeenvx e_1 : T@(\oton{r}) \\ \typeenvx e_2 : T \\ \forall i, writes(r_i) \in \Phi^*}{\typeenvx \text{write}(e_1, e_2) : T@(\oton{r})} &
-\infrule[REDUCE]{\Gamma(id) = (T_1, T_2), \emptyset, \emptyset \rightarrow T_1 \andalso \typeenvx e_2 : T_2 \\
+\infrule[T-Read]{\typeenvx e_1 : T@(\oton{r}) \\ \forall i, reads(r_i) \in \Phi^*}{\typeenvx \text{read}(e_1) : T} &
+\infrule[T-Write]{\typeenvx e_1 : T@(\oton{r}) \\ \typeenvx e_2 : T \\ \forall i, writes(r_i) \in \Phi^*}{\typeenvx \text{write}(e_1, e_2) : T@(\oton{r})} &
+\infrule[T-Reduce]{\Gamma(id) = (T_1, T_2), \emptyset, \emptyset \rightarrow T_1 \andalso \typeenvx e_2 : T_2 \\
  \typeenvx e_1 : T_1@(\oton{r}) \andalso \forall i, reduces_{id}(r_i) \in \Phi^*}{\typeenvx \text{reduce}(id, e_1, e_2) : T_1@(\oton{r})}
@@
 \begin{tabular}{m{5cm}m{5cm}m{8cm}}
-\infrule[PACK]{T_1 = \exists r'_1, \ldots r'_n.\ T_2\text{ where }\Omega_1 \\ \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^* \\ \typeenvx e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]}{\typeenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] : T_1} &
-\infrule[IF-ELSE]{\typeenvx e_1 : bool \\ \typeenvx e_2 : T \\ \typeenvx e_3 : T}{\typeenvx \text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3 : T} &
-\infrule[TASK-CALL]{\Gamma(id) = \forall r'_1, \ldots r'_k.(\oton{T}),\Phi', Q' \rightarrow T_r \\ \typeenvx e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\ \Phi'[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*}{\typeenvx id[\otok{k}{r}](\oton{e}) : T_r[r_1/r'_1,\ldots,r_k/r'_k]}
+\infrule[T-Pack]{T_1 = \exists r'_1, \ldots r'_n.\ T_2\text{ where }\Omega_1 \\ \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^* \\ \typeenvx e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]}{\typeenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] : T_1} &
+\infrule[T-IfElse]{\typeenvx e_1 : bool \\ \typeenvx e_2 : T \\ \typeenvx e_3 : T}{\typeenvx \text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3 : T} &
+\infrule[T-Call]{\Gamma(id) = \forall r'_1, \ldots r'_k.(\oton{T}),\Phi', Q' \rightarrow T_r \\ \typeenvx e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\ \Phi'[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*}{\typeenvx id[\otok{k}{r}](\oton{e}) : T_r[r_1/r'_1,\ldots,r_k/r'_k]}
 \end{tabular}
@@
 \begin{tabular}{m{10cm}m{8cm}}
-\infrule[UNPACK]{T_1 = \exists r'_1, \ldots, r'_n.\ T_2\text{ where }\Omega_1 \andalso \typeenvx e_1 : T_1 \\
+\infrule[T-Unpack]{T_1 = \exists r'_1, \ldots, r'_n.\ T_2\text{ where }\Omega_1 \andalso \typeenvx e_1 : T_1 \\
   \Gamma' = \Gamma[T_2[r_1/r'_1,\ldots,r_k/r'_k] / id] \andalso \Omega' = \Omega \cup \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \\
@@
   {\typeenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 : T_3} &
-\infrule[PARTITION]{\typeenvx e_1 : \text{coloring}(r_p) \\ \Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{1 \leq i < j \leq k} r_i * r_j \\ \typeenvx[O=\Omega'] e_2 : T \\ \{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T) = \emptyset}
+\infrule[T-Part]{\typeenvx e_1 : \text{coloring}(r_p) \\ \Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{1 \leq i < j \leq k} r_i * r_j \\ \typeenvx[O=\Omega'] e_2 : T \\ \{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T) = \emptyset}
   {\typeenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 : T}
@@
 \vspace{-0.5cm}
-\infrule[PROGRAM]{\text{for }1 \leq i \leq p, \\
+\infrule[T-Program]{\text{for }1 \leq i \leq p, \\
   \Gamma(id_i) = \forall r_{i1}, \ldots r_{ik_i}. (T_{i1}, \ldots, T_{in_i}), \Phi_i, Q_i \rightarrow T_{ir} \andalso
@@
   \typeenv[G={\Gamma_i},P={\Phi_i},O=\emptyset]{e_i}{T_{ir}}}
-  {\vdash \{  \text{task }id_1[r_{11}, \ldots, r_{1k_1}]( a_{11} : T_{11}, \ldots a_{1n_1} : T_{1n_1} ), \Phi_1, Q_1 : T_{1r} : e_1, 
+  {\vdash \{  \text{function }id_1[r_{11}, \ldots, r_{1k_1}]( a_{11} : T_{11}, \ldots a_{1n_1} : T_{1n_1} ), \Phi_1, Q_1 : T_{1r} : e_1, 
            \ldots
-           \text{task }id_p[r_{p1}, \ldots, r_{pk_1}]( a_{p1} : T_{p1}, \ldots a_{pn_p} : T_{pn_p} ), \Phi_p, Q_p : T_{pr} : e_p \} : \bullet}
+           \text{function }id_p[r_{p1}, \ldots, r_{pk_1}]( a_{p1} : T_{p1}, \ldots a_{pn_p} : T_{pn_p} ), \Phi_p, Q_p : T_{pr} : e_p \} : \bullet}
 \end{minipage}
@@
 
-\item $H$ is a {\em heap typing}, assigning an invariant type to every memory location.
+\item $H$ is a {\em heap typing}, assigning an invariant type to every memory location.  The heap typing
+serves two important roles.  First, it is used to bound the values that can be used for reads when 
+concurrent operation may make the exact value uncertain.  Second, it greatly simplifies the consistency
+check of pointer values, allowing a simple test against the heap typing for a location in lieu of a
+potentially unbounded walk through the store for complicated pointer data structures.
 
@@
 
-\item $mark\_coherence(E, \hat Q)$ records the coherence of operations in trace $E$ according to coherence annotations $Q$.
+\item $mark\_coh(E, \hat Q)$ records the coherence of operations in trace $E$ according to coherence annotations $\hat Q$.
 
@@
 {\em valid\_interleave} is given in Section~\ref{sec:coherence}.  For now, we need only that a valid interleaving
-is at least an interleaving, containing each constituent trace as a subequence and no other operations.
+is at least an interleaving, containing each constituent trace as a subsequence and no other operations.
 
@@
 in the clobber set $C$, the set of locations concurrent tasks may
-clobber (i.e., update).\footnote{Note that this is also the reason for
-including a global heap typing.}  The guess is eventually checked for
+clobber (i.e., update).  The guess is eventually checked for
 consistency by the $valid\_interleave$ predicate of some enclosing
@@
 
-The most involved rule in Core Legion's semantics is for task calls.
+The most involved rule in Core Legion's semantics is for function calls.
 Note that the only regions accessible to the callee (through $M'$) are the region arguments.
@@
 $
-\begin{array}{lll}
+\begin{array}{@{}l@{}}
+\begin{array}{@{}l@{\hspace{-0.25in}}l@{}} 
+\begin{array}{@{}lll}
 apply(S, []) & = & S \\
@@
 apply(S, E \dplus [ write(l, c, v) ]) & = & apply(S, E)[v/l] \\
-apply(S, E \dplus [ reduce_{id}(l, c, v) ]) & = & S'[id(S'(l), v)/l], \text{ where } S' = apply(S, E) \\
-mark\_coherence([], \hat Q) & = & [] \\
-mark\_coherence(E \dplus [ op(l, c, v) ], \hat Q) & = & mark\_coherence(E, \hat Q) \dplus \\
-& & op(l, c', v), \text{ where } \\
-& & c' =
+apply(S, E \dplus [ reduce_{id}(l, c, v) ]) & = & S'[id(S'(l), v)/l], \\
+& & \text{ where } S' = apply(S, E)
+\end{array} &
+\begin{array}{@{}ll}
+mark\_coh([], \hat Q) & = [] \\
+mark\_coh([ op(l, c, v) ] \dplus E, \hat Q) & = [ op(l, c', v) ] \dplus mark\_coh(E, \hat Q), \\
+\multicolumn{2}{@{}l}{ \hspace{1.0in} \text{ where } c' =
 \begin{cases}
@@
 excl, & \text{otherwise}
-\end{cases} \\
+\end{cases}}
+\end{array}
+\end{array} \\
+\begin{array}{@{}lll}
 any\_interleave([], [], \ldots, []) & = & true \\
@@
 valid\_interleave(S, C, E', \oton{E}) & \Rightarrow & any\_interleave(E', \oton{E}) \\
+\end{array}
 \end{array}
@@
 \vspace{1.5mm} \\
-\text{task }id[\otok{k}{r'}](a_1 : T_1, \ldots, a_n : T_n), \Phi', Q' : T_r = e_{n+1} \\
+\text{function }id[\otok{k}{r'}](a_1 : T_1, \ldots, a_n : T_n), \Phi', Q' : T_r = e_{n+1} \\
 M' = \{ (r'_1, M(r_1)), \ldots (r'_k, M(r_k)) \} \\
@@
 \begin{tabular}{m{6cm}m{6cm}m{6cm}}
-\infax[OP-BOOL]{\opsenvx bv \mapsto bv, []} &
-\infax[OP-INT]{\opsenvx iv \mapsto iv, []} &
-\infax[OP-NULL]{\opsenvx \text{null }T@r \mapsto null, []} 
+\infax[E-Bool]{\opsenvx bv \mapsto bv, []} &
+\infax[E-Int]{\opsenvx iv \mapsto iv, []} &
+\infax[E-Null]{\opsenvx \text{null }T@r \mapsto null, []} 
 \end{tabular}
@@
 \begin{tabular}{m{9cm}m{4.5cm}m{4.5cm}}
-\infrule[OP-MAKETUPLE]{\opsenvx e_1 \mapsto v_1, E_1 \andalso S' = apply(S, E_1) \\
+\infrule[E-MakeTuple]{\opsenvx e_1 \mapsto v_1, E_1 \andalso S' = apply(S, E_1) \\
   \opsenvx[S=S'] e_2 \mapsto v_2, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)}
   {\opsenvx \langle e_1, e_2 \rangle \mapsto \langle v_1, v_2 \rangle, E'} &
-\infrule[OP-TUPLE1]{\opsenvx e \mapsto \langle v_1, v_2 \rangle, E}{\opsenvx e\text{.1} \mapsto v_1, E} &
-\infrule[OP-TUPLE2]{\opsenvx e \mapsto \langle v_1, v_2 \rangle, E}{\opsenvx e\text{.2} \mapsto v_2, E}
+\infrule[E-Tuple1]{\opsenvx e \mapsto \langle v_1, v_2 \rangle, E}{\opsenvx e\text{.1} \mapsto v_1, E} &
+\infrule[E-Tuple2]{\opsenvx e \mapsto \langle v_1, v_2 \rangle, E}{\opsenvx e\text{.2} \mapsto v_2, E}
 \end{tabular}
@@
 \begin{tabular}{m{4cm}m{4.5cm}m{5cm}m{5cm}}
-\infrule[OP-NEW]{l \in M(r) \\ H(l) = M \llbracket T \rrbracket}{\opsenvx \text{new }T@r \mapsto l, []} &
-\infrule[OP-VAL]{L(id) = v}{\opsenvx id \mapsto v, []} &
-\infrule[OP-ISNULL-FALSE]{\opsenvx e \mapsto l, E}{\opsenvx \text{isnull}(e) \mapsto \textit{false}, E} &
-\infrule[OP-ISNULL-TRUE]{\opsenvx e \mapsto null, E}{\opsenvx \text{isnull}(e) \mapsto true, E}
+\infrule[E-New]{l \in M(r) \\ H(l) = M \llbracket T \rrbracket}{\opsenvx \text{new }T@r \mapsto l, []} &
+\infrule[E-Var]{L(id) = v}{\opsenvx id \mapsto v, []} &
+\infrule[E-IsNull-F]{\opsenvx e \mapsto l, E}{\opsenvx \text{isnull}(e) \mapsto \textit{false}, E} &
+\infrule[E-IsNull-T]{\opsenvx e \mapsto null, E}{\opsenvx \text{isnull}(e) \mapsto true, E}
 \end{tabular}
@@
 \begin{tabular}{m{6cm}m{6cm}m{6cm}}
-\infrule[OP-LET]{\opsenvx e_1 \mapsto v_1, E_1 \\ L' = L[v_1/id] \\ S' = \text{apply}(S, E_1) \\ \opsenvx[L=L',S=S'] e_2 \mapsto v_2, E_2 \\ valid\_interleave(S, C, E', E_1, E_2)}{\opsenvx \text{let }id : T = e_1\text{ in }e_2 \mapsto v_2, E'} &
-\infrule[OP-ADD]{\opsenvx e_1 \mapsto v_1, E_1 \\ S' = \text{apply}(S, E_1) \\ \opsenvx[S=S'] e_2 \mapsto v_2, E_2 \\ v' = v_1 + v_2 \\ valid\_interleave(S, C, E', E_1, E_2)}{\opsenvx e_1 + e_2 \mapsto v', E'} &
-\infrule[OP-COMPARE]{\opsenvx e_1 \mapsto v_1, E_1 \\ S' = \text{apply}(S, E_1) \\ \opsenvx[S=S'] e_2 \mapsto v_2, E_2 \\ v' = \begin{cases} true, & \text{if $v_1 < v_2$}.
+\infrule[E-Let]{\opsenvx e_1 \mapsto v_1, E_1 \\ L' = L[v_1/id] \\ S' = \text{apply}(S, E_1) \\ \opsenvx[L=L',S=S'] e_2 \mapsto v_2, E_2 \\ valid\_interleave(S, C, E', E_1, E_2)}{\opsenvx \text{let }id : T = e_1\text{ in }e_2 \mapsto v_2, E'} &
+\infrule[E-Add]{\opsenvx e_1 \mapsto v_1, E_1 \\ S' = \text{apply}(S, E_1) \\ \opsenvx[S=S'] e_2 \mapsto v_2, E_2 \\ v' = v_1 + v_2 \\ valid\_interleave(S, C, E', E_1, E_2)}{\opsenvx e_1 + e_2 \mapsto v', E'} &
+\infrule[E-Compare]{\opsenvx e_1 \mapsto v_1, E_1 \\ S' = \text{apply}(S, E_1) \\ \opsenvx[S=S'] e_2 \mapsto v_2, E_2 \\ v' = \begin{cases} true, & \text{if $v_1 < v_2$}.
 \\ \textit{false}, & \text{otherwise}.
@@
 \begin{tabular}{m{6.5cm}m{6.5cm}m{6cm}}
-\infrule[OP-IF-ELSE-TRUE]{\opsenvx e_1 \mapsto true, E_1 \\ S' = \text{apply}(S, E_1) \\ \opsenvx[S=S'] e_2 \mapsto v_2, E_2}{\opsenvx \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_2, E_1 \dplus E_2} &
-\infrule[OP-IF-ELSE-FALSE]{\opsenvx e_1 \mapsto \textit{false}, E_1 \\ S' = \text{apply}(S, E_1) \\ \opsenvx[S=S'] e_3 \mapsto v_3, E_3}{\opsenvx \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_3, E_1 \dplus E_3} &
-\infrule[OP-NEW-COLOR]{K = \{ (l_1, iv_1), \ldots, (l_p, iv_p) \}, \text{ where } \\ (\forall i \in [1,p].
+\infrule[E-IfElse-T]{\opsenvx e_1 \mapsto true, E_1 \\ S' = \text{apply}(S, E_1) \\ \opsenvx[S=S'] e_2 \mapsto v_2, E_2}{\opsenvx \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_2, E_1 \dplus E_2} &
+\infrule[E-IfElse-F]{\opsenvx e_1 \mapsto \textit{false}, E_1 \\ S' = \text{apply}(S, E_1) \\ \opsenvx[S=S'] e_3 \mapsto v_3, E_3}{\opsenvx \text{if }e_1\text{ then }e_2\text{ else }e_3 \mapsto v_3, E_1 \dplus E_3} &
+\infrule[E-Newcolor]{K = \{ (l_1, iv_1), \ldots, (l_p, iv_p) \}, \text{ where } \\ (\forall i \in [1,p].
 l_i \in M(r)) \wedge (\forall i,j \in [1,p].
@@
 \begin{tabular}{m{8.5cm}m{10.5cm}}
-\infrule[OP-PARTITION]{\opsenv{e_1}{K}{E_1} \andalso \rho_i = \{ l : (l, i) \in K \}, \text{ for } 1 \leq i \leq k \\
+\infrule[E-Part]{\opsenv{e_1}{K}{E_1} \andalso \rho_i = \{ l : (l, i) \in K \}, \text{ for } 1 \leq i \leq k \\
   M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \andalso S' = \text{apply}(S, E_1) \\
@@
   {\opsenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 \mapsto v, E'} &
-\infrule[OP-COLOR]{\opsenvx e_1 \mapsto K, E_1 \andalso S' = \text{apply}(S, E_1) \\
+\infrule[E-Color]{\opsenvx e_1 \mapsto K, E_1 \andalso S' = \text{apply}(S, E_1) \\
   \opsenvx[S=S'] e_2 \mapsto l, E_2 \andalso S'' = \text{apply}(S', E_2) \\
@@
 \begin{tabular}{m{6cm}m{6.5cm}m{6.5cm}}
-\infrule[OP-UPREG]{\opsenvx e \mapsto v, E}{\opsenvx \text{upregion}(e, \oton{r}) \mapsto v, E} &
-\infrule[OP-DOWNREG]{\opsenvx e \mapsto l, E \\ v = \begin{cases} l, & \text{if $\exists i, l \in M(r_i)$}.
+\infrule[E-UpRgn]{\opsenvx e \mapsto v, E}{\opsenvx \text{upregion}(e, \oton{r}) \mapsto v, E} &
+\infrule[E-DownRgn]{\opsenvx e \mapsto l, E \\ v = \begin{cases} l, & \text{if $\exists i, l \in M(r_i)$}.
 \\ null, & \text{otherwise}.
 \end{cases}}{\opsenvx \text{downregion}(e, \oton{r}) \mapsto v, E} &
-\infrule[OP-READ]{\opsenvx e \mapsto l, E \andalso S' = \text{apply}(S, E) \\ v = \begin{cases} S'(l), & \text{if } l \not\in C \\ v' : H(l), & \text{otherwise} \end{cases}}{\opsenvx \text{read}(e) \mapsto v, E \dplus [ read(l, excl, v) ]}
+\infrule[E-Read]{\opsenvx e \mapsto l, E \andalso S' = \text{apply}(S, E) \\ v = \begin{cases} S'(l), & \text{if } l \not\in C \\ v' : H(l), & \text{otherwise} \end{cases}}{\opsenvx \text{read}(e) \mapsto v, E \dplus [ read(l, excl, v) ]}
 \end{tabular}
@@
 \begin{tabular}{m{9cm}m{9cm}}
-\infrule[OP-WRITE]{\opsenvx e_1 \mapsto l, E_1 \andalso S' = \text{apply}(S, E_1) \\
+\infrule[E-Write]{\opsenvx e_1 \mapsto l, E_1 \andalso S' = \text{apply}(S, E_1) \\
   \opsenvx[S=S'] e_2 \mapsto v, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)}
   {\opsenvx \text{write}(e_1, e_2) \mapsto l, E' \dplus [ write(l, excl, v) ]} &
-\infrule[OP-REDUCE]{\opsenvx e_1 \mapsto l, E_1 \andalso S' = \text{apply}(S, E_1) \\
+\infrule[E-Reduce]{\opsenvx e_1 \mapsto l, E_1 \andalso S' = \text{apply}(S, E_1) \\
   \opsenvx[S=S'] e_2 \mapsto v, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)}
@@
 \begin{tabular}{m{8cm}m{10cm}}
-\infrule[OP-PACK]{\opsenvx e_1 \mapsto v, E \andalso \rho_i = M[r_i], \text{ for } 1 \leq i \leq k \\
+\infrule[E-Pack]{\opsenvx e_1 \mapsto v, E \andalso \rho_i = M[r_i], \text{ for } 1 \leq i \leq k \\
   v' = \langle \langle \otok{k}{\rho}, v \rangle \rangle}
   {\opsenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] \mapsto v', E} &
-\infrule[OP-UNPACK]{\opsenvx e_1 \mapsto \langle \langle \otok{k}{\rho} , v_1 \rangle \rangle, E_1 \andalso M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \\
+\infrule[E-Unpack]{\opsenvx e_1 \mapsto \langle \langle \otok{k}{\rho} , v_1 \rangle \rangle, E_1 \andalso M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \\
   L' = L[v_1/id] \andalso S' = \text{apply}(S, E_1) \\
@@
 %\vspace{-0.5cm}
-\infrule[OP-Task-CALL]{\opsenvx e_1 \mapsto v_1, E_1 \andalso S_1 = \text{apply}(S, E_1) \andalso \opsenvx e_2 \mapsto v_2, E_2 \andalso \ldots \andalso S_n = \text{apply}(S_{n-1}, E_n) \\
-  valid\_interleave(S, C, E', \oton{E}) \andalso \text{task }id[\otok{k}{r'}](a_1 : T_1, \ldots, a_n : T_n), \Phi', Q' : T_r = e_{n+1} \\
+\infrule[E-Call]{\opsenvx e_1 \mapsto v_1, E_1 \andalso S_1 = \text{apply}(S, E_1) \andalso \opsenvx e_2 \mapsto v_2, E_2 \andalso \ldots \andalso S_n = \text{apply}(S_{n-1}, E_n) \\
+  valid\_interleave(S, C, E', \oton{E}) \andalso \text{function }id[\otok{k}{r'}](a_1 : T_1, \ldots, a_n : T_n), \Phi', Q' : T_r = e_{n+1} \\
   M' = \{ (r'_1, M(r_1)), \ldots (r'_k, M(r_k)) \} \andalso L' = \{ (a_1, v_1), \ldots, (a_n, v_n) \} \andalso S' = \text{apply}(S, E') \\
@@
 
-Two additional properties are proven for each subexpression:
+\noindent Two additional properties are proven for each subexpression:
 \begin{itemize}
@@
 for the execution environment of subexpressions and one for each of
-the three properties of the execution's results we wish to show.
+the three properties of the execution's results we wish to show:
 
@@
 
-\begin{itemize}
-
-\item[$\mapconsist{\Omega}$] Three expressions have subexpressions that modify $M$ or $\Omega$ and therefore do not
+\vspace{4pt}
+\hangindent=2\parindent \hangafter=1 \noindent %
+$\mapconsist{\Omega}$ - Three expressions have subexpressions that modify $M$ or $\Omega$ and therefore do not
 trivially satisfy region mapping consistency.  For {\tt partition},
@@
 
-\item[$\localconsist{L}{\Gamma}$] Four expressions have subexpressions that modify $L$, $\Gamma$, or $M$.  For
+\vspace{4pt}
+\hangindent=2\parindent \hangafter=1 \noindent %
+$\localconsist{L}{\Gamma}$ -  Four expressions have subexpressions that modify $L$, $\Gamma$, or $M$.  For
 {\tt partition}, which only modifies $M$, the requirement that it not reuse existing
@@
 
-\item[$\storeconsist{S}{H}$] The heap typing consistency of all stores used in subexpressions follows directly from the following
+\vspace{4pt}
+\hangindent=2\parindent \hangafter=1 \noindent %
+$\storeconsist{S}{H}$ -  The heap typing consistency of all stores used in subexpressions follows directly from the following
 lemma: If $\storeconsist{S}{H}$ and $\traceconsist{E}$, then $\storeconsist{\text{apply}(S, E)}{H}$.
 
-\end{itemize}
-
+\vspace{4pt}
 With the necessary consistency conditions proven for a subexpression, the conclusion of the theorem follows for
@@
 conclusion for an entire expression:
-\begin{itemize}
-
-\item[$\valueconsist{v}{T}$]    The consistency of 
+
+\vspace{4pt}
+\hangindent=2\parindent \hangafter=1 \noindent %
+$\valueconsist{v}{T}$ - The consistency of 
 {\tt upregion} is guaranteed by the type checking requirement of appropriate subregion
@@
 
+\hangindent=2\parindent \hangafter=1 \indent\indent\indent %
 The remaining interesting cases arise from changes to the mapping $M$ rather than transformations on the
@@
 
-\item[$\traceconsist{E}$] The type consistency of the values in an expression's memory
+\vspace{4pt}
+\hangindent=2\parindent \hangafter=1 \noindent %
+$\traceconsist{E}$ -  The type consistency of the values in an expression's memory
 trace follows from the following lemma applied to the traces of subexpressions:
@@
 
-\item[$\privconsist{E}{\Phi}$] The proof of the crucial property of containment of effects within the available privileges
+\vspace{4pt}
+\hangindent=2\parindent \hangafter=1 \noindent %
+$\privconsist{E}{\Phi}$ -  The proof of the crucial property of containment of effects within the available privileges
 is similar in outline to the previous step.  The easy cases are covered by a lemma combining the corresponding results
@@
 Straightforward proofs cover {\tt read}, {\tt write}, {\tt reduce}, and one final special case for function calls.
-\end{itemize}
 
@@
 
-Theorem~\ref{thm:effects} guarantees that a task produces a value of the
+Theorem~\ref{thm:effects} guarantees that a function call produces a value of the
 correct type and does so within the regions of memory for which privileges are held.
@@
 }
-We are now able to state the key result that allows the Legion runtime to safely run subtasks in parallel.
+We are now able to state the key result that allows the Legion runtime to safely evaluate subexpressions as
+separate tasks in parallel.
 
@@
 making the worst-case assumption that all operations use $excl$ mode.  However, if two or more
-subexpressions contain calls to tasks using $atomic$ or $simult$ coherence modes
+subexpressions contain calls to functions using $atomic$ or $simult$ coherence modes
 on some regions, executions besides serial program order may 
@@
 captures the hierarchical nature of the scheduler used by the Legion runtime.  As the cost of
-a scheduling decision is quadratic in the number of tasks (more precisely, the number of 
-privileges for those tasks), it should be clear that a hierarchical scheduling algorithm will
+a scheduling decision is quadratic in the number of function calls (more precisely, the number of 
+privileges for those calls), it should be clear that a hierarchical scheduling algorithm will
 be vastly more efficient than a global algorithm that must consider the privileges of every
@@
 } The Legion task scheduler uses Theorem~\ref{thm:hiersched} in the
-following way.  Sibling task calls (tasks invoked within the same
+following way.  Sibling function calls (those invoked within the same
 function body) are checked for non-interference of their (dynamic)
 privileges.  If they interfere they are serialized; otherwise they are
-considered for parallel execution.  Although determining that $e_1$
+considered for parallel execution as parallel subtasks.  Although determining that $e_1$
 and $e_2$ are noninterfering guarantees $e_3$ and $e_4$ are

### 2012-07-10 doc/popl2013/example.tex
Major work on the example and some edits to the intro

@@
 \label{sec:example}
+
+Listing~\ref{lst:circuit_ex} shows code for a
+circuit simulation written in the core Legion language introduced in
+Section~\ref{sec:legioncore} (with one exception described below).  The
+circuit simulation takes as input a large graph of circuit elements and wires
+connecting the elements.  To perform the simulation in parallel the graph
+is partitioned into pieces.  After partitioning
+the graph the simulation is run for a number of time steps.  Each time step
+consists of three phases: calculating the current on each wire, distributing
+charge to the nodes to which each wire is connected, and updating the voltage
+of each node.
+
+The top level function in the application is {\tt simulate\_circuit} (line 15).  The {\tt simulate\_circuit} 
+function is parameterized on the regions {\tt rl}, {\tt rw}, and {\tt rn} (line 15) which 
+implies that the when the function is invoked it must have privileges for the regions 
+bound to those parameters.  Furthermore, these privileges must be for both reading 
+and writing these regions (line 17).  The only logical regions which {\tt simulate\_circuit} will 
+be allowed to access are these regions and their logical subregions.  This property will be 
+enforced by the type system introduced in Section~\ref{sec:legioncore}.  {\tt simulate\_circuit}
+takes pointers to the lists of nodes and wires that describe the graph.  The nodes
+are stored in the {\tt rn} region while wires are stored in the {\tt rw} region.  The
+lists live in the {\tt rl} region.
+
+The first step in {\tt simulate\_circuit} is to partition the graph into pieces.  Partitioning
+is done by constructing {\em colorings} using the {\tt color\_circuit} function (line 18-19).  A
+coloring is a mapping from region elements to colors with each color corresponding to a new subregion
+to be created.  The algorithm for performing a coloring is an application specific decision made by 
+the programmer.  Note, that in the case of circuit, it would be impossible to
+pick a single static partitioning scheme that would do well for all graphs.  Dynamic partitioning
+provides the programmer the flexibility to make partitioning decisions based on input data.
+
+The {\tt color\_circuit} function returns three colorings.  The first coloring is a total 
+coloring that maps each node in the graph the piece that owns it.  The third coloring is a total
+coloring that maps each wire in the graph to a piece which owns on of its nodes.  The second coloring captures a
+different property of the graph: those nodes which must be shared between different pieces of the
+graph to perform computations in parallel.  These are often referred to as {\em ghost nodes}.  We
+describe how ghost node regions are used in conjunction with the owned nodes later.
+Note that its possible some ghost nodes are shared between multiple pieces, and can be colored multiple ways.  Thus,
+the second coloring is referred to as a {\em multicoloring} because it might map the same element
+to multiple colors.
+
+After creating the colorings, the application partitions the region into {\em subregions} based on 
+the colorings (lines 21-25).  For simplicity, our example only partitions the circuit into two pieces.  Lines
+21 and 23 create two different partitions with two different subregions of the {\tt rn} region.  The first partition is
+{\em disjoint} because it is based on a coloring while the second partition is {\em aliased}
+because it is based on a multicoloring which does not imply disjointness of subregions.  While 
+aliased partitions based on multicolorings are not a part of core Legion described in Section~\ref{sec:legioncore}, 
+they are functionally equivalent to creating single partitions for each color in the multicoloring.
+
+Disjoint partitions introduce
+disjointness constraints into the static environment indicating that all the subregions in the partition are 
+disjoint (e.g. {\tt rn0}$*${\tt rn1} on line 21 says {\tt rn0} is disjoint from {\tt rn1}).
+Both disjoint and aliased partitions introduce constraints into the environment indicating that all
+created regions are logical subregions of the partitioned parent region (e.g. {\tt rn0}$<${\tt rn}).
+
+The circuit example illustrates an example of partitioning called {\em allocate-then-partition}
+where the data structure is allocated first and parititoned later.  Alternatively, a
+{\em partition-then-allocate} scheme would allow the creation and paritioning of regions that
+are later populated with data.  Sequoia supports a static form of the former while DPJ supports
+a static form of the later.  Legion enables both approaches dynamically.
+
+After partitioning the circuit into pieces, the application creates instances of {\tt CircuitPiece} 
+(defined on lines 9-12) for each piece (lines 28-33).  {\tt CircuitPiece} is a bounded existential 
+type called a {\em region relationship}.  
+Region relationships allow the programmer to {\em pack} a group of regions together and remember properties about
+them such as disjointness and sub-region relationships (lines 29 and 33).  The type system verifies 
+the properties hold statically when packing so when region relationships are {\em unpacked} the same properties can
+be reintroduced into the static checking environment (lines 40-41).  One important feature of Legion
+is that privileges cannot be packed in a region relationship but can only be passed through function calls.
+
+The {\tt execute\_time\_steps} function represents the primary loop of the simulation (lines 37-48).  Here
+we see the importance of having multiple views onto the same logical region.  Both the the 
+{\tt calc\_new\_currents} and {\tt distribute\_charge} function calls bind the owned region of a piece 
+and the ghost region of a piece.  In the case of {\tt calc\_new\_currents} these regions only need
+read privileges, allowing both instances of {\tt calc\_new\_currents} to be run in parallel.  In the
+case of {\tt distribute\_charge} the privilege is for a reduction which can also be done 
+in parallel because of the atomic and commutative nature of reductions.  If it were only possible to have a single
+partition of the region {\tt rn} it would have been impossible to describe these data sharing patterns.
+
+Listing~\ref{lst:circuit_leaf} shows the leaf functions for each phase of the computation.  Each
+function iterates over the list of wires or nodes for its piece.  Each function specifies the privileges
+that it must have for all of its regions (lines 2,14,23).  These privileges are statically checked
+to match the operations performed inside of each function (e.g. read and write).  In the case of {\em reduce}
+the privilege must also specify which reduction function is being used (line 14).
+
+In addition to privileges, functions can also specify {\em coherence} on regions.  Coherence specifies what 
+values the function is allowed to see from other functions using the same region.  If not otherwise specified,
+coherence defaults to {\em exclusive} which means the function must appear to execute as if the whole program
+were run sequentialy, which we refer to as {\em program order}.  Line 14 shows an example of a relaxed
+coherence mode called {\tt atomic} which says that reductions to {\tt rn} and {\tt rg} must appear atomic relative to other
+functions using those regions.  The most relaxed coherence mode is {\tt simult} which means multiple functions
+all with simultaneous coherence are allowed to access the region concurrently.
 
@@
 
-\begin{lstlisting}[float={t},label={lst:circuit_ex},caption={Circuilt Simulation}]
--- <voltage,current,charge,capacitance>
-type CircuitNode        = <float,float,float,float>
--- < owned node, owned or ghost node, resistance, current>
-type CircuitWire<rn,rg>  = <CircuitNode@rn, CircuitNode@(rn,rg),float,float>
-
-type node_list<rl,rn>       = < CircuitNode@rn, node_list<rl,rn>@rl >
-type wire_list<rl,rw,rn,rg>= < CircuitWire<rn,rg>@rw, wire_list<rl,rw,rn,rg>@rl >
-
-type CircuitPiece<rl,rw,rn> = rr[rpw,rpn,rg]
-                            < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl >         
-                            where rpn <= rn and rg <= rn and rpw <= rw and
-                                  rpn * rg and rn * rw and rl * rn and rl * rw
-
--- Simulation initialization and invocation
-function simulate_circuit[rl,rw,rn] ( all_nodes : node_list<rl,rn>@rl, 
-                            all_wires : wire_list<rl,rw,rn,rn>@rl, steps : int ), 
-      reads(rn,rw,rl), writes(rn,rw,rl) : bool = 
-  let pc : <coloring(rn),multicoloring(rn),coloring(rw)> 
-            = color_circuit[rn,rw,rl](all_nodes,all_wires) in
-  -- Disjoint partition for the owned nodes of each piece
-  partition rn using pc.1 as rn0,rn1 in
-  -- Aliased partition for ghost nodes of each piece
-  partition rn using pc.2 as rg0,rg1 in
-  -- Disjoint partition for the owned wires of each piece
-  partition rw using pc.3 as rw0,rw1 in
-  let lists0 : <wire_list<rl,rw0,rn0,rg0>@rl,node_list<rl,rn0>@rl> = 
-        build_lists[rl,rw,rn,rw0,rn0,rg0](all_nodes,all_wires,pc.1,pc.2,pc.3,0) in
-  let piece0 : CircuitPiece<rl,rw,rn> = 
-        pack lists0 as CircuitPiece<rl,rw,rn>[rw0,rn0,rg0] in
-  let lists1 : <wire_list<rl,rw1,rn1,rg1>@rl,node_list<rl,rn1>@rl> =
-        build_lists[rl,rw,rn,rw1,rn1,rg1](all_nodes,all_wires,pc.1,pc.2,pc.3,1) in
-  let piece1 : CircuitPiece<rl,rw,rn> = 
-        pack lists1 as CircuitPiece<rl,rw,rn>[rw1,rn1,rg1] in
-      execute_time_steps[rl,rw,rn](piece0,piece1,steps)
-
--- Time Step Loop
-function execute_time_steps[rl,rw,rn] ( p0 : CircuitPiece<rl,rw,rn>, 
-      p1 : CircuitPiece<rl,rw,rn>, steps : int ) , reads(rn,rw,rl), writes(rn,rw) : bool = 
-  if steps #$<$# 1 then true else
-  unpack p0 as piece0 : CircuitPiece<rl,rw,rn>[rw0,rn0,rg0] in 
-  unpack p1 as piece1 : CircuitPiece<rl,rw,rn>[rw1,rn1,rg1] in
-  let _ : bool = calc_new_currents[rl,rw0,rn0,rg0](piece0.1) in
-  let _ : bool = calc_new_currents[rl,rw1,rn1,rg1](piece1.1) in
-  let _ : bool = distribute_charge[rl,rw0,rn0,rg0](piece0.1) in
-  let _ : bool = distribute_charge[rl,rw1,rn1,rg1](piece1.1) in
-  let _ : bool = update_voltage[rl,rn0](piece0.2) in
-  let _ : bool = update_voltage[rl,rn1](piece1.2) in
-      execute_time_steps[rl,rw,rn](p0,p1,steps-1)
-
-function color_circuit[rn,rw,rl] ( all_nodes : node_list<rl,rn>@rl, 
-                               all_wires : wire_list<rl,rw,rn>@rl ), 
-        reads(rn,rw,rl) : <coloring(rn), multicoloring(rn), coloring(rw)> =  
-  -- Invoke programmer chosen coloring algorithm (e.g. METIS)
-  -- return owned, ghost, wire colorings
-
--- Helper method
-function build_lists[rl,rw,rn,rpw,rpn,rg] ( nodes : node_list<rl,rn>@rl, 
-       wires : wire_list<rl,rw,rn>@rl, oc : coloring(rn), gc : multicoloring(rn), 
-       wc : coloring(rw), c : int), reads(rn,rw,rl), writes(rl) 
-       : < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl > = 
-  -- Construct lists of node and wire pointers for the given colorings
-\end{lstlisting}
-
-\begin{lstlisting}[float={t},label={lst:circuit_leaf},caption={Circuilt Leaf Tasks}]
+\begin{lstlisting}[float={t},label={lst:circuit_leaf},caption={Circuilt Leaf Functions}]
 function calc_new_currents[rl,rw,rn,rg] ( ptr_list : wire_list<rl,rw,rn,rg>@rl ), 

### 2012-07-10 doc/popl2013/intro.tex
Major work on the example and some edits to the intro

--- 2012-07-10 doc/popl2013/intro.tex
@@
 
-% This is a description of how the listings should be formatted.
-% It can go anywhere before the listings.
+Legion programs organize data into a forest of {\em logical regions}.  Logical
+regions express {\em locality} (data that will be used together, and therefore colocated if possible in the same
+physical memory) and {\em independence} (disjoint data that may be operated on in parallel).
+Researchers have previously explored language designs with hierarchical decomposition of data to express
+locality and independence for high performance programming. Two recent examples, Sequoia \cite{Fatahalian06} and 
+Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a mechanism to partition the heap recursively
+into a tree of collections of data.  The two designs are different in many aspects, but agree that there is
+a single tree-shaped partitioning of data that can be fully checked statically (see Section~\ref{sec:related} for more
+discussion of related work).
+
+Our own experience with writing high-performance applications in both 
+Sequoia and the current industry standard mix of MPI, shared-memory
+threads, and CUDA, has taught us that for many practical situations a
+single, fairly static partitioning is insufficient.  In practice, the
+best way to partition data is often a function of the data
+itself---i.e., the partitions need to be dynamically computed.
+Furthermore, it is useful to allow mutliple partitions simultaneously of
+the same data, providing different views on to that data. Thus, in Legion the 
+handling of regions and partitioning of regions is much more dynamic than in 
+previous designs.  In particular,
+\begin{itemize}
+\item  logical regions are first-class values in Legion
+and may be dynamically allocated and stored in data structures;
+
+\item logical regions can be {\em partitioned} into {\em subregions}; the programmer can express arbitrary partitions of
+regions;
+
+\item  a logical region may be dynamically partitioned in multiple different ways.
+\end{itemize}
+A consequence of these decisions is that, through partitioning, a
+given datum may belong to multiple different regions.  For example, if
+a region $R$ participates in three separate disjoint partition
+operations and each partition allocates every element of $R$ to a
+subregion (partitions need not be total in Legion), then every element
+of $R$ is included in three different subregions.  This is why we use
+the term {\em logical} regions: language-level regions serve to name
+sets of data but do not imply anything about physical layout or placement
+in the memory hierarchy.  There
+is a separate system of {\em physical regions} at runtime that hold
+copies of the data in the language-level logical regions.  In fact,
+the Legion implementation routinely maintains multiple distinct
+physical regions, each with a copy of the data in the corresponding
+logical region, for performance reasons.
+
+A further consequence of placing the same data in multiple logical
+regions is that it introduces aliasing which makes it difficult 
+to statically figure out which computations can be run in parallel.  
+One approach is to outlaw aliasing which allows for a static analysis
+to discover all available parallelism and schedule it entirely at
+compile-time, avoiding any runtime overhead.  Languages such as
+Sequoia and DPJ take this approach and allow only a single static partition of any region 
+to eliminate region aliasing at the cost of the flexibility of their language.
+
+% Putting this here so that we can get the code in a good place
 \lstset{
@@
 }
-
 \begin{lstlisting}[float={t},label={lst:circuit_ex},caption={Circuilt Simulation}]
@@
 
-Legion programs organize data into a forest of {\em logical regions}.  Logical
-regions express {\em locality} (data that will be used together, and therefore colocated if possible in the same
-physical memory) and {\em independence} (disjoint data that may be operated on in parallel).
-Researchers have previously explored language designs with hierarchical decomposition of data to express
-locality and independence for high performance programming. Two recent examples, Sequoia \cite{Fatahalion06} and 
-Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a mechanism to partition the heap recursively
-into a tree of collections of data.  The two designs are different in many aspects, but agree that there is
-a single tree-shaped partitioning of data that can be checked fully statically (see Section~\ref{sec:related} for more
-discussion of related work).
-
-\begin{lstlisting}[float={t},label={lst:circuit_leaf},caption={Circuilt Leaf Tasks}]
-function calc_new_currents[rl,rw,rn,rg] ( ptr_list : wire_list<rl,rw,rn,rg>@rl ), 
-      reads(rl,rw,rn,rg), writes(rw) : bool =
-  if isnull(ptr_list) then true else
-  let wire_node : wire_list<rl,rw,rn,rg> = read(ptr_list) in
-  let wire : CircuitWire<rn,rg> = read(wire_node.1) in
-  let in_node : CircuitNode = read(wire.1) in
-  let out_node : CircuitNode = read(wire.2) in
-  let current : float = (in_node.1 - out_node.1) /  wire.3 in 
-  let new_wire : CircuitWire<rn,rg> = <wire.1,wire.2,wire.3,current> in
-  let _ : CircuitWire<rn,rg>@rw = write(wire_node.1, new_wire) in
-      calc_new_currents[rl,rw,rn,rg](wire_node.2)
-
-function distribute_charge[rl,rw,rn,rg] ( ptr_list : wire_list<rl,rw,rn,rg>@rl ), 
-      reads(rl,rw,rn), reduces(reduce_charge,rn,rg), atomic(rn,rg) : bool =
-  if isnull(ptr_list) then true else
-  let wire_node : wire_list<rl,rw,rn,rg> = read(ptr_list) in
-  let wire : CircuitWire<rn,rg> = read(wire_node.1) in
-  let _ : CircuitNode@rn = reduce(reduce_charge, wire.1, wire.4) in
-  let _ : CircuitNode@(rn,rg) = reduce(reduce_charge, wire.2, wire.4) in
-      distribute_charge[rl,rw,rn,rg](wire_node.2)
-
-function update_voltage[rl,rn] ( ptr_list : node_list<rl,rn>@rl ), 
-      reads(rl,rn), writes(rn) : bool = 
-  if isnull(ptr_list) then true else
-  let node_node : node_list<rl,rn> = read(ptr_list) in
-  let node : CircuitNode = read(node_node.1) in
-  let voltage : float = (node.3/node.4) in
-  let new_node : CircuitNode = <voltage,node.2,node.3,node.4> in
-  let _ : CircuitNode@rn = write(node_node.1, new_node) in
-      update_voltage[rl,rn](node_node.2)
-
--- Reduction function for distribute charge
-function reduce_charge ( node : CircuitNode, current : float ) : CircuitNode =
-    let new_charge : float = node.3 + current in
-        < node.1,new_charge,node.3,node.4>
-\end{lstlisting}
-
-Our own experience with Sequoia and many high-performance applications
-written in the current industry standard mix of MPI, shared-memory
-threads, and CUDA, has taught us that for many practical situations a
-single, fairly static partitioning is insufficient.  In practice, the
-best way to partition data is often a function of the data
-itself---i.e., the partitions need to be dynamically computed.
-Furthermore, it is useful to allow mutliple partitions simultaneously of
-the same data, providing different views on to that data. Thus, in Legion the 
-handling of regions and partitioning of regions is much more dynamic than in 
-previous designs.  In particular,
-\begin{itemize}
-\item  logical regions are first-class values in Legion
-and may be dynamically allocated and stored in data structures;
-
-\item logical regions can be {\em partitioned} into {\em subregions}; the programmer can express arbitrary partitions of
-regions;
-
-\item  a logical region may be dynamically partitioned in multiple different ways.
-\end{itemize}
-A consequence of these decisions is that, through partitioning, a
-given datum may belong to multiple different regions.  For example, if
-a region $R$ participates in three separate disjoint partition
-operations and each partition allocates every element of $R$ to a
-subregion (partitions need not be total in Legion), then every element
-of $R$ is included in three different subregions.  This is why we use
-the term {\em logical} regions: language-level regions serve to name
-sets of data but do not imply anything about physical layout.  There
-is a separate system of {\em physical regions} at runtime that hold
-copies of the data in the language-level logical regions.  In fact,
-the Legion implementation routinely maintains multiple distinct
-physical regions, each with a copy of the data in the corresponding
-logical region, for performance reasons.
-
-A further consequence of placing the same data in multiple logical
-regions as a result of runtime computation is that Legion has a very
-hard static alias analysis problem on regions, which poses a problem
-for parallel computation because knowing regions do not alias is the
-basis for safe parallel execution.  This is why languages such as
-Sequoia and DPJ allow only a single static partition of each
-region---it simplifies static analysis of regions to the point that it
-is practical.
-
-Legion's execution model is that by default, a program's semantics is
-its meaning as a sequential program, which we refer to as the {\em
-program order} execution.  While the
-implementation is not our focus here, if two functions use
-disjoint data (because all of the regions they use are disjoint), then
-Legion may execute them simultaneously.  Also like DPJ,
-Legion has a system of {\em permissions} ({\em read}, {\em write}, and
-{\em reduce}) on regions and an orthogonal system of region {\em
-coherence} modes ({\em excl}, {\em atomic}, and {\em simultaneous})
-that increase the number of situations in which functions can be
-executed in parallel.
+% Moving this to the circuit example section since it fits better there
+% and some of this is redundant
+%Legion's execution model is that by default, a program's semantics is
+%its meaning as a sequential program, which we refer to as the {\em
+%program order} execution.  While the
+%implementation is not our focus here, if two functions use
+%disjoint data (because all of the regions they use are disjoint), then
+%Legion may execute them simultaneously.  Also like DPJ,
+%Legion has a system of {\em permissions} ({\em read}, {\em write}, and
+%{\em reduce}) on regions and an orthogonal system of region {\em
+%coherence} modes ({\em excl}, {\em atomic}, and {\em simultaneous})
+%that increase the number of situations in which functions can be
+%executed in parallel.
 
 The important difference bewteen Legion and the more static approaches
-is that the hardest part of the tests to justify parallel execution is
-done dynamically instead of statically.  The key observation is that
-alias analysis is hard statically but very easy and relatively cheap
-dynamically.  Thus, Legion falls between fully static systems, such as
+is that Legion allows aliasing, but does so in a structured way that minimizes
+runtime overheads.  The key observation is that
+alias analysis is hard statically but very easy dynamically and is relatively cheap
+when done at the granularity of regions instead of individual memory
+locations.  Thus, Legion falls between fully static systems, such as
 Sequoia and DPJ, and fully dynamic approaches such as transactional
@@
 pointer dereferences are checked statically.  The dynamic alias tests
-are done at the granularity of regions and one check on a region
-itself often serves to prove the safety of many subsequent region
-accesses.  In contrast, because transactional memory has no mechanism
-for grouping heap data into larger collections, it much test the
+for independence are done at the granularity of regions and one check on a region
+itself often serves to prove the independence of many subsequent region
+uses.  In contrast, because transactional memory has no mechanism
+for grouping heap data into larger collections, it must test the
 overlap between two sets of data on each individual element, which is

### 2012-07-10 doc/popl2013/evaluation.tex
small edits to evaluation
Removing the circuit description from evaluation since it now has its own section

@@
 We evaluate Legion on three real-world applications.  To qualitatively gauge the 
-expressivity of the Legion type system, we introduce each of these applications
+expressivity of the Legion type system, we introduce these applications
 by describing features of the Legion type system used in their implementations.
+The Circuit example was already covered in detail in Section~\ref{sec:example}.
 
-\subsubsection{Circuit Simulation}
-\label{subsec:circuit}
-Circuit simulates the wires of a large integrated circuit.
+%\subsubsection{Circuit Simulation}
+%\label{subsec:circuit}
+%Circuit simulates the wires of a large integrated circuit.
 % using an RLC model.  
@@
 %an iterative model, updated the charge in each node, and compute the voltage of each node.
-The primary data structure is a large irregular graph, which is
-dynamically partitioned into pieces to enable parallel simulation.
-Our implementation creates separate regions for the wires and
-nodes,  which are recursively partitioned into
-subregions for pieces of the graph.  The node region is partitioned an additional way to 
-describe the sets of ghost nodes required for each piece.  
-The information about each piece of the graph is stored
-in a region relationship that remembers the disjointness information of each piece
-from other pieces.  The tasks for the three phases of the computation use different combinations of
-read, write, and reduce privileges as well as exclusive and simultaneous coherence.
+%The primary data structure is a large irregular graph, which is
+%dynamically partitioned into pieces to enable parallel simulation.
+%Our implementation creates separate regions for the wires and
+%nodes,  which are recursively partitioned into
+%subregions for pieces of the graph.  The node region is partitioned an additional way to 
+%describe the sets of ghost nodes required for each piece.  
+%The information about each piece of the graph is stored
+%in a region relationship that remembers the disjointness information of each piece
+%from other pieces.  The tasks for the three phases of the computation use different combinations of
+%read, write, and reduce privileges as well as exclusive and simultaneous coherence.
 
@@
 that data can be accessed (Circuit,AMR).  All the types of privileges and coherence we have
-described are needed in different combinations by the applications.  Region relationships
+described are needed in different applications.  Region relationships
 are used in all applications. %, region relationships convey disjointness information.

### 2012-07-10 doc/popl2013/intro.tex

@@
                             < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl >         
-                            where rpn <= rn and rg <= rn and rpw <= rw and
+                            where rpn #$\le$# rn and rg #$\le$# rn and rpw #$\le$# rw and
                                   rpn * rg and rn * rw and rl * rn and rl * rw

### 2012-07-10 apps/AMR/popl_plots.py
Updating colors for the popl plots

@@ -325,10 +325,10 @@ def make_plot2(fig,orig,checks,box,prob_size):
         boxlib.append(boxlib_best*o.nodes)
         # Keep the index up to date
         i = i + 1
-    plt.bar(offset+ind, overhead_checks, width, color=tableau2, bottom=overhead_accum, label='Checking Overhead')
-    plt.bar(offset+ind, comm_checks, width, color=tableau3, bottom=comm_accum, label='Communication')
-    plt.bar(offset+ind, runtime_checks, width, color=tableau5, bottom=runtime_accum, label='Runtime Overhead')
-    plt.bar(offset+ind, kernel_checks, width, color=tableau9, label='Kernel')
+    plt.bar(offset+ind, overhead_checks, width, color=tableau6, bottom=overhead_accum, label='Checking Overhead')
+    plt.bar(offset+ind, comm_checks, width, color=tableau2, bottom=comm_accum, label='Communication')
+    plt.bar(offset+ind, runtime_checks, width, color=tableau10, bottom=runtime_accum, label='Runtime Overhead')
+    plt.bar(offset+ind, kernel_checks, width, color=tableau18, label='Kernel')
 
     # Draw the lines for the baselines
     overlap = 0.1

### 2012-07-10 apps/fluid3d/popl_plots.py
Updating plot script with new colors

@@ -171,10 +171,10 @@ def make_plot(fig, orig_exprs, check_exprs, prob_size):
         assert prob_size==19200
         plt.axhline(y=379.432, xmin=0.0125+delta, xmax=0.0125+(1.0/len(zip_list))-delta, linewidth=3, color='k', linestyle='--', label='PARSEC')
 
-    plt.bar(offset+ind, overhead_checks, width, color=tableau2, bottom=overhead_accum, label='Checking Overhead')
-    plt.bar(offset+ind, comm_checks, width, color=tableau3, bottom=comm_accum, label='Communication')
-    plt.bar(offset+ind, runtime_checks, width, color=tableau5, bottom=runtime_accum, label='Runtime Overhead')
-    plt.bar(offset+ind, kernel_checks, width, color=tableau9, label='Kernel')
+    plt.bar(offset+ind, overhead_checks, width, color=tableau6, bottom=overhead_accum, label='Checking Overhead')
+    plt.bar(offset+ind, comm_checks, width, color=tableau2, bottom=comm_accum, label='Communication')
+    plt.bar(offset+ind, runtime_checks, width, color=tableau10, bottom=runtime_accum, label='Runtime Overhead')
+    plt.bar(offset+ind, kernel_checks, width, color=tableau18, label='Kernel')
     plt.xlabel('Total CPUs (8 CPUs/node)')
     plt.xticks(offset+ind+width/2.0,cpus16)
     plt.ylabel('Processor Time (seconds)')

### 2012-07-10 apps/circuit/popl_plots.py
Updating plot script with new colors

@@ -164,10 +164,10 @@ def make_plot(fig, orig_exprs, check_exprs, ps):
         overhead_accum.append(comm_accum[i] + comm_checks[i])
         # Update the index
         i = i + 1
-    plt.bar(offset+ind, overhead_checks, width, color=tableau2, bottom=overhead_accum, label='Checking Overhead')
-    plt.bar(offset+ind, comm_checks, width, color=tableau3, bottom=comm_accum, label='Communication')
-    plt.bar(offset+ind, runtime_checks, width, color=tableau5, bottom=runtime_accum, label='Runtime Overhead')
-    plt.bar(offset+ind, kernel_checks, width, color=tableau9, label='Kernel')
+    plt.bar(offset+ind, overhead_checks, width, color=tableau6, bottom=overhead_accum, label='Checking Overhead')
+    plt.bar(offset+ind, comm_checks, width, color=tableau2, bottom=comm_accum, label='Communication')
+    plt.bar(offset+ind, runtime_checks, width, color=tableau10, bottom=runtime_accum, label='Runtime Overhead')
+    plt.bar(offset+ind, kernel_checks, width, color=tableau18, label='Kernel')
     plt.xlabel('Total GPUs (3 GPUs/node)')
     if len(zip_list)==16:
         plt.xticks(offset+ind+width/2.0,gpus48)

### 2012-07-10 doc/popl2013/example.tex
Small edit to the example section

--- 2012-07-10 doc/popl2013/example.tex
@@
 \label{sec:example}
+
+Listing~\ref{lst:circuit_ex} shows a circuit simulation written (with
+one exception) in the core Legion language (see
+Section~\ref{sec:legioncore}).  The input is a graph of circuit
+elements and connecting wires.  To perform parallel
+simulation the graph is partitioned into pieces, after which the
+simulation is run for a number of time steps.  Each time step has
+three phases: calculate each wire's current, distribute charge to
+the nodes to which each wire is connected, and update each node's voltage.
+
+The top-level function is {\tt simulate\_circuit} (line 15), which is
+parameterized on regions {\tt rl}, {\tt rw}, and {\tt rn} (line 15).
+The wires and nodes live in regions {\tt rw} and {\tt rn},
+respectively.  The {\tt rl} region holds lists of the nodes and wires.
+When {\tt simulate\_circuit} is invoked it must have read and
+write privileges for the region parameters
+(line 17); the type system enforces that
+{\tt simulate\_circuit} only accesses these regions and their
+subregions.
+
+Partitioning of the graph is done using {\em colorings} built in the
+{\tt color\_circuit} function (line 18-19).  A coloring maps region
+elements to colors, with each color corresponding to a new subregion
+to be created.  Note that it is impossible to pick a single static
+partitioning scheme that would do well for all graphs.  Dynamic
+partitioning using colorings provides the programmer the flexibility
+to make partitioning decisions based on input data.
+
+The {\tt color\_circuit} function returns three colorings.  The first
+maps each node to the graph piece that owns it.
+The third maps each wire to a graph piece that
+owns one of its nodes.  The second coloring captures, for each piece $p$,
+the nodes in other pieces on the boundary of $p$, often referred to as
+{\em ghost nodes}.  
+%We describe how ghost node regions are used in
+%conjunction with the owned nodes shortly.  
+Note that 
+ghost nodes may be on the boundary of multiple pieces and therefore colored
+multiple ways.  The second coloring is a {\em multicoloring} because
+it may map the same element to multiple colors.
+
+After creating the colorings, the application partitions the region
+into {\em subregions} according to the colorings (lines 21-25).  For
+simplicity, our example only partitions the circuit into two pieces.
+Lines 21 and 23 create two different partitions with two different
+subregions of the {\tt rn} region.  The first partition uses a
+coloring and is therefore {\em disjoint}; the second partition uses a
+multicoloring and is {\em aliased}---the subregions may not be
+disjoint. While aliased partitions based on multicolorings are not a
+part of core Legion described in Section~\ref{sec:legioncore}, they
+are functionally equivalent to creating single partitions for each
+color in the multicoloring.
+Disjoint partitions introduce region disjointness constraints into the static
+environment indicating that all the subregions in the partition are
+disjoint (e.g. {\tt rn0}$*${\tt rn1} on line 21 says {\tt rn0} is
+disjoint from {\tt rn1}).  Both disjoint and aliased partitions
+introduce constraints into the environment indicating that all created
+regions are logical subregions of the partitioned parent region
+(e.g. {\tt rn0}$\leq${\tt rn}).
+
+The circuit simulation uses an {\em allocate-then-partition} style of
+computation, where a large data structure is first allocated and then
+partitioned.  Legion also supports {\em partition-then-allocate},
+where empty regions are first partitioned and later populated with
+data.  Both are useful; e.g., Sequoia \cite{Fatahalian06} and DPJ
+\cite{Bocchino09} support static forms of allocate-then-partition and
+partition-then-allocate, respectively.  Legion supports both
+approaches dynamically.
+
+After partitioning the circuit into pieces, the application creates
+instances of {\tt CircuitPiece} (defined on lines 9-12) for each piece
+(lines 28-33).  {\tt CircuitPiece} is a {\em region relationship}, a
+bounded existential type.  Region relationships allow the programmer
+to {\em pack} a group of regions together and remember properties
+about them such as disjointness and subregion relationships (lines 29
+and 33).  The type system verifies the properties hold statically when
+packing; when region relationships are {\em unpacked} the same
+properties are reintroduced into the static checking environment
+(lines 40-41).  A key feature underlying the soundness of the Legion
+type system is that privileges cannot be packed in a region
+relationship---privileges belong to functions. When a function unpacks
+a region relationship it must already hold privileges for the new
+regions it finds in the packed value.  For example, on line 40
+function {\tt execute\_time\_steps} has read/write privileges for
+region {\tt rn0} because it has read/write privileges for {\tt rn} and
+the {\tt CircuitPiece} region relationship constraints ensure $\tt rn0
+\leq rn$ (line 12).
+
+The {\tt execute\_time\_steps} function (lines 37-48)  
+also illustrates the importance of using different partitions to give
+multiple views onto the same logical region.  Both {\tt
+  calc\_new\_currents} and {\tt distribute\_charge} 
+use the owned and ghost regions of a piece, which are from different partitions. In
+{\tt calc\_new\_currents} these regions only need read
+privileges, allowing both instances of {\tt calc\_new\_currents} to be
+run in parallel.  In {\tt distribute\_charge} the
+privilege is for a reduction which can also be done in parallel
+because of the atomic and commutative nature of reductions.  No
+single partition of the nodes
+describes these data sharing patterns.
+
+Listing~\ref{lst:circuit_leaf} shows the leaf functions for each phase.
+Each function iterates over the list of wires or
+nodes for its piece.  Each function specifies the region privileges it
+requires (lines 2,14,23).  
+%These privileges
+%are statically checked to match the operations performed inside of
+%each function (e.g. read and write).  
+In the case of {\em reduce} the
+privilege must also specify which reduction function is used
+(line 14).
+
+In addition to privileges, functions can also specify {\em coherence}
+on regions.  Coherence specifies what updates the function may
+see from other functions using the same region.  If not otherwise
+specified, coherence defaults to {\em exclusive}, meaning the
+function must appear to execute in program order.  Line 14 shows
+an example of {\tt atomic}, a relaxed coherence mode requiring
+that reductions to {\tt rn} and {\tt rg} must appear atomic relative
+to other functions using those regions.  The most relaxed coherence
+mode is {\tt simult}; all functions with 
+simultaneous coherence are allowed to access the region concurrently.
 

### 2012-07-10 doc/popl2013/langdef.tex
Patched version of theorem 2.

@@
 % macros for consistency flavors
-\newcommand{\foo}[1]{\cancel{#1}}
 \newcommand{\simH}{\sim_{\!H}}
 % usually: \mapconsist{\Omega}
-\newcommand{\mapconsist}[2][M]{#1 \foo\sim #2}
+\newcommand{\mapconsist}[2][M]{#1 \sim #2}
 % usually: \localconsist{L}{\Gamma}
-\newcommand{\localconsist}[3][M]{#2 \foo\simH \@ifmtarg{#1}{#3}{#1 \llbracket #3 \rrbracket}}
+\newcommand{\localconsist}[3][M]{#2 \simH \@ifmtarg{#1}{#3}{#1 \llbracket #3 \rrbracket}}
 % usually: \storeconsist{S}{H}
-\newcommand{\storeconsist}[2]{#1 \foo\sim #2}
+\newcommand{\storeconsist}[2]{#1 \sim #2}
 % usually: \valueconsist{v}{T}
-\newcommand{\valueconsist}[3][M]{#2 \foo\simH \@ifmtarg{#1}{#3}{#1 \llbracket #3 \rrbracket}}
+\newcommand{\valueconsist}[3][M]{#2 \simH \@ifmtarg{#1}{#3}{#1 \llbracket #3 \rrbracket}}
 % usually: \traceconsist{E}
-\newcommand{\traceconsist}[2][H]{#2 \foo\sim #1}
+\newcommand{\traceconsist}[2][H]{#2 \sim #1}
 % usually: \privconsist{E}{\Phi}
@@
 
-The first option is to serialize the constituent traces in program
-order.  Although each constituent trace has been checked for coherence, the
-checking of some operations may have been deferred due to non-empty
+The first option is to serialize the evaluation of the subexpressions that
+make up an expression and attempt to generate a valid interleaving by simply
+concatenating the constituent traces in program order.  However, there is a
+complication related to the nondeterminism that is allowed for reads.  Although
+each subexpression's trace will have been checked for coherence, the checking
+of some operations may have been deferred due to non-empty
 clobber sets.  It is possible that when an interleaving of these traces
@@
 if the wrong guesses were made in the derivation of a subexpression.
-However, we can show that as long as any valid interleaving
-exists, the serialization of the constituent traces is valid.
-To prove this result, we use the following lemma, the proof
-of which follows directly from the definition of $apply$.
+The solution in the case of serial execution is to not expand the clobber
+set for subexpressions, instead using the parent expression's clobber set.
+
+Only a single rule is modified for this alternate execution - $e_{n+1}$ in
+{\sc E-Call} uses the original $C$ instead of $C'$.  Note that this change
+can result in there being no terminating execution for an expression in
+which the only terminating executions required a particular (non-program-)
+ordering of subexpression evaluation, but the Legion core language does not
+guarantee parallel execution in any case, so the evaluation of such an 
+expression is unlikely to terminate even under the original clobber set.
+We can, however, show that any terminating execution that results from using
+the restricted clobber set is a valid execution under the original:
 
 \begin{lem}
-\label{lemma:applychain}
+\label{lem:clobberreduce}
 \rm 
-For any initial store $S$ and memory traces $E_1$ and $E_2$,
-$$apply(S, E_1 \dplus E_2) = apply(apply(S, E_1), E_2).$$
+Let $\typeenv{e}{T}$ and $\opsenv{e}{v}{E}$ and $\storeconsist{S}{H}$.  If $C \subseteq C'$, then
+$\opsenv[C=C']{e}{v}{E}$.
 \end{lem}
+
+We now show that the concatenation of the memory traces from these serially
+executed subexpression is always a valid interleaving:
 
@@
 \rm
-Let $S_0$ be a store, $C$ a clobber set and $\oton{E}$ memory traces to be
-interleaved.  Let $E' = E_1 \dplus \ldots \dplus E_n$, in program order.  If there exists any $E''$ that is a
-valid interleaving of $\oton{E}$, then $E'$ is also valid interleaving of $\oton{E}$.
+Let $\oton{e}$ be expressions such that $\opsenv[S={S_{i-1}}]{e_i}{v_i}{E_i}$, 
+where $S_i = apply(S_{i-1}, E_i)$.  If $E' = E_1 \dplus \ldots \dplus E_n$, then
+$valid\_interleave(S_0, C, E', \oton{E})$.
 \end{thm}
 
-% FIX ME FIX ME FIX ME
 \begin{proof}
-To prove $valid\_interleave(S_0, C, E', \oton{E})$, we need to show the four conditions:
-\begin{enumerate}
-\item $any\_interleave(E', \oton{E})$ is proven by induction on
-the length of $E_1$, followed by $E_2$, and so on.
-\item $coherent(S_0, C, E')$ is shown by considering each constituent trace's subsequence in $E'$
-in order, using Lemma~\ref{lemma:applychain} with the coherence of each $E_i$.
-\item $seq\_equiv(S_0, C, E', \oton{E})$ follows trivially by matching the structure of $E'$
-to the corresponding  pieces $\oton{E}$.
-\item $coherent(S_0, C, E')$ is shown using the sequential equivalence of $E''$.  The value 
-returned by each read in $E''$ is equal to the value in the store that results
-from the application of all previous constituent traces and the operations before the read in its
-own trace.  This is exactly the sequence of preceding operations of the corresponding read in $E'$,
-and Lemma~\ref{lemma:applychain} allows us to transform the chain of $apply$ operations into a 
-single $apply$ operation on the whole sequence, satisfying the coherence condition for the read
-in $E'$.  This property applies for all reads in $E''$, and since $E'$ and $E''$ are
-interleavings of the same traces, they have the same reads, so all reads
-in $E'$ are coherent.
-\end{enumerate}
+First, we must show that $any\_interleave(E', \oton{E})$.  This is done by
+induction on the length of $E_1$, followed by $E_2$, and so on.  Next, we must
+show that $E'$ is coherent for both $L_{excl}$ and $L_{atomic}$.  This is done by
+matching the structure of $E'$ with $\oton{E}$ and the use of the following
+lemma: $apply(S, E_1 \dplus E_2) = apply(apply(S, E_1), E_2)$.  With the
+question of coherence addressed, the sequential equivalence of $E'$ is trivial
+for both the exclusive and atomic cases.
 \end{proof}

### 2012-07-10 doc/popl2013/langdef.tex
more tweaks

@@
   \typeenv[G={\Gamma_i},P={\Phi_i},O=\emptyset]{e_i}{T_{ir}}}
-  {\vdash \{  \text{function }id_1[r_{11}, \ldots, r_{1k_1}]( a_{11} : T_{11}, \ldots a_{1n_1} : T_{1n_1} ), \Phi_1, Q_1 : T_{1r} : e_1, 
-           \ldots
-           \text{function }id_p[r_{p1}, \ldots, r_{pk_1}]( a_{p1} : T_{p1}, \ldots a_{pn_p} : T_{pn_p} ), \Phi_p, Q_p : T_{pr} : e_p \} : \bullet}
+  {\begin{array}{@{}l@{}l@{}}
+   \vdash \{ & \text{function }id_1[r_{11}, \ldots, r_{1k_1}]( a_{11} : T_{11}, \ldots a_{1n_1} : T_{1n_1} ), \Phi_1, Q_1 : T_{1r} : e_1, \\
+           & \ldots \\
+           & \text{function }id_p[r_{p1}, \ldots, r_{pk_1}]( a_{p1} : T_{p1}, \ldots a_{pn_p} : T_{pn_p} ), \Phi_p, Q_p : T_{pr} : e_p \} : \bullet
+   \end{array}
+  }
 \end{minipage}
@@
 
-\section{Soundness of Effects}
+\section{Soundness of Privileges}
 \label{sec:soundness}
@@
 if $\traceconsist{E_1}$ and $\traceconsist{E_2}$ and $any\_interleave(E', E_1, E_2)$, then $\traceconsist{E'}$.
-New effects are added by {\tt write} and {\tt reduce} expressions, but consistency follows
+New memory operations are added by {\tt write} and {\tt reduce} expressions, but consistency follows
 directly from the induction hypothesis.  Finally, the consistency of the
@@
 \hangindent=2\parindent \hangafter=1 \noindent %
-$\privconsist{E}{\Phi}$ -  The proof of the crucial property of containment of effects within the available privileges
+$\privconsist{E}{\Phi}$ -  The proof of the crucial property of containment of heap accesses within the available privileges
 is similar in outline to the previous step.  The easy cases are covered by a lemma combining the corresponding results
@@
 Only a single rule is modified for this alternate execution - $e_{n+1}$ in
-{\sc E-Call} uses the original $C$ instead of $C'$.  Note that this change
+{\sc E-Call} uses the original $C$ instead of $C'$.  This change
 can result in there being no terminating execution for an expression in
 which the only terminating executions required a particular (non-program-)
-ordering of subexpression evaluation, but the Legion core language does not
-guarantee parallel execution in any case, so the evaluation of such an 
+ordering of subexpression evaluation.  However, the Legion core language never
+guarantees parallel execution, so the evaluation of such an 
 expression is unlikely to terminate even under the original clobber set.
@@
 \rm
-Let $\oton{e}$ be expressions such that $\opsenv[S={S_{i-1}}]{e_i}{v_i}{E_i}$, 
+Let $\oton{e}$ be expressions such that 
+%$\opsenv[S={S_{i-1}}]{e_i}{v_i}{E_i}$, 
+$M, L, H, S_{i-1},\linebreak C \vdash e_i \mapsto v_i, E_i$,
 where $S_i = apply(S_{i-1}, E_i)$.  If $E' = E_1 \dplus \ldots \dplus E_n$, then
@@
 
-To summarize this, we define a {\em noninterference} operator on two effects that have been
+To summarize this, we define a {\em noninterference} operator on two memory operations that have been
 annotated with which constituent trace they came from as follows:
@@
 implementation that is different from other systems and relies on the
-soundness of privileges.  Noninterference of memory effects is
+soundness of privileges.  Noninterference of memory operations is
 something that can only be determined after evaluation of an
@@
 at runtime, which is much simpler and more efficient than checking
-noninterference of dynamic effects.  Even though the privileges
+noninterference of dynamic memory traces.  Even though the privileges
 themselves are static, the region mapping $M$ is dynamic.  Testing
@@
 To convert the required analysis to one that can be performed on regions, we modify
-the noninterference operator for effects to consider only locations in $L_{excl}$ and show that
+the noninterference operator for memory operations to consider only locations in $L_{excl}$ and show that
 traces that are noninterfering under this modified check satisfy the requirement for atomic
@@
 privileges that are {\em atomically noninterfering} (under a dynamic region mapping
-$M$) always have effects compatible with atomically-reordered execution.  Finally,
+$M$) always have memory traces compatible with atomically-reordered execution.  Finally,
 in the special case where each subexpression has $atomic(r)$ or $simult(r)$ in $Q_i$ for every $r$
@@
 checks are trivially satisfied, and any interleaving of the constituent traces is valid.
-We again modify the noninterference criteria for effects, traces, and privileges, this time
+We again modify the noninterference criteria for memory operations, traces, and privileges, this time
 only allowing $simult$ coherence annotations. We show that subexpression with privileges

### 2012-07-10 doc/popl2013/evaluation.tex
lots of small changes

@@
 \begin{itemize}
-\item Expressivity: Can real applications be written (Section \ref{subsec:expressivity})?
-\item Overhead: Does it reduce checking costs (Section \ref{subsec:overhead})?
-\item Scalability: Can we leverage hierarchical scheduling (Section \ref{subsec:scalability})?
+\item Expressivity: Can real applications be written (Section~\ref{subsec:expressivity})?
+\item Overhead: Does it reduce checking costs (Section~\ref{subsec:overhead})?
+\item Scalability: Can it leverage hierarchical scheduling (Section~\ref{subsec:scalability})?
 \end{itemize}
@@
 Our prototype implementation has two components: a type checker
-for the language introduced in Section \ref{sec:legioncore} and a C++ runtime library
+for the language introduced in Section~\ref{sec:legioncore} and a C++ runtime library
 for executing programs written in the Legion programming model\cite{Legion12}.  All experiments
@@
 When testing applications we frequently encountered memory corruption due to
-illegal region accesses caused by a confluence of application bugs and the lack of checking.  Many times these corruptions 
+illegal region accesses caused by a confluence of application bugs and lack of checking.  Many times these corruptions 
 occurred on remote nodes or on GPUs, requiring heroic debugging efforts.
@@
 degradation due to the extra registers required for checking.  We also measured the overhead
-of the dynamic checks associated with unpack operations but found them to be negligible.
+of the dynamic checks associated with checking task call privileges but found them to be negligible.
 % relative to the overhead of region access checks.

### 2012-07-10 doc/popl2013/example.tex
lots of small changes

@@
 simplicity, our example only partitions the circuit into two pieces.
-Lines 21 and 23 create two different partitions with two different
+Lines 21 and 23 create two different partitions each with two different
 subregions of the {\tt rn} region.  The first partition uses a
@@
 an example of {\tt atomic}, a relaxed coherence mode requiring
-that reductions to {\tt rn} and {\tt rg} must appear atomic relative
+that operations to {\tt rn} and {\tt rg} must appear atomic relative
 to other functions using those regions.  The most relaxed coherence
 mode is {\tt simult}; all functions with 
-simultaneous coherence are allowed to access the region concurrently.
+simultaneous coherence are allowed to access the region concurrently
+with other functions having simultaneous coherence to the region.
 

### 2012-07-10 doc/popl2013/intro.tex
lots of small changes

@@
 recently, multiple accelerators (usually GPUs) per node, which 
-can themselves be decomposed into several levels.   We present the operational and static
-semantics of Legion, a programming model targeted at providing an
+can themselves be decomposed.  We present the operational and static
+semantics of Legion\cite{Legion12}, a programming model targeted at providing an
 appropriate level of abstraction for programming such machines, one
@@
 Legion programs organize data into a forest of {\em logical regions}.  Logical
-regions express {\em locality} (data that will be used together, and therefore colocated if possible in the same
-physical memory) and {\em independence} (disjoint data that may be operated on in parallel).
+regions express {\em locality} (data that will be used together, and therefore should be colocated) 
+and {\em independence} (disjoint data that may be operated on in parallel).
 Researchers have previously explored language designs with hierarchical decomposition of data to express
@@
 \end{itemize}
-A consequence of these decisions is that, through partitioning, a
+One consequence of these decisions is that, through partitioning, a
 given datum may belong to multiple different regions.  For example, if
-a region $R$ participates in three separate disjoint partition
-operations and each partition allocates every element of $R$ to a
+a region $R$ is disjointly partitioned three distinct ways 
+and each partition allocates every element of $R$ to a
 subregion (partitions need not be total in Legion), then every element
@@
 physical regions, each with a copy of the data in the corresponding
-logical region, for performance reasons.
+logical region, for performance reasons \cite{Legion12}.
 
@@
 Sequoia and DPJ take this approach and allow only a single static partition of any region 
-to eliminate region aliasing at the cost of the flexibility of their language.
+to eliminate aliasing at the cost of the flexibility of their language.
 
@@
 for grouping heap data into larger collections, it must test the
-overlap between two sets of data on each individual element, which is
-much more expensive.
+overlap between two sets of data on each individual memory access, which is
+significantly more expensive.
 
 We begin in Section~\ref{sec:example} with an example program that illustrates
-a typical style of programming in Legion as well as motivating the need for multiple, dynamically computed  partitions of
+a typical style of programming in Legion and motivates the need for multiple, dynamically computed partitions of
 a region.  We define Core Legion, a small language suitable for our formal development, in Section~\ref{sec:legioncore}.
@@
 $e_1'$ of $e_1$ and $e_2'$ of $e_2$ are also non-interfering (Section~\ref{sec:scheduling}).  This result is the basis
-for a hierarchical, distributed scheduler in the Legion implementation; on the target class of machines, any centralized
-scheduler would be a serious bottleneck.
+for a hierarchical, distributed scheduler in the Legion implementation which is crucial for high performance
+on the target class of machines.
+%; on the target class of machines, any centralized
+%scheduler would be a serious bottleneck.
 

### 2012-07-10 doc/popl2013/langdef.tex
lots of small changes

@@
 a Legion function - it is limited to privileges given to it by its caller.  (The
-top-level task of a Legion program begins with a single region from with other
+top-level task of a Legion program begins with a single, empty region from which other
 regions can be partitioned and full read and
@@
 The most involved rule in Core Legion's semantics is for function calls.
-Note that the only regions accessible to the callee (through $M'$) are the region arguments.
+The only regions accessible to the callee (through $M'$) are the region arguments.
 Also, if any regions are used with relaxed coherence (either {\em atomic} or {\em simult} instead of
@@
 a region relation instance guarantees consistency of region mapping.
-Finally, the body of a called function with an initially-empty set of
+Finally, the body of a called function uses an initially-empty set of
 constraints, which are trivially satisfied.
@@
 to the store.  (If $L_2 = \emptyset$ initially, atomic reads without a preceding write are not checked, effectively defering
-these checks to point where traces that might be ordered before $E$ are included.)  The $seq\_equiv$
+these checks to a point where traces that might be ordered before $E$ are included.)  The $seq\_equiv$
 test verifies every read in $E'$ with a location in $L_1$ 
@@
 }
-We are now able to state the key result that allows the Legion runtime to safely evaluate subexpressions as
+We can now state the key result that allows the Legion runtime to safely evaluate subexpressions as
 separate tasks in parallel.
@@
 
-We are now able to explain an important aspect of the Legion
+We now highlight an important aspect of the Legion
 implementation that is different from other systems and relies on the
@@
 memory \cite{Harris05}.  At the other extreme are systems
-that check privileges statically, such as Jade \cite{Rinard98} and
+that check noninterference statically, such as Jade \cite{Rinard98} and
 DPJ \cite{Bocchino11}.  In contrast, Legion checks noninterference of privileges
@@
 result in valid interleavings even if constituent traces have locations in common.  We discuss
-two cases of interest.  In both the previous sections noninterference % Is noninterference the right subject here?
+two cases of interest.  In both, the previous section's analysis 
 is repeated with minor modifications, so for brevity we limit

### 2012-07-10 doc/popl2013/related.tex
lots of small changes

@@
 Legion is most directly related to Sequoia \cite{Fatahalian06}.  Sequoia is array-based, with
-a limited repetoire of ways to partition arrays.  Sequoia is a very static language with a single unified hierarchy
+a limited repetoire of ways to partition arrays.  Sequoia is a static language with a single unified hierarchy
 of tasks and data; Legion is more dynamic with separate task and region hierarchies.
@@
 permissions system as our primary formalism because separation logic does not easily
-support the interleaving of effects to non-disjoint regions of memory.
+support the interleaving of operations to non-disjoint regions of memory.
 

### 2012-07-10 doc/popl2013/related.tex
More edits to the related work section

@@
 
-Deterministic Parallel Java (DPJ) is the only other region-based parallel system we are
-aware of \cite{Bocchino09}.  While there are similarities in the permission system and we have
-reused some DPJ notation, there are differences stemming from DPJ's more static approach.
+Deterministic Parallel Java (DPJ) is the only other region-based parallel system of which we are
+aware\cite{Bocchino09}.  While there are similarities in the permission system and we have
+reused some DPJ notation, there are differences stemming from DPJ's static approach.
 Regions in Legion are first-class and can be created, partitioned, packed, and unpacked 
 dynamically, allowing programmers to compute data organization at runtime; like Sequoia, DPJ
-partition schemes must be decided at compile-time.  Legion allows 
+partition schemes must be statically decided.  Legion allows 
 programmers to create multiple partitions of the same region to give different 
@@
 %while Legion provides safe execution even in simultaneous environments.  
-There is also a difference in emphasis: DPJ targets shared memory machines, while the Legion system
+There is also a difference in emphasis: DPJ requires shared memory, while Legion 
 is designed for distributed heterogenous machines.

### 2012-07-11 doc/popl2013/example.tex
changes to intro and code example sections

@@
 simulation is run for a number of time steps.  Each time step has
-three phases: calculate each wire's current, distribute charge to
-the nodes to which each wire is connected, and update each node's voltage.
+three phases: calculation of each wire's current, distribution of charge to
+the nodes to which each wire is connected, and updating of each node's voltage.
 
 The top-level function is {\tt simulate\_circuit} (line 15), which is
-parameterized on regions {\tt rl}, {\tt rw}, and {\tt rn} (line 15).
-The wires and nodes live in regions {\tt rw} and {\tt rn},
-respectively.  The {\tt rl} region holds lists of the nodes and wires.
-When {\tt simulate\_circuit} is invoked it must have read and
-write privileges for the region parameters
+parameterized on regions {\tt rl}, {\tt rw}, and {\tt rn}.
+The wires and nodes are contained in regions {\tt rw} and {\tt rn},
+respectively.  The {\tt rl} region holds lists of pointers to the nodes and wires.
+When {\tt simulate\_circuit} is invoked it must have both read and
+write privileges for the regions {\tt rl}, {\tt rw}, and {\tt rn}
 (line 17); the type system enforces that
@@
 {\tt color\_circuit} function (line 18-19).  A coloring maps region
-elements to colors, with each color corresponding to a new subregion
+elements to ``colors'' (i.e. integers), with each color corresponding
+ to a new subregion
 to be created.  Note that it is impossible to pick a single static
@@
 
-After creating the colorings, the application partitions the region
+After creating the colorings, the application partitions the node and wire
+regions
 into {\em subregions} according to the colorings (lines 21-25).  For
 simplicity, our example only partitions the circuit into two pieces.
-Lines 21 and 23 create two different partitions each with two different
-subregions of the {\tt rn} region.  The first partition uses a
-coloring and is therefore {\em disjoint}; the second partition uses a
-multicoloring and is {\em aliased}---the subregions may not be
-disjoint. While aliased partitions based on multicolorings are not a
-part of core Legion described in Section~\ref{sec:legioncore}, they
-are functionally equivalent to creating single partitions for each
-color in the multicoloring.
-Disjoint partitions introduce region disjointness constraints into the static
-environment indicating that all the subregions in the partition are
-disjoint (e.g. {\tt rn0}$*${\tt rn1} on line 21 says {\tt rn0} is
-disjoint from {\tt rn1}).  Both disjoint and aliased partitions
-introduce constraints into the environment indicating that all created
-regions are logical subregions of the partitioned parent region
-(e.g. {\tt rn0}$\leq${\tt rn}).
+Line 21 uses {\tt pc.1} to partition {\tt rn} into {\tt rn0}, and {\tt rn1}.
+Subregion {\tt rn0} will contain all locations with a color of 1 in {\tt pc.1},
+while {\tt rn1} will contain all locations with a color of 2.  This
+partitioning uses a coloring and results in {\em disjoint} subpartitions.
+Constraints are introduced into the static environment describing both the
+disjointness of the subregions (i.e. {\tt rn0 $*$ rn1}) and the inclusion of
+each subregion in the original region (e.g. {\tt rn0 $\leq$ rn}).  Line
+23 is similar, but uses a multicoloring, so the subregions may be {\em aliased},
+and only the inclusion constraints are introduced into the static environment.
+Although multicolorings are not part of the Core Legion described in 
+Section~\ref{sec:legioncore}, they can be easily implemented using a separate 
+coloring for each color in the multicoloring.
 
@@
 computation, where a large data structure is first allocated and then
-partitioned.  Legion also supports {\em partition-then-allocate},
+partitioned.  Legion also supports a {\em partition-then-allocate} style,
 where empty regions are first partitioned and later populated with
@@
 privilege is for a reduction which can also be done in parallel
-because of the atomic and commutative nature of reductions.  No
+because of the atomic and commutative nature of reductions.  Finally,
+the {\tt update\_voltage} function modifies only the owned regions, permitting
+each piece to be updated in parallel.  No
 single partition of the nodes
 describes these data sharing patterns.
+
+This function also demonstrates the benefit of being able to dynamically
+(re-)discover disjointness.  Safely running both instances of 
+{\tt calc\_new\_currents} in parallel depends on knowing that {\tt rw0} and
+{\tt rw1} in {\tt execute\_time\_steps} are disjoint.  Similarly, parallel
+execution of the instances of {\tt update\_voltage} requires knowing that
+{\tt rn0} is disjoint from {\tt rn1}.  In both cases, this knowledge was
+statically available in {\tt simulate\_circuit} and could have been captured
+in a region relationship, but it would have resulted in much more complicated
+code.
+Instead, an inexpensive dynamic disjointness check is performed by the Legion
+runtime, and the safety of parallel execution follows from the soundness of
+the type system.
 

### 2012-07-11 doc/popl2013/intro.tex
changes to intro and code example sections

@@
 have the same immediate parent task) $t_1$ and $t_2$ are 
-{\em independent} (have no ordering requirements for correctness), 
-then the unique ancestors of $t_1$ and $t_2$ in the task hierarchy 
-are also independent.  This theorem is the basis for correct distributed
+{\em non-interfering} (have no ordering requirements for correctness), 
+then the any descendent of $t_1$ in the task hierarchy is non-interfering
+with any descendent of $t_2$.  This theorem is the basis for correct distributed
 task scheduling in a Legion implementation, which is crucial for
@@
 itself---i.e., the partitions need to be dynamically computed.
-Furthermore, it is useful to allow mutliple partitions simultaneously of
+Furthermore, it is useful to allow multiple partitions simultaneously of
 the same data, providing different views on to that data. Thus, in Legion the 
@@
 a region $R$ is disjointly partitioned three distinct ways 
-and each partition allocates every element of $R$ to a
+and each partition allocates every element in $R$ to a
 subregion (partitions need not be total in Legion), then every element
-of $R$ is included in three different subregions.  This is why we use
+in $R$ is included in three different subregions.  This is why we use
 the term {\em logical} regions: language-level regions serve to name
@@
                             where rpn #$\le$# rn and rg #$\le$# rn and rpw #$\le$# rw and
-                                  rpn * rg and rn * rw and rl * rn and rl * rw
+                                  rn * rw and rl * rn and rl * rw
 
@@
 runtime overheads.  The key observation is that
-alias analysis is hard statically but very easy dynamically and is relatively cheap
+static alias analysis is hard, but dynamic analysis is very easy
+and relatively cheap
 when done at the granularity of regions instead of individual memory
@@
 pointer dereferences are checked statically.  The dynamic alias tests
-for independence are done at the granularity of regions and one check on a region
-itself often serves to prove the independence of many subsequent region
-uses.  In contrast, because transactional memory has no mechanism
+for independence are done at the granularity of regions and one check 
+on a region
+often serves to prove the independence of many subsequent uses of that region.
+In contrast, because transactional memory has no mechanism
 for grouping heap data into larger collections, it must test the

### 2012-07-11 doc/popl2013/related.tex

@@
 permissions system as our primary formalism because separation logic does not easily
-support the interleaving of operations to non-disjoint regions of memory.
+support reasoning about the interleaving of operations to non-disjoint regions of memory.
 

### 2012-07-11 doc/popl2013/langdef.tex
Removing a widow from langdef

@@
 that comes after a write to $l$, return the value most recently written
-to the store.  (If $L_2 = \emptyset$ initially, atomic reads without a preceding write are not checked, effectively defering
-these checks to a point where traces that might be ordered before $E$ are included.)  The $seq\_equiv$
+to the store.  (If $L_2 = \emptyset$ initially, atomic reads without a preceding write are not 
+checked, but are defered to a point where traces that might be ordered before $E$ are included.)  The $seq\_equiv$
 test verifies every read in $E'$ with a location in $L_1$ 

### 2012-07-11 doc/popl2013/intro.tex
s/permission/privilege/g

@@
 tasks} and subtasks operating on regions.  Regions and tasks interact
-through a static system of {\em region permissions} specifying 
+through a static system of {\em region privileges} specifying 
 operations a task may perform on a region argument (read,
@@
 the region (exclusive, atomic, or simultaneous).  We prove the
-soundness of the permissions and coherence system and use these two
+soundness of the privileges and coherence system and use these two
 results to prove a third result: if two {\em siblings} (tasks that
@@
 threads, and CUDA, has taught us that for many practical situations a
-single, fairly static partitioning is insufficient.  In practice, the
+single, static partitioning is insufficient.  In practice, the
 best way to partition data is often a function of the data
 itself---i.e., the partitions need to be dynamically computed.
-Furthermore, it is useful to allow multiple partitions simultaneously of
+Furthermore, it is useful to allow multiple simultaneous partitions of
 the same data, providing different views on to that data. Thus, in Legion the 
@@
 %Legion may execute them simultaneously.  Also like DPJ,
-%Legion has a system of {\em permissions} ({\em read}, {\em write}, and
+%Legion has a system of {\em privileges} ({\em read}, {\em write}, and
 %{\em reduce}) on regions and an orthogonal system of region {\em
@@
 memory.  Legion still has a significant static component; in
-particular, the safety of permissions is checked statically and region
+particular, the required privileges for function calls is checked statically
+and region
 pointer dereferences are checked statically.  The dynamic alias tests
@@
 \begin{itemize}
-\item We prove the soundness of Legion's permissions system (Section~\ref{sec:soundness}).
+\item We prove the soundness of Legion's privileges system (Section~\ref{sec:soundness}).
 
-\item We use the soundness of permissions to prove the soundness of Legion's region coherence modes (Section~\ref{sec:coherence}).
+\item We use the soundness of privileges to prove the soundness of Legion's region coherence modes (Section~\ref{sec:coherence}).
 
@@
 dynamic region pointer checks would be expensive, justifying checking this aspect of the type system statically.
-We also show that the cost of checking aliasing on region permissions is low, showing that a much more expressive
+We also show that the cost of checking for non-interference of region privileges is low, showing that a much more expressive
 and dynamic langauge is not incompatible with high performance.

### 2012-07-11 doc/popl2013/langdef.tex
s/permission/privilege/g

@@
 
-Our goal is to show $\privconsist{E}{\Phi}$ holds for any execution of an expression $e$ with static permissions $\Phi$.
+Our goal is to show $\privconsist{E}{\Phi}$ holds for any execution of an expression $e$ with static privileges $\Phi$.
 As usual, even stating such a claim requires that corresponding parts of the initial static and dynamic environments be consistent.
@@
 
-To summarize this, we define a {\em noninterference} operator on two memory operations that have been
+To summarize this, we define a {\em non-interference} operator on two memory operations that have been
 annotated with which constituent trace they came from as follows:
@@
 The proof is included in \cite{LegionTypes12}.
-We now extend the noninterference operator to entire memory traces, providing a way to say that
+We now extend the non-interference operator to entire memory traces, providing a way to say that
 no operation in a trace interferes with any operation in another:
@@
 trace, their misordering would mean that $E'$ was not an interleaving of $\oton{E}$).  Every pair of
-traces is noninterfering, so $\epsilon_1 \nonint \epsilon_2$ and they can be swapped while preserving
+traces is non-interfering, so $\epsilon_1 \nonint \epsilon_2$ and they can be swapped while preserving
 validity of the trace.  At most $n^2/2$ swaps are needed to eliminate all misordered pairs.
@@
 }
-We next extend noninterference to privileges.  We 
+We next extend non-interference to privileges.  We 
 use the region mapping $M$ to translate the logical regions named in privileges into sets of
-locations (i.e. physical regions), and define noninterference between individual privileges and
+locations (i.e. physical regions), and define non-interference between individual privileges and
 sets of privileges:
@@
 Let $M$ be a region mapping and $E_1$ and $E_2$ two memory traces such that $\privconsist{E_1}{\Phi_1}$ and $\privconsist{E_2}{\Phi_2}$.
-If $\Phi_1$ and $\Phi_2$ are noninterfering under $M$, then $E_1$ and $E_2$ must be noninterfering.
+If $\Phi_1$ and $\Phi_2$ are non-interfering under $M$, then $E_1$ and $E_2$ must be non-interfering.
 \end{lem}
@@
 implementation that is different from other systems and relies on the
-soundness of privileges.  Noninterference of memory operations is
+soundness of privileges.  Non-interference of memory operations is
 something that can only be determined after evaluation of an
@@
 memory \cite{Harris05}.  At the other extreme are systems
-that check noninterference statically, such as Jade \cite{Rinard98} and
-DPJ \cite{Bocchino11}.  In contrast, Legion checks noninterference of privileges
+that check non-interference statically, such as Jade \cite{Rinard98} and
+DPJ \cite{Bocchino11}.  In contrast, Legion checks non-interference of privileges
 at runtime, which is much simpler and more efficient than checking
-noninterference of dynamic memory traces.  Even though the privileges
+non-interference of dynamic memory traces.  Even though the privileges
 themselves are static, the region mapping $M$ is dynamic.  Testing
-noninterference on the privileges with physical regions allows
+non-interference on the privileges with physical regions allows
 parallel execution in many more cases than a purely static analysis
@@
 To convert the required analysis to one that can be performed on regions, we modify
-the noninterference operator for memory operations to consider only locations in $L_{excl}$ and show that
-traces that are noninterfering under this modified check satisfy the requirement for atomic
-execution.  We then modify the noninterference operator on privileges to allow 
+the non-interference operator for memory operations to consider only locations in $L_{excl}$ and show that
+traces that are non-interfering under this modified check satisfy the requirement for atomic
+execution.  We then modify the non-interference operator on privileges to allow 
 $M(r_1) \cap M(r_2) \not= \emptyset$ if either $atomic(r_1)$ or $simult(r_1)$ is in $Q_1$ and
 either $atomic(r_2)$ or $simult(r_2)$ is in $Q_2$. We then show that subexpressions with 
-privileges that are {\em atomically noninterfering} (under a dynamic region mapping
+privileges that are {\em atomically non-interfering} (under a dynamic region mapping
 $M$) always have memory traces compatible with atomically-reordered execution.  Finally,
@@
 checks are trivially satisfied, and any interleaving of the constituent traces is valid.
-We again modify the noninterference criteria for memory operations, traces, and privileges, this time
+We again modify the non-interference criteria for memory operations, traces, and privileges, this time
 only allowing $simult$ coherence annotations. We show that subexpression with privileges
-that are {\em simultaneously noninterfering} (under mapping $M$) can be executed
+that are {\em simultaneously non-interfering} (under mapping $M$) can be executed
 in parallel, which includes all the cases covered by Theorem~\ref{thm:parallelexec} as well
@@
 permissible.  We do so by proving that any memory traces that can result from the evaluation
-of the subexpression must be noninterfering (allowing parallel execution by
+of the subexpression must be non-interfering (allowing parallel execution by
 Lemma~\ref{lem:noninteffects}).
@@
 $e_1$ and $e_2$ (with the usual consistent $M$, $L$, $H$, and $S$), must be
-noninterfering.
+non-interfering.
 \end{thm}
@@
 privileges.  If they interfere they are serialized; otherwise they are
-considered for parallel execution as parallel subtasks.  Although determining that $e_1$
-and $e_2$ are noninterfering guarantees $e_3$ and $e_4$ are
-noninterfering, the converse is not true.  It is possible for $e_1$
+considered for execution as parallel subtasks.  Although determining that $e_1$
+and $e_2$ are non-interfering guarantees $e_3$ and $e_4$ are
+non-interfering, the converse is not true.  It is possible for $e_1$
 and $e_2$ to be judged to be potentially interfering and be serialized
-even when the subexpressions are noninterfering.  This loss of
+even when the subexpressions are non-interfering.  This loss of
 precision can impact performance, and did in a few instances in our
@@
 In cases where $e_1$ and $e_2$ are in fact
-noninterfering, region partitions can be further refined to make the
+non-interfering, region partitions can be further refined to make the
 disjointness visible in the privileges.  Alternately, the code can be
 refactored to lift the subexpressions up to the same level, allowing
-the dynamic test based on the region mapping to infer noninterference
+the dynamic test based on the region mapping to infer non-interference
 in more cases.  

### 2012-07-11 doc/popl2013/related.tex
s/permission/privilege/g

--- 2012-07-10 doc/popl2013/related.tex
@@
 Deterministic Parallel Java (DPJ) is the only other region-based parallel system of which we are
-aware\cite{Bocchino09}.  While there are similarities in the permission system and we have
+aware\cite{Bocchino09}.  While there are similarities between DPJ's effects and Legion privileges and we have
 reused some DPJ notation, there are differences stemming from DPJ's static approach.
@@
 While we have borrowed some separation logic notation, we ultimately chose to use a 
-permissions system as our primary formalism because separation logic does not easily
-support the interleaving of operations to non-disjoint regions of memory.
+privileges system as our primary formalism because separation logic does not easily
+support reasoning about the interleaving of operations to non-disjoint regions of memory.
 

### 2012-07-11 doc/popl2013/example.tex
Updating the example with the some changes

@@
 three phases: calculation of each wire's current, distribution of charge to
-the nodes to which each wire is connected, and updating of each node's voltage.
+the nodes connected to each wire, and updating of node voltages.
 
-The top-level function is {\tt simulate\_circuit} (line 15), which is
+The entry function is {\tt simulate\_circuit} (line 15), which is
 parameterized on regions {\tt rl}, {\tt rw}, and {\tt rn}.
@@
 {\tt simulate\_circuit} only accesses these regions and their
-subregions.
+subregions at compile-time.
 
@@
 {\tt color\_circuit} function (line 18-19).  A coloring maps region
-elements to ``colors'' (i.e. integers), with each color corresponding
+elements to colors (i.e. integers), with each color corresponding
  to a new subregion
@@
 After creating the colorings, the application partitions the node and wire
-regions
-into {\em subregions} according to the colorings (lines 21-25).  For
+regions into {\em subregions} according to the colorings (lines 21-25).  For
 simplicity, our example only partitions the circuit into two pieces.
-Line 21 uses {\tt pc.1} to partition {\tt rn} into {\tt rn0}, and {\tt rn1}.
+Line 21 uses coloring {\tt pc.1} to partition {\tt rn} into {\tt rn0}, and {\tt rn1}.
 Subregion {\tt rn0} will contain all locations with a color of 1 in {\tt pc.1},
@@
 
-This function also demonstrates the benefit of being able to dynamically
-(re-)discover disjointness.  Safely running both instances of 
-{\tt calc\_new\_currents} in parallel depends on knowing that {\tt rw0} and
-{\tt rw1} in {\tt execute\_time\_steps} are disjoint.  Similarly, parallel
-execution of the instances of {\tt update\_voltage} requires knowing that
-{\tt rn0} is disjoint from {\tt rn1}.  In both cases, this knowledge was
-statically available in {\tt simulate\_circuit} and could have been captured
-in a region relationship, but it would have resulted in much more complicated
-code.
-Instead, an inexpensive dynamic disjointness check is performed by the Legion
-runtime, and the safety of parallel execution follows from the soundness of
-the type system.
+%This function also demonstrates the benefit of being able to dynamically
+%(re-)discover disjointness.  Safely running both instances of 
+%{\tt calc\_new\_currents} in parallel depends on knowing that {\tt rw0} and
+%{\tt rw1} in {\tt execute\_time\_steps} are disjoint.  Similarly, parallel
+%execution of the instances of {\tt update\_voltage} requires knowing that
+%{\tt rn0} is disjoint from {\tt rn1}.  In both cases, this knowledge was
+%statically available in {\tt simulate\_circuit} and could have been captured
+%in a region relationship, but it would have resulted in much more complicated
+%code.
+%Instead, an inexpensive dynamic disjointness check is performed by the Legion
+%runtime, and the safety of parallel execution follows from the soundness of
+%the type system.
+The {\tt execute\_time\_steps} function also demonstrates the benefit of dynamically
+being able to discover disjointness.  To run the two intances of 
+{\tt calc\_new\_currents} in parallel requires knowing that {\tt rw0}
+and {\tt rw1} are disjoint.  There is a similar requirement for parallel
+execution of the two {\tt update\_voltage} calls with {\tt rn0} and {\tt rn1}.  In
+both cases, this knowledge was statically available in {\tt simulate\_circuit}
+and could have been captured in a region relationship at the cost
+of much more complicated code.  Instead, an inexpensive dynamic disjointness
+check by the Legion runtime will discover the parallelism.
 
@@
 an example of {\tt atomic}, a relaxed coherence mode requiring
-that operations to {\tt rn} and {\tt rg} must appear atomic relative
+that operations to {\tt rn} and {\tt rg} appear atomic relative
 to other functions using those regions.  The most relaxed coherence
-mode is {\tt simult}; all functions with 
-simultaneous coherence are allowed to access the region concurrently
-with other functions having simultaneous coherence to the region.
+mode is {\tt simult}; simultaneous coherence allows concurrent access 
+to the region by all functions with simultaneous coherence on the region.
 

### 2012-07-11 doc/popl2013/evaluation.tex

@@
 are conducted on the Keeneland supercomputer\cite{Keeneland}.  Each node of the Keeneland
-machine consists of two Xeon 5660 processors, three Tesla M2090 GPUs, and 24 GB of DRAM.  Nodes
+machine consists of two Xeon 5660 CPUs, three Tesla M2090 GPUs, and 24 GB of DRAM.  Nodes
 are connected by a QDR Infiniband interconnect.
@@
 using particles that move within a regular grid of cells.  To perform operations in 
-parallel the cells are partitioned.  Unlike
+parallel the array of cells is partitioned.  Unlike
 Circuit, Fluid first creates regions and partitions them before
@@
 \label{subsec:amr}
-The third application is an adaptive mesh refinement (AMR) benchmark 
+The third application is a Legion port of an adaptive mesh refinement (AMR) benchmark 
 from BoxLib \cite{BoxLib}.  The algorithm solves the two
@@
 %to the values of refined cells.  
-The AMR application uses separate regions for every level of
-refinement.  The regions at each level are partitioned several ways to provide multiple views of
+Every levelof refinement uses a separate region, which is partitioned several ways to support multiple views of
 the cells.  One partitioning separates cells into pieces that can be updated in
@@
 %These region relationships capture both intra- and inter-level disjointness information.  
-The dynamic nature of AMR requires that regions can be created and partitioned at runtime.  
+The dynamic nature of AMR requires that regions be created and partitioned at runtime.  
 %The many ways in which cells are used also requires multiple simultaneous
@@
 These applications show Legion is capable of 
-expressing applications with both regular (Fluid,AMR) and irregular (Circuit) pointer data structures.
+expressing applications with both regular (Fluid, AMR) and irregular (Circuit) pointer data structures.
 Being able to dynamically create and partition regions at runtime is crucial to Legion's
-ability to handle applications that make runtime decisions about data organization(AMR).  Legion
+ability to handle applications that make runtime decisions about data organization (AMR).  Legion
 is able to capture both allocate-then-partition (Circuit) and partition-then-allocate (Fluid) algorithms. 
 Having multiple partitions of regions is necessary for describing the many ways 
-that data can be accessed (Circuit,AMR).  All the types of privileges and coherence we have
+that data can be accessed (Circuit, AMR).  All the types of privileges and coherence we have
 described are needed in different applications.  Region relationships
@@
 When testing applications we frequently encountered memory corruption due to
-illegal region accesses caused by a confluence of application bugs and lack of checking.  Many times these corruptions 
-occurred on remote nodes or on GPUs, requiring heroic debugging efforts.
+illegal region accesses caused by application bugs.
+In many cases, this corruption occurred between nodes in the cluster or on 
+GPUs, environments for which debugging tools are primitive at best.
 %  Finding these bugs required either multiple {\tt gdb} sessions 
@@
 %GPUs.  
-To mitigate this problem and guarantee safety we added 
-dynamic checks on  all region accesses for both CPUs and GPUs.  However, these checks
-added considerable runtime overhead.  
-To avoid the overhead of dynamic checks we implemented the type, privilege, and coherence checker we have described. 
-We rewrote all of our applications in this language and used our type checker to verify them.  Once the applications type checked we were
+To location the application bugs causing these illegal accesses, we initially added 
+dynamic checks on  all region accesses for both CPUs and GPUs.  
+Unsurprisingly, these checks added considerable runtime overhead.  
+To preserve the benefit of checking every access, but eliminate the cost of dynamic checks we implemented the type, privilege, and coherence checker we have described. 
+We then rewrote all of our applications in this language and used our type checker to verify them.  Once the applications type checked we were
 able to safely elide all the dynamic region access checks.
@@
 memory pressure.  For the GPU kernels in the Circuit application checking required up to 8 additional 
-registers per thread.  While the GPU kernels in Circuit where not bound by 
+registers per thread.  While the GPU kernels in Circuit were not bound by 
 available on-chip memory, kernels that are would be susceptible to extreme performance 
@@
 The scalability of Legion derives directly from Theorem~\ref{thm:hiersched}.  This 
-theorem makes it safe for the runtime to only perform dependence checks between two
+theorem proves it is safe for the runtime to only perform dependence checks between two
 tasks that share the same immediate parent task instead of having to consider all
@@
 share the same immediate parent are all created on the node where the parent is executing,
-all dependence checks and scheduling decisions can be performed locally with no communication.  
+all dependence checks and scheduling decisions can be performed locally with no inter-processor communication.  
 Without this property the runtime would be forced to inform all other processors in the system 

### 2012-07-11 doc/popl2013/intro.tex
tweaks related to elements, locations

--- 2012-07-11 doc/popl2013/intro.tex
@@
 a region $R$ is disjointly partitioned three distinct ways 
-and each partition allocates every element in $R$ to a
+and each partition assigns every element in $R$ to a
 subregion (partitions need not be total in Legion), then every element
@@
 and relatively cheap
-when done at the granularity of regions instead of individual memory
+when done at the granularity of regions instead of individual heap
 locations.  Thus, Legion falls between fully static systems, such as
@@
 memory.  Legion still has a significant static component; in
-particular, the required privileges for function calls is checked statically
+particular, the required privileges for function calls are checked statically
 and region

### 2012-07-11 doc/popl2013/langdef.tex
added definition of program order
tweaks related to elements, locations

@@
 that still illustrates all of the important issues.  Types
-include booleans, integers, tuples, and pointers.  Note that pointers
-are annotated with a list of regions into which they may point; in a
-well-typed program, the pointer is guaranteed to point to at least one
-of those regions. There is a special type for {\em colorings}, which
+include booleans, integers, tuples, and pointers.  Pointers
+are annotated with a list of regions - a non-null pointer must point to a 
+location that is contained in at least one of the regions. There is a special
+type for {\em colorings}, which
 are used to specify partitions of regions into subregions.
@@
 matches the store resulting from applying the
-constituent traces in program order for all locations in $L_1$.  For
+constituent traces in program order for all locations in $L_1$.  Program order
+is implicitly defined in the operational semantics, and is equivalent to a 
+postorder traversal of an expression's tree.  For
 the atomic location set, we allow the constituent traces to be

### 2012-07-11 doc/popl2013/evaluation.tex
speel cheeking

@@
 from BoxLib \cite{BoxLib}.  The algorithm solves the two
-dimensional heat diffusion equation on a grid of cells using three levels of refinement with subrefinements
+dimensional heat diffusion equation on a grid of cells using three levels of refinement with sub-refinements
 randomly placed on the surface.  
@@
 %to the values of refined cells.  
-Every levelof refinement uses a separate region, which is partitioned several ways to support multiple views of
+Every level of refinement uses a separate region, which is partitioned several ways to support multiple views of
 the cells.  One partitioning separates cells into pieces that can be updated in

### 2012-07-11 doc/popl2013/intro.tex
speel cheeking

@@
 kinds of computing technologies: networks of nodes at the top level,
-multiple chips per node, mutiple cores within a chip, and, most
+multiple chips per node, multiple cores within a chip, and, most
 recently, multiple accelerators (usually GPUs) per node, which 
@@
 {\em non-interfering} (have no ordering requirements for correctness), 
-then the any descendent of $t_1$ in the task hierarchy is non-interfering
-with any descendent of $t_2$.  This theorem is the basis for correct distributed
+then the any descendant of $t_1$ in the task hierarchy is non-interfering
+with any descendant of $t_2$.  This theorem is the basis for correct distributed
 task scheduling in a Legion implementation, which is crucial for
@@
 itself---i.e., the partitions need to be dynamically computed.
-Furthermore, it is useful to allow multiple simultaneous partitions of
+Furthermore, it is useful to allow multiple, simultaneous partitions of
 the same data, providing different views on to that data. Thus, in Legion the 
@@
 }
-\begin{lstlisting}[float={t},label={lst:circuit_ex},caption={Circuilt Simulation}]
+\begin{lstlisting}[float={t},label={lst:circuit_ex},caption={Circuit Simulation}]
 --                        <voltage,current,charge,capacitance>
@@
 
-The important difference bewteen Legion and the more static approaches
+The important difference between Legion and the more static approaches
 is that Legion allows aliasing, but does so in a structured way that minimizes
@@
 We also show that the cost of checking for non-interference of region privileges is low, showing that a much more expressive
-and dynamic langauge is not incompatible with high performance.
+and dynamic language is not incompatible with high performance.
 

### 2012-07-11 doc/popl2013/langdef.tex
speel cheeking

@@
 provided the region relation's constraints hold.  The
-corresponding {\em unpack} expression gives (new) names to those regions.  For simplicitly, neither the 
+corresponding {\em unpack} expression gives (new) names to those regions.  For simplicity, neither the 
 {\em partition} nor the {\em unpack} expression allows shadowing of region names.
@@
 \caption{Legion Core Type System}
+\vspace{-5mm}
 \label{fig:types}
@@
 \caption{Privilege and Constraint Closure}
+\vspace{-5mm}
 \label{fig:closure}
@@
 memory locations and their colors for a partition.  These locations
-support partition-then-allocate, ensuring the memory mappping $M$ has
+support partition-then-allocate, ensuring the memory mapping $M$ has
 locations for the new subregions that can be allocated by {\tt new}.
@@
 \label{fig:opsemfns}
+\vspace{-5mm}
 \end{figure*}
@@
 guaranteed to access the heap only in ways permitted by the privileges under which it was typed.
-A judgement $\privconsist{E}{\Phi}$ holds if all the memory operations in memory trace $E$ are of the types and
+A judgment $\privconsist{E}{\Phi}$ holds if all the memory operations in memory trace $E$ are of the types and
 to locations covered by the privileges in $\Phi$:
@@
 \item $\localconsist{L}{\Gamma}$, local values have types consistent with the environment under which the expression was typed
-\item $\storeconsist{S}{H}$, the initial state of the store contains values whose types are consisten with the heap typing
+\item $\storeconsist{S}{H}$, the initial state of the store contains values whose types are consistent with the heap typing
 \end{itemize}
@@
 \label{fig:constprop}
+\vspace{-5mm}
 \end{figure*}
@@
 \item {\em coherent} - a read of location $l$ should see the effect of all writes and reductions to $l$
-occuring earlier in the interleaved trace, and
+occurring earlier in the interleaved trace, and
 \item {\em sequentially equivalent} - the value read by each read 
@@
 \label{fig:validinterleave}
+\vspace{-5mm}
 \end{figure*}
@@
 to the store.  (If $L_2 = \emptyset$ initially, atomic reads without a preceding write are not 
-checked, but are defered to a point where traces that might be ordered before $E$ are included.)  The $seq\_equiv$
+checked, but are deferred to a point where traces that might be ordered before $E$ are included.)  The $seq\_equiv$
 test verifies every read in $E'$ with a location in $L_1$ 
@@
 concatenating the constituent traces in program order.  However, there is a
-complication related to the nondeterminism that is allowed for reads.  Although
+complication related to the non-determinism that is allowed for reads.  Although
 each subexpression's trace will have been checked for coherence, the checking
@@
 clobber sets.  It is possible that when an interleaving of these traces
-is attemped in the parent expression's context, there may no valid interleaving
+is attempted in the parent expression's context, there may no valid interleaving
 if the wrong guesses were made in the derivation of a subexpression.
@@
 The swapping algorithm as follows: Associate with each operation in $E'$ the index of that operation
-in $E_1 \dplus \ldots \dplus E_n$.  Any consectutive pair of operations $\epsilon_1$ and $\epsilon_2$
+in $E_1 \dplus \ldots \dplus E_n$.  Any consecutive pair of operations $\epsilon_1$ and $\epsilon_2$
 for which $index(\epsilon_1) > index(\epsilon_2)$ is a {\em misordered pair}.  For any such pair,
@@
 evaluation of $e_3$ within the expression tree.  $E_1$ will be formed from a tree of valid
-interleavings, each of which must include all the memory operations of each consitituent trace.
+interleavings, each of which must include all the memory operations of each constituent trace.
 By induction over the number of intermediate subexpressions between $e_1$ and $e_3$, we

### 2012-07-11 doc/popl2013/related.tex
speel cheeking

@@
 Legion is most directly related to Sequoia \cite{Fatahalian06}.  Sequoia is array-based, with
-a limited repetoire of ways to partition arrays.  Sequoia is a static language with a single unified hierarchy
+a limited repertoire of ways to partition arrays.  Sequoia is a static language with a single unified hierarchy
 of tasks and data; Legion is more dynamic with separate task and region hierarchies.
@@
 There is also a difference in emphasis: DPJ requires shared memory, while Legion 
-is designed for distributed heterogenous machines.
+is designed for distributed heterogeneous machines.
 

### 2012-07-11 doc/popl2013/intro.tex
Getting the paper ready to submit

--- 2012-07-11 doc/popl2013/intro.tex
@@
 a region $R$ is disjointly partitioned three distinct ways 
-and each partition assigns every element in $R$ to a
+and each partition allocates every element in $R$ to a
 subregion (partitions need not be total in Legion), then every element
@@
 and relatively cheap
-when done at the granularity of regions instead of individual heap
+when done at the granularity of regions instead of individual memory
 locations.  Thus, Legion falls between fully static systems, such as
@@
 memory.  Legion still has a significant static component; in
-particular, the required privileges for function calls are checked statically
+particular, the required privileges for function calls is checked statically
 and region

### 2012-07-11 doc/popl2013/langdef.tex
Getting the paper ready to submit
last (hah!) tweaks to langdef

@@
 Core Legion is explicitly typed using judgments of the form
-\begin{center}
-$\typeenv{e}{T}$
-\end{center}
-where
+$$\typeenv{e}{T}$$
+where:
 \begin{itemize}
@@
 privileges or constraints on subregions.  Figure~\ref{fig:closure} defines closure
-operations for $\Phi^*$ and $\Omega^*$ that add implied constraints.
+operations for $\Phi^*$ and $\Omega^*$ that explicitly add these implied constraints.
 The type checking rules are in Figure~\ref{fig:types}.  Many of the rules are 
@@
 a Legion function - it is limited to privileges given to it by its caller.  (The
-top-level task of a Legion program begins with a single, empty region from which other
+top-level task of a Core Legion program begins with a single, empty region from which other
 regions can be partitioned and full read and
@@
 Core Legion's operational semantics rules have the following form:
-
-\begin{center}
-$\opsenv{e}{v}{E}$
-\end{center}
-
+$$\opsenv{e}{v}{E}$$
 where: 
@@
 
+The proof of this (and other lemmas in this section) is included in \cite{LegionTypes12}.
 We now show that the concatenation of the memory traces from these serially
-executed subexpression is always a valid interleaving:
+executed subexpressions is always a valid interleaving:
 
@@
 \item Finally, if both operations are reductions to the same location, they may be swapped if they
-use the same reduction function, as reduction functions are assumed to commute.  
+use the same reduction function, as each reduction function is assumed to commute with itself.  
 %(We 
@@
 }
-The proof is included in \cite{LegionTypes12}.
 We now extend the non-interference operator to entire memory traces, providing a way to say that
@@
 $$
-\begin{array}{l@{ }l}
-priv_1(r_1) & \nonint[M] priv_2(r_2) \Leftrightarrow (M(r_1) \cap M(r_2) = \emptyset)\ \vee \\
-& (priv_1 = reads \wedge priv_2 = reads)\ \vee \\
-& (priv_1 = reduces_{id_1} \wedge priv_2 = reduces_{id_2} \wedge id_1 = id_2)
+\begin{array}{@{}l@{}}
+priv_1(r_1) \nonint[M] priv_2(r_2) \Leftrightarrow (M(r_1) \cap M(r_2) = \emptyset)\ \vee \\
+\hspace{0.35in} (priv_1 = reads \wedge priv_2 = reads)\ \vee \\
+\hspace{0.35in} (priv_1 = reduces_{id_1} \wedge priv_2 = reduces_{id_2} \wedge id_1 = id_2)
 \end{array}
@@
 Let $e_1$ and $e_2$ be well-typed Legion expressions using privileges $\Phi_1$ and
-$\Phi_2$ respectively .  Let $e_3$ be a subexpression of $e_1$ and $e_4$ be a 
+$\Phi_2$ respectively, where $\Phi_1 \nonint[M] \Phi_2$.  Let $e_3$ be a subexpression of $e_1$ and $e_4$ be a 
 subexpression of $e_2$.  Any memory traces $E_3$ of $e_3$ and $E_4$ of $e_4$ resulting from evaluation of

### 2012-07-11 doc/popl2013/example.tex
speel cheeking
fixed caption typo

@@
 The {\tt execute\_time\_steps} function also demonstrates the benefit of dynamically
-being able to discover disjointness.  To run the two intances of 
+being able to discover disjointness.  To run the two instances of 
 {\tt calc\_new\_currents} in parallel requires knowing that {\tt rw0}
@@
 
-\begin{lstlisting}[float={t},label={lst:circuit_leaf},caption={Circuilt Leaf Functions}]
+\begin{lstlisting}[float={t},label={lst:circuit_leaf},caption={Circuit Leaf Functions}]
 function calc_new_currents[rl,rw,rn,rg] ( ptr_list : wire_list<rl,rw,rn,rg>@rl ), 

### 2012-07-11 doc/popl2013/intro.tex
intro

--- 2012-07-11 doc/popl2013/intro.tex
@@
 kinds of computing technologies: networks of nodes at the top level,
-multiple chips per node, multiple cores within a chip, and, most
+multiple chips per node, multiple cores within a chip, and, 
 recently, multiple accelerators (usually GPUs) per node, which 
-can themselves be decomposed.  We present the operational and static
+can themselves be further decomposed.  We present the operational and static
 semantics of Legion\cite{Legion12}, a programming model targeted at providing an
@@
 performance; a centralized scheduler would be a bottleneck
-because of the huge latencies involved in communicating in the target
+because of the large communication latencies in the target
 class of machines.

### 2012-07-11 doc/popl2013/example.tex
Minor edits

--- 2012-07-11 doc/popl2013/example.tex
@@
 mode is {\tt simult}; simultaneous coherence allows concurrent access 
-to the region by all functions with simultaneous coherence on the region.
+to the region by all functions that are using the region
+in a simultaneous mode.  The interaction between tasks using the same
+region with different coherence
+modes is formalized in Section~\ref{sec:coherence}.
 

### 2012-07-11 doc/popl2013/intro.tex
intro

--- 2012-07-11 doc/popl2013/intro.tex
@@
 regions is that it introduces aliasing which makes it difficult 
-to statically figure out which computations can be run in parallel.  
+to statically determine which computations can be run in parallel.  
 One approach is to outlaw aliasing which allows for a static analysis
@@
 Sequoia and DPJ take this approach and allow only a single static partition of any region 
-to eliminate aliasing at the cost of the flexibility of their language.
+to eliminate region aliasing, at some cost in expressiveness.
 
@@
 The important difference between Legion and the more static approaches
-is that Legion allows aliasing, but does so in a structured way that minimizes
+is that Legion allows region aliasing, but does so in a structured way that minimizes
 runtime overheads.  The key observation is that
@@
 memory.  Legion still has a significant static component; in
-particular, the required privileges for function calls are checked statically
+particular, the required privileges for function calls
 and region
-pointer dereferences are checked statically.  The dynamic alias tests
+pointer dereferences are checked statically.  The dynamic alias checks
 for independence are done at the granularity of regions and one check 

### 2012-07-11 doc/popl2013/related.tex
Getting everything on 12 pages

@@
 Deterministic Parallel Java (DPJ) is the only other region-based parallel system of which we are
-aware\cite{Bocchino09}.  While there are similarities between DPJ's effects and Legion privileges and we have
-reused some DPJ notation, there are differences stemming from DPJ's static approach.
+aware\cite{Bocchino09}.  While there are similarities between DPJ's effects and Legion privileges, 
+%and we have reused some DPJ notation, 
+there are differences stemming from DPJ's static approach.
 Regions in Legion are first-class and can be created, partitioned, packed, and unpacked 

### 2012-07-11 doc/popl2013/intro.tex
intro

@@
 for grouping heap data into larger collections, it must test the
-overlap between two sets of data on each individual memory access, which is
+overlap between two sets of data on each individual memory location, which is
 significantly more expensive.

### 2012-07-11 doc/popl2013/langdef.tex
Getting everything on 12 pages
langdef

@@
 \label{fig:langdef}
+\vspace{-5mm}
 \end{figure*}
@@
 include booleans, integers, tuples, and pointers.  Pointers
-are annotated with a list of regions - a non-null pointer must point to a 
+are annotated with a list of regions---a non-null pointer must point to a 
 location that is contained in at least one of the regions. There is a special
@@
 access privileges and coherence it requires.  A function may be executed
-within the current task or run as a separate subtask - this decision is made
+within the current task or run as a separate subtask---this decision is made
 by the Legion task scheduler.  Function types are

### 2012-07-11 doc/popl2013/evaluation.tex
eval

@@
 parallel the array of cells is partitioned.  Unlike
-Circuit, Fluid first creates regions and partitions them before
-allocating cells in regions.  Another difference is that Fluid
+Circuit, Fluid creates and partitions regions before
+allocating cells in them (i.e., partition-then-allocate).  Another difference is that Fluid
 maintains separate regions for ghost cells rather using multiple partitions of
@@
 %GPUs.  
-To location the application bugs causing these illegal accesses, we initially added 
+To locate the application bugs causing these illegal accesses, we initially added 
 dynamic checks on  all region accesses for both CPUs and GPUs.  

### 2012-07-11 doc/popl2013/langdef.tex
def
langdef

@@
 access privileges and coherence it requires.  A function may be executed
-within the current task or run as a separate subtask---this decision is made
+within the current task or run in parallel as a separate subtask---this decision is made
 by the Legion task scheduler.  Function types are
@@
 provided the region relation's constraints hold.  The
-corresponding {\em unpack} expression gives (new) names to those regions.  For simplicity, neither the 
+corresponding {\em unpack} expression gives (new) names to those regions.  For simplicity, in Core Legion neither the 
 {\em partition} nor the {\em unpack} expression allows shadowing of region names.

### 2012-07-11 doc/popl2013/evaluation.tex
Small edit to the evaluation

--- 2012-07-11 doc/popl2013/evaluation.tex
@@
 degradation due to the extra registers required for checking.  We also measured the overhead
-of the dynamic checks associated with checking task call privileges but found them to be negligible.
+of the dynamic checks associated with checking task call privileges but found them to be negligible,
+demonstrating that runtime non-interference checks are inexpensive in Legion.
 % relative to the overhead of region access checks.

### 2012-07-13 doc/popl2013/example.tex

@@
 to the region by all functions that are using the region
-in a simultaneous mode.  The interaction between tasks using the same
+in a simultaneous mode.  The interaction between {\em tasks} (which are functions chosen by the runtime to execute in parallel) using the same
 region with different coherence

### 2012-07-13 doc/popl2013/legion-types.tex
updates
more updates
Getting rid of the ugly boxes for the type system
Adding a separate example section
s/permission/privilege/g
Getting the paper ready to submit

@@
-\documentclass[9pt,nocopyrightspace,preprint]{sigplanconf}
-
+\documentclass[9pt,nocopyrightspace]{sigplanconf}
+% preprint for dates
 \bibliographystyle{plainnat}
@@
 \usepackage{url}
+\usepackage{bcprules}
+\usepackage{verbatim}
+\usepackage{graphicx}
+\usepackage{chngpage}
 
-\newcommand{\infrule}[2]{\displaystyle\frac{\displaystyle\strut{#1}}{\displaystyle\strut {#2}}}
+%\newcommand{\infrule}[2]{\displaystyle\frac{\displaystyle\strut{#1}}{\displaystyle\strut {#2}}}
 \newcommand{\deref}{\ast}
@@
 \newtheorem{lemma}[theorem]{Lemma}
+\newcommand{\texcomment}[1]{}
 
@@
 with complex memory hierarchies.  Logical regions are first-class 
-values in Legion, allowing programmers to specify independence 
-and locality properties of program data.  We present
-a type system with permissions and coherence properties for reasoning about logical regions.  
+values in Legion, and programmers may {\em partition} regions in multiple different
+ways into subregions.  We present
+a type system with privileges and coherence properties for reasoning about logical regions.  
 We prove the soundness of the type system and use it to show
@@
 
+\input{example}
 %\input{outline}

### 2012-07-13 doc/popl2013full/conclusion.tex
clone of popl2013 into popl2013full

\section{Conclusion}
\label{sect:conclusion}

Modern architectures have dramatically increased in complexity in recent years and we believe that new programming models
are needed to take advantage of the potential of such machines.  We have presented the semantics and type system
for Legion, which differs from previous efforts in having a much more flexible programming model (in particular,
regions are first-class values and data can be dynamically partitioned in as many ways as needed) while maintaining
soundness of privileges and coherence.  The key insight is that we can divide the problem between static and dynamic
checking differently, keeping what is expensive to check dynamically (region pointer checks) in the static system while incurring very little if any performance cost for the work deferred to runtime (the region aliasing checks).  We have also given
a formal basis for hierarchical, distributed scheduling in Legion, and provided experimental evidence  that the
expected benefits of the Legion design are realized in real-world programs.

### 2012-07-13 doc/asplos2013/legion-llr.tex
Adding an abstract for the asplos paper

@@
 
-\title{Legion: Expressing Locality and Independence with Logical Regions}
+\title{Asynchronous High-Performance Language Primitives for Distributed Heterogenous Architectures}
 \numberofauthors{3}
@@
 \begin{abstract}
-Last year's limit was 12 pages, including bibliography and appendices.
-Deadlines:
-\begin{tabular}{ll}
-Abstract deadline & July 16, 2012 3pm PDT \\
-Paper deadline & July 23, 2012 3pm PDT \\
-Author response & around October 15, 2012 \\
-Author notification & November 9, 2012
-\end{tabular}
+%The limit is 12 pages, including bibliography and appendices.
+%Deadlines:
+%\begin{tabular}{ll}
+%Abstract deadline & July 16, 2012 3pm PDT \\
+%Paper deadline & July 23, 2012 3pm PDT \\
+%Author response & around October 15, 2012 \\
+%Author notification & November 9, 2012
+%\end{tabular}
+Modern parallel architectures are composed of both heterogenous processors
+and complex distributed memory hierarchies.  We present a low-level programming
+interface that provides language primitives for targetting this class of
+machines.  Our interface provides mechanisms for the creation and movement
+of data through the memory hierarchy as well as a novel synchronization primitive 
+called a deferred lock.  To hide the large latencies associated with the target
+set of machines the interface is made asynchronous via an event system.  All
+operations return an event to be triggered when the operation completes.  Events
+can be chained together and used to describe dependences between operations.
+We describe an implementation of our interface capable of running on
+a large supercomputer with both CPU and GPU processors and a high speed interconnect.
+We employ several micro-benchmarks to show that our implementation is close to the metal.
+We also demonstrate that the interface is sufficiently powerful for the construction
+of an advanced runtime system.  We measure the performance of three real-world applications
+that use or system on the Keeneland supercomputer.
 \end{abstract}
@@
 Introduction goes here.
+
+\input{outline}
 

### 2012-07-13 doc/asplos2013/legion-llr.tex
Fixing some spelling

@@
 %\end{tabular}
-Modern parallel architectures are composed of both heterogenous processors
+Modern parallel architectures are composed of both heterogeneous processors
 and complex distributed memory hierarchies.  We present a low-level programming
-interface that provides language primitives for targetting this class of
+interface that provides language primitives for targeting this class of
 machines.  Our interface provides mechanisms for the creation and movement

### 2012-07-14 runtime/legion.cc
Fixing a couple of bugs in the high level runtime

@@ -5744,7 +5744,6 @@ namespace RegionRuntime {
 #ifdef DEBUG_HIGH_LEVEL
         current_taken = true;
 #endif
-        // We only need to compute the return size one time
         size_t buffer_size = 2*sizeof(Processor) + sizeof(size_t) + compute_task_size();
         for (typename std::vector<CT>::const_iterator it = chunks.begin();
               it != chunks.end(); it++)
@@ -10308,6 +10307,7 @@ namespace RegionRuntime {
       result += sizeof(size_t); // num valid instances
       result += sizeof(size_t); // num reduction instances
       result += (region_state.valid_instances.size() * (sizeof(InstanceID) + sizeof(bool)));
+      result += (region_state.reduction_instances.size() * (sizeof(InstanceID) + sizeof(bool)));
       result += sizeof(PartState);
       result += sizeof(DataState);
       result += sizeof(ReductionOpID);
@@ -11823,22 +11823,25 @@ namespace RegionRuntime {
         // First check to see if we're doing sanitizing or actual opening
         if (ren.sanitizing)
         {
-          // Create an instance info for each of the valid instances we can find 
-          std::set<Memory> locations;
-          get_physical_locations(ren.ctx_id,locations,true/*recurse*/,false/*reduction*/);
-#ifdef DEBUG_HIGH_LEVEL
-          //assert(!locations.empty());
-#endif
-          // For each location, make a physical instance here
-          for (std::set<Memory>::const_iterator it = locations.begin();
-                it != locations.end(); it++)
+          if (!IS_REDUCE(ren.usage))
           {
-            std::pair<InstanceInfo*,bool> result = find_physical_instance(ren.ctx_id,*it,true/*recurse*/,false/*reduction*/);
-            InstanceInfo *info = result.first;
-            if (info->handle != handle)
+            // Create an instance info for each of the valid instances we can find 
+            std::set<Memory> locations;
+            get_physical_locations(ren.ctx_id,locations,true/*recurse*/,false/*reduction*/);
+#ifdef DEBUG_HIGH_LEVEL
+            //assert(!locations.empty());
+#endif
+            // For each location, make a physical instance here
+            for (std::set<Memory>::const_iterator it = locations.begin();
+                  it != locations.end(); it++)
             {
-              InstanceInfo *new_info = ren.ctx->create_instance_info(handle,info);
-              update_valid_instances(ren.ctx_id, new_info, false/*write*/, true/*owned*/);
[…]

### 2012-07-14 doc/asplos2013/legion-llr.tex
Another small edit

@@
 a large supercomputer with both CPU and GPU processors and a high speed interconnect.
-We employ several micro-benchmarks to show that our implementation is close to the metal.
-We also demonstrate that the interface is sufficiently powerful for the construction
+We employ several micro-benchmarks to show that our implementation approaches
+the limits of the underlying hardware.
+We also demonstrate that the interface is sufficiently powerful to serve as the foundation 
 of an advanced runtime system.  We measure the performance of three real-world applications
-that use or system on the Keeneland supercomputer.
+that use our system on the Keeneland supercomputer.
 \end{abstract}

### 2012-07-14 doc/asplos2013/legion-llr.tex
More edits to the abstract

@@
 and complex distributed memory hierarchies.  We present a low-level programming
-interface that provides language primitives for targeting this class of
-machines.  Our interface provides mechanisms for the creation and movement
+interface that provides simple primitives for targeting this class of
+machines.  Our interface includes mechanisms for the creation and movement
 of data through the memory hierarchy as well as a novel synchronization primitive 
@@
 operations return an event to be triggered when the operation completes.  Events
-can be chained together and used to describe dependences between operations.
+can be chained together and used to create dependences between operations.
 We describe an implementation of our interface capable of running on

### 2012-07-14 doc/asplos2013/legion-llr.tex
Fixing a spelling error in the title

@@
 
-\title{Asynchronous High-Performance Language Primitives for Distributed Heterogenous Architectures}
+\title{Asynchronous High-Performance Language Primitives for Distributed Heterogeneous Architectures}
 \numberofauthors{3}
@@
 We employ several micro-benchmarks to show that our implementation approaches
-the limits of the underlying hardware.
+the performance limits of the underlying hardware.
 We also demonstrate that the interface is sufficiently powerful to serve as the foundation 

### 2012-07-15 doc/popl2013full/langdef.tex
proof

--- 2012-07-11 doc/popl2013/langdef.tex
@@
 
-We spare the reader the lengthy proof, which can be found in
-\cite{LegionTypes12}, and merely outline the general strategy.  The
-proof is a standard induction on the structure of the derivation in
-the operational semantics.  The proof is organized into six pieces,
+The proof is a standard induction on the structure of the derivation
+For each of 
+the Core Legion expressions there are six separate parts to the proof,
 one for each of the three consistency properties that must be shown
 for the execution environment of subexpressions and one for each of
-the three properties of the execution's results we wish to show:
-
-\texcomment{
-For each of 
-the 23 different forms a Legion expression can have, we assume the soundness of the subexpressions'
-evaluations (provided we can show the consistency of the mappings and store used by that
-subexpression) and prove that the result of the expression's evaluation is sound.  Although there
-are a multitude of cases to be proven, many of them are similar, and benefit from
-the use of the following lemmas:
+the three properties of the execution's results.
+Many of the cases are similar, and benefit from
+the use of the following straighforward lemmas:
 
@@
 \label{lemma:heapconst:apply}
+\rm
 If $\storeconsist{S}{H}$ and $\traceconsist{E}$, then $\storeconsist{\text{apply}(S, E)}{H}$.
@@
 \label{lemma:heapconst:effects1}
+\rm
 If $\traceconsist{E_1}$ and $\traceconsist{E_2}$, then $\traceconsist{E_1 \dplus E_2}$.
@@
 \label{lemma:heapconst:effects2}
-If $\traceconsist{E_1}$ and $\traceconsist{E_2}$ and $valid\_interleave(S, C, E', E_1, E_2)$, then $\traceconsist{E'}$.
+\rm
+If $\traceconsist{E_1}$ and $\traceconsist{E_2}$ and \\
+$valid\_interleave(S, C, E', E_1, E_2)$, then $\traceconsist{E'}$.
 \end{lem}
@@
 \label{lemma:effsound:effects1}
+\rm
 If $\privconsist{E_1}{\Phi}$ and $\privconsist{E_2}{\Phi}$, then $\privconsist{E_1 \dplus E_2}{\Phi}$.
@@
 \label{lemma:effsound:effects2}
-If $\privconsist{E_1}{\Phi}$ and $\privconsist{E_2}{\Phi}$ and $valid\_interleave(S, C, E', E_1, E_2)$, then $\privconsist{E'}{\Phi}$.
+\rm
+If $\privconsist{E_1}{\Phi}$ and $\privconsist{E_2}{\Phi}$ and \\ 
+$valid\_interleave(S, C, E', E_1, E_2)$, then $\privconsist{E'}{\Phi}$.
 \end{lem}
@@
 \label{lemma:omegaclosure}
+\rm
 $\mapconsist{\Omega^*}$ if and only if $\mapconsist{\Omega}$.
@@
 \label{lemma:phiclosure}
+\rm
 $\privconsist{E}{\Phi^*}$ if and only if $\privconsist{E}{\Phi}$.
@@
 
-The proofs of these lemmas can also be found here\cite{LegionTypes12}.  
-}
-
+\begin{proof}
+The proof is by cases on the structure of the derivation of $e$.
+\begin{itemize}
+
+\item
+Case [E-Bool]. By assumption, 
+$\opsenv{bv}{bv}{[]}$ and
+$\typeenv{bv}{T}$ and
+$\mapconsist{\Omega}$ and
+$\localconsist{L}{\Gamma}$ and 
+$\storeconsist{S}{H}$.
+The only type rule that can apply is [T-Bool], so $T = bool$.
+Therefore $\valueconsist{bv}{bool}$ and 
+$\traceconsist{[]}$ and 
+$\privconsist{[]}{\Phi}$ all follow trivially.
+
+\item Case [E-Int].  
+By assumption, 
+$\opsenv{iv}{iv}{[]}$ and 
+$\typeenv{iv}{T}$ and 
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and $\storeconsist{S}{H}$.
+The only type rule that can apply is [T-Int], so $T = int$.
+Therefore $\valueconsist{iv}{int}$ and 
+$\traceconsist{[]}$ and 
+$\privconsist{[]}{\Phi}$ all follow trivially.
+
+\item Case [E-Null].
+By assumption, 
+$\opsenv{\text{null\ }T@r}{null}{[]}$ and 
+$\typeenv{\text{null}}{T'}$ and 
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and $\storeconsist{S}{H}$.
+The only type rule that can apply is [T-Null], so $T' = T@r$.
+Therefore $\valueconsist{null}{T@r}$, $\traceconsist{[]}$ and $\privconsist{[]}{\Phi}$ all follow trivially.
+
+\item
+Case [E-MakeTuple].
+By assumption, 
+$\opsenv{\langle e_1, e_2 \rangle}{\langle v_1,v_2 \rangle}{E'}$ and 
+$\typeenv{\langle e_1, e_2 \rangle}{T}$ and 
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and $\storeconsist{S}{H}$.
+The only type rule that can apply is [T-MakeType], so $T = \langle T_1, T_2 \rangle$.
+
+For the subexpression $e_1$, 
+from the form of rule [E-MakeTuple], we know that $\opsenv{e_1}{v_1}{E_1}$.
+From the form of rule [T-MakeTuple], we know that $\typeenv{e_1}{T_1}$.
+Then, by induction, we have $\valueconsist{v_1}{T_1}$, $\traceconsist{E_1}$ and 
+      $\privconsist{E_1}{\Phi}$.
+
+For the subexpression $e_2$, 
+from the form of rule [E-MakeTuple], we know that $\opsenv[S=S']{e_2}{v_2}{E_2}$
+where $S' = apply(S,E_1)$.  By Lemma~\ref{lemma:heapconst:apply}, $\storeconsist{S' = \text{apply}(S,E)}{H}$.
+From the form of rule [T-MakeTuple], we know that $\typeenv{e_2}{T_2}$.
+Then, by induction, we have $\valueconsist{v_2}{T_2}$, $\traceconsist{E_2}$ and 
+      $\privconsist{E_2}{\Phi}$.
+
+Finally, from the precedinig steps we can conclude that $\valueconsist{\langle v_1, v_2\rangle}{\langle T_1, T_2 \rangle}$.
+By Lemma~\ref{lemma:heapconst:effects1}, $\traceconsist{E_1 \dplus E_2}$, and by Lemma~\ref{lemma:effsound:effects1}
+$\privconsist{E_1 \dplus E_2}{\Phi}$.
+
+\item Case [E-Tuple1].
+By assumption, 
+$\opsenv{e\text{.1}}{v_1}{E}$ and 
+$\typeenv{e\text{.1}}{T_1}$ and 
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and $\storeconsist{S}{H}$.
+From the form of [E-Tuple1] we know that
+$\opsenv{e}{\langle v_1, v_2 \rangle}{E}$.
+The only type rule that can apply is [T-Tuple1], and from the form of the rule we know
+$\typeenv{e}{\langle T_1, T_2 \rangle}$.
+By induction, we have $\valueconsist{\langle v_1, v_2 \rangle}{\langle T_1, T_2 \rangle}$, $\traceconsist{E}$ and 
+      $\privconsist{E}{\Phi}$.
+It follows immediately that $\valueconsist{v_1}{T_1}$ also holds.
+
+\item Case [E-Tuple2] is symmetric to case [E-Tuple1].
+
+\item[E-New]
+\item[E-Var]
+\item[E-IsNull-F]
+\item[E-IsNull-T]
+\item[E-Let]
+\item[E-Add]
+\item[E-Compare]
+\item[E-IfElse-T]
+\item[E-IfElse-F]
+\item[E-Newcolor]
+\item[E-Color]
+\item[E-UpRgn]
+\item[E-DownRgn]
+\item[E-Read]
+\item[E-Write]
+\item[E-Reduce]
+\item[E-Pack]
+\item[E-Unpack]
+\item[E-Call]
+\item[E-Program]
+
+\end{itemize}
+
+\end{proof}
 \vspace{4pt}
@@
 
-%% \begin{table*}
-%% \centering
-%% {\small
-%% \begin{tabular}{l|p{5cm}|p{5cm}|p{5cm}}
-%% {\bf Rule} & 
-%% \multicolumn{1}{|c|}{$\mathbf{v \sim_H M \llbracket T \rrbracket}$} &
-%% \multicolumn{1}{|c|}{$\mathbf{E \sim H}$} &
-%% \multicolumn{1}{|c|}{$\mathbf{E : \Phi}$} \\
-%% \hline
-%% BoolConst & & & \\
-%% IntConst & & & \\
-%% TupleBuild & & & \\
-%% TupleField & & & \\
-%% LocalVar & & & \\
-%% NullPtr & & & \\
-%% NewPtr & & & \\
-%% IsNull & & & \\
-%% UpRegion & & & \\
-%% DownRegion & & & \\
-%% Read & & & \\
-%% Write & & & \\
-%% Reduce & & & \\
-%% Newcolor & & & \\
-%% Color & & & \\
-%% IntMath & & & \\
-%% IntCmp & & & \\
-%% Let & & & \\
-%% IfElse & & & \\
-%% TaskCall & & & \\
-%% Partition & & & \\
-%% Pack & & & \\
-%% Unpack & & &
-%% \end{tabular}}
-%% \caption{Proof pieces}
-%% \end{table*}
+ \begin{table*}
+ \centering
+ {\small
+ \begin{tabular}{l|p{5cm}|p{5cm}|p{5cm}}
+ {\bf Rule} & 
+ \multicolumn{1}{|c|}{$\mathbf{v \sim_H M \llbracket T \rrbracket}$} &
+ \multicolumn{1}{|c|}{$\mathbf{E \sim H}$} &
+ \multicolumn{1}{|c|}{$\mathbf{E : \Phi}$} \\
+ \hline
+ BoolConst & & & \\
+ IntConst & & & \\
+ TupleBuild & & & \\
+ TupleField & & & \\
+ LocalVar & & & \\
+ NullPtr & & & \\
+ NewPtr & & & \\
+ IsNull & & & \\
+ UpRegion & & & \\
+ DownRegion & & & \\
+ Read & & & \\
+ Write & & & \\
+ Reduce & & & \\
+ Newcolor & & & \\
+ Color & & & \\
+ IntMath & & & \\
+ IntCmp & & & \\
+ Let & & & \\
+ IfElse & & & \\
+ TaskCall & & & \\
+ Partition & & & \\
+ Pack & & & \\
+ Unpack & & &
+ \end{tabular}}
+ \caption{Proof pieces}
+ \end{table*}
 
@@
 \end{lem}
-\texcomment{
 \begin{proof}
@@
 \end{proof}
-}
+
 We now extend the non-interference operator to entire memory traces, providing a way to say that
@@
 \end{proof}
-}
+
 We next extend non-interference to privileges.  We 
@@
 $$
+%% 
 %% \begin{center}
@@
 \end{lem}
-\texcomment{
+
 \begin{proof}
@@
 \end{proof}
-}
+
 We can now state the key result that allows the Legion runtime to safely evaluate subexpressions as
@@
 
-\texcomment{
 Let $e_1$ and $e_2$ be well-typed Legion expressions using privileges $\Phi_1$ and
@@
 Lemma~\ref{lem:noninteffects}).
-}
+
 \begin{thm}
@@
 \end{thm}
-\texcomment{
+
 \begin{proof}
@@
 \end{proof}
-} The Legion task scheduler uses Theorem~\ref{thm:hiersched} in the
+
+ The Legion task scheduler uses Theorem~\ref{thm:hiersched} in the
 following way.  Sibling function calls (those invoked within the same

### 2012-07-15 apps/circuit/bulksync_results/run_ckt_sim_experiments.sh
Adding results of bulk synchronous experiments on keeneland

@@ -1,0 +1,78 @@
+#!/bin/bash
+
+cd /nics/d/home/sequoia/mike_legion/region/apps/circuit/bulksync_results/
+cat $PBS_NODEFILE | uniq > unique.txt
+export GASNET_SSH_NODEFILE=/nics/d/home/sequoia/mike_legion/region/apps/circuit/bulksync_results//unique.txt
+
+#echo "48 pieces 1 nodes 1 cpus 1 gpus"
+#GASNET_BACKTRACE=1 gasnetrun_ibv -n 1 /nics/d/home/sequoia/mike_legion/region/apps/circuit/ckt_sim -ll:cpu 1 -ll:gpu 1 -ll:dma 2 -ll:zsize 1536 -ll:fsize 1024 -ll:csize 4096 -ll:gsize 1536 -npp 2500 -wpp 16384 -p 48 -l 10 -level 3 1> ckt_sim_48_1_1_1.stdio 2> ckt_sim_48_1_1_1.stderr 
+
+#echo "48 pieces 1 nodes 1 cpus 2 gpus"
+#GASNET_BACKTRACE=1 gasnetrun_ibv -n 1 /nics/d/home/sequoia/mike_legion/region/apps/circuit/ckt_sim -ll:cpu 1 -ll:gpu 2 -ll:dma 2 -ll:zsize 1536 -ll:fsize 1024 -ll:csize 4096 -ll:gsize 1536 -npp 2500 -wpp 16384 -p 48 -l 10 -level 3 1> ckt_sim_48_1_1_2.stdio 2> ckt_sim_48_1_1_2.stderr 
+
+echo "48 pieces 1 nodes 1 cpus 3 gpus"
+GASNET_BACKTRACE=1 gasnetrun_ibv -n 1 /nics/d/home/sequoia/mike_legion/region/apps/circuit/ckt_sim -ll:cpu 1 -ll:gpu 3 -ll:dma 2 -ll:zsize 1536 -ll:fsize 1024 -ll:csize 4096 -ll:gsize 1536 -npp 2500 -wpp 16384 -p 48 -l 10 -level 3 -sync 1 1> ckt_sim_48_1_1_3.stdio 2> ckt_sim_48_1_1_3.stderr 
+
+#echo "48 pieces 2 nodes 1 cpus 1 gpus"
[…]

### 2012-07-15 apps/circuit/bulksync_results/run_ckt_sim_experiments_large.sh

@@ -1,0 +1,33 @@
+#!/bin/bash
+
+cd /nics/d/home/sequoia/mike_legion/region/apps/circuit/bulksync_results/
+cat $PBS_NODEFILE | uniq > /nics/d/home/sequoia/mike_legion/region/apps/circuit/bulksync_results//unique_big2.txt
+export GASNET_SSH_NODEFILE=/nics/d/home/sequoia/mike_legion/region/apps/circuit/bulksync_results//unique_big2.txt
+
+#echo "48 pieces 16 nodes 1 cpus 1 gpus"
+#GASNET_BACKTRACE=1 gasnetrun_ibv -n 16 /nics/d/home/sequoia/mike_legion/region/apps/circuit/ckt_sim -ll:cpu 1 -ll:gpu 1 -ll:dma 2 -ll:zsize 1536 -ll:fsize 1024 -ll:csize 4096 -ll:gsize 1536 -npp 2500 -wpp 16384 -p 48 -l 10 -level 3 1> ckt_sim_48_16_1_1.stdio 2> ckt_sim_48_16_1_1.stderr 
+
+#echo "48 pieces 16 nodes 1 cpus 2 gpus"
+#GASNET_BACKTRACE=1 gasnetrun_ibv -n 16 /nics/d/home/sequoia/mike_legion/region/apps/circuit/ckt_sim -ll:cpu 1 -ll:gpu 2 -ll:dma 2 -ll:zsize 1536 -ll:fsize 1024 -ll:csize 4096 -ll:gsize 1536 -npp 2500 -wpp 16384 -p 48 -l 10 -level 3 1> ckt_sim_48_16_1_2.stdio 2> ckt_sim_48_16_1_2.stderr 
+
+#echo "48 pieces 16 nodes 1 cpus 3 gpus"
+#GASNET_BACKTRACE=1 gasnetrun_ibv -n 16 /nics/d/home/sequoia/mike_legion/region/apps/circuit/ckt_sim -ll:cpu 1 -ll:gpu 3 -ll:dma 2 -ll:zsize 1536 -ll:fsize 1024 -ll:csize 4096 -ll:gsize 1536 -npp 2500 -wpp 16384 -p 48 -l 10 -level 3 -sync 1 1> ckt_sim_48_16_1_3.stdio 2> ckt_sim_48_16_1_3.stderr 
+
[…]

### 2012-07-15 doc/popl2013full/langdef.tex

@@
 
-\item[E-New]
-\item[E-Var]
-\item[E-IsNull-F]
-\item[E-IsNull-T]
+\item Case[E-New].
+By assumption, 
+$\opsenv{\text{new }T@r}{l}{[]}$ and 
+$\typeenv{\text{new }T@r}{T@r}$ and 
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and 
+$\storeconsist{S}{H}$.
+The hypothesis of the [E-New] rule is that $l \in M(r)$ and $H(l) = M \llbracket T \rrbracket$,
+which establishes $\valueconsist{l}{T}$.  
+The other two conclusions,
+$\traceconsist{[]}$ and 
+$\privconsist{[]}{\Phi}$, follow trivially.
+\item Case [E-Var].
+By assumption, 
+$\opsenv{id}{v}{[]}$ and 
+$\typeenv{id}{T}$ and 
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and 
+$\storeconsist{S}{H}$.
+The hypothesis of the [E-Var] rule is $L(id) = v$.
+From $\localconsist{L}{\Gamma}$ we have $\valueconsist{L(id)}{\Gamma(id)}$.
+The hypothesis of the [T-Var] rule is $\Gamma(id) = T$.
+Putting these three facts together we conclude $\valueconsist{v}{T}$.
+The other two conclusions,
+$\traceconsist{[]}$ and 
+$\privconsist{[]}{\Phi}$, follow trivially.
+
+\item Case [E-IsNull-F].
+By assumption, 
+$\opsenv{\text{isnull}(e)}{\textit{false}}{E}$ and 
+$\typeenv{\text{isnull}(e)}{bool}$ and 
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and 
+$\storeconsist{S}{H}$.
+From the form of [E-IsNull-F] we know that
+$\opsenv{e}{l}{E}$, and
+and from the form of the type rule [T-IsNull] we know
+$\typeenv{e}{T@(\oton{r})}$.
+Clearly $\valueconsist{\textit{false}}{bool}$, and
+$\traceconsist{E}$ and 
+      $\privconsist{E}{\Phi}$ follow by induction.
+
+\item Case [E-IsNull-T] is almost identical to case [E-IsNull-F], with $l$ replaced by \textit{null} and
+\textit{true} replaced by \textit{false}.
+
 \item[E-Let]

### 2012-07-16 doc/popl2013full/langdef.tex
proof

@@
 \typeenvx e_1 : T_1 \\
-\typeenvx[G={\Gamma[id/T_1]}] e_2 : T_2
+\typeenvx[G={\Gamma[T_1/id]}] e_2 : T_2
 \end{array}}
@@
 \begin{tabular}{m{6cm}m{5cm}m{6cm}}
-\infrule[T-Let]{\typeenvx e_1 : T_1 \\ \typeenvx[G={\Gamma[id/T_1]}] e_2 : T_2}{\typeenvx : \text{let}\ id : T_1 \text{in}\ e_2 : T_2} &
+\infrule[T-Let]{\typeenvx e_1 : T_1 \\ \typeenvx[G={\Gamma[T_1/id]}] e_2 : T_2}{\typeenvx : \text{let}\ id : T_1 \text{in}\ e_2 : T_2} &
 \infrule[T-Add]{\typeenvx e_1 : int \\ \typeenvx e_2 : int}{\typeenvx e_1 + e_2 : int} &
@@
 & & (\forall id, l, v. reduce_{id}(l, v) \in E \Rightarrow \\
-& & (M \llbracket \Gamma \rrbracket (id) = (\hat T_1, \hat T_2), \emptyset, \emptyset \Rightarrow \hat T_1) \wedge H(l) = \hat T_1 \wedge \valueconsist[]{v}{\hat T_2}) \\
+& & (M \llbracket \Gamma \rrbracket (id) = (\hat T_1, \hat T_2), \emptyset, \emptyset \rightarrow \hat T_1) \wedge H(l) = \hat T_1 \wedge \valueconsist[]{v}{\hat T_2}) \\
 \end{array}
@@
 the use of the following straighforward lemmas:
-
 \begin{lem}
@@
 
-\begin{proof}
-The proof is by cases on the structure of the derivation of $e$.
-\begin{itemize}
-
-\item
-Case [E-Bool]. By assumption, 
-$\opsenv{bv}{bv}{[]}$ and
-$\typeenv{bv}{T}$ and
-$\mapconsist{\Omega}$ and
-$\localconsist{L}{\Gamma}$ and 
-$\storeconsist{S}{H}$.
-The only type rule that can apply is [T-Bool], so $T = bool$.
-Therefore $\valueconsist{bv}{bool}$ and 
-$\traceconsist{[]}$ and 
-$\privconsist{[]}{\Phi}$ all follow trivially.
-
-\item Case [E-Int].  
-By assumption, 
-$\opsenv{iv}{iv}{[]}$ and 
-$\typeenv{iv}{T}$ and 
-$\mapconsist{\Omega}$ and 
-$\localconsist{L}{\Gamma}$ and $\storeconsist{S}{H}$.
-The only type rule that can apply is [T-Int], so $T = int$.
-Therefore $\valueconsist{iv}{int}$ and 
-$\traceconsist{[]}$ and 
-$\privconsist{[]}{\Phi}$ all follow trivially.
-
-\item Case [E-Null].
-By assumption, 
-$\opsenv{\text{null\ }T@r}{null}{[]}$ and 
-$\typeenv{\text{null}}{T'}$ and 
-$\mapconsist{\Omega}$ and 
-$\localconsist{L}{\Gamma}$ and $\storeconsist{S}{H}$.
-The only type rule that can apply is [T-Null], so $T' = T@r$.
-Therefore $\valueconsist{null}{T@r}$, $\traceconsist{[]}$ and $\privconsist{[]}{\Phi}$ all follow trivially.
-
-\item
-Case [E-MakeTuple].
-By assumption, 
-$\opsenv{\langle e_1, e_2 \rangle}{\langle v_1,v_2 \rangle}{E'}$ and 
-$\typeenv{\langle e_1, e_2 \rangle}{T}$ and 
-$\mapconsist{\Omega}$ and 
-$\localconsist{L}{\Gamma}$ and $\storeconsist{S}{H}$.
-The only type rule that can apply is [T-MakeType], so $T = \langle T_1, T_2 \rangle$.
-
-For the subexpression $e_1$, 
-from the form of rule [E-MakeTuple], we know that $\opsenv{e_1}{v_1}{E_1}$.
-From the form of rule [T-MakeTuple], we know that $\typeenv{e_1}{T_1}$.
-Then, by induction, we have $\valueconsist{v_1}{T_1}$, $\traceconsist{E_1}$ and 
-      $\privconsist{E_1}{\Phi}$.
-
-For the subexpression $e_2$, 
-from the form of rule [E-MakeTuple], we know that $\opsenv[S=S']{e_2}{v_2}{E_2}$
-where $S' = apply(S,E_1)$.  By Lemma~\ref{lemma:heapconst:apply}, $\storeconsist{S' = \text{apply}(S,E)}{H}$.
-From the form of rule [T-MakeTuple], we know that $\typeenv{e_2}{T_2}$.
-Then, by induction, we have $\valueconsist{v_2}{T_2}$, $\traceconsist{E_2}$ and 
-      $\privconsist{E_2}{\Phi}$.
-
-Finally, from the precedinig steps we can conclude that $\valueconsist{\langle v_1, v_2\rangle}{\langle T_1, T_2 \rangle}$.
-By Lemma~\ref{lemma:heapconst:effects1}, $\traceconsist{E_1 \dplus E_2}$, and by Lemma~\ref{lemma:effsound:effects1}
-$\privconsist{E_1 \dplus E_2}{\Phi}$.
-
-\item Case [E-Tuple1].
-By assumption, 
-$\opsenv{e\text{.1}}{v_1}{E}$ and 
-$\typeenv{e\text{.1}}{T_1}$ and 
-$\mapconsist{\Omega}$ and 
-$\localconsist{L}{\Gamma}$ and $\storeconsist{S}{H}$.
-From the form of [E-Tuple1] we know that
-$\opsenv{e}{\langle v_1, v_2 \rangle}{E}$.
-The only type rule that can apply is [T-Tuple1], and from the form of the rule we know
-$\typeenv{e}{\langle T_1, T_2 \rangle}$.
-By induction, we have $\valueconsist{\langle v_1, v_2 \rangle}{\langle T_1, T_2 \rangle}$, $\traceconsist{E}$ and 
-      $\privconsist{E}{\Phi}$.
-It follows immediately that $\valueconsist{v_1}{T_1}$ also holds.
-
-\item Case [E-Tuple2] is symmetric to case [E-Tuple1].
-
-\item Case[E-New].
-By assumption, 
-$\opsenv{\text{new }T@r}{l}{[]}$ and 
-$\typeenv{\text{new }T@r}{T@r}$ and 
-$\mapconsist{\Omega}$ and 
-$\localconsist{L}{\Gamma}$ and 
-$\storeconsist{S}{H}$.
-The hypothesis of the [E-New] rule is that $l \in M(r)$ and $H(l) = M \llbracket T \rrbracket$,
-which establishes $\valueconsist{l}{T}$.  
-The other two conclusions,
-$\traceconsist{[]}$ and 
-$\privconsist{[]}{\Phi}$, follow trivially.
-\item Case [E-Var].
-By assumption, 
-$\opsenv{id}{v}{[]}$ and 
-$\typeenv{id}{T}$ and 
-$\mapconsist{\Omega}$ and 
-$\localconsist{L}{\Gamma}$ and 
-$\storeconsist{S}{H}$.
-The hypothesis of the [E-Var] rule is $L(id) = v$.
-From $\localconsist{L}{\Gamma}$ we have $\valueconsist{L(id)}{\Gamma(id)}$.
-The hypothesis of the [T-Var] rule is $\Gamma(id) = T$.
-Putting these three facts together we conclude $\valueconsist{v}{T}$.
-The other two conclusions,
-$\traceconsist{[]}$ and 
-$\privconsist{[]}{\Phi}$, follow trivially.
-
-\item Case [E-IsNull-F].
-By assumption, 
-$\opsenv{\text{isnull}(e)}{\textit{false}}{E}$ and 
-$\typeenv{\text{isnull}(e)}{bool}$ and 
-$\mapconsist{\Omega}$ and 
-$\localconsist{L}{\Gamma}$ and 
-$\storeconsist{S}{H}$.
-From the form of [E-IsNull-F] we know that
-$\opsenv{e}{l}{E}$, and
-and from the form of the type rule [T-IsNull] we know
-$\typeenv{e}{T@(\oton{r})}$.
-Clearly $\valueconsist{\textit{false}}{bool}$, and
-$\traceconsist{E}$ and 
-      $\privconsist{E}{\Phi}$ follow by induction.
-
-\item Case [E-IsNull-T] is almost identical to case [E-IsNull-F], with $l$ replaced by \textit{null} and
-\textit{true} replaced by \textit{false}.
-
-\item[E-Let]
-\item[E-Add]
-\item[E-Compare]
-\item[E-IfElse-T]
-\item[E-IfElse-F]
-\item[E-Newcolor]
-\item[E-Color]
-\item[E-UpRgn]
-\item[E-DownRgn]
-\item[E-Read]
-\item[E-Write]
-\item[E-Reduce]
-\item[E-Pack]
-\item[E-Unpack]
-\item[E-Call]
-\item[E-Program]
-
-\end{itemize}
-
-\end{proof}
+We first give an outline of the idea behindthe proof.  For each subexpression of a compound expression,
+there are three consistency properties we must prove to be able to apply induction: \\
+
 \vspace{4pt}
@@
 
- \begin{table*}
- \centering
- {\small
- \begin{tabular}{l|p{5cm}|p{5cm}|p{5cm}}
- {\bf Rule} & 
- \multicolumn{1}{|c|}{$\mathbf{v \sim_H M \llbracket T \rrbracket}$} &
- \multicolumn{1}{|c|}{$\mathbf{E \sim H}$} &
- \multicolumn{1}{|c|}{$\mathbf{E : \Phi}$} \\
- \hline
- BoolConst & & & \\
- IntConst & & & \\
- TupleBuild & & & \\
- TupleField & & & \\
- LocalVar & & & \\
- NullPtr & & & \\
- NewPtr & & & \\
- IsNull & & & \\
- UpRegion & & & \\
- DownRegion & & & \\
- Read & & & \\
- Write & & & \\
- Reduce & & & \\
- Newcolor & & & \\
- Color & & & \\
- IntMath & & & \\
- IntCmp & & & \\
- Let & & & \\
- IfElse & & & \\
- TaskCall & & & \\
- Partition & & & \\
- Pack & & & \\
- Unpack & & &
- \end{tabular}}
- \caption{Proof pieces}
- \end{table*}
+\begin{proof}
+The proof of the theorem is by cases on the structure of the derivation of $e$.
+\begin{itemize}
+
+\item
+Case [E-Bool]. By assumption, 
+$$\opsenv{bv}{bv}{[]}$$ and
+$$\typeenv{bv}{T}$$ and
+$\mapconsist{\Omega}$ and
+$\localconsist{L}{\Gamma}$ and 
+$\storeconsist{S}{H}$.
+The only type rule that can apply is [T-Bool], so $T = bool$.
+Therefore $\valueconsist{bv}{bool}$ and 
+$\traceconsist{[]}$ and 
+$\privconsist{[]}{\Phi}$ all follow trivially.
+
+\item Case [E-Int].  
+By assumption, 
+$$\opsenv{iv}{iv}{[]}$$ and 
+$$\typeenv{iv}{T}$$ and 
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and $\storeconsist{S}{H}$.
+The only type rule that can apply is [T-Int], so $T = int$.
+Therefore $\valueconsist{iv}{int}$ and 
+$\traceconsist{[]}$ and 
+$\privconsist{[]}{\Phi}$ all follow trivially.
+
+\item Case [E-Null].
+By assumption, 
+$$\opsenv{\text{null\ }T@r}{null}{[]}$$ and 
+$$\typeenv{\text{null}}{T'}$$ and 
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and $\storeconsist{S}{H}$.
+The only type rule that can apply is [T-Null], so $T' = T@r$.
+Therefore $\valueconsist{null}{T@r}$, $\traceconsist{[]}$ and $\privconsist{[]}{\Phi}$ all follow trivially.
+
+\item
+Case [E-MakeTuple].
+By assumption, 
+$$\opsenv{\langle e_1, e_2 \rangle}{\langle v_1,v_2 \rangle}{E'}$$ and 
+$$\typeenv{\langle e_1, e_2 \rangle}{T}$$ and 
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and $\storeconsist{S}{H}$.
+The only type rule that can apply is [T-MakeTuple], so $T = \langle T_1, T_2 \rangle$.
+
+For the subexpression $e_1$, 
+from the form of rule [E-MakeTuple], we know that $$\opsenv{e_1}{v_1}{E_1}$$
+From the form of rule [T-MakeTuple], we know that $$\typeenv{e_1}{T_1}$$
+Then, by induction, we have $\valueconsist{v_1}{T_1}$, $\traceconsist{E_1}$ and 
+      $\privconsist{E_1}{\Phi}$.
+
+For the subexpression $e_2$, 
+from the form of rule [E-MakeTuple], we know that $$\opsenv[S=S']{e_2}{v_2}{E_2}$$
+where $S' = apply(S,E_1)$.  By Lemma~\ref{lemma:heapconst:apply}, $\storeconsist{S' = \text{apply}(S,E)}{H}$.
+From the form of rule [T-MakeTuple], we know that $$\typeenv{e_2}{T_2}$$
+Then, by induction, we have $\valueconsist{v_2}{T_2}$, $\traceconsist{E_2}$ and 
+      $\privconsist{E_2}{\Phi}$.
+
+From the preceding steps we can conclude that $\valueconsist{\langle v_1, v_2\rangle}{\langle T_1, T_2 \rangle}$.
+Finally, from [E-MakeTuple] we also have $$valid\_interleave(S,C,E',E_1,E_2)$$ and
+Lemma~\ref{lemma:heapconst:effects2}, $\traceconsist{E'}$ and by Lemma~\ref{lemma:effsound:effects2}
+$\privconsist{E'}{\Phi}$.
+
+\item Case [E-Tuple1].
+By assumption, 
+$$\opsenv{e\text{.1}}{v_1}{E}$$ and 
+$$\typeenv{e\text{.1}}{T_1}$$ and 
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and $\storeconsist{S}{H}$.
+From the form of [E-Tuple1] we know that
+$$\opsenv{e}{\langle v_1, v_2 \rangle}{E}$$
+The only type rule that can apply is [T-Tuple1], and from the form of the rule we know
+$$\typeenv{e}{\langle T_1, T_2 \rangle}$$
+By induction, we have $\valueconsist{\langle v_1, v_2 \rangle}{\langle T_1, T_2 \rangle}$, $\traceconsist{E}$ and 
+      $\privconsist{E}{\Phi}$.
+It follows immediately that $\valueconsist{v_1}{T_1}$ also holds.
+
+\item Case [E-Tuple2] is symmetric to case [E-Tuple1].
+
+\item Case[E-New].
+By assumption, 
+$$\opsenv{\text{new }T@r}{l}{[]}$$ and 
+$$\typeenv{\text{new }T@r}{T@r}$$ and 
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and 
+$\storeconsist{S}{H}$.
+The hypothesis of the [E-New] rule is that $l \in M(r)$ and $H(l) = M \llbracket T \rrbracket$,
+which establishes $\valueconsist{l}{T}$.  
+The other two conclusions,
+$\traceconsist{[]}$ and 
+$\privconsist{[]}{\Phi}$, follow trivially.
+\item Case [E-Var].
+By assumption, 
+$$\opsenv{id}{v}{[]}$$ and 
+$$\typeenv{id}{T}$$ and 
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and 
+$\storeconsist{S}{H}$.
+The hypothesis of the [E-Var] rule is $L(id) = v$.
+From $\localconsist{L}{\Gamma}$ we have $\valueconsist{L(id)}{\Gamma(id)}$.
+The hypothesis of the [T-Var] rule is $\Gamma(id) = T$.
+Putting these three facts together we conclude $\valueconsist{v}{T}$.
+The other two conclusions,
+$\traceconsist{[]}$ and 
+$\privconsist{[]}{\Phi}$, follow trivially.
+
+\item Case [E-IsNull-F].
+By assumption, 
+$$\opsenv{\text{isnull}(e)}{\textit{false}}{E}$$ and 
+$$\typeenv{\text{isnull}(e)}{bool}$$ and 
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and 
+$\storeconsist{S}{H}$.
+From the form of [E-IsNull-F] we know that
+$\opsenv{e}{l}{E}$, and
+and from the form of the type rule [T-IsNull] we know
+$\typeenv{e}{T@(\oton{r})}$.
+Clearly $\valueconsist{\textit{false}}{bool}$, and
+$\traceconsist{E}$ and 
+      $\privconsist{E}{\Phi}$ follow by induction.
+
+\item Case [E-IsNull-T] is almost identical to case [E-IsNull-F], with $l$ replaced by \textit{null} and
+\textit{false} replaced by \textit{true}.
+
+\item Case [E-Let].
+By assumption, 
+$$\opsenv{\text{let } id:T_1 = e_1 \text{ in } e_2}{v_2}{E'}$$ and 
+$$\typeenv{\text{let } id:T_1 = e_1 \text{ in } e_2}{T_2}$$ and 
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and $\storeconsist{S}{H}$.
+
+For the subexpression $e_1$, 
+from [E-Let], we know that $$\opsenv{e_1}{v_1}{E_1}$$
+From type rule [T-Let], we know that $$\typeenv{e_1}{T_1}$$
+Then, by induction, we have $\valueconsist{v_1}{T_1}$, $\traceconsist{E_1}$ and 
+      $\privconsist{E_1}{\Phi}$.
+
+For the subexpression $e_2$, 
+from [E-Let], we have $$\opsenv[L=L',S=S']{e_2}{v_2}{E_2}$$
+where $L' = L[v_1/id]$ and $S' = apply(S,E_1)$.  
+From [T-Let], we have $\typeenv[G={\Gamma[T_1/id]}]{e_2}{T_2}$.
+By Lemma~\ref{lemma:heapconst:apply}, $\storeconsist{S' = \text{apply}(S,E)}{H}$.
+From $\localconsist{L}{\Gamma}$ and $\valueconsist{v_1}{T_1}$ we conclude $\localconsist{L[v_1/id]}{\Gamma[T_1/id]}$.
+Then, by induction, we have $\valueconsist{v_2}{T_2}$, $\traceconsist{E_2}$ and 
+      $\privconsist{E_2}{\Phi}$.
+
+Finally, from [E-Let] we also have $$valid\_interleave(S,C,E',E_1,E_2)$$
+Then, by Lemma~\ref{lemma:heapconst:effects2}, $\traceconsist{E'}$ and by 
+Lemma~\ref{lemma:effsound:effects2} $\privconsist{E'}{\Phi}$.
+
+\item Case [E-Add].
+By assumption, 
+$$\opsenv{e_1 + e_2}{v'}{E'}$$ and 
+$$\typeenv{e_1 + e_2}{T}$$ and 
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and $\storeconsist{S}{H}$.
+The only type rule that can apply is [T-Add], so $T = int$.
+
+For the subexpression $e_1$, 
+from [E-Add] we know $$\opsenv{e_1}{v_1}{E_1}$$
+From [T-Add], we know that $$\typeenv{e_1}{int}$$
+Then, by induction, we have $\valueconsist{v_1}{T_1}$, $\traceconsist{E_1}$ and 
+      $\privconsist{E_1}{\Phi}$.
+
+For the subexpression $e_2$, 
+from [E-Add], we know $$\opsenv[S=S']{e_2}{v_2}{E_2}$$
+where $S' = apply(S,E_1)$.  By Lemma~\ref{lemma:heapconst:apply}, $\storeconsist{S' = \text{apply}(S,E)}{H}$.
+From [T-Add] we know $$\typeenv{e_2}{int}$$
+Then, by induction, we have $\valueconsist{v_2}{T_2}$, $\traceconsist{E_2}$ and 
+      $\privconsist{E_2}{\Phi}$.
+
+It is immediate that $\valueconsist{v_1 + v_2}{int}$.
+Finally, from [E-Add] we also have $$valid\_interleave(S,C,E',E_1,E_2)$$ and
+Lemma~\ref{lemma:heapconst:effects2}, $\traceconsist{E'}$ and by Lemma~\ref{lemma:effsound:effects2},
+$\privconsist{E'}{\Phi}$.
+
+\item Case [E-Compare] is isomorphic to [E-Add].
+
+\item Case [E-IfElse-T].
+ By assumption, 
+$$\opsenv{\text{ if } e_1 \text{ then } e_2 \text{ else } e_3}{v_2}{E}$$ and 
+$$\typeenv{\text{ if } e_1 \text{ then } e_2 \text{ else } e_3}{T}$$ and
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and 
+$\storeconsist{S}{H}$.
+
+For the subexpression $e_1$, 
+from [E-IfElse-T] we know $$\opsenv{e_1}{true}{E_1 \dplus E_2}$$
+From [T-IfElse], we know that $$\typeenv{e_1}{bool}$$
+Clearly $\valueconsist{true}{bool}$, and by induction we have  $\traceconsist{E_1}$ and 
+      $\privconsist{E_1}{\Phi}$.
+
+For the subexpression $e_2$, 
+from [E-IfElse-T] we know $$\opsenv[S=S']{e_2}{v_2}{E_2}$$
+where $S' = apply(S,E_1)$.  By Lemma~\ref{lemma:heapconst:apply}, $\storeconsist{S' = \text{apply}(S,E)}{H}$.
+From [T-IfElse-T] we know $$\typeenv{e_2}{T}$$
+Then, by induction, we have $\valueconsist{v_2}{T}$, $\traceconsist{E_2}$ and 
+      $\privconsist{E_2}{\Phi}$.
+
+Finally, by Lemma~\ref{lemma:heapconst:effects1}, $\traceconsist{E_1 \dplus E_2}$ and by Lemma~\ref{lemma:effsound:effects1},
+$\privconsist{E_1 \dplus E_2}{\Phi}$.
+
+\item Case [E-IfElse-F] is nearly the same as [E-IfElse-T] with {\em false} replacing {\em true} and $e_3$ replacing $e_2$.
+
+\item Case [E-Newcolor].
+By assumption, 
+$$\opsenv{\text{newcolor } r}{K}{[]}$$ 
+where  $K = \{ (l_1,v_1),\ldots,(l_p,i_p) \}$ such that
+$l_i \in M(r)$ for all $i$ and $l_1,\ldots,l_p$ are pairwise distinct locations.
+In addition, we have the assumptions
+$$\typeenv{\text{newcolor } r}{T}$$ and
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and 
+$\storeconsist{S}{H}$.
+The only type rule that can apply if [T-Newcolor], so can infer that $T = \text{coloring}(r)$.
+From the structure of $K$ it follows that $\valueconsist{K}{\text{coloring(r)}}$.
+The other two conclusions,
+$\traceconsist{[]}$ and 
+$\privconsist{[]}{\Phi}$, follow trivially.
+
+\item  Case [E-Color].
+By assumption, 
+$$\opsenv{\text{color($e_1$, $e_2$, $e_3$)}}{K'}{E'}$$ and
+$$\typeenv{\text{color($e_1$, $e_2$,$e_3$)}}{T}$$ and
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and 
+$\storeconsist{S}{H}$.
+The only type rule that can apply if [T-Color], so we know $T = \text{coloring}(r)$.
+
+For the subexpression $e_1$, 
+from [E-Color] we know $$\opsenv{e_1}{K}{E_1}$$
+From [T-Color], we know that $$\typeenv{e_1}{\text{coloring}(r)}$$
+By induction it follows that $\valueconsist{K}{\text{coloring(r)}}$ and $\traceconsist{E_1}$ and 
+      $\privconsist{E_1}{\Phi}$.
+
+For the subexpression $e_2$, 
+from [E-Color] we know $$\opsenv[S=S']{e_2}{l}{E_2}$$ where $S' = \text{apply}(S,E_1)$.
+From [T-Color], we know that $$\typeenv{e_2}{T'@r}$$
+By Lemma~\ref{lemma:heapconst:apply}, $\storeconsist{S'}{H}$.
+By induction, $\valueconsist{l}{T'@r}$ and $\traceconsist{E_2}$ and 
+      $\privconsist{E_2}{\Phi}$.
+
+For the subexpression $e_3$, 
+from [E-Color] we know $$\opsenv[S=S'']{e_3}{v}{E_3}$$ where $S'' \text{apply}(S',E_2)$.
+From [T-Color], we know that $$\typeenv{e_3}{int}$$
+By Lemma~\ref{lemma:heapconst:apply}, $\storeconsist{S''}{H}$.
+By induction, $\valueconsist{v}{int}$ and $\traceconsist{E_3}$ and 
+      $\privconsist{E_3}{\Phi}$.
+
+Now $K'$ is built from $K$ by extending $K$ with a mapping for location $l$.  First, because $\valueconsist{l}{T'@r}$ and
+$\valueconsist{K}{\text{coloring(r))}$ and because $K'$ excludes any previous mapping for $l$ (guaranteeing the the coloring is a function), we conclude $\valueconsist{K'}{\text{coloring}(r)}}$.
+From [E-Color], we also know
+$$valid\_interleave(S, C, E', E_1, E_2, E_3)$$ and by applying Lemma~\ref{lemma:heapconst:effects2} twice
+we can conclude $\traceconsist{E'}$.  Similarly, by applying Lemma~\ref{lemma:effsound:effects2} twice
+we can conlude $\privconsist{E'}{\Phi}$.
+
+\item  Case [E-UpRgn].
+By assumption, $$\opsenv{\text{upregion}(e,\oton{r})}{v}{E}$$ and
+$$\typeenv{\text{upgregion}(e,\oton{r})}{T}$$ and
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and 
+$\storeconsist{S}{H}$.
+The type rule must be [T-UpRgn], so we also know $T = T' @ (\oton{r})$.
+
+From [E-UpRgn] we know $$\opsenv{e}{v}{E}$$ and
+from [T-UpRgn] we know $$\typeenv{e}{T' @ (r_1',\ldots,r_k')}$$
+By induction, we can conclude that $\valueconsist{v}{T'@(r_1',\ldots,r_k')}$,
+$\traceconsist{E}$ and $\privconsist{E}{\Phi}$.
+
+To show $\valueconsist{v}{T'@(\oton{r})}$, we first observe that from
+$\valueconsist{v}{T'@(r_1',\ldots,r_k')}$ we know
+there is an $i$ such that $v \in M(r_i')$ and 
+$H(v) = \hat T'$.   From [T-UpRgn]
+we know $\forall i. \exists j. r_i' \leq r_j \in \Omega^*$.  Combining this with
+$\mapconsist{\Omega}$ and Lemma~\ref{lemma:omegaclosure}, we know there
+is some $r_j$ such that $v \in M(r_j)$.
+Therefore $\valueconsist{v}{T'@(\oton{r})}$.
+
+\item Case [E-DownRgn].
+By assumption, $$\opsenv{\text{downregion}(e,\oton{r})}{v}{E}$$ and
+$$\typeenv{\text{downgregion}(e,\oton{r})}{T}$$ and
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and 
+$\storeconsist{S}{H}$.
+The type rule must be [T-DownRgn], so we also know $T = T' @ (\oton{r})$.
+
+From [E-DownRgn] we know $$\opsenv{e}{l}{E}$$ and
+from [T-DownRgn] we know $$\typeenv{e}{T' @ (r_1',\ldots,r_k')}$$
+By induction, we can conclude that $\valueconsist{l}{T'@(r_1',\ldots,r_k')}$,
+$\traceconsist{E}$ and $\privconsist{E}{\Phi}$.
+
+To show $\valueconsist{v}{T'@(\oton{r})}$ there are two cases.
+If $v = null$ then $\valueconsist{null}{T'@(\oton{r})}$ follows by definition.
+Otherwise, $v = l$ and $l \in M(r_i)$ for some $i$.  We also know $H(v) = \hat T'$
+because $\valueconsist{l}{T'@(r_1',\ldots,r_k')}$.  Therefore
+$\valueconsist{v}{T'@(\oton{r})}$.
+
+\item Case [E-Read].
+By assumption, $\opsenv{\text{read}(e)}{v}{E \dplus [ read(l,excl,v) ]}$ and
+$\typeenv{\text{read}(e)}{T}$ and
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and 
+$\storeconsist{S}{H}$.
+
+From [E-Read] we have $$\opsenv{e}{l}{E}$$ and
+from [T-Read] we know $$\typeenv{e}{T @ (\oton{r})}$$
+By induction, we can conclude that $\valueconsist{l}{T@(\oton{r})}$,
+$\traceconsist{E}$ and $\privconsist{E}{\Phi}$.
+
+To show $\valueconsist{v}{T}$ there are two cases.
+If $v = M(l)$ then the claim follows from $\valueconsist{l}{T@(\oton{r})}$.
+If $v$ is instead some $v'$ of type $H(l)$ the result also follows because $H(l) = \hat{T}$.
+Next, $\traceconsist{E \dplus [ read(l,excl,v)]}$ follows immediately from $\traceconsist{E}$ as
+any read operation is consistent (the resriction is only on writes and reductions).
+
+Finally, we must show $\privconsist{E \dplus [ read(l,excl,v)]}{\Phi}$.
+Since we have already shown $\privconsist{E}{\Phi}$, it suffices to show $\privconsist{[ read(l,excl,v) ]}{\Phi}$.
+By $\valueconsist{l}{T@(\oton{r})}$, there is some $r_i$ such that $l \in M(r_i)$.  Furthermore,
+from [T-Read] we have $reads(r_i) \in \Phi^*$, which shows $\privconsist{[ read(l,excl,v)]}{\Phi^*}$.
+The result follows from Lemma~\ref{lemma:phiclosure}.
+
+\item Case [E-Write].
+By assumption, $$\opsenv{\text{write}(e_1,e_2)}{l}{E' \dplus [ write(l,excl,v) ]}$$ and
+$$\typeenv{\text{write}(e_1,e_2)}{T@(\oton{r})}$$ and
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and 
+$\storeconsist{S}{H}$.
+
+For subexpression $e_1$,
+From [E-Write] we have $$\opsenv{e_1}{l}{E_1}$$ and
+from [T-Write] we know $$\typeenv{e_1}{T @ (\oton{r})}$$.
+By induction, we can conclude that $\valueconsist{l}{T@(\oton{r})}$,
+$\traceconsist{E_1}$ and $\privconsist{E_1}{\Phi}$.
+
+For subexpression $e_2$,
+From [E-Write] we have $$\opsenv[S=S']{e_2}{v}{E_2}$$ where
+$S' = \text{apply}(S,E_1)$.
+From [T-Write] we know $\typeenv{e_2}{T}$.
+Using Lemma~\ref{lemma:heapconst:apply} it follows that $\storeconsist{S'}{H}$. 
+Then by induction, we can conclude that $\valueconsist{v}{T}$,
+$\traceconsist{E_2}$ and $\privconsist{E_2}{\Phi}$.
+
+We have already shown $\valueconsist{l}{T@(\oton{r})}$, so it only remains to show
+$\traceconsist{E' \dplus [ write(l,excl,v) ]}$ and 
+$\privconsist{E' \dplus [ write(l,excl,v) ]}{\Phi}$.
+From [E-Write] we have $$valid\_interleave(S,C,E',E_1,E_2)$$ and therefore by Lemma~\ref{lemma:heapconst:effects2} we have
+$\traceconsist{E'}$.  To show $\traceconsist{[ write(l,excl,v)]}$ we must show $\valueconsist{v}{H(l)}$,
+which follows from $\valueconsist{v}{T}$ and $\valueconsist{l}{T@(\oton{r})}$ and $\storeconsist{S}{H}$.
+Thus, $\traceconsist{E' \dplus [ write(l,excl,v) ]}$.
+
+To show $\privconsist{E' \dplus [ write(l,excl,v) ]}{\Phi}$, we first observe that
+$valid\_interleave(S,C,E',E_1,E_2)$ and Lemma~\ref{lemma:effsound:effects2} imply
+$\privconsist{E'}{\Phi}$.  To show $\privconsist{[ write(l,excl,v) ]}{\Phi}$, we observe that because
+$\valueconsist{l}{T@(\oton{r})}$, there is some $r_i$ such that $l \in M(r_i)$.  Furthermore,
+from [T-Write] we have $writes(r_i) \in \Phi^*$, which shows $\privconsist{[ write(l,excl,v)]}{\Phi^*}$.
+The result follows from Lemma~\ref{lemma:phiclosure}.
+
+\item Case[E-Reduce].
+By assumption, $$\opsenv{\text{reduce}(id,e_1,e_2)}{l}{E' \dplus [ reduce_{id}(l,excl,v) ]}$$ and
+$$\typeenv{\text{reduce}(id,e_1,e_2)}{T_1@(\oton{r})}$$ and
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and 
+$\storeconsist{S}{H}$.
+
+For subexpression $e_1$,
+From [E-Reduce] we have $$\opsenv{e_1}{l}{E_1}$$ and
+from [T-Reduce] we know $$\typeenv{e_1}{T @ (\oton{r})}$$.
+By induction, we can conclude that $\valueconsist{l}{T@(\oton{r})}$,
+$\traceconsist{E_1}$ and $\privconsist{E_1}{\Phi}$.
+
+For subexpression $e_2$,
+From [E-Reduce] we have $$\opsenv[S=S']{e_2}{v}{E_2}$$ where
+$S' = \text{apply}(S,E_1)$.
+From [T-Reduce] we know $$\typeenv{e_2}{T_2}$$
+Using Lemma~\ref{lemma:heapconst:apply} it follows that $\storeconsist{S'}{H}$. 
+Then by induction, we can conclude that $\valueconsist{v}{T_2}$,
+$\traceconsist{E_2}$ and $\privconsist{E_2}{\Phi}$.
+
+We have already shown $\valueconsist{l}{T_1@(\oton{r})}$, so it only remains to show
+$\traceconsist{E' \dplus [ reduce_{id}(l,excl,v) ]}$ and 
+$\privconsist{E' \dplus [ reduce_{id}(l,excl,v) ]}{\Phi}$.
+From [E-Reduce] we have $valid\_interleave(S,C,E',E_1,E_2)$ and therefore by Lemma~\ref{lemma:heapconst:effects2} we have
+$\traceconsist{E'}$.  To show $\traceconsist{[ reduce_{id}(l,excl,v)]}$ we must show three things.
+First, we must show $M \llbracket \Gamma \rrbracket (id) = (\hat T_1, \hat T_2), \emptyset, \emptyset \rightarrow \hat T_1$ which
+follows from the assumption $\Gamma(id) = (T_1, T_2), \emptyset, \emptyset \rightarrow T_1$ in [T-Reduce].
+Second, we must show $H(l) = \hat T_1$, which follows from $\valueconsist{l}{T_1@(\oton{r})}$.
+Finally, we have already shown $\valueconsist{v}{\hat T_2}$.
+Thus, $$\traceconsist{E' \dplus [ reduce_{id}(l,excl,v) ]}$$.
+
+To show $$\privconsist{E' \dplus [ reduce_{id}(l,excl,v) ]}{\Phi}$$, we first observe that
+$valid\_interleave(S,C,E',E_1,E_2)$ and Lemma~\ref{lemma:effsound:effects2} imply
+$\privconsist{E'}{\Phi}$.  To show $\privconsist{[ reduce_{id}(l,excl,v) ]}{\Phi}$, we observe that because
+$\valueconsist{l}{T@(\oton{r})}$, there is some $r_i$ such that $l \in M(r_i)$.  Furthermore,
+from [T-Reduce] we have $reduce_{id}(r_i) \in \Phi^*$, which shows $\privconsist{[ reduce_{id}(l,excl,v)]}{\Phi^*}$.
+The result follows from Lemma~\ref{lemma:phiclosure}.
+
+\item Case [E-Pack].
+By assumption, $$\opsenv{\text{pack } e_1 \text{ as } T_1[\otok{k}{r}]}{v'}{E}$$ and
+$$\typeenv{\text{pack } e_1 \text{ as } T_1[\otok{k}{r}]}{T_1}$$ and
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and 
+$\storeconsist{S}{H}$.
+
+From [E-Pack] we have $\opsenv{e_1}{v}{E}$ and
+from [T-Pack] we know $\typeenv{e_1}{T_2[r_1/r'_1,\ldots,r_k/r'_k]}$.
+By induction, we can conclude that $\valueconsist{v}{T_2[r_1/r'_1,\ldots,r_k/r'_k]}$,
+$\traceconsist{E}$ and $\privconsist{E}{\Phi}$.
+
+The only thing remaining to show is $\valueconsist{v'}{T_1}$.
+From [E-Pack] we have   $v' = \langle \langle \otok{k}{\rho}, v \rangle \rangle$.
+From [T-Pack] we have $T_1 = \exists r'_1, \ldots r'_n.\ T_2\text{ where }\Omega_1$
+and $\Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^*$.  
+%
+% FIXME  The next sentence can't be made fully formal with the current setup.
+%
+The result
+follows from the definition of value consistency on region relationships.
+
+\item Case [E-Unpack].
+By assumption, $$\opsenv{\text{unpack } e_1 \text{ as } id : T_1[\otok{k}{r}] \text{ in } e_2}{v_2}{E'}$$ and
+$$\typeenv{\text{unpack } e_1 \text{ as } id : T_1[\otok{k}{r}] \text{ in } e_2}{T}$$ and
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and 
+$\storeconsist{S}{H}$.
+
+For subexpression $e_1$, from [E-UnPack] we have $$\opsenv{e_1}{\langle \langle \otok{k}{\rho},v \rangle\rangle}{E}$$ and
+from [T-UnPack] we know $$\typeenv{e_1}{T_1 = \exists r'_1,\ldots,r'_n.T_2 \text{ where }\Omega_1}$$
+By induction, we can conclude that $\valueconsist{\langle \langle \otok{k}{\rho},v \rangle\rangle}{T_1}$ and
+$\traceconsist{E}$ and $\privconsist{E}{\Phi}$.
+
+For subexpression $e_2$, from [E-UnPack] we have
+$$\opsenv[M=M',L=L',S=S']{e_2}{v_2}{E_2}$$ where
+$S' = \text{apply}(S,E_2)$, $L' = L[v_1/id]$ and $M' = M[\rho_1/r_1,\ldots,\rho_k/r_k]$.
+From [T-Unpack] we have
+$$\typeenv[G=\Gamma',O=\Omega']{e_2}{T_3}$$
+where
+$\Gamma' = \Gamma[T_2[r_1/r'_1,\ldots,r_k/r'_k] / id]$ and
+$\Omega' = \Omega \cup \Omega_1[r_1/r'_1,\ldots,r_k/r'_k]$.
+To establish the induction hypothesis for $e_2$, we must show
+$\mapconsist[M']{\Omega'}$ and
+$\localconsist[M']{L'}{\Gamma'}$ and
+$\storeconsist{S'}{H}$.
+
+To show $\mapconsist[M']{\Omega'}$, consider a constraint $r \leq r' \in \Omega'$.  There are two cases. If $r \leq r' \in \Omega_1[r_1/r'_1,\ldots,r_k/r'_k]$
+then $M'(r) \subseteq M'(r')$ follows from $\valueconsist{\langle \langle \otok{k}{\rho},v \rangle\rangle}{T_1}$.
+If $r \leq r' \not\in \Omega_1[r_1/r'_1,\ldots,r_k/r'_k]$ then $r \leq r' \in \Omega$ and $M'(r) = M(r) \subseteq M(r') = M'(r')$.
+The reasoning for constraints $r * r'$ is the same.
+
+To show $\localconsist[M']{L'}{\Gamma'}$, we must show
+$\valueconsist[M']{L'(x)}{\Gamma'(x)}$ for all identifiers $x$ in
+the domain of $\Gamma'$.  For any $id' \neq id$, the statement is equivalent
+to showing $\valueconsist[M']{L(x)}{\Gamma(x)}$, which must hold becuase $M'$ does 
+does not change any existing binding in $M$.  For $id$, we have
+$\valueconsist[M']{v_1}{T_1[r_1/r'_1,\ldots,r_k/r'_k]}$, which is implied by
+$\valueconsist{\langle \langle \otok{k}{\rho},v \rangle\rangle}{T_1}$.
+
+Finally, as in previous cases, $\storeconsist{S'}{H}$ follows from Lemma~\ref{lemma:heapconst:apply}.
+Thus, we can conclude that 
+$\valueconsist[M']{v_2}{T_3}$ and
+$\traceconsist{E_2}$ and 
+$\privconsist{E_2}{\Phi}$.
+
+Because $T_3$ cannot include any of the unpacked regions in it, $\valueconsist{v_2}{T_3}$.
+From [E-Unpack] we have $$valid\_interleave(S, C, E', E_1, E_2)$$ and using Lemma~\ref{lemma:heapconst:effects2} we
+prove $\traceconsist{E'}$.
+To finish the case, we note that from $\privconsist{E_1}{\Phi}$, $\privconsist{E_2}{\Phi}$, 
+$valid\_interleave(S, C, E', E_1, E_2)$, and using Lemma~\ref{lemma:effsound:effects2} we can conclude $\privconsist{E'}{\Phi}$.
+
+\item Case [E-Part].
+By assumption,
+$$\opsenv{\text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2}{v}{E'}$$
+and
+$$\typeenv{\text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2}{T}$$
+and
+$\mapconsist{\Omega}$ and 
+$\localconsist{L}{\Gamma}$ and 
+$\storeconsist{S}{H}$.
+
+For subexpression $e_1$, from [E-Part] we have $$\opsenv{e_1}{K}{E_1}$$ and
+from [T-Part] we have $$\typeenv{e_1}{\text{coloring}(r_p)}$$
+By induction, we conclude that $\valueconsist{K}{\text{coloring}(r_p)}$ and
+$\traceconsist{E_1}$ and 
+$\privconsist{E_1}{\Phi}$.
+For subexpression $e_2$, from [E-Part] we have
+$$\opsenv[M=M',S=S']{e_2}{v}{E_2}$$
+where 
+$M' = M[\rho_1/r_1, \ldots, \rho_k/r_k]$ and
+$S' = \text{apply}(S, E_1)$.
+From [T-Part] we have
+$$\typeenv[O=\Omega']{e_2}{T}$$ 
+where
+$\Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{1 \leq i < j \leq k} r_i * r_j$.    
+To establish the induction hypothesis for $e_2$, we must show
+$\mapconsist[M']{\Omega'}$ and
+$\localconsist[M']{L}{\Gamma}$ and
+$\storeconsist{S'}{H}$.
+
+For $\mapconsist[M']{\Omega'}$, we observe that all of the constraints are either present previously
+(hold in $\mapconsist{\Omega}$) or are established explicitly by the partition operation.  Note that
+the type of the coloring carries the name of the region being partitioned and that all introduced region names are
+not in use in $M$.  For $\localconsist[M']{L}{\Gamma}$, we simply observe that $\localconsist{L}{\Gamma}$ holds
+and $M'$ is an extension of $M$.  
+Finally, as in previous cases, $\storeconsist{S'}{H}$ follows from Lemma~\ref{lemma:heapconst:apply}.
+
+Therefore, we can conclude by induction
+$\valueconsist[M']{v}{T}$ and
+$\traceconsist{E_2}$ and 
+$\privconsist{E_2}{\Phi}$.
+Because $T$ may not mention any of the regions introduced in the partitioning, we have
+$\valueconsist{v}{T}$.
+From [E-Part] we have $$valid\_interleave(S, C, E', E_1, E_2)$$ and using Lemma~\ref{lemma:heapconst:effects2} we
+prove $\traceconsist{E'}$.
+To finish the case, we note that from $\privconsist{E_1}{\Phi}$, $\privconsist{E_2}{\Phi}$, 
+$valid\_interleave(S, C, E', E_1, E_2)$, and using Lemma~\ref{lemma:effsound:effects2} we can conclude $\privconsist{E'}{\Phi}$.
+
+\item Case [E-Call].
+From [E-Call] we have
+$$\opsenv{id[\otok{k}{r}](\oton{e})}{v_{n+1}}{E''}$$
+and from [T-Call] we have
+$$\typeenv{id[\otok{k}{r}](\oton{e})}{T_r[r_1/r'_1,\ldots,r_k/r'_k]}$$
+
+For subexpression $e_1$, from [E-Call] we have $\opsenv{e_1}{v_1}{E_1}$ and from [T-Call] we have
+$\typeenv{e_1}{T_1[r_1/r'_1,\ldots,r_k/r'_k]}$.
+By induction, 
+$\valueconsist{v_1}{T_1[r_1/r'_1,\ldots,r_k/r'_k]}$,
+$\traceconsist{E_1}$ and 
+$\privconsist{E_1}{\Phi}$.
+
+For the other subexpressions that are arguments to the call the reasoning is very similar; we show
+only the last argument $e_n$.  First, 
+$$S_n = \text{apply}(E_{n-1},\text{apply}(E_{n-2},\ldots \text{apply}(E_1,S) \ldots )$$
+From Lemma~\ref{lemma:heapconst:apply} it follows that $\storeconsist{S_n}{H}$.
+We also have $\opsenv{e_n}{v_n}{E_n}$ from [E-Call] and
+$\typeenv{e_n}{T_n[r_1/r'_1,\ldots,r_k/r'_k]}$ from [T-Call].
+By induction, 
+$\valueconsist{v_n}{T_n[r_1/r'_1,\ldots,r_k/r'_k]}$,
+$\traceconsist{E_n}$ and 
+$\privconsist{E_n}{\Phi}$.
+
+Now, from [E-Call] we have
+$valid\_interleave(S, C, E', \oton{E})$
+and furthermore
+$M' = \{ (r'_1, M(r_1)), \ldots (r'_k, M(r_k)) \}$  and
+$L' = \{ (a_1, v_1), \ldots, (a_n, v_n) \}$ and
+$S' = \text{apply}(S, E')$ and
+$\opsenv[M=M',L=L',S=S',C=C']{e_{n+1}}{v_{n+1}}{E_{n+1}}$.
+From [T-Program] we know
+$\typeenv[G=\Gamma',O=\emptyset,P=\Phi']{e_{n+1}}{T_r}$,
+where $e_{n+1}$ is the body of the function.
+To apply the induction hypothesis to $e_{n+1}$ we must
+show 
+\begin{itemize}
+\item $\mapconsist[M']{\emptyset}$,
+which is immediate.
+
+\item $\localconsist[M']{L'}{\Gamma'}$ where
+$\Gamma'(a_i) = T_i$and $\storeconsist{S'}{H}$.
+\end{itemize}
+
+$\Gamma(id) = \forall r'_1, \ldots r'_k.(\oton{T}),\Phi', Q' \rightarrow T_r$ 
+$\text{function }id[\otok{k}{r'}](a_1 : T_1, \ldots, a_n : T_n), \Phi', Q' : T_r = e_{n+1}$
+
+$E'_{n+1} = mark\_coherence(E_{n+1}, M' \llbracket Q' \rrbracket)$ 
+$valid\_interleave(S, C, E'', E', E_{n+1})$
+
+$\typeenv{e_i}{T_i[r_1/r'_1,\ldots,r_k/r'_k]}$
+$\Phi'[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*$
+
+\end{itemize}
+\end{proof}
 

### 2012-07-16 tools/lock_analyzer.py
Adding a script for analyzing lock trace dumps

@@ -1,0 +1,170 @@
+#!/usr/bin/python
+
+import struct
+import subprocess
+import sys, os, shutil
+import string, re
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.backends.backend_pdf import PdfPages
+from getopt import getopt
+
+tableau1 = (0.968,0.714,0.824)
+tableau2 = (0.882,0.478,0.470)
+tableau3 = (0.565,0.663,0.792)
+tableau4 = (0.635,0.635,0.635)
+tableau5 = (0.678,0.545,0.788)
+tableau6 = (1.000,0.620,0.290)
+tableau7 = (0.859,0.859,0.553)
+tableau8 = (0.769,0.612,0.580)
+tableau9 = (0.478,0.757,0.424)
+tableau10= (0.427,0.800,0.855)
+tableau11= (0.929,0.592,0.792)
+tableau12= (0.929,0.400,0.364)
+tableau13= (0.447,0.620,0.808)
+tableau14= (0.780,0.780,0.780)
+tableau15= (0.773,0.690,0.835)
+tableau16= (0.882,0.616,0.353)
+tableau17= (0.804,0.800,0.365)
+tableau18= (0.659,0.471,0.431)
+tableau18= (0.404,0.749,0.361)
+tableau19= (0.137,0.122,0.125)
+
+
+class LockItem(object):
+    def __init__(self,tupple):
+        assert len(tupple) == 6
+        self.time = tupple[0]
+        self.idy = tupple[3]
+        self.owner = tupple[4]
+        self.location = tupple[1]
+        self.action = tupple[5]
+
+class Lock(object):
+    def __init__(self,item):
+        self.idy = item.idy
+        self.requests = list()
+        self.forwards = list()
+        self.grants = list()
+        self.releases = list()
+
+    def add_item(self,item):
+        if item.action == 0 or item.action == 1:
+            self.requests.append(item)
+        elif item.action == 2:
+            self.forwards.append(item)
+        elif item.action == 3 or item.action == 4:
+            self.grants.append(item)
+        elif item.action == 5:
+            self.releases.append(item)
+        else:
+            print "Illegal lock action code "+str(item.action)
+            assert False
+
+    def get_remote_requests(self):
+        result = 0
+        for r in self.requests:
+            if r.owner <> r.location:
+                result = result + 1
+        return result
+
+    def get_total_messages(self):
+        result = self.get_remote_requests()
[…]

### 2012-07-16 doc/popl2013full/langdef.tex
proof

@@
 \item Case [E-Read].
-By assumption, $\opsenv{\text{read}(e)}{v}{E \dplus [ read(l,excl,v) ]}$ and
-$\typeenv{\text{read}(e)}{T}$ and
+By assumption, $$\opsenv{\text{read}(e)}{v}{E \dplus [ read(l,excl,v) ]}$$ and
+$$\typeenv{\text{read}(e)}{T}$$ and
 $\mapconsist{\Omega}$ and 

### 2012-07-17 doc/asplos2013/interface.tex
Getting the asplos paper off the ground

\section{Programming Interface}
\label{sec:interface}

### 2012-07-17 doc/asplos2013/intro.tex
Getting the asplos paper off the ground

\section{Introduction}
\label{sec:intro}
Introduction goes here\cite{Keeneland}.

### 2012-07-17 doc/asplos2013/micro.tex
Getting the asplos paper off the ground

\section{Microbenchmark Evaluation}
\label{sec:micro}

### 2012-07-17 doc/asplos2013/results.tex
Getting the asplos paper off the ground

\section{Application Evaluation}
\label{sec:apps}

### 2012-07-18 doc/asplos2013/interface.tex
Working on the interface section

@@
 \label{sec:interface}
+
+The Legion Low-Level Runtime (LLR) interface organizes
+functionality into objects.  Objects are grouped into
+four different categories: machine objects for describing
+the underlying architecture, events for capturing dependences,
+deferred locks for performing synchronization, and physical regions
+for managing data.  In all cases, the objects are light-weight
+{\em handles} that can be passed around by value.  Furthermore
+every handle is valid everywhere in the system.  This property allows 
+a programmer to pass handles by value between computations without
+having to reason about the distributed nature of the system.
+We describe how we support this property in more detail in 
+Section~\ref{sec:impl}.
+
+\lstset{
+  captionpos=b,
+  language=C++,
+  basicstyle=\scriptsize,
+  numbers=left,
+  numberstyle=\tiny,
+  columns=fullflexible,
+  stepnumber=1,
+  escapechar=\#,
+  keepspaces=true,
+  %literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
+  %morekeywords={region,coloring,partition,spawn,disjoint,aliased},
+  %deletekeywords=float,
+}
+\subsection{Machine Interface}
+\label{subsec:machmodel}
+Listing~\ref{lst:machineapi} shows the machine interface.  The machine interface 
+contains three different object types: {\tt Processor} (line 1), {\tt Memory} (line 17), and a 
+singleton {\tt Machine} object (line 22).  For every processor (i.e. CPU core or GPU) in 
+the system there exists a unique {\tt Processor} handle naming that processor.  
+{\tt Processor} objects support a single {\tt spawn} operation (line 13-14) that will
+launch a {\em task} on that processor.  Note that because the {\tt spawn} operation
+is invoked on a processor handle, it is possible to launch a task on any
+processor from anywhere in the system.  The {\tt spawn} call takes as arguments
+the {\tt TaskFuncID} of the task to be run, a pointer to and size of arguments to
+be passed to the task, and an optional {\tt Event} (described in Section~\ref{subsec:events})
+that must trigger before the task can begin.  The result of the {\tt spawn} call
+is an {\tt Event} that will trigger when the task completes.
+
+For every memory in the system there also exists a unique {\tt Memory} handle. 
+Different memories often, but not always, imply a different address space.  For
+example, in a large cluster each node would have a different {\tt Memory} handle
+corresponding to each node's memory.
+However, if the cluster supported multiple GPUs then there would be a different
+{\tt Memory} handle for each GPU's {\em zero-copy} memory which shares part of the 
+address space with the corresponding node memory.  Memories will be used for describing
+the placement of data described in Section~\ref{subsec:phyreg}.
+
+The {\tt Machine} object is a singleton object that serves two purposes.  First, it
+is the mechanism for initializing the runtime; at the start of a program the programmer
+creates a {\tt Machine} object and invokes the {\tt run} method (line 30-31).
+Second, the {\tt Machine} object provides an interface for the programmer to 
+discover properties of the underlying machine (lines 35-43).  The {\tt Machine} object maintains
+sets of all the {\tt Processor} and all the {\tt Memory} handles in the system.
+The {\tt Machine} also maintains two relations between these sets:
+
+\begin{itemize}
+\item Processor-Memory: for each {\tt Processor} and each {\tt Memory} whether
+the {\tt Processor} can directly access (read and write) the {\tt Memory} and 
+latency and bandwidth properties if access is possible
+\item Memory-Memory: for every pair of {\tt Memory} whether a copy can be
+performed between the two memories and latency 
+bandwidth properties if a copy is possible
+\end{itemize}
+
+The programmer can query these two relations by the {\tt get\_*\_affinity} calls
+(lines 40-43).  These calls populate a vector with affinity structures (not shown)
+that contain latency and bandwidth information.  The programmer can restrict the query
+by specifying a specific processor or memory, otherwise information is returned
+about all processors and memories in the system.
+
+\begin{lstlisting}[float={t},label={lst:machineapi},caption={Machine Interface.}]
+class Processor {
+public:
+  const unsigned id;
+  typedef unsigned TaskFuncID;
+  typedef void (*TaskFuncPtr)
+            (const void *args,size_t arglen,Processor p);
+  typedef map<TaskFuncID, TaskFuncPtr> TaskIDTable;
+
+  enum Kind {
+    CPU_PROC,GPU_PROC,// ... future processor types
+  };
+
+  Event spawn(TaskFuncID func_id,const void *args,size_t arglen,
+              Event wait_on = Event::NO_EVENT) const;
+};
+
+class Memory {
+public:
+  const unsigned id;
+};
+
+class Machine {
+public:
+  Machine(int *argc, char ***argv,
+          const Processor::TaskIDTable &task_table);
+  enum RunStyle {
+    ONE_TASK_ONLY,ONE_TASK_PER_NODE,
+    ONE_TASK_PER_PROCESSOR,
+  };
+  void run(Processor::TaskFuncID task_id = 0, 
+        RunStyle = ONE_TASK_ONLY, const void *args, size_t arglen);
+public:
+  static Machine* get_machine(void); // pointer to Machine
+  // References to all processors and memories
+  const set<Memory>& get_all_memories(void) const;
+  const set<Processor>& get_all_processors(void) const;
+  Processor::Kind get_processor_kind(Processor p) const;
+  size_t get_memory_size(Memory m) const;
+  // Query properties of underlying hardware
+  int get_proc_mem_affinity(vector<ProcMemAffinity &result,
+          Processor restrict_proc = 0, Memory restrict_mem = 0);
+  int get_mem_mem_affinity(vector<MemMemAffinity> &result,
+          Memory restrict_mem1 = 0, Memory restrict_mem2 = 0);
+};
+\end{lstlisting}
+
+\subsection{Events}
+\label{subsec:events}
+Listing~\ref{lst:eventapi} shows the interface for events.  
+{\tt Event} is the base type of events.  An instance of an {\tt Event} type 
+names a unique event in the system.  {\tt NO\_EVENT} (line 5) is a special instance
+of an event that by definition has always triggered.  The {\tt Event} interface
+also supports testing whether an event has triggered (line 7) and waiting on
+an event to trigger (line 9).  While these methods can be useful, the most common
+use of events is passing them as preconditions for operations such as the {\tt spawn}
+call described in Section~\ref{subsec:machmodel}.  The programmer can use the
+{\tt merge\_events} call (line 11) to create an event corresponding to the 
+conjunction of a set of events when multiple events are a precondition for an operation.
+
+In most cases events are created by the runtime as the result of other
+operations and the runtime is responsible for triggering these events.  Users 
+can create a special type of event called a {\tt UserEvent} (line 14) that the
+user has the power to trigger.  {\tt UserEvent} is a sub-type of {\tt Event} 
+that has {\tt trigger} method (line 17).  Users can also create another
+type of event called a {\tt Barrier} that will require multiple arrivals before
+the event is triggered.  The user can manage the number of arrivals that must
+be seen as well as how many arrivals occur at a time.  Note that since {\tt UserEvent}
+and {\tt Barrier} are sub-types of {\tt Event} they can be used wherever an
+{\tt Event} is required.
+
+\begin{lstlisting}[float={t},label={lst:eventapi},caption={Event Interface.}]
+class Event {
+public:
+  const unsigned id;
+  const unsigned gen;
+  static const Event NO_EVENT;
+  // check if an event has triggered
+  bool has_triggered() const;
+  // wait on the event
+  void wait() const;
+  // Merge events together to a new event
+  static Event merge_events(const set<Event> &to_merge);
+};
+
+class UserEvent : public Event {
+public:
+  static UserEvent create_user_event();
+  void trigger() const;
+};
+
+class Barrier : public Event {
+public:
+  static Barrier create_barrier(unsigned expected_arrivals);
+  void alter_arrival_count(int delta) const;
+  void arrive(unsigned count = 1) const;
+};
+\end{lstlisting}
+
+\subsection{Deferred Locks}
+\label{subsec:locks}
+Deferred locks are a new synchronization mechanism that has different semantics from
+blocking locks.  From here on we refer to locks and deferred locks interchangeably, 
+but will refer to blocking locks explicitly.  Listing~\ref{lst:lockapi} shows the 
+interface for locks.  Unlike blocking locks, the {\tt lock} method (line 4) doesn't
+block but instead always returns immediately with an event that will be triggered
+when the lock has been acquired.  If the lock can be acquired immediately then a {\tt NO\_EVENT}
+is returned.  The {\tt lock} method also takes an optional event
+parameter indicating that the lock acquire shouldn't be performed until the event
+has triggered.  Similarly, the {\tt unlock} method (line 5) takes an optional event parameter indicating
+an event to wait for before performing the unlock operation.
+
+Deferred locks provide a super-set of the functionality of blocking locks.  As their
+name suggests, deferred locks can use the event corresponding to their lock acquire
+operation to defer execution until the lock acquire has been granted.  Deferred
+locks can also be converted back into a blocking lock by immediately waiting on the event
+returned from a call to {\tt lock}.  In addition, deffered locks also provide {\tt mode} and
+and {\tt exclusive} parameters that allow the user to specify whether other requests can acquire
+the lock simultaneously.  A lock can only be acquired in one mode at a time.  The
+{\tt exclusive} parameter specifies whether other owners are permitted once the
+current lock request is granted.
+
+An important difference between deferred locks and blocking locks is that the processor
+that requests the lock doesn't have to be the one that uses it.  A common
+convention in writing code with deferred locks is to acquire locks for a task being
+launched.  Listing~\ref{lst:lockapi} illustrates this with a simple example.  The
+{\tt launcher\_task} is calling a sub-task that is going to require the {\tt needed}
+lock.  A lock request is issued and the resulting {\tt lock\_event} is used as
+a precondition for launching the {\tt SUB\_TASK} task.  The unlock operation is then
+precondition on the event corresponding to {\tt SUB\_TASK} task's completion.  The
+sub-task can run safely while holding the lock and the lock is released when the task
+sub-task completes.  Using deferred locks in this style minimizes the time
+compute resources spend waiting on locks by ensuring tasks only run when their needed
+locks have already been acquired.
+
+\begin{lstlisting}[float={t},label={lst:lockapi},caption={Deferred Lock Interface and Example.}]
+class Lock {
+public:
+  const unsigned id;
+  Event lock(unsigned mode = 0, bool exclusive, 
+              Event wait_on = Event::NO_EVENT) const;
+  void unlock(Event wait_on = Event::NO_EVENT) const;
+  // Create a new lock, destroy an existing lock
+  static Lock create_lock();
+  void destroy_lock();
+};
+
+void launcher_task(const void *args, size_t arglen, Processor p) {
+  ...
+  // Unpack lock from arguments
+  Lock needed = ...
+  // Acquire lock
+  Event lock_event = needed.lock();
+  // Launch task
+  Event task_event = p.spawn(SUB_TASK,NULL,0,lock_event);
+  // Release lock
+  needed.unlock(task_event);
+  ...
+}
+\end{lstlisting}
+
+\subsection{Physical Regions}
+\label{subsec:phyreg}

### 2012-07-18 doc/asplos2013/interface.tex
More work on the interface section

@@
 \label{subsec:phyreg}
+
+The LLR interface uses {\em physical regions} as an abstraction for reasoning about
+the placement and movement of data in the memory hierarchy.  Physical regions imply a specific
+allocation of data in a single memory in the memory hierarchy.  Listing~\ref{lst:regionapi} 
+shows a subset of the physical region interface.  The first object in the interface
+is a {\tt RegionMetaData} object (line 1).  {\tt RegionMetaData} objects 
+do not actually allocate data, but instead provide an interface for the
+creation of physical regions called {\tt RegionInstance} objects (line 17).  {\tt RegionMetaData}
+objects are crucial for the movement of data by the runtime.  The runtime is only able
+to copy data between two {\tt RegionInstance} objects that were created by the
+same {\tt RegionMetaData} object.  A runtime error will be generated if a copy is attempted
+between two {\tt RegionInstance} objects that were not created by the same {\tt RegionMetaData}.
+
+When creating a {\tt RegionInstance} object the programmer must specify the {\tt Memory}
+in which the instance is going to be allocated (line 10).  Once the {\tt RegionInstance} is created it
+cannot be moved.   There is no coherence of data between {\tt RegionInstance} objects of the
+same {\tt RegionMetaData}.  It is the programmer's responsiblity to manage data coherence
+using copies between {\tt RegionInstance} objects (line 22-23).  Just like other operations
+copies can be predicated on an {\tt Event} and return an {\tt Event} corresponding to
+completion.  A copy between two {\tt RegionInstance} objects can only be performed if copies
+are permitted between their two memories.  A test of whether copies can be performed can be performed 
+by a query of the {\tt Machine} object (see Section~\ref{subsec:machmodel}).
+
+\begin{lstlisting}[float={t},label={lst:regionapi},caption={Subset of Physical Region Interface.}]
+class RegionMetaData {
+public:
+  const unsigned id;
+  static const RegionMetaData NO_REGION;
+  // Create and destroy metadatas
+  static RegionMetaData create_region(size_t num_elmts, 
+                                      size_t elmt_size);
+  void destroy_region() const;
+  // Create and destroy instances
+  RegionInstance create_instance(Memory memory) const;
+  void destroy_instance(RegionInstance instance) const;
+  // Create a physical instance only for reductions
+  RegionInstance create_instance(Memory memory, 
+                        ReductionOpID redopid) const;
+};
+
+class RegionInstance {
+public:
+  const unsigned id;
+  static const RegionInstance NO_INST;
+  // Copy between instances
+  Event copy_to(RegionInstance target, 
+                Event wait_on = Event::NO_EVENT);
+};
+\end{lstlisting}

### 2012-07-18 doc/asplos2013/impl.tex
More writing for the asplos paper

\section{Runtime Implementation}
\label{sec:impl}

There are currently two implementations of our interface:
one that works only on shared memory machines and another that
runs on larges clusters with both CPUs and GPUs.  The shared-memory
implementation uses the POSIX threads library.
The implementation for heterogeneous clusters also uses POSIX
threads as well as the CUDA runtime library for GPUs\cite{CUDA} and the GASNet
API for large clusters\cite{GASNET07}.

The heterogeneous implementation
models a cluster with CPUs and GPUs as having two types of processors 
and four types of memory.  Every CPU core is a different CPU processor
and every GPU is a GPU processor.  For every node in the system
there is a memory corresponding to the DRAM associated with that node.  For
every GPU in the system there are two memories corresponding to the framebuffer
memory and the zero-copy memory associated with that GPU \footnote{Zero-copy 
memory contains pages that are mapped in both framebuffer and node memory with
coherence maintained by the PCI-E protocol.}. The last memory is 
a global GASNet memory that represents pages that have been registered 
with all nodes in the system for supporting remote memory accesses (RMAs).
The GASNet memory provides the illusion of global memory.  CPU processors
can directly access their node memory, all zero-copy memories for GPUs on
their node, and the GASNet memory.  GPU processors can only access their
framebuffer memory and their zero-copy memory.  Copies are permitted
between all pairs of memory except between any framebuffer memory and
the GASNet memory.

To support inter-node operations our implementation relies on GASNet active
messages for communication.  Active messages are a command and a payload
that are sent remotely and cause a handler to be run on the target node.
A simple example of our use of active messages can be illustrated by the 
{\tt spawn} call on a {\tt Processor} object.  Consider that {\tt spawn}
is invoked on a processor that is on a remote node from the node on which
the currently running task is executing.  This operation is converted into
an active message that contains all the information about the task launch.
When the active message is handled by the remote node, the task and all its
metadata are registered with the processor on which the task was launched.
While most of the implementation of our heterogeneous runtime is straight
forward, the next three sections describe in greater detail the nuances
of the implementation of events, locks, and reduction instances.

\subsection{Event Implementation}
\label{subsec:eventimpl}

\subsection{Lock Implementation}
\label{subsec:lockimpl}

\subsection{Reduction Instances}
\label{subsec:reducimpl}

### 2012-07-18 doc/asplos2013/interface.tex
More writing for the asplos paper

--- 2012-07-18 doc/asplos2013/interface.tex
@@
 
-The Legion Low-Level Runtime (LLR) interface organizes
+The runtime interface organizes
 functionality into objects.  Objects are grouped into
@@
 for managing data.  In all cases, the objects are light-weight
-{\em handles} that can be passed around by value.  Furthermore
+{\em handles} that can be copied and passed around by value.  Furthermore
 every handle is valid everywhere in the system.  This property allows 
@@
 singleton {\tt Machine} object (line 22).  For every processor (i.e. CPU core or GPU) in 
-the system there exists a unique {\tt Processor} handle naming that processor.  
+the system there exists a {\tt Processor} handle with a unique ID naming that processor.  
 {\tt Processor} objects support a single {\tt spawn} operation (line 13-14) that will
@@
 
-For every memory in the system there also exists a unique {\tt Memory} handle. 
+For every memory in the system there also exists a {\tt Memory} handle with a unique ID. 
 Different memories often, but not always, imply a different address space.  For
 example, in a large cluster each node would have a different {\tt Memory} handle
-corresponding to each node's memory.
+corresponding to each node's DRAM memory.
 However, if the cluster supported multiple GPUs then there would be a different
@@
 discover properties of the underlying machine (lines 35-43).  The {\tt Machine} object maintains
-sets of all the {\tt Processor} and all the {\tt Memory} handles in the system.
+sets of all the unique {\tt Processor} and {\tt Memory} handles in the system.
 The {\tt Machine} also maintains two relations between these sets:
@@
 \begin{itemize}
-\item Processor-Memory: for each {\tt Processor} and each {\tt Memory} whether
+\item Processor-Memory: for each {\tt Processor} - {\tt Memory} pair whether
 the {\tt Processor} can directly access (read and write) the {\tt Memory} and 
@@
 \item Memory-Memory: for every pair of {\tt Memory} whether a copy can be
-performed between the two memories and latency 
+performed between the two memories and latency and
 bandwidth properties if a copy is possible
@@
 the event is triggered.  The user can manage the number of arrivals that must
-be seen as well as how many arrivals occur at a time.  Note that since {\tt UserEvent}
+be seen as well as how many arrivals occur at a time.  Once a barrier's arrival
+count goes to zero, the barrier will trigger.  Note that since {\tt UserEvent}
 and {\tt Barrier} are sub-types of {\tt Event} they can be used wherever an
@@
 \label{subsec:phyreg}
+
+{\em Physical regions} are the mechanism for reasoning about
+the placement and movement of data in the runtime.  Physical regions are an 
+allocation of data in a single memory in the memory hierarchy.  Listing~\ref{lst:regionapi} 
+shows a subset of the physical region interface.  The first object in the interface
+is a {\tt RegionMetaData} object (line 1).  {\tt RegionMetaData} objects provide the interface for the
+creation of sets of physical regions that possess the same number and type of elements (but not necessarily
+the same data layout).  The runtime is only able
+to copy data between two physical regions that were created by the
+same {\tt RegionMetaData} object.  A runtime error will be generated if a copy is attempted
+between two physical regions that were not created by the same {\tt RegionMetaData}.
+
+An instance of a physical region is represented by a {\tt RegionInstance} object (line 17).
+When creating a physical region the programmer must specify the {\tt Memory}
+in which the physical region is going to be allocated (line 10).  Once the physical region is created it
+cannot be moved.   There is no coherence of data between physical regions created by the 
+same {\tt RegionMetaData}.  It is the programmer's responsiblity to manage data coherence
+using copies between physical regions (line 22-23).  Just like other operations,
+copies can be predicated on an {\tt Event} and return an {\tt Event} corresponding to
+completion.  A copy between two physical regions can only be performed if copies
+are permitted between the two memories where the regions were created.  
+The legality of copies can be discovered using queries to the {\tt Machine} object
+described in Section~\ref{subsec:machmodel}.
+
+The data held inside a physical region is accessed by another kind of object called an
+{\em instance accessor} (not shown).  Accessors support read, write, and reduction operations
+to elements inside the physical region.  Accessors can be specialized for the case where
+all memory is known to be directly accessible to the executing processor, which allows 
+for direct references to elements.  In cases where not all memory is directly accessable, 
+accessors supply the level of indirection necessary to convert operations into remote 
+memory operations (RMAs).  %An example of this is described in more detail in Section~\ref{sec:impl}.
+
+One special (but common) case for applications is when reductions need to be performed.
+For reductions it is usually more efficient to store reduction
+operations in a local reduction buffer and then merge reduction buffers together at a later
+point in time.  To support this case the interface allows for the creation of a special kind
+of physical instance called a {\em reduction instance}.  The call to {\tt create\_instance} on
+lines 13-14 takes a {\em reduction operation} to be associated with a physical instance which
+tells the runtime to create a reduction instance.  Reduction instances can only be accessed
+using a reduction of the given reduction type; any other accesses will result in a runtime
+error.  A reduction operation is a functor which must have at least one method of the type
+$(T_1 \rightarrow T_2 \rightarrow T_1)$ which takes an element of type $T_1$ and a value
+to be reduced of type $T_2$ and creates and new element.  Reduction operations can also
+support two additional methods: a method that supplies an initial value of type $T_2$ and a 
+fold method of type $(T_2 \rightarrow T_2 \rightarrow T_2)$.  The presence of these methods
+enable additional runtime optimizations for reductions described in Section~\ref{sec:impl}.
+
+Reduction instances can be copied using the same interface as regular physical instances.
+It is only legal to copy a reduction instance to a regular physical instance or another
+reduction instance.  Any attempt to copy from a regular physical instance to a reduction
+instance will result in a runtime error.  A copy from a reduction instance to a regular
+physical instance will apply all the reductions in the reduction buffer to the physical
+instance.  If a copy is from one reduction instance to another, the two reduction buffers
+will be merged.  The semantics of a merge are dependent upon the underlying implementation
+of reduction buffers which is discussed in more detail in Section~\ref{sec:impl}.
+
+\begin{lstlisting}[float={t},label={lst:regionapi},caption={Subset of Physical Region Interface.}]
+class RegionMetaData {
+public:
+  const unsigned id;
+  static const RegionMetaData NO_REGION;
+  // Create and destroy metadatas
+  static RegionMetaData create_region(size_t num_elmts, 
+                                      size_t elmt_size);
+  void destroy_region() const;
+  // Create and destroy instances
+  RegionInstance create_instance(Memory memory) const;
+  void destroy_instance(RegionInstance instance) const;
+  // Create a physical instance only for reductions
+  RegionInstance create_instance(Memory memory, 
+                        ReductionOpID redopid) const;
+};
+
+class RegionInstance {
+public:
+  const unsigned id;
+  static const RegionInstance NO_INST;
+  // Copy between instances
+  Event copy_to(RegionInstance target, 
+                Event wait_on = Event::NO_EVENT);
+};
+\end{lstlisting}

### 2012-07-18 doc/asplos2013/impl.tex
Text for the event implementation section

@@
 A simple example of our use of active messages can be illustrated by the 
-{\tt spawn} call on a {\tt Processor} object.  Consider that {\tt spawn}
-is invoked on a processor that is on a remote node from the node on which
-the currently running task is executing.  This operation is converted into
+{\tt spawn} call on a {\tt Processor} object.  Consider the case where {\tt spawn}
+is invoked on a processor that is on a remote node.  This operation is converted into
 an active message that contains all the information about the task launch.
@@
 \label{subsec:eventimpl}
+Events are created on demand by each node.  The ID of the creating node is encoded
+in the upper bits of each event's ID and the creating node is said to be the 
+event's {\em owner}.  If an event is used as an argument to an
+operation on a remote node then the remote node can determine the owner of 
+the event and send an active message to become a {\em subscriber} of the event.  When a
+remote node becomes a subscriber to an event it guarantees that it will receive an active
+message from the event's owner when the event triggers.  To minimize the amount
+of communication that occurs between nodes, each node remembers the events to which it
+is subscribed and guarantees that it only subscribes to an event once.  In the case
+where there may be many remote waiters on an event 
+this can dramatically reduce the number of inter-node active messages.
+
+When an event triggers, it automatically notifies all of the operations that are waiting
+on its local node.  The owner node will also send active messages to all of the subscriber nodes
+telling them that the event has triggered.  Even though the event has now notified all of
+its waiters, the memory on the owning node for storing the event cannot be reclaimed 
+because there still may be handles to the event in the system which will request to
+use this event in the future.  These requests will need to be satisfied saying that 
+the event has already triggered for the duration of the application.  Rather than waste
+the resources, we recycle event implementations.
+
+A {\em dynamic event} is a logical construct that will only be triggered once and is the level
+of abstraction at which the programmer reasons about events.  A {\em physical event} is an
+actual event implementation that consumes resources on its owner node.  Multiple dynamic events
+can be mapped to a single physical event.  However, for each physical event there can be at most
+one {\em active} event in the set of dynamic events that are mapped to it.  An active event is a 
+dynamic event that is yet to trigger.  By only having one active dyanmic event at a time 
+there is no need to disambiguate event triggers that are sent to the physical event.  Each
+physical event maintains a count of the number of trigger operations that it has observed.
+
+When the runtime needs to return a handle corresponding to a new dynamic event, it finds
+a physical event which currently has no active events (e.g. the trigger count equals the number
+of dynamic events mapped to the physical event).  The runtime increases the count
+of the number of dynamic events mapped to the physical event.  The handle that is returned 
+corresponds to a new dynamic event which contains an ID that refers to the physical
+event it is mapped to and a {\em generation} that is the dynamic event number for the particular
+physical event.  Note that by definition the generation will always be one greater than the
+trigger count at the time of dynamic event creation.
+
+Within this framework it is now very easy to test whether the dynamic event specified by a
+handle has triggered.  Using the ID that is contained within the handle we find the
+physical event implementation (which may require an active message if the owner node is
+remote from where the test is begin performed).  We can then compare the generation of
+the handle to the number of observed triggers contained by the physical event.  If the
+number of physical triggers is greater than or equal to the handle's generation, then
+the dynamic event has triggered.  We show in Section~\ref{sec:apps} that in practice
+the number of needed physical events is significantly less than the number of dynamic events.
+
+We also leverage the mapping of dynamic events onto physical events to improve the efficiency
+of subscriptions.  Nodes cache the last generation for every physical event for which they
+have seen a trigger notification.  If the trigger test described above detects a trigger based
+on the cached information then there would be no need for the subscription which would save an
+active message.  If the event has not been detected to have
+triggered locally then an active message is sent to the owner which would always have been necesary
+without the dynamic to physical event mapping.
 

### 2012-07-19 doc/popl2013full/langdef.tex
big proof

@@
 \end{itemize}
-Figure~\ref{fig:constprop} gives the precise definition of these properties.
+Figure~\ref{fig:constprop} gives the precise definition of these
+properties.\footnote{Note that in the definition of
+$\mapconsist{\Omega}$, the quantified variables $r_i$ and $r_j$
+range over both logical and concrete regions.  Since $M$'s domain is
+logical regions, we extend $M$ to constraints with mixed logical and
+concrete regions in the standard way by defining
+$M \llbracket \rho \rrbracket = \rho$.}
 
@@
 and $\Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^*$.
-%
-% FIXME  The next sentence can't be made fully formal with the current setup.
-%
-The result
-follows from the definition of value consistency on region relationships.
+
+Observe that 
+$$M \llbracket \exists r'_1, \ldots r'_n.\ T_2\text{ where }\Omega_1 \rrbracket$$
+replaces all free region names in $T_2$ and particularly $\Omega_1$ by concrete regions, so
+that the only region names remaining are $r'_1,\ldots,r'_n$.  It is then easy to show that
+$$\mapconsist[\{(r'_i,\rho_i)\}]{M \backslash \{r'_1,\ldots, r'_n \}\llbracket \Omega_1 \rrbracket}$$
+from which $\valueconsist{v'}{T_1}$ follows (see Figure~\ref{fig:constprop}).
 
@@
 To finish the case, we note that from $\privconsist{E_1}{\Phi}$, $\privconsist{E_2}{\Phi}$, 
-$valid\_interleave(S, C, E', E_1, E_2)$, and using Lemma~\ref{lemma:effsound:effects2} we can conclude $\privconsist{E'}{\Phi}$.
+$$valid\_interleave(S, C, E', E_1, E_2)$$, and using Lemma~\ref{lemma:effsound:effects2} we can conclude $\privconsist{E'}{\Phi}$.
 
@@
 
-For subexpression $e_1$, from [E-Call] we have $\opsenv{e_1}{v_1}{E_1}$ and from [T-Call] we have
-$\typeenv{e_1}{T_1[r_1/r'_1,\ldots,r_k/r'_k]}$.
+For subexpression $e_1$, from [E-Call] we have $$\opsenv{e_1}{v_1}{E_1}$$ and from [T-Call] we have
+$$\typeenv{e_1}{T_1[r_1/r'_1,\ldots,r_k/r'_k]}$$
 By induction, 
@@
 only the last argument $e_n$.  First, 
-$$S_n = \text{apply}(E_{n-1},\text{apply}(E_{n-2},\ldots \text{apply}(E_1,S) \ldots )$$
-From Lemma~\ref{lemma:heapconst:apply} it follows that $\storeconsist{S_n}{H}$.
-We also have $\opsenv{e_n}{v_n}{E_n}$ from [E-Call] and
-$\typeenv{e_n}{T_n[r_1/r'_1,\ldots,r_k/r'_k]}$ from [T-Call].
-By induction, 
+$$S_n
+= \text{apply}(E_{n-1},\text{apply}(E_{n-2},\ldots \text{apply}(E_1,S) \ldots
+)$$ An induction on the number of applications of {\em apply} using
+Lemma~\ref{lemma:heapconst:apply} shows that
+$\storeconsist{S_n}{H}$.  We also have $$\opsenv{e_n}{v_n}{E_n}$$ from
+[E-Call] and $$\typeenv{e_n}{T_n[r_1/r'_1,\ldots,r_k/r'_k]}$$ from
+[T-Call].  By induction,
 $\valueconsist{v_n}{T_n[r_1/r'_1,\ldots,r_k/r'_k]}$,
-$\traceconsist{E_n}$ and 
-$\privconsist{E_n}{\Phi}$.
+$\traceconsist{E_n}$ and $\privconsist{E_n}{\Phi}$.
 
 Now, from [E-Call] we have
-$valid\_interleave(S, C, E', \oton{E})$
+$$valid\_interleave(S, C, E', \oton{E})$$
 and furthermore
-$M' = \{ (r'_1, M(r_1)), \ldots (r'_k, M(r_k)) \}$  and
-$L' = \{ (a_1, v_1), \ldots, (a_n, v_n) \}$ and
-$S' = \text{apply}(S, E')$ and
-$\opsenv[M=M',L=L',S=S',C=C']{e_{n+1}}{v_{n+1}}{E_{n+1}}$.
+$$M' = \{ (r'_1, M(r_1)), \ldots (r'_k, M(r_k)) \}$$
+$$L' = \{ (a_1, v_1), \ldots, (a_n, v_n) \}$$
+$$S' = \text{apply}(S, E')$$
+We also know from [E-Call] that
+$$\opsenv[M=M',L=L',S=S',C=C']{e_{n+1}}{v_{n+1}}{E_{n+1}}$$
+where $e_{n+1}$ is the body of the called function.
 From [T-Program] we know
-$\typeenv[G=\Gamma',O=\emptyset,P=\Phi']{e_{n+1}}{T_r}$,
-where $e_{n+1}$ is the body of the function.
-To apply the induction hypothesis to $e_{n+1}$ we must
-show 
-\begin{itemize}
-\item $\mapconsist[M']{\emptyset}$,
-which is immediate.
-
-\item $\localconsist[M']{L'}{\Gamma'}$ where
-$\Gamma'(a_i) = T_i$and $\storeconsist{S'}{H}$.
-\end{itemize}
-
-$\Gamma(id) = \forall r'_1, \ldots r'_k.(\oton{T}),\Phi', Q' \rightarrow T_r$ 
-$\text{function }id[\otok{k}{r'}](a_1 : T_1, \ldots, a_n : T_n), \Phi', Q' : T_r = e_{n+1}$
-
-$E'_{n+1} = mark\_coherence(E_{n+1}, M' \llbracket Q' \rrbracket)$ 
-$valid\_interleave(S, C, E'', E', E_{n+1})$
-
-$\typeenv{e_i}{T_i[r_1/r'_1,\ldots,r_k/r'_k]}$
-$\Phi'[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*$
-
+$$\typeenv[G=\Gamma',O=\emptyset,P=\Phi']{e_{n+1}}{T_r}$$
+where $\Gamma'(a_i) = T_i$.
+To apply the induction hypothesis to $e_{n+1}$ we want to show
+\begin{enumerate}
+\item $\storeconsist{S'}{H}$, 
+\item $\mapconsist[M']{\emptyset}$, and
+\item $\localconsist[M']{L'}{\Gamma'}$ 
+\end{enumerate}
+
+For (1), using $\traceconsist{E_i}$ for $1 \leq i \leq n$ and applying Lemma~\ref{lemma:heapconst:effects2} $n - 1$ times
+we can conclude that $\traceconsist{E'}$.  Then because $\storeconsist{S}{H}$ and $S' = \text{apply}(S, E')$ by
+Lemma~\ref{lemma:heapconst:apply} we have $\storeconsist{S'}{H}$.
+
+Part (2) is immediate by the definition of mapping consistency.
+
+For (3), in previous steps we have shown  $$\valueconsist{v_i}{T_i[r_1/r'_1,\ldots,r_k/r'_k]}$$ for $1 \leq i \leq n$.
+Since $M \llbracket r_i \rrbracket = M' \llbracket r'_i \rrbracket$ it follows that
+$M \llbracket T_i[r_1/r'_1,\ldots,r_k/r'_k \rrbracket = M' \llbracket T_i \rrbracket$ and therefore
+$\valueconsist[M']{v_i}{T_i}$, from which the result follows.
+
+Thus, by induction, we conclude that
+$\valueconsist[M']{v_{n+1}}{T_r}$, $\traceconsist{E_{n+1}}$ (under
+$M'$) and $\privconsist[M']{E_{n+1}}{\Phi'}$.  Because 
+$$M \llbracket T_r[r_1/r'_1,\ldots,r_k/r'_k \rrbracket = M' \llbracket T_r \rrbracket$$ we can conclude
+$\valueconsist{v_{n+1}}{T_r[r_1/r'_1,\ldots,r_k/r'_k}$.
+
+To show that
+$\traceconsist{E_{n+1}}$ under $M'$ implies $\traceconsist{E_{n+1}}$
+under $M$, note that $H(\cdot)$ does not depend on the mapping (because it includes only concrete, not logical, regions) so the
+types of locations are unaffected by the mapping (see
+Figure~\ref{fig:constprop}).  Now, the environment
+$M' \llbracket \Gamma' \rrbracket$ is used to look up reduction
+functions, but the reduction function signatures are restricted to
+have no regions named in their arguments and so
+$$M \llbracket \Gamma' \rrbracket(id)  = M' \llbracket \Gamma' \rrbracket(id)$$
+Finally, from 
+$$valid\_interleave(S, C, E'', E', E_{n+1})$$
+and Lemma~\ref{lemma:heapconst:effects2} we conclude $\traceconsist{E''}$.
+
+For the last consistency property of the function call, we first observe
+that $\privconsist[M']{E_{n+1}}{\Phi'}$ is equivalent to 
+$$\privconsist{E_{n+1}}{\Phi[r_1/r'_1,\ldots,r_k/r'_k]}$$
+because, again, $E_{n+1}$ does not mention logical region names and
+$M' \llbracket r' \rrbracket = M \llbracket r \rrbracket$.
+From [T-Call] we have 
+$$\Phi[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*$$
+and so
+$\privconsist{E_{n+1}}{\Phi^*}$.  Applying Lemma~\ref{lemma:phiclosure} we conclude
+that $\privconsist{E_{n+1}}{\Phi}$.
+Finally, from Lemma~\ref{lemma:effsound:effects2} we conclude $\privconsist{E''}{\Phi}$.
 \end{itemize}
@@
 \begin{proof}
-The proof is by contradiction.
-Assume $E_1 \cancel\nonint E_2$.
-Then there must be some $\epsilon_1$ in $E_1$ and some $\epsilon_2$ in $E_2$ such that $\epsilon_1 \cancel\nonint \epsilon_2$.
-This requires that
+The proof is by contradiction.  Assume $E_1 \nonint E_2$.  Then there must be some $\epsilon_1$ in $E_1$ and some $\epsilon_2$ in $E_2$ such that $\epsilon_1 \nonint \epsilon_2$.  This requires that
 $\epsilon_1$ and $\epsilon_2$ be operations to the same location $l$ and not both be reads or both
@@
 Similarly, there must be some $r_2$ satisfying $l \in M(r_2) \wedge writes(r_2) \in \Phi_2$.
-Letting $\phi_1 = reads(r_1)$ and $\phi_2 = writes(r_2)$, we have $M(r_1) \cap M(r_2) \not= \emptyset$ (it contains at least $l$), so $\phi_1 \cancel{\nonint[M]} \phi_2$ and $\Phi_1 \cancel{\nonint[M]} \Phi_2$.
+Letting $\phi_1 = reads(r_1)$ and $\phi_2 = writes(r_2)$, we have $M(r_1) \cap M(r_2) \not= \emptyset$ (it contains at least $l$), so $\phi_1 {\nonint[M]} \phi_2$ and $\Phi_1 {\nonint[M]} \Phi_2$.
 There are seven other cases to consider (using $id_1 \not= id_2$ for the reduce-reduce case), but all yield the same result.

### 2012-07-19 doc/asplos2013/intro.tex
Working on the intro

@@
 \label{sec:intro}
-Introduction goes here\cite{Keeneland}.
+
+% Big machines -> latencies are growing
+% Processors can't be blocked, need to be able to continue handling work
+% Operations need to be composable
+% deferred execution model
+
+% Contributions:
+% - An event system where every operation can take an event and returns an event
+% - Handles are globally valid
+% - More flexible synchronization primitive for this environment: deferred locks
+% - Physical regions for data movement including special versions for reductions
+
+With the multicore revolution, modern supercomputers have continued to scale
+by increasing the number of processors available for computation.  The
+price of this scaling is complex interconnection networks and
+deep memory hierarchies.  These parts of the system are designed to
+minimize latency, but the growing scale of these machines has
+caused average and worst-case latencies to grow by orders of magnitdue.
+Recently, some of these latencies have grown to the point that supercomputers
+share many of the same performance properties as machines commonly
+classified as distributed systems.  Despite this major evolution in the
+performance of supercomputers, most programs for supercomputers are still
+written using the same old programming models and libraries that were designed
+years ago for machines many orders of magnitude smaller.
+
+In this paper we propose a new interface for programming modern supercomputers 
+that is designed to hide the large latencies associated with current 
+interconnection networks and memory hierarchies.  Our interface is built 
+around {\em events}.  Events provide a mechanism for naming a point in the future
+when an operation (i.e. task, copy, synchronization) will complete.  Every call to perform
+an operation in our runtime is asynchronous and returns immediately with an event that will 
+{\em trigger} when the operation has completed.  Furthermore, every call
+to perform an operation can take a precondition event that must trigger before
+the operation can begin.  Using events programmers can compose large
+chains of dependent operations.  The runtime is free to execute operations
+in any order that obeys the event dependences.  With this freedom the
+runtime can schedule operations in a way that optimizes throughput
+while automatically hiding the long latency memory and communication operations.
+
+For most supercomputer programmers effectively overlapping computation with
+memory and communication latency is the most difficult barrier to achieving high
+performance.  Current parallel programming APIs hinder this ability by either
+failing to support asynchronous constructs, or by supporting constructs that are
+not composable.  For example, MPI supports asynchronous send and receive operations
+but does so in a way that later requires blocking to know when the send and receive 
+operations have completed\cite{MPI}.  There is no way in MPI to launch a new 
+operation contingent on the result of a blocking receive operation.
+
+% Parallel programs have three parts
+%  Parallel control
+%  Data movement in the memory hierarchy
+%  Synchronization

### 2012-07-19 doc/popl2013full/langdef.tex
edits

@@
 $\mapconsist{\Omega}$, the quantified variables $r_i$ and $r_j$
-range over both logical and concrete regions.  Since $M$'s domain is
+range over both logical and physical regions.  Since $M$'s domain is
 logical regions, we extend $M$ to constraints with mixed logical and
-concrete regions in the standard way by defining
+physical regions in the standard way by defining
 $M \llbracket \rho \rrbracket = \rho$.}
@@
 \valueconsist[]{\tuple{v_1}{v_2}}{\tuple{\hat T_1}{\hat T_2}} & \Leftrightarrow & (\valueconsist[]{v_1}{\hat T_1}) \wedge (\valueconsist[]{v_2}{\hat T_2}) \\
-\valueconsist[]{\langle \langle \oton{\rho}, v \rangle \rangle}{\text{rr}[\oton{r}]\ \hat T\text{ where }\Omega} & \Leftrightarrow & (\valueconsist[]{v}{T[\rho_1/r_1, \ldots \rho_n/r_n]}) \wedge (\mapconsist[\{ (r_i, \rho_i) \}]{\Omega}) \\
+\valueconsist[]{\langle \langle \oton{\rho}, v \rangle \rangle}{\text{rr}[\oton{r}]\ \hat T\text{ where }\hat \Omega} & \Leftrightarrow & (\valueconsist[]{v}{T[\rho_1/r_1, \ldots \rho_n/r_n]}) \wedge (\mapconsist[\{ (r_i, \rho_i) \}]{\hat \Omega}) \\
 \valueconsist[]{K}{coloring(\rho)} & \Leftrightarrow & \forall l_1, v_1. (l_1, v_1) \in K \Rightarrow (l_1 \in \rho\ \wedge \\
@@
 Many of the cases are similar, and benefit from
-the use of the following straighforward lemmas:
+the use of the following straightforward lemmas:
 \begin{lem}
@@
 
-We first give an outline of the idea behindthe proof.  For each subexpression of a compound expression,
+We first give an outline of the idea behind the proof.  For each subexpression of a compound expression,
 there are three consistency properties we must prove to be able to apply induction: \\
@@
 \hangindent=2\parindent \hangafter=1 \noindent %
-$\storeconsist{S}{H}$ -  The heap typing consistency of all stores used in subexpressions follows directly from the following
-lemma: If $\storeconsist{S}{H}$ and $\traceconsist{E}$, then $\storeconsist{\text{apply}(S, E)}{H}$.
+$\storeconsist{S}{H}$ -  The heap typing consistency of all stores used in subexpressions follows directly from
+Lemma~\ref{lem:heapconst:apply}.
 
@@
 $\traceconsist{E}$ -  The type consistency of the values in an expression's memory
-trace follows from the following lemma applied to the traces of subexpressions:
-if $\traceconsist{E_1}$ and $\traceconsist{E_2}$ and $any\_interleave(E', E_1, E_2)$, then $\traceconsist{E'}$.
+trace follows from Lemma~\ref{lem:heapconst:effects1} and Lemma~\ref{lem:heapconst:effects2}.
 New memory operations are added by {\tt write} and {\tt reduce} expressions, but consistency follows
@@
 The proof of the theorem is by cases on the structure of the derivation of $e$.
-\begin{itemize}
-
+{\begin{itemize}
+\setlength{\itemsep}{4pt}
 \item
@@
 $$\opsenv{\text{null\ }T@r}{null}{[]}$$ and 
-$$\typeenv{\text{null}}{T'}$$ and 
+$$\typeenv{\text{null\ }T@r}{T'}$$ and 
 $\mapconsist{\Omega}$ and 
@@
  By assumption, 
-$$\opsenv{\text{ if } e_1 \text{ then } e_2 \text{ else } e_3}{v_2}{E}$$ and 
+$$\opsenv{\text{ if } e_1 \text{ then } e_2 \text{ else } e_3}{v_2}{E_1 \dplus E_2}$$ and 
 $$\typeenv{\text{ if } e_1 \text{ then } e_2 \text{ else } e_3}{T}$$ and
@@
 For the subexpression $e_1$, 
-from [E-IfElse-T] we know $$\opsenv{e_1}{true}{E_1 \dplus E_2}$$
+from [E-IfElse-T] we know $$\opsenv{e_1}{true}{E_1}$$
 From [T-IfElse], we know that $$\typeenv{e_1}{bool}$$
@@
 $\storeconsist{S}{H}$.
-The only type rule that can apply if [T-Newcolor], so can infer that $T = \text{coloring}(r)$.
+The only type rule that can apply is [T-Newcolor], so we can infer that $T = \text{coloring}(r)$.
 From the structure of $K$ it follows that $\valueconsist{K}{\text{coloring(r)}}$.
@@
 $\storeconsist{S}{H}$.
-The only type rule that can apply if [T-Color], so we know $T = \text{coloring}(r)$.
+The only type rule that can apply is [T-Color], so we know $T = \text{coloring}(r)$.
 
@@
 For the subexpression $e_3$, 
-from [E-Color] we know $$\opsenv[S=S'']{e_3}{v}{E_3}$$ where $S'' \text{apply}(S',E_2)$.
+from [E-Color] we know $$\opsenv[S=S'']{e_3}{v}{E_3}$$ where $S'' = \text{apply}(S',E_2)$.
 From [T-Color], we know that $$\typeenv{e_3}{int}$$
@@
 Now $K'$ is built from $K$ by extending $K$ with a mapping for location $l$.  First, because $\valueconsist{l}{T'@r}$ and
-$\valueconsist{K}{\text{coloring(r))}$ and because $K'$ excludes any previous mapping for $l$ (guaranteeing the the coloring is a function), we conclude $\valueconsist{K'}{\text{coloring}(r)}}$.
+$\valueconsist{K}{\text{coloring}(r)}$ and because $K'$ excludes any previous mapping for $l$ (guaranteeing the the coloring is a function), we conclude $\valueconsist{K'}{\text{coloring}(r)}$.
 From [E-Color], we also know
@@
 we can conclude $\traceconsist{E'}$.  Similarly, by applying Lemma~\ref{lemma:effsound:effects2} twice
-we can conlude $\privconsist{E'}{\Phi}$.
+we can conclude $\privconsist{E'}{\Phi}$.
 
@@
 By assumption, $$\opsenv{\text{upregion}(e,\oton{r})}{v}{E}$$ and
-$$\typeenv{\text{upgregion}(e,\oton{r})}{T}$$ and
+$$\typeenv{\text{upregion}(e,\oton{r})}{T}$$ and
 $\mapconsist{\Omega}$ and 
@@
 we know $\forall i. \exists j. r_i' \leq r_j \in \Omega^*$.  Combining this with
-$\mapconsist{\Omega}$ and Lemma~\ref{lemma:omegaclosure}, we know there
-is some $r_j$ such that $v \in M(r_j)$.
-Therefore $\valueconsist{v}{T'@(\oton{r})}$.
+$\mapconsist{\Omega}$ and 
+From Lemma~\ref{lemma:omegaclosure}, we have $mapconsist{\Omega^*}$, so there
+must be some $r_j$ such that $M(r_i') \subseteq M(r_j)$.  Therefore $v \in M(r_j)$
+and $\valueconsist{v}{T'@(\oton{r})}$.
 
@@
 By assumption, $$\opsenv{\text{downregion}(e,\oton{r})}{v}{E}$$ and
-$$\typeenv{\text{downgregion}(e,\oton{r})}{T}$$ and
+$$\typeenv{\text{downregion}(e,\oton{r})}{T}$$ and
 $\mapconsist{\Omega}$ and 
@@
 
-To show $\valueconsist{v}{T}$ there are two cases.
-If $v = M(l)$ then the claim follows from $\valueconsist{l}{T@(\oton{r})}$.
-If $v$ is instead some $v'$ of type $H(l)$ the result also follows because $H(l) = \hat{T}$.
+To show $\valueconsist{v}{T}$ there are two cases.  If $l \not\in C$, then $v = S'(l)$.
+Since $\storeconsist{S'}{H}$ from Lemma~\ref{lemma:heapconst:apply}, we have $\valueconsist[]{S'(l)}{H(l)}$.
+If $l \in C$, we have $v = \valueconsist[]{v'}{H(l)}$ from [E-Read] directly.  In
+both cases, $\valueconsist{l}{T@(\oton{r})}$ gives us $H(l) = \hat T$ and therefore
+$\valueconsist{v}{T}$.
+%If $v = M(l)$ then the claim follows from $\valueconsist{l}{T@(\oton{r})}$.
+%If $v$ is instead some $v'$ of type $H(l)$ the result also follows because $H(l) = \hat{T}$.
 Next, $\traceconsist{E \dplus [ read(l,excl,v)]}$ follows immediately from $\traceconsist{E}$ as
-any read operation is consistent (the resriction is only on writes and reductions).
+any read operation is consistent (the restriction is only on writes and reductions).
 
@@
 $$M \llbracket \exists r'_1, \ldots r'_n.\ T_2\text{ where }\Omega_1 \rrbracket$$
-replaces all free region names in $T_2$ and particularly $\Omega_1$ by concrete regions, so
+replaces all free region names in $T_2$ and particularly $\Omega_1$ by physical regions, so
 that the only region names remaining are $r'_1,\ldots,r'_n$.  It is then easy to show that
@@
 the domain of $\Gamma'$.  For any $id' \neq id$, the statement is equivalent
-to showing $\valueconsist[M']{L(x)}{\Gamma(x)}$, which must hold becuase $M'$ does 
+to showing $\valueconsist[M']{L(x)}{\Gamma(x)}$, which must hold because $M'$ does 
 does not change any existing binding in $M$.  For $id$, we have
@@
 $\traceconsist{E_{n+1}}$ under $M'$ implies $\traceconsist{E_{n+1}}$
-under $M$, note that $H(\cdot)$ does not depend on the mapping (because it includes only concrete, not logical, regions) so the
+under $M$, note that $H(\cdot)$ does not depend on the mapping (because it includes only physical, not logical, regions) so the
 types of locations are unaffected by the mapping (see
@@
 Finally, from Lemma~\ref{lemma:effsound:effects2} we conclude $\privconsist{E''}{\Phi}$.
-\end{itemize}
+\end{itemize}}
 \end{proof}

### 2012-07-19 doc/popl2013full/langdef.tex
connected dots in E-Pack proof

@@
 physical regions in the standard way by defining
-$M \llbracket \rho \rrbracket = \rho$.}
+$M \llbracket \rho \rrbracket = \rho$.  And although the definition of
+$\traceconsist{E}$ refers to $M$ and $\Gamma$, the restriction that reduction
+functions must be monomorphic and visible in all functions means that 
+$\traceconsist{E}$ will be insensistive to the particular $M$ and $\Gamma$.}
 
@@
 \valueconsist[]{\tuple{v_1}{v_2}}{\tuple{\hat T_1}{\hat T_2}} & \Leftrightarrow & (\valueconsist[]{v_1}{\hat T_1}) \wedge (\valueconsist[]{v_2}{\hat T_2}) \\
-\valueconsist[]{\langle \langle \oton{\rho}, v \rangle \rangle}{\text{rr}[\oton{r}]\ \hat T\text{ where }\hat \Omega} & \Leftrightarrow & (\valueconsist[]{v}{T[\rho_1/r_1, \ldots \rho_n/r_n]}) \wedge (\mapconsist[\{ (r_i, \rho_i) \}]{\hat \Omega}) \\
+\valueconsist[]{\langle \langle \oton{\rho}, v \rangle \rangle}{\text{rr}[\oton{r}]\ \hat T\text{ where }\hat \Omega} & \Leftrightarrow & (\valueconsist[]{v}{\hat T[\rho_1/r_1, \ldots \rho_n/r_n]}) \wedge (\mapconsist[\{ (r_i, \rho_i) \}]{\hat \Omega}) \\
 \valueconsist[]{K}{coloring(\rho)} & \Leftrightarrow & \forall l_1, v_1. (l_1, v_1) \in K \Rightarrow (l_1 \in \rho\ \wedge \\
@@
 $\privconsist{E}{\Phi^*}$ if and only if $\privconsist{E}{\Phi}$.
+\end{lem}
+
+\begin{lem}
+\label{lemma:mappedomega}
+\rm
+$\mapconsist{\Omega}$ if and only if $\mapconsist[\emptyset]{M \llbracket \Omega \rrbracket}$.
 \end{lem}
@@
 The only thing remaining to show is $\valueconsist{v'}{T_1}$.
-From [E-Pack] we have   $v' = \langle \langle \otok{k}{\rho}, v \rangle \rangle$.
-From [T-Pack] we have $T_1 = \exists r'_1, \ldots r'_n.\ T_2\text{ where }\Omega_1$
+From [E-Pack] we have   $v' = \langle \langle \otok{k}{\rho}, v \rangle \rangle$ and $M(r_i) = \rho_i$.
+From [T-Pack] we have $T_1 = \exists r'_1, \ldots, r'_n.\ T_2\text{ where }\Omega_1$
 and $\Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^*$.
 
+Let $M \llbracket T_1 \rrbracket = \exists r'_1, \ldots, r'_n.\ \hat T_2\text{ where }\hat \Omega_1$.
+We require that $\valueconsist[]{v}{\hat T_2[\rho_1/r'_1,\ldots,\rho_k/r'_k]}$ and $\mapconsist[\{(r'_i,\rho_i)\}]{\hat \Omega_1}$.
+The right hand side of the first condition can be rewritten as $\hat T_2[M(r_1)/r'_1,\ldots,M(r_k)/r'_k]$, which is equal
+to $M \llbracket T_2 [r_1/r'_1,\ldots,r_k/r'_k] \rrbracket$.
+
+Similarly, observe that
+$$\hat \Omega_1 [ \rho_1/r'_1,\ldots,\rho_k/r'_k ] = M \llbracket \Omega_1 [ r_1/r'_1,\ldots,r_k/r'_k ] \rrbracket$$
+.  With two applications of Lemma~\ref{lemma:mappedomega}, we have $\mapconsist[\{(r'_i,\rho_i)\}]{\hat \Omega_1}$
+if and only if $\mapconsist{\Omega_1 [ r_1/r'_1,\ldots,r_k/r'_k ]}$.  The latter constraints are a subset of
+$\Omega^*$, and must be satisfied if $\mapconsist{\Omega^*}$, which we have from Lemma~\ref{lemma:omegaclosure}.
+
+\begin{comment}
+For the second condition, consider each constraint $x_i \leq x_j \in \hat \Omega_1$, where each $x_i$ is either a 
+physical region $\rho_i$ or one of the bounded logical regions $r'_i$.
+
+We must therefore show $\valueconsist{v}{T_2[\rho_1/r'_1,\ldots,\rho_k/r'_k]}$ and $\mapconsist[\{(r_i',\rho_i)\}]{M \llbracket \Omega_1 \rrbracket}$.
+We have $\rho_i = M(r_i)$ from [E-Pack] and a partial application of $M$ satisfies
+the first condition.  To show the second condition, we apply the renaming $[r_1/r'_1,\ldots,r_k/r'_k]$ to both sides,
+yielding
+$$\mapconsist[\{(r_i,\rho_i)\}]{M \llbracket \Omega_1 [r_1/r'_1,\ldots,r_k/r'_k] \rrbracket}$$
+,  Observe that the left hand side is a subset of $M$
+By partially applying $M$ to $T_2[r_1/r'_1,\ldots,r_k/r'_k]$, we get $\valueconsist{v}{T_2[\rho_1/r'_1,\ldots,\rho_k/r'_k]}$, satisfying the first half o
 Observe that 
@@
 from which $\valueconsist{v'}{T_1}$ follows (see Figure~\ref{fig:constprop}).
+\end{comment}
 

### 2012-07-19 doc/popl2013full/langdef.tex
unpack, a few more typo fixes

@@
 
-The proof is a standard induction on the structure of the derivation
+The proof is a standard induction on the structure of the derivation.
 For each of 
@@
 $\storeconsist{S}{H}$ -  The heap typing consistency of all stores used in subexpressions follows directly from
-Lemma~\ref{lem:heapconst:apply}.
+Lemma~\ref{lemma:heapconst:apply}.
 
@@
 $\traceconsist{E}$ -  The type consistency of the values in an expression's memory
-trace follows from Lemma~\ref{lem:heapconst:effects1} and Lemma~\ref{lem:heapconst:effects2}.
+trace follows from Lemma~\ref{lemma:heapconst:effects1} and Lemma~\ref{lemma:heapconst:effects2}.
 New memory operations are added by {\tt write} and {\tt reduce} expressions, but consistency follows
@@
 $$\opsenv{\text{newcolor } r}{K}{[]}$$ 
-where  $K = \{ (l_1,v_1),\ldots,(l_p,i_p) \}$ such that
+where  $K = \{ (l_1,iv_1),\ldots,(l_p,iv_p) \}$ such that
 $l_i \in M(r)$ for all $i$ and $l_1,\ldots,l_p$ are pairwise distinct locations.
@@
 $H(v) = \hat T'$.   From [T-UpRgn]
-we know $\forall i. \exists j. r_i' \leq r_j \in \Omega^*$.  Combining this with
-$\mapconsist{\Omega}$ and 
-From Lemma~\ref{lemma:omegaclosure}, we have $mapconsist{\Omega^*}$, so there
-must be some $r_j$ such that $M(r_i') \subseteq M(r_j)$.  Therefore $v \in M(r_j)$
+we know $\forall i. \exists j. r_i' \leq r_j \in \Omega^*$.
+%  Combining this with $\mapconsist{\Omega}$ and 
+From Lemma~\ref{lemma:omegaclosure}, we have $\mapconsist{\Omega^*}$, so there
+must be some $r_j$ such that $M(r_i') \subseteq M(r_j)$.  Then $v \in M(r_j)$
 and $\valueconsist{v}{T'@(\oton{r})}$.
@@
 $$\hat \Omega_1 [ \rho_1/r'_1,\ldots,\rho_k/r'_k ] = M \llbracket \Omega_1 [ r_1/r'_1,\ldots,r_k/r'_k ] \rrbracket$$
-.  With two applications of Lemma~\ref{lemma:mappedomega}, we have $\mapconsist[\{(r'_i,\rho_i)\}]{\hat \Omega_1}$
-if and only if $\mapconsist{\Omega_1 [ r_1/r'_1,\ldots,r_k/r'_k ]}$.  The latter constraints are a subset of
+.  With two applications of Lemma~\ref{lemma:mappedomega}, we have
+$$\mapconsist[\{(r'_i,\rho_i)\}]{\hat \Omega_1} \Leftrightarrow \mapconsist{\Omega_1 [ r_1/r'_1,\ldots,r_k/r'_k ]}$$
+.  The latter constraints are a subset of
 $\Omega^*$, and must be satisfied if $\mapconsist{\Omega^*}$, which we have from Lemma~\ref{lemma:omegaclosure}.
@@
 
-For subexpression $e_1$, from [E-UnPack] we have $$\opsenv{e_1}{\langle \langle \otok{k}{\rho},v \rangle\rangle}{E}$$ and
-from [T-UnPack] we know $$\typeenv{e_1}{T_1 = \exists r'_1,\ldots,r'_n.T_2 \text{ where }\Omega_1}$$
+For subexpression $e_1$, from [E-UnPack] we have $$\opsenv{e_1}{\langle \langle \otok{k}{\rho},v_1 \rangle\rangle}{E}$$ and
+from [T-Unpack] we know $$\typeenv{e_1}{T_1 = \exists r'_1,\ldots,r'_n.T_2 \text{ where }\Omega_1}$$
 By induction, we can conclude that $\valueconsist{\langle \langle \otok{k}{\rho},v \rangle\rangle}{T_1}$ and
@@
 
+We can show $\mapconsist[M']{\Omega'}$ by showing both $\mapconsist[M']{\Omega}$ and $\mapconsist[M']{\Omega_1[r_1/r'_1,\ldots,r_k/r'_k]}$.
+The former follows directly from $\mapconsist{\Omega}$ as $M'$ does not change any existing
+mappings in $M$.  The latter requires an argument similar to the [E-Pack] case.  We let
+$M \llbracket T_1 \rrbracket = \exists r'_1, \ldots, r'_n.\ \hat T_2\text{ where }\hat \Omega_1$, and conclude that
+$\mapconsist[\{ \rho_1/r'_1,\ldots,\rho_k/r'_k \}]{\hat \Omega_1}$.  From Lemma~\ref{lemma:mappedomega} we have 
+$\mapconsist[\emptyset]{\hat \Omega_1[\rho_1/r'_1,\ldots,\rho_k/r'_k]}$.  Finally, we observe that 
+$$M' \llbracket \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \rrbracket = \hat \Omega_1[\rho_1/r'_1,\ldots,\rho_k/r'_k]$$
+and conclude $\mapconsist[M']{\Omega'}$ with another application of Lemma~\ref{lemma:mappedomega}.
+
+\begin{comment}
 To show $\mapconsist[M']{\Omega'}$, consider a constraint $r \leq r' \in \Omega'$.  There are two cases. If $r \leq r' \in \Omega_1[r_1/r'_1,\ldots,r_k/r'_k]$
@@
 The reasoning for constraints $r * r'$ is the same.
+\end{comment}
 
@@
 to showing $\valueconsist[M']{L(x)}{\Gamma(x)}$, which must hold because $M'$ does 
-does not change any existing binding in $M$.  For $id$, we have
+does not change any existing binding in $M$.
+For $id$, we have $\valueconsist[]{v_1}{\hat T_2[\rho_1/r'_1,\ldots,\rho_k/r'_k]}$
+and again observe that since $r'_1,\ldots,r'_k$ do not appear in $M$, the right hand side
+is equal to $M' \llbracket T_2[r_1/r'_1,\ldots,r_k/r'_k] \rrbracket$.
+
+\begin{comment}
 $\valueconsist[M']{v_1}{T_1[r_1/r'_1,\ldots,r_k/r'_k]}$, which is implied by
 $\valueconsist{\langle \langle \otok{k}{\rho},v \rangle\rangle}{T_1}$.
+\end{comment}
 

### 2012-07-20 doc/popl2013full/langdef.tex
few more edits, one new proof to coherence section

@@
 
-The proof of this (and other lemmas in this section) is included in \cite{LegionTypes12}.
+\begin{proof}
+The proof is by induction on the structure of the derivation of $e$.  The clobber set is only manipulated
+in one case and only examined in one other case.  In all other cases, the clobber set is unchanged and
+used only in the derivation of subexpressions.  The case where the clobber set is manipulated is [E-Call], but
+the use of a union operation still allows the inductive hypothesis to be applied to the subexpression $e_{n+1}$.
+The examination of the clobber set occurs in [E-Read] where it is used to choose the result $v$ of the read
+from the heap.  For each use of the [E-Read] rule, there are three possible cases:
+\begin{itemize}
+\item $l \in C$, for which $v = S'(l)$ in the derivation using $C$.  Since $C \subseteq C'$, we have $l \in C'$, and $S'(l)$ is the correct result for the $C'$ derivation as well.
+\item $l \not\in C'$, and therefore $l \not\in C$.  Both derivations may use any value $v' : H(l)$.
+\item $l \in C, l \not\in C'$, for which $v = S'(l)$ must be chosen for the derivation using $C$.
+From Theorem~\ref{thm:effects} and Lemma~\ref{lemma:heapconst:apply}, we have $\traceconsist{E}$ and $\storeconsist{S' = apply(S, E)}{H}$, so $\valueconsist[]{S'(l)}{H(l)}$, and $v$ is a legal choice for $v'$ in the derivation using $C'$.
+\end{itemize}
+\end{proof}
+
+For the following proofs, we will need one additional lemma, which follows directly
+from the definition of $apply$:
+
+\begin{lem}
+\label{lemma:applychain}
+\rm
+Let $S$ be an initial store, and $E_1$ and $E_2$ be memory traces.  Then:
+$$apply(S, E_1 \dplus E_2) = apply(apply(S, E_1), E_2)$$
+\end{lem}
+
 We now show that the concatenation of the memory traces from these serially
@@
 show that $E'$ is coherent for both $L_{excl}$ and $L_{atomic}$.  This is done by
-matching the structure of $E'$ with $\oton{E}$ and the use of the following
-lemma: $apply(S, E_1 \dplus E_2) = apply(apply(S, E_1), E_2)$.  With the
+matching the structure of $E'$ with $\oton{E}$ and the use of Lemma~\ref{lemma:applychain}.
+With the
 question of coherence addressed, the sequential equivalence of $E'$ is trivial
@@
 \begin{proof}
-The proof is by contradiction.  Assume $E_1 \nonint E_2$.  Then there must be some $\epsilon_1$ in $E_1$ and some $\epsilon_2$ in $E_2$ such that $\epsilon_1 \nonint \epsilon_2$.  This requires that
+The proof is by contradiction.
+Assume $E_1 \cancel\nonint E_2$.
+Then there must be some $\epsilon_1$ in $E_1$ and some $\epsilon_2$ in $E_2$ such that $\epsilon_1 \cancel\nonint \epsilon_2$.
+This requires that
 $\epsilon_1$ and $\epsilon_2$ be operations to the same location $l$ and not both be reads or both
@@
 Similarly, there must be some $r_2$ satisfying $l \in M(r_2) \wedge writes(r_2) \in \Phi_2$.
-Letting $\phi_1 = reads(r_1)$ and $\phi_2 = writes(r_2)$, we have $M(r_1) \cap M(r_2) \not= \emptyset$ (it contains at least $l$), so $\phi_1 {\nonint[M]} \phi_2$ and $\Phi_1 {\nonint[M]} \Phi_2$.
+Letting $\phi_1 = reads(r_1)$ and $\phi_2 = writes(r_2)$, we have $M(r_1) \cap M(r_2) \not= \emptyset$ (the intersection contains at least $l$), so $\phi_1 \cancel{\nonint[M]} \phi_2$ and $\Phi_1 \cancel{\nonint[M]} \Phi_2$.
 There are seven other cases to consider (using $id_1 \not= id_2$ for the reduce-reduce case), but all yield the same result.

### 2012-07-20 doc/asplos2013/conclusion.tex
Adding a complete introduction

\section{Conclusion}
\label{sec:conclusion}

Conclusion goes here.

### 2012-07-20 doc/asplos2013/intro.tex
Adding a complete introduction

--- 2012-07-17 doc/asplos2013/intro.tex
@@
 \label{sec:intro}
-Introduction goes here\cite{Keeneland}.
+
+% Big machines -> latencies are growing
+% Processors can't be blocked, need to be able to continue handling work
+% Operations need to be composable
+% deferred execution model
+
+% Contributions:
+% - An event system where every operation can take an event and returns an event
+% - Handles are globally valid
+% - More flexible synchronization primitive for this environment: deferred locks
+% - Physical regions for data movement including special versions for reductions
+
+% Parallel programs have three parts
+%  Parallel control
+%  Data movement in the memory hierarchy
+%  Synchronization
+
+% Bulk-synchronous
+% asynchronous components are not composable
+
+With the multicore revolution, modern supercomputers have continued to scale
+by increasing the number of processors available for computation.  The
+price of this scaling is complex interconnection networks and
+deep memory hierarchies.  These parts of the system are designed to
+minimize latency, but the growing scale of supercomputers has
+caused average and worst-case latencies to grow by orders of magnitdue.
+Recently, some of these latencies have grown to the point that supercomputers
+share many of the same performance properties as machines commonly
+classified as distributed systems.  Despite this major evolution in the
+performance of supercomputers, most programs for supercomputers are still
+written using programming models and libraries that were designed
+for machines many orders of magnitude smaller\cite{MPI,COARRAY_FORTRAN,UPC99}.
+
+The approach that current programming models provide is commonly reffered to
+as a {\em bulk-synchronous} model\cite{Valiant90}.  Bulk-synchronous models
+break computations down into three kinds of phases: parallel computation, data
+communication, and synchronization.  At any given point in time all the processors
+in the system are engaged in one of the three kinds of phase.  The bulk-synchronous
+approach to parallel programming makes it easy to reason about parallel execution, but
+has a fundamental limitation: it is very difficult to hide long-latency communication
+or memory movement phases.  In older machines this was not a problem because slower
+processors and smaller machines meant computation phases were much larger than
+communication or synchronization phases.  However, today's machines are much
+larger with faster processors which cause increasingly large percentages of time
+to be spend in the communication and synchronization phases.
+
+To combat this problem, parallel programming libraries have introduced
+asynchronous constructs which allow the programmer to overlap phases.  For example,
+MPI has asynchronous {\em send} and {\em receive} operations for sending
+messages in parallel with computation.  There are two problems with this approach.
+First, not all phases support asynchronous constructs.  Second,
+the asynchronous operations do not {\em compose}.  There is no way to express
+that an asynchronous computation should begin as soon as an asynchronous
+communication operation has completed.  The result is that
+at some point there must be blocking operations to coordinate synchronous
+and asynchronous phases.  If blocking calls are not accurately placed in the
+program they can lead to processor stalls that expose latency.  Correct
+placement of blocking calls can be dependent on input data, algorithmic decisions like blocking
+factors, or even the underlying hardware.  
+To hide as much latency as possible and to minimize processor stalls, regardless
+of inputs, algorithmic choices, or underlying hardware,
+all three major aspects of parallel programs (computation, data movement, synchronization)
+should be made asynchronous and composable.
+
+We propose a new interface for programming modern supercomputers 
+that is fully asynchronous to hide the large latencies associated with current 
+interconnection networks and memory hierarchies.  Our interface makes it possible
+to asynchronously compose computation, data movement, and synchronizatioan.  The 
+fundamental mechanism that enables us to compose all three aspects of parallel
+programs are {\em events}.  Events provide a mechanism for naming a point in the future
+when an operation (i.e. computation, data copy, synchronization) will complete.  
+Every call to perform an operation in our runtime returns immediately with an event that will 
+{\em trigger} when the operation has completed.  Furthermore, every call
+to perform an operation can take a precondition event that must trigger before
+the operation can begin.  Using events programmers can compose large
+chains of dependent operations.  The runtime is free to execute operations
+in any order that obeys the event dependences.  With this freedom the
+runtime can schedule operations in a way that optimizes throughput
+while automatically hiding the long latency operations.
+
+The following sections describe our contributions:
+\begin{itemize}
+\item We present an interface for asynchronously composing
+parallel computation, data movement, and synchronization based on events.  
+To support synchronization in an event system we describe a novel synchronization
+primitive called a {\em deffered lock}.  Our interface also supports efficient
+mechanisms for performing distributed reduction operations (Section~\ref{sec:interface}).
+
+\item We describe a distributed implementation of our interface that requires no
+global coordination and is capable of running on current supercomputers.  We
+describe optimizations for the implementation of events, deffered locks, and
+reductions (Section~\ref{sec:impl}).
+
+\item We show the result of micro-benchmarks that are designed to stress
+our implementation.  The results indiciate that our implementation approaches the performance of
+the underlying hardware (Section~\ref{sec:micro}).
+
+\item We demonstrate that our interface is powerful enough to support a higher-level
+runtime system and that real applications running on this stack of software
+are capable of achieving high-performance (Section~\ref{sec:apps}).
+\end{itemize}
+
+In Section~\ref{sec:related} we describe related work and in Section~\ref{sec:conclusion}
+we conclude.

### 2012-07-20 doc/popl2013full/langdef.tex

@@
 $$\hat \Omega_1 [ \rho_1/r'_1,\ldots,\rho_k/r'_k ] = M \llbracket \Omega_1 [ r_1/r'_1,\ldots,r_k/r'_k ] \rrbracket$$
-.  With two applications of Lemma~\ref{lemma:mappedomega}, we have
+With two applications of Lemma~\ref{lemma:mappedomega}, we have
 $$\mapconsist[\{(r'_i,\rho_i)\}]{\hat \Omega_1} \Leftrightarrow \mapconsist{\Omega_1 [ r_1/r'_1,\ldots,r_k/r'_k ]}$$
-.  The latter constraints are a subset of
+The latter constraints are a subset of
 $\Omega^*$, and must be satisfied if $\mapconsist{\Omega^*}$, which we have from Lemma~\ref{lemma:omegaclosure}.

### 2012-07-20 doc/asplos2013/interface.tex
Major edits to the interface section

@@
 
-The runtime interface organizes
-functionality into objects.  Objects are grouped into
-four different categories: machine objects for describing
-the underlying architecture, events for capturing dependences,
-deferred locks for performing synchronization, and physical regions
-for managing data.  In all cases, the objects are light-weight
-{\em handles} that can be copied and passed around by value.  Furthermore
+Our interface organizes functionality into objects.  We describe
+these objects in five parts: events for composing operations 
+(Section~\ref{subsec:events}), processors for parallel computation 
+(\ref{subsec:procs}), deffered locks for synchronization
+(\ref{subsec:locks}), physical regions for data movement
+(\ref{subsec:phyreg}), and an object for introspection of the underlying
+hardware (\ref{subsec:machmodel}).  In all cases (except for the
+machine which is a static singleton), instances of the objects
+are light-weight {\em handles} that provide a unique name for an underlying
+implementation.  Handles can be copied and passed by value.  Furthermore,
 every handle is valid everywhere in the system.  This property allows 
@@
 }
-\subsection{Machine Interface}
-\label{subsec:machmodel}
-Listing~\ref{lst:machineapi} shows the machine interface.  The machine interface 
-contains three different object types: {\tt Processor} (line 1), {\tt Memory} (line 17), and a 
-singleton {\tt Machine} object (line 22).  For every processor (i.e. CPU core or GPU) in 
-the system there exists a {\tt Processor} handle with a unique ID naming that processor.  
-{\tt Processor} objects support a single {\tt spawn} operation (line 13-14) that will
-launch a {\em task} on that processor.  Note that because the {\tt spawn} operation
-is invoked on a processor handle, it is possible to launch a task on any
-processor from anywhere in the system.  The {\tt spawn} call takes as arguments
-the {\tt TaskFuncID} of the task to be run, a pointer to and size of arguments to
-be passed to the task, and an optional {\tt Event} (described in Section~\ref{subsec:events})
-that must trigger before the task can begin.  The result of the {\tt spawn} call
-is an {\tt Event} that will trigger when the task completes.
-
-For every memory in the system there also exists a {\tt Memory} handle with a unique ID. 
-Different memories often, but not always, imply a different address space.  For
-example, in a large cluster each node would have a different {\tt Memory} handle
-corresponding to each node's DRAM memory.
-However, if the cluster supported multiple GPUs then there would be a different
-{\tt Memory} handle for each GPU's {\em zero-copy} memory which shares part of the 
-address space with the corresponding node memory.  Memories will be used for describing
-the placement of data described in Section~\ref{subsec:phyreg}.
-
-The {\tt Machine} object is a singleton object that serves two purposes.  First, it
-is the mechanism for initializing the runtime; at the start of a program the programmer
-creates a {\tt Machine} object and invokes the {\tt run} method (line 30-31).
-Second, the {\tt Machine} object provides an interface for the programmer to 
-discover properties of the underlying machine (lines 35-43).  The {\tt Machine} object maintains
-sets of all the unique {\tt Processor} and {\tt Memory} handles in the system.
-The {\tt Machine} also maintains two relations between these sets:
-
-\begin{itemize}
-\item Processor-Memory: for each {\tt Processor} - {\tt Memory} pair whether
-the {\tt Processor} can directly access (read and write) the {\tt Memory} and 
-latency and bandwidth properties if access is possible
-\item Memory-Memory: for every pair of {\tt Memory} whether a copy can be
-performed between the two memories and latency and
-bandwidth properties if a copy is possible
-\end{itemize}
-
-The programmer can query these two relations by the {\tt get\_*\_affinity} calls
-(lines 40-43).  These calls populate a vector with affinity structures (not shown)
-that contain latency and bandwidth information.  The programmer can restrict the query
-by specifying a specific processor or memory, otherwise information is returned
-about all processors and memories in the system.
-
-\begin{lstlisting}[float={t},label={lst:machineapi},caption={Machine Interface.}]
-class Processor {
-public:
-  const unsigned id;
-  typedef unsigned TaskFuncID;
-  typedef void (*TaskFuncPtr)
-            (const void *args,size_t arglen,Processor p);
-  typedef map<TaskFuncID, TaskFuncPtr> TaskIDTable;
-
-  enum Kind {
-    CPU_PROC,GPU_PROC,// ... future processor types
-  };
-
-  Event spawn(TaskFuncID func_id,const void *args,size_t arglen,
-              Event wait_on = Event::NO_EVENT) const;
-};
-
-class Memory {
-public:
-  const unsigned id;
-};
-
-class Machine {
-public:
-  Machine(int *argc, char ***argv,
-          const Processor::TaskIDTable &task_table);
-  enum RunStyle {
-    ONE_TASK_ONLY,ONE_TASK_PER_NODE,
-    ONE_TASK_PER_PROCESSOR,
-  };
-  void run(Processor::TaskFuncID task_id = 0, 
-        RunStyle = ONE_TASK_ONLY, const void *args, size_t arglen);
-public:
-  static Machine* get_machine(void); // pointer to Machine
-  // References to all processors and memories
-  const set<Memory>& get_all_memories(void) const;
-  const set<Processor>& get_all_processors(void) const;
-  Processor::Kind get_processor_kind(Processor p) const;
-  size_t get_memory_size(Memory m) const;
-  // Query properties of underlying hardware
-  int get_proc_mem_affinity(vector<ProcMemAffinity &result,
-          Processor restrict_proc = 0, Memory restrict_mem = 0);
-  int get_mem_mem_affinity(vector<MemMemAffinity> &result,
-          Memory restrict_mem1 = 0, Memory restrict_mem2 = 0);
-};
-\end{lstlisting}
 
@@
 \label{subsec:events}
-Listing~\ref{lst:eventapi} shows the interface for events.  
-{\tt Event} is the base type of events.  An instance of an {\tt Event} type 
+Events are the primary mechanism for describing dependences in our system.
+Listing~\ref{lst:eventapi} shows the interface for events.  An instance of the {\tt Event} type 
 names a unique event in the system.  {\tt NO\_EVENT} (line 5) is a special instance
-of an event that by definition has always triggered.  The {\tt Event} interface
-also supports testing whether an event has triggered (line 7) and waiting on
+of an event that by definition has always triggered.  The event interface
+supports testing whether an event has triggered (line 7) and waiting on
 an event to trigger (line 9).  While these methods can be useful, the most common
-use of events is passing them as preconditions for operations such as the {\tt spawn}
-call described in Section~\ref{subsec:machmodel}.  The programmer can use the
+use of events is passing them as preconditions for other operations.  The programmer can use the
 {\tt merge\_events} call (line 11) to create an event corresponding to the 
-conjunction of a set of events when multiple events are a precondition for an operation.
+conjunction of a set of events.
 
-In most cases events are created by the runtime as the result of other
-operations and the runtime is responsible for triggering these events.  Users 
+In most cases events are created as the result of other
+operations and the implementation is responsible for triggering these events.  Users 
 can create a special type of event called a {\tt UserEvent} (line 14) that the
-user has the power to trigger.  {\tt UserEvent} is a sub-type of {\tt Event} 
-that has {\tt trigger} method (line 17).  Users can also create another
+user has the power to trigger.  A user event is a sub-type of an event
+that has a {\tt trigger} method (line 17).  Users can also create another
 type of event called a {\tt Barrier} that will require multiple arrivals before
@@
 be seen as well as how many arrivals occur at a time.  Once a barrier's arrival
-count goes to zero, the barrier will trigger.  Note that since {\tt UserEvent}
-and {\tt Barrier} are sub-types of {\tt Event} they can be used wherever an
-{\tt Event} is required.
+count goes to zero, the barrier will trigger.  Note that barriers in our interface
+provide a superset of the functionality of traditional barriers since they can
+be used both in a blocking manner (via the {\tt wait} call), but can also be used
+asynchronously.  Lastly, since user events and barriers are sub-types of 
+an event they can be used wherever an event is required.
 
@@
 
+\subsection{Processors}
+\label{subsec:procs}
+Listing~\ref{lst:procapi} shows the interface for {\tt Processor} objects which allow for
+the creation of parallel computations.  Processor objects provide a way of naming 
+every computational unit in the machine.  In the case of our current implementation 
+processors are either individual cores or discrete GPUs, but this can be easily modified 
+to include new processor types (line 10).   Processors support a single {\tt spawn}
+operation (lines 13-14) that will launch a new {\em task} on that processor.
+Note that because the spawn operation
+is invoked on a processor handle, it is possible to launch a task on any
+processor from anywhere in the system.  To allow for the chaining of dependences
+through events the spawn call takes an 
+optional event that must trigger before the task can begin.  
+The result of the spawn call is an event that will trigger when the task completes
+which permits the creation of dependences upon this task.
+
+\begin{lstlisting}[float={t},label={lst:procapi},caption={Processor Interface.}]
+class Processor {
+public:
+  const unsigned id;
+  typedef unsigned TaskFuncID;
+  typedef void (*TaskFuncPtr)
+            (const void *args,size_t arglen,Processor p);
+  typedef map<TaskFuncID, TaskFuncPtr> TaskIDTable;
+
+  enum Kind {
+    CPU_PROC,GPU_PROC,// ... future processor types
+  };
+
+  Event spawn(TaskFuncID func_id,const void *args,size_t arglen,
+              Event wait_on = Event::NO_EVENT) const;
+};
+\end{lstlisting}
+
 \subsection{Deferred Locks}
 \label{subsec:locks}
-Deferred locks are a new synchronization mechanism that has different semantics from
-blocking locks.  From here on we refer to locks and deferred locks interchangeably, 
-but will refer to blocking locks explicitly.  Listing~\ref{lst:lockapi} shows the 
+
+Events and barriers express ordering properties between operations, but in many
+cases ordering is too strict.  For many applications access to data must only be atomic and
+not necessarily ordered.  Traditionally locks have been used to synchronize access
+to data.  However, all lock implementations that we are aware of require either blocking
+or spinning on locks, neither of which composes well with other asynchronous operations.
+Deferred locks are a new synchronization mechanism that allows for synchronization
+in an completely asynchronous environment.
+
+Listing~\ref{lst:lockapi} shows the 
 interface for locks.  Unlike blocking locks, the {\tt lock} method (line 4) doesn't
 block but instead always returns immediately with an event that will be triggered
-when the lock has been acquired.  If the lock can be acquired immediately then a {\tt NO\_EVENT}
-is returned.  The {\tt lock} method also takes an optional event
-parameter indicating that the lock acquire shouldn't be performed until the event
-has triggered.  Similarly, the {\tt unlock} method (line 5) takes an optional event parameter indicating
-an event to wait for before performing the unlock operation.
-
-Deferred locks provide a super-set of the functionality of blocking locks.  As their
-name suggests, deferred locks can use the event corresponding to their lock acquire
-operation to defer execution until the lock acquire has been granted.  Deferred
-locks can also be converted back into a blocking lock by immediately waiting on the event
-returned from a call to {\tt lock}.  In addition, deffered locks also provide {\tt mode} and
-and {\tt exclusive} parameters that allow the user to specify whether other requests can acquire
-the lock simultaneously.  A lock can only be acquired in one mode at a time.  The
-{\tt exclusive} parameter specifies whether other owners are permitted once the
-current lock request is granted.
+when the lock has been acquired.  Just like the spawn method, the lock method also 
+takes an optional event parameter as a precondition.  Similarly, the {\tt unlock} 
+method (line 5) takes an optional event precondition parameter.
 
@@
 that requests the lock doesn't have to be the one that uses it.  A common
-convention in writing code with deferred locks is to acquire locks for a task being
+convention in writing code with deferred locks is to acquire locks on behalf of a task being
 launched.  Listing~\ref{lst:lockapi} illustrates this with a simple example.  The
@@
 precondition on the event corresponding to {\tt SUB\_TASK} task's completion.  The
-sub-task can run safely while holding the lock and the lock is released when the task
-sub-task completes.  Using deferred locks in this style minimizes the time
-compute resources spend waiting on locks by ensuring tasks only run when their needed
+sub-task can run safely while holding the lock and the lock is released when the
+sub-task completes.  Using deferred locks in this style prevents 
+compute resources from waiting on locks by ensuring tasks only run when their needed
 locks have already been acquired.
+
+Between deffered locks and barriers (described in Section~\ref{subsec:events}) programmers
+have access to the same set of synchronization primitves that they traditionally have
+in both threading and bulk-synchronous interfaces, but in a distributed asynchronous
+environment.  Furthermore these operations can now be composed with other asynchronous
+operations which is not possible in any other interface.
+
+%Deferred locks provide a super-set of the functionality of blocking locks.  As their
+%name suggests, deferred locks can use the event corresponding to their lock acquire
+%operation to defer execution until the lock acquire has been granted.  Deferred
+%locks can also be converted back into a blocking lock by immediately waiting on the event
+%returned from a call to {\tt lock}.  In addition, deffered locks also provide {\tt mode} and
+%and {\tt exclusive} parameters that allow the user to specify whether other requests can acquire
+%the lock simultaneously.  A lock can only be acquired in one mode at a time.  The
+%{\tt exclusive} parameter specifies whether other owners are permitted once the
+%current lock request is granted.
 
@@
 {\em Physical regions} are the mechanism for reasoning about
-the placement and movement of data in the runtime.  Physical regions are an 
+the placement and movement of data in the interface.  Physical regions are an 
 allocation of data in a single memory in the memory hierarchy.  Listing~\ref{lst:regionapi} 
 shows a subset of the physical region interface.  The first object in the interface
-is a {\tt RegionMetaData} object (line 1).  {\tt RegionMetaData} objects provide the interface for the
+is a {\tt RegionMetaData} object (line 1).  Region meta-data objects provide the interface for the
 creation of sets of physical regions that possess the same number and type of elements (but not necessarily
-the same data layout).  The runtime is only able
+the same data layout).  The interface is only able
 to copy data between two physical regions that were created by the
-same {\tt RegionMetaData} object.  A runtime error will be generated if a copy is attempted
-between two physical regions that were not created by the same {\tt RegionMetaData}.
+same region meta-data object.  A runtime error will be generated if a copy is attempted
+between two physical regions that were not created by the same meta-data object.
 
@@
 When creating a physical region the programmer must specify the {\tt Memory}
-in which the physical region is going to be allocated (line 10).  Once the physical region is created it
+in which the physical region is going to be allocated (line 10).  Memories are described
+in more detail in Section~\ref{subsec:machmodel}.  Once the physical region is created it
 cannot be moved.   There is no coherence of data between physical regions created by the 
-same {\tt RegionMetaData}.  It is the programmer's responsiblity to manage data coherence
+same meta-data object.  It is the programmer's responsiblity to manage data coherence
 using copies between physical regions (line 22-23).  Just like other operations,
-copies can be predicated on an {\tt Event} and return an {\tt Event} corresponding to
-completion.  A copy between two physical regions can only be performed if copies
-are permitted between the two memories where the regions were created.  
-The legality of copies can be discovered using queries to the {\tt Machine} object
-described in Section~\ref{subsec:machmodel}.
+copies can be predicated on an event and return an event corresponding to
+completion.  
+%A copy between two physical regions can only be performed if copies
+%are permitted between the two memories where the regions were created.  
+%The legality of copies can be discovered using queries to the {\tt Machine} object
+%described in Section~\ref{subsec:machmodel}.
 
 The data held inside a physical region is accessed by another kind of object called an
-{\em instance accessor} (not shown).  Accessors support read, write, and reduction operations
+{\em accessor} (not shown).  Accessors support read, write, and reduction operations
 to elements inside the physical region.  Accessors can be specialized for the case where
@@
 accessors supply the level of indirection necessary to convert operations into remote 
-memory operations (RMAs).  %An example of this is described in more detail in Section~\ref{sec:impl}.
+memory operations (RMAs) if supported by the underlying hardware.  
+%An example of this is described in more detail in Section~\ref{sec:impl}.
 
@@
 lines 13-14 takes a {\em reduction operation} to be associated with a physical instance which
-tells the runtime to create a reduction instance.  Reduction instances can only be accessed
+tells the interface to create a reduction instance.  Reduction instances can only be accessed
 using a reduction of the given reduction type; any other accesses will result in a runtime
 error.  A reduction operation is a functor which must have at least one method of the type
-$(T_1 \rightarrow T_2 \rightarrow T_1)$ which takes an element of type $T_1$ and a value
-to be reduced of type $T_2$ and creates and new element.  Reduction operations can also
+$(T_1 \rightarrow T_2 \rightarrow T_1)$ which takes a region element of type $T_1$ and a value
+to be reduced of type $T_2$ and creates and new region element.  Reduction operations can optionally 
 support two additional methods: a method that supplies an initial value of type $T_2$ and a 
 fold method of type $(T_2 \rightarrow T_2 \rightarrow T_2)$.  The presence of these methods
-enable additional runtime optimizations for reductions described in Section~\ref{sec:impl}.
+enable additional optimizations for reductions described in Section~\ref{subsec:reducimpl}.
 
@@
 It is only legal to copy a reduction instance to a regular physical instance or another
-reduction instance.  Any attempt to copy from a regular physical instance to a reduction
-instance will result in a runtime error.  A copy from a reduction instance to a regular
-physical instance will apply all the reductions in the reduction buffer to the physical
-instance.  If a copy is from one reduction instance to another, the two reduction buffers
-will be merged.  The semantics of a merge are dependent upon the underlying implementation
-of reduction buffers which is discussed in more detail in Section~\ref{sec:impl}.
+reduction instance of the same reduction operator.  Any attempt to copy from a regular physical 
+instance to a reduction instance will result in a runtime error.  A copy from a reduction 
+instance to a regular physical instance will apply all the reductions in the reduction 
+buffer to the physical instance.  If a copy is from one reduction instance to another, 
+the two reduction buffers will be merged.  The semantics of a merge are dependent upon 
+the underlying implementation of reduction buffers which is discussed in more detail 
+in Section~\ref{subsec:reducimpl}.
 
@@
 \end{lstlisting}
+
+\subsection{Machine Interface}
+\label{subsec:machmodel}
+The last component of the interface allows for introspection of the current machine.
+Listing~\ref{lst:machineapi} shows the machine interface.
+The {\tt Machine} object is a singleton object that serves two purposes.  First, it
+is the mechanism for initializing the runtime; at the start of a program the programmer
+creates a machine object and invokes the {\tt run} method (line 30-31).
+Second, the machine object provides an interface for the programmer to 
+discover properties of the underlying machine (lines 35-43).  The machine object maintains
+sets of all the unique {\tt Processor} and {\tt Memory} handles in the system.  Processors
+were described in Section~\ref{subsec:procs}.  For every memory in the system there also 
+exists a {\tt Memory} handle with a unique ID.  Different memories often, but not always, 
+imply a different address space.  Examples of different memories are described in
+Section~\ref{sec:impl}.
+
+%For
+%example, in a large cluster each node would have a different {\tt Memory} handle
+%corresponding to each node's DRAM memory.
+%However, if the cluster supported multiple GPUs then there would be a different
+%{\tt Memory} handle for each GPU's {\em zero-copy} memory which shares part of the 
+%address space with the corresponding node memory.  Memories will be used for describing
+%the placement of data described in Section~\ref{subsec:phyreg}.
+
+%The machine interface 
+%contains three different object types: {\tt Processor} (line 1), {\tt Memory} (line 17), and a 
+%singleton {\tt Machine} object (line 22).  For every processor (i.e. CPU core or GPU) in 
+%the system there exists a {\tt Processor} handle with a unique ID naming that processor.  
+%{\tt Processor} objects support a single {\tt spawn} operation (line 13-14) that will
+%launch a {\em task} on that processor.
+
+The machine object also maintains two relations between the set of processors
+and the set of memories:
+\begin{itemize}
+\item Processor-Memory: for each processor-memory pair whether
+the processor can directly access (read and write) the memory and 
+latency and bandwidth properties if access is possible
+\item Memory-Memory: for every pair of memories whether a copy can be
+performed between the two memories and latency and
+bandwidth properties if a copy is possible
+\end{itemize}
+
+The programmer can query these two relations by the {\tt get\_*\_affinity} calls
+(lines 40-43).
+
+%These calls populate a vector with affinity structures (not shown)
+%that contain latency and bandwidth information.  The programmer can restrict the query
+%by specifying a specific processor or memory, otherwise information is returned
+%about all processors and memories in the system.
+
+\begin{lstlisting}[float={t},label={lst:machineapi},caption={Machine Interface.}]
+
+class Memory {
+public:
+  const unsigned id;
+};
+
+class Machine {
+public:
+  Machine(int *argc, char ***argv,
+          const Processor::TaskIDTable &task_table);
+  enum RunStyle {
+    ONE_TASK_ONLY,ONE_TASK_PER_NODE,
+    ONE_TASK_PER_PROCESSOR,
+  };
+  void run(Processor::TaskFuncID task_id = 0, 
+        RunStyle = ONE_TASK_ONLY, const void *args, size_t arglen);
+public:
+  static Machine* get_machine(void); // pointer to Machine
+  // References to all processors and memories
+  const set<Memory>& get_all_memories(void) const;
+  const set<Processor>& get_all_processors(void) const;
+  Processor::Kind get_processor_kind(Processor p) const;
+  size_t get_memory_size(Memory m) const;
+  // Query properties of underlying hardware
+  int get_proc_mem_affinity(vector<ProcMemAffinity &result,
+          Processor restrict_proc = 0, Memory restrict_mem = 0);
+  int get_mem_mem_affinity(vector<MemMemAffinity> &result,
+          Memory restrict_mem1 = 0, Memory restrict_mem2 = 0);
+};
+\end{lstlisting}

### 2012-07-20 doc/asplos2013/related.tex
Adding some related work and updating the bibliography

\section{Related Work}
\label{sec:related}

MPI is the current industry standard for programming large
supercomputers\cite{MPI}.  While MPI supports asynchronous
communication operations there is no mechanism for composing
asynchronous operations.  Co-array Fortran, UPC, and Titanium
are array based languages that implement bulk-synchronous
programming models similar to MPI that allow for asynchronous
exchanges of data\cite{COARRAY_FORTRAN,UPC99,JV:Yel98}.  Like
MPI however, none of these languages allow for general composition
of asynchronous tasks, communication, and synchronization.

Both POSIX threads and OpenMP\cite{OPENMP98} are used for intra-node parallel
programming in large machines, but neither support any asynchronous
operations.  CUDA\cite{CUDA} and OpenCL\cite{Khronos:OpenCL} support
the compostition of asynchronous kernel launches and asynchronous copies
between a host node and a single GPU.  However, the only synchronization
options available in both interfaces are blocking operations on either
a stream or a work queue.  OpenCL has events similar to our interface for
expressing ordering, but has no synchronization primitive for expressing
more relaxed properties such as atomicity that can be described by a
deferred lock.  OpenCL events are not valid globally, but can only be
used with a single GPU context.

Cilk is a parallel programming model that has demonstrated the power of
asynchronous function calls\cite{CILK95}.  The Cilk {\em work-first}
principle provides a compelling theoretical argument for parallel programming
interfaces such as ours that optimize for throughput at the potential
expense of adding additional latency to the critical 
dependence path\cite{Frigo98}.

Chapel\cite{Chamberlain:Chapel} and X10\cite{X1005} are high-level parallel
programming languages that support asynchronous operations.  The constructs
introduced in these languages are higher-level and have complex 
semantics.  We view our interface as providing a potential target
for implementing these higher-level asynchronous operations in an efficient
manner similar to Legion\cite{Legion12}.

The implementation of our interface shares many similarities with large
distributed systems.  Many distributed system implement a publish/subscribe
abstraction for supporting communication that is very similar to how
we manage events and event waiters\cite{Aguilera99,Carzaniga01}.  Work has
also been done on using object oriented languages to build 
distributed systems \cite{Eugster01,Harrison97,Chang91}.  In these cases
callbacks are registered to run when event operations are triggered
from remote locations.

### 2012-07-20 doc/asplos2013/intro.tex
updates

--- 2012-07-17 doc/asplos2013/intro.tex
@@
 \label{sec:intro}
-Introduction goes here\cite{Keeneland}.
+
+% Big machines -> latencies are growing
+% Processors can't be blocked, need to be able to continue handling work
+% Operations need to be composable
+% deferred execution model
+
+% Contributions:
+% - An event system where every operation can take an event and returns an event
+% - Handles are globally valid
+% - More flexible synchronization primitive for this environment: deferred locks
+% - Physical regions for data movement including special versions for reductions
+
+% Parallel programs have three parts
+%  Parallel control
+%  Data movement in the memory hierarchy
+%  Synchronization
+
+% Bulk-synchronous
+% asynchronous components are not composable
+
+A characteristic feature of current and future heterogeneous
+supercomputers is the enormous, variable and growing latencies between
+components.  The well-known mechanism for masking unpredictable or
+long latencies is to use {\em asynchronous operations} that do not block,
+thereby enabling other useful work to be done while the asynchronous
+operation completes. Current implementation primitives for parallel
+programming systems, however, rely on blocking constructs that were
+designed for machines many orders of magnitude
+smaller\cite{MPI,COARRAY_FORTRAN,UPC99}, and these primitives expose
+latency on today's large distributed, heterogeneous parallel machines.
+
+Most current programming systems for large-scale parallel processing provide some variant
+of the {\em bulk-synchronous} execution model\cite{Valiant90}.  Bulk-synchronous models
+partition computations into {\em phases} that are either parallel computation, data
+communication, or synchronization.  At any given point in time all processors
+are engaged in the same kind of phase.  Bulk-synchrony
+makes it easy to reason about parallel execution but
+is very difficult to hide the increasingly long-latency communication
+or memory movement phases.  
+%In older machines this was not a problem because slower
+%processors and smaller machines meant computation phases were much larger than
+%communication or synchronization phases.  However, today's machines are much
+%larger with faster processors, causing an increasing percentage of time
+%to be spent in the communication and synchronization phases.
+
+To combat this problem, parallel programming libraries have introduced
+asynchronous constructs allowing programmers to overlap phases.  For
+example, MPI has asynchronous {\em send} and {\em receive} operations
+for sending messages in parallel with computation \cite{MPI}.  There
+are two problems with this approach.  First, not all phases support
+asynchronous constructs.  Second, the asynchronous operations do not
+compose: There is no way to express that one asynchronous computation
+should begin as soon as another asynchronous communication operation
+has completed.  The result is that at some point there must be
+blocking operations to coordinate synchronous and asynchronous phases.
+If blocking calls are not accurately placed in the program they can
+lead to processor stalls that expose latency.  Correct placement of
+blocking calls can be dependent on input data, algorithmic decisions
+like blocking factors, and the underlying hardware.  To hide as much
+latency as possible and to minimize processor stalls, regardless of
+inputs, algorithmic choices, or underlying hardware, all three major
+aspects of parallel programs (computation, data movement,
+synchronization) should be asynchronous and composable.
+
+We propose a new low-level interface for programming modern
+supercomputers that is fully asynchronous. In our system, one can
+asynchronously compose computation, data movement, and
+synchronizatioan.  The basic mechanism that enables compositionality
+is an underlying {\em event} primitive.  Events provide a mechanism
+for naming a point in the future when an asynchronous operation
+(computation, data copy, or synchronization) will complete.  Every
+call to perform an operation $o$ in our runtime returns immediately
+with an event that {\em triggers} when $o$ has completed.
+Furthermore, every call to perform an operation $o$ can take as a
+precondition an event that must trigger before $o$ can begin.  Using
+events programmers can compose arbitrary chains of dependent
+operations, which the runtime is free to execute in any way that
+respects the event dependences.  With this freedom the runtime can
+schedule operations to optimize throughput while hiding the long
+latency operations.
+
+The following sections describe our contributions:
+\begin{itemize}
+\item We present an interface for asynchronously composing
+parallel computation, data movement, and synchronization using events.  
+To support synchronization in an event system we introduce {\em deferred locks}, 
+a novel synchronization.   Our interface also supports efficient
+mechanisms for performing distributed reduction operations (Section~\ref{sec:interface}).
+
+\item Our approach depends on the efficient handling of very large numbers of events.
+We describe a distributed implementation of our interface that passes events by value,
+requires no global coordination, and has modest local storage costs.  We
+describe optimizations for the implementation of events, deferred locks, and
+reductions (Section~\ref{sec:impl}).
+
+\item We report on micro-benchmarks that are designed to stress
+our implementation.  The results indiciate that our implementation approaches the performance of
+the underlying hardware (Section~\ref{sec:micro}).
+
+\item We demonstrate that our interface is powerful enough to support Legion \cite{}, a higher-level
+programming system,  and that real Legion applications programmed to our interface
+are capable of achieving high-performance (Section~\ref{sec:apps}).  Furthermore, we demostrate
+that the same Legion applications running on a modified version of our interface that enforces
+bulk-synchronous execution incur a performance penalty of XX-XX\%.
+
+\end{itemize}
+
+Finally, Section~\ref{sec:related} describes related work and in Section~\ref{sec:conclusion}
+concludes.

### 2012-07-20 runtime/shared_lowlevel.cc
Fixing a deadlock in the shared lowlevel runtime

@@ -895,7 +895,7 @@ namespace RegionRuntime {
 	Lock get_lock(void) const;
     private:
 	Event register_request(unsigned m, bool exc, TriggerHandle handle = 0);
-	void perform_unlock();
+	void perform_unlock(std::set<EventImpl*> &to_trigger);
     private:
 	class LockRecord {
 	public:
@@ -1037,6 +1037,7 @@ namespace RegionRuntime {
 	PTHREAD_SAFE_CALL(pthread_mutex_lock(mutex));
         log_lock(LEVEL_DEBUG,"unlock request: lock=%x mode=%d excl=%d event=%x/%d count=%d",
                  index, mode, exclusive, wait_on.id, wait_on.gen, holders);
+        std::set<EventImpl*> to_trigger;
 	if (wait_on.exists())
 	{
 		// Register this lock to be unlocked when the even triggers	
@@ -1046,25 +1047,32 @@ namespace RegionRuntime {
 		{
 			// The event didn't register which means it already triggered
 			// so go ahead and perform the unlock operation
-			perform_unlock();
+			perform_unlock(to_trigger);
 		}	
 	}
 	else
 	{
 		// No need to wait to perform the unlock
-		perform_unlock();		
+		perform_unlock(to_trigger);		
 	}
 	PTHREAD_SAFE_CALL(pthread_mutex_unlock(mutex));
+        // Don't perform any triggers while holding the lock's lock
+        for (std::set<EventImpl*>::const_iterator it = to_trigger.begin();
+              it != to_trigger.end(); it++)
+        {
+          (*it)->trigger();
+        }
     }
 
     void LockImpl::trigger(unsigned count, TriggerHandle handle)
     {
+        std::set<EventImpl*> to_trigger;
 	PTHREAD_SAFE_CALL(pthread_mutex_lock(mutex));
         // If the trigger handle is 0 then unlock the lock, 
         // otherwise find the lock request to wake up
         if (handle == 0)
         {
-          perform_unlock();
+          perform_unlock(to_trigger);
         }
         else
         {
@@ -1088,8 +1096,7 @@ namespace RegionRuntime {
                 {
                   holders++;
                   // Trigger the event saying we have the lock
-                  EventImpl *impl = Runtime::get_runtime()->get_event_impl(it->event);
-                  impl->trigger();
+                  to_trigger.insert(Runtime::get_runtime()->get_event_impl(it->event));
                   // Remove the request
                   requests.erase(it);
                 }
[…]

### 2012-07-20 doc/asplos2013/legion-llr.tex
Getting the asplos paper off the ground
Working on the interface section

@@
 
-\bibliographystyle{plainnat}
 \usepackage{amsmath}
@@
 \usepackage{cancel}
+\usepackage{graphics}
+\usepackage{url}
 
@@
 
-\title{Asynchronous High-Performance Language Primitives for Distributed Heterogeneous Architectures}
+\title{High-Performance Language Primitives for Distributed Architectures using Asynchronous Events}
 \numberofauthors{3}
-\author{}
+\author{
+\alignauthor Sean Treichler\\
+             \affaddr{Stanford University} \\
+             \email{sjt@cs.stanford.edu}
+\alignauthor Michael Bauer \\
+             \affaddr{Stanford University} \\
+             \email{mebauer@cs.stanford.edu}
+\alignauthor Alex Aiken \\
+             \affaddr{Stanford University} \\
+             \email{aiken@cs.stanford.edu}
+}
 \maketitle
@@
 %\end{tabular}
-Modern parallel architectures are composed of both heterogeneous processors
-and complex distributed memory hierarchies.  We present a low-level programming
-interface that provides simple primitives for targeting this class of
-machines.  Our interface includes mechanisms for the creation and movement
-of data through the memory hierarchy as well as a novel synchronization primitive 
-called a deferred lock.  To hide the large latencies associated with the target
-set of machines the interface is made asynchronous via an event system.  All
+We present a low-level runtime interface for current highly distributed, heterogeneous
+parallel machines that provides a fully asynchronous 
+programming model, allowing all types of runtime actions, including synchronization,
+to be non-blocking.  Our interface includes mechanisms for the creation and movement
+of data through the memory hierarchy as well as {\em deferred locks}, a novel synchronization primitive.
+Asynchrony is exposed by the the interface using a distributed event system.  All
 operations return an event to be triggered when the operation completes.  Events
 can be chained together and used to create dependences between operations.
+
 We describe an implementation of our interface capable of running on
@@
 We also demonstrate that the interface is sufficiently powerful to serve as the foundation 
-of an advanced runtime system.  We measure the performance of three real-world applications
+of an advanced parallel programming system.  We measure the performance of three real-world applications
 that use our system on the Keeneland supercomputer.
@@
 
-\section{Introduction}
-\label{sect:intro}
-Introduction goes here.
-
-\input{outline}
+%\input{outline}
+\input{intro}
+\input{interface}
+\input{impl}
+\input{micro}
+\input{results}
+\input{related}
 
@@
 \small
+\bibliographystyle{abbrv}
 \bibliography{bibliography}

### 2012-07-20 benchmarks/lock_chains/lock_chains.cc
Adding a new benchmark for lock chains

@@ -1,0 +1,320 @@
+
+#include <cstdio>
+#include <cstdlib>
+#include <cassert>
+#include <cstring>
+#include <set>
+#include <time.h>
+#include <stdlib.h>
+
+#include "lowlevel.h"
+
+using namespace RegionRuntime::LowLevel;
+
+// TASK IDs
+enum {
+  TOP_LEVEL_TASK = Processor::TASK_ID_FIRST_AVAILABLE+0,
+  MAKE_LOCKS_TASK = Processor::TASK_ID_FIRST_AVAILABLE+1,
+  RETURN_LOCKS_TASK = Processor::TASK_ID_FIRST_AVAILABLE+2,
+  LAUNCH_CHAIN_CREATION_TASK = Processor::TASK_ID_FIRST_AVAILABLE+3,
+  ADD_FINAL_EVENT_TASK = Processor::TASK_ID_FIRST_AVAILABLE+4,
+  DUMMY_TASK = Processor::TASK_ID_FIRST_AVAILABLE+5,
+};
+
+struct InputArgs {
+  int argc;
+  char **argv;
+};
+
+struct FairStruct {
+  Processor orig;
+  Lock lock;
+  Event precondition;
+  int depth;
+};
+
+// forward declaration
+void fair_locks_task(const void *args, size_t arglen, Processor p);
+
+InputArgs& get_input_args(void)
+{
+  static InputArgs args;
+  return args;
+}
+
+std::set<Event>& get_final_events(void)
+{
+  static std::set<Event> final_events;
+  return final_events;
+}
+
+std::set<Lock>& get_lock_set(void)
+{
+  static std::set<Lock> lock_set;
+  return lock_set;
+}
+
+Processor get_next_processor(Processor cur)
+{
+  Machine *machine = Machine::get_machine();
+  const std::set<Processor> &all_procs = machine->get_all_processors();
+  for (std::set<Processor>::const_iterator it = all_procs.begin();
+        it != all_procs.end(); it++)
+  {
+    if (*it == cur)
+    {
+      // Advance the iterator once to get the next, handle
+      // the wrap around case too
+      it++;
+      if (it == all_procs.end())
+      {
+        return *(all_procs.begin());
+      }
+      else
+      {
+        return *it;
+      }
+    }
+  }
+  // Should always find one
+  assert(false);
+  return Processor::NO_PROC;
+}
+
+void shutdown(void)
+{
+  Machine *machine = Machine::get_machine();
+  const std::set<Processor> &all_procs = machine->get_all_processors();
[…]

### 2012-07-21 doc/asplos2013/intro.tex
More modifications to the intro

@@
 A characteristic feature of current and future heterogeneous
-supercomputers is the enormous, variable and growing latencies between
+supercomputers is the large, variable, and growing latencies between
 components.  The well-known mechanism for masking unpredictable or
@@
 thereby enabling other useful work to be done while the asynchronous
-operation completes. Current implementation primitives for parallel
+operations complete. Current implementation primitives for parallel
 programming systems, however, rely on blocking constructs that were
 designed for machines many orders of magnitude
-smaller\cite{MPI,COARRAY_FORTRAN,UPC99}, and these primitives expose
+smaller\cite{MPI,COARRAY_FORTRAN,UPC99}; these primitives expose
 latency on today's large distributed, heterogeneous parallel machines.
@@
 makes it easy to reason about parallel execution but
-is very difficult to hide the increasingly long-latency communication
+hinders attempts to hide the increasingly long-latency communication
 or memory movement phases.  
@@
 asynchronous constructs.  Second, the asynchronous operations do not
-compose: There is no way to express that one asynchronous computation
+compose.  There is no way to express that one asynchronous computation
 should begin as soon as another asynchronous communication operation
@@
 like blocking factors, and the underlying hardware.  To hide as much
-latency as possible and to minimize processor stalls, regardless of
-inputs, algorithmic choices, or underlying hardware, all three major
-aspects of parallel programs (computation, data movement,
+latency as possible and to minimize processor stalls, 
+%regardless of inputs, algorithmic choices, or underlying hardware,
+all three major aspects of parallel programs (computation, data movement,
 synchronization) should be asynchronous and composable.
@@
 asynchronously compose computation, data movement, and
-synchronizatioan.  The basic mechanism that enables compositionality
+synchronization.  The basic mechanism that enables composition of operations
 is an underlying {\em event} primitive.  Events provide a mechanism
@@
 (computation, data copy, or synchronization) will complete.  Every
-call to perform an operation $o$ in our runtime returns immediately
-with an event that {\em triggers} when $o$ has completed.
-Furthermore, every call to perform an operation $o$ can take as a
-precondition an event that must trigger before $o$ can begin.  Using
+call to perform an operation $op$ in our runtime returns immediately
+with an event that {\em triggers} when $op$ has completed.
+Furthermore, every call to perform an operation $op$ can take as a
+precondition an event that must trigger before $op$ can begin.  Using
 events programmers can compose arbitrary chains of dependent
@@
 \item We present an interface for asynchronously composing
-parallel computation, data movement, and synchronization using events.  
-To support synchronization in an event system we introduce {\em deferred locks}, 
-a novel synchronization.   Our interface also supports efficient
-mechanisms for performing distributed reduction operations (Section~\ref{sec:interface}).
+parallel computation, data movement, and synchronization using events (Section~\ref{sec:interface}).
+
+\item To support synchronization in our event system we introduce {\em deferred locks}, 
+a novel synchronization primitive that operates in an asynchronous enviroment (Section~\ref{subsec:locks}).
+
+\item We describe a physical region system for the management of data that abstracts
+data layouts.  Abstracted data layouts enable composable reduction operations that can be
+deferred to further hide latency (Section~\ref{subsec:phyreg}).
 
 \item Our approach depends on the efficient handling of very large numbers of events.
-We describe a distributed implementation of our interface that passes events by value,
-requires no global coordination, and has modest local storage costs.  We
-describe optimizations for the implementation of events, deferred locks, and
-reductions (Section~\ref{sec:impl}).
+We describe a distributed implementation of our interface that requires no global coordination
+and has modest local storage costs.  We describe optimizations for the implementation of events, 
+deferred locks, and reductions (Section~\ref{sec:impl}).
 
@@
 
-\item We demonstrate that our interface is powerful enough to support Legion \cite{}, a higher-level
+\item We demonstrate that our interface is powerful enough to support Legion \cite{Legion12}, a higher-level
 programming system, and that real Legion applications programmed to our interface
 are capable of achieving high-performance (Section~\ref{sec:apps}).  Furthermore, we demostrate
-that the same Legion applications running on a modified version of our interface that enforces
-bulk-synchronous execution incur a performance penalty of XX-XX\%.
-
+that the same applications written in a bulk-synchronous model incur performance penalites between 22-135\%.
 \end{itemize}

### 2012-07-21 doc/asplos2013/interface.tex
Major edits to the interface section

@@
 \label{sec:interface}
+The target set of clients for our interface are higher-level language
+compilers and runtimes such as Legion\cite{Legion12} as well as 
+advanced systems programmers.  This class of users demand total control
+over the underlying hardware and transparent performance from an
+implementation.  To support these demands we have designed our interface
+to be as low-level as possible.  In many cases we have chosen to trade-off
+programmability for performance.  We eschewed any features that would
+automatically be performed without being directed by the client.
+Our goal in the design of this interface is to provide a set of low-level,
+high-performance primitives that are asynchronous and composable.
 
@@
 (Section~\ref{subsec:events}), processors for parallel computation 
-(\ref{subsec:procs}), deffered locks for synchronization
-(\ref{subsec:locks}), physical regions for data movement
-(\ref{subsec:phyreg}), and an object for introspection of the underlying
+(\ref{subsec:procs}), deferred locks for synchronization
+(\ref{subsec:locks}), physical regions for data layout and movement
+(\ref{subsec:phyreg}), and a machine object for introspection of the underlying
 hardware (\ref{subsec:machmodel}).  In all cases (except for the
-machine which is a static singleton), instances of the objects
+machine object which is a static singleton), instances of the objects
 are light-weight {\em handles} that provide a unique name for an underlying
-implementation.  Handles can be copied and passed by value.  Furthermore,
-every handle is valid everywhere in the system.  This property allows 
-a programmer to pass handles by value between computations without
-having to reason about the distributed nature of the system.
-We describe how we support this property in more detail in 
-Section~\ref{sec:impl}.
+implementation.  Multiple copies of handles are permitted and handles can be 
+passed by value.  Furthermore, every handle is valid everywhere in the system.  
+This property allows a client to pass handles by value between computations 
+without having to reason about the distributed nature of the system.
+%We describe how we support this property in more detail in 
+%Section~\ref{sec:impl}.
 
@@
 supports testing whether an event has triggered (line 7) and waiting on
-an event to trigger (line 9).  While these methods can be useful, the most common
-use of events is passing them as preconditions for other operations.  The programmer can use the
-{\tt merge\_events} call (line 11) to create an event corresponding to the 
+an event to trigger (line 9).  While these methods can be useful, the encouraged
+use of events is passing them as preconditions for other operations.  The client can use the
+{\tt merge\_events} call (lines 11-12) to create an event corresponding to the 
 conjunction of a set of events.
@@
 operations and the implementation is responsible for triggering these events.  Users 
-can create a special type of event called a {\tt UserEvent} (line 14) that the
+can create a special type of event called a {\tt UserEvent} (line 15) that the
 user has the power to trigger.  A user event is a sub-type of an event
-that has a {\tt trigger} method (line 17).  Users can also create another
+that has a {\tt trigger} method (line 18).  Users can also create another
 type of event called a {\tt Barrier} that will require multiple arrivals before
@@
 provide a superset of the functionality of traditional barriers since they can
-be used both in a blocking manner (via the {\tt wait} call), but can also be used
+be used in a blocking manner (via the {\tt wait} call), but are primarily used
 asynchronously.  Lastly, since user events and barriers are sub-types of 
@@
   void wait() const;
-  // Merge events together to a new event
+  // Merge events together to create a new event
+  static Event merge_events(Event ev1, Event ev2);
   static Event merge_events(const set<Event> &to_merge);
@@
 Listing~\ref{lst:procapi} shows the interface for {\tt Processor} objects which allow for
-the creation of parallel computations.  Processor objects provide a way of naming 
-every computational unit in the machine.  In the case of our current implementation 
+the creation of parallel computations.  Processor provide a way of naming 
+every computational unit in the machine.  We describe how to find processor handles
+in Section~\ref{subsec:machmodel}.  In the case of our current interface 
 processors are either individual cores or discrete GPUs, but this can be easily modified 
@@
 is invoked on a processor handle, it is possible to launch a task on any
-processor from anywhere in the system.  To allow for the chaining of dependences
-through events the spawn call takes an 
-optional event that must trigger before the task can begin.  
-The result of the spawn call is an event that will trigger when the task completes
-which permits the creation of dependences upon this task.
+processor from anywhere in the system.  
+The spawn call takes an optional event that must trigger before the task can begin.  
+The spawn call returns an event that will trigger when the task completes.
 
-\begin{lstlisting}[float={t},label={lst:procapi},caption={Processor Interface.}]
+Lines 17-23 of Listing~\ref{lst:procapi} show a simple example using 
+the processor and event interface.  The {\tt diamond\_task} launches four
+sub-tasks with a diamond dependence pattern.  The {\tt merge\_events} call
+is used to merge the events {\tt eb} and {\tt ec} to create the shared dependence
+for the last sub-task.
+
+\begin{lstlisting}[float={t},label={lst:procapi},caption={Processor Interface and Example.}]
 class Processor {
@@
 };
+
+void diamond_task(const void *args,size_t arglen,Processor p){
+  Event ea = p.spawn(TASK_A,NULL,0);
+  Event eb = p.spawn(TASK_B,NULL,0,ea);
+  Event ec = p.spawn(TASK_C,NULL,0,ea);
+  Event em = Event::merge_events(eb,ec);
+  Event ed = p.spawn(TASK_D,NULL,0,em);
+}
 \end{lstlisting}
@@
 Events and barriers express ordering properties between operations, but in many
-cases ordering is too strict.  For many applications access to data must only be atomic and
-not necessarily ordered.  Traditionally locks have been used to synchronize access
-to data.  However, all lock implementations that we are aware of require either blocking
-or spinning on locks, neither of which composes well with other asynchronous operations.
+cases ordering is too strict of a requirement.  For many applications access to data must only be atomic and
+not necessarily ordered.  Traditionally locks have been used to serialize access to
+data.  However, all lock implementations that we are aware of require either blocking
+or spinning, neither of which composes well with asynchronous operations.
 Deferred locks are a new synchronization mechanism that allows for synchronization
@@
 lock.  A lock request is issued and the resulting {\tt lock\_event} is used as
-a precondition for launching the {\tt SUB\_TASK} task.  The unlock operation is then
-precondition on the event corresponding to {\tt SUB\_TASK} task's completion.  The
-sub-task can run safely while holding the lock and the lock is released when the
-sub-task completes.  Using deferred locks in this style prevents 
+a precondition for launching the {\tt SUB} task.  The unlock operation is then
+precondition on the event corresponding to {\tt SUB} task's completion.  The
+{\tt SUB} task can run safely while holding the lock and the lock is released when the
+{\tt SUB} task completes.  Using deferred locks in this style prevents 
 compute resources from waiting on locks by ensuring tasks only run when their needed
@@
 
-Between deffered locks and barriers (described in Section~\ref{subsec:events}) programmers
+Between deffered locks and barriers (described in Section~\ref{subsec:events}) clients 
 have access to the same set of synchronization primitves that they traditionally have
-in both threading and bulk-synchronous interfaces, but in a distributed asynchronous
-environment.  Furthermore these operations can now be composed with other asynchronous
-operations which is not possible in any other interface.
+in both threading and bulk-synchronous interfaces.  However, these operations can now 
+be composed with other asynchronous operations which is not possible in any other interface.
+of which we are aware.
 
@@
   Event lock_event = needed.lock();
-  // Launch task
-  Event task_event = p.spawn(SUB_TASK,NULL,0,lock_event);
+  // Launch sub-task
+  Event task_event = p.spawn(SUB,NULL,0,lock_event);
   // Release lock
@@
 \label{subsec:phyreg}
+In distributed machines with discrete memories, data movement often consists of more
+than a simple copy operation.  Many applications perform operations and transformations on
+their data in conjunction with data movement for higher performance.  One common example is 
+that applications accumulate reduction operations in a buffer and as part of a copy operation 
+the reduction operations are applied to a destination buffer on the receiving side of the copy.  
+In order to support these kinds of conjoined operations in an asynchronous environment, our interface 
+must be aware of the structure of data.  To describe the structure of data our interface 
+uses {\em physical regions.}
 
-{\em Physical regions} are the mechanism for reasoning about
-the placement and movement of data in the interface.  Physical regions are an 
-allocation of data in a single memory in the memory hierarchy.  Listing~\ref{lst:regionapi} 
-shows a subset of the physical region interface.  The first object in the interface
-is a {\tt RegionMetaData} object (line 1).  Region meta-data objects provide the interface for the
-creation of sets of physical regions that possess the same number and type of elements (but not necessarily
-the same data layout).  The interface is only able
-to copy data between two physical regions that were created by the
-same region meta-data object.  A runtime error will be generated if a copy is attempted
-between two physical regions that were not created by the same meta-data object.
+A physical region is an allocation of data in a single memory in the memory hierarchy.  Physical
+regions are grouped into {\em classes} which are sets of physical regions that share the
+same number and size of elements, but don't necessarily share the same data layout.  Our
+interface supports sub-classing, but the details are omitted due to space constraints.  The
+interface only supports data movement operations between physical regions that are of
+the same class (or between super- and sub-classes).
+% Say something about dynamically modifying the number of elements in a class
 
-An instance of a physical region is represented by a {\tt RegionInstance} object (line 17).
-When creating a physical region the programmer must specify the {\tt Memory}
-in which the physical region is going to be allocated (line 10).  Memories are described
-in more detail in Section~\ref{subsec:machmodel}.  Once the physical region is created it
-cannot be moved.   There is no coherence of data between physical regions created by the 
-same meta-data object.  It is the programmer's responsiblity to manage data coherence
-using copies between physical regions (line 22-23).  Just like other operations,
-copies can be predicated on an event and return an event corresponding to
-completion.  
+Listing~\ref{lst:regionapi} shows a subset of the interface for physical regions.  The
+first object in the interface is a {\tt RegionMetaData} object (line 1).  Physical
+regions created by the same region meta-data object belong to the same class.  Physical
+regions are represented by {\tt RegionInstance} objects (line 17).  When creating a physical
+region the client must specify the {\tt Memory} in which the physical region is going
+to be allocated.  Memories are described in more detail in Section~\ref{subsec:machmodel}.
+If there is not enough space for the physical region allocation a {\tt NO\_INST} will
+be returned.  There is no automatic coherence of data between physical regions in the
+same class.  It is the client's responsiblity to manage data coherence using copies
+between physical regions (line 22-23).  Just like other operations copies can be
+predicated on an event and return an event corresponding to completion.
+
+Clients can also associate an operation as part of a physical region.  One example of
+associating an operation with a physical region is shown on lines 13-14.  This method
+will create a physical region called a {\em reduction instance} that will be assoicated 
+with the given {\tt ReductionOp}.  The interface for a {\tt ReductionOp} is shown on 
+lines 26-33.  A reduction operation must have an {\tt apply} method, and can optionally 
+implement two additional methods, {\tt init} and {\tt fold} that support additional 
+optimizations described in Section~\ref{subsec:reducimpl}.  The two template parameters 
+describe the base element type of the region ({\tt LHS}:left-hand-side) and the type of 
+the argument to the reduction ({\tt RHS}:right-hand-side).  The reduction operation has 
+total control over the choice of data layout for the reduction instance (interface not shown).
+Reduction instances can be copied to other reduction instances with the same operation
+or to regular physical regions of the same class.
+
+An example of using reduction instances is shown on lines 35-49 of Listing~\ref{lst:regionapi}.
+In this example, two reduction instances, {\tt inst1} and {\tt inst2}, are created in different memories to
+be used with reduction operation 2 (lines 40-41).  A physical region of the same class {\tt inst3}
+is also created in a different memory (line 42).  Two tasks are launched that will apply
+reductions locally into their respective reduction buffers (lines 43-44).  The resulting
+events are then used to chain copies back to {\tt inst3} after the tasks complete (45-46).  As
+part of these copies the reductions stored in {\tt inst1} and {\tt inst2} will be applied
+to {\tt inst3}.  Since reductions are done atomically, there is no race between the reduction
+applications.  Finally, a merge on the events from the movement and reduction applications is 
+performed and a final task using {\tt inst3} is launched (lines 47-48).
+
+Reductions and reduction instances illustrate only a single case where physical regions provide 
+the structure necessary to conjoin an operation with data movement.  Physcal regions make our
+interface easily extensible to arbitrary transformations and operations as a part of data movment.
+While physical regions are a slightly higher-level construct than an annonymous
+buffer of bits, we believe that the optimizations they enable are important in an asynchronous
+environment.  Clients also still have access to an interface for controlling data layout through
+the operations associated with physical regions which mitigates the loss in generality.
+
+%{\em Physical regions} are the mechanism for reasoning about
+%the placement and movement of data in the interface.  Physical regions are an 
+%allocation of data in a single memory in the memory hierarchy.  Listing~\ref{lst:regionapi} 
+%shows a subset of the physical region interface.  The first object in the interface
+%is a {\tt RegionMetaData} object (line 1).  Region meta-data objects provide the interface for the
+%creation of sets of physical regions that possess the same number and type of elements (but not necessarily
+%the same data layout).  The interface is only able
+%to copy data between two physical regions that were created by the
+%same region meta-data object.  A runtime error will be generated if a copy is attempted
+%between two physical regions that were not created by the same meta-data object.
+
+%An instance of a physical region is represented by a {\tt RegionInstance} object (line 17).
+%When creating a physical region the programmer must specify the {\tt Memory}
+%in which the physical region is going to be allocated (line 10).  Memories are described
+%in more detail in Section~\ref{subsec:machmodel}.  Once the physical region is created it
+%cannot be moved.   There is no coherence of data between physical regions created by the 
+%same meta-data object.  It is the programmer's responsiblity to manage data coherence
+%using copies between physical regions (line 22-23).  Just like other operations,
+%copies can be predicated on an event and return an event corresponding to
+%completion.  
 %A copy between two physical regions can only be performed if copies
@@
 
-The data held inside a physical region is accessed by another kind of object called an
-{\em accessor} (not shown).  Accessors support read, write, and reduction operations
-to elements inside the physical region.  Accessors can be specialized for the case where
-all memory is known to be directly accessible to the executing processor, which allows 
-for direct references to elements.  In cases where not all memory is directly accessable, 
-accessors supply the level of indirection necessary to convert operations into remote 
-memory operations (RMAs) if supported by the underlying hardware.  
+%The data held inside a physical region is accessed by another kind of object called an
+%{\em accessor} (not shown).  Accessors support read, write, and reduction operations
+%to elements inside the physical region.  Accessors can be specialized for the case where
+%all memory is known to be directly accessible to the executing processor, which allows 
+%for direct references to elements.  In cases where not all memory is directly accessable, 
+%accessors supply the level of indirection necessary to convert operations into remote 
+%memory operations (RMAs) if supported by the underlying hardware.  
 %An example of this is described in more detail in Section~\ref{sec:impl}.
 
-One special (but common) case for applications is when reductions need to be performed.
-For reductions it is usually more efficient to store reduction
-operations in a local reduction buffer and then merge reduction buffers together at a later
-point in time.  To support this case the interface allows for the creation of a special kind
-of physical instance called a {\em reduction instance}.  The call to {\tt create\_instance} on
-lines 13-14 takes a {\em reduction operation} to be associated with a physical instance which
-tells the interface to create a reduction instance.  Reduction instances can only be accessed
-using a reduction of the given reduction type; any other accesses will result in a runtime
-error.  A reduction operation is a functor which must have at least one method of the type
-$(T_1 \rightarrow T_2 \rightarrow T_1)$ which takes a region element of type $T_1$ and a value
-to be reduced of type $T_2$ and creates and new region element.  Reduction operations can optionally 
-support two additional methods: a method that supplies an initial value of type $T_2$ and a 
-fold method of type $(T_2 \rightarrow T_2 \rightarrow T_2)$.  The presence of these methods
-enable additional optimizations for reductions described in Section~\ref{subsec:reducimpl}.
+%One special (but common) case for applications is when reductions need to be performed.
+%For reductions it is usually more efficient to store reduction
+%operations in a local reduction buffer and then merge reduction buffers together at a later
+%point in time.  To support this case the interface allows for the creation of a special kind
+%of physical instance called a {\em reduction instance}.  The call to {\tt create\_instance} on
+%lines 13-14 takes a {\em reduction operation} to be associated with a physical instance which
+%tells the interface to create a reduction instance.  Reduction instances can only be accessed
+%using a reduction of the given reduction type; any other accesses will result in a runtime
+%error.  A reduction operation is a functor which must have at least one method of the type
+%$(T_1 \rightarrow T_2 \rightarrow T_1)$ which takes a region element of type $T_1$ and a value
+%to be reduced of type $T_2$ and creates and new region element.  Reduction operations can optionally 
+%support two additional methods: a method that supplies an initial value of type $T_2$ and a 
+%fold method of type $(T_2 \rightarrow T_2 \rightarrow T_2)$.  The presence of these methods
+%enable additional optimizations for reductions described in Section~\ref{subsec:reducimpl}.
 
-Reduction instances can be copied using the same interface as regular physical instances.
-It is only legal to copy a reduction instance to a regular physical instance or another
-reduction instance of the same reduction operator.  Any attempt to copy from a regular physical 
-instance to a reduction instance will result in a runtime error.  A copy from a reduction 
-instance to a regular physical instance will apply all the reductions in the reduction 
-buffer to the physical instance.  If a copy is from one reduction instance to another, 
-the two reduction buffers will be merged.  The semantics of a merge are dependent upon 
-the underlying implementation of reduction buffers which is discussed in more detail 
-in Section~\ref{subsec:reducimpl}.
+%Reduction instances can be copied using the same interface as regular physical instances.
+%It is only legal to copy a reduction instance to a regular physical instance or another
+%reduction instance of the same reduction operator.  Any attempt to copy from a regular physical 
+%instance to a reduction instance will result in a runtime error.  A copy from a reduction 
+%instance to a regular physical instance will apply all the reductions in the reduction 
+%buffer to the physical instance.  If a copy is from one reduction instance to another, 
+%the two reduction buffers will be merged.  The semantics of a merge are dependent upon 
+%the underlying implementation of reduction buffers which is discussed in more detail 
+%in Section~\ref{subsec:reducimpl}.
 
-\begin{lstlisting}[float={t},label={lst:regionapi},caption={Subset of Physical Region Interface.}]
+\begin{lstlisting}[float={t},label={lst:regionapi},caption={Physical Region Interface and Example.}]
 class RegionMetaData {
@@
 };
+
+template<typename LHS, typename RHS>
+class ReductionOp {
+public:
+  void apply(LHS *lhs, RHS rhs);
+  // Optional methods for optimizations
+  void init(RHS *rhs, size_t num_elmts);
+  void fold(RHS *rhs1, RHS rhs);
+};
+
+void reduction_ex(const void *args,size_t arglen,Processor p) {
+  Memory m1,m2,m3; Processor p1,p2,p3; ReductionOpID redop = 2;
+  ... // Choose memory and processors
+  RegionMetaData meta = RegionMetaData::
+                        create_region(num_elmts,elmt_size);
+  Instance inst1 = meta.create_instance(m1,redop);
+  Instance inst2 = meta.create_instance(m2,redop);
+  Instance inst3 = meta.create_instance(m3);
+  Event t1 = p1.spawn(REDUC,&inst1,sizeof(RegionInstance));
+  Event t2 = p2.spawn(REDUC,&inst2,sizeof(RegionInstance));
+  Event c1 = inst1.copy_to(inst3,t1);
+  Event c2 = inst2.copy_to(inst3,t2);
+  Event e0 = Event::merge_events(c1,c2);
+  Event t3 = p3.spawn(USE,&inst3,sizeof(RegionInstance),e0);
+}
 \end{lstlisting}

### 2012-07-21 doc/asplos2013/interface.tex
Major edits to the interface section

@@
 The {\tt Machine} object is a singleton object that serves two purposes.  First, it
-is the mechanism for initializing the runtime; at the start of a program the programmer
-creates a machine object and invokes the {\tt run} method (line 30-31).
-Second, the machine object provides an interface for the programmer to 
+is the mechanism for initializing the runtime; at the start of a program the client 
+creates a machine object, instantiates a task table mapping task IDs to function pointers,
+and invokes the {\tt run} method (line 30-31).
+Second, the machine object provides an interface for the client to 
 discover properties of the underlying machine (lines 35-43).  The machine object maintains
@@
 latency and bandwidth properties if access is possible
-\item Memory-Memory: for every pair of memories whether a copy can be
-performed between the two memories and latency and
-bandwidth properties if a copy is possible
+\item Memory-Memory: for every pair of memories whether copies can be
+performed and latency and bandwidth properties if copies are possible
 \end{itemize}
 
-The programmer can query these two relations by the {\tt get\_*\_affinity} calls
+The client can query these two relations by the {\tt get\_*\_affinity} calls
 (lines 40-43).

### 2012-07-21 doc/asplos2013/intro.tex
Small edits to the intro

@@
 
-\item We describe a physical region system for the management of data that abstracts
-data layouts.  Abstracted data layouts enable composable reduction operations that can be
-deferred to further hide latency (Section~\ref{subsec:phyreg}).
+\item We describe a physical region system that provides structural information about
+data layout.  The structural information provided by physical regions enables an extensible
+interface for conjoining operations and transformations with asynchronous data movement for 
+further lantecy hiding (Section~\ref{subsec:phyreg}).
 
@@
 
-Finally, Section~\ref{sec:related} describes related work and in Section~\ref{sec:conclusion}
+Section~\ref{sec:related} describes related work and Section~\ref{sec:conclusion}
 concludes.

### 2012-07-21 doc/asplos2013/interface.tex
minor edits

@@
   Event c2 = inst2.copy_to(inst3,t2);
-  Event e0 = Event::merge_events(c1,c2);
-  Event t3 = p3.spawn(USE,&inst3,sizeof(RegionInstance),e0);
+  Event em = Event::merge_events(c1,c2);
+  Event t3 = p3.spawn(USE,&inst3,sizeof(RegionInstance),em);
 }

### 2012-07-21 doc/asplos2013/intro.tex
minor edits

@@
 precondition an event that must trigger before $op$ can begin.  Using
-events programmers can compose arbitrary chains of dependent
+events clients can compose arbitrary chains of dependent
 operations, which the runtime is free to execute in any way that

### 2012-07-21 doc/asplos2013/related.tex
Updating related work

@@
 communication operations there is no mechanism for composing
-asynchronous operations.  Co-array Fortran, UPC, and Titanium
+asynchronous operations.  GASNet is another interface for programming
+large clusters that centers around active messages\cite{GASNET07}.  GASNet
+is a part of the heterogeneous implementation of our interface, but like
+MPI, does not support composition of asynchronous active messages with
+other constructs.  Co-array Fortran, UPC, and Titanium
 are array based languages that implement bulk-synchronous
@@
 
+The Threaded Abstract Machine (TAM) is a programming model designed to
+make it easy to port dataflow programs onto a parallel machine\cite{CullerGSvE93}.  
+Conceptually our interface is very similar to the dataflow languages TAM supported.
+The implementation of TAM is designed for a class of much smaller machines
+and therefore is based on much heavier communication and threading mechanisms.
+While TAM supported composable asynchronous computation and communication, its
+synchronization mechanisms still required blocking operations.
+
 Chapel\cite{Chamberlain:Chapel} and X10\cite{X1005} are high-level parallel
@@
 distributed systems.  Many distributed system implement a publish/subscribe
-abstraction for supporting communication that is very similar to how
+abstraction for supporting communication that is similar in some ways to how
 we manage events and event waiters\cite{Aguilera99,Carzaniga01}.  Work has
 also been done on using object oriented languages to build 
-distributed systems \cite{Eugster01,Harrison97,Chang91}.  In these cases
+distributed systems\cite{Eugster01,Harrison97,Chang91}.  In these cases
 callbacks are registered to run when event operations are triggered
-from remote locations.
+from remote locations.  Events in these systems are much heavier weight
+and often carry large data payloads.  In many distributed event systems
+the focus is on resiliency instead of performance\cite{Ostrowski09}.

### 2012-07-21 doc/asplos2013/interface.tex
Minor edits

@@
 the creation of parallel computations.  Processor provide a way of naming 
-every computational unit in the machine.  We describe how to find processor handles
+every computational unit in the machine.  We describe how to discover processor handles
 in Section~\ref{subsec:machmodel}.  In the case of our current interface 
-processors are either individual cores or discrete GPUs, but this can be easily modified 
+processors are either individual cores or discrete GPUs, but this is easily modified 
 to include new processor types (line 10).   Processors support a single {\tt spawn}
@@
 in both threading and bulk-synchronous interfaces.  However, these operations can now 
-be composed with other asynchronous operations which is not possible in any other interface.
+be composed with other asynchronous operations which is not possible in any other interface
 of which we are aware.
@@
 the reduction operations are applied to a destination buffer on the receiving side of the copy.  
-In order to support these kinds of conjoined operations in an asynchronous environment, our interface 
-must be aware of the structure of data.  To describe the structure of data our interface 
-uses {\em physical regions.}
+To support these kinds of conjoined operations in an asynchronous environment, our interface 
+must be aware of the structure of data so it can perform accompanying operations with 
+data movement operations.  To describe the structure of data our interface uses {\em physical regions.}
 
@@
 interface only supports data movement operations between physical regions that are of
-the same class (or between super- and sub-classes).
+the same class (or between super- and sub-classes).  Note, clients can still pass arbitrary arrays
+of bits via the processor {\tt spawn} call.
 % Say something about dynamically modifying the number of elements in a class
@@
 Clients can also associate an operation as part of a physical region.  One example of
-associating an operation with a physical region is shown on lines 13-14.  This method
+associating an operation with a physical region is shown on lines 13-14.  This instance
+of the {\tt create\_instance} method
 will create a physical region called a {\em reduction instance} that will be assoicated 
-with the given {\tt ReductionOp}.  The interface for a {\tt ReductionOp} is shown on 
+with the given {\tt ReductionOp} specified by the ID.  The interface for a {\tt ReductionOp} is shown on 
 lines 26-33.  A reduction operation must have an {\tt apply} method, and can optionally 
@@
 the argument to the reduction ({\tt RHS}:right-hand-side).  The reduction operation has 
-total control over the choice of data layout for the reduction instance (interface not shown).
+total control over the choice of data layout for the reduction instance (not shown).
 Reduction instances can be copied to other reduction instances with the same operation
@@
 
-\subsection{Machine Interface}
+\subsection{Machine}
 \label{subsec:machmodel}
@@
 The client can query these two relations by the {\tt get\_*\_affinity} calls
-(lines 40-43).
+(lines 40-43) to discover properties of the machine.
 

### 2012-07-21 doc/asplos2013/intro.tex
Minor edits

@@
 lead to processor stalls that expose latency.  Correct placement of
-blocking calls can be dependent on input data, algorithmic decisions
-like blocking factors, and the underlying hardware.  To hide as much
+blocking calls can be dependent on input data, algorithmic decisions, 
+and the underlying hardware.  To hide as much
 latency as possible and to minimize processor stalls, 
@@
 data layout.  The structural information provided by physical regions enables an extensible
-interface for conjoining operations and transformations with asynchronous data movement for 
+interface for conjoining operations on data (e.g. reductions) with asynchronous data movement for 
 further lantecy hiding (Section~\ref{subsec:phyreg}).

### 2012-07-22 doc/asplos2013/impl.tex
redid event impl section - on to locks next

--- 2012-07-18 doc/asplos2013/impl.tex
@@
 The implementation for heterogeneous clusters also uses POSIX
-threads as well as the CUDA runtime library for GPUs\cite{CUDA} and the GASNet
+threads\cite{PTHREADS} (commonly referred to as Pthreads) as 
+well as the CUDA runtime library for GPUs\cite{CUDA} and the GASNet
 API for large clusters\cite{GASNET07}.
@@
 models a cluster with CPUs and GPUs as having two types of processors 
-and four types of memory.  Every CPU core is a different CPU processor
-and every GPU is a GPU processor.  For every node in the system
-there is a memory corresponding to the DRAM associated with that node.  For
-every GPU in the system there are two memories corresponding to the framebuffer
-memory and the zero-copy memory associated with that GPU \footnote{Zero-copy 
-memory contains pages that are mapped in both framebuffer and node memory with
-coherence maintained by the PCI-E protocol.}. The last memory is 
-a global GASNet memory that represents pages that have been registered 
-with all nodes in the system for supporting remote memory accesses (RMAs).
-The GASNet memory provides the illusion of global memory.  CPU processors
-can directly access their node memory, all zero-copy memories for GPUs on
-their node, and the GASNet memory.  GPU processors can only access their
-framebuffer memory and their zero-copy memory.  Copies are permitted
-between all pairs of memory except between any framebuffer memory and
-the GASNet memory.
+and four types of memory.  Every CPU core is presented as a different CPU processor
+and every GPU is a GPU processor.  This matches the scheduling granularity available
+in the Pthreads and CUDA APIs respectively.  The first type of memory is the 
+system memory that is accessible by every CPU core on a given node.  The second is the 
+framebuffer memory on each GPU (and accessible only by that GPU).  The third type
+of memory is system memory on a node that has been made accessible to the GPU(s) on
+that node as well as the CPU cores, commonly referred to as {\em zero-copy memory}.
+The final type of memory is the portion of system memory on each node that has been registered with
+the GASNet runtime to allow {\em remote memory access} (RMA) by other nodes in the cluster.
+%% For every node in the system
+%% there is a memory corresponding to the DRAM associated with that node.  For
+%% every GPU in the system there are two memories corresponding to the framebuffer
+%% memory and the zero-copy memory associated with that GPU \footnote{Zero-copy 
+%% memory contains pages that are mapped in both framebuffer and node memory with
+%% coherence maintained by the PCI-E protocol.}. The last memory is 
+%% a global GASNet memory that represents pages that have been registered 
+%% with all nodes in the system for supporting remote memory accesses (RMAs).
+%% The GASNet memory provides the illusion of global memory.  CPU processors
+%% can directly access their node memory, all zero-copy memories for GPUs on
+%% their node, and the GASNet memory.  GPU processors can only access their
+%% framebuffer memory and their zero-copy memory.  Copies are permitted
+%% between all pairs of memory except between any framebuffer memory and
+%% the GASNet memory.
 
-To support inter-node operations our implementation relies on GASNet active
-messages for communication.  Active messages are a command and a payload
-that are sent remotely and cause a handler to be run on the target node.
-A simple example of our use of active messages can be illustrated by the 
-{\tt spawn} call on a {\tt Processor} object.  Consider that {\tt spawn}
-is invoked on a processor that is on a remote node from the node on which
-the currently running task is executing.  This operation is converted into
-an active message that contains all the information about the task launch.
-When the active message is handled by the remote node, the task and all its
-metadata are registered with the processor on which the task was launched.
-While most of the implementation of our heterogeneous runtime is straight
-forward, the next three sections describe in greater detail the nuances
-of the implementation of events, locks, and reduction instances.
+In addition to the RMA capabilities, GASNet provides {\em active messages} for 
+inter-node communication.  Active messages consist of a command and payload that
+are sent by one node to another without any previous coordination.  Upon arrival
+at the destination node, a handler routine is invoked to process the payload.
+The handler may issue a response to the sender, but this is optional, and avoided
+as much as possible in our implementation due to the latencies involved in waiting
+for a response.  As an example, consider the {\tt spawn} call on a {\tt Processor}
+object.  If the processor is not on the same node as the requestor, the information
+about the task (e.g. the task ID and arguments, the actual target {\tt Processor},
+the prerequisite event (if any), and the completion event ID) is sent in an
+active message to the node containing the target processor.  The recipient node
+unpacks the message, and either places the task on the ready queue for the target
+processor or marks it as being deferred until the prerequisite event has triggered.
+No response is sent to the original requestor.  The requestor already knows the event
+that corresponds to the completion of this task (since it was provided in the active
+message) - there is no utility in knowing exactly when the active message itself has
+been processed.
+
+%% To support inter-node operations our implementation relies on GASNet active
+%% messages for communication.  Active messages are a command and a payload
+%% that are sent remotely and cause a handler to be run on the target node.
+%% A simple example of our use of active messages can be illustrated by the 
+%% {\tt spawn} call on a {\tt Processor} object.  Consider the case where {\tt spawn}
+%% is invoked on a processor that is on a remote node.  This operation is converted into
+%% an active message that contains all the information about the task launch.
+%% When the active message is handled by the remote node, the task and all its
+%% metadata are registered with the processor on which the task was launched.
+%% While most of the implementation of our heterogeneous runtime is straight
+%% forward, the next three sections describe in greater detail the nuances
+%% of the implementation of events, locks, and reduction instances.
+
+%% Our implementation uses a hierarchical model, in which data structures are shared by
+%% all threads on the same node (taking advantage of the shared address space and low
+%% latencies of using Pthreads mutexes and/or lock-free data structures).
+
+Due to space constraints, we will not cover every part of our implementation in
+detail, instead limiting our discussion to the three most interesting aspects:
+events, locks, and reduction instances.
 
@@
 \label{subsec:eventimpl}
+
+Events may be created on demand, and are always {\em owned} by the node on which the
+creation request was made.  The creation of an event occurs with no inter-node communcation.
+The set of event
+handles is divided across the nodes at startup time, allowing each node to assign handles
+to new events without fear of conflicting with another node's events.  No broadcast of
+the event creation is required either, as the static division is sufficient to allow any
+node to correctly determine an event's owning node when (and only if) an operation on that
+event is performed on the other node.
+
+At event creation time, the owning node allocates a data structure to track the state of the event
+(i.e. triggered or not) as well as record a list of the operations that are known to be dependent
+on the event (e.g. initiation of a copy operation, placing a task on a processor's ready queue, waking
+up a waiting task).  However, the owning node only keeps a list of the dependent operations from the same
+node.  Every other node also allocates a corresponding data structure (the first time the event
+is referenced) to remember if the event has already triggered and to record its local dependent operations.
+Arbitrarily many dependent operations from a
+single node are aggregated into a single {\em event subscription} active message that is sent to the
+owning node.  The owning node keeps a bitmask of which other nodes have subscribed to the event, and when the
+event finally does trigger, a single {\em event trigger} active message is sent to each subscribing node,
+which notes that the event has triggered (in case queries come after the trigger) and executes the list
+of local operations that were dependent on that event.  (In the case that the triggering of the event has
+occurred before the reception of an event subscription, a trigger message is sent immediatel to the 
+new subscriber.)
+
+The actual triggering of an event may occur on any node.  If it occurs on a node other than the owning
+node, an {\em event trigger} active message is sent from the triggering node back to the owning node, which
+then forwards that message to all subscribed nodes, with the exception of the triggering node, which was
+able to execute its locally dependent operations (if any) immediately.  Although this can often result in
+the latency of a triggering operation including two active messages, it bounds the number of active
+messages required per event to $2N$ where $N$ is the number of nodes (which can be much smaller than the
+number of dependent operations) interested in the event.  An alternative
+would have been to share the subscriber bitmask so that the triggering node can notify all interested
+nodes directly, but such an algorithm is much more complicated due to race conditions, and requires $O(N^2)$
+active messages.  As we will see in section~\ref{sec:microbenchmarks}, the latency of a single extra
+active message is very small, and any cost that is super-linear with the node count either is or will soon
+be an scalability issue for large systems.
+
+The data structures used to track an event cannot be freed until all operations on that event have been
+performed.  Creation and triggering can happen only once each, but there can be any number of operations
+that are dependent on an event, and some of those operations may not be requested until long after the
+event has triggered.  Other systems incorporating events address this by reference counting event
+handles\cite{Khronos:OpenCL}, but such reference counting adds both programmer and runtime overhead even when
+limited to a single node -- further overhead and complexity would be expected for a cluster-level
+reference-counting approach.
+
+Instead of focusing on freeing event data structures, our implementation aggressively recycles them, 
+needing fewer total event data structures than a referencing counting implementation while also eliminating
+the overhead of reference counting.  The key observation is that one {\em generational event} data structure
+ can efficiently capture the state of one {\em untriggered} event and a very large number (e.g. $2^{32}-1$)
+of already-triggered events.  An event's handle is expanded to include its generation number as well as the
+identifier for the underlying generational event.  In addition to the triggered-or-not state and list of
+dependent operations
+for the most recent generation, the owning node also remembers how many previous generations have already
+been triggered.  (There is no need to keep a list of dependent operations on previous generations.  Any
+new operation that is dependent on a generation that is known to be already triggered can immediately be
+executed.)  When an event is created, any generational event for whom the most recent generation has 
+triggered can be reused - the generation is increased by one and the state is returned to ``untriggered.''
+As before, this can be done with no inter-node communication.
+
+A remote node's data structure is also efficient - the boolean values for whether the event
+has triggered and whether the event has already been subscribed to are replaced with the numbers of the most
+recent generation known to have triggered (this is received in the event trigger active message) and of the
+most recent generation the remote node has subscribed to (this is sent in the event subscription active
+message).  And as with the owner node's data structure, only
+one list of dependent operations is needed.  Although the latencies of a large system can delay the
+reception of an event trigger active message, if an operation is created that depends on a later generation
+of an event than what the current list of events is dependent on, that serves as a roundabout indication that
+the previous generation has indeed triggered, allowing the existing list of operations to be executed.
+
+%% Events are created on demand by each node.  The ID of the creating node is encoded
+%% in the upper bits of each event's ID and the creating node is said to be the 
+%% event's {\em owner}.  If an event is used as an argument to an
+%% operation on a remote node then the remote node can determine the owner of 
+%% the event and send an active message to become a {\em subscriber} of the event.  When a
+%% remote node becomes a subscriber to an event it guarantees that it will receive an active
+%% message from the event's owner when the event triggers.  To minimize the amount
+%% of communication that occurs between nodes, each node remembers the events to which it
+%% is subscribed and guarantees that it only subscribes to an event once.  In the case
+%% where there may be many remote waiters on an event 
+%% this can dramatically reduce the number of inter-node active messages.
+
+%% When an event triggers, it automatically notifies all of the operations that are waiting
+%% on its local node.  The owner node will also send active messages to all of the subscriber nodes
+%% telling them that the event has triggered.  Even though the event has now notified all of
+%% its waiters, the memory on the owning node for storing the event cannot be reclaimed 
+%% because there still may be handles to the event in the system which will request to
+%% use this event in the future.  These requests will need to be satisfied saying that 
+%% the event has already triggered for the duration of the application.  Rather than waste
+%% the resources, we recycle event implementations.
+
+%% A {\em dynamic event} is a logical construct that will only be triggered once and is the level
+%% of abstraction at which the programmer reasons about events.  A {\em physical event} is an
+%% actual event implementation that consumes resources on its owner node.  Multiple dynamic events
+%% can be mapped to a single physical event.  However, for each physical event there can be at most
+%% one {\em active} event in the set of dynamic events that are mapped to it.  An active event is a 
+%% dynamic event that is yet to trigger.  By only having one active dyanmic event at a time 
+%% there is no need to disambiguate event triggers that are sent to the physical event.  Each
+%% physical event maintains a count of the number of trigger operations that it has observed.
+
+%% When the runtime needs to return a handle corresponding to a new dynamic event, it finds
+%% a physical event which currently has no active events (e.g. the trigger count equals the number
+%% of dynamic events mapped to the physical event).  The runtime increases the count
+%% of the number of dynamic events mapped to the physical event.  The handle that is returned 
+%% corresponds to a new dynamic event which contains an ID that refers to the physical
+%% event it is mapped to and a {\em generation} that is the dynamic event number for the particular
+%% physical event.  Note that by definition the generation will always be one greater than the
+%% trigger count at the time of dynamic event creation.
+
+%% Within this framework it is now very easy to test whether the dynamic event specified by a
+%% handle has triggered.  Using the ID that is contained within the handle we find the
+%% physical event implementation (which may require an active message if the owner node is
+%% remote from where the test is begin performed).  We can then compare the generation of
+%% the handle to the number of observed triggers contained by the physical event.  If the
+%% number of physical triggers is greater than or equal to the handle's generation, then
+%% the dynamic event has triggered.  We show in Section~\ref{sec:apps} that in practice
+%% the number of needed physical events is significantly less than the number of dynamic events.
+
+%% We also leverage the mapping of dynamic events onto physical events to improve the efficiency
+%% of subscriptions.  Nodes cache the last generation for every physical event for which they
+%% have seen a trigger notification.  If the trigger test described above detects a trigger based
+%% on the cached information then there would be no need for the subscription which would save an
+%% active message.  If the event has not been detected to have
+%% triggered locally then an active message is sent to the owner which would always have been necesary
+%% without the dynamic to physical event mapping.
 

### 2012-07-22 doc/asplos2013/impl.tex
first draft of implementation, playing around a bit in interface too

@@
 
+Like events, locks are created on demand by the node on which the creation request was made.  The lock
+handles are also statically divided across the nodes, allowing the creation to be done without inter-node
+communication.  Unlike events whose ownership is static, locks may migrate, so although the {\em creating}
+node is the initial owner of a lock, that ownership can be transferred to other nodes.
+
+Locks are often used to mediate access to data, and the size of that data is often relatively small.  To
+avoid adding the latency of inter-node data copies to critical sections, a lock may also be created with a
+{\em payload}, a small (i.e. less than 4KB - the limit for an active message's payload) piece of data that is moved around with the lock and guaranteed to be coherent if
+the lock is held.
+
+Since any node may at some point be the owner of a lock, all nodes use the same data structure to track the
+state of the lock (they differ only in when the structure is allocated -
+nodes other than the creator still wait until the first reference to the lock on that node).  The structure
+tracks the following:
+\begin{itemize}
+\item {\em owner node} - the most recently known owner of the lock.  If the current node is the owner, this
+information is always accurate.  If not, this information may be stale, but at least the recorded ``owner''
+is sure to have more up-to-date information.
+\item {\em lock status} - records whether the lock is currently held.  Only the current lock owner has
+guaranteed-current information here.
+\item {\em local waiters} - a list of pending local lock requests.  This data is always valid as well.
+\item {\em remote waiters} - a bitmask of which other nodes have pending requests.  This bitmask is only
+used when the current node is the owner of the lock.
+\item {\em local payload pointer and size} - the local node's copy of the lock payload
+\end{itemize}
+
+When a lock request is made, an event is created to track when the grant occurs.  The current node then
+examines its copy of the lock data structure (creating it if
+this the first operation for that lock on that node).  If the current node is the owner and the lock isn't
+held, the lock is granted immediately and the event is triggered.  If the lock is held, the event is added
+to the list of local waiters.  (This is the only thing that is stored.  The actual ``requestor'' of the lock
+is implicitly captured in which operations are dependent on the lock grant event.)  If the current node
+isn't the owner, a {\em lock request} active message is
+sent to the most recently known owner.  (If that information proves to be stale, the request is forwarded
+by the receiving node, eventually catching up to the current owner.)  If the current owner's status shows
+the lock is currently held, the requesting node is added to the remote waiters bitmask.  If the lock is
+not held, the ownership of the lock is given to the requesting node via a {\em lock transfer} active
+message.  A lock transfer message includes the current remote waiters bitmask (in case other nodes are
+have also requested the lock) and an up-to-date copy of the lock's payload.
+
+Similarly, an unlock request is first checked against the local node's lock state.  If the local node is
+not the owner, an {\em unlock} active message is sent to the most recently known owner, which is forwarded
+if necessary.  Once the unlock request is on the lock's current owning node, the local waiter list is
+examined.  If the list is non-empty, the lock remains in the ``locked'' state and the first lock grant
+event is pulled off the local waiter list and triggered.  If instead the local waiter list is empty, the
+lock state is changed to ``unlocked'', and the remote waiter bitmask is examined.  If any bits are set, then
+one of them is chosen and the corresponding node becomes the new owner via a lock transfer active message.
+
+The unfairness that results from a lock favoring local waiters over remote waiters is intentional.  When the
+contention on a lock is high (the only time when the question of fairness is even relevant), the latencies
+involved in transferring a lock between nodes can be the limiter on throughput.  By minimizing the number
+of lock transfers, the throughput on the lock is maximized.
+
+A similar issue can arise with the forwarding of lock request active messages.  If two nodes are
+transferring a lock very quickly back and forth, a third node's request could be forwarded an arbitrary
+number of times as it chases the lock around.  Although this can result in unfairness, it doesn't 
+prevent forward progress.  A lock is only transferred to a node with at least one local waiter, so if 
+the third node's request ever became the only one left, the lock transferring would stop and the request
+would finally catch up.
+
 \subsection{Reduction Instances}
 \label{subsec:reducimpl}
+
+In addition to normal physical regions, our implementation supports two different kinds of reduction-only
+physical regions, designed to optimize the common pattern in which multiple tasks make commutative 
+{\em reductions} into shared data.  Instead of having to move the existing data to a suitable memory and
+then applying individual reduction operations to the existing data 
+right away (as would happen in a normal physical region), reduction instances allow a large number of
+reduction operations to be batched up and transferred efficiently to the original location of the data.
+This decoupling from the original location of the data can allow allow greater parallelism
+in cases where there are available processors that cannot all access the same memory.
+
+The first kind of reduction-only physical region is a {\em reduction fold instance}, which may only be
+used for reduction operations that support a {\em fold} operation.  A reduction fold instance is similar
+to a normal physical region in that it is implemented as an array, indexed by the same element indices used
+for the normal physical regions.  The difference is that each array element is a value of the reduction's
+{\tt RHS} type instead of the actual element type of the region (i.e. the {\tt LHS} type).  When the client
+requests a reduction operation to a particular location, the supplied right-hand-side value is folded in
+to the corresponding location in the array.  An arbitrary number of reductions can be done to each location.
+When the reduction fold instance is copied back to a normal instance, the contents of the reduction instance
+(which are generally the same size or smaller than the normal region contents) are transferred in a large
+block to the location of the normal instance, where an application of the reduction function is performed
+once for each location in the region via a cache-friendly linear sweep over the memory.
+
+The second kind of reduction instance is a {\em reduction list instance}.  Although it generally doesn't
+perform as well as a reduction fold instance, it can be better for cases with sparse updates.  It is also
+the only option when a reduction operation doesn't support a {\em fold} operation.  Instead of folding
+together multiple reductions to the same location, a reduction list instance keeps a log of every
+reduction operation requested (i.e. the location and the right-hand-side value).  When the reduction list 
+instance is copied back to a normal physical region, the whole list is transferred and then replayed at
+the target region's location.  Except in sparse update cases, this list is often larger than the target
+region, and any randomness in the sequence of locations in the list can impact the memory access efficiency
+of the accesses to the target region.  However, as we will show in section~\ref{sec:microbenchmarks}, 
+the benefits of enabling parallelism and hiding latency usually more than make up for these drawbacks.

### 2012-07-22 doc/asplos2013/interface.tex
first draft of implementation, playing around a bit in interface too

@@
   const unsigned id;
-  Event lock(unsigned mode = 0, bool exclusive, 
-              Event wait_on = Event::NO_EVENT) const;
+  Event lock(Event wait_on = Event::NO_EVENT) const;
   void unlock(Event wait_on = Event::NO_EVENT) const;
   // Create a new lock, destroy an existing lock
-  static Lock create_lock();
+  static Lock create_lock(size_t payload_size = 0);
+  void *payload_ptr();
   void destroy_lock();
@@
 
-Clients can also associate an operation as part of a physical region.  One example of
-associating an operation with a physical region is shown on lines 13-14.  This instance
-of the {\tt create\_instance} method
-will create a physical region called a {\em reduction instance} that will be assoicated 
-with the given {\tt ReductionOp} specified by the ID.  The interface for a {\tt ReductionOp} is shown on 
-lines 26-33.  A reduction operation must have an {\tt apply} method, and can optionally 
-implement two additional methods, {\tt init} and {\tt fold} that support additional 
-optimizations described in Section~\ref{subsec:reducimpl}.  The two template parameters 
-describe the base element type of the region ({\tt LHS}:left-hand-side) and the type of 
-the argument to the reduction ({\tt RHS}:right-hand-side).  The reduction operation has 
-total control over the choice of data layout for the reduction instance (not shown).
-Reduction instances can be copied to other reduction instances with the same operation
-or to regular physical regions of the same class.
+If the only operations that will be performed on locations in a {\tt RegionInstance} object,
+a special reduction version can be created by providing the ID of a {\em reduction operator}
+when the object is created (line 13).  Reduction-only {\tt RegionInstance} objects can 
+increase the available parallelism and hide latency better, as we will show in section~\ref{sec:microbenchmarks}.  The interface supports any reduction operation that
+is commutative (i.e. multiple operations to the same location can be reordered without 
+changing the final result) and is coded as an {\tt apply} method as shown on line 29.
+Many commonly-used reduction operations are also {\em foldable} (e.g. $(l \text{ += } r_1) \text{ += } r_2$ can
+be folded as $l \text{ += } r_1 + r_2$), and if the {\tt fold} operation and a {\em right identity}
+(e.g. $0$ is a right identity for subtraction because $x - 0 = x$ for all $x$) are provided,
+the runtime can use a {\em reduction fold instance}, described in section~\ref{subsec:reducimpl},
+for ever better performance.  Although the way they store data is very different,
+reduction-only instances can be copied to other reduction-only instances (or normal 
+instances) of the same class with the same asynchronous copy operations.
+
+%% Clients can also associate an operation as part of a physical region.  One example of
+%% associating an operation with a physical region is shown on lines 13-14.  This instance
+%% of the {\tt create\_instance} method
+%% will create a physical region called a {\em reduction instance} that will be assoicated 
+%% with the given {\tt ReductionOp} specified by the ID.  The interface for a {\tt ReductionOp} is shown on 
+%% lines 26-33.  A reduction operation must have an {\tt apply} method, and can optionally 
+%% implement two additional methods, {\tt init} and {\tt fold} that support additional 
+%% optimizations described in Section~\ref{subsec:reducimpl}.  The two template parameters 
+%% describe the base element type of the region ({\tt LHS}:left-hand-side) and the type of 
+%% the argument to the reduction ({\tt RHS}:right-hand-side).  The reduction operation has 
+%% total control over the choice of data layout for the reduction instance (not shown).
+%% Reduction instances can be copied to other reduction instances with the same operation
+%% or to regular physical regions of the same class.
 
@@
 public:
-  void apply(LHS *lhs, RHS rhs);
-  // Optional methods for optimizations
-  void init(RHS *rhs, size_t num_elmts);
-  void fold(RHS *rhs1, RHS rhs);
+  static void apply(LHS *lhs, RHS rhs);
+  // Optional methods for folding optimization
+  static void fold(RHS *rhs1, RHS rhs2);
+  static const RHS right_identity;
 };

### 2012-07-22 doc/asplos2013/micro.tex
First draft of the micro benchmark section

@@
-\section{Microbenchmark Evaluation}
+\section{Benchmark Evaluation}
 \label{sec:micro}
+
+In this section we evaluate the heterogeneous implementation of our 
+interface using several small micro-benchmarks designed to illustrate
+that our implementation approaches the underlying performance of the
+hardware, and that the optimizations described in Section~\ref{sec:impl}
+are important.  All exeperiments are run on the Keeneland
+supercomputer\cite{Keeneland}.  Each Keeneland node is composed of
+two Xeon 5660 CPUs, three Tesla M2090 GPUs, and 24 GB of DRAM.  Nodes
+are connected by an Infiniband QDR interconnect.
+
+\subsection{Event Latency and Throughput}
+\label{subsec:eventmicro}
+We implemented two micro-benchmarks for evaluating the performance of
+events.  The first micro-benchmark tests the latency required to trigger
+events both within a node and between nodes.  Processors are organized
+into a ring and the first processor creates a user event and passes it
+to the next processor.  The next processor creates a new event dependent
+on the first, and passes the new event along. A long chain of dependent events 
+is constructed.  The experiment measures the time from the triggering
+of the first event to the triggering of the last event.  In the one node
+configuration there is only one processor and all events are local to the
+node.  In all other cases, there is only one processor per node which
+guarantees that all event triggers result in an active message.  
+Table~\ref{tab:eventlat} shows the results for both configurations.
+
+\begin{table}
+\begin{center}
+{ \small
+\begin{tabular}{m{2cm}|c|c|c|c|c}
+Nodes & 1 & 2 & 4 & 8 & 16 \\ \hline
+Mean Trigger Time ($\mu$s) & 0.329 & 3.259 & 3.799 & 3.862 & 4.013 \\
+\end{tabular}
+}
+\end{center}
+\vspace{-6mm}
+\caption{Event Latency Results.\label{tab:eventlat}}
+\vspace{-4mm}
+\end{table}
+
+In the single node case the latency of triggering an event is only 
+329 nano-seconds.  For the multi-node cases, the triggering time is
+between 3-4 micro-seconds which is the same as the cost of GASNet
+active messages on Infiniband\cite{GASNET07}.  The gradual increase in latency with
+node count is because each node must poll the Infiniband connection with all other nodes.
+This demonstrates that the total cost of performing event triggers is in
+the hardware and the underlying software stack.
+
+Our second micro-benchmark for events measures the maximum throughput of event
+triggers.  This benchmark creates {\em levels} of events where every event at a level
+corresponds to a merge of all of the events at the previous level.  The benchmark is therefore
+characterized by the number events at each level called the fan-in/out factor.  
+The events at each level are evenly distributed between the processors in the machine.
+In the case of an $n$-node experiment, only $n-1$ active messages will be required 
+per event trigger because the event subscription mechanism described in 
+Section~\ref{subsec:eventimpl}.  Figure~\ref{fig:eventthroo} shows the event triggering
+throughput versus total nodes for a range of fan-in/out factors.
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.33]{figs/event_throughput.pdf}
+\end{center}
+\vspace{-6mm}
+\caption{Event Throughput Micro-Benchmark.\label{fig:eventthroo}}
+\vspace{-4mm}
+\end{figure}
+
+For small fan-in/out factors, throughput falls off initially going to two nodes as active
+message become necessary, but starts increasing again at larger node counts.  Higher
+fan-in/out factors require more messages and have lower throughput that increases with
+node count.  In both cases the scaling of throughput is because the benchmark is
+bound by the computational costs of active messages and not the bandwidth of the
+interconnection network.  With more processors, more active messages can be processed
+in parallel and performance improves.  
+%Note we can further improve performance by increasing
+%the number of threads per node for handling active messages (in these experiments there
+%is one per node), but in real applications these threads would consume hardware cores and
+%take away computational resources from the primary application.  This is only a viable
+%option for memory or communication bound programs.
+
+The compute-bound nature of the benchmark demonstrate that active messages do not tax 
+the interconnection network and leave bandwidth available for application data movement.  
+The event throughputs sustained in this micro-benchmark are between one and two orders
+of magnitude larger than the actual throughputs required by the real applications
+described in Section~\ref{sec:apps}.
+
+\subsection{Lock Throughput}
+\label{subsec:lockmicro}
+Our lock micro-benchmark is designed to measure the maximum throughput of lock
+acquisitions.  The benchmark is characterized by the number of nodes, the number
+of locks initially created by each node, and the number of {\em chains} of
+lock requests per node.  A chain is 1024 lock acquire and release operations that
+are serialized by event dependences.  Each node creates the specified number of locks and
+informs all other nodes of its lock handles.  All nodes create the specified number
+of chains by randomly selecting from the set of all locks in the system.  All chains
+are predicated on a user event.  The experiment measures the time from the start of
+the user event to the completion of all chains.
+
+Figure~\ref{fig:fixedlock} shows the lock grant throughput versus nodes in the system
+for a fixed number of total chains in the system (256 total).  Each line corresponds
+to a number of locks per node.  At larger node counts there will be more total locks
+in the system.  At small node counts, small numbers of locks per node cause high
+contention and lock unfairness allows for very high throughput as many requests are
+handled locally before locks are migrated.  Large values have more total locks which
+results in lower contention and lower throughput because more lock migrations are
+required per satisfied request.  At larger node counts, there are enough locks
+in the system to remove contention and throughput increases.
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.33]{figs/fixed_lock_chains.pdf}
+\end{center}
+\vspace{-6mm}
+\caption{Lock Benchmark for Fixed Total Chains.\label{fig:fixedlock}}
+\vspace{-4mm}
+\end{figure}
+
+To demonstrate that lock unfairness is also beneficial at larger node counts we show an alternative
+cut of the experiment space.  In this case we fix the node count at 16 and plot
+grant throughput as a function of chains per node for many different locks per node (also
+total nodes in the system since node count is fixed).
+The results can be seen in Figure~\ref{fig:fixednode}.  At very small chains per node
+there is little contention and throughput is high.  At larger chains per node there is
+more contention for locks and throughput drops because of higher lock migrations.  Note that 
+as soon as chains per node is greater than total locks, the expected number of users per lock 
+per node becomes more than one.  It is exactly at this point on each line that throughput
+begins to improve again!  This is because as soon as there are multiple expected users on
+a node, the unfairness property of the lock can amortize the cost of lock migration which
+yields better throughput.
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.33]{figs/fixed_node_lock.pdf}
+\end{center}
+\vspace{-6mm}
+\caption{Lock Benchmark for Fixed Node Count.\label{fig:fixednode}}
+\vspace{-4mm}
+\end{figure}
+
+\subsection{Reduction Throughput}
+\label{subsec:reducmicro}
+To show the benefits of associating reduction operations with physical regions we designed
+a histogram micro-benchmark that requires all nodes to perform an addition reduction to a
+physical region located in the globally visible GASNet memory (described in 
+Section~\ref{sec:impl}).  Using the reduction interface described in Section~\ref{subsec:phyreg}
+reductions are performed in five ways:
+
+\begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
+\item Single Reads/Writes - Each node takes a lock on the global physical region and then for each reduction
+reads a values from the physical region, performs the reduction, and then write the result back.  The lock is then released.
+\item Single Reductions - The reduction operation implementation performs each reduction individually
+using an underlying active message implementation.
+\item Localized Instance - Each node takes a lock on the the global physical region, copies it to a local physical
+region, performs reductions to the localized region, copies the region back, and releases the lock.
+\item Fold Instance - Each node creates a local reduction instance that folds reductions to the same element.
+Reductions are applied locally and then the local reduction instance is merged back into the global region.
+\item List Instance - Each node creates a local reduction instance that stores a list of reduction operations.
+Reductions are applied locally and then the local reduction instance is merged back into the global region.
+\end{itemize}
+
+We run two experiments, one corresponding with dense reductions and one corresponding with sparse
+reductions.  Each experiment runs 8 reduction tasks per node, one for each processor.  The dense
+case has 256K buckets and each reduction task performs 4M reductions to random locations.  The results for the
+dense experiment are shown in Figure~\ref{fig:reducdense}.  The sparse case has 4M buckets and
+each reduction task performs 64K reductions to random locations.  Results for the sparse case
+be seen in Figure~\ref{fig:reducsparse}.
+
+The dense experiment demonstrates that using folding reduction regions perform 
+best and scale with the number of nodes, achieving over a billion reductions
+per second in the 16 node case.  List regions also scale, but perform about an
+order of mangitude worse than folding regions in the dense case.  Not surprisingly
+the other algorithms all have bottlenecks that result in poor performance.  The
+sparse case is very similar with additional interesting behavior that at 16 nodes
+the list regions begins performing better than the fold regions.  This shows that
+list regions are better suited for scaling sparse reduction operations.
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.33]{figs/reduce_dense.pdf}
+\end{center}
+\vspace{-6mm}
+\caption{Dense Reduction Benchmark.\label{fig:reducdense}}
+\vspace{-4mm}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.33]{figs/reduce_sparse.pdf}
+\end{center}
+\vspace{-6mm}
+\caption{Sparse Reduction Benchmark.\label{fig:reducsparse}}
+\vspace{-4mm}
+\end{figure}

### 2012-07-22 doc/asplos2013/micro.tex
edits to the micro benchmark section

@@
 active messages on Infiniband\cite{GASNET07}.  The gradual increase in latency with
-node count is because each node must poll the Infiniband connection with all other nodes.
-This demonstrates that the total cost of performing event triggers is in
-the hardware and the underlying software stack.
+node count is because each node must poll all the Infiniband connections for other nodes.
+This cost of performing event triggers is in entirely in the hardware and the underlying software stack.
 
@@
 interconnection network.  With more processors, more active messages can be processed
-in parallel and performance improves.  
+in parallel and throughput improves.  
 %Note we can further improve performance by increasing
@@
 Our lock micro-benchmark is designed to measure the maximum throughput of lock
-acquisitions.  The benchmark is characterized by the number of nodes, the number
+grants.  The benchmark is characterized by the number of nodes, the number
 of locks initially created by each node, and the number of {\em chains} of
@@
 of chains by randomly selecting from the set of all locks in the system.  All chains
-are predicated on a user event.  The experiment measures the time from the start of
+are predicated on a user event.  The experiment measures the time from the triggering of
 the user event to the completion of all chains.
@@
 Figure~\ref{fig:fixedlock} shows the lock grant throughput versus nodes in the system
-for a fixed number of total chains in the system (256 total).  Each line corresponds
-to a number of locks per node.  At larger node counts there will be more total locks
-in the system.  At small node counts, small numbers of locks per node cause high
+for 256 total chains in the system.  Each line corresponds
+to a different number of initial locks per node.  At small node counts, 
+small numbers of locks per node cause high
 contention and lock unfairness allows for very high throughput as many requests are
-handled locally before locks are migrated.  Large values have more total locks which
-results in lower contention and lower throughput because more lock migrations are
-required per satisfied request.  At larger node counts, there are enough locks
-in the system to remove contention and throughput increases.
+handled locally before locks are migrated.   When there are more total locks there
+is lower contention and lower throughput because the ratio of lock grants to migrated
+locks decreases.  At larger node counts, there are enough locks
+in the system to remove contention and throughput improves.
 
@@
 cut of the experiment space.  In this case we fix the node count at 16 and plot
-grant throughput as a function of chains per node for many different locks per node (also
-total nodes in the system since node count is fixed).
+grant throughput as a function of chains per node for many different locks per node.
 The results can be seen in Figure~\ref{fig:fixednode}.  At very small chains per node
-there is little contention and throughput is high.  At larger chains per node there is
-more contention for locks and throughput drops because of higher lock migrations.  Note that 
+there is little contention and throughput is high.  As the number of chains per node increases there is
+more contention for locks and throughput drops because of more lock migrations.  Note that 
 as soon as chains per node is greater than total locks, the expected number of users per lock 
 per node becomes more than one.  It is exactly at this point on each line that throughput
-begins to improve again!  This is because as soon as there are multiple expected users on
+begins to improve again.  This is because as soon as there are multiple expected users on
 a node, the unfairness property of the lock can amortize the cost of lock migration which
@@
 \item Single Reads/Writes - Each node takes a lock on the global physical region and then for each reduction
-reads a values from the physical region, performs the reduction, and then write the result back.  The lock is then released.
-\item Single Reductions - The reduction operation implementation performs each reduction individually
+reads a value from the physical region, performs the reduction, and then writes the result back.  The lock is then released.
+\item Single Reductions - The reduction operation performs each reduction individually
 using an underlying active message implementation.
@@
 
-The dense experiment demonstrates that using folding reduction regions perform 
+The dense experiment demonstrates that folding reduction regions perform 
 best and scale with the number of nodes, achieving over a billion reductions
@@
 the other algorithms all have bottlenecks that result in poor performance.  The
-sparse case is very similar with additional interesting behavior that at 16 nodes
-the list regions begins performing better than the fold regions.  This shows that
+sparse case is very similar with the additional interesting occurence that at 16 nodes
+list regions begin performing better than fold regions.  This shows that
 list regions are better suited for scaling sparse reduction operations.

### 2012-07-22 doc/asplos2013/results.tex
Working on the beginning of the results section

@@
 \label{sec:apps}
+Legion is a high-level runtime system that has been implemented on top
+of our interface\cite{Legion12}.  
+%The Legion runtime is distributed so 
+%that there is an independent scheduler for each processor in the machine.  
+The Legion programming model is built around the abstraction of 
+{\em logical regions} which express locality and independence properties 
+of data.  Computation in Legion is organized into a tree of tasks where 
+each task must specify which logical regions will be accessed.
+When executing a Legion task, the Legion runtime receives requests
+to execute sub-tasks along with their region requirements.
+This stream of tasks with region requirements is analogous to a stream
+of instructions with register requirements that are executed by 
+a hardware processor.  Hardware out-of-order processors are designed to
+run ahead of the actual execution of a stream of instructions
+to ensure that the processor's pipeline is fully utilized.  Similarly, on
+every processor of a supercomputer there is an instance of the Legion runtime
+that takes a stream of tasks with region requirements and asynchronously
+runs ahead of the actual execution using a {\em software-out-of-order} (SOOP)
+scheduler.  The details of the SOOP are beyond the scope of this paper
+but are described in \cite{Legion12}.
+
+The fully asynchronous design of the interface that we've proposed in
+this paper is crucial to Legion's SOOP implementation.  Without a fully
+asynchronous interface, the SOOP would have to periodically invoke
+blocking operations that would cause processors to stall and severly
+limit Legion's ability to run-ahead and keep the machine fully
+occupied.  With a fully asynchronous interface the SOOP can run
+far ahead of the actual execution, limited only by the physical resources
+available in the machine, exactly like a hardware out-of-order processor.
+
+In this section we demonstrate the performance properties of three
+real-world applications that use the Legion runtime and consequently the heterogeneous
+implementation of our interface to run on the Keeneland supercomputer.
+The three applications that we characterize are all multi-phase
+applications that require parallel computation, data exchange, and
+synchronization between phases. 
+\begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
+\item Circuit - 
+\item Fluid -
+\item AMR - 
+\end{itemize}
+In Section~\ref{subsec:eventlife} we show properties of events in
+the Fluid application.  Section~\ref{subsec:lockmig} examines
+the migration of locks in the Fluid application.  We give
+a brief case study of reductions for the Circuit application
+in Section~\ref{subsec:reduccase}.  Finally, we demonstrate
+that the asynchronous nature of our interface is essential to the
+performance of Legion by comparing to bulk-synchronous
+implementations of our target applications in Section~\ref{subsec:bulkcomp}.
+
+\subsection{Event Lifetimes}
+\label{subsec:eventlife}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.33]{figs/event_lifetimes.pdf}
+\end{center}
+\vspace{-6mm}
+\caption{Event Liftimes - Fluid Application.\label{fig:eventlife}}
+\vspace{-4mm}
+\end{figure}
+
+Figure~\ref{fig:eventlife} shows the event lifetimes in a run of the fluid application.  The total number
+of events created during the application is over 260,000.
+
+\subsection{Lock Migration}
+\label{subsec:lockmig}
+
+\subsection{Reduction Case Study}
+\label{subsec:reduccase}
+
+\subsection{Bulk-Synchronous Comparison}
+\label{subsec:bulkcomp}

### 2012-07-22 doc/asplos2013/results.tex
added table for lock request/transfer counts
Working on the results section

@@
 run ahead of the actual execution of a stream of instructions
-to ensure that the processor's pipeline is fully utilized.  Similarly, on
-every processor of a supercomputer there is an instance of the Legion runtime
+to ensure that the processor's pipeline is fully utilized.  Similarly, in a
+Legion application, for every processor there is an instance of the Legion runtime
 that takes a stream of tasks with region requirements and asynchronously
 runs ahead of the actual execution using a {\em software-out-of-order} (SOOP)
-scheduler.  The details of the SOOP are beyond the scope of this paper
-but are described in \cite{Legion12}.
+scheduler.  The SOOP leverages our interface by asynchronously issuing task launches, copy
+operations, and synchronization operations that are chained together by
+event dependences.  The full details of the SOOP are beyond the scope of this paper
+and are described in \cite{Legion12}.
 
@@
 blocking operations that would cause processors to stall and severly
-limit Legion's ability to run-ahead and keep the machine fully
+limit Legion's ability to run ahead and keep the machine fully
 occupied.  With a fully asynchronous interface the SOOP can run
@@
 In this section we demonstrate the performance properties of three
-real-world applications that use the Legion runtime and consequently the heterogeneous
+real-world applications that use the Legion runtime and the heterogeneous
 implementation of our interface to run on the Keeneland supercomputer.
@@
 \begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
-\item Circuit - 
-\item Fluid -
-\item AMR - 
+\item {\em Circuit} is simulation of an irregular graph of integrated
+circuit components connected by wires.  The graph is paritioned
+and distributed across the machine.  Each time step performs is composed
+of three phases that calculate currents on the wires, distribute charge between
+nodes, and update the voltages of all ndoes.
+\item Fluid - Is an incompressible fluid flow simulation from the PARSEC
+benchmark suite\cite{bienia11benchmarking}.  When written in Legion, Fluid can
+be run on distributed machines in addition to shared memory.  The Fluid simulation
+models particles that flow through cells.  Each time requires multiple phases
+that update different proprerties of the particles contingent upon neighboring
+particles.
+\item AMR - AMR is an adaptive mesh refinement benchmark based on the third heat
+equation example from the Berkeley Labs BoxLib project\cite{BoxLib}.  AMR
+simulates the two dimensional heat diffusion equation using three different levels
+of refinement.  Each level is partitioned and distributed.  Time steps require
+both intra- and inter-level communication and synchronization.
 \end{itemize}
-In Section~\ref{subsec:eventlife} we show properties of events in
+In Section~\ref{subsec:eventlife} we study the lifetime of events in
 the Fluid application.  Section~\ref{subsec:lockmig} examines
-the migration of locks in the Fluid application.  We give
+the migration of locks in the Circuit application.  We give
 a brief case study of reductions for the Circuit application
@@
 \label{subsec:eventlife}
+We instrumented the heterogenous implementation of our interface to 
+capture information about events.  Figure~\ref{fig:eventlife} shows
+the lifetimes of different types of events from a single run of the Fluid
+application.  Dynamic events is a monotonically increasing line that
+corresponds to the total number of events created in the system.  By
+the end of the run, the simulation created over 260,000 dynamic
+events.  In contrast, the physical events line corresponds to the total number
+of generational events required to represent the dynamic events 
+across all the nodes in the system.  Less than 5,000
+generational events were required to represent all the dynamic events
+illustrating that the technique of mapping from dynamic events to generational 
+events described in Section~\ref{subsec:eventimpl} was effective.
+
+The live event line shows the total number of {\em live events} at a point in time.  Dynamic events
+are considered {\em live events} until their last query has been performed.  The number of live
+events is equivalent to the number of physical events that would be required
+in a reference counted scheme.  In this case we see that by reusing
+generational events, we require up to 4X fewer generational events than would be
+required by a reference counted scheme.
+
+In an ideal world the total number of generational events required would be
+equivalent to the maximum number of generational events in the untriggered
+state at any point.  The untriggered event line shows the number of generational
+events in the untriggered state.  We see that the total number of generational
+events is actually slightly higher than the maximum number of generational
+events in the untriggered state at any point.  The reason for this is that unused
+generational events cannot be shared across nodes, so in some cases nodes must
+make new generational events eventhough there are unused generational events
+on other nodes.  The small difference in these lines shows that in practice
+this is only a minor inefficiency.
 
@@
 \label{subsec:lockmig}
+\begin{figure}
+{
+\renewcommand{\arraystretch}{1.2}
+\begin{center}
+\small
+\begin{tabular}{c|rrrrrrr}
+\multicolumn{1}{l}{\multirow{2}{*}{{\renewcommand{\arraystretch}{1}\begin{tabular}{@{}l}\bf Lock\\\bf Xfers\end{tabular}}}}
+& \multicolumn{7}{c}{\bf Total Lock Requests} \\
+     & 
+\multicolumn{1}{c}{\bf 0} &
+\multicolumn{1}{c}{\bf 1} &
+\multicolumn{1}{c}{\bf 2-8} &
+\multicolumn{1}{c}{\bf 9-16} &
+\multicolumn{1}{c}{\bf 17-32} &
+\multicolumn{1}{c}{\bf 33-64} &
+\multicolumn{1}{c}{\bf 65-128} \\
+{\bf 0    } & 45 & 2611 & 9   & 26   & 79    & 78    & 30     \\
+{\bf 1    } &  -  & 450  &  -   &   -   &   -    &    -   &  -      \\
+{\bf 2-8  } &  - & - & - & - & - & - & -\\
+{\bf 9-16 } &  -  &  -    &   -  & 8 & - & - & -\\
+\end{tabular}
+\end{center}
+}
+\vspace{-6mm}
+\caption{Locks by Request and Transfer Counts - Circuit Application.\label{fig:lockcount}}
+\vspace{-4mm}
+\end{figure}
 

### 2012-07-22 doc/asplos2013/results.tex
More work on the applications section

@@
 synchronization between phases.
-\begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
-\item {\em Circuit} is simulation of an irregular graph of integrated
+
+{\em Circuit} is simulation of an irregular graph of integrated
 circuit components connected by wires.  The graph is paritioned
-and distributed across the machine.  Each time step performs is composed
+and distributed across the machine.  Each time step is composed
 of three phases that calculate currents on the wires, distribute charge between
-nodes, and update the voltages of all ndoes.
-\item Fluid - Is an incompressible fluid flow simulation from the PARSEC
+nodes, and update the voltages of all ndoes.  The distribute charge
+phase leverages reduction instances to perform reductions locally in GPU
+framebuffer memeory before applying all the reductions to a global instance
+residing in GASNet memory.  Without reduction instances this phase
+couldn't be performed using GPUs.
+
+{\em Fluid} is an incompressible fluid flow simulation from the PARSEC
 benchmark suite\cite{bienia11benchmarking}.  When written in Legion, Fluid can
 be run on distributed machines in addition to shared memory.  The Fluid simulation
-models particles that flow through cells.  Each time requires multiple phases
+models particles that flow through cells.  Each time step requires multiple phases
 that update different proprerties of the particles contingent upon neighboring
-particles.
-\item AMR - AMR is an adaptive mesh refinement benchmark based on the third heat
+particles in the same cell.  The space of cells is partitioned and neighboring
+cells in different partitions must exchange data between phases.  Legion ensures
+that these exchanges are done point-to-point by chaining copies and tasks
+using events rather than employing global synchronization.
+
+{\em AMR} is an adaptive mesh refinement benchmark based on the third heat
 equation example from the Berkeley Labs BoxLib project\cite{BoxLib}.  AMR
@@
 of refinement.  Each level is partitioned and distributed.  Time steps require
-both intra- and inter-level communication and synchronization.
-\end{itemize}
+both intra- and inter-level communication and synchronization.  Legion again
+expresses dependences between tasks from the same and different levels through
+events.
+
 In Section~\ref{subsec:eventlife} we study the lifetime of events in
 the Fluid application.  Section~\ref{subsec:lockmig} examines
-the migration of locks in the Circuit application.  We give
-a brief case study of reductions for the Circuit application
-in Section~\ref{subsec:reduccase}.  Finally, we demonstrate
+the migration of locks in the Circuit application.  Finally, we demonstrate
 that the asynchronous nature of our interface is essential to the
@@
 We instrumented the heterogenous implementation of our interface to 
-capture information about events.  Figure~\ref{fig:eventlife} shows
+capture information about different kinds of events.  Figure~\ref{fig:eventlife} shows
 the lifetimes of different types of events from a single run of the Fluid
-application.  Dynamic events is a monotonically increasing line that
+application on 16 nodes using 8 processors per node.  Dynamic events 
+is a monotonically increasing line that
 corresponds to the total number of events created in the system.  By
@@
 
-Figure~\ref{fig:eventlife} shows the event lifetimes in a run of the fluid application.  The total number
-of events created during the application is over 260,000.
+%Figure~\ref{fig:eventlife} shows the event lifetimes in a run of the fluid application.  The total number
+%of events created during the application is over 260,000.
 
@@
 \label{subsec:lockmig}
+
+We also instrumented our heterogeneous implementation to profile the usage of 
+locks in real applications.  We measured how many locks were used in the application.  For
+each lock we measured the number of lock requests and the number of times that the lock
+had to be transferred from one node to another.  Figure~\ref{fig:lockcount} shows the results of 
+running the Circuit application on 16 nodes with 3 GPUs per node.  Of the 3336 total locks,
+86\% are not transferred.  13\% of locks are transferred a single time.  Only 8 total locks
+are migrated around in a way that touches multiple nodes.  This shows in practice
+that locks do need to be transferred between nodes, but the need to transfer locks is
+infrequent.
+
 \begin{figure}
@@
 
-\subsection{Reduction Case Study}
-\label{subsec:reduccase}
+%\subsection{Reduction Case Study}
+%\label{subsec:reduccase}
+
+%In this section we give a brief case study of how reductions make it possible
+%to achieve high performance on a real application.  One of the three phases
+%of the Circuit application involves reducing the current flowing in the wires
 
@@
 \label{subsec:bulkcomp}
+
+Our last set of experiments demonstrates that the asynchronous nature of our interface
+is essential to the high performance of the Legion runtime system.  The Legion
+implementations of our three applications have already been shown to outperform 
+pre-existing implementations\cite{Legion12}.  To quantify the benefit of the asynchronous
+nature of our interface, we implemented versions of our applications in
+Legion in a bulk-synchronous style to compare against our existing implementations.  
+Note that the bulk-synchronous versions of our applications still run on top of 
+the Legion runtime, but invoke blocking operations between computational phases,
+communication phases, and synchronization.
+
+Figures~\ref{fig:cktbulk}, \ref{fig:fluidbulk}, and \ref{fig:amrbulk} compare
+the performance of the bulk-synchronous implementations versus the original
+implementations for Circuit, Fluid, and AMR respectively.  Each plot contains
+performance curves for both implementations on two different problem sizes.
+Circuit is a compute-bound application and the bulk-synchronous implementation
+performs reasonably well.  By 16 nodes however the overhead grows to 19\%
+and 22\% on the 48 and 96 piece circuit inputs respectively.  The Fluid application 
+has a more evenly balanced computation-to-communication ratio.  As a result,
+Fluid suffers significantly worse performance degredation by switching to
+a bulk-synchronous model.  At 16 nodes, performance is 2.35X and 1.52X worse
+than the original Legion implementations on the small and large problem sizes
+respectively.  The AMR application is a memory bound application.  On 16 nodes
+the bulk-synchronous version is 1.82X and 2.02X slower on the small and
+large problem sizes.
+
+These results show that the asynchronous nature of our interface allows
+the Legion runtime system to run ahead of actual execution which
+provides significant latency hiding.  More
+importantly, the effect of this latency hiding grows as the number of
+nodes increases.  In all cases, the overhead of the bulk-synchronous
+implementations grew with the node count.  As we continue to scale
+applications, the ability of runtimes like Legion to run ahead
+will be essential to hiding the large latencies on this class of machines.
+The fully asynchronous nature of the interface presented in this paper
+will be crucial for supporting this ability.
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.33]{figs/circuit_bulk_sync.pdf}
+\end{center}
+\vspace{-6mm}
+\caption{Circuit Bulk-Synchronous Performance.\label{fig:cktbulk}}
+\vspace{-4mm}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.33]{figs/fluid_bulk_sync.pdf}
+\end{center}
+\vspace{-6mm}
+\caption{Fluid Bulk-Synchronous Performance.\label{fig:fluidbulk}}
+\vspace{-4mm}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.33]{figs/amr_bulk_sync.pdf}
+\end{center}
+\vspace{-6mm}
+\caption{AMR Bulk-Synchronous Performance.\label{fig:amrbulk}}
+\vspace{-4mm}
+\end{figure}

### 2012-07-22 doc/asplos2013/impl.tex

@@
 the previous generation has indeed triggered, allowing the existing list of operations to be executed.
+
+Barriers are implemented as an extension of events.  In place of an event's boolean triggered-or-not state,
+a barrier's owner tracks how many arrivals are still expected for the event.  When the arrival count 
+reaches zero, the barrier is considered to have ``triggered''.  Other nodes send {\em barrier update}
+active messages containing an adjustment value that is positive for {\tt alter\_arrival\_count} operations 
+and negative for {\tt arrive} operations.  A simplified version of vector clocks\cite{Fidge1998} is used
+to detect the race condition in which the barrier owner receives an {\tt arrive} operation before the
+corresponding {\tt alter\_arrival\_count} and defer the decrement of the arrival count to avoid a spurious
+triggering of the barrier.
 

### 2012-07-22 doc/asplos2013/micro.tex

--- 2012-07-17 doc/asplos2013/micro.tex
@@
-\section{Microbenchmark Evaluation}
+\section{Benchmark Evaluation}
 \label{sec:micro}
+
+In this section we evaluate the heterogeneous implementation of our 
+interface using several micro-benchmarks that test whether our implementation
+approaches the underlying performance of the
+hardware, and show that the optimizations described in Section~\ref{sec:impl}
+are important.  All experiments were run on the Keeneland
+supercomputer\cite{Keeneland}.  Each Keeneland node is composed of
+two Xeon 5660 CPUs, three Tesla M2090 GPUs, and 24 GB of DRAM.  Nodes
+are connected by an Infiniband QDR interconnect.
+
+\subsection{Event Latency and Trigger Rates}
+\label{subsec:eventmicro}
+We implemented two micro-benchmarks for evaluating the performance of
+events.  The first micro-benchmark tests the latency of event triggering,
+both within a node and between nodes.  Processors are organized
+into a ring and each processor in turn creates a user event that is dependent
+on the previous processor's event.  The first event in this long chain of
+dependent events is then triggered, and the time until the triggering of the
+last event in the chain is measured.  The total time is divided by the number
+of events in the chain to yield the mean trigger times.  In the single-node case, all events are local to
+that node, so no active messages are required.  For all other cases, the
+ring uses only a single processor per node so that every trigger requires
+the transmission (and reception) of an event trigger active message.
+
+%% is constructed.  The experiment measures the time from the triggering
+%% of the first event to the triggering of the last event.  In the one node
+%% configuration there is only one processor and all events are local to the
+%% node.  In all other cases, there is only one processor per node which
+%% guarantees that all event triggers result in an active message.  
+%% Table~\ref{tab:eventlat} shows the results for both configurations.
+
+\begin{figure}
+\begin{center}
+{ \small
+\begin{tabular}{m{2cm}|c|c|c|c|c}
+Nodes & 1 & 2 & 4 & 8 & 16 \\ \hline
+Mean Trigger Time ($\mu$s) & 0.329 & 3.259 & 3.799 & 3.862 & 4.013 \\
+\end{tabular}
+}
+\end{center}
+\vspace{-6mm}
+\caption{Event Latency Results.\label{tab:eventlat}}
+\vspace{-4mm}
+\end{figure}
+
+\hyphenation{GASNet}
+
+The calculated mean trigger times are shown in Table~\ref{tab:eventlat}.
+The cost of manipulating the data structures themselves and running dependent
+operations is shown by the single-node case, which experienced an average
+latency of only 329 nanoseconds.  The addition of nearly 3 microseconds when
+going from one node to two can be attributed to the latency of a GASNet
+active message - others have measured similar latencies\cite{GASNET06}.
+The gradual increase in latency with increasing node count is likely 
+related to the underlying point-to-point nature of Infiniband communication,
+which requires GASNet to poll a separate connection for every other node in
+the system.
+%% In the single node case the latency of triggering an event is only 
+%% 329 nano-seconds.  For the multi-node cases, the triggering time is
+%% between 3-4 micro-seconds which is the same as the cost of GASNet
+%% active messages on Infiniband\cite{GASNET07}.  The gradual increase in latency with
+%% node count is because each node must poll all the Infiniband connections for other nodes.
+%% This cost of performing event triggers is in entirely in the hardware and the underlying software stack.
+
+Our second micro-benchmark for events measures the maximum rate at which events can
+be triggered by our implementation.  Instead of a single chain, many parallel chains
+are created.  Additionally, each step of the chain can consist of parameterized number
+of events, each one dependent on all the events from the previous step of the chain.
+This parameter is called the {\em fan-in/out factor}.  The events within each step of
+a chain are distributed across the nodes.  Recall from Section~\ref{subsec:eventimpl}
+that the aggregation of event subscriptions
+will limit the number of event trigger active messages to one per node (per event trigger)
+even when the fan-in/out factor exceeds the node count.
+
+%% triggers.  This benchmark creates {\em levels} of events where every event at a level
+%% corresponds to a merge of all of the events at the previous level.  The benchmark is therefore
+%% characterized by the number events at each level called the fan-in/out factor.  
+%% The events at each level are evenly distributed between the processors in the machine.
+%% In the case of an $n$-node experiment, only $n-1$ active messages will be required 
+%% per event trigger because the event subscription mechanism described in 
+%% Section~\ref{subsec:eventimpl}.  Figure~\ref{fig:eventthroo} shows the event triggering
+%% throughput versus total nodes for a range of fan-in/out factors.
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.33]{figs/event_throughput.pdf}
+\end{center}
+\vspace{-6mm}
+\caption{Event Trigger Rate Micro-Benchmark.\label{fig:eventthroo}}
+\vspace{-4mm}
+\end{figure}
+
+Figure~\ref{fig:eventthroo} shows the event trigger rates achieved by our implementation
+for a variety of node counts and fan-in/out factors.
+For small fan-in/out factors, the total rate falls off initially going to two nodes as active
+message become necessary, but starts increasing slightly again at larger node counts.  Higher
+fan-in/out factors require more messages and have lower throughput that also increases with
+node count.  Although the number of events waiting on each node decreases with an increasing
+node code, the minimal scaling indicates that the main bottleneck is in the processing of the
+active message that each node must receive rather than the local redistribution of the triggering
+notification.
+%Note we can further improve performance by increasing
+%the number of threads per node for handling active messages (in these experiments there
+%is one per node), but in real applications these threads would consume hardware cores and
+%take away computational resources from the primary application.  This is only a viable
+%option for memory or communication bound programs.
+
+The compute-bound nature of the benchmark demonstrates that active messages do not tax 
+the interconnection network and leave bandwidth available for application data movement.  
+The event trigger rates achieved in this micro-benchmark are between one and two orders
+of magnitude larger than the actual trigger rates required by the real applications
+described in Section~\ref{sec:apps}.
+
+\subsection{Lock Grant Rates}
+\label{subsec:lockmicro}
+
+Our lock micro-benchmark is designed to measure the rate at which locks can be granted.
+A parameterized number of locks are created per node and their handles are made available to
+every node.  Each node then creates a parameterized number of {\em chains} of lock/unlock
+request pairs, where each request pair is to a random lock, and is made dependent on the 
+previous lock/unlock pair's completion.  The chains are made long enough that startup
+and cleanup times are negligible.  The total number of chains across all nodes gives the total
+number of lock requests that can exist in the system at any given time.  All chains are
+started at the same time (via a dependence on a single user event), and the total time to
+process all the chains is divided into the total number of lock requests to yield and average
+lock grant rate.
+
+%% maximum throughput of lock
+%% grants.  The benchmark is characterized by the number of nodes, the number
+%% of locks initially created by each node, and the number of {\em chains} of
+%% lock requests per node.  A chain is 1024 lock acquire and release operations that
+%% are serialized by event dependences.  Each node creates the specified number of locks and
+%% informs all other nodes of its lock handles.  All nodes create the specified number
+%% of chains by randomly selecting from the set of all locks in the system.  All chains
+%% are predicated on a user event.  The experiment measures the time from the triggering of
+%% the user event to the completion of all chains.
+
+Figure~\ref{fig:fixedlock} shows the lock grant rate for a variety of node counts and locks
+per node.  The number of chains per node is varied so that the total number of chains in the system
+is 1024 in all cases.  For the single-node cases, the insensitivity to the number of locks indicates
+that the bottleneck is in the computational ability of the node to process the requests. 
+For larger numbers of nodes, especially for the larger numbers of locks per node (where contention
+for any given lock is low), the speedup with increasing node count suggests the limiting factor is
+the rate at which lock-related active messages can be sent (nearly every request will require a
+transfer of ownership).  Especially in the middle of the node count range, the performance actually
+increases with
+decreasing lock counts.  Although contention increases, the favoring of local lock requestors makes
+this an advantage, reducing the number of lock-related active messages that must be sent.
+
+%%  When going
+%% from one node to two nodes, the bottleneck becomes the latency of transferring ownership of locks
+%% between the nodes
+%%  versus nodes in the system
+%% for 256 total chains in the system.  Each line corresponds
+%% to a different number of initial locks per node.  At small node counts, 
+%% small numbers of locks per node cause high
+%% contention and lock unfairness allows for very high throughput as many requests are
+%% handled locally before locks are migrated.   When there are more total locks there
+%% is lower contention and lower throughput because the ratio of lock grants to migrated
+%% locks decreases.  At larger node counts, there are enough locks
+%% in the system to remove contention and throughput improves.
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.33]{figs/fixed_lock_chains.pdf}
+\end{center}
+\vspace{-6mm}
+\caption{Lock Benchmark for Fixed Total Chains.\label{fig:fixedlock}}
+\vspace{-4mm}
+\end{figure}
+
+To more clearly demonstrate the benefit of lock unfairness, even at large node counts, we show an alternative
+cut of the experiment space in Figure~\ref{fig:fixednode}.
+In this case the node count is fixed at 8 and lock grant rates are shown for a variety of total lock counts
+and number of chains per node.
+With only 32 chains per node (a total of 256 chains)
+there is little contention and the grant rate is high.  As the number of chains per node increases there is
+more contention for locks and the rate drops as a result.  For smaller lock counts, further increases in
+the number of chains result in improved rates.  Note that the point on each line where this occurs
+ is when
+the chains per node is greater than total locks, i.e. where the expected number of requests per lock 
+per node becomes more than one.  As soon as there are multiple expected requestors for the same lock on
+a node, the unfairness property of our implementation is able to reduce the number of lock ownership 
+transfers, yielding better performance.
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.33]{figs/fixed_node_lock.pdf}
+\end{center}
+\vspace{-6mm}
+\caption{Lock Benchmark for Fixed Node Count.\label{fig:fixednode}}
+\vspace{-4mm}
+\end{figure}
+
+\subsection{Reduction Throughput}
+\label{subsec:reducmicro}
+To show the benefits of associating reduction operations with physical regions we designed
+a histogram micro-benchmark that requires all nodes to perform an addition reduction to a
+physical region located in the globally visible GASNet memory (described in 
+Section~\ref{sec:impl}).  Using the reduction interface described in Section~\ref{subsec:phyreg}
+reductions are performed in five ways:
+
+\begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
+\item Single Reads/Writes - Each node takes a lock on the global physical region and then for each reduction
+reads a value from the physical region, performs the addition, and then writes the result back.  The lock is then released.
+\item Single Reductions - Each reduction is individually sent as an active message to the node whose
+system memory holds the target of the reduction.
+\item Localized Instance - Each node takes a lock on the the global physical region, copies it to a new physical
+region in its own system memory, performs all reductions to the localized region, copies the region back, and releases the lock.
+\item Fold Instance - Each node creates a local reduction instance that folds reductions to the same element.
+All reductions are folded into the instance, which is then copied back to the global region.
+\item List Instance - Each node creates a local reduction instance that stores a list of reduction operations.
+All reductions are appended to the list, which is then copied back to the global region.
+\end{itemize}
+
+We ran two experiments, one corresponding with dense reductions and the other corresponding with sparse
+reductions.  In each case a large random source of data is divided across into chunks, which are given to
+separate reduction tasks.  There are 8 reduction tasks created for each node, one per processor.  The dense
+case uses a histogram with 256K buckets and each reduction task processes a chunk of 4M entries (i.e.
+performs 4M reductions).  The results for the
+dense experiment are shown in Figure~\ref{fig:reducdense}.  The sparse case use a histogram with 4M buckets,
+but only 64K elements are processed by each task.  Results for the sparse case
+be seen in Figure~\ref{fig:reducsparse}.
+
+The dense experiment demonstrates that reduction fold regions perform 
+best and scale well with the number of nodes, achieving over a billion reductions
+per second in the 16 node case.  List regions also scale well, but perform about an
+order of mangitude worse than folding regions in the dense case.  The use of a separate active message for
+each reduction operation is another two orders of magnitude worse, demonstrating that data must be
+transferred in larger blocks to be efficient on modern interconnect hardware.  The localized instance approach
+works well for a single node, but its inherently serial nature doesn't benefit at all from increasing node counts.
+Finally, the only time the latency of performing individual RMA reads and writes isn't disastrous is
+on a single node (where the reads and writes aren't really remote).
+
+The sparse case is very similar, except the scalability of the reduction fold case suffers from the
+overhead of transferring a value for every bucket when most buckets are unmodified by a given task.
+The reduction list case continues to scale well, surpassing the reduction fold performance at large
+node counts.  This shows that
+list regions are better suited for scaling sparse reduction operations.
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.33]{figs/reduce_dense.pdf}
+\end{center}
+\vspace{-6mm}
+\caption{Dense Reduction Benchmark.\label{fig:reducdense}}
+\vspace{-4mm}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.33]{figs/reduce_sparse.pdf}
+\end{center}
+\vspace{-6mm}
+\caption{Sparse Reduction Benchmark.\label{fig:reducsparse}}
+\vspace{-4mm}
+\end{figure}

### 2012-07-22 doc/asplos2013/conclusion.tex
Updated conclusion

@@
 
-Conclusion goes here.
+As large supercomputers continue to scale, the latencies associated
+with common operations such as communication and synchronization will
+become larger and more variable.  To effectively hide these latencies
+future programming systems will need to leverage asynchronous operations
+to overlap long latency operations with useful work.  We have
+presented a low-level programming interface that is fully asynchronous.
+Our interface uses events to compose asynchronous operations.  Clients
+can use events to explicitly specify dependences between operations,
+leaving the runtime to schedule them in a way that maximizes latency 
+hiding while still obeying dependences.
+
+We've presented an implementation of interface that is capable of running
+on a large heterogeneous cluster containing both CPUs and GPUs.  We
+employed micro-benchmarks to show that our implementation is efficient.
+We also demonstrated that our interface is capable of supporting Legion,
+an advanced high-level runtime system that fully leverages the asynchronous
+features of our interface.  We showed that for three real-world applications
+the asynchronous Legion implementations were between 22\% to 2.35X faster than
+implementations which contained blocking operations.

### 2012-07-23 doc/asplos2013/results.tex
minor edit
edits to results section (except events)

@@
 scheduler.  The SOOP leverages our interface by asynchronously issuing task launches, copy
-operations, and synchronization operations that are chained together by
-event dependences.  The full details of the SOOP are beyond the scope of this paper
+operations, and synchronization operations with the dependencies between them expressed
+as events.  The full details of the SOOP are beyond the scope of this paper
 and are described in \cite{Legion12}.
 
-The fully asynchronous design of the interface that we've proposed in
+The fully asynchronous design of the interface proposed in
 this paper is crucial to Legion's SOOP implementation.  Without a fully
@@
 limit Legion's ability to run ahead and keep the machine fully
-occupied.  With a fully asynchronous interface the SOOP can run
+occupied.  With a fully asynchronous interface the SOOP can hide latencies by 
+running
 far ahead of the actual execution, limited only by the physical resources
@@
 
-{\em Circuit} is simulation of an irregular graph of integrated
-circuit components connected by wires.  The graph is paritioned
-and distributed across the machine.  Each time step is composed
+{\em Circuit} is simulation of an electric circuit that is described as an
+irregular graph of integrated
+circuit components connected by wires.  The graph is partitioned
+and distributed across the machine.  The computation for each time step consists
 of three phases that calculate currents on the wires, distribute charge between
-nodes, and update the voltages of all ndoes.  The distribute charge
-phase leverages reduction instances to perform reductions locally in GPU
-framebuffer memeory before applying all the reductions to a global instance
+nodes, and update the voltages of all ndoes.  In the distribute charge
+phase, a single node may be updated by wires from multiple partitions.  Reduction instances 
+are used to allow each partition to perform reductions in parallel and locally in GPU
+framebuffer memeory before merging all the reductions to a global instance
 residing in GASNet memory.  Without reduction instances this phase
@@
 {\em Fluid} is an incompressible fluid flow simulation from the PARSEC
-benchmark suite\cite{bienia11benchmarking}.  When written in Legion, Fluid can
-be run on distributed machines in addition to shared memory.  The Fluid simulation
+benchmark suite\cite{bienia11benchmarking}.  The original implementation can 
+only be run on a single shared memory machine, but when written in Legion, Fluid can
+be run on large distributed machines as well.  The Fluid simulation
 models particles that flow through cells.  Each time step requires multiple phases
@@
 that these exchanges are done point-to-point by chaining copies and tasks
-using events rather than employing global synchronization.
+using events rather than employing a global bulk-synchronous approach.
 
@@
 simulates the two dimensional heat diffusion equation using three different levels
-of refinement.  Each level is partitioned and distributed.  Time steps require
+of refinement.  Each level is partitioned and distributed across the machine.  Time steps require
 both intra- and inter-level communication and synchronization.  Legion again
@@
 \vspace{-6mm}
-\caption{Event Liftimes - Fluid Application.\label{fig:eventlife}}
+\caption{Event Lifetimes - Fluid Application.\label{fig:eventlife}}
 \vspace{-4mm}
@@
 each lock we measured the number of lock requests and the number of times that the lock
-had to be transferred from one node to another.  Figure~\ref{fig:lockcount} shows the results of 
-running the Circuit application on 16 nodes with 3 GPUs per node.  Of the 3336 total locks,
-86\% are not transferred.  13\% of locks are transferred a single time.  Only 8 total locks
-are migrated around in a way that touches multiple nodes.  This shows in practice
-that locks do need to be transferred between nodes, but the need to transfer locks is
-infrequent.
+had to be transferred from one node to another.  Figure~\ref{fig:lockcount} shows the results 
+for the Circuit and AMR application, both running on 16 nodes.  The Circuit application uses
+a larger number of locks (3336 vs. 1393), but most of them (86\%) are only used on the node on 
+which they were created.  The AMR application exhibits more varied locking behavior, and in 
+particular has locks that are requested more often than they are transferred.  Without the ability
+for locks to migrate, more of those requests would have required inter-node communication.
 
 \begin{figure}
+\centering
+
+\subfigure[Circuit application]{
+\label{fig:lockcount:ckt}
 {
 \renewcommand{\arraystretch}{1.2}
-\begin{center}
 \small
@@
 \end{tabular}
-\end{center}
-}
-\vspace{-6mm}
-\caption{Locks by Request and Transfer Counts - Circuit Application.\label{fig:lockcount}}
+}}
+\vspace{-2mm}
+\subfigure[AMR application]{
+{
+\renewcommand{\arraystretch}{1.2}
+\small
+\begin{tabular}{c|rrrrrrr}
+\multicolumn{1}{l}{\multirow{2}{*}{{\renewcommand{\arraystretch}{1}\begin{tabular}{@{}l}\bf Lock\\\bf Xfers\end{tabular}}}}
+& \multicolumn{7}{c}{\bf Total Lock Requests} \\
+     & 
+\multicolumn{1}{c}{\bf 0} &
+\multicolumn{1}{c}{\bf 1} &
+\multicolumn{1}{c}{\bf 2} &
+\multicolumn{1}{c}{\bf 3-4} &
+\multicolumn{1}{c}{\bf 5-8} &
+\multicolumn{1}{c}{\bf 9-16} &
+\multicolumn{1}{c}{\bf 17-32} \\
+{\bf 0   } & - & 94  & 592 & 414 & 94 & 3 & 41 \\
+{\bf 1   } & - & 122 & -   & -   & -  & - & - \\
+{\bf 2   } & - & -   & 6   & 7   & -  & - & - \\
+{\bf 3-4 } & - & -   & -   & -   & 9  & 3 & - \\
+{\bf 5-8 } & - & -   & -   & -   & -  & - & - \\
+{\bf 9-16} & - & -   & -   & -   & -  & 7 & 1
+\end{tabular}
+}}
+\vspace{-2mm}
+\caption{Locks by Request and Transfer Counts.\label{fig:lockcount}}
 \vspace{-4mm}
@@
 Legion in a bulk-synchronous style to compare against our existing implementations.  
-Note that the bulk-synchronous versions of our applications still run on top of 
+These bulk-synchronous versions of still run on top of 
 the Legion runtime, but invoke blocking operations between computational phases,
@@
 performs reasonably well.  By 16 nodes however the overhead grows to 19\%
-and 22\% on the 48 and 96 piece circuit inputs respectively.  The Fluid application 
+and 22\% on the small and large inputs respectively.  The Fluid application 
 has a more evenly balanced computation-to-communication ratio.  As a result,
-Fluid suffers significantly worse performance degredation by switching to
-a bulk-synchronous model.  At 16 nodes, performance is 2.35X and 1.52X worse
+Fluid performance suffers significantly more by switching to
+a bulk-synchronous model.  At 16 nodes, performance is 135\% and 52\% worse
 than the original Legion implementations on the small and large problem sizes
-respectively.  The AMR application is a memory bound application.  On 16 nodes
-the bulk-synchronous version is 1.82X and 2.02X slower on the small and
+respectively.  Finally, the AMR application is a memory bound application, and on 16 nodes
+the bulk-synchronous version is 82\% and 102\% slower on the small and
 large problem sizes.
@@
 the Legion runtime system to run ahead of actual execution which
-provides significant latency hiding.  More
-importantly, the effect of this latency hiding grows as the number of
-nodes increases.  In all cases, the overhead of the bulk-synchronous
+provides significant latency hiding capability.  
+The benefit of this latency hiding grows as the number of
+nodes increases and as the computation-to-communication ratio decreases.
+In all cases, the overhead of the bulk-synchronous
 implementations grew with the node count.  As we continue to scale
-applications, the ability of runtimes like Legion to run ahead
-will be essential to hiding the large latencies on this class of machines.
+applications to larger and larger machines, the ability of runtimes like Legion to run ahead
+will be essential to hiding the large latencies inherent in such machine architectures.
 The fully asynchronous nature of the interface presented in this paper

### 2012-07-23 doc/asplos2013/impl.tex
Some edits to implementation

--- 2012-07-22 doc/asplos2013/impl.tex
@@
 runs on larges clusters with both CPUs and GPUs.  The shared-memory
-implementation uses the POSIX threads library.
-The implementation for heterogeneous clusters also uses POSIX
-threads\cite{PTHREADS} (commonly referred to as Pthreads) as 
+implementation uses the POSIX threads library\cite{PTHREADS} (Pthreads)
+and is primarily used for debugging.
+The implementation for heterogeneous clusters also uses Pthreads as 
 well as the CUDA runtime library for GPUs\cite{CUDA} and the GASNet
@@
 models a cluster with CPUs and GPUs as having two types of processors 
-and four types of memory.  Every CPU core is presented as a different CPU processor
-and every GPU is a GPU processor.  This matches the scheduling granularity available
+and four types of memory.  Every CPU core is presented as a different {\tt CPU} processor
+and every GPU is a {\tt GPU} processor.  This matches the scheduling granularity available
 in the Pthreads and CUDA APIs respectively.  The first type of memory is the 
@@
 the GASNet runtime to allow {\em remote memory access} (RMA) by other nodes in the cluster.
+We refer to this memory as {\em GASNet memory}.
 %% For every node in the system
@@
 are sent by one node to another without any previous coordination.  Upon arrival
-at the destination node, a handler routine is invoked to process the payload.
-The handler may issue a response to the sender, but this is optional, and avoided
-as much as possible in our implementation due to the latencies involved in waiting
-for a response.  As an example, consider the {\tt spawn} call on a {\tt Processor}
+at a destination node, a handler routine is invoked to process an active message.
+%The handler may issue a response to the sender, but this is optional, and avoided
+%as much as possible in our implementation due to the latencies involved in waiting
+%for a response.  
+One example of a case where we use active messages involves the {\tt spawn} call on a {\tt Processor}
 object.  If the processor is not on the same node as the requestor, the information
-about the task (e.g. the task ID and arguments, the actual target {\tt Processor},
-the prerequisite event (if any), and the completion event ID) is sent in an
+%about the task (e.g. the task ID and arguments, the actual target {\tt Processor},
+%the prerequisite event (if any), and the completion event ID) 
+for the task is sent in an
 active message to the node containing the target processor.  The recipient node
@@
 processor or marks it as being deferred until the prerequisite event has triggered.
-No response is sent to the original requestor.  The requestor already knows the event
-that corresponds to the completion of this task (since it was provided in the active
-message) - there is no utility in knowing exactly when the active message itself has
-been processed.
+%No response is sent to the original requestor.  The requestor already knows the event
+%that corresponds to the completion of this task (since it was provided in the active
+%message) - there is no utility in knowing exactly when the active message itself has
+%been processed.
 
@@
 
-Due to space constraints, we will not cover every part of our implementation in
-detail, instead limiting our discussion to the three most interesting aspects:
+%Due to space constraints, we will not cover every part of our implementation in
+%detail, instead limiting our discussion to the three most interesting aspects:
+%events, locks, and reduction instances.
+Due to space constraints we cover only a small part of the implementation in detail.
+We limit our discussion to three interesting aspects:
 events, locks, and reduction instances.
@@
 
-Events may be created on demand, and are always {\em owned} by the node on which the
-creation request was made.  The creation of an event occurs with no inter-node communcation.
-The set of event
+Events are created on demand and are {\em owned} by the node on which they
+were created.  The creation of an event occurs with no inter-node communcation.
+The space of event
 handles is divided across the nodes at startup time, allowing each node to assign handles
-to new events without fear of conflicting with another node's events.  No broadcast of
-the event creation is required either, as the static division is sufficient to allow any
-node to correctly determine an event's owning node when (and only if) an operation on that
-event is performed on the other node.
+to new events without the risk of conflicting with another node's assignments.  The static division
+of event handles permits any node to determine the owning node of an event
+without any communication.
+%No broadcast of
+%the event creation is required either, as the static division is sufficient to allow any
+%node to correctly determine an event's owning node when (and only if) an operation on that
+%event is performed on the other node.
 
-At event creation time, the owning node allocates a data structure to track the state of the event
-(i.e. triggered or not) as well as record a list of the operations that are known to be dependent
-on the event (e.g. initiation of a copy operation, placing a task on a processor's ready queue, waking
-up a waiting task).  However, the owning node only keeps a list of the dependent operations from the same
-node.  Every other node also allocates a corresponding data structure (the first time the event
-is referenced) to remember if the event has already triggered and to record its local dependent operations.
-Arbitrarily many dependent operations from a
-single node are aggregated into a single {\em event subscription} active message that is sent to the
-owning node.  The owning node keeps a bitmask of which other nodes have subscribed to the event, and when the
-event finally does trigger, a single {\em event trigger} active message is sent to each subscribing node,
-which notes that the event has triggered (in case queries come after the trigger) and executes the list
-of local operations that were dependent on that event.  (In the case that the triggering of the event has
-occurred before the reception of an event subscription, a trigger message is sent immediatel to the 
-new subscriber.)
+%At event creation time, the owning node allocates a data structure to track the state of the event
+%(i.e. triggered or not) as well as record a list of the operations that are known to be dependent
+%on the event (e.g. initiation of a copy operation, placing a task on a processor's ready queue, waking
+%up a waiting task).  However, the owning node only keeps a list of the dependent operations from the same
+%node.  Every other node also allocates a corresponding data structure (the first time the event
+%is referenced) to remember if the event has already triggered and to record its local dependent operations.
+%Arbitrarily many dependent operations from a
+%single node are aggregated into a single {\em event subscription} active message that is sent to the
+%owning node.  The owning node keeps a bitmask of which other nodes have subscribed to the event, and when the
+%event finally does trigger, a single {\em event trigger} active message is sent to each subscribing node,
+%which notes that the event has triggered (in case queries come after the trigger) and executes the list
+%of local operations that were dependent on that event.  (In the case that the triggering of the event has
+%occurred before the reception of an event subscription, a trigger message is sent immediatel to the 
+%new subscriber.)
+When a new event is created, the owning node allocates a data structure to track the state of
+the event ({\em triggered} or {\em untriggered}) as well as to record a list of operations dependent on the
+event from the same node called {\em waiters} (e.g. copy operation or task launch).  The first operation contingent on an event
+from a remote node will create the same data structure on the remote node.  A single
+{\em event subscription} active message is then sent from the remote node to the owning node indicating
+that the remote node should be informed when the event triggers.  Arbitrarily many additional operations on
+the remote node can then be aggregated as waiters onto the remote data structure without any additional communication.
+When the event does trigger, the owner node notifies all local waiters and
+sends a single {\em event trigger} active message to all nodes from which it has seen an event subscription
+to notify remote waiters.
+In the case where an event triggers while a subscription message is in flight from a remote node, the owner node will
+immediately respond with a trigger message.
 
@@
 node, an {\em event trigger} active message is sent from the triggering node back to the owning node, which
-then forwards that message to all subscribed nodes, with the exception of the triggering node, which was
-able to execute its locally dependent operations (if any) immediately.  Although this can often result in
-the latency of a triggering operation including two active messages, it bounds the number of active
-messages required per event to $2N$ where $N$ is the number of nodes (which can be much smaller than the
-number of dependent operations) interested in the event.  An alternative
-would have been to share the subscriber bitmask so that the triggering node can notify all interested
-nodes directly, but such an algorithm is much more complicated due to race conditions, and requires $O(N^2)$
-active messages.  As we will see in section~\ref{sec:microbenchmarks}, the latency of a single extra
-active message is very small, and any cost that is super-linear with the node count either is or will soon
-be an scalability issue for large systems.
+then forwards that message to all the other subscribed nodes.  The triggering node will automatically
+notify its waiters and no message is sent from the owning node back to the triggering node.
+%with the exception of the triggering node, which was
+%able to execute its locally dependent operations (if any) immediately.  
+While a remote trigger of an event can result in
+the latency of a triggering operation being two active messages, it bounds the number of active
+messages required per event to $2N-2$ where $N$ is the number of nodes 
+%(which can be much smaller than the number of dependent operations) 
+interested in the event.  An alternative
+would have been to share the list of subscribers so that the triggering node could notify all interested
+nodes directly.  However, such an algorithm is both more complicated due to race conditions, and requires $O(N^2)$
+active messages.  As we will see in Section~\ref{subsec:eventmicro}, the latency of a single event trigger
+active message is very small, and any algorithm that is super-linear with the node count will not scale well.
+%either is or will soon be an scalability issue for large systems.
 
 The data structures used to track an event cannot be freed until all operations on that event have been
-performed.  Creation and triggering can happen only once each, but there can be any number of operations
-that are dependent on an event, and some of those operations may not be requested until long after the
+performed.  Creation and triggering can each happen only once, but there can be an arbitrary number of operations
+that are dependent on an event.   Furthermore, some operations may not be requested until long after the
 event has triggered.  Other systems incorporating events address this by reference counting event
 handles\cite{Khronos:OpenCL}, but such reference counting adds both programmer and runtime overhead even when
-limited to a single node -- further overhead and complexity would be expected for a cluster-level
+limited to a single node; further overhead and complexity would be expected for a cluster-level
 reference-counting approach.
 
-Instead of focusing on freeing event data structures, our implementation aggressively recycles them, 
-needing fewer total event data structures than a referencing counting implementation while also eliminating
-the overhead of reference counting.  The key observation is that one {\em generational event} data structure
- can efficiently capture the state of one {\em untriggered} event and a very large number (e.g. $2^{32}-1$)
-of already-triggered events.  An event's handle is expanded to include its generation number as well as the
-identifier for the underlying generational event.  In addition to the triggered-or-not state and list of
-dependent operations
-for the most recent generation, the owning node also remembers how many previous generations have already
-been triggered.  (There is no need to keep a list of dependent operations on previous generations.  Any
-new operation that is dependent on a generation that is known to be already triggered can immediately be
-executed.)  When an event is created, any generational event for whom the most recent generation has 
-triggered can be reused - the generation is increased by one and the state is returned to ``untriggered.''
-As before, this can be done with no inter-node communication.
+%Instead of focusing on freeing event data structures, our implementation aggressively recycles them, 
+%needing fewer total event data structures than a referencing counting implementation while also eliminating
+%the overhead of reference counting.  The key observation is that one {\em generational event} data structure
+% can efficiently capture the state of one {\em untriggered} event and a very large number (e.g. $2^{32}-1$)
+%of already-triggered events.  An event's handle is expanded to include its generation number as well as the
+%identifier for the underlying generational event.  In addition to the triggered-or-not state and list of
+%dependent operations
+%for the most recent generation, the owning node also remembers how many previous generations have already
+%been triggered.  (There is no need to keep a list of dependent operations on previous generations.  Any
+%new operation that is dependent on a generation that is known to be already triggered can immediately be
+%executed.)  When an event is created, any generational event for whom the most recent generation has 
+%triggered can be reused - the generation is increased by one and the state is returned to ``untriggered.''
+%As before, this can be done with no inter-node communication.
+Instead of attempting to free event data structures, our implementation aggressively recyles them.
+Compared to a reference counting scheme our implementation requires fewer total event data structures
+and has no programmer/runtime overhead.  The key observations is that one {\em generational event}
+data structure can efficiently capture the state of one untriggered event and a very large 
+number (e.g. $2^{32}-1$) of already-triggered events.  To accomplish this, we extend each event
+handle to include its generation number as well as the identifier for its generational event.  Each
+generational event remembers how many previous generations have already been triggered.  Any
+new operation that is dependent on a generation that is known to already be triggered can immediately
+be executed.  A generational event can be recycled as soon as the current generation has triggered. To
+create a new event, a node finds a generational event in the triggered state,
+increases the generation by one, and sets the generational event's state to untriggered.  As before, 
+this can be done with no communication.
 
-A remote node's data structure is also efficient - the boolean values for whether the event
-has triggered and whether the event has already been subscribed to are replaced with the numbers of the most
-recent generation known to have triggered (this is received in the event trigger active message) and of the
-most recent generation the remote node has subscribed to (this is sent in the event subscription active
-message).  And as with the owner node's data structure, only
-one list of dependent operations is needed.  Although the latencies of a large system can delay the
-reception of an event trigger active message, if an operation is created that depends on a later generation
-of an event than what the current list of events is dependent on, that serves as a roundabout indication that
-the previous generation has indeed triggered, allowing the existing list of operations to be executed.
+%A remote node's data structure is also efficient - the boolean values for whether the event
+%has triggered and whether the event has already been subscribed to are replaced with the numbers of the most
+%recent generation known to have triggered (this is received in the event trigger active message) and of the
+%most recent generation the remote node has subscribed to (this is sent in the event subscription active
+%message).  And as with the owner node's data structure, only
+%one list of dependent operations is needed.  Although the latencies of a large system can delay the
+%reception of an event trigger active message, if an operation is created that depends on a later generation
+%of an event than what the current list of events is dependent on, that serves as a roundabout indication that
+%the previous generation has indeed triggered, allowing the existing list of operations to be executed.
+Nodes also maintain generational event data structures corresponding to remote events that they have
+observed.  These data structures maintain the most recent known generation to have triggered as well
+as the generation of the most recent subscription message sent (if any).  The distributed nature
+of the system allows remote generational events to perform an interesting optimization.  If at any point a
+remote generational event observes a request to wait on a later generation than its current generation,
+then it is safe to infer that all generations up to the requested generation have already 
+been triggered because a new generation of the event was already created by the event's owner node.
+All local waiters can then be notified even before receiving an event triggered message for the current generation.
+
+Barriers are implemented as an extension of events.  In place of an event's triggered/untriggered state,
+a barrier's owner tracks the number of outstanding arrivals.  When the arrival count 
+reaches zero, the barrier is considered to have triggered.  Remote nodes send {\em barrier update}
+active messages containing an adjustment value that is positive for {\tt alter\_arrival\_count} operations 
+and negative for {\tt arrive} operations.  A version of vector clocks\cite{Fidge1998} is used
+to detect the race condition in which the barrier owner receives an {\tt arrive} operation before the
+corresponding {\tt alter\_arrival\_count}.  When detected, the decrement of the arrival count 
+is delayed to avoid a spurious triggering of the barrier.
 
@@
 handles are also statically divided across the nodes, allowing the creation to be done without inter-node
-communication.  Unlike events whose ownership is static, locks may migrate, so although the {\em creating}
-node is the initial owner of a lock, that ownership can be transferred to other nodes.
+communication.  Unlike events whose ownership is static, locks may migrate; the {\em creating}
+node is the initial owner of a lock, but that ownership can be transferred to other nodes.
 
-Locks are often used to mediate access to data, and the size of that data is often relatively small.  To
-avoid adding the latency of inter-node data copies to critical sections, a lock may also be created with a
-{\em payload}, a small (i.e. less than 4KB - the limit for an active message's payload) piece of data that is moved around with the lock and guaranteed to be coherent if
+Locks are often used to mediate access to data, and the size of that data is often small.  To
+support this common case we allow locks to be created with an associated
+%avoid adding the latency of inter-node data copies to critical sections, a lock may also be created with a
+{\em payload}, a small (i.e. less than 4KB - the limit for an active message payload) 
+piece of data that is moved around with the lock and guaranteed to be coherent while 
 the lock is held.
@@
 Since any node may at some point be the owner of a lock, all nodes use the same data structure to track the
-state of the lock (they differ only in when the structure is allocated -
-nodes other than the creator still wait until the first reference to the lock on that node).  The structure
-tracks the following:
-\begin{itemize}
+state of a lock on the node.
+%(they differ only in when the structure is allocated -
+%nodes other than the creator still wait until the first reference to the lock on that node).  
+The structure tracks the following:
+\begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
 \item {\em owner node} - the most recently known owner of the lock.  If the current node is the owner, this
-information is always accurate.  If not, this information may be stale, but at least the recorded ``owner''
-is sure to have more up-to-date information.
-\item {\em lock status} - records whether the lock is currently held.  Only the current lock owner has
-guaranteed-current information here.
-\item {\em local waiters} - a list of pending local lock requests.  This data is always valid as well.
+information is always accurate.  If not, this information may be stale, but by induction the recorded 
+owner is guaranteed to know the actual owner or the next node to query about ownership.
+\item {\em lock status} - records whether the lock is currently held.  This is only valid if the node
+is the current owner.
+\item {\em local waiters} - a list of pending local lock requests.  This data is always valid on all nodes.
 \item {\em remote waiters} - a bitmask of which other nodes have pending requests.  This bitmask is only
-used when the current node is the owner of the lock.
+valid on the current owner.
 \item {\em local payload pointer and size} - the local node's copy of the lock payload
@@
 When a lock request is made, an event is created to track when the grant occurs.  The current node then
-examines its copy of the lock data structure (creating it if
-this the first operation for that lock on that node).  If the current node is the owner and the lock isn't
+examines its copy of the lock data structure to determine if it is the owner.
+%(creating it if this the first operation for that lock on that node).  
+If the current node is the owner and the lock isn't
 held, the lock is granted immediately and the event is triggered.  If the lock is held, the event is added
-to the list of local waiters.  (This is the only thing that is stored.  The actual ``requestor'' of the lock
-is implicitly captured in which operations are dependent on the lock grant event.)  If the current node
-isn't the owner, a {\em lock request} active message is
-sent to the most recently known owner.  (If that information proves to be stale, the request is forwarded
-by the receiving node, eventually catching up to the current owner.)  If the current owner's status shows
-the lock is currently held, the requesting node is added to the remote waiters bitmask.  If the lock is
+to the list of local waiters.  Note that the event associated with
+the lock request is the only data that must be stored.
+%(This is the only thing that is stored.  The actual ``requestor'' of the lock
+%is implicitly captured in which operations are dependent on the lock grant event.)  
+If the current node isn't the owner, a {\em lock request} active message is
+sent to the most recently known owner.  If the receiving node is no
+longer the owner, it will forward the message on to the next node known
+to be the owner of the lock.
+%(If that information proves to be stale, the request is forwarded
+%by the receiving node, eventually catching up to the current owner.)  
+If the current owner's status shows
+the lock is currently held, the bit for the requesting node is set in remote waiters bitmask.  If the lock is
 not held, the ownership of the lock is given to the requesting node via a {\em lock transfer} active
-message.  A lock transfer message includes the current remote waiters bitmask (in case other nodes are
-have also requested the lock) and an up-to-date copy of the lock's payload.
+message.  A lock transfer message includes the bitmask of remote waiters and an up-to-date copy 
+of the lock's payload.
 
@@
 if necessary.  Once the unlock request is on the lock's current owning node, the local waiter list is
-examined.  If the list is non-empty, the lock remains in the ``locked'' state and the first lock grant
+examined.  If the list is non-empty, the lock remains in the locked state and the first lock grant
 event is pulled off the local waiter list and triggered.  If instead the local waiter list is empty, the
-lock state is changed to ``unlocked'', and the remote waiter bitmask is examined.  If any bits are set, then
+lock state is changed to unlocked, and the bitmask of remote waiters is examined.  If there are any, then
 one of them is chosen and the corresponding node becomes the new owner via a lock transfer active message.
@@
 The unfairness that results from a lock favoring local waiters over remote waiters is intentional.  When the
-contention on a lock is high (the only time when the question of fairness is even relevant), the latencies
+contention on a lock is high (the only time when the question of fairness is relevant), the latencies
 involved in transferring a lock between nodes can be the limiter on throughput.  By minimizing the number
@@
 the third node's request ever became the only one left, the lock transferring would stop and the request
-would finally catch up.
+would eventually catch up.
 

### 2012-07-23 doc/asplos2013/impl.tex
Some edits to implementation

@@
 
-In addition to normal physical regions, our implementation supports two different kinds of reduction-only
-physical regions, designed to optimize the common pattern in which multiple tasks make commutative 
-{\em reductions} into shared data.  Instead of having to move the existing data to a suitable memory and
-then applying individual reduction operations to the existing data 
-right away (as would happen in a normal physical region), reduction instances allow a large number of
-reduction operations to be batched up and transferred efficiently to the original location of the data.
-This decoupling from the original location of the data can allow allow greater parallelism
-in cases where there are available processors that cannot all access the same memory.
+In addition to normal physical regions, our implementation supports two different layouts of data for
+reduction-only physical regions called {\em reduction instances}.   
+These layouts are designed to optimize for the common pattern where 
+multiple tasks make commutative {\em reductions} into shared data.  Instead of having to copy the existing 
+physical region to a suitable memory and then applying individual reduction operations, reduction instances 
+allow a large number of reduction operations to be accumulated and transferred efficiently to the original 
+location of the data. This decoupling from the original location of the data can allow allow greater parallelism
+in cases where processors cannot all access the same memory (i.e. many GPUs).
 
 The first kind of reduction-only physical region is a {\em reduction fold instance}, which may only be
-used for reduction operations that support a {\em fold} operation.  A reduction fold instance is similar
+used for reduction operations that support a {\em fold} operation (see Listing~\ref{lst:regionapi}).  
+A reduction fold instance is similar
 to a normal physical region in that it is implemented as an array, indexed by the same element indices used
 for the normal physical regions.  The difference is that each array element is a value of the reduction's
-{\tt RHS} type instead of the actual element type of the region (i.e. the {\tt LHS} type).  When the client
-requests a reduction operation to a particular location, the supplied right-hand-side value is folded in
+{\tt RHS} type which is generally the same size or smaller than the actual element type of the region 
+(i.e. the {\tt LHS} type).  When the client requests a reduction operation to a particular location, 
+the supplied right-hand-side value is folded in
 to the corresponding location in the array.  An arbitrary number of reductions can be done to each location.
 When the reduction fold instance is copied back to a normal instance, the contents of the reduction instance
-(which are generally the same size or smaller than the normal region contents) are transferred in a large
-block to the location of the normal instance, where an application of the reduction function is performed
-once for each location in the region via a cache-friendly linear sweep over the memory.
+are transferred in a large block to the location of the normal instance.  On the receiving side
+the runtime automatically applies the reduction instance to the destination region
+by invoking the reduction function once for each location in the region via a cache-friendly 
+linear sweep over the memory.
 
@@
 perform as well as a reduction fold instance, it can be better for cases with sparse updates.  It is also
-the only option when a reduction operation doesn't support a {\em fold} operation.  Instead of folding
+valid option for when a reduction operation doesn't support a {\em fold} operation.  Instead of folding
 together multiple reductions to the same location, a reduction list instance keeps a log of every
@@
 region, and any randomness in the sequence of locations in the list can impact the memory access efficiency
-of the accesses to the target region.  However, as we will show in section~\ref{sec:microbenchmarks}, 
+of the reductions to the target region.  However, as we will show in section~\ref{subsec:reducmicro}, 
 the benefits of enabling parallelism and hiding latency usually more than make up for these drawbacks.

### 2012-07-23 doc/asplos2013/interface.tex

@@
 to be as low-level as possible.  In many cases we have chosen to trade-off
-programmability for performance.  We eschewed any features that would
+ease of programmability for performance.  We eschewed any features that would
 automatically be performed without being directed by the client.
@@
 In most cases events are created as the result of other
-operations and the implementation is responsible for triggering these events.  Users 
-can create a special type of event called a {\tt UserEvent} (line 15) that the
-user has the power to trigger.  A user event is a sub-type of an event
-that has a {\tt trigger} method (line 18).  Users can also create another
-type of event called a {\tt Barrier} that will require multiple arrivals before
-the event is triggered.  The user can manage the number of arrivals that must
-be seen as well as how many arrivals occur at a time.  Once a barrier's arrival
-count goes to zero, the barrier will trigger.  Note that barriers in our interface
-provide a superset of the functionality of traditional barriers since they can
-be used in a blocking manner (via the {\tt wait} call), but are primarily used
-asynchronously.  Lastly, since user events and barriers are sub-types of 
-an event they can be used wherever an event is required.
+operations and the implementation is responsible for triggering these events.  Clients
+can also create a {\tt UserEvent} (line 15) that are triggered explicitly by the client.
+
+The common synchronization pattern in which multiple consumers are dependent on a 
+(potentially different) set of producers is supported by a {\tt Barrier} object,
+which is similar to a {\tt UserEvent}, but does not trigger until the expected
+number of {\tt arrive} operations have occurred.  Barrier constructs exist in other
+programming models\cite{MPI}, but in addition to the decoupling of the waiters from
+the arrivers, our barrier interface has two further improvements to better support
+composability and asynchronous operation:
+\begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
+\item An arrival operation can be made dependent on another event, allowing a launch 
+of a subtask or a copy of data to be completed before the barrier can trigger.
+\item The expected arrival count of a barrier can be dynamically increased, allowing
+a producer task to be split into multiple subtasks without the original task having to wait for
+their completion to signal its arrival.
+\end{itemize}
+
+Although {\tt UserEvent}s and {\tt Barrier}s are created and triggered differently,
+any operation that expresses a dependence on an {\tt Event} can transparently use
+a {\tt UserEvent} or {\tt Barrier} as well.
+
+%%   Users can also create another
+%% type of event called a {\tt Barrier} that will require multiple arrivals before
+%% the event is triggered.  The user can manage the number of arrivals that must
+%% be seen as well as how many arrivals occur at a time.  Once a barrier's arrival
+%% count goes to zero, the barrier will trigger.  Note that barriers in our interface
+%% provide a superset of the functionality of traditional barriers since they can
+%% be used in a blocking manner (via the {\tt wait} call), but are primarily used
+%% asynchronously.  Lastly, since user events and barriers are sub-types of 
+%% an event they can be used wherever an event is required.
 
@@
   void alter_arrival_count(int delta) const;
-  void arrive(unsigned count = 1) const;
+  void arrive(unsigned count = 1,
+                  Event wait_for = Event::NO_EVENT) const;
 };
@@
 to include new processor types (line 10).   Processors support a single {\tt spawn}
-operation (lines 13-14) that will launch a new {\em task} on that processor.
+operation (lines 14-15) that will launch a new {\em task} on that processor.
 Note that because the spawn operation
@@
 
-Lines 17-23 of Listing~\ref{lst:procapi} show a simple example using 
+Lines 18-24 of Listing~\ref{lst:procapi} show a simple example using 
 the processor and event interface.  The {\tt diamond\_task} launches four
@@
   };
+  Kind kind() const;
 
@@
 
-Between deffered locks and barriers (described in Section~\ref{subsec:events}) clients 
+Between deferred locks and barriers (described in Section~\ref{subsec:events}) clients 
 have access to the same set of synchronization primitves that they traditionally have
@@
 regions are grouped into {\em classes} which are sets of physical regions that share the
-same number and size of elements, but don't necessarily share the same data layout.  Our
+same names (i.e. addresses) for elements, but don't necessarily share the same data layout.  This
+allows pointers to elements within regions to be used, even if the regions are copied from one memory
+to another.  Our
 interface supports sub-classing, but the details are omitted due to space constraints.  The
-interface only supports data movement operations between physical regions that are of
-the same class (or between super- and sub-classes).  Note, clients can still pass arbitrary arrays
-of bits via the processor {\tt spawn} call.
+interface allows copies between any pair of physical regions that are of
+the same class (or between super- and sub-classes).
+
+%Note, clients can still pass arbitrary arrays
+%of bits via the processor {\tt spawn} call.
+
 % Say something about dynamically modifying the number of elements in a class
@@
 
-Reductions and reduction instances illustrate only a single case where physical regions provide 
-the structure necessary to conjoin an operation with data movement.  Physcal regions make our
-interface easily extensible to arbitrary transformations and operations as a part of data movment.
-While physical regions are a slightly higher-level construct than an annonymous
-buffer of bits, we believe that the optimizations they enable are important in an asynchronous
-environment.  Clients also still have access to an interface for controlling data layout through
-the operations associated with physical regions which mitigates the loss in generality.
+%% SJT - redundant with first paragraph in this subsection?
+%% Reductions and reduction instances illustrate only a single case where physical regions provide 
+%% the structure necessary to conjoin an operation with data movement.  Physcal regions make our
+%% interface easily extensible to arbitrary transformations and operations as a part of data movment.
+%% While physical regions are a slightly higher-level construct than an annonymous
+%% buffer of bits, we believe that the optimizations they enable are important in an asynchronous
+%% environment.  Clients also still have access to an interface for controlling data layout through
+%% the operations associated with physical regions which mitigates the loss in generality.
 
@@
 \label{subsec:machmodel}
-The last component of the interface allows for introspection of the current machine.
-Listing~\ref{lst:machineapi} shows the machine interface.
-The {\tt Machine} object is a singleton object that serves two purposes.  First, it
-is the mechanism for initializing the runtime; at the start of a program the client 
-creates a machine object, instantiates a task table mapping task IDs to function pointers,
-and invokes the {\tt run} method (line 30-31).
-Second, the machine object provides an interface for the client to 
-discover properties of the underlying machine (lines 35-43).  The machine object maintains
-sets of all the unique {\tt Processor} and {\tt Memory} handles in the system.  Processors
-were described in Section~\ref{subsec:procs}.  For every memory in the system there also 
-exists a {\tt Memory} handle with a unique ID.  Different memories often, but not always, 
-imply a different address space.  Examples of different memories are described in
-Section~\ref{sec:impl}.
+The last component of the interface is shown in Listing~\ref{lst:machineapi} and allows the
+initialization and run-time introspection
+of the current machine.
+The {\tt Machine} object is a singleton object that the client creates at the start of a
+program, providing a task table that maps task IDs to function pointers,
+and then invoking the {\tt run} method (line 15-16).
+Any running task can get a pointer to the {\tt Machine} object with the
+{\tt get\_machine} method (line 18) and use it to obtain lists of all the {\tt Memory} (line 19) or 
+{\tt Processor} (line 20) handles in the system.  Using the affinity methods (lines 22-25),
+the application can also determine:
+\begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
+\item for each {\tt Processor}, which {\tt Memory} objects are accessible and with what performance, and
+\item which pairs of {\tt Memory} objects are able to efficiently copy data between them.
+\end{itemize}
+
+%% in the machine.
+%% Second, the machine object provides an interface for the client to 
+%% discover properties of the underlying machine (lines 35-43).  The machine object maintains
+%% sets of all the unique {\tt Processor} and {\tt Memory} handles in the system.  Processors
+%% were described in Section~\ref{subsec:procs}.  For every memory in the system there also 
+%% exists a {\tt Memory} handle with a unique ID.  Different memories often, but not always, 
+%% imply a different address space.  Examples of different memories are described in
+%% Section~\ref{sec:impl}.
 
@@
 
-The machine object also maintains two relations between the set of processors
-and the set of memories:
-\begin{itemize}
-\item Processor-Memory: for each processor-memory pair whether
-the processor can directly access (read and write) the memory and 
-latency and bandwidth properties if access is possible
-\item Memory-Memory: for every pair of memories whether copies can be
-performed and latency and bandwidth properties if copies are possible
-\end{itemize}
-
-The client can query these two relations by the {\tt get\_*\_affinity} calls
-(lines 40-43) to discover properties of the machine.
+%% The machine object also maintains two relations between the set of processors
+%% and the set of memories:
+%% \begin{itemize}
+%% \item Processor-Memory: for each processor-memory pair whether
+%% the processor can directly access (read and write) the memory and 
+%% latency and bandwidth properties if access is possible
+%% \item Memory-Memory: for every pair of memories whether copies can be
+%% performed and latency and bandwidth properties if copies are possible
+%% \end{itemize}
+%%
+%% The client can query these two relations by the {\tt get\_*\_affinity} calls
+%% (lines 40-43) to discover properties of the machine.
 
@@
 \begin{lstlisting}[float={t},label={lst:machineapi},caption={Machine Interface.}]
-
 class Memory {
@@
   const unsigned id;
+  size_t size() const;
 };
@@
 public:
-  static Machine* get_machine(void); // pointer to Machine
-  // References to all processors and memories
+  static Machine* get_machine(void);
   const set<Memory>& get_all_memories(void) const;
   const set<Processor>& get_all_processors(void) const;
-  Processor::Kind get_processor_kind(Processor p) const;
-  size_t get_memory_size(Memory m) const;
-  // Query properties of underlying hardware
+
   int get_proc_mem_affinity(vector<ProcMemAffinity &result,

### 2012-07-23 doc/asplos2013/legion-llr.tex

@@
 of data through the memory hierarchy as well as {\em deferred locks}, a novel synchronization primitive.
-Asynchrony is exposed by the the interface using a distributed event system.  All
+Asynchrony is exposed by the interface using a distributed event system.  All
 operations return an event to be triggered when the operation completes.  Events
-can be chained together and used to create dependences between operations.
+can be chained together and used to capture dependences between operations.
 

### 2012-07-23 doc/asplos2013/results.tex

@@
 \label{subsec:eventlife}
+
 We instrumented the heterogenous implementation of our interface to 
-capture information about different kinds of events.  Figure~\ref{fig:eventlife} shows
-the lifetimes of different types of events from a single run of the Fluid
-application on 16 nodes using 8 processors per node.  Dynamic events 
-is a monotonically increasing line that
-corresponds to the total number of events created in the system.  By
-the end of the run, the simulation created over 260,000 dynamic
-events.  In contrast, the physical events line corresponds to the total number
-of generational events required to represent the dynamic events 
-across all the nodes in the system.  Less than 5,000
-generational events were required to represent all the dynamic events
-illustrating that the technique of mapping from dynamic events to generational 
-events described in Section~\ref{subsec:eventimpl} was effective.
+capture event usage information.  Figure~\ref{fig:eventlife} shows
+a timeline of the execution of the Fluid application on 16 nodes.
+The dynamic events line measures the total number of event creations.
+A large number of events are created - over 260,000 in less than 15
+seconds - and allocating separate storage for every event would clearly
+be difficult for long-running applications.  The {\em live events} line
+measures the number of events that are live at any given time - after
+the last operation on an event, a reference counting implementation would
+be able to recover the storage associated with the event.  Such an
+algorithm would reduce the storage needed by over 10X at the cost of
+computation and communication overhead for performing the reference 
+counting.  As discussed in Section~\ref{subsec:eventimpl}, our implementation
+requires storage that grows with the number of {\em untriggered events}, a number
+that is 10X smaller than even the maximal live event count.  The actual storage
+requirements of our implementation are shown by the {\em physical events} line,
+which is larger than the peak number of untriggered events because of the
+decision to always create an event owned by the node from which the request
+was made.  Despite this inefficiency, our implementation uses 5X less storage
+than a reference counting implementation and avoids any overhead related to 
+reference counting.  These savings would likely be even more dramatic for longer 
+runs of the application, as the number of live events is steadily growing as the
+application runs, while the peak number of generational events needed appears to 
+occur during the startup of the application.
 
-The live event line shows the total number of {\em live events} at a point in time.  Dynamic events
-are considered {\em live events} until their last query has been performed.  The number of live
-events is equivalent to the number of physical events that would be required
-in a reference counted scheme.  In this case we see that by reusing
-generational events, we require up to 4X fewer generational events than would be
-required by a reference counted scheme.
+%% much storage would be required for events using
 
-In an ideal world the total number of generational events required would be
-equivalent to the maximum number of generational events in the untriggered
-state at any point.  The untriggered event line shows the number of generational
-events in the untriggered state.  We see that the total number of generational
-events is actually slightly higher than the maximum number of generational
-events in the untriggered state at any point.  The reason for this is that unused
-generational events cannot be shared across nodes, so in some cases nodes must
-make new generational events eventhough there are unused generational events
-on other nodes.  The small difference in these lines shows that in practice
-this is only a minor inefficiency.
+%%  about different kinds of events.  Figure~\ref{fig:eventlife} shows
+%% the lifetimes of different types of events from a single run of the Fluid
+%% application on 16 nodes using 8 processors per node.  Dynamic events 
+%% is a monotonically increasing line that
+%% corresponds to the total number of events created in the system.  By
+%% the end of the run, the simulation created over 260,000 dynamic
+%% events.  In contrast, the physical events line corresponds to the total number
+%% of generational events required to represent the dynamic events 
+%% across all the nodes in the system.  Less than 5,000
+%% generational events were required to represent all the dynamic events
+%% illustrating that the technique of mapping from dynamic events to generational 
+%% events described in Section~\ref{subsec:eventimpl} was effective.
+
+%% The live event line shows the total number of {\em live events} at a point in time.  Dynamic events
+%% are considered {\em live events} until their last query has been performed.  The number of live
+%% events is equivalent to the number of physical events that would be required
+%% in a reference counted scheme.  In this case we see that by reusing
+%% generational events, we require up to 4X fewer generational events than would be
+%% required by a reference counted scheme.
+
+%% In an ideal world the total number of generational events required would be
+%% equivalent to the maximum number of generational events in the untriggered
+%% state at any point.  The untriggered event line shows the number of generational
+%% events in the untriggered state.  We see that the total number of generational
+%% events is actually slightly higher than the maximum number of generational
+%% events in the untriggered state at any point.  The reason for this is that unused
+%% generational events cannot be shared across nodes, so in some cases nodes must
+%% make new generational events eventhough there are unused generational events
+%% on other nodes.  The small difference in these lines shows that in practice
+%% this is only a minor inefficiency.
 

### 2012-07-23 doc/asplos2013/micro.tex
small edits to the microbenchmarks

@@
 For small fan-in/out factors, the total rate falls off initially going to two nodes as active
-message become necessary, but starts increasing slightly again at larger node counts.  Higher
+messages become necessary, but starts increasing slightly again at larger node counts.  Higher
 fan-in/out factors require more messages and have lower throughput that also increases with
@@
 the number of chains result in improved rates.  Note that the point on each line where this occurs
- is when
-the chains per node is greater than total locks, i.e. where the expected number of requests per lock 
+is when the chains per node is greater than total locks, which corresponds to the point where the expected 
+number of requests per lock 
 per node becomes more than one.  As soon as there are multiple expected requestors for the same lock on
-a node, the unfairness property of our implementation is able to reduce the number of lock ownership 
+a node, the unfairness property of our lock implementation is able to reduce the number of lock ownership 
 transfers, yielding better performance.
@@
 \label{subsec:reducmicro}
-To show the benefits of associating reduction operations with physical regions we designed
+To show the benefits of associating reduction instances we designed
 a histogram micro-benchmark that requires all nodes to perform an addition reduction to a
@@
 region in its own system memory, performs all reductions to the localized region, copies the region back, and releases the lock.
-\item Fold Instance - Each node creates a local reduction instance that folds reductions to the same element.
+\item Fold Instance - Each node creates a local fold reduction instance.
 All reductions are folded into the instance, which is then copied back to the global region.
-\item List Instance - Each node creates a local reduction instance that stores a list of reduction operations.
-All reductions are appended to the list, which is then copied back to the global region.
+\item List Instance - Each node creates a local list reduction instance.
+All reductions are appended to the list reduction instance, which is then copied back to the global region.
 \end{itemize}
@@
 separate reduction tasks.  There are 8 reduction tasks created for each node, one per processor.  The dense
-case uses a histogram with 256K buckets and each reduction task processes a chunk of 4M entries (i.e.
-performs 4M reductions).  The results for the
+case uses a histogram with 256K buckets and each reduction task performs 4M reductions.  The results for the
 dense experiment are shown in Figure~\ref{fig:reducdense}.  The sparse case use a histogram with 4M buckets,
-but only 64K elements are processed by each task.  Results for the sparse case
+but only 64K reductions are performed by each task.  Results for the sparse case
 be seen in Figure~\ref{fig:reducsparse}.
@@
 Finally, the only time the latency of performing individual RMA reads and writes isn't disastrous is
-on a single node (where the reads and writes aren't really remote).
+on a single node (where the reads and writes are local).
 

### 2012-07-23 doc/asplos2013/interface.tex
event dependence graphs for code examples

@@
 the processor and event interface.  The {\tt diamond\_task} launches four
-sub-tasks with a diamond dependence pattern.  The {\tt merge\_events} call
+sub-tasks with a diamond dependence pattern, as shown in Figure~\ref{fig:procevents}.
+The {\tt merge\_events} call
 is used to merge the events {\tt eb} and {\tt ec} to create the shared dependence
@@
 
+\begin{figure}
+\centering
+\scalebox{0.8}{
+\begin{tikzpicture}
+  %\path (1,3) node (t1) [shape=rectangle,draw] {run TASK\_A} -- node [right=1pt] {ea} +(0,-0.5) node[auto] {eax};
+  \node (t1) at (0,0) [shape=rectangle,draw] {\begin{tabular}{c}{\bf run}\\TASK\_A\end{tabular}};
+  \draw (t1) to node[auto] {ea} (1.4,0);
+
+  \node (t2) at (2.8,1) [shape=rectangle,draw] {\begin{tabular}{c}{\bf run}\\TASK\_B\end{tabular}};
+  \draw (t2) to node[auto] {eb} (4.2,1);
+  \draw [->] (1.4,0) to (t2.west);
+
+  \node (t3) at (2.8,-1) [shape=rectangle,draw] {\begin{tabular}{c}{\bf run}\\TASK\_C\end{tabular}};
+  \draw (t3) to node[auto] {ec} (4.2,-1);
+  \draw [->] (1.4,0) to (t3.west);
+
+  \node (t4j) at (4.6,0) [shape=circle,inner sep=2pt,draw] {+};
+  \draw (t4j) to node[auto] {em} (5.4,0);
+  \draw [->] (4.2,1) to (t4j);
+  \draw [->] (4.2,-1) to (t4j);
+
+  \node (t4) at (6.4,0) [shape=rectangle,draw] {\begin{tabular}{c}{\bf run}\\TASK\_D\end{tabular}};
+  \draw [->] (t4) to node[auto] {ed} (7.9,0);
+  \draw [->] (5.4,0) to (t4.west);
+  %
+\end{tikzpicture}
+}
+\caption{Processor Example Event Graph\label{fig:procevents}}
+\end{figure}
+
 \subsection{Deferred Locks}
@@
 {\tt SUB} task can run safely while holding the lock and the lock is released when the
-{\tt SUB} task completes.  Using deferred locks in this style prevents 
+{\tt SUB} task completes.  Figure~\ref{fig:lockevents} shows the event dependence graph, with
+dotted lines representing a lock operation's dependence on a previous unlock.
+Using deferred locks in this style prevents 
 compute resources from waiting on locks by ensuring tasks only run when their needed
@@
 
+\begin{figure}
+\centering
+\scalebox{0.8}{
+\begin{tikzpicture}
+  %\path (1,3) node (t1) [shape=rectangle,draw] {run TASK\_A} -- node [right=1pt] {ea} +(0,-0.5) node[auto] {eax};
+  \node (t1) at (0,0) [shape=rectangle,draw] {\begin{tabular}{c}{\bf lock}\\needed\end{tabular}};
+  \draw (t1) to node[auto] {lock\_event} (2.4,0);
+
+  \node (t2) at (3.2,0) [shape=rectangle,draw] {\begin{tabular}{c}{\bf run}\\SUB\end{tabular}};
+  \draw (t2) to node[auto] {task\_event} (5.4,0);
+  \draw [->] (2.4,0) to (t2.west);
+
+  \node (t3) at (6.4,0) [shape=rectangle,draw] {\begin{tabular}{c}{\bf unlock}\\needed\end{tabular}};
+  \draw [->] (5.4,0) to (t3.west);
+
+  \draw [->,dotted,thick] (-1,1) to (t1.135);
+  \draw [->,dotted,thick] (t3.45) to (7.4,1);
+\end{tikzpicture}
+}
+\caption{Lock Example Event Graph\label{fig:lockevents}}
+\end{figure}
+
 \subsection{Physical Regions}
@@
 
+\begin{lstlisting}[float={t},label={lst:regionapi},caption={Physical Region Interface and Example.}]
+class RegionMetaData {
+public:
+  const unsigned id;
+  static const RegionMetaData NO_REGION;
+  // Create and destroy metadatas
+  static RegionMetaData create_region(size_t num_elmts, 
+                                      size_t elmt_size);
+  void destroy_region() const;
+  // Create and destroy instances
+  RegionInstance create_instance(Memory memory) const;
+  void destroy_instance(RegionInstance instance) const;
+  // Create a physical instance only for reductions
+  RegionInstance create_instance(Memory memory, 
+                        ReductionOpID redopid) const;
+};
+
+class RegionInstance {
+public:
+  const unsigned id;
+  static const RegionInstance NO_INST;
+  // Copy between instances
+  Event copy_to(RegionInstance target, 
+                Event wait_on = Event::NO_EVENT);
+};
+
+template<typename LHS, typename RHS>
+class ReductionOp {
+public:
+  static void apply(LHS *lhs, RHS rhs);
+  // Optional methods for folding optimization
+  static void fold(RHS *rhs1, RHS rhs2);
+  static const RHS right_identity;
+};
+
+void reduction_ex(const void *args,size_t arglen,Processor p) {
+  Memory m1,m2,m3; Processor p1,p2,p3; ReductionOpID redop = 2;
+  ... // Choose memory and processors
+  RegionMetaData meta = RegionMetaData::
+                        create_region(num_elmts,elmt_size);
+  Instance inst1 = meta.create_instance(m1,redop);
+  Instance inst2 = meta.create_instance(m2,redop);
+  Instance inst3 = meta.create_instance(m3);
+  Event t1 = p1.spawn(REDUC,&inst1,sizeof(RegionInstance));
+  Event t2 = p2.spawn(REDUC,&inst2,sizeof(RegionInstance));
+  Event c1 = inst1.copy_to(inst3,t1);
+  Event c2 = inst2.copy_to(inst3,t2);
+  Event em = Event::merge_events(c1,c2);
+  Event t3 = p3.spawn(USE,&inst3,sizeof(RegionInstance),em);
+}
+\end{lstlisting}
+
+\begin{figure}
+\centering
+\scalebox{0.8}{
+\begin{tikzpicture}
+  %\path (1,3) node (t1) [shape=rectangle,draw] {run TASK\_A} -- node [right=1pt] {ea} +(0,-0.5) node[auto] {eax};
+  \node (t1) at (0,1) [shape=rectangle,draw] {\begin{tabular}{c}{\bf run}\\REDUC(inst1)\end{tabular}};
+  \draw (t1) to node[auto] {t1} (1.8,1);
+
+  \node (t2) at (0,-1) [shape=rectangle,draw] {\begin{tabular}{c}{\bf run}\\REDUC(inst2)\end{tabular}};
+  \draw (t2) to node[auto] {t2} (1.8,-1);
+
+  \node (t3) at (3.2,1) [shape=rectangle,draw] {\begin{tabular}{c}{\bf copy}\\inst1 $\rightarrow$ inst3\end{tabular}};
+  \draw (t3) to node[auto] {c1} (4.9,1);
+  \draw [->] (1.8,1) to (t3.west);
+
+  \node (t4) at (3.2,-1) [shape=rectangle,draw] {\begin{tabular}{c}{\bf copy}\\inst2 $\rightarrow$ inst3\end{tabular}};
+  \draw (t4) to node[auto] {c2} (4.9,-1);
+  \draw [->] (1.8,-1) to (t4.west);
+
+  \node (t4j) at (5.3,0) [shape=circle,inner sep=2pt,draw] {+};
+  \draw (t4j) to node[auto] {em} (6.1,0);
+  \draw [->] (4.9,1) to (t4j);
+  \draw [->] (4.9,-1) to (t4j);
+
+  \node (t5) at (7.3,0) [shape=rectangle,draw] {\begin{tabular}{c}{\bf run}\\USE(inst3)\end{tabular}};
+  \draw [->] (t5) to node[auto] {t3} (9,0);
+  \draw [->] (6.1,0) to (t5.west);
+
+\end{tikzpicture}
+}
+\caption{Reduction Example Event Graph\label{fig:reducevents}}
+\end{figure}
+
 Listing~\ref{lst:regionapi} shows a subset of the interface for physical regions.  The
@@
 when the object is created (line 13).  Reduction-only {\tt RegionInstance} objects can 
-increase the available parallelism and hide latency better, as we will show in section~\ref{sec:microbenchmarks}.  The interface supports any reduction operation that
+increase the available parallelism and hide latency better, as we will show in 
+section~\ref{subsec:reducmicro}.  The interface supports any reduction operation that
 is commutative (i.e. multiple operations to the same location can be reordered without 
@@
 applications.  Finally, a merge on the events from the movement and reduction applications is 
-performed and a final task using {\tt inst3} is launched (lines 47-48).
+performed and a final task using {\tt inst3} is launched (lines 47-48).  The resulting event
+graph is shown in Figure~\ref{fig:reducevents}.
 
@@
 %in Section~\ref{subsec:reducimpl}.
-
-\begin{lstlisting}[float={t},label={lst:regionapi},caption={Physical Region Interface and Example.}]
-class RegionMetaData {
-public:
-  const unsigned id;
-  static const RegionMetaData NO_REGION;
-  // Create and destroy metadatas
-  static RegionMetaData create_region(size_t num_elmts, 
-                                      size_t elmt_size);
-  void destroy_region() const;
-  // Create and destroy instances
-  RegionInstance create_instance(Memory memory) const;
-  void destroy_instance(RegionInstance instance) const;
-  // Create a physical instance only for reductions
-  RegionInstance create_instance(Memory memory, 
-                        ReductionOpID redopid) const;
-};
-
-class RegionInstance {
-public:
-  const unsigned id;
-  static const RegionInstance NO_INST;
-  // Copy between instances
-  Event copy_to(RegionInstance target, 
-                Event wait_on = Event::NO_EVENT);
-};
-
-template<typename LHS, typename RHS>
-class ReductionOp {
-public:
-  static void apply(LHS *lhs, RHS rhs);
-  // Optional methods for folding optimization
-  static void fold(RHS *rhs1, RHS rhs2);
-  static const RHS right_identity;
-};
-
-void reduction_ex(const void *args,size_t arglen,Processor p) {
-  Memory m1,m2,m3; Processor p1,p2,p3; ReductionOpID redop = 2;
-  ... // Choose memory and processors
-  RegionMetaData meta = RegionMetaData::
-                        create_region(num_elmts,elmt_size);
-  Instance inst1 = meta.create_instance(m1,redop);
-  Instance inst2 = meta.create_instance(m2,redop);
-  Instance inst3 = meta.create_instance(m3);
-  Event t1 = p1.spawn(REDUC,&inst1,sizeof(RegionInstance));
-  Event t2 = p2.spawn(REDUC,&inst2,sizeof(RegionInstance));
-  Event c1 = inst1.copy_to(inst3,t1);
-  Event c2 = inst2.copy_to(inst3,t2);
-  Event em = Event::merge_events(c1,c2);
-  Event t3 = p3.spawn(USE,&inst3,sizeof(RegionInstance),em);
-}
-\end{lstlisting}
 

### 2012-07-23 doc/asplos2013/results.tex
Some minor edits to the results

@@
 that takes a stream of tasks with region requirements and asynchronously
-runs ahead of the actual execution using a {\em software-out-of-order} (SOOP)
-scheduler.  The SOOP leverages our interface by asynchronously issuing task launches, copy
+runs ahead of the actual execution using a scheduler called a {\em Software Out-of-Order Processor} 
+(SOOP).  The SOOP leverages our interface by asynchronously issuing task launches, copy
 operations, and synchronization operations with the dependencies between them expressed
@@
 
-{\em Circuit} is simulation of an electric circuit that is described as an
-irregular graph of integrated
-circuit components connected by wires.  The graph is partitioned
+{\em Circuit} is a simulation of an integrated circuit that is described as an
+irregular graph of components connected by wires.  The graph is partitioned
 and distributed across the machine.  The computation for each time step consists
 of three phases that calculate currents on the wires, distribute charge between
-nodes, and update the voltages of all ndoes.  In the distribute charge
+nodes, and update the voltages of all nodes.  In the distribute charge
 phase, a single node may be updated by wires from multiple partitions.  Reduction instances 
@@
 seconds - and allocating separate storage for every event would clearly
-be difficult for long-running applications.  The {\em live events} line
-measures the number of events that are live at any given time - after
+be difficult for long-running applications.
+
+Dynamic events are considered {\em live events} until their last operation 
+(i.e. wait, trigger) has been performed.  The live events line
+measures the number of events that are live at any given time.  After
 the last operation on an event, a reference counting implementation would
-be able to recover the storage associated with the event.  Such an
-algorithm would reduce the storage needed by over 10X at the cost of
-computation and communication overhead for performing the reference 
-counting.  As discussed in Section~\ref{subsec:eventimpl}, our implementation
-requires storage that grows with the number of {\em untriggered events}, a number
+be able to recover the storage associated with the event.  In this example 
+a reference counting algorithm would reduce the storage needed for dynanmic events
+by over 10X, but could only do so at the cost of
+computation and communication for performing the reference counting.
+
+As discussed in Section~\ref{subsec:eventimpl}, our implementation
+requires storage that grows with the maximum number of {\em untriggered events}, a number
 that is 10X smaller than even the maximal live event count.  The actual storage
 requirements of our implementation are shown by the {\em physical events} line,
-which is larger than the peak number of untriggered events because of the
-decision to always create an event owned by the node from which the request
-was made.  Despite this inefficiency, our implementation uses 5X less storage
+which is slightly larger than the peak number of untriggered events. 
+The reason for this difference is because nodes create events locally if they have
+no free generational events, even if there are unused generational events on remote nodes.
+Despite this inefficiency, our implementation uses 5X less storage
 than a reference counting implementation and avoids any overhead related to 
@@
 which they were created.  The AMR application exhibits more varied locking behavior, and in 
-particular has locks that are requested more often than they are transferred.  Without the ability
-for locks to migrate, more of those requests would have required inter-node communication.
+particular has locks that are requested more often than they are transferred.  Without the
+ability for locks to migrate, these applications couldn't have made use of asynchronous
+locks.
 
@@
 Legion in a bulk-synchronous style to compare against our existing implementations.  
-These bulk-synchronous versions of still run on top of 
+These bulk-synchronous versions still run on top of 
 the Legion runtime, but invoke blocking operations between computational phases,
@@
 The benefit of this latency hiding grows as the number of
-nodes increases and as the computation-to-communication ratio decreases.
+nodes increases. % and as the computation-to-communication ratio decreases.
 In all cases, the overhead of the bulk-synchronous

### 2012-07-23 doc/asplos2013/related.tex
Minor edits

@@
 introduced in these languages are higher-level and have complex 
-semantics.  We view our interface as providing a potential target
+semantics.  We view our interface as providing a target
 for implementing these higher-level asynchronous operations in an efficient
 manner similar to Legion\cite{Legion12}.
+
+Physical regions in our interface are related to
+arrays from the Sequoia runtime interface\cite{Houston08} in that they
+both allow the runtime to do specialized operations on data by being
+aware of the structure of data.  Sequoia's runtime interface supported
+asynchronous tasks and copies, but did not permit composition of
+asynchronous operations.
 

### 2012-07-23 doc/asplos2013/impl.tex

@@
 to new events without the risk of conflicting with another node's assignments.  The static division
-of event handles permits any node to determine the owning node of an event
+of event handles also permits any node to determine the owning node of an event
 without any communication.
@@
 When the event does trigger, the owner node notifies all local waiters and
-sends a single {\em event trigger} active message to all nodes from which it has seen an event subscription
+sends a single {\em event trigger} active message to each node from which it has seen an event subscription
 to notify remote waiters.
@@
 While a remote trigger of an event can result in
-the latency of a triggering operation being two active messages, it bounds the number of active
+the latency of a triggering operation including two active message flight times, it bounds the number of active
 messages required per event to $2N-2$ where $N$ is the number of nodes 
@@
 %As before, this can be done with no inter-node communication.
-Instead of attempting to free event data structures, our implementation aggressively recyles them.
+Instead of attempting to free event data structures, our implementation aggressively recycles them.
 Compared to a reference counting scheme our implementation requires fewer total event data structures
@@
 new operation that is dependent on a generation that is known to already be triggered can immediately
-be executed.  A generational event can be recycled as soon as the current generation has triggered. To
+be executed.  A generational event can be reused for a new generation as soon as the current generation has triggered. To
 create a new event, a node finds a generational event in the triggered state,
 increases the generation by one, and sets the generational event's state to untriggered.  As before, 
-this can be done with no communication.
+this can be done with no inter-node communication.
 
@@
 of the system allows remote generational events to perform an interesting optimization.  If at any point a
-remote generational event observes a request to wait on a later generation than its current generation,
+remote generational event receives a request to wait on a later generation than its current
+generation,
 then it is safe to infer that all generations up to the requested generation have already 
 been triggered because a new generation of the event was already created by the event's owner node.
-All local waiters can then be notified even before receiving an event triggered message for the current generation.
+All local waiters can then be notified even before receiving the event trigger message for the current generation.
 
@@
 active messages containing an adjustment value that is positive for {\tt alter\_arrival\_count} operations 
-and negative for {\tt arrive} operations.  A version of vector clocks\cite{Fidge1998} is used
+and negative for {\tt arrive} operations.  A simplified version of vector clocks\cite{Fidge1998} is used
 to detect the race condition in which the barrier owner receives an {\tt arrive} operation before the
@@
 involved in transferring a lock between nodes can be the limiter on throughput.  By minimizing the number
-of lock transfers, the throughput on the lock is maximized.
+of lock transfers, the throughput on the lock is maximized.  This effect will be demonstrated in
+Section~\ref{subsec:lockmicro}.
 
@@
 location of the data. This decoupling from the original location of the data can allow allow greater parallelism
-in cases where processors cannot all access the same memory (i.e. many GPUs).
+in cases where processors cannot all access the same memory (e.g. many GPUs).
 

### 2012-07-23 doc/asplos2013/impl.tex
move discussion of lock payloads to interface section

@@
 
-Locks are often used to mediate access to data, and the size of that data is often small.  To
-support this common case we allow locks to be created with an associated
-%avoid adding the latency of inter-node data copies to critical sections, a lock may also be created with a
-{\em payload}, a small (i.e. less than 4KB - the limit for an active message payload) 
-piece of data that is moved around with the lock and guaranteed to be coherent while 
-the lock is held.
-
 Since any node may at some point be the owner of a lock, all nodes use the same data structure to track the
@@
 message.  A lock transfer message includes the bitmask of remote waiters and an up-to-date copy 
-of the lock's payload.
+of the lock's payload.  (The inclusion of the payload in the active message is the reason for the 
+4KB limit on a payload's size.)
 

### 2012-07-23 doc/asplos2013/interface.tex
move discussion of lock payloads to interface section

@@
 
-Listing~\ref{lst:lockapi} shows the 
-interface for locks.  Unlike blocking locks, the {\tt lock} method (line 4) doesn't
-block but instead always returns immediately with an event that will be triggered
-when the lock has been acquired.  Just like the spawn method, the lock method also 
-takes an optional event parameter as a precondition.  Similarly, the {\tt unlock} 
-method (line 5) takes an optional event precondition parameter.
-
-An important difference between deferred locks and blocking locks is that the processor
-that requests the lock doesn't have to be the one that uses it.  A common
-convention in writing code with deferred locks is to acquire locks on behalf of a task being
-launched.  Listing~\ref{lst:lockapi} illustrates this with a simple example.  The
-{\tt launcher\_task} is calling a sub-task that is going to require the {\tt needed}
-lock.  A lock request is issued and the resulting {\tt lock\_event} is used as
-a precondition for launching the {\tt SUB} task.  The unlock operation is then
-precondition on the event corresponding to {\tt SUB} task's completion.  The
-{\tt SUB} task can run safely while holding the lock and the lock is released when the
-{\tt SUB} task completes.  Figure~\ref{fig:lockevents} shows the event dependence graph, with
-dotted lines representing a lock operation's dependence on a previous unlock.
-Using deferred locks in this style prevents 
-compute resources from waiting on locks by ensuring tasks only run when their needed
-locks have already been acquired.
-
-Between deferred locks and barriers (described in Section~\ref{subsec:events}) clients 
-have access to the same set of synchronization primitves that they traditionally have
-in both threading and bulk-synchronous interfaces.  However, these operations can now 
-be composed with other asynchronous operations which is not possible in any other interface
-of which we are aware.
-
-%Deferred locks provide a super-set of the functionality of blocking locks.  As their
-%name suggests, deferred locks can use the event corresponding to their lock acquire
-%operation to defer execution until the lock acquire has been granted.  Deferred
-%locks can also be converted back into a blocking lock by immediately waiting on the event
-%returned from a call to {\tt lock}.  In addition, deffered locks also provide {\tt mode} and
-%and {\tt exclusive} parameters that allow the user to specify whether other requests can acquire
-%the lock simultaneously.  A lock can only be acquired in one mode at a time.  The
-%{\tt exclusive} parameter specifies whether other owners are permitted once the
-%current lock request is granted.
-
 \begin{lstlisting}[float={t},label={lst:lockapi},caption={Deferred Lock Interface and Example.}]
@@
 \end{figure}
+
+Listing~\ref{lst:lockapi} shows the 
+interface for locks.  Unlike blocking locks, the {\tt lock} method (line 4) doesn't
+block but instead always returns immediately with an event that will be triggered
+when the lock has been acquired.  Just like the spawn method, the lock method also 
+takes an optional event parameter as a precondition.  Similarly, the {\tt unlock} 
+method (line 5) takes an optional event precondition parameter.
+
+An important difference between deferred locks and blocking locks is that the processor
+that requests the lock doesn't have to be the one that uses it.  A common
+convention in writing code with deferred locks is to acquire locks on behalf of a task being
+launched.  Listing~\ref{lst:lockapi} illustrates this with a simple example.  The
+{\tt launcher\_task} is calling a sub-task that is going to require the {\tt needed}
+lock.  A lock request is issued and the resulting {\tt lock\_event} is used as
+a precondition for launching the {\tt SUB} task.  The unlock operation is then
+precondition on the event corresponding to {\tt SUB} task's completion.  The
+{\tt SUB} task can run safely while holding the lock and the lock is released when the
+{\tt SUB} task completes.  Figure~\ref{fig:lockevents} shows the event dependence graph, with
+dotted lines representing a lock operation's dependence on a previous unlock.
+Using deferred locks in this style prevents 
+compute resources from waiting on locks by ensuring tasks only run when their needed
+locks have already been acquired.
+
+When a lock is used to mediate access to a piece of data, the first thing a task will usually
+do after being granted the lock is access the data.  To be able to hide the latency of that
+access as well, an additional feature of locks in our interface is the association of a lock with a
+{\em payload}, a small (i.e. less than 4KB) 
+piece of data that is moved around with the lock and guaranteed to be coherent while 
+the lock is held.  The size of the payload is specified when the lock is created (line 7) and
+the pointer to the local copy of the payload can be obtained from the {\tt payload\_ptr} method
+(line 8).
+
+Locks are often used to mediate access to data, and the size of that data is often small.  To
+support this common case we allow locks to be created with an associated
+%avoid adding the latency of inter-node data copies to critical sections, a lock may also be created with a
+{\em payload}, a small (i.e. less than 4KB - the limit for an active message payload) 
+piece of data that is moved around with the lock and guaranteed to be coherent while 
+the lock is held.
+
+Between deferred locks and barriers (described in Section~\ref{subsec:events}) clients 
+have access to the same set of synchronization primitves that they traditionally have
+in both threading and bulk-synchronous interfaces.  However, these operations can now 
+be composed with other asynchronous operations which is not possible in any other interface
+of which we are aware.
+
+%Deferred locks provide a super-set of the functionality of blocking locks.  As their
+%name suggests, deferred locks can use the event corresponding to their lock acquire
+%operation to defer execution until the lock acquire has been granted.  Deferred
+%locks can also be converted back into a blocking lock by immediately waiting on the event
+%returned from a call to {\tt lock}.  In addition, deffered locks also provide {\tt mode} and
+%and {\tt exclusive} parameters that allow the user to specify whether other requests can acquire
+%the lock simultaneously.  A lock can only be acquired in one mode at a time.  The
+%{\tt exclusive} parameter specifies whether other owners are permitted once the
+%current lock request is granted.
 
@@
 {\tt Processor} (line 20) handles in the system.  Using the affinity methods (lines 22-25),
-the application can also determine:
-\begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
-\item for each {\tt Processor}, which {\tt Memory} objects are accessible and with what performance, and
-\item which pairs of {\tt Memory} objects are able to efficiently copy data between them.
-\end{itemize}
+the application can also determine which {\tt Memory} objects are accessible by each
+{\tt Processor},and with what performance, and which pairs of {\tt Memory} objects are able to
+efficiently copy data between them.
 

### 2012-07-23 doc/asplos2013/conclusion.tex

@@
 features of our interface.  We showed that for three real-world applications
-the asynchronous Legion implementations were between 22\% to 2.35X faster than
+the asynchronous Legion implementations were between 22\% to 135\% faster than
 implementations which contained blocking operations.

### 2012-07-23 doc/asplos2013/results.tex
put all bulk synchronous figures into a single figure

@@
 
-Figures~\ref{fig:cktbulk}, \ref{fig:fluidbulk}, and \ref{fig:amrbulk} compare
+Figure~\ref{fig:bulksync} compare
 the performance of the bulk-synchronous implementations versus the original
@@
 
-\begin{figure}
-\begin{center}
+\begin{figure}[h!]
+\centering
+\subfigure[Circuit Application]{
+\label{fig:cktbulk}
 \includegraphics[scale=0.33]{figs/circuit_bulk_sync.pdf}
-\end{center}
-\vspace{-6mm}
-\caption{Circuit Bulk-Synchronous Performance.\label{fig:cktbulk}}
+}
+
+\subfigure[Fluid Application]{
+\label{fig:fluidbulk}
+\includegraphics[scale=0.33]{figs/fluid_bulk_sync.pdf}
+}
+
+\subfigure[AMR Application]{
+\label{fig:amrbulk}
+\includegraphics[scale=0.33]{figs/amr_bulk_sync.pdf}
+}
+\vspace{-4mm}
+\caption{Bulk-Synchronous Performance.\label{fig:bulksync}}
 \vspace{-4mm}
 \end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.33]{figs/fluid_bulk_sync.pdf}
-\end{center}
-\vspace{-6mm}
-\caption{Fluid Bulk-Synchronous Performance.\label{fig:fluidbulk}}
-\vspace{-4mm}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.33]{figs/amr_bulk_sync.pdf}
-\end{center}
-\vspace{-6mm}
-\caption{AMR Bulk-Synchronous Performance.\label{fig:amrbulk}}
-\vspace{-4mm}
-\end{figure}

### 2012-07-23 doc/asplos2013/impl.tex
space compaction, including the nuking of a duplicate paragraph

@@
 to detect the race condition in which the barrier owner receives an {\tt arrive} operation before the
-corresponding {\tt alter\_arrival\_count}.  When detected, the decrement of the arrival count 
+corresponding {\tt alter\_arrival\_count}.  When detected, the decrement of the count 
 is delayed to avoid a spurious triggering of the barrier.
@@
 sent to the most recently known owner.  If the receiving node is no
-longer the owner, it will forward the message on to the next node known
-to be the owner of the lock.
+longer the owner, it forwards the message on to the node it has recorded as the owner.
 %(If that information proves to be stale, the request is forwarded
@@
 of the lock's payload.  (The inclusion of the payload in the active message is the reason for the 
-4KB limit on a payload's size.)
+4KB size limit.)
 
@@
 The unfairness that results from a lock favoring local waiters over remote waiters is intentional.  When the
-contention on a lock is high (the only time when the question of fairness is relevant), the latencies
+contention on a lock is high (the only time the question of fairness is relevant), the latencies
 involved in transferring a lock between nodes can be the limiter on throughput.  By minimizing the number

### 2012-07-23 doc/asplos2013/interface.tex
space compaction, including the nuking of a duplicate paragraph

@@
   keepspaces=true,
+  belowskip=-10pt,
   %literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
@@
 
-Although {\tt UserEvent}s and {\tt Barrier}s are created and triggered differently,
-any operation that expresses a dependence on an {\tt Event} can transparently use
-a {\tt UserEvent} or {\tt Barrier} as well.
+Although UserEvents and Barriers are created and triggered differently,
+operations that express dependences on Events can transparently use
+UserEvents or Barriers as well.
 
@@
 
-\begin{lstlisting}[float={t},label={lst:procapi},caption={Processor Interface and Example.}]
+\begin{lstlisting}[float={t},label={lst:procapi},caption={Processor Interface and Example.},belowskip=0pt]
 class Processor {
@@
 }
+\vspace{-2mm}
 \caption{Processor Example Event Graph\label{fig:procevents}}
+\vspace{-4mm}
 \end{figure}
@@
 }
+\vspace{-2mm}
 \caption{Lock Example Event Graph\label{fig:lockevents}}
+\vspace{-4mm}
 \end{figure}
@@
 (line 8).
-
-Locks are often used to mediate access to data, and the size of that data is often small.  To
-support this common case we allow locks to be created with an associated
-%avoid adding the latency of inter-node data copies to critical sections, a lock may also be created with a
-{\em payload}, a small (i.e. less than 4KB - the limit for an active message payload) 
-piece of data that is moved around with the lock and guaranteed to be coherent while 
-the lock is held.
 
@@
 }
+\vspace{-6mm}
 \caption{Reduction Example Event Graph\label{fig:reducevents}}
+\vspace{-4mm}
 \end{figure}

### 2012-07-23 doc/asplos2013/impl.tex

@@
 event has triggered.  Other systems incorporating events address this by reference counting event
-handles\cite{Khronos:OpenCL}, but such reference counting adds both programmer and runtime overhead even when
+handles\cite{Khronos:OpenCL}, but such reference counting adds both client and runtime overhead even when
 limited to a single node; further overhead and complexity would be expected for a cluster-level
@@
 Compared to a reference counting scheme our implementation requires fewer total event data structures
-and has no programmer/runtime overhead.  The key observations is that one {\em generational event}
+and has no client/runtime overhead.  The key observations is that one {\em generational event}
 data structure can efficiently capture the state of one untriggered event and a very large 

### 2012-07-23 doc/asplos2013/intro.tex

@@
 If blocking calls are not accurately placed in the program they can
-lead to processor stalls that expose latency.  Correct placement of
-blocking calls can be dependent on input data, algorithmic decisions, 
-and the underlying hardware.  To hide as much
+lead to processor stalls that expose latency.  Placement of blocking
+calls is difficult to reason about because 
+they can be dependent on input data, algorithmic decisions, 
+and the underlying hardware.  Instead, to hide as much
 latency as possible and to minimize processor stalls, 
@@
 (computation, data copy, or synchronization) will complete.  Every
-call to perform an operation $op$ in our runtime returns immediately
-with an event that {\em triggers} when $op$ has completed.
-Furthermore, every call to perform an operation $op$ can take as a
-precondition an event that must trigger before $op$ can begin.  Using
+call to perform an operation $p$ in our runtime returns immediately
+with an event that {\em triggers} when $p$ has completed.
+Furthermore, every call to perform an operation $p$ can take as a
+precondition an event that must trigger before $p$ can begin.  Using
 events clients can compose arbitrary chains of dependent
@@
 The following sections describe our contributions:
-\begin{itemize}
+\begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
 \item We present an interface for asynchronously composing

### 2012-07-23 doc/asplos2013/results.tex

@@
 
-Figure~\ref{fig:bulksync} compare
+Figure~\ref{fig:bulksync} compares
 the performance of the bulk-synchronous implementations versus the original

### 2012-07-23 doc/asplos2013/impl.tex
speeling

--- 2012-07-23 doc/asplos2013/impl.tex
@@
 Events are created on demand and are {\em owned} by the node on which they
-were created.  The creation of an event occurs with no inter-node communcation.
+were created.  The creation of an event occurs with no inter-node communication.
 The space of event
-handles is divided across the nodes at startup time, allowing each node to assign handles
+handles is divided across the nodes at start-up time, allowing each node to assign handles
 to new events without the risk of conflicting with another node's assignments.  The static division

### 2012-07-23 doc/asplos2013/interface.tex
speeling

@@
 Between deferred locks and barriers (described in Section~\ref{subsec:events}) clients 
-have access to the same set of synchronization primitves that they traditionally have
+have access to the same set of synchronization primitives that they traditionally have
 in both threading and bulk-synchronous interfaces.  However, these operations can now 
@@
 be returned.  There is no automatic coherence of data between physical regions in the
-same class.  It is the client's responsiblity to manage data coherence using copies
+same class.  It is the client's responsibility to manage data coherence using copies
 between physical regions (line 22-23).  Just like other operations copies can be

### 2012-07-23 doc/asplos2013/intro.tex
speeling

@@
 \item To support synchronization in our event system we introduce {\em deferred locks}, 
-a novel synchronization primitive that operates in an asynchronous enviroment (Section~\ref{subsec:locks}).
+a novel synchronization primitive that operates in an asynchronous environment (Section~\ref{subsec:locks}).
 
@@
 interface for conjoining operations on data (e.g. reductions) with asynchronous data movement for 
-further lantecy hiding (Section~\ref{subsec:phyreg}).
+further latency hiding (Section~\ref{subsec:phyreg}).
 
@@
 \item We report on micro-benchmarks that are designed to stress
-our implementation.  The results indiciate that our implementation approaches the performance of
+our implementation.  The results indicate that our implementation approaches the performance of
 the underlying hardware (Section~\ref{sec:micro}).
@@
 programming system, and that real Legion applications programmed to our interface
-are capable of achieving high-performance (Section~\ref{sec:apps}).  Furthermore, we demostrate
-that the same applications written in a bulk-synchronous model incur performance penalites between 22-135\%.
+are capable of achieving high-performance (Section~\ref{sec:apps}).  Furthermore, we demonstrate
+that the same applications written in a bulk-synchronous model incur performance penalties between 22-135\%.
 \end{itemize}

### 2012-07-23 doc/asplos2013/micro.tex
speeling

@@
 request pairs, where each request pair is to a random lock, and is made dependent on the 
-previous lock/unlock pair's completion.  The chains are made long enough that startup
+previous lock/unlock pair's completion.  The chains are made long enough that start-up
 and cleanup times are negligible.  The total number of chains across all nodes gives the total
@@
 per second in the 16 node case.  List regions also scale well, but perform about an
-order of mangitude worse than folding regions in the dense case.  The use of a separate active message for
+order of magnitude worse than folding regions in the dense case.  The use of a separate active message for
 each reduction operation is another two orders of magnitude worse, demonstrating that data must be

### 2012-07-23 doc/asplos2013/related.tex
speeling

@@
 operations.  CUDA\cite{CUDA} and OpenCL\cite{Khronos:OpenCL} support
-the compostition of asynchronous kernel launches and asynchronous copies
+the composition of asynchronous kernel launches and asynchronous copies
 between a host node and a single GPU.  However, the only synchronization

### 2012-07-23 doc/asplos2013/results.tex
Removing a latex warning
speeling

@@
 asynchronous interface, the SOOP would have to periodically invoke
-blocking operations that would cause processors to stall and severly
+blocking operations that would cause processors to stall and severely
 limit Legion's ability to run ahead and keep the machine fully
@@
 are used to allow each partition to perform reductions in parallel and locally in GPU
-framebuffer memeory before merging all the reductions to a global instance
+framebuffer memory before merging all the reductions to a global instance
 residing in GASNet memory.  Without reduction instances this phase
@@
 models particles that flow through cells.  Each time step requires multiple phases
-that update different proprerties of the particles contingent upon neighboring
+that update different properties of the particles contingent upon neighboring
 particles in the same cell.  The space of cells is partitioned and neighboring
@@
 
-We instrumented the heterogenous implementation of our interface to 
+We instrumented the heterogeneous implementation of our interface to 
 capture event usage information.  Figure~\ref{fig:eventlife} shows
@@
 be able to recover the storage associated with the event.  In this example 
-a reference counting algorithm would reduce the storage needed for dynanmic events
+a reference counting algorithm would reduce the storage needed for dynamic events
 by over 10X, but could only do so at the cost of
@@
 application runs, while the peak number of generational events needed appears to 
-occur during the startup of the application.
+occur during the start-up of the application.
 
@@
 
-\begin{figure}[h!]
+\begin{figure}[!ht]
 \centering

### 2012-07-23 doc/asplos2013/impl.tex
Lots of very small changes

@@
 While a remote trigger of an event can result in
-the latency of a triggering operation including two active message flight times, it bounds the number of active
+the latency of a triggering operation being at least two active message flight times, it bounds the number of active
 messages required per event to $2N-2$ where $N$ is the number of nodes 
 %(which can be much smaller than the number of dependent operations) 
-interested in the event.  An alternative
+monitoring the event.  An alternative
 would have been to share the list of subscribers so that the triggering node could notify all interested
 nodes directly.  However, such an algorithm is both more complicated due to race conditions, and requires $O(N^2)$
-active messages.  As we will see in Section~\ref{subsec:eventmicro}, the latency of a single event trigger
-active message is very small, and any algorithm that is super-linear with the node count will not scale well.
+active messages.  Any algorithm that is super-linear with the node count will not scale well, 
+and as we will see in Section~\ref{subsec:eventmicro}, the latency of a single event trigger
+active message is very small. 
+%and any algorithm that is super-linear with the node count will not scale well.
 %either is or will soon be an scalability issue for large systems.
@@
 
-\subsection{Lock Implementation}
+\subsection{Deferred Lock Implementation}
 \label{subsec:lockimpl}
 
-Like events, locks are created on demand by the node on which the creation request was made.  The lock
+Like events, deferred locks, or locks from here on, are created on demand by the node on which the creation request was made.  The lock
 handles are also statically divided across the nodes, allowing the creation to be done without inter-node
 communication.  Unlike events whose ownership is static, locks may migrate; the {\em creating}
-node is the initial owner of a lock, but that ownership can be transferred to other nodes.
+node is the initial owner, but that ownership can be transferred to other nodes.
 
@@
 If the current owner's status shows
-the lock is currently held, the bit for the requesting node is set in remote waiters bitmask.  If the lock is
+the lock is currently held, the bit for the requesting node is set in the remote waiters bitmask.  If the lock is
 not held, the ownership of the lock is given to the requesting node via a {\em lock transfer} active
 message.  A lock transfer message includes the bitmask of remote waiters and an up-to-date copy 
-of the lock's payload.  (The inclusion of the payload in the active message is the reason for the 
-4KB size limit.)
+of the lock's payload.  The inclusion of the payload in the active message is the reason for the 
+4KB size limit specified on payloads in Section~\ref{subsec:locks}.
 

### 2012-07-23 doc/asplos2013/interface.tex
Lots of very small changes

@@
 operations and the implementation is responsible for triggering these events.  Clients
-can also create a {\tt UserEvent} (line 15) that are triggered explicitly by the client.
+can also create a {\tt UserEvent} (line 15) that is triggered explicitly by the client.
 
-The common synchronization pattern in which multiple consumers are dependent on a 
-(potentially different) set of producers is supported by a {\tt Barrier} object,
-which is similar to a {\tt UserEvent}, but does not trigger until the expected
-number of {\tt arrive} operations have occurred.  Barrier constructs exist in other
-programming models\cite{MPI}, but in addition to the decoupling of the waiters from
-the arrivers, our barrier interface has two further improvements to better support
-composability and asynchronous operation:
+The common synchronization pattern in which multiple consumers are dependent on
+a set of producers is supported by a {\tt Barrier} object, which is similar
+to a user event, but does not trigger until the expected number of {\tt arrive}
+operations have occurred.  Our interface differs from barriers in other programming
+models\cite{MPI} in three fundamental ways:
+%The common synchronization pattern in which multiple consumers are dependent on a 
+%(potentially different) set of producers is supported by a {\tt Barrier} object,
+%which is similar to a {\tt UserEvent}, but does not trigger until the expected
+%number of {\tt arrive} operations have occurred.  Barrier constructs exist in other
+%programming models\cite{MPI}, but in addition to the decoupling of the waiters from
+%the arrivers, our barrier interface has two further improvements to better support
+%composability and asynchronous operation:
 \begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
-\item An arrival operation can be made dependent on another event, allowing a launch 
-of a subtask or a copy of data to be completed before the barrier can trigger.
-\item The expected arrival count of a barrier can be dynamically increased, allowing
-a producer task to be split into multiple subtasks without the original task having to wait for
-their completion to signal its arrival.
+\item The set of producers of arrival calls can be different from the set of consumers waiting on the barrier.
+\item An arrival operation can be made dependent on another event, allowing a client to
+express that an operation should be completed before the barrier can trigger.
+\item The expected arrival count of a barrier can be dynamically modified, allowing
+barrier creation before needing to know the exact number of arrivals.
 \end{itemize}
 
-Although UserEvents and Barriers are created and triggered differently,
-operations that express dependences on Events can transparently use
-UserEvents or Barriers as well.
+Both user events and barriers are a sub-type of events and can therefore
+be used transparently to express dependences wherever events are used.
+%Although user events and barriers are created and triggered differently,
+%operations that express dependences on Events can transparently use
+%UserEvents or Barriers as well.
 
@@
 Listing~\ref{lst:procapi} shows the interface for {\tt Processor} objects which allow for
-the creation of parallel computations.  Processor provide a way of naming 
+the creation of parallel computations.  Processors provide a way of naming 
 every computational unit in the machine.  We describe how to discover processor handles
@@
 their data in conjunction with data movement for higher performance.  One common example is 
-that applications accumulate reduction operations in a buffer and as part of a copy operation 
+that applications accumulate reduction operations in a buffer and then, as part of a copy operation,
 the reduction operations are applied to a destination buffer on the receiving side of the copy.  
@@
 
-If the only operations that will be performed on locations in a {\tt RegionInstance} object,
+If reductions are the only operations that will be performed on locations in a region instance,
 a special reduction version can be created by providing the ID of a {\em reduction operator}
-when the object is created (line 13).  Reduction-only {\tt RegionInstance} objects can 
+when the object is created (line 13).  Reduction-only region instances can 
 increase the available parallelism and hide latency better, as we will show in 
@@
 be folded as $l \text{ += } r_1 + r_2$), and if the {\tt fold} operation and a {\em right identity}
-(e.g. $0$ is a right identity for subtraction because $x - 0 = x$ for all $x$) are provided,
+(e.g. $0$ is a right identity for addition because $x + 0 = x$ for all $x$) are provided,
 the runtime can use a {\em reduction fold instance}, described in section~\ref{subsec:reducimpl},
@@
 reduction-only instances can be copied to other reduction-only instances (or normal 
-instances) of the same class with the same asynchronous copy operations.
+instances) of the same class with the same asynchronous copy operations.  Copying between
+instances with different layouts is only possible because physical regions provide 
+the necessary information to understand the structure of the data.
 
@@
 
-An example of using reduction instances is shown on lines 35-49 of Listing~\ref{lst:regionapi}.
-In this example, two reduction instances, {\tt inst1} and {\tt inst2}, are created in different memories to
-be used with reduction operation 2 (lines 40-41).  A physical region of the same class {\tt inst3}
+An example using reduction-only regions is shown on lines 35-49 of Listing~\ref{lst:regionapi}.
+Two reduction-only regions, {\tt inst1} and {\tt inst2}, are created in different memories to
+be used with a reduction operation (lines 40-41).  A physical region of the same class {\tt inst3}
 is also created in a different memory (line 42).  Two tasks are launched that will apply
@@
 the application can also determine which {\tt Memory} objects are accessible by each
-{\tt Processor},and with what performance, and which pairs of {\tt Memory} objects are able to
+{\tt Processor} and with what performance, and which pairs of {\tt Memory} objects are able to
 efficiently copy data between them.

### 2012-07-23 doc/asplos2013/intro.tex
Lots of very small changes

@@
 smaller\cite{MPI,COARRAY_FORTRAN,UPC99}; these primitives expose
-latency on today's large distributed, heterogeneous parallel machines.
+latency on today's large, distributed parallel machines.
 

### 2012-07-23 doc/asplos2013/micro.tex
Lots of very small changes

@@
 going from one node to two can be attributed to the latency of a GASNet
-active message - others have measured similar latencies\cite{GASNET06}.
+active message; others have measured similar latencies\cite{GASNET06}.
 The gradual increase in latency with increasing node count is likely 
@@
 
-\subsection{Lock Grant Rates}
+\subsection{Deferred Lock Grant Rates}
 \label{subsec:lockmicro}
 
-Our lock micro-benchmark is designed to measure the rate at which locks can be granted.
+Our lock micro-benchmark is designed to measure the rate at which deferred locks can be granted
+in a distributed environment.
 A parameterized number of locks are created per node and their handles are made available to
@@
 request pairs, where each request pair is to a random lock, and is made dependent on the 
-previous lock/unlock pair's completion.  The chains are made long enough that start-up
+previous lock/unlock pair's completion by the event returned from the lock request.  
+The chains are made long enough that start-up
 and cleanup times are negligible.  The total number of chains across all nodes gives the total
@@
 started at the same time (via a dependence on a single user event), and the total time to
-process all the chains is divided into the total number of lock requests to yield and average
+process all the chains is divided into the total number of lock requests to yield an average
 lock grant rate.
@@
 decreasing lock counts.  Although contention increases, the favoring of local lock requestors makes
-this an advantage, reducing the number of lock-related active messages that must be sent.
+the contention an advantage, reducing the number of lock-related active messages that must be sent
+per lock grant.
 
@@
 more contention for locks and the rate drops as a result.  For smaller lock counts, further increases in
-the number of chains result in improved rates.  Note that the point on each line where this occurs
+the number of chains result in improved rates.  Note that the point on each line where the increase occurs
 is when the chains per node is greater than total locks, which corresponds to the point where the expected 
@@
 The dense experiment demonstrates that reduction fold regions perform 
-best and scale well with the number of nodes, achieving over a billion reductions
+best and scales well with the number of nodes, achieving over a billion reductions
 per second in the 16 node case.  List regions also scale well, but perform about an
@@
 transferred in larger blocks to be efficient on modern interconnect hardware.  The localized instance approach
-works well for a single node, but its inherently serial nature doesn't benefit at all from increasing node counts.
+works well for a single node, but its inherently serial nature doesn't benefit from increasing node counts.
 Finally, the only time the latency of performing individual RMA reads and writes isn't disastrous is
-on a single node (where the reads and writes are local).
+on a single node (where the reads and writes are all local).
 

### 2012-07-23 doc/asplos2013/related.tex
Lots of very small changes

@@
 
-The implementation of our interface shares many similarities with large
-distributed systems.  Many distributed system implement a publish/subscribe
+The implementation of our interface shares similarities with large
+distributed systems.  Many distributed systems implement a publish/subscribe
 abstraction for supporting communication that is similar in some ways to how
@@
 callbacks are registered to run when event operations are triggered
-from remote locations.  Events in these systems are much heavier weight
+remotely.  Events in these systems are much heavier weight
 and often carry large data payloads.  In many distributed event systems

### 2012-07-23 doc/asplos2013/results.tex
Lots of very small changes

@@
 In this section we demonstrate the performance properties of three
-real-world applications that use the Legion runtime and the heterogeneous
+real-world applications that use the Legion runtime and by extension the heterogeneous
 implementation of our interface to run on the Keeneland supercomputer.
@@
 of three phases that calculate currents on the wires, distribute charge between
-nodes, and update the voltages of all nodes.  In the distribute charge
+circuit nodes, and update the voltages of all nodes.  In the distribute charge
 phase, a single node may be updated by wires from multiple partitions.  Reduction instances 
-are used to allow each partition to perform reductions in parallel and locally in GPU
+are used to allow each partition to perform reductions in parallel and to do so locally in GPU
 framebuffer memory before merging all the reductions to a global instance
@@
 {\em Fluid} is an incompressible fluid flow simulation from the PARSEC
-benchmark suite\cite{bienia11benchmarking}.  The original implementation can 
+benchmark suite\cite{bienia11benchmarking}.  The reference implementation can 
 only be run on a single shared memory machine, but when written in Legion, Fluid can
@@
 measures the number of events that are live at any given time.  After
-the last operation on an event, a reference counting implementation would
-be able to recover the storage associated with the event.  In this example 
+the last operation on a dynamic event, a reference counting implementation would
+be able to recover the storage associated with the event.  The live events line
+is therefore equivalent to the number of needed events in a refernce counting
+scheme.  In this example 
 a reference counting algorithm would reduce the storage needed for dynamic events
@@
 that is 10X smaller than even the maximal live event count.  The actual storage
-requirements of our implementation are shown by the {\em physical events} line,
-which is slightly larger than the peak number of untriggered events. 
+requirements of our implementation are shown by the generational events line,
+which shows the total number of generational events in the system (see Section~\ref{subsec:eventimpl}).
+The maximum number of generational events needed is slightly larger than the peak number of untriggered events. 
 The reason for this difference is because nodes create events locally if they have
@@
 
-\subsection{Lock Migration}
+\subsection{Deferred Lock Migration}
 \label{subsec:lockmig}
@@
 We also instrumented our heterogeneous implementation to profile the usage of 
-locks in real applications.  We measured how many locks were used in the application.  For
+deferred locks in real applications.  We measured how many locks were used in the application.  For
 each lock we measured the number of lock requests and the number of times that the lock
@@
 which they were created.  The AMR application exhibits more varied locking behavior, and in 
-particular has locks that are requested more often than they are transferred.  Without the
-ability for locks to migrate, these applications couldn't have made use of asynchronous
-locks.
+particular has locks that are requested more often than they are transferred.  In addition
+to allowing synchronization in an asynchronous environment, the ability of deferred locks to
+migrate is necessary for implementing real applications on distributed machines.
+
+%Without the
+%ability for locks to migrate, these applications couldn't have made use of asynchronous
+%locks.
 

### 2012-07-23 doc/asplos2013/interface.tex
changes through 2.3

@@
 compilers and runtimes such as Legion\cite{Legion12} as well as 
-advanced systems programmers.  This class of users demand total control
+advanced systems programmers.  This class of clients expect total control
 over the underlying hardware and transparent performance from an
@@
 are light-weight {\em handles} that provide a unique name for an underlying
-implementation.  Multiple copies of handles are permitted and handles can be 
-passed by value.  Furthermore, every handle is valid everywhere in the system.  
-This property allows a client to pass handles by value between computations 
+implementation.
+Every handle is valid everywhere in the system, allowing handles to be freely copied,
+passed as arguments to other tasks, or stored in the heap
 without having to reason about the distributed nature of the system.
+%%   Multiple copies of handles are permitted and handles can be 
+%% passed by value.  Furthermore, every handle is valid everywhere in the system.  
+%% This property allows a client to pass handles by value between computations 
+%% without having to reason about the distributed nature of the system.
 %We describe how we support this property in more detail in 
@@
   belowskip=-10pt,
-  %literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
+  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
   %morekeywords={region,coloring,partition,spawn,disjoint,aliased},
@@
 use of events is passing them as preconditions for other operations.  The client can use the
-{\tt merge\_events} call (lines 11-12) to create an event corresponding to the 
-conjunction of a set of events.
+{\tt merge\_events} call (lines 11-12) to create an event that triggers only when every
+event in a given set of events has triggered.
+%corresponding to the conjunction of a set of events.
 
@@
 to a user event, but does not trigger until the expected number of {\tt arrive}
-operations have occurred.  Our interface differs from barriers in other programming
+operations have occurred.  To better support
+composability and asynchronous operation, our interface differs from barriers in other programming
 models\cite{MPI} in three fundamental ways:
@@
 
-Events and barriers express ordering properties between operations, but in many
-cases ordering is too strict of a requirement.  For many applications access to data must only be atomic and
+Events and barriers express ordering properties between operations, but 
+%in many cases 
+ordering is often too strict a requirement.  For many applications access to data must only be atomic and
 not necessarily ordered.  Traditionally locks have been used to serialize access to
@@
 a precondition for launching the {\tt SUB} task.  The unlock operation is then
-precondition on the event corresponding to {\tt SUB} task's completion.  The
+made dependent
+%precondition
+on the event corresponding to {\tt SUB} task's completion.  The
 {\tt SUB} task can run safely while holding the lock and the lock is released when the
@@
 
-  int get_proc_mem_affinity(vector<ProcMemAffinity &result,
+  int get_proc_mem_affinity(vector<ProcMemAffinity> &result,
           Processor restrict_proc = 0, Memory restrict_mem = 0);

### 2012-07-23 doc/asplos2013/intro.tex
changes through 2.3

@@
 asynchronously compose computation, data movement, and
-synchronization.  The basic mechanism that enables composition of operations
-is an underlying {\em event} primitive.  Events provide a mechanism
+synchronization.  The basic mechanism that enables composition of 
+asynchronous operations
+is an {\em event} primitive.  Events provide a mechanism
 for naming a point in the future when an asynchronous operation
@@
 Furthermore, every call to perform an operation $p$ can take as a
-precondition an event that must trigger before $p$ can begin.  Using
+precondition an event that must trigger before $p$ can begin.
+Any number of operations can be depedendent on a single event, and
+a events can be merged to allow an operation to wait for multiple
+events before it begins.
+  Using
 events clients can compose arbitrary chains of dependent
@@
 \item We present an interface for asynchronously composing
-parallel computation, data movement, and synchronization using events (Section~\ref{sec:interface}).
+parallel computation, data movement, and synchronization using events (Section~\ref{subsec:events}).
 

### 2012-07-23 doc/asplos2013/interface.tex
2.4 and 2.5

@@
 \begin{lstlisting}[float={t},label={lst:regionapi},caption={Physical Region Interface and Example.}]
+class Memory {
+public:
+  const unsigned id;
+  size_t size() const;
+};
+
 class RegionMetaData {
@@
 
-Listing~\ref{lst:regionapi} shows a subset of the interface for physical regions.  The
-first object in the interface is a {\tt RegionMetaData} object (line 1).  Physical
-regions created by the same region meta-data object belong to the same class.  Physical
-regions are represented by {\tt RegionInstance} objects (line 17).  When creating a physical
-region the client must specify the {\tt Memory} in which the physical region is going
-to be allocated.  Memories are described in more detail in Section~\ref{subsec:machmodel}.
-If there is not enough space for the physical region allocation a {\tt NO\_INST} will
-be returned.  There is no automatic coherence of data between physical regions in the
+Listing~\ref{lst:regionapi} shows a subset of the interface for physical regions.  
+Each memory in the machine's memory hierarchy is named by a {\tt Memory} object (line 1).
+The next object is a {\tt RegionMetaData} object (line 7), which defines a class of
+regions.  Physical regions that belong to that class are created with the {\tt create\_instance}
+method (line 16) and represented by {\tt RegionInstance} objects (line 23).  When creating a physical
+region the client specifies the memory in which the physical region is going
+to be allocated.  To maintain performance transparency, a new physical region can only be allocated
+if the memory has sufficient capacity.
+There is no automatic coherence of data between physical regions in the
 same class.  It is the client's responsibility to manage data coherence using copies
-between physical regions (line 22-23).  Just like other operations copies can be
+between physical regions (line 28-29).  Like other operations copies can be
 predicated on an event and return an event corresponding to completion.
@@
 a special reduction version can be created by providing the ID of a {\em reduction operator}
-when the object is created (line 13).  Reduction-only region instances can 
+when the object is created (line 19).  Reduction-only region instances can 
 increase the available parallelism and hide latency better, as we will show in 
@@
 is commutative (i.e. multiple operations to the same location can be reordered without 
-changing the final result) and is coded as an {\tt apply} method as shown on line 29.
+changing the final result) and is coded as an {\tt apply} method as shown on line 35.
 Many commonly-used reduction operations are also {\em foldable} (e.g. $(l \text{ += } r_1) \text{ += } r_2$ can
@@
 
-An example using reduction-only regions is shown on lines 35-49 of Listing~\ref{lst:regionapi}.
+An example using reduction-only regions is shown on lines 41-55 of Listing~\ref{lst:regionapi}.
 Two reduction-only regions, {\tt inst1} and {\tt inst2}, are created in different memories to
-be used with a reduction operation (lines 40-41).  A physical region of the same class {\tt inst3}
-is also created in a different memory (line 42).  Two tasks are launched that will apply
-reductions locally into their respective reduction buffers (lines 43-44).  The resulting
-events are then used to chain copies back to {\tt inst3} after the tasks complete (45-46).  As
+be used with a reduction operation (lines 46-47).  A physical region of the same class {\tt inst3}
+is also created in a different memory (line 48).  Two tasks are launched that will apply
+reductions locally into their respective reduction buffers (lines 49-50).  The resulting
+events are then used to chain copies back to {\tt inst3} after the tasks complete (lines 51-52).  As
 part of these copies the reductions stored in {\tt inst1} and {\tt inst2} will be applied
@@
 applications.  Finally, a merge on the events from the movement and reduction applications is 
-performed and a final task using {\tt inst3} is launched (lines 47-48).  The resulting event
+performed and a final task using {\tt inst3} is launched (lines 53-54).  The resulting event
 graph is shown in Figure~\ref{fig:reducevents}.
@@
 program, providing a task table that maps task IDs to function pointers,
-and then invoking the {\tt run} method (line 15-16).
+and then invoking the {\tt run} method (lines 9-10).
 Any running task can get a pointer to the {\tt Machine} object with the
-{\tt get\_machine} method (line 18) and use it to obtain lists of all the {\tt Memory} (line 19) or 
-{\tt Processor} (line 20) handles in the system.  Using the affinity methods (lines 22-25),
+{\tt get\_machine} method (line 12) and use it to obtain lists of all the {\tt Memory} (line 13) or 
+{\tt Processor} (line 14) handles in the system.  Using the affinity methods (lines 16-19),
 the application can also determine which {\tt Memory} objects are accessible by each
@@
 \begin{lstlisting}[float={t},label={lst:machineapi},caption={Machine Interface.}]
-class Memory {
-public:
-  const unsigned id;
-  size_t size() const;
-};
-
 class Machine {

### 2012-07-23 doc/asplos2013/impl.tex
minor edits to section 3

@@
 one that works only on shared memory machines and another that
-runs on larges clusters with both CPUs and GPUs.  The shared-memory
+runs on large clusters with both CPUs and GPUs.  The shared-memory
 implementation uses the POSIX threads library\cite{PTHREADS} (Pthreads)
 and is primarily used for debugging.
-The implementation for heterogeneous clusters also uses Pthreads as 
+The implementation for heterogeneous clusters uses Pthreads as 
 well as the CUDA runtime library for GPUs\cite{CUDA} and the GASNet
@@
 The heterogeneous implementation
-models a cluster with CPUs and GPUs as having two types of processors 
-and four types of memory.  Every CPU core is presented as a different {\tt CPU} processor
+models a cluster with CPUs and GPUs as having two kinds of processors 
+and four kinds of memory.  Every CPU core is presented as a different {\tt CPU} processor
 and every GPU is a {\tt GPU} processor.  This matches the scheduling granularity available
-in the Pthreads and CUDA APIs respectively.  The first type of memory is the 
+in the Pthreads and CUDA APIs respectively.  The first kind of memory is the 
 system memory that is accessible by every CPU core on a given node.  The second is the 
-framebuffer memory on each GPU (and accessible only by that GPU).  The third type
+framebuffer memory on each GPU (and accessible only by that GPU).  The third kind
 of memory is system memory on a node that has been made accessible to the GPU(s) on
 that node as well as the CPU cores, commonly referred to as {\em zero-copy memory}.
-The final type of memory is the portion of system memory on each node that has been registered with
+The final kind of memory is the portion of system memory on each node that has been registered with
 the GASNet runtime to allow {\em remote memory access} (RMA) by other nodes in the cluster.
@@
 are sent by one node to another without any previous coordination.  Upon arrival
-at a destination node, a handler routine is invoked to process an active message.
+at a destination node, a handler routine is invoked to process the message.
 %The handler may issue a response to the sender, but this is optional, and avoided
@@
 %events, locks, and reduction instances.
-Due to space constraints we cover only a small part of the implementation in detail.
-We limit our discussion to three interesting aspects:
+Due to space constraints we cover only the most interesting parts of the implementation in detail.
+We limit our discussion to three aspects:
 events, locks, and reduction instances.
@@
 When a new event is created, the owning node allocates a data structure to track the state of
-the event ({\em triggered} or {\em untriggered}) as well as to record a list of operations dependent on the
-event from the same node called {\em waiters} (e.g. copy operation or task launch).  The first operation contingent on an event
-from a remote node will create the same data structure on the remote node.  A single
+the event ({\em triggered} or {\em untriggered}) as well as to record a list of dependent operations
+(e.g. copy operation or task launch) from the same node (called {\em waiters}).
+The first operation dependent on an event
+from a remote node will allocate the same data structure on the remote node.  A single
 {\em event subscription} active message is then sent from the remote node to the owning node indicating
 that the remote node should be informed when the event triggers.  Arbitrarily many additional operations on
-the remote node can then be aggregated as waiters onto the remote data structure without any additional communication.
+the remote node can then be added to the list of waiters without any additional communication.
 When the event does trigger, the owner node notifies all local waiters and
@@
 
-The data structures used to track an event cannot be freed until all operations on that event have been
+The data structure used to track an event cannot be freed until all operations on that event have been
 performed.  Creation and triggering can each happen only once, but there can be an arbitrary number of operations
@@
 Compared to a reference counting scheme our implementation requires fewer total event data structures
-and has no client/runtime overhead.  The key observations is that one {\em generational event}
+and has no client/runtime overhead.  The key observation is that one {\em generational event}
 data structure can efficiently capture the state of one untriggered event and a very large 
@@
 handles are also statically divided across the nodes, allowing the creation to be done without inter-node
-communication.  Unlike events whose ownership is static, locks may migrate; the {\em creating}
+communication.  Whereas events ownership is static, locks may migrate; the {\em creating}
 node is the initial owner, but that ownership can be transferred to other nodes.
@@
 If the current node isn't the owner, a {\em lock request} active message is
-sent to the most recently known owner.  If the receiving node is no
-longer the owner, it forwards the message on to the node it has recorded as the owner.
+sent to the most recently known owner.  (If the receiver of a lock request message is no
+longer the owner, it forwards the message on to the node it has recorded as the owner.)
 %(If that information proves to be stale, the request is forwarded
@@
 the supplied right-hand-side value is folded in
-to the corresponding location in the array.  An arbitrary number of reductions can be done to each location.
+to the corresponding location in the array.  An arbitrary number of reductions can be folded into each location.
 When the reduction fold instance is copied back to a normal instance, the contents of the reduction instance

### 2012-07-23 doc/asplos2013/micro.tex
section 4 edits

@@
 last event in the chain is measured.  The total time is divided by the number
-of events in the chain to yield the mean trigger times.  In the single-node case, all events are local to
+of events in the chain to yield the mean trigger time.  In the single-node case, all events are local to
 that node, so no active messages are required.  For all other cases, the
@@
 be triggered by our implementation.  Instead of a single chain, many parallel chains
-are created.  Additionally, each step of the chain can consist of parameterized number
-of events, each one dependent on all the events from the previous step of the chain.
+are created.  Additionally, each step of the chain consists of a parameterized number
+of events, each one dependent on every event from the previous step of the chain.
 This parameter is called the {\em fan-in/out factor}.  The events within each step of
@@
 node count.  Although the number of events waiting on each node decreases with an increasing
-node code, the minimal scaling indicates that the main bottleneck is in the processing of the
+node count, the minimal scaling indicates that the main bottleneck is in the processing of the
 active message that each node must receive rather than the local redistribution of the triggering
@@
 \label{subsec:reducmicro}
-To show the benefits of associating reduction instances we designed
+To show the benefits of reduction instances we designed
 a histogram micro-benchmark that requires all nodes to perform an addition reduction to a
@@
 We ran two experiments, one corresponding with dense reductions and the other corresponding with sparse
-reductions.  In each case a large random source of data is divided across into chunks, which are given to
+reductions.  In each case a large random source of data is divided into chunks, which are given to
 separate reduction tasks.  There are 8 reduction tasks created for each node, one per processor.  The dense
 case uses a histogram with 256K buckets and each reduction task performs 4M reductions.  The results for the
-dense experiment are shown in Figure~\ref{fig:reducdense}.  The sparse case use a histogram with 4M buckets,
+dense experiment are shown in Figure~\ref{fig:reducdense}.  The sparse case uses a histogram with 4M buckets,
 but only 64K reductions are performed by each task.  Results for the sparse case
@@
 The dense experiment demonstrates that reduction fold regions perform 
-best and scales well with the number of nodes, achieving over a billion reductions
+best and scale well with the number of nodes, achieving over a billion reductions
 per second in the 16 node case.  List regions also scale well, but perform about an

### 2012-07-23 doc/asplos2013/results.tex
section 5 edits

@@
 In this section we demonstrate the performance properties of three
-real-world applications that use the Legion runtime and by extension the heterogeneous
+real-world applications that use the Legion runtime and the heterogeneous
 implementation of our interface to run on the Keeneland supercomputer.
@@
 the Fluid application.  Section~\ref{subsec:lockmig} examines
-the migration of locks in the Circuit application.  Finally, we demonstrate
+the migration of locks in the Circuit and AMR applications.  Finally, we demonstrate
 that the asynchronous nature of our interface is essential to the
 performance of Legion by comparing to bulk-synchronous
-implementations of our target applications in Section~\ref{subsec:bulkcomp}.
+implementations of these applications in Section~\ref{subsec:bulkcomp}.
 
@@
 is essential to the high performance of the Legion runtime system.  The Legion
-implementations of our three applications have already been shown to outperform 
+implementations of the three applications have already been shown to outperform 
 pre-existing implementations\cite{Legion12}.  To quantify the benefit of the asynchronous
-nature of our interface, we implemented versions of our applications in
-Legion in a bulk-synchronous style to compare against our existing implementations.  
+nature of our interface, we re-implemented the applications in
+a bulk-synchronous style to compare against the existing Legion implementations.  
 These bulk-synchronous versions still run on top of 
@@
 the performance of the bulk-synchronous implementations versus the original
-implementations for Circuit, Fluid, and AMR respectively.  Each plot contains
+implementations for Circuit, Fluid, and AMR.  Each plot contains
 performance curves for both implementations on two different problem sizes.

### 2012-07-23 doc/asplos2013/conclusion.tex
Lots of very small changes
section 6 and 7 edits

@@
 to overlap long latency operations with useful work.  We have
-presented a low-level programming interface that is fully asynchronous.
+presented a low-level programming interface that is fully asynchronous
+including a new synchronization primitive called a deferred lock.
 Our interface uses events to compose asynchronous operations.  Clients
 can use events to explicitly specify dependences between operations,
-leaving the runtime to schedule them in a way that maximizes latency 
+giving the runtime the flexibility to schedule them in a way that maximizes latency 
 hiding while still obeying dependences.
 
-We've presented an implementation of interface that is capable of running
+We've presented an implementation of the interface that is capable of running
 on a large heterogeneous cluster containing both CPUs and GPUs.  We

### 2012-07-23 doc/asplos2013/related.tex
section 6 and 7 edits

@@
 make it easy to port dataflow programs onto a parallel machine\cite{CullerGSvE93}.  
-Conceptually our interface is very similar to the dataflow languages TAM supported.
-The implementation of TAM is designed for a class of much smaller machines
+Conceptually our interface is similar to the dataflow languages TAM supported.
+However, the implementation of TAM is designed for a class of much smaller machines
 and therefore is based on much heavier communication and threading mechanisms.

### 2012-07-23 doc/asplos2013/interface.tex
tiny edits

@@
 Our interface organizes functionality into objects.  We describe
-these objects in five parts: events for composing operations 
-(Section~\ref{subsec:events}), processors for parallel computation 
-(\ref{subsec:procs}), deferred locks for synchronization
-(\ref{subsec:locks}), physical regions for data layout and movement
-(\ref{subsec:phyreg}), and a machine object for introspection of the underlying
+these objects in five parts: {\em events} for composing operations 
+(Section~\ref{subsec:events}), {\em processors} for parallel computation 
+(\ref{subsec:procs}), {\em deferred locks} for synchronization
+(\ref{subsec:locks}), {\em physical regions} for data layout and movement
+(\ref{subsec:phyreg}), and a {\em machine} object for introspection of the underlying
 hardware (\ref{subsec:machmodel}).  In all cases (except for the
@@
 \label{subsec:events}
-Events are the primary mechanism for describing dependences in our system.
+Events are the primary mechanism for describing dependences between operations in our system.
 Listing~\ref{lst:eventapi} shows the interface for events.  An instance of the {\tt Event} type 
@@
 {\tt merge\_events} call (lines 11-12) to create an event that triggers only when every
-event in a given set of events has triggered.
+event in a given set of events has triggered.  An example using events follows in
+Section~\ref{subsec:procs}.
 %corresponding to the conjunction of a set of events.
@@
 to include new processor types (line 10).   Processors support a single {\tt spawn}
-operation (lines 14-15) that will launch a new {\em task} on that processor.
+method (lines 14-15) that launches a new {\em task} on that processor.
 Note that because the spawn operation
@@
 The {\tt merge\_events} call
-is used to merge the events {\tt eb} and {\tt ec} to create the shared dependence
+merges the events {\tt eb} and {\tt ec} to create the shared dependence
 for the last sub-task.
@@
 %in many cases 
-ordering is often too strict a requirement.  For many applications access to data must only be atomic and
+ordering is often too strict a requirement.  For many applications access to data need only be atomic and
 not necessarily ordered.  Traditionally locks have been used to serialize access to
@@
 or spinning, neither of which composes well with asynchronous operations.
-Deferred locks are a new synchronization mechanism that allows for synchronization
+{\em Deferred locks} are a new synchronization mechanism that allows for synchronization
 in an completely asynchronous environment.
@@
 Between deferred locks and barriers (described in Section~\ref{subsec:events}) clients 
-have access to the same set of synchronization primitives that they traditionally have
-in both threading and bulk-synchronous interfaces.  However, these operations can now 
+have access to the same set of synchronization primitives that they have
+in threading and bulk-synchronous interfaces.  However, these operations can now 
 be composed with other asynchronous operations which is not possible in any other interface
@@
 the runtime can use a {\em reduction fold instance}, described in section~\ref{subsec:reducimpl},
-for ever better performance.  Although the way they store data is very different,
+for even better performance.  Although the way they store data is very different,
 reduction-only instances can be copied to other reduction-only instances (or normal 
@@
 and then invoking the {\tt run} method (lines 9-10).
-Any running task can get a pointer to the {\tt Machine} object with the
+Any running task can get a pointer to the machine object with the
 {\tt get\_machine} method (line 12) and use it to obtain lists of all the {\tt Memory} (line 13) or 
 {\tt Processor} (line 14) handles in the system.  Using the affinity methods (lines 16-19),
-the application can also determine which {\tt Memory} objects are accessible by each
-{\tt Processor} and with what performance, and which pairs of {\tt Memory} objects are able to
-efficiently copy data between them.
+the application can also determine which memories are accessible by each
+processor and with what performance, as well as which pairs of memories support copy operations.
+%are able to efficiently copy data between themselves.
 

### 2012-07-23 doc/asplos2013/micro.tex
tiny edits

@@
 related to the underlying point-to-point nature of Infiniband communication,
-which requires GASNet to poll a separate connection for every other node in
-the system.
+which requires GASNet to poll a separate connection for every other node.
 %% In the single node case the latency of triggering an event is only 
@@
 previous lock/unlock pair's completion by the event returned from the lock request.  
-The chains are made long enough that start-up
-and cleanup times are negligible.  The total number of chains across all nodes gives the total
+%The chains are made long enough that start-up
+%and cleanup times are negligible.  
+The total number of chains across all nodes gives the total
 number of lock requests that can exist in the system at any given time.  All chains are
@@
 \item Single Reads/Writes - Each node takes a lock on the global physical region and then for each reduction
-reads a value from the physical region, performs the addition, and then writes the result back.  The lock is then released.
+reads a value from the physical region, performs the addition, and then writes the result back.  
+The lock is released after all reductions are performed.
 \item Single Reductions - Each reduction is individually sent as an active message to the node whose
@@
 \item Fold Instance - Each node creates a local fold reduction instance.
-All reductions are folded into the instance, which is then copied back to the global region.
+All reductions are folded into the instance, which is then copied and applied to the global region.
 \item List Instance - Each node creates a local list reduction instance.
-All reductions are appended to the list reduction instance, which is then copied back to the global region.
+All reductions are appended to the list reduction instance, which is then copied and applied to the global region.
 \end{itemize}

### 2012-07-23 doc/asplos2013/results.tex
moved legion citation
tiny edits

@@
 \label{sec:apps}
-Legion is a high-level runtime system that has been implemented on top
-of our interface\cite{Legion12}.  
+Legion\cite{Legion12} is a high-level runtime system that has been implemented on top
+of our interface.  
 %The Legion runtime is distributed so 
@@
 of three phases that calculate currents on the wires, distribute charge between
-circuit nodes, and update the voltages of all nodes.  In the distribute charge
-phase, a single node may be updated by wires from multiple partitions.  Reduction instances 
+components, and update the voltages of all components.  In the distribute charge
+phase, a single component may be updated by wires from multiple partitions.  Reduction instances 
 are used to allow each partition to perform reductions in parallel and to do so locally in GPU
@@
 requirements of our implementation are shown by the generational events line,
-which shows the total number of generational events in the system (see Section~\ref{subsec:eventimpl}).
+which shows the total number of generational events (defined in Section~\ref{subsec:eventimpl}) 
+in the system.
 The maximum number of generational events needed is slightly larger than the peak number of untriggered events. 
@@
 particular has locks that are requested more often than they are transferred.  In addition
-to allowing synchronization in an asynchronous environment, the ability of deferred locks to
-migrate is necessary for implementing real applications on distributed machines.
+to allowing synchronization in an asynchronous environment, deferred locks must be able to 
+migrate to implement real applications on distributed machines.
 

### 2012-07-23 doc/asplos2013/impl.tex
speeling

@@
 the event ({\em triggered} or {\em untriggered}) as well as to record a list of dependent operations
-(e.g. copy operation or task launch) from the same node (called {\em waiters}).
+(e.g. copy operation or task launch) from the same node called {\em waiters}.
 The first operation dependent on an event
@@
 {\em event subscription} active message is then sent from the remote node to the owning node indicating
-that the remote node should be informed when the event triggers.  Arbitrarily many additional operations on
+that the remote node should be informed when the event triggers.  Arbitrarily many dependent operations on
 the remote node can then be added to the list of waiters without any additional communication.
@@
 handles are also statically divided across the nodes, allowing the creation to be done without inter-node
-communication.  Whereas events ownership is static, locks may migrate; the {\em creating}
+communication.  Whereas event ownership is static, locks may migrate; the {\em creating}
 node is the initial owner, but that ownership can be transferred to other nodes.
@@
 If the current node isn't the owner, a {\em lock request} active message is
-sent to the most recently known owner.  (If the receiver of a lock request message is no
-longer the owner, it forwards the message on to the node it has recorded as the owner.)
+sent to the most recently known owner.  If the receiver of a lock request message is no
+longer the owner, it forwards the message on to the node it has recorded as the owner.
 %(If that information proves to be stale, the request is forwarded
@@
 These layouts are designed to optimize for the common pattern where 
-multiple tasks make commutative {\em reductions} into shared data.  Instead of having to copy the existing 
-physical region to a suitable memory and then applying individual reduction operations, reduction instances 
+multiple tasks make commutative {\em reductions} into shared data.  Instead of having to copy an existing 
+physical region to a suitable memory and then apply individual reduction operations, reduction instances 
 allow a large number of reduction operations to be accumulated and transferred efficiently to the original 

### 2012-07-23 doc/asplos2013/impl.tex
Lots of little changes

--- 2012-07-23 doc/asplos2013/impl.tex
@@
 {\em event subscription} active message is then sent from the remote node to the owning node indicating
-that the remote node should be informed when the event triggers.  Arbitrarily many additional operations on
+that the remote node should be informed when the event triggers.  Arbitrarily many depedent operations on
 the remote node can then be added to the list of waiters without any additional communication.

### 2012-07-23 doc/asplos2013/related.tex
Lots of little changes

@@
 deferred lock.  OpenCL events are not valid globally, but can only be
-used with a single GPU context.
+used within a single GPU context.
 
@@
 and therefore is based on much heavier communication and threading mechanisms.
-While TAM supported composable asynchronous computation and communication, its
-synchronization mechanisms still required blocking operations.
+While TAM supports composable asynchronous computation and communication, its
+synchronization mechanisms still require blocking operations.
 
@@
 we manage events and event waiters\cite{Aguilera99,Carzaniga01}.  Work has
-also been done on using object oriented languages to build 
+also been done on using object oriented languages to build event-based
 distributed systems\cite{Eugster01,Harrison97,Chang91}.  In these cases

### 2012-07-23 doc/asplos2013/results.tex
speeling
Lots of little changes

@@
 which shows the total number of generational events (defined in Section~\ref{subsec:eventimpl}) 
-in the system.
+in the system
 The maximum number of generational events needed is slightly larger than the peak number of untriggered events. 

### 2012-07-23 tools/event_analyzer.py
dashed lines for the straight ones

@@ -217,9 +217,9 @@ def make_event_lifetimes_plot(outdir,
                               liveness_time, liveness_list):
     fig = plt.figure(figsize=(10,7))
     lw = 1.0
-    plt.plot(dynamic_time,dynamic_event_list,'--',color=tableau12,linestyle='solid',label='Dynamic Events',linewidth=lw)
+    plt.plot(dynamic_time,dynamic_event_list,'--',color=tableau12,linestyle='dashed',label='Dynamic Events',linewidth=lw)
     plt.plot(liveness_time,liveness_list,'--',color=tableau13,linestyle='solid',label='Live Events',linewidth=lw)
-    plt.plot(dynamic_time,physical_event_list,'--',color=tableau9,linestyle='solid',label='Generational Events',linewidth=lw)
+    plt.plot(dynamic_time,physical_event_list,'--',color=tableau9,linestyle='dashed',label='Generational Events',linewidth=lw)
     plt.plot(active_time, active_list, '--', color=tableau1, linestyle='solid', label='Untriggered Events', linewidth=lw)
     plt.legend(loc=2,ncol=1)
     plt.xlabel('Time (seconds)')

### 2012-08-04 doc/sc2012/legion.tex
Adding some stuff for the final version of the SC paper

@@
 \documentclass[conference]{IEEEtran}
-
+\IEEEoverridecommandlockouts 
 \bibliographystyle{IEEEtran}
@@
 \usepackage{url}
+\usepackage{minibox}
 
@@
     policies, either expressed or implied, of the U.S. Government.}}
-\author{ \IEEEauthorblockN{Michael Bauer\IEEEauthorrefmark{1},
-%\authorinfo(Stanford University, email: \mbox{mebauer@cs.stanford.edu}
-Sean Treichler\IEEEauthorrefmark{1}, 
-%\authorinfo(Stanford University, email: \mbox{sjt@cs.stanford.edu}
-Elliott Slaughter\IEEEauthorrefmark{1}, 
-% \authorinfo(Stanford University, email: \mbox{slaughter@cs.stanford.edu}
-Alex Aiken\IEEEauthorrefmark{1}  
-%\authorinfo(Stanford University, email: \mbox{aiken@cs.stanford.edu}
-%\email{aiken@cs.stanford.edu}
+%\author{ \IEEEauthorblockN{Michael Bauer\IEEEauthorrefmark{1},
+%%\authorinfo(Stanford University, email: \mbox{mebauer@cs.stanford.edu}
+%Sean Treichler\IEEEauthorrefmark{1}, 
+%%\authorinfo(Stanford University, email: \mbox{sjt@cs.stanford.edu}
+%Elliott Slaughter\IEEEauthorrefmark{1}, 
+%% \authorinfo(Stanford University, email: \mbox{slaughter@cs.stanford.edu}
+%Alex Aiken\IEEEauthorrefmark{1}  
+%%\authorinfo(Stanford University, email: \mbox{aiken@cs.stanford.edu}
+%%\email{aiken@cs.stanford.edu}
+%}
+%\IEEEauthorblockA{
+%\IEEEauthorrefmark{1}{Computer Science Department, Stanford University}
+%}
+%}
+
+\author{
+\IEEEauthorblockN{Michael Bauer}
+\IEEEauthorblockA{Stanford University \\ mebauer@cs.stanford.edu}
+\and
+\IEEEauthorblockN{Sean Treichler}
+\IEEEauthorblockA{StanfordUniversity \\ sjt@cs.stanford.edu}
+\and
+\IEEEauthorblockN{Elliott Slaughter}
+\IEEEauthorblockA{Stanford University \\ slaughter@cs.stanford.edu}
+\and
+\IEEEauthorblockN{Alex Aiken}
+\IEEEauthorblockA{Stanford University \\ aiken@cs.stanford.edu}
 }
-\IEEEauthorblockA{
-\IEEEauthorrefmark{1}{Computer Science Department, Stanford University}
-}
+
+\IEEEpubid{\makebox[\columnwidth][l]{\minibox{SC12, November 10-16, 2012, Salt Lake City, Utah, USA \\ 978-1-4673-0806-9/12/ \$31.00~\copyright~2012 IEEE}}
+\hspace{\columnsep}
+\makebox[\columnwidth]{}
 }

### 2012-08-06 doc/sc2012/conclusion.tex
Adding stuff for the camera ready version of the SC paper

\section{Conclusion}
\label{sec:conc}

We have presented Legion, a programming model for expressing
the abstract locality and independence properties of program
data by using logical regions.  The Legion programming model enabled us to implement
a distributed Software-Out-of-Order Processor (SOOP) scheduling 
algorithm.  The SOOP leveraged the locality and independence properties
capture by logical regions to efficiently execute programs.  We also
described a novel mapping interface that guides the SOOP execution
and allows tuning of program performance independent of correctness.
We demonstrated that our SOOP runtime implementation
of Legion was faster than existing implementations of
three benchmarks and that it was able to scale to distributed, heterogeneous
machines.

### 2012-08-06 doc/sc2012/legion.tex
Adding stuff for the camera ready version of the SC paper

@@
 \IEEEauthorblockN{Sean Treichler}
-\IEEEauthorblockA{StanfordUniversity \\ sjt@cs.stanford.edu}
+\IEEEauthorblockA{Stanford University \\ sjt@cs.stanford.edu}
 \and
@@
 \input{related}
-%\input{conclusion}
+\input{conclusion}
+
+\section*{Acknowledgements}
+This research used resources of the Keeneland Computing Facility at the
+Georgia Institute of Technology, which is supported by the National
+Science Foundation under Contract OCI-0910735.  Michael Bauer was supported
+by an NVIDIA Graduate Research Fellowship.
 

### 2012-08-08 doc/sc2012/code_ex.tex
Lots of small edits for the camera ready version of the SC paper

@@
 }
-
+                           // ROE = Read-Only-Exclusive
 void calc_new_currents(CircuitPiece piece):
@@
 }
-
+                          // RdA = Reduce-Atomic
 void distribute_charge(CircuitPiece piece, float dt):

### 2012-08-08 doc/sc2012/conclusion.tex
Lots of small edits for the camera ready version of the SC paper

@@
 the abstract locality and independence properties of program
-data by using logical regions.  The Legion programming model enabled us to implement
+data by using logical regions.  The Legion programming model enables 
 a distributed Software-Out-of-Order Processor (SOOP) scheduling 
-algorithm.  The SOOP leveraged the locality and independence properties
-capture by logical regions to efficiently execute programs.  We also
+algorithm.  The SOOP algorithm leverages the locality and independence properties
+captured by logical regions to efficiently execute programs.  We also
 described a novel mapping interface that guides the SOOP execution
@@
 of Legion was faster than existing implementations of
-three benchmarks and that it was able to scale to distributed, heterogeneous
+three benchmarks and that it was able to scale on distributed, heterogeneous
 machines.

### 2012-08-08 doc/sc2012/experiments.tex
Lots of small edits for the camera ready version of the SC paper

@@
 able to keep the entire simulation state in fast framebuffer memory.  Each line shows the scaling of
-a particular problem size as the number of nodes is varied.  Our results demonstrate
+a particular problem size as the number of compute nodes is varied.  Our results demonstrate
 excellent strong scaling, with speedups of 39.0X for the small problem on 48 GPUs and 
@@
 SOOP runtime.  As the node count increases, the non-communication overhead stays relatively constant.
-As expected the communication overhead grow linearly with number of nodes.
+As expected, the communication overhead grows quadratically with the number of nodes.
 
@@
 high communication-to-computation ratio which, for a fixed problem
-size, begins as being memory bound and with increasing node count
+size, begins as being memory bound and, with increasing node count,
 becomes network bound as the perimeter-to-area ratio of cell grids

### 2012-08-08 doc/sc2012/legion.tex
Lots of small edits for the camera ready version of the SC paper

@@
   fluid-flow on a regular grid, a three-level AMR code solving a
-  heat diffusion equation and a circuit simulation.
+  heat diffusion equation, and a circuit simulation.
   We compare each application with the best reference

### 2012-08-08 doc/sc2012/related.tex
Lots of small edits for the camera ready version of the SC paper

@@
 all access is exclusive, there is a static mapping of tasks and data (though
-extensions to Sequoia make this mapping more dynamic \cite{BCSA11}) and, most fundamentally, the decomposition of tasks and the decomposition
+extensions to Sequoia make this mapping more dynamic \cite{BCSA11}) and, most 
+fundamentally, the decomposition of tasks and the decomposition
 of data is one-to-one.  Legion generalizes the Sequoia model by allowing for
@@
 properties.  Legion's decoupling of the task tree from the region tree leads directly to the scheduling problem solved by our software out-of-order processor for tasks with region arguments.
+
+The SSMP programming model is the most similar work to Legion \cite{Perez10}.  
+Like Legion, SSMP supports dynamic detection of dependences between tasks based 
+on data requirements.  However, SSMP only supports a single disjoint rectilinear 
+partition of an array, unlike Legion which supports multiple
+arbitrary partitions of regions.  Furthermore, the SSMP runtime must perform dependence
+checks between every pair of tasks created in the system.  Legion's programming model only
+requires dependence checks between tasks with the same parent task which enables scalable nested
+parallelism in distributed machines.  SSMP only operates on shared memory machines.
 
@@
 mandates that data movement be explicitly managed by user level code
-or implicitly by the compiler or runtime \cite{X1008}.  Recently X10
+or implicitly by the compiler \cite{X1008}.  Recently X10
 has introduced regions into the compiler's
@@
 DPJ efforts have so far focused on JVM implementations on shared memory machines.
-There have also been several other recent efforts that propose task schedulers
+
+There have also been several other efforts that use partitions to either avoid
+or detect conflicts in shared memory programs dynamically.
+Object assemblies are a mechanism for partitioning a shared memory
+heap to enable parallel execution \cite{Lublinerman09}.  Object assemblies
+only support a disjoint partitioning of the heap unlike Legion which
+allows for multiple partitions of any data structure.  Legion further uses 
+the information from partitions to operate on distributed architectures.
+
+Galois is programming system based on optimistic parallelism that dynamically detects 
+memory conflicts between concurrent threads at the very
+fine granularity of object accesses.  To achieve reasonable performance
+Galois supports a partitioning interface for breaking up data structures \cite{Kulkarni08}.  
+Galois relies on shared memory
+and only leverages partitioning for coarser locking and conflict detection.
+Several other recent efforts have proposed task schedulers
 based on dynamic detection of memory conflicts, but these also have assumed

### 2012-08-08 doc/sc2012/soop.tex
Spelling corrections
Lots of small edits for the camera ready version of the SC paper

@@
 the SOOP is constrained by region dependences.
-The SOOP  is pipelined, distributed and extracts nested parallelism from subtasks.
+The SOOP  is pipelined, distributed, and extracts nested parallelism from subtasks.
 There are two major challenges in implementing an efficient task scheduler:
@@
 use the same physical instance of the logical region.  Since tasks
-haven't mapped physical instances at this point in the pipeline,
+have not mapped physical instances at this point in the pipeline,
 {\tt Same} is always a mapping dependence.
@@
 task $t$.  First, the mapper must ensure there is at least one {\em valid}
-physical instance of $r$ for $t$ to use.  Not all intances have up-to-date
+physical instance of $r$ for $t$ to use.  Not all instances have up-to-date
 data at all times.  For example, if an instance of a subregion of $r$ has been written by a previous task,

### 2012-08-08 doc/sc2012/soop.tex
soop

@@
 \begin{enumerate}
-\item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rw\_pvt} do not alias as they are subregions of a disjoint partition.
+\item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rw\_pvt} do not alias as they are different subregions of a disjoint partition.
 %\item {\tt pieces[0].rn\_pvt} and {\tt pieces[1].rn\_pvt} do not alias for the same reason.
@@
 %  partitions of the same region.
-\item {\tt pieces[0].rn\_ghost} and {\tt pieces[1].rn\_ghost} alias because they are in different subregions of
-an aliased partition.
+\item Subregions of an aliased partition always alias (e.g., {\tt pieces[0].rn\_ghost} and {\tt pieces[1].rn\_ghost}).
 \item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rn\_pvt}

### 2012-08-09 doc/sc2012/legion.tex
Fixing some contract stuff for darpa

@@
 \title{Legion: Expressing Locality and Independence with Logical
-  Regions\thanks{This work was supported by the Army High Performance
-    Research Center (DARPA contract HR0011-10-9-0008), a subcontract
+  Regions\thanks{This work was supported by the DARPA UHPC project
+    through a subcontract of NVIDIA (DARPA contract HR0011-10-9-0008),
+    and by the Army High Performance Research Center, through a subcontract
     from Los Alamos National Laboratory for the DOE Office of Science,
-    Advanced Scientific Computing Research (DE-AC52-06NA25396), and NVIDIA.  The views and
+    Advanced Scientific Computing Research (DE-AC52-06NA25396).  The views and
     conclusions contained in this document are those of the authors

### 2012-08-09 doc/sc2012/related.tex
Messing with DPJ paragraph in related work

@@
 Deterministic Parallel Java (DPJ) is a parallel extension of Java
-that, like Legion, uses regions to express locality and does
-dependence analysis at the granularity of conflicts on region
-arguments to functions \cite{Bocchino09}.  As the name suggests, one
-of the primary goals of DPJ is to provide a programming model that
+that, like Legion, uses regions to express locality, but does
+static dependence analysis on region
+arguments to functions to find dependences\cite{Bocchino09}.  The primary goal of DPJ 
+is to provide a programming model that
 guarantees determinism while also supporting parallelism.
-As a result the programming model is more restrictive than Legion, as
-there is no notion of weaker coherence properties of regions than
-exclusive access (which is essential to determinism).  DPJ can express
+As a result, the DPJ programming model is more restrictive than Legion.  
+DPJ only supports a static form of Legion's exclusive and atomic 
+coherence modes which mandates the same coherence for the lifetime
+of a region.  DPJ can express
 the populate-then-partition style of using regions, but not the
-partition-then-populate; overall, the language is more static, and
+partition-then-populate.  Overall, DPJ is more static, and
 thus provides more guarantees and less flexibility than Legion.  

### 2012-08-09 doc/sc2012/conclusion.tex
Minor edits

@@
 and allows tuning of program performance independent of correctness.
-We demonstrated that our SOOP runtime implementation
-of Legion was faster than existing implementations of
-three benchmarks and that it was able to scale on distributed, heterogeneous
-machines.
+We demonstrated that for three benchmarks our SOOP runtime 
+enabled Legion implementations that were faster than existing codes
+and that the SOOP was able to scale on distributed, heterogenous machines.

### 2012-08-10 doc/sc2012/experiments.tex
Working on some formatting

@@
 \includegraphics[scale=0.48]{figs/CircuitMem.pdf}
-\caption{Tasks and data for the circuit simulation on a cluster of GPUs.}
-\label{fig:gpumapping}
+\caption{Tasks and data for the circuit simulation on a cluster of GPUs. \label{fig:gpumapping}}
+\vspace{-6mm}
 \end{figure}
@@
 }
-\vspace{-2mm}
+\vspace{-3mm}
 \caption{Circuit simulation results.}
@@
 To measure scaling beyond a single node, three different problem sizes
-were run for each of the three systems (Figure~\ref{fig:fluid_multi}).
+were run on each of the three systems.
+The results are presented in Figure~\ref{fig:fluid_multi}.
 For the smallest problem (300K particles), we observe a 20\% speedup
 from 1 to 2 nodes (16 threads total), but slow
-down beyond that due to communication overhead---at 4 nodes there are
+down beyond that due to communication overhead; at 4 nodes there are
 twice as many ghost cells as interior grid cells.  The larger problem
-sizes (2.4M and 19M particles) do much better, with scaling of up to
-5.4x when going from 1 node up to 16 because of a lower
+sizes (2.4M and 19M particles) perform much better, with scaling of up to
+5.4x when going from 1 node to 16 because of a lower
 communication-to-computation ratio.
@@
 %}
-\vspace{-2mm}
+\vspace{-0mm}
 \caption{Fluid simulation results.}
-\vspace{-6mm}
+\vspace{2mm}
 \end{figure}

### 2012-08-13 doc/sc2012/related.tex
Fixing minor grammar

@@
 
-Galois is programming system based on optimistic parallelism that dynamically detects 
+Galois is a programming system based on optimistic parallelism that dynamically detects 
 memory conflicts between concurrent threads at the very

### 2012-08-14 doc/sc2012/circuit.tex
Small edits to the circuit example
Making some colors better in gray scale
Lots of tiny edits for the camera ready version

@@
 
+We begin by describing an example program
+written in the Legion programming model.
 Listing~\ref{lst:code_ex} shows code for an electrical
@@
 attached by wires only to nodes in the same piece, are 
-{\em private} to the piece.  Orange (darker) nodes, on the boundary of a
+{\em private} to the piece.  Red (darker) nodes, on the boundary of a
 piece, are {\em shared} with (connected to) other pieces.
@@
 
-Lines 31-57 form the bulk of the actual simulation, performing three
-passes over the circuit in each iteration.  Each pass loops over an
+Lines 31-57 perform the actual simulation by making three
+passes over the circuit for each time step.  Each pass loops over an
 array of pieces (constructed on lines 27-30 from the partitions),
@@
 parallel execution ({\tt spawn} simply indicates a task call and does
-not mandate parallel execution) nor is there synchronization 
+not mandate parallel execution) nor is there explicit synchronization 
 between the passes.  Which tasks can be run in

### 2012-08-14 doc/sc2012/conclusion.tex
Lots of tiny edits for the camera ready version

--- 2012-08-08 doc/sc2012/conclusion.tex
@@
 and allows tuning of program performance independent of correctness.
-We demonstrated that our SOOP runtime implementation
-of Legion was faster than existing implementations of
-three benchmarks and that it was able to scale on distributed, heterogeneous
-machines.
+We demonstrated that for three benchmarks our SOOP runtime 
+enabled Legion implementations that were faster than existing codes
+and that the SOOP was able to scale on distributed, heterogeneous machines.

### 2012-08-14 doc/sc2012/legion.tex
Lots of tiny edits for the camera ready version

@@
   several stages, our SOOP is distributed across the machine and is
-  also hierarchical and naturally extracts nested
+  also hierarchical to naturally extract nested
   parallelism (because tasks may recursively spawn subtasks).
@@
 
-\section*{Acknowledgements}
+\section*{Acknowledgments}
 This research used resources of the Keeneland Computing Facility at the

### 2012-08-14 doc/sc2012/mapping.tex
Lots of tiny edits for the camera ready version

@@
 making possible application- or machine-specific mapping decisions that would be difficult for a 
-general-purpose programming system to infer.
+general-purpose programming system to infer.  Furthermore, this interface is invoked at runtime which
+allows for dynamic mapping decisions based on program input data.
 We describe the interface (Section \ref{sec:mapinterface}),
@@
 for inspecting properties of the machine, including a list of
-processors and their type (i.e. CPU, GPU), a list of
+processors and their type (e.g. CPU, GPU), a list of
 memories visible to each processor, and their latencies and bandwidths.
@@
 To optimize a Legion program or library, programmers can create one or more custom mappers.  
-For each Legion task call the programmer specifies which mapper should be invoked by the runtime for
-mapping that particular task.  This allows for the composition of Legion applications and libraries
-each with their own custom mappers.
+Each custom mapper extends the default mapper.  A programmer need only override the mapper functions 
+%necessary for creating a custom mapping.  
+he wishes to customize.  Mappers are registered with the runtime and given unique handles.
+When a task is launched, the programmer specifies the handle for the mapper that should be invoked by the runtime for
+mapping that particular task.
 
-Each custom mapper extends the default mapper.  A programmer need only override the functions 
-%necessary for creating a custom mapping.  
-he wishes to customize.  Custom mappers can be used to create totally static mappings, 
-mappings that memoize their results, or even totally dynamic mappings.  We describe examples of 
+Supporting custom mappers has two benefits.  First, it allows for the composition of Legion applications
+and Legion libraries each with their own custom mappers.
+Second, custom mappers can be used to create totally static mappings, 
+mappings that memoize their results, or even totally dynamic mappings for different 
+subsets of tasks in Legion applications.  We describe examples of 
 custom mappers in Section \ref{sec:exp}.
+
+%  This allows for the composition of Legion applications and libraries
+%each with their own custom mappers.

### 2012-08-14 doc/sc2012/soop.tex
Lots of tiny edits for the camera ready version

@@
 Legion uses a {\em  software out-of-order processor}, or SOOP, to schedule tasks.  A SOOP 
-dynamically schedules a stream of tasks in a manner akin to an out-of-order processor scheduling a stream of instructions:
-just as an instruction scheduler is constrained by register dependences,
+dynamically schedules a stream of tasks in a manner akin to an out-of-order processor scheduling a stream of instructions.
+Just as an instruction scheduler is constrained by register dependences,
 the SOOP is constrained by region dependences.

### 2012-09-10 doc/popl2013/response.txt
initial draft of response to PLDI reviewers

We'd first like to thank the reviewers for their detailed and useful feedback.
Rather than addressing each reviewer's questions separately, we have organized
our response based on the areas the questions covered.

First, there were multiple requests for more detail on the code example, the
use of colorings to describe partitions, and the algorithm used to perform 
efficient dynamic disjointness checks.  Each of these was covered in significant
detail in [1], {which we understand the reviewers may not have had access to.}

As described in [1], Legion does not attempt to solve the (in many cases still-open)
problem of partitioning data for any given application.  The body of the "color_circuit"
task would likely consist of a call to an external library (e.g. PARMETIS for graph-based
data structures).  The coloring is merely a way to capture the result of that external
computation for use by the Legion runtime.  The "build_lists" task is not strictly required.
While a task could iterate over the elements of a subregion through repeated uses of "downregion" on
each element of the parent region, by building a list up front, the cost of those "downregion"
checks can be amortized over multiple iterations of the simulation.

In Core Legion, colorings are explicit mappings of region elements (named by their location)
to integer "colors".  This represents the most general case of describing the partitioning of
a region.  More concise descriptions exist in the full Legion language (e.g. using a pure
function to determine the color of an element on demand), but a single form was selected for
Core Legion for simplicity.

Legion supports the "partition then allocate" model in which new elements allocated in a subregion
are guaranteed to exist in parent regions as well and preserve that subregion's disjointness with
any other regions.  The use of a non-empty initial coloring in the operational semantics simplifies
this in Core Legion by allowing extra unnamed locations to exist in every region to allow these
allocations without changing the sets of locations assigned to physical regions.  {Is this opening a
can of worms?}

As several reviewers noted, the Legion type system is similar in many ways to existing region-based
typed systems, and this is by design.  Additions to the "standard" region-based type system were 
done just to capture the new features of Legion: colorings as a description of an arbitrary partitioning
of a region, a hierarchy of regions in which a location may be contained in multiple regions, the 
ability to capture the relationship between regions, and the ability to relax the determinism of 
execution via coherence modes.  Although these changes were made as contained as
possible, a proof that soundness was preserved by these additions seemed prudent.

The use of coherence modes in Legion differs from existing usages, so we tried to reduce confusion by
not reusing existing terminology.  While "exclusive" coherence corresponds to apparently-sequential
execution, "atomic" corresponds to transactional-memory-like execution, and "simultaneous" corresponds to
"all bets are off" pthreads-style parallelism, an important difference is that these coherence modes apply
locally in Legion, i.e. only between subtasks of the same parent task.  All three of these modes have
their uses in real-world applications - to exclude any of them would limit the applicability of the Legion
programming model.  We felt the additional complexity in the type system and operational semantics was
justified by the results that both soundness of the type system and ability to make local scheduling decisions
were noy compromised by the inclusion of these coherence modes.

Finally, the choice of large-step operational semantics was made to match the hierachical and distributed
scheduling used by the Legion runtime.  When coherence modes other than exclusive are used, there may be
multiple tasks modifying the same location of memory - the clobber set is the set of all such locations for
a given task.  The clobber set is used to bound the allowed non-determinism within a given task, and the
fact that the top-level task (which has no sibling tasks) has an empty clobber set guarantees that any 
execution of the program is consistent with a standard sequentially consistent memory model.

An advantage of the large-step semantics with the use of clobber sets is that our soundness result holds for
any scheduling of tasks that satisfies the "valid_interleave" rules.  We considered the use of small-step semantics
that explicitly track the state of every parallel task's execution, but the complexity of the constraints used
to decide which task(s) were able to take a step at any given point were onerous and would limit our soundness
guarantees to only that scheduling algorithm.

### 2012-09-10 doc/popl2013/response.txt
more edits

@@
 efficient dynamic disjointness checks.  Each of these was covered in significant
-detail in [1], {which we understand the reviewers may not have had access to.}
+detail in [1], which may have been unavailable to the reviewers previously but can
+now be found at http://theory.stanford.edu/~aiken/publications/papers/sc12.pdf .
 
@@
 
-Legion supports the "partition then allocate" model in which new elements allocated in a subregion
+[Legion supports the "partition then allocate" model in which new elements allocated in a subregion
 are guaranteed to exist in parent regions as well and preserve that subregion's disjointness with
@@
 allocations without changing the sets of locations assigned to physical regions.  {Is this opening a
-can of worms?}
+can of worms?}]
 
 As several reviewers noted, the Legion type system is similar in many ways to existing region-based
-typed systems, and this is by design.  Additions to the "standard" region-based type system were 
-done just to capture the new features of Legion: colorings as a description of an arbitrary partitioning
+typed systems; this is by design.  Additions to the "standard" region-based type system were 
+done to capture the new features of Legion: colorings as a description of an arbitrary partitioning
 of a region, a hierarchy of regions in which a location may be contained in multiple regions, the 
 ability to capture the relationship between regions, and the ability to relax the determinism of 
-execution via coherence modes.  Although these changes were made as contained as
-possible, a proof that soundness was preserved by these additions seemed prudent.
+execution via coherence modes.  With these additional features a proof that soundness was preserved 
+was a prerequisite for later proofs in our paper.
 
-The use of coherence modes in Legion differs from existing usages, so we tried to reduce confusion by
-not reusing existing terminology.  While "exclusive" coherence corresponds to apparently-sequential
-execution, "atomic" corresponds to transactional-memory-like execution, and "simultaneous" corresponds to
-"all bets are off" pthreads-style parallelism, an important difference is that these coherence modes apply
-locally in Legion, i.e. only between subtasks of the same parent task.  All three of these modes have
-their uses in real-world applications - to exclude any of them would limit the applicability of the Legion
-programming model.  We felt the additional complexity in the type system and operational semantics was
-justified by the results that both soundness of the type system and ability to make local scheduling decisions
-were noy compromised by the inclusion of these coherence modes.
+A description of the usage and intuition behind coherence modes was fully presented in [1] and
+was omitted from this paper due to space constraints.
+"Exclusive" coherence corresponds to apparently-sequential execution, "atomic" corresponds 
+to transactional-memory-like execution, and "simultaneous" corresponds to
+"all bets are off" pthreads-style parallelism.  However, unlike other implementations, coherence modes 
+only allow for reordering of subtasks with the same parent task, permitting the containment of non-determinism
+to a portion of the task hierarchy.  This paper contains two new
+results about coherence modes not contained in [1].  First, we give a formal semantics of the possible
+interleavings allowed by 
+different coherence modes and their interactions.  Second, we prove that coherence modes do not
+impact the soundness of our type system or the correctness of our hierarchical scheduling algorithm.
 

### 2012-09-11 doc/popl2013/response.txt
editting popl response

@@
 to transactional-memory-like execution, and "simultaneous" corresponds to
-"all bets are off" pthreads-style parallelism.  However, unlike other implementations, coherence modes 
-only allow for reordering of subtasks with the same parent task, permitting the containment of non-determinism
-to a portion of the task hierarchy.  This paper contains two new
-results about coherence modes not contained in [1].  First, we give a formal semantics of the possible
-interleavings allowed by 
+"all bets are off" pthreads-style parallelism. However, unlike previous work, coherence modes 
+only allow for reordering of subtasks with the same parent task.  This paper contains two useful
+results about coherence modes.  First, we give a formal semantics of the possible interleavings allowed by 
 different coherence modes and their interactions.  Second, we prove that coherence modes do not

### 2012-09-11 doc/popl2013/response.txt
More edits to the response

--- 2012-09-10 doc/popl2013/response.txt
@@
-We'd first like to thank the reviewers for their detailed and useful feedback.
+We would first like to thank the reviewers for their detailed and useful feedback.
 Rather than addressing each reviewer's questions separately, we have organized
@@
 only allow for reordering of subtasks with the same parent task, permitting the containment of non-determinism
-to a portion of the task hierarchy.  This paper contains two new
-results about coherence modes not contained in [1].  First, we give a formal semantics of the possible
-interleavings allowed by 
+to a portion of the task hierarchy.  This paper contains two new results about coherence modes 
+not presented in [1].  First, we give a formal semantics of the possible interleavings allowed by 
 different coherence modes and their interactions.  Second, we prove that coherence modes do not

### 2012-09-11 doc/popl2013/response.txt
small edit

@@
 efficient dynamic disjointness checks.  Each of these was covered in significant
-detail in [1], which may have been unavailable to the reviewers previously but can
+detail in our prior work [1], which may have been unavailable to the reviewers previously but can
 now be found at http://theory.stanford.edu/~aiken/publications/papers/sc12.pdf .

### 2012-09-11 doc/popl2013/response.txt
Lots of edits to the popl response

--- 2012-09-11 doc/popl2013/response.txt
@@
-We'd first like to thank the reviewers for their detailed and useful feedback.
-Rather than addressing each reviewer's questions separately, we have organized
-our response based on the areas the questions covered.
+We thank the reviewers for their thoughtful, extensive, and very detailed comments.
+Most of the comments are specifically about presentation; we acknowledge the issue,
+and if the paper is accepted we will address the concerns raised.
 
-First, there were multiple requests for more detail on the code example, the
-use of colorings to describe partitions, and the algorithm used to perform 
-efficient dynamic disjointness checks.  Each of these was covered in significant
-detail in [1], which may have been unavailable to the reviewers previously but can
+Many of the reviewers raise similar questions, which we have grouped together in a
+single response below in addition to answering issues raised by individual reviewers.
+To keep the response to a reasonable length we have tried to address the most common
+and significant issues.  In many cases, questions were asked which were addressed
+in our prior published work [1], which may have been unavailable to the reviewers previously but can
 now be found at http://theory.stanford.edu/~aiken/publications/papers/sc12.pdf .
 
-As described in [1], Legion does not attempt to solve the (in many cases still-open)
-problem of partitioning data for any given application.  The body of the "color_circuit"
-task would likely consist of a call to an external library (e.g. PARMETIS for graph-based
-data structures).  The coloring is merely a way to capture the result of that external
-computation for use by the Legion runtime.  The "build_lists" task is not strictly required.
-While a task could iterate over the elements of a subregion through repeated uses of "downregion" on
-each element of the parent region, by building a list up front, the cost of those "downregion"
-checks can be amortized over multiple iterations of the simulation.
-
+Reviewers A, B, and C all asked for additional details about colorings.
 In Core Legion, colorings are explicit mappings of region elements (named by their location)
@@
 function to determine the color of an element on demand), but a single form was selected for
-Core Legion for simplicity.
+Core Legion for simplicity.  As described in [1], Legion does not attempt to solve the (in many cases still-open)
+problem of partitioning data for any given application.  In our example, the body of the "color_circuit"
+task would require the programmer to compute a partition of the circuit graph and return a coloring
+that assigns each circuit element a color corresponding to its piece of the partition.  
+In production scientific codes this is often done by a call 
+to an external library (e.g. PARMETIS for graph-based data structures).  
+A coloring is merely a way to capture the result of the partitioning 
+computation in a general format understood by the Legion runtime.
+
+Reviewer A asked how "downregion" operations are performed and used.  The "downregion" operation
+requires a dynamic check that the given location is valid in the provided logical region.  To
+perform these checks the runtime remembers which locations have been allocated in each logical
+region and dynamically checks that the location has been allocated for the given region.  The 
+"build_lists" task is an example of an optimization done to reduce the number of "downregion" calls.
+While a task could iterate over the elements of a subregion through repeated uses of "downregion" on
+each element of the parent region, by building a list typed on the subregion up front, the cost 
+of those "downregion" checks can be amortized over multiple iterations of the simulation.
+
+Both reviewers A and B inquired how the dynamic disjointness tests are performed.  This
+is the primary subject of our prior work [1].
 
@@
 
-As several reviewers noted, the Legion type system is similar in many ways to existing region-based
-typed systems; this is by design.  Additions to the "standard" region-based type system were 
-done to capture the new features of Legion: colorings as a description of an arbitrary partitioning
-of a region, a hierarchy of regions in which a location may be contained in multiple regions, the 
-ability to capture the relationship between regions, and the ability to relax the determinism of 
-execution via coherence modes.  With these additional features a proof that soundness was preserved 
-was a prerequisite for later proofs in our paper.
+Reviewer B commented that the Legion type system is similar in several ways to existing region-based
+typed systems.  While we reuse many features of prior region languages, we believe that Legion
+represents a fundamentally new composition of these features.  Primarily Legion allows for first-class
+logical regions which permit elements to exist in many regions simultaneously via multiple dynamic
+partitions of regions.  Furthermore, the way the region information is leveraged both at compile-time 
+and at runtime represent a new point in the design space of region languages.  While the proof of
+soundness is not novel, we felt it necessary given the difference in semantics of logical regions 
+from previous region languages.  The soundness proof was also a prerequisite for our later proofs
+concerning coherence modes and hierarchical scheduling.
 
+Both reviewers B and C thought coherence modes were an interesting technical contribution, but
+would have liked a higher-level description to build intuition.
 A description of the usage and intuition behind coherence modes was fully presented in [1] and
-was omitted from this paper due to space constraints.
-"Exclusive" coherence corresponds to apparently-sequential execution, "atomic" corresponds 
+was omitted from this paper due to space constraints.  Our goal in this paper was to present
+two new results about coherence modes not presented in [1].  First, we present a formal semantics 
+of the possible interleavings allowed by different coherence modes and their interactions.  
+Second, we prove that coherence modes do not impact the soundness of our type system or the 
+correctness of our hierarchical scheduling algorithm.  A brief summary of coherence modes is:
+"exclusive" coherence corresponds to apparently-sequential execution, "atomic" corresponds 
 to transactional-memory-like execution, and "simultaneous" corresponds to
-"all bets are off" pthreads-style parallelism. However, unlike previous work, coherence modes 
-only allow for reordering of subtasks with the same parent task.  This paper contains two useful
-results about coherence modes.  First, we give a formal semantics of the possible interleavings allowed by 
-different coherence modes and their interactions.  Second, we prove that coherence modes do not
-impact the soundness of our type system or the correctness of our hierarchical scheduling algorithm.
+"all bets are off" pthreads-style parallelism.  Unlike other implementations, coherence modes 
+only allow for reordering of subtasks with the same parent task, permitting the containment of non-determinism
+to a portion of the task hierarchy.
 
-Finally, the choice of large-step operational semantics was made to match the hierachical and distributed
+Reviewer D asked what is currently implemented.  Legion currently consists of a static type 
+checker for programs written in Core-Legion and a C++ runtime library for fully-featured Legion programs.
+
+Finally, reviewers B, C, and D expressed confusion about the nature of the clobber set.
+The choice of large-step operational semantics was made to match the hierachical and distributed
 scheduling used by the Legion runtime.  When coherence modes other than exclusive are used, there may be

### 2012-09-11 doc/popl2013/response.txt
update

@@
-We thank the reviewers for their thoughtful, extensive, and very detailed comments.
-Most of the comments are specifically about presentation; we acknowledge the issue,
-and if the paper is accepted we will address the concerns raised.
+We thank the reviewers for their very detailed comments.  Most of the
+comments are specifically about presentation; we acknowledge the
+issue, and if the paper is accepted we will address the concerns
+raised.
 
-Many of the reviewers raise similar questions, which we have grouped together in a
-single response below in addition to answering issues raised by individual reviewers.
-To keep the response to a reasonable length we have tried to address the most common
-and significant issues.  In many cases, questions were asked which were addressed
-in our prior published work [1], which may have been unavailable to the reviewers previously but can
-now be found at http://theory.stanford.edu/~aiken/publications/papers/sc12.pdf .
+Many of the reviewers raise similar questions, which we have grouped
+together in a single response below in addition to answering issues
+raised by individual reviewers.  To keep the response to a reasonable
+length we have tried to address the most common and significant
+issues.  In many cases, questions were asked which were addressed in
+our prior published work [1], which may have been unavailable to the
+reviewers previously but can now be found at
+http://theory.stanford.edu/~aiken/publications/papers/sc12.pdf .
 
@@
 Both reviewers A and B inquired how the dynamic disjointness tests are performed.  This
-is the primary subject of our prior work [1].
+is a primary subject of our prior work [1].
 
@@
 and at runtime represent a new point in the design space of region languages.  While the proof of
-soundness is not novel, we felt it necessary given the difference in semantics of logical regions 
-from previous region languages.  The soundness proof was also a prerequisite for our later proofs
+soundness does not introduce novel proof techniques, it is also far from immediate that the type system
+is sound and we feel the proof is necessary given large differences in the semantics of logical regions 
+from previous region systems.  The soundness proof was also a prerequisite for our later proofs
 concerning coherence modes and hierarchical scheduling.
@@
 would have liked a higher-level description to build intuition.
-A description of the usage and intuition behind coherence modes was fully presented in [1] and
+A description of the usage and intuition behind coherence modes is presented in [1] and
 was omitted from this paper due to space constraints.  Our goal in this paper was to present
-two new results about coherence modes not presented in [1].  First, we present a formal semantics 
+two new results about coherence modes that are not in [1].  First, we present a formal semantics 
 of the possible interleavings allowed by different coherence modes and their interactions.  

### 2012-09-12 doc/popl2013/response.txt
couple more edits, answers to the last open questions, spell check

@@
 task would require the programmer to compute a partition of the circuit graph and return a coloring
-that assigns each circuit element a color corresponding to its piece of the partition.  
+that assigns each circuit element a color corresponding to its piece of the partition.  The same
+set of colors are used to identify the subregions that result from the partitioning.
 In production scientific codes this is often done by a call 
@@
 computation in a general format understood by the Legion runtime.
+
+The principle reason to support multiple partitions of the same data is to allow precise descriptions
+of the sets of elements that will be accessed by a given task.  These sets are often different in 
+different phases of an algorithm, and the increased precision is necessary to safely parallelize
+applications such as the circuit example.  Although the subregions of different partitions are often
+accessed with different privileges, this is a property of an algorithm rather than a Legion requirement.
 
@@
 is a primary subject of our prior work [1].
-
-[Legion supports the "partition then allocate" model in which new elements allocated in a subregion
-are guaranteed to exist in parent regions as well and preserve that subregion's disjointness with
-any other regions.  The use of a non-empty initial coloring in the operational semantics simplifies
-this in Core Legion by allowing extra unnamed locations to exist in every region to allow these
-allocations without changing the sets of locations assigned to physical regions.  {Is this opening a
-can of worms?}]
 
@@
 to transactional-memory-like execution, and "simultaneous" corresponds to
-"all bets are off" pthreads-style parallelism.  Unlike other implementations, coherence modes 
-only allow for reordering of subtasks with the same parent task, permitting the containment of non-determinism
-to a portion of the task hierarchy.
+"all bets are off" pthreads-style parallelism.  However, unlike other implementations, 
+Legion's coherence modes only allow for reordering of subtasks with the same parent task, 
+permitting the containment of non-determinism to a portion of the task hierarchy.
+
+Reviewer A asked if the operational semantics would be simpler if the only relaxation of sequential
+semantics came from an explicit "par(e1,e2)" construct.  While the semantics would indeed be simpler,
+the use of a static construct (rather than one able to consider dynamic region mappings) would limit
+the expressible forms of parallelism to a (well-studied) subset of what Legion programs can achieve.
 
 Reviewer D asked what is currently implemented.  Legion currently consists of a static type 
-checker for programs written in Core-Legion and a C++ runtime library for fully-featured Legion programs.
+checker for programs written in Core Legion and a C++ runtime library for fully-featured Legion programs.
 
 Finally, reviewers B, C, and D expressed confusion about the nature of the clobber set.
-The choice of large-step operational semantics was made to match the hierachical and distributed
+The choice of large-step operational semantics was made to match the hierarchical and distributed
 scheduling used by the Legion runtime.  When coherence modes other than exclusive are used, there may be
@@
 to decide which task(s) were able to take a step at any given point were onerous and would limit our soundness
-guarantees to only that scheduling algorithm.
+guarantees to only a particular scheduling algorithm.

### 2012-09-12 doc/popl2013/response.txt
re-added an answer to why E-NewColor uses a non-empty initial mapping

@@
 
-Many of the reviewers raise similar questions, which we have grouped
+Many of the reviewers raised similar questions, which we have grouped
 together in a single response below in addition to answering issues
@@
 
-Reviewers A, B, and C all asked for additional details about colorings.
+Reviewers A, B, and C all asked for additional details about colorings which were covered in [1].
 In Core Legion, colorings are explicit mappings of region elements (named by their location)
@@
 set of colors are used to identify the subregions that result from the partitioning.
-In production scientific codes this is often done by a call 
+In production scientific codes partitioning is often done by a call 
 to an external library (e.g. PARMETIS for graph-based data structures).  
@@
 
+Reviewer C asked for clarification concerning the the relationship between multiple partitions and privileges.
 The principle reason to support multiple partitions of the same data is to allow precise descriptions
@@
 and at runtime represent a new point in the design space of region languages.  While the proof of
-soundness does not introduce novel proof techniques, it is also far from immediate that the type system
-is sound and we feel the proof is necessary given large differences in the semantics of logical regions 
+soundness does not introduce novel proof techniques, it is far from immediate that the type system
+is sound and we felt the proof was necessary given the large differences in the semantics of logical regions 
 from previous region systems.  The soundness proof was also a prerequisite for our later proofs
@@
 A description of the usage and intuition behind coherence modes is presented in [1] and
-was omitted from this paper due to space constraints.  Our goal in this paper was to present
+was omitted from this paper due to space constraints.  In this paper we present
 two new results about coherence modes that are not in [1].  First, we present a formal semantics 
@@
 guarantees to only a particular scheduling algorithm.
+
+Reviewer A asked why the E-NewColor rule allows the initial coloring to be non-empty.  There are two
+reasons for this.  First, in the large-step semantics we use, the region mapping and heap typing cannot
+be changed by an E-New rule, so a partitioning of a region must include extra elements of the appropriate
+heap type to support allocation within a subregion.  Second, it allows implementations in which 
+an existing element that was not colored by the programmer is included in (at most one) arbitrary
+subregion, which can improve the efficiency of data transfers at runtime.

### 2012-09-12 doc/popl2013/response.txt
Some minor edits

@@
 
-Reviewer C asked for clarification concerning the the relationship between multiple partitions and privileges.
+Reviewer C ask for clarification concerning the the relationship between multiple partitions and privileges.
 The principle reason to support multiple partitions of the same data is to allow precise descriptions
@@
 guarantees to only a particular scheduling algorithm.
-
-Reviewer A asked why the E-NewColor rule allows the initial coloring to be non-empty.  There are two
-reasons for this.  First, in the large-step semantics we use, the region mapping and heap typing cannot
-be changed by an E-New rule, so a partitioning of a region must include extra elements of the appropriate
-heap type to support allocation within a subregion.  Second, it allows implementations in which 
-an existing element that was not colored by the programmer is included in (at most one) arbitrary
-subregion, which can improve the efficiency of data transfers at runtime.

### 2012-09-13 doc/popl2013/response.txt
small edit to response
udpate

--- 2012-09-12 doc/popl2013/response.txt
@@
 We thank the reviewers for their very detailed comments.  Most of the
-comments are specifically about presentation; we acknowledge the
-issue, and if the paper is accepted we will address the concerns
-raised.
+comments are specifically about presentation.  If the paper is accepted 
+we will certainly address all the concerns raised; below we focus on what
+appear to be the most common or important points.
 
-Many of the reviewers raised similar questions, which we have grouped
-together in a single response below in addition to answering issues
-raised by individual reviewers.  To keep the response to a reasonable
-length we have tried to address the most common and significant
-issues.  In many cases, questions were asked which were addressed in
-our prior published work [1], which may have been unavailable to the
-reviewers previously but can now be found at
+In many cases, questions were asked which were addressed in our prior
+published work [1], which may have been unavailable to the reviewers
+previously but can now be found at
 http://theory.stanford.edu/~aiken/publications/papers/sc12.pdf .
 
-Reviewers A, B, and C all asked for additional details about colorings which were covered in [1].
-In Core Legion, colorings are explicit mappings of region elements (named by their location)
-to integer "colors".  This represents the most general case of describing the partitioning of
-a region.  More concise descriptions exist in the full Legion language (e.g. using a pure
-function to determine the color of an element on demand), but a single form was selected for
-Core Legion for simplicity.  As described in [1], Legion does not attempt to solve the (in many cases still-open)
-problem of partitioning data for any given application.  In our example, the body of the "color_circuit"
-task would require the programmer to compute a partition of the circuit graph and return a coloring
-that assigns each circuit element a color corresponding to its piece of the partition.  The same
-set of colors are used to identify the subregions that result from the partitioning.
-In production scientific codes partitioning is often done by a call 
-to an external library (e.g. PARMETIS for graph-based data structures).  
-A coloring is merely a way to capture the result of the partitioning 
+Reviewers A, B, and C asked for additional details about colorings.
+In Core Legion, colorings are explicit mappings of region elements
+(named by their location) to integers ("colors"), which represents the
+most general case of describing the partitioning of a region.  More
+concise descriptions exist in the full Legion language (e.g. using a
+pure function to determine the color of an element on demand), but a
+single form was selected for Core Legion for simplicity.  As described
+in [1], Legion does not attempt to solve the (in many cases
+still-open) problem of optimally partitioning data for any given
+application, but instead provides a mechanism for programmers to write
+their own partitioning code and communicate the partitioning decisions
+to the Legion runtime.  In our example, the body of the
+"color_circuit" task would require the programmer to compute a
+partition of the circuit graph and return a coloring that assigns each
+circuit element a color; each color identifies a separate subregion of
+the partition consisting of all the circuit elements with that color.
+In production scientific codes partitioning is often done by a call to
+an external library (e.g. PARMETIS for graph-based data structures).
+A coloring is a way to capture the result of the partitioning
 computation in a general format understood by the Legion runtime.
+More information on colorings can be found in [1].
 
-Reviewer C asked for clarification concerning the the relationship between multiple partitions and privileges.
-The principle reason to support multiple partitions of the same data is to allow precise descriptions
-of the sets of elements that will be accessed by a given task.  These sets are often different in 
-different phases of an algorithm, and the increased precision is necessary to safely parallelize
-applications such as the circuit example.  Although the subregions of different partitions are often
-accessed with different privileges, this is a property of an algorithm rather than a Legion requirement.
+Reviewer C correctly points out that regions are central to
+performance as well as for the type structure of Legion programs.
+Regions are in Legion to allow programmers to describe locality and
+independence for structuring parallel computations.  The programmer's
+choice of egions can affect load balancing, communication, and the
+granularity of parallel tasks.  In the case of the circuit example, no
+partitioning scheme that does not look at the actual structure of the
+graph being partitioned can pick a partition that will lead to good
+performance; that is, for performance reasons the partitioning should
+be done dynamically.  Assuming Legion has a sound type system,
+however, any partitioning should be correct.  The paper focuses on the
+correctness issues for the Legion implementation: that there are no
+wild pointers to the wrong regions and that tasks observe both
+privilege and coherence requirements.
 
-Reviewer A asked how "downregion" operations are performed and used.  The "downregion" operation
-requires a dynamic check that the given location is valid in the provided logical region.  To
-perform these checks the runtime remembers which locations have been allocated in each logical
-region and dynamically checks that the location has been allocated for the given region.  The 
-"build_lists" task is an example of an optimization done to reduce the number of "downregion" calls.
-While a task could iterate over the elements of a subregion through repeated uses of "downregion" on
-each element of the parent region, by building a list typed on the subregion up front, the cost 
-of those "downregion" checks can be amortized over multiple iterations of the simulation.
+Reviewer C asked for clarification concerning the the relationship
+between multiple partitions and privileges.  The principle reason to
+support multiple partitions of the same data is to allow precise
+descriptions of the sets of elements that will be accessed by a given
+task.  These sets are often different in different phases of an
+algorithm, and both naming the precise set of nodes and the privileges
+required are needed to safely parallelize applications such as the
+circuit code.  In some cases (like the circuit code) it is possible in
+theory to have a single partitioning of the data by taking the product
+of all the overlapping partitions, but this results in many, many
+unintuitive distinct subregions and is extremely unpleasant
+to program with (we tried it once).
 
-Both reviewers A and B inquired how the dynamic disjointness tests are performed.  This
-is a primary subject of our prior work [1].
+Reviewer A asked how "downregion" operations are performed and used.
+The "downregion" operation requires a dynamic check that the given
+location is valid in the provided logical region.  To perform these
+checks the runtime remembers which locations have been allocated in
+each logical region and dynamically checks that the location has been
+allocated for the given region.  The "build_lists" task is an example
+of an optimization done to reduce the number of "downregion" calls.
+While a task could iterate over the elements of a subregion through
+repeated uses of "downregion" on each element of the parent region, by
+building a list typed on the subregion up front, the cost of those
+"downregion" checks can be amortized over multiple iterations of the
+simulation.
 
-Reviewer B commented that the Legion type system is similar in several ways to existing region-based
-typed systems.  While we reuse many features of prior region languages, we believe that Legion
-represents a fundamentally new composition of these features.  Primarily Legion allows for first-class
-logical regions which permit elements to exist in many regions simultaneously via multiple dynamic
-partitions of regions.  Furthermore, the way the region information is leveraged both at compile-time 
-and at runtime represent a new point in the design space of region languages.  While the proof of
-soundness does not introduce novel proof techniques, it is far from immediate that the type system
-is sound and we felt the proof was necessary given the large differences in the semantics of logical regions 
-from previous region systems.  The soundness proof was also a prerequisite for our later proofs
+Both reviewers A and B inquired how the dynamic disjointness tests are
+performed.  This is a primary subject of our prior work [1].  Briefly,
+the runtime keeps track of the tree of region partitions and the tasks
+that are currently using those regions (and with what
+privileges/coherence).  When a new task is launched its region
+arguments (and required privileges/coherence) are checked against the
+current region tree to decide which, if any, of the already running
+tasks must complete before the new task can execute.  This test can be
+made provably efficient, even when scheduling is concurrent and
+distributed, but relies crucially on the soundness of the type system
+for correctness.
+
+Reviewer B commented that the Legion type system is similar in several
+ways to existing region-based typed systems.  While we reuse many
+features of prior region languages, we believe that Legion represents
+a new composition of these features.  Primarily Legion allows for
+first-class logical regions which permit elements to exist in many
+regions simultaneously via multiple dynamic partitions of regions.
+Furthermore, the way the region information is leveraged both at
+compile-time and at runtime represent a new point in the design space
+of region languages.  We would like to point out specifically the way
+in which existential types are used to read (packed) regions from the
+heap and open them with permission to read/write/reduce values in
+those regions - there is a subtle interaction between free variables
+in existential types and bounded quantification of privileges, but
+this is what allows Legion to check statically that a task can, say,
+iterate over an unbounded list L of subregions and access elements of
+those regions so long as there is a constraint that shows the task has
+the same permissions on some ancestor region of every subregion in L.
+And while the proof of soundness does not introduce novel proof
+techniques, it is far from immediate that the type system is sound and
+we felt the proof was necessary given the large differences in the
+semantics of logical regions from previous region systems.  The
+soundness proof was also a prerequisite for our later proofs
 concerning coherence modes and hierarchical scheduling.
 
-Both reviewers B and C thought coherence modes were an interesting technical contribution, but
-would have liked a higher-level description to build intuition.
-A description of the usage and intuition behind coherence modes is presented in [1] and
-was omitted from this paper due to space constraints.  In this paper we present
-two new results about coherence modes that are not in [1].  First, we present a formal semantics 
-of the possible interleavings allowed by different coherence modes and their interactions.  
-Second, we prove that coherence modes do not impact the soundness of our type system or the 
-correctness of our hierarchical scheduling algorithm.  A brief summary of coherence modes is:
-"exclusive" coherence corresponds to apparently-sequential execution, "atomic" corresponds 
-to transactional-memory-like execution, and "simultaneous" corresponds to
-"all bets are off" pthreads-style parallelism.  However, unlike other implementations, 
-Legion's coherence modes only allow for reordering of subtasks with the same parent task, 
-permitting the containment of non-determinism to a portion of the task hierarchy.
+Both reviewers B and C thought coherence modes were an interesting
+technical contribution, but would have liked a higher-level
+description to build intuition.  A description of the usage and
+intuition behind coherence modes is presented in [1] and was omitted
+from this paper due to space constraints.  In this paper we present
+two new results about coherence modes that are not in [1].  First, we
+present a formal semantics of the possible interleavings allowed by
+different coherence modes and their interactions.  Second, we prove
+that coherence modes do not impact the soundness of our type system or
+the correctness of our hierarchical scheduling algorithm.  A brief
+summary of coherence modes is: "exclusive" coherence corresponds to
+apparently-sequential execution, "atomic" corresponds to
+transactional-memory-like execution, and "simultaneous" corresponds to
+"all bets are off" pthreads-style parallelism.  However, unlike other
+implementations, Legion's coherence modes only allow for reordering of
+subtasks with the same parent task, permitting the containment of
+non-determinism to a portion of the task hierarchy.
 
-Reviewer A asked if the operational semantics would be simpler if the only relaxation of sequential
-semantics came from an explicit "par(e1,e2)" construct.  While the semantics would indeed be simpler,
-the use of a static construct (rather than one able to consider dynamic region mappings) would limit
-the expressible forms of parallelism to a (well-studied) subset of what Legion programs can achieve.
+Reviewer A asked if the operational semantics would be simpler if the
+only relaxation of sequential semantics came from an explicit
+"par(e1,e2)" construct.  While the semantics would indeed be simpler,
+the use of a static construct (rather than one able to consider
+dynamic region mappings) would limit the expressible forms of
+parallelism to a (well-studied) subset of what Legion programs can
+achieve.
 
-Reviewer D asked what is currently implemented.  Legion currently consists of a static type 
-checker for programs written in Core Legion and a C++ runtime library for fully-featured Legion programs.
+Reviewer D asked what is currently implemented.  Legion currently
+consists of a static type checker for programs written in Core Legion
+and a C++ runtime library for fully-featured Legion programs.
 
-Finally, reviewers B, C, and D expressed confusion about the nature of the clobber set.
-The choice of large-step operational semantics was made to match the hierarchical and distributed
-scheduling used by the Legion runtime.  When coherence modes other than exclusive are used, there may be
-multiple tasks modifying the same location of memory - the clobber set is the set of all such locations for
-a given task.  The clobber set is used to bound the allowed non-determinism within a given task, and the
-fact that the top-level task (which has no sibling tasks) has an empty clobber set guarantees that any 
-execution of the program is consistent with a standard sequentially consistent memory model.
+Finally, reviewers B, C, and D expressed confusion about the nature of
+the clobber set.  The choice of large-step operational semantics was
+made to match the hierarchical and distributed scheduling used by the
+Legion runtime.  When coherence modes other than exclusive are used,
+multiple tasks may modify the same location of memory - the clobber
+set for a task is the set of all such locations that may be updated by
+other tasks.  The clobber set is used to bound the allowed
+non-determinism within a given task, and the fact that the top-level
+task (which has no sibling tasks) has an empty clobber set guarantees
+that any execution of the program is consistent with a standard
+sequentially consistent memory model.
 
-An advantage of the large-step semantics with the use of clobber sets is that our soundness result holds for
-any scheduling of tasks that satisfies the "valid_interleave" rules.  We considered the use of small-step semantics
-that explicitly track the state of every parallel task's execution, but the complexity of the constraints used
-to decide which task(s) were able to take a step at any given point were onerous and would limit our soundness
-guarantees to only a particular scheduling algorithm.
+An advantage of the large-step semantics with the use of clobber sets
+is that our soundness result holds for any scheduling of tasks that
+satisfies the "valid_interleave" rules.  We considered the use of
+small-step semantics that explicitly track the state of every parallel
+task's execution, but the complexity of the constraints used to decide
+which task(s) were able to take a step at any given point were
+onerous; to manage this complexity we were concerned we would have to
+limit our soundness guarantees to a particular scheduling algorithm.
 
-Reviewer A asked why the E-NewColor rule allows the initial coloring to be non-empty.  There are two
-reasons for this.  First, in the large-step semantics we use, the region mapping and heap typing cannot
-be changed by an E-New rule, so a partitioning of a region must include extra elements of the appropriate
-heap type to support allocation within a subregion.  Second, it allows implementations in which 
-an existing element that was not colored by the programmer is included in (at most one) arbitrary
-subregion, which can improve the efficiency of data transfers at runtime.
+Reviewer A asked why the E-NewColor rule allows the initial coloring
+to be non-empty.  There are two reasons for this.  First, in the
+large-step semantics we use, the region mapping and heap typing cannot
+be changed by an E-New rule, so a partitioning of a region must
+include extra elements of the appropriate heap type to support
+allocation within a subregion.  Second, it allows implementations in
+which an existing element that was not colored by the programmer is
+included in (at most one) arbitrary subregion, which can improve the
+efficiency of data transfers at runtime.

### 2012-09-13 doc/popl2013/response.txt
some small edits to the response

@@
 independence for structuring parallel computations.  The programmer's
-choice of egions can affect load balancing, communication, and the
+choice of regions can affect load balancing, communication, and the
 granularity of parallel tasks.  In the case of the circuit example, no
@@
 made provably efficient, even when scheduling is concurrent and
-distributed, but relies crucially on the soundness of the type system
-for correctness.
+distributed, but relies crucially on the proof of correctness of hierarhical
+scheduling presented in this paper and is only possible with a sound type system.
 
@@
 Furthermore, the way the region information is leveraged both at
-compile-time and at runtime represent a new point in the design space
+compile-time and at runtime represents a new point in the design space
 of region languages.  We would like to point out specifically the way
@@
 
-Finally, reviewers B, C, and D expressed confusion about the nature of
+Reviewers B, C, and D expressed confusion about the nature of
 the clobber set.  The choice of large-step operational semantics was
@@
 
-Reviewer A asked why the E-NewColor rule allows the initial coloring
+Finally, reviewer A asked why the E-NewColor rule allows the initial coloring
 to be non-empty.  There are two reasons for this.  First, in the
@@
 allocation within a subregion.  Second, it allows implementations in
-which an existing element that was not colored by the programmer is
+which an existing element, that was not colored by the programmer, is
 included in (at most one) arbitrary subregion, which can improve the

### 2012-09-13 doc/popl2013/response.txt
More small edits

@@
 correctness issues for the Legion implementation: that there are no
-wild pointers to the wrong regions and that tasks observe both
-privilege and coherence requirements.
+wild pointers to the wrong regions and that tasks observe 
+privilege requirements.
 

### 2012-09-13 doc/popl2013/response.txt
an attempt at a response to "what is a trace?"

@@
 
+Reviewer C requested clarification of the purpose of the memory trace.  Because
+memory trace captures the heap accesses (i.e. side effects) that occur during
+the evaluation of an expression, interactions between expressions being
+evaluated in parallel are completely described by the way in which their
+traces are interleaved.
+
 Reviewer D asked what is currently implemented.  Legion currently

### 2012-09-14 runtime/legion.cc
More hacking on the high level runtime

@@ -34,7 +34,7 @@ namespace RegionRuntime {
 
     //--------------------------------------------------------------------------
     Task::Task(void)
-      : unique_id(0), task_id(0), args(NULL), arglen(0), map_id(0), tag(0),
+      : task_id(0), args(NULL), arglen(0), map_id(0), tag(0),
         orig_proc(Processor::NO_PROC), steal_count(0), must(false),
         is_index_space(false), index_space(IndexSpace::NO_SPACE),
         index_point(NULL), index_element_size(0), 
@@ -515,7 +515,7 @@ namespace RegionRuntime {
 
     //--------------------------------------------------------------------------
     PhysicalRegion::PhysicalRegion(void)
-      : is_impl(false), map_set(false), accessor_map(0) // note this is an invalid configuration
+      : is_impl(false), map_set(false), accessor_map(0), gen_id(0) // note this is an invalid configuration
     //--------------------------------------------------------------------------
     {
       op.map = NULL;
@@ -523,15 +523,15 @@ namespace RegionRuntime {
 
     //--------------------------------------------------------------------------
     PhysicalRegion::PhysicalRegion(PhysicalRegionImpl *i)
-      : is_impl(true), map_set(false), accessor_map(0)
+      : is_impl(true), map_set(false), accessor_map(0), gen_id(0)
     //--------------------------------------------------------------------------
     {
       op.impl = i;
     }
 
     //--------------------------------------------------------------------------
-    PhysicalRegion::PhysicalRegion(MappingOperation *map_op)
-      : is_impl(false), map_set(false), accessor_map(0)
+    PhysicalRegion::PhysicalRegion(MappingOperation *map_op, GenerationID id)
+      : is_impl(false), map_set(false), accessor_map(0), gen_id(id)
     //--------------------------------------------------------------------------
     {
       op.map = map_op;
@@ -545,7 +545,10 @@ namespace RegionRuntime {
       if (is_impl)
         op.impl = rhs.op.impl;
       else
+      {
         op.map = rhs.op.map;
+        gen_id = rhs.gen_id;
+      }
     }
 
     //--------------------------------------------------------------------------
@@ -556,7 +559,10 @@ namespace RegionRuntime {
       if (this->is_impl)
         op.impl = rhs.op.impl;
       else
+      {
         op.map = rhs.op.map;
+        gen_id = rhs.gen_id;
+      }
       this->map_set = rhs.map_set;
[…]

### 2012-09-14 runtime/legion.h

@@ -76,7 +76,6 @@ namespace RegionRuntime {
      */
     class Task {
     public:
-      UniqueID unique_id; // Unique id for the task in the system
       Processor::TaskFuncID task_id; // Id for the task to perform
       std::vector<IndexSpaceRequirement> indexes;
       std::vector<FieldSpaceRequirement> fields;
@@ -100,11 +99,6 @@ namespace RegionRuntime {
       // Get the index point if it is an index point
       template<typename PT, unsigned DIM>
       void get_index_point(PT buffer[DIM]) const;
-    public:
-      bool operator==(const Task &task) const
-        { return unique_id == task.unique_id; }
-      bool operator<(const Task &task) const
-        { return unique_id < task.unique_id; }
     protected:
       // Only the high level runtime should be able to make these
       friend class HighLevelRuntime;
@@ -475,7 +469,7 @@ namespace RegionRuntime {
       friend class SingleTask;
       friend class MappingOperation;
       PhysicalRegion(PhysicalRegionImpl *impl);
-      PhysicalRegion(MappingOperation *op);
+      PhysicalRegion(MappingOperation *op, GenerationID id);
     public:
       PhysicalRegion(void);
       PhysicalRegion(const PhysicalRegion &rhs);
@@ -498,6 +492,7 @@ namespace RegionRuntime {
       bool is_impl;
       bool map_set;
       unsigned accessor_map;
+      GenerationID gen_id; // for checking validity of inline mappings
     };
 
     /////////////////////////////////////////////////////////////
@@ -662,6 +657,7 @@ namespace RegionRuntime {
                                           TaskID uid, const char *name, bool index_space,
                                           Processor::Kind proc_kind, bool leaf);
       static void register_runtime_tasks(Processor::TaskIDTable &table);
+      static TaskVariantCollection* find_collection(Processor::TaskFuncID tid);
     protected:
       static bool is_subtype(TypeHandle parent, TypeHandle child);
     private:
@@ -1378,10 +1374,11 @@ namespace RegionRuntime {
     protected:
       LogicalRegion get_logical_region(void) const;
       PhysicalInstance get_physical_instance(void) const;
+      void invalidate(void);
     protected:
+      bool valid;
       unsigned idx;
       LogicalRegion handle;
-      // Only for non-inline-mapped
       PhysicalInstance instance;
     };
 

### 2012-09-14 runtime/legion_ops.cc

@@ -203,22 +203,36 @@ namespace RegionRuntime {
               parent_ctx->get_unique_id(),unique_id,0,requirement.region.index_space.id,requirement.region.field_space.id,
               requirement.region.tree_id,PRINT_REG(requirement.parent),requirement.privilege,requirement.prop);
 #endif
-      ctx->register_child_map(this);
+      ctx->register_child_map(this, idx);
     }
 
     //--------------------------------------------------------------------------
-    bool MappingOperation::is_valid(void) const
+    bool MappingOperation::is_valid(GenerationID gen_id) const
     //--------------------------------------------------------------------------
     {
+#ifdef DEBUG_HIGH_LEVEL
+      if (gen_id != generation)
+      {
+        log_region(LEVEL_ERROR,"Accessing stale inline mapping operation that has been invalided");
+        exit(ERROR_STALE_INLINE_MAPPING_ACCESS);
+      }
+#endif
       if (mapped_event.has_triggered())
         return ready_event.has_triggered();
       return false;
     }
 
     //--------------------------------------------------------------------------
-    void MappingOperation::wait_until_valid(void)
+    void MappingOperation::wait_until_valid(GenerationID gen_id)
     //--------------------------------------------------------------------------
     {
+#ifdef DEBUG_HIGH_LEVEL
+      if (gen_id != generation)
+      {
+        log_region(LEVEL_ERROR,"Accessing stale inline mapping operation that has been invalided");
+        exit(ERROR_STALE_INLINE_MAPPING_ACCESS);
+      }
+#endif
       // Make sure to wait until we've mapped
       mapped_event.wait();
       // Then wait until we're ready
@@ -226,17 +240,29 @@ namespace RegionRuntime {
     }
 
     //--------------------------------------------------------------------------
-    LogicalRegion MappingOperation::get_logical_region(void) const
+    LogicalRegion MappingOperation::get_logical_region(GenerationID gen_id) const
     //--------------------------------------------------------------------------
     {
+#ifdef DEBUG_HIGH_LEVEL
+      if (gen_id != generation)
+      {
+        log_region(LEVEL_ERROR,"Accessing stale inline mapping operation that has been invalided");
+        exit(ERROR_STALE_INLINE_MAPPING_ACCESS);
+      }
+#endif
       return requirement.region;
     }
 
     //--------------------------------------------------------------------------
-    PhysicalInstance MappingOperation::get_physical_instance(void) const
+    PhysicalInstance MappingOperation::get_physical_instance(GenerationID gen_id) const
[…]

### 2012-09-14 runtime/legion_ops.h

@@ -69,11 +69,11 @@ namespace RegionRuntime {
       void initialize(Context ctx, const RegionRequirement &req, MapperID id, MappingTagID tag);
       void initialize(Context ctx, unsigned idx, MapperID id, MappingTagID tag);
     public:
-      bool is_valid(void) const;
-      void wait_until_valid(void);
-      LogicalRegion get_logical_region(void) const; 
-      PhysicalInstance get_physical_instance(void) const;
-      bool has_accessor(AccessorType at) const;
+      bool is_valid(GenerationID gen_id) const;
+      void wait_until_valid(GenerationID gen_id);
+      LogicalRegion get_logical_region(GenerationID gen_id) const; 
+      PhysicalInstance get_physical_instance(GenerationID gen_id) const;
+      bool has_accessor(GenerationID gen_id, AccessorType at) const;
       PhysicalRegion get_physical_region(void);
     public:
       // Functions from GenerlizedOperation
@@ -162,6 +162,11 @@ namespace RegionRuntime {
      */
     class TaskContext : public Task, public GeneralizedOperation {
     public:
+      TaskContext(HighLevelRuntime *rt, ContextID id);
+    public:
+      bool activate_task(GeneralizedOperation *parent);
+      void deactivate_task(void);
+    public:
       void initialize_task(Context parent, Processor::TaskFuncID tid,
                       void *args, size_t arglen, 
                       const Predicate &predicate,
@@ -177,8 +182,8 @@ namespace RegionRuntime {
                             const std::vector<RegionRequirement> &regions, bool perform_checks);
     public:
       // Functions from GeneralizedOperation
-      virtual bool activate(GeneralizedOperation *parent = NULL);
-      virtual void deactivate(void);
+      virtual bool activate(GeneralizedOperation *parent = NULL) = 0;
+      virtual void deactivate(void) = 0;
       virtual void perform_dependence_analysis(void);
       virtual bool perform_operation(void) = 0;
       virtual void trigger(void) = 0;
@@ -187,11 +192,13 @@ namespace RegionRuntime {
       virtual bool is_locally_mapped(void) = 0;
       virtual bool is_stealable(void) = 0;
       virtual bool is_remote(void) = 0;
+      virtual bool is_partially_unpacked(void) = 0;
     public:
       virtual bool distribute_task(void) = 0; // Return true if still local
       virtual bool perform_mapping(void) = 0; // Return if mapping was successful
       virtual void launch_task(void) = 0;
       virtual void sanitize_region_forest(void) = 0;
[…]

### 2012-09-14 runtime/legion_types.h

@@ -54,6 +54,11 @@ namespace RegionRuntime {
       ERROR_LEAF_TASK_VIOLATION = 29,
       ERROR_INVALID_REDOP_ID = 30,
       ERROR_REDUCTION_INITIAL_VALUE_MISMATCH = 31,
+      ERROR_INVALID_UNMAP_OP = 32,
+      ERROR_INVALID_DUPLICATE_MAPPING = 33,
+      ERROR_INVALID_REGION_ARGUMENT_INDEX = 34,
+      ERROR_INVALID_MAPPING_ACCESS = 35,
+      ERROR_STALE_INLINE_MAPPING_ACCESS = 36,
     };
 
     // enum and namepsaces don't really get along well

### 2012-09-14 doc/popl2013/response.txt

@@
 (named by their location) to integers ("colors"), which represents the
-most general case of describing the partitioning of a region.  More
-concise descriptions exist in the full Legion language (e.g. using a
-pure function to determine the color of an element on demand), but a
-single form was selected for Core Legion for simplicity.  As described
-in [1], Legion does not attempt to solve the (in many cases
-still-open) problem of optimally partitioning data for any given
-application, but instead provides a mechanism for programmers to write
-their own partitioning code and communicate the partitioning decisions
-to the Legion runtime.  In our example, the body of the
-"color_circuit" task would require the programmer to compute a
-partition of the circuit graph and return a coloring that assigns each
-circuit element a color; each color identifies a separate subregion of
-the partition consisting of all the circuit elements with that color.
-In production scientific codes partitioning is often done by a call to
-an external library (e.g. PARMETIS for graph-based data structures).
-A coloring is a way to capture the result of the partitioning
-computation in a general format understood by the Legion runtime.
-More information on colorings can be found in [1].
+general case of describing the partitioning of a region.  More concise
+descriptions exist in the full Legion language (e.g. using a pure
+function to determine the color of an element on demand), but a single
+form was selected for Core Legion for simplicity.  As described in
+[1], Legion does not attempt to solve the (in many cases still-open)
+problem of optimally partitioning data for any given application, but
+instead provides a mechanism for programmers to write their own
+partitioning code and communicate the partitioning decisions to the
+Legion runtime.  In our example, the body of the "color_circuit" task
+would require the programmer to compute a partition of the circuit
+graph and return a coloring that assigns each circuit element a color;
+each color identifies a separate subregion of the partition consisting
+of all the circuit elements with that color.  In production scientific
+codes partitioning is often done by a call to an external library
+(e.g. PARMETIS for graph-based data structures).  A coloring is a way
+to capture the result of the partitioning computation in a general
+format understood by the Legion runtime.  More information on
+colorings can be found in [1].
 
 Reviewer C correctly points out that regions are central to
-performance as well as for the type structure of Legion programs.
-Regions are in Legion to allow programmers to describe locality and
+performance as well as the type structure of Legion programs.  Regions
+are in Legion to allow programmers to describe locality and
 independence for structuring parallel computations.  The programmer's
@@
 correctness issues for the Legion implementation: that there are no
-wild pointers to the wrong regions and that tasks observe 
-privilege requirements.
+wild pointers to the wrong regions and that tasks observe privilege
+requirements.
 
-Reviewer C asked for clarification concerning the the relationship
-between multiple partitions and privileges.  The principle reason to
-support multiple partitions of the same data is to allow precise
-descriptions of the sets of elements that will be accessed by a given
-task.  These sets are often different in different phases of an
-algorithm, and both naming the precise set of nodes and the privileges
-required are needed to safely parallelize applications such as the
-circuit code.  In some cases (like the circuit code) it is possible in
-theory to have a single partitioning of the data by taking the product
-of all the overlapping partitions, but this results in many, many
-unintuitive distinct subregions and is extremely unpleasant
-to program with (we tried it once).
+Reviewer C asked for clarification concerning the relationship between
+multiple partitions and privileges.  The principle reason to support
+multiple partitions of the same data is to allow precise descriptions
+of the sets of elements that will be accessed by a given task.  These
+sets are often different in different phases of an algorithm, and both
+naming the precise set of nodes and the privileges required are needed
+to safely parallelize applications such as the circuit code.  In some
+cases (like the circuit code) it is possible in theory to have a
+single partitioning of the data by taking the product of all the
+overlapping partitions, but this results in many, many unintuitive
+distinct subregions and is extremely unpleasant to program with (we
+tried it once).
 
@@
 made provably efficient, even when scheduling is concurrent and
-distributed, but relies crucially on the proof of correctness of hierarhical
-scheduling presented in this paper and is only possible with a sound type system.
+distributed, but relies crucially on the proof of correctness of
+hierarhical scheduling presented in this paper and is only possible
+with a sound type system.
 
@@
 iterate over an unbounded list L of subregions and access elements of
-those regions so long as there is a constraint that shows the task has
-the same permissions on some ancestor region of every subregion in L.
-And while the proof of soundness does not introduce novel proof
+those subregions so long as there is a constraint that shows the task
+has the same permissions on some ancestor region of every subregion in
+L.  And while the proof of soundness does not introduce novel proof
 techniques, it is far from immediate that the type system is sound and
@@
 
-Reviewer C requested clarification of the purpose of the memory trace.  Because
-memory trace captures the heap accesses (i.e. side effects) that occur during
-the evaluation of an expression, interactions between expressions being
-evaluated in parallel are completely described by the way in which their
-traces are interleaved.
+Reviewer C requested clarification of the purpose of the memory trace.
+Because a memory trace captures the heap accesses (i.e. side effects)
+that occur during the evaluation of an expression, interactions
+between expressions being evaluated in parallel are completely
+described by the way in which their traces are interleaved.
 
@@
 consists of a static type checker for programs written in Core Legion
-and a C++ runtime library for fully-featured Legion programs.
+and a C++ runtime library for fully-featured Legion programs.  The
+runtime library is itself divided into a high-level runtime that manages
+regions and task scheduling and a low-level runtime that provides
+a portability layer across the different types of hardware (multicore chips,
+clusters, GPUs) that we currently support.  More details are in [1].
 
@@
 
-Finally, reviewer A asked why the E-NewColor rule allows the initial coloring
-to be non-empty.  There are two reasons for this.  First, in the
+Finally, reviewer A asked why the E-NewColor rule allows the initial
+coloring to be non-empty.  There are two reasons.  First, in the
 large-step semantics we use, the region mapping and heap typing cannot

### 2012-10-10 runtime/default_mapper.cc
Some more bug fixes for the high level runtime

@@ -417,7 +417,8 @@ namespace RegionRuntime {
     }
 
     //--------------------------------------------------------------------------------------------
-    void Mapper::rank_copy_targets(const Task *task, const RegionRequirement &req,
+    void Mapper::rank_copy_targets(const Task *task, MappingTagID tag, bool inline_mapping,
+                                   const RegionRequirement &req, unsigned index,
                                    const std::set<Memory> &current_instances,
                                    std::set<Memory> &to_reuse,
                                    std::vector<Memory> &to_create,

### 2012-10-10 runtime/legion.h

@@ -52,6 +52,7 @@ namespace RegionRuntime {
       inline bool operator==(const LogicalRegion &rhs) const;
       inline bool operator<(const LogicalRegion &rhs) const;
     public:
+      inline RegionTreeID get_tree_id(void) const { return tree_id; }
       inline IndexSpace get_index_space(void) const;
       inline FieldSpace get_field_space(void) const;
     private:
@@ -1318,7 +1319,7 @@ namespace RegionRuntime {
 
       /**
        * A copy-up operation is occuring to write dirty data back to a parent physical
-       * instance.  To perform the copy-up, the compiler is asking for a target location to
+       * instance.  To perform the copy-up, the runtime is asking for a target location to
        * perform the copy-up operation.  The mapper must give at least one memory to perform
        * the close operation to.  The mapper can chose to re-use an existing instance or
        * to create a new one in one or memories.  In order for the closer operation to
@@ -1328,7 +1329,8 @@ namespace RegionRuntime {
        * last parameter lets the mapper say whether the runtime should only try to create
        * one or all of the instances (default is true).
        */
-      virtual void rank_copy_targets(const Task *task, const RegionRequirement &req,
+      virtual void rank_copy_targets(const Task *task, MappingTagID tag, bool inline_mapping,
+                                    const RegionRequirement &req, unsigned index,
                                     const std::set<Memory> &current_instances,
                                     std::set<Memory> &to_reuse,
                                     std::vector<Memory> &to_create,

### 2012-10-10 runtime/legion_ops.cc

@@ -207,7 +207,7 @@ namespace RegionRuntime {
           (IS_WRITE(req)))
       {
         lock_context();
-        if (forest_ctx->is_disjoint(req.partition))
+        if (!forest_ctx->is_disjoint(req.partition))
         {
           unlock_context();
           return ERROR_NON_DISJOINT_PARTITION;

### 2012-10-10 runtime/region_tree.cc

@@ -409,7 +409,7 @@ namespace RegionRuntime {
         part_color = parent_node->generate_color();
       else
         part_color = unsigned(color);
-      IndexPartNode *new_part = create_node(pid, parent_node, disjoint, part_color, true/*add*/);
+      IndexPartNode *new_part = create_node(pid, parent_node, part_color, disjoint, true/*add*/);
       // Now do all of the child nodes
       for (std::map<Color,IndexSpace>::const_iterator it = coloring.begin();
             it != coloring.end(); it++)
@@ -3922,7 +3922,12 @@ namespace RegionRuntime {
         gstate.added_states.insert(gstate.added_states.end(),
                         new_states.begin(),new_states.end());
       }
+#if 0
+      // Actually this is no longer a valid check since children can be
+      // open in different modes if they are disjoint even if they're
+      // both using the same field.
 #ifdef DEBUG_HIGH_LEVEL
+      if (gstate.field_states.size() > 1)
       {
         // Each field should appear in at most one of these states
         // at any point in time
@@ -3935,11 +3940,13 @@ namespace RegionRuntime {
         }
       }
 #endif
+#endif
     }
 
     //--------------------------------------------------------------------------
     RegionTreeNode::FieldState RegionTreeNode::perform_close_operations(TreeCloser &closer,
-        const GenericUser &user, const FieldMask &closing_mask, FieldState &state, int next_child/*= -1*/)
+        const GenericUser &user, const FieldMask &closing_mask, FieldState &state, 
+        bool allow_same_child, bool &close_successful, int next_child/*= -1*/)
     //--------------------------------------------------------------------------
     {
       std::vector<Color> to_delete;
@@ -3952,8 +3959,10 @@ namespace RegionRuntime {
         // Check field disjointnes
         if (it->second * closing_mask)
           continue;
-        // Check for same child 
-        if ((next_child >= 0) && (next_child == int(it->first)))
+        // Check for same child, only allow upgrades in some cases
+        // such as read-only -> exclusive.  This is calling context
+        // sensitive hence the parameter.
+        if (allow_same_child && (next_child >= 0) && (next_child == int(it->first)))
         {
           FieldMask open_users = it->second & closing_mask;
           result.open_children[it->first] = open_users;
@@ -3968,6 +3977,12 @@ namespace RegionRuntime {
         // Now we need to close this child 
         FieldMask close_mask = it->second & closing_mask;
         RegionTreeNode *child_node = get_tree_child(it->first);
+        // Check to see if the closer is ready to do the close
[…]

### 2012-10-10 runtime/region_tree.h

@@ -444,7 +444,8 @@ namespace RegionRuntime {
       bool siphon_open_children(TreeCloser &closer, GenericState &state, 
             const GenericUser &user, const FieldMask &current_mask, int next_child = -1);
       FieldState perform_close_operations(TreeCloser &closer, const GenericUser &user, 
-                    const FieldMask &closing_mask, FieldState &state, int next_child=-1);
+                    const FieldMask &closing_mask, FieldState &state, 
+                    bool allow_same_child, bool &close_successful, int next_child=-1);
     protected:
       // Logical region helper functions
       FieldMask perform_dependence_checks(const LogicalUser &user, 

### 2012-10-13 doc/types-pldi2013/langdef.tex
checkpoint

--- 2012-07-11 doc/popl2013/langdef.tex
@@
 
+As Legion attempts to benefit from both static and dynamic checks
+to achieve performance while maintaining safety, both the type system and the operational
+semantics have been extended in interesting ways from a basic expression language.  We
+will describe these extensions in the next two subsections, and explain how these extensions
+were chosen to preserve a straightforward proof of the soundness of the type system.
+
 \subsection{Core Type System}
@@
 $$\typeenv{e}{T}$$
-where:
-\begin{itemize}
-\item $\Gamma$ is the type environment, mapping the names of local variables and global functions
-to their types,
-\item $\Phi$ is the region access privileges under which the expression is evaluated, and
-\item $\Omega$ is the set of constraints that must hold between regions
-\end{itemize}
-As regions are hierarchical, privileges and/or constraints on a region often imply similar
-privileges or constraints on subregions.  Figure~\ref{fig:closure} defines closure
-operations for $\Phi^*$ and $\Omega^*$ that explicitly add these implied constraints.
-The type checking rules are in Figure~\ref{fig:types}.  Many of the rules are 
-straightforward, but several deserve more explanation.
-
-Core Legion has no subtyping. A pointer's region is changed only via explicit
-{\em upregion} and {\em downregion} operations.  {\em Upregion} calls can be inferred,
-but {\em downregion} must be accompanied by a run-time check that the pointer really does
-point into the subregion(s), so {\em upregion} is made explicit as well for simplicity.
-
-Accesses to heap memory are performed via the {\em read}, {\em write}, and {\em reduce}
-operations, which are only legal if the regions in a pointer's type are ones for which the
-enclosing function has privileges.
-Similarly, a call to another function is permitted only if the privileges required by the 
-callee are a subset of those held by the caller.  There is no way to create privileges within
-a Legion function - it is limited to privileges given to it by its caller.  (The
-top-level task of a Core Legion program begins with a single, empty region from which other
-regions can be partitioned and full read and
-write privileges.)
-
-The {\em partition} operation uses a {\em coloring} to split a region into 
-disjoint subregions.  The subregions are given names within the body of the partition,
-as well as constraints asserting subregions are pairwise disjoint and contained within the parent region.
-%These constraints can be used locally, or packed into region relation instances that can be unpacked and used later.
-
-The {\em pack} expression packs together a value and the corresponding logical regions,
-provided the region relation's constraints hold.  The
-corresponding {\em unpack} expression gives (new) names to those regions.  For simplicity, in Core Legion neither the 
-{\em partition} nor the {\em unpack} expression allows shadowing of region names.
+
+In addition to the traditional type environment $\Gamma$, a Legion type judgment includes the
+set of access privileges $\Phi$ for the logical regions in the expression as well as a set of
+constraints $\Omega$ that must hold between those logical regions.
+
+Both $\Phi$ and $\Omega$ are used in the heap access expressions: {\em read}, {\em write}, and {\em reduce}.
+For a heap access to be valid, the the appropriate permission must exist for logical region(s) in the
+pointer's type.  As regions are hierarchical, the exact region need not be named in $\Phi$ if permissions
+exist for a logical region that is known to contain the pointer's region.  To simplify this check, we
+define closure operations in Figure~\ref{fig:closure} that expand $\Omega$ into an $\Omega^*$, which is
+then used to expand $\Phi$ into a $\Phi^*$ that does explicitly name every logical region for which
+privileges are known to exist.
+
+Core Legion provides no automatic pointer type conversions.  Valid pointers are created via the {\em new}
+expression in specified region and invalid (i.e. {\em null}) pointers must also be typed.  Pointers can be
+``upcast'' via the {\em upregion} expression, which uses the expanded constraints $\Omega^*$ to verify that
+every possible region a pointer might point into is covered by at least one of the regions in the desired
+pointer type.  In contrast, a ``downcast'' using the {\em downregion} expression must perform a dynamic
+test to verify which region a run-time pointer value points into.  A static test similar to the one used
+for {\em upregion} could be used to catch cases in which the downcast can never succeed at runtime, but we
+have left it out for simplicity.
+
+The introduction of constraints into $\Omega$ occurs through the use of the {\em partition} expression,
+which makes use of a new type specific to Legion, a {\em coloring}.  A coloring is an opaque mapping from
+pointers (which much point into the region being colored) to integer ``color'' values.  (For simplicity, 
+Core Legion allows only an iterative construction of a coloring via a {\em newcolor} expression that generates
+an empty map and a {\em color} expression which adds a new entry to an existing map, but the opacity of the
+coloring allows for more efficient generation (and storage) of colorings as needed.)  The {\em partition} 
+expression introduces names for subregions corresponding to each of those ``colors'', guaranteeing that each
+subregion is disjoint from the other subregions and all are included in the original region.
+
+The {\em pack} and {\em unpack} expressions allow the storage of pointers in the heap.  Logical region names
+are lexically scoped and if stored in the heap as-is, can only be read within the same scope.  A region relationship
+is used instead to capture the (dynamic) physical regions while preserving the statically known constraints
+between those regions.  Because the {\em pack} expression is able to verify these constraints hold when the 
+region relationship value is created, the {\em unpack} expression is able to reintroduce these constraints on
+the (fresh) logical region names given to the contents of an unwrapped region relationship value.
+
+Finally, the task call expression handles the aliasing of logical region names between the caller and callee and
+verifies the property that the privileges held by a called task are always a subset of those held
+by the caller.  In addition to being necessary for the proof of soundness, this property is critical to
+enable the hierarchical scheduling model used by the Legion runtime.
 

### 2012-10-15 runtime/region_tree.h
Fixing some more packing and unpacking of state bugs

@@ -108,13 +108,16 @@ namespace RegionRuntime {
       void unpack_region_tree_updates_return(Deserializer &derez);
     public:
       // Packing and unpacking state return
-      size_t compute_region_tree_state_return(const RegionRequirement &req, ContextID ctx, bool all, SendingMode mode);
+      size_t compute_region_tree_state_return(const RegionRequirement &req, unsigned idx, 
+                                              ContextID ctx, bool overwrite, SendingMode mode);
       size_t post_compute_region_tree_state_return(void);
       void begin_pack_region_tree_state_return(Serializer &rez);
-      void pack_region_tree_state_return(const RegionRequirement &req, ContextID ctx, bool all, SendingMode mode, Serializer &rez);
+      void pack_region_tree_state_return(const RegionRequirement &req, unsigned idx, 
+                              ContextID ctx, bool overwrite, SendingMode mode, Serializer &rez);
       void end_pack_region_tree_state_return(Serializer &rez);
       void begin_unpack_region_tree_state_return(Deserializer &derez);
-      void unpack_region_tree_state_return(const RegionRequirement &req, ContextID ctx, bool all, SendingMode mode, Deserializer &derez);
+      void unpack_region_tree_state_return(const RegionRequirement &req, ContextID ctx, 
+                                            bool overwrite, SendingMode mode, Deserializer &derez);
       void end_unpack_region_tree_state_return(Deserializer &derez);
     public:
       size_t compute_created_state_return(ContextID ctx);
@@ -207,8 +210,8 @@ namespace RegionRuntime {
       std::map<InstanceView*,FieldMask> unique_views; // points to the top instance view
       std::vector<InstanceView*>        ordered_views;
       std::vector<bool>                 overwrite_views; // for knowing when to overwrite views when returning 
-      std::vector<RegionNode*>          diff_regions;
-      std::vector<PartitionNode*>       diff_partitions;
+      std::map<unsigned,std::vector<RegionNode*> >  diff_region_maps;
+      std::map<unsigned,std::vector<PartitionNode*> > diff_part_maps;
       std::vector<InstanceManager*>     returning_managers;
       std::vector<InstanceView*>        returning_views;
       std::map<EscapedUser,unsigned>    escaped_users;
@@ -908,7 +911,7 @@ namespace RegionRuntime {
       std::map<UniqueID,FieldMask> epoch_users;
       std::map<Event,FieldMask> epoch_copy_users;
       std::map<Event,FieldMask> valid_events;
[…]

### 2012-10-15 doc/types-pldi2013/evaluation.tex

--- 2012-07-11 doc/popl2013/evaluation.tex
@@
 
-We evaluate the Legion type system on three criteria:
+We evaluate the Legion type system on four criteria:
 \begin{itemize}
@@
 \item Scalability: Can it leverage hierarchical scheduling (Section~\ref{subsec:scalability})?
+\item Performance: Does the performance increase from relaxed coherence modes (Section~\ref{subsec:relaxedperf})
+warrant the increased complexity in the operational semantics?
 \end{itemize}
@@
 runtime is able to achieve a 62.5X speedup on 96 GPUs over a hand-coded single GPU implementation\cite{Legion12}.
+
+\subsection{Performance}
+\label{subsec:relaxedperf}
+
+To demonstrate the benefit of relaxed coherence modes, we modified the circuit example from Section~\ref{sec:example}
+to use exclusive coherence in the {\em distribute\_charge} task and compared the performance of the two versions.
+The results are shown in Figure~\ref{fig:circuit_excl}.  Even though the {\em distribute\_charge} tasks
+are a small fraction of the total work, the serialization that results from requiring exclusive access to the 
+overlapping ghost regions limits the scalability of the application, as would be expected from Amdahl's Law.

### 2012-10-16 runtime/legion.cc
Lots of support for legion spy debugging in the high level runtime

@@ -1,6 +1,7 @@
 
 #include "legion.h"
 #include "legion_utilities.h"
+#include "legion_logging.h"
 #include "legion_ops.h"
 #include "region_tree.h"
 
@@ -20,10 +21,14 @@ namespace RegionRuntime {
     Logger::Category log_field("field_spaces");
     Logger::Category log_region("regions");
     Logger::Category log_inst("instances");
-    Logger::Category log_spy("legion_spy");
     Logger::Category log_garbage("gc");
     Logger::Category log_leak("leaks");
     Logger::Category log_variant("variants");
+#ifdef LEGION_SPY
+    namespace LegionSpy {
+      Logger::Category log_spy("legion_spy");
+    };
+#endif
 
     const LogicalRegion LogicalRegion::NO_REGION = LogicalRegion();
     const LogicalPartition LogicalPartition::NO_PART = LogicalPartition();
@@ -2853,9 +2858,9 @@ namespace RegionRuntime {
         top->initialize_task(NULL/*no parent*/, HighLevelRuntime::legion_main_id,
                               &HighLevelRuntime::get_input_args(), sizeof(InputArgs),
                               Predicate::TRUE_PRED, 0/*map id*/, 0/*mapping tag*/, get_mapper(0), get_mapper_lock(0));
-#ifndef LOG_EVENT_ONLY
-        log_spy(LEVEL_INFO,"Top Task %d %d",top->get_unique_id(),HighLevelRuntime::legion_main_id);
-#endif 
+#ifdef LEGION_SPY
+        LegionSpy::log_top_level_task(top->get_unique_id(), top->ctx_id, HighLevelRuntime::legion_main_id);
+#endif
         // Pack up the future and a pointer to the context so we can deactivate the task
         // context when we're done.  This will make sure that everything gets
         // cleanedup and will help capture any leaks.
@@ -3476,8 +3481,8 @@ namespace RegionRuntime {
       log_index(LEVEL_DEBUG, "Creating index space %x in task %s (ID %d) with %ld maximum elements", space.id,
                               ctx->variants->name,ctx->get_unique_id(), max_num_elmts);
 #endif
-#ifndef LOG_EVENT_ONLY
-      log_spy(LEVEL_INFO,"Index Space %x",space.id);
+#ifdef LEGION_SPY
+      LegionSpy::log_top_index_space(space.id);
 #endif
       ctx->create_index_space(space);
       return space;
[…]

### 2012-10-16 runtime/legion_logging.h

@@ -1,0 +1,97 @@
+
+#ifndef __LEGION_LOGGING_H__
+#define __LEGION_LOGGING_H__
+
+#include "utilities.h"
+
+namespace RegionRuntime {
+  namespace HighLevel {
+    namespace LegionSpy {
+
+      extern Logger::Category log_spy;
+
+      // Logger calls for the shape of region trees
+      static inline void log_top_index_space(unsigned unique_id)
+      {
+        log_spy(LEVEL_INFO,"Index Space %d", unique_id);
+      }
+
+      static inline void log_index_partition(unsigned parent_id, unsigned unique_id, bool disjoint, unsigned color)
+      {
+        log_spy(LEVEL_INFO,"Index Partition %d %d %d %d", parent_id, unique_id, disjoint, color);
+      }
+
+      static inline void log_index_subspace(unsigned parent_id, unsigned unique_id, unsigned color)
+      {
+        log_spy(LEVEL_INFO,"Index Subspace %d %d %d", parent_id, unique_id, color);
+      }
+
+      static inline void log_field_space(unsigned unique_id)
+      {
+        log_spy(LEVEL_INFO,"Field Space %d", unique_id);
+      }
+
+      static inline void log_field_creation(unsigned unique_id, unsigned field_id)
+      {
+        log_spy(LEVEL_INFO,"Field Creation %d %d", unique_id, field_id);
+      }
+
+      static inline void log_top_region(unsigned index_space, unsigned field_space, unsigned tree_id)
+      {
+        log_spy(LEVEL_INFO,"Region %d %d %d", index_space, field_space, tree_id);
+      }
+
+      // Logger calls for operations 
+      static inline void log_top_level_task(unsigned unique_id, unsigned context, unsigned top_id)
+      {
+        log_spy(LEVEL_INFO,"Top Task %d %d", unique_id, top_id);
+      }
+
+      static inline void log_task_operation(unsigned unique_id, unsigned task_id, unsigned parent_id, unsigned parent_ctx)
+      {
+        log_spy(LEVEL_INFO,"Task Operation %d %d %d %d", unique_id, task_id, parent_id, parent_ctx);
+      }
+
+      static inline void log_mapping_operation(unsigned unique_id, unsigned parent_id, unsigned parent_ctx)
+      {
+        log_spy(LEVEL_INFO,"Mapping Operation %d %d %d", unique_id, parent_id, parent_ctx);
+      }
+
[…]

### 2012-10-16 runtime/legion_ops.cc

@@ -1,6 +1,7 @@
 
 #include "legion_ops.h"
 #include "region_tree.h"
+#include "legion_logging.h"
 #include <algorithm>
 
 #define PRINT_REG(reg) (reg).index_space.id,(reg).field_space.id, (reg).tree_id
@@ -300,11 +301,11 @@ namespace RegionRuntime {
       tag = t;
       // Check privileges for the region requirement
       check_privilege();
-#ifndef LOG_EVENT_ONLY
-      log_spy(LEVEL_INFO,"Map %d Parent %d",unique_id,parent_ctx->get_unique_id());
-      log_spy(LEVEL_INFO,"Context %d Task %d Region %d Handle (%x,%x,%x) Parent (%x,%x,%x) Privilege %d Coherence %d",
-              parent_ctx->get_unique_id(),unique_id,0,req.region.index_space.id,req.region.field_space.id,req.region.tree_id,
-              PRINT_REG(req.parent),req.privilege,req.prop);
+#ifdef LEGION_SPY
+      LegionSpy::log_mapping_operation(unique_id, parent_ctx->get_unique_id(), parent_ctx->ctx_id);
+      LegionSpy::log_logical_requirement(unique_id,0,true,req.region.index_space.id, req.region.field_space.id,
+                                        req.region.tree_id, req.privilege, req.prop, req.redop);
+      LegionSpy::log_requirement_fields(unique_id, 0, req.privilege_fields);
 #endif
       parent_ctx->register_child_map(this);
     }
@@ -327,11 +328,11 @@ namespace RegionRuntime {
       tag = t;
       // Check privileges for the region requirement
       check_privilege();
-#ifndef LOG_EVENT_ONLY
-      log_spy(LEVEL_INFO,"Map %d Parent %d",unique_id,parent_ctx->get_unique_id());
-      log_spy(LEVEL_INFO,"Context %d Task %d Region %d Handle (%x,%x,%x) Parent (%x,%x,%x) Privilege %d Coherence %d",
-              parent_ctx->get_unique_id(),unique_id,0,requirement.region.index_space.id,requirement.region.field_space.id,
-              requirement.region.tree_id,PRINT_REG(requirement.parent),requirement.privilege,requirement.prop);
+#ifdef LEGION_SPY
+      LegionSpy::log_mapping_operation(unique_id, parent_ctx->get_unique_id(), parent_ctx->ctx_id);
+      LegionSpy::log_logical_requirement(unique_id,0,true,requirement.region.index_space.id, requirement.region.field_space.id,
[…]

### 2012-10-16 runtime/legion_ops.h

@@ -353,7 +353,7 @@ namespace RegionRuntime {
       // Operations on index space trees
       void create_index_space(IndexSpace space);
       void destroy_index_space(IndexSpace space); 
-      void create_index_partition(IndexPartition pid, IndexSpace parent, bool disjoint, int color,
+      Color create_index_partition(IndexPartition pid, IndexSpace parent, bool disjoint, int color,
                                   const std::map<Color,IndexSpace> &coloring); 
       void destroy_index_partition(IndexPartition pid);
       IndexPartition get_index_partition(IndexSpace parent, Color color);

### 2012-10-16 runtime/region_tree.cc

@@ -396,7 +396,7 @@ namespace RegionRuntime {
     }
 
     //--------------------------------------------------------------------------
-    void RegionTreeForest::create_index_partition(IndexPartition pid, IndexSpace parent, bool disjoint,
+    Color RegionTreeForest::create_index_partition(IndexPartition pid, IndexSpace parent, bool disjoint,
                                 int color, const std::map<Color,IndexSpace> &coloring)
     //--------------------------------------------------------------------------
     {
@@ -416,6 +416,7 @@ namespace RegionRuntime {
       {
         create_node(it->second, new_part, it->first, true/*add*/);
       }
+      return part_color;
     }
 
     //--------------------------------------------------------------------------
@@ -3770,11 +3771,11 @@ namespace RegionRuntime {
               it++; // Not empty so keep going
           }
         }
-        // Add ourselves to the current epoch
-        state.curr_epoch_users.push_back(user);
         // Close up any partitions which we might have dependences on below
         LogicalCloser closer(user, az.ctx, state.prev_epoch_users, are_closing_partition());
         siphon_open_children(closer, state, user, user.field_mask);
+        // Add ourselves to the current epoch
+        state.curr_epoch_users.push_back(user);
       }
       else
       {

### 2012-10-16 runtime/region_tree.h

@@ -36,7 +36,7 @@ namespace RegionRuntime {
       // Index Space operations
       void create_index_space(IndexSpace space);
       void destroy_index_space(IndexSpace space);
-      void create_index_partition(IndexPartition pid, IndexSpace parent, bool disjoint, int color,
+      Color create_index_partition(IndexPartition pid, IndexSpace parent, bool disjoint, int color,
                                   const std::map<Color,IndexSpace> &coloring); 
       void destroy_index_partition(IndexPartition pid);
       IndexPartition get_index_partition(IndexSpace parent, Color color);

### 2012-10-16 tools/legion_spy.py
Reworking the legion spy infrastructure for debugging

@@ -1,15 +1,58 @@
 #!/usr/bin/python
 
 import subprocess
-import pygame, sys, os, shutil
+#import pygame, sys, os, shutil
+import sys, os, shutil
 import string
-from pygame.locals import *
+#from pygame.locals import *
 from getopt import getopt
+from spy_parser import parse_log_file
+from spy_state import *
 
-from parser import parse_log_file 
+#from parser import parse_log_file 
 
 temp_dir = ".spy/"
 
+
+def usage():
+    print "Usage: "+ sys.argv[0] +" [-c (check dependences)] file_name"
+    sys.exit(1)
+
+def main():
+    if len(sys.argv) < 2:
+        usage()
+
+    opts, args = getopt(sys.argv[1:],'l')
+    opts = dict(opts)
+    if len(args) <> 1:
+        usage()
+
+    logical_checks = False
+    if '-l' in opts:
+        logical_checks = True 
+
+    file_name = args[0]
+
+    tree_state = TreeState()
+    ops_state  = OpState()
+
+    print 'Loading log file '+file_name+'...'
+    total_matches = parse_log_file(file_name, tree_state, ops_state)
+    print 'Matched '+str(total_matches)+' lines'
+    if total_matches == 0:
+        print "No matches. Exiting..."
+        return
+
+    if logical_checks:
+        print "Checking all mapping dependences..."
+        ops_state.check_logical(tree_state)
+
+    print "Legion Spy analysis complete.  Exiting..."
+
+
+# Everything below here is the old version of legion spy except
+# for the if-statement at the end that calls main
+'''
 class ImageWrapper(object):
     def __init__(self,name,file_name,fontObj=None,blackColor=None,scale=10):
         self.name = name
@@ -274,11 +317,11 @@ def main():
 
         pygame.display.update()
         fpsClock.tick(30)
-
+'''
 
 if __name__ == "__main__":
     try:
-        assert pygame.image.get_extended()
+        #assert pygame.image.get_extended()
         os.mkdir(temp_dir)
         main()
         shutil.rmtree(temp_dir)

### 2012-10-16 tools/spy_parser.py

@@ -1,0 +1,99 @@
+#!/usr/bin/python
+
+from spy_state import *
+import sys, re
+
+# All these calls are based on the print statements in legion_logging.h
+
+prefix = "\[[0-9]+ - [0-9a-f]+\] \{\w+\}\{legion_spy\}: "
+# Calls for the shape of region trees
+top_index_pat           = re.compile(prefix+"Index Space (?P<uid>[0-9]+)")
+index_part_pat          = re.compile(prefix+"Index Partition (?P<pid>[0-9]+) (?P<uid>[0-9]+) (?P<disjoint>[0-1]) (?P<color>[0-9]+)")
+index_subspace_pat      = re.compile(prefix+"Index Subspace (?P<pid>[0-9]+) (?P<uid>[0-9]+) (?P<color>[0-9]+)")
+field_space_pat         = re.compile(prefix+"Field Space (?P<uid>[0-9]+)")
+field_create_pat        = re.compile(prefix+"Field Creation (?P<uid>[0-9]+) (?P<fid>[0-9]+)")
+region_pat              = re.compile(prefix+"Region (?P<iid>[0-9]+) (?P<fid>[0-9]+) (?P<tid>[0-9]+)")
+
+# Logger calls for operations
+top_task_pat            = re.compile(prefix+"Top Task (?P<uid>[0-9]+) (?P<tid>[0-9]+)")
+task_pat                = re.compile(prefix+"Task Operation (?P<uid>[0-9]+) (?P<tid>[0-9]+) (?P<pid>[0-9]+) (?P<ctx>[0-9]+)")
+mapping_pat             = re.compile(prefix+"Mapping Operation (?P<uid>[0-9]+) (?P<pid>[0-9]+) (?P<ctx>[0-9]+)")
+deletion_pat            = re.compile(prefix+"Deletion Operation (?P<uid>[0-9]+) (?P<pid>[0-9]+) (?P<ctx>[0-9]+)")
+task_name_pat           = re.compile(prefix+"Task Name (?P<uid>[0-9]+) (?P<name>\w+)")
+
+# Logger calls for logical mapping dependence analysis
[…]

### 2012-10-16 tools/spy_state.py

@@ -1,0 +1,501 @@
+#!/usr/bin/python
+
+import subprocess
+import string
+
+# These are imported from legion_types.h
+NO_DEPENDENCE = 0
+TRUE_DEPENDENCE = 1
+ANTI_DEPENDENCE = 2
+ATOMIC_DEPENDENCE = 3
+SIMULTANEOUS_DEPENDENCE = 4
+
+NO_ACCESS  = 0x00000000
+READ_ONLY  = 0x00000001
+READ_WRITE = 0x00000111
+WRITE_ONLY = 0x00000010
+REDUCE     = 0x00000100
+
+EXCLUSIVE = 0
+ATOMICE = 1
+SIMULTANEOUS = 2
+RELAXED = 3
+
+
+class Requirement(object):
+    def __init__(self, index, is_reg, ispace, fspace, tid, priv, coher, redop):
+        self.index = index
+        self.is_reg = is_reg
+        self.ispace = ispace
+        self.fspace = fspace
+        self.tid = tid
+        self.priv = priv
+        self.coher = coher
+        self.redop = redop
+        self.fields = set()
+
+    def add_field(self, fid):
+        assert fid not in self.fields
+        self.fields.add(fid)
+
+    def is_read_only(self):
+        return (self.priv == NO_ACCESS) or (self.priv == READ_ONLY)
+
+    def has_write(self):
+        return (self.priv == READ_WRITE) or (self.priv == REDUCE) or (self.priv == WRITE_ONLY)
+
+    def is_write(self):
+        return (self.priv == READ_WRITE) or (self.priv == WRITE_ONLY)
+
+    def is_write_only(self):
+        return self.priv == WRITE_ONLY
+
+    def is_reduce(self):
+        return self.priv == REDUCE
+
+    def is_exclusive(self):
+        return self.coher == EXCLUSIVE
+
+    def is_atomic(self):
+        return self.coher == ATOMIC
+
+    def is_simult(self):
+        return self.coher == SIMULTANEOUS
+
+    def is_relaxed(self):
+        return self.coher == RELAXED
+
+
+def is_mapping_dependence(dtype):
+    if dtype == NO_DEPENDENCE:
+        return False
+    # All other types of dependences are mapping dependences
+    return True
+
+def check_for_anti_dependence(req1, req2, actual):
+    if req1.is_read_only():
+        assert req1.has_write()
+        return ANTI_DEPENDENCE
+    else:
+        if req2.is_write_only():
+            return ANTI_DEPENDENCE
[…]

### 2012-10-16 doc/types-pldi2013/langdef.tex

@@
 \begin{itemize}
-\item $\mapconsist{\Omega}$, a region mapping satisfies the region constraints under which the expression was typed
-\item $\localconsist{L}{\Gamma}$, local values have types consistent with the environment under which the expression was typed
-\item $\storeconsist{S}{H}$, the initial state of the store contains values whose types are consistent with the heap typing
+\item mapping consistency, written $\mapconsist{\Omega}$, guarantees a region mapping $M$ satisfies the region constraints $\Omega$ under
+which an expression was typed
+\item local value consistency, written $\localconsist{L}{\Gamma}$, guarantees local values in $L$ have types consistent with the environment under which the expression was typed
+\item store consistency, written $\storeconsist{S}{H}$, guarantees every location in the store $S$ contains a value consistent with the location's type in the heap typing $H$
 \end{itemize}
@@
 \begin{itemize}
-\item $\valueconsist{v}{T}$, the expression yields a value of the right type
-\item $\traceconsist{E}$, values written to the store match the heap types of their locations
+\item result value consistency, written $\valueconsist{v}{T}$, guarantees any evaluation of an expression yields a value of the right type
+\item memory trace consistency, written $\traceconsist{E}$, guarantees that all writes and reductions use values of the right types
 \end{itemize}
@@
 
+The proof, which can be found in \cite{LegionTypes12}, is lengthy but
+straightforward due to the way in which the operational semantics were
+constructed.  It proceeds by standard induction on the structure of the
+derivation in the operational semantics, at each step showing that
+subexpressions have consistent starting environments and that the
+value and trace consistency of the subexpressions is sufficient to prove
+the consistency of the parent expression.
+
+\texcomment{
 We spare the reader the lengthy proof, which can be found in
@@
 Straightforward proofs cover {\tt read}, {\tt write}, {\tt reduce}, and one final special case for function calls. 
-
-%% \begin{table*}
-%% \centering
-%% {\small
-%% \begin{tabular}{l|p{5cm}|p{5cm}|p{5cm}}
-%% {\bf Rule} & 
-%% \multicolumn{1}{|c|}{$\mathbf{v \sim_H M \llbracket T \rrbracket}$} &
-%% \multicolumn{1}{|c|}{$\mathbf{E \sim H}$} &
-%% \multicolumn{1}{|c|}{$\mathbf{E : \Phi}$} \\
-%% \hline
-%% BoolConst & & & \\
-%% IntConst & & & \\
-%% TupleBuild & & & \\
-%% TupleField & & & \\
-%% LocalVar & & & \\
-%% NullPtr & & & \\
-%% NewPtr & & & \\
-%% IsNull & & & \\
-%% UpRegion & & & \\
-%% DownRegion & & & \\
-%% Read & & & \\
-%% Write & & & \\
-%% Reduce & & & \\
-%% Newcolor & & & \\
-%% Color & & & \\
-%% IntMath & & & \\
-%% IntCmp & & & \\
-%% Let & & & \\
-%% IfElse & & & \\
-%% TaskCall & & & \\
-%% Partition & & & \\
-%% Pack & & & \\
-%% Unpack & & &
-%% \end{tabular}}
-%% \caption{Proof pieces}
-%% \end{table*}
+}
 

### 2012-10-16 apps/fdtd/fdtd.cc
Checkpoint FDTD partitioning code.

@@ -109,6 +109,10 @@ static inline unsigned block_id(unsigned bx, unsigned by, unsigned bz,
   return (bx*nby + by)*nbz + bz;
 }
 
+static inline unsigned border_block_id(unsigned nbx, unsigned nby, unsigned nbz) {
+  return nbx*nby*nbz;
+}
+
 static inline unsigned cell_id(unsigned x, unsigned y, unsigned z,
                                unsigned nx, unsigned ny, unsigned nz) {
   return (x*(ny + 2) + y)*(nz + 2) + z;
@@ -133,9 +137,12 @@ public:
     unsigned next_index = 0;
     unsigned nbx = x_divs.size(), nby = y_divs.size(), nbz = z_divs.size();
 
+    for (unsigned id = 0; id < nbx*nby*nbz + 1; id++) {
+      coloring[id] = ColoredPoints<unsigned>();
+    }
+
     // Color points for plane of points at x == 0 boundary.
-    unsigned border = nbx*nby*nbz;
-    coloring[border] = ColoredPoints<unsigned>();
+    unsigned border = border_block_id(nbx, nby, nbz);
     unsigned x_plane_size = (ny + 2)*(nz + 2);
     log_app.debug("Assigning points %d..%d to x == 0 plane",
                   next_index, next_index + x_plane_size);
@@ -168,7 +175,6 @@ public:
         for (unsigned bz = 0; bz < nbz; bz++) {
           unsigned id = block_id(bx, by, bz, nbx, nby, nbz);
           unsigned block_size = z_divs[bz].second - z_divs[bz].first;
-          coloring[id] = ColoredPoints<unsigned>();
           log_app.debug("Assigning points %d..%d to block %d x %d x %d (id %d)",
                         next_index, next_index + block_size, bx, by, bz, id);
           coloring[id].ranges.insert(
@@ -208,17 +214,19 @@ private:
   const std::vector<std::pair<unsigned, unsigned> > x_divs, y_divs, z_divs;
 };
 
-enum dim_t {
+enum dimension_t {
   DIM_X,
   DIM_Y,
   DIM_Z,
 };
+const unsigned NUM_DIMENSIONS = 3; // Must equal number of entries in above enum.
 
 enum ghost_t {
   COLOR_NEGATIVE,
   COLOR_POSITIVE,
   COLOR_UNSHARED,
 };
+const unsigned NUM_GHOST_COLORS = 3; // Must equal number of entries in above enum.
 
 // Colors each block into three pieces, one of which is not shared any
[…]

### 2012-10-16 doc/asplos2013/response.txt

We would like to thank the reviewers for the detailed feedback.  Below we will...

Reviewers 1, 2, and 4 requested clarification of which components of the the API are
novel and which are similar to prior work.  Although events, locks, barriers, and reductions
have appeared in many forms, we believe the following contributions to be novel:
  a) the use of generational events to efficiently describe the very large number of dependencies
that exist in a distributed application
  b) the deferred lock construct in which access to a critical resource can be mediated without
blocking operations that can waste execution resources
  c) the ability to modify a barrier's expected arrival count which allows nested task subdivision
without requiring a parent task to explicitly wait for subtasks to finish
  d) a general treatment of reductions that allows arbitrary reduction operations and improved
efficiency through specialized reduction instances that allow individual reduction operations to be
deferred and applied in bulk
  e) the ability to asychronously compose all of these constructs

Reviewers 1 and 5 had several questions about our handling of reductions.  One source of confusion
was the use of "commutativity".  The traditional use of these terms is limited to binary operations
in which both the left and right operands are of the same type.  Our implementation allows reductions
that have different left and right operand types (e.g. scaling of a vector by a scalar) and treats
each reduction operation as a unary function applied (atomically) to the current value (e.g. V *= 3
becomes (*3)(V) ).  When multiple reductions are applied to the same value, these unary functions
are essentially composed, and it's the commutativity of that composition
(e.g. [(*3) o (*2)](X) = [(*2) o (*3)](X) for all X) that allows the deferred application of these
reduction operations.  Nearly all "reasonable" reduction operations
that one might propose satisfy this form of commutativity.  Most also satisfy our criterion of foldability
(e.g. [(*3) o (*2)](X) = (*6)(X) for all X), allowing the use of our reduction fold instances to further
improve performance in most cases.

Reviewer 1 noted that use "Distributed Architectures" in our title might be confusing.  Our goal was
to distinguish our implementation from those that are limited to shared-memory systems.  A more
accurate description would be "Distributed Memory Architectures", and we'd be happy to change the 
title should our submission be accepted.

### 2012-10-17 doc/asplos2013/response.txt
More work on the asplos response

@@
 to distinguish our implementation from those that are limited to shared-memory systems.  A more
-accurate description would be "Distributed Memory Architectures", and we'd be happy to change the 
-title should our submission be accepted.
+accurate description would be "Distributed Memory Architectures."  We would be willing to change our
+title to clarify our meaning if the submission is accepted.
+
+Reviewer 2 noted that OpenCL supports events for asynchronously launching tasks and moving data. 
+As we noted in our related work, OpenCL events are only valid in a single address space (e.g. node) unlike our 
+events which are valid in every address space on distributed memory machines.  Furthermore, the
+synchronization primitives in OpenCL do not use events and cannot compose with other operations.
+
+Reviewer 3 expressed doubts about the performance of the Legion version of the applications 
+relative to hand-coded applications and whether the applications fit well into the bulk-synchronous model.  
+The fourth citation in our paper [4] (which has been accepted and will appear at Supercomputing 2012) 
+reports speedups over hand-coded versions of the applications.  In this paper, we aimed to quantify 
+exactly what fraction of the speedup from [4] results from the composable, asynchronous nature of our 
+runtime system.  To eliminate any performance benefits achieved by the Legion programming model 
+independent from our runtime, we compared against bulk-synchronous version of the applications from [4] 
+written in Legion instead of hand-coded implementations.  The Fluid and AMR applications had hand-coded 
+reference implementations written by others.  In both cases, the architects of those programs chose to 
+express their computation in a bulk-synchronous model indicating that the programs are a good fit
+for bulk-synchronous execution.
+
+Reviewer 4 requested a comparison to existing dataflow programming models.  A more apt comparison
+for dataflow models would be the Legion programming system which is beyond the scope of this paper.  
+Instead, we view the runtime presented in this paper as orthogonal work to dataflow programming models.   
+The runtime presented here provides a lower-level layer of abstraction that could serve as the 
+foundation for many different dataflow programming systems.

### 2012-10-17 doc/asplos2013/response.txt
More additions to the asplos response

@@
 novel and which are similar to prior work.  Although events, locks, barriers, and reductions
-have appeared in many forms, we believe the following contributions to be novel:
+have appeared in many forms, we believe the following contributions are novel:
   a) the use of generational events to efficiently describe the very large number of dependencies
@@
 becomes (*3)(V) ).  When multiple reductions are applied to the same value, these unary functions
-are essentially composed, and it's the commutativity of that composition
+can be composed, and it's the commutativity of that composition
 (e.g. [(*3) o (*2)](X) = [(*2) o (*3)](X) for all X) that allows the deferred application of these
@@
 to distinguish our implementation from those that are limited to shared-memory systems.  A more
-accurate description would be "Distributed Memory Architectures."  We would be willing to change our
+accurate description would be "Distributed Memory Architectures."  We would be willing to change the 
 title to clarify our meaning if the submission is accepted.
@@
 foundation for many different dataflow programming systems.
+
+Reviewer 4 described deferred locks as having the same semantics as I-structures from the dataflow
+literature.  This is incorrect.  I-structures support a producer-consumer semantics which is isomorphic
+to the functionality provided by events in our runtime (we do not claim the semantics of events to
+be novel).  Unlike the producer-consumer semantics supported by I-structures, deferred locks support
+unordered atomic semantics, which allows programmers to describe properties like serializability
+without the requirement of ordering.  We are unaware of any feature of dataflow implementations
+that support a similar semantics to deferred locks.
+
+Reviewer 1 commented that blocking APIs do not necessarily stall processors if there are enough 
+runnable threads to keep the processor busy.  While this is true, the overhead of having many threads
+to support this kind of computation is high due to the cost of software threads (except on GPUs where 
+threads are lightweight).  Instead our system provides lighter-weight tasks which the programmer
+can launch at a much finer granularity with incurring the overhead of a thread.  Due to the minimal
+overhead of tasks, programmers can express a much finer granularity of parallelism which the runtime
+can then leverage if there are available hardware resources.  Existing programming systems like MPI
+and P-threads entail too much overhead to leverage this granularity of task-parallelism.
+
+Reviewer 1 and reviewer 5 asked for clarification regarding barriers and the need for a dynamic change
+in arrival count.  In our experiencing writing programs to our runtime system, we've found that
+the number of expected arrivals at a barrier is not known at barrier creation-time (intuitively this
+is necessary to support nested parallelism in Legion).  To support this paradigm, we added the feature of
+being able to dynamically alter the arrival count.  To implement this feature on the distributed 
+machines we're targetting, we use the well-studied vector clock algorithm from the distributed systems
+literature (see citation [16] for additional details on the semantics and implementation of vector clocks).
+
+Reviewer 1 and reviwer 5 both inquired about how the runtime works with GPUs.  GPUs are just another kind
+of processor in our system and GPU memories (global, zero-copy, etc) are just another kind of memory with
+regards to the runtime.  Copies of data between GPU and CPU memories are no different than copies between
+any other pair of memories and must be explicitly made by the programmer.  Tasks launched on a GPU processor
+invoked a kernel on the GPU but otherwise have the same semantics as CPU task.  One restriction is 
+runtime calls are not currently available inside of tasks running on the GPU.  We left this aspect of 
+the implementation for future work and owing to this asymmetry opted not claim heterogeneity as one
+of our primary contributions.

### 2012-10-17 doc/asplos2013/response.txt
edits

@@
 can be composed, and it's the commutativity of that composition
-(e.g. [(*3) o (*2)](X) = [(*2) o (*3)](X) for all X) that allows the deferred application of these
-reduction operations.  Nearly all "reasonable" reduction operations
+(e.g. [(*3) o (*2)](X) = [(*2) o (*3)](X) for all X) that allows the application of these
+reduction operations to be deferred and reordered.  Nearly all "reasonable" reduction operations
 that one might propose satisfy this form of commutativity.  Most also satisfy our criterion of foldability
@@
 Reviewer 2 noted that OpenCL supports events for asynchronously launching tasks and moving data. 
-As we noted in our related work, OpenCL events are only valid in a single address space (e.g. node) unlike our 
-events which are valid in every address space on distributed memory machines.  Furthermore, the
+As we noted in our related work, OpenCL events are only valid in a single address space (e.g. node),
+whereas our events which are valid in every address space on distributed memory machines.  Furthermore, the
 synchronization primitives in OpenCL do not use events and cannot compose with other operations.
@@
 relative to hand-coded applications and whether the applications fit well into the bulk-synchronous model.  
-The fourth citation in our paper [4] (which has been accepted and will appear at Supercomputing 2012) 
-reports speedups over hand-coded versions of the applications.  In this paper, we aimed to quantify 
-exactly what fraction of the speedup from [4] results from the composable, asynchronous nature of our 
-runtime system.  To eliminate any performance benefits achieved by the Legion programming model 
+The fourth citation in our paper [4] (which will appear at Supercomputing 2012) compares these same 
+applications written in Legion to hand-coded reference implementations, and shows the Legion versions outperforming
+the existing implementations on larger machine sizes.  All three reference implementations, two of which
+were written by third parties, were coded in a bulk-synchronous style.  In this paper, we aimed to quantify 
+what portion of that speedup results from the composable, asynchronous nature of our 
+runtime system.  To factor out any performance benefits gained by the Legion programming model 
 independent from our runtime, we compared against bulk-synchronous version of the applications from [4] 
-written in Legion instead of hand-coded implementations.  The Fluid and AMR applications had hand-coded 
-reference implementations written by others.  In both cases, the architects of those programs chose to 
-express their computation in a bulk-synchronous model indicating that the programs are a good fit
-for bulk-synchronous execution.
+written in Legion instead of the reference implementations.
 
@@
 
-Reviewer 4 described deferred locks as having the same semantics as I-structures from the dataflow
-literature.  This is incorrect.  I-structures support a producer-consumer semantics which is isomorphic
-to the functionality provided by events in our runtime (we do not claim the semantics of events to
-be novel).  Unlike the producer-consumer semantics supported by I-structures, deferred locks support
-unordered atomic semantics, which allows programmers to describe properties like serializability
-without the requirement of ordering.  We are unaware of any feature of dataflow implementations
-that support a similar semantics to deferred locks.
+Reviewer 4 was curious about the difference between our deferred locks and I-structures from the dataflow
+literature.  I-structures support a producer-consumer, write-once semantics which is much closer
+to the functionality provided by events in our runtime.  In contrast, deferred locks provide
+unordered atomic semantics, permitting multiple writes to an object and allowing programmers to describe
+properties like serializability without the requirement of ordering.  We are unaware of any feature of
+dataflow implementations that support a similar semantics to deferred locks.
 
@@
 runnable threads to keep the processor busy.  While this is true, the overhead of having many threads
-to support this kind of computation is high due to the cost of software threads (except on GPUs where 
-threads are lightweight).  Instead our system provides lighter-weight tasks which the programmer
+to support this kind of computation is high due to the cost of software threads on CPUs.  (On GPUs, threads
+are much lighter-weight, but current programming models do not allow these threads to be suspended.)
+Instead our system provides lighter-weight tasks which the programmer
 can launch at a much finer granularity with incurring the overhead of a thread.  Due to the minimal
@@
 
-Reviewer 1 and reviwer 5 both inquired about how the runtime works with GPUs.  GPUs are just another kind
-of processor in our system and GPU memories (global, zero-copy, etc) are just another kind of memory with
-regards to the runtime.  Copies of data between GPU and CPU memories are no different than copies between
-any other pair of memories and must be explicitly made by the programmer.  Tasks launched on a GPU processor
-invoked a kernel on the GPU but otherwise have the same semantics as CPU task.  One restriction is 
-runtime calls are not currently available inside of tasks running on the GPU.  We left this aspect of 
-the implementation for future work and owing to this asymmetry opted not claim heterogeneity as one
-of our primary contributions.
+Reviewer 1 and reviwer 5 both inquired about how the runtime works with GPUs.  Our runtime uses CUDA to 
+control the GPUs in the cluster.  A GPU is exposed as another kind of processor in our system, and launches
+asynchronous CUDA grids for each task.  GPU memories (framebuffer and zero-copy) are exposed in the same way
+as CPU main memory and the globally accessible "GASNet" memory.  Copies of data between GPU and other
+memories are explicitly requested by the programmer (like all other copies), and are performed with
+asynchronous DMA requests to the CUDA driver.  The CUDA driver provides device events, which are exposed
+in our runtime as general events that allow GPU operations to be composeable as well.

### 2012-10-17 doc/asplos2013/response.txt
More edits to the response

@@
-We would like to thank the reviewers for the detailed feedback.  Below we will...
+We would like to thank the reviewers for the detailed feedback.  Below we address
+the comments and questions most frequently raised by the reviewers.
 
@@
 reduction operations to be deferred and reordered.  Nearly all "reasonable" reduction operations
-that one might propose satisfy this form of commutativity.  Most also satisfy our criterion of foldability
+that one might propose satisfy this form of commutativity.  Most reductions also satisfy our criterion of "foldability"
 (e.g. [(*3) o (*2)](X) = (*6)(X) for all X), allowing the use of our reduction fold instances to further
@@
 accurate description would be "Distributed Memory Architectures."  We would be willing to change the 
-title to clarify our meaning if the submission is accepted.
+title to clarify our meaning.
 
-Reviewer 2 noted that OpenCL supports events for asynchronously launching tasks and moving data. 
-As we noted in our related work, OpenCL events are only valid in a single address space (e.g. node),
-whereas our events which are valid in every address space on distributed memory machines.  Furthermore, the
-synchronization primitives in OpenCL do not use events and cannot compose with other operations.
+Reviewer 2 noted that APIs like OpenCL supports events for asynchronously launching tasks and moving data. 
+As we noted in our related work, OpenCL events (and similar constructs in other APIs) are only valid in a 
+single address space (e.g. node), whereas our events which are valid in every address space on 
+distributed memory machines.  Furthermore, the synchronization primitives in APIs like OpenCL do not use 
+events and cannot compose with other operations the way the synchronization mechanisms in our runtime can.
 
-Reviewer 3 expressed doubts about the performance of the Legion version of the applications 
+Reviewer 3 expressed doubts about the performance of the Legion versions of the applications 
 relative to hand-coded applications and whether the applications fit well into the bulk-synchronous model.  
@@
 Instead our system provides lighter-weight tasks which the programmer
-can launch at a much finer granularity with incurring the overhead of a thread.  Due to the minimal
-overhead of tasks, programmers can express a much finer granularity of parallelism which the runtime
-can then leverage if there are available hardware resources.  Existing programming systems like MPI
-and P-threads entail too much overhead to leverage this granularity of task-parallelism.
+can launch at a much finer granularity without incurring the overhead of a thread.  Due to the minimal
+overhead of tasks, programmers can express a much finer granularity of parallelism which our runtime
+can then efficiently map onto available hardware resources.  Existing programming systems like MPI, OpenMP,
+and P-threads entail too much overhead to leverage this finer-grained task parallelism.
 
@@
 Reviewer 1 and reviwer 5 both inquired about how the runtime works with GPUs.  Our runtime uses CUDA to 
-control the GPUs in the cluster.  A GPU is exposed as another kind of processor in our system, and launches
-asynchronous CUDA grids for each task.  GPU memories (framebuffer and zero-copy) are exposed in the same way
-as CPU main memory and the globally accessible "GASNet" memory.  Copies of data between GPU and other
-memories are explicitly requested by the programmer (like all other copies), and are performed with
-asynchronous DMA requests to the CUDA driver.  The CUDA driver provides device events, which are exposed
-in our runtime as general events that allow GPU operations to be composeable as well.
+control the GPUs in the cluster.  A GPU is exposed as another kind of processor in our system, and each task
+launched on a GPU processor results in an asyncrhonous CUDA grid launch.  GPU memories (framebuffer and zero-copy) 
+are exposed in the same way as CPU main memory and the globally accessible "GASNet" memory.  Copies of 
+data between GPU and other memories are explicitly requested by the programmer (like all other copies), 
+and are performed with asynchronous DMA requests to the CUDA driver.  The CUDA driver provides device events, 
+which are exposed in our runtime as general events that allow GPU operations to be composeable as well.

### 2012-10-17 doc/asplos2013/response.txt
minor edits

@@
 distributed memory machines.  Furthermore, the synchronization primitives in APIs like OpenCL do not use 
-events and cannot compose with other operations the way the synchronization mechanisms in our runtime can.
+events and consequently cannot compose with other operations the way the synchronization mechanisms in our runtime can.
 
@@
 
-Reviewer 4 was curious about the difference between our deferred locks and I-structures from the dataflow
+Reviewer 4 was unsure about the difference between our deferred locks and I-structures from the dataflow
 literature.  I-structures support a producer-consumer, write-once semantics which is much closer
@@
 properties like serializability without the requirement of ordering.  We are unaware of any feature of
-dataflow implementations that support a similar semantics to deferred locks.
+dataflow language implementations that support a similar semantics to deferred locks.
 
@@
 in arrival count.  In our experiencing writing programs to our runtime system, we've found that
-the number of expected arrivals at a barrier is not known at barrier creation-time (intuitively this
+the number of expected arrivals at a barrier is often not known at barrier creation-time (intuitively this
 is necessary to support nested parallelism in Legion).  To support this paradigm, we added the feature of

### 2012-10-17 doc/asplos2013/response.txt
more edits

@@
   a) the use of generational events to efficiently describe the very large number of dependencies
-that exist in a distributed application
+that exist in a distributed memory application
   b) the deferred lock construct in which access to a critical resource can be mediated without
@@
 Reviewers 1 and 5 had several questions about our handling of reductions.  One source of confusion
-was the use of "commutativity".  The traditional use of these terms is limited to binary operations
+was use of the term "commutativity".  One traditional use of this term is limited to binary operations
 in which both the left and right operands are of the same type.  Our implementation allows reductions
@@
 to distinguish our implementation from those that are limited to shared-memory systems.  A more
-accurate description would be "Distributed Memory Architectures."  We would be willing to change the 
+accurate description would be "Distributed Memory Architectures."  We would be happy to change the 
 title to clarify our meaning.
@@
 runtime system.  To factor out any performance benefits gained by the Legion programming model 
-independent from our runtime, we compared against bulk-synchronous version of the applications from [4] 
+independent from our runtime, we compared against bulk-synchronous versions of the applications from [4] 
 written in Legion instead of the reference implementations.
+
+Reviewer 3 commented that the experimental sections of the paper failed to evaluate the composability claim
+of the runtime API.  It is unclear to us what kind of experiments would be useful to quantify the composability
+of the API, but we would be open to suggests from the reviewers.  Our argument for compsability is a qualitative
+one, centered around the ability of events to express dependences between all operations supported by the
+runtime: computation, data movement, and synchronization.  As far as we are aware, no other API currently used
+for parallel programming shares this property.
 
@@
 for dataflow models would be the Legion programming system which is beyond the scope of this paper.  
-Instead, we view the runtime presented in this paper as orthogonal work to dataflow programming models.   
+Instead, we view the runtime presented in this paper as orthogonal to work on dataflow programming models.   
 The runtime presented here provides a lower-level layer of abstraction that could serve as the 
@@
 Reviewer 4 was unsure about the difference between our deferred locks and I-structures from the dataflow
-literature.  I-structures support a producer-consumer, write-once semantics which is much closer
-to the functionality provided by events in our runtime.  In contrast, deferred locks provide
+literature.  I-structures support a producer-consumer, write-once semantics that has some of the
+properties of events in our runtime.  In contrast, deferred locks provide
 unordered atomic semantics, permitting multiple writes to an object and allowing programmers to describe
@@
 are much lighter-weight, but current programming models do not allow these threads to be suspended.)
-Instead our system provides lighter-weight tasks which the programmer
-can launch at a much finer granularity without incurring the overhead of a thread.  Due to the minimal
-overhead of tasks, programmers can express a much finer granularity of parallelism which our runtime
-can then efficiently map onto available hardware resources.  Existing programming systems like MPI, OpenMP,
-and P-threads entail too much overhead to leverage this finer-grained task parallelism.
+This cost is magnified by a programmer's need to conservatively create new threads whenever an API
+call _might_ block or risk performance loss from the serialization of independent subtasks that are
+launched from the same parent task.
 
-Reviewer 1 and reviewer 5 asked for clarification regarding barriers and the need for a dynamic change
-in arrival count.  In our experiencing writing programs to our runtime system, we've found that
+Reviewers 1 and 5 asked for clarification regarding barriers and the need for a dynamic change
+in arrival count.  In our experience, we've found that
 the number of expected arrivals at a barrier is often not known at barrier creation-time (intuitively this
-is necessary to support nested parallelism in Legion).  To support this paradigm, we added the feature of
-being able to dynamically alter the arrival count.  To implement this feature on the distributed 
+is necessary to support nested parallelism in Legion).  To support this paradigm, we added the ability to 
+dynamically alter the arrival count.  To implement this feature on the distributed 
 machines we're targetting, we use the well-studied vector clock algorithm from the distributed systems
@@
 
-Reviewer 1 and reviwer 5 both inquired about how the runtime works with GPUs.  Our runtime uses CUDA to 
+Reviewers 1 and 5 both inquired about how the runtime works with GPUs.  Our runtime uses CUDA to 
 control the GPUs in the cluster.  A GPU is exposed as another kind of processor in our system, and each task

### 2012-10-17 doc/asplos2013/response.txt
more edits

@@
 are much lighter-weight, but current programming models do not allow these threads to be suspended.)
-This cost is magnified by a programmer's need to conservatively create new threads whenever an API
-call _might_ block or risk performance loss from the serialization of independent subtasks that are
-launched from the same parent task.
+Instead our system provides lighter-weight tasks which the programmer
+can launch at a much finer granularity without incurring the overhead of a thread.  Due to the minimal
+overhead of tasks, programmers can express a much finer granularity of parallelism which our runtime
+can then efficiently map onto available hardware resources.  Existing programming systems like MPI, OpenMP,
+and P-threads entail too much overhead to leverage this finer-grained task parallelism.
 
-Reviewers 1 and 5 asked for clarification regarding barriers and the need for a dynamic change
+Reviewer 1 and reviewer 5 asked for clarification regarding barriers and the need for a dynamic change
 in arrival count.  In our experience, we've found that
@@
 
-Reviewers 1 and 5 both inquired about how the runtime works with GPUs.  Our runtime uses CUDA to 
+Reviewer 1 and reviwer 5 both inquired about how the runtime works with GPUs.  Our runtime uses CUDA to 
 control the GPUs in the cluster.  A GPU is exposed as another kind of processor in our system, and each task

### 2012-10-17 doc/asplos2013/response.txt
sorting and editing

--- 2012-10-17 doc/asplos2013/response.txt
@@
 deferred and applied in bulk
-  e) the ability to asychronously compose all of these constructs
+  e) the ability to asynchronously compose all of these constructs
+
+Reviewer 3 expressed doubts about the performance of the Legion versions of the applications 
+relative to hand-coded applications and whether the applications fit well into the bulk-synchronous model.  
+The fourth citation in our paper [4] (which will appear at Supercomputing 2012) compares these same 
+applications written in Legion to hand-coded reference implementations, and shows the Legion versions outperforming
+the existing implementations on larger machine sizes.  All three reference implementations, two of which
+were written by third parties, were coded in a bulk-synchronous style.  In this paper, we aimed to quantify 
+what portion of that speedup results from the composable, asynchronous nature of our 
+runtime system.  To factor out any performance benefits gained by the Legion programming model 
+independent from our runtime, we compared against bulk-synchronous versions of the applications from [4] 
+written in Legion instead of the reference implementations.
+
+Reviewer 4 was unsure about the difference between our deferred locks and I-structures from the dataflow
+literature.  I-structures support a producer-consumer, write-once semantics that has some of the
+properties of events in our runtime.  In contrast, deferred locks provide
+unordered atomic semantics, permitting multiple writes to an object and allowing programmers to describe
+properties like serializability without the requirement of ordering.  We are unaware of any feature of
+dataflow language implementations that support a similar semantics to deferred locks.
+
+Reviewer 3 commented that the experimental sections of the paper failed to evaluate the composability claim
+of the runtime API.  It is unclear to us what kind of experiments would be useful to quantify the composability
+of the API, but we would be open to suggests from the reviewers.  Our argument for composability is a qualitative
+one, centered around the ability of events to express dependences between all operations supported by the
+runtime: computation, data movement, and synchronization.  As far as we are aware, no other API currently used
+for parallel programming shares this property.
+
+Reviewer 2 noted that APIs like OpenCL supports events for asynchronously launching tasks and moving data. 
+As we noted in our related work, OpenCL events (and similar constructs in other APIs) are only valid in a 
+single address space (e.g. node), whereas our events are valid in every address space on 
+distributed memory machines.  Furthermore, the synchronization primitives in APIs like OpenCL do not use 
+events and consequently cannot compose with other operations the way the synchronization mechanisms in our runtime can.
+
+Reviewers 1 and 5 asked for clarification regarding barriers and the need for a dynamic change
+in arrival count.  In programming models that support nested parallelism,
+the number of expected arrivals at a barrier is often not known at barrier creation-time.
+To allow tasks to be dynamically subdivided, we added the ability to 
+alter the arrival count after the barrier has been created.  We use the well-studied vector 
+clock algorithm from the distributed systems literature to
+ensure that every node has observed all barrier alterations before completing the barrier, 
+eliminating race conditions between alterations to and arrivals at the barrier.
+(See citation [16] for additional details on the semantics and implementation of vector clocks.)
 
@@
 
-Reviewer 1 noted that use "Distributed Architectures" in our title might be confusing.  Our goal was
-to distinguish our implementation from those that are limited to shared-memory systems.  A more
-accurate description would be "Distributed Memory Architectures."  We would be happy to change the 
-title to clarify our meaning.
-
-Reviewer 2 noted that APIs like OpenCL supports events for asynchronously launching tasks and moving data. 
-As we noted in our related work, OpenCL events (and similar constructs in other APIs) are only valid in a 
-single address space (e.g. node), whereas our events which are valid in every address space on 
-distributed memory machines.  Furthermore, the synchronization primitives in APIs like OpenCL do not use 
-events and consequently cannot compose with other operations the way the synchronization mechanisms in our runtime can.
-
-Reviewer 3 expressed doubts about the performance of the Legion versions of the applications 
-relative to hand-coded applications and whether the applications fit well into the bulk-synchronous model.  
-The fourth citation in our paper [4] (which will appear at Supercomputing 2012) compares these same 
-applications written in Legion to hand-coded reference implementations, and shows the Legion versions outperforming
-the existing implementations on larger machine sizes.  All three reference implementations, two of which
-were written by third parties, were coded in a bulk-synchronous style.  In this paper, we aimed to quantify 
-what portion of that speedup results from the composable, asynchronous nature of our 
-runtime system.  To factor out any performance benefits gained by the Legion programming model 
-independent from our runtime, we compared against bulk-synchronous versions of the applications from [4] 
-written in Legion instead of the reference implementations.
-
-Reviewer 3 commented that the experimental sections of the paper failed to evaluate the composability claim
-of the runtime API.  It is unclear to us what kind of experiments would be useful to quantify the composability
-of the API, but we would be open to suggests from the reviewers.  Our argument for compsability is a qualitative
-one, centered around the ability of events to express dependences between all operations supported by the
-runtime: computation, data movement, and synchronization.  As far as we are aware, no other API currently used
-for parallel programming shares this property.
+Reviewer 1 commented that blocking APIs do not necessarily stall processors if there are enough 
+runnable threads to keep the processor busy.  While this is true, the overhead of having many threads
+to support this kind of computation is high due to the cost of software threads on CPUs.  (On GPUs, threads
+are much lighter-weight, but current programming models do not allow these threads to be suspended.)
+This cost is magnified by a programmer's need to conservatively create new threads whenever an API
+call might block.  Failing to do this can result in performance loss from the serialization of 
+independent subtasks that are launched from the same parent task.
 
@@
 
-Reviewer 4 was unsure about the difference between our deferred locks and I-structures from the dataflow
-literature.  I-structures support a producer-consumer, write-once semantics that has some of the
-properties of events in our runtime.  In contrast, deferred locks provide
-unordered atomic semantics, permitting multiple writes to an object and allowing programmers to describe
-properties like serializability without the requirement of ordering.  We are unaware of any feature of
-dataflow language implementations that support a similar semantics to deferred locks.
-
-Reviewer 1 commented that blocking APIs do not necessarily stall processors if there are enough 
-runnable threads to keep the processor busy.  While this is true, the overhead of having many threads
-to support this kind of computation is high due to the cost of software threads on CPUs.  (On GPUs, threads
-are much lighter-weight, but current programming models do not allow these threads to be suspended.)
-This cost is magnified by a programmer's need to conservatively create new threads whenever an API
-call _might_ block or risk performance loss from the serialization of independent subtasks that are
-launched from the same parent task.
-
-Reviewers 1 and 5 asked for clarification regarding barriers and the need for a dynamic change
-in arrival count.  In our experience, we've found that
-the number of expected arrivals at a barrier is often not known at barrier creation-time (intuitively this
-is necessary to support nested parallelism in Legion).  To support this paradigm, we added the ability to 
-dynamically alter the arrival count.  To implement this feature on the distributed 
-machines we're targetting, we use the well-studied vector clock algorithm from the distributed systems
-literature (see citation [16] for additional details on the semantics and implementation of vector clocks).
+Reviewer 1 noted that use "Distributed Architectures" in our title might be confusing.  Our goal was
+to distinguish our implementation from those that are limited to shared-memory systems.  A more
+accurate description would be "Distributed Memory Architectures."  We would be happy to change the 
+title to clarify our meaning.
 
@@
 control the GPUs in the cluster.  A GPU is exposed as another kind of processor in our system, and each task
-launched on a GPU processor results in an asyncrhonous CUDA grid launch.  GPU memories (framebuffer and zero-copy) 
+launched on a GPU processor results in an asynchronous CUDA grid launch.  GPU memories (framebuffer and zero-copy) 
 are exposed in the same way as CPU main memory and the globally accessible "GASNet" memory.  Copies of 
@@
 and are performed with asynchronous DMA requests to the CUDA driver.  The CUDA driver provides device events, 
-which are exposed in our runtime as general events that allow GPU operations to be composeable as well.
+which are exposed in our runtime as general events that allow GPU operations to be composable as well.

### 2012-10-17 doc/asplos2013/response.txt
added an alternative attempt at the composabilty response

@@
 
+@@@ Version 1
 Reviewer 3 commented that the experimental sections of the paper failed to evaluate the composability claim
@@
 for parallel programming shares this property.
+
+@@@ Version 2
+Reviewer 3 had concerns related to the composability of the constructs and the performance benefits of that
+composability.  They key to the composability of our constructs is the event object.  Every operation in
+our API (task launch, barrier arrival, lock request/release, data copy) accepts an event as a 'wait_for'
+parameter, allowing that action to be deferred until some the provided event has occurred.  Every operation
+provides, either directly or indirectly, an event that indicates when that operation has been completed.  By
+linking operations together with events, a task can create an arbitrary dependency graph of operations to be
+executed by the runtime, all asynchronously to the execution of the original task.  As clarified above, our results
+section demonstrates that this decoupling of the dependency graph from the original requesting thread provides
+significant speedup for applications written in Legion, especially for larger cluster sizes where the number of 
+tasks and the latency of communication grow.
 

### 2012-10-18 doc/asplos2013/response.txt
more edits with sean

--- 2012-10-17 doc/asplos2013/response.txt
@@
 
-Reviewer 3 commented that the experimental sections of the paper failed to evaluate the composability claim
-of the runtime API.  It is unclear to us what kind of experiments would be useful to quantify the composability
-of the API, but we would be open to suggests from the reviewers.  Our argument for composability is a qualitative
-one, centered around the ability of events to express dependences between all operations supported by the
-runtime: computation, data movement, and synchronization.  As far as we are aware, no other API currently used
-for parallel programming shares this property.
+Reviewer 3 had concerns related to the composability of the constructs and the performance benefits of that
+composability.  They key to the composability of our constructs is the event object.  Unlike other 
+parallel programming APIs, every operation in
+our API (task launch, barrier arrival, lock/unlock, data copy) accepts an event as a 'wait_for'
+parameter, causing that action to be deferred until the specified event has occurred.  With the exception of
+'unlock' which completes immediately, every operation provides an event that indicates when that operation has been completed.  By
+linking operations together with events, an arbitrary dependency graph of operations can be created and executed by the runtime
+with no further interaction with the requesting task.  Without the composability of operations provided by events this would 
+not be possible.  As our experimental results
+demonstrate, this decoupling of the dependency graph from the original requesting task provides
+significant speedup for applications written in Legion, especially when scaling to larger cluster sizes 
+with increased communication latency.
 

### 2012-10-18 doc/asplos2013/response.txt
(hopefully) last set of edits from me

@@
 Reviewer 3 expressed doubts about the performance of the Legion versions of the applications 
-relative to hand-coded applications and whether the applications fit well into the bulk-synchronous model.  
+relative to hand-coded applications and whether bulk-synchronous implementations were appropriate 
+for those applications.
 The fourth citation in our paper [4] (which will appear at Supercomputing 2012) compares these same 
@@
 composability.  They key to the composability of our constructs is the event object.  Unlike other 
-parallel programming APIs, every operation in
-our API (task launch, barrier arrival, lock/unlock, data copy) accepts an event as a 'wait_for'
-parameter, causing that action to be deferred until the specified event has occurred.  With the exception of
-'unlock' which completes immediately, every operation provides an event that indicates when that operation has been completed.  By
-linking operations together with events, an arbitrary dependency graph of operations can be created and executed by the runtime
-with no further interaction with the requesting task.  Without the composability of operations provided by events this would 
-not be possible.  As our experimental results
-demonstrate, this decoupling of the dependency graph from the original requesting task provides
-significant speedup for applications written in Legion, especially when scaling to larger cluster sizes 
-with increased communication latency.
+parallel programming APIs, every operation in our API (task launch, barrier arrival, lock/unlock, data copy)
+accepts an event as a 'wait_for' parameter, causing that action to be deferred until the specified event has
+occurred.  With the exception of 'unlock' which completes immediately, every operation also provides an event
+that indicates when that operation has been completed.  By linking operations together with events, an arbitrary
+dependency graph of operations can be created and then executed by the runtime with no further interaction with
+the requesting task.  Without the composability of operations provided by events this would not be possible.
+As our experimental results demonstrate, this decoupling of the dependency graph from the original requesting
+task provides significant speedup for applications written in Legion, especially when scaling to larger cluster
+sizes with increased communication latency.
 
@@
 distributed memory machines.  Furthermore, the synchronization primitives in APIs like OpenCL do not use 
-events and consequently cannot compose with other operations the way the synchronization mechanisms in our runtime can.
+events and consequently cannot compose with other operations the way the synchronization mechanisms in
+our runtime can.
 
-Reviewers 1 and 5 asked for clarification regarding barriers and the need for a dynamic change
-in arrival count.  In programming models that support nested parallelism,
-the number of expected arrivals at a barrier is often not known at barrier creation-time.
-To allow tasks to be dynamically subdivided, we added the ability to 
-alter the arrival count after the barrier has been created.  We use the well-studied vector 
-clock algorithm from the distributed systems literature to
-ensure that every node has observed all barrier alterations before completing the barrier, 
-eliminating race conditions between alterations to and arrivals at the barrier.
+Reviewers 1 and 5 asked for clarification regarding barriers and the need for dynamic changes
+to a barrier's arrival count.  In programming models that support nested parallelism,
+the number of expected arrivals at a barrier is often not known at the time the barrier is created.
+To allow tasks to be dynamically subdivided, we added the ability to alter the arrival count after
+the barrier has been created.  We use the well-studied vector clock algorithm from the distributed
+systems literature to ensure that every node has observed all barrier alterations before completing
+the barrier, eliminating race conditions between alterations to and arrivals at the barrier.
 (See citation [16] for additional details on the semantics and implementation of vector clocks.)
@@
 reduction operations to be deferred and reordered.  Nearly all "reasonable" reduction operations
-that one might propose satisfy this form of commutativity.  Most reductions also satisfy our criterion of "foldability"
-(e.g. [(*3) o (*2)](X) = (*6)(X) for all X), allowing the use of our reduction fold instances to further
-improve performance in most cases.
+that one might propose satisfy this form of commutativity.  Most reductions also satisfy our criterion of
+"foldability" (e.g. [(*3) o (*2)](X) = (*6)(X) for all X), allowing the use of our reduction fold
+instances to further improve performance in most cases.
 

### 2012-10-18 doc/asplos2013/response.txt
added Alex's edits

@@
 without requiring a parent task to explicitly wait for subtasks to finish
-  d) a general treatment of reductions that allows arbitrary reduction operations and improved
+  d) a general treatment of reductions that permits arbitrary reduction operations and improved
 efficiency through specialized reduction instances that allow individual reduction operations to be
@@
 
-Reviewer 2 noted that APIs like OpenCL supports events for asynchronously launching tasks and moving data. 
+Reviewer 2 noted that APIs like OpenCL support events for asynchronously launching tasks and moving data. 
 As we noted in our related work, OpenCL events (and similar constructs in other APIs) are only valid in a 
@@
 
-Reviewer 4 requested a comparison to existing dataflow programming models.  A more apt comparison
-for dataflow models would be the Legion programming system which is beyond the scope of this paper.  
+Reviewer 4 requested a comparison to existing dataflow programming models.  A more apt comparison for
+dataflow models would be the higher-level Legion programming system which is beyond the scope of this paper.  
 Instead, we view the runtime presented in this paper as orthogonal to work on dataflow programming models. 
-The runtime presented here provides a lower-level layer of abstraction that could serve as the 
-foundation for many different dataflow programming systems.
+The runtime presented here provides a lower-level layer of abstraction that can serve as the basis for
+implementing systems with deferred execution, potentially including dataflow systems.
 

### 2012-10-18 doc/asplos2013/response.txt
last few minor edits to the response

@@
 unordered atomic semantics, permitting multiple writes to an object and allowing programmers to describe
-properties like serializability without the requirement of ordering.  We are unaware of any feature of
+properties like serializability without the requirement of ordering.  We are unaware of any feature from 
 dataflow language implementations that support a similar semantics to deferred locks.
@@
 the number of expected arrivals at a barrier is often not known at the time the barrier is created.
-To allow tasks to be dynamically subdivided, we added the ability to alter the arrival count after
-the barrier has been created.  We use the well-studied vector clock algorithm from the distributed
+To support this paradigm for higher-level programming models that target our runtime, 
+we added the ability to alter the arrival count after
+a barrier has been created.  We use the well-studied vector clock algorithm from the distributed
 systems literature to ensure that every node has observed all barrier alterations before completing
@@
 call might block.  Failing to do this can result in performance loss from the serialization of 
-independent subtasks that are launched from the same parent task.
+independent operations that are launched from the same parent task.
 

### 2012-10-23 doc/types-pldi2013/langdef.tex
operational semantics rewrite

@@
 $$\opsenv{e}{v}{E}$$
-where: 
-\begin{itemize}
-\item $M$ is a mapping from logical regions $r_i$ to {\em physical regions} $\rho_i$, which are sets of concrete memory locations.
-We write $\hat x = M \llbracket x \rrbracket$ for the application of $M$ to other objects (e.g., types), 
-replacing all free logical region names $r_i$ in $x$ with corresponding physical regions $\rho_i$.  
-(Note $M \llbracket x \rrbracket$ is ill-formed if $x$ contains a free logical region name $r_j \not\in dom(M)$.)
-
-\item $L$ is a mapping of local variables to (immutable) values.
-
-\item $H$ is a {\em heap typing}, assigning an invariant type to every memory location.  The heap typing
-serves two important roles.  First, it is used to bound the values that can be used for reads when 
-concurrent operation may make the exact value uncertain.  Second, it greatly simplifies the consistency
-check of pointer values, allowing a simple test against the heap typing for a location in lieu of a
-potentially unbounded walk through the store for complicated pointer data structures.
-
-\item $S$ is a store, mapping memory locations $l_i$ to values.
-
-\item $C$ is a {\em clobber set}, memory locations that may be accessed by concurrently executing operations.
-
-\item $e$ is an expression.
-
-\item $v$ is the result of evaluating $e$.
-
-\item $E$ is a {\em memory trace} of the execution, a list of {\em memory operations}
-in the order performed.  A memory operation $\epsilon$ is either
-$read(l, c, v)$ (a read of location $l$ with coherence mode $c$, returning value $v$),
-$write(l, c, v)$ (a write to location $l$ with coherence mode $c$, storing value $v$),
-or $reduce_{id}(l, c, v)$ (a reduction into location $l$ with coherence mode $c$ using an
-associative function $id$ with value $v$).  Coherence modes are one of ({\em excl}, {\em atomic}, {\em simult}).
-\end{itemize}
-
-Helper functions are used in some rules:
-\begin{itemize}
-\item $apply(S, E)$ computes the store that results from applying memory trace $E$ to initial store $S$.
-
-\item $mark\_coh(E, \hat Q)$ records the coherence of operations in trace $E$ according to coherence annotations $\hat Q$.
-
-\item $valid\_interleave(S, C, E', \oton{E})$ is true if $E'$ is a valid interleaving
-of the memory traces $\oton{E}$.
-\end{itemize}
-Figure~\ref{fig:opsemfns} defines {\em apply} and {\em mark\_coherence}.  The precise definition of
-{\em valid\_interleave} is given in Section~\ref{sec:coherence}.  For now, we need only that a valid interleaving
-is at least an interleaving, containing each constituent trace as a subsequence and no other operations.
-
-Some rules modeling concurrent execution are locally
-non-deterministic, guessing values that result from
-concurrent tasks.  For example, the rule for {\tt read(e)} returns an
-arbitrary value $v'$ of the correct type if the location being read is
-in the clobber set $C$, the set of locations concurrent tasks may
-clobber (i.e., update).  The guess is eventually checked for
-consistency by the $valid\_interleave$ predicate of some enclosing
-expression, at a point where the task that wrote $v'$ is also local to
-the expression being evaluated.
-
-Another prescient operation is {\tt newcolor r}, which guesses initial
-memory locations and their colors for a partition.  These locations
-support partition-then-allocate, ensuring the memory mapping $M$ has
-locations for the new subregions that can be allocated by {\tt new}.
-As is usual, memory allocation is also non-deterministic, choosing any
-location of the correct type in the region.  For simplicity we do not
-require locations to be ``fresh''; our soundness results hold even if
-{\tt new T@r} returns locations already in use, provided the heap
-typing of the location is correct.
-
-The most involved rule in Core Legion's semantics is for function calls.
-The only regions accessible to the callee (through $M'$) are the region arguments.
-Also, if any regions are used with relaxed coherence (either {\em atomic} or {\em simult} instead of
-{\em excl}), those locations are added to the clobber set, allowing the
-impact of concurrent modifications of the store to be handled by the caller.  
-%This is done by
-%annotating the called function's memory trace with the relaxed coherence on a per-location basis
-%and using the {\em valid\_interleave} predicate to control what parallel execution is allowed.
+and specify that, given a certain environment, the evaluation of expression $e$ yields a value $v$ and a
+{\em dynamic memory trace} $E$ which is a sequence of all heap operations (reads, writes, reductions) performed
+during the evaluation, including their location and the value read or written.
+
+In addition to the usual value mappings for local variables $L$ and store (i.e. heap) locations $S$, the Legion
+operational semantics environments includes an immutable heap typing $H$ that maps locations to types, and a region
+mapping $M$ that maps logical regions $r_i$ to physical regions $\rho_i$, which are sets of concrete memory
+locations.
+
+The final element of the environtment is the {\em clobber set} $C$.  Our goal is to show that parallel Legion
+programs can
+be scheduled in a hierarchical (and therefore scalable) fashion, we require a formulation of operational semantics
+that is both parallel and hierarchical.  Because Legion allows constrained nondeterminism (in the order in which
+reductions are applied and in cases of relaxed coherence), it is possible for the heap accesses performed by
+one expression's evaluation to observe interference from other concurrent evaluations.  The traditional way of
+handling this in operational semantics \cite{XYZ} is to use a top-down approach with small-step semantics and
+a global view of all
+partially-evaluated expressions.  Rules are specified for which expression(s) are valid choices for the next small
+step of execution.  This approach has the benefit of being constructive but the implicit global ``scheduler'' is
+difficult to map into a hierarchical one in which decisions must necessarily be made with limited information.
+
+Instead we use a bottom-up approach with large-step semantics in which each expression is evaluated with limited 
+information about the overall environment.  The information consists of the initial state of the heap $S$, and a
+set of memory locations that might be modified by other concurrent evaluations.  This is the clobber set $C$,
+and is used to relax the constraints the evaluation of an expression.  Whereas the requirement for locations not
+in the clobber set is the usual one of {\em consistency} (i.e. a read from a location observes the effect of all 
+previous modifications applied in order), the value read from a location in the clobber set may be hypothesized
+to be any value at all.  The validation of these hypotheses is deferred until all potentially-concurrent
+modifications to a location can be observed (i.e. the evaluation of a parent expression whose clobber set does
+not include the location) and a consistent interleaving of the subexpressions' memory traces can be determined.
+With this formulation, the criteria for when a hierarchical scheduler may execute subexpressions in parallel can be
+made clear, but they are no longer constructive.  Instead, we will have to show that the scheduling algorithms
+used by the Legion runtime satisfy these constraints, guaranteeing the soundness of the Legion programming model.
+
+The complete operational semantics rules can be found in \cite{LegionTypes12} - we show the most interesting ones
+in Figure~\ref{fig:semantics}.  The rules for the {\em read}, {\em write}, and {\em reduce} show how dynamic memory
+traces are constructed and how the result of a read expression is relaxed when the location falls within the
+clobber set.  Additionally, the write and reduce operations, with their two subexpressions, demonstrate how the
+traces of subexpressions are interleaved within the environment (and clobber set) of the parent expression.
+
+The first use of the region mapping $M$ occurs in the {\em new} and {\em downregion} expressions.  The new
+expression is constrained to select a location (with the right heap typing) from the set that the specified logical
+region has been mapped to.  (Although one would expect the new expression to also return a location that is not
+currently in use, our soundness result does not require this additional constraint, so we have omitted it for
+simplicity.)  Similarly, the downregion expression checks whether a location is within the set of locations the
+logical region(s) have been mapped to, returning {\em null} if not.  As discussed above, the corresponding 
+{\em upregion} expression checks are performed statically.
+
+The general form of a coloring being an abitrary mapping of locations to ``colors'' (integers) is shown in the
+operational semantics rules for the {\em newcolor} and {\em color} expressions.  The color expression accepts an
+existing coloring and returns one in which the specified location is given the specified color, replacing an
+existing coloring for that location (if present).  The behavior of the newcolor expression is a bit trickier.
+Since Legion supports the {\em partition-then-allocate} model in which the {\em new} expression may be called
+on subregions, each subregion that results from a partitioning may need an arbitrary number of locations for which
+the program does not know the name at partitioning time.  The semantics for the newcolor expression allow this while
+guaranteeing that any explicit colorings performed by the program are honored.
+
+The semantics for the {\em pack} and {\em unpack} expressions are straightforward.  Packing a region relationship
+is just a matter of capturing the physical regions to which the logical regions are currently mapped, whereas 
+unpacking takes those physical regions and gives them fresh logical names within the evaluation of the body.
+
+Finally, although the semantics for calls to subtasks appear complicated at first glance, the bulk of the rules
+are just handling the evaluation of each actual argument expression and the creation of the called subtask's 
+environment (with potentially different names for logical regions).  The key difference in Legion task calls is the
+handling of relaxed coherence.  Any location that falls within
+a logical region whose coherence has been relaxed (via {\em atomic} or {\em simult}) is added to the clobber set
+$C'$, allowing non-deterministic evaluation within the subtask with respect to those locations.  All actual accesses
+to those locations within the resulting memory trace ($E_{n+1}$) are marked with the corresponding coherence mode,
+allowing the {\em valid\_interleave} check within the caller's environment to verify the consistency of the overall
+memory trace.
 

### 2012-10-24 runtime/legion_ops.cc
Fixing bugs in the invalidation of physical contexts which will help the garbage collector

@@ -1096,6 +1096,9 @@ namespace RegionRuntime {
       created_field_spaces.clear();
       created_regions.clear();
       created_fields.clear();
+      deleted_regions.clear();
+      deleted_partitions.clear();
+      deleted_fields.clear();
       launch_preconditions.clear();
       mapped_preconditions.clear();
       map_dependent_waiters.clear();
@@ -1413,17 +1416,17 @@ namespace RegionRuntime {
     }
 
     //--------------------------------------------------------------------------
-    void TaskContext::return_privileges(const std::list<IndexSpace> &new_indexes,
-                                        const std::list<FieldSpace> &new_fields,
-                                        const std::list<LogicalRegion> &new_regions,
-                                        const std::list<FieldID> &new_field_ids)
+    void TaskContext::return_privileges(const std::set<IndexSpace> &new_indexes,
+                                        const std::set<FieldSpace> &new_fields,
+                                        const std::set<LogicalRegion> &new_regions,
+                                        const std::map<FieldID,FieldSpace> &new_field_ids)
     //--------------------------------------------------------------------------
     {
       lock();
-      created_index_spaces.insert(created_index_spaces.end(),new_indexes.begin(),new_indexes.end());
-      created_field_spaces.insert(created_field_spaces.end(),new_fields.begin(),new_fields.end());
-      created_regions.insert(created_regions.end(),new_regions.begin(),new_regions.end());
-      created_fields.insert(created_fields.end(),new_field_ids.begin(),new_field_ids.end());
+      created_index_spaces.insert(new_indexes.begin(),new_indexes.end());
+      created_field_spaces.insert(new_fields.begin(),new_fields.end());
+      created_regions.insert(new_regions.begin(),new_regions.end());
+      created_fields.insert(new_field_ids.begin(),new_field_ids.end());
       unlock();
     }
 
@@ -1431,52 +1434,28 @@ namespace RegionRuntime {
     bool TaskContext::has_created_index_space(IndexSpace handle) const
     //--------------------------------------------------------------------------
     {
-      for (std::list<IndexSpace>::const_iterator it = created_index_spaces.begin();
-            it != created_index_spaces.end(); it++)
-      {
-        if (handle == (*it))
-          return true;
-      }
-      return false;
+      return (created_index_spaces.find(handle) != created_index_spaces.end());
     }
 
     //--------------------------------------------------------------------------
[…]

### 2012-10-24 runtime/region_tree.cc

@@ -397,6 +397,22 @@ namespace RegionRuntime {
     }
 
     //--------------------------------------------------------------------------
+    void RegionTreeForest::get_destroyed_regions(IndexSpace space, std::vector<LogicalRegion> &new_deletions)
+    //--------------------------------------------------------------------------
+    {
+#ifdef DEBUG_HIGH_LEVEL
+      assert(lock_held);
+#endif
+      IndexSpaceNode *target_node = get_node(space);
+      // First destroy all the logical regions trees that use this index space  
+      for (std::list<RegionNode*>::const_iterator it = target_node->logical_nodes.begin();
+            it != target_node->logical_nodes.end(); it++)
+      {
+        new_deletions.push_back((*it)->handle);
+      }
+    }
+
+    //--------------------------------------------------------------------------
     Color RegionTreeForest::create_index_partition(IndexPartition pid, IndexSpace parent, bool disjoint,
                                 int color, const std::map<Color,IndexSpace> &coloring)
     //--------------------------------------------------------------------------
@@ -439,6 +455,21 @@ namespace RegionRuntime {
       // Now we delete the index partition
       deleted_index_parts.push_back(target_node->handle);
       destroy_node(target_node, true/*top*/);
+    }
+
+    //--------------------------------------------------------------------------
+    void RegionTreeForest::get_destroyed_partitions(IndexPartition pid, std::vector<LogicalPartition> &new_deletions)
+    //--------------------------------------------------------------------------
+    {
+#ifdef DEBUG_HIGH_LEVEL
+      assert(lock_held);
+#endif
+      IndexPartNode *target_node = get_node(pid);
+      for (std::list<PartitionNode*>::const_iterator it = target_node->logical_nodes.begin();
+            it != target_node->logical_nodes.end(); it++)
+      {
+        new_deletions.push_back((*it)->handle);
+      }
     }
 
     //--------------------------------------------------------------------------
@@ -527,6 +558,21 @@ namespace RegionRuntime {
       {
         created_field_spaces.erase(finder);
         deleted_field_spaces.pop_back();
+      }
+    }
+
+    //--------------------------------------------------------------------------
+    void RegionTreeForest::get_destroyed_regions(FieldSpace space, std::vector<LogicalRegion> &new_deletions)
+    //--------------------------------------------------------------------------
+    {
[…]

### 2012-10-24 runtime/region_tree.h

@@ -36,15 +36,18 @@ namespace RegionRuntime {
       // Index Space operations
       void create_index_space(IndexSpace space);
       void destroy_index_space(IndexSpace space);
+      void get_destroyed_regions(IndexSpace space, std::vector<LogicalRegion> &new_deletions);
       Color create_index_partition(IndexPartition pid, IndexSpace parent, bool disjoint, int color,
                                   const std::map<Color,IndexSpace> &coloring); 
       void destroy_index_partition(IndexPartition pid);
+      void get_destroyed_partitions(IndexPartition pid, std::vector<LogicalPartition> &new_deletions);
       IndexPartition get_index_partition(IndexSpace parent, Color color);
       IndexSpace get_index_subspace(IndexPartition p, Color color);
     public:
       // Field Space operations
       void create_field_space(FieldSpace space);
       void destroy_field_space(FieldSpace space);
+      void get_destroyed_regions(FieldSpace space, std::vector<LogicalRegion> &new_deletions);
       void allocate_fields(FieldSpace space, const std::map<FieldID,size_t> &field_allocations);
       void free_fields(FieldSpace space, const std::set<FieldID> &to_free);
       bool has_field(FieldSpace space, FieldID fid);
@@ -74,6 +77,10 @@ namespace RegionRuntime {
       InstanceRef initialize_physical_context(const RegionRequirement &req, InstanceRef ref, UniqueID uid, ContextID ctx);
       void map_region(RegionMapper &rm, LogicalRegion start_region);
       Event close_to_instance(const InstanceRef &ref, RegionMapper &rm);
+      void invalidate_physical_context(const RegionRequirement &req, const std::vector<FieldID> &new_fields, ContextID ctx, bool new_only);
+      void invalidate_physical_context(LogicalRegion handle, ContextID ctx);
+      void invalidate_physical_context(LogicalPartition handle, ContextID ctx);
+      void invalidate_physical_context(LogicalRegion handle, ContextID ctx, const std::vector<FieldID> &fields);
     public:
       // Packing and unpacking send
       size_t compute_region_forest_shape_size(const std::vector<IndexSpaceRequirement> &indexes,

### 2012-10-24 doc/types-pldi2013/langdef.tex

@@
 
-The final element of the environtment is the {\em clobber set} $C$.  Our goal is to show that parallel Legion
+The final element of the environment is the {\em clobber set} $C$.  Our goal is to show that parallel Legion
 programs can
@@
 traces of subexpressions are interleaved within the environment (and clobber set) of the parent expression.
+(A formal definition of the {\em valid\_interleave} predicate can be found in Figure~\ref{fig:validinterleave}, and
+will be discussed in Section~\ref{sec:coherence}.)
 
@@
 Theorem~\ref{thm:effects} guarantees that a function call produces a value of the
-correct type and does so within the regions of memory for which privileges are held.
-It does not, however, address {\em coherence}, what concurrent operations on the regions are permitted.
-An innovation in Legion is that programmers can control coherence on a per-region basis.  Intuitively,
-the different coherence modes permit different interleavings of memory traces.  For example,
-in addition to being an interleaving of the constituent traces, exclusive access requires interleaved memory traces to be:
+correct type and does so within the regions of memory for which privileges are held.  In fact, this results holds
+even if the {\em valid\_interleave} constraint is weakened to just require an arbitrary interleaving of the memory
+traces of subexpressions (i.e. the {\em any\_interleave} predicated defined in Figure~\ref{fig:opsecfns}).  The 
+additional constraints on {\em valid\_interleave} in Figure~\ref{fig:validinterleave} address the coherence of 
+heap accesses and allow for the per-region relaxation of coherence that is introduced in Legion.
+
+To determine whether an interleaving of two or more memory traces is valid, we consider three sets of addresses:
 \begin{itemize}
-\item {\em coherent} - a read of location $l$ should see the effect of all writes and reductions to $l$
-occurring earlier in the interleaved trace, and
-\item {\em sequentially equivalent} - the value read by each read 
-operation and the result of applying the interleaved trace to the original store should be the same as if each constituent trace
-was applied to the store in a valid sequential execution of the program.
+\item exclusive locations ($l \in L_{excl}$) are those which have at least once access in exclusive mode in the traces
+and are not in the clobber set.  For these locations, we require the standard {\em sequential consistency} - all
+reads to these locations see the effect of previous writes and reductions, and the resulting state of the store is
+as if all writes and reductions were applied from each trace in order
+\item atomic locations ($l \in L_{atomic}$) are those which have at least one access in atomic mode in the traces and
+are in neither $L_{excl}$ nor the clobber set.  For these locations, we allow permutations of the original
+subexpression trace order.
+\item for locations that have only access in {\em simult} mode or are in the clobber set, no constraints are
+enforced.  The validity of the interleaving of these accesses will be performed in the an enclosing expression.
 \end{itemize}
-These requirements are relaxed for either a non-empty clobber set (representing
-operations that may occur concurrently as part of evaluating other functions) or for coherence
-modes other than {\em excl}.
 
@@
 
-Figure~\ref{fig:validinterleave} formally defines {\em
-valid\_interleave}, which we briefly explain.  Consider
-traces $\oton{E}$.  The {\em exclusive location set} $L_{excl}$ is the
-set of locations used in an operation with coherence mode $excl$ in
-some trace $E_i$, excepting any locations in the clobber set $C$.
-Similarly, the {\em atomic location set} $L_{atomic}$ is the set of
-locations in an operation with coherence mode $atomic$, but no
-operations with coherence mode $excl$, again excepting any location in
-the clobber set.  The $coherent(S,L_1,L_2,E)$ test checks that all
-reads of locations in $L_2$, and any read of a location $l \in L_1$
-that comes after a write to $l$, return the value most recently written
-to the store.  (If $L_2 = \emptyset$ initially, atomic reads without a preceding write are not 
-checked, but are deferred to a point where traces that might be ordered before $E$ are included.)  The $seq\_equiv$
-test verifies every read in $E'$ with a location in $L_1$ 
-is coherent if operations are performed in 
-program order and that the store that results from applying $E'$
-matches the store resulting from applying the
-constituent traces in program order for all locations in $L_1$.  Program order
-is implicitly defined in the operational semantics, and is equivalent to a 
-postorder traversal of an expression's tree.  For
-the atomic location set, we allow the constituent traces to be
-permuted, but all locations must be checked with the same permutation.
-
-Flexibility in interleaving operations is the key to achieving parallel execution, but
-implementing the precise criteria for checking a valid interleaving is
-much too expensive.  We now consider practical mechanisms that generate valid interleavings and 
-under what conditions these approaches may be used.
-
-The first option is to serialize the evaluation of the subexpressions that
-make up an expression and attempt to generate a valid interleaving by simply
-concatenating the constituent traces in program order.  However, there is a
-complication related to the non-determinism that is allowed for reads.  Although
-each subexpression's trace will have been checked for coherence, the checking
-of some operations may have been deferred due to non-empty
-clobber sets.  It is possible that when an interleaving of these traces
-is attempted in the parent expression's context, there may no valid interleaving
-if the wrong guesses were made in the derivation of a subexpression.
-The solution in the case of serial execution is to not expand the clobber
-set for subexpressions, instead using the parent expression's clobber set.
-
-Only a single rule is modified for this alternate execution - $e_{n+1}$ in
-{\sc E-Call} uses the original $C$ instead of $C'$.  This change
-can result in there being no terminating execution for an expression in
-which the only terminating executions required a particular (non-program-)
-ordering of subexpression evaluation.  However, the Legion core language never
-guarantees parallel execution, so the evaluation of such an 
-expression is unlikely to terminate even under the original clobber set.
-We can, however, show that any terminating execution that results from using
-the restricted clobber set is a valid execution under the original:
+Although it is our goal to enable parallel execution whenever possible, it will help to briefly
+consider serialized execution of a Legion program.  We will show that a serial execution trivially
+satisfies the interleaving criteria required by the Legion operational semantics.  In addition to
+serial execution being the fallback behavior of the hierarchical schedule, our proof of the
+soundness of the parallel scheduling will depend on this result.
+
+Serial execution is achieved by ignoring the coherence mode $Q$ in all function calls, using $Q' = \emptyset$
+instead, and by forming ``interleaved'' traces by simply concatenating the subexpressions' traces.  By ignoring
+the coherence modes, the clobber set remains empty, making the result of all {\em read} expressions fully
+determined.  The following lemma and theorem (proofs of which can be found in \cite{LegionTypes12}) show that the
+value and memory trace that result from a serialized execution of an expression are always valid executions of
+Legion programs.
 
@@
 \end{lem}
-
-The proof of this (and other lemmas in this section) is included in \cite{LegionTypes12}.
-We now show that the concatenation of the memory traces from these serially
-executed subexpressions is always a valid interleaving:
 
@@
 
-\begin{proof}
-First, we must show that $any\_interleave(E', \oton{E})$.  This is done by
-induction on the length of $E_1$, followed by $E_2$, and so on.  Next, we must
-show that $E'$ is coherent for both $L_{excl}$ and $L_{atomic}$.  This is done by
-matching the structure of $E'$ with $\oton{E}$ and the use of the following
-lemma: $apply(S, E_1 \dplus E_2) = apply(apply(S, E_1), E_2)$.  With the
-question of coherence addressed, the sequential equivalence of $E'$ is trivial
-for both the exclusive and atomic cases.
-\end{proof}
+CURRENT EDIT POINT
 

### 2012-10-25 apps/fdtd/fdtd.cc
FDTD: Work on math kernels.

@@ -195,6 +195,11 @@ static inline int cell_id(int x, int y, int z,
   return (x*(ny + 2) + y)*(nz + 2) + z;
 }
 
+static inline unsigned cell_stride(unsigned x, unsigned y, unsigned z,
+                                   unsigned nx, unsigned ny, unsigned nz) {
+  return cell_id(x, y, z, nx, ny, nz) - cell_id(0, 0, 0, nx, ny, nz);
+}
+
 ////////////////////////////////////////////////////////////////////////
 // This coloring is only used for initialization of the grid
 // contents. In this scheme each block on the outer surface of the
@@ -818,64 +823,30 @@ void step_task(const void * input_global_args, size_t input_global_arglen,
   RegionRuntime::LowLevel::RegionAccessor<RegionRuntime::LowLevel::AccessorGeneric> ghost1 = ghost1_cells.get_accessor<AccessorGeneric>(field_read1);
   RegionRuntime::LowLevel::RegionAccessor<RegionRuntime::LowLevel::AccessorGeneric> ghost2 = ghost2_cells.get_accessor<AccessorGeneric>(field_read2);
 
+  // hx[x, y, z] += (ez[x, y + 1, z] - ez[x, y, z]) -
+  //                (ey[x, y, z + 1] - ey[x, y, z])
+  int d = dir == DIR_POS ? 1 : -1;
+  int s1 = 0, s2 = 0;
+  if (dim == DIM_X) {
+    s1 = cell_stride(0, d, 0, nx, ny, nz);
+    s2 = cell_stride(0, 0, d, nx, ny, nz);
+  } else if (dim == DIM_Y) {
+    s1 = cell_stride(0, 0, d, nx, ny, nz);
+    s2 = cell_stride(d, 0, 0, nx, ny, nz);
+  } else if (dim == DIM_Z) {
+    s1 = cell_stride(d, 0, 0, nx, ny, nz);
+    s2 = cell_stride(0, d, 0, nx, ny, nz);
+  } else {
+    assert(0 && "unreachable");
+  }
+
   for (int x = x_min; x < x_max; x++) {
     for (int y = y_min; y < y_max; y++) {
       for (int z = z_min; z < z_max; z++) {
-        int c = cell_id(x, y, z, nx, ny, nz);
-        assert(fabs(write.read(ptr_t<double>(c)) - 1.2345678*c*field_write) < 0.000001);
[…]

### 2012-10-25 doc/types-pldi2013/intro.tex

\section{Introduction}
\label{sec:intro}

In the past decade machine architectures, particularly at the high
performance end of the spectrum, have undergone a revolution.  The
latest supercomputers are now composed of heterogeneous processors
and deep memory hierarchies.  Programming this class of machines is
becoming increasingly difficult.  Current programming systems for these
machines have intricate features for describing parallelism.  However,
these systems relegate to the programmer the responsibility for:
\begin{itemize}
\item the correctness of their parallel declarations
\item the movement and placement of data in the memory hierarchy
\end{itemize}
Coupled with the complexity of these machines, this places a heavy
burden on the programmer.  The reason that the programmer must bear
this load is because current programming systems have few or no
insights into properties of program data (e.g. locality and independence).  
To enable more efficient programming of this class of machines we 
need programming systems capable of reasoning both statically and dynamically
about the structure of program data.

Researchers have previously explored language designs with hierarchical 
decomposition of data to express locality and independence for high 
performance programming. Two recent examples, Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
mechanism to partition the heap recursively into a tree of collections 
of data.  The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that must be fully checked statically 
(see Section~\ref{sec:related} for more discussion of related work).

Our own experience with writing high-performance applications in both 
Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow multiple, simultaneous partitions of
the same data, providing different views on to that data.

To support reasoning about the structure
of data in applications with dynamically instantiated data structures, 
we present a type system and operational semantics for Legion\cite{Legion12}.  
Legion is a programming model that presents {\em logical regions} as an
abstraction for capturing {\em locality} (data that will be used together,
and therefore should be colocated) and {\em independence} (disjoint data
that may be operated on in parallel).  Logical regions have the following
properties that enable programmers to dynamically communicate information
about the structure of program data to Legion:
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

\item logical regions can be dynamically {\em partitioned} into {\em subregions}; 
the programmer can express arbitrary partitions of regions;

\item  a logical region may be dynamically partitioned in multiple different ways.
\end{itemize}

While data in Legion is organized into a hierarchy of logical region forests, 
computation is organized into a hierarchy of tasks.  Logical regions and tasks interact
through a system of privileges specifying operations that a task
can perform on a logical region argument (read, write, or reduce).  
Tasks are only permitted to access logical regions for which they 
possess privileges.  
%Privileges are passed to a task by their calling task
%or by creating new logical regions.  
%The Legion programming model 
%mandates a functional restriction: a sub-task can only be passed a subset
%of the privileges possesed by its parent task.  As we describe below,
%our type system relies on a combination of static and dynamic checks to 
%enforce this restriction owing to the first-class nature of logical regions.

Legion's ability to describe dynamically instantiated data structures with
multiple dynamic partitions of data does have an important consequence:
aliasing.  Through multiple partitions
of a logical region, a given datum may belong to multiple different regions.  
For example, if a region $R$ is disjointly partitioned three distinct ways 
and each partition assigns every element in $R$ to a
subregion (partitions need not be total in Legion), then every element
in $R$ is included in three different subregions.  This is why we use
the term {\em logical} regions: language-level regions serve to name
sets of data but do not imply anything about physical layout or placement
in the memory hierarchy.\footnote{There
is a separate system of {\em physical regions} at runtime that hold
copies of the data in the language-level logical regions.  Details can
be found in \cite{Legion12}.}
%In fact,
%the Legion implementation routinely maintains multiple distinct
%physical regions, each with a copy of the data in the corresponding
%logical region, for performance reasons \cite{Legion12}.

One consequence of aliasing is that it significantly hinders 
static analyses for proving properties of parallel programs.  
Approaches to dealing with aliasing in the domain
of parallelism have traditionally bifurcated to extremes: Sequoia and DPJ both disallow aliasing
which enables their static analyses at the coarse granularity of regions;  
approaches like transactional memory\cite{} and thread-level speculation\cite{} perform 
fully dynamic analysis at the granularity of individual heap locations.
Alternatively, the type system we present for Legion leverages a hybrid
approach to handling aliasing that enables a combination of static and dynamic
analyses to prove important properties of Legion programs.

The key observation behind our type system is that dynamic alias analysis
is very easy and relatively cheap when done at the granularity of regions 
instead of individual heap locations.  To that end, our type system supports
static verification of all pointer dereferences and passing of region privileges at 
task calls boundaries.  The Legion runtime then leverages its knowledge of logical
region aliasing to perform dynamic safety checks for the unpacking
and packing of logical regions to/from heap with regards to task privileges.  
These checks are then amortized over all the uses of the logical region 
which significantly reduces their cost.

Another important consequence of aliasing is that it often limits 
parallelism.  Legion currently does not attempt to statically prove any two tasks
parallel, but instead relies on dynamic region aliasing checks to discover
parallelism at runtime.  If regions contain many elements, then these checks are
amortized as we will show in Section~\ref{sec:evaluation}.  In addition
to supporting parallel execution in alias-free tasks, Legion also enables
programmer controlled parallel execution in the presence of aliasing.
In many cases, programmers often want to enable parallel execution 
even in the presence of aliased data.  For example, two tasks 
accessing aliased data may only need to be serialized
with respect to each other (e.g. run atomically) or are permitted to run
simultaneously with the programmer using separate synchronization mechanisms
to control access (e.g. locks).  To support these relaxed access modes
Legion provides {\em cohrence modes} on logical regions.  For each region
argument to a task, a programmer specifies the coherence in which other tasks
can access aliased data while the task is running.  Our operational semantics
provides a formal framework with which to reason about the interaction of tasks
accessing aliased in different coherence modes.

We prove the soundess of both the privilges system and the coherence system.
We then use these results to prove a third result: if two {\em siblings} (tasks that
have the same immediate parent task) $t_1$ and $t_2$ are 
{\em non-interfering} (have no ordering requirements for correctness), 
then the any descendant of $t_1$ in the task hierarchy is non-interfering
with any descendant of $t_2$.  This theorem is the basis for correct, distributed
task scheduling in a Legion implementation, which is crucial for
performance; a centralized scheduler would be a bottleneck
because of the large communication latencies in the target
class of machines.

The crucial aspect of this result is that the tasks $t_1$ and $t_2$ only 
need be non-interfering instead of non-aliased.  Thus, through a combination
of static and dynamic analysis, our privilge type system and coherence system
enable us to implement a hierarchical, distributed scheduling algorithm even
in the presence of aliasing introduced by the need for dynamic descriptions of
data via first-class logical regions.  
%All of this is possible only because
%our programming system supports an abstraction for understanding the structure
%of program data.

%The important difference between Legion and the more static approaches
%is that Legion allows region aliasing, but does so in a structured way that minimizes
%runtime overheads.  
%The key observation is that static alias analysis is hard, but dynamic 
%alias analysis is very easy and relatively cheap when done at the granularity 
%of regions instead of individual heap locations.  
%Thus, Legion falls between fully static systems, such as
%Sequoia and DPJ, and fully dynamic approaches such as transactional
%memory.  
%Legion still has a significant static component; in
%particular, the required privileges for function calls
%and region
%pointer dereferences are checked statically.  The dynamic alias checks
%for independence are done at the granularity of regions and one check 
%on a region
%often serves to prove the independence of many subsequent uses of that region.
%In contrast, because transactional memory has no mechanism
%for grouping heap data into larger collections, it must test the
%overlap between two sets of data on each individual memory location, which is
%significantly more expensive.

%which makes it difficult 
%to statically determine which computations can be run in parallel.  
%One approach is to outlaw aliasing which allows for a static analysis
%to discover all available parallelism and schedule it entirely at
%compile-time, avoiding any runtime overhead.  Languages such as
%Sequoia and DPJ take this approach and allow only a single static partition of any region 
%to eliminate region aliasing, at some cost in expressiveness.

%While many programming systems for
%these machines have intricate features for describing parallelism,
%few have any abstractions for capturing properties of a program's data (e.g.
%independence and locality).  Understanding the structure of a program's data
%is critical for allowing programming systems to validate/discover
%parallelism or support placement of data in the memory hierarchy.  
%The few programming systems that do support data abstraction 
%features\cite{Fatahalian06,Bocchino09} permit only
%statically analyzable abstractions which restrict expressivity.  We
%present a type system for Legion\cite{Legion12}, a programming model
%that permits both static and dynamic descriptions of the structure of 
%program data.  Despite the dual nature of Legion, our type system allows us 
%to statically prove several useful properties about the structure and usage 
%of data in Legion programs.  We show how to leverage these properties 
%to enable a scalable, hierarchical scheduling algorithm based on a 
%dynamic analysis of program data.

%complex hierarchies of many different
%kinds of computing technologies: networks of nodes at the top level,
%multiple chips per node, multiple cores within a chip, and, 
%recently, multiple accelerators (usually GPUs) per node, which 
%can themselves be further decomposed.  We present the operational and static
%semantics of Legion\cite{Legion12}, a programming model targeted at providing an
%appropriate level of abstraction for programming such machines, one
%that is both sufficiently high-level to be portable while still
%exposing aspects that are crucial to performance.

%The primary abstraction for capturing the structure of data in Legion is {\em logical regions}. Logical
%regions express {\em locality} (data that will be used together, and therefore should be colocated) 
%and {\em independence} (disjoint data that may be operated on in parallel).
%Legion programs organize data into a forest of logical regions.  Logical regions can be
%recursively partitioned into sub-regions.

%In Legion data is organized in a hierarchy of {\em regions}
%and subregions while computation is organized in a hierarchy of {\em
%tasks} and subtasks operating on regions.  Regions and tasks interact
%through a static system of {\em region privileges} specifying 
%operations a task may perform on a region argument (read,
%write, or reduce) and  {\em region coherence} annotations that
%express what other tasks may do concurrently with
%the region (exclusive, atomic, or simultaneous).  We prove the
%soundness of the privileges and coherence system and use these two
%results to prove a third result: if two {\em siblings} (tasks that
%have the same immediate parent task) $t_1$ and $t_2$ are 
%{\em non-interfering} (have no ordering requirements for correctness), 
%then the any descendant of $t_1$ in the task hierarchy is non-interfering
%with any descendant of $t_2$.  This theorem is the basis for correct distributed
%task scheduling in a Legion implementation, which is crucial for
%performance; a centralized scheduler would be a bottleneck
%because of the large communication latencies in the target
%class of machines.

% Putting this here so that we can get the code in a good place
\lstset{
  captionpos=b,
  language=Haskell,
  basicstyle=\scriptsize,
  numbers=left,
  numberstyle=\tiny,
  columns=fullflexible,
  stepnumber=1,
  escapechar=\#,
  keepspaces=true,
  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
  morekeywords={function,rr,int,float,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,multicoloring,color,newcolor,atomic,simultaneous},
  deletekeywords={float,head,min,max}
}
\begin{lstlisting}[float={t},label={lst:circuit_ex},caption={Circuit Simulation}]
--                        <voltage,current,charge,capacitance>
type CircuitNode        = <float,float,float,float>
--                     < owned node, owned or ghost node, resistance, current>
type CircuitWire<rn,rg>  = <CircuitNode@rn, CircuitNode@(rn,rg),float,float>

type node_list<rl,rn>       = < CircuitNode@rn, node_list<rl,rn>@rl >
type wire_list<rl,rw,rn,rg>= < CircuitWire<rn,rg>@rw, wire_list<rl,rw,rn,rg>@rl >

type CircuitPiece<rl,rw,rn> = rr[rpw,rpn,rg]
                            < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl >         
                            where rpn #$\le$# rn and rg #$\le$# rn and rpw #$\le$# rw and
                                  rn * rw and rl * rn and rl * rw

-- Simulation initialization and invocation
function simulate_circuit[rl,rw,rn] ( all_nodes : node_list<rl,rn>@rl, 
                            all_wires : wire_list<rl,rw,rn,rn>@rl, steps : int ), 
      reads(rn,rw,rl), writes(rn,rw,rl) : bool = 
  let pc : <coloring(rn),multicoloring(rn),coloring(rw)> 
            = color_circuit[rn,rw,rl](all_nodes,all_wires) in
  -- Disjoint partition for the owned nodes of each piece
  partition rn using pc.1 as rn0,rn1 in
  -- Aliased partition for ghost nodes of each piece
  partition rn using pc.2 as rg0,rg1 in
  -- Disjoint partition for the owned wires of each piece
  partition rw using pc.3 as rw0,rw1 in
  let lists0 : <wire_list<rl,rw0,rn0,rg0>@rl,node_list<rl,rn0>@rl> = 
        build_lists[rl,rw,rn,rw0,rn0,rg0](all_nodes,all_wires,pc.1,pc.2,pc.3,0) in
  let piece0 : CircuitPiece<rl,rw,rn> = 
        pack lists0 as CircuitPiece<rl,rw,rn>[rw0,rn0,rg0] in
  let lists1 : <wire_list<rl,rw1,rn1,rg1>@rl,node_list<rl,rn1>@rl> =
        build_lists[rl,rw,rn,rw1,rn1,rg1](all_nodes,all_wires,pc.1,pc.2,pc.3,1) in
  let piece1 : CircuitPiece<rl,rw,rn> = 
        pack lists1 as CircuitPiece<rl,rw,rn>[rw1,rn1,rg1] in
      execute_time_steps[rl,rw,rn](piece0,piece1,steps)

-- Time Step Loop
function execute_time_steps[rl,rw,rn] ( p0 : CircuitPiece<rl,rw,rn>, 
      p1 : CircuitPiece<rl,rw,rn>, steps : int ) , reads(rn,rw,rl), writes(rn,rw) : bool = 
  if steps #$<$# 1 then true else
  unpack p0 as piece0 : CircuitPiece<rl,rw,rn>[rw0,rn0,rg0] in 
  unpack p1 as piece1 : CircuitPiece<rl,rw,rn>[rw1,rn1,rg1] in
  let _ : bool = calc_new_currents[rl,rw0,rn0,rg0](piece0.1) in
  let _ : bool = calc_new_currents[rl,rw1,rn1,rg1](piece1.1) in
  let _ : bool = distribute_charge[rl,rw0,rn0,rg0](piece0.1) in
  let _ : bool = distribute_charge[rl,rw1,rn1,rg1](piece1.1) in
  let _ : bool = update_voltage[rl,rn0](piece0.2) in
  let _ : bool = update_voltage[rl,rn1](piece1.2) in
      execute_time_steps[rl,rw,rn](p0,p1,steps-1)

function color_circuit[rn,rw,rl] ( all_nodes : node_list<rl,rn>@rl, 
                               all_wires : wire_list<rl,rw,rn>@rl ), 
        reads(rn,rw,rl) : <coloring(rn), multicoloring(rn), coloring(rw)> =  
  -- Invoke programmer chosen coloring algorithm (e.g. METIS)
  -- return owned, ghost, wire colorings

-- Helper method
function build_lists[rl,rw,rn,rpw,rpn,rg] ( nodes : node_list<rl,rn>@rl, 
       wires : wire_list<rl,rw,rn>@rl, oc : coloring(rn), gc : multicoloring(rn), 
       wc : coloring(rw), c : int), reads(rn,rw,rl), writes(rl) 
       : < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl > = 
  -- Construct lists of node and wire pointers for the given colorings
\end{lstlisting}

% Moving this to the circuit example section since it fits better there
% and some of this is redundant
%Legion's execution model is that by default, a program's semantics is
%its meaning as a sequential program, which we refer to as the {\em
%program order} execution.  While the
%implementation is not our focus here, if two functions use
%disjoint data (because all of the regions they use are disjoint), then
%Legion may execute them simultaneously.  Also like DPJ,
%Legion has a system of {\em privileges} ({\em read}, {\em write}, and
%{\em reduce}) on regions and an orthogonal system of region {\em
%coherence} modes ({\em excl}, {\em atomic}, and {\em simultaneous})
%that increase the number of situations in which functions can be
%executed in parallel.

The rest of the paper is organized as follows.
We begin in Section~\ref{sec:example} with an example program that illustrates
a typical style of programming in Legion and motivates the need for multiple, dynamically computed partitions of
a region that introduce aliased data.  We define Core Legion, a small language suitable 
for our formal development, in Section~\ref{sec:legioncore}.
The next four sections each state and prove one of our main results and contributions:
\begin{itemize}
\item We prove the soundness of Legion's privileges system (Section~\ref{sec:soundness}).

\item We use the soundness of privileges to prove the soundness of Legion's region coherence modes (Section~\ref{sec:coherence}).

\item We show that if expressions $e_1$ and $e_2$ are {\em non-interfering} (can be executed in parallel), then subexpressions
$e_1'$ of $e_1$ and $e_2'$ of $e_2$ are also non-interfering (Section~\ref{sec:scheduling}).  
%This result is the basis
%for a hierarchical, distributed scheduler in the Legion implementation which is crucial for high performance
%on the target class of machines.
%; on the target class of machines, any centralized
%scheduler would be a serious bottleneck.

\item We give experimental evidence for the Legion design.  On three real-world applications, we show that 
dynamic region pointer checks would be expensive, justifying checking this aspect of the type system statically.
We also show that the cost of checking for non-interference of region privileges is low, showing that a much more expressive
and dynamic language with aliasing is not incompatible with high performance (Section~\ref{sec:evaluation}.

\end{itemize}
Finally, Section~\ref{sec:related} discusses the most closely related work and
Section~\ref{sect:conclusion} concludes.

### 2012-10-26 doc/types-pldi2013/langdef.tex
checkpoint

@@
 
+\subsection{Serial Execution}
+\label{subsec:coherenceserial}
+
 Although it is our goal to enable parallel execution whenever possible, it will help to briefly
@@
 \end{thm}
+
+\subsection{Parallel Execution}
+\label{subsec:coherenceparallel}
+
+To determine when parallel execution is safe, we start from the serial execution trace and allow the reordering of 
+adjacent heap operations that do not change the behavior of the application.  If we can show that reordering is allowed
+between any two operations that come from different constituent traces, any interleaving of the constituent traces will
+be equivalent to the serialized execution and parallel execution was safe.  Finally, to allow the safety of parallel
+execution to be determined {\em before} the execution is actually performed, we show that a test based on the subtask
+privileges and the current region mapping can soundly predict when this property will hold.
+
+We define a {\em non-interference} operator $\epsilon_1 \nonint \epsilon_2$ on two memory operations as follows:
+$$
+\begin{array}{l@{ }l}
+op_1(l_1, c_1,& v_1)\ \nonint\ op_2(l_2, c_2, v_2) \Leftrightarrow l_1 \not= l_2 \vee \\
+& (op_1 = read \wedge op_2 = read)\ \vee \\
+& (op_1 = reduce_{id_1} \wedge op_2 = reduce_{id_2} \wedge id_1 = id_2)
+\end{array}
+$$
+Accesses to different locations cannot effect each other, and reads have no side-effects.  The safety of the third case
+follows from the requirement that reduction operations be commutative.
+
+The next step is to show that this non-interference operator is strong enough to preserve the validity of an interleaving
+of two or more traces when reordering a pair of memory operations.
+\begin{lem}
+\label{lem:nonintswap}
+\rm
+Let $S$ be a store, $C$ a clobber set, $\oton{E}, E'_a, E'_b$ memory traces, and
+$\epsilon_1, \epsilon_2$ be two memory operations from $E_i$ and $E_j$ ($i \not= j$).  Then,
+$$\begin{array}{l@{}l}
+valid\_interleave&(S, C, E'_a \dplus [ \epsilon_1, \epsilon_2 ] \dplus E'_b, \oton{E}) \wedge \epsilon_1 \nonint \epsilon_2 \Rightarrow \\
+ & valid\_interleave(S, C, E'_a \dplus [ \epsilon_2, \epsilon_1 ] \dplus E'_b, \oton{E}).
+\end{array}$$
+\end{lem}
+The proof follows from the definitions of {\em coherent} and {\em seq\_equiv} in Figure~\ref{fig:validinterleave}.
+
+We now extend the notion of non-interference to whole memory traces, with two traces being non-interfering if no operation from one
+trace interferes with any operation from the other:
+$$E_1 \nonint E_2 \Leftrightarrow \displaystyle\bigwedge_{\epsilon_1 \text{ in } E_1, \epsilon_2 \text{ in } E_2} \epsilon_1 \nonint \epsilon_2$$
+
+If whole memory traces are non-interfering, then any interleaving of them can be sorted via pairwise swaps to match the serial memory
+trace and we get the result that shows parallel execution is safe:
+
+\begin{lem}
+\label{lem:noninteffects}
+\rm
+Let $S$ be an initial store, $C$ be a clobber set, $\oton{E}$ be memory traces such that $E_i \nonint E_j$
+for every $1 \leq i < j \leq n$.  Then,
+$any\_interleave(E', \oton{E}) \Rightarrow$ \\
+\hspace*{0pt} \hfill $valid\_interleave(S, C, E', \oton{E})$.
+\end{lem}
+
+The final piece is to turn this test into something that can be performed before the subtasks are executed, ideally as efficiently as
+possible.  The key to this is the privileges on Legion tasks and their ability to provide bounds on the accesses that may be performed
+during execution.  Our non-interference operator is extended to work on privileges and sets of privileges:
+$
+\begin{array}{l@{}l}
+\multicolumn{2}{l}{priv_1(r_1) \nonint[M] priv_2(r_2) \Leftrightarrow} \\
+& (priv_1 = reads \wedge priv_2 = reads)\ \vee \\
+& (priv_1 = reduces_{id_1} \wedge priv_2 = reduces_{id_2} \wedge id_1 = id_2)\ \vee \\
+& (r_1 * r_2)\ \vee \\
+& (M(r_1) \cap M(r_2) = \emptyset) \\
+\\
+\multicolumn{2}{l}{\Phi_1 \nonint[M] \Phi_2 \Leftrightarrow \displaystyle\bigwedge_{\phi_1 \in \Phi_1, \phi_2 \in \Phi_2} \phi_1 \nonint[M] \phi_2}
+\end{array}$
+Note although the
 

### 2012-10-28 doc/types-pldi2013/intro.tex
More edits to the intro for pldi

@@
 and deep memory hierarchies.  Programming this class of machines is
-becoming increasingly difficult.  Current programming systems for these
-machines have intricate features for describing parallelism.  However,
-these systems relegate to the programmer the responsibility for:
-\begin{itemize}
-\item the correctness of their parallel declarations
-\item the movement and placement of data in the memory hierarchy
-\end{itemize}
-Coupled with the complexity of these machines, this places a heavy
-burden on the programmer.  The reason that the programmer must bear
-this load is because current programming systems have few or no
-insights into properties of program data (e.g. locality and independence).  
-To enable more efficient programming of this class of machines we 
-need programming systems capable of reasoning both statically and dynamically
-about the structure of program data.
+increasingly difficult.  Current programming systems for these
+machines have intricate features for describing parallelism, but contain
+few abstractions for describing properties of program data
+(e.g. independence).  Consequently, the burden of managing
+the correctness of parallel computations and movement of data
+through the memory hierarchy is placed on the programmer.
 
-Researchers have previously explored language designs with hierarchical 
-decomposition of data to express locality and independence for high 
-performance programming. Two recent examples, Sequoia \cite{Fatahalian06} and 
+%However,
+%these systems relegate to the programmer the responsibility for:
+%\begin{itemize}
+%\item the correctness of their parallel declarations
+%\item the movement and placement of data in the memory hierarchy
+%\end{itemize}
+%Coupled with the complexity of these machines, this places a heavy
+%burden on the programmer.  The reason that the programmer must bear
+%this load is because current programming systems have few or no
+%insights into properties of program data (e.g. locality and independence).  
+%To enable more efficient programming of this class of machines we 
+%need programming systems capable of reasoning both statically and dynamically
+%about the structure of program data.
+
+To combat this problem, researchers have previously explored programming
+systems that are capable of reasoning about the structure of data in
+parallel programs.  Traditionally, these attempts have taken two approaches.   
+The first approach relies on fully static analyses that
+have no runtime overhead, but limited expressivity.  Two recent examples, 
+Sequoia \cite{Fatahalian06} and 
 Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
-mechanism to partition the heap recursively into a tree of collections 
+mechanism to statically partition the heap recursively into a tree of collections 
 of data.  The two designs are different in many aspects, but agree that there is
 a single tree-shaped partitioning of data that must be fully checked statically 
-(see Section~\ref{sec:related} for more discussion of related work).
+(see Section~\ref{sec:related} for more discussion of related work).  The
+second approach relies on a dynamic analysis of program data that doesn't
+limit expressivity, but incurs runtime overhead.  Examples of this approach 
+include transactional memory and thread-level speculation.
 
-Our own experience with writing high-performance applications in both 
-Sequoia and the current industry standard mix of MPI, shared-memory
+These two different approaches illustrate different mechanisms for dealing with
+the problem of data aliasing in parallel programs.  The fully static approach outlaws aliasing
+by requiring a single static partitioning of data.  Our own experience with writing high-performance 
+applications in both Sequoia and the current industry standard mix of MPI, shared-memory
 threads, and CUDA, has taught us that for many practical situations a
@@
 Furthermore, it is useful to allow multiple, simultaneous partitions of
-the same data, providing different views on to that data.
+the same data, providing different views on to that data.  Alternatively, 
+the fully dynamic approach enables arbitrary data aliasing, but requires 
+centralized control logic to perform conflict detection between every pair 
+of parallel computations. This results in runtime overhead and is 
+not scalable on distributed-memory machines.
 
-To support reasoning about the structure
-of data in applications with dynamically instantiated data structures, 
-we present a type system and operational semantics for Legion\cite{Legion12}.  
-Legion is a programming model that presents {\em logical regions} as an
+In this work, we present a type system and operational semantics for
+Legion\cite{Legion12}.  Legion is a programming model that supports multiple,
+dynamic partitions of program data which can result in aliasing.  We prove
+the soundness of our type system and show that through a combination of
+static and dynamic analysis we can implement it in a way that minimizes dynamic
+runtime checks and results in low actual runtime overhead.  We then show 
+that the soundness of our type system enables a scalable, hierarchical 
+distributed scheduling algorithm that enables Legion programs to achieve
+high performance on current supercomputers.
+
+Legion presents {\em logical regions} as an
 abstraction for capturing {\em locality} (data that will be used together,
@@
 \end{itemize}
-
-While data in Legion is organized into a hierarchy of logical region forests, 
-computation is organized into a hierarchy of tasks.  Logical regions and tasks interact
-through a system of privileges specifying operations that a task
-can perform on a logical region argument (read, write, or reduce).  
-Tasks are only permitted to access logical regions for which they 
-possess privileges.  
-%Privileges are passed to a task by their calling task
-%or by creating new logical regions.  
-%The Legion programming model 
-%mandates a functional restriction: a sub-task can only be passed a subset
-%of the privileges possesed by its parent task.  As we describe below,
-%our type system relies on a combination of static and dynamic checks to 
-%enforce this restriction owing to the first-class nature of logical regions.
-
-Legion's ability to describe dynamically instantiated data structures with
-multiple dynamic partitions of data does have an important consequence:
-aliasing.  Through multiple partitions
+Through multiple partitions
 of a logical region, a given datum may belong to multiple different regions.  
@@
 be found in \cite{Legion12}.}
-%In fact,
-%the Legion implementation routinely maintains multiple distinct
-%physical regions, each with a copy of the data in the corresponding
-%logical region, for performance reasons \cite{Legion12}.
 
-One consequence of aliasing is that it significantly hinders 
-static analyses for proving properties of parallel programs.  
-Approaches to dealing with aliasing in the domain
-of parallelism have traditionally bifurcated to extremes: Sequoia and DPJ both disallow aliasing
-which enables their static analyses at the coarse granularity of regions;  
-approaches like transactional memory\cite{} and thread-level speculation\cite{} perform 
-fully dynamic analysis at the granularity of individual heap locations.
-Alternatively, the type system we present for Legion leverages a hybrid
-approach to handling aliasing that enables a combination of static and dynamic
-analyses to prove important properties of Legion programs.
+While data in Legion is organized into a hierarchy of logical region forests, 
+computation is organized into a hierarchy of tasks.  Logical regions and tasks interact
+through a system of privileges specifying operations that a task
+can perform on a logical region argument (read, write, or reduce).  
+Tasks are only permitted to access logical regions for which they 
+possess privileges.  Privileges are given to a task by their calling task
+or by creating new logical regions.  The Legion programming model 
+mandates a functional restriction: a sub-task can only be passed a subset
+of the privileges possesed by its parent task.
 
-The key observation behind our type system is that dynamic alias analysis
-is very easy and relatively cheap when done at the granularity of regions 
-instead of individual heap locations.  To that end, our type system supports
+To enforce this restriction, our type system relies on a combination of static 
+and dynamic analyses.  The key observation behind our type system is that dynamic
+analysis is very easy and relatively cheap when done at the granularity of
+regions instead of individual heap locations.  To that end, our type system supports
 static verification of all pointer dereferences and passing of region privileges at 
@@
 
-Another important consequence of aliasing is that it often limits 
-parallelism.  Legion currently does not attempt to statically prove any two tasks
-parallel, but instead relies on dynamic region aliasing checks to discover
-parallelism at runtime.  If regions contain many elements, then these checks are
-amortized as we will show in Section~\ref{sec:evaluation}.  In addition
-to supporting parallel execution in alias-free tasks, Legion also enables
-programmer controlled parallel execution in the presence of aliasing.
-In many cases, programmers often want to enable parallel execution 
-even in the presence of aliased data.  For example, two tasks 
+To enable parallelism, Legion again relies on a combination of static and dynamic
+analysis to prove that two tasks operate on independent sets of data.  If two tasks
+cannot be proven independent statically, the Legion runtime performs dynamic dependence
+tests to detect region aliasing.  Since these tests are performed at the granularity of
+regions instead of individual heap locations, they are significantly less costly.
+In addition to supporting parallel execution for independent tasks, Legion also enables
+programmer specified parallel execution in the presence of aliasing.  For example, two tasks 
 accessing aliased data may only need to be serialized
@@
 argument to a task, a programmer specifies the coherence in which other tasks
-can access aliased data while the task is running.  Our operational semantics
-provides a formal framework with which to reason about the interaction of tasks
-accessing aliased in different coherence modes.
+can access aliased data while the task is running.  We present an operational semantics
+for coherence modes and show that it is sound.
 
-We prove the soundess of both the privilges system and the coherence system.
-We then use these results to prove a third result: if two {\em siblings} (tasks that
+Using our proofs of the soundess of the privilegs and coherence systems, we then 
+prove a third result: if two {\em siblings} (tasks that
 have the same immediate parent task) $t_1$ and $t_2$ are 
@@
 
-The crucial aspect of this result is that the tasks $t_1$ and $t_2$ only 
-need be non-interfering instead of non-aliased.  Thus, through a combination
-of static and dynamic analysis, our privilge type system and coherence system
-enable us to implement a hierarchical, distributed scheduling algorithm even
-in the presence of aliasing introduced by the need for dynamic descriptions of
-data via first-class logical regions.  
+%The crucial aspect of this result is that the tasks $t_1$ and $t_2$ only 
+%need be non-interfering instead of non-aliased.  Thus, through a combination
+%of static and dynamic analysis, our privilge type system and coherence system
+%enable us to implement a hierarchical, distributed scheduling algorithm even
+%in the presence of aliasing introduced by the need for dynamic descriptions of
+%data via first-class logical regions.  
 %All of this is possible only because
@@
 We also show that the cost of checking for non-interference of region privileges is low, showing that a much more expressive
-and dynamic language with aliasing is not incompatible with high performance (Section~\ref{sec:evaluation}.
+and dynamic language with aliasing is not incompatible with high performance (Section~\ref{sec:evaluation}).
 

### 2012-10-29 doc/types-pldi2013/legion-types.tex
Updating the abstract and title

--- 2012-07-13 doc/popl2013/legion-types.tex
@@
-\documentclass[9pt,nocopyrightspace]{sigplanconf}
+\documentclass[9pt,nocopyrightspace,preprint]{sigplanconf}
 % preprint for dates
@@
 
-\title{A Type System for Safe and Scalable Parallelism using Logical Regions}
+\title{A Type System for Structured Aliasing in Parallel Programs}
 %\title{Scalable and Safe Parallelism using Logical Regions}
@@
 \begin{abstract}
-
-Legion is a programming model that introduces
-{\em logical regions} as a mechanism for targeting large heterogeneous machines 
-with complex memory hierarchies.  Logical regions are first-class 
-values in Legion, and programmers may {\em partition} regions in multiple different
-ways into subregions.  We present
-a type system with privileges and coherence properties for reasoning about logical regions.  
-We prove the soundness of the type system and use it to show
-that all memory accesses in Legion are safe.  Furthermore, we prove that
-type soundness enables a scalable, hierarchical
-scheduling algorithm for Legion programs.  We show that an implementation of the type checker
-for Legion programs improves performance up to 71\% by eliding
-memory checks the type system proves safe.  We also demonstrate that an implementation of the Legion 
-runtime based on the hierarchical scheduling algorithm scales well for
-three real-world applications.
+Aliasing introduces a significant impediment to reasoning about the
+structure and usage of data in parallel programs.  Traditional approaches
+have either disallowed aliasing to enable static analyses, or have permitted
+aliasing at the cost of dynamic runtime analyses.  In this paper we introduce a
+type system for the Legion programming model that uses a hybrid static and
+dynamic approach to manage aliasing with minimal runtime overhead.  We prove 
+that our type system can soundly reason about the privilege and coherence 
+properties in Legion programs. Furthermore, we prove that type soundness 
+enables a hierarchical, distributed scheduling algorithm for Legion programs 
+even in the presence of aliasing. We show that an implementation of a Legion 
+type checker improves the performance of Legion programs up to 71\% by 
+eliding memory checks that the type system proves safe.  We also demonstrate 
+that an implementation of the Legion runtime based on the hierarchical 
+scheduling algorithm scales well for three real-world applications, achieving 
+up to 62.5X speedup on 96 GPUs on the Keeneland supercomputer.
+%Legion is a programming model that introduces
+%{\em logical regions} as a mechanism for targeting large heterogeneous machines 
+%with complex memory hierarchies.  Logical regions are first-class 
+%values in Legion, and programmers may {\em partition} regions in multiple different
+%ways into subregions.  We present
+%a type system with privileges and coherence properties for reasoning about logical regions.  
+%We prove the soundness of the type system and use it to show
+%that all memory accesses in Legion are safe.  Furthermore, we prove that
+%type soundness enables a scalable, hierarchical
+%scheduling algorithm for Legion programs.  We show that an implementation of the type checker
+%for Legion programs improves performance up to 71\% by eliding
+%memory checks the type system proves safe.  We also demonstrate that an implementation of the Legion 
+%runtime based on the hierarchical scheduling algorithm scales well for
+%three real-world applications.
 

### 2012-10-29 doc/types-pldi2013/intro.tex
some small edits

@@
 systems that are capable of reasoning about the structure of data in
-parallel programs.  Traditionally, these attempts have taken two approaches.   
-The first approach relies on fully static analyses that
+parallel programs.  Traditionally, these attempts have taken one of two approaches.   
+The first approach relies on static analyses that
 have no runtime overhead, but limited expressivity.  Two recent examples, 
@@
 These two different approaches illustrate different mechanisms for dealing with
-the problem of data aliasing in parallel programs.  The fully static approach outlaws aliasing
+the problem of data aliasing in parallel programs.  The fully static approach disallows aliasing
 by requiring a single static partitioning of data.  Our own experience with writing high-performance 
@@
 {\em non-interfering} (have no ordering requirements for correctness), 
-then the any descendant of $t_1$ in the task hierarchy is non-interfering
+then any descendant of $t_1$ in the task hierarchy is non-interfering
 with any descendant of $t_2$.  This theorem is the basis for correct, distributed

### 2012-10-29 doc/types-pldi2013/legion-types.tex
some small edits

@@
 that an implementation of the Legion runtime based on the hierarchical 
-scheduling algorithm scales well for three real-world applications, achieving 
-up to 62.5X speedup on 96 GPUs on the Keeneland supercomputer.
+scheduling algorithm scales well for three real-world applications running
+on distributed memory machines, achieving up to 62.5X speedup on 96 GPUs 
+on the Keeneland supercomputer.
 %Legion is a programming model that introduces

### 2012-10-29 doc/types-pldi2013/intro.tex
Adding some references

@@
 parallel programs.  Traditionally, these attempts have taken one of two approaches.   
-The first approach relies on static analyses that
+The first approach relies on fully static analyses that
 have no runtime overhead, but limited expressivity.  Two recent examples, 
@@
 (see Section~\ref{sec:related} for more discussion of related work).  The
-second approach relies on a dynamic analysis of program data that doesn't
+second approach relies on a totally dynamic analysis of program data that doesn't
 limit expressivity, but incurs runtime overhead.  Examples of this approach 
-include transactional memory and thread-level speculation.
+include transactional memory \cite{Harris05} and thread-level speculation \cite{Steffan00}.
 
-These two different approaches illustrate different mechanisms for dealing with
+These two approaches illustrate different mechanisms for dealing with
 the problem of data aliasing in parallel programs.  The fully static approach disallows aliasing

### 2012-10-29 doc/types-pldi2013/example.tex
More edits to sections 2 3 and 4

--- 2012-07-11 doc/popl2013/example.tex
@@
 on regions.  Coherence specifies what updates the function may
-see from other functions using the same region.  If not otherwise
-specified, coherence defaults to {\em exclusive}, meaning the
-function must appear to execute in program order.  Line 14 shows
+observe from other functions using aliased regions.  
+If not otherwise specified, coherence defaults to {\em exclusive}, 
+meaning the function must appear to execute in program order.  
+Line 14 in Listing~\ref{lst:circuit_leaf} shows
 an example of {\tt atomic}, a relaxed coherence mode requiring
 that operations to {\tt rn} and {\tt rg} appear atomic relative
-to other functions using those regions.  The most relaxed coherence
+to other functions using regions which may alias.  The most relaxed coherence
 mode is {\tt simult}; simultaneous coherence allows concurrent access 
 to the region by all functions that are using the region
-in a simultaneous mode.  The interaction between tasks using the same
+in a simultaneous mode.  The interaction between {\em tasks} (which are functions chosen to execute in parallel) using the same
 region with different coherence

### 2012-10-29 doc/types-pldi2013/langdef.tex
More edits to sections 2 3 and 4

@@
 
-As Legion attempts to benefit from both static and dynamic checks
-to achieve performance while maintaining safety, both the type system and the operational
+To achieve high performance in the presence of aliasing, the Legion type system has 
+been designed to efficiently leverage both static and dynamic checks.
+Consequently, both the type system and the operational
 semantics have been extended in interesting ways from a basic expression language.  We
-will describe these extensions in the next two subsections, and explain how these extensions
-were chosen to preserve a straightforward proof of the soundness of the type system.
+describe these extensions in the next two subsections, and explain how they 
+were chosen to preserve a straightforward proof of soundness for the type system.
 
@@
 which makes use of a new type specific to Legion, a {\em coloring}.  A coloring is an opaque mapping from
-pointers (which much point into the region being colored) to integer ``color'' values.  (For simplicity, 
+pointers (which must point into the region being colored) to integer {\em color} values.  (For simplicity, 
 Core Legion allows only an iterative construction of a coloring via a {\em newcolor} expression that generates
-an empty map and a {\em color} expression which adds a new entry to an existing map, but the opacity of the
+an empty map, and a {\em color} expression which adds a new entry to an existing map. The opacity of the
 coloring allows for more efficient generation (and storage) of colorings as needed.)  The {\em partition} 
-expression introduces names for subregions corresponding to each of those ``colors'', guaranteeing that each
+expression introduces names for subregions corresponding to each color in the coloring, guaranteeing that each
 subregion is disjoint from the other subregions and all are included in the original region.
@@
 Finally, the task call expression handles the aliasing of logical region names between the caller and callee and
-verifies the property that the privileges held by a called task are always a subset of those held
+verifies the functional requirement that the privileges held by a called task are always a subset of those held
 by the caller.  In addition to being necessary for the proof of soundness, this property is critical to
@@
 programs can
-be scheduled in a hierarchical (and therefore scalable) fashion, we require a formulation of operational semantics
+be scheduled in a hierarchical fashion.  We therefore require a formulation of operational semantics
 that is both parallel and hierarchical.  Because Legion allows constrained nondeterminism (in the order in which
-reductions are applied and in cases of relaxed coherence), it is possible for the heap accesses performed by
+reductions are applied and in cases of relaxed coherence with aliasing), it is possible for the heap accesses performed by
 one expression's evaluation to observe interference from other concurrent evaluations.  The traditional way of
@@
 partially-evaluated expressions.  Rules are specified for which expression(s) are valid choices for the next small
-step of execution.  This approach has the benefit of being constructive but the implicit global ``scheduler'' is
+step of execution.  This approach has the benefit of being constructive but the implicit global scheduler is
 difficult to map into a hierarchical one in which decisions must necessarily be made with limited information.
@@
 With this formulation, the criteria for when a hierarchical scheduler may execute subexpressions in parallel can be
-made clear, but they are no longer constructive.  Instead, we will have to show that the scheduling algorithms
-used by the Legion runtime satisfy these constraints, guaranteeing the soundness of the Legion programming model.
-
-The complete operational semantics rules can be found in \cite{LegionTypes12} - we show the most interesting ones
+made clear, but they are no longer constructive.  Instead, we will have to show that the scheduling algorithm
+used by the Legion runtime satisfies these constraints, guaranteeing the soundness of the Legion programming model.
+
+The complete operational semantics rules can be found in \cite{LegionTypes12}; we show the most interesting ones
 in Figure~\ref{fig:semantics}.  The rules for the {\em read}, {\em write}, and {\em reduce} show how dynamic memory
@@
 existing coloring and returns one in which the specified location is given the specified color, replacing an
-existing coloring for that location (if present).  The behavior of the newcolor expression is a bit trickier.
-Since Legion supports the {\em partition-then-allocate} model in which the {\em new} expression may be called
+existing coloring for that location (if present).  The behavior of the newcolor expression is more subtle.
+Since Legion supports the {\em partition-then-allocate} model (see Section~\ref{sec:example}) in which the {\em new} expression may be called
 on subregions, each subregion that results from a partitioning may need an arbitrary number of locations for which
@@
 correct type and does so within the regions of memory for which privileges are held.  In fact, this results holds
-even if the {\em valid\_interleave} constraint is weakened to just require an arbitrary interleaving of the memory
+even if the {\em valid\_interleave} constraint is weakened to only require an arbitrary interleaving of the memory
 traces of subexpressions (i.e. the {\em any\_interleave} predicated defined in Figure~\ref{fig:opsecfns}).  The 
 additional constraints on {\em valid\_interleave} in Figure~\ref{fig:validinterleave} address the coherence of 
-heap accesses and allow for the per-region relaxation of coherence that is introduced in Legion.
+heap accesses and allow for the per-region relaxation of coherence that is introduced in Legion to enable parallelism
+in the presence of aliasing.
 

### 2012-10-29 doc/types-pldi2013/conclusion.tex
more edits

\section{Conclusion}
\label{sect:conclusion}

Modern architectures have dramatically increased in complexity in recent years.  To program this class
of machines, new programming systems will be required that are capable about reasoning about the structure
of data including aliasing.  We have presented a type system for the Legion programming system that
reasons about aliasing using a combination of static and dynamic analyses.  Furthermore, we've proven
the soundness of our type system enables a hierarchical, distributed scheduling algorithm capable
of scaling on large, distribute memory machines.  This result demonstrates that allowing programmers
to provide descriptions of data that include aliasing is not prohibitive to high performance.

%Modern architectures have dramatically increased in complexity in recent years and we believe that new programming models
%are needed to take advantage of the potential of such machines.  We have presented the semantics and type system
%for Legion, which differs from previous efforts in having a much more flexible programming model (in particular,
%regions are first-class values and data can be dynamically partitioned in as many ways as needed) while maintaining
%soundness of privileges and coherence.  The key insight is that we can divide the problem between static and dynamic
%checking differently, keeping what is expensive to check dynamically (region pointer checks) in the static system while incurring very little if any performance cost for the work deferred to runtime (the region aliasing checks).  We have also given
%a formal basis for hierarchical, distributed scheduling in Legion, and provided experimental evidence  that the
%expected benefits of the Legion design are realized in real-world programs.

### 2012-10-29 doc/types-pldi2013/evaluation.tex
more edits

--- 2012-07-11 doc/popl2013/evaluation.tex
@@
 
-We evaluate the Legion type system on three criteria:
+We evaluate the Legion type system on four criteria:
 \begin{itemize}
 \item Expressivity: Can real applications be written (Section~\ref{subsec:expressivity})?
-\item Overhead: Does it reduce checking costs (Section~\ref{subsec:overhead})?
-\item Scalability: Can it leverage hierarchical scheduling (Section~\ref{subsec:scalability})?
+\item Overhead: Does it reduce dynamic checking costs (Section~\ref{subsec:overhead})?
+\item Scalability: Can it enable hierarchical scheduling (Section~\ref{subsec:scalability})?
+\item Performance: Does the performance increase from relaxed coherence modes 
+warrant the increased complexity in the operational semantics (Section~\ref{subsec:relaxedperf})?
 \end{itemize}
@@
 
-These applications show Legion is capable of 
-expressing applications with both regular (Fluid, AMR) and irregular (Circuit) pointer data structures.
+%These applications directly demonstrate the need to for multiple, dynamic partitions
+%of program data.
+%These applications demonstrate Legion is capable of 
+%expressing real applications with both regular (Fluid, AMR) and 
+%irregular (Circuit) pointer data structures.
 Being able to dynamically create and partition regions at runtime is crucial to Legion's
@@
 described are needed in different applications.  Region relationships
-are used in all applications. %, region relationships convey disjointness information.
+are used in all applications.   Finally, all applications introduce aliasing of
+data through both multi-colorings or having multiple partitions.  Our implementations of
+these applications both type check and execute, proving that our type system is
+sufficiently expressive to handle real-world applications written in the Legion programming model.
+%, region relationships convey disjointness information.
 % despite
@@
 \label{subsec:scalability}
-The scalability of Legion derives directly from Theorem~\ref{thm:hiersched}.  This 
-theorem proves it is safe for the runtime to only perform dependence checks between two
-tasks that share the same immediate parent task instead of having to consider all
-possible interactions with other tasks anywhere in the system.  Since all tasks that
-share the same immediate parent are all created on the node where the parent is executing,
-all dependence checks and scheduling decisions can be performed locally with no inter-processor communication.  
-Without this property the runtime would be forced to inform all other processors in the system 
-of every task that it launches using expensive broadcasts.
+%The scalability of Legion derives directly from Theorem~\ref{thm:hiersched}.  This 
+%theorem proves it is safe for the runtime to only perform dependence checks between two
+%tasks that share the same immediate parent task instead of having to consider all
+%possible interactions with other tasks anywhere in the system.  Since all tasks that
+%share the same immediate parent are all created on the node where the parent is executing,
+%all dependence checks and scheduling decisions can be performed locally with no inter-processor communication.  
+%Without this property the runtime would be forced to inform all other processors in the system 
+%of every task that it launches using expensive broadcasts.
 
@@
 the overhead of the Legion runtime is always less than 7\% of the total execution time
-of the applications.  In some applications communication doesn't scale well, but this is
-a result of the algorithm chosen by the programmer, not the Legion runtime.
+of the applications.  In some applications communication does not scale well, but this is
+a result of the algorithm required by the application, not the Legion runtime.
 Even in the case of the Circuit application, which exhibits quadratic communication scaling, the Legion
 runtime is able to achieve a 62.5X speedup on 96 GPUs over a hand-coded single GPU implementation\cite{Legion12}.
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.33]{figs/circuit_coherence.pdf}
+\end{center}
+\vspace{-2mm}
+\caption{Performance gains from relaxed coherence modes.\label{fig:circuit_excl}}
+\vspace{-6mm}
+\end{figure}
+
+\subsection{Performance}
+\label{subsec:relaxedperf}
+
+To demonstrate the benefit of relaxed coherence modes, we modified the circuit example from Section~\ref{sec:example}
+to use exclusive coherence instead of atomic coherence in the {\em distribute\_charge} task and compared 
+the performance of the two versions.
+The results are shown in Figure~\ref{fig:circuit_excl}.  Slow-downs ranged from 34\% on 48-GPUs to 67\% on 
+96 GPUs and more importantly scaled with node count.  This is a direct consequence of Amdahl's Law.  
+Even though the {\em distribute\_charge} tasks are a small fraction of the total work, the serialization 
+that results from requiring exclusive access to the 
+overlapping ghost regions limits the scalability of the application.  Relaxed coherence modes will be 
+imperative for enabling the scalability of applications with aliased data on distributed memory machines.

### 2012-10-29 doc/types-pldi2013/langdef.tex
more edits

@@
 \end{proof}
-} The Legion task scheduler uses Theorem~\ref{thm:hiersched} in the
-following way.  Sibling function calls (those invoked within the same
-function body) are checked for non-interference of their (dynamic)
-privileges.  If they interfere they are serialized; otherwise they are
-considered for execution as parallel subtasks.  Although determining that $e_1$
+} The Legion task scheduler uses Theorem~\ref{thm:hiersched} as follows: 
+sibling function calls (those invoked within the same
+function body) $e_1$ and $e_2$ are checked for non-interference of their (dynamic)
+privileges.  Since $e_1$ and $e_2$ are guaranteed to be created on the node on which their parent
+is running, no communication is required to perform the non-interference
+test.  If they interfere they are executed in program order or serialized
+depending on their coherence specifications; otherwise they are
+considered for execution as parallel subtasks.    If $e_1$ and $e_2$ are 
+determined to be non-interfering and are scheduled in parallel on different 
+remote processors then Theorem~\ref{thm:hiersched}
+guarantees that there is no communication required between $e_1$ and $e_2$
+to perform non-interference tests between their sub-tasks.  Therefore,
+the runtime requires no communication for scheduling.
+
+Although determining that $e_1$
 and $e_2$ are non-interfering guarantees $e_3$ and $e_4$ are

### 2012-10-29 doc/types-pldi2013/related.tex
more edits

--- 2012-07-11 doc/popl2013/related.tex
@@
 programmers to create multiple partitions of the same region to give different 
-views onto the same data, which is not possible in DPJ.  
+views onto the same data, which is not possible in DPJ.  DPJ supports both
+exclusive and atomic tasks which are similar to Legion's coherence modes, but only
+allows specification at the coarser granularity of tasks and not individual regions.
+DPJ does not support any modes analogous to simultaneous coherence.
 %DPJ only supports the 
@@
 privileges system as our primary formalism because separation logic does not easily
-support reasoning about the interleaving of operations to non-disjoint regions of memory.
+support reasoning about the interleaving of operations to aliased regions of memory.
 

### 2012-10-29 doc/types-pldi2013/langdef.tex
checkpoint

@@
 
+\newcommand{\placeat}[3]{%
+\makebox[0pt]{%
+\hspace*{#1in} \raisebox{#2in}{#3}%
+}%
+}
+
+\newcommand{\infrulez}[2]{\displaystyle\frac{\displaystyle\strut{#1}}{\displaystyle\strut {#2}}}
+\newcommand{\cinfrulez}[3]{\parbox{14cm}{\hfil$\infrulez{#1}{#2}$\hfil}\parbox{4cm}{$\,#3$\hfil}}
+\newcommand{\finfrulez}[2]{\framebox{$\infrulez{#1}{#2}$}}
+
+\newcommand{\ruleat}[5]{
+\node[below right] at (#1) {$\infrulez{#4}{#5}$};
+\node[below] at (#2) {[#3]};
+}
+
+\newcommand{\ruleatx}[5]{
+\node[below right] at (#1) {$\infrulez{\begin{array}{l}#4\end{array}}{#5}$};
+\node[below] at (#2) {[#3]};
+}
+
+\newcommand{\axiomat}[4]{
+\node[below right] at (#1) {$#4$};
+\node[below] at (#2) {[#3]};
+}
+
+\begin{figure*}
+{
+\begin{tikzpicture}[x=1in,y={(0,-1in)}]
+%\draw[thin] (0,0) to (7,0);
+%\draw[thin] (0,0) to (0,6);
+
+% start with read/write/reduce rules
+\ruleat{0.1,0.1}{2.5,0.2}{T-Read}
+{\begin{array}{l}
+\typeenvx e_1 : T@(\oton{r}) \\
+\forall i, reads(r_i) \in \Phi^*\end{array}}
+{\typeenvx \text{read}(e_1) : T}
+
+\ruleat{0.1,1.1}{2.5,1.2}{T-Write}
+{\begin{array}{l}
+\typeenvx e_1 : T@(\oton{r}) \\
+\typeenvx e_2 : T \\
+\forall i, writes(r_i) \in \Phi^*
+\end{array}}
+{\typeenvx \text{write}(e_1, e_2) : T@(\oton{r})}
+
+\ruleat{0.1,2.1}{2.5,2.2}{T-Reduce}
+{\begin{array}{l}
+\Gamma(id) = (T_1, T_2), \emptyset, \emptyset \rightarrow T_1 \\
+\typeenvx e_1 : T_1@(\oton{r}) \\
+\typeenvx e_2 : T_2 \\
+\forall i, reduces_{id}(r_i) \in \Phi^*
+\end{array}}
+{\typeenvx \text{reduce}(id, e_1, e_2) : T_1@(\oton{r})}
+
+\axiomat{0.1,2.8}{2.5,2.9}{T-New}
+{\typeenvx \text{new }T@r : T@r}
+
+\ruleat{0.1,3.2}{2.5,3.3}{T-UpRgn}
+{\begin{array}{l}
+\typeenvx e : T@(r'_1, \ldots r'_k) \\
+\forall i. \exists j, r'_i \leq r_j \in \Omega^* \\
+\end{array}}
+{\typeenvx upregion(e,\oton{r}) : T@(\oton{r})}
+
+\ruleat{0.1,3.8}{2.5,3.9}{T-DnRgn}
+{\begin{array}{l}
+\typeenvx e : T@(r'_1, \ldots r'_k) \\
+%\forall j. \exists i, r_j \leq r'_i \in \Omega^* \\
+\end{array}}
+{\typeenvx downregion(e,\oton{r}) : T@(\oton{r})}
+
+\axiomat{0.1,4.1}{2.5,4.2}{T-NewColor}
+{\typeenvx \text{newcolor }r : \text{coloring}(r)}
+
+\ruleat{0.1,4.5}{2.5,4.5}{T-Color}
+{\begin{array}{l}
+\typeenvx e_1 : \text{coloring}(r) \\
+\typeenvx e_2 : T@r \\
+\typeenvx e_3 : int
+\end{array}}
+{\typeenvx \text{color}(e_1, e_2, e_3) : \text{coloring}(r)}
+
+\ruleat{0.1,5.0}{2.5,5.1}{T-Partition}
+{\begin{array}{l}
+\typeenvx e_1 : \text{coloring}(r_p) \\
+\Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{1 \leq i < j \leq k} r_i * r_j \\
+\typeenvx[O=\Omega'] e_2 : T \\
+\{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T) = \emptyset
+\end{array}}
+{\typeenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 : T}
+
+\ruleat{0,5.3}{2.5,5.4}{T-Pack}
+{\begin{array}{l}
+T_1 = \exists r'_1, \ldots r'_n.\ T_2\text{ where }\Omega_1 \\
+\Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^* \\
+\typeenvx e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]
+\end{array}}
+{\typeenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] : T_1}
+
+\ruleat{0.0,6.0}{2.5,6.1}{T-Unpack}
+{\begin{array}{l}
+T_1 = \exists r'_1, \ldots, r'_n.\ T_2\text{ where }\Omega_1 \\
+\typeenvx e_1 : T_1 \\
+\Gamma' = \Gamma[T_2[r_1/r'_1,\ldots,r_k/r'_k] / id] \\
+\Omega' = \Omega \cup \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \\
+\typeenvx[G=\Gamma',O=\Omega'] e_2 : T_3 \\
+\{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T_1, T_3) = \emptyset
+\end{array}}
+{\typeenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 : T_3}
+
+\ruleat{0.0,6.5}{2.5,6.6}{T-Call}
+{\begin{array}{l}
+\Gamma(id) = \forall r'_1, \ldots r'_k.(\oton{T}),\Phi', Q' \rightarrow T_r \\
+\typeenvx e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\
+\Phi'[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*
+\end{array}}
+{\typeenvx id[\otok{k}{r}](\oton{e}) : T_r[r_1/r'_1,\ldots,r_k/r'_k]}
+
+% now the operational semantics versions
+
+\ruleatx{3.0,0.0}{6.5,0.0}{E-Read}{
+\opsenvx e \mapsto l, E \andalso S' = \text{apply}(S, E) \\
+v = \begin{cases} S'(l), & \text{if } l \not\in C \\
+v' : H(l), & \text{otherwise} \end{cases}
+}{\opsenvx \text{read}(e) \mapsto v, E \dplus [ read(l, excl, v) ]}
+
+\ruleatx{3.0,1.0}{6.5,1.1}{E-Write}{
+\opsenvx e_1 \mapsto l, E_1 \andalso S' = \text{apply}(S, E_1) \\
+\opsenvx[S=S'] e_2 \mapsto v, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)
+}{\opsenvx \text{write}(e_1, e_2) \mapsto l, E' \dplus [ write(l, excl, v) ]}
+
+\ruleatx{3.0,1.5}{6.5,1.6}{E-Reduce}{
+\opsenvx e_1 \mapsto l, E_1 \andalso S' = \text{apply}(S, E_1) \\
+\opsenvx[S=S'] e_2 \mapsto v, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)
+}{\opsenvx \text{reduce}(id, e_1, e_2) \mapsto l, E' \dplus [ reduce_{id}(l, excl, v) ]}
+
+\ruleatx{3.0,2.0}{6.5,2.1}{E-New}{
+l \in M(r) \\
+H(l) = M \llbracket T \rrbracket
+}{\opsenvx \text{new }T@r \mapsto l, []}
+
+\ruleat{3.0,3.0}{6.5,3.1}{E-UpRgn}{
+\opsenvx e \mapsto v, E
+}{\opsenvx \text{upregion}(e, \oton{r}) \mapsto v, E}
+
+\ruleatx{3.0,4.0}{6.5,4.1}{E-DnRgn}{
+\opsenvx e \mapsto l, E \\
+v = \begin{cases} l, & \text{if $\exists i, l \in M(r_i)$}. \\ null, & \text{otherwise}. \end{cases}
+}{\opsenvx \text{downregion}(e, \oton{r}) \mapsto v, E}
+
+\ruleatx{3.0,5.0}{6.5,5.1}{E-NewColor}{
+K = \{ (l_1, iv_1), \ldots, (l_p, iv_p) \}, \text{ where } \\
+(\forall i \in [1,p]. l_i \in M(r)) \wedge (\forall i,j \in [1,p]. l_i \not= l_j)
+}{\opsenvx \text{newcolor }r \mapsto K, []}
+
+\ruleatx{3.0,5.5}{6.5,5.6}{E-Color}{
+\opsenvx e_1 \mapsto K, E_1 \andalso S' = \text{apply}(S, E_1) \\
+\opsenvx[S=S'] e_2 \mapsto l, E_2 \andalso S'' = \text{apply}(S', E_2) \\
+\opsenvx[S=S''] e_3 \mapsto v, E_3 \\
+K' = \{ (l,v) \} \cup \{ (l_i,v_i) : (l_i,v_i) \in K \wedge l \not= l_i \} \\
+valid\_interleave(S, C, E', E_1, E_2, E_3)
+}{\opsenvx \text{color}(e_1, e_2, e_3) \mapsto K', E'}
+
+\ruleatx{3.0,6.0}{6.5,6.1}{E-Partition}{
+\opsenv{e_1}{K}{E_1} \andalso \rho_i = \{ l : (l, i) \in K \}, \text{ for } 1 \leq i \leq k \\
+M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \andalso S' = \text{apply}(S, E_1) \\
+\opsenvx[M=M',S=S'] e_2 \mapsto v, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)
+}{\opsenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 \mapsto v, E'}
+
+\ruleatx{3.0,6.5}{6.5,6.6}{E-Pack}{
+\opsenvx e_1 \mapsto v, E \andalso \rho_i = M[r_i], \text{ for } 1 \leq i \leq k \\
+v' = \langle \langle \otok{k}{\rho}, v \rangle \rangle
+}{\opsenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] \mapsto v', E}
+
+\ruleatx{3.0,7.0}{6.5,7.1}{E-Unpack}{
+\opsenvx e_1 \mapsto \langle \langle \otok{k}{\rho} , v_1 \rangle \rangle, E_1 \andalso M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \\
+L' = L[v_1/id] \andalso S' = \text{apply}(S, E_1) \\
+\opsenvx[M=M',L=L',S=S'] e_2 \mapsto v_2, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)
+}{\opsenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 \mapsto v_2, E'}
+
+\ruleatx{3.0,7.5}{6.5,7.6}{E-Call}{
+\opsenvx e_1 \mapsto v_1, E_1 \andalso S_1 = \text{apply}(S, E_1) \andalso \opsenvx e_2 \mapsto v_2, E_2 \andalso \ldots \andalso S_n = \text{apply}(S_{n-1}, E_n) \\
+valid\_interleave(S, C, E', \oton{E}) \andalso \text{function }id[\otok{k}{r'}](a_1 : T_1, \ldots, a_n : T_n), \Phi', Q' : T_r = e_{n+1} \\
+M' = \{ (r'_1, M(r_1)), \ldots (r'_k, M(r_k)) \} \andalso L' = \{ (a_1, v_1), \ldots, (a_n, v_n) \} \andalso S' = \text{apply}(S, E') \\
+C' = C \cup \{ l : \exists \rho. atomic(\rho) \in M' \llbracket Q' \rrbracket \vee simult(\rho) \in M' \llbracket Q' \rrbracket \} \\
+\opsenvx[M=M',L=L',S=S',C=C'] e_{n+1} \mapsto v_{n+1}, E_{n+1} \andalso E'_{n+1} = mark\_coherence(E_{n+1}, M' \llbracket Q' \rrbracket) \andalso valid\_interleave(S, C, E'', E', E_{n+1})
+}{\opsenvx id[\otok{k}{r}](\oton{e}) \mapsto v_{n+1}, E''}
+
+\end{tikzpicture}
+\texcomment{
+\framebox[7in]{
+\placeat{2}{3}{bar}%
+\placeat{3}{1}{3,1}%
+\placeat{-3}{1}{-3,1} %
+\makebox[0pt]{%
+\hspace*{2in} \raisebox{3in}{---}%
+} %
+\makebox[0pt]{%
+\hspace*{1in} \raisebox{1in}{foo}%
+} %
+\makebox[0pt]{%
+\hspace*{2in} \raisebox{3in}{bar}%
+}%
+\rule{0.1pt}{6in}
+}}
+}
+\caption{Type System and Operational Semantics Rules}
+\label{fig:langrules}
+\end{figure*}
+
 %\newcommand{\infrule}[2]{\displaystyle\frac{\displaystyle\strut{#1}}{\displaystyle\strut {#2}}}
@@
 
+\texcomment{
 \begin{figure*}
@@
 \end{figure*}
+}
 
@@
 
-The final piece is to turn this test into something that can be performed before the subtasks are executed, ideally as efficiently as
-possible.  The key to this is the privileges on Legion tasks and their ability to provide bounds on the accesses that may be performed
-during execution.  Our non-interference operator is extended to work on privileges and sets of privileges:
+The final piece is to turn this test into something that can be performed before the subtasks are executed,
+ideally as efficiently as possible.  The key to this is the privileges on Legion tasks and their ability to
+provide bounds on the accesses that may be performed during execution.  Our non-interference operator is
+extended to work on privileges and sets of privileges:
 $
@@
 \end{array}$
-Note although the
-
-CURRENT EDIT POINT
-
-Next we examine how to modify the serialized interleaving to produce valid interleavings allowing parallel execution. 
-We consider when we may swap the order of two adjacent memory operations in a valid interleaving.
-\begin{itemize}
-\item The two operations must have come from different constituent
-traces, or the swap would violate $any\_interleave$.
-\item If one is a read of
-a location $l$ (that is not in the clobber set), the other may not be a write or reduce to $l$, 
-as changing the order would almost certainly change the value read.
-\item If one is a write of a location $l$, the other may not be a write or reduce to $l$,
-as the store that would result from applying those operations in the opposite order may
-differ in $l$.
-\item Finally, if both operations are reductions to the same location, they may be swapped if they
-use the same reduction function, as each reduction function is assumed to commute with itself.  
-%(We 
-%follow the example of other work (i.e. DPJ) and leave proofs of commutativity to the programmer,
-%possibly with the aid of other tools.)  
-However, if the reduction functions have different names, swapping the two operations might result in a different store.
-\end{itemize}
-
-To summarize this, we define a {\em non-interference} operator on two memory operations that have been
-annotated with which constituent trace they came from as follows:
-$$
-\begin{array}{l@{ }l}
-op_1^{E_1}(l_1, c_1,& v_1)\ \nonint\ op_2^{E_2}(l_2, c_2, v_2) \Leftrightarrow E_1 \not= E_2\ \wedge \\
-& \big( l_1 \not= l_2 \vee (op_1 = read \wedge op_2 = read)\ \vee \\
-& (op_1 = reduce_{id_1} \wedge op_2 = reduce_{id_2} \wedge id_1 = id_2) \big)
-\end{array}
-$$
-
-\begin{lem}
-\label{lem:nonintswap}
-\rm
-Let $S$ be a store, $C$ a clobber set, $\oton{E}, E'_a, E'_b$ memory traces, and
-$\epsilon_1, \epsilon_2$ two annotated memory operations such that $\epsilon_1 \nonint \epsilon_2$.
-Then,
-$$valid\_interleave(S, C, E'_a \dplus [ \epsilon_1, \epsilon_2 ] \dplus E'_b, \oton{E})$$
-if and only if
-$$valid\_interleave(S, C, E'_a \dplus [ \epsilon_2, \epsilon_1 ] \dplus E'_b, \oton{E}).$$
-\end{lem}
-\texcomment{
-\begin{proof}
-As before, we consider the three conditions needed to be a valid interleaving.  We describe the
-implication in the ``forward'' direction, but the arguments are all symmetric.
-\begin{enumerate}
-\item If $E'_a \dplus [ \epsilon_1 ] \dplus [ \epsilon_2 ] \dplus E'_b$ is an interleaving of $\oton{E}$,
-the proof must first ``pop off'' all of $E'_a$, leaving $[ \epsilon_1 ] \dplus [ \epsilon_2 ] \dplus E'_b$ for
-the interleaved subtrace.  The next two steps must pop $\epsilon_1$ and $\epsilon_2$, leaving $E'_b$.
-They cannot be in the same subtrace because $\epsilon_1 \nonint \epsilon_2$, so we could also pop
-$\epsilon_2$ and then $\epsilon_1$, resulting in a proof that $E'_a \dplus [ \epsilon_2 ] \dplus [ \epsilon_1 ] \dplus E'_b$ is also an interleaving of the constituent traces.
-\item For coherence, we must show the value read (if any) by $\epsilon_1$ does not change, the value
-read (if any) by $\epsilon_2$ does not change, and that any read in $E'_b$ sees no change.  With the
-help of Lemma~\ref{lemma:applychain}, we can eliminate the common $E'_a$ from all sequences by
-using $S' = apply(S, E'_a)$.
-Similarly to show that $apply(S', [ \epsilon_1  ] \dplus [ \epsilon_2 ] \dplus E) = apply(S', [ \epsilon_2 ] \dplus [ \epsilon_1 ] \dplus E)$ for any $E$, we need only to show that it holds
-for $E = []$.  The proof is now reduced to showing the following:
-\begin{enumerate}
-\item $\epsilon_1 = read(l, c, v) \Rightarrow S'(l) = apply(S',\epsilon_2)(l)$ - since $\epsilon_1$
-is a read, either $\epsilon_2$ must be a read (in which case $apply(S, \epsilon_2) = S'$) or must be
-to a different location, so the result of applying just it to $S'$ cannot change the value in 
-location $l$.
-\item $\epsilon_2 = read(l, c, v) \Rightarrow apply(S', \epsilon_1)(l) = S'(l)$ - this is the same
-as above, with $\epsilon_1$ and $\epsilon_2$ switched.
-\item $apply(S', \epsilon_1 \dplus \epsilon_2) = apply(S', \epsilon_2 \dplus \epsilon_1)$ - we must
-show that the two store operations $apply(\bullet, \epsilon_1)$ and $apply(\bullet, \epsilon_2)$
-commute.  Modifications to two different locations commute, as do two reads (which make no change
-to the store).  Finally, two reductions using the same function to the same location also commute.
-\end{enumerate}
-\item The proof that sequential equivalence is maintained comes down to showing the final store
-is identical after the swap of the two operations, and this is merely a special case of the third
-piece of the coherence proof.
-\end{enumerate}
-\end{proof}
-}
-We now extend the non-interference operator to entire memory traces, providing a way to say that
-no operation in a trace interferes with any operation in another:
-$$E_1 \nonint E_2 \Leftrightarrow \displaystyle\bigwedge_{\epsilon_1 \text{ in } E_1, \epsilon_2 \text{ in } E_2} \epsilon_1 \nonint \epsilon_2$$
-and observe that this property permits arbitrary ordering (i.e. parallel execution) of the 
-corresponding subexpressions.
-
-\begin{lem}
-\label{lem:noninteffects}
-\rm
-Let $S$ be an initial store, $C$ be a clobber set, $\oton{E}$ be memory traces such that $E_i \nonint E_j$
-for every $1 \leq i < j \leq n$.  Then any trace $E'$ satisfying $any\_interleave(E', \oton{E})$
-is a valid interleaving.
-\end{lem}
-\texcomment{
-\begin{proof}
-We will use Theorem~\ref{thm:sequential} and Lemma~\ref{lem:nonintswap} to make repeated swaps
-of adjacent operations to transform $E'$ into $E_1 \dplus \ldots \dplus E_n$.  The symmetry of
-Lemma~\ref{lem:nonintswap} is then used to show that the validity of $E_1 \dplus \ldots \dplus E_n$
-under Theorem~\ref{thm:sequential} implies the validity of $E'$.
-
-The swapping algorithm as follows: Associate with each operation in $E'$ the index of that operation
-in $E_1 \dplus \ldots \dplus E_n$.  Any consecutive pair of operations $\epsilon_1$ and $\epsilon_2$
-for which $index(\epsilon_1) > index(\epsilon_2)$ is a {\em misordered pair}.  For any such pair,
-$\epsilon_1$ and $\epsilon_2$ will be from different constituent traces (if they were from the same
-trace, their misordering would mean that $E'$ was not an interleaving of $\oton{E}$).  Every pair of
-traces is non-interfering, so $\epsilon_1 \nonint \epsilon_2$ and they can be swapped while preserving
-validity of the trace.  At most $n^2/2$ swaps are needed to eliminate all misordered pairs.
-\end{proof}
-}
-We next extend non-interference to privileges.  We 
-use the region mapping $M$ to translate the logical regions named in privileges into sets of
-locations (i.e. physical regions), and define non-interference between individual privileges and
-sets of privileges:
-$$
-\begin{array}{@{}l@{}}
-priv_1(r_1) \nonint[M] priv_2(r_2) \Leftrightarrow (M(r_1) \cap M(r_2) = \emptyset)\ \vee \\
-\hspace{0.35in} (priv_1 = reads \wedge priv_2 = reads)\ \vee \\
-\hspace{0.35in} (priv_1 = reduces_{id_1} \wedge priv_2 = reduces_{id_2} \wedge id_1 = id_2)
-\end{array}
-$$
-%% \begin{center}
-%% \begin{tabular}{llc}
-%% \multicolumn{1}{c}{$\phi_1$} & \multicolumn{1}{c}{$\phi_2$} & $\phi_1 \nonint[M] \phi_2$? \\
-%% \multirow{3}{*}{$reads(r_1)$} & $reads(r_2)$ & true \\
-%% & $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
-%% & $reduces_{id_2}(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
-%% \cline{1-1}
-%% \multirow{3}{*}{$writes(r_1)$} & $reads(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
-%% & $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
-%% & $reduces_{id_2}(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
-%% \cline{1-1}
-%% \multirow{3}{*}{$reduces_{id_1}(r_1)$} & $reads(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
-%% & $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
-%% & $reduces_{id_2}(r_2)$ & $\begin{array}{c}id_1 = id_2 \vee \\ M(r_1) \cap M(r_2) = \emptyset\end{array}$ \\
-%% \end{tabular}
-%% \end{center}
-
-$$\Phi_1 \nonint[M] \Phi_2 \Leftrightarrow \displaystyle\bigwedge_{\phi_1 \in \Phi_1, \phi_2 \in \Phi_2} \phi_1 \nonint[M] \phi_2$$
+The ``two readers'' and ``two reducers'' conditions for memory operation non-interference have exact equivalents
+for regions, which can be performed at compile-time.  The ``different address'' condition is approximated by two
+tests.  The first uses (logical) region disjointness constraints from the type system to statically infer
+non-interference.  The second uses the region mapping $M$ to determine
+the two physical regions and performs a run-time test for overlapping.  Although still a dynamic test, it is performed
+once per pair of regions rather than for for every pair of memory operations, and often need not be performed at all
+(if one of the other three conditions is shown to hold at compile-time).
+
+As region non-interference is an approximation of memory trace non-interference, we must show that it is sound.  We
+state the Lemma here, and give the proof in \cite{LegionTypes12}.
 
@@
 \end{lem}
-\texcomment{
-\begin{proof}
-The proof is by contradiction.
-Assume $E_1 \cancel\nonint E_2$.
-Then there must be some $\epsilon_1$ in $E_1$ and some $\epsilon_2$ in $E_2$ such that $\epsilon_1 \cancel\nonint \epsilon_2$.
-This requires that
-$\epsilon_1$ and $\epsilon_2$ be operations to the same location $l$ and not both be reads or both
-reductions using the same function name.
-Assume $\epsilon_1 = read(l, c_1, v_1)$ and $\epsilon_2 = write(l, c_2, v_2)$.
-The definition of $\privconsist{E_1}{\Phi_1}$ tells us that there must be some $r_1$ satisfying $l \in M(r_1) \wedge reads(r_1) \in \Phi_1$.
-Similarly, there must be some $r_2$ satisfying $l \in M(r_2) \wedge writes(r_2) \in \Phi_2$.
-Letting $\phi_1 = reads(r_1)$ and $\phi_2 = writes(r_2)$, we have $M(r_1) \cap M(r_2) \not= \emptyset$ (it contains at least $l$), so $\phi_1 \cancel{\nonint[M]} \phi_2$ and $\Phi_1 \cancel{\nonint[M]} \Phi_2$.
-There are seven other cases to consider (using $id_1 \not= id_2$ for the reduce-reduce case), but all yield the same result.
-\end{proof}
-}
-We can now state the key result that allows the Legion runtime to safely evaluate subexpressions as
-separate tasks in parallel.
+
+We are now able to put all the pieces together and state the theorem that allows the Legion runtime to perform hierarchical
+and parallel scheduling of non-interfering tasks:
 

### 2012-10-29 doc/types-pldi2013/langdef.tex
relaxed coherence modes

@@
 
-\subsection{Relaxed Coherence}
-\label{subsec:relaxed}
-
-The analysis above does not consider the coherence modes of operations,
-making the worst-case assumption that all operations use $excl$ mode.  However, if two or more
-subexpressions contain calls to functions using $atomic$ or $simult$ coherence modes
-on some regions, executions besides serial program order may 
-result in valid interleavings even if constituent traces have locations in common.  We discuss
-two cases of interest.  In both, the previous section's analysis 
-is repeated with minor modifications, so for brevity we limit
-discussion to a description of the differences rather than complete proofs.
-
-The first important case occurs when all apparently-interfering operations have coherence modes
-of either $atomic$ or $simult$.  In this case, the conditions required for
-coherence and sequential equivalence are trivially satisfied for locations in $L_{excl}$.  The
-requirements for locations in $L_{atomic}$ allow any permutation of the constituent traces,
-so the runtime is free to serialize the execution of the subexpressions in any order, not just
-program order.
-
-To convert the required analysis to one that can be performed on regions, we modify
-the non-interference operator for memory operations to consider only locations in $L_{excl}$ and show that
-traces that are non-interfering under this modified check satisfy the requirement for atomic
-execution.  We then modify the non-interference operator on privileges to allow 
-$M(r_1) \cap M(r_2) \not= \emptyset$ if either $atomic(r_1)$ or $simult(r_1)$ is in $Q_1$ and
-either $atomic(r_2)$ or $simult(r_2)$ is in $Q_2$. We then show that subexpressions with 
-privileges that are {\em atomically non-interfering} (under a dynamic region mapping
-$M$) always have memory traces compatible with atomically-reordered execution.  Finally,
-in the special case where each subexpression has $atomic(r)$ or $simult(r)$ in $Q_i$ for every $r$
-in $\Phi_i$, we can eliminate the dynamic check entirely.
-
-The second case is a subset of the first, in which all apparently-interfering operations have
-the $simult$ coherence mode.  In this case, all the coherence and sequential equivalence 
-checks are trivially satisfied, and any interleaving of the constituent traces is valid.
-We again modify the non-interference criteria for memory operations, traces, and privileges, this time
-only allowing $simult$ coherence annotations. We show that subexpression with privileges
-that are {\em simultaneously non-interfering} (under mapping $M$) can be executed
-in parallel, which includes all the cases covered by Theorem~\ref{thm:parallelexec} as well
-as those where the programmer has used the $simult$ coherence mode to explicitly allow parallel
-execution.  Again, a special (but important) case exists where all privileges for each
-subexpression are paired with $simult$ coherence mode declarations, allowing the safety of
-parallel execution to be determined at compile time.
+\subsection{Atomic Coherence}
+\label{subsec:coherenceatomic}
+
+In cases where the Legion compiler and runtime cannot safely infer non-interference of
+privileges (either because the programmer did not partition the regions enough or because the two
+tasks really do access the same location), the fallback behavior of serialized execution can
+be avoided through the use of relaxed coherence.
+
+The first form of relaxed coherence is the {\em atomic} mode, which specifies that although two
+tasks accessing the same region cannot run concurrently, they may execute in either order, allowing
+a ``later'' task to run as soon as its ready, even if the ``earlier'' task is not ready.  This
+relaxation only applies if all regions of overlap are designated as atomic coherence.
+
+To show that this is safe, we define a relaxed version of non-interference for atomic coherence as
+follows:
+$$
+\begin{array}{l@{}l}
+op_1(l_1, c_1, v_1)\ \nonint^A\ op_2(l_2, c_2, v_2) \Leftrightarrow &op_1(l_1, c_1, v_1)\ \nonint^A\ op_2(l_2, c_2, v_2) \vee \\
+& (c_1 = atomic \wedge c_2 = atomic)
+\end{array}$$
+
+We repeat the steps in Section~\ref{subsec:coherenceparallel} using the $\nonint^A$ operator and reach another
+result used by the Legion runtime scheduler:
+
+\begin{thm}
+\label{thm:atomicexec}
+\rm
+Let $\oton{e}$ be well-typed Legion expressions, each with its own privileges $\Phi_i$.
+Let $M$ be a region mapping, $L$ a local value mapping,
+$H$ a heap typing, and $S$ be an initial store satisfying
+$\mapconsist{\Omega}$, $\localconsist{L}{\Gamma}$, and $\storeconsist{S}{H}$.
+If $\Phi_i \nonint[M]^A \Phi_j$ for $1 \leq i < j \leq n$, then for any permuatation $(\oton{\pi})$ of
+$(1, \ldots, n)$, $E_{\pi_1} \dplus \ldots \dplus E_{\pi_n}$ is a valid interleaving.
+\end{thm}
+
+\subsection{Simultaneous Coherence}
+\label{subsec:coherencesimult}
+
+The other way in which coherence can be relaxed is using the {\em simult} mode, which allows
+multiple tasks to access the same region concurrently.  The simult coherence mode is appropriate in two
+important cases:
+\begin{enumerate}
+\item When subtasks are accessing disjoint data, but the disjointness is hard to describe (e.g. walking
+separate linked lists that have been allocated in the same region)
+\item When the algorithm is tolerant of non-determinism (e.g. in a breadth-first search, setting the parent
+pointer of a node with multiple equally-short paths to the root).
+\end{enumerate}
+
+When one or more regions have been marked as using the simult coherence mode, the non-interference operators from
+the non-interference test is relaxed analagously to the atomic coherence case, yielding a $\nonint^S$ operator.
+Because the rules for valid interleavings exclude locations that are only accessed in simult mode, it is straightforward
+to extend Theorem~\ref{thm:parallelexec} to show that parallel execution is safe as long as $\Phi_i \nonint[M]^S \Phi_j$.
+
+It is also possible to have both atomic and simult coherence modes in use at the same time (although for different regions),
+in which case the non-interference test $\nonint^{AS}$ uses both the atomic and simult relaxations, and
+Theorem~\ref{thm:atomicexec} is extended to allow arbitrary reordering (but not simultaneous execution) of subtasks when
+$\Phi_i \nonint[M]^{AS} \Phi_j$.
 

### 2012-10-30 doc/types-pldi2013/intro.tex
minor edits

@@
 the same data, providing different views on to that data.  Alternatively, 
-the fully dynamic approach enables arbitrary data aliasing, but requires 
+the fully dynamic approach permits arbitrary data aliasing, but requires 
 centralized control logic to perform conflict detection between every pair 
@@
 runtime checks and results in low actual runtime overhead.  We then show 
-that the soundness of our type system enables a scalable, hierarchical 
+that the soundness of our type system permits a scalable, hierarchical 
 distributed scheduling algorithm that enables Legion programs to achieve
@@
 that may be operated on in parallel).  Logical regions have the following
-properties that enable programmers to dynamically communicate information
+properties that allows programmers to dynamically communicate information
 about the structure of program data to Legion:

### 2012-10-30 doc/types-pldi2013/intro.tex
More edits

@@
 To enforce this restriction, our type system relies on a combination of static 
-and dynamic analyses.  The key observation behind our type system is that dynamic
+and dynamic analyses.  The key observation is that dynamic
 analysis is very easy and relatively cheap when done at the granularity of
@@
 simultaneously with the programmer using separate synchronization mechanisms
-to control access (e.g. locks).  To support these relaxed access modes
+to control access (e.g. locks).  To support these relaxed access paradigms 
 Legion provides {\em cohrence modes} on logical regions.  For each region

### 2012-10-30 doc/types-pldi2013/langdef.tex
minor edits
untangled the rules figure
More edits

@@
 $}
-\caption{Operational Semantics Helper Functions}
-\label{fig:opsemfns}
+\caption{Operational Semantics Helper Functions \label{fig:opsemfns}}
 \vspace{-5mm}
@@
 
-Theorem~\ref{thm:effects} guarantees that a function call produces a value of the
-correct type and does so within the regions of memory for which privileges are held.  In fact, this results holds
-even if the {\em valid\_interleave} constraint is weakened to only require an arbitrary interleaving of the memory
-traces of subexpressions (i.e. the {\em any\_interleave} predicated defined in Figure~\ref{fig:opsecfns}).  The 
-additional constraints on {\em valid\_interleave} in Figure~\ref{fig:validinterleave} address the coherence of 
-heap accesses and allow for the per-region relaxation of coherence that is introduced in Legion to enable parallelism
+An important insight in the proof of Theorem~\ref{thm:effects} was that 
+it did not rely on the definition of {\em valid\_interleave}.  Weakening 
+{\em valid\_interleave} to {\em any\_interleave} (Figure~\ref{fig:opsemfns}), 
+which only requires any arbitrary interleaving of memory
+traces from subexpressions, 
+would have been a sufficient condition for the proof.  
+The additional constraints in {\em valid\_interleave}
+(Figure~\ref{fig:validinterleave}) address the {\em coherence} of heap accesses.
+Coherence defines the permissible ordering of operations to each individual heap 
+location by parallel computations.  The constraints enforced by {\em valid\_interleave}
+specify exactly what interleavings of memory traces are permitted based on
+the per-region coherence modes introduced by Legion for supporting parallelism
 in the presence of aliasing.
@@
 \begin{itemize}
-\item exclusive locations ($l \in L_{excl}$) are those which have at least once access in exclusive mode in the traces
+\item exclusive locations ($l \in L_{excl}$) are those which have at least one access in exclusive mode in the traces
 and are not in the clobber set.  For these locations, we require the standard {\em sequential consistency} - all
 reads to these locations see the effect of previous writes and reductions, and the resulting state of the store is
-as if all writes and reductions were applied from each trace in order
+as if all writes and reductions were applied from each trace in order.
 \item atomic locations ($l \in L_{atomic}$) are those which have at least one access in atomic mode in the traces and
@@
 Serial execution is achieved by ignoring the coherence mode $Q$ in all function calls, using $Q' = \emptyset$
-instead, and by forming ``interleaved'' traces by simply concatenating the subexpressions' traces.  By ignoring
+instead, and simply interleaving traces by concatenating the subexpressions' traces in program order.  By ignoring
 the coherence modes, the clobber set remains empty, making the result of all {\em read} expressions fully
@@
 To determine when parallel execution is safe, we start from the serial execution trace and allow the reordering of 
-adjacent heap operations that do not change the behavior of the application.  If we can show that reordering is allowed
-between any two operations that come from different constituent traces, any interleaving of the constituent traces will
-be equivalent to the serialized execution and parallel execution was safe.  Finally, to allow the safety of parallel
-execution to be determined {\em before} the execution is actually performed, we show that a test based on the subtask
-privileges and the current region mapping can soundly predict when this property will hold.
+adjacent heap operations that do not change the behavior of the application.  If we can show that it is safe to 
+reorder any pair of operations that come from two different constituent traces, then any interleaving of the constituent
+traces will be equivalent to a sequential execution, thereby enabling parallel execution.  To
+efficiently implement this requirement at runtime, we require a test that can determine this property prior
+to the actual execution of the tasks that create the traces.  We show that a test based on the subtask privileges and current region 
+mapping can soundly predict when this property will hold.
+%If we can show that reordering is allowed
+%between any two operations that come from different constituent traces, any interleaving of the constituent traces will
+%be equivalent to the serialized execution and parallel execution was safe.  Finally, to allow the safety of parallel
+%execution to be determined {\em before} the execution is actually performed, we show that a test based on the subtask
+%privileges and the current region mapping can soundly predict when this property will hold.
 
@@
 The ``two readers'' and ``two reducers'' conditions for memory operation non-interference have exact equivalents
-for regions, which can be performed at compile-time.  The ``different address'' condition is approximated by two
+for regions, which can be performed statically.  The ``different address'' condition is approximated by two
 tests.  The first uses (logical) region disjointness constraints from the type system to statically infer
@@
 the two physical regions and performs a run-time test for overlapping.  Although still a dynamic test, it is performed
-once per pair of regions rather than for for every pair of memory operations, and often need not be performed at all
-(if one of the other three conditions is shown to hold at compile-time).
+once per pair of regions rather than for for every pair of memory operations.  Details on how to perform
+this test efficiently at runtime can be found in \cite{Legion12}.
+%and often need not be performed at all
+%(if one of the other three conditions is shown to hold at compile-time).
 

### 2012-10-30 doc/types-pldi2013/langdef.tex
More edits to the coherence section

@@
 \newcommand{\ruleatx}[5]{
-\node[below right] at (#1) {$\infrulez{\begin{array}{l}#4\end{array}}{#5}$};
+\node[below right,fill=white!50] at (#1) {$\infrulez{\begin{array}{l}#4\end{array}}{#5}$};
 \node[below] at (#2) {[#3]};
@@
 %\draw[thin] (0,0) to (0,6);
+\foreach \y in {0,...,80}{
+%  \draw[thin] (0,0.1*\y) to (7,0.1*\y);
+}
+\foreach \y in {0,...,16}{
+%  \draw[red,thick] (0,0.5*\y) to (7,0.5*\y);
+}
+\foreach \y in {0,...,8}{
+%  \draw[blue,very thick] (0,\y) to (7,\y);
+}
+
+\draw[very thin] (3.0,0.1) -- (3.0,7.6);
 
 % start with read/write/reduce rules
-\ruleat{0.1,0.1}{2.5,0.2}{T-Read}
+\ruleat{0.0,0.0}{2.6,0.2}{T-Read}
 {\begin{array}{l}
@@
 
-\ruleat{0.1,1.1}{2.5,1.2}{T-Write}
+\ruleat{0.0,0.6}{2.6,0.8}{T-Write}
 {\begin{array}{l}
@@
 
-\ruleat{0.1,2.1}{2.5,2.2}{T-Reduce}
+\ruleat{0.0,1.3}{2.6,1.5}{T-Reduce}
 {\begin{array}{l}
@@
 
-\axiomat{0.1,2.8}{2.5,2.9}{T-New}
+\axiomat{0.0,2.2}{2.6,2.2}{T-New}
 {\typeenvx \text{new }T@r : T@r}
 
-\ruleat{0.1,3.2}{2.5,3.3}{T-UpRgn}
+\ruleat{0.0,2.5}{2.6,2.55}{T-UpRgn}
 {\begin{array}{l}
@@
 
-\ruleat{0.1,3.8}{2.5,3.9}{T-DnRgn}
+\ruleat{0.0,3.1}{2.6,3.05}{T-DnRgn}
 {\begin{array}{l}
@@
 
-\axiomat{0.1,4.1}{2.5,4.2}{T-NewColor}
+\axiomat{0.0,3.6}{2.6,3.6}{T-NewColor}
 {\typeenvx \text{newcolor }r : \text{coloring}(r)}
 
-\ruleat{0.1,4.5}{2.5,4.5}{T-Color}
+\ruleat{0.0,3.9}{2.6,4.1}{T-Color}
 {\begin{array}{l}
@@
 
-\ruleat{0.1,5.0}{2.5,5.1}{T-Partition}
+\ruleat{0.0,4.6}{2.6,4.55}{T-Partition}
 {\begin{array}{l}
@@
 
-\ruleat{0,5.3}{2.5,5.4}{T-Pack}
+\ruleat{0,5.5}{2.6,5.7}{T-Pack}
 {\begin{array}{l}
@@
 
-\ruleat{0.0,6.0}{2.5,6.1}{T-Unpack}
+\ruleat{0.0,6.2}{2.6,6.4}{T-Unpack}
 {\begin{array}{l}
@@
 
-\ruleat{0.0,6.5}{2.5,6.6}{T-Call}
+\ruleat{0.0,7.4}{2.6,7.6}{T-Call}
 {\begin{array}{l}
@@
 
-\ruleatx{3.0,0.0}{6.5,0.0}{E-Read}{
+\ruleatx{3.1,0.0}{6.7,0.2}{E-Read}{
 \opsenvx e \mapsto l, E \andalso S' = \text{apply}(S, E) \\
@@
 
-\ruleatx{3.0,1.0}{6.5,1.1}{E-Write}{
+\ruleatx{3.1,0.9}{6.7,0.85}{E-Write}{
 \opsenvx e_1 \mapsto l, E_1 \andalso S' = \text{apply}(S, E_1) \\
@@
 
-\ruleatx{3.0,1.5}{6.5,1.6}{E-Reduce}{
+\ruleatx{3.1,1.5}{6.7,1.45}{E-Reduce}{
 \opsenvx e_1 \mapsto l, E_1 \andalso S' = \text{apply}(S, E_1) \\
@@
 
-\ruleatx{3.0,2.0}{6.5,2.1}{E-New}{
-l \in M(r) \\
+\ruleatx{3.1,2.1}{6.7,2.2}{E-New}{
+l \in M(r) \andalso
 H(l) = M \llbracket T \rrbracket
@@
 
-\ruleat{3.0,3.0}{6.5,3.1}{E-UpRgn}{
+\ruleat{3.1,2.6}{6.7,2.7}{E-UpRgn}{
 \opsenvx e \mapsto v, E
@@
 
-\ruleatx{3.0,4.0}{6.5,4.1}{E-DnRgn}{
+\ruleatx{3.1,3.1}{6.7,3.3}{E-DnRgn}{
 \opsenvx e \mapsto l, E \\
@@
 
-\ruleatx{3.0,5.0}{6.5,5.1}{E-NewColor}{
+\ruleatx{3.1,4.0}{6.7,4.2}{E-NewColor}{
 K = \{ (l_1, iv_1), \ldots, (l_p, iv_p) \}, \text{ where } \\
@@
 
-\ruleatx{3.0,5.5}{6.5,5.6}{E-Color}{
+\ruleatx{3.1,4.7}{6.7,4.9}{E-Color}{
 \opsenvx e_1 \mapsto K, E_1 \andalso S' = \text{apply}(S, E_1) \\
@@
 
-\ruleatx{3.0,6.0}{6.5,6.1}{E-Partition}{
+\ruleatx{3.1,5.7}{6.7,5.55}{E-Partition}{
 \opsenv{e_1}{K}{E_1} \andalso \rho_i = \{ l : (l, i) \in K \}, \text{ for } 1 \leq i \leq k \\
@@
 
-\ruleatx{3.0,6.5}{6.5,6.6}{E-Pack}{
+\ruleatx{3.1,6.5}{6.7,6.6}{E-Pack}{
 \opsenvx e_1 \mapsto v, E \andalso \rho_i = M[r_i], \text{ for } 1 \leq i \leq k \\
@@
 
-\ruleatx{3.0,7.0}{6.5,7.1}{E-Unpack}{
+\ruleatx{3.1,7.1}{6.7,6.95}{E-Unpack}{
 \opsenvx e_1 \mapsto \langle \langle \otok{k}{\rho} , v_1 \rangle \rangle, E_1 \andalso M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \\
@@
 
-\ruleatx{3.0,7.5}{6.5,7.6}{E-Call}{
+\ruleatx{0.0,8.1}{6.7,8.4}{E-Call}{
 \opsenvx e_1 \mapsto v_1, E_1 \andalso S_1 = \text{apply}(S, E_1) \andalso \opsenvx e_2 \mapsto v_2, E_2 \andalso \ldots \andalso S_n = \text{apply}(S_{n-1}, E_n) \\
@@
 \end{tikzpicture}
+\vspace*{-0.5cm}
 \texcomment{
@@
 
-An important insight in the proof of Theorem~\ref{thm:effects} was that 
-it did not rely on the definition of {\em valid\_interleave}.  Weakening 
+An important insight from the proof of Theorem~\ref{thm:effects} is that 
+it does not rely on the definition of {\em valid\_interleave}.  Weakening 
 {\em valid\_interleave} to {\em any\_interleave} (Figure~\ref{fig:opsemfns}), 
@@
 
-Although it is our goal to enable parallel execution whenever possible, it will help to briefly
-consider serialized execution of a Legion program.  We will show that a serial execution trivially
-satisfies the interleaving criteria required by the Legion operational semantics.  In addition to
-serial execution being the fallback behavior of the hierarchical schedule, our proof of the
-soundness of the parallel scheduling will depend on this result.
+Although it is our goal to enable parallel execution whenever possible, we briefly
+consider sequential execution of a Legion program.  We show that a serial execution trivially
+satisfies the interleaving criteria required by the Legion operational semantics.  Our proof
+of the soundness of parallel scheduling depends on this result.
+%In addition to
+%sequential execution being the default behavior of the hierarchical scheduler, our proof of the
+%soundness of the parallel scheduling will depend on this result.
 
@@
 determined.  The following lemma and theorem (proofs of which can be found in \cite{LegionTypes12}) show that the
-value and memory trace that result from a serialized execution of an expression are always valid executions of
+value and memory trace that result from a sequential execution of an expression are always valid executions of
 Legion programs.
@@
 reorder any pair of operations that come from two different constituent traces, then any interleaving of the constituent
-traces will be equivalent to a sequential execution, thereby enabling parallel execution.  To
-efficiently implement this requirement at runtime, we require a test that can determine this property prior
-to the actual execution of the tasks that create the traces.  We show that a test based on the subtask privileges and current region 
-mapping can soundly predict when this property will hold.
+traces will be equivalent to a sequential execution and parallel execution is safe.  To
+efficiently discover these cases at runtime, we require a test that can determine this property prior
+to the actual execution of the tasks that create the traces.  We show that a test based on the subtask privileges and 
+the current region mapping can soundly predict when this property will hold.
 %If we can show that reordering is allowed
@@
 
-The final piece is to turn this test into something that can be performed before the subtasks are executed,
-ideally as efficiently as possible.  The key to this is the privileges on Legion tasks and their ability to
+We now show that we can turn this result into an efficient runtime test that can be performed before 
+the subtasks are executed.  The key to this is the privileges on Legion tasks and their ability to
 provide bounds on the accesses that may be performed during execution.  Our non-interference operator is
@@
 \end{array}$
-The ``two readers'' and ``two reducers'' conditions for memory operation non-interference have exact equivalents
-for regions, which can be performed statically.  The ``different address'' condition is approximated by two
-tests.  The first uses (logical) region disjointness constraints from the type system to statically infer
-non-interference.  The second uses the region mapping $M$ to determine
-the two physical regions and performs a run-time test for overlapping.  Although still a dynamic test, it is performed
+
+The cases where both subtasks have read-only privileges or both subtasks have reduce-only privileges have exact equivalents
+for regions, which can be tested statically.  Detecting the case where the two sets of memory addresses are disjoint is 
+approximated by twotests.  The first uses (logical) region disjointness constraints from the type system to 
+statically infer non-interference.  The second uses the region mapping $M$ to determine
+the two physical regions and performs a run-time test for disjointness.  Although still a dynamic test, it is performed
 once per pair of regions rather than for for every pair of memory operations.  Details on how to perform
@@
 
-We are now able to put all the pieces together and state the theorem that allows the Legion runtime to perform hierarchical
+%We are now able to put all the pieces together and 
+We now state the theorem that allows the Legion runtime to perform hierarchical
 and parallel scheduling of non-interfering tasks:
@@
 and \ref{lem:nonintpriv}.  This result holds even if the
-clobber set $C$ is non-empty, allowing two ``locally independent''
+clobber set $C$ is non-empty, allowing two locally independent
 subtasks to run in parallel even if one or both of them is interacting
@@
 implementation that is different from other systems and relies on the
-soundness of privileges.  Non-interference of memory operations is
-something that can only be determined after evaluation of an
-expression is completed, and only at great expense, as illustrated at
-one extreme by work on transactional
-memory \cite{Harris05}.  At the other extreme are systems
-that check non-interference statically, such as Jade \cite{Rinard98} and
-DPJ \cite{Bocchino11}.  In contrast, Legion checks non-interference of privileges
+soundness of privileges.  Dynamic non-interference of 
+memory operations is something that can only be determined after 
+evaluation of an expression is completed, and only at great expense, 
+as illustrated by work on transactional
+memory \cite{Harris05}.  However, many systems are willing to pay this cost to
+extract parallelism from operations that access aliased data. At the other extreme are systems like 
+Jade \cite{Rinard98} and DPJ \cite{Bocchino11} that check non-interference 
+statically by disallowing aliasing.  In contrast, Legion checks non-interference of privileges
 at runtime, which is much simpler and more efficient than checking
 non-interference of dynamic memory traces.  Even though the privileges
-themselves are static, the region mapping $M$ is dynamic.  Testing
-non-interference on the privileges with physical regions allows
-parallel execution in many more cases than a purely static analysis
-can achieve.
+themselves are static, the region mapping $M$ is dynamic.  Dynamically testing non-interference 
+on the privileges of physical regions allows parallel execution in many more cases than a purely 
+static analysis can achieve in the presence of aliasing.
 
@@
 
-In cases where the Legion compiler and runtime cannot safely infer non-interference of
-privileges (either because the programmer did not partition the regions enough or because the two
-tasks really do access the same location), the fallback behavior of serialized execution can
-be avoided through the use of relaxed coherence.
+In cases where Legion cannot safely infer non-interference of
+privileges, either because the programmer did not sufficiently partition the regions to expose disjointness or because the two
+tasks actually access the same data in aliased regions, parallel execution can
+still be requested by the programmer through the use of relaxed coherence 
+annotations on individual regions passed to a task.
+%the default behavior of sequential execution can
+%be avoided through the use of relaxed coherence.
 
 The first form of relaxed coherence is the {\em atomic} mode, which specifies that although two
-tasks accessing the same region cannot run concurrently, they may execute in either order, allowing
-a ``later'' task to run as soon as its ready, even if the ``earlier'' task is not ready.  This
-relaxation only applies if all regions of overlap are designated as atomic coherence.
+tasks accessing aliased regions cannot run concurrently, they may execute in either order, allowing
+the task issued later in program order to run as soon as its ready, even before the task issued earlier 
+in program order.  This relaxation only applies if all aliased regions are annotated with atomic coherence.
+%or a more relaxed coherence mode (see Section~\ref{subsec:coherencesimult}).
 
@@
 
-When one or more regions have been marked as using the simult coherence mode, the non-interference operators from
-the non-interference test is relaxed analagously to the atomic coherence case, yielding a $\nonint^S$ operator.
+To support the {\em simult} coherence mode, the non-interference test is extended 
+%When one or more regions have been marked as using the simult coherence mode, 
+%the non-interference test is relaxed analagously to the atomic coherence case, yielding a $\nonint^S$ operator.
+with a $\nonint^S$ operator analogous to the one for the atomic coherence case.
 Because the rules for valid interleavings exclude locations that are only accessed in simult mode, it is straightforward
@@
 
-It is also possible to have both atomic and simult coherence modes in use at the same time (although for different regions),
-in which case the non-interference test $\nonint^{AS}$ uses both the atomic and simult relaxations, and
+It is also possible to have both atomic and simult coherence modes in use at the same time for different regions in a task call.
+In this case the non-interference test $\nonint^{AS}$ uses both the atomic and simult relaxations, and
 Theorem~\ref{thm:atomicexec} is extended to allow arbitrary reordering (but not simultaneous execution) of subtasks when

### 2012-10-30 doc/types-pldi2013/langdef.tex
More updates to the coherence section

@@
 
-\subsection{Serial Execution}
-\label{subsec:coherenceserial}
+\subsection{Sequential Execution}
+\label{subsec:coherenceseq}
 
 Although it is our goal to enable parallel execution whenever possible, we briefly
-consider sequential execution of a Legion program.  We show that a serial execution trivially
+consider sequential execution of a Legion program.  We show that a sequential execution trivially
 satisfies the interleaving criteria required by the Legion operational semantics.  Our proof
@@
 
-Serial execution is achieved by ignoring the coherence mode $Q$ in all function calls, using $Q' = \emptyset$
+Sequential execution is achieved by ignoring the coherence mode $Q$ in all function calls, using $Q' = \emptyset$
 instead, and simply interleaving traces by concatenating the subexpressions' traces in program order.  By ignoring
@@
 
-To determine when parallel execution is safe, we start from the serial execution trace and allow the reordering of 
+To determine when parallel execution is safe, we start from the sequential execution trace and allow the reordering of 
 adjacent heap operations that do not change the behavior of the application.  If we can show that it is safe to 
@@
 
-If whole memory traces are non-interfering, then any interleaving of them can be sorted via pairwise swaps to match the serial memory
-trace and we get the result that shows parallel execution is safe:
+If whole memory traces are non-interfering, then any interleaving of them can be sorted via 
+pairwise swaps to match the sequential memory
+trace. This gives us the following result which permits safe parallel execution:
 
@@
 
-We now show that we can turn this result into an efficient runtime test that can be performed before 
-the subtasks are executed.  The key to this is the privileges on Legion tasks and their ability to
+We now show that we can turn this result into an efficient runtime test for non-interface that can be 
+performed task execution.  The key to this test is the privileges on Legion tasks and their ability to
 provide bounds on the accesses that may be performed during execution.  Our non-interference operator is
 extended to work on privileges and sets of privileges:
+
 $
@@
 for regions, which can be tested statically.  Detecting the case where the two sets of memory addresses are disjoint is 
-approximated by twotests.  The first uses (logical) region disjointness constraints from the type system to 
-statically infer non-interference.  The second uses the region mapping $M$ to determine
-the two physical regions and performs a run-time test for disjointness.  Although still a dynamic test, it is performed
+approximated by two tests.  The first uses (logical) region disjointness constraints from the type system to 
+statically infer non-interference.  The second uses the region mapping $M$ to dynamically determine the
+disjointness of the two physical regions.  Although still a dynamic test, it is performed
 once per pair of regions rather than for for every pair of memory operations.  Details on how to perform
@@
 The first form of relaxed coherence is the {\em atomic} mode, which specifies that although two
-tasks accessing aliased regions cannot run concurrently, they may execute in either order, allowing
+tasks access aliased regions cannot run concurrently, they may execute in either order, allowing
 the task issued later in program order to run as soon as its ready, even before the task issued earlier 

### 2012-10-30 doc/types-pldi2013/langdef.tex
more edits to coherence

@@
 \item exclusive locations ($l \in L_{excl}$) are those which have at least one access in exclusive mode in the traces
-and are not in the clobber set.  For these locations, we require the standard {\em sequential consistency} - all
+and are not in the clobber set.  For these locations, we require sequential execution semantics - all
 reads to these locations see the effect of previous writes and reductions, and the resulting state of the store is
@@
 \item for locations that have only access in {\em simult} mode or are in the clobber set, no constraints are
-enforced.  The validity of the interleaving of these accesses will be performed in the an enclosing expression.
+enforced.  The validity of the interleaving of these accesses will be performed in the enclosing expression.
 \end{itemize}
@@
 statically infer non-interference.  The second uses the region mapping $M$ to dynamically determine the
-disjointness of the two physical regions.  Although still a dynamic test, it is performed
-once per pair of regions rather than for for every pair of memory operations.  Details on how to perform
-this test efficiently at runtime can be found in \cite{Legion12}.
+disjointness of the two regions.  Although still a dynamic test, it is performed
+once per pair of regions rather than for every pair of memory operations.  Details on how to perform
+the dynamic test efficiently at runtime can be found in \cite{Legion12}.
 %and often need not be performed at all
@@
 The first form of relaxed coherence is the {\em atomic} mode, which specifies that although two
-tasks access aliased regions cannot run concurrently, they may execute in either order, allowing
+tasks access interfere due to accessing aliased regions, they may execute in either order, allowing
 the task issued later in program order to run as soon as its ready, even before the task issued earlier 
@@
 \begin{array}{l@{}l}
-op_1(l_1, c_1, v_1)\ \nonint^A\ op_2(l_2, c_2, v_2) \Leftrightarrow &op_1(l_1, c_1, v_1)\ \nonint^A\ op_2(l_2, c_2, v_2) \vee \\
+op_1(l_1, c_1, v_1)\ \nonint^A\ op_2(l_2, c_2, v_2) \Leftrightarrow &op_1(l_1, c_1, v_1)\ \nonint\ op_2(l_2, c_2, v_2) \vee \\
 & (c_1 = atomic \wedge c_2 = atomic)

### 2012-10-30 doc/types-pldi2013/example.tex
Lots of small changes

@@
 write privileges for the regions {\tt rl}, {\tt rw}, and {\tt rn}
-(line 17); the type system enforces that
-{\tt simulate\_circuit} only accesses these regions and their
-subregions at compile-time.
+(line 17); the type system statically enforces the functional restriciton that
+{\tt simulate\_circuit} only accesses these regions and their subregions.
 
@@
 Line 21 uses coloring {\tt pc.1} to partition {\tt rn} into {\tt rn0}, and {\tt rn1}.
-Subregion {\tt rn0} will contain all locations with a color of 1 in {\tt pc.1},
-while {\tt rn1} will contain all locations with a color of 2.  This
+Subregion {\tt rn0} will contain all locations with a color of 0 in {\tt pc.1},
+while {\tt rn1} will contain all locations with a color of 1.  This
 partitioning uses a coloring and results in {\em disjoint} subpartitions.
@@
 a region relationship it must already hold privileges for the new
-regions it finds in the packed value.  For example, on line 40
+regions it finds in the packed value.  For example, on line 40 the
 function {\tt execute\_time\_steps} has read/write privileges for

### 2012-10-30 doc/types-pldi2013/intro.tex
Lots of small changes

@@
 the problem of data aliasing in parallel programs.  The fully static approach disallows aliasing
-by requiring a single static partitioning of data.  Our own experience with writing high-performance 
+by requiring a single disjoint partitioning of data.  Our own experience with writing high-performance 
 applications in both Sequoia and the current industry standard mix of MPI, shared-memory
@@
 runtime checks and results in low actual runtime overhead.  We then show 
-that the soundness of our type system permits a scalable, hierarchical 
-distributed scheduling algorithm that enables Legion programs to achieve
+that the soundness of our type system permits a hierarchical,
+distributed scheduling algorithm that enables Legion programs to scale to 
 high performance on current supercomputers.
@@
 that may be operated on in parallel).  Logical regions have the following
-properties that allows programmers to dynamically communicate information
+properties that allow programmers to dynamically communicate information
 about the structure of program data to Legion:
@@
 region aliasing to perform dynamic safety checks for the unpacking
-and packing of logical regions to/from heap with regards to task privileges.  
+and packing of logical regions to/from the heap with regards to task privileges.  
 These checks are then amortized over all the uses of the logical region 

### 2012-10-30 doc/types-pldi2013/langdef.tex
Lots of small changes

@@
 set of memory locations that might be modified by other concurrent evaluations.  This is the clobber set $C$,
-and is used to relax the constraints the evaluation of an expression.  Whereas the requirement for locations not
-in the clobber set is the usual one of {\em consistency} (i.e. a read from a location observes the effect of all 
+and is used to relax the constraints on the evaluation of an expression.  Whereas the requirement for locations not
+in the clobber set is the usual one of {\em coherence} (i.e. a read from a location observes the effect of all 
 previous modifications applied in order), the value read from a location in the clobber set may be hypothesized
@@
 non-interfering, the converse is not true.  It is possible for $e_1$
-and $e_2$ to be judged to be potentially interfering and be serialized
+and $e_2$ to be judged to be interfering and be executed sequentially 
 even when the subexpressions are non-interfering.  This loss of

### 2012-10-30 apps/fdtd/fdtd.cc
FDTD: Add dtdx factor.

@@ -198,7 +198,7 @@ struct InitGlobalArgs {
 // Arguments to step_task.
 ////////////////////////////////////////////////////////////////////////
 struct StepGlobalArgs {
-  StepGlobalArgs(int nx, int ny, int nz, dim_t dim, dir_t dir,
+  StepGlobalArgs(int nx, int ny, int nz, dim_t dim, dir_t dir, double dtdx,
                  FieldID field_write, FieldID field_read1, FieldID field_read2)
     : n(nx, ny, nz), dim(dim), dir(dir),
       field_write(field_write), field_read1(field_read1), field_read2(field_read2) {}
@@ -208,6 +208,7 @@ struct StepGlobalArgs {
   dim_t dim;
   // Direction to look for ghost cells.
   dir_t dir;
+  double dtdx;
   // Fields to read and write.
   FieldID field_write, field_read1, field_read2;
 };
@@ -550,6 +551,7 @@ void main_task(const void *input_args, size_t input_arglen,
   double t_sim = 1.0;     // FIXME (Elliott): Full simulation takes forever.
   double courant = 0.5;
   double dt = courant/a;
+  double dtdx = courant;
 
   printf("+---------------------------------------------+\n");
   printf("| FDTD simulation parameters                  |\n");
@@ -736,7 +738,7 @@ void main_task(const void *input_args, size_t input_arglen,
                                           as_set<FieldID>(ghost2_fields), ghost2_fields,
                                           READ_ONLY, EXCLUSIVE, cells));
 
-      StepGlobalArgs global_args(nx, ny, nz, (dim_t)dim, DIR_POS,
+      StepGlobalArgs global_args(nx, ny, nz, (dim_t)dim, DIR_POS, dtdx,
                                  field_e[dim], field_h[(dim + 1)%NDIMS], field_h[(dim + 2)%NDIMS]);
 
       fs.push_back(
@@ -768,7 +770,7 @@ void main_task(const void *input_args, size_t input_arglen,
                                           as_set<FieldID>(ghost2_fields), ghost2_fields,
                                           READ_ONLY, EXCLUSIVE, cells));
 
-      StepGlobalArgs global_args(nx, ny, nz, (dim_t)dim, DIR_NEG,
+      StepGlobalArgs global_args(nx, ny, nz, (dim_t)dim, DIR_NEG, dtdx,
[…]

### 2012-10-30 apps/fdtd/fdtd.cc
FDTD: That should be minus, not plus....

@@ -893,7 +893,7 @@ void step_task(const void * input_global_args, size_t input_global_arglen,
       for (int z = rmin[2]; z < rmax[2]; z++) {
         vec3 v = rot3(vec3(x, y, z), -(dim + 1));
         ptr_t<double> i = cell_id(v, n);
-        f.write(i, f.read(i) + dtdx*(g1b.read(i + s1) - g1a.read(i)) - (g2b.read(i + s2) - g2a.read(i)));
+        f.write(i, f.read(i) - dtdx*(g1b.read(i + s1) - g1a.read(i)) - (g2b.read(i + s2) - g2a.read(i)));
       }
     }
   }

### 2012-10-31 doc/types-pldi2013/intro.tex
intro, abstract

@@
 
-In the past decade machine architectures, particularly at the high
-performance end of the spectrum, have undergone a revolution.  The
+Recently machine architecture, particularly at the high
+performance end of the spectrum, has undergone a revolution.  The
 latest supercomputers are now composed of heterogeneous processors
-and deep memory hierarchies.  Programming this class of machines is
-increasingly difficult.  Current programming systems for these
-machines have intricate features for describing parallelism, but contain
-few abstractions for describing properties of program data
-(e.g. independence).  Consequently, the burden of managing
-the correctness of parallel computations and movement of data
-through the memory hierarchy is placed on the programmer.
+and deep memory hierarchies.  Current programming systems for these
+machines have elaborate features for describing parallelism, but 
+few abstractions for describing the organization of data.  However, as the architecture
+of distributed memory has become increasingly complex, the placement and movement of dat has
+become a first order programming concern.
+
+%Consequently, the burden of managing
+%the correctness of parallel computations and movement of data
+%through the memory hierarchy is placed on the programmer.
 
@@
 
-To combat this problem, researchers have previously explored programming
-systems that are capable of reasoning about the structure of data in
-parallel programs.  Traditionally, these attempts have taken one of two approaches.   
-The first approach relies on fully static analyses that
-have no runtime overhead, but limited expressivity.  Two recent examples, 
+Previous work has explored two approaches to expressing the organization of data
+in distributed memory parallel programs.
+One class of work uses fully static analysis 
+with no runtime overhead but limited expressivity.  Two recent examples, 
 Sequoia \cite{Fatahalian06} and 
 Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
-mechanism to statically partition the heap recursively into a tree of collections 
+mechanism to statically partition the heap into a tree of collections 
 of data.  The two designs are different in many aspects, but agree that there is
 a single tree-shaped partitioning of data that must be fully checked statically 
-(see Section~\ref{sec:related} for more discussion of related work).  The
-second approach relies on a totally dynamic analysis of program data that doesn't
-limit expressivity, but incurs runtime overhead.  Examples of this approach 
+(see Section~\ref{sec:related}).  The
+second approach relies on fully dynamic analysis of program data that doesn't
+limit expressivity but incurs runtime overhead;  examples 
 include transactional memory \cite{Harris05} and thread-level speculation \cite{Steffan00}.
 
-These two approaches illustrate different mechanisms for dealing with
-the problem of data aliasing in parallel programs.  The fully static approach disallows aliasing
-by requiring a single disjoint partitioning of data.  Our own experience with writing high-performance 
+Our own experience writing high-performance 
 applications in both Sequoia and the current industry standard mix of MPI, shared-memory
-threads, and CUDA, has taught us that for many practical situations a
-single, static partitioning is insufficient.  In practice, the
+threads, and CUDA, has taught us that a
+single, static data partitioning is insufficient.  In practice, the
 best way to partition data is often a function of the data
-itself---i.e., the partitions need to be dynamically computed.
-Furthermore, it is useful to allow multiple, simultaneous partitions of
+itself---i.e., the partitions must be dynamically computed.
+Furthermore, applications often need multiple, simultaneous partitions of
 the same data, providing different views on to that data.  Alternatively, 
-the fully dynamic approach permits arbitrary data aliasing, but requires 
+the fully dynamic approach permits arbitrary partitioning, but requires 
 centralized control logic to perform conflict detection between every pair 
-of parallel computations. This results in runtime overhead and is 
+of parallel computations, resulting in runtime overhead that isn
 not scalable on distributed-memory machines.

### 2012-10-31 doc/types-pldi2013/legion-types.tex
intro, abstract

--- 2012-10-29 doc/types-pldi2013/legion-types.tex
@@
 
-\title{A Type System for Structured Aliasing in Parallel Programs}
+%\title{A Type System for Structured Aliasing in Parallel Programs}
+\title{Supporting Dynamic Data Partitioning in Parallel Programs}
 %\title{Scalable and Safe Parallelism using Logical Regions}
@@
 \begin{abstract}
-Aliasing introduces a significant impediment to reasoning about the
-structure and usage of data in parallel programs.  Traditional approaches
-have either disallowed aliasing to enable static analyses, or have permitted
-aliasing at the cost of dynamic runtime analyses.  In this paper we introduce a
-type system for the Legion programming model that uses a hybrid static and
-dynamic approach to manage aliasing with minimal runtime overhead.  We prove 
-that our type system can soundly reason about the privilege and coherence 
-properties in Legion programs. Furthermore, we prove that type soundness 
-enables a hierarchical, distributed scheduling algorithm for Legion programs 
-even in the presence of aliasing. We show that an implementation of a Legion 
-type checker improves the performance of Legion programs up to 71\% by 
-eliding memory checks that the type system proves safe.  We also demonstrate 
-that an implementation of the Legion runtime based on the hierarchical 
-scheduling algorithm scales well for three real-world applications, achieving 
-up to 62.5X speedup on 96 GPUs on the Keeneland supercomputer.
+Many distributed-memeory parallel applications dynamically partition 
+data, but current language designs either 
+require fully static partitioning or rely on entirely dynamic techniques.
+We describe Legion, a programming system with a combination
+of language features and static and dynamic checks to support 
+dynamic data partitions with minimal overhead.
+We describe the novel aspects of the Legion design, prove the soundness of the Legion type system,
+ and show Legion type checking improves performance by up to 71\% by eliding provably
+safe memory checks.  Furthermore, we show how the language design and type safety
+enables a hierarchical, distributed runtime scheduling algorithm, and how
+a system of data coherence annotations improves expressivity without compromising
+soundness. We report results for three real-world applications running
+on distributed memory machines, achieving up to 62.5X speedup on 96 GPUs 
+on the Keeneland supercomputer.
 %Legion is a programming model that introduces

### 2012-10-31 doc/types-pldi2013/langdef.tex
fun with figures

@@
 {
+\scalebox{0.9}{
 \begin{tikzpicture}[x=1in,y={(0,-1in)}]
@@
 
-\draw[very thin] (3.0,0.1) -- (3.0,7.6);
+% column positions
+% \XA - left edge of type rules
+% \XB - center of type rule names
+% \XC - vertical line
+% \XD - left edge of semantics rules
+% \XE - center of semantics rule names
+
+\newcommand{\XA}{0.0}
+\newcommand{\XB}{2.8}
+\newcommand{\XC}{3.4}
+\newcommand{\XD}{3.65}
+\newcommand{\XE}{7.3}
+
+\draw[very thin] (\XC,0.1) -- (\XC,7.6);
 
 % start with read/write/reduce rules
-\ruleat{0.0,0.0}{2.6,0.2}{T-Read}
+\ruleat{\XA,0.0}{\XB,0.2}{T-Read}
 {\begin{array}{l}
@@
 
-\ruleat{0.0,0.6}{2.6,0.8}{T-Write}
+\ruleat{\XA,0.6}{\XB,0.8}{T-Write}
 {\begin{array}{l}
@@
 
-\ruleat{0.0,1.3}{2.6,1.5}{T-Reduce}
+\ruleat{\XA,1.3}{\XB,1.5}{T-Reduce}
 {\begin{array}{l}
@@
 
-\axiomat{0.0,2.2}{2.6,2.2}{T-New}
+\axiomat{\XA,2.2}{\XB,2.2}{T-New}
 {\typeenvx \text{new }T@r : T@r}
 
-\ruleat{0.0,2.5}{2.6,2.55}{T-UpRgn}
+\ruleat{\XA,2.5}{\XB,2.55}{T-UpRgn}
 {\begin{array}{l}
@@
 
-\ruleat{0.0,3.1}{2.6,3.05}{T-DnRgn}
+\ruleat{\XA,3.1}{\XB,3.05}{T-DnRgn}
 {\begin{array}{l}
@@
 
-\axiomat{0.0,3.6}{2.6,3.6}{T-NewColor}
+\axiomat{\XA,3.6}{\XB,3.6}{T-NewColor}
 {\typeenvx \text{newcolor }r : \text{coloring}(r)}
 
-\ruleat{0.0,3.9}{2.6,4.1}{T-Color}
+\ruleat{\XA,3.9}{\XB,4.1}{T-Color}
 {\begin{array}{l}
@@
 
-\ruleat{0.0,4.6}{2.6,4.55}{T-Partition}
+\ruleat{\XA,4.6}{\XB,4.55}{T-Partition}
 {\begin{array}{l}
@@
 
-\ruleat{0,5.5}{2.6,5.7}{T-Pack}
+\ruleat{0,5.5}{\XB,5.7}{T-Pack}
 {\begin{array}{l}
@@
 
-\ruleat{0.0,6.2}{2.6,6.4}{T-Unpack}
+\ruleat{\XA,6.2}{\XB,6.4}{T-Unpack}
 {\begin{array}{l}
@@
 
-\ruleat{0.0,7.4}{2.6,7.6}{T-Call}
+\ruleat{\XA,7.4}{\XB,7.6}{T-Call}
 {\begin{array}{l}
@@
 
-\ruleatx{3.1,0.0}{6.7,0.2}{E-Read}{
+\ruleatx{\XD,0.0}{\XE,0.2}{E-Read}{
 \opsenvx e \mapsto l, E \andalso S' = \text{apply}(S, E) \\
@@
 
-\ruleatx{3.1,0.9}{6.7,0.85}{E-Write}{
+\ruleatx{\XD,0.9}{\XE,0.85}{E-Write}{
 \opsenvx e_1 \mapsto l, E_1 \andalso S' = \text{apply}(S, E_1) \\
@@
 
-\ruleatx{3.1,1.5}{6.7,1.45}{E-Reduce}{
+\ruleatx{\XD,1.5}{\XE,1.45}{E-Reduce}{
 \opsenvx e_1 \mapsto l, E_1 \andalso S' = \text{apply}(S, E_1) \\
@@
 
-\ruleatx{3.1,2.1}{6.7,2.2}{E-New}{
+\ruleatx{\XD,2.1}{\XE,2.2}{E-New}{
 l \in M(r) \andalso
@@
 
-\ruleat{3.1,2.6}{6.7,2.7}{E-UpRgn}{
+\ruleat{\XD,2.6}{\XE,2.7}{E-UpRgn}{
 \opsenvx e \mapsto v, E
@@
 
-\ruleatx{3.1,3.1}{6.7,3.3}{E-DnRgn}{
+\ruleatx{\XD,3.1}{\XE,3.3}{E-DnRgn}{
 \opsenvx e \mapsto l, E \\
@@
 
-\ruleatx{3.1,4.0}{6.7,4.2}{E-NewColor}{
+\ruleatx{\XD,4.0}{\XE,4.2}{E-NewColor}{
 K = \{ (l_1, iv_1), \ldots, (l_p, iv_p) \}, \text{ where } \\
@@
 
-\ruleatx{3.1,4.7}{6.7,4.9}{E-Color}{
+\ruleatx{\XD,4.7}{\XE,4.9}{E-Color}{
 \opsenvx e_1 \mapsto K, E_1 \andalso S' = \text{apply}(S, E_1) \\
@@
 
-\ruleatx{3.1,5.7}{6.7,5.55}{E-Partition}{
+\ruleatx{\XD,5.7}{\XE,5.55}{E-Partition}{
 \opsenv{e_1}{K}{E_1} \andalso \rho_i = \{ l : (l, i) \in K \}, \text{ for } 1 \leq i \leq k \\
@@
 
-\ruleatx{3.1,6.5}{6.7,6.6}{E-Pack}{
+\ruleatx{\XD,6.5}{\XE,6.6}{E-Pack}{
 \opsenvx e_1 \mapsto v, E \andalso \rho_i = M[r_i], \text{ for } 1 \leq i \leq k \\
@@
 
-\ruleatx{3.1,7.1}{6.7,6.95}{E-Unpack}{
+\ruleatx{\XD,7.1}{\XE,6.95}{E-Unpack}{
 \opsenvx e_1 \mapsto \langle \langle \otok{k}{\rho} , v_1 \rangle \rangle, E_1 \andalso M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \\
@@
 
-\ruleatx{0.0,8.1}{6.7,8.4}{E-Call}{
+\ruleatx{\XA,8.1}{\XE,8.4}{E-Call}{
 \opsenvx e_1 \mapsto v_1, E_1 \andalso S_1 = \text{apply}(S, E_1) \andalso \opsenvx e_2 \mapsto v_2, E_2 \andalso \ldots \andalso S_n = \text{apply}(S_{n-1}, E_n) \\
@@
 
+% try to fit helper rules at the bottom instead of in a separate figure
+
+\draw (0.05,9.12) -- (7.6,9.12);
+
+\node[below right] at (\XA,9.15) {
+$
+\begin{array}{@{}l@{}}
+\begin{array}{@{}l@{}l@{}} 
+\begin{array}{@{}lll}
+apply(S, []) & = & S \\
+apply(S, E \dplus [ read(l, c, v) ]) & = & apply(S, E) \\
+apply(S, E \dplus [ write(l, c, v) ]) & = & apply(S, E)[v/l] \\
+apply(S, E \dplus [ reduce_{id}(l, c, v) ]) & = & S'[id(S'(l), v)/l], \\
+& & \text{ where } S' = apply(S, E)
+\end{array} &
+\begin{array}{@{}ll}
+mark\_coherence([], \hat Q) & = [] \\
+mark\_coherence([ op(l, c, v) ] \dplus E, \hat Q) & = [ op(l, c', v) ] \dplus mark\_coherence(E, \hat Q), \\
+\multicolumn{2}{@{}l}{ \hspace{1.0in} \text{ where } c' =
+\begin{cases}
+simult, & \text{if }\exists \rho. l \in \rho \wedge simult(\rho) \in \hat Q \\
+atomic, & \text{if }\exists \rho. l \in \rho \wedge atomic(\rho) \in \hat Q \\
+excl, & \text{otherwise}
+\end{cases}}
+\end{array}
+\end{array}
+\end{array}
+$};
+
 \end{tikzpicture}
+}
 \vspace*{-0.5cm}
@@
 $$
-\begin{array}{lll}
-valid\_interleave(S, C, E', \oton{E}) & = & any\_interleave(E', \oton{E}) \wedge \\
-& & coherent(S, L_{excl}(E', C), L_{excl}(E', C), E')\ \wedge \\
-& & seq\_equiv(S, L_{excl}(E', C), L_{excl}(E', C), E', \oton{E})\ \wedge \\
-& & coherent(S, L_{atomic}(E', C), \emptyset, E')\ \wedge \\
-& & \exists \oton{\pi}, (\oton{\pi}) \text{ is a permutation of } (1,\ldots,n)\ \wedge \\
-& & \ \ \ seq\_equiv(S, L_{atomic}(E', C), \emptyset, E', E_{\pi_1}, \ldots, E_{\pi_n}) \\
-coherent(S, L_1, L_2, []) & = & \text{true} \\
-coherent(S, L_1, L_2, [ \epsilon ] \dplus E) & = &\
+\begin{array}{@{}l@{}}
+\begin{array}{@{}lll@{}}
+any\_interleave([], [], \ldots, []) & = & true \\
+any\_interleave([ \epsilon ] \dplus E', E_1, \ldots, [ \epsilon ] \dplus E_i, \ldots, E_n) & = & any\_interleave(E', E_1, \ldots, E_i, \ldots, E_n) \\
+\\
+\multicolumn{3}{@{}l}{
+\begin{array}{@{}l|l@{}}
+\begin{array}{@{}l@{}}
+valid\_interleave(S, C, E', \oton{E}) = \vspace{0.03in} \\
+\hspace{0.2in} any\_interleave(E', \oton{E})\ \wedge \vspace{0.03in} \\
+\hspace{0.2in} coherent(S, L_{excl}(E', C), L_{excl}(E', C), E')\ \wedge \vspace{0.03in} \\
+\hspace{0.2in} seq\_equiv(S, L_{excl}(E', C), L_{excl}(E', C), E', \oton{E})\ \wedge \vspace{0.03in} \\
+\hspace{0.2in} coherent(S, L_{atomic}(E', C), \emptyset, E')\ \wedge \vspace{0.03in} \\
+\hspace{0.2in} \exists \oton{\pi}, (\oton{\pi}) \text{ is a permutation of } (1,\ldots,n)\ \wedge \\
+\hspace{0.3in} seq\_equiv(S, L_{atomic}(E', C), \emptyset, E', E_{\pi_1}, \ldots, E_{\pi_n})
+\end{array} &
+\begin{array}{@{}l@{}}
+coherent(S, L_1, L_2, []) = \text{true} \\
+coherent(S, L_1, L_2, [ \epsilon ] \dplus E) = \\
+\hspace{0.2in}
 \begin{cases}
@@
 \ \ \ coherent(S, L_1, L_2, E), & \text{if $\epsilon = read(l, c, v)$} \\
-coherent(apply(S, \epsilon), L_1, L_2 \cup \{ l \}, E), & \text{if $\epsilon = write(l, c, v)$ and $l \in L_1$} \\
+coherent(apply(S, \epsilon), L_1, L_2 \cup \{ l \}, E), & \text{if $\epsilon = write(l, c, v)$} \\
+& \hspace{0.25in} \text{and $l \in L_1$} \\
 coherent(apply(S, \epsilon), L_1, L_2, E), & \text{otherwise}
-\end{cases} \\
-seq\_equiv(S, L_1, L_2, E', \oton{E}) & = & coherent(S, L_1, L_2, E_1 \dplus \ldots \dplus E_2)\ \wedge \\
-& & \forall l \in L_1. apply(S, E')(l) = apply(S, E_1 \dplus \ldots \dplus E_n)(l) \\
+\end{cases} 
+\end{array}
+\end{array}
+} \\
+\\
+\end{array} \\
+\begin{array}{@{}ll@{}}
+\begin{array}{@{}l@{}}
+seq\_equiv(S, L_1, L_2, E', \oton{E}) = \\
+\hspace{0.2in} coherent(S, L_1, L_2, E_1 \dplus \ldots \dplus E_2)\ \wedge \\
+\hspace{0.2in} \forall l \in L_1. apply(S, E')(l) = apply(S, E_1 \dplus \ldots \dplus E_n)(l)
+\end{array} &
+\begin{array}{@{\hspace{0.2in}}lll@{}}
 L_{excl}(E, C) & = & \{ l : op(l, excl, v) \text{ in } E  \} \setminus C \\
 L_{atomic}(E, C) & = & \{ l : op(l, atomic, v) \text{ in } E \} \setminus (C \cup L_{excl}(E, C)) \\
+\end{array}
+\end{array}
 \end{array}

### 2012-11-01 doc/types-pldi2013/langdef.tex
added reference for parallel operational semantics (concurrent ML)

@@
 one expression's evaluation to observe interference from other concurrent evaluations.  The traditional way of
-handling this in operational semantics \cite{XYZ} is to use a top-down approach with small-step semantics and
-a global view of all
+handling this in operational semantics (\cite{CML} is a well-known example) is to use a top-down approach with
+small-step semantics and a global view of all
 partially-evaluated expressions.  Rules are specified for which expression(s) are valid choices for the next small

### 2012-11-01 doc/types-pldi2013/intro.tex
more intro

@@
 
-Recently machine architecture, particularly at the high
-performance end of the spectrum, has undergone a revolution.  The
+Machine architecture, particularly at the high
+performance end of the spectrum, has recently undergone a revolution.  The
 latest supercomputers are now composed of heterogeneous processors
@@
 few abstractions for describing the organization of data.  However, as the architecture
-of distributed memory has become increasingly complex, the placement and movement of dat has
+of distributed memory has become increasingly complex, the placement and movement of data has
 become a first order programming concern.
@@
 second approach relies on fully dynamic analysis of program data that doesn't
-limit expressivity but incurs runtime overhead;  examples 
+limit expressivity but incurs runtime overhead and centralized control logic that is
+difficult to scale to distributed memory machines;  examples 
 include transactional memory \cite{Harris05} and thread-level speculation \cite{Steffan00}.
@@
 Our own experience writing high-performance 
-applications in both Sequoia and the current industry standard mix of MPI, shared-memory
+using the current industry standard mix of MPI, shared-memory
 threads, and CUDA, has taught us that a
@@
 Furthermore, applications often need multiple, simultaneous partitions of
-the same data, providing different views on to that data.  Alternatively, 
-the fully dynamic approach permits arbitrary partitioning, but requires 
-centralized control logic to perform conflict detection between every pair 
-of parallel computations, resulting in runtime overhead that isn
-not scalable on distributed-memory machines.
+the same data, providing different views on to that data.
 
-In this work, we present a type system and operational semantics for
-Legion\cite{Legion12}.  Legion is a programming model that supports multiple,
-dynamic partitions of program data which can result in aliasing.  We prove
-the soundness of our type system and show that through a combination of
-static and dynamic analysis we can implement it in a way that minimizes dynamic
-runtime checks and results in low actual runtime overhead.  We then show 
-that the soundness of our type system permits a hierarchical,
-distributed scheduling algorithm that enables Legion programs to scale to 
-high performance on current supercomputers.
-
-Legion presents {\em logical regions} as an
-abstraction for capturing {\em locality} (data that will be used together,
+In this work, we present the design of Legion \cite{Legion12},
+a parallel programming model that efficiently supports multiple, dynamic partitions
+of data. Legion's {\em logical regions} are an
+abstraction capturing {\em locality} (data that will be used together,
 and therefore should be colocated) and {\em independence} (disjoint data
-that may be operated on in parallel).  Logical regions have the following
+that may be operated on in parallel).  Logical regions have 
 properties that allow programmers to dynamically communicate information
@@
 
-\item  a logical region may be dynamically partitioned in multiple different ways.
+\item  a logical region may be dynamically partitioned in multiple different ways;
+
+\item programmers give logical regions {\em privilege} and {\em coherence} properties that
+describe how the region's data may be used by computations.
 \end{itemize}
-Through multiple partitions
-of a logical region, a given datum may belong to multiple different regions.  
-For example, if a region $R$ is disjointly partitioned three distinct ways 
-and each partition assigns every element in $R$ to a
-subregion (partitions need not be total in Legion), then every element
-in $R$ is included in three different subregions.  This is why we use
-the term {\em logical} regions: language-level regions serve to name
-sets of data but do not imply anything about physical layout or placement
-in the memory hierarchy.\footnote{There
-is a separate system of {\em physical regions} at runtime that hold
-copies of the data in the language-level logical regions.  Details can
-be found in \cite{Legion12}.}
 
-While data in Legion is organized into a hierarchy of logical region forests, 
-computation is organized into a hierarchy of tasks.  Logical regions and tasks interact
+While data in Legion is organized in a hierarchy of logical region forests, 
+computation is organized in a hierarchy of tasks.  Logical regions and tasks interact
 through a system of privileges specifying operations that a task
 can perform on a logical region argument (read, write, or reduce).  
-Tasks are only permitted to access logical regions for which they 
-possess privileges.  Privileges are given to a task by their calling task
-or by creating new logical regions.  The Legion programming model 
-mandates a functional restriction: a sub-task can only be passed a subset
-of the privileges possesed by its parent task.
+Tasks can only access logical regions for which they 
+possess privileges.  Tasks acquire privileges from their calling task
+or by creating new logical regions.  A key restriction is that 
+a sub-task can only be passed a subset of the privileges of its parent task.
 
-To enforce this restriction, our type system relies on a combination of static 
-and dynamic analyses.  The key observation is that dynamic
-analysis is very easy and relatively cheap when done at the granularity of
-regions instead of individual heap locations.  To that end, our type system supports
-static verification of all pointer dereferences and passing of region privileges at 
-task calls boundaries.  The Legion runtime then leverages its knowledge of logical
-region aliasing to perform dynamic safety checks for the unpacking
-and packing of logical regions to/from the heap with regards to task privileges.  
-These checks are then amortized over all the uses of the logical region 
-which significantly reduces their cost.
+Through multiple partitions of a logical region, a given datum may
+belong to multiple different regions.  Thus, allowing multiple
+simultaneous partitions of the same data introduces
+aliasing---distinct regions may have overlapping data.\footnote{This
+  is why we use the term {\em logical} regions: language-level regions
+  name sets of data but do not imply a physical layout or placement in
+  the memory hierarchy. A separate system of {\em physical regions} at
+  runtime holds copies of data in the language-level logical regions
+  \cite{Legion12}.}  We believe permitting very flexible aliasing is
+necessary, but it is also presents a major challenge. For example, to
+detect that tasks can execute in parallel it is sufficient to prove
+that they work on disjoint data.
 
-To enable parallelism, Legion again relies on a combination of static and dynamic
-analysis to prove that two tasks operate on independent sets of data.  If two tasks
-cannot be proven independent statically, the Legion runtime performs dynamic dependence
-tests to detect region aliasing.  Since these tests are performed at the granularity of
-regions instead of individual heap locations, they are significantly less costly.
-In addition to supporting parallel execution for independent tasks, Legion also enables
-programmer specified parallel execution in the presence of aliasing.  For example, two tasks 
-accessing aliased data may only need to be serialized
-with respect to each other (e.g. run atomically) or are permitted to run
-simultaneously with the programmer using separate synchronization mechanisms
-to control access (e.g. locks).  To support these relaxed access paradigms 
-Legion provides {\em cohrence modes} on logical regions.  For each region
-argument to a task, a programmer specifies the coherence in which other tasks
-can access aliased data while the task is running.  We present an operational semantics
-for coherence modes and show that it is sound.
+Legion also provides {\em coherence} modes for regions, which allows
+programmers to permit parallel execution even in cases where tasks in
+fact do access the same data in conflicting ways.  The programmer may
+specify that tasks conflicting on a region  must execute serially (in program
+order, the default), {\em atomically} with respect to each other, or even
+{\em simultaneously} if the programmer uses their own synchronization mechanisms
+to control access (e.g., locks).
 
-Using our proofs of the soundess of the privilegs and coherence systems, we then 
-prove a third result: if two {\em siblings} (tasks that
-have the same immediate parent task) $t_1$ and $t_2$ are 
-{\em non-interfering} (have no ordering requirements for correctness), 
-then any descendant of $t_1$ in the task hierarchy is non-interfering
-with any descendant of $t_2$.  This theorem is the basis for correct, distributed
-task scheduling in a Legion implementation, which is crucial for
-performance; a centralized scheduler would be a bottleneck
-because of the large communication latencies in the target
-class of machines.
+To efficiently enforce restrictions that guarantee the safety and
+correctness of program executions for uses of first-class regions,
+their privilege and coherence properties, and arbitrary region
+aliasing, the Legion runtime relies on a combination of static and
+dynamic checks.  A key theme that runs through the design is that very
+hard static analysis problems (e.g., alias analysis) are often both easy and
+inexpensive if they can be done dynamically at the granularity of regions instead of individual
+heap locations, while checks that must be done on individual heap locations are only
+cheap if done statically.  Explicating and evaluating the design of Legion's
+static and dynamic semantics and its implications is the main topic of this paper,
+which we present as a series of contributions:
+\begin{itemize}
+\item We present a type system for {\em Core Legion} programs that statically
+  verifies the safety of individual pointer references and region
+  privileges at task call boundaries (Section~\ref{subsec:coretypes}).
+  A subtle issue arises in checking privileges when regions are
+  stored in and later retrieved from heap data structures.  The type
+  system provides enough static information that the Legion runtime
+  can leverage its dynamic information about region aliasing to
+  perform dynamic privelege checks guaranteeing safety.
+
+\item We present a parallel operational semantics for Core Legion that
+  may be of independent interest.  This semantics is compositional,
+  hierarchical, and asynchronous, reflecting the way such programs
+  actually execute on the hardware.  However, the semantics also
+  directly models conflicting memory operations between tasks in cases of relaxed
+  coherence, making explicit the boundary between deterministic and
+  possible non-deterministic behavior (Section~\ref{subsec:opsemantics}).
+
+\item We prove the soundness of Legion's privelege system (Section~\ref{sec:soundness}).
+
+\item We use the soundness of privileges to prove the soundness of
+  Legion's region coherence modes (Section~\ref{sec:coherence}).
+
+\item We show that if expressions $e_1$ and $e_2$ are {\em non-interfering} (can be executed in parallel), then subexpressions
+$e_1'$ of $e_1$ and $e_2'$ of $e_2$ are also non-interfering (Section~\ref{sec:scheduling}).  
+This result is the basis
+for a hierarchical, distributed scheduler in the Legion implementation which is crucial for high performance
+on the target class of machines;  any form of centralized scheduling would be a bottleneck because of the
+large communication latencies in the target class of machines.
+%; on the target class of machines, any centralized
+%scheduler would be a serious bottleneck.
+
+\item We give experimental evidence for the Legion design.  On three
+  real-world applications, we show that dynamic region pointer checks
+  would be expensive, justifying checking this aspect of the type
+  system statically.  We also show that the cost of region aliasing
+  checks is low, showing that a much more expressive and dynamic
+  language with aliasing is not incompatible with both high
+  performance and safety (Section~\ref{sec:evaluation}).
+
+\end{itemize}
 
@@
 
-The rest of the paper is organized as follows.
-We begin in Section~\ref{sec:example} with an example program that illustrates
-a typical style of programming in Legion and motivates the need for multiple, dynamically computed partitions of
-a region that introduce aliased data.  We define Core Legion, a small language suitable 
-for our formal development, in Section~\ref{sec:legioncore}.
-The next four sections each state and prove one of our main results and contributions:
-\begin{itemize}
-\item We prove the soundness of Legion's privileges system (Section~\ref{sec:soundness}).
-
-\item We use the soundness of privileges to prove the soundness of Legion's region coherence modes (Section~\ref{sec:coherence}).
-
-\item We show that if expressions $e_1$ and $e_2$ are {\em non-interfering} (can be executed in parallel), then subexpressions
-$e_1'$ of $e_1$ and $e_2'$ of $e_2$ are also non-interfering (Section~\ref{sec:scheduling}).  
-%This result is the basis
-%for a hierarchical, distributed scheduler in the Legion implementation which is crucial for high performance
-%on the target class of machines.
-%; on the target class of machines, any centralized
-%scheduler would be a serious bottleneck.
-
-\item We give experimental evidence for the Legion design.  On three real-world applications, we show that 
-dynamic region pointer checks would be expensive, justifying checking this aspect of the type system statically.
-We also show that the cost of checking for non-interference of region privileges is low, showing that a much more expressive
-and dynamic language with aliasing is not incompatible with high performance (Section~\ref{sec:evaluation}).
-
-\end{itemize}
-Finally, Section~\ref{sec:related} discusses the most closely related work and
-Section~\ref{sect:conclusion} concludes.
+%The rest of the paper is organized as follows.
+%We begin in Section~\ref{sec:example} with an example program that illustrates
+%a typical style of programming in Legion and motivates the need for multiple, dynamically computed partitions of
+%a region that introduce aliased data.  We define Core Legion, a small language suitable 
+%for our formal development, in Section~\ref{sec:legioncore}.
+%Finally, Section~\ref{sec:related} discusses the most closely related work and
+%Section~\ref{sect:conclusion} concludes.

### 2012-11-01 doc/types-pldi2013/legion-types.tex
more intro

@@
 data, but current language designs either 
-require fully static partitioning or rely on entirely dynamic techniques.
+require fully static partitioning or rely on expensive dynamic checks.
 We describe Legion, a programming system with a combination

### 2012-11-01 doc/types-pldi2013/intro.tex
intro

@@
 
-Our own experience writing high-performance 
+Our own experience writing high-performance applications
 using the current industry standard mix of MPI, shared-memory
@@
 Furthermore, applications often need multiple, simultaneous partitions of
-the same data, providing different views on to that data.
+the same data, each providing different views on to that data.
 
-In this work, we present the design of Legion \cite{Legion12},
+In this work, we present the static and dynamic semantics of Legion \cite{Legion12},
 a parallel programming model that efficiently supports multiple, dynamic partitions
@@
 that may be operated on in parallel).  Logical regions have 
-properties that allow programmers to dynamically communicate information
-about the structure of program data to Legion:
+properties that allow programmers to communicate information
+about the structure of program data to the Legion implementation:
 \begin{itemize}
@@
 
-\item programmers give logical regions {\em privilege} and {\em coherence} properties that
-describe how the region's data may be used by computations.
+\item {\em privilege} and {\em coherence} properties that
+describe how a logical region's data may be used by computations.
 \end{itemize}
 
-While data in Legion is organized in a hierarchy of logical region forests, 
+While data in Legion is organized in a hierarchy of logical regions,
 computation is organized in a hierarchy of tasks.  Logical regions and tasks interact
-through a system of privileges specifying operations that a task
+through a system of {\em privileges} specifying operations that a task
 can perform on a logical region argument (read, write, or reduce).  
 Tasks can only access logical regions for which they 
-possess privileges.  Tasks acquire privileges from their calling task
-or by creating new logical regions.  A key restriction is that 
-a sub-task can only be passed a subset of the privileges of its parent task.
+possess privileges.  
+%Tasks acquire privileges from their calling task
+%or by creating new logical regions.  A key restriction is that 
+%a sub-task can only be passed a subset of the privileges of its parent task.
 
@@
 simultaneous partitions of the same data introduces
-aliasing---distinct regions may have overlapping data.\footnote{This
-  is why we use the term {\em logical} regions: language-level regions
+{\em aliasing}---distinct regions may have overlapping data.\footnote{Thus
+  the term {\em logical} regions: language-level regions
   name sets of data but do not imply a physical layout or placement in
@@
   runtime holds copies of data in the language-level logical regions
-  \cite{Legion12}.}  We believe permitting very flexible aliasing is
-necessary, but it is also presents a major challenge. For example, to
-detect that tasks can execute in parallel it is sufficient to prove
-that they work on disjoint data.
+  \cite{Legion12}.}
 
@@
 
-To efficiently enforce restrictions that guarantee the safety and
-correctness of program executions for uses of first-class regions,
-their privilege and coherence properties, and arbitrary region
-aliasing, the Legion runtime relies on a combination of static and
+To efficiently guarantee the safety and
+correctness of programs using first-class regions,
+privilege and coherence properties, and arbitrary region
+aliasing, Legion relies on a combination of static and
 dynamic checks.  A key theme that runs through the design is that very
-hard static analysis problems (e.g., alias analysis) are often both easy and
+hard static analysis problems (specifically, alias analysis) are often both easy and
 inexpensive if they can be done dynamically at the granularity of regions instead of individual
-heap locations, while checks that must be done on individual heap locations are only
+heap locations, while checks that must be done on individual values are only
 cheap if done statically.  Explicating and evaluating the design of Legion's
-static and dynamic semantics and its implications is the main topic of this paper,
+static and dynamic semantics and its implications are the topics of this paper,
 which we present as a series of contributions:
@@
 \item We present a type system for {\em Core Legion} programs that statically
-  verifies the safety of individual pointer references and region
+  verifies the safety of individual pointers and region
   privileges at task call boundaries (Section~\ref{subsec:coretypes}).
-  A subtle issue arises in checking privileges when regions are
-  stored in and later retrieved from heap data structures.  The type
-  system provides enough static information that the Legion runtime
-  can leverage its dynamic information about region aliasing to
-  perform dynamic privelege checks guaranteeing safety.
+%  A subtle issue arises in checking privileges when regions are
+%  stored in and later retrieved from heap data structures.  The type
+%  system provides enough static information that the Legion runtime
+%  can leverage its dynamic information about region aliasing to
+%  perform dynamic privelege checks guaranteeing safety.
 
@@
   hierarchical, and asynchronous, reflecting the way such programs
-  actually execute on the hardware.  However, the semantics also
-  directly models conflicting memory operations between tasks in cases of relaxed
+  actually execute on the hardware.  The semantics also
+  models conflicting memory operations between tasks in cases of relaxed
   coherence, making explicit the boundary between deterministic and
@@
 
-\item We prove the soundness of Legion's privelege system (Section~\ref{sec:soundness}).
+\item We prove the soundness of Legion's static type and privelege system (Section~\ref{sec:soundness}).
 
-\item We use the soundness of privileges to prove the soundness of
+\item We use soundness of the type system to prove the soundness of
   Legion's region coherence modes (Section~\ref{sec:coherence}).
 
-\item We show that if expressions $e_1$ and $e_2$ are {\em non-interfering} (can be executed in parallel), then subexpressions
+\item Again using the sounds of the type system,  we show that if expressions $e_1$ and $e_2$ are {\em non-interfering} (can be executed in parallel), then subexpressions
 $e_1'$ of $e_1$ and $e_2'$ of $e_2$ are also non-interfering (Section~\ref{sec:scheduling}).  
@@
 for a hierarchical, distributed scheduler in the Legion implementation which is crucial for high performance
-on the target class of machines;  any form of centralized scheduling would be a bottleneck because of the
-large communication latencies in the target class of machines.
+on the target class of machines.
+%  any form of centralized scheduling would be a bottleneck because of the
+%large communication latencies in the target class of machines.
 %; on the target class of machines, any centralized
@@
   system statically.  We also show that the cost of region aliasing
-  checks is low, showing that a much more expressive and dynamic
+  checks is low, showing that an expressive and dynamic
   language with aliasing is not incompatible with both high

### 2012-11-01 doc/types-pldi2013/langdef.tex
updates

--- 2012-10-31 doc/types-pldi2013/langdef.tex
@@
 }
-\caption{Type System and Operational Semantics Rules}
+\caption{Type System and Operational Semantics}
 \label{fig:langrules}
@@
 Figure~\ref{fig:langdef} defines Core Legion, a subset of the full Legion language
-that still illustrates all of the important issues.  Types
-include booleans, integers, tuples, and pointers.  Pointers
-are annotated with a list of regions---a non-null pointer must point to a 
+that still illustrates all of the important issues.  
+%Types include booleans, integers, tuples, and pointers.  
+Pointers are annotated with a list of regions---a non-null pointer must point to a 
 location that is contained in at least one of the regions. There is a special
-type for {\em colorings}, which
-are used to specify partitions of regions into subregions.
-Functions in Legion are named, accept one or more arguments and
-return a value of some type.  A function also declares the region
-access privileges and coherence it requires.  A function may be executed
-within the current task or run in parallel as a separate subtask---this decision is made
+{\em coloring} type, used to specify partitions of regions into subregions.
+%Functions in Legion are named, accept one or more arguments and
+%return a value of some type.  
+Functions declare the region
+access privileges and coherence they require.  Some function instances are {\em tasks}; a function may be executed
+within the current task or run in parallel as a separate (sub)task---this decision is made
 by the Legion task scheduler.  Function types are
 universally quantified over all region names appearing in the type.
-The final type is a region relationship, an instance of which captures a value and
+The other important type is a {\em region relationship}, an instance of which captures a value and
 one or more regions satisfying associated constraints.  A region relation instance
@@
 
-To achieve high performance in the presence of aliasing, the Legion type system has 
-been designed to efficiently leverage both static and dynamic checks.
-Consequently, both the type system and the operational
-semantics have been extended in interesting ways from a basic expression language.  We
-describe these extensions in the next two subsections, and explain how they 
-were chosen to preserve a straightforward proof of soundness for the type system.
+To achieve high performance in the presence of aliasing, the Legion type system
+leverages both static and dynamic checks, resulting in interesting extensions beyond
+a basic expression language to both the type system and the operational semantics.
+The following two subsections illustrate these decisions and how they have been
+chosen to preserve a straightforward proof of type soundness.
 
@@
 $$\typeenv{e}{T}$$
-
-In addition to the traditional type environment $\Gamma$, a Legion type judgment includes the
-set of access privileges $\Phi$ for the logical regions in the expression as well as a set of
+Besides a type environment $\Gamma$, type judgments include the
+access privileges $\Phi$ for the logical regions in the expression $e$ as well as 
 constraints $\Omega$ that must hold between those logical regions.
 
-Both $\Phi$ and $\Omega$ are used in the heap access expressions: {\em read}, {\em write}, and {\em reduce}.
-For a heap access to be valid, the the appropriate permission must exist for logical region(s) in the
-pointer's type.  As regions are hierarchical, the exact region need not be named in $\Phi$ if permissions
-exist for a logical region that is known to contain the pointer's region.  To simplify this check, we
-define closure operations in Figure~\ref{fig:closure} that expand $\Omega$ into an $\Omega^*$, which is
-then used to expand $\Phi$ into a $\Phi^*$ that does explicitly name every logical region for which
-privileges are known to exist.
+A representative selection of the type rules is given in Figure~\ref{fig:langrules}.  Both $\Phi$ and $\Omega$ are used in the heap access expressions {\em read}, {\em write}, and {\em reduce}.
+A valid heap access has the needed permission for logical region(s) in the
+pointer's type.  Note the exact region need not be named in $\Phi$ if permissions
+exist for a logical region that provably contains the pointer's region.  To simplify this check, Figure~\ref{fig:closure}
+defines closure operations $\Omega^*$ (all constraints implied by $\Omega$) and
+$\Phi^*$ (all privileges implied by $\Phi$ and $\Omega^*$).
 
 Core Legion provides no automatic pointer type conversions.  Valid pointers are created via the {\em new}
-expression in specified region and invalid (i.e. {\em null}) pointers must also be typed.  Pointers can be
-``upcast'' via the {\em upregion} expression, which uses the expanded constraints $\Omega^*$ to verify that
-every possible region a pointer might point into is covered by at least one of the regions in the desired
-pointer type.  In contrast, a ``downcast'' using the {\em downregion} expression must perform a dynamic
-test to verify which region a run-time pointer value points into.  A static test similar to the one used
-for {\em upregion} could be used to catch cases in which the downcast can never succeed at runtime, but we
-have left it out for simplicity.
-
-The introduction of constraints into $\Omega$ occurs through the use of the {\em partition} expression,
-which makes use of a new type specific to Legion, a {\em coloring}.  A coloring is an opaque mapping from
-pointers (which must point into the region being colored) to integer {\em color} values.  (For simplicity, 
-Core Legion allows only an iterative construction of a coloring via a {\em newcolor} expression that generates
-an empty map, and a {\em color} expression which adds a new entry to an existing map. The opacity of the
-coloring allows for more efficient generation (and storage) of colorings as needed.)  The {\em partition} 
-expression introduces names for subregions corresponding to each color in the coloring, guaranteeing that each
-subregion is disjoint from the other subregions and all are included in the original region.
-
-The {\em pack} and {\em unpack} expressions allow the storage of pointers in the heap.  Logical region names
-are lexically scoped and if stored in the heap as-is, can only be read within the same scope.  A region relationship
-is used instead to capture the (dynamic) physical regions while preserving the statically known constraints
-between those regions.  Because the {\em pack} expression is able to verify these constraints hold when the 
-region relationship value is created, the {\em unpack} expression is able to reintroduce these constraints on
-the (fresh) logical region names given to the contents of an unwrapped region relationship value.
-
-Finally, the task call expression handles the aliasing of logical region names between the caller and callee and
-verifies the functional requirement that the privileges held by a called task are always a subset of those held
+expression in the specified region.
+% and invalid (i.e. {\em null}) pointers must also be typed.  
+Pointers can be
+``upcast'' via the {\em upregion} expression; the type rule verifies 
+every possible region the pointer might point into is covered by at least one of the regions in the target
+pointer type.  A ``downcast'' using the {\em downregion} expression performs a dynamic
+test to verify the region a run-time pointer value points into.  
+%A static test similar to the one used
+%for {\em upregion} could be used to catch cases in which the downcast can never succeed at runtime, but we
+%have left it out for simplicity.
+
+Region constraints are introduced by the {\em partition} expression,
+which makes use of {\em colorings}, a new type specific to Legion.  A coloring is an opaque mapping from
+pointers (which must point into the region being colored) to integer {\em color} values.  
+%(For simplicity, 
+%Core Legion allows only an iterative construction of a coloring via a {\em newcolor} expression that generates
+%an empty map, and a {\em color} expression which adds a new entry to an existing map. The opacity of the
+%coloring allows for more efficient generation (and storage) of colorings as needed.)  
+Partitioning introduces names for subregions corresponding to each color in the coloring, guaranteeing that each
+subregion is disjoint from the others and all are subregions of the original region.
+
+The {\em pack} and {\em unpack} expressions allow regions to be stored in the heap.
+%  Logical region names are lexically scoped and if stored in the heap as-is, can only be read within the same scope.  
+A {\em region relationship} captures statically known constraints between regions;
+because the {\em pack} rule verifies these constraints hold when the 
+region relationship value is created,{\em unpack} can reintroduce these constraints on
+the (fresh) logical region names when the value is read and unpacked from the heap.
+Crucially for soundness,  region relationships do not capture privileges.
+%
+% FIXME  It would be nice to add some comments about unpack here, and how the constraints introduced by unpack
+% must be sufficient to allow access to the new regions if they are to be used, which generally requires the type
+% have a free region variable capturing subregioning relationships between the packed regions and some existing region
+% in scope for which privileges are held.
+% 
+Finally, task calls verify the requirement that the privileges held by a called task are a subset of those held
 by the caller.  In addition to being necessary for the proof of soundness, this property is critical to
-enable the hierarchical scheduling model used by the Legion runtime.
+enabling hierarchical and distributed scheduling of tasks.
 
@@
 
-Core Legion's operational semantics rules have the following form:
+Core Legion's operational semantics rules have the form
 $$\opsenv{e}{v}{E}$$
-and specify that, given a certain environment, the evaluation of expression $e$ yields a value $v$ and a
-{\em dynamic memory trace} $E$ which is a sequence of all heap operations (reads, writes, reductions) performed
-during the evaluation, including their location and the value read or written.
-
-In addition to the usual value mappings for local variables $L$ and store (i.e. heap) locations $S$, the Legion
-operational semantics environments includes an immutable heap typing $H$ that maps locations to types, and a region
-mapping $M$ that maps logical regions $r_i$ to physical regions $\rho_i$, which are sets of concrete memory
+and specify the evaluation of expression $e$ yields a value $v$ and a
+{\em dynamic memory trace} $E$, which is a sequence of all heap operations (reads, writes, reductions) $e$ performs.
+%during the evaluation, including their location and the value read or written.
+In addition to  mappings for local variables $L$ and the store (i.e. heap) $S$, the 
+environment includes an immutable heap typing $H$ assigning types to heap locations, and a region
+mapping $M$ mapping logical regions $r_i$ to physical regions $\rho_i$, which are sets of concrete memory
 locations.
 
-The final element of the environment is the {\em clobber set} $C$.  Our goal is to show that parallel Legion
-programs can
-be scheduled in a hierarchical fashion.  We therefore require a formulation of operational semantics
-that is both parallel and hierarchical.  Because Legion allows constrained nondeterminism (in the order in which
-reductions are applied and in cases of relaxed coherence with aliasing), it is possible for the heap accesses performed by
-one expression's evaluation to observe interference from other concurrent evaluations.  The traditional way of
-handling this in operational semantics \cite{XYZ} is to use a top-down approach with small-step semantics and
-a global view of all
-partially-evaluated expressions.  Rules are specified for which expression(s) are valid choices for the next small
-step of execution.  This approach has the benefit of being constructive but the implicit global scheduler is
-difficult to map into a hierarchical one in which decisions must necessarily be made with limited information.
-
-Instead we use a bottom-up approach with large-step semantics in which each expression is evaluated with limited 
-information about the overall environment.  The information consists of the initial state of the heap $S$, and a
-set of memory locations that might be modified by other concurrent evaluations.  This is the clobber set $C$,
-and is used to relax the constraints on the evaluation of an expression.  Whereas the requirement for locations not
-in the clobber set is the usual one of {\em coherence} (i.e. a read from a location observes the effect of all 
-previous modifications applied in order), the value read from a location in the clobber set may be hypothesized
-to be any value at all.  The validation of these hypotheses is deferred until all potentially-concurrent
-modifications to a location can be observed (i.e. the evaluation of a parent expression whose clobber set does
-not include the location) and a consistent interleaving of the subexpressions' memory traces can be determined.
-With this formulation, the criteria for when a hierarchical scheduler may execute subexpressions in parallel can be
-made clear, but they are no longer constructive.  Instead, we will have to show that the scheduling algorithm
-used by the Legion runtime satisfies these constraints, guaranteeing the soundness of the Legion programming model.
-
-The complete operational semantics rules can be found in \cite{LegionTypes12}; we show the most interesting ones
-in Figure~\ref{fig:semantics}.  The rules for the {\em read}, {\em write}, and {\em reduce} show how dynamic memory
-traces are constructed and how the result of a read expression is relaxed when the location falls within the
-clobber set.  Additionally, the write and reduce operations, with their two subexpressions, demonstrate how the
-traces of subexpressions are interleaved within the environment (and clobber set) of the parent expression.
-(A formal definition of the {\em valid\_interleave} predicate can be found in Figure~\ref{fig:validinterleave}, and
-will be discussed in Section~\ref{sec:coherence}.)
-
-The first use of the region mapping $M$ occurs in the {\em new} and {\em downregion} expressions.  The new
-expression is constrained to select a location (with the right heap typing) from the set that the specified logical
-region has been mapped to.  (Although one would expect the new expression to also return a location that is not
-currently in use, our soundness result does not require this additional constraint, so we have omitted it for
-simplicity.)  Similarly, the downregion expression checks whether a location is within the set of locations the
-logical region(s) have been mapped to, returning {\em null} if not.  As discussed above, the corresponding 
-{\em upregion} expression checks are performed statically.
-
-The general form of a coloring being an abitrary mapping of locations to ``colors'' (integers) is shown in the
-operational semantics rules for the {\em newcolor} and {\em color} expressions.  The color expression accepts an
-existing coloring and returns one in which the specified location is given the specified color, replacing an
-existing coloring for that location (if present).  The behavior of the newcolor expression is more subtle.
-Since Legion supports the {\em partition-then-allocate} model (see Section~\ref{sec:example}) in which the {\em new} expression may be called
-on subregions, each subregion that results from a partitioning may need an arbitrary number of locations for which
-the program does not know the name at partitioning time.  The semantics for the newcolor expression allow this while
-guaranteeing that any explicit colorings performed by the program are honored.
-
-The semantics for the {\em pack} and {\em unpack} expressions are straightforward.  Packing a region relationship
-is just a matter of capturing the physical regions to which the logical regions are currently mapped, whereas 
-unpacking takes those physical regions and gives them fresh logical names within the evaluation of the body.
-
-Finally, although the semantics for calls to subtasks appear complicated at first glance, the bulk of the rules
-are just handling the evaluation of each actual argument expression and the creation of the called subtask's 
-environment (with potentially different names for logical regions).  The key difference in Legion task calls is the
+The environment also has a {\em clobber set} $C$.
+We will show that parallel Legion programs can be scheduled in
+a hierarchical fashion.  We therefore require an
+operational semantics that is parallel and hierarchical.  Now
+Legion allows limited nondeterminism, specifically because of relaxed coherence on
+regions shared between parallel tasks and choices of reduction orderings.
+Thus, heap accesses performed by one expression may observe side-effects from 
+concurrent evaluations.  The traditional solution to this issue
+is to use a top-down, small-step semantics with a global view of all concurrently executing expressions (e.g., \cite{CML}).  
+%Rules are specified for which
+%expression(s) are valid choices for the next small step of execution.
+This approach is incompatible with a hierarchical view in which local
+decisions are made with limited information.
+
+Instead we use a bottom-up approach with big-step semantics.
+Each expression has a clobber set $C$ of memory locations that might 
+be modified by concurrent evaluations.  While a read from a location not in $C$ is {\em coherent}
+(returning the most recently written value), the value read from a location $l$ in $C$ may be any
+value.  Determing which value is deferred until all potentially-concurrent
+modifications to $l$ are observable in the evaluation of a parent expression whose clobber set does
+not include $l$ and a consistent interleaving of the subexpressions' memory traces is determined.
+%With this formulation, the criteria for when a hierarchical scheduler may execute subexpressions in parallel can be
+%made clear, but they are no longer constructive.  Instead, we will have to show that the scheduling algorithm
+%used by the Legion runtime satisfies these constraints, guaranteeing the soundness of the Legion programming model.
+
+%FIXME  need to add much earlier that the complete report has all missing type and operational rules and proofs
+
+We show the most interesting operational rules for Core Legion in
+Figure~\ref{fig:langrules}.  The rules for the {\em read}, {\em
+write}, and {\em reduce} show how dynamic memory traces are
+constructed and how the result of a read expression is relaxed when
+the location falls within the clobber set.  Additionally, the write
+and reduce operations, with their two subexpressions, demonstrate how
+the traces of subexpressions are interleaved within the environment
+(and clobber set) of the parent expression.  (A formal definition of
+the {\em valid\_interleave} predicate can be found in
+Figure~\ref{fig:validinterleave}, and will be discussed in
+Section~\ref{sec:coherence}.)
+
+%The first use of the region mapping $M$ occurs in the {\em new} and {\em downregion} expressions.  
+The {\em new} expression selects a location (with the right heap typing) from those assigned to 
+the logical region argument.
+%(Although one would expect the new expression to also return a location that is not
+%currently in use, our soundness result does not require this additional constraint, so we have omitted it for
+%simplicity.)  
+Similarly, {\em downregion} checks whether a location is within the set assigned to the logical region.
+%logical region(s) have been mapped to, returning {\em null} if not.  
+As discussed above, 
+{\em upregion} checks are performed statically.
+
+% FIXME The explanation of newcolor is confusing.
+%
+%The general form of a coloring being an abitrary mapping of locations to ``colors'' (integers) is shown in the
+%operational semantics rules for the {\em newcolor} and {\em color} expressions.  
+The {\em color} expression updates an existing coloring with a new color for a location, replacing any
+existing coloring for that location.  The behavior of {\em newcolor} is subtler.
+Legion supports first partitioning regions and then allocating data in the subregions.
+%the {\em partition-then-allocate} model (see Section~\ref{sec:example}) in which the {\em new} expression may be called
+%on subregions, 
+Thus, each subregion resulting from a partitioning may need an arbitrary number of locations that are
+unknown at partitioning time.  The semantics of newcolor allows this while
+guaranteeing all explicit colorings performed by the program are honored.
+
+%The semantics for the {\em pack} and {\em unpack} expressions are straightforward.  
+Packing a region relationship captures the physical regions to which the logical regions are mapped;
+unpacking simply gives those physical regions fresh logical names.
+
+Finally, calls to subtasks evaluate actual arguments and create the subtasks's
+environment (with potentially different names for logical regions).  A key point in Legion task calls is the
 handling of relaxed coherence.  Any location that falls within
@@
 
-A key property of the Legion type system is that any expression that is well-typed is
-guaranteed to access the heap only in ways permitted by the privileges under which it was typed.
-A judgment $\privconsist{E}{\Phi}$ holds if all the memory operations in memory trace $E$ are of the types and
-to locations covered by the privileges in $\Phi$:
+Our first results shows a well-typed expression accesses the heap in ways consistent with its static privileges.
+A judgment $\privconsist{E}{\Phi}$ holds if memory operations in memory trace $E$ have types and
+locations covered by privileges $\Phi$:
 \[ 
@@
 
-Our goal is to show $\privconsist{E}{\Phi}$ holds for any execution of an expression $e$ with static privileges $\Phi$.
-As usual, even stating such a claim requires that corresponding parts of the initial static and dynamic environments be consistent.
+\noindent
+%We must show $\privconsist{E}{\Phi}$ holds for any execution of an expression $e$ with static privileges $\Phi$.
+As usual, the soundness claim is proven assuming the initial type and execution environments are consistent.
 For our results, three consistency properties are needed:
 \begin{itemize}
-\item mapping consistency, written $\mapconsist{\Omega}$, guarantees a region mapping $M$ satisfies the region constraints $\Omega$ under
-which an expression was typed
-\item local value consistency, written $\localconsist{L}{\Gamma}$, guarantees local values in $L$ have types consistent with the environment under which the expression was typed
-\item store consistency, written $\storeconsist{S}{H}$, guarantees every location in the store $S$ contains a value consistent with the location's type in the heap typing $H$
+\item mapping consistency, written $\mapconsist{\Omega}$, guarantees a region mapping $M$ satisfies the region constraints $\Omega$% under which an expression was typed
+\item local value consistency, written $\localconsist{L}{\Gamma}$, guarantees local values in $L$ have types consistent with the environment $\Gamma$
+%under which the expression was typed
+\item store consistency, written $\storeconsist{S}{H}$, guarantees locations in $S$ have values consistent with heap typing $H$
 \end{itemize}
@@
 \end{itemize}
-Figure~\ref{fig:constprop} gives the precise definition of these properties.
+Figure~\ref{fig:constprop} defines these properties.
 
@@
 
-The proof, which can be found in \cite{LegionTypes12}, is lengthy but
-straightforward due to the way in which the operational semantics were
-constructed.  It proceeds by standard induction on the structure of the
-derivation in the operational semantics, at each step showing that
-subexpressions have consistent starting environments and that the
-value and trace consistency of the subexpressions is sufficient to prove
-the consistency of the parent expression.
+The proof, which can be found in the submitted appendix, is lengthy but
+straightforward due to the compositional operational semantics.
+%It proceeds by standard induction on the structure of the
+%derivation in the operational semantics, at each step showing that
+%subexpressions have consistent starting environments and that the
+%value and trace consistency of the subexpressions is sufficient to prove
+%the consistency of the parent expression.
 

### 2012-11-01 apps/fdtd/fdtd.cc
FDTD: Checkpoint for debugging.

@@ -90,6 +90,7 @@ enum {
   TOP_LEVEL_TASK,
   MAIN_TASK,
   INIT_TASK,
+  SOURCE_TASK,
   STEP_TASK,
   DUMP_TASK,
 };
@@ -109,6 +110,15 @@ enum dim_t {
   NDIMS = 3, // Must be last entry in enum.
 };
 
+static inline const char * dim_name(dim_t dim) {
+  switch (dim) {
+  case DIM_X: return "x";
+  case DIM_Y: return "y";
+  case DIM_Z: return "z";
+  default: assert(0 && "unreachable");
+  }
+}
+
 ////////////////////////////////////////////////////////////////////////
 // During the computation, each block will only require access to
 // ghost cells in one direction. Thus each block will partitioned six
@@ -119,6 +129,14 @@ enum dir_t {
   DIR_NEG = 1,
   NDIRS = 2, // Must be last entry in enum.
 };
+
+static inline const char * dir_name(dir_t dir) {
+  switch (dir) {
+  case DIR_POS: return "+";
+  case DIR_NEG: return "-";
+  default: assert(0 && "unreachable");
+  }
+}
 
 static inline int sign(dir_t dir) {
   return dir == DIR_POS ? 1 : -1;
@@ -197,6 +215,29 @@ struct InitGlobalArgs {
 };
 
 ////////////////////////////////////////////////////////////////////////
+// Arguments to source_task.
+////////////////////////////////////////////////////////////////////////
+struct SourceGlobalArgs {
+  SourceGlobalArgs(int nx, int ny, int nz, dim_t dim, FieldID field)
+    : n(nx, ny, nz), dim(dim), field(field) {}
+  // Number of cells.
+  vec3 n;
+  // Dimension to source.
+  dim_t dim;
+  // Field to update.
+  FieldID field;
+};
+
+struct PositionLocalArgs {
+  PositionLocalArgs(std::pair<int, int> &x_span,
+                    std::pair<int, int> &y_span,
+                    std::pair<int, int> &z_span)
+    : min(x_span.first, y_span.first, z_span.first),
+      max(x_span.second, y_span.second, z_span.second) {}
+  vec3 min /* inclusive */, max /* exclusive */;
+};
+
+////////////////////////////////////////////////////////////////////////
 // Arguments to step_task.
 ////////////////////////////////////////////////////////////////////////
 struct StepGlobalArgs {
@@ -213,15 +254,6 @@ struct StepGlobalArgs {
   double dtdx;
[…]

### 2012-11-01 doc/types-pldi2013/legion-types.tex

@@
 \begin{abstract}
-Many distributed-memeory parallel applications dynamically partition 
+Many distributed-memory parallel applications dynamically partition 
 data, but current language designs either 

### 2012-11-01 apps/fdtd/fdtd.cc
FDTD: Update comment.

@@ -746,7 +746,7 @@ void main_task(const void *input_args, size_t input_arglen,
     f.wait_all_results();
   }
 
-  // Preload argument map for step task.
+  // Preload argument map for source and step tasks.
   ArgumentMap position_arg_map = runtime->create_argument_map(ctx);
   for (int bx = 0; bx < nbx; bx++) {
     for (int by = 0; by < nby; by++) {

### 2012-11-02 doc/types-pldi2013/langdef.tex
coherence

@@
 
-The proof, which can be found in the submitted appendix, is lengthy but
-straightforward due to the compositional operational semantics.
+The proof, which can be found in the submitted appendix, is made
+straightforward by the compositional operational semantics.
 %It proceeds by standard induction on the structure of the
@@
 it does not rely on the definition of {\em valid\_interleave}.  Weakening 
-{\em valid\_interleave} to {\em any\_interleave} (Figure~\ref{fig:opsemfns}), 
-which only requires any arbitrary interleaving of memory
-traces from subexpressions, 
-would have been a sufficient condition for the proof.  
-The additional constraints in {\em valid\_interleave}
-(Figure~\ref{fig:validinterleave}) address the {\em coherence} of heap accesses.
-Coherence defines the permissible ordering of operations to each individual heap 
-location by parallel computations.  The constraints enforced by {\em valid\_interleave}
-specify exactly what interleavings of memory traces are permitted based on
-the per-region coherence modes introduced by Legion for supporting parallelism
-in the presence of aliasing.
+{\em valid\_interleave} to {\em any\_interleave} (Figure~\ref{fig:validinterleave}), 
+which  allows arbitrary interleavings of memory traces from subexpressions, 
+is sufficient for the proof of soundness.  
+The stronger constraints in {\em valid\_interleave}
+(Figure~\ref{fig:validinterleave}) address the coherence of heap accesses;
+{\em valid\_interleave} specifies permitted interleavings of memory operations for
+the particular coherence modes on logical regions.
 
@@
 \item for locations that have only access in {\em simult} mode or are in the clobber set, no constraints are
-enforced.  The validity of the interleaving of these accesses will be performed in the enclosing expression.
+enforced.  The validity of the interleaving of these accesses is performed by some enclosing expression.
 \end{itemize}
@@
 
-Although it is our goal to enable parallel execution whenever possible, we briefly
-consider sequential execution of a Legion program.  We show that a sequential execution trivially
-satisfies the interleaving criteria required by the Legion operational semantics.  Our proof
+%Although it is our goal to enable parallel execution whenever possible, we briefly
+%consider sequential execution of a Legion program.  
+We now show that a sequential execution trivially
+satisfies the interleaving criteria required by the operational semantics.  Our proof
 of the soundness of parallel scheduling depends on this result.
@@
 
-Sequential execution is achieved by ignoring the coherence mode $Q$ in all function calls, using $Q' = \emptyset$
-instead, and simply interleaving traces by concatenating the subexpressions' traces in program order.  By ignoring
-the coherence modes, the clobber set remains empty, making the result of all {\em read} expressions fully
-determined.  The following lemma and theorem (proofs of which can be found in \cite{LegionTypes12}) show that the
-value and memory trace that result from a sequential execution of an expression are always valid executions of
-Legion programs.
+Sequential execution ignores the coherence mode $Q$ in all function calls, using $Q' = \emptyset$
+instead, and interleaves traces by concatenating the subexpressions' traces in program order.  By ignoring
+the coherence modes, the clobber set remains empty and the result of all {\em read} expressions is fully
+determined.  The following lemma and theorem show that the
+value and memory trace that result from a sequential execution are always valid executions.
 
@@
 %privileges and the current region mapping can soundly predict when this property will hold.
-
-We define a {\em non-interference} operator $\epsilon_1 \nonint \epsilon_2$ on two memory operations as follows:
+We define a {\em non-interference} operator $\epsilon_1 \nonint \epsilon_2$ on two memory operations:
 $$
@@
 $$
-Accesses to different locations cannot effect each other, and reads have no side-effects.  The safety of the third case
+Accesses to different locations cannot affect each other, and reads have no side-effects.  The safety of the third case
 follows from the requirement that reduction operations be commutative.  
-
-The next step is to show that this non-interference operator is strong enough to preserve the validity of an interleaving
+We next show that this non-interference operator preserves the validity of an interleaving
 of two or more traces when reordering a pair of memory operations.
@@
 \end{lem}
-The proof follows from the definitions of {\em coherent} and {\em seq\_equiv} in Figure~\ref{fig:validinterleave}.
-
-We now extend the notion of non-interference to whole memory traces, with two traces being non-interfering if no operation from one
+%The proof follows from the definitions of {\em coherent} and {\em seq\_equiv} in Figure~\ref{fig:validinterleave}.
+
+\noindent Two whole memory traces are non-interfering if no operation from one
 trace interferes with any operation from the other:
 $$E_1 \nonint E_2 \Leftrightarrow \displaystyle\bigwedge_{\epsilon_1 \text{ in } E_1, \epsilon_2 \text{ in } E_2} \epsilon_1 \nonint \epsilon_2$$
-
-If whole memory traces are non-interfering, then any interleaving of them can be sorted via 
+If whole memory traces are non-interfering, any interleaving can be sorted via 
 pairwise swaps to match the sequential memory
-trace. This gives us the following result which permits safe parallel execution:
+trace. This gives us a result permitting safe parallel execution:
 
@@
 
-We now show that we can turn this result into an efficient runtime test for non-interface that can be 
-performed task execution.  The key to this test is the privileges on Legion tasks and their ability to
-provide bounds on the accesses that may be performed during execution.  Our non-interference operator is
-extended to work on privileges and sets of privileges:
+\noindent 
+We now use the bounds that static privileges place on 
+runtime accesses to give an efficient runtime test for non-interference.  We first extend the non-interference operator 
+to work on privileges:
 
@@
 
-The cases where both subtasks have read-only privileges or both subtasks have reduce-only privileges have exact equivalents
+\noindent The cases where both subtasks have read-only privileges or both subtasks have reduce-only privileges have equivalents
 for regions, which can be tested statically.  Detecting the case where the two sets of memory addresses are disjoint is 
@@
 statically infer non-interference.  The second uses the region mapping $M$ to dynamically determine the
-disjointness of the two regions.  Although still a dynamic test, it is performed
-once per pair of regions rather than for every pair of memory operations.  Details on how to perform
-the dynamic test efficiently at runtime can be found in \cite{Legion12}.
+disjointness of the two regions.  Although a dynamic test, it is performed
+once per pair of regions rather than for every pair of memory operations.  An algorithm
+to perform the dynamic test efficiently is given in \cite{Legion12}.
 %and often need not be performed at all
 %(if one of the other three conditions is shown to hold at compile-time).
-
-As region non-interference is an approximation of memory trace non-interference, we must show that it is sound.  We
-state the Lemma here, and give the proof in \cite{LegionTypes12}.
+As region non-interference is an approximation of memory trace non-interference, we must show that it is sound.  
+%We state the Lemma here, and give the proof in \cite{LegionTypes12}.
 
@@
 We now state the theorem that allows the Legion runtime to perform hierarchical
-and parallel scheduling of non-interfering tasks:
+and parallel scheduling of non-interfering tasks.
 
@@
 
-The proof follows directly from Lemmas~\ref{lem:noninteffects}
-and \ref{lem:nonintpriv}.  This result holds even if the
-clobber set $C$ is non-empty, allowing two locally independent
-subtasks to run in parallel even if one or both of them is interacting
-(in a programmer-permitted way) with a third subtask.
-
-We now highlight an important aspect of the Legion
+%The proof follows directly from Lemmas~\ref{lem:noninteffects} and \ref{lem:nonintpriv}.  
+\noindent This result holds even if the
+clobber set $C$ is non-empty, allowing locally independent
+subtasks to run in parallel even if they interact
+(in a programmer-permitted way) with another  subtask.
+
+We highlight an important aspect of a Legion
 implementation that is different from other systems and relies on the
 soundness of privileges.  Dynamic non-interference of 
-memory operations is something that can only be determined after 
+memory operations can only be determined after 
 evaluation of an expression is completed, and only at great expense, 
 as illustrated by work on transactional
-memory \cite{Harris05}.  However, many systems are willing to pay this cost to
-extract parallelism from operations that access aliased data. At the other extreme are systems like 
+memory \cite{Harris05}.  
+%However, many systems are willing to pay this cost to
+%extract parallelism from operations that access aliased data. 
+At the other extreme are systems like 
 Jade \cite{Rinard98} and DPJ \cite{Bocchino11} that check non-interference 
@@
 In cases where Legion cannot safely infer non-interference of
-privileges, either because the programmer did not sufficiently partition the regions to expose disjointness or because the two
-tasks actually access the same data in aliased regions, parallel execution can
+privileges (perhaps because two tasks actually access the same data in aliased regions) 
+parallel execution can
 still be requested by the programmer through the use of relaxed coherence 
@@
 %be avoided through the use of relaxed coherence.
-
-The first form of relaxed coherence is the {\em atomic} mode, which specifies that although two
-tasks access interfere due to accessing aliased regions, they may execute in either order, allowing
-the task issued later in program order to run as soon as its ready, even before the task issued earlier 
+The {\em atomic} coherence mode specifies that although two
+tasks interfere due to accessing aliased regions, they may execute in either order, allowing
+the task issued later in program order to possibly run before the task issued earlier 
 in program order.  This relaxation only applies if all aliased regions are annotated with atomic coherence.
 %or a more relaxed coherence mode (see Section~\ref{subsec:coherencesimult}).
-
-To show that this is safe, we define a relaxed version of non-interference for atomic coherence as
-follows:
+To show this is safe, we define a relaxed version of non-interference for atomic coherence:
 $$
@@
 
-The other way in which coherence can be relaxed is using the {\em simult} mode, which allows
+Coherence also can be relaxed using the {\em simult} mode, which allows
 multiple tasks to access the same region concurrently.  The simult coherence mode is appropriate in two
@@
 \begin{enumerate}
-\item When subtasks are accessing disjoint data, but the disjointness is hard to describe (e.g. walking
-separate linked lists that have been allocated in the same region)
+\item When subtasks are accessing disjoint data, but the disjointness is difficult to describe (e.g. walking
+separate linked lists that have been allocated in the same region).
 \item When the algorithm is tolerant of non-determinism (e.g. in a breadth-first search, setting the parent
@@
 
-To support the {\em simult} coherence mode, the non-interference test is extended 
+\noindent To support the {\em simult} coherence, the non-interference test is extended 
 %When one or more regions have been marked as using the simult coherence mode, 
 %the non-interference test is relaxed analagously to the atomic coherence case, yielding a $\nonint^S$ operator.
-with a $\nonint^S$ operator analogous to the one for the atomic coherence case.
+with a $\nonint^S$ operator, analogous to $\nonint^A$ for atomic coherence.
 Because the rules for valid interleavings exclude locations that are only accessed in simult mode, it is straightforward
@@
 
-It is also possible to have both atomic and simult coherence modes in use at the same time for different regions in a task call.
+It is also possible to have both atomic and simult coherence modes at the same time for different regions in a task call.
 In this case the non-interference test $\nonint^{AS}$ uses both the atomic and simult relaxations, and

### 2012-11-02 doc/types-pldi2013/conclusion.tex
minor edit
updates

@@
 
-Modern architectures have dramatically increased in complexity in recent years.  To program this class
-of machines, new programming systems will be required that are capable about reasoning about the structure
-of data including aliasing.  We have presented a type system for the Legion programming system that
-reasons about aliasing using a combination of static and dynamic analyses.  Furthermore, we've proven
-the soundness of our type system enables a hierarchical, distributed scheduling algorithm capable
-of scaling on large, distribute memory machines.  This result demonstrates that allowing programmers
-to provide descriptions of data that include aliasing is not prohibitive to high performance.
+%Modern architectures have dramatically increased in complexity in recent years.  To program this class
+%of machines, new programming systems will be required that are capable about reasoning about the structure
+%of data including aliasing.  
+We have presented static and dymamic semantics for the Legion programming system,
+showing how a combination of static and dynamic checks can be used to soundly and efficiently schedule
+Legion programs in parallel, even in the presence of dynamically partitioned and aliased data.  We have
+also given a novel compositional parallel semantics, permiting a precise treatment of relaxed coherence modes;
+in particular we have shown the Legion design is sound even with relaxed coherence.
 

### 2012-11-02 doc/types-pldi2013/evaluation.tex
updates

@@
 
-We evaluate the Legion type system on four criteria:
-\begin{itemize}
-\item Expressivity: Can real applications be written (Section~\ref{subsec:expressivity})?
-\item Overhead: Does it reduce dynamic checking costs (Section~\ref{subsec:overhead})?
-\item Scalability: Can it enable hierarchical scheduling (Section~\ref{subsec:scalability})?
-\item Performance: Does the performance increase from relaxed coherence modes 
-warrant the increased complexity in the operational semantics (Section~\ref{subsec:relaxedperf})?
-\end{itemize}
+We evaluate the design of Legion's static and dynamic semantics on four criteria:
+expressivity (can real applications be written---Section~\ref{subsec:expressivity}),
+overhead  (what are the dynamic checking costs---Section~\ref{subsec:overhead}),
+scalability (can it enable hierarchical scheduling---Section~\ref{subsec:scalability}),
+and performance  (does the performance increase from relaxed coherence modes 
+warrant the increased semantic complexity---Section~\ref{subsec:relaxedperf}).
 %Our evaluation uses a prototype implementation of the Legion language and programming model.
@@
 for the language introduced in Section~\ref{sec:legioncore} and a C++ runtime library
-for executing programs written in the Legion programming model\cite{Legion12}.  All experiments
+for executing programs written in the Legion programming model.  All experiments
 are conducted on the Keeneland supercomputer\cite{Keeneland}.  Each node of the Keeneland
@@
 
-\subsubsection{Fluid Simulation}
-\label{subsec:fluid}
+%\subsubsection{Fluid Simulation}
+%\label{subsec:fluid}
 Fluid is based on the {\tt fluidanimate} benchmark from the PARSEC 
@@
 Circuit, Fluid creates and partitions regions before
-allocating cells in them (i.e., partition-then-allocate).  Another difference is that Fluid
+allocating cells in them. 
+%(i.e., partition-then-allocate).  
+Another difference is that Fluid
 maintains separate regions for ghost cells rather using multiple partitions of
@@
 
-\subsubsection{Adaptive Mesh Refinement}
-\label{subsec:amr}
+%\subsubsection{Adaptive Mesh Refinement}
+%\label{subsec:amr}
+
 The third application is a Legion port of an adaptive mesh refinement (AMR) benchmark 
@@
 %irregular (Circuit) pointer data structures.
+
 Being able to dynamically create and partition regions at runtime is crucial to Legion's
@@
 Having multiple partitions of regions is necessary for describing the many ways 
-that data can be accessed (Circuit, AMR).  All the types of privileges and coherence we have
-described are needed in different applications.  Region relationships
+that data can be accessed (Circuit, AMR).  All the types of privileges and coherence 
+are needed in some application;  region relationships
 are used in all applications.   Finally, all applications introduce aliasing of
@@
 these applications both type check and execute, proving that our type system is
-sufficiently expressive to handle real-world applications written in the Legion programming model.
+sufficiently expressive to handle real-world applications.
+% written in the Legion programming model.
 %, region relationships convey disjointness information.

### 2012-11-02 doc/types-pldi2013/langdef.tex
updates

@@
 
-The enforcement of a (locally) valid interleaving at every step in the operational semantics
-captures the hierarchical nature of the scheduler used by the Legion runtime.  As the cost of
-a scheduling decision is quadratic in the number of function calls (more precisely, the number of 
-privileges for those calls), it should be clear that a hierarchical scheduling algorithm will
-be vastly more efficient than a global algorithm that must consider the privileges of every
-concurrently executing expression.
-However, a scheduling algorithm that makes decisions hierarchically must do so with limited
-information, and must respect all required scheduling dependencies while ideally imposing as
-few extra dependencies as possible.  The following theorem forms the basis of the 
-Legion scheduler.
+Because testing non-interference of tasks is a pairwise operation, scheduling $n$ tasks can require ${\cal O}(n^2)$
+tests.  Thus, a scheduler that must globally consider all pairs of tasks will be impractical for large machines 
+and large numbers of tasks.  The following theorem, however, shows that Legion programs enjoy a locality property that
+limits the scope of needed non-interference tests.
+
+%The enforcement of a (locally) valid interleaving at every step in the operational semantics
+%captures the hierarchical nature of the scheduler used by the Legion runtime.  As the cost of
+%a scheduling decision is quadratic in the number of function calls (more precisely, the number of 
+%privileges for those calls), it should be clear that a hierarchical scheduling algorithm will
+%be vastly more efficient than a global algorithm that must consider the privileges of every
+%concurrently executing expression.
+%However, a scheduling algorithm that makes decisions hierarchically must do so with limited
+%information, and must respect all required scheduling dependencies while ideally imposing as
+%few extra dependencies as possible.  The following theorem forms the basis of the 
+%Legion scheduler.
 
@@
 \rm
-Let $e_1$ and $e_2$ be well-typed Legion expressions using privileges $\Phi_1$ and
+Let $e_1$ and $e_2$ be well-typed expressions using privileges $\Phi_1$ and
 $\Phi_2$ respectively, where $\Phi_1 \nonint[M] \Phi_2$.  Let $e_3$ be a subexpression of $e_1$ and $e_4$ be a 
 subexpression of $e_2$.  Any memory traces $E_3$ of $e_3$ and $E_4$ of $e_4$ resulting from evaluation of
-$e_1$ and $e_2$ (with the usual consistent $M$, $L$, $H$, and $S$), must be
+$e_1$ and $e_2$ (with the usual consistent $M$, $L$, $H$, and $S$) are 
 non-interfering.
@@
 
-Although determining that $e_1$
-and $e_2$ are non-interfering guarantees $e_3$ and $e_4$ are
-non-interfering, the converse is not true.  It is possible for $e_1$
-and $e_2$ to be judged to be interfering and be executed sequentially 
-even when the subexpressions are non-interfering.  This loss of
-precision can impact performance, and did in a few instances in our
-experiments.  
+%
+% FIXME
+% This is a nice and useful discussion, but also not on the main track and good candidate for space savings.
+% Marking this with a FIXME so we can reconsider if we like.
+%
+% 
+%Although determining that $e_1$
+%and $e_2$ are non-interfering guarantees $e_3$ and $e_4$ are
+%non-interfering, the converse is not true.  It is possible for $e_1$
+%and $e_2$ to be judged to be interfering and be executed sequentially 
+%even when the subexpressions are non-interfering.  This loss of
+%precision can impact performance, and did in a few instances in our
+%experiments.  
 %When the performance impact is significant, two options
 %exist to improve matters.  
-In cases where $e_1$ and $e_2$ are in fact
-non-interfering, region partitions can be further refined to make the
-disjointness visible in the privileges.  Alternately, the code can be
-refactored to lift the subexpressions up to the same level, allowing
-the dynamic test based on the region mapping to infer non-interference
-in more cases.  
+%In cases where $e_1$ and $e_2$ are in fact
+%non-interfering, region partitions can be further refined to make the
+%disjointness visible in the privileges.  Alternately, the code can be
+%refactored to lift the subexpressions up to the same level, allowing
+%the dynamic test based on the region mapping to infer non-interference
+%in more cases.  
 %Although this can increase the overhead of scheduling

### 2012-11-02 doc/types-pldi2013/related.tex
updates

@@
 dynamically, allowing programmers to compute data organization at runtime; like Sequoia, DPJ
-partition schemes must be statically decided.  Legion allows 
+partition schemes are static.  Legion allows 
 programmers to create multiple partitions of the same region to give different 
@@
 allows specification at the coarser granularity of tasks and not individual regions.
-DPJ does not support any modes analogous to simultaneous coherence.
+%DPJ does not support any modes analogous to simultaneous coherence.
 %DPJ only supports the 
@@
 %while Legion provides safe execution even in simultaneous environments.  
-There is also a difference in emphasis: DPJ requires shared memory, while Legion 
-is designed for distributed heterogeneous machines.
+%There is also a difference in emphasis: DPJ requires shared memory, while Legion 
+%is designed for distributed heterogeneous machines.
 
@@
 \cite{Boyapati03} including ones that leverage nested regions for describing
-relationships \cite{Clarke02,Cameron07}.  However, ownership type and effect systems
+relationships (e.g., \cite{Clarke02}).  However, ownership type and effect systems
 are primarily used for reasoning about determinism in object oriented languages and
 don't capture the range of disjointness properties that can be specified in Legion.
-
-Reasoning about disjoint heap data is the strong suit of separation logic \cite{Reynolds02}.  
-Concurrent separation logic\cite{Brookes04} has been 
-used both to parallelize sequential programs\cite{Raza09,Gotsman07} and to provide 
-a mechanism for reasoning about independence\cite{Hayman06}.
+Reasoning about disjoint data is the strong suit of separation logic \cite{Reynolds02}.  
+%Concurrent separation logic\cite{Brookes04} has been 
+%used both to parallelize sequential programs\cite{Raza09,Gotsman07} and to provide 
+%a mechanism for reasoning about independence\cite{Hayman06}.
 While we have borrowed some separation logic notation, we ultimately chose to use a 

### 2012-11-02 doc/types-pldi2013/evaluation.tex
end

@@
 Our prototype implementation has two components: a type checker
-for the language introduced in Section~\ref{sec:legioncore} and a C++ runtime library
+for the language of Section~\ref{sec:legioncore} and a C++ runtime library
 for executing programs written in the Legion programming model.  All experiments
-are conducted on the Keeneland supercomputer\cite{Keeneland}.  Each node of the Keeneland
-machine consists of two Xeon 5660 CPUs, three Tesla M2090 GPUs, and 24 GB of DRAM.  Nodes
+are conducted on the Keeneland supercomputer\cite{Keeneland}.  Each node of Keeneland
+consists of two Xeon 5660 CPUs, three Tesla M2090 GPUs, and 24 GB of DRAM.  Nodes
 are connected by a QDR Infiniband interconnect.
@@
 We evaluate Legion on three real-world applications.  To qualitatively gauge the 
-expressivity of the Legion type system, we introduce these applications
-by describing features of the Legion type system used in their implementations.
+expressivity of Legion, we introduce these applications
+by describing features used in their implementations.
 The Circuit example was already covered in detail in Section~\ref{sec:example}.
@@
 parallel.  Additional partitions are created for viewing data from coarser and finer levels of
-the simulation.  Two types of region relationships are created: one for describing the pieces at 
-each level of refinement, and another for describing the relationship between pieces at different
+the simulation.  Two types of region relationships are created: one describes pieces at 
+each level of refinement, and another describes relationships between pieces at different
 levels of refinement.  
@@
 
-Being able to dynamically create and partition regions at runtime is crucial to Legion's
-ability to handle applications that make runtime decisions about data organization (AMR).  Legion
-is able to capture both allocate-then-partition (Circuit) and partition-then-allocate (Fluid) algorithms. 
+Dynamically creating and partitioning regions at runtime is crucial to Legion's
+ability to handle applications that make runtime decisions about data organization (AMR).  
+% FIXME
+% I've deemphasized the point about allocate-then-partition vs. partition-then-allocate.
+% If we want to put it back in it needs to be go in multiple places, not just here.
+%Legion is able to capture both allocate-then-partition (Circuit) and partition-then-allocate (Fluid) algorithms. 
 Having multiple partitions of regions is necessary for describing the many ways 
@@
 \label{subsec:overhead}
-Our first Legion implementation consisted of a C++ library of Legion primitives
+The first Legion implementation consisted of a C++ library of Legion primitives
 \cite{Legion12} with no checking of region memory accesses.  
-When testing applications we frequently encountered memory corruption due to
+When testing applications on this system we frequently encountered memory corruption due to
 illegal region accesses caused by application bugs.
@@
 To locate the application bugs causing these illegal accesses, we initially added 
-dynamic checks on  all region accesses for both CPUs and GPUs.  
-Unsurprisingly, these checks added considerable runtime overhead.  
+dynamic checks on  all region accesses for both CPUs and GPUs which  
+added considerable runtime overhead.  
 To preserve the benefit of checking every access, but eliminate the cost of dynamic checks we implemented the type, privilege, and coherence checker we have described. 
-We then rewrote all of our applications in this language and used our type checker to verify them.  Once the applications type checked we were
-able to safely elide all the dynamic region access checks.
+We then rewrote the applications in this language and type checked them, at which point the
+dynamic region access checks could be safely elided.
 
@@
 \ref{fig:amr4096}) checking overhead is the difference
-between better and worse performance than the baseline.  For an explanation of overall
-performance relative to the baseline implementations see \cite{Legion12}.
+between better and worse performance than the baseline.  The overall
+performance relative to the baseline implementations is discussed in \cite{Legion12}.
 

### 2012-11-02 doc/types-pldi2013/related.tex
end

@@
 are primarily used for reasoning about determinism in object oriented languages and
-don't capture the range of disjointness properties that can be specified in Legion.
+don't capture the range of disjointness properties in Legion.
 Reasoning about disjoint data is the strong suit of separation logic \cite{Reynolds02}.  
@@
 %a mechanism for reasoning about independence\cite{Hayman06}.
-While we have borrowed some separation logic notation, we ultimately chose to use a 
-privileges system as our primary formalism because separation logic does not easily
+While we have borrowed some separation logic notation, we chose to use a 
+privileges system because separation logic does not easily
 support reasoning about the interleaving of operations to aliased regions of memory.

### 2012-11-02 doc/types-pldi2013/intro.tex
updates
Edits to the intro and abstract

@@
 machines have elaborate features for describing parallelism, but 
-few abstractions for describing the organization of data.  However, as the architecture
-of distributed memory has become increasingly complex, the placement and movement of data has
-become a first order programming concern.
+few abstractions for describing the organization of data.  However, as 
+distributed memory machines increase in complexity,
+reasoning about the organization of program data will be imperative
+for supporting the management of parallelism and the movement of data.
 
@@
 
-Previous work has explored two approaches to expressing the organization of data
-in distributed memory parallel programs.
-One class of work uses fully static analysis 
-with no runtime overhead but limited expressivity.  Two recent examples, 
+Previous work has explored two approaches for describing the organization of data
+in distributed memory parallel programs.  One class of work disallows
+potential data aliasing to enable fully static analyses with no
+runtime overhead, but limited expressivity. Two recent examples, 
 Sequoia \cite{Fatahalian06} and 
@@
 a single tree-shaped partitioning of data that must be fully checked statically 
-(see Section~\ref{sec:related}).  The
-second approach relies on fully dynamic analysis of program data that doesn't
-limit expressivity but incurs runtime overhead and centralized control logic that is
-difficult to scale to distributed memory machines;  examples 
-include transactional memory \cite{Harris05} and thread-level speculation \cite{Steffan00}.
+(see Section~\ref{sec:related}).  The second approach permits potential data
+aliasing and uses fully dynamic analyses of program data that do not 
+limit expressivity.  However, this approach incurs significant runtime overhead and requires 
+centralized control logic that is difficult to scale to distributed 
+memory machines; examples include transactional memory \cite{Harris05} 
+and thread-level speculation \cite{Steffan00}.
 
@@
 In this work, we present the static and dynamic semantics of Legion \cite{Legion12},
-a parallel programming model that efficiently supports multiple, dynamic partitions
-of data. Legion's {\em logical regions} are an
-abstraction capturing {\em locality} (data that will be used together,
-and therefore should be colocated) and {\em independence} (disjoint data
-that may be operated on in parallel).  Logical regions have 
-properties that allow programmers to communicate information
+a parallel programming model that efficiently reason about the data
+aliasing introduced by multiple, dynamic data partitions. 
+Legion's {\em logical regions} are an abstraction capturing {\em locality} 
+(data that will be used together, and therefore should be colocated) 
+and {\em independence} (disjoint data that may be operated on in parallel).  
+Logical regions have properties that allow programmers to communicate information
 about the structure of program data to the Legion implementation:
@@
 \item {\em privilege} and {\em coherence} properties that
-describe how a logical region's data may be used by computations.
+describe how a logical region's data may be used by computations (explained below).
 \end{itemize}
+
+Both dynamically computed partitions
+and simultaneous partitions of the same data introduce the possibility
+of {\em aliasing}---distinct regions may have overlapping data.\footnote{Thus
+  the term {\em logical} regions: language-level regions
+  name sets of data but do not imply a physical layout or placement in
+  the memory hierarchy. A separate system of {\em physical regions} at
+  runtime holds copies of data in the language-level logical regions
+  \cite{Legion12}.}
 
@@
 
-Through multiple partitions of a logical region, a given datum may
-belong to multiple different regions.  Thus, allowing multiple
-simultaneous partitions of the same data introduces
-{\em aliasing}---distinct regions may have overlapping data.\footnote{Thus
-  the term {\em logical} regions: language-level regions
-  name sets of data but do not imply a physical layout or placement in
-  the memory hierarchy. A separate system of {\em physical regions} at
-  runtime holds copies of data in the language-level logical regions
-  \cite{Legion12}.}
+%Through multiple partitions of a logical region, a given datum may
+%belong to multiple different regions.
 
+To support additional parallelism in the presence of aliased data,
 Legion also provides {\em coherence} modes for regions, which allows
-programmers to permit parallel execution even in cases where tasks in
-fact do access the same data in conflicting ways.  The programmer may
+programmers to permit relaxed models execution even in cases where tasks in
+fact do access the same data.  The programmer may
 specify that tasks conflicting on a region  must execute serially (in program
@@
 correctness of programs using first-class regions,
-privilege and coherence properties, and arbitrary region
+privilege and coherence properties in the presence of arbitrary region
 aliasing, Legion relies on a combination of static and
-dynamic checks.  A key theme that runs through the design is that very
-hard static analysis problems (specifically, alias analysis) are often both easy and
-inexpensive if they can be done dynamically at the granularity of regions instead of individual
+dynamic checks.  A key theme that runs through the design is that 
+alias analysis is a very hard static analysis problem, but is both easy and
+inexpensive if they done dynamically at the granularity of regions instead of individual
 heap locations, while checks that must be done on individual values are only
@@
 %  can leverage its dynamic information about region aliasing to
-%  perform dynamic privelege checks guaranteeing safety.
+%  perform dynamic privilege checks guaranteeing safety.
 
@@
 
-\item We prove the soundness of Legion's static type and privelege system (Section~\ref{sec:soundness}).
+\item We prove the soundness of Legion's static type and privilege system (Section~\ref{sec:soundness}).
 

### 2012-11-02 doc/types-pldi2013/legion-types.tex
Edits to the intro and abstract

@@
 Many distributed-memory parallel applications dynamically partition 
-data, but current language designs either 
-require fully static partitioning or rely on expensive dynamic checks.
-We describe Legion, a programming system with a combination
-of language features and static and dynamic checks to support 
-dynamic data partitions with minimal overhead.
-We describe the novel aspects of the Legion design, prove the soundness of the Legion type system,
- and show Legion type checking improves performance by up to 71\% by eliding provably
-safe memory checks.  Furthermore, we show how the language design and type safety
+data in a way that makes static reasoning of data organization difficult 
+due to potential data aliasing. Current language designs either 
+disallow aliasing to enable static analyses of program data, 
+or have permitted aliasing at the cost of expensive dynamic analyses.
+We describe Legion, a programming system that relies on a hybrid system
+of static and dynamic checks to support reasoning about the aliasing introduced
+by dynamic data partitions with minimal overhead.  We describe the novel aspects of the Legion design, 
+prove the soundness of the Legion type system, and show Legion type checking 
+improves performance by up to 71\% by eliding provably safe memory checks.  
+Furthermore, we show how the language design and type safety
 enables a hierarchical, distributed runtime scheduling algorithm, and how
-a system of data coherence annotations improves expressivity without compromising
-soundness. We report results for three real-world applications running
+a system of data coherence annotations improves performance in the presence
+of aliased data.  We report results for three real-world applications running
 on distributed memory machines, achieving up to 62.5X speedup on 96 GPUs 

### 2012-11-02 doc/types-pldi2013/intro.tex
more tweaks to abstract/intro

@@
 Previous work has explored two approaches for describing the organization of data
-in distributed memory parallel programs.  One class of work disallows
-potential data aliasing to enable fully static analyses with no
-runtime overhead, but limited expressivity. Two recent examples, 
-Sequoia \cite{Fatahalian06} and 
+in distributed memory parallel programs.
+
+One class of work uses fully static analyses with no runtime overhead, but must
+disallow potential data aliasing to make the analyis tractable, limiting expressivity.
+Two recent examples, Sequoia \cite{Fatahalian06} and 
 Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
-mechanism to statically partition the heap into a tree of collections 
-of data.  The two designs are different in many aspects, but agree that there is
-a single tree-shaped partitioning of data that must be fully checked statically 
+mechanism to statically partition the heap.
+% into a tree of collections of data.
+The two designs are different in many aspects, but agree that there is
+a single tree-shaped partitioning of data that must be checked statically 
 (see Section~\ref{sec:related}).  The second approach permits potential data
-aliasing and uses fully dynamic analyses of program data that do not 
-limit expressivity.  However, this approach incurs significant runtime overhead and requires 
+aliasing, allowing greater expressivity, but must fully dynamic analyses of
+program execution to detect aliasing situations and guarantee correct results.
+However, this approach incurs significant runtime overhead and requires 
 centralized control logic that is difficult to scale to distributed 
@@
 using the current industry standard mix of MPI, shared-memory
-threads, and CUDA, has taught us that a
-single, static data partitioning is insufficient.  In practice, the
+threads, and CUDA has taught us that a
+single, static data partitioning is often insufficient.  In many cases, the
 best way to partition data is often a function of the data
@@
 In this work, we present the static and dynamic semantics of Legion \cite{Legion12},
-a parallel programming model that efficiently reason about the data
-aliasing introduced by multiple, dynamic data partitions. 
+a parallel programming model that supports multiple, dynamic data partitions and is
+able to efficiently reason about potential aliasing.
 Legion's {\em logical regions} are an abstraction capturing {\em locality} 
@@
 \item logical regions can be dynamically {\em partitioned} into {\em subregions}; 
-the programmer can express arbitrary partitions of regions;
-
-\item  a logical region may be dynamically partitioned in multiple different ways;
-
-\item {\em privilege} and {\em coherence} properties that
-describe how a logical region's data may be used by computations (explained below).
-\end{itemize}
-
-Both dynamically computed partitions
-and simultaneous partitions of the same data introduce the possibility
-of {\em aliasing}---distinct regions may have overlapping data.\footnote{Thus
+the programmer can express arbitrary partitions of regions
+\footnote{Thus
   the term {\em logical} regions: language-level regions
@@
   runtime holds copies of data in the language-level logical regions
-  \cite{Legion12}.}
+  \cite{Legion12}.};
+
+\item  a logical region may be dynamically partitioned in multiple different ways.
+Subregions from multiple partitions may include the same data, introducing aliasing.
+
+%\item {\em privilege} and {\em coherence} properties that
+%describe how a logical region's data may be used by computations (explained below).
+\end{itemize}
+
+%Both dynamically computed partitions
+%and simultaneous partitions of the same data introduce the possibility
+%of {\em aliasing}---distinct regions may have overlapping data.
 
@@
 computation is organized in a hierarchy of tasks.  Logical regions and tasks interact
-through a system of {\em privileges} specifying operations that a task
-can perform on a logical region argument (read, write, or reduce).  
-Tasks can only access logical regions for which they 
-possess privileges.  
+through a system of {\em privileges} that specify which operations (read, write, or reduce)
+a task (or any of its subtasks) may perform on each logical region argument.  
+%Tasks can only access logical regions for which they 
+%possess privileges.  
 %Tasks acquire privileges from their calling task
@@
 To support additional parallelism in the presence of aliased data,
-Legion also provides {\em coherence} modes for regions, which allows
-programmers to permit relaxed models execution even in cases where tasks in
-fact do access the same data.  The programmer may
+Legion provides {\em coherence} modes at the granularity of regions, which allows
+relaxed constraints on the execution order of tasks, even in case where they
+do access the same region.  The programmer may
 specify that tasks conflicting on a region  must execute serially (in program
 order, the default), {\em atomically} with respect to each other, or even
-{\em simultaneously} if the programmer uses their own synchronization mechanisms
-to control access (e.g., locks).
+{\em simultaneously} if the application has alternative methods for ensuring
+determinism or can tolerate non-deterministic behavior in that region..
 
@@
 
-\item Again using the sounds of the type system,  we show that if expressions $e_1$ and $e_2$ are {\em non-interfering} (can be executed in parallel), then subexpressions
+\item Again using the soundness of the type system,  we show that if expressions $e_1$ and $e_2$ are {\em non-interfering} (can be executed in parallel), then subexpressions
 $e_1'$ of $e_1$ and $e_2'$ of $e_2$ are also non-interfering (Section~\ref{sec:scheduling}).  
@@
 
-\item We give experimental evidence for the Legion design.  On three
+\item We give experimental evidence that supports the Legion design choices.  On three
   real-world applications, we show that dynamic region pointer checks

### 2012-11-02 doc/types-pldi2013/legion-types.tex
more tweaks to abstract/intro

@@
 \begin{abstract}
-Many distributed-memory parallel applications dynamically partition 
-data in a way that makes static reasoning of data organization difficult 
-due to potential data aliasing. Current language designs either 
-disallow aliasing to enable static analyses of program data, 
-or have permitted aliasing at the cost of expensive dynamic analyses.
+Applications written for distributed-memory architectures must
+partition their data in order to enable parallel execution.
+Current language designs either analyze the partitioning of data
+statically, disallowing any aliasing of data, or use expensive dynamic
+checks to detect aliasing and ensure correct execution.
 We describe Legion, a programming system that relies on a hybrid system
-of static and dynamic checks to support reasoning about the aliasing introduced
-by dynamic data partitions with minimal overhead.  We describe the novel aspects of the Legion design, 
+of static and dynamic analyses to support reasoning about potential aliasing introduced
+by dynamic data partitions with minimal overhead.  We present the novel aspects of the Legion design, 
 prove the soundness of the Legion type system, and show Legion type checking 

### 2012-11-03 doc/types-pldi2013/example.tex
edits to intro, saving space

@@
   keepspaces=true,
+  belowskip=-10pt,
   literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,

### 2012-11-03 doc/types-pldi2013/intro.tex
edits to intro, saving space

@@
 in distributed memory parallel programs.
-
-One class of work uses fully static analyses with no runtime overhead, but must
-disallow potential data aliasing to make the analyis tractable, limiting expressivity.
+One class of work uses fully static analyses with no runtime overhead, but 
+disallows potential data aliasing to make the analyis tractable, limiting expressivity.
 Two recent examples, Sequoia \cite{Fatahalian06} and 
@@
 (see Section~\ref{sec:related}).  The second approach permits potential data
-aliasing, allowing greater expressivity, but must fully dynamic analyses of
+aliasing, allowing greater expressivity, but uses fully dynamic analyses of
 program execution to detect aliasing situations and guarantee correct results.
-However, this approach incurs significant runtime overhead and requires 
+This approach incurs significant runtime overhead and requires 
 centralized control logic that is difficult to scale to distributed 
@@
 
-\item  a logical region may be dynamically partitioned in multiple different ways.
-Subregions from multiple partitions may include the same data, introducing aliasing.
+\item  a logical region may be dynamically partitioned in multiple different ways;
+subregions from multiple partitions may include the same data, introducing aliasing.
 
@@
 
-To efficiently guarantee the safety and
-correctness of programs using first-class regions,
-privilege and coherence properties in the presence of arbitrary region
-aliasing, Legion relies on a combination of static and
-dynamic checks.  A key theme that runs through the design is that 
-alias analysis is a very hard static analysis problem, but is both easy and
-inexpensive if they done dynamically at the granularity of regions instead of individual
-heap locations, while checks that must be done on individual values are only
-cheap if done statically.  Explicating and evaluating the design of Legion's
+To guarantee the safety and correctness of programs, Legion relies on
+a combination of static and dynamic analysis based on the privilege
+and coherence systems.  
+%To efficiently guarantee the safety and
+%correctness of programs using first-class regions,
+%privilege and coherence properties in the presence of arbitrary region
+%aliasing, Legion relies on a combination of static and
+%dynamic checks.  
+The critical feature in Legion that makes static analysis tractable
+and dynamic analysis efficient in the presence of potential aliasing is logical regions.
+Alias analysis is a very hard static analysis problem, but is both easy and
+inexpensive if done dynamically at the granulatiry of logical regions instead
+of individual memory locations; similarly
+checks on individual memory accesses are expensive if done dynamically, but
+cheap if done statically within the context of local logical region privileges.
+%A key theme that runs through the design is that 
+%alias analysis is a very hard static analysis problem, but is both easy and
+%inexpensive if done dynamically at the granularity of regions instead of individual
+%heap locations, while checks that must be done on individual values are only
+%cheap if done statically.  
+Explicating and evaluating the design of Legion's
 static and dynamic semantics and its implications are the topics of this paper,
@@
 
-\item Again using the soundness of the type system,  we show that if expressions $e_1$ and $e_2$ are {\em non-interfering} (can be executed in parallel), then subexpressions
+\item Again using the soundness of the type system,  we show that if expressions $e_1$ and $e_2$ 
+are {\em non-interfering} (can be executed in parallel), then subexpressions
 $e_1'$ of $e_1$ and $e_2'$ of $e_2$ are also non-interfering (Section~\ref{sec:scheduling}).  

### 2012-11-03 doc/types-pldi2013/legion-types.tex
title, abstract modifications, and making things annonymous

@@
 %\title{A Type System for Structured Aliasing in Parallel Programs}
-\title{Supporting Dynamic Data Partitioning in Parallel Programs}
+%\title{Supporting Dynamic Data Partitioning in Parallel Programs}
 %\title{Scalable and Safe Parallelism using Logical Regions}
-\authorinfo{Sean Treichler \and Michael Bauer \and Alex Aiken}{Stanford University}{\{sjt,mebauer,aiken\}@cs.stanford.edu}
+\title{Supporting Dynamic Data Partitioning in Distributed Memory Architectures with Logical Regions}
+%\authorinfo{Sean Treichler \and Michael Bauer \and Alex Aiken}{Stanford University}{\{sjt,mebauer,aiken\}@cs.stanford.edu}
 %\authorinfo{Sean Treichler}{Stanford Unviersity}{sjt@cs.stanford.edu}
@@
 %\authorinfo{Alex Aiken}{Stanford University}{aiken@cs.stanford.edu}
+\authorinfo{}{}{}
 \maketitle
@@
 We describe Legion, a programming system that relies on a hybrid system
-of static and dynamic analyses to support reasoning about potential aliasing introduced
-by dynamic data partitions with minimal overhead.  We present the novel aspects of the Legion design, 
+of static and dynamic analyses on logical regions to support efficient
+reasoning about potential aliasing introduced by dynamic data partitions.  
+We present the novel aspects of the Legion design, 
 prove the soundness of the Legion type system, and show Legion type checking 

### 2012-11-03 doc/types-pldi2013/intro.tex
minor edits

@@
 reasoning about the organization of program data will be imperative
-for supporting the management of parallelism and the movement of data.
+for supporting the management of parallelism and the movement of data
+through the memory hierarchy.
 
@@
 Previous work has explored two approaches for describing the organization of data
-in distributed memory parallel programs.
+in distributed memory programs.
 One class of work uses fully static analyses with no runtime overhead, but 
@@
 aliasing, allowing greater expressivity, but uses fully dynamic analyses of
-program execution to detect aliasing situations and guarantee correct results.
+program execution to detect data aliasing and guarantee correct results.
 This approach incurs significant runtime overhead and requires 
@@
 \item logical regions can be dynamically {\em partitioned} into {\em subregions}; 
-the programmer can express arbitrary partitions of regions
-\footnote{Thus
+the programmer can express arbitrary partitions of regions;
+
+\item  a logical region may be dynamically partitioned in multiple different ways;
+subregions from multiple partitions may include the same data, introducing 
+aliasing\footnote{Thus
   the term {\em logical} regions: language-level regions
@@
   runtime holds copies of data in the language-level logical regions
-  \cite{Legion12}.};
-
-\item  a logical region may be dynamically partitioned in multiple different ways;
-subregions from multiple partitions may include the same data, introducing aliasing.
+  \cite{Legion12}.}.
 
@@
 To support additional parallelism in the presence of aliased data,
-Legion provides {\em coherence} modes at the granularity of regions, which allows
+Legion provides {\em coherence} modes at the granularity of regions, which allow
 relaxed constraints on the execution order of tasks, even in case where they

### 2012-11-03 doc/types-pldi2013/langdef.tex
bunch of little edits - pushed over 10 pages, but fixable

@@
 %Types include booleans, integers, tuples, and pointers.  
-Pointers are annotated with a list of regions---a non-null pointer must point to a 
+Pointers are annotated with a list of logical regions---a non-null pointer must point to a 
 location that is contained in at least one of the regions. There is a special
@@
 pointer's type.  Note the exact region need not be named in $\Phi$ if permissions
-exist for a logical region that provably contains the pointer's region.  To simplify this check, Figure~\ref{fig:closure}
+exist for logical regions that provably contain the pointer's region(s).  To simplify this check, Figure~\ref{fig:closure}
 defines closure operations $\Omega^*$ (all constraints implied by $\Omega$) and
@@
 %coloring allows for more efficient generation (and storage) of colorings as needed.)  
-Partitioning introduces names for subregions corresponding to each color in the coloring, guaranteeing that each
+Partitioning introduces names for the subregions corresponding to each color in the coloring, guaranteeing that each
 subregion is disjoint from the others and all are subregions of the original region.
 
-The {\em pack} and {\em unpack} expressions allow regions to be stored in the heap.
+The {\em pack} and {\em unpack} expressions allow regions and pointers into those regions to be stored in the heap.
 %  Logical region names are lexically scoped and if stored in the heap as-is, can only be read within the same scope.  
@@
 constructed and how the result of a read expression is relaxed when
-the location falls within the clobber set.  Additionally, the write
-and reduce operations, with their two subexpressions, demonstrate how
+the location falls within the clobber set.  The write
+and reduce operations, with their two subexpressions, also demonstrate how
 the traces of subexpressions are interleaved within the environment
@@
 
-Our first results shows a well-typed expression accesses the heap in ways consistent with its static privileges.
+Our first result shows a well-typed expression accesses the heap in ways consistent with its static privileges.
 A judgment $\privconsist{E}{\Phi}$ holds if memory operations in memory trace $E$ have types and
@@
 An important insight from the proof of Theorem~\ref{thm:effects} is that 
-it does not rely on the definition of {\em valid\_interleave}.  Weakening 
+it does not rely on the full definition of {\em valid\_interleave}.  Weakening 
 {\em valid\_interleave} to {\em any\_interleave} (Figure~\ref{fig:validinterleave}), 
@@
 %privileges and the current region mapping can soundly predict when this property will hold.
-We define a {\em non-interference} operator $\epsilon_1 \nonint \epsilon_2$ on two memory operations:
+We begin by defining a {\em non-interference} operator on two memory operations
+$\epsilon_1 = op_1(l_1, c_1, v_1)$ and $\epsilon_2 = op_2(l_2, c_2, v_2)$:
 $$
 \begin{array}{l@{ }l}
-op_1(l_1, c_1,& v_1)\ \nonint\ op_2(l_2, c_2, v_2) \Leftrightarrow l_1 \not= l_2 \vee \\
-& (op_1 = read \wedge op_2 = read)\ \vee \\
-& (op_1 = reduce_{id_1} \wedge op_2 = reduce_{id_2} \wedge id_1 = id_2)
+\epsilon_1\ \nonint\ \epsilon_2 \Leftrightarrow & (op_1 = read \wedge op_2 = read)\ \vee \\
+& (op_1 = reduce_{id_1} \wedge op_2 = reduce_{id_2} \wedge id_1 = id_2) \vee \\
+& l_1 \not= l_2
 \end{array}
 $$
-Accesses to different locations cannot affect each other, and reads have no side-effects.  The safety of the third case
+Reads have no side effects, and cannot change what another read returns.  The safety of the second case
 follows from the requirement that reduction operations be commutative.  
+Finally, accesses to different locations cannot affect each other.
 We next show that this non-interference operator preserves the validity of an interleaving
@@
 Jade \cite{Rinard98} and DPJ \cite{Bocchino11} that check non-interference 
-statically by disallowing aliasing.  In contrast, Legion checks non-interference of privileges
+statically, but must disallow aliasing to do so.  In contrast, Legion 
+can verify non-interference of privileges
 at runtime, which is much simpler and more efficient than checking
@@
 In cases where Legion cannot safely infer non-interference of
-privileges (perhaps because two tasks actually access the same data in aliased regions) 
-parallel execution can
-still be requested by the programmer through the use of relaxed coherence 
+privileges (perhaps because two tasks actually access the same data in aliased regions),
+relaxation of the constraints on execution order can
+still be requested by the programmer through the use of coherence 
 annotations on individual regions passed to a task.
@@
 Let $e_1$ and $e_2$ be well-typed Legion expressions using privileges $\Phi_1$ and
-$\Phi_2$.  Now let $e_3$ be some subexpression of $e_1$ and $e_4$ be some 
-subexpression of $e_2$, with corresponding privileges $\Phi_3$ and $\Phi_4$.
+$\Phi_2$.  Now let $e_1'$ be some subexpression of $e_1$ and $e_2'$ be some 
+subexpression of $e_2$, with corresponding privileges $\Phi_1'$ and $\Phi_2'$.
 By Theorem~\ref{thm:parallelexec}, if the runtime can verify $\Phi_1 \nonint[M] \Phi_2$, then 
 expressions $e_1$ and $e_2$ may be evaluated in parallel.  The evaluation of the subexpressions
-$e_3$ and $e_4$ will therefore be allowed to run in parallel, and we must show that this is
+$e_1'$ and $e_2'$ will therefore be allowed to run in parallel, and we must show that this is
 permissible.  We do so by proving that any memory traces that can result from the evaluation
@@
 Let $e_1$ and $e_2$ be well-typed expressions using privileges $\Phi_1$ and
-$\Phi_2$ respectively, where $\Phi_1 \nonint[M] \Phi_2$.  Let $e_3$ be a subexpression of $e_1$ and $e_4$ be a 
-subexpression of $e_2$.  Any memory traces $E_3$ of $e_3$ and $E_4$ of $e_4$ resulting from evaluation of
+$\Phi_2$ respectively, where $\Phi_1 \nonint[M] \Phi_2$.  Let $e_1'$ be a subexpression of $e_1$ and $e_2'$ be a 
+subexpression of $e_2$.  Any memory traces $E_1'$ of $e_1'$ and $E_2'$ of $e_2'$ resulting from evaluation of
 $e_1$ and $e_2$ (with the usual consistent $M$, $L$, $H$, and $S$) are 
@@
 \begin{proof}
-Let $E_1$ be the result of an evaluation of $e_1$, with $E_3$ being the memory trace from the
-evaluation of $e_3$ within the expression tree.  $E_1$ will be formed from a tree of valid
+Let $E_1$ be the result of an evaluation of $e_1$, with $E_1'$ being the memory trace from the
+evaluation of $e_1'$ within the expression tree.  $E_1$ will be formed from a tree of valid
 interleavings, each of which must include all the memory operations of each constituent trace.
-By induction over the number of intermediate subexpressions between $e_1$ and $e_3$, we
-can show that any memory operation $\epsilon_3$ that is in $E_3$ must also be in $E_1$.
-Similarly, any memory operation $\epsilon_4$ that is in $E_4$ must also be in $E_2$.  We have
-$E_1 \nonint E_2$ from Lemma~\ref{lem:nonintpriv}, so and therefore $\epsilon_3 \nonint \epsilon_4$.  Since
-this holds for all $\epsilon_3$ in $E_3$ and all $\epsilon_4$ in $E_4$, we have $E_3 \nonint E_4$.
+By induction over the number of intermediate subexpressions between $e_1$ and $e_1'$, we
+can show that any memory operation $\epsilon_1'$ that is in $E_1'$ must also be in $E_1$.
+Similarly, any memory operation $\epsilon_2'$ that is in $E_2'$ must also be in $E_2$.  We have
+$E_1 \nonint E_2$ from Lemma~\ref{lem:nonintpriv}, so and therefore $\epsilon_1' \nonint \epsilon_2'$.  Since
+this holds for all $\epsilon_1'$ in $E_1'$ and all $\epsilon_2'$ in $E_2'$, we have $E_1' \nonint E_2'$.
 \end{proof}

### 2012-11-03 runtime/common.h
Lots of updates to the runtime for having the high level runtime manager multiple processors

@@ -7,8 +7,54 @@
 // of the program including the application code
 // as well as both of the runtimes.
 
+// At this point most of the code has been switched over to untyped pointers
+#ifndef TYPED_POINTERS
+struct ptr_t
+{ 
+public:
+  ptr_t(void) : value(0) { }
+  ptr_t(const ptr_t &p) : value(p.value) { }
+  ptr_t(unsigned v) : value(v) { }
+public:
+  unsigned value; 
+public: 
+#ifdef __CUDACC__
+  __host__ __device__
+#endif
+  inline ptr_t& operator=(const ptr_t &ptr) { value = ptr.value; return *this; }
+
+  inline bool operator==(const ptr_t &ptr) const { return (ptr.value == this->value); }
+  inline bool operator!=(const ptr_t &ptr) const { return (ptr.value != this->value); }
+  inline bool operator< (const ptr_t &ptr) const { return (ptr.value <  this->value); }
+  inline operator bool(void) const { return (value != (unsigned)-1); }
+  inline bool operator!(void) const { return (value == (unsigned)-1); }
+
+  // Addition operation on pointers
+  inline ptr_t operator+(const ptr_t &ptr) const { return ptr_t(value + ptr.value); }
+  inline ptr_t operator+(unsigned offset) const { return ptr_t(value + offset); }
+  inline ptr_t operator+(int offset) const { return ptr_t(value + offset); }
+
+  // Subtraction operation on pointers
+  inline ptr_t operator-(const ptr_t &ptr) const { return ptr_t(value - ptr.value); }
+  inline ptr_t operator-(unsigned offset) const { return ptr_t(value - offset); }
+  inline ptr_t operator-(int offset) const { return ptr_t(value - offset); }
+  
+  inline ptr_t& operator++(void) { value++; return *this; }
+  inline ptr_t operator++(int) { value++; return *this; }
+  inline ptr_t& operator--(void) { value--; return *this; }
+  inline ptr_t operator--(int) { value--; return *this; }
+
+  // Thank you Eric for type cast operators!
+  inline operator unsigned(void) const { return value; } 
+  inline operator int(void) const { return int(value); }
+
+  static inline ptr_t nil(void) { ptr_t p; p.value = (unsigned)-1; return p; }
+};
+
+#else // TYPED_POINTERS
+
 // Forware declaration
-template<typename T> struct ptr_t;
[…]

### 2012-11-03 runtime/default_mapper.cc

@@ -21,8 +21,8 @@ namespace LegionRuntime {
     Logger::Category log_mapper("defmapper");
 
     //--------------------------------------------------------------------------------------------
-    DefaultMapper::DefaultMapper(Machine *m, HighLevelRuntime *rt, ProcessorGroup group) 
-      : runtime(rt), local_group(group), machine(m),
+    DefaultMapper::DefaultMapper(Machine *m, HighLevelRuntime *rt, Processor local) 
+      : runtime(rt), local_proc(local), local_kind(machine->get_processor_kind(local)), machine(m),
         max_steals_per_theft(STATIC_MAX_PERMITTED_STEALS),
         max_steal_count(STATIC_MAX_STEAL_COUNT),
         splitting_factor(STATIC_SPLIT_FACTOR),
@@ -31,11 +31,11 @@ namespace LegionRuntime {
         max_schedule_count(STATIC_MAX_SCHEDULE_COUNT)
     //--------------------------------------------------------------------------------------------
     {
-      log_mapper(LEVEL_SPEW,"Initializing the default mapper for processor group %x",local_group.id);
+      log_mapper(LEVEL_SPEW,"Initializing the default mapper for processor %x",local_proc.id);
       // Get the kind of processor that this mapper is managing
-      const std::set<Processor> &group_procs = group.get_all_processors(); 
-      for (std::set<Processor>::const_iterator it = group_procs.begin();
-            it != group_procs.end(); it++)
+      const std::set<Processor> &locals = local_proc.get_local_processors(); 
+      for (std::set<Processor>::const_iterator it = locals.begin();
+            it != locals.end(); it++)
       {
         local_procs[*it] = machine->get_processor_kind(*it);
       }
@@ -72,20 +72,15 @@ namespace LegionRuntime {
       {
         // Optimize CPUs for latency and GPUs for throughput
         compute_memory_stack(it->first, memory_stacks[it->first], machine, (it->second == Processor::TOC_PROC));
+        assert(!memory_stacks[it->first].empty());
       }
       // Now build our set of similar processors and our alternative processor map
       {
-        const std::set<ProcessorGroup> &all_groups = machine->get_all_groups();
-        for (std::set<ProcessorGroup>::const_iterator it = all_groups.begin();
-              it != all_groups.end(); it++)
+        const std::set<Processor> &all_procs = machine->get_all_processors();
+        for (std::set<Processor>::const_iterator it = all_procs.begin();
[…]

### 2012-11-03 runtime/default_mapper.h

@@ -13,32 +13,19 @@ namespace LegionRuntime {
 
     class DefaultMapper : public Mapper {
     public:
-      // Some Mapping Tags that will be respected
-      // by the default mapper, but not necessarily 
-      // by anything derived from it.
-      enum {
-        MAPTAG_VIRTUAL_MAP_REGION   = (1U << 0),
-        MAPTAG_VIRTUAL_MAP_REGION_0 = (1U << 0),
-        MAPTAG_VIRTUAL_MAP_REGION_1 = (1U << 1),
-        MAPTAG_VIRTUAL_MAP_REGION_2 = (1U << 2),
-        MAPTAG_VIRTUAL_MAP_REGION_3 = (1U << 3),
-        MAPTAG_VIRTUAL_MAP_REGION_4 = (1U << 4),
-        MAPTAG_VIRTUAL_MAP_ANY_REGION = (1U << 5) - 1,
-      };
-    public:
-      DefaultMapper(Machine *machine, HighLevelRuntime *rt, ProcessorGroup group);
+      DefaultMapper(Machine *machine, HighLevelRuntime *rt, Processor local);
       virtual ~DefaultMapper(void);
     public:
       virtual void select_tasks_to_schedule(const std::list<Task*> &ready_tasks, std::vector<bool> &ready_mask);
       virtual bool map_task_locally(const Task *task);
       virtual bool spawn_task(const Task *task);
-      virtual ProcessorGroup select_target_group(const Task *task);
-      virtual Processor select_final_processor(const Task *task, const std::set<Processor> &options);
-      virtual ProcessorGroup target_task_steal(const std::set<ProcessorGroup> &blacklisted);
-      virtual void permit_task_steal(ProcessorGroup thief, const std::vector<const Task*> &tasks,
+      virtual Processor select_target_processor(const Task *task);
+      virtual Processor target_task_steal(const std::set<Processor> &blacklisted);
+      virtual void permit_task_steal(Processor thief, const std::vector<const Task*> &tasks,
                                       std::set<const Task*> &to_steal);
       virtual void slice_index_space(const Task *task, const IndexSpace &index_space,
                                       std::vector<Mapper::IndexSplit> &slice);
+      virtual VariantID select_task_variant(const Task *task, Processor target);
       virtual bool map_region_virtually(const Task *task, Processor target,
                                         const RegionRequirement &req, unsigned index);
       virtual void map_task_region(const Task *task, Processor target, 
@@ -71,7 +58,8 @@ namespace LegionRuntime {
[…]

### 2012-11-03 runtime/legion.cc

@@ -10,8 +10,8 @@
 #define MAX_NUM_PROCS           1024
 #define DEFAULT_MAPPER_SLOTS    8
 #define DEFAULT_OPS             4
-#define MAX_TASK_MAPS_PER_STEP  4
 #define MAX_TASK_WINDOW         1024
+#define MIN_TASKS_TO_PERFORM_SCHEDULING 1
 
 namespace LegionRuntime {
   namespace HighLevel {
@@ -309,38 +309,71 @@ namespace LegionRuntime {
     bool TaskVariantCollection::has_variant(Processor::Kind kind, bool index_space)
     //--------------------------------------------------------------------------
     {
-      bool result = false;
-      for (std::vector<Variant>::const_iterator it = variants.begin();
+      for (std::map<VariantID,Variant>::const_iterator it = variants.begin();
             it != variants.end(); it++)
       {
-        if ((it->proc_kind == kind) && (it->index_space == index_space))
+        if ((it->second.proc_kind == kind) && (it->second.index_space == index_space))
         {
-          result = true;
-          break;
+          return true;
         }
       }
-      return result;
+      return false;
     }
 
     //--------------------------------------------------------------------------
-    void TaskVariantCollection::add_variant(Processor::TaskFuncID low_id, Processor::Kind kind, bool index, bool leaf)
+    VariantID TaskVariantCollection::get_variant(Processor::Kind kind, bool index_space)
+    //--------------------------------------------------------------------------
+    {
+      for (std::map<VariantID,Variant>::const_iterator it = variants.begin();
+            it != variants.end(); it++)
+      {
+        if ((it->second.proc_kind == kind) && (it->second.index_space == index_space))
+        {
+          return it->first;
+        }
+      }
+      log_variant(LEVEL_ERROR,"User task %s (ID %d) has no registered variants for "
+          "processors of kind %d and index space %d",name, user_id, kind, index_space);
+      exit(ERROR_UNREGISTERED_VARIANT);
+      return 0;
+    }
+
+    //--------------------------------------------------------------------------
+    bool TaskVariantCollection::has_variant(VariantID vid)
+    //--------------------------------------------------------------------------
+    {
+      return (variants.find(vid) != variants.end());
+    }
+
+    //--------------------------------------------------------------------------
+    const TaskVariantCollection::Variant& TaskVariantCollection::get_variant(VariantID vid)
     //--------------------------------------------------------------------------
     {
 #ifdef DEBUG_HIGH_LEVEL
-      assert(!has_variant(kind, index));
[…]

### 2012-11-03 runtime/legion.h

@@ -52,9 +52,9 @@ namespace LegionRuntime {
       inline bool operator==(const LogicalRegion &rhs) const;
       inline bool operator<(const LogicalRegion &rhs) const;
     public:
-      inline RegionTreeID get_tree_id(void) const { return tree_id; }
       inline IndexSpace get_index_space(void) const;
       inline FieldSpace get_field_space(void) const;
+      inline RegionTreeID get_tree_id(void) const;
     private:
       // These are private so you can't just arbitrarily change them
       RegionTreeID tree_id;
@@ -82,6 +82,7 @@ namespace LegionRuntime {
     public:
       inline IndexPartition get_index_partition(void) const;
       inline FieldSpace get_field_space(void) const;
+      inline RegionTreeID get_tree_id(void) const;
     private:
       // These are private so you can't just arbitrary change them
       RegionTreeID tree_id;
@@ -152,27 +153,32 @@ namespace LegionRuntime {
         Processor::TaskFuncID low_id;
         Processor::Kind proc_kind;
         bool index_space;
-        bool leaf;
       public:
-        Variant(Processor::TaskFuncID id, Processor::Kind k, bool index, bool lf)
-          : low_id(id), proc_kind(k), index_space(index), leaf(lf) { }
+        Variant(void)
+          : low_id(0) { }
+        Variant(Processor::TaskFuncID id, Processor::Kind k, bool index)
+          : low_id(id), proc_kind(k), index_space(index) { }
       };
     public:
-      TaskVariantCollection(Processor::TaskFuncID uid, const char *n)
-        : user_id(uid), name(n) { }
+      TaskVariantCollection(Processor::TaskFuncID uid, const char *n, const bool lf)
+        : user_id(uid), name(n), leaf(lf) { }
     public:
       bool has_variant(Processor::Kind kind, bool index_space);
+      VariantID get_variant(Processor::Kind kind, bool index_space);
+      bool has_variant(VariantID vid);
+      const Variant& get_variant(VariantID vid);
     protected:
       friend class HighLevelRuntime;
       friend class TaskContext;
       friend class SingleTask;
-      void add_variant(Processor::TaskFuncID low_id, Processor::Kind kind, bool index, bool leaf);
+      void add_variant(Processor::TaskFuncID low_id, Processor::Kind kind, bool index);
       const Variant& select_variant(bool index, Processor::Kind kind);
     public:
       const Processor::TaskFuncID user_id;
       const char *name;
[…]

### 2012-11-03 runtime/legion_ops.cc

@@ -27,7 +27,7 @@ namespace LegionRuntime {
 
     //--------------------------------------------------------------------------
     GeneralizedOperation::GeneralizedOperation(HighLevelRuntime *rt)
-      : Lockable(), active(false), context_owner(false), forest_ctx(NULL), 
+      : Lockable(), Mappable(), active(false), context_owner(false), forest_ctx(NULL), 
         generation(0), runtime(rt)
     //--------------------------------------------------------------------------
     {
@@ -85,6 +85,10 @@ namespace LegionRuntime {
       }
       forest_ctx = NULL;
       context_owner = false;
+      mapper = NULL;
+#ifdef LOW_LEVEL_LOCKS
+      mapper_lock = Lock::NO_LOCK;
+#endif
       active = false;
     }
 
@@ -314,8 +318,6 @@ namespace LegionRuntime {
       mapped_event = UserEvent::create_user_event();
       ready_event = Event::NO_EVENT;
       unmapped_event = UserEvent::create_user_event();
-      mapper = runtime->get_mapper(id);
-      mapper_lock = runtime->get_mapper_lock(id);
       tag = t;
       // Check privileges for the region requirement
       check_privilege();
@@ -341,8 +343,6 @@ namespace LegionRuntime {
       mapped_event = UserEvent::create_user_event();
       ready_event = Event::NO_EVENT;
       unmapped_event = UserEvent::create_user_event();
-      mapper = runtime->get_mapper(id);
-      mapper_lock = runtime->get_mapper_lock(id);
       tag = t;
       // Check privileges for the region requirement
       check_privilege();
@@ -492,12 +492,6 @@ namespace LegionRuntime {
       Context parent = parent_ctx;
 #endif
       parent_ctx = NULL;
-      mapper = NULL;
-#ifdef LOW_LEVEL_LOCKS
-      mapper_lock = Lock::NO_LOCK;
-#else
-      mapper_lock.clear();
-#endif
       tag = 0;
       map_dependent_waiters.clear();
 #ifndef INORDER_EXECUTION
@@ -664,7 +658,14 @@ namespace LegionRuntime {
     //--------------------------------------------------------------------------
     {
       // Enqueue this operation with the runtime
-      runtime->add_to_ready_queue(this);
+      runtime->add_to_ready_queue(parent_ctx->get_executing_processor(), this);
+    }
+
+    //--------------------------------------------------------------------------
+    MapperID MappingOperation::get_mapper_id(void) const
[…]

### 2012-11-03 runtime/legion_ops.h

@@ -17,7 +17,7 @@ namespace LegionRuntime {
      * interface for doing mapping dependence analysis as well
      * as operation activation and deactivation.
      */
-    class GeneralizedOperation : public Lockable { // include Lockable for fine-grained locking inside object
+    class GeneralizedOperation : public Lockable, public Mappable { // include Lockable for fine-grained locking inside object
     public:
       GeneralizedOperation(HighLevelRuntime *rt);
       virtual ~GeneralizedOperation(void);
@@ -45,6 +45,7 @@ namespace LegionRuntime {
       virtual void deactivate(void) = 0; 
       virtual void perform_dependence_analysis(void) = 0;
       virtual bool perform_operation(void) = 0;
+      virtual MapperID get_mapper_id(void) const = 0;
     protected:
       // Called once the task is ready to map
       virtual void trigger(void) = 0;
@@ -99,6 +100,7 @@ namespace LegionRuntime {
       virtual void perform_dependence_analysis(void);
       virtual bool perform_operation(void);
       virtual void trigger(void);
+      virtual MapperID get_mapper_id(void) const;
     private:
       void check_privilege(void);
     private:
@@ -111,13 +113,6 @@ namespace LegionRuntime {
     private:
       std::set<GeneralizedOperation*> map_dependent_waiters;
       std::vector<InstanceRef> source_copy_instances;
-    private:
-      Mapper *mapper;
-#ifdef LOW_LEVEL_LOCKS
-      Lock mapper_lock;
-#else
-      ImmovableLock mapper_lock;
-#endif
       MappingTagID tag;
     };
 
@@ -155,6 +150,7 @@ namespace LegionRuntime {
       virtual void perform_dependence_analysis(void);
       virtual bool perform_operation(void);
       virtual void trigger(void);
+      virtual MapperID get_mapper_id(void) const;
     public:
       Event flush(void);
     private:
@@ -200,13 +196,7 @@ namespace LegionRuntime {
       void initialize_task(Context parent, Processor::TaskFuncID tid,
                       void *args, size_t arglen, 
                       const Predicate &predicate,
-                      MapperID mid, MappingTagID tag, Mapper *mapper,
-#ifdef LOW_LEVEL_LOCKS
-                      Lock map_lock
-#else
-                      ImmovableLock map_lock
-#endif
-                                             );
+                      MapperID mid, MappingTagID tag);
       void set_requirements(const std::vector<IndexSpaceRequirement> &indexes,
[…]

### 2012-11-03 runtime/legion_types.h

@@ -96,6 +96,10 @@ namespace LegionRuntime {
       ERROR_INVALID_REGION_HANDLE = 61,
       ERROR_INVALID_PARTITION_HANDLE = 62,
       ERROR_VIRTUAL_MAP_IN_LEAF_TASK = 63,
+      ERROR_LEAF_MISMATCH = 64,
+      ERROR_INVALID_PROCESSOR_SELECTION = 65,
+      ERROR_INVALID_VARIANT_SELECTION = 66,
+      ERROR_INVALID_MAPPER_OUTPUT = 67,
     };
 
     // enum and namepsaces don't really get along well
@@ -180,10 +184,11 @@ namespace LegionRuntime {
     class TaskArgument;
     class ArgumentMap;
     class FutureMap;
-    class ColoringFunctor;
     class PhysicalRegion;
     class IndexAllocator;
     class FieldAllocator;
+    class IndexIterator;
+    template<typename T> struct ColoredPoints; 
     class Structure;
     class HighLevelRuntime;
     class Mapper;
@@ -194,6 +199,7 @@ namespace LegionRuntime {
     class Collectable;
     class Notifiable;
     class Waitable;
+    class Mappable; // For things that need to have a mapper and lock selelcted 
     class PredicateImpl;
     class FutureImpl;
     class FutureMapImpl;
@@ -262,8 +268,6 @@ namespace LegionRuntime {
     typedef LowLevel::RegionInstance PhysicalInstance;
     typedef LowLevel::Memory Memory;
     typedef LowLevel::Processor Processor;
-    typedef LowLevel::UtilityProcessor UtilityProcessor;
-    typedef LowLevel::ProcessorGroup ProcessorGroup;
     typedef LowLevel::Event Event;
     typedef LowLevel::UserEvent UserEvent;
     typedef LowLevel::Lock Lock;
@@ -272,6 +276,7 @@ namespace LegionRuntime {
     typedef LowLevel::ReductionOpUntyped ReductionOp;
     typedef LowLevel::Machine::ProcessorMemoryAffinity ProcessorMemoryAffinity;
     typedef LowLevel::Machine::MemoryMemoryAffinity MemoryMemoryAffinity;
+    typedef LowLevel::ElementMask::Enumerator Enumerator;
     typedef unsigned int Color;
     typedef unsigned int IndexPartition;
     typedef unsigned int FieldID;
@@ -281,14 +286,16 @@ namespace LegionRuntime {
     typedef unsigned int InstanceID;
[…]

### 2012-11-03 runtime/lowlevel.h

@@ -18,8 +18,6 @@ namespace LegionRuntime {
     class Lock;
     class Memory;
     class Processor;
-    class UtilityProcessor;
-    class ProcessorGroup;
 
     class IndexSpace;
     class IndexSpaceAllocator;
@@ -133,8 +131,8 @@ namespace LegionRuntime {
 
       bool exists(void) const { return id != 0; }
 
-      UtilityProcessor get_utility_processor(void) const;
-      ProcessorGroup get_processor_group(void) const;
+      Processor get_utility_processor(void) const;
+      const std::set<Processor>& get_local_processors(void) const;
 
       typedef unsigned TaskFuncID;
       typedef void (*TaskFuncPtr)(const void *args, size_t arglen, Processor proc);
@@ -146,6 +144,9 @@ namespace LegionRuntime {
 	LOC_PROC, // Latency core
 	UTIL_PROC, // Utility core
       };
+
+      void enable_idle_task(void);
+      void disable_idle_task(void);
 
       // special task IDs
       enum {
@@ -159,27 +160,6 @@ namespace LegionRuntime {
 
       Event spawn(TaskFuncID func_id, const void *args, size_t arglen,
 		  Event wait_on = Event::NO_EVENT) const;
-    };
-
-    class UtilityProcessor : public Processor {
-    public:
-      void enable_idle_task(void);
-      void disable_idle_task(void);
-    };
-
-    class ProcessorGroup {
-    public:
-      typedef unsigned id_t;
-      id_t id;
-      bool operator<(const ProcessorGroup& rhs) const { return id < rhs.id; }
-      bool operator==(const ProcessorGroup& rhs) const { return id == rhs.id; }
-      bool operator!=(const ProcessorGroup& rhs) const { return id != rhs.id; }
-
-      static const ProcessorGroup NO_GROUP;
-      bool exists(void) const { return id != 0; }
-    public:
-      const std::set<Processor>& get_all_processors(void) const;
-      UtilityProcessor get_utility_processor(void) const;
     };
 
     class Memory {
@@ -323,7 +303,7 @@ namespace LegionRuntime {
 
     template <class LHS, class RHS>
     struct ReductionListEntry {
-      ptr_t<LHS> ptr;
+      ptr_t ptr;
       RHS rhs;
     };
 
[…]

### 2012-11-03 runtime/region_tree.cc

@@ -796,6 +796,36 @@ namespace LegionRuntime {
     }
 
     //--------------------------------------------------------------------------
+    LogicalPartition RegionTreeForest::get_partition_subtree(IndexPartition handle, FieldSpace space, RegionTreeID tid)
+    //--------------------------------------------------------------------------
+    {
+#ifdef DEBUG_HIGH_LEVEL
+      assert(lock_held);
+#endif
+      LogicalPartition result(tid, handle, space);
+      if (!has_node(result))
+      {
+        get_node(result);
+      }
+      return result;
+    }
+
+    //--------------------------------------------------------------------------
+    LogicalRegion RegionTreeForest::get_region_subtree(IndexSpace handle, FieldSpace space, RegionTreeID tid)
+    //--------------------------------------------------------------------------
+    {
+#ifdef DEBUG_HIGH_LEVEL
+      assert(lock_held);
+#endif
+      LogicalRegion result(tid, handle, space);
+      if (!has_node(result))
+      {
+        get_node(result);
+      }
+      return result;
+    }
+
+    //--------------------------------------------------------------------------
     void RegionTreeForest::initialize_logical_context(LogicalRegion handle, ContextID ctx)
     //--------------------------------------------------------------------------
     {
@@ -5474,6 +5504,14 @@ namespace LegionRuntime {
         rm.mapper->map_task_region(rm.task, rm.target, rm.tag, rm.inline_mapping,
                                    rm.req, rm.idx, valid_memories, chosen_order, enable_WAR);
       }
+#ifdef DEBUG_HIGH_LEVEL
+      if (valid_memories.empty() && chosen_order.empty())
+      {
+        log_region(LEVEL_ERROR,"Illegal mapper output, no memories specified for first instance of region %d of task %s (ID %d)",
+                                rm.idx, rm.task->variants->name, rm.task->get_unique_task_id());
+        exit(ERROR_INVALID_MAPPER_OUTPUT);
+      }
+#endif
       InstanceView *result = NULL;
       FieldMask needed_fields; 
       // Go through each of the memories provided by the mapper

### 2012-11-03 runtime/region_tree.h

@@ -61,6 +61,8 @@ namespace LegionRuntime {
       LogicalRegion get_partition_subregion(LogicalPartition parent, IndexSpace handle);
       LogicalPartition get_region_subcolor(LogicalRegion parent, Color c);
       LogicalRegion get_partition_subcolor(LogicalPartition parent, Color c);
+      LogicalPartition get_partition_subtree(IndexPartition handle, FieldSpace space, RegionTreeID tid);
+      LogicalRegion get_region_subtree(IndexSpace handle, FieldSpace space, RegionTreeID tid);
     public:
       // Logical Region contexts 
       void initialize_logical_context(LogicalRegion handle, ContextID ctx);
@@ -213,6 +215,7 @@ namespace LegionRuntime {
       FieldSpaceNode* get_node(FieldSpace space);
       RegionNode*     get_node(LogicalRegion handle);
       PartitionNode * get_node(LogicalPartition handle);
+      RegionNode*     get_top_node(RegionTreeID tid);
     public:
       bool has_node(IndexSpace space) const;
       bool has_node(IndexPartition part) const;

### 2012-11-03 runtime/shared_lowlevel.cc

@@ -410,7 +410,8 @@ namespace LegionRuntime {
         ProcessorImpl(pthread_barrier_t *init, const Processor::TaskIDTable &table,
                       Processor p, size_t stacksize, ProcessorImpl *util)
           : init_bar(init), task_table(table), proc(p), 
-            utility(util->get_utility_processor()), utility_proc(util), has_scheduler(false),
+            utility(util->get_utility_processor()), utility_proc(util), 
+            has_scheduler(table.find(Processor::TASK_ID_PROCESSOR_IDLE) != table.end()),
             is_utility_proc(false), remaining_stops(0),
             scheduler_invoked(false), util_shutdown(false)
         {
@@ -439,12 +440,11 @@ namespace LegionRuntime {
         void initialize_state(size_t stacksize);
     public:
         // Operations for utility processors
-        UtilityProcessor get_utility_processor(void) const;
-        ProcessorGroup get_processor_group(void) const;
+        Processor get_utility_processor(void) const;
         void release_user(void);
         void utility_finish(void);
         const std::set<Processor>& get_utility_users(void) const;
-        void add_utility_user(Processor p);
+        void add_utility_user(Processor p, ProcessorImpl *impl);
     public:
 	Event spawn(Processor::TaskFuncID func_id, const void * args,
 				size_t arglen, Event wait_on);
@@ -457,6 +457,7 @@ namespace LegionRuntime {
         void disable_idle_task(void);
     private:
 	void execute_task(bool permit_shutdown);
+        bool perform_scheduling(bool need_lock);
     private:
 	class TaskDesc {
 	public:
@@ -472,7 +473,7 @@ namespace LegionRuntime {
         pthread_barrier_t *init_bar;
 	Processor::TaskIDTable task_table;
 	Processor proc;
-        UtilityProcessor utility;
+        Processor utility;
         ProcessorImpl *utility_proc;
 	std::list<TaskDesc> ready_queue;
 	std::list<TaskDesc> waiting_queue;
@@ -488,6 +489,7 @@ namespace LegionRuntime {
         bool scheduler_invoked;   // for traking if we've invoked the scheduler
         bool util_shutdown;       // for knowing when our utility processor is done
         std::set<Processor> util_users;// Users of the utility processor to know when it's safe to finish
[…]

### 2012-11-03 runtime/utilities.h

@@ -33,6 +33,7 @@ enum {
     LEVEL_SPEW,
     LEVEL_DEBUG,
     LEVEL_INFO,
+    LEVEL_PRINT,
     LEVEL_WARNING,
     LEVEL_ERROR,
     LEVEL_NONE,
@@ -273,6 +274,8 @@ namespace LegionRuntime {
           return "DEBUG";
         case LEVEL_INFO:
           return "INFO";
+        case LEVEL_PRINT:
+          return "PRINT";
         case LEVEL_WARNING:
           return "WARNING";
         case LEVEL_ERROR:

### 2012-11-03 apps/saxpy/saxpy.cc
Updating saxpy to the new interface

@@ -37,7 +37,7 @@ struct Entry {
 struct Block {
   float alpha;
   LogicalRegion r_x, r_y, r_z;
-  unsigned entry_x[BLOCK_SIZE], entry_y[BLOCK_SIZE], entry_z[BLOCK_SIZE];
+  ptr_t entry_x[BLOCK_SIZE], entry_y[BLOCK_SIZE], entry_z[BLOCK_SIZE];
   unsigned id;
 };
 
@@ -48,18 +48,6 @@ struct MainArgs {
   IndexSpace ispace;
   FieldSpace fspace;
   LogicalRegion r_x, r_y, r_z;
-};
-
-class Coloring : public ColoringFunctor {
-public:
-  virtual bool is_disjoint(void) { return true; }
-  virtual void perform_coloring(IndexSpace color_space, IndexSpace parent_space,
-                                std::map<Color,ColoredPoints<unsigned> > &coloring) {
-    for (unsigned i = 0; i < *get_num_blocks(); i++) {
-      coloring[i] = ColoredPoints<unsigned>();
-      coloring[i].ranges.insert(std::pair<unsigned, unsigned>(BLOCK_SIZE*i, BLOCK_SIZE*(i + 1)-1));
-    }
-  }
 };
 
 float get_rand_float() {
@@ -101,7 +89,7 @@ void top_level_task(const void *args, size_t arglen,
 
   Future f = runtime->execute_task(ctx, TASKID_MAIN, indexes, fields, main_regions,
 				   TaskArgument(&main_args, sizeof(MainArgs)));
-  //f.get_void_result();
+  f.get_void_result();
 
   // Destroy our logical regions clean up the region trees
   runtime->destroy_logical_region(ctx, main_args.r_x);
@@ -130,9 +118,8 @@ void main_task(const void *args, size_t arglen,
 
   // Allocate space in the regions
   std::vector<Block> blocks(*get_num_blocks());
-  printf("Allocating...");
-  unsigned initial_index = alloc.alloc(main_args->num_elems);
-  unsigned next_index = initial_index;
+  ptr_t initial_index = alloc.alloc(main_args->num_elems);
+  ptr_t next_index = initial_index;
   for (unsigned i = 0; i < *get_num_blocks(); i++) {
     blocks[i].alpha = main_args->alpha;
     blocks[i].id = i;
@@ -150,7 +137,10 @@ void main_task(const void *args, size_t arglen,
[…]

### 2012-11-03 apps/fdtd/fdtd.cc
Updating fdtd to the new interface

@@ -303,7 +303,7 @@ static inline int cell_stride(vec3 v, vec3 n) {
 // the computation these cells will be read-only, but they need to be
 // initialized once to hold valid values.
 ////////////////////////////////////////////////////////////////////////
-class InitBlockColoring : public ColoringFunctor {
+class InitBlockColoring {
 public:
   InitBlockColoring(int nx, int ny, int nz,
                     std::vector<std::pair<int, int> > x_divs,
@@ -311,15 +311,14 @@ public:
                     std::vector<std::pair<int, int> > z_divs)
     : nx(nx), ny(ny), nz(nz), x_divs(x_divs), y_divs(y_divs), z_divs(z_divs) {}
 
-  virtual bool is_disjoint(void) { return true; }
+  bool is_disjoint(void) { return true; }
 
-  virtual void perform_coloring(IndexSpace color_space, IndexSpace parent_space,
-                                std::map<Color,ColoredPoints<unsigned> > &coloring) {
+  void perform_coloring(std::map<Color,ColoredPoints<ptr_t> > &coloring) {
     unsigned next_index = 0;
     int nbx = x_divs.size(), nby = y_divs.size(), nbz = z_divs.size();
 
     for (int id = 0; id < nbx*nby*nbz; id++) {
-      coloring[id] = ColoredPoints<unsigned>();
+      coloring[id] = ColoredPoints<ptr_t>();
     }
 
     // Color points in xyz cube.
@@ -342,7 +341,7 @@ public:
           log_app.debug("Assigning points %d..%d to block %d x %d x %d (id %d)",
                         next_index, next_index + block_size, bx, by, bz, id);
           coloring[id].ranges.insert(
-            std::pair<unsigned, unsigned>(next_index, next_index + block_size - 1));
+            std::pair<ptr_t, ptr_t>(next_index, next_index + block_size - 1));
           next_index += block_size;
         }
       }
@@ -364,7 +363,7 @@ private:
 // ghost cells. Each block will then be further sub-divided; see
 // below.
 ////////////////////////////////////////////////////////////////////////
-class OwnedBlockColoring : public ColoringFunctor {
+class OwnedBlockColoring {
 public:
   OwnedBlockColoring(int nx, int ny, int nz,
[…]

### 2012-11-03 doc/types-pldi2013/evaluation.tex
minor evaluation edit

@@
 %\label{subsec:fluid}
-Fluid is based on the {\tt fluidanimate} benchmark from the PARSEC 
+Fluid is a distributed memory version of the {\tt fluidanimate} benchmark from the PARSEC 
 suite\cite{bienia11benchmarking}.  Fluid simulates the flow of an incompressible fluid
@@
 maintains separate regions for ghost cells rather using multiple partitions of
-the regions containing shared data.  Region relationships are used for remembering
+the regions containing shared data.  Region relationships are used to capture
 which regions are required for each grid.
@@
 baseline performance.  In a few cases (Figures~\ref{fig:fluid_overhead} and 
-\ref{fig:amr4096}) checking overhead is the difference
+\ref{fig:amr4096}), the checking overhead is the difference
 between better and worse performance than the baseline.  The overall
@@
 In addition to total processor overhead, we also measured performance gain from eliding checks in 
-terms of wall-clock time.  Since most region accesses occur in leaf tasks the checks parallelize 
+terms of wall-clock time.  Since most region accesses occur in leaf tasks, the checks parallelize 
 well.  Wall-clock performance gains from eliding memory checks ranged from 1-10\%, 1-15\%, 

### 2012-11-03 doc/types-pldi2013/example.tex

@@
 
-Listing~\ref{lst:circuit_ex} shows a circuit simulation written (with
-one exception) in the core Legion language (see
+Listing~\ref{lst:circuit_ex} shows a circuit simulation written
+in the core Legion language (see
 Section~\ref{sec:legioncore}).  The input is a graph of circuit
@@
 
-The entry function is {\tt simulate\_circuit} (line 15), which is
-parameterized on regions {\tt rl}, {\tt rw}, and {\tt rn}.
-The wires and nodes are contained in regions {\tt rw} and {\tt rn},
-respectively.  The {\tt rl} region holds lists of pointers to the nodes and wires.
+The entry function is {\tt simulate\_circuit} (line 15), which accepts a 
+linked list of nodes contained in region {\tt rn} and a linked list of 
+wires contained in region {\tt rw}.  The linked list elements themselves
+are in region {\tt rl}.
 When {\tt simulate\_circuit} is invoked it must have both read and
 write privileges for the regions {\tt rl}, {\tt rw}, and {\tt rn}
-(line 17); the type system statically enforces the functional restriciton that
+(line 17); the type system statically verifies that
 {\tt simulate\_circuit} only accesses these regions and their subregions.
@@
 {\tt color\_circuit} function (line 18-19).  A coloring maps region
-elements to colors (i.e. integers), with each color corresponding
+elements to colors (Core Legion uses integers), with each color corresponding
  to a new subregion
@@
 
-The {\tt color\_circuit} function returns three colorings.  The first
-maps each node to the graph piece that owns it.
+The {\tt color\_circuit} function is responsible for specifying how the
+circuit is to be partitioned.  The body of this function is not shown as the
+ideal algorithm for partitioning data will be application specific.  The 
+only restriction imposed by the Legion programming model is that the result of
+the algorithm be provided in the form of colorings.  For our circuit example,
+there are three colorings.  The first maps each node to the graph piece that owns it.
 The third maps each wire to a graph piece that
@@
 Note that 
-ghost nodes may be on the boundary of multiple pieces and therefore colored
-multiple ways.  The second coloring is a {\em multicoloring} because
-it may map the same element to multiple colors.
+ghost nodes may be on the boundary of multiple pieces and therefore be included in
+multiple ghost regions.  For conciseness, a bit of syntactic sugar is used here the
+form of a {\em multicoloring}, which allows the mapping of an element to multiple 
+colors.  Although multicolorings are not part of the Core Legion described in 
+Section~\ref{sec:legioncore}, they can be easily implemented using a separate 
+coloring for each color in the multicoloring.
 
@@
 and only the inclusion constraints are introduced into the static environment.
-Although multicolorings are not part of the Core Legion described in 
-Section~\ref{sec:legioncore}, they can be easily implemented using a separate 
-coloring for each color in the multicoloring.
 
@@
 bounded existential type.  Region relationships allow the programmer
-to {\em pack} a group of regions together and remember properties
+to {\em pack} a group of regions and pointers into those regions together
+and remember properties
 about them such as disjointness and subregion relationships (lines 29
@@
 of much more complicated code.  Instead, an inexpensive dynamic disjointness
-check by the Legion runtime will discover the parallelism.
+check by the Legion runtime will (re)discover the parallelism.
 
@@
 If not otherwise specified, coherence defaults to {\em exclusive}, 
-meaning the function must appear to execute in program order.  
+meaning the function must appear to execute in program order relative to
+other function calls.  
 Line 14 in Listing~\ref{lst:circuit_leaf} shows

### 2012-11-04 doc/types-pldi2013/intro.tex
intro

@@
 To support additional parallelism in the presence of aliased data,
-Legion provides {\em coherence} modes at the granularity of regions, which allow
-relaxed constraints on the execution order of tasks, even in case where they
-do access the same region.  The programmer may
-specify that tasks conflicting on a region  must execute serially (in program
-order, the default), {\em atomically} with respect to each other, or even
-{\em simultaneously} if the application has alternative methods for ensuring
-determinism or can tolerate non-deterministic behavior in that region..
+Legion provides {\em coherence} modes at the granularity of regions,
+which allow relaxed constraints on task execution order, even
+in case where tasks do access the same region.  The programmer may
+specify that tasks conflicting on a region must execute serially (in
+program order, the default), {\em atomically} with respect to each
+other, or even {\em simultaneously} if the application has alternative
+methods for ensuring determinism or can tolerate non-deterministic
+behavior in that region.
 
-To guarantee the safety and correctness of programs, Legion relies on
-a combination of static and dynamic analysis based on the privilege
-and coherence systems.  
+%To guarantee the safety and correctness of programs, Legion relies on
+%a combination of static and dynamic analysis based on the privilege
+%and coherence systems.  
 %To efficiently guarantee the safety and

### 2012-11-04 doc/types-pldi2013/conclusion.tex
spell checking corrections and space saving measures

@@
 %of data including aliasing.  
-We have presented static and dymamic semantics for the Legion programming system,
+We have presented static and dynamic semantics for the Legion programming system,
 showing how a combination of static and dynamic checks can be used to soundly and efficiently schedule
 Legion programs in parallel, even in the presence of dynamically partitioned and aliased data.  We have
-also given a novel compositional parallel semantics, permiting a precise treatment of relaxed coherence modes;
+also given a novel compositional parallel semantics, permitting a precise treatment of relaxed coherence modes;
 in particular we have shown the Legion design is sound even with relaxed coherence.

### 2012-11-04 doc/types-pldi2013/evaluation.tex
spell checking corrections and space saving measures

@@
 added considerable runtime overhead.  
-To preserve the benefit of checking every access, but eliminate the cost of dynamic checks we implemented the type, privilege, and coherence checker we have described. 
+To preserve the benefit of checking every access without the cost of dynamic checks, we implemented the type, privilege, and coherence checker we have described. 
 We then rewrote the applications in this language and type checked them, at which point the

### 2012-11-04 doc/types-pldi2013/example.tex
spell checking corrections and space saving measures

@@
 {\tt color\_circuit} function (line 18-19).  A coloring maps region
-elements to colors (Core Legion uses integers), with each color corresponding
+elements to colors (integers in Core Legion), with each color corresponding
  to a new subregion
@@
 partitioning scheme that would do well for all graphs.  Dynamic
-partitioning using colorings provides the programmer the flexibility
+partitioning using colorings enables the programmer 
 to make partitioning decisions based on input data.
@@
 while {\tt rn1} will contain all locations with a color of 1.  This
-partitioning uses a coloring and results in {\em disjoint} subpartitions.
+partitioning results in {\em disjoint} subpartitions.
 Constraints are introduced into the static environment describing both the
@@
 relationship---privileges belong to functions. When a function unpacks
-a region relationship it must already hold privileges for the new
-regions it finds in the packed value.  For example, on line 40 the
+a region relationship it must already hold privileges for the unpacked regions. 
+%new regions it finds in the packed value.  
+For example, on line 40 the
 function {\tt execute\_time\_steps} has read/write privileges for

### 2012-11-04 doc/types-pldi2013/intro.tex
spell checking corrections and space saving measures

@@
 One class of work uses fully static analyses with no runtime overhead, but 
-disallows potential data aliasing to make the analyis tractable, limiting expressivity.
+disallows potential data aliasing to make the analysis tractable, limiting expressivity.
 Two recent examples, Sequoia \cite{Fatahalian06} and 
@@
 which allow relaxed constraints on task execution order, even
-in case where tasks do access the same region.  The programmer may
-specify that tasks conflicting on a region must execute serially (in
-program order, the default), {\em atomically} with respect to each
+in cases where tasks access the same region.  The programmer may
+specify that tasks conflicting on a region run: in
+program order (the default), {\em atomically} with respect to each
 other, or even {\em simultaneously} if the application has alternative
@@
 Alias analysis is a very hard static analysis problem, but is both easy and
-inexpensive if done dynamically at the granulatiry of logical regions instead
+inexpensive if done dynamically at the granularity of logical regions instead
 of individual memory locations; similarly
@@
 Explicating and evaluating the design of Legion's
-static and dynamic semantics and its implications are the topics of this paper,
+static and dynamic semantics is the topic of this paper, 
+%and its implications are the topics of this paper,
 which we present as a series of contributions:

### 2012-11-04 doc/types-pldi2013/langdef.tex
spell checking corrections and space saving measures

@@
 because the {\em pack} rule verifies these constraints hold when the 
-region relationship value is created,{\em unpack} can reintroduce these constraints on
-the (fresh) logical region names when the value is read and unpacked from the heap.
+region relationship value is created, {\em unpack} can reintroduce these constraints on
+the (fresh) logical region names when the value is unpacked from the heap.
 Crucially for soundness, region relationships do not capture privileges.
@@
 % 
-Finally, task calls verify the requirement that the privileges held by a called task are a subset of those held
-by the caller.  In addition to being necessary for the proof of soundness, this property is critical to
-enabling hierarchical and distributed scheduling of tasks.
+Task calls verify the requirement that the privileges held by a called task are a subset of those held
+by the caller.  In addition to being necessary for the proof of soundness, this property is critical for 
+hierarchical and distributed scheduling of tasks.
 
@@
 (returning the most recently written value), the value read from a location $l$ in $C$ may be any
-value.  Determing which value is deferred until all potentially-concurrent
+value.  Determining which value is deferred until all potentially-concurrent
 modifications to $l$ are observable in the evaluation of a parent expression whose clobber set does
@@
 
-The proof, which can be found in the submitted appendix, is made
-straightforward by the compositional operational semantics.
+The proof is made straightforward by the compositional operational
+semantics.  The proof, along with proofs of all theorems and lemmas
+in this paper, can be found in the submitted appendix.
+%The proof, which can be found in the submitted appendix, is made
+%straightforward by the compositional operational semantics.
 %It proceeds by standard induction on the structure of the
@@
 $\mapconsist{\Omega}$, $\localconsist{L}{\Gamma}$, and $\storeconsist{S}{H}$.
-If $\Phi_i \nonint[M]^A \Phi_j$ for $1 \leq i < j \leq n$, then for any permuatation $(\oton{\pi})$ of
+If $\Phi_i \nonint[M]^A \Phi_j$ for $1 \leq i < j \leq n$, then for any permutation $(\oton{\pi})$ of
 $(1, \ldots, n)$, $E_{\pi_1} \dplus \ldots \dplus E_{\pi_n}$ is a valid interleaving.

### 2012-11-04 doc/llr-pldi2013/related.tex
First commit on new low-level runtime paper for pldi

--- 2012-07-23 doc/asplos2013/related.tex
@@
 However, the implementation of TAM is designed for a class of much smaller machines
-and therefore is based on much heavier communication and threading mechanisms.
+and therefore is based on heavier communication and threading mechanisms.
 While TAM supports composable asynchronous computation and communication, its

### 2012-11-05 runtime/region_tree.cc
Major bug fix in the detecting of dependences from multiple views accessing the same physical instance

@@ -1345,18 +1345,36 @@ namespace LegionRuntime {
       if (req.handle_type == SINGULAR)
       {
         RegionNode *top_node = get_node(req.region);
+        std::set<InstanceManager*> needed_managers;
         result += top_node->compute_state_size(ctx, packing_mask, 
-                    unique_managers, unique_views, ordered_views, false/*mark invalid views*/, true/*recurse*/);
+                    needed_managers, //unique_views, ordered_views, 
+                    false/*mark invalid views*/, true/*recurse*/);
+        // Now for each of the managers get the set of required views
+        for (std::set<InstanceManager*>::const_iterator it = needed_managers.begin();
+              it != needed_managers.end(); it++)
+        {
+          unique_managers.insert(*it);
+          (*it)->find_views_from(req.region, unique_views, ordered_views, packing_mask);
+        }
       }
       else
       {
         PartitionNode *top_node = get_node(req.partition);
         // Pack the parent state without recursing
+        std::set<InstanceManager*> needed_managers;
         result += top_node->parent->compute_state_size(ctx, packing_mask, 
-                        unique_managers, unique_views, ordered_views,
-                        false/*mark invalid views*/, false/*recurse*/, top_node->row_source->color);
+                        needed_managers, //unique_managers, unique_views, ordered_views,
+                        false/*mark invalid views*/, false/*recurse*/); //, top_node->row_source->color);
         result += top_node->compute_state_size(ctx, packing_mask,
-                        unique_managers, unique_views, ordered_views, false/*mark invalid views*/, true/*recurse*/);
+                        needed_managers, //unique_managers, unique_views, ordered_views, 
+                        false/*mark invalid views*/, true/*recurse*/);
+        for (std::set<InstanceManager*>::const_iterator it = needed_managers.begin();
+              it != needed_managers.end(); it++)
+        {
+          unique_managers.insert(*it);
+          (*it)->find_views_from(top_node->parent->handle, unique_views, ordered_views, 
+                                  packing_mask, top_node->row_source->color);
+        }
       }
       return result;
     }
@@ -1906,8 +1924,16 @@ namespace LegionRuntime {
 #endif
         // Pack the entire state of the tree
         RegionNode *top_node = get_node(req.region);
+        std::set<InstanceManager*> needed_managers;
         result += top_node->compute_state_size(ctx, packing_mask,
[…]

### 2012-11-05 runtime/region_tree.h

@@ -640,17 +640,17 @@ namespace LegionRuntime {
     public:
       size_t compute_state_size(ContextID ctx, const FieldMask &pack_mask,
                                 std::set<InstanceManager*> &unique_managers, 
-                                std::map<InstanceView*,FieldMask> &unique_views,
-                                std::vector<InstanceView*> &ordered_views,
-                                bool mark_invalid_views, bool recurse, int sub = -1);
+                                //std::map<InstanceView*,FieldMask> &unique_views,
+                                //std::vector<InstanceView*> &ordered_views,
+                                bool mark_invalid_views, bool recurse); //, int sub = -1);
       void pack_physical_state(ContextID ctx, const FieldMask &pack_mask,
                                 Serializer &rez, bool invalidate_views, bool recurse);
       void unpack_physical_state(ContextID ctx, Deserializer &derez, bool recurse, unsigned shift = 0);
     public:
       size_t compute_diff_state_size(ContextID, const FieldMask &pack_mask,
                                 std::set<InstanceManager*> &unique_managers,
-                                std::map<InstanceView*,FieldMask> &unique_views,
-                                std::vector<InstanceView*> &ordered_views,
+                                //std::map<InstanceView*,FieldMask> &unique_views,
+                                //std::vector<InstanceView*> &ordered_views,
                                 std::vector<RegionNode*> &diff_regions,
                                 std::vector<PartitionNode*> &diff_partitions,
                                 bool invalidate_views, bool recurse);
@@ -718,8 +718,8 @@ namespace LegionRuntime {
     public:
       size_t compute_state_size(ContextID ctx, const FieldMask &pack_mask,
                                 std::set<InstanceManager*> &unique_managers, 
-                                std::map<InstanceView*,FieldMask> &unique_views,
-                                std::vector<InstanceView*> &ordered_views,
+                                //std::map<InstanceView*,FieldMask> &unique_views,
+                                //std::vector<InstanceView*> &ordered_views,
                                 bool mark_invalid_views, bool recurse);
       void pack_physical_state(ContextID ctx, const FieldMask &mask,
                                 Serializer &rez, bool invalidate_views, bool recurse);
@@ -727,8 +727,8 @@ namespace LegionRuntime {
     public:
[…]

### 2012-11-05 runtime/region_tree.cc
Removing dead code

@@ -1347,8 +1347,7 @@ namespace LegionRuntime {
         RegionNode *top_node = get_node(req.region);
         std::set<InstanceManager*> needed_managers;
         result += top_node->compute_state_size(ctx, packing_mask, 
-                    needed_managers, //unique_views, ordered_views, 
-                    false/*mark invalid views*/, true/*recurse*/);
+                    needed_managers, false/*mark invalid views*/, true/*recurse*/);
         // Now for each of the managers get the set of required views
         for (std::set<InstanceManager*>::const_iterator it = needed_managers.begin();
               it != needed_managers.end(); it++)
@@ -1363,11 +1362,9 @@ namespace LegionRuntime {
         // Pack the parent state without recursing
         std::set<InstanceManager*> needed_managers;
         result += top_node->parent->compute_state_size(ctx, packing_mask, 
-                        needed_managers, //unique_managers, unique_views, ordered_views,
-                        false/*mark invalid views*/, false/*recurse*/); //, top_node->row_source->color);
+                        needed_managers, false/*mark invalid views*/, false/*recurse*/);
         result += top_node->compute_state_size(ctx, packing_mask,
-                        needed_managers, //unique_managers, unique_views, ordered_views, 
-                        false/*mark invalid views*/, true/*recurse*/);
+                        needed_managers, false/*mark invalid views*/, true/*recurse*/);
         for (std::set<InstanceManager*>::const_iterator it = needed_managers.begin();
               it != needed_managers.end(); it++)
         {
@@ -1926,8 +1923,7 @@ namespace LegionRuntime {
         RegionNode *top_node = get_node(req.region);
         std::set<InstanceManager*> needed_managers;
         result += top_node->compute_state_size(ctx, packing_mask,
-                          needed_managers, //unique_managers, unique_views, ordered_views, 
-                          true/*mark invalide views*/, true/*recurse*/);
+                          needed_managers, true/*mark invalide views*/, true/*recurse*/);
         for (std::set<InstanceManager*>::const_iterator it = needed_managers.begin();
               it != needed_managers.end(); it++)
         {
@@ -1952,8 +1948,8 @@ namespace LegionRuntime {
           RegionNode *top_node = get_node(req.region);
           std::set<InstanceManager*> needed_managers;
           result += top_node->compute_diff_state_size(ctx, packing_mask,
-                          needed_managers, //unique_managers, unique_views, ordered_views, 
-                          diff_regions, diff_partitions, true/*invalidate views*/, true/*recurse*/);
[…]

### 2012-11-05 runtime/region_tree.h

@@ -640,8 +640,6 @@ namespace LegionRuntime {
     public:
       size_t compute_state_size(ContextID ctx, const FieldMask &pack_mask,
                                 std::set<InstanceManager*> &unique_managers, 
-                                //std::map<InstanceView*,FieldMask> &unique_views,
-                                //std::vector<InstanceView*> &ordered_views,
                                 bool mark_invalid_views, bool recurse); //, int sub = -1);
       void pack_physical_state(ContextID ctx, const FieldMask &pack_mask,
                                 Serializer &rez, bool invalidate_views, bool recurse);
@@ -649,8 +647,6 @@ namespace LegionRuntime {
     public:
       size_t compute_diff_state_size(ContextID, const FieldMask &pack_mask,
                                 std::set<InstanceManager*> &unique_managers,
-                                //std::map<InstanceView*,FieldMask> &unique_views,
-                                //std::vector<InstanceView*> &ordered_views,
                                 std::vector<RegionNode*> &diff_regions,
                                 std::vector<PartitionNode*> &diff_partitions,
                                 bool invalidate_views, bool recurse);
@@ -718,8 +714,6 @@ namespace LegionRuntime {
     public:
       size_t compute_state_size(ContextID ctx, const FieldMask &pack_mask,
                                 std::set<InstanceManager*> &unique_managers, 
-                                //std::map<InstanceView*,FieldMask> &unique_views,
-                                //std::vector<InstanceView*> &ordered_views,
                                 bool mark_invalid_views, bool recurse);
       void pack_physical_state(ContextID ctx, const FieldMask &mask,
                                 Serializer &rez, bool invalidate_views, bool recurse);
@@ -727,8 +721,6 @@ namespace LegionRuntime {
     public:
       size_t compute_diff_state_size(ContextID, const FieldMask &pack_mask,
                                 std::set<InstanceManager*> &unique_managers,
-                                //std::map<InstanceView*,FieldMask> &unique_views,
-                                //std::vector<InstanceView*> &ordered_views,
                                 std::vector<RegionNode*> &diff_regions,
                                 std::vector<PartitionNode*> &diff_partitions,
                                 bool invalidate_views, bool recurse);
@@ -1039,20 +1031,6 @@ namespace LegionRuntime {
                               std::vector<AliasedCopyUser> &add_aliased_copies,
                               std::vector<AliasedEvent> &add_aliased_events,
[…]

### 2012-11-06 doc/types-pldi2013/legion-types.tex
intro
spell checking corrections and space saving measures

@@
 %\title{Scalable and Safe Parallelism using Logical Regions}
-\title{Supporting Dynamic Data Partitioning in Distributed Memory Architectures with Logical Regions}
+%\title{Supporting Dynamic Data Partitioning in Distributed Memory Architectures with Logical Regions}
+\title{The Design of an Aliasing-Tolerant Parallel Programming Model}
 %\authorinfo{Sean Treichler \and Michael Bauer \and Alex Aiken}{Stanford University}{\{sjt,mebauer,aiken\}@cs.stanford.edu}
@@
 Applications written for distributed-memory architectures must
-partition their data in order to enable parallel execution.
+partition their data to enable parallel execution.
 Current language designs either analyze the partitioning of data
@@
 \input{conclusion}
-
+\vspace{-3mm}
 {
-\small
 \bibliography{bibliography}

### 2012-11-06 apps/fdtd/fdtd.cc
FDTD: Return to "real" math in step kernel.

@@ -972,6 +972,7 @@ void source_task(const void * input_global_args, size_t input_global_arglen,
     for (int y = min[1]; y < max[1]; y++) {
       for (int z = min[2]; z < max[2]; z++) {
         ptr_t i = cell_id(vec3(x, y, z), n);
+        // TODO (Elliott): Real math here.
         f.write(i, f.read<double>(i) + 0.1);
       }
     }
@@ -1035,9 +1036,7 @@ void step_task(const void * input_global_args, size_t input_global_arglen,
       for (int z = rmin[2]; z < rmax[2]; z++) {
         vec3 v = rot3(vec3(x, y, z), -(dim + 1));
         ptr_t i = cell_id(v, n);
-        // Elliott: Debugging
-        f.write(i, 1.0);
-        //f.write(i, f.read(i) - dtdx*(g1b.read(i + s1) - g1a.read(i)) - (g2b.read(i + s2) - g2a.read(i)));
+        f.write(i, f.read<double>(i) - dtdx*(g1b.read<double>(i + s1) - g1a.read<double>(i)) - (g2b.read<double>(i + s2) - g2a.read<double>(i)));
       }
     }
   }

### 2012-11-07 tools/lock_analyzer.py
think this was needed even for the original asplos submission

@@ -127,6 +127,15 @@ def parse_log_file(file_name,items):
     finally:
         f.close()
 
+def bucket(v):
+    if v <= 1 : return v
+    minv = 2
+    maxv = 2
+    while v > maxv:
+        minv = maxv + 1
+        maxv = maxv * 2
+    return "%d-%d" % (minv, maxv)
+
 def sort_lock_items(items,lock_table):
     latest_time = 0.0
     for item in sorted(items,key=lambda i: i.time):
@@ -136,6 +145,11 @@ def sort_lock_items(items,lock_table):
         lock_table[item.idy].add_item(item)
         if item.time > latest_time:
             latest_time = item.time
+    for idy, lck in lock_table.iteritems():
+        #if lck.get_remote_grants() == 0: continue
+        #print idy, ("%x" % (idy,)), len(lck.requests), len(lck.forwards), len(lck.grants), len(lck.releases), lck.get_local_grants(), lck.get_remote_grants()
+        print bucket(len(lck.requests)), bucket((lck.get_remote_grants() if len(lck.requests) > lck.get_remote_grants() else len(lck.requests)))
+    exit(0)
     return latest_time
 
 def plot_request_message_ratios(lock_table,outdir):

### 2012-11-07 benchmarks/reductions/asplos_plot.py
simplified reduction benchmark graph - text to come next

@@ -140,6 +140,31 @@ def make_plots(show = True, save = True, outdir="figs/"):
     fig = plt.figure(figsize = (10,7))
     plt.loglog(basex = 2, basey = 10)
     for perfvals, label, color, marker in (
+        (dense_redfold, "Reduction Instance", tableau5, "o"),
+        (dense_localize, "Localized Instance", tableau18, "d"),
+        (any_redsingle, "Single Reductions", tableau6, "^"),
+        ):
+        print "(%s, %s, %s, %s)" % (perfvals, label, color, marker)
+        plt.plot(nodes, perfvals, "--",
+                 label = label,
+                 color = color, markerfacecolor = color,
+                 linestyle = "solid", markersize = 8, marker = marker,
+                 linewidth = 1)
+    plt.ylim(ymax=14000)
+    plt.legend(loc=2,ncol=1)
+    plt.xlabel('Number of Nodes')
+    plt.ylabel('Reductions per Second (in Millions)')
+    plt.xticks(nodes, nodes)
+    plt.gca().set_xlim([1/1.1,16*1.1])
+    #plt.axis([0,17,100,1000])
+    plt.grid(True)
+
+    if save:
+        fig.savefig(outdir+'/reduce_simple.pdf',format='pdf',bbox_inches='tight')
+
+    fig = plt.figure(figsize = (10,7))
+    plt.loglog(basex = 2, basey = 10)
+    for perfvals, label, color, marker in (
         (dense_redfold, "Fold Instance", tableau5, "o"),
         (dense_redlist, "List Instance", tableau12, "s"),
         (dense_localize, "Localized Instance", tableau18, "d"),

### 2012-11-08 doc/types-pldi2013/example.tex
Adding a new version of the circuit example

@@
+\lstset{
+  captionpos=b,
+  language=Haskell,
+  basicstyle=\scriptsize,
+  numbers=left,
+  numberstyle=\tiny,
+  columns=fullflexible,
+  stepnumber=1,
+  escapechar=\#,
+  keepspaces=true,
+  belowskip=-10pt,
+  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
+  morekeywords={function,rr,int,float,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,multicoloring,color,newcolor,atomic,simultaneous},
+  deletekeywords={float,head,min,max}
+}
+\begin{lstlisting}[float={t},label={lst:circuit_ex},caption={Circuit Simulation}]
+--                        <voltage,current,charge,capacitance>
+type CircuitNode        = <float,float,float,float>
+--                     < owned node, owned or ghost node, resistance, current>
+type CircuitWire<rn,rg>  = <CircuitNode@rn, CircuitNode@(rn,rg),float,float>
+
+type node_list<rl,rn>       = < CircuitNode@rn, node_list<rl,rn>@rl >
+type wire_list<rl,rw,rn,rg>= < CircuitWire<rn,rg>@rw, wire_list<rl,rw,rn,rg>@rl >
+
+type CircuitPiece<rl,rw,rn> = rr[rpw,rpn,rg]
+                            < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl >         
+                            where rpn #$\le$# rn and rg #$\le$# rn and rpw #$\le$# rw and
+                                  rn * rw and rl * rn and rl * rw
+
+-- Simulation initialization and invocation
+function simulate_circuit[rl,rw,rn] ( all_nodes : node_list<rl,rn>@rl, 
+                            all_wires : wire_list<rl,rw,rn,rn>@rl, steps : int ), 
+      reads(rn,rw,rl), writes(rn,rw,rl) : bool = 
+  let pc : <coloring(rn),multicoloring(rn),coloring(rw)> 
+            = color_circuit[rn,rw,rl](all_nodes,all_wires) in
+  -- Disjoint partition for the owned nodes of each piece
+  partition rn using pc.1 as rn0,rn1 in
+  -- Aliased partition for ghost nodes of each piece
+  partition rn using pc.2 as rg0,rg1 in
+  -- Disjoint partition for the owned wires of each piece
+  partition rw using pc.3 as rw0,rw1 in
+  let lists0 : <wire_list<rl,rw0,rn0,rg0>@rl,node_list<rl,rn0>@rl> = 
+        build_lists[rl,rw,rn,rw0,rn0,rg0](all_nodes,all_wires,pc.1,pc.2,pc.3,0) in
+  let piece0 : CircuitPiece<rl,rw,rn> = 
+        pack lists0 as CircuitPiece<rl,rw,rn>[rw0,rn0,rg0] in
+  let lists1 : <wire_list<rl,rw1,rn1,rg1>@rl,node_list<rl,rn1>@rl> =
+        build_lists[rl,rw,rn,rw1,rn1,rg1](all_nodes,all_wires,pc.1,pc.2,pc.3,1) in
+  let piece1 : CircuitPiece<rl,rw,rn> = 
+        pack lists1 as CircuitPiece<rl,rw,rn>[rw1,rn1,rg1] in
+      execute_time_steps[rl,rw,rn](piece0,piece1,steps)
+
+-- Time Step Loop
+function execute_time_steps[rl,rw,rn] ( p0 : CircuitPiece<rl,rw,rn>, 
+      p1 : CircuitPiece<rl,rw,rn>, steps : int ) , reads(rn,rw,rl), writes(rn,rw) : bool = 
+  if steps #$<$# 1 then true else
+  unpack p0 as piece0 : CircuitPiece<rl,rw,rn>[rw0,rn0,rg0] in 
+  unpack p1 as piece1 : CircuitPiece<rl,rw,rn>[rw1,rn1,rg1] in
+  let _ : bool = calc_new_currents[rl,rw0,rn0,rg0](piece0.1) in
+  let _ : bool = calc_new_currents[rl,rw1,rn1,rg1](piece1.1) in
+  let _ : bool = distribute_charge[rl,rw0,rn0,rg0](piece0.1) in
+  let _ : bool = distribute_charge[rl,rw1,rn1,rg1](piece1.1) in
+  let _ : bool = update_voltage[rl,rn0](piece0.2) in
+  let _ : bool = update_voltage[rl,rn1](piece1.2) in
+      execute_time_steps[rl,rw,rn](p0,p1,steps-1)
+
+function color_circuit[rn,rw,rl] ( all_nodes : node_list<rl,rn>@rl, 
+                               all_wires : wire_list<rl,rw,rn>@rl ), 
+        reads(rn,rw,rl) : <coloring(rn), multicoloring(rn), coloring(rw)> =  
+  -- Invoke programmer chosen coloring algorithm (e.g. METIS)
+  -- return owned, ghost, wire colorings
+
+-- Helper method
+function build_lists[rl,rw,rn,rpw,rpn,rg] ( nodes : node_list<rl,rn>@rl, 
+       wires : wire_list<rl,rw,rn>@rl, oc : coloring(rn), gc : multicoloring(rn), 
+       wc : coloring(rw), c : int), reads(rn,rw,rl), writes(rl) 
+       : < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl > = 
+  -- Construct lists of node and wire pointers for the given colorings
+\end{lstlisting}
+
 \section{Circuit Example}

### 2012-11-08 doc/types-pldi2013/intro.tex
Adding a new version of the circuit example

@@
 % Putting this here so that we can get the code in a good place
-\lstset{
-  captionpos=b,
-  language=Haskell,
-  basicstyle=\scriptsize,
-  numbers=left,
-  numberstyle=\tiny,
-  columns=fullflexible,
-  stepnumber=1,
-  escapechar=\#,
-  keepspaces=true,
-  belowskip=-10pt,
-  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
-  morekeywords={function,rr,int,float,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,multicoloring,color,newcolor,atomic,simultaneous},
-  deletekeywords={float,head,min,max}
-}
-\begin{lstlisting}[float={t},label={lst:circuit_ex},caption={Circuit Simulation}]
---                        <voltage,current,charge,capacitance>
-type CircuitNode        = <float,float,float,float>
---                     < owned node, owned or ghost node, resistance, current>
-type CircuitWire<rn,rg>  = <CircuitNode@rn, CircuitNode@(rn,rg),float,float>
-
-type node_list<rl,rn>       = < CircuitNode@rn, node_list<rl,rn>@rl >
-type wire_list<rl,rw,rn,rg>= < CircuitWire<rn,rg>@rw, wire_list<rl,rw,rn,rg>@rl >
-
-type CircuitPiece<rl,rw,rn> = rr[rpw,rpn,rg]
-                            < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl >         
-                            where rpn #$\le$# rn and rg #$\le$# rn and rpw #$\le$# rw and
-                                  rn * rw and rl * rn and rl * rw
-
--- Simulation initialization and invocation
-function simulate_circuit[rl,rw,rn] ( all_nodes : node_list<rl,rn>@rl, 
-                            all_wires : wire_list<rl,rw,rn,rn>@rl, steps : int ), 
-      reads(rn,rw,rl), writes(rn,rw,rl) : bool = 
-  let pc : <coloring(rn),multicoloring(rn),coloring(rw)> 
-            = color_circuit[rn,rw,rl](all_nodes,all_wires) in
-  -- Disjoint partition for the owned nodes of each piece
-  partition rn using pc.1 as rn0,rn1 in
-  -- Aliased partition for ghost nodes of each piece
-  partition rn using pc.2 as rg0,rg1 in
-  -- Disjoint partition for the owned wires of each piece
-  partition rw using pc.3 as rw0,rw1 in
-  let lists0 : <wire_list<rl,rw0,rn0,rg0>@rl,node_list<rl,rn0>@rl> = 
-        build_lists[rl,rw,rn,rw0,rn0,rg0](all_nodes,all_wires,pc.1,pc.2,pc.3,0) in
-  let piece0 : CircuitPiece<rl,rw,rn> = 
-        pack lists0 as CircuitPiece<rl,rw,rn>[rw0,rn0,rg0] in
-  let lists1 : <wire_list<rl,rw1,rn1,rg1>@rl,node_list<rl,rn1>@rl> =
-        build_lists[rl,rw,rn,rw1,rn1,rg1](all_nodes,all_wires,pc.1,pc.2,pc.3,1) in
-  let piece1 : CircuitPiece<rl,rw,rn> = 
-        pack lists1 as CircuitPiece<rl,rw,rn>[rw1,rn1,rg1] in
-      execute_time_steps[rl,rw,rn](piece0,piece1,steps)
-
--- Time Step Loop
-function execute_time_steps[rl,rw,rn] ( p0 : CircuitPiece<rl,rw,rn>, 
-      p1 : CircuitPiece<rl,rw,rn>, steps : int ) , reads(rn,rw,rl), writes(rn,rw) : bool = 
-  if steps #$<$# 1 then true else
-  unpack p0 as piece0 : CircuitPiece<rl,rw,rn>[rw0,rn0,rg0] in 
-  unpack p1 as piece1 : CircuitPiece<rl,rw,rn>[rw1,rn1,rg1] in
-  let _ : bool = calc_new_currents[rl,rw0,rn0,rg0](piece0.1) in
-  let _ : bool = calc_new_currents[rl,rw1,rn1,rg1](piece1.1) in
-  let _ : bool = distribute_charge[rl,rw0,rn0,rg0](piece0.1) in
-  let _ : bool = distribute_charge[rl,rw1,rn1,rg1](piece1.1) in
-  let _ : bool = update_voltage[rl,rn0](piece0.2) in
-  let _ : bool = update_voltage[rl,rn1](piece1.2) in
-      execute_time_steps[rl,rw,rn](p0,p1,steps-1)
-
-function color_circuit[rn,rw,rl] ( all_nodes : node_list<rl,rn>@rl, 
-                               all_wires : wire_list<rl,rw,rn>@rl ), 
-        reads(rn,rw,rl) : <coloring(rn), multicoloring(rn), coloring(rw)> =  
-  -- Invoke programmer chosen coloring algorithm (e.g. METIS)
-  -- return owned, ghost, wire colorings
-
--- Helper method
-function build_lists[rl,rw,rn,rpw,rpn,rg] ( nodes : node_list<rl,rn>@rl, 
-       wires : wire_list<rl,rw,rn>@rl, oc : coloring(rn), gc : multicoloring(rn), 
-       wc : coloring(rw), c : int), reads(rn,rw,rl), writes(rl) 
-       : < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl > = 
-  -- Construct lists of node and wire pointers for the given colorings
-\end{lstlisting}
-
 % Moving this to the circuit example section since it fits better there

### 2012-11-08 doc/types-pldi2013/alt_example.tex
Minor edits

\section{Circuit Example}
\label{sec:example}

\lstset{
  captionpos=b,
  language=Haskell,
  basicstyle=\scriptsize,
  numbers=left,
  numberstyle=\tiny,
  columns=fullflexible,
  stepnumber=1,
  escapechar=\#,
  keepspaces=true,
  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
  morekeywords={function,rr,int,float,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,multicoloring,color,newcolor,atomic,simultaneous},
  deletekeywords={head,min,max}
}

We begin by introducing an example program 
that motivates many of the features in Legion programming model
and provides an intuition for many of our design decisions.  The example program
executes a simulation of an arbitrary graph of electrical circuit elements.  
The simulation runs for many time steps and performs three computations 
for each time step: calculate new currents, distribute charge, and update voltages.
Due to space constraints we only introduce small code examples. 
The source code for the full circuit example written in the core Legion
language (see Section~\ref{sec:legioncore}) can be found in the
submitted appendix.

The simulation takes as input an arbitrary graph of circuit elements
that have been dynamically allocated into two logical regions: {\tt all\_nodes} 
and {\tt all\_wires}.  The types of nodes and wires are the tuples
shown in Listing~\ref{lst:tuples}.  The first two fields in the
{\tt CircuitWire} tuple are pointers that specify the two end points of
a wire.  Pointers are denoted by the @ symbol
followed by the set of region(s) into which a pointer can point.
The {\tt CircuitWire} type is parametrized by the local region
names {\tt rn} and {\tt rg} which contain {\tt CircuitNode} elements.
Our type system enforces the correctness of pointer types.

\begin{lstlisting}[label={lst:tuples},caption={Tuples and Pointers Example}]
--                        voltage,current,charge,capacitance,piece ID
type CircuitNode        = <float,float,float,float,int>
--                      owned node, owned or ghost node, resistance, current
type CircuitWire<rn,rg>  = <CircuitNode@rn, CircuitNode@(rn,rg),float,float>
\end{lstlisting}

The most important decision in any Legion program is the choice of
how to partition program data.  The choice of a partitioning algorithm
can depend on a multitude of factors including the type of data structure
and input data.  Therefore
Legion requires applications to select their own partitioning algorithms.
In this example we invoke a third-party dynamic graph partitioning library
called METIS\cite{Metis98} as shown in Listing~\ref{lst:metis}.

\begin{lstlisting}[label={lst:metis},caption={Partition Computation}]
type NodeList<rl,rn>       = < CircuitNode@rn, node_list<rl,rn>@rl >
type WireList<rl,rw,rn,rg>= < CircuitWire<rn,rg>@rw, wire_list<rl,rw,rn,rg>@rl >
function extern_metis[rl,rn,rw](node_list : NodeList<rl,rn>@rl,
          wires_list : WireList<rl,rw,rn,rn>@rl), reads(rl,rn,rw), writes(rn) : bool;
\end{lstlisting}

METIS takes as arguments lists of all the nodes and all the wires in
the graph.  It then breaks the graph into pieces by annotating each 
{\tt CircuitNode} with a piece ID (the last field in the tuple).
The function {\tt extern\_metis} is parametrized on the regions it will
access; in this case {\tt rn} the region containing nodes, {\tt rw} the region
contining wires, and {\tt rl} the region containing the lists of pointers.  The function
must also specify the {\em privileges} (e.g read) with which it will access
these regions.  {\tt extern\_metis} will read all three regions and write
the {\tt rn} region since it will be writing to the last field of each
{\tt CircuitNode}.  Our type system enforces the Legion 
requirement that each function can only be invoked with a subset of the 
privileges contained in its calling context.

\begin{figure}[t]
\centering
\subfigure[Node region tree.]{
\label{sfig:part_fig:tree}
\includegraphics[scale=0.50]{figs/OwnedTree.pdf}
}
\subfigure[Circuit piece coloring.]{
\label{sfig:part_fig:pieces}
\includegraphics[scale=0.67]{figs/Owned_Nodes.pdf}
}
\subfigure[Ghost {\tt rg0} coloring.]{
\label{sfig:part_fig:ghost_zero}
\includegraphics[scale=0.67]{figs/Ghost_RN0.pdf}
}
\subfigure[Ghost {\tt rg1} coloring.]{
\label{sfig:part_fig:ghost_one}
\includegraphics[scale=0.67]{figs/Ghost_RN1.pdf}
}
\vspace{-2mm} 
\caption{Partitions of $r{\_}all{\_}nodes$. \label{fig:part_fig}}
\vspace{-6mm}
%  \caption[My caption
\end{figure}

Figure~\ref{sfig:part_fig:pieces} shows one possible output from
METIS with different colors identifing different piece IDs.  To capture
the result of a partitioning algorithm, Legion uses {\em colorings}.
Colorings are a mapping from pointers to {\em colors} (e.g. integers).  In
the circuit example, the piece IDs are the designated colors. 
Listing~\ref{lst:coloring} shows the function {\tt create\_piece\_coloring}
that constructs a coloring from a list of nodes which have been 
annotated with piece IDs.

\begin{lstlisting}[label={lst:coloring},caption={Coloring Construction}]
function create_piece_coloring[rl,rn](node_list: NodeList<rl,rn>@rl),
            reads(rl,rn) : coloring(rn) = 
    if isnull(node_list) then newcolor(rn) else
    let list_elem : NodeList<rl,rn> = read(node_list) in
    let part_coloring : coloring(rn) = create_node_coloring(list_elem.2) in
    let node : CircuitNode = read(list_elem.1) in
        color(part_coloring, list_elem.1, node.5)
\end{lstlisting}

Colorings are the input for partition operations in Legion.  The partition
statement shown in Listing~\ref{lst:partition} partitions the region
{\tt all\_nodes} into logical subregions with a separate output region
for each color in the coloring.  Due to space constraints we, only show
code for {\tt rn0} and {\tt rn1} pieces for the remainder of the example.  We also
do not show the coloring or partitioning of the {\tt all\_wires} region as
it can be computed in a straightforward manner from the partitioning of
nodes into pieces.

\begin{lstlisting}[label={lst:partition},caption={Partition Operation Example}]
let piece_coloring : coloring(all_nodes) = create_piece_coloring[rl,all_nodes](node_list) in
partition all_nodes using piece_coloring as rn0,rn1... in ...
\end{lstlisting}

Since colorings can only color each pointer at most with a single color
(dynamically checked), the partition operation results in 
{\em disjoint} subregions.  Furthermore, the partition operation
introduces constraints into the static checking enviornment describing
both the disjointness of the subregions (i.e. {\tt rn0 $*$ rn1})
and the inclusion of each subregion in the original region
(i.e. {\tt rn0 $\leq$ all\_nodes}).

In addition, to partitioning the graph to describe {\em owned} nodes
for each piece, the circuit example also
needs to be able to name the {\em ghost nodes} for each piece.  Ghost node
regions are used for accessing the nodes of wires that span
two different pieces. (This is also the reason {\tt CircuitWire} is parametrized on
two regions).  Figure~\ref{sfig:part_fig:ghost_zero} and 
Figure~\ref{sfig:part_fig:ghost_one} show colorings for the ghost regions
corresponding to {\tt rn0} and {\tt rn1} respectively.  Note that two 
nodes are colored in both colorings.  Since nodes can appear in multiple
subregions, we use a {\em multicoloring} (implemented in core Legion
by multiple single colorings) to describe {\em aliased} partitions.  Aliased
partitions still introduce subregion constraints into the static checking
environment, but not disjointness constraints.  Figure~\ref{sfig:part_fig:tree}
shows the resulting logical region tree for node regions.

The first-class nature of logical regions in Legion allows them to
be written into the heap.  To avoid losing important static properties of
logical regions in this scenario, we use {\em region
relationships}.  Region relationships are a bounded existential type
that allows programmers to {\em pack} a group of regions together and remember
properties about them such as disjointness and subregion relationships.
Listing~\ref{lst:rr} shows an example region relationship for a {\tt CircuitPiece} 
describing its region of wires {\tt rpw}, region of 
nodes {\tt rpn}, region of ghost nodes {\tt rg}, and important constraints.

\begin{lstlisting}[label={lst:rr},caption={Region Relationship Example}]
type CircuitPiece<rl,rw,rn> = rr[rpw,rpn,rg]
                            < WireList<rl,rpw,rpn,rg>@rl, NodeList<rl,rpn>@rl >         
                            where rpn #$\le$# rn and rg #$\le$# rn and rpw #$\le$# rw and
                                  rn * rw and rl * rn and rl * rw
\end{lstlisting}

The Legion type system verifies properties of region relationships when packing;
when region relationships are unpacked the same properties are
reintroduced into the static checking environment.
A key feature underlying the soundness of the Legion
type system is that privileges cannot be packed in a region
relationship---privileges belong to functions. When a function unpacks
a region relationship it must already hold privileges for the unpacked regions.
Listing~\ref{lst:loop} shows the main time step loop run by the 
recursive function {\em execute\_time\_steps} (line 11)
as well the leaf task functions (lines 2-7).  
For each time step, the loop unpacks 
two previously packed circuit pieces, giving new names to the subregions introduced
by each region relationship.  The {\em execute\_time\_steps} function
will have read/write privileges for the newly named regions, such as {\tt rn0},
because it has read/write privileges for {\tt rn} and the {\tt CircuitPiece} 
region relationship declares {\tt rn0 $\leq$ rn}.  Note this does require
a dynamic check that {\tt rn0} is a subregion of {\tt rn}, but the cost of the 
check will be ammortized over all the uses of {\tt rn0}.

The {\tt execute\_time\_steps} function   
illustrates the importance of having different partitions provide 
multiple views onto the same logical region.  Both {\tt calc\_new\_currents} 
and {\tt distribute\_charge} (lines 2-5)
use the owned and ghost regions of a piece, which are from different partitions. In
{\tt calc\_new\_currents} these regions only need read
privileges, allowing both instances of the task to be
run in parallel.  In {\tt distribute\_charge} the
privilege is for a reduction which can also be done in parallel
because of the atomic and commutative nature of reductions.  Finally,
the {\tt update\_voltage} function (line 6) modifies only the disjoint owned regions, 
permitting each piece to be updated in parallel.  No single partition of the nodes
describes these data sharing patterns.

\lstset{
  captionpos=b,
  language=Haskell,
  basicstyle=\scriptsize,
  numbers=left,
  numberstyle=\tiny,
  columns=fullflexible,
  stepnumber=1,
  escapechar=\#,
  keepspaces=true,
  belowskip=-10pt,
  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
  morekeywords={function,rr,int,float,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,multicoloring,color,newcolor,atomic,simultaneous},
  deletekeywords={head,min,max}
}
\begin{lstlisting}[float={t},label={lst:loop},caption={Main Simulation Loop}]
-- Leaf Task Declarations (implementations in appendix)
function calc_new_currents[rl,rw,rn,rg] ( ptr_list : WireList<rl,rw,rn,rg>@rl ), 
      reads(rl,rw,rn,rg), writes(rw) : bool
function distribute_charge[rl,rw,rn,rg] ( ptr_list : WireList<rl,rw,rn,rg>@rl ), 
      reads(rl,rw,rn), reduces(reduce_charge,rn,rg), atomic(rn,rg) : bool 
function update_voltage[rl,rn] ( ptr_list : NodeList<rl,rn>@rl ), 
      reads(rl,rn), writes(rn) : bool
-- Reduction function for distribute charge
function reduce_charge ( node : CircuitNode, current : float ) : CircuitNode

-- Time Step Loop
function execute_time_steps[rl,rw,rn] ( p0 : CircuitPiece<rl,rw,rn>, 
      p1 : CircuitPiece<rl,rw,rn>, steps : int ) , reads(rn,rw,rl), writes(rn,rw) : bool = 
  if steps #$<$# 1 then true else
  unpack p0 as piece0 : CircuitPiece<rl,rw,rn>[rw0,rn0,rg0] in 
  unpack p1 as piece1 : CircuitPiece<rl,rw,rn>[rw1,rn1,rg1] in
  let _ : bool = calc_new_currents[rl,rw0,rn0,rg0](piece0.1) in
  let _ : bool = calc_new_currents[rl,rw1,rn1,rg1](piece1.1) in
  let _ : bool = distribute_charge[rl,rw0,rn0,rg0](piece0.1) in
  let _ : bool = distribute_charge[rl,rw1,rn1,rg1](piece1.1) in
  let _ : bool = update_voltage[rl,rn0](piece0.2) in
  let _ : bool = update_voltage[rl,rn1](piece1.2) in
      execute_time_steps[rl,rw,rn](p0,p1,steps-1)
\end{lstlisting}

The {\tt execute\_time\_steps} function also demonstrates the need to 
dynamically discover parallelism through region disjointess.  
To run the two instances of 
{\tt calc\_new\_currents} in parallel requires knowing that {\tt rw0}
and {\tt rw1} are disjoint (lines 17-18).  There is a similar requirement for parallel
execution of the two {\tt update\_voltage} calls with {\tt rn0} and {\tt rn1} (lines 21-22).  In
both cases, this knowledge was statically available when the subregions were created 
and could have been captured in a region relationship at the cost
of much more complicated code.  Instead, we use a simpler region relationship
and rely on an inexpensive dynamic disjointness
check by the Legion runtime to (re)discover the parallelism.

In addition to privileges, functions can also specify {\em coherence}
on regions.  By default, coherence on all regions is {\em exclusive}, meaning
the function must appear to execute in program order relative to other function
calls that access non-disjoint data.  Programmers can allow other
execution orderings using relaxed coherence.  Line 5 of Listing~\ref{lst:loop} shows
an example of {\tt atomic}, a relaxed coherence mode requiring
that operations to {\tt rn} and {\tt rg} appear atomic relative
to other functions using regions which may alias.  The most relaxed coherence
mode is {\tt simult}; simultaneous coherence allows concurrent access 
to the region by all functions that are using the region
in a simultaneous mode.  The interaction between {\em tasks} (which are functions 
chosen to execute in parallel) using the same region with different coherence
modes is formalized in Section~\ref{sec:coherence}.

%The most important decision in any Legion program is the choice of
%how to partition program data into logical regions.  Logical regions
%are the granularity at which the Legion runtime will reason
%about the independence of tasks, and therefore how much parallelism
%is available.  Logical regions also define the granularity at which
%the Legion runtime will move data.  Coarser regions will lead
%to more efficient bulk copies of data throgh the memory hierarchy due
%to better locality.  There is therefore a natural tension: partitioning 
%should be fine-grained enough to express desired parallelism, but not 
%so fine-grained as to result in inefficient data movement.

%The first step in the circuit example is to partition the arbitrary graph
%of circuit elements into pieces that can be handled independently.
%In the case of this example all nodes in the graph have initially
%been allocated in the region {\tt all\_nodes} (Figure~\label{sfig:part_fig:tree}).
%The decision of how to partition the graph is the responsibility
%of the programmer and can be done by an external library (e.g. METIS
%in this case).  Once the decision of how to partition is made, the
%result must be communicated to the Legion runtime to 
%generate the corresponding logical regions.  Partitioning decisions
%are described via {\em colorings}.  Colorings are parametrized by
%the region they are partitioning and capture a mapping from pointers
%into the region to specific colors (e.g. integers).  Figure~\ref{sfig:part_fig:pvs}
%illustrates the coloring used for generating the {\tt p\_nodes\_pvs}
%partition.

### 2012-11-08 doc/llr-pldi2013/impl.tex
redid reduction stuff, monkeyed with figures to put examples with event graphs

--- 2012-07-23 doc/asplos2013/impl.tex
@@
 
+\texcomment{
 A similar issue can arise with the forwarding of lock request active messages.  If two nodes are
@@
 would eventually catch up.
+}
 
@@
 
-In addition to normal physical regions, our implementation supports two different layouts of data for
-reduction-only physical regions called {\em reduction instances}.   
-These layouts are designed to optimize for the common pattern where 
-multiple tasks make commutative {\em reductions} into shared data.  Instead of having to copy an existing 
-physical region to a suitable memory and then apply individual reduction operations, reduction instances 
-allow a large number of reduction operations to be accumulated and transferred efficiently to the original 
-location of the data. This decoupling from the original location of the data can allow allow greater parallelism
-in cases where processors cannot all access the same memory (e.g. many GPUs).
-
-The first kind of reduction-only physical region is a {\em reduction fold instance}, which may only be
-used for reduction operations that support a {\em fold} operation (see Listing~\ref{lst:regionapi}).  
-A reduction fold instance is similar
-to a normal physical region in that it is implemented as an array, indexed by the same element indices used
-for the normal physical regions.  The difference is that each array element is a value of the reduction's
-{\tt RHS} type which is generally the same size or smaller than the actual element type of the region 
-(i.e. the {\tt LHS} type).  When the client requests a reduction operation to a particular location, 
-the supplied right-hand-side value is folded in
-to the corresponding location in the array.  An arbitrary number of reductions can be folded into each location.
-When the reduction fold instance is copied back to a normal instance, the contents of the reduction instance
-are transferred in a large block to the location of the normal instance.  On the receiving side
-the runtime automatically applies the reduction instance to the destination region
-by invoking the reduction function once for each location in the region via a cache-friendly 
-linear sweep over the memory.
-
-The second kind of reduction instance is a {\em reduction list instance}.  Although it generally doesn't
-perform as well as a reduction fold instance, it can be better for cases with sparse updates.  It is also
-valid option for when a reduction operation doesn't support a {\em fold} operation.  Instead of folding
-together multiple reductions to the same location, a reduction list instance keeps a log of every
-reduction operation requested (i.e. the location and the right-hand-side value).  When the reduction list 
-instance is copied back to a normal physical region, the whole list is transferred and then replayed at
-the target region's location.  Except in sparse update cases, this list is often larger than the target
-region, and any randomness in the sequence of locations in the list can impact the memory access efficiency
-of the reductions to the target region.  However, as we will show in section~\ref{subsec:reducmicro}, 
-the benefits of enabling parallelism and hiding latency usually more than make up for these drawbacks.
+In addition to normal physical region instances, our implementation supports a special reduction-only
+instance that differ in two important ways.  First, the per-element storage in a reduction-only instance
+is sized to hold the ``right-hand side'' of the reduction operation (e.g. the $v$ in an operation like
+$struct.field\ \text{+=}\ v$), allowing a more compact representation.  Second, individual reductions are applied
+(atomically) to the local instance, which can then be sent as a large batched reduction to the target
+instance.  When multiple reductions are made to the same element, they can be {\em folded} together locally,
+further reducing the inter-node communication needed.  Note that the folding operation is not always
+identical to the reduction operation itself.  For example, if the reduction operation is exponentiation,
+the corresponding fold operation is multiplication, according to the following identity:
+$$(r[i]\ \text{**=}\ a)\ \text{**=}\ b \quad \Leftrightarrow \quad r[i]\ \text{**=}\ (a * b).$$

### 2012-11-08 doc/llr-pldi2013/interface.tex
fixed up some of the negative vspaces that are overkill in the PLDI layout
redid reduction stuff, monkeyed with figures to put examples with event graphs

--- 2012-07-23 doc/asplos2013/interface.tex
@@
 class Event {
-public:
   const unsigned id;
@@
 class UserEvent : public Event {
-public:
   static UserEvent create_user_event();
@@
 class Barrier : public Event {
-public:
   static Barrier create_barrier(unsigned expected_arrivals);
@@
 
-\begin{lstlisting}[float={t},label={lst:procapi},caption={Processor Interface and Example.},belowskip=0pt]
+\begin{lstlisting}[float={t},label={lst:procapi},caption={Processor Interface.},belowskip=0pt]
 class Processor {
-public:
   const unsigned id;
@@
 };
+\end{lstlisting}
 
+\begin{figure}
+\begin{lrbox}{\mylistingbox}
+\begin{lstlisting}
 void diamond_task(const void *args,size_t arglen,Processor p){
@@
 \end{lstlisting}
+\end{lrbox}
+\subfigure{\usebox{\mylistingbox}}\\
 
-\begin{figure}
 \centering
@@
 }
-\vspace{-2mm}
-\caption{Processor Example Event Graph\label{fig:procevents}}
+%\vspace{-2mm}
+\caption{Processor Example and Event Graph.\label{fig:procevents}}
 \vspace{-4mm}
@@
 
-\begin{lstlisting}[float={t},label={lst:lockapi},caption={Deferred Lock Interface and Example.}]
+\begin{lstlisting}[float={t},label={lst:lockapi},caption={Deferred Lock Interface.}]
 class Lock {
-public:
   const unsigned id;
@@
 };
-
-void launcher_task(const void *args, size_t arglen, Processor p) {
-  ...
-  // Unpack lock from arguments
-  Lock needed = ...
-  // Acquire lock
-  Event lock_event = needed.lock();
-  // Launch sub-task
-  Event task_event = p.spawn(SUB,NULL,0,lock_event);
-  // Release lock
-  needed.unlock(task_event);
-  ...
-}
 \end{lstlisting}
@@
 \begin{figure}
+\begin{lrbox}{\mylistingbox}
+\begin{lstlisting}
+void launcher_task(const void *args, size_t arglen, Processor p) {
+  // Unpack lock, input events from arguments
+  Lock needed = ...
+  // Acquire lock, launch first subtask, release lock
+  Event lock_event_1 = needed.lock(prev_event_1);
+  Event task_event_1 = p.spawn(SUB, args_1, sizeof(args_1), lock_event_1);
+  needed.unlock(task_event_1);
+  // same for second subtask
+  Event lock_event_2 = needed.lock(prev_event_2);
+  Event task_event_2 = p.spawn(SUB, NULL, 0, lock_event_2);
+  needed.unlock(task_event_2);
+}
+\end{lstlisting}
+\end{lrbox}
+\subfigure{\usebox{\mylistingbox}}\\
+
 \centering
@@
 \begin{tikzpicture}
+  \draw (1.0,1.5) rectangle (6.2,2.4);
+  \node (l1) at (3.6,1.95) {\begin{tabular}{c}{\bf lock}\\needed\end{tabular}};
+
+  \draw (6.7,1.5) rectangle (8.6,2.4);
+  \node (t1) at (7.65,1.95) {\begin{tabular}{c}{\bf run}\\SUB(args\_1)\end{tabular}};
+
+  \draw (9.1,1.5) rectangle (10.5,2.4);
+  \node (u1) at (9.8,1.95) {\begin{tabular}{c}{\bf unlock}\\needed\end{tabular}};
+
+  \draw [->] (0.0,1.95) to (1.0,1.95);
+  \draw [->] (6.2,1.95) to (6.7,1.95);
+  \draw [->] (8.6,1.95) to (9.1,1.95);
+
+  \draw (0.3,0.1) rectangle (1.7,1.0);
+  \node (l2) at (1.0,0.55) {\begin{tabular}{c}{\bf lock}\\needed\end{tabular}};
+
+  \draw (2.2,0.1) rectangle (4.1,1.0);
+  \node (t2) at (3.15,0.55) {\begin{tabular}{c}{\bf run}\\SUB(args\_2)\end{tabular}};
+
+  \draw (4.6,0.1) rectangle (6.0,1.0);
+  \node (u2) at (5.3,0.55) {\begin{tabular}{c}{\bf unlock}\\needed\end{tabular}};
+
+  \draw [->] (0.0,0.55) to (0.3,0.55);
+  \draw [->] (1.7,0.55) to (2.2,0.55);
+  \draw [->] (4.1,0.55) to (4.6,0.55);
+
+  \draw [->,dotted,thick] (0.2,2.5) to (0.5,1.0);
+  \draw [->,dotted,thick] (5.8,1.0) to (6.0,1.5);
+  \draw [->,dotted,thick] (10.3,1.5) to (10.6,0.0);
+
   %\path (1,3) node (t1) [shape=rectangle,draw] {run TASK\_A} -- node [right=1pt] {ea} +(0,-0.5) node[auto] {eax};
-  \node (t1) at (0,0) [shape=rectangle,draw] {\begin{tabular}{c}{\bf lock}\\needed\end{tabular}};
-  \draw (t1) to node[auto] {lock\_event} (2.4,0);
+  %% \node (t1) at (0,0) [shape=rectangle,draw] {\begin{tabular}{c}{\bf lock}\\needed\end{tabular}};
+  %% \draw (t1) to node[auto] {lock\_event} (2.4,0);
 
-  \node (t2) at (3.2,0) [shape=rectangle,draw] {\begin{tabular}{c}{\bf run}\\SUB\end{tabular}};
-  \draw (t2) to node[auto] {task\_event} (5.4,0);
-  \draw [->] (2.4,0) to (t2.west);
+  %% \node (t2) at (3.2,0) [shape=rectangle,draw] {\begin{tabular}{c}{\bf run}\\SUB\end{tabular}};
+  %% \draw (t2) to node[auto] {task\_event} (5.4,0);
+  %% \draw [->] (2.4,0) to (t2.west);
 
-  \node (t3) at (6.4,0) [shape=rectangle,draw] {\begin{tabular}{c}{\bf unlock}\\needed\end{tabular}};
-  \draw [->] (5.4,0) to (t3.west);
+  %% \node (t3) at (6.4,0) [shape=rectangle,draw] {\begin{tabular}{c}{\bf unlock}\\needed\end{tabular}};
+  %% \draw [->] (5.4,0) to (t3.west);
 
-  \draw [->,dotted,thick] (-1,1) to (t1.135);
-  \draw [->,dotted,thick] (t3.45) to (7.4,1);
+  %% \draw [->,dotted,thick] (-1,1) to (t1.135);
+  %% \draw [->,dotted,thick] (t3.45) to (7.4,1);
 \end{tikzpicture}
 }
-\vspace{-2mm}
-\caption{Lock Example Event Graph\label{fig:lockevents}}
+%\vspace{-2mm}
+\caption{Lock Example and Event Graph.\label{fig:lockevents}}
 \vspace{-4mm}
@@
 class Memory {
-public:
   const unsigned id;
@@
 class RegionMetaData {
-public:
   const unsigned id;
@@
   void destroy_instance(RegionInstance instance) const;
-  // Create a physical instance only for reductions
+  // Create a reduction-only instance
   RegionInstance create_instance(Memory memory, 
@@
 class RegionInstance {
-public:
   const unsigned id;
@@
                 Event wait_on = Event::NO_EVENT);
+  Event reduce_to(RegionInstance target, ReductionOpID redopid,
+                Event wait_on = Event::NO_EVENT);
 };
-
-template<typename LHS, typename RHS>
-class ReductionOp {
-public:
-  static void apply(LHS *lhs, RHS rhs);
-  // Optional methods for folding optimization
-  static void fold(RHS *rhs1, RHS rhs2);
-  static const RHS right_identity;
-};
-
-void reduction_ex(const void *args,size_t arglen,Processor p) {
-  Memory m1,m2,m3; Processor p1,p2,p3; ReductionOpID redop = 2;
-  ... // Choose memory and processors
-  RegionMetaData meta = RegionMetaData::
-                        create_region(num_elmts,elmt_size);
-  Instance inst1 = meta.create_instance(m1,redop);
-  Instance inst2 = meta.create_instance(m2,redop);
-  Instance inst3 = meta.create_instance(m3);
-  Event t1 = p1.spawn(REDUC,&inst1,sizeof(RegionInstance));
-  Event t2 = p2.spawn(REDUC,&inst2,sizeof(RegionInstance));
-  Event c1 = inst1.copy_to(inst3,t1);
-  Event c2 = inst2.copy_to(inst3,t2);
-  Event em = Event::merge_events(c1,c2);
-  Event t3 = p3.spawn(USE,&inst3,sizeof(RegionInstance),em);
-}
 \end{lstlisting}
@@
 \begin{figure}
-\centering
+\begin{lrbox}{\mylistingbox}
+\begin{lstlisting}
+void reduction_ex(const void *args,size_t arglen,Processor p) {
+  Memory m1,m2,m3; Processor p1,p2,p3;
+  ... // Choose memory and processors, get redopid from args
+  ReductionOpID redopid = ...;
+  RegionMetaData meta = RegionMetaData::
+                        create_region(num_elmts, elmt_size);
+  Instance inst1 = meta.create_instance(m1, redop);
+  Instance inst2 = meta.create_instance(m2, redop);
+  Instance inst3 = meta.create_instance(m3);
+  Event t1 = p1.spawn(REDUC, &inst1, sizeof(RegionInstance));
+  Event t2 = p2.spawn(REDUC, &inst2, sizeof(RegionInstance));
+  Event c1 = inst1.reduce_to(inst3, redopid, t1);
+  Event c2 = inst2.reduce_to(inst3, redopid, t2);
+  Event em = Event::merge_events(c1, c2);
+  Event t3 = p3.spawn(USE, &inst3, sizeof(RegionInstance), em);
+}
+\end{lstlisting}
+\end{lrbox}
+\subfigure{\usebox{\mylistingbox}} \\
+
+\subfigure{
 \scalebox{0.8}{
@@
 
-  \node (t3) at (3.2,1) [shape=rectangle,draw] {\begin{tabular}{c}{\bf copy}\\inst1 $\rightarrow$ inst3\end{tabular}};
+  \node (t3) at (3.2,1) [shape=rectangle,draw] {\begin{tabular}{c}{\bf copy-reduce}\\inst1 $\rightarrow$ inst3\end{tabular}};
   \draw (t3) to node[auto] {c1} (4.9,1);
@@
 
-  \node (t4) at (3.2,-1) [shape=rectangle,draw] {\begin{tabular}{c}{\bf copy}\\inst2 $\rightarrow$ inst3\end{tabular}};
+  \node (t4) at (3.2,-1) [shape=rectangle,draw] {\begin{tabular}{c}{\bf copy-reduce}\\inst2 $\rightarrow$ inst3\end{tabular}};
   \draw (t4) to node[auto] {c2} (4.9,-1);
@@
 }
-\vspace{-6mm}
-\caption{Reduction Example Event Graph\label{fig:reducevents}}
+}
+  %\vspace{-6mm}
+  \caption{Reduction Example and Event Graph.\label{fig:reducevents}}
   \vspace{-4mm}
@@
 
+If the only operations a task will perform on a region are reductions, a special reduction-only
+instance can be created.  Rather than applying each reduction separately to the target instance,
+the reduction operations can be accumulated in the reduction instance, and applied in bulk to 
+the target using the {\em reduce\_to} operation.  This has several benefits.  First, it results
+in much better performance, as we will show in section~\ref{subsec:reducmicro}), by reducing both
+the total inter-node traffic and the overhead of that traffic while still allowing parallelism.
+Second, a reduction-only instance is often smaller that the target instance, potentially improving
+cache performance and helping address capacity issues for smaller memories (e.g. a GPU's device
+memory).  For example, if reductions are being made to a single field in a larger structure, the 
+reduction-only instance only needs to store that field and not the other fields that won't be 
+modified by the reduction operation.
+
+\texcomment{
 If reductions are the only operations that will be performed on locations in a region instance,
@@
 the necessary information to understand the structure of the data.
+}
 
@@
 Two reduction-only regions, {\tt inst1} and {\tt inst2}, are created in different memories to
-be used with a reduction operation (lines 46-47).  A physical region of the same class {\tt inst3}
+be used with a reduction operation (lines 46-47).  (Reduction operations are registered at startup with integer IDs
+as function pointers may not be portable across nodes.)  A physical region of the same class {\tt inst3}
 is also created in a different memory (line 48).  Two tasks are launched that will apply
 reductions locally into their respective reduction buffers (lines 49-50).  The resulting
-events are then used to chain copies back to {\tt inst3} after the tasks complete (lines 51-52).  As
-part of these copies the reductions stored in {\tt inst1} and {\tt inst2} will be applied
-to {\tt inst3}.  Since reductions are done atomically, there is no race between the reduction
-applications.  Finally, a merge on the events from the movement and reduction applications is 
-performed and a final task using {\tt inst3} is launched (lines 53-54).  The resulting event
+events are then used to chain reduction copies back to {\tt inst3} after the tasks complete (lines 51-52).
+The reduction operations accumulated in each of {\tt inst1} and {\tt inst2} are applied atomically
+to {\tt inst3}.  The order in which the reduction operations are performed will be different from
+the original order, but the requirement that reduction operations be commutative ensures the final
+result is correct.  Finally, a merge on the events from the two reduction copies is 
+performed and a final task using the results in {\tt inst3} is launched (lines 53-54).  This event
 graph is shown in Figure~\ref{fig:reducevents}.
@@
 class Machine {
-public:
   Machine(int *argc, char ***argv,
@@
         RunStyle = ONE_TASK_ONLY, const void *args, size_t arglen);
-public:
+
   static Machine* get_machine(void);

### 2012-11-08 doc/llr-pldi2013/micro.tex
fixed up some of the negative vspaces that are overkill in the PLDI layout
simplified reduction benchmark graph - text to come next
redid reduction stuff, monkeyed with figures to put examples with event graphs

--- 2012-07-23 doc/asplos2013/micro.tex
@@
 { \small
-\begin{tabular}{m{2cm}|c|c|c|c|c}
+\begin{tabular}{l|c|c|c|c|c}
 Nodes & 1 & 2 & 4 & 8 & 16 \\ \hline
@@
 \end{center}
-\vspace{-6mm}
+\vspace{-2mm}
 \caption{Event Latency Results.\label{tab:eventlat}}
@@
 \end{center}
-\vspace{-6mm}
+\vspace{-2mm}
 \caption{Event Trigger Rate Micro-Benchmark.\label{fig:eventthroo}}
@@
 \end{center}
-\vspace{-6mm}
+\vspace{-2mm}
 \caption{Lock Benchmark for Fixed Total Chains.\label{fig:fixedlock}}
@@
 \end{center}
-\vspace{-6mm}
+\vspace{-2mm}
 \caption{Lock Benchmark for Fixed Node Count.\label{fig:fixednode}}
@@
 \label{subsec:reducmicro}
-To show the benefits of reduction instances we designed
-a histogram micro-benchmark that requires all nodes to perform an addition reduction to a
-physical region located in the globally visible GASNet memory (described in 
+To demonstrate the benefits of reduction-only instances we designed
+a histogram micro-benchmark that requires all nodes to perform an addition reduction to various histogram
+buckets in a physical region located in the globally visible GASNet memory (described in 
 Section~\ref{sec:impl}).  Using the reduction interface described in Section~\ref{subsec:phyreg}
-reductions are performed in five ways:
+reductions are performed in three ways:
 
 \begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
-\item Single Reads/Writes - Each node takes a lock on the global physical region and then for each reduction
-reads a value from the physical region, performs the addition, and then writes the result back.  
-The lock is released after all reductions are performed.
 \item Single Reductions - Each reduction is individually sent as an active message to the node whose
@@
 region in its own system memory, performs all reductions to the localized region, copies the region back, and releases the lock.
-\item Fold Instance - Each node creates a local fold reduction instance.
-All reductions are folded into the instance, which is then copied and applied to the global region.
-\item List Instance - Each node creates a local list reduction instance.
-All reductions are appended to the list reduction instance, which is then copied and applied to the global region.
+\item Reduction Instance - Each node creates a local reduction-only instance.
+All reductions are accumulated into the instance, which is then reduced into the global region.
 \end{itemize}
 
-We ran two experiments, one corresponding with dense reductions and the other corresponding with sparse
-reductions.  In each case a large random source of data is divided into chunks, which are given to
-separate reduction tasks.  There are 8 reduction tasks created for each node, one per processor.  The dense
-case uses a histogram with 256K buckets and each reduction task performs 4M reductions.  The results for the
-dense experiment are shown in Figure~\ref{fig:reducdense}.  The sparse case uses a histogram with 4M buckets,
-but only 64K reductions are performed by each task.  Results for the sparse case
-be seen in Figure~\ref{fig:reducsparse}.
+For our experiment, there are 8 reduction tasks created for each node, one per processor.  Each task performs 4M reductions
+into a histogram with 256K buckets.  The results for a variety of node counts are shown in Figure~\ref{fig:reducbench}.
+They show that the version using reduction-only instance performs 
+best and scales nearly linearly with the number of nodes, achieving over a billion reductions
+per second in the 16 node case.  The localized instance approach works well for a single node, 
+but its inherently serial nature doesn't benefit from increasing node counts.  Finally, the version using individual
+reduction messages is scalable, but the per-message overhead makes it nearly 1000x slower than the versions that are
+able to move data in bulk.
 
-The dense experiment demonstrates that reduction fold regions perform 
-best and scale well with the number of nodes, achieving over a billion reductions
-per second in the 16 node case.  List regions also scale well, but perform about an
-order of magnitude worse than folding regions in the dense case.  The use of a separate active message for
-each reduction operation is another two orders of magnitude worse, demonstrating that data must be
-transferred in larger blocks to be efficient on modern interconnect hardware.  The localized instance approach
-works well for a single node, but its inherently serial nature doesn't benefit from increasing node counts.
-Finally, the only time the latency of performing individual RMA reads and writes isn't disastrous is
-on a single node (where the reads and writes are all local).
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.33]{figs/reduce_simple.pdf}
+\end{center}
+\vspace{-2mm}
+\caption{Reduction Benchmark.\label{fig:reducbench}}
+\vspace{-4mm}
+\end{figure}
 
-The sparse case is very similar, except the scalability of the reduction fold case suffers from the
-overhead of transferring a value for every bucket when most buckets are unmodified by a given task.
-The reduction list case continues to scale well, surpassing the reduction fold performance at large
-node counts.  This shows that
-list regions are better suited for scaling sparse reduction operations.
-
+\texcomment{
 \begin{figure}
@@
 \end{center}
-\vspace{-6mm}
+\vspace{-2mm}
 \caption{Dense Reduction Benchmark.\label{fig:reducdense}}
@@
 \end{center}
-\vspace{-6mm}
+\vspace{-2mm}
 \caption{Sparse Reduction Benchmark.\label{fig:reducsparse}}
@@
 \end{figure}
+}

### 2012-11-08 doc/types-pldi2013/intro.tex
intro

@@
 a single tree-shaped partitioning of data that must be checked statically 
-(see Section~\ref{sec:related}).  The second approach permits potential data
+(see Section~\ref{sec:related}).  The second approach permits data
 aliasing, allowing greater expressivity, but uses fully dynamic analyses of
-program execution to detect data aliasing and guarantee correct results.
+program execution to detect aliasing and guarantee correct results.
 This approach incurs significant runtime overhead and requires 
@@
 best way to partition data is often a function of the data
-itself---i.e., the partitions must be dynamically computed.
+itself---i.e., the partitions must be dynamically computed and cannot be statically described.
 Furthermore, applications often need multiple, simultaneous partitions of
-the same data, each providing different views on to that data.
+the same data, which introduces aliasing.
 
@@
 in cases where tasks access the same region.  The programmer may
-specify that tasks conflicting on a region run: in
+specify that tasks conflicting on a region run in
 program order (the default), {\em atomically} with respect to each
@@
 
-\item We present a parallel operational semantics for Core Legion that
-  may be of independent interest.  This semantics is compositional,
+\item We present a novel parallel operational semantics for Core Legion.
+ This semantics is compositional,
   hierarchical, and asynchronous, reflecting the way such programs
@@
   checks is low, showing that an expressive and dynamic
-  language with aliasing is not incompatible with both high
+  language with aliasing is compatible with both high
   performance and safety (Section~\ref{sec:evaluation}).

### 2012-11-08 doc/types-pldi2013/alt_example.tex
bunch o' changes to section 2

@@
 a wire.  Pointers are denoted by the @ symbol
-followed by the set of region(s) into which a pointer can point.
+followed by the region(s) into which the pointer may point.
 The {\tt CircuitWire} type is parametrized by the local region
@@
 The most important decision in any Legion program is the choice of
-how to partition program data.  The choice of a partitioning algorithm
-can depend on a multitude of factors including the type of data structure
-and input data.  Therefore
-Legion requires applications to select their own partitioning algorithms.
+how to partition program data.  The ideal partitioning algorithm
+depends on a multitude of factors including the type of data structure,
+the input data, and the desired number of partitions (which often depends on the
+size of the machine running the application).  Therefore
+Legion permits applications to select their own partitioning algorithms.
 In this example we invoke a third-party dynamic graph partitioning library
@@
 {\tt CircuitNode} with a piece ID (the last field in the tuple).
-The function {\tt extern\_metis} is parametrized on the regions it will
-access; in this case {\tt rn} the region containing nodes, {\tt rw} the region
-contining wires, and {\tt rl} the region containing the lists of pointers.  The function
-must also specify the {\em privileges} (e.g read) with which it will access
-these regions.  {\tt extern\_metis} will read all three regions and write
-the {\tt rn} region since it will be writing to the last field of each
-{\tt CircuitNode}.  Our type system enforces the Legion 
-requirement that each function can only be invoked with a subset of the 
-privileges contained in its calling context.
+The function {\tt extern\_metis} is parameterized on the regions it will
+access; in this case {\tt rn} is the region containing nodes, {\tt rw} is the region
+contining wires, and {\tt rl} is the region containing the lists of pointers.  The function
+must also specify the access {\em privileges} (e.g read, write) it requires for each of
+these regions.  All three regions will be read by {\tt extern\_metis}, but only the {\tt rn}
+region will be written to (since it will be modifying the last field of each {\tt CircuitNode}).
+A function may only be invoked if the caller possesses all the privileges needed by the function
+being called.  This is verified statically by the Legion type system.
 
@@
 
+To efficiently support the access patterns of the circuit simulation,
+multiple partitionings of the {\tt all\_nodes} region are performed.
+The desired {\em region tree} is shown in Figure~\ref{sfig:part_fig:tree}.
+First, we need subregions that describe the set of nodes ``owned'' by
+each piece.  These will be called {\tt rn0}, {\tt rn1}, ... and since each
+node has a single owner, they form the {\em disjoint} partition (indicated with a *)
+on the left side of the tree.
 Figure~\ref{sfig:part_fig:pieces} shows one possible output from
-METIS with different colors identifing different piece IDs.  To capture
+METIS with different colors identifing different piece IDs.
+Second, each piece of the circuit will need to 
+access the nodes that border its own.  These are commonly referred as {\em ghost nodes},
+and the set of ghost nodes for the first two circuit pieces (named {\tt rg0} and {\tt rg1})
+are shown in Figures~\ref{sfig:part_fig:ghost_zero}
+and~\ref{sfig:part_fig:ghost_one}.  Because a node may neighbor more than one other circuit
+piece, this second partition is an {\em aliased} partition.
+Without the ability to have multiple partitionings of the same data, one would have two
+possible alternatives:
+\begin{enumerate}
+\item Make a single partition with $2^N$ subregions, one for each possible case of sharing, or
+\item Have all pieces use the {\tt all\_nodes} for their ghost cell region, greatly increasing the 
+amount of data movement required
+\end{enumerate}
+
+To describe
 the result of a partitioning algorithm, Legion uses {\em colorings}.
-Colorings are a mapping from pointers to {\em colors} (e.g. integers).  In
+A coloring is a mapping from pointers to {\em colors} (represented in Core Legion as integers).  In
 the circuit example, the piece IDs are the designated colors. 
@@
 that constructs a coloring from a list of nodes which have been 
-annotated with piece IDs.
+annotated with piece IDs.  The coloring of wires is similar, with each wire
+using the color of its first endpoint.  For the ``ghost nodes'', a {\em multicoloring}
+is used, which allows each pointer to be assigned multiple colorings.  (Multicolorings are
+implemented in Core Legion with separate colorings for each color.)
 
@@
             reads(rl,rn) : coloring(rn) = 
-    if isnull(node_list) then newcolor(rn) else
+    if isnull(node_list) then
+        newcolor(rn) 
+    else
         let list_elem : NodeList<rl,rn> = read(node_list) in
@@
 Colorings are the input for partition operations in Legion.  The partition
-statement shown in Listing~\ref{lst:partition} partitions the region
-{\tt all\_nodes} into logical subregions with a separate output region
-for each color in the coloring.  Due to space constraints we, only show
-code for {\tt rn0} and {\tt rn1} pieces for the remainder of the example.  We also
-do not show the coloring or partitioning of the {\tt all\_wires} region as
-it can be computed in a straightforward manner from the partitioning of
-nodes into pieces.
+statement shown in Listing~\ref{lst:partition} performs the ``owned nodes''
+partitioning.  The partitioning of wires is identical, while the partitioning
+of the ghost nodes is performed in Core Legion as a separate partitioning operation
+for each color.
 
@@
 (dynamically checked), the partition operation results in 
-{\em disjoint} subregions.  Furthermore, the partition operation
-introduces constraints into the static checking enviornment describing
-both the disjointness of the subregions (i.e. {\tt rn0 $*$ rn1})
+{\em disjoint} subregions, each of which contains a subset of the original region.
+The partition operation introduces constraints into the type system describing
+both the disjointness of the subregions (e.g. {\tt rn0 $*$ rn1})
 and the inclusion of each subregion in the original region
-(i.e. {\tt rn0 $\leq$ all\_nodes}).
+(e.g. {\tt rn0 $\leq$ all\_nodes}).
 
-In addition, to partitioning the graph to describe {\em owned} nodes
-for each piece, the circuit example also
-needs to be able to name the {\em ghost nodes} for each piece.  Ghost node
-regions are used for accessing the nodes of wires that span
-two different pieces. (This is also the reason {\tt CircuitWire} is parametrized on
-two regions).  Figure~\ref{sfig:part_fig:ghost_zero} and 
-Figure~\ref{sfig:part_fig:ghost_one} show colorings for the ghost regions
-corresponding to {\tt rn0} and {\tt rn1} respectively.  Note that two 
-nodes are colored in both colorings.  Since nodes can appear in multiple
-subregions, we use a {\em multicoloring} (implemented in core Legion
-by multiple single colorings) to describe {\em aliased} partitions.  Aliased
-partitions still introduce subregion constraints into the static checking
-environment, but not disjointness constraints.  Figure~\ref{sfig:part_fig:tree}
-shows the resulting logical region tree for node regions.
+%% In addition, to partitioning the graph to describe {\em owned} nodes
+%% for each piece, the circuit example also
+%% needs to be able to name the {\em ghost nodes} for each piece.  Ghost node
+%% regions are used for accessing the nodes of wires that span
+%% two different pieces. (This is also the reason {\tt CircuitWire} is parametrized on
+%% two regions).  Figure~\ref{sfig:part_fig:ghost_zero} and 
+%% Figure~\ref{sfig:part_fig:ghost_one} show colorings for the ghost regions
+%% corresponding to {\tt rn0} and {\tt rn1} respectively.  Note that two 
+%% nodes are colored in both colorings.  Since nodes can appear in multiple
+%% subregions, we use a {\em multicoloring} (implemented in core Legion
+%% by multiple single colorings) to describe {\em aliased} partitions.  Aliased
+%% partitions still introduce subregion constraints into the static checking
+%% environment, but not disjointness constraints.  Figure~\ref{sfig:part_fig:tree}
+%% shows the resulting logical region tree for node regions.
 
 The first-class nature of logical regions in Legion allows them to
-be written into the heap.  To avoid losing important static properties of
-logical regions in this scenario, we use {\em region
+be written into the heap.  To avoid losing information about these regions and pointers
+into those regions, Legion provides {\em region
 relationships}.  Region relationships are a bounded existential type
-that allows programmers to {\em pack} a group of regions together and remember
+that allows programmers to {\em pack} a group of regions and pointers together and capture
 properties about them such as disjointness and subregion relationships.
@@
 describing its region of wires {\tt rpw}, region of 
-nodes {\tt rpn}, region of ghost nodes {\tt rg}, and important constraints.
+nodes {\tt rpn}, region of ghost nodes {\tt rg}, and important constraints.  Although the
+``names'' of {\tt rpw}, {\tt rpn}, and {\tt rg} are lost when packed into the region relationship,
+the knowledge that they are subregions of the {\tt rw} and {\tt rn} regions is not.
 
@@
 
-The Legion type system verifies properties of region relationships when packing;
-when region relationships are unpacked the same properties are
-reintroduced into the static checking environment.
-A key feature underlying the soundness of the Legion
-type system is that privileges cannot be packed in a region
+The Legion type system verifies properties of region relationships when packing.  When
+a region relationship is unpacked later, fresh names are given to the regions that were
+captured by the region relationship, and the properties that are known to hold for these
+regions are re-introduced to the type environment.
+In constrast, region access privileges cannot be packed in a region
 relationship---privileges belong to functions. When a function unpacks
-a region relationship it must already hold privileges for the unpacked regions.
+a region relationship it must already hold privileges for the unpacked regions it wishes to access.
+
 Listing~\ref{lst:loop} shows the main time step loop run by the 
@@
 because it has read/write privileges for {\tt rn} and the {\tt CircuitPiece} 
-region relationship declares {\tt rn0 $\leq$ rn}.  Note this does require
-a dynamic check that {\tt rn0} is a subregion of {\tt rn}, but the cost of the 
-check will be ammortized over all the uses of {\tt rn0}.
+region relationship ensures that {\tt rn0 $\leq$ rn}.
 

### 2012-11-09 doc/types-pldi2013/langdef.tex
trying to be consistent with math/text font choices

@@
 \typeenvx e_1 : T@(\oton{r}) \\
-\forall i, reads(r_i) \in \Phi^*\end{array}}
+\forall i.\ \text{reads}(r_i) \in \Phi^*\end{array}}
 {\typeenvx \text{read}(e_1) : T}
@@
 \typeenvx e_2 : T \\
-\forall i, writes(r_i) \in \Phi^*
+\forall i.\ \text{writes}(r_i) \in \Phi^*
 \end{array}}
@@
 \typeenvx e_2 : T_2 \\
-\forall i, reduces_{id}(r_i) \in \Phi^*
+\forall i.\ \text{reduces}_{id}(r_i) \in \Phi^*
 \end{array}}
@@
 \end{array}}
-{\typeenvx upregion(e,\oton{r}) : T@(\oton{r})}
+{\typeenvx \text{upregion}(e,\oton{r}) : T@(\oton{r})}
 
@@
 \end{array}}
-{\typeenvx downregion(e,\oton{r}) : T@(\oton{r})}
+{\typeenvx \text{downregion}(e,\oton{r}) : T@(\oton{r})}
 
@@
 \typeenvx e_2 : T@r \\
-\typeenvx e_3 : int
+\typeenvx e_3 : \text{int}
 \end{array}}
@@
 \ruleatx{\XD,0.0}{\XE,0.2}{E-Read}{
-\opsenvx e \mapsto l, E \andalso S' = \text{apply}(S, E) \\
+\opsenvx e \mapsto l, E \andalso S' = apply(S, E) \\
 v = \begin{cases} S'(l), & \text{if } l \not\in C \\
@@
 \ruleatx{\XD,0.9}{\XE,0.85}{E-Write}{
-\opsenvx e_1 \mapsto l, E_1 \andalso S' = \text{apply}(S, E_1) \\
+\opsenvx e_1 \mapsto l, E_1 \andalso S' = apply(S, E_1) \\
 \opsenvx[S=S'] e_2 \mapsto v, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)
@@
 \ruleatx{\XD,1.5}{\XE,1.45}{E-Reduce}{
-\opsenvx e_1 \mapsto l, E_1 \andalso S' = \text{apply}(S, E_1) \\
+\opsenvx e_1 \mapsto l, E_1 \andalso S' = apply(S, E_1) \\
 \opsenvx[S=S'] e_2 \mapsto v, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)
@@
 \ruleatx{\XD,4.7}{\XE,4.9}{E-Color}{
-\opsenvx e_1 \mapsto K, E_1 \andalso S' = \text{apply}(S, E_1) \\
-\opsenvx[S=S'] e_2 \mapsto l, E_2 \andalso S'' = \text{apply}(S', E_2) \\
+\opsenvx e_1 \mapsto K, E_1 \andalso S' = apply(S, E_1) \\
+\opsenvx[S=S'] e_2 \mapsto l, E_2 \andalso S'' = apply(S', E_2) \\
 \opsenvx[S=S''] e_3 \mapsto v, E_3 \\
@@
 \opsenv{e_1}{K}{E_1} \andalso \rho_i = \{ l : (l, i) \in K \}, \text{ for } 1 \leq i \leq k \\
-M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \andalso S' = \text{apply}(S, E_1) \\
+M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \andalso S' = apply(S, E_1) \\
 \opsenvx[M=M',S=S'] e_2 \mapsto v, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)
@@
 \opsenvx e_1 \mapsto \langle \langle \otok{k}{\rho} , v_1 \rangle \rangle, E_1 \andalso M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \\
-L' = L[v_1/id] \andalso S' = \text{apply}(S, E_1) \\
+L' = L[v_1/id] \andalso S' = apply(S, E_1) \\
 \opsenvx[M=M',L=L',S=S'] e_2 \mapsto v_2, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)
@@
 \ruleatx{\XA,8.1}{\XE,8.4}{E-Call}{
-\opsenvx e_1 \mapsto v_1, E_1 \andalso S_1 = \text{apply}(S, E_1) \andalso \opsenvx e_2 \mapsto v_2, E_2 \andalso \ldots \andalso S_n = \text{apply}(S_{n-1}, E_n) \\
+\opsenvx e_1 \mapsto v_1, E_1 \andalso S_1 = apply(S, E_1) \andalso \opsenvx e_2 \mapsto v_2, E_2 \andalso \ldots \andalso S_n = apply(S_{n-1}, E_n) \\
 valid\_interleave(S, C, E', \oton{E}) \andalso \text{function }id[\otok{k}{r'}](a_1 : T_1, \ldots, a_n : T_n), \Phi', Q' : T_r = e_{n+1} \\
-M' = \{ (r'_1, M(r_1)), \ldots (r'_k, M(r_k)) \} \andalso L' = \{ (a_1, v_1), \ldots, (a_n, v_n) \} \andalso S' = \text{apply}(S, E') \\
+M' = \{ (r'_1, M(r_1)), \ldots (r'_k, M(r_k)) \} \andalso L' = \{ (a_1, v_1), \ldots, (a_n, v_n) \} \andalso S' = apply(S, E') \\
 C' = C \cup \{ l : \exists \rho. atomic(\rho) \in M' \llbracket Q' \rrbracket \vee simult(\rho) \in M' \llbracket Q' \rrbracket \} \\
@@
 \begin{cases}
-simult, & \text{if }\exists \rho. l \in \rho \wedge simult(\rho) \in \hat Q \\
-atomic, & \text{if }\exists \rho. l \in \rho \wedge atomic(\rho) \in \hat Q \\
+simult, & \text{if }\exists \rho. l \in \rho \wedge \text{simult}(\rho) \in \hat Q \\
+atomic, & \text{if }\exists \rho. l \in \rho \wedge \text{atomic}(\rho) \in \hat Q \\
 excl, & \text{otherwise}
@@
 \Phi \subseteq \Phi^* \\
-r_i \leq r_j \in \Omega^* \wedge reads(r_j) \in \Phi^* \Rightarrow reads(r_i) \in \Phi^* \\
-r_i \leq r_j \in \Omega^* \wedge writes(r_j) \in \Phi^* \Rightarrow writes(r_i) \in \Phi^* \\
-r_i \leq r_j \in \Omega^* \wedge reduces_{id}(r_j) \in \Phi^* \Rightarrow reduces_{id}(r_i) \in \Phi^* \\
-reads(r) \in \Phi^* \wedge writes(r) \in \Phi^* \Rightarrow reduces_{id}(r) \in \Phi^* \\
+r_i \leq r_j \in \Omega^* \wedge \text{reads}(r_j) \in \Phi^* \Rightarrow \text{reads}(r_i) \in \Phi^* \\
+r_i \leq r_j \in \Omega^* \wedge \text{writes}(r_j) \in \Phi^* \Rightarrow \text{writes}(r_i) \in \Phi^* \\
+r_i \leq r_j \in \Omega^* \wedge \text{reduces}_{id}(r_j) \in \Phi^* \Rightarrow \text{reduces}_{id}(r_i) \in \Phi^* \\
+\text{reads}(r) \in \Phi^* \wedge \text{writes}(r) \in \Phi^* \Rightarrow \text{reduces}_{id}(r) \in \Phi^* \\
 \ \ \ \mbox{for every function identifier $id$}
@@
 \privconsist{E}{\Phi} \Leftrightarrow \forall \epsilon \in E. \\
-\quad\quad (\epsilon = read(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reads(r) \in \Phi)\ \wedge \\
-\quad\quad (\epsilon = write(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge writes(r) \in \Phi)\ \wedge \\
-\quad\quad (\epsilon = reduce_{id}(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reduces_{id}(r) \in \Phi)
+\quad\quad (\epsilon = read(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge \text{reads}(r) \in \Phi)\ \wedge \\
+\quad\quad (\epsilon = write(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge \text{writes}(r) \in \Phi)\ \wedge \\
+\quad\quad (\epsilon = reduce_{id}(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge \text{reduces}_{id}(r) \in \Phi)
 \end{array}
@@
 \multicolumn{2}{l}{priv_1(r_1) \nonint[M] priv_2(r_2) \Leftrightarrow} \\
-& (priv_1 = reads \wedge priv_2 = reads)\ \vee \\
-& (priv_1 = reduces_{id_1} \wedge priv_2 = reduces_{id_2} \wedge id_1 = id_2)\ \vee \\
+& (priv_1 = \text{reads} \wedge priv_2 = \text{reads})\ \vee \\
+& (priv_1 = \text{reduces}_{id_1} \wedge priv_2 = \text{reduces}_{id_2} \wedge id_1 = id_2)\ \vee \\
 & (r_1 * r_2)\ \vee \\

### 2012-11-09 doc/types-pldi2013/alt_example.tex
updates

@@
 We begin by introducing an example program 
-that motivates many of the features in Legion programming model
-and provides an intuition for many of our design decisions.  The example program
-executes a simulation of an arbitrary graph of electrical circuit elements.  
-The simulation runs for many time steps and performs three computations 
-for each time step: calculate new currents, distribute charge, and update voltages.
-Due to space constraints we only introduce small code examples. 
-The source code for the full circuit example written in the core Legion
-language (see Section~\ref{sec:legioncore}) can be found in the
-submitted appendix.
+motivating the novel features of Legion.
+Due to space constraints we present only code fragments;
+the full example is in the submitted appendix.
 
+Consider a simulation of an electrical circuit.
+The simulation runs for many time steps, each of which performs three computations:
+ calculate new currents, distribute charges, and update voltages.
 The simulation takes as input an arbitrary graph of circuit elements
-that have been dynamically allocated into two logical regions: {\tt all\_nodes} 
-and {\tt all\_wires}.  The types of nodes and wires are the tuples
+that have been dynamically allocated in two logical regions: {\tt all\_nodes} 
+and {\tt all\_wires}.  The types of nodes and wires are 
 shown in Listing~\ref{lst:tuples}.  The first two fields in the
@@
 
-The most important decision in any Legion program is the choice of
-how to partition program data.  The ideal partitioning algorithm
-depends on a multitude of factors including the type of data structure,
-the input data, and the desired number of partitions (which often depends on the
-size of the machine running the application).  Therefore
-Legion permits applications to select their own partitioning algorithms.
-In this example we invoke a third-party dynamic graph partitioning library
-called METIS\cite{Metis98} as shown in Listing~\ref{lst:metis}.
+The most important decision in any Legion program is how to partition
+program data.  An ideal partitioning depends on many factors,
+including the shape of data structures, the input, and the number of
+partitions, which usually varies with the target machine.  To allow
+the necessary flexibility, in Legion partitioning is simply programmed,
+rather than expressed in a more structured and restricted
+sublanguage.  In the example we invoke METIS\cite{Metis98}, a dynamic graph
+partitioning library called (see Listing~\ref{lst:metis}).
 
@@
 
-METIS takes as arguments lists of all the nodes and all the wires in
-the graph.  It then breaks the graph into pieces by annotating each 
-{\tt CircuitNode} with a piece ID (the last field in the tuple).
-The function {\tt extern\_metis} is parameterized on the regions it will
-access; in this case {\tt rn} is the region containing nodes, {\tt rw} is the region
-contining wires, and {\tt rl} is the region containing the lists of pointers.  The function
-must also specify the access {\em privileges} (e.g read, write) it requires for each of
-these regions.  All three regions will be read by {\tt extern\_metis}, but only the {\tt rn}
-region will be written to (since it will be modifying the last field of each {\tt CircuitNode}).
-A function may only be invoked if the caller possesses all the privileges needed by the function
-being called.  This is verified statically by the Legion type system.
+METIS takes as arguments lists of the nodes and wires in
+the graph.  It breaks the graph into pieces by annotating each 
+{\tt CircuitNode} with a piece ID (the last field of the tuple).
+The function {\tt extern\_metis} is parameterized on the regions it 
+accesses: {\tt rn} is a region of nodes, {\tt rw} is a region of wires, 
+and {\tt rl} is a region with the lists of pointers.  The function
+also specifies the access {\em privileges} (e.g read, write) it requires for each 
+region.  All three regions are read by {\tt extern\_metis}, but only the {\tt rn}
+region is written (since it writes the piece ID).  A function can only 
+have privileges that its caller also possesses; this property is verified statically. 
+% by the Legion type system.
 
@@
 To efficiently support the access patterns of the circuit simulation,
-multiple partitionings of the {\tt all\_nodes} region are performed.
-The desired {\em region tree} is shown in Figure~\ref{sfig:part_fig:tree}.
-First, we need subregions that describe the set of nodes ``owned'' by
-each piece.  These will be called {\tt rn0}, {\tt rn1}, ... and since each
-node has a single owner, they form the {\em disjoint} partition (indicated with a *)
-on the left side of the tree.
-Figure~\ref{sfig:part_fig:pieces} shows one possible output from
-METIS with different colors identifing different piece IDs.
-Second, each piece of the circuit will need to 
-access the nodes that border its own.  These are commonly referred as {\em ghost nodes},
-and the set of ghost nodes for the first two circuit pieces (named {\tt rg0} and {\tt rg1})
-are shown in Figures~\ref{sfig:part_fig:ghost_zero}
-and~\ref{sfig:part_fig:ghost_one}.  Because a node may neighbor more than one other circuit
-piece, this second partition is an {\em aliased} partition.
-Without the ability to have multiple partitionings of the same data, one would have two
-possible alternatives:
-\begin{enumerate}
-\item Make a single partition with $2^N$ subregions, one for each possible case of sharing, or
-\item Have all pieces use the {\tt all\_nodes} for their ghost cell region, greatly increasing the 
-amount of data movement required
-\end{enumerate}
+the {\tt all\_nodes} region is partitioned in two different ways.  The
+desired {\em region tree} is shown in Figure~\ref{sfig:part_fig:tree}.
+First, there are {\em subregions} of {\tt all\_nodes} that describe
+the set of nodes ``owned'' by each piece, called {\tt rn0}, {\tt rn1},
+$\ldots$. Since each node is in one piece, this partition is {\em
+disjoint} (indicated by a $*$ on the left subtree).
+Figure~\ref{sfig:part_fig:pieces} shows one possible node partition;
+different colors identify different piece IDs.  Second, each piece of
+the circuit needs access to the nodes on its border, commonly referred
+to as {\em ghost nodes}.  The ghost nodes for two circuit pieces are
+shown in Figures~\ref{sfig:part_fig:ghost_zero}
+and~\ref{sfig:part_fig:ghost_one}; note two nodes are in both sets.
+Because a node may neighbor more than one other circuit piece, this
+second partition of {\tt all\_nodes} is {\em aliased}.  Thus, there
+are two sources of aliasing in the region tree: the two distinct partitions
+divide the {\tt all\_nodes} region in different ways, and the ghost
+node subregions are not disjoint.
 
-To describe
-the result of a partitioning algorithm, Legion uses {\em colorings}.
-A coloring is a mapping from pointers to {\em colors} (represented in Core Legion as integers).  In
-the circuit example, the piece IDs are the designated colors. 
+Note that there are two alternatives to using multiple partitions of the same data, both of which avoid introducing aliasing.
+We could use a single partition with $2^n$ subregions, one for each
+possible case of sharing; alternatively, all pieces could just use the
+{\tt all\_nodes} region to access their ghost nodes.  Neither option
+is attractive: the former significantly complicates programming, the
+latter greatly increases data movement required at
+runtime.
+
+Programmers exoress partitions using {\em colorings},
+maps from region elements to integers (``colors'').
 Listing~\ref{lst:coloring} shows the function {\tt create\_piece\_coloring}
-that constructs a coloring from a list of nodes which have been 
-annotated with piece IDs.  The coloring of wires is similar, with each wire
-using the color of its first endpoint.  For the ``ghost nodes'', a {\em multicoloring}
-is used, which allows each pointer to be assigned multiple colorings.  (Multicolorings are
-implemented in Core Legion with separate colorings for each color.)
-
+that constructs a coloring from the piece IDs of a list of nodes.
+%The coloring of wires is similar, with each wire
+%using the color of its first endpoint.  
+For aliased partitions such as the ghost nodes, a {\em multicoloring}
+is used, which assigns each region element (possibly) multiple colors.  
+%(Multicolorings are
+%implemented in Core Legion with separate colorings for each color.)
 \begin{lstlisting}[label={lst:coloring},caption={Coloring Construction}]
@@
 
-Colorings are the input for partition operations in Legion.  The partition
+Colorings are the input to partition operations in Legion.  The partition
 statement shown in Listing~\ref{lst:partition} performs the ``owned nodes''
-partitioning.  The partitioning of wires is identical, while the partitioning
-of the ghost nodes is performed in Core Legion as a separate partitioning operation
-for each color.
+partitioning.    Since colorings assign each region element at most one color
+(dynamically checked), the partition operation results in 
+{\em disjoint} subregions, each a subset of the {\em parent} region.
+A partition operation introduces constraints into the static type environment describing
+both the disjointness of subregions (e.g., {\tt rn0 $*$ rn1})
+and the subregion relationships
+(e.g., {\tt rn0 $\leq$ all\_nodes}).  A partition created from a multicoloring
+only introduces the subregion relationships, as the parition may be aliased.
+
+%The partitioning of wires is identical, while the partitioning
+%of the ghost nodes is performed in Core Legion as a separate partitioning operation
+%for each color.
 
@@
 \end{lstlisting}
-
-Since colorings can only color each pointer at most with a single color
-(dynamically checked), the partition operation results in 
-{\em disjoint} subregions, each of which contains a subset of the original region.
-The partition operation introduces constraints into the type system describing
-both the disjointness of the subregions (e.g. {\tt rn0 $*$ rn1})
-and the inclusion of each subregion in the original region
-(e.g. {\tt rn0 $\leq$ all\_nodes}).
 

### 2012-11-09 doc/types-pldi2013/langdef.tex
langdef

--- 2012-11-04 doc/types-pldi2013/langdef.tex
@@
 Figure~\ref{fig:langdef} defines Core Legion, a subset of the full Legion language
-that still illustrates all of the important issues.  
+that illustrates the important issues.  
 %Types include booleans, integers, tuples, and pointers.  
@@
 The following two subsections illustrate these decisions and how they have been
-chosen to preserve a straightforward proof of type soundness.
+chosen to allow a straightforward proof of type soundness.
 
@@
 subexpression trace order.
-\item for locations that have only access in {\em simult} mode or are in the clobber set, no constraints are
+\item for locations with only access in {\em simult} mode or in the clobber set, no constraints are
 enforced.  The validity of the interleaving of these accesses is performed by some enclosing expression.

### 2012-11-09 doc/types-pldi2013/related.tex
Lots of small changes and moving figures around

@@
 \label{sec:related}
-Legion is most directly related to Sequoia \cite{Fatahalian06}.  Sequoia is array-based, with
-a limited repertoire of ways to partition arrays.  Sequoia is a static language with a single unified hierarchy
+Legion is most directly related to Sequoia \cite{Fatahalian06}. 
+Sequoia is a static language with a single unified hierarchy
 of tasks and data; Legion is more dynamic with separate task and region hierarchies. 
+Both Sequoia and Titanium\cite{TIT98} are array-based languages, with
+a limited repertoire of ways to partition arrays.
 
@@
 Chapel \cite{Chamberlain:Chapel} and X10 \cite{X1005} also provide some Legion-like facilities.
-Chapel's domains and X10's places provide the programmer with a 
-mechanism for expressing locality, similar to regions in Legion.  However, domains
+Chapel's locales and X10's places provide the programmer with a 
+mechanism for expressing locality, similar to regions in Legion.  However, locales 
 and places are not used for independence analysis to discover parallelism.

### 2012-11-09 apps/fdtd/fdtd.cc
FDTD: More work on source kernel math.

@@ -72,6 +72,7 @@ block of cubes in +z direction for Ey, and in the +y direction for Ez.
 #include <cassert>
 #include <cmath>
 #include <cstdio>
+#include <complex>
 
 #include "legion.h"
 #include "lowlevel.h"
@@ -81,6 +82,8 @@ using namespace LegionRuntime::HighLevel;
 typedef LegionRuntime::LowLevel::RegionAccessor<LegionRuntime::LowLevel::AccessorGeneric> Accessor;
 
 LegionRuntime::Logger::Category log_app("app");
+
+const double pi = 3.141592653589793238462643383276;
 
 ////////////////////////////////////////////////////////////////////////
 // Global task ID list. Each of these tasks will be registered with
@@ -119,6 +122,13 @@ static inline const char * dim_name(dim_t dim) {
   }
 }
 
+static inline int mod3(int dim) {
+  return dim % NDIMS;
+  if (dim < 0) {
+    dim += NDIMS;
+  }
+}
+
 ////////////////////////////////////////////////////////////////////////
 // During the computation, each block will only require access to
 // ghost cells in one direction. Thus each block will partitioned six
@@ -154,10 +164,17 @@ class vec {
 public:
   vec() { x[0] = x[1] = x[2] = 0; }
   vec(T a, T b, T c) { x[0] = a; x[1] = b; x[2] = c; }
+  vec(const vec &a) { x[0] = a[0]; x[1] = a[1]; x[2] = a[2]; }
+  template<typename U>
+  vec(const vec<U> &a) { x[0] = (T)a[0]; x[1] = (T)a[1]; x[2] = (T)a[2]; }
   T & operator[] (int i) { return x[i]; }
   const T & operator[] (int i) const { return x[i]; }
   vec operator+ (const vec &a) const { return vec(x[0] + a[0], x[1] + a[1], x[2] + a[2]); }
   vec operator- (const vec &a) const { return vec(x[0] - a[0], x[1] - a[1], x[2] - a[2]); }
+  vec operator+ (T a) const { return vec(x[0] + a, x[1] + a, x[2] + a); }
+  vec operator- (T a) const { return vec(x[0] - a, x[1] - a, x[2] - a); }
+  vec operator* (T a) const { return vec(x[0] * a, x[1] * a, x[2] * a); }
+  vec operator/ (T a) const { return vec(x[0] / a, x[1] / a, x[2] / a); }
 private:
   T x[3];
 };
[…]

### 2012-11-09 doc/types-pldi2013/alt_example.tex

@@
 {\tt CircuitNode} with a piece ID (the last field of the tuple).
-The function {\tt extern\_metis} is parameterized on the regions it 
+The function {\tt extern\_metis} is parametrized on the regions it 
 accesses: {\tt rn} is a region of nodes, {\tt rw} is a region of wires, 
@@
 
-Programmers exoress partitions using {\em colorings},
+Programmers express partitions using {\em colorings},
 maps from region elements to integers (``colors'').
@@
 (e.g., {\tt rn0 $\leq$ all\_nodes}).  A partition created from a multicoloring
-only introduces the subregion relationships, as the parition may be aliased.
+only introduces the subregion relationships, as the partition may be aliased.
 
@@
 regions are re-introduced to the type environment.
-In constrast, region access privileges cannot be packed in a region
+In contrast, region access privileges cannot be packed in a region
 relationship---privileges belong to functions. When a function unpacks

### 2012-11-09 doc/llr-pldi2013/interface.tex
monkeyed with figures in section 2 - alot

@@
 
-Our interface organizes functionality into objects.  We describe
-these objects in five parts: {\em events} for composing operations 
+Our interface is shown in Figure~\ref{fig:runtimeapi} and organizes functionality into objects.  We describe
+these objects in six parts: {\em events} for composing operations 
 (Section~\ref{subsec:events}), {\em processors} for parallel computation 
-(\ref{subsec:procs}), {\em deferred locks} for synchronization
+(\ref{subsec:procs}), {\em barriers} for producer/consumer coordination (\ref{subsec:barriers}),
+{\em deferred locks} for more general synchronization
 (\ref{subsec:locks}), {\em physical regions} for data layout and movement
@@
 are light-weight {\em handles} that provide a unique name for an underlying
-implementation.
+implementation object.
 Every handle is valid everywhere in the system, allowing handles to be freely copied,
@@
 Events are the primary mechanism for describing dependences between operations in our system.
-Listing~\ref{lst:eventapi} shows the interface for events.  An instance of the {\tt Event} type 
-names a unique event in the system.  {\tt NO\_EVENT} (line 5) is a special instance
+Lines 1-13 of Figure~\ref{fig:runtimeapi} show the interface for events.  An instance of the {\tt Event} type 
+names a unique event in the system.  {\tt NO\_EVENT} (line 3)
+is a special instance
 of an event that by definition has always triggered.  The event interface
-supports testing whether an event has triggered (line 7) and waiting on
-an event to trigger (line 9).  While these methods can be useful, the encouraged
+supports testing whether an event has triggered (line 5) and waiting on
+an event to trigger (line 6).  While these methods can be useful, the preferred
 use of events is passing them as preconditions for other operations.  The client can use the
-{\tt merge\_events} call (lines 11-12) to create an event that triggers only when every
-event in a given set of events has triggered.  An example using events follows in
+{\tt merge\_events} call (line 7) to create an event that triggers only when every
+event in a given set of events has triggered.  A code example using events follows in
 Section~\ref{subsec:procs}.
@@
 operations and the implementation is responsible for triggering these events.  Clients
-can also create a {\tt UserEvent} (line 15) that is triggered explicitly by the client.
+can also create a {\tt UserEvent} (line 10) that is triggered explicitly by the client,
+and may be used as precondition like any other event.
 
-The common synchronization pattern in which multiple consumers are dependent on
-a set of producers is supported by a {\tt Barrier} object, which is similar
-to a user event, but does not trigger until the expected number of {\tt arrive}
-operations have occurred.  To better support
-composability and asynchronous operation, our interface differs from barriers in other programming
-models\cite{MPI} in three fundamental ways:
-%The common synchronization pattern in which multiple consumers are dependent on a 
-%(potentially different) set of producers is supported by a {\tt Barrier} object,
-%which is similar to a {\tt UserEvent}, but does not trigger until the expected
-%number of {\tt arrive} operations have occurred.  Barrier constructs exist in other
-%programming models\cite{MPI}, but in addition to the decoupling of the waiters from
-%the arrivers, our barrier interface has two further improvements to better support
-%composability and asynchronous operation:
-\begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
-\item The set of producers of arrival calls can be different from the set of consumers waiting on the barrier.
-\item An arrival operation can be made dependent on another event, allowing a client to
-express that an operation should be completed before the barrier can trigger.
-\item The expected arrival count of a barrier can be dynamically modified, allowing
-barrier creation before needing to know the exact number of arrivals.
-\end{itemize}
-
-Both user events and barriers are a sub-type of events and can therefore
-be used transparently to express dependences wherever events are used.
 %Although user events and barriers are created and triggered differently,
@@
 
-\begin{lstlisting}[float={t},label={lst:eventapi},caption={Event Interface.}]
+\begin{figure}
+\begin{lrbox}{\mylistingbox}
+\begin{lstlisting}
 class Event {
-  const unsigned id;
-  const unsigned gen;
+  const unsigned id, gen;
   static const Event NO_EVENT;
-  // check if an event has triggered
+
   bool has_triggered() const;
-  // wait on the event
   void wait() const;
-  // Merge events together to create a new event
-  static Event merge_events(Event ev1, Event ev2);
   static Event merge_events(const set<Event> &to_merge);
@@
 };
+\end{lstlisting}
+\end{lrbox}
+\subfigure{\usebox{\mylistingbox}} \\
 
+\begin{lrbox}{\mylistingbox}
+\begin{lstlisting}[firstnumber=14]
+class Processor {
+  const unsigned id;
+  typedef unsigned TaskFuncID;
+  typedef void (*TaskFuncPtr)(void *args,size_t arglen,Processor p);
+  typedef map<TaskFuncID, TaskFuncPtr> TaskIDTable;
+
+  enum Kind { CPU_PROC,GPU_PROC /* ... */ };
+  Kind kind() const;
+
+  Event spawn(TaskFuncID func_id,const void *args,size_t arglen, Event wait_on) const;
+};
+\end{lstlisting}
+\end{lrbox}
+\subfigure{\usebox{\mylistingbox}} \\
+
+\begin{lrbox}{\mylistingbox}
+\begin{lstlisting}[firstnumber=25]
 class Barrier : public Event {
@@
   void alter_arrival_count(int delta) const;
-  void arrive(unsigned count = 1,
-                  Event wait_for = Event::NO_EVENT) const;
+  void arrive(unsigned count = 1, Event wait_for = NO_EVENT) const;
 };
 \end{lstlisting}
+\end{lrbox}
+\subfigure{\usebox{\mylistingbox}} \\
+
+\begin{lrbox}{\mylistingbox}
+\begin{lstlisting}[firstnumber=30]
+class Lock {
+  const unsigned id;
+  Event lock(Event wait_on = NO_EVENT) const;
+  void unlock(Event wait_on = NO_EVENT) const;
+
+  static Lock create_lock(size_t payload_size = 0);
+  void *payload_ptr();
+  void destroy_lock();
+};
+\end{lstlisting}
+\end{lrbox}
+\subfigure{\usebox{\mylistingbox}} \\
+
+\begin{lrbox}{\mylistingbox}
+\begin{lstlisting}[firstnumber=39]
+class Memory {
+  const unsigned id;
+  size_t size() const;
+};
+
+class RegionMetaData {
+  const unsigned id;
+  static const RegionMetaData NO_REGION;
+
+  static RegionMetaData create_region(size_t num_elmts, size_t elmt_size);
+  void destroy_region() const;
+
+  RegionInstance create_instance(Memory memory) const;
+  RegionInstance create_instance(Memory memory, ReductionOpID redopid) const;
+  void destroy_instance(RegionInstance instance) const;
+};
+
+class RegionInstance {
+  const unsigned id;
+
+  Event copy_to(RegionInstance target, Event wait_on = NO_EVENT);
+  Event reduce_to(RegionInstance target, ReductionOpID redopid, Event wait_on = NO_EVENT);
+};
+\end{lstlisting}
+\end{lrbox}
+\subfigure{\usebox{\mylistingbox}} \\
+
+\begin{lrbox}{\mylistingbox}
+\begin{lstlisting}[firstnumber=62]
+class Machine {
+  Machine(int *argc, char ***argv, const Processor::TaskIDTable &task_table);
+
+  void run(Processor::TaskFuncID task_id, const void *args, size_t arglen);
+
+  static Machine* get_machine(void);
+  const set<Memory>& get_all_memories(void) const;
+  const set<Processor>& get_all_processors(void) const;
+
+  int get_proc_mem_affinity(vector<ProcMemAffinity> &result, ...);
+  int get_mem_mem_affinity(vector<MemMemAffinity> &result, ...);
+};
+\end{lstlisting}
+\end{lrbox}
+\subfigure{\usebox{\mylistingbox}} \\
+
+\caption{Runtime Interface.\label{fig:runtimeapi}}
+\vspace{-4mm}
+\end{figure}
 
@@
 \label{subsec:procs}
-Listing~\ref{lst:procapi} shows the interface for {\tt Processor} objects which allow for
+Lines 14-24 of Figure~\ref{fig:runtimeapi} show the interface for {\tt Processor} objects which allow for
 the creation of parallel computations.  Processors provide a way of naming 
@@
 in Section~\ref{subsec:machmodel}.  In the case of our current interface 
-processors are either individual cores or discrete GPUs, but this is easily modified 
-to include new processor types (line 10).   Processors support a single {\tt spawn}
-method (lines 14-15) that launches a new {\em task} on that processor.
+processors are either individual CPU cores or discrete GPUs, but this is easily modified 
+to include new processor types (line 20).   Processors support a single {\tt spawn}
+method (line 23) that launches a new {\em task} on that processor.
 Note that because the spawn operation
@@
 processor from anywhere in the system.  
-The spawn call takes an optional event that must trigger before the task can begin.  
-The spawn call returns an event that will trigger when the task completes.
+The spawn call takes an optional event that must trigger before the task can begin, and
+always returns an event that will trigger when the task completes.
 
-Lines 18-24 of Listing~\ref{lst:procapi} show a simple example using 
+Figure~\ref{fig:procevents} shows a simple example using 
 the processor and event interface.  The {\tt diamond\_task} launches four
-sub-tasks with a diamond dependence pattern, as shown in Figure~\ref{fig:procevents}.
+sub-tasks with a diamond dependence pattern.
 The {\tt merge\_events} call
@@
 for the last sub-task.
-
-\begin{lstlisting}[float={t},label={lst:procapi},caption={Processor Interface.},belowskip=0pt]
-class Processor {
-  const unsigned id;
-  typedef unsigned TaskFuncID;
-  typedef void (*TaskFuncPtr)
-            (const void *args,size_t arglen,Processor p);
-  typedef map<TaskFuncID, TaskFuncPtr> TaskIDTable;
-
-  enum Kind {
-    CPU_PROC,GPU_PROC,// ... future processor types
-  };
-  Kind kind() const;
-
-  Event spawn(TaskFuncID func_id,const void *args,size_t arglen,
-              Event wait_on = Event::NO_EVENT) const;
-};
-\end{lstlisting}
 
@@
 
+\subsection{Barriers}
+\label{subsec:barriers}
+The common synchronization pattern in which multiple consumers are dependent on
+a set of producers is supported by a {\tt Barrier} object (lines 25-29 of Figure~\ref{fig:runtimeapi}), which is similar
+to a user event, but does not trigger until the expected number of {\tt arrive}
+operations have occurred.  To better support
+composability and asynchronous operation, our interface differs from barriers in other programming
+models\cite{MPI} in three fundamental ways:
+%The common synchronization pattern in which multiple consumers are dependent on a 
+%(potentially different) set of producers is supported by a {\tt Barrier} object,
+%which is similar to a {\tt UserEvent}, but does not trigger until the expected
+%number of {\tt arrive} operations have occurred.  Barrier constructs exist in other
+%programming models\cite{MPI}, but in addition to the decoupling of the waiters from
+%the arrivers, our barrier interface has two further improvements to better support
+%composability and asynchronous operation:
+\begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
+\item The set of producers arriving at the barrier can be different from the set of consumers waiting on the barrier.
+\item An arrival operation can be made dependent on another event, allowing a client to
+express that an operation should be completed before the barrier can trigger.
+\item The expected arrival count of a barrier can be dynamically modified, allowing
+producer tasks to be subdivided as necessary, even after the barrier has been created.
+\end{itemize}
+
+Like user events, barriers are a sub-type of events and can therefore
+be used transparently to express dependences wherever events are used.
+
 \subsection{Deferred Locks}
@@
 
-\begin{lstlisting}[float={t},label={lst:lockapi},caption={Deferred Lock Interface.}]
-class Lock {
-  const unsigned id;
-  Event lock(Event wait_on = Event::NO_EVENT) const;
-  void unlock(Event wait_on = Event::NO_EVENT) const;
-  // Create a new lock, destroy an existing lock
-  static Lock create_lock(size_t payload_size = 0);
-  void *payload_ptr();
-  void destroy_lock();
-};
-\end{lstlisting}
-
 \begin{figure}
@@
 \begin{lstlisting}
-void launcher_task(const void *args, size_t arglen, Processor p) {
-  // Unpack lock, input events from arguments
+void lock_example(const void *args, size_t arglen, Processor p) {
+  // Unpack lock, input events, target processors from arguments
   Lock needed = ...
@@
   Event lock_event_1 = needed.lock(prev_event_1);
-  Event task_event_1 = p.spawn(SUB, args_1, sizeof(args_1), lock_event_1);
+  Event task_event_1 = proc1.spawn(SUB, args_1, sizeof(args_1), lock_event_1);
   needed.unlock(task_event_1);
@@
   Event lock_event_2 = needed.lock(prev_event_2);
-  Event task_event_2 = p.spawn(SUB, NULL, 0, lock_event_2);
+  Event task_event_2 = proc2.spawn(SUB, args_2, sizeof(args_2), lock_event_2);
   needed.unlock(task_event_2);
@@
 \begin{tikzpicture}
-  \draw (1.0,1.5) rectangle (6.2,2.4);
-  \node (l1) at (3.6,1.95) {\begin{tabular}{c}{\bf lock}\\needed\end{tabular}};
+  \node (l1) at (2,1.95) [shape=rectangle,draw] {\begin{tabular}{c}{\bf lock}\\needed\end{tabular}};
+  \node (t1) at (4.5,1.95) [shape=rectangle,draw] {\begin{tabular}{c}{\bf run}\\SUB(args\_1)\end{tabular}};
+  \node (u1) at (7,1.95) [shape=rectangle,draw] {\begin{tabular}{c}{\bf unlock}\\needed\end{tabular}};
 
-  \draw (6.7,1.5) rectangle (8.6,2.4);
-  \node (t1) at (7.65,1.95) {\begin{tabular}{c}{\bf run}\\SUB(args\_1)\end{tabular}};
+  \draw [->] (0,1.95) to (l1.west);
+  \draw [->] (l1.east) to (t1.west);
+  \draw [->] (t1.east) to (u1.west);
 
-  \draw (9.1,1.5) rectangle (10.5,2.4);
-  \node (u1) at (9.8,1.95) {\begin{tabular}{c}{\bf unlock}\\needed\end{tabular}};
+  \node (l2) at (2,0.25) [shape=rectangle,draw] {\begin{tabular}{c}{\bf lock}\\needed\end{tabular}};
+  \node (t2) at (4.5,0.25) [shape=rectangle,draw] {\begin{tabular}{c}{\bf run}\\SUB(args\_2)\end{tabular}};
+  \node (u2) at (7,0.25) [shape=rectangle,draw] {\begin{tabular}{c}{\bf unlock}\\needed\end{tabular}};
 
-  \draw [->] (0.0,1.95) to (1.0,1.95);
-  \draw [->] (6.2,1.95) to (6.7,1.95);
-  \draw [->] (8.6,1.95) to (9.1,1.95);
+  \draw [->] (0,0.25) to (l2.west);
+  \draw [->] (l2.east) to (t2.west);
+  \draw [->] (t2.east) to (u2.west);
 
-  \draw (0.3,0.1) rectangle (1.7,1.0);
-  \node (l2) at (1.0,0.55) {\begin{tabular}{c}{\bf lock}\\needed\end{tabular}};
+  \draw [->,dotted,thick] (0.5,2.5) to (l2.140);
+  \draw [->,dotted,thick] (u2.40) to (l1.220);
+  \draw [->,dotted,thick] (u1.320) to (8.5,-0.5);
 
-  \draw (2.2,0.1) rectangle (4.1,1.0);
-  \node (t2) at (3.15,0.55) {\begin{tabular}{c}{\bf run}\\SUB(args\_2)\end{tabular}};
+  %% \draw (1.0,1.5) rectangle (6.2,2.4);
+  %% \node (l1) at (3.6,1.95) {\begin{tabular}{c}{\bf lock}\\needed\end{tabular}};
 
-  \draw (4.6,0.1) rectangle (6.0,1.0);
-  \node (u2) at (5.3,0.55) {\begin{tabular}{c}{\bf unlock}\\needed\end{tabular}};
+  %% \draw (6.7,1.5) rectangle (8.6,2.4);
+  %% \node (t1) at (7.65,1.95) {\begin{tabular}{c}{\bf run}\\SUB(args\_1)\end{tabular}};
 
-  \draw [->] (0.0,0.55) to (0.3,0.55);
-  \draw [->] (1.7,0.55) to (2.2,0.55);
-  \draw [->] (4.1,0.55) to (4.6,0.55);
+  %% \draw (9.1,1.5) rectangle (10.5,2.4);
+  %% \node (u1) at (9.8,1.95) {\begin{tabular}{c}{\bf unlock}\\needed\end{tabular}};
 
-  \draw [->,dotted,thick] (0.2,2.5) to (0.5,1.0);
-  \draw [->,dotted,thick] (5.8,1.0) to (6.0,1.5);
-  \draw [->,dotted,thick] (10.3,1.5) to (10.6,0.0);
+  %% \draw [->] (0.0,1.95) to (1.0,1.95);
+  %% \draw [->] (6.2,1.95) to (6.7,1.95);
+  %% \draw [->] (8.6,1.95) to (9.1,1.95);
+
+  %% \draw (0.3,0.1) rectangle (1.7,1.0);
+  %% \node (l2) at (1.0,0.55) {\begin{tabular}{c}{\bf lock}\\needed\end{tabular}};
+
+  %% \draw (2.2,0.1) rectangle (4.1,1.0);
+  %% \node (t2) at (3.15,0.55) {\begin{tabular}{c}{\bf run}\\SUB(args\_2)\end{tabular}};
+
+  %% \draw (4.6,0.1) rectangle (6.0,1.0);
+  %% \node (u2) at (5.3,0.55) {\begin{tabular}{c}{\bf unlock}\\needed\end{tabular}};
+
+  %% \draw [->] (0.0,0.55) to (0.3,0.55);
+  %% \draw [->] (1.7,0.55) to (2.2,0.55);
+  %% \draw [->] (4.1,0.55) to (4.6,0.55);
+
+  %% \draw [->,dotted,thick] (0.2,2.5) to (0.5,1.0);
+  %% \draw [->,dotted,thick] (5.8,1.0) to (6.0,1.5);
+  %% \draw [->,dotted,thick] (10.3,1.5) to (10.6,0.0);
 
@@
 
-Listing~\ref{lst:lockapi} shows the 
-interface for locks.  Unlike blocking locks, the {\tt lock} method (line 4) doesn't
-block but instead always returns immediately with an event that will be triggered
-when the lock has been acquired.  Just like the spawn method, the lock method also 
+The interface for deferred locks is shown in lines 30-38 of Figure~\ref{fig:runtimeapi}.
+Unlike traditional blocking locks, the {\tt lock} method (line 32) doesn't
+wait if the lock is in use but instead always returns immediately with an event that will be triggered
+when the lock has been acquired.  Like a processor's spawn method, the lock method also 
 takes an optional event parameter as a precondition.  Similarly, the {\tt unlock} 
-method (line 5) takes an optional event precondition parameter.
+method (line 33) takes an optional event precondition parameter.
 
@@
 convention in writing code with deferred locks is to acquire locks on behalf of a task being
-launched.  Listing~\ref{lst:lockapi} illustrates this with a simple example.  The
-{\tt launcher\_task} is calling a sub-task that is going to require the {\tt needed}
-lock.  A lock request is issued and the resulting {\tt lock\_event} is used as
-a precondition for launching the {\tt SUB} task.  The unlock operation is then
-made dependent
-%precondition
-on the event corresponding to {\tt SUB} task's completion.  The
-{\tt SUB} task can run safely while holding the lock and the lock is released when the
-{\tt SUB} task completes.  Figure~\ref{fig:lockevents} shows the event dependence graph, with
-dotted lines representing a lock operation's dependence on a previous unlock.
-Using deferred locks in this style prevents 
-compute resources from waiting on locks by ensuring tasks only run when their needed
-locks have already been acquired.
+launched.  Figure~\ref{fig:lockevents} illustrates this with a simple example.  The {\tt lock\_example}
+function needs to launch two subtasks on two othe processors, each with exclusive access to some resource protected by a
+lock named {\tt mutex}.  For each subtask, it first requests the lock on the subtaks behalf, making the
+subtask wait until the lock has been acquired.  The parent task also requests that the runtime automatically
+release the lock after each subtask has completed.  In the event dependency graph, solid arrows represent
+the explicit events used in the code, while the dotted line represents the dynamic ordering that results
+from the mutual exclusion guarantee of the lock.  (In this example, the second subtask is granted the lock first.)
 
@@
 piece of data that is moved around with the lock and guaranteed to be coherent while 
-the lock is held.  The size of the payload is specified when the lock is created (line 7) and
+the lock is held.  The size of the payload is specified when the lock is created (line 35 of Figure~\ref{fig:runtimeapi}) and
 the pointer to the local copy of the payload can be obtained from the {\tt payload\_ptr} method
-(line 8).
+(line 36).
 
@@
 % Say something about dynamically modifying the number of elements in a class
-
-\begin{lstlisting}[float={t},label={lst:regionapi},caption={Physical Region Interface and Example.}]
-class Memory {
-  const unsigned id;
-  size_t size() const;
-};
-
-class RegionMetaData {
-  const unsigned id;
-  static const RegionMetaData NO_REGION;
-  // Create and destroy metadatas
-  static RegionMetaData create_region(size_t num_elmts, 
-                                      size_t elmt_size);
-  void destroy_region() const;
-  // Create and destroy instances
-  RegionInstance create_instance(Memory memory) const;
-  void destroy_instance(RegionInstance instance) const;
-  // Create a reduction-only instance
-  RegionInstance create_instance(Memory memory, 
-                        ReductionOpID redopid) const;
-};
-
-class RegionInstance {
-  const unsigned id;
-  static const RegionInstance NO_INST;
-  // Copy between instances
-  Event copy_to(RegionInstance target, 
-                Event wait_on = Event::NO_EVENT);
-  Event reduce_to(RegionInstance target, ReductionOpID redopid,
-                Event wait_on = Event::NO_EVENT);
-};
-\end{lstlisting}
 
@@
 
-Listing~\ref{lst:regionapi} shows a subset of the interface for physical regions.  
-Each memory in the machine's memory hierarchy is named by a {\tt Memory} object (line 1).
-The next object is a {\tt RegionMetaData} object (line 7), which defines a class of
+Lines 39-61 of Figure~\ref{fig:runtimeapi} show a subset of the interface for physical regions.  
+Each memory in the machine's memory hierarchy is named by a {\tt Memory} object (line 39).
+The next object is a {\tt RegionMetaData} object (line 44), which defines a class of
 regions.  Physical regions that belong to that class are created with the {\tt create\_instance}
-method (line 16) and represented by {\tt RegionInstance} objects (line 23).  When creating a physical
+method (line 51) and represented by {\tt RegionInstance} objects (line 56).  When creating a physical
 region the client specifies the memory in which the physical region is going
@@
 same class.  It is the client's responsibility to manage data coherence using copies
-between physical regions (line 28-29).  Like other operations copies can be
+between physical regions (line 59).  Like other operations copies can be
 predicated on an event and return an event corresponding to completion.
@@
 If the only operations a task will perform on a region are reductions, a special reduction-only
-instance can be created.  Rather than applying each reduction separately to the target instance,
+instance can be created (line 52).  Rather than applying each reduction separately to the target instance,
 the reduction operations can be accumulated in the reduction instance, and applied in bulk to 
-the target using the {\em reduce\_to} operation.  This has several benefits.  First, it results
+the target using the {\em reduce\_to} operation (line 60).  This has several benefits.  First, it results
 in much better performance, as we will show in section~\ref{subsec:reducmicro}), by reducing both
@@
 
-An example using reduction-only regions is shown on lines 41-55 of Listing~\ref{lst:regionapi}.
+An example using reduction-only regions is given in Figure~\ref{fig:reducevents} along with the resulting event graph.
 Two reduction-only regions, {\tt inst1} and {\tt inst2}, are created in different memories to
-be used with a reduction operation (lines 46-47).  (Reduction operations are registered at startup with integer IDs
+be used with a reduction operation (lines 7-8).  (Reduction operations are registered at startup with integer IDs
 as function pointers may not be portable across nodes.)  A physical region of the same class {\tt inst3}
-is also created in a different memory (line 48).  Two tasks are launched that will apply
-reductions locally into their respective reduction buffers (lines 49-50).  The resulting
-events are then used to chain reduction copies back to {\tt inst3} after the tasks complete (lines 51-52).
+is also created in a different memory (line 9).  Two tasks are launched that will apply
+reductions locally into their respective reduction buffers (lines 10-11).  The resulting
+events are then used to chain reduction copies back to {\tt inst3} after the tasks complete (lines 12-13).
 The reduction operations accumulated in each of {\tt inst1} and {\tt inst2} are applied atomically
@@
 result is correct.  Finally, a merge on the events from the two reduction copies is 
-performed and a final task using the results in {\tt inst3} is launched (lines 53-54).  This event
-graph is shown in Figure~\ref{fig:reducevents}.
+performed and a final task using the results in {\tt inst3} is launched (lines 14-15).
 
@@
 \label{subsec:machmodel}
-The last component of the interface is shown in Listing~\ref{lst:machineapi} and allows the
-initialization and run-time introspection
+The last component of the interface is the {\tt Machine} object, shown in lines 62-73 of Figure~\ref{fig:runtimeapi},
+which provides initialization and run-time introspection
 of the current machine.
@@
 program, providing a task table that maps task IDs to function pointers,
-and then invoking the {\tt run} method (lines 9-10).
+and then invoking the {\tt run} method (lines 65).
 Any running task can get a pointer to the machine object with the
-{\tt get\_machine} method (line 12) and use it to obtain lists of all the {\tt Memory} (line 13) or 
-{\tt Processor} (line 14) handles in the system.  Using the affinity methods (lines 16-19),
+{\tt get\_machine} method (line 67) and use it to obtain lists of all the {\tt Memory} (line 68) or 
+{\tt Processor} (line 69) handles in the system.  Using the affinity methods (lines 71-72),
 the application can also determine which memories are accessible by each
@@
 %about all processors and memories in the system.
-
-\begin{lstlisting}[float={t},label={lst:machineapi},caption={Machine Interface.}]
-class Machine {
-  Machine(int *argc, char ***argv,
-          const Processor::TaskIDTable &task_table);
-  enum RunStyle {
-    ONE_TASK_ONLY,ONE_TASK_PER_NODE,
-    ONE_TASK_PER_PROCESSOR,
-  };
-  void run(Processor::TaskFuncID task_id = 0, 
-        RunStyle = ONE_TASK_ONLY, const void *args, size_t arglen);
-
-  static Machine* get_machine(void);
-  const set<Memory>& get_all_memories(void) const;
-  const set<Processor>& get_all_processors(void) const;
-
-  int get_proc_mem_affinity(vector<ProcMemAffinity> &result,
-          Processor restrict_proc = 0, Memory restrict_mem = 0);
-  int get_mem_mem_affinity(vector<MemMemAffinity> &result,
-          Memory restrict_mem1 = 0, Memory restrict_mem2 = 0);
-};
-\end{lstlisting}

### 2012-11-10 doc/llr-pldi2013/interface.tex
some edits and working on the reduction story

@@
 express that an operation should be completed before the barrier can trigger.
-\item The expected arrival count of a barrier can be dynamically modified, allowing
-producer tasks to be subdivided as necessary, even after the barrier has been created.
+\item The expected arrival count of a barrier can be dynamically modified, which we
+have found useful in supporting nested parallelism in higher-level languages.
+%allowing
+%producer tasks to be subdivided as necessary, even after the barrier has been created.
 \end{itemize}
@@
 {\em Deferred locks} are a new synchronization mechanism that allows for synchronization
-in an completely asynchronous environment.
+without blocking or spinning.
+%in an completely non-blocking environment.
 
@@
 wait if the lock is in use but instead always returns immediately with an event that will be triggered
-when the lock has been acquired.  Like a processor's spawn method, the lock method also 
-takes an optional event parameter as a precondition.  Similarly, the {\tt unlock} 
-method (line 33) takes an optional event precondition parameter.
+when the lock has been acquired.  Like a processor's spawn method, both the {\tt lock} 
+and {\tt unlock} methods take an optional event paramenter as a precondition (lines 32-33).
 
@@
 When a lock is used to mediate access to a piece of data, the first thing a task will usually
-do after being granted the lock is access the data.  To be able to hide the latency of that
+do after being granted the lock is access the data.  To hide the latency of that
 access as well, an additional feature of locks in our interface is the association of a lock with a
@@
 \label{subsec:phyreg}
-In distributed machines with discrete memories, data movement often consists of more
-than a simple copy operation.  Many applications perform operations and transformations on
-their data in conjunction with data movement for higher performance.  One common example is 
-that applications accumulate reduction operations in a buffer and then, as part of a copy operation,
-the reduction operations are applied to a destination buffer on the receiving side of the copy.  
-To support these kinds of conjoined operations in an asynchronous environment, our interface 
-must be aware of the structure of data so it can perform accompanying operations with 
-data movement operations.  To describe the structure of data our interface uses {\em physical regions.}
+%In distributed machines with discrete memories, data movement often consists of more
+%than a simple copy operation.  Many applications perform operations and transformations on
+%their data in conjunction with data movement for higher performance.  One common example is 
+%that applications accumulate reduction operations in a buffer and then, as part of a copy operation,
+%the reduction operations are applied to a destination buffer on the receiving side of the copy.  
+%To support these kinds of conjoined operations in an asynchronous environment, our interface 
+%must be aware of the structure of data so it can perform accompanying operations with 
+%data movement operations.  To describe the structure of data our interface uses {\em physical regions.}
+Current APIs for moving data in machines with discrete memories only support untyped
+copies of a buffer of data.  This restriction requires programmers to manage both the layout of
+data and the marshalling of data for communication.  {\em Physical regions} are a
+mechanism for decoupling the placement of data from its layout.  Physical regions make
+the runtime aware of the structure of elements in the allocation and thereby enable multiple data
+layouts, such as specialized layouts for buffering individual reduction operations.
 

### 2012-11-10 doc/llr-pldi2013/intro.tex
some edits and working on the reduction story

--- 2012-07-23 doc/asplos2013/intro.tex
@@
 hinders attempts to hide the increasingly long-latency communication
-or memory movement phases.  
+or synchronization phases.  
 %In older machines this was not a problem because slower
@@
 precondition an event that must trigger before $p$ can begin.
-Any number of operations can be depedendent on a single event, and
-a events can be merged to allow an operation to wait for multiple
-events before it begins.
+%Any number of operations can be depedendent on a single event, and
+%a events can be merged to allow an operation to wait for multiple
+%events before it begins.
+Events can be merged into new events and multiple operations
+can use an event as a precondition.
   Using
@@
 respects the event dependences.  With this freedom the runtime can
-schedule operations to optimize throughput while hiding the long
+schedule operations to optimize throughput while hiding long
 latency operations.
@@
 
-\item We describe a physical region system that provides structural information about
-data layout.  The structural information provided by physical regions enables an extensible
-interface for conjoining operations on data (e.g. reductions) with asynchronous data movement for 
-further latency hiding (Section~\ref{subsec:phyreg}).
+%\item We describe a physical region system that provides structural information about
+%data layout.  The structural information provided by physical regions enables an extensible
+%interface for conjoining operations on data (e.g. reductions) with asynchronous data movement for 
+%further latency hiding (Section~\ref{subsec:phyreg}).
+\item We describe a physical region system that encourages efficient movement of data
+via bulk transfers.  The structural information provided by physical regions also enables
+special reduction instances that buffer individual reduction operations for bulk
+transfer and application (Section~\ref{subsec:phyreg}).
 
@@
 We describe a distributed implementation of our interface that requires no global coordination
-and has modest local storage costs.  We describe optimizations for the implementation of events, 
+and has modest local storage costs.  We describe an optimized implementation of events, 
 deferred locks, and reductions (Section~\ref{sec:impl}).
 
-\item We report on micro-benchmarks that are designed to stress
+\item We report on micro-benchmarks that stress
 our implementation.  The results indicate that our implementation approaches the performance of
@@
 are capable of achieving high-performance (Section~\ref{sec:apps}).  Furthermore, we demonstrate
-that the same applications written in a bulk-synchronous model incur performance penalties between 22-135\%.
+that the same applications written in a bulk-synchronous model incur slowdowns between 22-135\%.
 \end{itemize}

### 2012-11-10 doc/llr-pldi2013/legion-llr.tex
First commit on new low-level runtime paper for pldi
redid reduction stuff, monkeyed with figures to put examples with event graphs
some edits and working on the reduction story

--- 2012-07-23 doc/asplos2013/legion-llr.tex
@@
-\documentclass{sig-alternate}
-
+\documentclass[9pt,nocopyrightspace,preprint]{sigplanconf}
+\bibliographystyle{plainnat}
 \usepackage{amsmath}
@@
 \newtheorem{lemma}[theorem]{Lemma}
+\newcommand{\texcomment}[1]{}
+
+\setlength{\pdfpagewidth}{8.5in}
+\setlength{\pdfpageheight}{11in}
+
+\pagenumbering{arabic}
+
+\newsavebox{\mylistingbox}
 
@@
 \title{High-Performance Language Primitives for Distributed Architectures using Asynchronous Events}
-\numberofauthors{3}
-\author{
-\alignauthor Sean Treichler\\
-             \affaddr{Stanford University} \\
-             \email{sjt@cs.stanford.edu}
-\alignauthor Michael Bauer \\
-             \affaddr{Stanford University} \\
-             \email{mebauer@cs.stanford.edu}
-\alignauthor Alex Aiken \\
-             \affaddr{Stanford University} \\
-             \email{aiken@cs.stanford.edu}
-}
+% Commented out for double-blind goodness
+%% \numberofauthors{3}
+%% \author{
+%% \alignauthor Sean Treichler\\
+%%              \affaddr{Stanford University} \\
+%%              \email{sjt@cs.stanford.edu}
+%% \alignauthor Michael Bauer \\
+%%              \affaddr{Stanford University} \\
+%%              \email{mebauer@cs.stanford.edu}
+%% \alignauthor Alex Aiken \\
+%%              \affaddr{Stanford University} \\
+%%              \email{aiken@cs.stanford.edu}
+%% }
+\authorinfo{}{}{}
 \maketitle
@@
 parallel machines that provides a fully asynchronous 
-programming model, allowing all types of runtime actions, including synchronization,
+programming model, allowing all runtime actions, including synchronization,
 to be non-blocking.  Our interface includes mechanisms for the creation and movement
 of data through the memory hierarchy as well as {\em deferred locks}, a novel synchronization primitive.
-Asynchrony is exposed by the interface using a distributed event system.  All
-operations return an event to be triggered when the operation completes.  Events
+Asynchrony is exposed by the interface via a distributed event system.  All
+operations return an event that will be triggered when the operation completes.  Events
 can be chained together and used to capture dependences between operations.
@@
 {
-\small
-\bibliographystyle{abbrv}
 \bibliography{bibliography}

### 2012-11-11 doc/llr-pldi2013/impl.tex
updates

@@
 In addition to the RMA capabilities, GASNet provides {\em active messages} for 
-inter-node communication.  Active messages consist of a command and payload that
+inter-node communication.  Implementation of all runtime features requiring communication
+rely heavily on active messages.  Active messages consist of a command and payload that
 are sent by one node to another without any previous coordination.  Upon arrival
@@
 %for a response.  
-One example of a case where we use active messages involves the {\tt spawn} call on a {\tt Processor}
-object.  If the processor is not on the same node as the requestor, the information
+%One example of a case where we use active messages involves the {\tt spawn} call on a {\tt Processor}
+%object.  If the processor is not on the same node as the requestor, the information
 %about the task (e.g. the task ID and arguments, the actual target {\tt Processor},
 %the prerequisite event (if any), and the completion event ID) 
-for the task is sent in an
-active message to the node containing the target processor.  The recipient node
-unpacks the message, and either places the task on the ready queue for the target
-processor or marks it as being deferred until the prerequisite event has triggered.
+%for the task is sent in an
+%active message to the node containing the target processor.  The recipient node
+%unpacks the message, and either places the task on the ready queue for the target
+%processor or marks it as being deferred until the prerequisite event has triggered.
 %No response is sent to the original requestor.  The requestor already knows the event
@@
 Events are created on demand and are {\em owned} by the node on which they
-were created.  The creation of an event occurs with no inter-node communication.
+are created.  Creating an event requires no inter-node communication.
 The space of event
 handles is divided across the nodes at start-up time, allowing each node to assign handles
-to new events without the risk of conflicting with another node's assignments.  The static division
+to new events without conflicting with another node's assignments.  The static division
 of event handles also permits any node to determine the owning node of an event
-without any communication.
+without communication.
 %No broadcast of
@@
 The first operation dependent on an event
-from a remote node will allocate the same data structure on the remote node.  A single
+from a remote node allocates the same data structure on the remote node.  A single
 {\em event subscription} active message is then sent from the remote node to the owning node indicating
@@
 would have been to share the list of subscribers so that the triggering node could notify all interested
-nodes directly.  However, such an algorithm is both more complicated due to race conditions, and requires $O(N^2)$
+nodes directly.  However, such an algorithm is both more complicated (due to race conditions) and requires $O(N^2)$
 active messages.  Any algorithm that is super-linear with the node count will not scale well, 
@@
 
-Like events, deferred locks, or locks from here on, are created on demand by the node on which the creation request was made.  The lock
-handles are also statically divided across the nodes, allowing the creation to be done without inter-node
-communication.  Whereas event ownership is static, locks may migrate; the {\em creating}
-node is the initial owner, but that ownership can be transferred to other nodes.
+Like events, deferred locks (henceforth called simply locks) are
+created on demand by the node of the creation request. 
+Lock handles are also statically divided across the nodes,
+allowing creation to be done without inter-node communication.
+Whereas event ownership is static, locks may migrate; the {\em
+  creating} node is the initial owner, but ownership can be
+transferred to other nodes.
 
@@
 not held, the ownership of the lock is given to the requesting node via a {\em lock transfer} active
-message.  A lock transfer message includes the bitmask of remote waiters and an up-to-date copy 
+message, which includes the bitmask of remote waiters and an up-to-date copy 
 of the lock's payload.  The inclusion of the payload in the active message is the reason for the 

### 2012-11-11 doc/llr-pldi2013/interface.tex
updates

@@
 \label{sec:interface}
-The target set of clients for our interface are higher-level language
+The clients for our interface are higher-level language
 compilers and runtimes such as Legion\cite{Legion12} as well as 
-advanced systems programmers.  This class of clients expect total control
+advanced systems programmers.  These clients expect total control
 over the underlying hardware and transparent performance from an
@@
 supports testing whether an event has triggered (line 5) and waiting on
-an event to trigger (line 6).  While these methods can be useful, the preferred
+an event to trigger (line 6), but the preferred
 use of events is passing them as preconditions for other operations.  The client can use the
@@
 every computational unit in the machine.  We describe how to discover processor handles
-in Section~\ref{subsec:machmodel}.  In the case of our current interface 
-processors are either individual CPU cores or discrete GPUs, but this is easily modified 
+in Section~\ref{subsec:machmodel}.  In the current interface 
+processors are either individual CPU cores or discrete GPUs, but this is easily extended
 to include new processor types (line 20).   Processors support a single {\tt spawn}
@@
 Note that because the spawn operation
-is invoked on a processor handle, it is possible to launch a task on any
+is invoked on a processor handle, a task can be launched on a 
 processor from anywhere in the system.  
-The spawn call takes an optional event that must trigger before the task can begin, and
-always returns an event that will trigger when the task completes.
+The spawn call takes an optional event that must trigger before the task begins and
+always returns an event that triggers when the task completes.
 
-Figure~\ref{fig:procevents} shows a simple example using 
+Figure~\ref{fig:procevents} shows an example using 
 the processor and event interface.  The {\tt diamond\_task} launches four
@@
 launched.  Figure~\ref{fig:lockevents} illustrates this with a simple example.  The {\tt lock\_example}
-function needs to launch two subtasks on two othe processors, each with exclusive access to some resource protected by a
-lock named {\tt mutex}.  For each subtask, it first requests the lock on the subtaks behalf, making the
+function needs to launch two subtasks on two other processors, each with exclusive access to some resource protected by a
+lock named {\tt mutex}.  For each subtask, it first requests the lock on the subtask's behalf, making the
 subtask wait until the lock has been acquired.  The parent task also requests that the runtime automatically
@@
 {\em payload}, a small (i.e. less than 4KB) 
-piece of data that is moved around with the lock and guaranteed to be coherent while 
+piece of data that is moved with the lock and guaranteed to be coherent while 
 the lock is held.  The size of the payload is specified when the lock is created (line 35 of Figure~\ref{fig:runtimeapi}) and
@@
 
-Between deferred locks and barriers (described in Section~\ref{subsec:events}) clients 
-have access to the same set of synchronization primitives that they have
+With deferred locks and barriers (see Section~\ref{subsec:events}) clients 
+have the same synchronization primitives as
 in threading and bulk-synchronous interfaces.  However, these operations can now 
@@
 A physical region is an allocation of data in a single memory in the memory hierarchy.  Physical
-regions are grouped into {\em classes} which are sets of physical regions that share the
-same names (i.e. addresses) for elements, but don't necessarily share the same data layout.  This
+regions are grouped into {\em classes}, which are sets of physical regions sharing the
+same names (i.e. addresses) for elements, but not necessarily sharing the same data layout.  This
 allows pointers to elements within regions to be used, even if the regions are copied from one memory
@@
 interface supports sub-classing, but the details are omitted due to space constraints.  The
-interface allows copies between any pair of physical regions that are of
+interface allows copies between physical regions of
 the same class (or between super- and sub-classes).
@@
 method (line 51) and represented by {\tt RegionInstance} objects (line 56).  When creating a physical
-region the client specifies the memory in which the physical region is going
+region the client specifies the memory in which the physical region is
 to be allocated.  To maintain performance transparency, a new physical region can only be allocated
@@
 the target using the {\em reduce\_to} operation (line 60).  This has several benefits.  First, it results
-in much better performance, as we will show in section~\ref{subsec:reducmicro}), by reducing both
+in much better performance, as we show in Section~\ref{subsec:reducmicro}, by reducing both
 the total inter-node traffic and the overhead of that traffic while still allowing parallelism.

### 2012-11-11 doc/llr-pldi2013/results.tex
First commit on new low-level runtime paper for pldi
fixed up some of the negative vspaces that are overkill in the PLDI layout
update

--- 2012-07-23 doc/asplos2013/results.tex
@@
 operations, and synchronization operations with the dependencies between them expressed
-as events.  The full details of the SOOP are beyond the scope of this paper
-and are described in \cite{Legion12}.
+as events.  For details the interested reader is referred to the work of \cite{Legion12}.
 
@@
 be able to recover the storage associated with the event.  The live events line
-is therefore equivalent to the number of needed events in a refernce counting
+is therefore equivalent to the number of needed events in a reference counting
 scheme.  In this example 
@@
 \end{center}
-\vspace{-6mm}
+\vspace{-2mm}
 \caption{Event Lifetimes - Fluid Application.\label{fig:eventlife}}
@@
 }}
-\vspace{-2mm}
+%\vspace{-2mm}
 \caption{Locks by Request and Transfer Counts.\label{fig:lockcount}}
@@
 }
-\vspace{-4mm}
+\vspace{-2mm}
 \caption{Bulk-Synchronous Performance.\label{fig:bulksync}}

### 2012-11-11 doc/llr-pldi2013/conclusion.tex
Edits to the final sections of the low level paper

--- 2012-07-23 doc/asplos2013/conclusion.tex
@@
 become larger and more variable.  To effectively hide these latencies
-future programming systems will need to leverage asynchronous operations
+future programming systems will need to compose asynchronous operations
 to overlap long latency operations with useful work.  We have
@@
 
-We've presented an implementation of the interface that is capable of running
+We've presented an implementation of our interface that is capable of running
 on a large heterogeneous cluster containing both CPUs and GPUs.  We

### 2012-11-11 doc/llr-pldi2013/related.tex
Edits to the final sections of the low level paper

@@
 MPI is the current industry standard for programming large
-supercomputers\cite{MPI}.  While MPI supports asynchronous
+super\-computers\cite{MPI}.  While MPI supports asynchronous
 communication operations there is no mechanism for composing
@@
 Both POSIX threads and OpenMP\cite{OPENMP98} are used for intra-node parallel
-programming in large machines, but neither support any asynchronous
+programming on large machines, but neither support asynchronous
 operations.  CUDA\cite{CUDA} and OpenCL\cite{Khronos:OpenCL} support
@@
 more relaxed properties such as atomicity that can be described by a
-deferred lock.  OpenCL events are not valid globally, but can only be
-used within a single GPU context.
+deferred lock.  OpenCL events are not valid globally in distributed memory
+spaces, and can only be used within a single GPU context.
 
@@
 synchronization mechanisms still require blocking operations.
+
+The design of coarse-grained dataflow languages such as Lucid \cite{Lucid95}
+and CGD \cite{CGD09} are orthogonal to the design of this work.  We view
+the interface presented here as a potential implementation target for any
+number of higher-level programming systems, including coarse-grained
+dataflow programming models.  The implementation of many dataflow languages
+have features that are related to our interface, but are not implemented
+for distributed memory machines.  I-structures from the Id language \cite{Arvind89}
+have similar semantics to events in our interface, but have a much simpler implementation
+because they are limited to use in a shared address space.
 

### 2012-11-11 doc/llr-pldi2013/results.tex
Edits to the final sections of the low level paper

@@
 
-The fully asynchronous design of the interface proposed in
-this paper is crucial to Legion's SOOP implementation.  Without a fully
-asynchronous interface, the SOOP would have to periodically invoke
+The composable design of the interface proposed in
+this paper is crucial to Legion's SOOP implementation.  If all
+operations in the interface could not be composed via events, 
+the SOOP would have to periodically invoke
 blocking operations that would cause processors to stall and severely
 limit Legion's ability to run ahead and keep the machine fully
-occupied.  With a fully asynchronous interface the SOOP can hide latencies by 
-running
-far ahead of the actual execution, limited only by the physical resources
+occupied.  With a fully asynchronous, composable interface the SOOP can hide latencies by 
+running far ahead of the actual execution, limited only by the physical resources
 available in the machine, exactly like a hardware out-of-order processor.
@@
 and distributed across the machine.  The computation for each time step consists
-of three phases that calculate currents on the wires, distribute charge between
+of three phases: calculate currents on the wires, distribute charge between
 components, and update the voltages of all components.  In the distribute charge
@@
 {\em Fluid} is an incompressible fluid flow simulation from the PARSEC
-benchmark suite\cite{bienia11benchmarking}.  The reference implementation can 
-only be run on a single shared memory machine, but when written in Legion, Fluid can
-be run on large distributed machines as well.  The Fluid simulation
+benchmark suite\cite{bienia11benchmarking}.  The reference implementation 
+only runs on shared memory machines, but when written in Legion, Fluid
+runs on large distributed machines as well.  The Fluid simulation
 models particles that flow through cells.  Each time step requires multiple phases
@@
 the Fluid application.  Section~\ref{subsec:lockmig} examines
-the migration of locks in the Circuit and AMR applications.  Finally, we demonstrate
+the performance of deferred locks in the Circuit and AMR applications.  
+%We describe how the Circuit application leverages reduction instances
+%to operate on GPUs in Section~\ref{subsec:reduccase}.  
+Finally, we demonstrate
 that the asynchronous nature of our interface is essential to the
@@
 a reference counting algorithm would reduce the storage needed for dynamic events
-by over 10X, but could only do so at the cost of
+by over 10X, but only at the cost of
 computation and communication for performing the reference counting.
@@
 
-\subsection{Deferred Lock Migration}
+%\subsection{Deferred Lock Migration}
+\subsection{Deferred Lock Performance}
 \label{subsec:lockmig}
+We also instrumented our heterogeneous implementation to profile
+the usage of deferred locks in real applications.  The Circuit and AMR 
+applications both made use of deferred locks, creating 3336 and 1393 
+deferred locks respectively.  Of all created locks in both applications, 
+14\% were migrated at least once during their lifetime.  We measured the 
+lock grant rates for both applications and found them to be many orders of magnitude smaller
+than the maximum lock grant rates achieved by our locking micro-benchmarks
+in Section~\ref{subsec:lockmicro}.  This demonstrates that real applications need deferred
+locks to compose synchronization in a non-blocking environment, and that the
+ability of deferred locks to migrate is necessary for implementing real applications
+on distributed machines.
 
-We also instrumented our heterogeneous implementation to profile the usage of 
-deferred locks in real applications.  We measured how many locks were used in the application.  For
-each lock we measured the number of lock requests and the number of times that the lock
-had to be transferred from one node to another.  Figure~\ref{fig:lockcount} shows the results 
-for the Circuit and AMR application, both running on 16 nodes.  The Circuit application uses
-a larger number of locks (3336 vs. 1393), but most of them (86\%) are only used on the node on 
-which they were created.  The AMR application exhibits more varied locking behavior, and in 
-particular has locks that are requested more often than they are transferred.  In addition
-to allowing synchronization in an asynchronous environment, deferred locks must be able to 
-migrate to implement real applications on distributed machines.
+%We also instrumented our heterogeneous implementation to profile the usage of 
+%deferred locks in real applications.  We measured how many locks were used in the application.  For
+%each lock we measured the number of lock requests and the number of times that the lock
+%had to be transferred from one node to another.  Figure~\ref{fig:lockcount} shows the results 
+%for the Circuit and AMR application, both running on 16 nodes.  The Circuit application uses
+%a larger number of locks (3336 vs. 1393), but most of them (86\%) are only used on the node on 
+%which they were created.  The AMR application exhibits more varied locking behavior, and in 
+%particular has locks that are requested more often than they are transferred.  In addition
+%to allowing synchronization in an asynchronous environment, deferred locks must be able to 
+%migrate to implement real applications on distributed machines.
 
@@
 
+\texcomment{
 \begin{figure}
@@
 \end{figure}
+}
 
-%\subsection{Reduction Case Study}
+%\subsection{Reduction Instances}
 %\label{subsec:reduccase}
@@
 
-Our last set of experiments demonstrates that the asynchronous nature of our interface
-is essential to the high performance of the Legion runtime system.  The Legion
-implementations of the three applications have already been shown to outperform 
-pre-existing implementations\cite{Legion12}.  To quantify the benefit of the asynchronous
-nature of our interface, we re-implemented the applications in
-a bulk-synchronous style to compare against the existing Legion implementations.  
-These bulk-synchronous versions still run on top of 
+Our last set of experiments demonstrates that the composable, asynchronous nature of our interface
+is essential to the high performance of the Legion runtime system.  The work
+of \cite{Legion12} has already shown that Legion implementations perform well compared
+to hand-coded reference implementations.  Rather than compare directly to hand-coded
+reference implementations, we control for any performance advantages gained by using
+the Legion runtime by re-implementing the applications in a bulk-synchronous
+style within the Legion framework.  These bulk-synchronous versions still run on top of 
 the Legion runtime, but invoke blocking operations between computational phases,
-communication phases, and synchronization.
+communication phases, and synchronization.  By controlling for any advantages derived
+from the Legion programming system, we can directly quantify the benefits
+reaped by the asynchronous and composable nature of our interface.
 
@@
 will be essential to hiding the large latencies inherent in such machine architectures.
-The fully asynchronous nature of the interface presented in this paper
-will be crucial for supporting this ability.
+The composable and asynchronous nature of the interface presented in this paper
+will be crucial for supporting this ability in future high-level runtimes.
 

### 2012-11-11 doc/types-pldi2013/alt_example.tex
Working on figures to make the prettier and easier to understand

@@
 rather than expressed in a more structured and restricted
-sublanguage.  In the example we invoke METIS\cite{Metis98}, a dynamic graph
-partitioning library called (see Listing~\ref{lst:metis}).
+sublanguage.  In the example we invoke METIS\cite{Metis98}, a 
+third-party dynamic graph partitioning library (see Listing~\ref{lst:metis}).
 
 \begin{lstlisting}[label={lst:metis},caption={Partition Computation}]
-type NodeList<rl,rn>       = < CircuitNode@rn, node_list<rl,rn>@rl >
-type WireList<rl,rw,rn,rg>= < CircuitWire<rn,rg>@rw, wire_list<rl,rw,rn,rg>@rl >
+type NodeList<rl,rn>       = < CircuitNode@rn, NodeList<rl,rn>@rl >
+type WireList<rl,rw,rn,rg>= < CircuitWire<rn,rg>@rw, WireList<rl,rw,rn,rg>@rl >
 function extern_metis[rl,rn,rw](node_list : NodeList<rl,rn>@rl,
-          wires_list : WireList<rl,rw,rn,rn>@rl), reads(rl,rn,rw), writes(rn) : bool;
+          wires_list : WireList<rl,rw,rn,rn>@rl), reads(rl,rn,rw), writes(rn) : bool
 \end{lstlisting}
@@
 
-Note that there are two alternatives to using multiple partitions of the same data, both of which avoid introducing aliasing.
-We could use a single partition with $2^n$ subregions, one for each
-possible case of sharing; alternatively, all pieces could just use the
+Note there are two alternative approaches to using multiple partitions,
+both of which avoid introducing aliasing:
+first, we could create a single partition with $2^n$ subregions, one for each
+possible case of sharing; second, tasks for each piece could use the
 {\tt all\_nodes} region to access their ghost nodes.  Neither option
 is attractive: the former significantly complicates programming, the
-latter greatly increases data movement required at
-runtime.
+latter greatly increases runtime data movement and would limit parallelism.
 
@@
 For aliased partitions such as the ghost nodes, a {\em multicoloring}
-is used, which assigns each region element (possibly) multiple colors.  
+is used, which assigns each region element to one or more colors.  
 %(Multicolorings are
@@
         newcolor(rn) 
-    else
+    else                                 -- tuple fields accessed by .(field number)
         let list_elem : NodeList<rl,rn> = read(node_list) in
@@
 \begin{lstlisting}[label={lst:rr},caption={Region Relationship Example}]
-type CircuitPiece<rl,rw,rn> = rr[rpw,rpn,rg]
+type CircuitPiece<rl,rw,rn> = rr[rpw,rpn,rg]  -- rr is keyword for region relationship 
                             < WireList<rl,rpw,rpn,rg>@rl, NodeList<rl,rpn>@rl >         
@@
 a region relationship is unpacked later, fresh names are given to the regions that were
-captured by the region relationship, and the properties that are known to hold for these
+captured by the region relationship (passed in {\tt []} parameters), and the properties that are known to hold for these
 regions are re-introduced to the type environment.

### 2012-11-11 doc/types-pldi2013/intro.tex
Working on figures to make the prettier and easier to understand

@@
 
-In this work, we present the static and dynamic semantics of Legion \cite{Legion12},
+In this work, we present the static and dynamic semantics for Legion \cite{Legion12},
 a parallel programming model that supports multiple, dynamic data partitions and is
 able to efficiently reason about potential aliasing.
-Legion's {\em logical regions} are an abstraction capturing {\em locality} 
+Legion's {\em logical regions} are an abstraction that captures {\em locality} 
 (data that will be used together, and therefore should be colocated) 
@@
 
-To support additional parallelism in the presence of aliased data,
+To support additional parallelism when tasks access aliased regions,
 Legion provides {\em coherence} modes at the granularity of regions, 
-which allow relaxed constraints on task execution order, even
-in cases where tasks access the same region.  The programmer may
-specify that tasks conflicting on a region run in
+which allow relaxed constraints on task execution order.  The programmer may
+specify that tasks accessing aliased regions run in
 program order (the default), {\em atomically} with respect to each
 other, or even {\em simultaneously} if the application has alternative
-methods for ensuring determinism or can tolerate non-deterministic
+methods for managing non-deterministic
 behavior in that region.

### 2012-11-11 doc/types-pldi2013/langdef.tex
Lots of small changes and moving figures around
Fixing some typesetting issues for inline math modes

@@
 
-\begin{figure*}
-{
-\scalebox{0.9}{
-\begin{tikzpicture}[x=1in,y={(0,-1in)}]
-%\draw[thin] (0,0) to (7,0);
-%\draw[thin] (0,0) to (0,6);
-\foreach \y in {0,...,80}{
-%  \draw[thin] (0,0.1*\y) to (7,0.1*\y);
-}
-\foreach \y in {0,...,16}{
-%  \draw[red,thick] (0,0.5*\y) to (7,0.5*\y);
-}
-\foreach \y in {0,...,8}{
-%  \draw[blue,very thick] (0,\y) to (7,\y);
-}
-
-% column positions
-% \XA - left edge of type rules
-% \XB - center of type rule names
-% \XC - vertical line
-% \XD - left edge of semantics rules
-% \XE - center of semantics rule names
-
-\newcommand{\XA}{0.0}
-\newcommand{\XB}{2.8}
-\newcommand{\XC}{3.4}
-\newcommand{\XD}{3.65}
-\newcommand{\XE}{7.3}
-
-\draw[very thin] (\XC,0.1) -- (\XC,7.6);
-
-% start with read/write/reduce rules
-\ruleat{\XA,0.0}{\XB,0.2}{T-Read}
-{\begin{array}{l}
-\typeenvx e_1 : T@(\oton{r}) \\
-\forall i.\ \text{reads}(r_i) \in \Phi^*\end{array}}
-{\typeenvx \text{read}(e_1) : T}
-
-\ruleat{\XA,0.6}{\XB,0.8}{T-Write}
-{\begin{array}{l}
-\typeenvx e_1 : T@(\oton{r}) \\
-\typeenvx e_2 : T \\
-\forall i.\ \text{writes}(r_i) \in \Phi^*
-\end{array}}
-{\typeenvx \text{write}(e_1, e_2) : T@(\oton{r})}
-
-\ruleat{\XA,1.3}{\XB,1.5}{T-Reduce}
-{\begin{array}{l}
-\Gamma(id) = (T_1, T_2), \emptyset, \emptyset \rightarrow T_1 \\
-\typeenvx e_1 : T_1@(\oton{r}) \\
-\typeenvx e_2 : T_2 \\
-\forall i.\ \text{reduces}_{id}(r_i) \in \Phi^*
-\end{array}}
-{\typeenvx \text{reduce}(id, e_1, e_2) : T_1@(\oton{r})}
-
-\axiomat{\XA,2.2}{\XB,2.2}{T-New}
-{\typeenvx \text{new }T@r : T@r}
-
-\ruleat{\XA,2.5}{\XB,2.55}{T-UpRgn}
-{\begin{array}{l}
-\typeenvx e : T@(r'_1, \ldots r'_k) \\
-\forall i. \exists j, r'_i \leq r_j \in \Omega^* \\
-\end{array}}
-{\typeenvx \text{upregion}(e,\oton{r}) : T@(\oton{r})}
-
-\ruleat{\XA,3.1}{\XB,3.05}{T-DnRgn}
-{\begin{array}{l}
-\typeenvx e : T@(r'_1, \ldots r'_k) \\
-%\forall j. \exists i, r_j \leq r'_i \in \Omega^* \\
-\end{array}}
-{\typeenvx \text{downregion}(e,\oton{r}) : T@(\oton{r})}
-
-\axiomat{\XA,3.6}{\XB,3.6}{T-NewColor}
-{\typeenvx \text{newcolor }r : \text{coloring}(r)}
-
-\ruleat{\XA,3.9}{\XB,4.1}{T-Color}
-{\begin{array}{l}
-\typeenvx e_1 : \text{coloring}(r) \\
-\typeenvx e_2 : T@r \\
-\typeenvx e_3 : \text{int}
-\end{array}}
-{\typeenvx \text{color}(e_1, e_2, e_3) : \text{coloring}(r)}
-
-\ruleat{\XA,4.6}{\XB,4.55}{T-Partition}
-{\begin{array}{l}
-\typeenvx e_1 : \text{coloring}(r_p) \\
-\Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{1 \leq i < j \leq k} r_i * r_j \\
-\typeenvx[O=\Omega'] e_2 : T \\
-\{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T) = \emptyset
-\end{array}}
-{\typeenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 : T}
-
-\ruleat{0,5.5}{\XB,5.7}{T-Pack}
-{\begin{array}{l}
-T_1 = \exists r'_1, \ldots r'_n.\ T_2\text{ where }\Omega_1 \\
-\Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^* \\
-\typeenvx e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]
-\end{array}}
-{\typeenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] : T_1}
-
-\ruleat{\XA,6.2}{\XB,6.4}{T-Unpack}
-{\begin{array}{l}
-T_1 = \exists r'_1, \ldots, r'_n.\ T_2\text{ where }\Omega_1 \\
-\typeenvx e_1 : T_1 \\
-\Gamma' = \Gamma[T_2[r_1/r'_1,\ldots,r_k/r'_k] / id] \\
-\Omega' = \Omega \cup \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \\
-\typeenvx[G=\Gamma',O=\Omega'] e_2 : T_3 \\
-\{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T_1, T_3) = \emptyset
-\end{array}}
-{\typeenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 : T_3}
-
-\ruleat{\XA,7.4}{\XB,7.6}{T-Call}
-{\begin{array}{l}
-\Gamma(id) = \forall r'_1, \ldots r'_k.(\oton{T}),\Phi', Q' \rightarrow T_r \\
-\typeenvx e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\
-\Phi'[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*
-\end{array}}
-{\typeenvx id[\otok{k}{r}](\oton{e}) : T_r[r_1/r'_1,\ldots,r_k/r'_k]}
-
-% now the operational semantics versions
-
-\ruleatx{\XD,0.0}{\XE,0.2}{E-Read}{
-\opsenvx e \mapsto l, E \andalso S' = apply(S, E) \\
-v = \begin{cases} S'(l), & \text{if } l \not\in C \\
-v' : H(l), & \text{otherwise} \end{cases}
-}{\opsenvx \text{read}(e) \mapsto v, E \dplus [ read(l, excl, v) ]}
-
-\ruleatx{\XD,0.9}{\XE,0.85}{E-Write}{
-\opsenvx e_1 \mapsto l, E_1 \andalso S' = apply(S, E_1) \\
-\opsenvx[S=S'] e_2 \mapsto v, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)
-}{\opsenvx \text{write}(e_1, e_2) \mapsto l, E' \dplus [ write(l, excl, v) ]}
-
-\ruleatx{\XD,1.5}{\XE,1.45}{E-Reduce}{
-\opsenvx e_1 \mapsto l, E_1 \andalso S' = apply(S, E_1) \\
-\opsenvx[S=S'] e_2 \mapsto v, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)
-}{\opsenvx \text{reduce}(id, e_1, e_2) \mapsto l, E' \dplus [ reduce_{id}(l, excl, v) ]}
-
-\ruleatx{\XD,2.1}{\XE,2.2}{E-New}{
-l \in M(r) \andalso
-H(l) = M \llbracket T \rrbracket
-}{\opsenvx \text{new }T@r \mapsto l, []}
-
-\ruleat{\XD,2.6}{\XE,2.7}{E-UpRgn}{
-\opsenvx e \mapsto v, E
-}{\opsenvx \text{upregion}(e, \oton{r}) \mapsto v, E}
-
-\ruleatx{\XD,3.1}{\XE,3.3}{E-DnRgn}{
-\opsenvx e \mapsto l, E \\
-v = \begin{cases} l, & \text{if $\exists i, l \in M(r_i)$}. \\ null, & \text{otherwise}. \end{cases}
-}{\opsenvx \text{downregion}(e, \oton{r}) \mapsto v, E}
-
-\ruleatx{\XD,4.0}{\XE,4.2}{E-NewColor}{
-K = \{ (l_1, iv_1), \ldots, (l_p, iv_p) \}, \text{ where } \\
-(\forall i \in [1,p]. l_i \in M(r)) \wedge (\forall i,j \in [1,p]. l_i \not= l_j)
-}{\opsenvx \text{newcolor }r \mapsto K, []}
-
-\ruleatx{\XD,4.7}{\XE,4.9}{E-Color}{
-\opsenvx e_1 \mapsto K, E_1 \andalso S' = apply(S, E_1) \\
-\opsenvx[S=S'] e_2 \mapsto l, E_2 \andalso S'' = apply(S', E_2) \\
-\opsenvx[S=S''] e_3 \mapsto v, E_3 \\
-K' = \{ (l,v) \} \cup \{ (l_i,v_i) : (l_i,v_i) \in K \wedge l \not= l_i \} \\
-valid\_interleave(S, C, E', E_1, E_2, E_3)
-}{\opsenvx \text{color}(e_1, e_2, e_3) \mapsto K', E'}
-
-\ruleatx{\XD,5.7}{\XE,5.55}{E-Partition}{
-\opsenv{e_1}{K}{E_1} \andalso \rho_i = \{ l : (l, i) \in K \}, \text{ for } 1 \leq i \leq k \\
-M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \andalso S' = apply(S, E_1) \\
-\opsenvx[M=M',S=S'] e_2 \mapsto v, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)
-}{\opsenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 \mapsto v, E'}
-
-\ruleatx{\XD,6.5}{\XE,6.6}{E-Pack}{
-\opsenvx e_1 \mapsto v, E \andalso \rho_i = M[r_i], \text{ for } 1 \leq i \leq k \\
-v' = \langle \langle \otok{k}{\rho}, v \rangle \rangle
-}{\opsenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] \mapsto v', E}
-
-\ruleatx{\XD,7.1}{\XE,6.95}{E-Unpack}{
-\opsenvx e_1 \mapsto \langle \langle \otok{k}{\rho} , v_1 \rangle \rangle, E_1 \andalso M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \\
-L' = L[v_1/id] \andalso S' = apply(S, E_1) \\
-\opsenvx[M=M',L=L',S=S'] e_2 \mapsto v_2, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)
-}{\opsenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 \mapsto v_2, E'}
-
-\ruleatx{\XA,8.1}{\XE,8.4}{E-Call}{
-\opsenvx e_1 \mapsto v_1, E_1 \andalso S_1 = apply(S, E_1) \andalso \opsenvx e_2 \mapsto v_2, E_2 \andalso \ldots \andalso S_n = apply(S_{n-1}, E_n) \\
-valid\_interleave(S, C, E', \oton{E}) \andalso \text{function }id[\otok{k}{r'}](a_1 : T_1, \ldots, a_n : T_n), \Phi', Q' : T_r = e_{n+1} \\
-M' = \{ (r'_1, M(r_1)), \ldots (r'_k, M(r_k)) \} \andalso L' = \{ (a_1, v_1), \ldots, (a_n, v_n) \} \andalso S' = apply(S, E') \\
-C' = C \cup \{ l : \exists \rho. atomic(\rho) \in M' \llbracket Q' \rrbracket \vee simult(\rho) \in M' \llbracket Q' \rrbracket \} \\
-\opsenvx[M=M',L=L',S=S',C=C'] e_{n+1} \mapsto v_{n+1}, E_{n+1} \andalso E'_{n+1} = mark\_coherence(E_{n+1}, M' \llbracket Q' \rrbracket) \andalso valid\_interleave(S, C, E'', E', E_{n+1})
-}{\opsenvx id[\otok{k}{r}](\oton{e}) \mapsto v_{n+1}, E''}
-
-% try to fit helper rules at the bottom instead of in a separate figure
-
-\draw (0.05,9.12) -- (7.6,9.12);
-
-\node[below right] at (\XA,9.15) {
-$
-\begin{array}{@{}l@{}}
-\begin{array}{@{}l@{}l@{}} 
-\begin{array}{@{}lll}
-apply(S, []) & = & S \\
-apply(S, E \dplus [ read(l, c, v) ]) & = & apply(S, E) \\
-apply(S, E \dplus [ write(l, c, v) ]) & = & apply(S, E)[v/l] \\
-apply(S, E \dplus [ reduce_{id}(l, c, v) ]) & = & S'[id(S'(l), v)/l], \\
-& & \text{ where } S' = apply(S, E)
-\end{array} &
-\begin{array}{@{}ll}
-mark\_coherence([], \hat Q) & = [] \\
-mark\_coherence([ op(l, c, v) ] \dplus E, \hat Q) & = [ op(l, c', v) ] \dplus mark\_coherence(E, \hat Q), \\
-\multicolumn{2}{@{}l}{ \hspace{1.0in} \text{ where } c' =
-\begin{cases}
-simult, & \text{if }\exists \rho. l \in \rho \wedge \text{simult}(\rho) \in \hat Q \\
-atomic, & \text{if }\exists \rho. l \in \rho \wedge \text{atomic}(\rho) \in \hat Q \\
-excl, & \text{otherwise}
-\end{cases}}
-\end{array}
-\end{array}
-\end{array}
-$};
-
-\end{tikzpicture}
-}
-\vspace*{-0.5cm}
-\texcomment{
-\framebox[7in]{
-\placeat{2}{3}{bar}%
-\placeat{3}{1}{3,1}%
-\placeat{-3}{1}{-3,1} %
-\makebox[0pt]{%
-\hspace*{2in} \raisebox{3in}{---}%
-} %
-\makebox[0pt]{%
-\hspace*{1in} \raisebox{1in}{foo}%
-} %
-\makebox[0pt]{%
-\hspace*{2in} \raisebox{3in}{bar}%
-}%
-\rule{0.1pt}{6in}
-}}
-}
-\caption{Type System and Operational Semantics}
-\label{fig:langrules}
-\end{figure*}
-
 %\newcommand{\infrule}[2]{\displaystyle\frac{\displaystyle\strut{#1}}{\displaystyle\strut {#2}}}
@@
 }
+
+\begin{figure*}
+{
+\scalebox{0.9}{
+\begin{tikzpicture}[x=1in,y={(0,-1in)}]
+%\draw[thin] (0,0) to (7,0);
+%\draw[thin] (0,0) to (0,6);
+\foreach \y in {0,...,80}{
+%  \draw[thin] (0,0.1*\y) to (7,0.1*\y);
+}
+\foreach \y in {0,...,16}{
+%  \draw[red,thick] (0,0.5*\y) to (7,0.5*\y);
+}
+\foreach \y in {0,...,8}{
+%  \draw[blue,very thick] (0,\y) to (7,\y);
+}
+
+% column positions
+% \XA - left edge of type rules
+% \XB - center of type rule names
+% \XC - vertical line
+% \XD - left edge of semantics rules
+% \XE - center of semantics rule names
+
+\newcommand{\XA}{0.0}
+\newcommand{\XB}{2.8}
+\newcommand{\XC}{3.4}
+\newcommand{\XD}{3.65}
+\newcommand{\XE}{7.3}
+
+\draw[very thin] (\XC,0.1) -- (\XC,7.6);
+
+% start with read/write/reduce rules
+\ruleat{\XA,0.0}{\XB,0.2}{T-Read}
+{\begin{array}{l}
+\typeenvx e_1 : T@(\oton{r}) \\
+\forall i.\ \text{reads}(r_i) \in \Phi^*\end{array}}
+{\typeenvx \text{read}(e_1) : T}
+
+\ruleat{\XA,0.6}{\XB,0.8}{T-Write}
+{\begin{array}{l}
+\typeenvx e_1 : T@(\oton{r}) \\
+\typeenvx e_2 : T \\
+\forall i.\ \text{writes}(r_i) \in \Phi^*
+\end{array}}
+{\typeenvx \text{write}(e_1, e_2) : T@(\oton{r})}
+
+\ruleat{\XA,1.3}{\XB,1.5}{T-Reduce}
+{\begin{array}{l}
+\Gamma(id) = (T_1, T_2), \emptyset, \emptyset \rightarrow T_1 \\
+\typeenvx e_1 : T_1@(\oton{r}) \\
+\typeenvx e_2 : T_2 \\
+\forall i.\ \text{reduces}_{id}(r_i) \in \Phi^*
+\end{array}}
+{\typeenvx \text{reduce}(id, e_1, e_2) : T_1@(\oton{r})}
+
+\axiomat{\XA,2.2}{\XB,2.2}{T-New}
+{\typeenvx \text{new }T@r : T@r}
+
+\ruleat{\XA,2.5}{\XB,2.55}{T-UpRgn}
+{\begin{array}{l}
+\typeenvx e : T@(r'_1, \ldots r'_k) \\
+\forall i. \exists j, r'_i \leq r_j \in \Omega^* \\
+\end{array}}
+{\typeenvx \text{upregion}(e,\oton{r}) : T@(\oton{r})}
+
+\ruleat{\XA,3.1}{\XB,3.05}{T-DnRgn}
+{\begin{array}{l}
+\typeenvx e : T@(r'_1, \ldots r'_k) \\
+%\forall j. \exists i, r_j \leq r'_i \in \Omega^* \\
+\end{array}}
+{\typeenvx \text{downregion}(e,\oton{r}) : T@(\oton{r})}
+
+\axiomat{\XA,3.6}{\XB,3.6}{T-NewColor}
+{\typeenvx \text{newcolor }r : \text{coloring}(r)}
+
+\ruleat{\XA,3.9}{\XB,4.1}{T-Color}
+{\begin{array}{l}
+\typeenvx e_1 : \text{coloring}(r) \\
+\typeenvx e_2 : T@r \\
+\typeenvx e_3 : \text{int}
+\end{array}}
+{\typeenvx \text{color}(e_1, e_2, e_3) : \text{coloring}(r)}
+
+\ruleat{\XA,4.6}{\XB,4.55}{T-Partition}
+{\begin{array}{l}
+\typeenvx e_1 : \text{coloring}(r_p) \\
+\Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{1 \leq i < j \leq k} r_i * r_j \\
+\typeenvx[O=\Omega'] e_2 : T \\
+\{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T) = \emptyset
+\end{array}}
+{\typeenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 : T}
+
+\ruleat{0,5.5}{\XB,5.7}{T-Pack}
+{\begin{array}{l}
+T_1 = \exists r'_1, \ldots r'_n.\ T_2\text{ where }\Omega_1 \\
+\Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^* \\
+\typeenvx e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]
+\end{array}}
+{\typeenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] : T_1}
+
+\ruleat{\XA,6.2}{\XB,6.4}{T-Unpack}
+{\begin{array}{l}
+T_1 = \exists r'_1, \ldots, r'_n.\ T_2\text{ where }\Omega_1 \\
+\typeenvx e_1 : T_1 \\
+\Gamma' = \Gamma[T_2[r_1/r'_1,\ldots,r_k/r'_k] / id] \\
+\Omega' = \Omega \cup \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \\
+\typeenvx[G=\Gamma',O=\Omega'] e_2 : T_3 \\
+\{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T_1, T_3) = \emptyset
+\end{array}}
+{\typeenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 : T_3}
+
+\ruleat{\XA,7.4}{\XB,7.6}{T-Call}
+{\begin{array}{l}
+\Gamma(id) = \forall r'_1, \ldots r'_k.(\oton{T}),\Phi', Q' \rightarrow T_r \\
+\typeenvx e_i : T_i[r_1/r'_1,\ldots,r_k/r'_k] \\
+\Phi'[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*
+\end{array}}
+{\typeenvx id[\otok{k}{r}](\oton{e}) : T_r[r_1/r'_1,\ldots,r_k/r'_k]}
+
+% now the operational semantics versions
+
+\ruleatx{\XD,0.0}{\XE,0.2}{E-Read}{
+\opsenvx e \mapsto l, E \andalso S' = apply(S, E) \\
+v = \begin{cases} S'(l), & \text{if } l \not\in C \\
+v' : H(l), & \text{otherwise} \end{cases}
+}{\opsenvx \text{read}(e) \mapsto v, E \dplus [ read(l, excl, v) ]}
+
+\ruleatx{\XD,0.9}{\XE,0.85}{E-Write}{
+\opsenvx e_1 \mapsto l, E_1 \andalso S' = apply(S, E_1) \\
+\opsenvx[S=S'] e_2 \mapsto v, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)
+}{\opsenvx \text{write}(e_1, e_2) \mapsto l, E' \dplus [ write(l, excl, v) ]}
+
+\ruleatx{\XD,1.5}{\XE,1.45}{E-Reduce}{
+\opsenvx e_1 \mapsto l, E_1 \andalso S' = apply(S, E_1) \\
+\opsenvx[S=S'] e_2 \mapsto v, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)
+}{\opsenvx \text{reduce}(id, e_1, e_2) \mapsto l, E' \dplus [ reduce_{id}(l, excl, v) ]}
+
+\ruleatx{\XD,2.1}{\XE,2.2}{E-New}{
+l \in M(r) \andalso
+H(l) = M \llbracket T \rrbracket
+}{\opsenvx \text{new }T@r \mapsto l, []}
+
+\ruleat{\XD,2.6}{\XE,2.7}{E-UpRgn}{
+\opsenvx e \mapsto v, E
+}{\opsenvx \text{upregion}(e, \oton{r}) \mapsto v, E}
+
+\ruleatx{\XD,3.1}{\XE,3.3}{E-DnRgn}{
+\opsenvx e \mapsto l, E \\
+v = \begin{cases} l, & \text{if $\exists i, l \in M(r_i)$}. \\ null, & \text{otherwise}. \end{cases}
+}{\opsenvx \text{downregion}(e, \oton{r}) \mapsto v, E}
+
+\ruleatx{\XD,4.0}{\XE,4.2}{E-NewColor}{
+K = \{ (l_1, iv_1), \ldots, (l_p, iv_p) \}, \text{ where } \\
+(\forall i \in [1,p]. l_i \in M(r)) \wedge (\forall i,j \in [1,p]. l_i \not= l_j)
+}{\opsenvx \text{newcolor }r \mapsto K, []}
+
+\ruleatx{\XD,4.7}{\XE,4.9}{E-Color}{
+\opsenvx e_1 \mapsto K, E_1 \andalso S' = apply(S, E_1) \\
+\opsenvx[S=S'] e_2 \mapsto l, E_2 \andalso S'' = apply(S', E_2) \\
+\opsenvx[S=S''] e_3 \mapsto v, E_3 \\
+K' = \{ (l,v) \} \cup \{ (l_i,v_i) : (l_i,v_i) \in K \wedge l \not= l_i \} \\
+valid\_interleave(S, C, E', E_1, E_2, E_3)
+}{\opsenvx \text{color}(e_1, e_2, e_3) \mapsto K', E'}
+
+\ruleatx{\XD,5.7}{\XE,5.55}{E-Partition}{
+\opsenv{e_1}{K}{E_1} \andalso \rho_i = \{ l : (l, i) \in K \}, \text{ for } 1 \leq i \leq k \\
+M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \andalso S' = apply(S, E_1) \\
+\opsenvx[M=M',S=S'] e_2 \mapsto v, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)
+}{\opsenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 \mapsto v, E'}
+
+\ruleatx{\XD,6.5}{\XE,6.6}{E-Pack}{
+\opsenvx e_1 \mapsto v, E \andalso \rho_i = M[r_i], \text{ for } 1 \leq i \leq k \\
+v' = \langle \langle \otok{k}{\rho}, v \rangle \rangle
+}{\opsenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] \mapsto v', E}
+
+\ruleatx{\XD,7.1}{\XE,6.95}{E-Unpack}{
+\opsenvx e_1 \mapsto \langle \langle \otok{k}{\rho} , v_1 \rangle \rangle, E_1 \andalso M' = M[\rho_1/r_1, \ldots, \rho_k/r_k] \\
+L' = L[v_1/id] \andalso S' = apply(S, E_1) \\
+\opsenvx[M=M',L=L',S=S'] e_2 \mapsto v_2, E_2 \andalso valid\_interleave(S, C, E', E_1, E_2)
+}{\opsenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 \mapsto v_2, E'}
+
+\ruleatx{\XA,8.1}{\XE,8.4}{E-Call}{
+\opsenvx e_1 \mapsto v_1, E_1 \andalso S_1 = apply(S, E_1) \andalso \opsenvx e_2 \mapsto v_2, E_2 \andalso \ldots \andalso S_n = apply(S_{n-1}, E_n) \\
+valid\_interleave(S, C, E', \oton{E}) \andalso \text{function }id[\otok{k}{r'}](a_1 : T_1, \ldots, a_n : T_n), \Phi', Q' : T_r = e_{n+1} \\
+M' = \{ (r'_1, M(r_1)), \ldots (r'_k, M(r_k)) \} \andalso L' = \{ (a_1, v_1), \ldots, (a_n, v_n) \} \andalso S' = apply(S, E') \\
+C' = C \cup \{ l : \exists \rho. atomic(\rho) \in M' \llbracket Q' \rrbracket \vee simult(\rho) \in M' \llbracket Q' \rrbracket \} \\
+\opsenvx[M=M',L=L',S=S',C=C'] e_{n+1} \mapsto v_{n+1}, E_{n+1} \andalso E'_{n+1} = mark\_coherence(E_{n+1}, M' \llbracket Q' \rrbracket) \andalso valid\_interleave(S, C, E'', E', E_{n+1})
+}{\opsenvx id[\otok{k}{r}](\oton{e}) \mapsto v_{n+1}, E''}
+
+% try to fit helper rules at the bottom instead of in a separate figure
+
+\draw (0.05,9.12) -- (7.6,9.12);
+
+\node[below right] at (\XA,9.15) {
+$
+\begin{array}{@{}l@{}}
+\begin{array}{@{}l@{}l@{}} 
+\begin{array}{@{}lll}
+apply(S, []) & = & S \\
+apply(S, E \dplus [ read(l, c, v) ]) & = & apply(S, E) \\
+apply(S, E \dplus [ write(l, c, v) ]) & = & apply(S, E)[v/l] \\
+apply(S, E \dplus [ reduce_{id}(l, c, v) ]) & = & S'[id(S'(l), v)/l], \\
+& & \text{ where } S' = apply(S, E)
+\end{array} &
+\begin{array}{@{}ll}
+mark\_coherence([], \hat Q) & = [] \\
+mark\_coherence([ op(l, c, v) ] \dplus E, \hat Q) & = [ op(l, c', v) ] \dplus mark\_coherence(E, \hat Q), \\
+\multicolumn{2}{@{}l}{ \hspace{1.0in} \text{ where } c' =
+\begin{cases}
+simult, & \text{if }\exists \rho. l \in \rho \wedge \text{simult}(\rho) \in \hat Q \\
+atomic, & \text{if }\exists \rho. l \in \rho \wedge \text{atomic}(\rho) \in \hat Q \\
+excl, & \text{otherwise}
+\end{cases}}
+\end{array}
+\end{array}
+\end{array}
+$};
+
+\end{tikzpicture}
+}
+\vspace*{-0.5cm}
+\texcomment{
+\framebox[7in]{
+\placeat{2}{3}{bar}%
+\placeat{3}{1}{3,1}%
+\placeat{-3}{1}{-3,1} %
+\makebox[0pt]{%
+\hspace*{2in} \raisebox{3in}{---}%
+} %
+\makebox[0pt]{%
+\hspace*{1in} \raisebox{1in}{foo}%
+} %
+\makebox[0pt]{%
+\hspace*{2in} \raisebox{3in}{bar}%
+}%
+\rule{0.1pt}{6in}
+}}
+}
+\caption{Type System and Operational Semantics}
+\label{fig:langrules}
+\end{figure*}
 
@@
 $\epsilon_1, \epsilon_2$ be two memory operations from $E_i$ and $E_j$ ($i \not= j$).  Then,
-$$\begin{array}{l@{}l}
-valid\_interleave&(S, C, E'_a \dplus [ \epsilon_1, \epsilon_2 ] \dplus E'_b, \oton{E}) \wedge \epsilon_1 \nonint \epsilon_2 \Rightarrow \\
- & valid\_interleave(S, C, E'_a \dplus [ \epsilon_2, \epsilon_1 ] \dplus E'_b, \oton{E}).
-\end{array}$$
+%%$$\begin{array}{l}
+\begin{align*}
+valid&\_interleave(S, C, E'_a \dplus [ \epsilon_1, \epsilon_2 ] \dplus E'_b, \oton{E}) \wedge \epsilon_1 \nonint \epsilon_2 \\
+&\Rightarrow valid\_interleave(S, C, E'_a \dplus [ \epsilon_2, \epsilon_1 ] \dplus E'_b, \oton{E}).
+\end{align*}
+%%\end{array}$$
 \end{lem}
@@
 To show this is safe, we define a relaxed version of non-interference for atomic coherence:
-$$
-\begin{array}{l@{}l}
-op_1(l_1, c_1, v_1)\ \nonint^A\ op_2(l_2, c_2, v_2) \Leftrightarrow &op_1(l_1, c_1, v_1)\ \nonint\ op_2(l_2, c_2, v_2) \vee \\
-& (c_1 = atomic \wedge c_2 = atomic)
-\end{array}$$
+%$$
+%\begin{array}{l@{}l}
+\begin{align*}
+op_1 &(l_1, c_1, v_1)\ \nonint^A\ op_2(l_2, c_2, v_2) \Leftrightarrow  \\
+& op_1(l_1, c_1, v_1)\ \nonint\ op_2(l_2, c_2, v_2) \vee (c_1 = atomic \wedge c_2 = atomic)
+\end{align*}
+%\end{array}
+%$$
 

### 2012-11-11 doc/types-pldi2013/example_full.tex
added appendix - did first pass of cleaning up label references

--- 2012-11-08 doc/types-pldi2013/example.tex
@@
-\lstset{
-  captionpos=b,
-  language=Haskell,
-  basicstyle=\scriptsize,
-  numbers=left,
-  numberstyle=\tiny,
-  columns=fullflexible,
-  stepnumber=1,
-  escapechar=\#,
-  keepspaces=true,
-  belowskip=-10pt,
-  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
-  morekeywords={function,rr,int,float,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,multicoloring,color,newcolor,atomic,simultaneous},
-  deletekeywords={float,head,min,max}
-}
-\begin{lstlisting}[float={t},label={lst:circuit_ex},caption={Circuit Simulation}]
+\section{Circuit Example}
+\label{sec:examplefull}
+
+Listing~\ref{lst:circuit_ex_full} shows a circuit simulation written (with
+one exception) in the core Legion language (see
+Section~\ref{sec:legioncorefull}).  The input is a graph of circuit
+elements and connecting wires.  To perform parallel
+simulation the graph is partitioned into pieces, after which the
+simulation is run for a number of time steps.  Each time step has
+three phases: calculation of each wire's current, distribution of charge to
+the nodes connected to each wire, and updating of node voltages.
+
+The entry function is {\tt simulate\_circuit} (line 15), which is
+parameterized on regions {\tt rl}, {\tt rw}, and {\tt rn}.
+The wires and nodes are contained in regions {\tt rw} and {\tt rn},
+respectively.  The {\tt rl} region holds lists of pointers to the nodes and wires.
+When {\tt simulate\_circuit} is invoked it must have both read and
+write privileges for the regions {\tt rl}, {\tt rw}, and {\tt rn}
+(line 17); the type system enforces that
+{\tt simulate\_circuit} only accesses these regions and their
+subregions at compile-time.
+
+Partitioning of the graph is done using {\em colorings} built in the
+{\tt color\_circuit} function (line 18-19).  A coloring maps region
+elements to colors (i.e. integers), with each color corresponding
+ to a new subregion
+to be created.  Note that it is impossible to pick a single static
+partitioning scheme that would do well for all graphs.  Dynamic
+partitioning using colorings provides the programmer the flexibility
+to make partitioning decisions based on input data.
+
+The {\tt color\_circuit} function returns three colorings.  The first
+maps each node to the graph piece that owns it.
+The third maps each wire to a graph piece that
+owns one of its nodes.  The second coloring captures, for each piece $p$,
+the nodes in other pieces on the boundary of $p$, often referred to as
+{\em ghost nodes}.  
+%We describe how ghost node regions are used in
+%conjunction with the owned nodes shortly.  
+Note that 
+ghost nodes may be on the boundary of multiple pieces and therefore colored
+multiple ways.  The second coloring is a {\em multicoloring} because
+it may map the same element to multiple colors.
+
+After creating the colorings, the application partitions the node and wire
+regions into {\em subregions} according to the colorings (lines 21-25).  For
+simplicity, our example only partitions the circuit into two pieces.
+Line 21 uses coloring {\tt pc.1} to partition {\tt rn} into {\tt rn0}, and {\tt rn1}.
+Subregion {\tt rn0} will contain all locations with a color of 1 in {\tt pc.1},
+while {\tt rn1} will contain all locations with a color of 2.  This
+partitioning uses a coloring and results in {\em disjoint} subpartitions.
+Constraints are introduced into the static environment describing both the
+disjointness of the subregions (i.e. {\tt rn0 $*$ rn1}) and the inclusion of
+each subregion in the original region (e.g. {\tt rn0 $\leq$ rn}).  Line
+23 is similar, but uses a multicoloring, so the subregions may be {\em aliased},
+and only the inclusion constraints are introduced into the static environment.
+Although multicolorings are not part of the Core Legion described in 
+Section~\ref{sec:legioncorefull}, they can be easily implemented using a separate 
+coloring for each color in the multicoloring.
+
+The circuit simulation uses an {\em allocate-then-partition} style of
+computation, where a large data structure is first allocated and then
+partitioned.  Legion also supports a {\em partition-then-allocate} style,
+where empty regions are first partitioned and later populated with
+data.  Both are useful; e.g., Sequoia \cite{Fatahalian06} and DPJ
+\cite{Bocchino09} support static forms of allocate-then-partition and
+partition-then-allocate, respectively.  Legion supports both
+approaches dynamically.
+
+After partitioning the circuit into pieces, the application creates
+instances of {\tt CircuitPiece} (defined on lines 9-12) for each piece
+(lines 28-33).  {\tt CircuitPiece} is a {\em region relationship}, a
+bounded existential type.  Region relationships allow the programmer
+to {\em pack} a group of regions together and remember properties
+about them such as disjointness and subregion relationships (lines 29
+and 33).  The type system verifies the properties hold statically when
+packing; when region relationships are {\em unpacked} the same
+properties are reintroduced into the static checking environment
+(lines 40-41).  A key feature underlying the soundness of the Legion
+type system is that privileges cannot be packed in a region
+relationship---privileges belong to functions. When a function unpacks
+a region relationship it must already hold privileges for the new
+regions it finds in the packed value.  For example, on line 40
+function {\tt execute\_time\_steps} has read/write privileges for
+region {\tt rn0} because it has read/write privileges for {\tt rn} and
+the {\tt CircuitPiece} region relationship constraints ensure $\tt rn0
+\leq rn$ (line 12).
+
+\begin{lstlisting}[float={t},label={lst:circuit_ex_full},caption={Circuit Simulation}]
 --                        <voltage,current,charge,capacitance>
@@
 
-\section{Circuit Example}
-\label{sec:example}
-
-Listing~\ref{lst:circuit_ex} shows a circuit simulation written
-in the core Legion language (see
-Section~\ref{sec:legioncore}).  The input is a graph of circuit
-elements and connecting wires.  To perform parallel
-simulation the graph is partitioned into pieces, after which the
-simulation is run for a number of time steps.  Each time step has
-three phases: calculation of each wire's current, distribution of charge to
-the nodes connected to each wire, and updating of node voltages.
-
-The entry function is {\tt simulate\_circuit} (line 15), which accepts a 
-linked list of nodes contained in region {\tt rn} and a linked list of 
-wires contained in region {\tt rw}.  The linked list elements themselves
-are in region {\tt rl}.
-When {\tt simulate\_circuit} is invoked it must have both read and
-write privileges for the regions {\tt rl}, {\tt rw}, and {\tt rn}
-(line 17); the type system statically verifies that
-{\tt simulate\_circuit} only accesses these regions and their subregions.
-
-Partitioning of the graph is done using {\em colorings} built in the
-{\tt color\_circuit} function (line 18-19).  A coloring maps region
-elements to colors (integers in Core Legion), with each color corresponding
- to a new subregion
-to be created.  Note that it is impossible to pick a single static
-partitioning scheme that would do well for all graphs.  Dynamic
-partitioning using colorings enables the programmer 
-to make partitioning decisions based on input data.
-
-The {\tt color\_circuit} function is responsible for specifying how the
-circuit is to be partitioned.  The body of this function is not shown as the
-ideal algorithm for partitioning data will be application specific.  The 
-only restriction imposed by the Legion programming model is that the result of
-the algorithm be provided in the form of colorings.  For our circuit example,
-there are three colorings.  The first maps each node to the graph piece that owns it.
-The third maps each wire to a graph piece that
-owns one of its nodes.  The second coloring captures, for each piece $p$,
-the nodes in other pieces on the boundary of $p$, often referred to as
-{\em ghost nodes}.  
-%We describe how ghost node regions are used in
-%conjunction with the owned nodes shortly.  
-Note that 
-ghost nodes may be on the boundary of multiple pieces and therefore be included in
-multiple ghost regions.  For conciseness, a bit of syntactic sugar is used here the
-form of a {\em multicoloring}, which allows the mapping of an element to multiple 
-colors.  Although multicolorings are not part of the Core Legion described in 
-Section~\ref{sec:legioncore}, they can be easily implemented using a separate 
-coloring for each color in the multicoloring.
-
-After creating the colorings, the application partitions the node and wire
-regions into {\em subregions} according to the colorings (lines 21-25).  For
-simplicity, our example only partitions the circuit into two pieces.
-Line 21 uses coloring {\tt pc.1} to partition {\tt rn} into {\tt rn0}, and {\tt rn1}.
-Subregion {\tt rn0} will contain all locations with a color of 0 in {\tt pc.1},
-while {\tt rn1} will contain all locations with a color of 1.  This
-partitioning results in {\em disjoint} subpartitions.
-Constraints are introduced into the static environment describing both the
-disjointness of the subregions (i.e. {\tt rn0 $*$ rn1}) and the inclusion of
-each subregion in the original region (e.g. {\tt rn0 $\leq$ rn}).  Line
-23 is similar, but uses a multicoloring, so the subregions may be {\em aliased},
-and only the inclusion constraints are introduced into the static environment.
-
-The circuit simulation uses an {\em allocate-then-partition} style of
-computation, where a large data structure is first allocated and then
-partitioned.  Legion also supports a {\em partition-then-allocate} style,
-where empty regions are first partitioned and later populated with
-data.  Both are useful; e.g., Sequoia \cite{Fatahalian06} and DPJ
-\cite{Bocchino09} support static forms of allocate-then-partition and
-partition-then-allocate, respectively.  Legion supports both
-approaches dynamically.
-
-After partitioning the circuit into pieces, the application creates
-instances of {\tt CircuitPiece} (defined on lines 9-12) for each piece
-(lines 28-33).  {\tt CircuitPiece} is a {\em region relationship}, a
-bounded existential type.  Region relationships allow the programmer
-to {\em pack} a group of regions and pointers into those regions together
-and remember properties
-about them such as disjointness and subregion relationships (lines 29
-and 33).  The type system verifies the properties hold statically when
-packing; when region relationships are {\em unpacked} the same
-properties are reintroduced into the static checking environment
-(lines 40-41).  A key feature underlying the soundness of the Legion
-type system is that privileges cannot be packed in a region
-relationship---privileges belong to functions. When a function unpacks
-a region relationship it must already hold privileges for the unpacked regions. 
-%new regions it finds in the packed value.  
-For example, on line 40 the
-function {\tt execute\_time\_steps} has read/write privileges for
-region {\tt rn0} because it has read/write privileges for {\tt rn} and
-the {\tt CircuitPiece} region relationship constraints ensure $\tt rn0
-\leq rn$ (line 12).
-
 The {\tt execute\_time\_steps} function (lines 37-48)  
@@
 of much more complicated code.  Instead, an inexpensive dynamic disjointness
-check by the Legion runtime will (re)discover the parallelism.
+check by the Legion runtime will discover the parallelism.
 
-Listing~\ref{lst:circuit_leaf} shows the leaf functions for each phase.
+Listing~\ref{lst:circuit_leaf_full} shows the leaf functions for each phase.
 Each function iterates over the list of wires or
@@
 on regions.  Coherence specifies what updates the function may
-observe from other functions using aliased regions.  
-If not otherwise specified, coherence defaults to {\em exclusive}, 
-meaning the function must appear to execute in program order relative to
-other function calls.  
-Line 14 in Listing~\ref{lst:circuit_leaf} shows
+see from other functions using the same region.  If not otherwise
+specified, coherence defaults to {\em exclusive}, meaning the
+function must appear to execute in program order.  Line 14 shows
 an example of {\tt atomic}, a relaxed coherence mode requiring
 that operations to {\tt rn} and {\tt rg} appear atomic relative
-to other functions using regions which may alias.  The most relaxed coherence
+to other functions using those regions.  The most relaxed coherence
 mode is {\tt simult}; simultaneous coherence allows concurrent access 
 to the region by all functions that are using the region
-in a simultaneous mode.  The interaction between {\em tasks} (which are functions chosen to execute in parallel) using the same
+in a simultaneous mode.  The interaction between {\em tasks} (which are functions chosen by the runtime to execute in parallel) using the same
 region with different coherence
-modes is formalized in Section~\ref{sec:coherence}.
+modes is formalized in Section~\ref{sec:coherencefull}.
 
@@
   keepspaces=true,
-  belowskip=-10pt,
   literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
@@
 
-\begin{lstlisting}[float={t},label={lst:circuit_leaf},caption={Circuit Leaf Functions}]
+\begin{lstlisting}[float={t},label={lst:circuit_leaf_full},caption={Circuit Leaf Functions}]
 function calc_new_currents[rl,rw,rn,rg] ( ptr_list : wire_list<rl,rw,rn,rg>@rl ), 

### 2012-11-11 doc/types-pldi2013/langdef_full.tex
added appendix - did first pass of cleaning up label references

--- 2012-07-20 doc/popl2013full/langdef.tex
@@
+\setcounter{thm}{0}
+\setcounter{lem}{0}
+\texcomment{
 \newtheorem{thm}{Theorem}
@@
 \makeatother
-
+}
+
+\texcomment{
 \begin{figure*}
@@
 \caption{Core Legion}
-\label{fig:langdef}
+\label{fig:langdeffull}
 \vspace{-5mm}
 \end{figure*}
+}
 
@@
 
-\section{Legion}
-\label{sec:legioncore}
+\section{Core Legion}
+\label{sec:legioncorefull}
 
@@
 \subsection{Core Type System}
-\label{subsec:coretypes}
+\label{subsec:coretypesfull}
 
@@
 operations for $\Phi^*$ and $\Omega^*$ that explicitly add these implied constraints.
-The type checking rules are in Figure~\ref{fig:types}.  Many of the rules are 
+The type checking rules are in Figure~\ref{fig:typesfull}.  Many of the rules are 
 straightforward, but several deserve more explanation.
@@
 \caption{Legion Core Type System}
-\label{fig:types}
+\label{fig:typesfull}
 \end{figure*}
@@
 \vspace{-5mm}
-\label{fig:types}
+\label{fig:typesfull}
 \end{figure*}
 
+\texcomment{
 \begin{figure}
@@
 \end{figure}
+}
 
 \subsection{Operational Semantics}
-\label{subsec:opsemantics}
+\label{subsec:opsemanticsfull}
 
@@
 \end{itemize}
-Figure~\ref{fig:opsemfns} defines {\em apply} and {\em mark\_coherence}.  The precise definition of
-{\em valid\_interleave} is given in Section~\ref{sec:coherence}.  For now, we need only that a valid interleaving
+Figure~\ref{fig:langdef} defines {\em apply} and {\em mark\_coherence}.  The precise definition of
+{\em valid\_interleave} is given in Section~\ref{sec:coherencefull}.  For now, we need only that a valid interleaving
 is at least an interleaving, containing each constituent trace as a subsequence and no other operations.
@@
 \caption{Operational Semantics Helper Functions}
-\label{fig:opsemfns}
+\label{fig:opsemfnsfull}
 \vspace{-5mm}
 \end{figure*}
+}
 
@@
 \caption{Legion Core Operational Semantics}
-\label{fig:semantics}
+\label{fig:semanticsfull}
 \end{figure*}
@@
 \caption{Legion Core Operational Semantics}
-\label{fig:semantics}
+\label{fig:semanticfull}
 \end{figure*}
@@
 \section{Soundness of Privileges}
-\label{sec:soundness}
+\label{sec:soundnessfull}
 
@@
 \caption{Consistency Properties}
-\label{fig:constprop}
+\label{fig:constpropfull}
 \vspace{-5mm}
 \end{figure*}
+}
 
 \begin{thm}
-\label{thm:effects}
+\label{thm:effectsfull}
 \rm
@@
 \section{Coherence}
-\label{sec:coherence}
-
-Theorem~\ref{thm:effects} guarantees that a function call produces a value of the
+\label{sec:coherencefull}
+
+Theorem~\ref{thm:effectsfull} guarantees that a function call produces a value of the
 correct type and does so within the regions of memory for which privileges are held.
@@
 \caption{Valid Interleaving Test}
-\label{fig:validinterleave}
+\label{fig:validinterleavefull}
 \vspace{-5mm}
 \end{figure*}
+}
 
@@
 \begin{lem}
-\label{lem:clobberreduce}
+\label{lem:clobberreducefull}
 \rm 
@@
 \item $l \in C, l \not\in C'$, for which $v = S'(l)$ must be chosen for the derivation using $C$.
-From Theorem~\ref{thm:effects} and Lemma~\ref{lemma:heapconst:apply}, we have $\traceconsist{E}$ and $\storeconsist{S' = apply(S, E)}{H}$, so $\valueconsist[]{S'(l)}{H(l)}$, and $v$ is a legal choice for $v'$ in the derivation using $C'$.
+From Theorem~\ref{thm:effectsfull} and Lemma~\ref{lemma:heapconst:apply}, we have $\traceconsist{E}$ and $\storeconsist{S' = apply(S, E)}{H}$, so $\valueconsist[]{S'(l)}{H(l)}$, and $v$ is a legal choice for $v'$ in the derivation using $C'$.
 \end{itemize}
@@
 \begin{thm}
-\label{thm:sequential}
+\label{thm:sequentialfull}
 \rm
@@
 \begin{lem}
-\label{lem:nonintswap}
+\label{lem:nonintswapfull}
 \rm
@@
 \begin{lem}
-\label{lem:noninteffects}
+\label{lem:noninteffectsfull}
 \rm
@@
 \begin{proof}
-We will use Theorem~\ref{thm:sequential} and Lemma~\ref{lem:nonintswap} to make repeated swaps
+We will use Theorem~\ref{thm:sequentialfull} and Lemma~\ref{lem:nonintswapfull} to make repeated swaps
 of adjacent operations to transform $E'$ into $E_1 \dplus \ldots \dplus E_n$.  The symmetry of
-Lemma~\ref{lem:nonintswap} is then used to show that the validity of $E_1 \dplus \ldots \dplus E_n$
-under Theorem~\ref{thm:sequential} implies the validity of $E'$.
+Lemma~\ref{lem:nonintswapfull} is then used to show that the validity of $E_1 \dplus \ldots \dplus E_n$
+under Theorem~\ref{thm:sequentialfull} implies the validity of $E'$.
 
@@
 \begin{lem}
-\label{lem:nonintpriv}
+\label{lem:nonintprivfull}
 \rm
@@
 \begin{thm}
-\label{thm:parallelexec}
+\label{thm:parallelexecfull}
 \rm
@@
 
-The proof follows directly from Lemmas~\ref{lem:noninteffects}
-and \ref{lem:nonintpriv}.  This result holds even if the
+The proof follows directly from Lemmas~\ref{lem:noninteffectsfull}
+and \ref{lem:nonintprivfull}.  This result holds even if the
 clobber set $C$ is non-empty, allowing two ``locally independent''
@@
 \subsection{Relaxed Coherence}
-\label{subsec:relaxed}
+\label{subsec:relaxedfull}
 
@@
 that are {\em simultaneously non-interfering} (under mapping $M$) can be executed
-in parallel, which includes all the cases covered by Theorem~\ref{thm:parallelexec} as well
+in parallel, which includes all the cases covered by Theorem~\ref{thm:parallelexecfull} as well
 as those where the programmer has used the $simult$ coherence mode to explicitly allow parallel
@@
 \section{Hierarchical Scheduling}
-\label{sec:scheduling}
+\label{sec:schedulingfull}
 
@@
 subexpression of $e_2$, with corresponding privileges $\Phi_3$ and $\Phi_4$.
-By Theorem~\ref{thm:parallelexec}, if the runtime can verify $\Phi_1 \nonint[M] \Phi_2$, then 
+By Theorem~\ref{thm:parallelexecfull}, if the runtime can verify $\Phi_1 \nonint[M] \Phi_2$, then 
 expressions $e_1$ and $e_2$ may be evaluated in parallel.  The evaluation of the subexpressions
@@
 of the subexpression must be non-interfering (allowing parallel execution by
-Lemma~\ref{lem:noninteffects}).
+Lemma~\ref{lem:noninteffectsfull}).
 
 \begin{thm}
-\label{thm:hiersched}
+\label{thm:hierschedfull}
 \rm
@@
 Similarly, any memory operation $\epsilon_4$ that is in $E_4$ must also be in $E_2$.  We have
-$E_1 \nonint E_2$ from Lemma~\ref{lem:nonintpriv}, so and therefore $\epsilon_3 \nonint \epsilon_4$.  Since
+$E_1 \nonint E_2$ from Lemma~\ref{lem:nonintprivfull}, so and therefore $\epsilon_3 \nonint \epsilon_4$.  Since
 this holds for all $\epsilon_3$ in $E_3$ and all $\epsilon_4$ in $E_4$, we have $E_3 \nonint E_4$.
@@
 
- The Legion task scheduler uses Theorem~\ref{thm:hiersched} in the
+ The Legion task scheduler uses Theorem~\ref{thm:hierschedfull} in the
 following way.  Sibling function calls (those invoked within the same

### 2012-11-12 doc/llr-pldi2013/interface.tex
redid physical region part a bit

@@
 \subfigure{\usebox{\mylistingbox}} \\
+\vspace{-0.2cm}
 
@@
 
-class RegionMetaData {
+class PhysicalRegion {
   const unsigned id;
-  static const RegionMetaData NO_REGION;
+  static const PhysicalRegion NO_REGION;
 
-  static RegionMetaData create_region(size_t num_elmts, size_t elmt_size);
+  static PhysicalRegion create_region(size_t num_elmts, size_t elmt_size);
   void destroy_region() const;
+
+  ptr_t alloc();
+  void free(ptr_t p);
 
@@
   RegionInstance create_instance(Memory memory, ReductionOpID redopid) const;
-  void destroy_instance(RegionInstance instance) const;
+  void destroy_instance(RegionInstance instance, Event wait_on = NO_EVENT) const;
 };
@@
 
+  void *element_data_ptr(ptr_t p);
   Event copy_to(RegionInstance target, Event wait_on = NO_EVENT);
@@
 \subfigure{\usebox{\mylistingbox}} \\
+\vspace{-0.2cm}
 
 \begin{lrbox}{\mylistingbox}
-\begin{lstlisting}[firstnumber=62]
+\begin{lstlisting}[firstnumber=last]
 class Machine {
@@
   %\path (1,3) node (t1) [shape=rectangle,draw] {run TASK\_A} -- node [right=1pt] {ea} +(0,-0.5) node[auto] {eax};
-  \node (t1) at (0,0) [shape=rectangle,draw] {\begin{tabular}{c}{\bf run}\\TASK\_A\end{tabular}};
-  \draw (t1) to node[auto] {ea} (1.4,0);
+  \node (t1) at (-0.2,0) [shape=rectangle,draw] {\begin{tabular}{c}{\bf run}\\TASK\_A\end{tabular}};
+  \draw (t1) to node[auto] {ea} (1.2,0);
 
@@
   \draw (t2) to node[auto] {eb} (4.2,1);
-  \draw [->] (1.4,0) to (t2.west);
+  \draw [->] (1.2,0) to (1.6,1) to (t2.west);
 
@@
   \draw (t3) to node[auto] {ec} (4.2,-1);
-  \draw [->] (1.4,0) to (t3.west);
+  \draw [->] (1.2,0) to (1.6,-1) to (t3.west);
 
@@
 Current APIs for moving data in machines with discrete memories only support untyped
-copies of a buffer of data.  This restriction requires programmers to manage both the layout of
+copies of a buffer of data and force the programmer to work with pointers in a variety of address
+spaces.  This restriction requires programmers to manage both the layout of
 data and the marshalling of data for communication.  {\em Physical regions} are a
-mechanism for decoupling the placement of data from its layout.  Physical regions make
-the runtime aware of the structure of elements in the allocation and thereby enable multiple data
+mechanism for decoupling the placement of data from its layout.  Physical regions allow portability
+of ``pointers'' as data is moved around the memory hierarchy, and enable multiple data
 layouts, such as specialized layouts for buffering individual reduction operations.
 
-A physical region is an allocation of data in a single memory in the memory hierarchy.  Physical
-regions are grouped into {\em classes}, which are sets of physical regions sharing the
-same names (i.e. addresses) for elements, but not necessarily sharing the same data layout.  This
-allows pointers to elements within regions to be used, even if the regions are copied from one memory
-to another.  Our
-interface supports sub-classing, but the details are omitted due to space constraints.  The
-interface allows copies between physical regions of
-the same class (or between super- and sub-classes).
+%% A physical region is an allocation of data in a single memory in the memory hierarchy.  Physical
+%% regions are grouped into {\em classes}, which are sets of physical regions sharing the
+%% same names (i.e. addresses) for elements, but not necessarily sharing the same data layout.  This
+%% allows pointers to elements within regions to be used, even if the regions are copied from one memory
+%% to another.  Our
+%% interface supports sub-classing, but the details are omitted due to space constraints.  The
+%% interface allows copies between physical regions of
+%% the same class (or between super- and sub-classes).
 
@@
   ReductionOpID redopid = ...;
-  RegionMetaData meta = RegionMetaData::
+  PhysicalRegion meta = PhysicalRegion::
                         create_region(num_elmts, elmt_size);
@@
 Each memory in the machine's memory hierarchy is named by a {\tt Memory} object (line 39).
-The next object is a {\tt RegionMetaData} object (line 44), which defines a class of
+The next object is a {\tt PhysicalRegion} object (line 44), which contains information about the
+maximum number and size of the elements within the region (line 48) and handles dynamic allocation
+within the region (lines 51-52), but does not contain any storage for the element data.  Instead,
+element data is stored in on or more {\tt RegionInstance} objects (line 59), each of which can be
+created in an arbitrary {\tt Memory} (line 54).
+
+A {\tt ptr\_t} value returned by the {\tt alloc}
+method may be used to access the corresponding element in any of the instances of the region (line 62).
+This
+allows the creation of irregular data structures within physical regions.  The {\tt ptr\_t} values can
+be safely stored in a region, and need not be changed when data is copied from one instance to another.
+In common cases, this is as cheap as array-addressing calculations, so the overhead is minimal.
+When more than one instance exists for a region, the client is responsible for maintaining coherence
+of the data between the regions.  This is typically achieved through the use of instance-to-instance
+copy operations (line 63) that use the producer task's completion event as a prerequisite and provide
+their own completion event as a prerequisite for the consumer task.
+defines a class of
 regions.  Physical regions that belong to that class are created with the {\tt create\_instance}
@@
 between physical regions (line 59).  Like other operations copies can be
-predicated on an event and return an event corresponding to completion.
+predicated on an event and return an event corresponding to completion.  When a copy operation is
+the last use of a particular instance, the destruction of that instance can also be requested and made
+dependent on the copy operation's event, allowing the runtime to reclaim the memory as soon as (but no
+sooner than) possible.
 
 If the only operations a task will perform on a region are reductions, a special reduction-only
-instance can be created (line 52).  Rather than applying each reduction separately to the target instance,
+instance can be created (line 55).  Rather than applying each reduction separately to the target instance,
 the reduction operations can be accumulated in the reduction instance, and applied in bulk to 
-the target using the {\em reduce\_to} operation (line 60).  This has several benefits.  First, it results
+the target using the {\em reduce\_to} operation (line 64).  This has several benefits.  First, it results
 in much better performance, as we show in Section~\ref{subsec:reducmicro}, by reducing both
@@
 %shows a subset of the physical region interface.  The first object in the interface
-%is a {\tt RegionMetaData} object (line 1).  Region meta-data objects provide the interface for the
+%is a {\tt PhysicalRegion} object (line 1).  Region meta-data objects provide the interface for the
 %creation of sets of physical regions that possess the same number and type of elements (but not necessarily
@@
 \label{subsec:machmodel}
-The last component of the interface is the {\tt Machine} object, shown in lines 62-73 of Figure~\ref{fig:runtimeapi},
+The last component of the interface is the {\tt Machine} object, shown in lines 66-77 of Figure~\ref{fig:runtimeapi},
 which provides initialization and run-time introspection
@@
 program, providing a task table that maps task IDs to function pointers,
-and then invoking the {\tt run} method (lines 65).
+and then invoking the {\tt run} method (line 69).
 Any running task can get a pointer to the machine object with the
-{\tt get\_machine} method (line 67) and use it to obtain lists of all the {\tt Memory} (line 68) or 
-{\tt Processor} (line 69) handles in the system.  Using the affinity methods (lines 71-72),
+{\tt get\_machine} method (line 71) and use it to obtain lists of all the {\tt Memory} (line 72) or 
+{\tt Processor} (line 73) handles in the system.  Using the affinity methods (lines 75-76),
 the application can also determine which memories are accessible by each

### 2012-11-12 doc/llr-pldi2013/conclusion.tex
conc

--- 2012-07-23 doc/asplos2013/conclusion.tex
@@
 
-As large supercomputers continue to scale, the latencies associated
-with common operations such as communication and synchronization will
-become larger and more variable.  To effectively hide these latencies
-future programming systems will need to leverage asynchronous operations
-to overlap long latency operations with useful work.  We have
-presented a low-level programming interface that is fully asynchronous.
-Our interface uses events to compose asynchronous operations.  Clients
-can use events to explicitly specify dependences between operations,
-leaving the runtime to schedule them in a way that maximizes latency 
+As supercomputers continue to grow, the latencies of communication and
+synchronization will become larger and more variable.  Effectively
+hiding latency in future programming systems will require composing
+asynchronous operations to overlap long latency operations with useful
+work.  We have presented a low-level programming interface that is
+fully asynchronous, including deferred locks, a new synchronization
+primitive.  Asynchronous operations are composed using events, which
+specify dependences between operations, giving the
+runtime the flexibility to schedule operations to maximize latency
 hiding while still obeying dependences.
 
-We've presented an implementation of interface that is capable of running
-on a large heterogeneous cluster containing both CPUs and GPUs.  We
-employed micro-benchmarks to show that our implementation is efficient.
-We also demonstrated that our interface is capable of supporting Legion,
-an advanced high-level runtime system that fully leverages the asynchronous
-features of our interface.  We showed that for three real-world applications
-the asynchronous Legion implementations were between 22\% to 135\% faster than
-implementations which contained blocking operations.
+%We've presented an implementation of our interface that is capable of running
+%on a large heterogeneous cluster containing both CPUs and GPUs.  We
+%employed micro-benchmarks to show that our implementation is efficient.
+%We also demonstrated that our interface is capable of supporting Legion,
+%an advanced high-level runtime system that fully leverages the asynchronous
+%features of our interface.  We showed that for three real-world applications
+%the asynchronous Legion implementations were between 22\% to 135\% faster than
+%implementations which contained blocking operations.

### 2012-11-12 doc/llr-pldi2013/interface.tex
got rid of an old paragraph that was in the middle of a new one

@@
 within the region (lines 51-52), but does not contain any storage for the element data.  Instead,
-element data is stored in on or more {\tt RegionInstance} objects (line 59), each of which can be
-created in an arbitrary {\tt Memory} (line 54).
+element data is stored in one or more {\tt RegionInstance} objects (line 59), each of which can be
+created in an arbitrary {\tt Memory} (line 54).  To maintain performance transparency, there is no
+use of virtual memory - each {\tt Memory} is sized based on physical capacity and a new physical
+instance can only be allocated if sufficient space remains.
 
@@
 copy operations (line 63) that use the producer task's completion event as a prerequisite and provide
-their own completion event as a prerequisite for the consumer task.
-defines a class of
-regions.  Physical regions that belong to that class are created with the {\tt create\_instance}
-method (line 51) and represented by {\tt RegionInstance} objects (line 56).  When creating a physical
-region the client specifies the memory in which the physical region is
-to be allocated.  To maintain performance transparency, a new physical region can only be allocated
-if the memory has sufficient capacity.
-There is no automatic coherence of data between physical regions in the
-same class.  It is the client's responsibility to manage data coherence using copies
-between physical regions (line 59).  Like other operations copies can be
-predicated on an event and return an event corresponding to completion.  When a copy operation is
+their own completion event as a prerequisite for the consumer task.  When a copy operation is
 the last use of a particular instance, the destruction of that instance can also be requested and made

### 2012-11-12 doc/llr-pldi2013/interface.tex
more minor fixes

@@
 
-If the only operations a task will perform on a region are reductions, a special reduction-only
+If the only operations a task will perform on a instance are reductions, a special reduction-only
 instance can be created (line 55).  Rather than applying each reduction separately to the target instance,
-the reduction operations can be accumulated in the reduction instance, and applied in bulk to 
-the target using the {\em reduce\_to} operation (line 64).  This has several benefits.  First, it results
+the reduction operations can be accumulated in the reduction instance.  Using the {\tt reduce\_to} 
+method (line 64), these accumulated reductions can be applied in bulk to another instance, which may be either
+a normal instance or another reduction-only instance.  First, it results
 in much better performance, as we show in Section~\ref{subsec:reducmicro}, by reducing both
 the total inter-node traffic and the overhead of that traffic while still allowing parallelism.
-Second, a reduction-only instance is often smaller that the target instance, potentially improving
+Second, a reduction-only instance is often smaller than a normal instance, potentially improving
 cache performance and helping address capacity issues for smaller memories (e.g. a GPU's device
@@
 
-An example using reduction-only regions is given in Figure~\ref{fig:reducevents} along with the resulting event graph.
-Two reduction-only regions, {\tt inst1} and {\tt inst2}, are created in different memories to
+An example using reduction-only instances is given in Figure~\ref{fig:reducevents} along with the resulting event graph.
+Two reduction-only instances, {\tt inst1} and {\tt inst2}, are created in different memories to
 be used with a reduction operation (lines 7-8).  (Reduction operations are registered at startup with integer IDs
-as function pointers may not be portable across nodes.)  A physical region of the same class {\tt inst3}
+as function pointers may not be portable across nodes.)  A normal instance {\tt inst3}
 is also created in a different memory (line 9).  Two tasks are launched that will apply
-reductions locally into their respective reduction buffers (lines 10-11).  The resulting
+reductions locally into their respective reduction instances (lines 10-11).  The resulting
 events are then used to chain reduction copies back to {\tt inst3} after the tasks complete (lines 12-13).

### 2012-11-12 doc/llr-pldi2013/micro.tex
more minor fixes

@@
 For our experiment, there are 8 reduction tasks created for each node, one per processor.  Each task performs 4M reductions
-into a histogram with 256K buckets.  The results for a variety of node counts are shown in Figure~\ref{fig:reducbench}.
+into a histogram with 256K buckets.
+The input data is random, so although the average number of reductions per bucket is 16, there will be some buckets with more and some with fewer.
+The results for a variety of node counts are shown in Figure~\ref{fig:reducbench}.
 They show that the version using reduction-only instance performs 
@@
 but its inherently serial nature doesn't benefit from increasing node counts.  Finally, the version using individual
-reduction messages is scalable, but the per-message overhead makes it nearly 1000x slower than the versions that are
-able to move data in bulk.
+reduction messages is scalable, but the need to send each reduction separately and the per-message overhead make it nearly 1000x slower than the reduction instance version that is
+able to accumulate multiple reductions to the same location locally and move data in bulk.
 

### 2012-11-12 doc/llr-pldi2013/interface.tex
Minor edits to the low level paper

@@
 When more than one instance exists for a region, the client is responsible for maintaining coherence
-of the data between the regions.  This is typically achieved through the use of instance-to-instance
+of the data between instances of a region.  This is typically achieved through the use of instance-to-instance
 copy operations (line 63) that use the producer task's completion event as a prerequisite and provide
@@
 the last use of a particular instance, the destruction of that instance can also be requested and made
-dependent on the copy operation's event, allowing the runtime to reclaim the memory as soon as (but no
+dependent on the copy operation's event (line 56), allowing the runtime to reclaim the memory as soon as (but no
 sooner than) possible.
@@
 method (line 64), these accumulated reductions can be applied in bulk to another instance, which may be either
-a normal instance or another reduction-only instance.  First, it results
+a normal instance or another reduction-only instance.  The use of reduction instances can result 
 in much better performance, as we show in Section~\ref{subsec:reducmicro}, by reducing both
 the total inter-node traffic and the overhead of that traffic while still allowing parallelism.
-Second, a reduction-only instance is often smaller than a normal instance, potentially improving
+Furthermore, reduction-only instances are often smaller than normal instances, potentially improving
 cache performance and helping address capacity issues for smaller memories (e.g. a GPU's device

### 2012-11-12 doc/llr-pldi2013/micro.tex
Minor edits to the low level paper

@@
 but its inherently serial nature doesn't benefit from increasing node counts.  Finally, the version using individual
-reduction messages is scalable, but the need to send each reduction separately and the per-message overhead make it nearly 1000x slower than the reduction instance version that is
+reduction messages is scalable, but the need to send each reduction separately and the per-message overhead make it nearly 
+1000x slower than the reduction instance version, which is
 able to accumulate multiple reductions to the same location locally and move data in bulk.

### 2012-11-13 doc/llr-pldi2013/impl.tex
update
Lots of small edits

@@
 %events, locks, and reduction instances.
-Due to space constraints we cover only the most interesting parts of the implementation in detail.
-We limit our discussion to three aspects:
-events, locks, and reduction instances.
+%Due to space constraints we cover only the most interesting parts of the implementation in detail.
+%We limit our discussion to three aspects:
+%events, locks, and reduction instances.
+Due to space constraints, we only describe the implementation of features
+that are technical contributions.  We cover in detail
+the implementation of events, deferred locks, and reduction instances.
 
@@
 In addition to normal physical region instances, our implementation supports a special reduction-only
-instance that differ in two important ways.  First, the per-element storage in a reduction-only instance
+instance that differs in two important ways.  First, the per-element storage in a reduction-only instance
 is sized to hold the ``right-hand side'' of the reduction operation (e.g. the $v$ in an operation like

### 2012-11-13 doc/llr-pldi2013/interface.tex
Lots of small edits

@@
 not necessarily ordered.  Traditionally locks have been used to serialize access to
-data.  However, all lock implementations that we are aware of require either blocking
+data.  However, lock implementations in existing parallel programming models require either blocking
 or spinning, neither of which composes well with asynchronous operations.
@@
 
-If the only operations a task will perform on a instance are reductions, a special reduction-only
+If the only operation a task will perform on an instance is a reduction, a special reduction-only
 instance can be created (line 55).  Rather than applying each reduction separately to the target instance,

### 2012-11-13 doc/llr-pldi2013/intro.tex
Lots of small edits

@@
 can use an event as a precondition.
-  Using
-events clients can compose arbitrary chains of dependent
+Using events, clients can compose arbitrary chains of dependent
 operations, which the runtime is free to execute in any way that
@@
 
-\item Our approach depends on the efficient handling of very large numbers of events.
-We describe a distributed implementation of our interface that requires no global coordination
-and has modest local storage costs.  We describe an optimized implementation of events, 
-deferred locks, and reductions (Section~\ref{sec:impl}).
+\item We describe an implmentation of events based on a {\em generational event}
+data structure that compactly represents a large number of events.  Generational
+events also amortize the cost of inter-node communication when multiple operations are dependent
+on the same event (Section~\ref{sec:impl}).
+%%\item Our approach depends on the efficient handling of very large numbers of events.
+%%We describe a distributed implementation of our interface that requires no global coordination
+%%and has modest local storage costs.  We describe an optimized implementation of events, 
+%%deferred locks, and reductions (Section~\ref{sec:impl}).
 

### 2012-11-13 doc/types-pldi2013/alt_example.tex
fixed typos found by Elliott - tried to tweak multicoloring and clobber set paragraphs

@@
 Programmers express partitions using {\em colorings},
-maps from region elements to integers (``colors'').
+maps from region elements to integers identifying which subregion an element
+should be in (``colors'').
 Listing~\ref{lst:coloring} shows the function {\tt create\_piece\_coloring}
@@
 is used, which assigns each region element to one or more colors.  
-%(Multicolorings are
-%implemented in Core Legion with separate colorings for each color.)
+\footnote{For simplicity, multicolorings are
+implemented in Core Legion as separate colorings for each color.}
+
 \begin{lstlisting}[label={lst:coloring},caption={Coloring Construction}]

### 2012-11-13 doc/types-pldi2013/langdef.tex
fixed typos found by Elliott - tried to tweak multicoloring and clobber set paragraphs

@@
 \begin{tabular}{m{5cm}m{5cm}m{8cm}}
-\infrule[T-Pack]{T_1 = \exists r'_1, \ldots r'_n.\ T_2\text{ where }\Omega_1 \\ \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^* \\ \typeenvx e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]}{\typeenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] : T_1} &
+\infrule[T-Pack]{T_1 = \exists r'_1, \ldots r'_k.\ T_2\text{ where }\Omega_1 \\ \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^* \\ \typeenvx e_1 : T_2[r_1/r'_1,\ldots,r_k/r'_k]}{\typeenvx \text{pack}\ e_1\ \text{as}\ T_1[\otok{k}{r}] : T_1} &
 \infrule[T-IfElse]{\typeenvx e_1 : bool \\ \typeenvx e_2 : T \\ \typeenvx e_3 : T}{\typeenvx \text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3 : T} &
@@
 \begin{tabular}{m{10cm}m{8cm}}
-\infrule[T-Unpack]{T_1 = \exists r'_1, \ldots, r'_n.\ T_2\text{ where }\Omega_1 \andalso \typeenvx e_1 : T_1 \\
+\infrule[T-Unpack]{T_1 = \exists r'_1, \ldots, r'_k.\ T_2\text{ where }\Omega_1 \andalso \typeenvx e_1 : T_1 \\
   \Gamma' = \Gamma[T_2[r_1/r'_1,\ldots,r_k/r'_k] / id] \andalso \Omega' = \Omega \cup \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \\
-  \typeenvx[G=\Gamma',O=\Omega'] e_2 : T_3 \andalso \{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T_1, T_3) = \emptyset}
-  {\typeenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 : T_3} &
-\infrule[T-Part]{\typeenvx e_1 : \text{coloring}(r_p) \\ \Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{1 \leq i < j \leq k} r_i * r_j \\ \typeenvx[O=\Omega'] e_2 : T \\ \{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T) = \emptyset}
+  \typeenvx[G=\Gamma',O=\Omega'] e_2 : T_3 \andalso \{ \otok{k}{r} \} \cap \textit{regions\_of}(\Gamma, T_1, T_3) = \emptyset}
+  {\typeenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}]\ \text{in}\ e_2 : T_3} &
+\infrule[T-Part]{\typeenvx e_1 : \text{coloring}(r_p) \\ \Omega' = \Omega \wedge \bigwedge_{i \in [1,k]} r_i \leq r_p \wedge \bigwedge_{1 \leq i < j \leq k} r_i * r_j \\ \typeenvx[O=\Omega'] e_2 : T \\ \{ \otok{k}{r} \} \cap \textit{regions\_of}(\Gamma, T) = \emptyset}
   {\typeenvx \text{partition}\ r_p\text{ using }e_1\text{ as }\otok{k}{r}\text{ in }e_2 : T}
@@
 
-The environment also has a {\em clobber set} $C$.
-We will show that parallel Legion programs can be scheduled in
-a hierarchical fashion.  We therefore require an
-operational semantics that is parallel and hierarchical.  Now
-Legion allows limited nondeterminism, specifically because of relaxed coherence on
+The environment also has a {\em clobber set} $C$, which is the key to handling
+limited nondeterminism  while still having an operational
+semantics that is parallel and hierarchical.  Nondeterminism arises in a Legion
+program because of relaxed coherence on
 regions shared between parallel tasks and choices of reduction orderings.
@@
 %\vspace{-0.5cm}
-\infrule[E-Call]{\opsenvx e_1 \mapsto v_1, E_1 \andalso S_1 = \text{apply}(S, E_1) \andalso \opsenvx e_2 \mapsto v_2, E_2 \andalso \ldots \andalso S_n = \text{apply}(S_{n-1}, E_n) \\
+\infrule[E-Call]{\opsenvx e_1 \mapsto v_1, E_1 \andalso S_1 = \text{apply}(S, E_1) \andalso \opsenvx[S=S_1] e_2 \mapsto v_2, E_2 \andalso \ldots \andalso S_n = \text{apply}(S_{n-1}, E_n) \\
   valid\_interleave(S, C, E', \oton{E}) \andalso \text{function }id[\otok{k}{r'}](a_1 : T_1, \ldots, a_n : T_n), \Phi', Q' : T_r = e_{n+1} \\
@@
   C' = C \cup \{ l : \exists \rho. atomic(\rho) \in M' \llbracket Q' \rrbracket \vee simult(\rho) \in M' \llbracket Q' \rrbracket \} \\
-  \opsenvx[M=M',L=L',S=S',C=C'] e_{n+1} \mapsto v_{n+1}, E_{n+1} \andalso E'_{n+1} = mark\_coherence(E_{n+1}, M' \llbracket Q' \rrbracket) \andalso valid\_interleave(S, C, E'', E', E_{n+1})}
+  \opsenvx[M=M',L=L',S=S',C=C'] e_{n+1} \mapsto v_{n+1}, E_{n+1} \andalso E'_{n+1} = mark\_coherence(E_{n+1}, M' \llbracket Q' \rrbracket) \andalso valid\_interleave(S, C, E'', E', E'_{n+1})}
   {\opsenvx id[\otok{k}{r}](\oton{e}) \mapsto v_{n+1}, E''}
@@
 \typeenvx[O=\Omega'] e_2 : T \\
-\{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T) = \emptyset
+\{ \otok{k}{r} \} \cap \textit{regions\_of}(\Gamma, T) = \emptyset
 \end{array}}
@@
 {\begin{array}{l}
-T_1 = \exists r'_1, \ldots r'_n.\ T_2\text{ where }\Omega_1 \\
+T_1 = \exists r'_1, \ldots r'_k.\ T_2\text{ where }\Omega_1 \\
 \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^* \\
@@
 {\begin{array}{l}
-T_1 = \exists r'_1, \ldots, r'_n.\ T_2\text{ where }\Omega_1 \\
+T_1 = \exists r'_1, \ldots, r'_k.\ T_2\text{ where }\Omega_1 \\
 \typeenvx e_1 : T_1 \\
@@
 \typeenvx[G=\Gamma',O=\Omega'] e_2 : T_3 \\
-\{ \oton{r} \} \cap \textit{regions\_of}(\Gamma, T_1, T_3) = \emptyset
-\end{array}}
-{\typeenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}] \text{in}\ e_2 : T_3}
+\{ \otok{k}{r} \} \cap \textit{regions\_of}(\Gamma, T_1, T_3) = \emptyset
+\end{array}}
+{\typeenvx \text{unpack}\ e_1\ \text{as}\ id : T_1[\otok{k}{r}]\ \text{in}\ e_2 : T_3}
 
@@
 \ruleatx{\XA,8.1}{\XE,8.4}{E-Call}{
-\opsenvx e_1 \mapsto v_1, E_1 \andalso S_1 = apply(S, E_1) \andalso \opsenvx e_2 \mapsto v_2, E_2 \andalso \ldots \andalso S_n = apply(S_{n-1}, E_n) \\
+\opsenvx e_1 \mapsto v_1, E_1 \andalso S_1 = apply(S, E_1) \andalso \opsenvx[S=S_1] e_2 \mapsto v_2, E_2 \andalso \ldots \andalso S_n = apply(S_{n-1}, E_n) \\
 valid\_interleave(S, C, E', \oton{E}) \andalso \text{function }id[\otok{k}{r'}](a_1 : T_1, \ldots, a_n : T_n), \Phi', Q' : T_r = e_{n+1} \\
@@
 C' = C \cup \{ l : \exists \rho. atomic(\rho) \in M' \llbracket Q' \rrbracket \vee simult(\rho) \in M' \llbracket Q' \rrbracket \} \\
-\opsenvx[M=M',L=L',S=S',C=C'] e_{n+1} \mapsto v_{n+1}, E_{n+1} \andalso E'_{n+1} = mark\_coherence(E_{n+1}, M' \llbracket Q' \rrbracket) \andalso valid\_interleave(S, C, E'', E', E_{n+1})
+\opsenvx[M=M',L=L',S=S',C=C'] e_{n+1} \mapsto v_{n+1}, E_{n+1} \andalso E'_{n+1} = mark\_coherence(E_{n+1}, M' \llbracket Q' \rrbracket) \andalso valid\_interleave(S, C, E'', E', E'_{n+1})
 }{\opsenvx id[\otok{k}{r}](\oton{e}) \mapsto v_{n+1}, E''}

### 2012-11-13 doc/types-pldi2013/intro.tex
intro

@@
 
-In this work, we present the static and dynamic semantics for Legion \cite{Legion12},
+In this work, we present static and dynamic semantics for Legion  \cite{Legion12},
 a parallel programming model that supports multiple, dynamic data partitions and is

### 2012-11-13 doc/types-pldi2013/evaluation.tex
Working on figures to make the prettier and easier to understand
mods

@@
 \cite{Legion12} with no checking of region memory accesses.  
-When testing applications on this system we frequently encountered memory corruption due to
+When using this system we frequently encountered memory corruption due to
 illegal region accesses caused by application bugs.
@@
 \begin{center}
-\includegraphics[scale=0.33]{figs/circuit_96_popl.pdf}
+\includegraphics[scale=0.30]{figs/circuit_96_popl.pdf}
 \end{center}
@@
 \caption{Overhead for the Circuit simulation with 96 total pieces.\label{fig:ckt_overhead}}
-\vspace{-6mm}
+\vspace{-5mm}
 \end{figure}
@@
 \begin{center}
-\includegraphics[scale=0.33]{figs/fluid_19200_popl.pdf}
+\includegraphics[scale=0.30]{figs/fluid_19200_popl.pdf}
 \end{center}
@@
 \caption{Overhead for the Fluid simulation with 19200 cells.\label{fig:fluid_overhead}}
-\vspace{-6mm}
+\vspace{-5mm}
 \end{figure}
@@
 {
-\includegraphics[scale=0.33]{figs/amr_4096_popl.pdf}
+\includegraphics[scale=0.30]{figs/amr_4096_popl.pdf}
 \label{fig:amr4096}
@@
 {
-\includegraphics[scale=0.33]{figs/amr_8192_popl.pdf}
+\includegraphics[scale=0.30]{figs/amr_8192_popl.pdf}
 \label{fig:amr8192}
@@
 {
-\includegraphics[scale=0.33]{figs/amr_16384_popl.pdf}
+\includegraphics[scale=0.30]{figs/amr_16384_popl.pdf}
 \label{fig:amr16384}
@@
 \caption{Overhead for the AMR application.\label{fig:amr_overhead}}
-\vspace{-6mm}
+\vspace{-8mm}
 \end{figure}
@@
 \begin{center}
-\includegraphics[scale=0.33]{figs/circuit_coherence.pdf}
+\subfigure[48 Piece Problem Size]
+{
+\includegraphics[scale=0.18]{figs/circuit_coherence_48.pdf}
+\label{fig:circuit_excl_48}
+}
+\subfigure[96 Piece Problem Size]
+{
+\includegraphics[scale=0.18]{figs/circuit_coherence_96.pdf}
+\label{fig:circuit_excl_96}
+}
 \end{center}
-\vspace{-2mm}
-\caption{Performance gains from relaxed coherence modes.\label{fig:circuit_excl}}
+\vspace{-4mm}
+\caption{Performance of relaxed coherence modes.\label{fig:circuit_excl}}
 \vspace{-6mm}
 \end{figure}
+%\begin{figure}
+%\begin{center}
+%\includegraphics[scale=0.33]{figs/circuit_coherence.pdf}
+%\end{center}
+%\vspace{-2mm}
+%\caption{Performance gains from relaxed coherence modes.\label{fig:circuit_excl}}
+%\vspace{-6mm}
+%\end{figure}
 

### 2012-11-14 doc/types-pldi2013/alt_example.tex
hopefully last wave of edits, spell check

@@
 \begin{lstlisting}[label={lst:tuples},caption={Tuples and Pointers Example}]
---                        voltage,current,charge,capacitance,piece ID
-type CircuitNode        = <float,float,float,float,int>
+--                        voltage, current, charge, capacitance, piece ID
+type CircuitNode        = <float, float, float, float, int>
 --                      owned node, owned or ghost node, resistance, current
-type CircuitWire<rn,rg>  = <CircuitNode@rn, CircuitNode@(rn,rg),float,float>
+type CircuitWire<rn,rg>  = <CircuitNode@rn, CircuitNode@(rn,rg), float, float>
 \end{lstlisting}
@@
 program data.  An ideal partitioning depends on many factors,
-including the shape of data structures, the input, and the number of
-partitions, which usually varies with the target machine.  To allow
+including the shape of data structures, the input, and the desired number of
+partitions (which usually varies with the target machine).  To allow
 the necessary flexibility, in Legion partitioning is simply programmed,
@@
 \caption{Partitions of $r{\_}all{\_}nodes$. \label{fig:part_fig}}
-\vspace{-6mm}
+\vspace{-4mm}
 %  \caption[My caption
@@
 statement shown in Listing~\ref{lst:partition} performs the ``owned nodes''
-partitioning.    Since colorings assign each region element at most one color
-(dynamically checked), the partition operation results in 
+partitioning.    Since colorings assign each region element at most one color,
+the partition operation results in 
 {\em disjoint} subregions, each a subset of the {\em parent} region.
@@
 \begin{lstlisting}[label={lst:partition},caption={Partition Operation Example}]
-let piece_coloring : coloring(all_nodes) = create_piece_coloring[rl,all_nodes](node_list) in
-partition all_nodes using piece_coloring as rn0,rn1... in ...
+let owned_map : coloring(all_nodes) = create_piece_coloring[rl,all_nodes](node_list) in
+partition all_nodes using owned_map as rn0,rn1... in ...
 \end{lstlisting}
@@
 Listing~\ref{lst:loop} shows the main time step loop run by the 
-recursive function {\em execute\_time\_steps} (line 11)
+recursive function {\em execute\_time\_steps} (lines 13-24)
 as well the leaf task functions (lines 2-7).  
@@
 because of the atomic and commutative nature of reductions.  Finally,
-the {\tt update\_voltage} function (line 6) modifies only the disjoint owned regions, 
+the {\tt update\_voltage} function (lines 6-7) modifies only the disjoint owned regions, 
 permitting each piece to be updated in parallel.  No single partition of the nodes
@@
       reads(rl,rn), writes(rn) : bool
+
 -- Reduction function for distribute charge
@@
 The {\tt execute\_time\_steps} function also demonstrates the need to 
-dynamically discover parallelism through region disjointess.  
+dynamically discover parallelism through region disjointness.  
 To run the two instances of 
@@
 and could have been captured in a region relationship at the cost
-of much more complicated code.  Instead, we use a simpler region relationship
+of much more complicated code.  Instead, we are able to use a simpler region relationship
 and rely on an inexpensive dynamic disjointness

### 2012-11-14 doc/types-pldi2013/evaluation.tex
hopefully last wave of edits, spell check

@@
 using particles that move within a regular grid of cells.  To perform operations in 
-parallel the array of cells is partitioned.  Unlike
+parallel, the array of cells is partitioned.  Unlike
 Circuit, Fluid creates and partitions regions before
@@
 are used in all applications.   Finally, all applications introduce aliasing of
-data through both multi-colorings or having multiple partitions.  Our implementations of
+data through both multicolorings or having multiple partitions.  Our implementations of
 these applications both type check and execute, proving that our type system is
@@
 a result of the algorithm required by the application, not the Legion runtime.
-Even in the case of the Circuit application, which exhibits quadratic communication scaling, the Legion
+Even in the case of the Circuit application, which exhibits quadratic increases in communication cost, the Legion
 runtime is able to achieve a 62.5X speedup on 96 GPUs over a hand-coded single GPU implementation\cite{Legion12}.

### 2012-11-14 doc/types-pldi2013/intro.tex
hopefully last wave of edits, spell check

@@
 %dynamic checks.  
-The critical feature in Legion that makes static analysis tractable
+The critical feature in Legion that makes both static analysis tractable
 and dynamic analysis efficient in the presence of potential aliasing is logical regions.

### 2012-11-14 doc/types-pldi2013/langdef.tex
hopefully last wave of edits, spell check

@@
 %Types include booleans, integers, tuples, and pointers.  
-Pointers are annotated with a list of logical regions---a non-null pointer must point to a 
+Pointer types are annotated with a list of logical regions---a non-null pointer value must point to a 
 location that is contained in at least one of the regions. There is a special
@@
 chosen to allow a straightforward proof of type soundness.
+
+\begin{figure}
+\centering{
+$\begin{array}{l}
+\Omega \subseteq \Omega^* \\
+r_i \leq r_j \in \Omega^*  \Rightarrow r_i \leq r_i \in \Omega^*\wedge r_j \leq r_j \in \Omega^* \\
+r_i \leq r_j \in \Omega^* \wedge r_j \leq r_k \in \Omega^* \Rightarrow r_i \leq r_k \in \Omega^* \\
+r_i \leq r_j \in \Omega^* \wedge r_j * r_k \in \Omega^* \Rightarrow r_i * r_k \in \Omega^* \\
+r_i * r_j \in \Omega^* \Rightarrow r_j * r_i \in \Omega^* \\
+\\
+\Phi \subseteq \Phi^* \\
+r_i \leq r_j \in \Omega^* \wedge \text{reads}(r_j) \in \Phi^* \Rightarrow \text{reads}(r_i) \in \Phi^* \\
+r_i \leq r_j \in \Omega^* \wedge \text{writes}(r_j) \in \Phi^* \Rightarrow \text{writes}(r_i) \in \Phi^* \\
+r_i \leq r_j \in \Omega^* \wedge \text{reduces}_{id}(r_j) \in \Phi^* \Rightarrow \text{reduces}_{id}(r_i) \in \Phi^* \\
+\text{reads}(r) \in \Phi^* \wedge \text{writes}(r) \in \Phi^* \Rightarrow \text{reduces}_{id}(r) \in \Phi^* \\
+\ \ \ \mbox{for every function identifier $id$}
+\end{array}$
+}
+\caption{Privilege and Constraint Closure}
+\vspace{-5mm}
+\label{fig:closure}
+\end{figure}
 
@@
 
-\begin{figure}
-\centering{
-$\begin{array}{l}
-\Omega \subseteq \Omega^* \\
-r_i \leq r_j \in \Omega^*  \Rightarrow r_i \leq r_i \in \Omega^*\wedge r_j \leq r_j \in \Omega^* \\
-r_i \leq r_j \in \Omega^* \wedge r_j \leq r_k \in \Omega^* \Rightarrow r_i \leq r_k \in \Omega^* \\
-r_i \leq r_j \in \Omega^* \wedge r_j * r_k \in \Omega^* \Rightarrow r_i * r_k \in \Omega^* \\
-r_i * r_j \in \Omega^* \Rightarrow r_j * r_i \in \Omega^* \\
-\\
-\Phi \subseteq \Phi^* \\
-r_i \leq r_j \in \Omega^* \wedge \text{reads}(r_j) \in \Phi^* \Rightarrow \text{reads}(r_i) \in \Phi^* \\
-r_i \leq r_j \in \Omega^* \wedge \text{writes}(r_j) \in \Phi^* \Rightarrow \text{writes}(r_i) \in \Phi^* \\
-r_i \leq r_j \in \Omega^* \wedge \text{reduces}_{id}(r_j) \in \Phi^* \Rightarrow \text{reduces}_{id}(r_i) \in \Phi^* \\
-\text{reads}(r) \in \Phi^* \wedge \text{writes}(r) \in \Phi^* \Rightarrow \text{reduces}_{id}(r) \in \Phi^* \\
-\ \ \ \mbox{for every function identifier $id$}
-\end{array}$
-}
-\caption{Privilege and Constraint Closure}
-\vspace{-5mm}
-\label{fig:closure}
-\end{figure}
-
 \subsection{Operational Semantics}
@@
 Each expression has a clobber set $C$ of memory locations that might 
-be modified by concurrent evaluations.  While a read from a location not in $C$ is {\em coherent}
-(returning the most recently written value), the value read from a location $l$ in $C$ may be any
-value.  Determining which value is deferred until all potentially-concurrent
-modifications to $l$ are observable in the evaluation of a parent expression whose clobber set does
-not include $l$ and a consistent interleaving of the subexpressions' memory traces is determined.
+be modified by concurrent evaluations.  Reads of locations in $C$ may be
+chosen to return a value other than what was most recently written by the current
+expression, as long a
+valid interleaving of the operations with other concurrent operations can be found within
+an enclosing expression.
+%% While a read from a location not in $C$ is {\em coherent}
+%% (returning the most recently written value), the value read from a location $l$ in $C$ may be any
+%% value.  Determining which value is deferred until all potentially-concurrent
+%% modifications to $l$ are observable in the evaluation of a parent expression whose clobber set does
+%% not include $l$ and a consistent interleaving of the subexpressions' memory traces is determined.
 %With this formulation, the criteria for when a hierarchical scheduler may execute subexpressions in parallel can be
@@
 Thus, each subregion resulting from a partitioning may need an arbitrary number of locations that are
-unknown at partitioning time.  The semantics of newcolor allows this while
+unknown at partitioning time.  The semantics of newcolor allows extra locations to be placed in every
+subregion while
 guaranteeing all explicit colorings performed by the program are honored.
@@
 handling of relaxed coherence.  Any location that falls within
-a logical region whose coherence has been relaxed (via {\em atomic} or {\em simult}) is added to the clobber set
-$C'$, allowing non-deterministic evaluation within the subtask with respect to those locations.  All actual accesses
-to those locations within the resulting memory trace ($E_{n+1}$) are marked with the corresponding coherence mode,
-allowing the {\em valid\_interleave} check within the caller's environment to verify the consistency of the overall
-memory trace.
+a logical region whose coherence has been relaxed (via {\em atomic} or {\em simult}) is added to the
+subtask's clobber set $C'$, allowing non-deterministic evaluation within the subtask with respect to
+those locations.  All actual accesses to those locations within the resulting memory trace ($E_{n+1}$)
+are marked with the corresponding coherence mode, allowing the {\em valid\_interleave} check within
+the caller's environment to verify the consistency of the overall memory trace.
 
@@
 \privconsist{E}{\Phi} \Leftrightarrow \forall \epsilon \in E. \\
-\quad\quad (\epsilon = read(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge \text{reads}(r) \in \Phi)\ \wedge \\
-\quad\quad (\epsilon = write(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge \text{writes}(r) \in \Phi)\ \wedge \\
-\quad\quad (\epsilon = reduce_{id}(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge \text{reduces}_{id}(r) \in \Phi)
+\quad\quad (\epsilon = read(l, c, v) \Rightarrow \exists r, l \in M(r) \wedge \text{reads}(r) \in \Phi)\ \wedge \\
+\quad\quad (\epsilon = write(l, c, v) \Rightarrow \exists r, l \in M(r) \wedge \text{writes}(r) \in \Phi)\ \wedge \\
+\quad\quad (\epsilon = reduce_{id}(l, c, v) \Rightarrow \exists r, l \in M(r) \wedge \text{reduces}_{id}(r) \in \Phi)
 \end{array}
@@
 \item mapping consistency, written $\mapconsist{\Omega}$, guarantees a region mapping $M$ satisfies the region constraints $\Omega$% under which an expression was typed
-\item local value consistency, written $\localconsist{L}{\Gamma}$, guarantees local values in $L$ have types consistent with the environment $\Gamma$
+\item local value consistency, written $\localconsist{L}{\Gamma}$, guarantees local values in $L$ have types consistent with the environment $\Gamma$ (with logical regions mapped to physical regions via $M$)
 %under which the expression was typed
@@
 is sufficient for the proof of soundness.  
-The stronger constraints in {\em valid\_interleave}
-(Figure~\ref{fig:validinterleave}) address the coherence of heap accesses;
+The stronger constraints address the coherence of heap accesses;
 {\em valid\_interleave} specifies permitted interleavings of memory operations for
@@
 \item for locations with only access in {\em simult} mode or in the clobber set, no constraints are
-enforced.  The validity of the interleaving of these accesses is performed by some enclosing expression.
+enforced.  The validity of the interleaving of these accesses is determined within the context of some enclosing expression.
 \end{itemize}
@@
 Finally, accesses to different locations cannot affect each other.
-We next show that this non-interference operator preserves the validity of an interleaving
-of two or more traces when reordering a pair of memory operations.
+Therefore, an adjacent pair of non-interfering memory operations in a memory trace can be reordered while
+preserving the validity of an interleaving.
+%We next show that this non-interference operator preserves the validity of an interleaving
+%of two or more traces when reordering a pair of memory operations.
 \begin{lem}
@@
 
-$
-\begin{array}{l@{}l}
-\multicolumn{2}{l}{priv_1(r_1) \nonint[M] priv_2(r_2) \Leftrightarrow} \\
-& (priv_1 = \text{reads} \wedge priv_2 = \text{reads})\ \vee \\
+$\begin{array}{@{}l@{}l@{}}
+\multicolumn{2}{@{}l}{priv_1(r_1) \nonint[M] priv_2(r_2) \Leftrightarrow} \\
+\hspace*{0.5cm} & (priv_1 = \text{reads} \wedge priv_2 = \text{reads})\ \vee \\
 & (priv_1 = \text{reduces}_{id_1} \wedge priv_2 = \text{reduces}_{id_2} \wedge id_1 = id_2)\ \vee \\
@@
 \\
-\multicolumn{2}{l}{\Phi_1 \nonint[M] \Phi_2 \Leftrightarrow \displaystyle\bigwedge_{\phi_1 \in \Phi_1, \phi_2 \in \Phi_2} \phi_1 \nonint[M] \phi_2}
+\multicolumn{2}{@{}l}{\Phi_1 \nonint[M] \Phi_2 \Leftrightarrow \displaystyle\bigwedge_{\phi_1 \in \Phi_1, \phi_2 \in \Phi_2} \phi_1 \nonint[M] \phi_2}
 \end{array}$
@@
 
-Because testing non-interference of tasks is a pairwise operation, scheduling $n$ tasks can require ${\cal O}(n^2)$
-tests.  Thus, a scheduler that must globally consider all pairs of tasks will be impractical for large machines 
-and large numbers of tasks.  The following theorem, however, shows that Legion programs enjoy a locality property that
+Because testing non-interference of tasks is a pairwise operation, scheduling $n$ tasks
+can require ${\cal O}(n^2)$ tests.  Thus, a scheduler that must globally consider all
+pairs of tasks will be impractical for large machines and large numbers of tasks.  The
+following theorem, however, shows that Legion programs enjoy a locality property that
 limits the scope of needed non-interference tests.
@@
 \end{proof}
-} The Legion task scheduler uses Theorem~\ref{thm:hiersched} as follows: 
+}
+
+The Legion task scheduler uses Theorem~\ref{thm:hiersched} as follows: 
 sibling function calls (those invoked within the same
 function body) $e_1$ and $e_2$ are checked for non-interference of their (dynamic)
-privileges.  Since $e_1$ and $e_2$ are guaranteed to be created on the node on which their parent
-is running, no communication is required to perform the non-interference
+privileges.  Since $e_1$ and $e_2$ are called on the parent task's node, no communication
+is required to perform the non-interference
 test.  If they interfere they are executed in program order or serialized

### 2012-11-14 doc/types-pldi2013/langdef_full.tex
fixed typos found by Elliott - tried to tweak multicoloring and clobber set paragraphs
hopefully last wave of edits, spell check

@@
 functions must be monomorphic and visible in all functions means that 
-$\traceconsist{E}$ will be insensistive to the particular $M$ and $\Gamma$.}
+$\traceconsist{E}$ will be insensitive to the particular $M$ and $\Gamma$.}
 

### 2012-11-14 doc/types-pldi2013/langdef.tex
two more minor tweaks

@@
 chosen to return a value other than what was most recently written by the current
-expression, as long a
+expression, as long as a
 valid interleaving of the operations with other concurrent operations can be found within
@@
 \item mapping consistency, written $\mapconsist{\Omega}$, guarantees a region mapping $M$ satisfies the region constraints $\Omega$% under which an expression was typed
-\item local value consistency, written $\localconsist{L}{\Gamma}$, guarantees local values in $L$ have types consistent with the environment $\Gamma$ (with logical regions mapped to physical regions via $M$)
+\item local value consistency, written $\localconsist{L}{\Gamma}$, guarantees local values in $L$ have types consistent with the environment $\Gamma$ (using $M$ to map logical regions in $\Gamma$ to physical regions)
 %under which the expression was typed

### 2012-11-14 doc/llr-pldi2013/impl.tex
last-minute edits, spell check

@@
 The final kind of memory is the portion of system memory on each node that has been registered with
-the GASNet runtime to allow {\em remote memory access} (RMA) by other nodes in the cluster.
+the GASNet runtime to allow {\em remote direct memory access} (RDMA) by other nodes in the cluster.
 We refer to this memory as {\em GASNet memory}.
@@
 %% a global GASNet memory that represents pages that have been registered 
-%% with all nodes in the system for supporting remote memory accesses (RMAs).
+%% with all nodes in the system for supporting remote memory accesses (RDMAs).
 %% The GASNet memory provides the illusion of global memory.  CPU processors
@@
 
-In addition to the RMA capabilities, GASNet provides {\em active messages} for 
+In addition to the RDMA capabilities, GASNet provides {\em active messages} for 
 inter-node communication.  Implementation of all runtime features requiring communication
@@
 The first operation dependent on an event
-from a remote node allocates the same data structure on the remote node.  A single
-{\em event subscription} active message is then sent from the remote node to the owning node indicating
+from a remote node allocates the same data structure on the remote node.  An
+{\em event subscription} active message is then sent to the owning node indicating
 that the remote node should be informed when the event triggers.  Arbitrarily many dependent operations on
-the remote node can then be added to the list of waiters without any additional communication.
-When the event does trigger, the owner node notifies all local waiters and
-sends a single {\em event trigger} active message to each node from which it has seen an event subscription
+the subscribing node can be added to the list of waiters without any additional communication.
+When the event triggers, the owner node notifies all local waiters and
+sends an {\em event trigger} active message to each node from which it has seen an event subscription
 to notify remote waiters.
-In the case where an event triggers while a subscription message is in flight from a remote node, the owner node will
-immediately respond with a trigger message.
+In the case where an event triggers while a subscription message is in flight from a remote node, the owner
+node will immediately respond with a trigger message upon receipt of the subscription message.
 
@@
 Like events, deferred locks (henceforth called simply locks) are
-created on demand by the node of the creation request. 
-Lock handles are also statically divided across the nodes,
-allowing creation to be done without inter-node communication.
-Whereas event ownership is static, locks may migrate; the {\em
-  creating} node is the initial owner, but ownership can be
+created on demand, using a space of handles that has been divided
+across the nodes.  As with events, lock creation requires no communication
+and the handle alone is sufficient for any node to determine the creator
+of a lock.  However, whereas event ownership is static, locks may migrate; the
+creating node is the initial owner, but ownership can be
 transferred to other nodes.
@@
 
-When a lock request is made, an event is created to track when the grant occurs.  The current node then
+Each time a lock request is made, a new event is created to track when the grant occurs.  The current node then
 examines its copy of the lock data structure to determine if it is the owner.

### 2012-11-14 doc/llr-pldi2013/interface.tex
last-minute edits, spell check

@@
 use of events is passing them as preconditions for other operations.  The client can use the
-{\tt merge\_events} call (line 7) to create an event that triggers only when every
+{\tt merge\_events} call (line 7) to create an aggregate event that triggers once every
 event in a given set of events has triggered.  A code example using events follows in
@@
   // Unpack lock, input events, target processors from arguments
-  Lock needed = ...
+  Lock mutex = ...
   // Acquire lock, launch first subtask, release lock
-  Event lock_event_1 = needed.lock(prev_event_1);
+  Event lock_event_1 = mutex.lock(prev_event_1);
   Event task_event_1 = proc1.spawn(SUB, args_1, sizeof(args_1), lock_event_1);
-  needed.unlock(task_event_1);
+  mutex.unlock(task_event_1);
   // same for second subtask
-  Event lock_event_2 = needed.lock(prev_event_2);
+  Event lock_event_2 = mutex.lock(prev_event_2);
   Event task_event_2 = proc2.spawn(SUB, args_2, sizeof(args_2), lock_event_2);
-  needed.unlock(task_event_2);
+  mutex.unlock(task_event_2);
 }
@@
 \begin{tikzpicture}
-  \node (l1) at (2,1.95) [shape=rectangle,draw] {\begin{tabular}{c}{\bf lock}\\needed\end{tabular}};
+  \node (l1) at (2,1.95) [shape=rectangle,draw] {\begin{tabular}{c}{\bf lock}\\mutex\end{tabular}};
   \node (t1) at (4.5,1.95) [shape=rectangle,draw] {\begin{tabular}{c}{\bf run}\\SUB(args\_1)\end{tabular}};
-  \node (u1) at (7,1.95) [shape=rectangle,draw] {\begin{tabular}{c}{\bf unlock}\\needed\end{tabular}};
+  \node (u1) at (7,1.95) [shape=rectangle,draw] {\begin{tabular}{c}{\bf unlock}\\mutex\end{tabular}};
 
+  \node[auto] (x1) at (-0.3,2.15) {prev\_event\_1};
+  %\draw [->] (x1) to (l1.west);
   \draw [->] (0,1.95) to (l1.west);
@@
 
-  \node (l2) at (2,0.25) [shape=rectangle,draw] {\begin{tabular}{c}{\bf lock}\\needed\end{tabular}};
+  \node (l2) at (2,0.25) [shape=rectangle,draw] {\begin{tabular}{c}{\bf lock}\\mutex\end{tabular}};
   \node (t2) at (4.5,0.25) [shape=rectangle,draw] {\begin{tabular}{c}{\bf run}\\SUB(args\_2)\end{tabular}};
-  \node (u2) at (7,0.25) [shape=rectangle,draw] {\begin{tabular}{c}{\bf unlock}\\needed\end{tabular}};
+  \node (u2) at (7,0.25) [shape=rectangle,draw] {\begin{tabular}{c}{\bf unlock}\\mutex\end{tabular}};
 
+  \node[auto] (x2) at (-0.3,0.45) {prev\_event\_2};
   \draw [->] (0,0.25) to (l2.west);
@@
 when the lock has been acquired.  Like a processor's spawn method, both the {\tt lock} 
-and {\tt unlock} methods take an optional event paramenter as a precondition (lines 32-33).
+and {\tt unlock} methods take an optional event parameter as a precondition (lines 32-33).
 
@@
 function needs to launch two subtasks on two other processors, each with exclusive access to some resource protected by a
-lock named {\tt mutex}.  For each subtask, it first requests the lock on the subtask's behalf, making the
-subtask wait until the lock has been acquired.  The parent task also requests that the runtime automatically
-release the lock after each subtask has completed.  In the event dependency graph, solid arrows represent
+lock named {\tt mutex}.  For each subtask, it first requests the lock on the subtask's behalf.  Using the
+event returned by the lock request as a precondition, it then spawns the subtask.  The runtime will guarantee
+that the subtask is not started until the lock has been acquired.  The parent task also requests that the
+runtime automatically release the lock after each subtask has completed by using the subtask's completion
+event as a precondition for the unlock.  In the event dependency graph (Figure~\ref{fig:lockevents}),
+solid arrows represent
 the explicit events used in the code, while the dotted line represents the dynamic ordering that results
@@
 spaces.  This restriction requires programmers to manage both the layout of
-data and the marshalling of data for communication.  {\em Physical regions} are a
+data and the marshaling of data for communication.  {\em Physical regions} are a
 mechanism for decoupling the placement of data from its layout.  Physical regions allow portability
@@
 %accessors supply the level of indirection necessary to convert operations into remote 
-%memory operations (RMAs) if supported by the underlying hardware.  
+%memory operations (RDMAs) if supported by the underlying hardware.  
 %An example of this is described in more detail in Section~\ref{sec:impl}.

### 2012-11-14 doc/llr-pldi2013/micro.tex
Lots of small edits
last-minute edits, spell check

@@
 be triggered by our implementation.  Instead of a single chain, many parallel chains
-are created.  Additionally, each step of the chain consists of a parameterized number
+are created.  Additionally, each step of the chain consists of a parametrized number
 of events, each one dependent on every event from the previous step of the chain.
@@
 in a distributed environment.
-A parameterized number of locks are created per node and their handles are made available to
-every node.  Each node then creates a parameterized number of {\em chains} of lock/unlock
+A parametrized number of locks are created per node and their handles are made available to
+every node.  Each node then creates a parametrized number of {\em chains} of lock/unlock
 request pairs, where each request pair is to a random lock, and is made dependent on the 
@@
 increases with
-decreasing lock counts.  Although contention increases, the favoring of local lock requestors makes
+decreasing lock counts.  Although contention increases, the favoring of local lock requesters makes
 the contention an advantage, reducing the number of lock-related active messages that must be sent
@@
 more contention for locks and the rate drops as a result.  For smaller lock counts, further increases in
-the number of chains result in improved rates.  Note that the point on each line where the increase occurs
+the number of chains result in improved performance.  Note that the point on each line where the increase occurs
 is when the chains per node is greater than total locks, which corresponds to the point where the expected 
 number of requests per lock 
-per node becomes more than one.  As soon as there are multiple expected requestors for the same lock on
+per node becomes more than one.  As soon as there are multiple expected requesters for the same lock on
 a node, the unfairness property of our lock implementation is able to reduce the number of lock ownership 
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.33]{figs/reduce_simple.pdf}
-\end{center}
-\vspace{-2mm}
-\caption{Reduction Benchmark.\label{fig:reducbench}}
-\vspace{-4mm}
-\end{figure}
-
 \texcomment{

### 2012-11-14 doc/llr-pldi2013/results.tex
last-minute edits, spell check

@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.33]{figs/reduce_simple.pdf}
+\end{center}
+\vspace{-2mm}
+\caption{Reduction Benchmark.\label{fig:reducbench}}
+\vspace{-4mm}
+\end{figure}
+
 The composable design of the interface proposed in
@@
 
+{\em Fluid} is an incompressible fluid flow simulation from the PARSEC
+benchmark suite\cite{bienia11benchmarking}.  The reference implementation 
+only runs on shared memory machines, but when written in Legion, Fluid
+runs on large distributed machines as well.  The Fluid simulation
+models particles that flow through cells.  Each time step requires multiple phases
+that update different properties of the particles contingent upon neighboring
+particles in the same cell.  The space of cells is partitioned and neighboring
+cells in different partitions must exchange data between phases.  Legion ensures
+that these exchanges are done point-to-point by chaining copies and tasks
+using events rather than employing a global bulk-synchronous approach.
+
 {\em Circuit} is a simulation of an integrated circuit that is described as an
@@
 couldn't be performed using GPUs.
-
-{\em Fluid} is an incompressible fluid flow simulation from the PARSEC
-benchmark suite\cite{bienia11benchmarking}.  The reference implementation 
-only runs on shared memory machines, but when written in Legion, Fluid
-runs on large distributed machines as well.  The Fluid simulation
-models particles that flow through cells.  Each time step requires multiple phases
-that update different properties of the particles contingent upon neighboring
-particles in the same cell.  The space of cells is partitioned and neighboring
-cells in different partitions must exchange data between phases.  Legion ensures
-that these exchanges are done point-to-point by chaining copies and tasks
-using events rather than employing a global bulk-synchronous approach.
 
@@
 \caption{Locks by Request and Transfer Counts.\label{fig:lockcount}}
-\vspace{-4mm}
+\vspace{-6mm}
 \end{figure}
@@
 the performance of the bulk-synchronous implementations versus the original
-implementations for Circuit, Fluid, and AMR.  Each plot contains
+implementations for Fluid, Circuit, and AMR.  Each plot contains
 performance curves for both implementations on two different problem sizes.
@@
 \centering
+\subfigure[Fluid Application]{
+\label{fig:fluidbulk}
+\includegraphics[scale=0.33]{figs/fluid_bulk_sync.pdf}
+\vspace{-3mm}
+}
 \subfigure[Circuit Application]{
@@
 \includegraphics[scale=0.33]{figs/circuit_bulk_sync.pdf}
+\vspace{-3mm}
 }
-
-\subfigure[Fluid Application]{
-\label{fig:fluidbulk}
-\includegraphics[scale=0.33]{figs/fluid_bulk_sync.pdf}
-}
-
 \subfigure[AMR Application]{
@@
 \caption{Bulk-Synchronous Performance.\label{fig:bulksync}}
-\vspace{-4mm}
+\vspace{-6mm}
 \end{figure}

### 2012-11-14 doc/llr-pldi2013/interface.tex
minor edits for the last part of the low level paper

@@
   // Unpack lock, input events, target processors from arguments
-  Lock mutex = ...
+  Lock mutex = Lock::create_lock(); 
   // Acquire lock, launch first subtask, release lock
@@
 function needs to launch two subtasks on two other processors, each with exclusive access to some resource protected by a
-lock named {\tt mutex}.  For each subtask, it first requests the lock on the subtask's behalf.  Using the
-event returned by the lock request as a precondition, it then spawns the subtask.  The runtime will guarantee
-that the subtask is not started until the lock has been acquired.  The parent task also requests that the
+lock named {\tt mutex} (created on line 3).  For each subtask, it first requests the lock on the subtask's behalf.  Using the
+event returned by the lock request as a precondition, it then spawns the subtask.  The event dependence guarantees that the runtime will 
+not start the subtask until the lock has been acquired.  The parent task also requests that the
 runtime automatically release the lock after each subtask has completed by using the subtask's completion
